### 2011-03-24 main.tex
Première version

\typeout{Batch, Off-Policy Apprenticeship Learning}
\documentclass{article}
\usepackage{ijcai11}
\usepackage{times}
% the following package is optional:
%\usepackage{latexsym} 
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{natbib}
\usepackage{graphicx}

\title{Batch, Off-Policy Apprencticeship Learning}
\author{Edouard Klein \and Matthieu Geist \and Olivier Pietquin \\
Supélec, IMS\\
Metz, France \\
\{edouard.klein|matthieu.geist|olivier.pietquin\}@supelec.fr}
\begin{document}
\maketitle
\tableofcontents
\begin{abstract}
We consider the IRL framework for apprenticeship learning. %1
Current IRL algorithms use a simple Monte-Carlo estimation to approximate the feature expectation of both the expert and the policy under evaluation. %2
In this paper, we propose another method for estimating the feature expectation based on the Least Square Temporal Difference (LSTD) family of algorithms. %3
This new method allows for batch, off-policy appreticeship learning in situations where the dynamics of the system remains unknown. %4
By testing it on a classical problem, we show that our method is empirically sound.%5
We also provide some theoretical results.%6
\end{abstract} 
\section{Introduction}
When a system can be empirically controlled by an expert whose behaviour is difficult to describe formally, apprenticeship learning is a way to let a machine control the system by mimicking the expert.\\

This is a restriction of the more general problem of task transfer, where the goal of the expert is to be guessed and then pursued with the machine own means.\\

\emph{Inverse Reinforcement Learning} is a framework created to tackle both of these problems. \citet{russell1998learning} first defined it in the terms used today, drawing from animal behavior research and more loosely from macroeconomics research : given an expert's sensory input and decisions, determine the reward (i.e. the \emph{goal}) optimized by that expert. This reward is then to be optimized via tools developped whithin the \emph{Reinforcement Learning} framework \citep{sutton1998reinforcement}.\\

\citet{ng2000algorithms} proposed new algorithms to solve the IRL problem, followed by an article \citep{abbeel2004apprenticeship} about the more specific problem of apprenticeship learning, in which imitation metrics are introduced. The method proposed in this work is built upon this paper, and compared against it.\\

The main limitation in \citet{abbeel2004apprenticeship}'s work is the need for a model of the environnement or, at least, a simulator. \\

FIXME:Autres approches qu'Abbeel, notamment celles qui cherchent à contourner le même problème.\\

With the LSTD \citep{bradtke1996linear} and LSPI \citep{lagoudakis2003least} algorithms, one can work out the Reinforcement Learning problem without the need for a model. We extend these capabilities to \citet{abbeel2004apprenticeship}'s algorithm by assessing the imitation metrics through LSTD. We also do not need a simulator, the data from the expert being used to feed LSPI.\\
\section{Background}
A \emph{Markov Decision Process} is a tuple $\{S,A,\gamma,R\}$ with :
\begin{itemize}
\item $S$ the state space,
\item $A$ the action space,
\item $\gamma$ the discount factor,
\item $R$ the reward function
\end{itemize}

A policy $\pi$ is a map from the states to the actions : $\pi:S\rightarrow A$.\\

The \emph{Reinforcement Learning} problem seeks the policy that maximize the expected discounted reward. When following a policy $\pi$ from state $s\in S$, the expected discounted reward is described by the \emph{value function} $V^\pi$ : \\
\begin{equation}
V^\pi(s) = E\left[\left.\sum\limits_{t=0}^{\infty}\gamma^tr_t\right|s_0=s,\pi\right]
\label{eqn:V}
\end{equation}

The state space $S$ is projected into a feature space generated by the function $\psi : S \rightarrow \mathbb{R}^p$. Both the state space and the action space are projected into a feature space generated by the function $\phi : S\times A \rightarrow \mathbb{R}^k$.\\

The value function is approximated via a matrix of coefficients $\omega$, such as :
\begin{equation}
\hat V^\pi(s) = \arg\max_{a\in A}\omega_\pi^T\phi(s,a)
\end{equation}

The \emph{expert} is an agent whose interactions with the world are recorded as a set of transitions : $D_E = \left\{\{s_i,a_i,s'_i,r_i\}_i\right\}$.\\

$D$ is a random transition source.\\

The feature expectation is a mapping from a policy $\pi$ to a function $\mu_\pi : S \rightarrow \mathbb{R}^p$ which gives for a state $s\in S$ the expected discounted value for the features of the states visited when following policy $\pi$ from state $s$ :
\begin{equation}
\mu_\pi(s) = E\left[\left.\sum\limits_{t=0}^{\infty}\gamma^t\psi(s_t)\right|s_0=s,\pi\right]
\label{eqn:mu}
\end{equation}\\
The feature expectation of the expert is denoted $\mu_E$.\\

The IRL problem seeks the reward the expert is maximizing. The imitation problem seeks the reward which, maximized by one of the many RL tools, will give a behavior as close to the expert's as possible.\\

\citet{abbeel2004apprenticeship}'s algorithm solves the imitation pproblem by approximating this reward with a matrix of coefficients $\theta$ such as :
\begin{equation}
\hat R(s) = \theta^T\phi(s)
\end{equation}

The quality of the imitation is assessed via the error term $||\mu_\pi(s_0)-\mu_E(s_0)||_2$. Given that the norm of $\theta$ is bounded, this error term also gives a feeling of the difference of performance under the estimated reward function $\hat R$ :
\begin{equation}
|V^\pi(s_0)-V^E(s_0)| \leq ||\theta||_2||\mu_\pi(s_0)-\mu_E(s_0)||_2
\end{equation}

The main shortcoming of this algorithm lies in its need for either a model of the environnement or at least a sample source $D$, in order to approximate the feature expectation of both the expert and the current policy via a Monte-Carlo estimator.
\section{LSTD-$\mu$}
The similarity between equations \ref{eqn:mu} and \ref{eqn:V} (both are Bellman equations) allows the use of the LSTD algorithm in order to estimate the feature expectation of a policy.\\

We introduce the matrix $\zeta$ such as :
\begin{equation}
\hat \mu_\pi (s) = \zeta^T\phi(s,\pi(s))
\end{equation}
$\zeta$ is given by :
\begin{equation}
\zeta = A^{-1}b
\end{equation}
with :
\begin{eqnarray}
FIXME: copier depuis le code
\end{eqnarray}

This approach is superior to the Monte-Carlo estimation suggested by \citet{abbeel2004apprenticeship}, as it allows for a batch, off policy estimation of $\mu_\pi$, even when no model of the environment is known and no simulator is available whereas the Monte-Carlo estimation needs either a model or a simulator, and only allow for an estimation of $\mu_\pi(s_0)$.\\

Coupled with the LSPI algorithm, this allows an adaptation of \citet{abbeel2004apprenticeship} algorithm to "real-life" settings where the only data available come from the expert's trajectories.\\

FIXME: Un mot sur pourquoi il est inutile de faire des itérations sur mu directement ?

FIXME: Dérouler les deux algorithmes complets côte à côte dans une forme lisible, car c'est un peu la galère en lisant le papier d'Abbeel.
\section{Analysis}
FIXME: Matthieu ?
\section{Experimental benchmark}
The experimental benchmark chosen here is the same as in \citep{ng2000algorithms}, a 5x5 grid world. The expert policy maximizes the following reward : 0 everywhere but in the upper-right square, where it is 1.\\

Both \cite{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side. We use the projection method in both algorithm. As a MDP solver, mc\_ANIRL is given LSPI with a (very) big matrix $D$ as a sample source. $\mu_\pi(s_0)$ and $\mu_E(s_0)$ are computed thanks to a (very) big Monte-Carlo estimation. Both are, for all intended purpose on this toy problem, as good as a perfect theoretical solver.\\

On the other hand lstd\_ANIRL uses LSPI and LSTD$\mu$, fed only with the expert's transitions. This corresponds to a real-life setting where data generation is expensive and the system can not be controlled by an untrained machine.\\

Figure \ref{fig:E} shows the performance of both algorithms with repect to the number of samples from the expert made available.\\

With respect to computing power, our algorithm needed far less iteration to get the job done.
\begin{figure*}
\includegraphics[width=\textwidth]{../Code/courbe_e.pdf}
\caption{$||\mu_E(s_0)-\mu_\pi(s_0)||_2$ with respect to the number of samples available from the expert. Our algorithm is not as good as \citet{abbeel2004apprenticeship} with very few samples, but make up for it very soon.}
\label{fig:E}
\end{figure*}
\section{Quality criterion}
FIXME: Discussion sur les critères utilisés, ouverture vers le transfert de tâche.
\begin{figure*}
\includegraphics[width=\textwidth]{../Code/courbe_a.pdf}
\caption{Different criteria with respect to the number of iterations. See how decorelated they can be.}
\label{fig:A}
\end{figure*}
\section{Future Work}
FIXME: Une seule itération, est-ce normal ? Abbeel reste-t-il meilleur sur un problème plus complexe ? Transfert de tâche ? Est-il nécessaire de choisir sa politique avec LSTDMu, puisqu'il finit en une seule fois ?
\section{Conclusion}
\bibliographystyle{named}
\bibliography{../Biblio/Biblio}
\end{document}

### 2011-03-25 main.tex
Nouvelle intro

@@
 \usepackage{amssymb}
+\usepackage{natbib}
+\usepackage{graphicx}
+\usepackage{subfigure}
+%\usepackage{algorithm2e}
 \usepackage{algorithm}
 \usepackage{algorithmic}
-\usepackage{natbib}
-\usepackage{graphicx}
 
@@
 Metz, France \\
-\{edouard.klein|matthieu.geist|olivier.pietquin\}@supelec.fr}
+\{edouard.klein$|$matthieu.geist$|$olivier.pietquin\}@supelec.fr}
 \begin{document}
 \maketitle
-\tableofcontents
 \begin{abstract}
@@
 \section{Introduction}
-When a system can be empirically controlled by an expert whose behaviour is difficult to describe formally, apprenticeship learning is a way to let a machine control the system by mimicking the expert.\\
+The optimal control problem can be expressed as having a machine control a system, with the goal of fulfilling a specific task.
+Various solutions to this problem exist.
+Things can get hard to deal with, however, when precisely defining the task is in itself a challenge.
+When the system can be empirically controlled by an expert, even when his behaviour is difficult to describe formally, apprenticeship learning is a way to let a machine control the system by mimicking the expert.\\
 
-This is a restriction of the more general problem of task transfer, where the goal of the expert is to be guessed and then pursued with the machine own means.\\
+When the task at hand can be formally described, the problem of optimal control can be solved by tools from the \emph{Reinforcement Learning} (RL) framework.
+An extensive introduction to this matter can be found in \citep{sutton1998reinforcement} : the control problem is projected into a sequential decision problem inside a \emph{Markov Decision Process}.
+The agent transits from \emph{state} to \emph{state}, the decision taking the form of choosing an \emph{action} at each visited state.
+For each \emph{transition}, the agent receives a \emph{reward}.
+The \emph{reward function} (a mapping from each state to a real value) defines the task.
+The agent's goal is to maximize the long-term reward by maximizing what is called the \emph{expected discounted reward} : the expectation of the sum of all rewards to be obtained, multiplied by a deacreasing-over-time factor (called the \emph{discount factor}).
+For a given \emph{policy} (a mapping from each state to an action) that the agent follows, the mapping from each state to the expected discounted reward is called the \emph{value function}.
+The \emph{optimal policy}, hopefully found by the agent, is the unique policy of which the value fuction is, for every state, superior or equal to all the other policie's value functions.\\
 
-\emph{Inverse Reinforcement Learning} is a framework created to tackle both of these problems.
-\citet{russell1998learning} first defined it in the terms used today, drawing from animal behavior research and more loosely from macroeconomics research : given an expert's sensory input and decisions, determine the reward (i.e.
-the \emph{goal}) optimized by that expert.
-This reward is then to be optimized via tools developped whithin the \emph{Reinforcement Learning} framework \citep{sutton1998reinforcement}.\\
+It is not necessary for a model of the environment or for a simulator to be available, as for example the LSTD (\citep{bradtke1996linear}) algorithm can approximate the value function in a batch, off-policy, model-free setting.
+Building on it, the LSPI (\citep{lagoudakis2003least}) algorithm can be used to actually solve the RL problem in such settings : given a set of random or arbitrary transitions and the corresponding rewards, find the optimal policy.\\
 
-\citet{ng2000algorithms} proposed new algorithms to solve the IRL problem, followed by an article \citep{abbeel2004apprenticeship} about the more specific problem of apprenticeship learning, in which imitation metrics are introduced.
-The method proposed in this work is built upon this paper, and compared against it.\\
+\citet{russell1998learning} first defined the \emph{Inverse Reinforcement Learning} (IRL) problem in the terms used today, drawing from animal behavior research and more loosely from macroeconomics research.
+What is looked for is not the policy as it is in the RL problem, but the reward, which defines the task and of which the expected discounted sum is maximized by an expert.
+IRL is a framework in which it could be possible to attack the apprenticeship learning problem mentionned earlier.\\
 
-The main limitation in \citet{abbeel2004apprenticeship}'s work is the need for a model of the environnement or, at least, a simulator. \\
+\citet{ng2000algorithms} first proposed an algorithm to solve the IRL problem, sheding light on a problem foreseen in \citep{russell1998learning} : the IRL problem is mathematically ill-defined as it admits a infinity of solutions, including degenerative ones (e.g.
+the null reward accepts any policy as an optimal policy).\\
+
+This work was followed by an article \citep{abbeel2004apprenticeship} focussing on the more specific, but better defined problem of apprenticeship learning using the IRL framework.
+This work makes use of the IRL framework by using an approximator for the reward function, and looking for the parameters for which the approximated reward function is the best.
+This however only concerns the apprenticeship learning (a.k.a.
+imitation learning) because the metric used, although intended to discriminate the expert's performance and the current policy's performance \emph{under the current approximate reward}, can be seen as a measure of proximity of the different states visitation frequencies.
+
+This metric, called the \emph{feature expectation}, is akin to the expected discounted reward discussed when describing the RL framework.
+It does however not sum the reward but the features of the visited states, features defined by the choice of the reward function approximator.
+It can be assessed either by analytical resolution (when the model is available) of a Monte-Carlo estimation (when a simulator is available).\\
+
+The main limitation in \citet{abbeel2004apprenticeship}'s work (from now on reffered to as the ANIRL algorithm) is the need for a model of the environnement or, at least, a simulator.
+We build directly upon this work by presenting a new algorithm, LSTD$\mu$, which along with the use of LSPI and LSTD allows the use of the ANIRL algorithm in a batch, off-policy, model-free setting.\\
 
@@
 
-With the LSTD \citep{bradtke1996linear} and LSPI \citep{lagoudakis2003least} algorithms, one can work out the Reinforcement Learning problem without the need for a model.
-We extend these capabilities to \citet{abbeel2004apprenticeship}'s algorithm by assessing the imitation metrics through LSTD.
-We also do not need a simulator, the data from the expert being used to feed LSPI.\\
 \section{Background}
@@
 \begin{equation}
-\zeta = A^{-1}b
+\zeta = \tilde A^{-1}\tilde b
 \end{equation}
@@
 \begin{eqnarray}
-FIXME: copier depuis le code
+\tilde A = \sum_{\{s,a,s'\} \in D}\phi(s,a)\left(\phi(s,a) - \gamma \phi(s',\pi(s'))\right)^T\\
+\tilde b = \sum_{\{s,a,s'\} \in D} \phi(s,a)\psi(s)^T
 \end{eqnarray}
@@
 FIXME: Un mot sur pourquoi il est inutile de faire des itérations sur mu directement ?
+\begin{figure*}
+ \begin{minipage}[t]{.45\textwidth}
+ \begin{algorithm}[H]
+   \caption{lstd\_ANIRL} 
+   \label{alg:lstd_anirl} 
+   \begin{algorithmic}[1]
+     \STATE $\zeta \leftarrow 0$
+     \STATE $D_\mu \leftarrow D $
+     \STATE $D_\mu.r \leftarrow \psi(D.s)$
+     \STATE $\mu \leftarrow LSTD\mu( D_\mu )$
+     \STATE $D_E.r \leftarrow \psi(D_E.s)$
+     \STATE $\mu_E \leftarrow LSTD\mu( D_E )$
+     \STATE $\theta \leftarrow {\mu_E - \mu\over ||\mu_E - \mu||_2}$
+     \STATE $ \bar\mu \leftarrow \mu $
+     \STATE $ t \leftarrow ||\mu_E - \bar\mu||_2 $
+     \WHILE{$ t > \epsilon $}
+     \STATE $D.r \leftarrow \theta^T\psi(D.s)$
+     \STATE $ \omega \leftarrow LSPI(D) $
+     \STATE $ \mu \leftarrow LSTD\mu( D_\mu )$
+     \ENDWHILE
+   \end{algorithmic}
+ \end{algorithm}
+ \end{minipage}
+ \hfill
+ \begin{minipage}[t]{.45\textwidth}
+   \begin{algorithm}[H]
+     \caption{Mission-side2 Greedy} 
+     \label{alg2} 
+     \begin{algorithmic}[1]
+       \STATE whatever your alg does
+       \FOR{each mission $M_{j}$ in sorted order} 
+       \STATE do something
+       \ENDFOR
+     \end{algorithmic}
+   \end{algorithm}
+ \end{minipage}
+ \hfill
+\end{figure*}
+
+%% \begin{figure*}
+%%   \subfigure{
+%%     \begin{algorithm}
+%%       $A\leftarrow 0$
+%%     \end{algorithm}
+%%   }
+  %% \qquad
+  %% \subfigure{
+    %% \begin{algorithm}[H]
+    %%   \caption{Algo deux}
+    %%   \begin{algorithmic}
+    %%     $b\leftarrow 0$
+    %%   \end{algorithmic}
+    %% \end{algorithm}
+    %%   } 
+%% \end{figure*}
 
@@
 With respect to computing power, our algorithm needed far less iteration to get the job done.
-\begin{figure*}
-\includegraphics[width=\textwidth]{../Code/courbe_e.pdf}
+\begin{figure}
+\includegraphics[width=.5\textwidth]{../Code/courbe_e.pdf}
 \caption{$||\mu_E(s_0)-\mu_\pi(s_0)||_2$ with respect to the number of samples available from the expert.
@@
 \label{fig:E}
-\end{figure*}
+\end{figure}
 \section{Quality criterion}
 FIXME: Discussion sur les critères utilisés, ouverture vers le transfert de tâche.
-\begin{figure*}
-\includegraphics[width=\textwidth]{../Code/courbe_a.pdf}
+\begin{figure}
+\includegraphics[width=.5\textwidth]{../Code/courbe_a.pdf}
 \caption{Different criteria with respect to the number of iterations. See how decorelated they can be.}
 \label{fig:A}
-\end{figure*}
+\end{figure}
 \section{Future Work}

### 2011-03-26 main.tex
Partie expérimentale et partie sur le critère de qualitay. Plus un bout de l'algo

@@
 |V^\pi(s_0)-V^E(s_0)| \leq ||\theta||_2||\mu_\pi(s_0)-\mu_E(s_0)||_2
+\label{eqn:vmu}
 \end{equation}
@@
    \begin{algorithmic}[1]
-     \STATE $\zeta \leftarrow 0$
+     \STATE $\omega \leftarrow 0$
      \STATE $D_\mu \leftarrow D $
      \STATE $D_\mu.r \leftarrow \psi(D.s)$
-     \STATE $\mu \leftarrow LSTD\mu( D_\mu )$
+     \STATE $\zeta \leftarrow LSTD\mu( D_\mu )$
+     \STATE $\mu \leftarrow \zeta^T\psi(s,\pi_\omega(s))$
      \STATE $D_E.r \leftarrow \psi(D_E.s)$
@@
      \STATE $ \omega \leftarrow LSPI(D) $
-     \STATE $ \mu \leftarrow LSTD\mu( D_\mu )$
+     \STATE $ \mu \leftarrow LSTD\mu( D_\mu, \pi_\omega )$
+     \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $
+     \STATE $\theta \leftarrow 
+       {\mu_E - \bar\mu\over ||\mu_E - \bar\mu||_2}$
      \ENDWHILE
@@
 The experimental benchmark chosen here is the same as in \citep{ng2000algorithms}, a 5x5 grid world.
-The expert policy maximizes the following reward : 0 everywhere but in the upper-right square, where it is 1.\\
+The agent is in one of the case of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
+With probability 0.9, the agent moves in the intended direction.
+With probability 0.1, the direction is randomly drawn (and thus have probability 0.25 to match the original direction).
+The expert policy maximizes the following reward : 0 everywhere but in the upper-right square, where it is 1.
+An episodes ends when the upper right square is attained, or after 20 steps.
+At the start of each episode, the expert is in the lower-left corner of the grid (the opposite of where the reward is).\\
 
-Both \cite{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side.
-We use the projection method in both algorithm.
+Both the state and action spaces are finite and of small cardinality.
+Hence, the chosen feature functions $\phi$ and $\psi$ are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
+There is a bijective mapping from the input space elements ($S$ for $\psi$, $S \times A$ for $\phi$) to the indexes $i$ for which the coordinate of the feature is 1.\\
+
+Both \citet{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side. We use the projection method in both algorithms.\\
+
 As a MDP solver, mc\_ANIRL is given LSPI with a (very) big matrix $D$ as a sample source.
 $\mu_\pi(s_0)$ and $\mu_E(s_0)$ are computed thanks to a (very) big Monte-Carlo estimation.
-Both are, for all intended purpose on this toy problem, as good as a perfect theoretical solver.\\
+Both are, for all intended purpose on this toy problem, as good as a perfect theoretical solver.
+We thus are in the case intended by \citet{abbeel2004apprenticeship}.\\
 
-On the other hand lstd\_ANIRL uses LSPI and LSTD$\mu$, fed only with the expert's transitions.
+On the other hand lstd\_ANIRL is used as if it could not have access to the simulator.
+It uses LSPI and LSTD$\mu$, fed only with the expert's transitions.
 This corresponds to a real-life setting where data generation is expensive and the system can not be controlled by an untrained machine.\\
 
-Figure \ref{fig:E} shows the performance of both algorithms with repect to the number of samples from the expert made available.\\
+We want to compare the porformance of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
+The discussion about the choice of the performance metric has its own dedicated section (section \ref{sec:perf}).
+We use here the $||\mu(s_0) - \mu_E(s_0)||_2$ error term.\\
 
-With respect to computing power, our algorithm needed far less iteration to get the job done.
+Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $|\mu(s_0)-\mu_E(s_0)||_2$ after the algorithm has converged, or at least did 40 iterations.
+The best policy is found by lstd\_ANIRL after one iteration only (this is being investigated), whereas in mc\_ANIRL convergence happens after at least ten iterations, and the best policy is not always the last.
+The score of the best policy (not the last) is what is plotted here.\\
+
+We can see that although lstd\_ANIRL is not as good as mc\_ANIRL when very few samples are available, both algorithm quickly converge to almost the same value (we don't know yet if the seemingly slightly best performance of lstd\_ANIRL is an artifact).
+The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version.
+\section{Quality criterion}
+
+Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion.
+These plots show four different quality ciretria during a run of mc\_ANIRL and several runs of lstd\_ANIRL.
+The $||\mu(s_0) - \mu_E(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
+It bears an interesting relation with the difference between the expert's value function and the current value function with respect \emph{to the current reward} (equation \ref{eqn:vmu}).\\
+
+The fact that  the curves of  $||\mu(s_0) - \mu_E(s_0)||_2$ and $||\hat\mu(s_0) - \hat\mu_E(s_0)||_2$ are indistinguishable in figure \ref{fig:A} means that for it has access to a cheap simulator, mc\_ANIRL is as if it worked within an analitycal framework.
+This however can not be said of lstd\_ANIRL, for which the two curves are indeed different (Figure \ref{fig:B}).
+Not knowing the true value of $\mu_E(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm, $t = ||\bar\mu-\hat\mu_E(s_0)||$.
+After the algorithm has converged (it systematically does after only one iteration, this is being investigated), there is no way to know whether the policy is actually good or not (note that lstd\_ANIRL normally does not have acces to what was plotted in figure \ref{fig:E}).\\
+
+It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in mc\_ANIRL either, as it can be low when the policy is bad.
+The best policy, however, can be easily chosen from the $||\mu(s_0) - \mu_E(s_0)||_2$ term, which mc\_ANIRL can compute.
+When this term is low, the objective performance (that is, $V^E(s_0)-V^\pi(s_0)$ with respect to the unknow true reward function) is low too.
 \begin{figure}
@@
 \end{figure}
-\section{Quality criterion}
-FIXME: Discussion sur les critères utilisés, ouverture vers le transfert de tâche.
+\label{sec:perf}
 \begin{figure}
 \includegraphics[width=.5\textwidth]{../Code/courbe_a.pdf}
-\caption{Different criteria with respect to the number of iterations. See how decorelated they can be.}
+\caption{Different criteria with respect to the number of iterations for a run of mc\_ANIRL.}
 \label{fig:A}
+\end{figure}
+\begin{figure}
+\includegraphics[width=.5\textwidth]{../Code/courbe_b.pdf}
+\caption{Different criteria with respect to the number of samples from the expert, for several runs of lstd\_ANIRL.}
+\label{fig:B}
 \end{figure}

### 2011-03-28 main.tex
Ajout des deux algos, les différences en bleu

@@
 \usepackage{times}
+\usepackage{color}
 % the following package is optional:
@@
      \STATE $\omega \leftarrow 0$
+     {\color{blue}
      \STATE $D_\mu \leftarrow D $
      \STATE $D_\mu.r \leftarrow \psi(D.s)$
-     \STATE $\zeta \leftarrow LSTD\mu( D_\mu )$
-     \STATE $\mu \leftarrow \zeta^T\psi(s,\pi_\omega(s))$
+     \STATE $\zeta \leftarrow LSTD\mu( D_\mu, \pi_\omega )$
+     \STATE $\mu \leftarrow \zeta^T\phi(s_0,\pi_\omega(s_0))$
      \STATE $D_E.r \leftarrow \psi(D_E.s)$
-     \STATE $\mu_E \leftarrow LSTD\mu( D_E )$
+     \STATE $\zeta \leftarrow LSTD\mu( D_E )$ 
+     \STATE $\mu_E \leftarrow \zeta^T\phi(D_E.s_0, D_E.a_0)$
+     }
      \STATE $\theta \leftarrow {\mu_E - \mu\over ||\mu_E - \mu||_2}$
@@
      \STATE $D.r \leftarrow \theta^T\psi(D.s)$
-     \STATE $ \omega \leftarrow LSPI(D) $
-     \STATE $ \mu \leftarrow LSTD\mu( D_\mu, \pi_\omega )$
+     \STATE $ \omega \leftarrow LSPI( D, \omega_0 ) $
+     {\color{blue}
+     \STATE $ \zeta \leftarrow LSTD\mu( D_\mu, \pi_\omega )$   
+     \STATE $ \mu \leftarrow \zeta^T\phi(s_0,\pi_\omega(s_0))$}
      \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $
@@
        {\mu_E - \bar\mu\over ||\mu_E - \bar\mu||_2}$
+     \STATE $ t\leftarrow ||\mu_E - \bar\mu||_2$
      \ENDWHILE
@@
    \begin{algorithm}[H]
-     \caption{Mission-side2 Greedy} 
-     \label{alg2} 
+   \caption{mc\_ANIRL} 
+   \label{alg:mc_anirl} 
    \begin{algorithmic}[1]
-       \STATE whatever your alg does
-       \FOR{each mission $M_{j}$ in sorted order} 
-       \STATE do something
-       \ENDFOR
+     \STATE $\omega \leftarrow 0$
+     {\color{blue}
+     \STATE $D_\pi \leftarrow simulator( \omega )$
+     \STATE $\mu \leftarrow \sum\limits_{s\in D_\pi.s}\psi(s)$
+     \STATE $\mu_E \leftarrow \sum\limits_{s\in D_E.s}\psi(s)$
+     }
+     \STATE $\theta \leftarrow {\mu_E - \mu\over ||\mu_E - \mu||_2}$
+     \STATE $ \bar\mu \leftarrow \mu $
+     \STATE $ t \leftarrow ||\mu_E - \bar\mu||_2 $
+     \WHILE{$ t > \epsilon $}
+     \STATE $D.r \leftarrow \theta^T\psi(D.s)$
+     \STATE $ \omega \leftarrow LSPI( D, \omega_0 ) $
+     {\color{blue}
+     \STATE $D_\pi \leftarrow simulator( \omega )$
+     \STATE $\mu \leftarrow \sum\limits_{s\in D_\pi.s}\psi(s)$
+     }
+     \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $
+     \STATE $\theta \leftarrow 
+       {\mu_E - \bar\mu\over ||\mu_E - \bar\mu||_2}$
+     \STATE $ t\leftarrow ||\mu_E - \bar\mu||_2$
+     \ENDWHILE
      \end{algorithmic}  
@@
  \hfill
+ \caption{Both algorithms side by side, with specific parts in blue}
 \end{figure*}
@@
 With probability 0.1, the direction is randomly drawn (and thus have probability 0.25 to match the original direction).
-The expert policy maximizes the following reward : 0 everywhere but in the upper-right square, where it is 1.
-An episodes ends when the upper right square is attained, or after 20 steps.
-At the start of each episode, the expert is in the lower-left corner of the grid (the opposite of where the reward is).\\
+The reward optimized by the expert is 0 everywhere but in the upper-right square, where it is 1.
+For every policy, an episodes ends when the upper right square is attained, or after 20 steps.
+At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is).\\
 

### 2011-03-28 main.tex
Modifications Lulu

@@
 \section{Background}
+\label{sec:back}
 A \emph{Markov Decision Process} is a tuple $\{S,A,\gamma,R\}$ with :
@@
      \WHILE{$ t > \epsilon $}
-     \STATE $D.r \leftarrow \theta^T\psi(D.s)$
-     \STATE $ \omega \leftarrow LSPI( D, \omega_0 ) $
+     \STATE $D.r \leftarrow \theta^T\psi(D.s)$  \COMMENT{The estimated reward is propagated in the data}
+     \STATE $ \omega \leftarrow LSPI( D, \omega_0 ) $  \COMMENT{The MDP is solved}
      {\color{blue}
@@
      }
-     \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $
+     \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $ \COMMENT{We use the projection method, and not the QP solver}
      \STATE $\theta \leftarrow 
@@
  \caption{Both algorithms side by side, with specific parts in blue}
+ \label{fig:alg}
 \end{figure*}
@@
 Both the state and action spaces are finite and of small cardinality.
-Hence, the chosen feature functions $\phi$ and $\psi$ are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
+Hence, the chosen feature functions $\phi$ and $\psi$ (see section \ref{sec:back}) are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
 There is a bijective mapping from the input space elements ($S$ for $\psi$, $S \times A$ for $\phi$) to the indexes $i$ for which the coordinate of the feature is 1.\\
 
-Both \citet{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side. We use the projection method in both algorithms.\\
+Both \citet{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side.
+The pseudo code of the implementation we used can be found figure \ref{fig:alg}.\\
 
-As a MDP solver, mc\_ANIRL is given LSPI with a (very) big matrix $D$ as a sample source.
+The MDP solver of mc\_ANIRL is LSPI with a (very) big matrix $D$ as a sample source.
 $\mu_\pi(s_0)$ and $\mu_E(s_0)$ are computed thanks to a (very) big Monte-Carlo estimation.
-Both are, for all intended purpose on this toy problem, as good as a perfect theoretical solver.
+Both these computations are  as good as perfect theoretical solvers for all intended purpose on this toy problem.
 We thus are in the case intended by \citet{abbeel2004apprenticeship}.\\
@@
 
-Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $|\mu(s_0)-\mu_E(s_0)||_2$ after the algorithm has converged, or at least did 40 iterations.
+Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu(s_0)-\mu_E(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
 The best policy is found by lstd\_ANIRL after one iteration only (this is being investigated), whereas in mc\_ANIRL convergence happens after at least ten iterations, and the best policy is not always the last.
@@
 
-We can see that although lstd\_ANIRL is not as good as mc\_ANIRL when very few samples are available, both algorithm quickly converge to almost the same value (we don't know yet if the seemingly slightly best performance of lstd\_ANIRL is an artifact).
-The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version.
+We can see that although lstd\_ANIRL is not as good as mc\_ANIRL when very few samples are available, both algorithm quickly converge to almost the same value (we don't know yet if the seemingly slightly better performance of lstd\_ANIRL is an artifact).
+The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version (for example controlling a machine tool working on expensive materials).
 \section{Quality criterion}
@@
 Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion.
-These plots show four different quality ciretria during a run of mc\_ANIRL and several runs of lstd\_ANIRL.
+These plots show four different quality ciretria during a run of mc\_ANIRL and several runs of lstd\_ANIRL, for which the abscissa can not be the number of iterations.\\
+
 The $||\mu(s_0) - \mu_E(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
@@
 Not knowing the true value of $\mu_E(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm, $t = ||\bar\mu-\hat\mu_E(s_0)||$.
-After the algorithm has converged (it systematically does after only one iteration, this is being investigated), there is no way to know whether the policy is actually good or not (note that lstd\_ANIRL normally does not have acces to what was plotted in figure \ref{fig:E}).\\
+After the algorithm has converged (it systematically does after only one iteration, this is being investigated), there is no way to know whether the policy is actually good or not.\\
 

### 2011-03-28 main.tex
Suppression des algos.

@@
 FIXME: Un mot sur pourquoi il est inutile de faire des itérations sur mu directement ?
-\begin{figure*}
- \begin{minipage}[t]{.45\textwidth}
- \begin{algorithm}[H]
-   \caption{lstd\_ANIRL} 
-   \label{alg:lstd_anirl} 
-   \begin{algorithmic}[1]
-     \STATE $\omega \leftarrow 0$
-     {\color{blue}
-     \STATE $D_\mu \leftarrow D $
-     \STATE $D_\mu.r \leftarrow \psi(D.s)$
-     \STATE $\zeta \leftarrow LSTD\mu( D_\mu, \pi_\omega )$
-     \STATE $\mu \leftarrow \zeta^T\phi(s_0,\pi_\omega(s_0))$
-     \STATE $D_E.r \leftarrow \psi(D_E.s)$
-     \STATE $\zeta \leftarrow LSTD\mu( D_E )$ 
-     \STATE $\mu_E \leftarrow \zeta^T\phi(D_E.s_0, D_E.a_0)$
-     }
-     \STATE $\theta \leftarrow {\mu_E - \mu\over ||\mu_E - \mu||_2}$
-     \STATE $ \bar\mu \leftarrow \mu $
-     \STATE $ t \leftarrow ||\mu_E - \bar\mu||_2 $
-     \WHILE{$ t > \epsilon $}
-     \STATE $D.r \leftarrow \theta^T\psi(D.s)$
-     \STATE $ \omega \leftarrow LSPI( D, \omega_0 ) $
-     {\color{blue}
-     \STATE $ \zeta \leftarrow LSTD\mu( D_\mu, \pi_\omega )$   
-     \STATE $ \mu \leftarrow \zeta^T\phi(s_0,\pi_\omega(s_0))$}
-     \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $ 
-     \STATE $\theta \leftarrow 
-       {\mu_E - \bar\mu\over ||\mu_E - \bar\mu||_2}$
-     \STATE $ t\leftarrow ||\mu_E - \bar\mu||_2$
-     \ENDWHILE
-   \end{algorithmic}
- \end{algorithm}
- \end{minipage}
- \hfill
- \begin{minipage}[t]{.45\textwidth}
-   \begin{algorithm}[H]
-   \caption{mc\_ANIRL} 
-   \label{alg:mc_anirl} 
-   \begin{algorithmic}[1]
-     \STATE $\omega \leftarrow 0$
-     {\color{blue}
-     \STATE $D_\pi \leftarrow simulator( \omega )$
-     \STATE $\mu \leftarrow \sum\limits_{s\in D_\pi.s}\psi(s)$
-     \STATE $\mu_E \leftarrow \sum\limits_{s\in D_E.s}\psi(s)$
-     }
-     \STATE $\theta \leftarrow {\mu_E - \mu\over ||\mu_E - \mu||_2}$
-     \STATE $ \bar\mu \leftarrow \mu $
-     \STATE $ t \leftarrow ||\mu_E - \bar\mu||_2 $
-     \WHILE{$ t > \epsilon $}
-     \STATE $D.r \leftarrow \theta^T\psi(D.s)$  \COMMENT{The estimated reward is propagated in the data}
-     \STATE $ \omega \leftarrow LSPI( D, \omega_0 ) $  \COMMENT{The MDP is solved}
-     {\color{blue}
-     \STATE $D_\pi \leftarrow simulator( \omega )$
-     \STATE $\mu \leftarrow \sum\limits_{s\in D_\pi.s}\psi(s)$
-     }
-     \STATE $\bar\mu \leftarrow \bar\mu + { (\mu-\bar\mu)^T (\mu_E-\bar\mu) \over (\mu-\bar\mu)^T (\mu-\bar\mu) } (\mu-\bar\mu) $ \COMMENT{We use the projection method, and not the QP solver}
-     \STATE $\theta \leftarrow 
-       {\mu_E - \bar\mu\over ||\mu_E - \bar\mu||_2}$
-     \STATE $ t\leftarrow ||\mu_E - \bar\mu||_2$
-     \ENDWHILE
-     \end{algorithmic}  
-   \end{algorithm}
- \end{minipage}
- \hfill
- \caption{Both algorithms side by side, with specific parts in blue}
- \label{fig:alg}
-\end{figure*}
 
-%% \begin{figure*}
-%%   \subfigure{
-%%     \begin{algorithm}
-%%       $A\leftarrow 0$
-%%     \end{algorithm}
-%%   }
-  %% \qquad
-  %% \subfigure{
-    %% \begin{algorithm}[H]
-    %%   \caption{Algo deux}
-    %%   \begin{algorithmic}
-    %%     $b\leftarrow 0$
-    %%   \end{algorithmic}
-    %% \end{algorithm}
-    %%   } 
-%% \end{figure*}
-
-FIXME: Dérouler les deux algorithmes complets côte à côte dans une forme lisible, car c'est un peu la galère en lisant le papier d'Abbeel.
 \section{Analysis}

### 2011-03-28 main.tex
Intro de Matthieu

@@
 \section{Introduction}
-The optimal control problem can be expressed as having a machine control a system, with the goal of fulfilling a specific task.
-Various solutions to this problem exist.
-Things can get hard to deal with, however, when precisely defining the task is in itself a challenge.
-When the system can be empirically controlled by an expert, even when his behaviour is difficult to describe formally, apprenticeship learning is a way to let a machine control the system by mimicking the expert.\\
+The optimal control problem can be expressed as having a machine
+control a system, with the goal of fulfilling a specific task.
+Various solutions to this problem exist. Notably, the reinforcement
+learning (RL) paradigm~\citep{sutton1998reinforcement} is a general framework in
+which an agent learns to control optimally a dynamic system through
+interactions with it. The task is specified through a reward
+function, the agent objective being to take sequential decisions so
+as to maximize the expected cumulative reward.
 
-When the task at hand can be formally described, the problem of optimal control can be solved by tools from the \emph{Reinforcement Learning} (RL) framework.
-An extensive introduction to this matter can be found in \citep{sutton1998reinforcement} : the control problem is projected into a sequential decision problem inside a \emph{Markov Decision Process}.
-The agent transits from \emph{state} to \emph{state}, the decision taking the form of choosing an \emph{action} at each visited state.
-For each \emph{transition}, the agent receives a \emph{reward}.
-The \emph{reward function} (a mapping from each state to a real value) defines the task.
-The agent's goal is to maximize the long-term reward by maximizing what is called the \emph{expected discounted reward} : the expectation of the sum of all rewards to be obtained, multiplied by a deacreasing-over-time factor (called the \emph{discount factor}).
-For a given \emph{policy} (a mapping from each state to an action) that the agent follows, the mapping from each state to the expected discounted reward is called the \emph{value function}.
-The \emph{optimal policy}, hopefully found by the agent, is the unique policy of which the value fuction is, for every state, superior or equal to all the other policie's value functions.\\
+However, things can get hard to deal with  when precisely defining
+the task is in itself a challenge. When the system can be
+empirically controlled by an expert, even when its behavior is
+difficult to describe formally, apprenticeship learning is a way to
+let a machine control the system by mimicking the expert. Rather
+than directly mimicking the expert with some supervised learning
+approach, inverse reinforcement learning (IRL)~\citep{ng2000algorithms}
+consists in learning a reward function under which the policy
+demonstrated by an expert is optimal. Mimicking the expert therefore
+ends up to learning an optimal policy according to this reward
+function. A significant advantage of such an approach is that
+expert's actions can be guessed in states which have not been
+encountered during demonstration.
 
-It is not necessary for a model of the environment or for a simulator to be available, as for example the LSTD (\citep{bradtke1996linear}) algorithm can approximate the value function in a batch, off-policy, model-free setting.
-Building on it, the LSPI (\citep{lagoudakis2003least}) algorithm can be used to actually solve the RL problem in such settings : given a set of random or arbitrary transitions and the corresponding rewards, find the optimal policy.\\
+Basically, there are three families of IRL algorithms:
+feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008game,syed2008apprenticeship,ziebart2008maximum,boularias2011bootstrapping},
+marge-maximization-based~\citep{ratliff2006maximum,ratliff2007boosting,ratliff2007imitation,kolter2008hierarchical}
+and approaches based on the parameterization of the policy by the
+reward function~\citep{Ramachandran2007bayesian,Neu2007apprenticeship}. The first family
+assumes a linearly parameterized reward function. This naturally
+leads to a linearly parameterized value function, the associated
+feature vector being the so-called feature expectation (defined as
+the cumulative discounted reward's feature vector under a given
+policy, see Section~\ref{sec:back} for a formal definition).
+These approaches learn a reward function such that the feature
+expectation of the optimal policy (according to the learnt reward
+function) is close to the feature expectation of the expert policy.
+This is a sufficient condition to have close value functions. The
+second family expresses IRL as a constrained optimization problem in
+which expert's examples have higher expected reward than all other
+policies by a certain margin. Moreover, suboptimality of the expert
+can be considered through the introduction of slack variables. The
+last family parameterizes policies with a reward function. Assuming
+that the expert acts according to a Gibbs policy (respectively to
+the optimal value function related to the reward function which is
+optimized), it is possible to estimate the likelihood of a set of
+state-action pairs provided by the expert. The algorithms differs in
+the way this likelihood is maximized.
 
-\citet{russell1998learning} first defined the \emph{Inverse Reinforcement Learning} (IRL) problem in the terms used today, drawing from animal behavior research and more loosely from macroeconomics research.
-What is looked for is not the policy as it is in the RL problem, but the reward, which defines the task and of which the expected discounted sum is maximized by an expert.
-IRL is a framework in which it could be possible to attack the apprenticeship learning problem mentionned earlier.\\
+This paper focuses on the first family of algorithms, and more
+precisely on its founding algorithm~\citep{abbeel2004apprenticeship}. All of them
+rely on the computation of the feature expectation (which depends on
+policies but not on rewards) of (i) the expert and (ii) some
+intermediate policies. The expert's feature expectation is computed
+using a simple Monte Carlo approach (which requires full
+trajectories of the expert). Other feature expectations are either
+computed exactly (which requires knowing analytically the dynamic of
+the system) or with a Monte Carlo approach (which requires to
+simulate the system). The contribution of this paper is LSTD-$\mu$,
+a new temporal-difference-based algorithm to estimate these feature
+expectations. It relaxes the predeceasing assumptions: transitions
+of the expert are sufficient (rather than full trajectories) and nor
+the model neither a simulator are necessary to compute intermediate
+feature expectations. This paper focuses on the algorithm introduced
+in~\citep{abbeel2004apprenticeship}, but we are confident that the proposed
+approach can be extended to other algorithms based on feature
+expectation computation~\citep{syed2008game,syed2008apprenticeship,ziebart2008maximum}.
 
-\citet{ng2000algorithms} first proposed an algorithm to solve the IRL problem, sheding light on a problem foreseen in \citep{russell1998learning} : the IRL problem is mathematically ill-defined as it admits a infinity of solutions, including degenerative ones (e.g.
-the null reward accepts any policy as an optimal policy).\\
-
-This work was followed by an article \citep{abbeel2004apprenticeship} focussing on the more specific, but better defined problem of apprenticeship learning using the IRL framework.
-This work makes use of the IRL framework by using an approximator for the reward function, and looking for the parameters for which the approximated reward function is the best.
-This however only concerns the apprenticeship learning (a.k.a.
-imitation learning) because the metric used, although intended to discriminate the expert's performance and the current policy's performance \emph{under the current approximate reward}, can be seen as a measure of proximity of the different states visitation frequencies.
-
-This metric, called the \emph{feature expectation}, is akin to the expected discounted reward discussed when describing the RL framework.
-It does however not sum the reward but the features of the visited states, features defined by the choice of the reward function approximator.
-It can be assessed either by analytical resolution (when the model is available) of a Monte-Carlo estimation (when a simulator is available).\\
-
-The main limitation in \citet{abbeel2004apprenticeship}'s work (from now on reffered to as the ANIRL algorithm) is the need for a model of the environnement or, at least, a simulator.
-We build directly upon this work by presenting a new algorithm, LSTD$\mu$, which along with the use of LSPI and LSTD allows the use of the ANIRL algorithm in a batch, off-policy, model-free setting.\\
-
-FIXME:Autres approches qu'Abbeel, notamment celles qui cherchent à contourner le même problème.\\
+The rest of this paper is organized as follows.
+Section~\ref{sec:back} provides the necessary background,
+notably the definition of feature expectation and its use in the
+seminal IRL algorithm of~\citep{abbeel2004apprenticeship}.
+Section~\ref{sec:lstdmu} presents LSTD-$\mu$, our main contribution.
+Section~\ref{sec:exp} provides some preliminary experiments and
+Section~\ref{sec:conclusion} opens perspectives.
 
@@
 \section{LSTD-$\mu$}
+\label{sec:lstdmu}
 The similarity between equations \ref{eqn:mu} and \ref{eqn:V} (both are Bellman equations) allows the use of the LSTD algorithm in order to estimate the feature expectation of a policy.\\
@@
 \section{Experimental benchmark}
+\label{sec:exp}
 The experimental benchmark chosen here is the same as in \citep{ng2000algorithms}, a 5x5 grid world.
@@
 \end{figure}
-\section{Future Work}
-FIXME: Une seule itération, est-ce normal ?
-Abbeel reste-t-il meilleur sur un problème plus complexe ?
-Transfert de tâche ?
-Est-il nécessaire de choisir sa politique avec LSTDMu, puisqu'il finit en une seule fois ?
 \section{Conclusion}
+\label{sec:conclusion}
 \bibliographystyle{named}

### 2011-03-28 main.tex
Background de Matthieu

@@
 \usepackage{algorithmic}
+\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
 
@@
 \section{Background}
-\label{sec:back}
-A \emph{Markov Decision Process} is a tuple $\{S,A,\gamma,R\}$ with :
-\begin{itemize}
-\item $S$ the state space,
-\item $A$ the action space,
-\item $\gamma$ the discount factor,
-\item $R$ the reward function
-\end{itemize}
+A sequential decision problem is often framed as a Markov Decision
+Process (MDP)~\citep{puterman1994markov}. A MDP is a tuple
+$\{S,A,P,R,\gamma\}$ with $S$ being the state space, $A$ the action
+space, $P\in\mathcal{P}(S)^{S\times A}$ the set of Markovian
+transition probabilities, $R\in\mathbb{R}^S$ the reward function
+(assumed to be absolutely bounded by 1) and $\gamma\in[0,1[$ a
+discounting factor. A policy $\pi\in A^S$ maps states to action. The
+quality of a policy is quantified by the associated value function
+$V^\pi$, which associates to each state the expected and discounted
+cumulative reward:
+\begin{equation}
+  V^\pi(s) = E[\sum_{t=0}^\infty \gamma^t R(s_t)|s_0=s, \pi]
+\end{equation}
+Dynamic programming aims at finding the optimal policy $\pi^*$, that
+is one of the policies associated to the optimal value function,
+$V^* = \argmax_\pi V^\pi$, which maximizes the value for each state.
+If the model (that is transition probabilities and the reward
+function) is unknown, learning the optimal control through
+interactions is addressed by reinforcement learning.
 
-A policy $\pi$ is a map from the states to the actions : $\pi:S\rightarrow A$.\\
+For inverse reinforcement learning, the problem is reversed. It is
+assume that an expert acts according to an optimal policy $\pi_E$,
+this policy being optimal according to some unknown reward function
+$R^*$. The goal of inverse reinforcement learning is to learn this
+reward function from sampled trajectories of the expert. This is a
+difficult and ill-posed problem~\citep{ng2000algorithms}. Apprenticeship
+learning through IRL, which is the problem at hand in this paper,
+has a somehow weaker objective: it aims at learning a policy
+$\hat{\pi}$ which is (approximately) as good as the expert policy
+$\pi_E$ under the unknown reward function $R^*$, for a known initial
+state $s_0$ (this condition can be weakened by assuming a
+distribution for initial states, this is not done here for clarity
+of exposition). Now, the approach proposed in~\citep{abbeel2004apprenticeship} is
+presented.
 
-The \emph{Reinforcement Learning} problem seeks the policy that maximize the expected discounted reward.
-When following a policy $\pi$ from state $s\in S$, the expected discounted reward is described by the \emph{value function} $V^\pi$ : \\
+We assume that the true reward function belongs to some hypothesis
+space $\mathcal{H}_\phi = \{\theta^T \phi(s),
+\theta\in\mathbb{R}^p\}$, of which we assume the basis functions to
+be bounded by 1: $|\phi_i(s)|\leq 1, \forall s\in S, 1\leq i \leq
+p$. Therefore, there exists a parameter vector $\theta^*$ such that:
 \begin{equation}
-V^\pi(s) = E\left[\left.\sum\limits_{t=0}^{\infty}\gamma^tr_t\right|s_0=s,\pi\right]
-\label{eqn:V}
+  R^*(s) = (\theta^*)^T \phi(s)
 \end{equation}
+In order to ensure that rewards are bounded, we impose that
+$\|\theta\|_2\leq 1$. For any reward function belonging to
+$\mathcal{H}_\phi$ and for any policy $\pi$, the related value
+function $V^\pi(s)$ can be expressed as follows:
+\begin{align}
+  V^\pi(s) &= E[\sum_{t=0}^\infty \gamma^t \theta^T \phi(s_t)|s_0=s, \pi]
+  \\
+  &= \theta^T  E[\sum_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
+\end{align}
+Therefore, the value function is also linearly parameterized, with
+the same weights and with basis functions being grouped into the
+so-called \emph{feature expectation} $\mu^\pi$:
+\begin{equation}
+  \mu^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
+\end{equation}
+Recall that the problem is to find a policy whose performance is
+close to that of the expert's for the starting state $s_0$, on the
+unknown reward function $R^*$. In order to achieve this goal, it is
+proposed in~\citep{abbeel2004apprenticeship} to find a policy $\tilde{\pi}$ such
+that $\|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2\leq \epsilon$
+for some (small) $\epsilon>0$. Actually, this ensures that the value
+of the expert's policy and the value of the estimated policy (for
+the starting state $s_0$) are close for \emph{any} reward function
+of $\mathcal{H}_\phi$:
+\begin{align}
+  |V^{\pi_E}(s_0) - V^{\tilde{\pi}}(s_0)| &=
+  |\theta^T(\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0))|
+  \\
+  &\leq \|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2
+\end{align}
+This last equation uses the Cauchy-Schwartz inequality and the
+assumption that $\|\theta\|_2\leq 1$. Therefore, the approach
+described here does not ensure to retrieve the true reward function
+$R^*$, but to act as well as the expert under this reward function
+(and actually under any reward function, notably the unknown optimal
+one).
 
-The state space $S$ is projected into a feature space generated by the function $\psi : S \rightarrow \mathbb{R}^p$.
-Both the state space and the action space are projected into a feature space generated by the function $\phi : S\times A \rightarrow \mathbb{R}^k$.\\
+Let us now describe the algorithm proposed in~\citep{abbeel2004apprenticeship} to
+achieve this goal:
+\begin{enumerate}
+  \item Starts with some initial policy $\pi^{(0)}$ and compute
+  $\mu^{\pi^{(0)}}(s_0)$. Set $j=1$;
+  %
+  \item Compute $t^{(j)} = \max_{\theta: \|\theta\|_2\leq 1}
+  \min_{k\in\{0,j-1\}}\theta^T(\mu^{\pi_E}(s_0) -
+  \mu^{\pi^{(k)}}(s_0))$ and let $\theta^{(j)}$ be the value
+  attaining this maximum. At this step, one searches for the reward
+  function which maximizes the distance between the value of the
+  expert at $s_0$ and the value of \emph{any} policy computed so far
+  (still at $s_0$). This optimization problem can be solved using a
+  quadratic programming approach or a projection
+  algorithm~\citep{abbeel2004apprenticeship};
+  %
+  \item if $t^{(j)}\leq \epsilon$, terminate. The algorithm outputs a
+  set of policies $\{\pi^{(0)}, \dots, \pi^{(j-1)}\}$ among which
+  the user chooses manually or automatically the closest to the
+  expert (see~\citep{abbeel2004apprenticeship} for details on how to choose this
+  policy\footnote{A simple solution, not mentioned in~\citep{abbeel2004apprenticeship},
+  is to choose the policy of which the feature expectation is the
+  closest to the one of the expert.}).
+  Notice that the last policy is not necessary the best;
+  %
+  \item solve the MDP with the reward function $R^{(j)}(s) =
+  (\theta^{(j)})^T\phi(s)$ and denote $\pi^{(j)}$ the associated
+  optimal policy. Compute $\mu^{\pi^{(j)}}(s_0)$;
+  %
+  \item set $j\leftarrow j+1$ and go back to step 2.
+\end{enumerate}
+There remains three problems: computing the feature expectation of
+the expert, solving the MDP and computing feature expectations of
+intermediate policies.
 
-The value function is approximated via a matrix of coefficients $\omega$, such as :
+As suggested in~\citep{abbeel2004apprenticeship}, solving the MDP can be done
+approximately by using any appropriate reinforcement learning
+algorithm. In this paper, we use the Least-Squares Policy Iteration
+(LSPI) algorithm~\citep{lagoudakis2003least}. There remains to estimate
+feature expectations. In~\citep{abbeel2004apprenticeship}, $\mu^{\pi_E}(s_0)$ is
+estimated using a Monte Carlo approach over $m$ trajectories:
 \begin{equation}
-\hat V^\pi(s) = \arg\max_{a\in A}\omega_\pi^T\phi(s,a)
+  \hat{\mu}_E(s_0) = \frac{1}{m} \sum_{h=1}^m \sum_{t=0}^\infty
+  \gamma^i \phi(s_t^{(h)})
 \end{equation}
+This approach does not hold if only transitions of the expert are
+available, or if trajectories are too long (in this case, it is
+still possible to truncate them). For intermediate policies, it is
+also suggested to estimate associated feature expectations using a
+Monte Carlo approach (if they cannot be computed exactly). This is
+more constraining than for the expert, as this assumes that a
+simulator of the system is available. In order to address these
+problems, we introduce a temporal-difference-based algorithm to
+estimate these feature expectations.
 
-The \emph{expert} is an agent whose interactions with the world are recorded as a set of transitions : $D_E = \left\{\{s_i,a_i,s'_i,r_i\}_i\right\}$.\\
-
-$D$ is a random transition source.\\
-
-The feature expectation is a mapping from a policy $\pi$ to a function $\mu_\pi : S \rightarrow \mathbb{R}^p$ which gives for a state $s\in S$ the expected discounted value for the features of the states visited when following policy $\pi$ from state $s$ :
-\begin{equation}
-\mu_\pi(s) = E\left[\left.\sum\limits_{t=0}^{\infty}\gamma^t\psi(s_t)\right|s_0=s,\pi\right]
-\label{eqn:mu}
-\end{equation}\\
-The feature expectation of the expert is denoted $\mu_E$.\\
-
-The IRL problem seeks the reward the expert is maximizing.
-The imitation problem seeks the reward which, maximized by one of the many RL tools, will give a behavior as close to the expert's as possible.\\
-
-\citet{abbeel2004apprenticeship}'s algorithm solves the imitation pproblem by approximating this reward with a matrix of coefficients $\theta$ such as :
-\begin{equation}
-\hat R(s) = \theta^T\phi(s)
-\end{equation}
-
-The quality of the imitation is assessed via the error term $||\mu_\pi(s_0)-\mu_E(s_0)||_2$.
-Given that the norm of $\theta$ is bounded, this error term also gives a feeling of the difference of performance under the estimated reward function $\hat R$ :
-\begin{equation}
-|V^\pi(s_0)-V^E(s_0)| \leq ||\theta||_2||\mu_\pi(s_0)-\mu_E(s_0)||_2
-\label{eqn:vmu}
-\end{equation}
-
-The main shortcoming of this algorithm lies in its need for either a model of the environnement or at least a sample source $D$, in order to approximate the feature expectation of both the expert and the current policy via a Monte-Carlo estimator.
 \section{LSTD-$\mu$}

### 2011-03-28 main.tex
LSTDmu de matthieu

@@
 \section{LSTD-$\mu$}
-\label{sec:lstdmu}
-The similarity between equations \ref{eqn:mu} and \ref{eqn:V} (both are Bellman equations) allows the use of the LSTD algorithm in order to estimate the feature expectation of a policy.\\
+Let us write the definition of the $i^\text{th}$ component of a
+feature expectation $\mu^\pi(s)$ for some policy $\pi$:
+\begin{equation}
+  \mu_i^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi_i(s_t)|s_0=s,\pi]
+\end{equation}
+This is exactly the definition of the value function of the policy
+$\pi$ for the MDP considered with the $i^\text{th}$ basis function
+$\phi_i(s)$ as the reward function. There exists many algorithms to
+estimate a value function, any of them can be used to estimate
+$\mu_i^\pi$. Based on this remark, we propose to use specifically
+the least-squares temporal difference (LSTD)
+algorithm~\citep{bradtke1996linear} to estimate each component of the
+feature expectation (as each of these components can be understood
+as a value function related to a specific and known reward
+function).
 
-We introduce the matrix $\zeta$ such as :
+Assume that a set of transitions $\{(s_t,r_t,s_{t+1})_{1\leq t \leq
+n}\}$ sampled according to the policy $\pi$ is available. We assume
+that value functions are searched for in some  hypothesis space
+$\mathcal{H}_\psi = \{ \hat{V}_\xi(s) = \sum_{i=1}^q \xi_i \psi_i(s)
+= \xi^T \psi(s), \xi\in\mathbb{R}^q\}$. As reward and value
+functions are possibly quite different, another hypothesis space is
+considered for value function estimation. But if $\mathcal{H}_\phi$
+is rich enough, one can still consider
+$\mathcal{H}_\psi=\mathcal{H}_\phi$. Therefore, we are looking for
+an approximation of the following form:
 \begin{equation}
-\hat \mu_\pi (s) = \zeta^T\phi(s,\pi(s))
+  \hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)
 \end{equation}
-$\zeta$ is given by :
+The parameter vector $\xi_i^*$ is here the LSTD estimate:
 \begin{equation}
-\zeta = \tilde A^{-1}\tilde b
+  \xi_i^* = \left(\sum_{t=1}^n
+  \psi(s_t)(\psi(s_t)-\gamma\psi(s'_{t}))^T\right)^{-1}
+  \sum_{t=1}^n \psi(s_t) \phi_i(s_t)
 \end{equation}
-with :
-\begin{eqnarray}
-\tilde A = \sum_{\{s,a,s'\} \in D}\phi(s,a)\left(\phi(s,a) - \gamma \phi(s',\pi(s'))\right)^T\\
-\tilde b = \sum_{\{s,a,s'\} \in D} \phi(s,a)\psi(s)^T
-\end{eqnarray}
+For apprenticeship learning, we are interested more particularly in
+$\hat{\mu}^\pi(s_0)$. Let $\Psi = (\psi_i(s_t))_{t,i}$ be the
+$n\times q$ matrix of values predictors, $\Delta\Psi = (\psi_i(s_t)
+- \gamma\psi_i(s'_t))_{t,i}$ be the related $n\times q$ matrix and
+$\Phi = (\phi_i(s_t))_{t,i}$ the $n\times p$ matrix of reward
+predictors. It can be easily checked that $\hat{\mu}^\pi(s_0)$
+satisfies:
+\begin{equation}
+  (\hat{\mu}^\pi(s_0))^T = \psi(s_0)^T (\Psi^T \Delta\Psi)^{-1} \Psi^T \Phi
+\end{equation}
+This provides an efficient way to estimate the feature expectation
+of the expert in $s_0$.
 
-This approach is superior to the Monte-Carlo estimation suggested by \citet{abbeel2004apprenticeship}, as it allows for a batch, off policy estimation of $\mu_\pi$, even when no model of the environment is known and no simulator is available whereas the Monte-Carlo estimation needs either a model or a simulator, and only allow for an estimation of $\mu_\pi(s_0)$.\\
+There remains to compute the feature expectations of intermediate
+policies, which should be done in an off-policy manner (that is
+without explicitly sampling trajectories according to the policy of
+interest). To do so, still interpreting each component of the
+feature expectation as a value function, we introduce a state-action
+feature expectation defined as follows (much as the classic
+$Q$-function extends the value function):
+\begin{equation}
+  \mu^\pi(s,a) = E[\sum_{t=0}^\infty \gamma^t
+  \phi(s_t)|s_0=s,a_0=a,\pi]
+\end{equation}
+Compared to the classic feature expectation, this definition adds a
+degree of freedom on the first action to be chosen before following
+the policy $\pi$. With a slightly different definition of the
+related hypothesis space, each component of this feature expectation
+can still be estimated using the LSTD algorithm (namely using the
+LSTD-Q algorithm~\citep{lagoudakis2003least}). The clear advantage of
+introducing the state-action feature expectation is that this
+additional degree of freedom allows off-policy learning. Assuming
+that $\mathcal{H}_\psi$ defines state-action feature vectors and
+that a set of transitions $\{(s_t,a_t,r_t,s_{t+1})_{1\leq t \leq
+n}\}$ sampled according to some behaviorial policy $\pi_0$ is
+available, the parameter vector $\xi_i^*$ related to the
+$i^\text{th}$ component of the state-action feature expectation
+$\mu^\pi(s)$ (policy $\pi$ being different from policy $\pi_0$) is
+given by:
+\begin{equation}
+  \xi_i^* = \left(\sum_{t=1}^n
+  \psi(s_t,a_t)(\psi(s_t,a_t)-\gamma\psi(s'_{t},\pi(s'_t)))^T\right)^{-1}
+  \sum_{t=1}^n \psi(s_t,a_t) \phi_i(s_t)
+\end{equation}
+Extending LSTD-$\mu$ to state-action LSTD-$\mu$ is done in the same
+manner as LSTD is extended to LSTD-Q~\citep{lagoudakis2003least},
+technical details are not provided here for the clarity of
+exposition.
 
-Coupled with the LSPI algorithm, this allows an adaptation of \citet{abbeel2004apprenticeship} algorithm to "real-life" settings where the only data available come from the expert's trajectories.\\
+Given the (state-action) LSTD-$\mu$ algorithm, the apprenticeship
+learning algorithm of~\citep{abbeel2004apprenticeship} (see
+Section~\ref{sec:back}) can be easily extended to a batch and
+off-policy setting. The solely available data is a set of
+transitions sampled according to the expert policy. The
+corresponding feature expectation for the starting state $s_0$ is
+estimated with the LSTD-$\mu$ algorithm. At step~4 of this
+algorithm, the MDP is (approximately) solved using
+LSPI~\citep{lagoudakis2003least} (an approximate policy iteration
+algorithm using LSTD-Q as the off-policy $Q$-function estimator).
+The corresponding feature expectation at state $s_0$ is estimated
+using the proposed state-action LSTD-$\mu$.
 
-FIXME: Un mot sur pourquoi il est inutile de faire des itérations sur mu directement ?
+Before presenting some experimental results, let us stress that
+LSTD-$\mu$ is simply the LSTD algorithm applied to a specific reward
+function. Although quite clear, the idea of using a temporal
+difference algorithm to estimate the feature expectation is new, as
+far as we know. A clear advantage of the proposed approach is that
+any theoretical result holding for LSTD also holds for LSTD-$\mu$,
+such as convergence~\citep{nedic2003least} or
+finite-sample~\citep{lazaric2010finiteLSTD} analysis for example.
 
-\section{Analysis}
-FIXME: Matthieu ?
 \section{Experimental benchmark}

### 2011-03-28 main.tex
Modifié ma section pour coller à ce que MAtthieu a écrit

@@
 \section{Background}
+\label{sec:back}
 A sequential decision problem is often framed as a Markov Decision
@@
   |\theta^T(\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0))|
+  \label{eqn:vs0}
   \\
@@
   \mu_i^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi_i(s_t)|s_0=s,\pi]
+  \label{eqn:phi}
 \end{equation}
@@
   \hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)
+  \label{eqn:psi}
 \end{equation}
@@
 \label{sec:exp}
+\subsection{Experiment description and results}
 The experimental benchmark chosen here is the same as in \citep{ng2000algorithms}, a 5x5 grid world.
@@
 The reward optimized by the expert is 0 everywhere but in the upper-right square, where it is 1.
-For every policy, an episodes ends when the upper right square is attained, or after 20 steps.
+For every policy, an episode ends when the upper right square is attained, or after 20 steps.
 At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is).\\
@@
 Both the state and action spaces are finite and of small cardinality.
-Hence, the chosen feature functions $\phi$ and $\psi$ (see section \ref{sec:back}) are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
-There is a bijective mapping from the input space elements ($S$ for $\psi$, $S \times A$ for $\phi$) to the indexes $i$ for which the coordinate of the feature is 1.\\
+Hence, the chosen feature functions $\phi$ and $\psi$ (see equations \ref{eqn:phi} and \ref{eqn:psi} for a remainder of their meaning) are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
+There is a bijective mapping from the input space elements to the indexes $i$ for which the coordinate of the feature is 1.\\
 
 Both \citet{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side.
-The pseudo code of the implementation we used can be found figure \ref{fig:alg}.\\
-
 The MDP solver of mc\_ANIRL is LSPI with a (very) big matrix $D$ as a sample source.
@@
 We thus are in the case intended by \citet{abbeel2004apprenticeship}.\\
-
 On the other hand lstd\_ANIRL is used as if it could not have access to the simulator.
@@
 We want to compare the porformance of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
-The discussion about the choice of the performance metric has its own dedicated section (section \ref{sec:perf}).
+The discussion about the choice of the performance metric has its own dedicated subsection (subsection \ref{sec:perf}).
 We use here the $||\mu(s_0) - \mu_E(s_0)||_2$ error term.\\
@@
 Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu(s_0)-\mu_E(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
-The best policy is found by lstd\_ANIRL after one iteration only (this is being investigated), whereas in mc\_ANIRL convergence happens after at least ten iterations, and the best policy is not always the last.
+The best policy is found by lstd\_ANIRL after one iteration only\footnote{Precise reasons for what it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in mc\_ANIRL convergence happens after at least ten iterations and the best policy is not always the last.
 The score of the best policy (not the last) is what is plotted here.\\
@@
 The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version (for example controlling a machine tool working on expensive materials).
-\section{Quality criterion}
-
+\subsection{Discussion about the quality criterion}
 Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion.
@@
 The $||\mu(s_0) - \mu_E(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
-It bears an interesting relation with the difference between the expert's value function and the current value function with respect \emph{to the current reward} (equation \ref{eqn:vmu}).\\
+It bears an interesting relation with the difference between the expert's value function and the current value function in the initial state with respect \emph{to the current reward} (equation \ref{eqn:vs0}).\\
 
@@
 This however can not be said of lstd\_ANIRL, for which the two curves are indeed different (Figure \ref{fig:B}).
-Not knowing the true value of $\mu_E(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm, $t = ||\bar\mu-\hat\mu_E(s_0)||$.
-After the algorithm has converged (it systematically does after only one iteration, this is being investigated), there is no way to know whether the policy is actually good or not.\\
+Not knowing the true value of $\mu_E(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm\footnote{This equation differs from what has been given as the second step of the algorithm in section \ref{sec:back}.
+This is because we implemented the version of the algorithm that uses a projection (see \citet{abbeel2004apprenticeship}'s supplementary material for an in-depth discussion of both versions).
+Although the mathematical expressions are different, both convey the same meaning and are exchangeable.}, $t = ||\bar\mu-\hat\mu_E(s_0)||$.
+After the algorithm has converged there is no way to know whether the policy is actually good or not other than actually testing it.\\
 
@@
 \section{Conclusion}
+
 \label{sec:conclusion}

### 2011-03-28 main.tex
Intro d'Olivier

@@
 \section{Introduction}
-The optimal control problem can be expressed as having a machine
-control a system, with the goal of fulfilling a specific task.
-Various solutions to this problem exist. Notably, the reinforcement
-learning (RL) paradigm~\citep{sutton1998reinforcement} is a general framework in
-which an agent learns to control optimally a dynamic system through
-interactions with it. The task is specified through a reward
-function, the agent objective being to take sequential decisions so
-as to maximize the expected cumulative reward.
 
-However, things can get hard to deal with  when precisely defining
-the task is in itself a challenge. When the system can be
-empirically controlled by an expert, even when its behavior is
-difficult to describe formally, apprenticeship learning is a way to
-let a machine control the system by mimicking the expert. Rather
-than directly mimicking the expert with some supervised learning
-approach, inverse reinforcement learning (IRL)~\citep{ng2000algorithms}
+Optimal control consist in having a machine controlling a system
+with the goal of fulfilling a specific task, optimality being
+defined as how well the task is performed. Various solutions to this
+problem exist from automation to planification. Notably, the
+reinforcement learning (RL) paradigm~\citep{sutton1998reinforcement} is a general
+machine learning framework in which an agent learns to control
+optimally a dynamic system through interactions with it. The task is
+specified through a reward function, the agent objective being to
+take sequential decisions so as to maximize the expected cumulative
+reward.
+
+However, defining optimality (and thus the reward function) can
+itself be a challenge. If the system can be empirically controlled
+by an expert, even though his/her behavior can be difficult to
+describe formally, apprenticeship learning is a way to have a
+machine controlling the system by mimicking the expert. Rather than
+directly mimicking the expert with some supervised learning
+approach, Inverse Reinforcement Learning (IRL)~\citep{ng2000algorithms}
 consists in learning a reward function under which the policy
@@
 expert's actions can be guessed in states which have not been
-encountered during demonstration.
+encountered during demonstration. Firstly introduced in \citep{russell1998learning}, another advantage claimed by the author would be to
+find a compact and complete representation of the task in the form
+of the reward function.
 
-Basically, there are three families of IRL algorithms:
-feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008game,syed2008apprenticeship,ziebart2008maximum,boularias2011bootstrapping},
-marge-maximization-based~\citep{ratliff2006maximum,ratliff2007boosting,ratliff2007imitation,kolter2008hierarchical}
+There roughly exist three families of IRL algorithms:
+feature-expectation-based~\cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
+marge-maximization-based~\cite{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
 and approaches based on the parameterization of the policy by the
-reward function~\citep{Ramachandran2007bayesian,Neu2007apprenticeship}. The first family
+reward function~\cite{ramachandran2007bayesian,neu2007apprenticeship}. The first family
 assumes a linearly parameterized reward function. This naturally
@@
 This paper focuses on the first family of algorithms, and more
-precisely on its founding algorithm~\citep{abbeel2004apprenticeship}. All of them
+precisely on the seminal work of~\cite{abbeel2004apprenticeship}. All of them
 rely on the computation of the feature expectation (which depends on
@@
 trajectories of the expert). Other feature expectations are either
-computed exactly (which requires knowing analytically the dynamic of
-the system) or with a Monte Carlo approach (which requires to
+computed exactly (which requires knowing analytically the dynamics
+of the system) or with a Monte Carlo approach (which requires to
 simulate the system). The contribution of this paper is LSTD-$\mu$,
@@
 feature expectations. This paper focuses on the algorithm introduced
-in~\citep{abbeel2004apprenticeship}, but we are confident that the proposed
+in~\cite{abbeel2004apprenticeship}, but we are confident that the proposed
 approach can be extended to other algorithms based on feature
-expectation computation~\citep{syed2008game,syed2008apprenticeship,ziebart2008maximum}.
+expectation computation~\cite{syed2008apprenticeship,syed2008game,ziebart2008maximum}.
 
@@
 notably the definition of feature expectation and its use in the
-seminal IRL algorithm of~\citep{abbeel2004apprenticeship}.
+seminal IRL algorithm of~\cite{abbeel2004apprenticeship}.
 Section~\ref{sec:lstdmu} presents LSTD-$\mu$, our main contribution.

### 2011-03-28 main.tex
Conclusion

@@
 \section{LSTD-$\mu$}
+\label{sec:lstdmu}
 Let us write the definition of the $i^\text{th}$ component of a
@@
 When this term is low, the objective performance (that is, $V^E(s_0)-V^\pi(s_0)$ with respect to the unknow true reward function) is low too.
+\section{Conclusion}
+\label{sec:conclusion}
+Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning vie inverse reinforcement learning to a batch, model-free, off-policy setting at the cost of the need for a bigger number of samples from the expert.
+We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach.
+Furthermore, our method requires less computing power as it converges after only one iteration (that is, as far as we witnessed in our experimental setting).\\
+
+The simple idea of using LSTD to estimate the feature expectation can be applied to other algorithms as well, for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
+
+The exact reason why LSTD$\mu$ converges so fast should be investigated and a bigger experiment should be run, possibly on a continuous problem.\\
 \begin{figure}
@@
 \end{figure}
-\section{Conclusion}
-
-\label{sec:conclusion}
 \bibliographystyle{named}

### 2011-03-28 main.tex
Stye JFPDA

@@
-\typeout{Batch, Off-Policy Apprenticeship Learning}
-\documentclass{article}
-\usepackage{ijcai11}
-\usepackage{times}
+\documentclass{jfpda2011}
+% Ne pas modifier ========================================================================
+\usepackage{amssymb}
+\usepackage{natbib}
+
+% Typographie française ==================================================================
+\usepackage[english,frenchb]{babel} % typographie francaise
+\usepackage[utf8]{inputenc} % saisie de caracteres accentues, etc.
+\usepackage{lmodern} % polices vectorielles
+\usepackage{xspace} % gestion des espaces a la francaise
+\AddThinSpaceBeforeFootnotes % appel de note legerement decale du mot prec.
+\FrenchFootnotes % numeros de notes a la francaise
+
+% Vos environnements =====================================================================
+\newtheorem{prop}{Proposition}
+\newenvironment{pf}{{\bf Proof }}{\hfill$\Box$\par}
+\newcommand{\spaceafterproof}{\vspace{1em}}
+
+% Corps de l'article =====================================================================
 \usepackage{color}
@@
 %\usepackage{latexsym} 
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
+%\usepackage{amsthm}
+%% \newtheorem{theorem}{Theorem}
+%% \newtheorem{lemma}{Lemma}
 \usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{natbib}
 \usepackage{graphicx}
@@
 
-\title{Batch, Off-Policy Apprencticeship Learning}
-\author{Edouard Klein \and Matthieu Geist \and Olivier Pietquin \\
-Supélec, IMS\\
-Metz, France \\
-\{edouard.klein$|$matthieu.geist$|$olivier.pietquin\}@supelec.fr}
+\title{Batch, Off-Policy, Model-free Apprencticeship Learning}
+%\author{Robert}
+%   1. Supelec-Metz Campus, IMS Reasearch group, France. \texttt{prenom.nom@supelec.fr}}
+\author{Edouard Klein\inst{1}\inst{3} \and Matthieu Geist\inst{1} \and Olivier Pietquin\inst{1}\inst{2}}
+ \institute{
+ 1. Supélec-Metz Campus, IMS Research group, France, \texttt{nom.prenom@supelec.fr}\\
+ 2. UMI 2958 CNRS - GeorgiaTech, France\\
+ 3. Equipe ABC, Loria, INRIA Grand-Est, Nancy, France.
+ }
+\date{\today}
+
 \begin{document}
+
 \maketitle
@@
 \begin{figure}
-\includegraphics[width=.5\textwidth]{../Code/courbe_e.pdf}
+\includegraphics[width=\textwidth]{../Code/courbe_e.pdf}
 \caption{$||\mu_E(s_0)-\mu_\pi(s_0)||_2$ with respect to the number of samples available from the expert.
@@
 \begin{figure}
-\includegraphics[width=.5\textwidth]{../Code/courbe_a.pdf}
+\includegraphics[width=\textwidth]{../Code/courbe_a.pdf}
 \caption{Different criteria with respect to the number of iterations for a run of mc\_ANIRL.}
@@
 \begin{figure}
-\includegraphics[width=.5\textwidth]{../Code/courbe_b.pdf}
+\includegraphics[width=\textwidth]{../Code/courbe_b.pdf}
 \caption{Different criteria with respect to the number of samples from the expert, for several runs of lstd\_ANIRL.}
@@
 \end{figure}
-\bibliographystyle{named}
+%\bibliographystyle{named}
 \bibliography{../Biblio/Biblio}

### 2011-03-28 main.tex
Abstract Olivier et Matthieu

@@
 \begin{abstract}
-We consider the IRL framework for apprenticeship learning. %1
-Current IRL algorithms use a simple Monte-Carlo estimation to approximate the feature expectation of both the expert and the policy under evaluation. %2
-In this paper, we propose another method for estimating the feature expectation based on the Least Square Temporal Difference (LSTD) family of algorithms. %3
-This new method allows for batch, off-policy appreticeship learning in situations where the dynamics of the system remains unknown. %4
-By testing it on a classical problem, we show that our method is empirically sound.%5
-We also provide some theoretical results.%6
+  This paper adresses the problem of apprenticeship learning, that
+  is learning control policies from demonstration by an expert. An
+  efficient framework for it is inverse reinforcement learning
+  (IRL). Based on the assumption that the expert maximizes a utility
+  function, IRL aims at learning the reward which defines this
+  utility function from example trajectories. Many IRL algorithms
+  assume that the reward function is linearly parameterized and rely
+  on the computation of some associated \emph{feature expectations},
+  which is done through Monte Carlo simulation. However, this
+  assumes to have full trajectories for the expert policy as well as
+  at least a generative model for intermediate policies. In this
+  paper, we introduce a temporal difference method, namely LSTD-$\mu$, to compute these
+  feature expectations. This allows extending apprenticeship
+  learning to a batch and off-policy setting.
 \end{abstract}

### 2011-03-28 main.tex
Traduction française de l'abstract

--- 2011-03-28 main.tex
@@
 \begin{abstract}
-We consider the IRL framework for apprenticeship learning. %1
-Current IRL algorithms use a simple Monte-Carlo estimation to approximate the feature expectation of both the expert and the policy under evaluation. %2
-In this paper, we propose another method for estimating the feature expectation based on the Least Square Temporal Difference (LSTD) family of algorithms. %3
-This new method allows for batch, off-policy appreticeship learning in situations where the dynamics of the system remains unknown. %4
-By testing it on a classical problem, we show that our method is empirically sound.%5
-We also provide some theoretical results.%6
+ Ce papier s'intéresse au problème de l'apprentissage par imitation, c'est à dire la résolution du problème du contrôle optimal à partir de données tirées d'une desmonstration d'expert.
+L'apprentissage par renforcement inverse (IRL) propose un cadre essicace pour résoudre ce problème.
+En se basant sur l'hypothèse que l'expert maximise un critère, l'IRL essaie d'apprendre la récompense qui définit ce critère à partir de trajectoires d'exemple.
+Beaucoup d'algorithmes d'IRL font l'hypothèse de l'existence d'un bon approximateur linéaire pour la fonction de récompense et calculent l'attribut moyen (le cumul moyen pondéré des fonctions de base, relatives à la paramétrisation linéaire supposée de la récompense, évaluées en les états d'une trajectoire associée à une certaine politique)  via une estimation de Monte-Carlo.
+Cela implique d'avoir accès à des trajectoires complète de l'expert ainsi qu'à au moins un modèle génératif pour tester les politiques intermédiaires.
+Dans ce papier nous introduisons une méthode de différence temporelle, LSTD-$\mu$, pour calculer cet attribut moyen.
+Cela permet d'étendre l'apprentissage par imitation aux cas \emph{batch} et \emph{off-policy}.
 \end{abstract}

### 2011-03-29 main.tex
Ce qui a été soumis à JFPDA

@@
 
-\title{Batch, Off-Policy, Model-free Apprencticeship Learning}
+\title{Apprentissage par imitation étendu au cas batch, off-policy et sans modèle}
 %\author{Robert}
@@
  Ce papier s'intéresse au problème de l'apprentissage par imitation, c'est à dire la résolution du problème du contrôle optimal à partir de données tirées d'une desmonstration d'expert.
-L'apprentissage par renforcement inverse (IRL) propose un cadre essicace pour résoudre ce problème.
+L'apprentissage par renforcement inverse (IRL) propose un cadre efficace pour résoudre ce problème.
 En se basant sur l'hypothèse que l'expert maximise un critère, l'IRL essaie d'apprendre la récompense qui définit ce critère à partir de trajectoires d'exemple.
@@
 There roughly exist three families of IRL algorithms:
-feature-expectation-based~\cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
-marge-maximization-based~\cite{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
+feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
+marge-maximization-based~\citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
 and approaches based on the parameterization of the policy by the
-reward function~\cite{ramachandran2007bayesian,neu2007apprenticeship}. The first family
+reward function~\citep{ramachandran2007bayesian,neu2007apprenticeship}. The first family
 assumes a linearly parameterized reward function. This naturally
@@
 This paper focuses on the first family of algorithms, and more
-precisely on the seminal work of~\cite{abbeel2004apprenticeship}. All of them
+precisely on the seminal work of~\citet{abbeel2004apprenticeship}. All of them
 rely on the computation of the feature expectation (which depends on
@@
 feature expectations. This paper focuses on the algorithm introduced
-in~\cite{abbeel2004apprenticeship}, but we are confident that the proposed
+in~\citep{abbeel2004apprenticeship}, but we are confident that the proposed
 approach can be extended to other algorithms based on feature
-expectation computation~\cite{syed2008apprenticeship,syed2008game,ziebart2008maximum}.
+expectation computation~\citep{syed2008apprenticeship,syed2008game,ziebart2008maximum}.
 
@@
 notably the definition of feature expectation and its use in the
-seminal IRL algorithm of~\cite{abbeel2004apprenticeship}.
+seminal IRL algorithm of~\citet{abbeel2004apprenticeship}.
 Section~\ref{sec:lstdmu} presents LSTD-$\mu$, our main contribution.
@@
 A sequential decision problem is often framed as a Markov Decision
-Process (MDP)~\citep{puterman1994markov}. A MDP is a tuple
+Process (MDP)~\citep{puterman1994markov}. An MDP is a tuple
 $\{S,A,P,R,\gamma\}$ with $S$ being the state space, $A$ the action
@@
 For inverse reinforcement learning, the problem is reversed. It is
-assume that an expert acts according to an optimal policy $\pi_E$,
+assumed that an expert acts according to an optimal policy $\pi_E$,
 this policy being optimal according to some unknown reward function
@@
 \label{sec:exp}
+Preliminary experimental results have been obtained.
+Subsection \ref{ssec:exp} details the protocol and the results while subsection \ref{ssec:quality} inspects the meaning of the different quality criteria.
 \subsection{Experiment description and results}
-The experimental benchmark chosen here is the same as in \citep{ng2000algorithms}, a 5x5 grid world.
+\label{ssec:exp}
+The experimental benchmark chosen here is one of those proposed in \citep{ng2000algorithms}, a 5x5 grid world.
 The agent is in one of the case of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
@@
 With probability 0.1, the direction is randomly drawn (and thus have probability 0.25 to match the original direction).
-The reward optimized by the expert is 0 everywhere but in the upper-right square, where it is 1.
+The reward optimized by the expert is 0 everywhere except in the upper-right square, where it is 1.
 For every policy, an episode ends when the upper right square is attained, or after 20 steps.
@@
 Both the state and action spaces are finite and of small cardinality.
-Hence, the chosen feature functions $\phi$ and $\psi$ (see equations \ref{eqn:phi} and \ref{eqn:psi} for a remainder of their meaning) are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
+Hence, the chosen feature functions $\phi$ and $\psi$ (see equations (\ref{eqn:phi}) and (\ref{eqn:psi}) for a remainder of their meaning) are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
 There is a bijective mapping from the input space elements to the indexes $i$ for which the coordinate of the feature is 1.\\
 
-Both \citet{abbeel2004apprenticeship}'s algorithm (mc\_ANIRL) and our adaptation (lstd\_ANIRL) are tested side by side.
-The MDP solver of mc\_ANIRL is LSPI with a (very) big matrix $D$ as a sample source.
-$\mu_\pi(s_0)$ and $\mu_E(s_0)$ are computed thanks to a (very) big Monte-Carlo estimation.
-Both these computations are  as good as perfect theoretical solvers for all intended purpose on this toy problem.
+Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as mc\_ANIRL\footnote{Abbeel \& Ng IRL}) and our adaptation (referred t as lstd\_ANIRL) are tested side by side.
+The MDP solver of mc\_ANIRL is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly.
+$\mu^{\pi^(j)}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to reduce the variance to less than $0.0001\%$ of the mean.
+We consider both these computations as perfect theoretical solvers for all intended purpose on this toy problem.
 We thus are in the case intended by \citet{abbeel2004apprenticeship}.\\
@@
 We want to compare the porformance of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
-The discussion about the choice of the performance metric has its own dedicated subsection (subsection \ref{sec:perf}).
-We use here the $||\mu(s_0) - \mu_E(s_0)||_2$ error term.\\
+Indeed as they are quite costly (in both means and human time) to generate they are often not in abundance hence the critical need for an algorithm to be expert-sample efficient.
+The discussion about the choice of the performance metric has its own dedicated subsection (subsection \ref{ssec:quality}).
+We use here the $||\mu^{\pi^(j)}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term.\\
 
-Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu(s_0)-\mu_E(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
+Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^(j)}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
 The best policy is found by lstd\_ANIRL after one iteration only\footnote{Precise reasons for what it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in mc\_ANIRL convergence happens after at least ten iterations and the best policy is not always the last.
-The score of the best policy (not the last) is what is plotted here.\\
+The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is what is plotted here.\\
 
-We can see that although lstd\_ANIRL is not as good as mc\_ANIRL when very few samples are available, both algorithm quickly converge to almost the same value (we don't know yet if the seemingly slightly better performance of lstd\_ANIRL is an artifact).
-The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version (for example controlling a machine tool working on expensive materials).
+We can see that although lstd\_ANIRL is not as good as mc\_ANIRL when very few samples are available, both algorithm quickly converge to almost the same value (although not low, the number of run so far is not statistically significant enough to allow us to state that the seemingly slightly better performance of lstd\_ANIRL is not an artifact).
+The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version.
 \subsection{Discussion about the quality criterion}
-Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion.
-These plots show four different quality ciretria during a run of mc\_ANIRL and several runs of lstd\_ANIRL, for which the abscissa can not be the number of iterations.\\
+\label{ssec:quality}
+Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion ; these plots show four different quality ciretria during a run of mc\_ANIRL and several runs of lstd\_ANIRL.
+For the latter, the abscissa can not be the number of iterations as it always converges in one iteration.\\
 
-The $||\mu(s_0) - \mu_E(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
+The $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
 It bears an interesting relation with the difference between the expert's value function and the current value function in the initial state with respect \emph{to the current reward} (equation \ref{eqn:vs0}).\\
 
-The fact that  the curves of  $||\mu(s_0) - \mu_E(s_0)||_2$ and $||\hat\mu(s_0) - \hat\mu_E(s_0)||_2$ are indistinguishable in figure \ref{fig:A} means that for it has access to a cheap simulator, mc\_ANIRL is as if it worked within an analitycal framework.
+The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in figure \ref{fig:A} means that for it has access to a cheap simulator, mc\_ANIRL works as if it had access to the exact values.
 This however can not be said of lstd\_ANIRL, for which the two curves are indeed different (Figure \ref{fig:B}).
-Not knowing the true value of $\mu_E(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm\footnote{This equation differs from what has been given as the second step of the algorithm in section \ref{sec:back}.
+Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm\footnote{This equation differs from what has been given as the second step of the algorithm in section \ref{sec:back}.
 This is because we implemented the version of the algorithm that uses a projection (see \citet{abbeel2004apprenticeship}'s supplementary material for an in-depth discussion of both versions).
-Although the mathematical expressions are different, both convey the same meaning and are exchangeable.}, $t = ||\bar\mu-\hat\mu_E(s_0)||$.
+Although the mathematical expressions are different, both convey the same meaning and are exchangeable.}, $t = ||\bar\mu^{(j)}-\hat\mu^{\pi_E}(s_0)||$ where $\bar\mu^{(j)}$ is the projection of $\mu^{\pi_E}$ on $(\mu^{\pi^{(j-1)}},\mu^{\pi^{(j)}})$.
 After the algorithm has converged there is no way to know whether the policy is actually good or not other than actually testing it.\\
@@
 It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in mc\_ANIRL either, as it can be low when the policy is bad.
-The best policy, however, can be easily chosen from the $||\mu(s_0) - \mu_E(s_0)||_2$ term, which mc\_ANIRL can compute.
-When this term is low, the objective performance (that is, $V^E(s_0)-V^\pi(s_0)$ with respect to the unknow true reward function) is low too.
+The best policy, however, can be easily chosen from the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which mc\_ANIRL can compute.
+When this term is low, the objective performance (that is, $V^{\pi_E}(s_0)-V^{\pi^{(j)}}(s_0)$ with respect to the unknow true reward function) is low too.
 \section{Conclusion}
@@
 
-The simple idea of using LSTD to estimate the feature expectation can be applied to other algorithms as well, for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
-
+The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
 The exact reason why LSTD$\mu$ converges so fast should be investigated and a bigger experiment should be run, possibly on a continuous problem.\\
 \begin{figure}
-\includegraphics[width=\textwidth]{../Code/courbe_e.pdf}
-\caption{$||\mu_E(s_0)-\mu_\pi(s_0)||_2$ with respect to the number of samples available from the expert.
+\centering
+\input{../Code/courbe_e}
+\caption{$||\mu^{\pi}(s_0) - \mu^{\pi_E}(s_0)||_2$ where $\pi$ is the policy for which this term is the lowest for the whole run, with respect to the number of samples available from the expert.
 Our algorithm is not as good as \citet{abbeel2004apprenticeship} with very few samples, but make up for it very soon.}
@@
 \end{figure}
-\label{sec:perf}
 \begin{figure}
-\includegraphics[width=\textwidth]{../Code/courbe_a.pdf}
+\centering
+\input{../Code/courbe_a}
 \caption{Different criteria with respect to the number of iterations for a run of mc\_ANIRL.}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../Code/courbe_b.pdf}
+\centering
+\input{../Code/courbe_b}
 \caption{Different criteria with respect to the number of samples from the expert, for several runs of lstd\_ANIRL.}

### 2011-03-31 GridWorld_generator.c
Courbes de la première soumission à JFPDA

@@ -5,7 +5,7 @@
 #include <gsl/gsl_matrix.h>
 #include "utils.h"
 #include "GridWorld.h"
-#define NUMBER_OF_WALKS 100
+#define NUMBER_OF_WALKS 500
 #define MAX_WALK_LENGTH  (GRID_HEIGHT+GRID_WIDTH)
 
 int main(void){

### 2011-03-31 IRL_Globals.h

@@ -36,3 +36,9 @@ extern unsigned int g_iNb_samples;
    policy because only a handful of state action
    couples are visited, thus making the A matrix singular */
 extern double g_dLambda_lstdmu; 
+
+/* When using MC in ANIRL, we store the best score when
+   the measured error is at its lowest yet*/
+extern double g_dBest_error;
+extern double g_dBest_true_error;
+extern double g_dBest_diff;

### 2011-03-31 LSTDmu.c

@@ -263,11 +263,11 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
     gsl_matrix_submatrix( D_mu, 0, g_iS+g_iA+g_iS+g_iP, 
 			  D->size1, 1 ); 
   gsl_matrix_memcpy( &Dmueoe.matrix, &Deoe.matrix );
-  /* \mu \leftarrow LSTD_\mu( D_\mu, k, p, s, a, \phi,
+  /* \mu \leftarrow LSTD\mu( D_\mu, k, p, s, a, \phi,
      \psi, \gamma, \pi ) */
   g_mOmega = omega;
   gsl_matrix* mu = lstd_mu( D_mu, &greedy_policy, s_0 );
-  /* D_E.r \leftarrow \psi(s) */
+  /* D_E.r \leftarrow \psi(D_E.s) */
   gsl_matrix* D_E_mu = 
     gsl_matrix_alloc( D_E->size1, g_iS+g_iA+g_iS+g_iP+1 );
   gsl_matrix_view DEsas_dst = 
@@ -308,9 +308,10 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
   /* \bar\mu \leftarrow \mu*/
   gsl_matrix* bar_mu = gsl_matrix_alloc( g_iP, 1 );
   gsl_matrix_memcpy( bar_mu, mu );
-  /* t = ||\mu_E - \bar\mu||_2*/
+  /* t \leftarrow ||\mu_E - \bar\mu||_2*/
   double t = diff_norm( mu_E, bar_mu );
   unsigned int nb_it = 0;
+  /* while t > \epsilon */
   while( t > g_dEpsilon_anirl ){
     /* Output of the different criteria */
     double empirical_err = diff_norm( mu_E, mu );

### 2011-03-31 abbeel2004apprenticeship.c

@@ -10,6 +10,12 @@
 /* #include "greedy.h" */
 /* #include "GridWorld_simulator.h" */
 
+/* When using MC in ANIRL, we store the best score when
+   the measured error is at its lowest yet*/
+double g_dBest_error;
+double g_dBest_true_error;
+double g_dBest_diff;
+
 /* Compute an estimate of \mu using the monte carlo method,
    given a set of trajectories 
    \hat\mu = {1\over m }\sum_{i=0}^m
@@ -77,6 +83,9 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
   /* t \leftarrow ||\mu_E - \bar\mu||_2*/
   double t = diff_norm( mu_E, bar_mu );
   unsigned int nb_it = 0;
+  g_dBest_error = diff_norm( mu_E, mu );
+  g_dBest_true_error = true_diff_norm( omega );
+  g_dBest_diff = true_V_diff( omega );
   while( t > g_dEpsilon_anirl && nb_it < g_iIt_max_anirl ){
     /* Output of the different criteria */
     double empirical_err = diff_norm( mu_E, mu );
@@ -85,6 +94,11 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
     printf( "%d %d %d %lf %lf %lf %lf\n", 
 	    m, nb_it, g_iNb_samples, 
 	    t, empirical_err, true_err, true_V );
+    if( empirical_err <= g_dBest_error ){
+      g_dBest_error = empirical_err;
+      g_dBest_true_error = true_err;
+      g_dBest_diff = true_V;
+    }
     /* D.r \leftarrow \theta^T\psi(D.s) */
     for( unsigned int i = 0 ; i < D->size1 ; i++ ){
       gsl_matrix_view state = 
@@ -156,6 +170,11 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
   printf( "%d %d %d %lf %lf %lf %lf\n", 
 	  m, nb_it, g_iNb_samples, 
 	  t, empirical_err, true_err, true_V );
+  if( empirical_err <= g_dBest_error ){
+    g_dBest_error = empirical_err;
+    g_dBest_true_error = true_err;
+    g_dBest_diff = true_V;
+  }
   gsl_matrix_free( omega_0 );
   gsl_matrix_free( mu );
   gsl_matrix_free( mu_E );

### 2011-03-31 courbe_b.c

@@ -66,7 +66,7 @@ int main( void ){
   gsl_matrix* omega_expert = lspi( D, omega_0 );
   g_mOmega_E = omega_expert;
   expert_just_set();
-  unsigned int M = 35;
+  unsigned int M = 40;
   g_iNb_samples = D->size1;
   g_mOmega =  omega_expert;
   gsl_matrix* D_expert = gridworld_simulator( M );

### 2011-04-06 LSTDmu.c
I'm about to get rid of some code I am not sur I won't need in the future

@@ -2,6 +2,7 @@
 #include <gsl/gsl_blas.h>
 #include <gsl/gsl_permutation.h>
 #include <gsl/gsl_linalg.h>
+#include <math.h>
 #include "greedy.h"
 #include "utils.h"
 #include "criteria.h"
@@ -116,6 +117,7 @@ gsl_matrix* lstd_mu_op(  gsl_matrix* D_mu ){
 		  omega_pi, phi_s_pi_s, 0.0, mu );
   gsl_matrix_free( omega_pi );
   gsl_matrix_free( s_pi_s );
+  gsl_matrix_free( phi_s_pi_s );
   gsl_matrix_free( a_0 );
   gsl_matrix_free( s_0 );
   return mu;
@@ -216,6 +218,7 @@ gsl_matrix* lstd_mu(  gsl_matrix* D_mu,
   gsl_matrix_free( omega_pi );
   gsl_matrix_free( s_pi_s );
   gsl_matrix_free( pi_s_src );
+  gsl_matrix_free( phi_s_pi_s );
   return mu;
 }
 
@@ -297,6 +300,7 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
   /* \mu_E \leftarrow on-LSTD_\mu( D_E, k, p, s, a, \psi,
      \phi, \gamma) */
   gsl_matrix* mu_E = lstd_mu_op( D_E_mu );
+  gsl_matrix_free( D_E_mu );
   /* \theta \leftarrow {\mu_E - \mu\over ||\mu_E - \mu||_2} */
   gsl_matrix* theta = gsl_matrix_alloc( g_iP, 1 );
   gsl_matrix_memcpy( theta, mu_E );
@@ -311,15 +315,25 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
   /* t \leftarrow ||\mu_E - \bar\mu||_2*/
   double t = diff_norm( mu_E, bar_mu );
   unsigned int nb_it = 0;
+  g_dBest_error = diff_norm( mu_E, mu );
+  g_dBest_true_error = true_diff_norm( omega );
+  g_dBest_diff = true_V_diff( omega );
+  g_dBest_t = t;
   /* while t > \epsilon */
-  while( t > g_dEpsilon_anirl ){
+  while( t > g_dEpsilon_anirl && nb_it < g_iIt_max_anirl ){
     /* Output of the different criteria */
[…]

### 2011-04-07 ChainWalk/lagoudakis2003least_figure10.c
2nde soumission JFPDA

@@ -17,7 +17,7 @@
                     /* compute the mean and the var*/
 #define D_PREFIX "Samples" /* Prefix for the files containing */
                            /* the transitions*/
-#define ACTION_FILE "actions_lagoudakis.mat"
+#define ACTION_FILE "actions.mat"
 
 gsl_matrix* phi( gsl_matrix* sa ){
   gsl_matrix* answer = gsl_matrix_calloc( 6, 1 );

### 2011-04-07 GridWorld/conf.py

@@ -1,0 +1,3 @@
+g_nlInput = ['criteria_lstd.dat', ['-d" " -f\'1 2\'', '-d" " -f\'1 3\'', '-d" " -f\'1 4\'','-d" " -f\'1 5\''], [[5,13],[70,110],[270,330],[470,530],[700,800],[900,1100],[1800,2200]],
+             'both_error.dat', ['-d" " -f\'1 2\'','-d" " -f\'3 4\''], [[5,13],[70,110],[270,330],[470,530],[700,800],[900,1100],[1800,2200]]]
+g_iN = 100

### 2011-04-07 GridWorld/plot.c

@@ -1,6 +1,6 @@
 #include <gsl/gsl_matrix.h>
 #include "GridWorld.h"
-#include "GridWorld_simulator.h"
+#include "simulator.h"
 #include "utils.h"
 #include "LSPI.h"
 #include "abbeel2004apprenticeship.h"
@@ -10,16 +10,7 @@
 #include "IRL_Globals.h"
 #define D_FILE_NAME "Samples.dat"
 #define TRANS_WIDTH 7
-#define ACTION_FILE "actions_abbeel.mat"
-/* #include "greedy.h" */
-/* #define K (GRID_HEIGHT*GRID_WIDTH*4) /\* dim(\phi) *\/ */
-/* #define P (GRID_HEIGHT*GRID_WIDTH) /\* dim(\psi) *\/ */
-/* #define S 2 */
-/* #define A 1 */
-/* #define GAMMA 0.9 */
-/* #define GAMMA_LSPI 0.9 */
-/* #define EPSILON 0.1 */
-/* #define EPSILON_LSPI 0.1 */
+#define ACTION_FILE "actions.mat"
 
 unsigned int g_iK = (GRID_HEIGHT*GRID_WIDTH*4); /* dim(\phi) */
 unsigned int g_iP = (GRID_HEIGHT*GRID_WIDTH); /* dim(\psi) */
@@ -66,7 +57,8 @@ int main( void ){
   gsl_matrix* omega_expert = lspi( D, omega_0 );
   g_mOmega_E = omega_expert;
   expert_just_set();
-  unsigned int M = 40;
+  /* Courbe A : différences entre les mesures pour MC */
+  unsigned int M = 500;
   g_iNb_samples = D->size1;
   g_mOmega =  omega_expert;
   gsl_matrix* D_expert = gridworld_simulator( M );
@@ -74,13 +66,28 @@ int main( void ){
     proj_mc_lspi_ANIRL( D_expert, D, M );
   gsl_matrix_free( omega_imitation );
   gsl_matrix_free( D_expert );
-  for(M = 1; M<=100; M+=20 ){
+  /*Courbes B & C : Performance de MC et LSTDmu*/
+  M = 501; //To differentiate from above when greping.
+  //See the Makefile
+  int m_exp[] = {1,10,30,50,75,100,200};
+  for( int i=0; i<7 ; i++ ){
     g_iNb_samples = 0;
     g_mOmega =  omega_expert;
[…]

### 2011-04-07 IRL_Globals.h

@@ -37,8 +37,14 @@ extern unsigned int g_iNb_samples;
    couples are visited, thus making the A matrix singular */
 extern double g_dLambda_lstdmu; 
 
-/* When using MC in ANIRL, we store the best score when
-   the measured error is at its lowest yet*/
+/* When using ANIRL, we store the best score when
+   the measured error is at its lowest yet. This does not
+   mean that the other criteria (especially t) also are at 
+   their lowest.
+   These are declared in abbeel2004apprenticeship.c
+   but also used in LSTDmu.c
+*/
 extern double g_dBest_error;
 extern double g_dBest_true_error;
 extern double g_dBest_diff;
+extern double g_dBest_t;

### 2011-04-07 abbeel2004apprenticeship.c

@@ -10,11 +10,12 @@
 /* #include "greedy.h" */
 /* #include "GridWorld_simulator.h" */
 
-/* When using MC in ANIRL, we store the best score when
+/* When using ANIRL, we store the best score when
    the measured error is at its lowest yet*/
 double g_dBest_error;
 double g_dBest_true_error;
 double g_dBest_diff;
+double g_dBest_t;
 
 /* Compute an estimate of \mu using the monte carlo method,
    given a set of trajectories 
@@ -86,18 +87,20 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
   g_dBest_error = diff_norm( mu_E, mu );
   g_dBest_true_error = true_diff_norm( omega );
   g_dBest_diff = true_V_diff( omega );
+  g_dBest_t = t;
   while( t > g_dEpsilon_anirl && nb_it < g_iIt_max_anirl ){
     /* Output of the different criteria */
     double empirical_err = diff_norm( mu_E, mu );
     double true_err = true_diff_norm( omega );
     double true_V = true_V_diff( omega );
-    printf( "%d %d %d %lf %lf %lf %lf\n", 
-	    m, nb_it, g_iNb_samples, 
+    printf( "%d %d %lf %lf %lf %lf\n", 
+	    m, nb_it,
 	    t, empirical_err, true_err, true_V );
     if( empirical_err <= g_dBest_error ){
       g_dBest_error = empirical_err;
       g_dBest_true_error = true_err;
       g_dBest_diff = true_V;
+      g_dBest_t = t;
     }
     /* D.r \leftarrow \theta^T\psi(D.s) */
     for( unsigned int i = 0 ; i < D->size1 ; i++ ){
@@ -167,13 +170,14 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
   double empirical_err = diff_norm( mu_E, mu );
   double true_err = true_diff_norm( omega );
   double true_V = true_V_diff( omega );
-  printf( "%d %d %d %lf %lf %lf %lf\n", 
-	  m, nb_it, g_iNb_samples, 
+  printf( "%d %d %lf %lf %lf %lf\n", 
+	  m, nb_it, 
 	  t, empirical_err, true_err, true_V );
   if( empirical_err <= g_dBest_error ){
     g_dBest_error = empirical_err;
     g_dBest_true_error = true_err;
[…]

### 2011-04-07 courbe_a.c

@@ -1,71 +1,0 @@
-#include <gsl/gsl_matrix.h>
-#include "GridWorld.h"
-#include "GridWorld_simulator.h"
-#include "utils.h"
-#include "LSPI.h"
-#include "abbeel2004apprenticeship.h"
-#include "LSTDmu.h"
-#include "criteria.h"
-#include "RL_Globals.h"
-#include "IRL_Globals.h"
-#define D_FILE_NAME "Samples.dat"
-#define TRANS_WIDTH 7
-#define ACTION_FILE "actions_abbeel.mat"
-
-unsigned int g_iK = (GRID_HEIGHT*GRID_WIDTH*4); /* dim(\phi) */
-unsigned int g_iP = (GRID_HEIGHT*GRID_WIDTH); /* dim(\psi) */
-
-gsl_matrix* phi( gsl_matrix* sa ){
-  gsl_matrix* answer = gsl_matrix_calloc( g_iK, 1 );
-  unsigned int x = (unsigned int)gsl_matrix_get( sa, 0, 0 );
-  unsigned int y = (unsigned int)gsl_matrix_get( sa, 0, 1 );
-  unsigned int a = (unsigned int)gsl_matrix_get( sa, 0, 2 );
-  unsigned int index = (y-1)*GRID_WIDTH*4 + (x-1)*4 + a-1;
-  gsl_matrix_set( answer, index, 0, 1.0 );
-  return answer;
-}
-
-gsl_matrix* psi( gsl_matrix* s ){
-  gsl_matrix* answer = gsl_matrix_calloc( g_iP, 1 );
-  unsigned int x = (unsigned int)gsl_matrix_get( s, 0, 0 );
-  unsigned int y = (unsigned int)gsl_matrix_get( s, 0, 1 );
-  unsigned int index = (y-1)*GRID_WIDTH + (x-1);
-  gsl_matrix_set( answer, index, 0, 1.0 );
-  return answer;
-}
-
-unsigned int g_iS = 2;
-unsigned int g_iA = 1;
-unsigned int g_iIt_max_lspi = 20;
-gsl_matrix* (*g_fPhi)(gsl_matrix*) = &phi;
-gsl_matrix* g_mOmega = NULL;
-double g_dLambda_lstdQ = 0.1;
-double g_dGamma_lstdq =  0.9;
-double g_dEpsilon_lspi = 0.1;
-double g_dLambda_lstdmu = 0.1;
-double g_dGamma_anirl = 0.9;
[…]

### 2011-04-07 courbe_c.c

@@ -1,73 +1,0 @@
-#include <gsl/gsl_matrix.h>
-#include "GridWorld.h"
-#include "GridWorld_simulator.h"
-#include "utils.h"
-#include "LSPI.h"
-#include "abbeel2004apprenticeship.h"
-#include "LSTDmu.h"
-#include "criteria.h"
-#include "RL_Globals.h"
-#include "IRL_Globals.h"
-#define D_FILE_NAME "Samples.dat"
-#define TRANS_WIDTH 7
-#define ACTION_FILE "actions_abbeel.mat"
-
-unsigned int g_iK = (GRID_HEIGHT*GRID_WIDTH*4); /* dim(\phi) */
-unsigned int g_iP = (GRID_HEIGHT*GRID_WIDTH); /* dim(\psi) */
-
-gsl_matrix* phi( gsl_matrix* sa ){
-  gsl_matrix* answer = gsl_matrix_calloc( g_iK, 1 );
-  unsigned int x = (unsigned int)gsl_matrix_get( sa, 0, 0 );
-  unsigned int y = (unsigned int)gsl_matrix_get( sa, 0, 1 );
-  unsigned int a = (unsigned int)gsl_matrix_get( sa, 0, 2 );
-  unsigned int index = (y-1)*GRID_WIDTH*4 + (x-1)*4 + a-1;
-  gsl_matrix_set( answer, index, 0, 1.0 );
-  return answer;
-}
-
-gsl_matrix* psi( gsl_matrix* s ){
-  gsl_matrix* answer = gsl_matrix_calloc( g_iP, 1 );
-  unsigned int x = (unsigned int)gsl_matrix_get( s, 0, 0 );
-  unsigned int y = (unsigned int)gsl_matrix_get( s, 0, 1 );
-  unsigned int index = (y-1)*GRID_WIDTH + (x-1);
-  gsl_matrix_set( answer, index, 0, 1.0 );
-  return answer;
-}
-
-unsigned int g_iS = 2;
-unsigned int g_iA = 1;
-unsigned int g_iIt_max_lspi = 20;
-gsl_matrix* (*g_fPhi)(gsl_matrix*) = &phi;
-gsl_matrix* g_mOmega = NULL;
-double g_dLambda_lstdQ = 0.1;
-double g_dGamma_lstdq =  0.9;
-double g_dEpsilon_lspi = 0.1;
-double g_dLambda_lstdmu = 0.1;
-double g_dGamma_anirl = 0.9;
-double g_dEpsilon_anirl = 0.1;
[…]

### 2011-04-07 criteria.c

@@ -6,7 +6,7 @@
 #include "abbeel2004apprenticeship.h"
 #include "utils.h"
 
-#define PERFECT_MC_LENGTH 1000
+#define PERFECT_MC_LENGTH 2000
 
 /* Public global where the omega of the expert is stored */
 gsl_matrix* g_mOmega_E = NULL;
@@ -66,7 +66,15 @@ void expert_just_set(){
   g_iNb_samples = nb_samples_backup;
   g_mMu_E = monte_carlo_mu( D );
   g_dV_E = value_func( D );
+  fprintf(stderr,"V_E(s_0) = %lf\n",g_dV_E);
   gsl_matrix_free( D );
+}
+
+/* Must call this one if expert_just_set has been
+   called before, in order to free the memory
+*/
+void expert_free(){
+  gsl_matrix_free( g_mMu_E );
 }
 
 

### 2011-04-07 criteria.h

@@ -15,6 +15,11 @@ double true_diff_norm( gsl_matrix* omega );
 */
 void expert_just_set();
 
+/* Must call this one if expert_just_set has been
+   called before, in order to free the memory
+*/
+void expert_free();
+
 /* Compute V(s0) for the given set of trajectories */
 double value_func( gsl_matrix* D );
 

### 2011-04-07 greedy.c

@@ -36,5 +36,44 @@ gsl_matrix* greedy_policy( gsl_matrix* state ){
     gsl_matrix_free( phi_sa );
   }
   gsl_matrix_free( mQ );
+  gsl_matrix_free( sa );
   return best_action;
+}
+
+double greedy_value_function( gsl_matrix* state ){
+  gsl_matrix* sa = gsl_matrix_alloc( 1, g_iS + g_iA );
+  gsl_matrix_view s_dst = gsl_matrix_submatrix( sa, 0, 0, 
+						1, g_iS );
+  gsl_matrix_memcpy( &s_dst.matrix, state );
+  gsl_matrix_view a_dst = gsl_matrix_submatrix( sa, 0, g_iS,
+						1, g_iA );
+  gsl_matrix_view a_src = 
+    gsl_matrix_submatrix( g_mActions, 0, 0, 1, g_iA );
+  gsl_matrix* best_action = gsl_matrix_alloc( 1, g_iA );
+  gsl_matrix_memcpy( &a_dst.matrix, &a_src.matrix );
+  gsl_matrix_memcpy( best_action, &a_src.matrix );
+  gsl_matrix* phi_sa = g_fPhi( sa );
+  gsl_matrix* mQ = gsl_matrix_alloc( 1, 1 );
+  gsl_blas_dgemm( CblasTrans, CblasNoTrans, 
+		  1.0, g_mOmega, phi_sa, 
+		  0.0, mQ );
+  gsl_matrix_free( phi_sa );
+  double Q_max = gsl_matrix_get( mQ, 0, 0 );
+  for( unsigned int i = 0 ; i < g_mActions->size1 ; i++ ){
+    a_src = gsl_matrix_submatrix( g_mActions, i, 0, 1, g_iA );
+    gsl_matrix_memcpy( &a_dst.matrix, &a_src.matrix );
+    gsl_matrix* phi_sa = g_fPhi( sa );
+    gsl_blas_dgemm( CblasTrans, CblasNoTrans, 
+		    1.0, g_mOmega, phi_sa, 
+		    0.0, mQ );
+    double Q_current = gsl_matrix_get( mQ, 0, 0 );
+    if( Q_current > Q_max ){
+      Q_max = Q_current;
+      gsl_matrix_memcpy( best_action, &a_src.matrix );
+    }
+    gsl_matrix_free( phi_sa );
+  }
+  gsl_matrix_free( mQ );
+  gsl_matrix_free( best_action );
+  return Q_max;
 }

### 2011-04-07 greedy.h

@@ -1,2 +1,5 @@
-/* Returns the action a for which Q(state,a) is max */
+/* Return the action a for which Q(state,a) is max */
 gsl_matrix* greedy_policy( gsl_matrix* state );
+
+/* Return V(s) found by maximizing Q(s,a) over a */
+double greedy_value_function( gsl_matrix* state );

### 2011-04-08 criteria.c
Correction de l'inversion des légendes de GridWorld/both_error juste avant la deadline de JFPDA

@@ -66,7 +66,6 @@ void expert_just_set(){
   g_iNb_samples = nb_samples_backup;
   g_mMu_E = monte_carlo_mu( D );
   g_dV_E = value_func( D );
-  fprintf(stderr,"V_E(s_0) = %lf\n",g_dV_E);
   gsl_matrix_free( D );
 }
 

### 2011-04-08 IRL_Globals.h
Travail en cours sur le problème continu

@@ -48,3 +48,4 @@ extern double g_dBest_error;
 extern double g_dBest_true_error;
 extern double g_dBest_diff;
 extern double g_dBest_t;
+extern gsl_matrix* g_mBest_omega;

### 2011-04-08 InvertedPendulum/InvertedPendulum.c

@@ -1,0 +1,60 @@
+#include <stdlib.h>
+#include <math.h>
+#include <gsl/gsl_matrix.h>
+#include "InvertedPendulum.h"
+#include "utils.h"
+
+void iv_init( double* pos, double* speed ){
+  *pos = 0.04;
+  //(double)rand()/(double)RAND_MAX*0.01;
+  *speed = -0.02;
+  //  (double)rand()/(double)RAND_MAX*0.01;
+  int sign = random_int( 0, 1 );
+  if (sign == 0){
+    //*pos=-*pos;
+  }
+  sign = random_int( 0, 1 );
+  if (sign == 0){
+    //*speed=-*speed;
+  }
+}
+
+void iv_step( double state_p, double state_v, 
+	      unsigned int action,
+	      double* next_state_p, double* next_state_v, 
+	      double* reward, int* eoe ){
+  unsigned int noise = random_int( -1, 1 );
+  int iControl;
+  switch( action ){
+  case LEFT:
+    iControl = -50 + noise;
+    break;
+  case NONE:
+    iControl = 0 + noise;
+    break;
+  case RIGHT:
+    iControl = 50 + noise;
+    break;
+  }
+  double g = 9.8;
+  double m = 2.0;
+  double M = 8.0;
+  double l = 0.50;
+  double alpha = 1./(m+M);
+  double step = 0.1;
+  double control = (double)iControl;
+  double acceleration = 
+    (g*sin(state_p) - 
+     alpha*m*l*pow(state_v,2)*sin(2*state_p)/2. - 
+     alpha*cos(state_p)*control) / 
+    (4.*l/3. - alpha*m*l*pow(cos(state_p),2));
+  *next_state_p = state_p +state_v*step;
+  *next_state_v = state_v + acceleration*step;
+  if( *next_state_p > PI/2. || *next_state_p < -PI/2. ){
+    *reward = -1;
+    *eoe = 0; 
+  }else{
+    *reward = 0;
+    *eoe = 1;
+  }
+}

### 2011-04-08 InvertedPendulum/InvertedPendulum.h

@@ -1,0 +1,9 @@
+#define LEFT 0
+#define NONE 1
+#define RIGHT 2
+#define PI (3.1415926536)
+void iv_init( double* pos, double* speed );
+void iv_step( double state_p, double state_v, 
+	      unsigned int action,
+	      double* next_state_p, double* next_state_v, 
+	      double* reward, int* eoe );

### 2011-04-08 InvertedPendulum/generator.c

@@ -1,0 +1,37 @@
+#include <stdlib.h>
+#include <time.h>
+#include <unistd.h>
+#include <math.h>
+#include "InvertedPendulum.h"
+#include <gsl/gsl_matrix.h>
+#include "utils.h"
+#define NUMBER_OF_WALKS (500)
+#define MAX_WALK_LENGTH (3000)
+
+int main( void ){
+  srand(time(NULL)+getpid()); rand(); rand();rand();
+  for( unsigned int i = 0 ; i < NUMBER_OF_WALKS ; i++ ){
+    double state_p;//position
+    double state_v;//vitesse
+    iv_init( &state_p, &state_v );
+    int eoe = 1;
+    for( unsigned int j = 0 ; j < MAX_WALK_LENGTH && eoe == 1 ; 
+	 j++ ){
+      double next_state_p;
+      double next_state_v;
+      double reward;
+      unsigned int action = random_int( LEFT, RIGHT );
+      iv_step( state_p, state_v, action, 
+	       &next_state_p, &next_state_v, &reward, &eoe );
+      if( j == MAX_WALK_LENGTH - 1 ){
+	eoe = 0;
+      }
+      printf("%lf %lf %d %lf %lf %lf %d\n",
+	     state_p, state_v, action, 
+	     next_state_p, next_state_v, reward, eoe );
+      state_p = next_state_p;
+      state_v = next_state_v;
+    }
+  }
+  return 0;
+}

### 2011-04-08 InvertedPendulum/plot.c

@@ -1,0 +1,160 @@
+#define _POSIX_C_SOURCE 1
+#include <gsl/gsl_matrix.h>
+#include <math.h>
+#include "InvertedPendulum.h"
+#include "simulator.h"
+#include "utils.h"
+#include "LSPI.h"
+#include "abbeel2004apprenticeship.h"
+#include "LSTDmu.h"
+#include "criteria.h"
+#include "RL_Globals.h"
+#include "IRL_Globals.h"
+#include "greedy.h"
+#define D_FILE_NAME "Samples.dat"
+#define TRANS_WIDTH 7
+#define ACTION_FILE "actions.mat"
+
+unsigned int g_iK = (30); /* dim(\phi) */
+unsigned int g_iP = (10); /* dim(\psi) */
+
+gsl_matrix* phi( gsl_matrix* sa ){
+  gsl_matrix* answer = gsl_matrix_calloc( g_iK, 1 );
+  double position = gsl_matrix_get( sa, 0, 0 );
+  double speed = gsl_matrix_get( sa, 0, 1 );
+  unsigned int action = (unsigned int)gsl_matrix_get( sa, 0, 2 );
+  unsigned int index = action*10;
+  gsl_matrix_set( answer, index, 0, 1.0 );
+  index++;
+  for( int i = -1. ; i <= 1 ; i++ ){
+    for( int j = -1 ; j <= 1 ; j++ ){
+      double d_i = (double)i * PI/4.;
+      double d_j = (double)j;
+      gsl_matrix_set( answer, index, 0, 
+		      exp(-(pow(position-d_i,2) + 
+			    pow(speed-d_j,2))/2.));
+      index++;
+    }
+  }
+  return answer;
+}
+
+gsl_matrix* psi( gsl_matrix* s ){
+  gsl_matrix* answer = gsl_matrix_calloc( g_iP, 1 );
+  double position = gsl_matrix_get( s, 0, 0 );
+  double speed = gsl_matrix_get( s, 0, 1 );
+  unsigned int index = 0;
+  gsl_matrix_set( answer, index, 0, 1.0 );
+  index++;
+  for( int i = -1. ; i <= 1 ; i++ ){
+    for( int j = -1 ; j <= 1 ; j++ ){
+      double d_i = (double)i * PI/4.;
+      double d_j = (double)j;
+      gsl_matrix_set( answer, index, 0,
+		      exp(-(pow(position-d_i,2) + 
[…]

### 2011-04-08 InvertedPendulum/simulator.c

@@ -1,0 +1,67 @@
+#include <time.h>
+#include <unistd.h>
+#include <gsl/gsl_matrix.h> 
+#include <math.h>
+#include "utils.h"
+#include "simulator.h"
+#include "greedy.h"
+#include "InvertedPendulum.h"
+#define TRANS_WIDTH 7 /* s,a,s',r,e = 7*/
+
+unsigned int g_iNb_samples = 0;
+
+/* Simulate nbEpisodes episodes using greedy_policy()*/
+gsl_matrix* inverted_pendulum_simulator( int nbEpisodes ){
+  srand(time(NULL)+getpid()); rand(); rand();rand();
+  gsl_matrix* transitions = gsl_matrix_alloc( nbEpisodes*g_iMax_episode_len, TRANS_WIDTH );
+  // set the value of the pendulum in a state near the 
+  // equilibrium
+  double state_p; //position
+  double state_v; //vitesse
+  iv_init( &state_p, &state_v );
+  gsl_matrix* state = gsl_matrix_alloc( 1, 2 );
+  unsigned int j = 0; //Index in transitions
+  for( unsigned int i = 0 ; i < nbEpisodes ; i++ ){
+    unsigned int nb_steps = 0;
+    int eoe = 1;
+    while( eoe==1 ){
+      nb_steps++;
+      g_iNb_samples++;
+      gsl_matrix_set( state, 0, 0, (double)state_p );
+      gsl_matrix_set( state, 0, 1, (double)state_v );
+      double next_state_p;
+      double next_state_v;
+      gsl_matrix* mAction = greedy_policy( state );
+      unsigned int action = 
+	(unsigned int)gsl_matrix_get( mAction, 0, 0 );
+      gsl_matrix_free( mAction );
+      double reward;
+      iv_step( state_p, state_v, action, 
+	       &next_state_p, &next_state_v, &reward, &eoe );
+      if( nb_steps == g_iMax_episode_len ){
+	eoe = 0;
+      }
+      gsl_matrix_set( transitions, j, 0, (double)state_p );
+      gsl_matrix_set( transitions, j, 1, (double)state_v );
+      gsl_matrix_set( transitions, j, 2, (double)action );
+      gsl_matrix_set( transitions, j, 3, (double)next_state_p );
+      gsl_matrix_set( transitions, j, 4, (double)next_state_v );
[…]

### 2011-04-08 InvertedPendulum/simulator.h

@@ -1,0 +1,4 @@
+extern unsigned int g_iMax_episode_len;
+
+/* Simulate nbEpisodes episodes using greedy_policy()*/
+gsl_matrix* inverted_pendulum_simulator( int nbEpisodes );

### 2011-04-08 LSTDmu.c

@@ -319,6 +319,8 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
   g_dBest_true_error = true_diff_norm( omega );
   g_dBest_diff = true_V_diff( omega );
   g_dBest_t = t;
+  g_mBest_omega = gsl_matrix_alloc( omega->size1, omega->size2 );
+  gsl_matrix_memcpy( g_mBest_omega, omega );
   /* while t > \epsilon */
   while( t > g_dEpsilon_anirl && nb_it < g_iIt_max_anirl ){
     /* Output of the different criteria */
@@ -328,11 +330,13 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
     printf( "%d %d %lf %lf %lf %lf\n", 
 	    m, nb_it, 
 	    t, empirical_err, true_err, true_V );
-    if( empirical_err <= g_dBest_error ){
+    //if( empirical_err <= g_dBest_error ){
+    if( true_err <= g_dBest_true_error ){
       g_dBest_error = empirical_err;
       g_dBest_true_error = true_err;
       g_dBest_diff = true_V;
       g_dBest_t = t;
+      gsl_matrix_memcpy( g_mBest_omega, omega );
     }
     /* D.r \leftarrow \theta^T\psi(D.s) */
     for( unsigned int i = 0 ; i < D->size1 ; i++ ){
@@ -416,11 +420,13 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
   printf( "%d %d %lf %lf %lf %lf\n", 
 	  m, nb_it, 
 	  t, empirical_err, true_err, true_V );
-  if( empirical_err <= g_dBest_error ){
+  //  if( empirical_err <= g_dBest_error ){
+  if( true_err <= g_dBest_true_error ){
     g_dBest_error = empirical_err;
     g_dBest_true_error = true_err;
     g_dBest_diff = true_V;
     g_dBest_t = t;
+    gsl_matrix_memcpy( g_mBest_omega, omega );
   }
   gsl_matrix_free( s_0 );
   gsl_matrix_free( D_mu );
@@ -429,6 +435,7 @@ gsl_matrix* proj_lstd_lspi_ANIRL( gsl_matrix* D_E,
   gsl_matrix_free( mu_E );
   gsl_matrix_free( bar_mu );
   gsl_matrix_free( theta );
[…]

### 2011-04-08 abbeel2004apprenticeship.c

@@ -16,6 +16,7 @@ double g_dBest_error;
 double g_dBest_true_error;
 double g_dBest_diff;
 double g_dBest_t;
+gsl_matrix* g_mBest_omega;
 
 /* Compute an estimate of \mu using the monte carlo method,
    given a set of trajectories 
@@ -88,6 +89,8 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
   g_dBest_true_error = true_diff_norm( omega );
   g_dBest_diff = true_V_diff( omega );
   g_dBest_t = t;
+  g_mBest_omega = gsl_matrix_alloc( omega->size1, omega->size2 );
+  gsl_matrix_memcpy( g_mBest_omega, omega );
   while( t > g_dEpsilon_anirl && nb_it < g_iIt_max_anirl ){
     /* Output of the different criteria */
     double empirical_err = diff_norm( mu_E, mu );
@@ -96,11 +99,13 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
     printf( "%d %d %lf %lf %lf %lf\n", 
 	    m, nb_it,
 	    t, empirical_err, true_err, true_V );
-    if( empirical_err <= g_dBest_error ){
+    //if( empirical_err <= g_dBest_error ){
+    if( true_err <= g_dBest_true_error ){
       g_dBest_error = empirical_err;
       g_dBest_true_error = true_err;
       g_dBest_diff = true_V;
       g_dBest_t = t;
+      gsl_matrix_memcpy( g_mBest_omega, omega );
     }
     /* D.r \leftarrow \theta^T\psi(D.s) */
     for( unsigned int i = 0 ; i < D->size1 ; i++ ){
@@ -173,17 +178,20 @@ gsl_matrix* proj_mc_lspi_ANIRL( gsl_matrix* D_E,
   printf( "%d %d %lf %lf %lf %lf\n", 
 	  m, nb_it, 
 	  t, empirical_err, true_err, true_V );
-  if( empirical_err <= g_dBest_error ){
+  //  if( empirical_err <= g_dBest_error ){
+  if( true_err <= g_dBest_true_error ){
     g_dBest_error = empirical_err;
     g_dBest_true_error = true_err;
     g_dBest_diff = true_V;
     g_dBest_t = t;
+    gsl_matrix_memcpy( g_mBest_omega, omega );
   }
[…]

### 2011-04-08 InvertedPendulum/plot.c
J'arrive à faire marcher lstdmu sur le problème continu simplifé, en lui donnant une matrice D

@@ -88,39 +88,57 @@ int main( void ){
   expert_just_set();
   fprintf(stderr,"done\n");
   
-  /* g_iNb_samples = 0; */
-  /* g_mOmega = omega_0; */
-  /* gsl_matrix* discard = inverted_pendulum_simulator( 100 ); */
-  /* gsl_matrix_free( discard ); */
-  /* unsigned int mean_control_steps = g_iNb_samples/100; */
-  /* fprintf(stderr,"Performance of default policy : %d\n",mean_control_steps); */
-
-  //g_mOmega =  omega_expert;
-  /* g_mOmega = gsl_matrix_calloc( g_iK, 1 ); */
-  /* for( int ii = 0; ii<30;ii++){ */
-  /*   gsl_matrix_set( g_mOmega,ii,0,1.0); */
-  /* } */
-  /* gsl_matrix* state = gsl_matrix_alloc( 1, 2 ); */
-  /* for( double i = -1. ; i <= 1 ; i+=0.05 ){ */
-  /*   for( double j = -1 ; j <= 1 ; j+=0.05 ){ */
-  /*     gsl_matrix_set( state, 0, 0, i*PI/2.); */
-  /*     gsl_matrix_set( state, 0, 1, j*PI/2.); */
-  /*     double v = greedy_value_function( state ); */
-  /*     printf("%lf %lf %lf\n", i*PI/2., j*PI/2.,v ); */
-  /*   } */
-  /*   printf("\n"); */
-  /* } */
 
   gsl_matrix* D_expert;
-
   gsl_matrix* omega_imitation;
-  unsigned int M = 1001; //To differentiate from above when greping.
-  //See the Makefile
+  //La première courbe montre que sans matrice D, 
+  //quelque soit la taille de la matrice D_E,
+  //la variante LSTD ne fonctionne pas tandis que
+  //la variante MC fonctionne
   int m_exp[] = {1,10,100};
   for( int i=0; i<3 ; i++ ){
     g_iNb_samples = 0;
     g_mOmega =  omega_expert;
-    g_iMax_episode_len = 30;
+    g_iMax_episode_len = 100;
+    D_expert = inverted_pendulum_simulator( m_exp[i] );
+    unsigned int nb_samples_exp = g_iNb_samples;
+    g_iMax_episode_len = 100;
+    gsl_matrix* omega_lstd =
+      proj_lstd_lspi_ANIRL( D_expert, D_expert );
[…]

### 2011-04-12 GridWorld/conf.py
Modifié LSTDmu on policy pour qu'il ressemble à LSTD et non à LSTDQ. Tout semble marcher.

@@ -1,3 +1,4 @@
 g_nlInput = ['criteria_lstd.dat', ['-d" " -f\'1 2\'', '-d" " -f\'1 3\'', '-d" " -f\'1 4\'','-d" " -f\'1 5\''], [[5,13],[70,110],[270,330],[470,530],[700,800],[900,1100],[1800,2200]],
              'both_error.dat', ['-d" " -f\'1 2\'','-d" " -f\'3 4\''], [[5,13],[70,110],[270,330],[470,530],[700,800],[900,1100],[1800,2200]]]
 g_iN = 100
+

### 2011-04-12 LSTDmu.c

@@ -17,69 +17,46 @@
    in its on policy form. Typically used for the expert.
    The D_mu matrix has the form
    s a s' psi(s) eoe
-   s_0,a_0 are taken from the first transition of D_\mu
+   s_0 is taken from the first transition of D_\mu
 */
 gsl_matrix* lstd_mu_op(  gsl_matrix* D_mu ){
   gsl_matrix* s_0 = gsl_matrix_alloc( 1, g_iS );
-  gsl_matrix* a_0 = gsl_matrix_alloc( 1, g_iA );
   gsl_matrix_view s_0_src = gsl_matrix_submatrix( D_mu, 0, 0,
 						  1, g_iS );
-  gsl_matrix_view a_0_src = gsl_matrix_submatrix( D_mu, 0, g_iS,
-						  1, g_iA );
   gsl_matrix_memcpy( s_0, &s_0_src.matrix );
-  gsl_matrix_memcpy( a_0, &a_0_src.matrix );
   // \tilde A \leftarrow 0
-  gsl_matrix* A = gsl_matrix_calloc( g_iK, g_iK );
+  gsl_matrix* A = gsl_matrix_calloc( g_iP, g_iP );
   // \tilde b \leftarrow 0
-  gsl_matrix* b = gsl_matrix_calloc( g_iK, g_iP );
+  gsl_matrix* b = gsl_matrix_calloc( g_iP, g_iP );
   // for each (s,a,s',\psi(s),eoe) \in D_mu
   for( unsigned int i=0; i < D_mu->size1 ; i++ ){
-    // \tilde A \leftarrow \tilde A + \phi(s,a)\left(\phi(s,a) 
-    // - \gamma \phi(s',a')\right)^T
-    gsl_matrix_view sa = gsl_matrix_submatrix(D_mu,i,0, 1, 
-					      g_iS+g_iA);
-    gsl_matrix* phi_sa = g_fPhi( &sa.matrix );
-    gsl_matrix* sa_dash = gsl_matrix_calloc( 1, g_iS+g_iA );
-    gsl_matrix_view sdash_dst = gsl_matrix_submatrix( sa_dash, 
-						      0, 0,
-						      1, g_iS);
-    gsl_matrix_view sdash_src = 
-      gsl_matrix_submatrix( D_mu, i, g_iS+g_iA, 1, g_iS );
[…]

### 2011-04-12 GridWorld/plot.c
s_0 est échantilloné selon une distribution pour LSTDmu

@@ -34,6 +34,13 @@ gsl_matrix* psi( gsl_matrix* s ){
   return answer;
 }
 
+gsl_matrix* initial_state( void ){
+  gsl_matrix* answer = gsl_matrix_alloc( 1, 2 );
+  gsl_matrix_set( answer, 0, 0, 1.0 );
+  gsl_matrix_set( answer, 0, 1, 1.0 );
+  return answer;
+}
+
 unsigned int g_iS = 2;
 unsigned int g_iA = 1;
 unsigned int g_iIt_max_lspi = 20;
@@ -49,6 +56,7 @@ unsigned int g_iIt_max_anirl = 40;
 gsl_matrix* g_mActions = NULL; 
 gsl_matrix* (*g_fPsi)(gsl_matrix*) = &psi;
 gsl_matrix* (*g_fSimulator)(int) = &gridworld_simulator;
+gsl_matrix* (*g_fS_0)(void) = &initial_state;
 
 int main( void ){
   gsl_matrix* D = file2matrix( D_FILE_NAME, TRANS_WIDTH );

### 2011-04-12 IRL_Globals.h

@@ -49,3 +49,9 @@ extern double g_dBest_true_error;
 extern double g_dBest_diff;
 extern double g_dBest_t;
 extern gsl_matrix* g_mBest_omega;
+
+/* The initial state(s) for which \mu is to be computed is 
+   given to LSTDmu through this function. One state per line.
+   When using Monte-Carlo, it is assumed that the first sample
+   or each trajectory is among the initial state(s)*/
+extern gsl_matrix* (*g_fS_0)( void );

### 2011-04-12 LSTDmu.c

@@ -20,10 +20,6 @@
    s_0 is taken from the first transition of D_\mu
 */
 gsl_matrix* lstd_mu_op(  gsl_matrix* D_mu ){
-  gsl_matrix* s_0 = gsl_matrix_alloc( 1, g_iS );
-  gsl_matrix_view s_0_src = gsl_matrix_submatrix( D_mu, 0, 0,
-						  1, g_iS );
-  gsl_matrix_memcpy( s_0, &s_0_src.matrix );
   // \tilde A \leftarrow 0
   gsl_matrix* A = gsl_matrix_calloc( g_iP, g_iP );
   // \tilde b \leftarrow 0
@@ -78,13 +74,23 @@ gsl_matrix* lstd_mu_op(  gsl_matrix* D_mu ){
   gsl_permutation_free( p );
   gsl_matrix_free( A );
   gsl_matrix_free( b );
-   //\mu_\pi(s) \leftarrow \tilde\omega_\pi^T\psi(s)
-  gsl_matrix* psi_s_0 = g_fPsi( s_0 );
-  gsl_matrix* mu = gsl_matrix_alloc( g_iP, 1 );
-  gsl_blas_dgemm( CblasTrans, CblasNoTrans, 1.0,
-		  omega_pi, psi_s_0, 0.0, mu );
+  //\mu_\pi(s_0) \leftarrow 
+  //\sum_{s_0}\tilde\omega_\pi^T\psi(s_0)
+  gsl_matrix* s_0 = g_fS_0();
+  gsl_matrix* mu = gsl_matrix_calloc( g_iP, 1 );
+  gsl_matrix* delta_mu = gsl_matrix_alloc( g_iP, 1 );
+  for( unsigned int i = 0; i<s_0->size1;i++){
+    gsl_matrix_view mS_0 = gsl_matrix_submatrix( s_0, i, 0, 
+						 1, g_iS );
+    gsl_matrix* psi_s_0 = g_fPsi( &mS_0.matrix );
+    gsl_blas_dgemm( CblasTrans, CblasNoTrans, 1.0,
+		    omega_pi, psi_s_0, 0.0, delta_mu );
+    gsl_matrix_add( mu, delta_mu );
+    gsl_matrix_free( psi_s_0 );
+  }
+  gsl_matrix_scale( mu, 1./((double)s_0->size1) );
+  gsl_matrix_free( delta_mu );
   gsl_matrix_free( omega_pi );
-  gsl_matrix_free( psi_s_0 );
[…]

### 2011-04-12 InvertedPendulum/plot.c
Resultats sur le problème continu reproduits avec la nouvelle architecture

@@ -59,6 +59,17 @@ gsl_matrix* psi( gsl_matrix* s ){
   return answer;
 }
 
+gsl_matrix* initial_state( void ){
+  gsl_matrix* answer = gsl_matrix_alloc( 1, 2 );
+  double pos;
+  double speed;
+  iv_init( &pos, &speed );
+  gsl_matrix_set( answer, 0, 0, pos );
+  gsl_matrix_set( answer, 0, 1, speed );
+  return answer;
+}
+
+
 unsigned int g_iS = 2;
 unsigned int g_iA = 1;
 unsigned int g_iIt_max_lspi = 50;
@@ -74,6 +85,7 @@ unsigned int g_iIt_max_anirl = 2;
 gsl_matrix* g_mActions = NULL; 
 gsl_matrix* (*g_fPsi)(gsl_matrix*) = &psi;
 gsl_matrix* (*g_fSimulator)(int) = &inverted_pendulum_simulator;
+gsl_matrix* (*g_fS_0)(void) = &initial_state;
 unsigned int g_iMax_episode_len = 30; //Shorter episodes are 
 //better for MC, as s_0 is seen more often
 

### 2011-04-12 InvertedPendulum/InvertedPendulum.c
Résultats précédents sur le problème continu reproduits avec l'aléatoire en plus

@@ -5,17 +5,17 @@
 #include "utils.h"
 
 void iv_init( double* pos, double* speed ){
-  *pos = 0.04;
-  //(double)rand()/(double)RAND_MAX*0.01;
-  *speed = -0.02;
-  //  (double)rand()/(double)RAND_MAX*0.01;
+  *pos = 
+    (double)rand()/(double)RAND_MAX*0.1;
+  *speed = 
+    (double)rand()/(double)RAND_MAX*0.1;
   int sign = random_int( 0, 1 );
   if (sign == 0){
-    //*pos=-*pos;
+    *pos=-*pos;
   }
   sign = random_int( 0, 1 );
   if (sign == 0){
-    //*speed=-*speed;
+    *speed=-*speed;
   }
 }
 

### 2011-04-12 InvertedPendulum/plot.c

@@ -63,9 +63,11 @@ gsl_matrix* initial_state( void ){
   gsl_matrix* answer = gsl_matrix_alloc( 1, 2 );
   double pos;
   double speed;
-  iv_init( &pos, &speed );
-  gsl_matrix_set( answer, 0, 0, pos );
-  gsl_matrix_set( answer, 0, 1, speed );
+  for( unsigned int i=0; i<1; i++ ){
+    iv_init( &pos, &speed );
+    gsl_matrix_set( answer, i, 0, pos );
+    gsl_matrix_set( answer, i, 1, speed );
+  }
   return answer;
 }
 

### 2011-04-13 InvertedPendulum/InvertedPendulum.c
Courbes "threshold" et "threshold_EB" pour le problème continu

@@ -35,6 +35,9 @@ void iv_step( double state_p, double state_v,
   case RIGHT:
     iControl = 50 + noise;
     break;
+  default:
+    fprintf(stderr,"Dying\n");
+    exit(-1);
   }
   double g = 9.8;
   double m = 2.0;

### 2011-04-13 InvertedPendulum/first.c

@@ -104,37 +104,86 @@ int main( void ){
   
 
   gsl_matrix* D_expert;
-  int m_exp[] = {1,100};
-  int D_len[] = {1,200,300,400,500,600,700,800,900,1000,1100,1200,1300,1400,1500,1600,1700,1800,1900,2000,3000};
-  //Cette courbe illustre l'influence des
-  //paramètres que sont la taille de D et la taille 
-  //de D_E sur la variante LSTDmu
-  for( int i=0; i<2 ; i++ ){
-    for( int j=0; j<21;j++){
-      g_iNb_samples = 0;
-      g_mOmega =  omega_expert;
-      g_iMax_episode_len = 100;
-      D_expert = inverted_pendulum_simulator( m_exp[i] );
-      unsigned int nb_samples_exp = g_iNb_samples;
-      gsl_matrix_view sub_D = 
-	gsl_matrix_submatrix( D, 0, 0, D_len[j], TRANS_WIDTH );
-      g_iMax_episode_len = 100;
-      gsl_matrix* omega_lstd =
-	proj_lstd_lspi_ANIRL( D_expert, &sub_D.matrix );
-      g_mOmega = omega_lstd;
-      g_iNb_samples = 0;
-      g_iMax_episode_len = 3000;
-      gsl_matrix* discard = inverted_pendulum_simulator( 100 );
-      gsl_matrix_free( discard );
-      unsigned int mean_control_steps = g_iNb_samples/100;
-      gsl_matrix_free( omega_lstd );
-      printf("%d-LSTD %d %d %d\n", m_exp[i], nb_samples_exp,
-	     D_len[j],  mean_control_steps );
-    }
+  gsl_matrix* omega_imitation;
+  int m_exp[] = {1,10,100};
+  int D_len[] = {1000,1100,1200,1250,1300,1350,1400,1500,1600};
+  //Premières observations :  sans matrice D,
+  //quelque soit la taille de la matrice D_E,
+  //la variante LSTD ne fonctionne pas tandis que
+  //la variante MC fonctionne
+  for( int i=0; i<3 ; i++ ){
+    g_iNb_samples = 0;
+    g_mOmega =  omega_expert;
+    g_iMax_episode_len = 100;
+    D_expert = inverted_pendulum_simulator( m_exp[i] );
+    unsigned int nb_samples_exp = g_iNb_samples;
+    g_iMax_episode_len = 100;
[…]

### 2011-04-13 main.tex
Seconde soumission à JFPDA : correction de la mention de l'INRIA
Changements pour le nom des courbes
Dernière soumission JFPDA (sauf la dernière figure)

@@
 
-\title{Apprentissage par imitation étendu au cas batch, off-policy et sans modèle}
+\title{Apprentissage par imitation dans un cadre \emph{batch}, \emph{off-policy} et sans modèle}
 %\author{Robert}
@@
  2. UMI 2958 CNRS - GeorgiaTech, France\\
- 3. Equipe ABC, Loria, INRIA Grand-Est, Nancy, France.
+ 3. Equipe ABC, LORIA-CNRS, France.
  }
@@
 \begin{abstract}
- Ce papier s'intéresse au problème de l'apprentissage par imitation, c'est à dire la résolution du problème du contrôle optimal à partir de données tirées d'une desmonstration d'expert.
+ Ce papier traite le problème de l'apprentissage par imitation, c'est à dire la résolution du problème du contrôle optimal à partir de données tirées de démonstrations d'un expert.
 L'apprentissage par renforcement inverse (IRL) propose un cadre efficace pour résoudre ce problème.
-En se basant sur l'hypothèse que l'expert maximise un critère, l'IRL essaie d'apprendre la récompense qui définit ce critère à partir de trajectoires d'exemple.
-Beaucoup d'algorithmes d'IRL font l'hypothèse de l'existence d'un bon approximateur linéaire pour la fonction de récompense et calculent l'attribut moyen (le cumul moyen pondéré des fonctions de base, relatives à la paramétrisation linéaire supposée de la récompense, évaluées en les états d'une trajectoire associée à une certaine politique)  via une estimation de Monte-Carlo.
-Cela implique d'avoir accès à des trajectoires complète de l'expert ainsi qu'à au moins un modèle génératif pour tester les politiques intermédiaires.
-Dans ce papier nous introduisons une méthode de différence temporelle, LSTD-$\mu$, pour calculer cet attribut moyen.
+En se basant sur l'hypothèse que l'expert maximise une fonction de valeur,  l'IRL essaie d'apprendre la récompense qui définit cette dernière à partir de trajectoires d'exemple.
+Beaucoup d'algorithmes d'IRL font l'hypothèse de l'existence d'unn approximateur linéaire pour la fonction de récompense et calculent l'attribut moyen (le cumul moyen pondéré des fonctions de base, relatives à la paramétrisation linéaire supposée de la récompense, évaluées en les états d'une trajectoire associée à une certaine politique)  via une estimation de Monte-Carlo.
+Cela implique d'avoir accès à des trajectoires complètes de l'expert ainsi qu'à au moins un modèle génératif pour tester les politiques intermédiaires.
+Dans ce papier nous introduisons une méthode de différences temporelles, LSTD-$\mu$, pour calculer cet attribut moyen.
 Cela permet d'étendre l'apprentissage par imitation aux cas \emph{batch} et \emph{off-policy}.
@@
 
-Optimal control consist in having a machine controlling a system
+Optimal control consists in putting a machine in control of a system
 with the goal of fulfilling a specific task, optimality being
@@
 
-However, defining optimality (and thus the reward function) can
+However, defining optimality (through the reward function) can
 itself be a challenge. If the system can be empirically controlled
@@
 consists in learning a reward function under which the policy
-demonstrated by an expert is optimal. Mimicking the expert therefore
+demonstrated by the expert is optimal. Mimicking the expert therefore
 ends up to learning an optimal policy according to this reward
@@
 
-There roughly exist three families of IRL algorithms:
+There roughly exists three families of IRL algorithms:
 feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
@@
 function) is close to the feature expectation of the expert policy.
-This is a sufficient condition to have close value functions. The
+This is a sufficient condition to have close value functions (for any parameterized reward function, and therefore particularly the optimized one). The
 second family expresses IRL as a constrained optimization problem in
-which expert's examples have higher expected reward than all other
+which expert's examples have higher expected cumulative reward than all other
 policies by a certain margin. Moreover, suboptimality of the expert
@@
 computed exactly (which requires knowing analytically the dynamics
-of the system) or with a Monte Carlo approach (which requires to
-simulate the system). The contribution of this paper is LSTD-$\mu$,
+of the system) or with a Monte Carlo approach (which requires simulating the system). The contribution of this paper is LSTD-$\mu$,
 a new temporal-difference-based algorithm to estimate these feature
@@
 approach can be extended to other algorithms based on feature
-expectation computation~\citep{syed2008apprenticeship,syed2008game,ziebart2008maximum}.
+expectation computation~\citep{syed2008apprenticeship,syed2008game,ziebart2008maximum}, or even \citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting}.
 
@@
 has a somehow weaker objective: it aims at learning a policy
-$\hat{\pi}$ which is (approximately) as good as the expert policy
+$\tilde{\pi}$ which is (approximately) as good as the expert policy
 $\pi_E$ under the unknown reward function $R^*$, for a known initial
 state $s_0$ (this condition can be weakened by assuming a
-distribution for initial states, this is not done here for clarity
+distribution over initial states ; this is not done here for clarity
 of exposition). Now, the approach proposed in~\citep{abbeel2004apprenticeship} is
@@
 
-We assume that the true reward function belongs to some hypothesis
+We assume that the true reward funnotiction belongs to some hypothesis
 space $\mathcal{H}_\phi = \{\theta^T \phi(s),
@@
   closest to the one of the expert.}).
-  Notice that the last policy is not necessary the best;
+  Notice that the last policy is not necessary the best (as illustrated in Section \ref{sec:exp});
   %
@@
   \hat{\mu}_E(s_0) = \frac{1}{m} \sum_{h=1}^m \sum_{t=0}^\infty
-  \gamma^i \phi(s_t^{(h)})
+  \gamma^t \phi(s_t^{(h)})
 \end{equation}
@@
 off-policy setting. The solely available data is a set of
-transitions sampled according to the expert policy. The
+transitions sampled according to the expert policy  (and possibly a set of sub-optimal but good trajectories). The
 corresponding feature expectation for the starting state $s_0$ is
@@
 such as convergence~\citep{nedic2003least} or
-finite-sample~\citep{lazaric2010finiteLSTD} analysis for example.
+finite-sample~\citep{lazaric2010finiteLSTD} analysis for example, also, the algorithm learns from transitions and not trajectories.
 
@@
 \label{sec:exp}
-Preliminary experimental results have been obtained.
+This section provides some preliminary experimental results on a simple maze benchmark.
 Subsection \ref{ssec:exp} details the protocol and the results while subsection \ref{ssec:quality} inspects the meaning of the different quality criteria.
@@
 The experimental benchmark chosen here is one of those proposed in \citep{ng2000algorithms}, a 5x5 grid world.
-The agent is in one of the case of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
+The agent is in one of the cell of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
 With probability 0.9, the agent moves in the intended direction.
 With probability 0.1, the direction is randomly drawn (and thus have probability 0.25 to match the original direction).
-The reward optimized by the expert is 0 everywhere except in the upper-right square, where it is 1.
-For every policy, an episode ends when the upper right square is attained, or after 20 steps.
+The reward optimized by the expert is 0 everywhere except in the upper-right cell, where it is 1.
+For every policy, an episode ends when the upper right cell is attained, or after 20 steps.
 At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is).\\
@@
 Both the state and action spaces are finite and of small cardinality.
-Hence, the chosen feature functions $\phi$ and $\psi$ (see equations (\ref{eqn:phi}) and (\ref{eqn:psi}) for a remainder of their meaning) are the typical features of a tabular representation : 0 everywhere but at a given index $i$ for which the value is 1.
-There is a bijective mapping from the input space elements to the indexes $i$ for which the coordinate of the feature is 1.\\
+Hence, the chosen feature functions $\phi$ and $\psi$ (see equations (\ref{eqn:phi}) and (\ref{eqn:psi}) for a remainder of their meaning) are the typical features of a tabular representation :0 everywhere except the component corresponding to the current state (-action pair).\\
 
-Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as mc\_ANIRL\footnote{Abbeel \& Ng IRL}) and our adaptation (referred t as lstd\_ANIRL) are tested side by side.
-The MDP solver of mc\_ANIRL is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly.
-$\mu^{\pi^(j)}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to reduce the variance to less than $0.0001\%$ of the mean.
+Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as the MC variant) and our adaptation (referred to as the LSTD variant) are tested side by side.
+The MDP solver of the mc variant is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly.
+Both $\mu^{\pi^(j)}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to make the variance negligible.
 We consider both these computations as perfect theoretical solvers for all intended purpose on this toy problem.
 We thus are in the case intended by \citet{abbeel2004apprenticeship}.\\
-On the other hand lstd\_ANIRL is used as if it could not have access to the simulator.
-It uses LSPI and LSTD$\mu$, fed only with the expert's transitions.
+On the other hand the LSTD variant is used without accessing a simulator.
+It uses LSPI and LSTD$\mu$, fed only with the expert's transitions (although we could also use non expert transitions if available).
 This corresponds to a real-life setting where data generation is expensive and the system can not be controlled by an untrained machine.\\
 
-We want to compare the porformance of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
+We want to compare the efficiency of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
 Indeed as they are quite costly (in both means and human time) to generate they are often not in abundance hence the critical need for an algorithm to be expert-sample efficient.
+Having a simulator at sake can also be difficult.
+For the simple problem we use this is however not an issue.
 The discussion about the choice of the performance metric has its own dedicated subsection (subsection \ref{ssec:quality}).
-We use here the $||\mu^{\pi^(j)}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term.\\
+We use here the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term.\\
 
-Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^(j)}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
-The best policy is found by lstd\_ANIRL after one iteration only\footnote{Precise reasons for what it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in mc\_ANIRL convergence happens after at least ten iterations and the best policy is not always the last.
-The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is what is plotted here.\\
+Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^{(j)}}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
+The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons for what it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations.
+The best policy is not always the last, and although it experimentally always have been with the LSTD variant, there is absolutely no way to tell whether this is a feature.
+The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is plotted here.\\
 
-We can see that although lstd\_ANIRL is not as good as mc\_ANIRL when very few samples are available, both algorithm quickly converge to almost the same value (although not low, the number of run so far is not statistically significant enough to allow us to state that the seemingly slightly better performance of lstd\_ANIRL is not an artifact).
-The fact that lstd\_ANIRL can work in a batch, off-policy and model-free setting, while requesting less computing power than mc\_ANIRL makes it suitable to a range of task inaccessible to the mc\_ANIRL version.
+We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithm quickly converge to almost the same value ; our version converges to a slightly lower error value.
+The error bars shown correspond to the standard deviation over 100 runs.
+The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of task inaccessible to the MC variant.
 \subsection{Discussion about the quality criterion}
 \label{ssec:quality}
-Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion ; these plots show four different quality ciretria during a run of mc\_ANIRL and several runs of lstd\_ANIRL.
-For the latter, the abscissa can not be the number of iterations as it always converges in one iteration.\\
-
+Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion ; Figure \ref{fig:A} shows four different quality ciretria during a run of the MC variant.
+Figure \ref{fig:B} shows the same criteria for several runs of the LSTD variant, as the abscissa can not be the number of iterations (as in Figure \ref{fig:A}) because it always converges in one iteration.
 The $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
@@
 
-The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in figure \ref{fig:A} means that for it has access to a cheap simulator, mc\_ANIRL works as if it had access to the exact values.
-This however can not be said of lstd\_ANIRL, for which the two curves are indeed different (Figure \ref{fig:B}).
-Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for lstd\_ANIRL, as it can introduce an error int the halt criterion of the algorithm\footnote{This equation differs from what has been given as the second step of the algorithm in section \ref{sec:back}.
+The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in Figure \ref{fig:A} means that, for it has access to a cheap simulator, the MC variant works as if it had access to the exact values.
+This however can not be said of the LSTD variant, for which the two curves are indeed different (Figure \ref{fig:B}).
+Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for our variant, as it can introduce an error in the stopping criterion of the algorithm\footnote{This equation differs from what has been given as the second step of the algorithm in section \ref{sec:back}.
 This is because we implemented the version of the algorithm that uses a projection (see \citet{abbeel2004apprenticeship}'s supplementary material for an in-depth discussion of both versions).
@@
 
-It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in mc\_ANIRL either, as it can be low when the policy is bad.
-The best policy, however, can be easily chosen from the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which mc\_ANIRL can compute.
+It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in the MC variant either, as it can be low (and thus halt the algorithm) when the policy is bad.
+The best policy, however, can be easily chosen among all those created during the execution of the algorithm thanks to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which the MC variant can compute.
 When this term is low, the objective performance (that is, $V^{\pi_E}(s_0)-V^{\pi^{(j)}}(s_0)$ with respect to the unknow true reward function) is low too.
@@
 \label{sec:conclusion}
-Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning vie inverse reinforcement learning to a batch, model-free, off-policy setting at the cost of the need for a bigger number of samples from the expert.
+Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning vie inverse reinforcement learning to a batch, model-free, off-policy setting.
+Experimentally, there is a need for a bigger number of samples from the expert.
 We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach.
-Furthermore, our method requires less computing power as it converges after only one iteration (that is, as far as we witnessed in our experimental setting).\\
-
-The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
-The exact reason why LSTD$\mu$ converges so fast should be investigated and a bigger experiment should be run, possibly on a continuous problem.\\
+The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
 \begin{figure}
 \centering
-\input{../Code/courbe_e}
+\input{../Code/GridWorld/both_error_EB}
 \caption{$||\mu^{\pi}(s_0) - \mu^{\pi_E}(s_0)||_2$ where $\pi$ is the policy for which this term is the lowest for the whole run, with respect to the number of samples available from the expert.
-Our algorithm is not as good as \citet{abbeel2004apprenticeship} with very few samples, but make up for it very soon.}
+Our algorithm is not as good as \citet{abbeel2004apprenticeship} with very few samples, but make up for it very soon and converges to a lower value.
+The error bars represent the standard deviation over 100 runs.}
 \label{fig:E}
@@
 \centering
-\input{../Code/courbe_a}
-\caption{Different criteria with respect to the number of iterations for a run of mc\_ANIRL.}
+\input{../Code/GridWorld/criteria_mc}
+\caption{Different criteria with respect to the number of iterations for a run of the MC variant.}
 \label{fig:A}
@@
 \centering
-\input{../Code/courbe_b}
-\caption{Different criteria with respect to the number of samples from the expert, for several runs of lstd\_ANIRL.}
+\input{../Code/GridWorld/criteria_lstd_EB}
+\caption{Different criteria with respect to the number of samples from the expert, for several runs of the LSTD variant.
+We can see that the algorithm is blind, as all it has access to is always zero.
+The true error values, however, smoothly converge to something small.
+Knowing how many expert samples are actually needed in a real world problem is an open question.
+The error bars represent the standard deviation over 100 runs.}
 \label{fig:B}
+\end{figure}
+\begin{figure}
+\centering
+\input{../Code/InvertedPendulum/threshold_EB}
+\caption{Continuous}
+\label{fig:DSKLFLFK}
+\end{figure}
+\begin{figure}
+\centering
+\input{../Code/InvertedPendulum/threshold}
+\caption{Continuous}
+\label{fig:DSKLFsdfdf}
 \end{figure}

### 2011-04-14 InvertedPendulum/plot.c
Courbes threshol et threshold_EB modiiées pour comparer LSTD et MC, telles qu'utilisées dans un drat d'IJCAI

@@ -104,36 +104,44 @@ int main( void ){
   
 
   gsl_matrix* D_expert;
-  int m_exp[] = {1,100};
   int D_len[] = {1,200,300,400,500,600,700,800,900,1000,1100,1200,1300,1400,1500,1600,1700,1800,1900,2000,3000};
-  //Cette courbe illustre l'influence des
-  //paramètres que sont la taille de D et la taille 
-  //de D_E sur la variante LSTDmu
-  for( int i=0; i<2 ; i++ ){
-    for( int j=0; j<21;j++){
-      g_iNb_samples = 0;
-      g_mOmega =  omega_expert;
-      g_iMax_episode_len = 100;
-      D_expert = inverted_pendulum_simulator( m_exp[i] );
-      unsigned int nb_samples_exp = g_iNb_samples;
-      gsl_matrix_view sub_D = 
-	gsl_matrix_submatrix( D, 0, 0, D_len[j], TRANS_WIDTH );
-      g_iMax_episode_len = 100;
-      gsl_matrix* omega_lstd =
-	proj_lstd_lspi_ANIRL( D_expert, &sub_D.matrix );
-      g_mOmega = omega_lstd;
-      g_iNb_samples = 0;
-      g_iMax_episode_len = 3000;
-      gsl_matrix* discard = inverted_pendulum_simulator( 100 );
-      gsl_matrix_free( discard );
-      unsigned int mean_control_steps = g_iNb_samples/100;
-      gsl_matrix_free( omega_lstd );
-      printf("%d-LSTD %d %d %d\n", m_exp[i], nb_samples_exp,
-	     D_len[j],  mean_control_steps );
-    }
-  }
-
-  
+  //Cette courbe illustre l'influence de
+  // la taille de D sur les deux variantes
+  for( int j=0; j<21;j++){
+    g_iNb_samples = 0;
+    g_mOmega =  omega_expert;
+    g_iMax_episode_len = 100;
+    D_expert = inverted_pendulum_simulator( 1 );
+    unsigned int nb_samples_exp = g_iNb_samples;
+    gsl_matrix_view sub_D = 
+      gsl_matrix_submatrix( D, 0, 0, D_len[j], TRANS_WIDTH );
+    g_iMax_episode_len = 100;
+    gsl_matrix* omega_lstd =
+      proj_lstd_lspi_ANIRL( D_expert, &sub_D.matrix );
+    g_mOmega = omega_lstd;
+    g_iNb_samples = 0;
[…]

### 2011-04-14 LSPI.c

@@ -1,28 +1,26 @@
 #include <gsl/gsl_matrix.h>
-/* #include <gsl/gsl_blas.h> */
 #include "LSTDQ.h"
 #include "utils.h"
 #include "greedy.h"
-/* #include "LSPI.h" */
-/* #include "criteria.h" */
+#include "LSPI.h"
 #include "RL_Globals.h"
 
 gsl_matrix* lspi( gsl_matrix* D, gsl_matrix* omega_0 ){
   unsigned int nb_iterations = 0;
   gsl_matrix* omega = gsl_matrix_alloc( g_iK, 1 );
   g_mOmega = omega;
-  //\omega'\leftarrow \omega_0
+  //$\omega'\leftarrow \omega_0$
   gsl_matrix* omega_dash = gsl_matrix_alloc( g_iK, 1 );
   gsl_matrix_memcpy( omega_dash, omega_0 );
   double norm;
   //Repeat
   do{
-    //\omega \leftarrow \omega'
+    //$\omega \leftarrow \omega'$
     gsl_matrix_memcpy( omega, omega_dash );
-    //\omega' \leftarrow lstd_q(D,k,\phi,\gamma,\omega)
+    //$\omega' \leftarrow lstd_q(D,k,\phi,\gamma,\omega)$
     gsl_matrix_free( omega_dash );
     omega_dash = lstd_q( D, &greedy_policy );
-    //until ( ||\omega'-\omega|| < \epsilon )
+    //until ( $||\omega'-\omega|| < \epsilon$ )
     norm = diff_norm( omega_dash, omega );
     nb_iterations++;
   }while( norm >= g_dEpsilon_lspi && 

### 2011-06-06 main.tex
Mise à niveau IJCAI->JFPDA

@@
 feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
-marge-maximization-based~\citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
+margin-maximization-based~\citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
 and approaches based on the parameterization of the policy by the
@@
 of the system) or with a Monte Carlo approach (which requires simulating the system). The contribution of this paper is LSTD-$\mu$,
-a new temporal-difference-based algorithm to estimate these feature
-expectations. It relaxes the predeceasing assumptions: transitions
+a new temporal-difference-based algorithm for estimating these feature
+expectations. It relaxes the preceding assumptions: transitions
 of the expert are sufficient (rather than full trajectories) and nor
@@
 learning through IRL, which is the problem at hand in this paper,
-has a somehow weaker objective: it aims at learning a policy
+has a somewhat weaker objective: it aims at learning a policy
 $\tilde{\pi}$ which is (approximately) as good as the expert policy
@@
 that value functions are searched for in some  hypothesis space
-$\mathcal{H}_\psi = \{ \hat{V}_\xi(s) = \sum_{i=1}^q \xi_i \psi_i(s)
+$\mathcal{H}_\psi = \{ \hat{V}_\xi(s) = \sum\limits_{i=1}^q \xi_i \psi_i(s)
 = \xi^T \psi(s), \xi\in\mathbb{R}^q\}$. As reward and value
@@
 
-Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as the MC variant) and our adaptation (referred to as the LSTD variant) are tested side by side.
-The MDP solver of the mc variant is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly.
-Both $\mu^{\pi^(j)}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to make the variance negligible.
-We consider both these computations as perfect theoretical solvers for all intended purpose on this toy problem.
-We thus are in the case intended by \citet{abbeel2004apprenticeship}.\\
-On the other hand the LSTD variant is used without accessing a simulator.
-It uses LSPI and LSTD$\mu$, fed only with the expert's transitions (although we could also use non expert transitions if available).
-This corresponds to a real-life setting where data generation is expensive and the system can not be controlled by an untrained machine.\\
-
 We want to compare the efficiency of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
 Indeed as they are quite costly (in both means and human time) to generate they are often not in abundance hence the critical need for an algorithm to be expert-sample efficient.
-Having a simulator at sake can also be difficult.
+Having a simulator at one's disposal can also be difficult.
 For the simple problem we use this is however not an issue.
-The discussion about the choice of the performance metric has its own dedicated subsection (subsection \ref{ssec:quality}).
-We use here the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term.\\
+The discussion about the choice of the performance metric has its own dedicated subsection (subsection~\ref{ssec:quality}).
+We use here the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term.
+Fig.~\ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^{(j)}}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
+The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons for why it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations.
+The best policy is not always the last, and although it experimentally always have been with the LSTD variant, there is absolutely no way to tell whether this is a feature.
+The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is plotted here.
+We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithm quickly converge to almost the same value ; our version converges to a slightly lower error value.
+The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of task inaccessible to the MC variant, the requirement of a simulator being often constraining.
+\subsubsection{Inverted Pendulum}
+\begin{figure}
+\centering
+\resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold}}
+\caption{Number of balancing steps for the policies found by both variants with respect to the number of samples from a sub-optimal policy given to LSPI.
+Note the absence of middle ground in the quality.}
+\label{fig:threshold}
+\end{figure}
+\begin{figure}
+\centering
+\resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold_EB}}
+\caption{Same plot as Fig.~\ref{fig:threshold}, with mean and standard deviation over 100 runs.
+The big standard deviation stems from the absence of middle ground in the quality of the policies.
+The increasing mean w.r.t.
+the abscissa means that the proportion of good policies increases with the number of samples from a sub-optimal policy given to LSPI.}
+\label{fig:threshold_EB}
+\end{figure}
+Another classic toy problem is the \emph{inverted pendulum}, used for example in \citep{lagoudakis2003least} from where we drew the exact configuration (mass of the pendulum, length of the arm, definition of the feature space, etc.).
+In this setting, the machine must imitate an expert who maintains the unstable equilibrium of an inverted pendulum allowed to move along only one axis.
+A balancing step is considered successful (reward $0$) when the angle is less than $\pi\over 2$.
+It is considered a failure (reward $-1$) when the pendulum has fallen (i.e.
+the angle is more than $\pi \over 2$).
+A run is stopped after a failure or after 3000 successful balancing steps.\\
 
-Figure \ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^{(j)}}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
-The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons for what it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations.
-The best policy is not always the last, and although it experimentally always have been with the LSTD variant, there is absolutely no way to tell whether this is a feature.
-The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is plotted here.\\
-
-We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithm quickly converge to almost the same value ; our version converges to a slightly lower error value.
-The error bars shown correspond to the standard deviation over 100 runs.
-The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of task inaccessible to the MC variant.
 \subsection{Discussion about the quality criterion}
@@
 \end{figure}
-\begin{figure}
-\centering
-\input{../Code/InvertedPendulum/threshold_EB}
-\caption{Continuous}
-\label{fig:DSKLFLFK}
-\end{figure}
-\begin{figure}
-\centering
-\input{../Code/InvertedPendulum/threshold}
-\caption{Continuous}
-\label{fig:DSKLFsdfdf}
-\end{figure}
+
+Fig.~\ref{fig:A}~and~\ref{fig:B} illustrate (using the GridWorld problem) the difficulty of choosing the quality criterion ; Fig.~\ref{fig:A} shows four different quality criteria during a run of the MC variant.
+Fig.~\ref{fig:B} shows the same criteria for several runs of the LSTD variant with a varying number of samples from the expert.
+The $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
+It bears an interesting relation with the difference between the expert's value function and the current value function in the initial state with respect \emph{to the current reward} (Eq.~\ref{eqn:vs0}).\\
+
+The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in Fig.~\ref{fig:A} means that, for it has access to a cheap simulator, the MC variant works as if it had access to the exact values.
+This however cannot be said of the LSTD variant, for which the two curves are indeed different (Fig.~\ref{fig:B}).
+Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for our variant, as it can introduce an error in the stopping criterion of the algorithm.\\
+
+It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in the MC variant either, as it can be low (and thus halt the algorithm) when the policy is bad.
+The best policy, however, can be easily chosen among all those created during the execution of the algorithm thanks to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which the MC variant can compute.
+When this term is low, the objective performance (that is, $V^{\pi_E}(s_0)-V^{\pi^{(j)}}(s_0)$ with respect to the unknown true reward function) is low too.
+\section{Conclusion}
+\label{sec:conclusion}
+Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning via inverse reinforcement learning to a batch, model-free, off-policy setting.
+Experimentally, there is a need for either a slightly bigger number of samples from the expert or some samples from a sub-optimal policy.
+We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach.
+Furthermore, tranferring the reward and not he policy may overcome this difficulty.
+We intend to do this in a real life setting.
+The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
+
+\bibliographystyle{named}
 %\bibliographystyle{named}

### 2011-06-06 main.tex
Corrections suggérées par Yann

@@
 En se basant sur l'hypothèse que l'expert maximise une fonction de valeur,  l'IRL essaie d'apprendre la récompense qui définit cette dernière à partir de trajectoires d'exemple.
-Beaucoup d'algorithmes d'IRL font l'hypothèse de l'existence d'unn approximateur linéaire pour la fonction de récompense et calculent l'attribut moyen (le cumul moyen pondéré des fonctions de base, relatives à la paramétrisation linéaire supposée de la récompense, évaluées en les états d'une trajectoire associée à une certaine politique)  via une estimation de Monte-Carlo.
+Beaucoup d'algorithmes d'IRL font l'hypothèse de l'existence d'un approximateur linéaire pour la fonction de récompense et calculent l'attribut moyen (le cumul moyen pondéré des fonctions de base, relatives à la paramétrisation linéaire supposée de la récompense, évaluées en les états d'une trajectoire associée à une certaine politique)  via une estimation de Monte-Carlo.
 Cela implique d'avoir accès à des trajectoires complètes de l'expert ainsi qu'à au moins un modèle génératif pour tester les politiques intermédiaires.
@@
 leads to a linearly parameterized value function, the associated
-feature vector being the so-called feature expectation (defined as
-the cumulative discounted reward's feature vector under a given
-policy, see Section~\ref{sec:back} for a formal definition).
+feature vector being the so-called feature expectation (see Section~\ref{sec:back} for a formal definition).
 These approaches learn a reward function such that the feature
@@
 optimized), it is possible to estimate the likelihood of a set of
-state-action pairs provided by the expert. The algorithms differs in
+state-action pairs provided by the expert. The algorithms differ in
 the way this likelihood is maximized.
@@
 feature expectations. This paper focuses on the algorithm introduced
-in~\citep{abbeel2004apprenticeship}, but we are confident that the proposed
-approach can be extended to other algorithms based on feature
-expectation computation~\citep{syed2008apprenticeship,syed2008game,ziebart2008maximum}, or even \citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting}.
+in~\citep{abbeel2004apprenticeship}, but the proposed
+approach can be used in other algorithms based on feature
+expectation computation.
 
@@
 A sequential decision problem is often framed as a Markov Decision
-Process (MDP)~\citep{puterman1994markov}. An MDP is a tuple
+Process (MDP)~\citep{puterman1994markov}, a tuple
 $\{S,A,P,R,\gamma\}$ with $S$ being the state space, $A$ the action
@@
 (assumed to be absolutely bounded by 1) and $\gamma\in[0,1[$ a
-discounting factor. A policy $\pi\in A^S$ maps states to action. The
+discounting factor. A policy $\pi\in A^S$ maps states to actions. The
 quality of a policy is quantified by the associated value function
@@
 Dynamic programming aims at finding the optimal policy $\pi^*$, that
-is one of the policies associated to the optimal value function,
+is one of the policies associated with the optimal value function,
 $V^* = \argmax_\pi V^\pi$, which maximizes the value for each state.
@@
 function) is unknown, learning the optimal control through
-interactions is addressed by reinforcement learning.
+interactions is addressed by RL.
 
-For inverse reinforcement learning, the problem is reversed. It is
+For IRL, the problem is reversed. It is
 assumed that an expert acts according to an optimal policy $\pi_E$,
 this policy being optimal according to some unknown reward function
-$R^*$. The goal of inverse reinforcement learning is to learn this
+$R^*$. The goal is to learn this
 reward function from sampled trajectories of the expert. This is a
@@
 $\pi_E$ under the unknown reward function $R^*$, for a known initial
-state $s_0$ (this condition can be weakened by assuming a
-distribution over initial states ; this is not done here for clarity
-of exposition). Now, the approach proposed in~\citep{abbeel2004apprenticeship} is
+state $s_0$ (this condition can be easily weakened by assuming a
+distribution over initial states). Now, the approach proposed in~\citep{abbeel2004apprenticeship} is
 presented.
 
-We assume that the true reward funnotiction belongs to some hypothesis
+We assume that the true reward function belongs to some hypothesis
 space $\mathcal{H}_\phi = \{\theta^T \phi(s),
@@
 \end{align}
-This last equation uses the Cauchy-Schwartz inequality and the
+This equation uses the Cauchy-Schwarz inequality and the
 assumption that $\|\theta\|_2\leq 1$. Therefore, the approach
@@
 $R^*$, but to act as well as the expert under this reward function
-(and actually under any reward function, notably the unknown optimal
-one).
+(and actually under any reward function).
 
@@
   expert (see~\citep{abbeel2004apprenticeship} for details on how to choose this
-  policy\footnote{A simple solution, not mentioned in~\citep{abbeel2004apprenticeship},
-  is to choose the policy of which the feature expectation is the
-  closest to the one of the expert.}).
-  Notice that the last policy is not necessary the best (as illustrated in Section \ref{sec:exp});
+  policy).
+  Notice that the last policy is not necessarily the best (as illustrated in Section~\ref{sec:exp});
   %
@@
 \end{enumerate}
-There remains three problems: computing the feature expectation of
+There remain three problems: computing the feature expectation of
 the expert, solving the MDP and computing feature expectations of
@@
 $\pi$ for the MDP considered with the $i^\text{th}$ basis function
-$\phi_i(s)$ as the reward function. There exists many algorithms to
+$\phi_i(s)$ as the reward function. There exist many algorithms to
 estimate a value function, any of them can be used to estimate
@@
 introducing the state-action feature expectation is that this
-additional degree of freedom allows off-policy learning. Assuming
-that $\mathcal{H}_\psi$ defines state-action feature vectors and
-that a set of transitions $\{(s_t,a_t,r_t,s_{t+1})_{1\leq t \leq
-n}\}$ sampled according to some behaviorial policy $\pi_0$ is
-available, the parameter vector $\xi_i^*$ related to the
-$i^\text{th}$ component of the state-action feature expectation
-$\mu^\pi(s)$ (policy $\pi$ being different from policy $\pi_0$) is
-given by:
-\begin{equation}
-  \xi_i^* = \left(\sum_{t=1}^n
-  \psi(s_t,a_t)(\psi(s_t,a_t)-\gamma\psi(s'_{t},\pi(s'_t)))^T\right)^{-1}
-  \sum_{t=1}^n \psi(s_t,a_t) \phi_i(s_t)
-\end{equation}
+additional degree of freedom allows off-policy learning.
 Extending LSTD-$\mu$ to state-action LSTD-$\mu$ is done in the same
@@
 Given the (state-action) LSTD-$\mu$ algorithm, the apprenticeship
-learning algorithm of~\citep{abbeel2004apprenticeship} (see
-Section~\ref{sec:back}) can be easily extended to a batch and
+learning algorithm presented in Section~\ref{sec:back} can be easily extended to a batch and
 off-policy setting. The solely available data is a set of
-transitions sampled according to the expert policy  (and possibly a set of sub-optimal but good trajectories). The
+transitions sampled according to the expert policy  (and possibly a set of sub-optimal trajectories). The
 corresponding feature expectation for the starting state $s_0$ is
@@
 such as convergence~\citep{nedic2003least} or
-finite-sample~\citep{lazaric2010finiteLSTD} analysis for example, also, the algorithm learns from transitions and not trajectories.
-
+finite-sample~\citep{lazaric2010finiteLSTD} analysis for example.
 \section{Experimental benchmark}
 \label{sec:exp}
-This section provides some preliminary experimental results on a simple maze benchmark.
-Subsection \ref{ssec:exp} details the protocol and the results while subsection \ref{ssec:quality} inspects the meaning of the different quality criteria.
+This section provides experimental results on two complementary problems.
+Subsection~\ref{ssec:exp} details the protocol and the results while subsection~\ref{ssec:quality} inspects the meaning of the different quality criteria.
 \subsection{Experiment description and results}
 \label{ssec:exp}
-The experimental benchmark chosen here is one of those proposed in \citep{ng2000algorithms}, a 5x5 grid world.
-The agent is in one of the cell of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
+\subsubsection{GridWorld}
+\begin{figure}
+\centering
+\resizebox{\columnwidth}{!}{\input{../Code/GridWorld/both_error_EB}}
+\caption{$||\mu^{\pi}(s_0) - \mu^{\pi_E}(s_0)||_2$ with respect to the number of samples available from the expert. The error bars represent the standard deviation over 100 runs.}
+\label{fig:E}
+\end{figure}
+The first experimental benchmark chosen here is one of those proposed in \citep{ng2000algorithms}, a 5x5 grid world.
+The agent is in one of the cells of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
 With probability 0.9, the agent moves in the intended direction.
-With probability 0.1, the direction is randomly drawn (and thus have probability 0.25 to match the original direction).
+With probability 0.1, the direction is randomly drawn.
 The reward optimized by the expert is 0 everywhere except in the upper-right cell, where it is 1.
 For every policy, an episode ends when the upper right cell is attained, or after 20 steps.
-At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is).\\
+At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is).
+Both the state and action spaces are finite and of small cardinality.
+Hence, the chosen feature functions $\phi$ and $\psi$ are the typical features of a tabular representation: 0 everywhere except for the component corresponding to the current state (-action pair).\\
 
-Both the state and action spaces are finite and of small cardinality.
-Hence, the chosen feature functions $\phi$ and $\psi$ (see equations (\ref{eqn:phi}) and (\ref{eqn:psi}) for a remainder of their meaning) are the typical features of a tabular representation :0 everywhere except the component corresponding to the current state (-action pair).\\
+Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as the MC variant) and our adaptation (referred to as the LSTD variant) are tested side by side.
+The MDP solver of the MC variant is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly.
+Both $\mu^{\pi^{(j)}}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to make the variance negligible.
+We consider both these computations as perfect theoretical solvers for all intended purpose on this toy problem.
+We thus are in the case intended by \citet{abbeel2004apprenticeship}.
+On the other hand the LSTD variant is used without accessing a simulator.
+It uses LSPI and LSTD-$\mu$, fed only with the expert's transitions (although we could also use non expert transitions  to compute intermediate policies, if available).
+This corresponds to a real-life setting where data generation is expensive and the system cannot be controlled by an untrained machine.\\
 
-We want to compare the efficiency of both versions of the algorithm with repect to the number of samples available from the expert, as these samples usually are the bottleneck.
+We want to compare the efficiency of both versions of the algorithm with respect to the number of samples available from the expert, as these samples usually are the bottleneck.
 Indeed as they are quite costly (in both means and human time) to generate they are often not in abundance hence the critical need for an algorithm to be expert-sample efficient.
@@
 Fig.~\ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^{(j)}}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
-The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons for why it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations.
-The best policy is not always the last, and although it experimentally always have been with the LSTD variant, there is absolutely no way to tell whether this is a feature.
+The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons why it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations.
+The best policy is not always the last, and although it has always been observed so far with the LSTD variant, there is absolutely no way to tell whether this is a feature.
 The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is plotted here.
-We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithm quickly converge to almost the same value ; our version converges to a slightly lower error value.
-The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of task inaccessible to the MC variant, the requirement of a simulator being often constraining.
+We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithms quickly converge to almost the same value ; our version converges to a slightly lower error value.
+The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of tasks inaccessible to the MC variant, the requirement of a simulator being often constraining.
 \subsubsection{Inverted Pendulum}
@@
 It is considered a failure (reward $-1$) when the pendulum has fallen (i.e.
-the angle is more than $\pi \over 2$).
+the angle exceeds $\pi \over 2$).
 A run is stopped after a failure or after 3000 successful balancing steps.\\
 
+This problem bears two fundamuntal differences with the previous one and thus comes as complementary to it.
+First, it presents a continuous state space whereas in the GridWorld a tabular representation was possible.
+Then, the initial state is randomly drawn from a non singleton subspace of the state space.
+This last point is addressed differently by the LSTD variant and the MC variant.
+the MC variant will naturally sample the initial state distribution at each new trajectory from the expert.
+The LSTD variant, on the other hand, still does not need complete trajectories from the expert but mere samples.
+As it approximates the whole $\mu^{\pi}$ function and not just $\mu^\pi(s_0)$, it is possible to compute the mean of the approximation $\hat\mu^\pi$ over a few states drawn with the same distribution as the initial states.\\
+
+A striking finding on this toy problem is that the parameter controlling the success or the failure of the experiment is not the number of samples from the expert, but the quality of the samples available to LSPI to solve the MDP.
+\emph{A contrario} to what happened in the previous toy problem, there are fewer samples from the expert than what LSPI needs to solve the MDP as they do not cover the whole state space.
+When given only these samples, the LSTD variant fails.
+The MC variant, however, is not bound by such restrictions and can use as many samples as it needs from a simulator.
+Thus, with only one trajectory from the expert (100 samples) the MC variant is able to successfully balance the pendulum for more than 3000 steps.
+The problem here stems from the fact that we don't learn the reward, but a policy that maximizes the reward.
+If the optimal control problem was solved separately, learning the reward only from the samples of the expert would be possible.\\
+
+The sensitivity of both algorithms to the number of samples available to LSPI is extreme, as is shown in Fig.~\ref{fig:threshold}.
+It may seem nonsensical to restrict the number of samples available to LSPI for the MC variant as it can access a simulator, but this has been done to show that both variants exhibit the same behavior, hence allowing us to locate the source of this behavior in what both algorithms have in common (the use of LSPI inside the structure of \citep{abbeel2004apprenticeship}) excluding the point where they differ (LSTD-$\mu$ and Monte-Carlo).\\
+
+Fig.~\ref{fig:threshold_EB} shows that when given samples from a sub-optimal policy, the LSTD variant can sort the problem out, statistically almost always with 3000 sub-optimal samples, and sometimes with as low as 200 sub-optimal samples.
+Such a setting is still batch, off-policy and model-free.
+When given enough sub-optimal samples, both variants are successful with only one trajectory from the expert (i.e.
+100 samples, this is what is plotted here).
+Giving more trajectories from the expert does not improve the success rate.
 \subsection{Discussion about the quality criterion}
 \label{ssec:quality}
-Figure \ref{fig:A} and \ref{fig:B} illustrate the difficulty of choosing the quality criterion ; Figure \ref{fig:A} shows four different quality ciretria during a run of the MC variant.
-Figure \ref{fig:B} shows the same criteria for several runs of the LSTD variant, as the abscissa can not be the number of iterations (as in Figure \ref{fig:A}) because it always converges in one iteration.
-The $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
-It bears an interesting relation with the difference between the expert's value function and the current value function in the initial state with respect \emph{to the current reward} (equation \ref{eqn:vs0}).\\
-
-The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in Figure \ref{fig:A} means that, for it has access to a cheap simulator, the MC variant works as if it had access to the exact values.
-This however can not be said of the LSTD variant, for which the two curves are indeed different (Figure \ref{fig:B}).
-Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for our variant, as it can introduce an error in the stopping criterion of the algorithm\footnote{This equation differs from what has been given as the second step of the algorithm in section \ref{sec:back}.
-This is because we implemented the version of the algorithm that uses a projection (see \citet{abbeel2004apprenticeship}'s supplementary material for an in-depth discussion of both versions).
-Although the mathematical expressions are different, both convey the same meaning and are exchangeable.}, $t = ||\bar\mu^{(j)}-\hat\mu^{\pi_E}(s_0)||$ where $\bar\mu^{(j)}$ is the projection of $\mu^{\pi_E}$ on $(\mu^{\pi^{(j-1)}},\mu^{\pi^{(j)}})$.
-After the algorithm has converged there is no way to know whether the policy is actually good or not other than actually testing it.\\
-
-It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in the MC variant either, as it can be low (and thus halt the algorithm) when the policy is bad.
-The best policy, however, can be easily chosen among all those created during the execution of the algorithm thanks to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which the MC variant can compute.
-When this term is low, the objective performance (that is, $V^{\pi_E}(s_0)-V^{\pi^{(j)}}(s_0)$ with respect to the unknow true reward function) is low too.
-\section{Conclusion}
-\label{sec:conclusion}
-Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning vie inverse reinforcement learning to a batch, model-free, off-policy setting.
-Experimentally, there is a need for a bigger number of samples from the expert.
-We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach.
-The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
-\begin{figure}
-\centering
-\input{../Code/GridWorld/both_error_EB}
-\caption{$||\mu^{\pi}(s_0) - \mu^{\pi_E}(s_0)||_2$ where $\pi$ is the policy for which this term is the lowest for the whole run, with respect to the number of samples available from the expert.
-Our algorithm is not as good as \citet{abbeel2004apprenticeship} with very few samples, but make up for it very soon and converges to a lower value.
-The error bars represent the standard deviation over 100 runs.}
-\label{fig:E}
-\end{figure}
 \begin{figure}
@@
 We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach.
-Furthermore, tranferring the reward and not he policy may overcome this difficulty.
+Furthermore, tranferring the reward and not the policy may overcome this difficulty.
 We intend to do this in a real life setting.
@@
 
-\bibliographystyle{named}
+%\bibliographystyle{named}
 %\bibliographystyle{named}

### 2011-06-16 main.tex
Correction dans les adresses mail
Added the EXRL style
Abstract en anglais

@@
-\documentclass{jfpda2011}
-% Ne pas modifier ========================================================================
+\documentclass{llncs}
+%
+\usepackage{makeidx}  % allows for indexgeneration
+
 \usepackage{amssymb}
@@
 % Typographie française ==================================================================
-\usepackage[english,frenchb]{babel} % typographie francaise
+\usepackage[english]{babel} % typographie francaise
 \usepackage[utf8]{inputenc} % saisie de caracteres accentues, etc.
-\usepackage{lmodern} % polices vectorielles
-\usepackage{xspace} % gestion des espaces a la francaise
-\AddThinSpaceBeforeFootnotes % appel de note legerement decale du mot prec.
-\FrenchFootnotes % numeros de notes a la francaise
 
@@
  \institute{
- 1. Supélec-Metz Campus, IMS Research group, France, \texttt{nom.prenom@supelec.fr}\\
+ 1. Supélec-Metz Campus, IMS Research group, France, \texttt{prenom.nom@supelec.fr}\\
  2. UMI 2958 CNRS - GeorgiaTech, France\\
@@
 \begin{abstract}
- Ce papier traite le problème de l'apprentissage par imitation, c'est à dire la résolution du problème du contrôle optimal à partir de données tirées de démonstrations d'un expert.
-L'apprentissage par renforcement inverse (IRL) propose un cadre efficace pour résoudre ce problème.
-En se basant sur l'hypothèse que l'expert maximise une fonction de valeur,  l'IRL essaie d'apprendre la récompense qui définit cette dernière à partir de trajectoires d'exemple.
-Beaucoup d'algorithmes d'IRL font l'hypothèse de l'existence d'un approximateur linéaire pour la fonction de récompense et calculent l'attribut moyen (le cumul moyen pondéré des fonctions de base, relatives à la paramétrisation linéaire supposée de la récompense, évaluées en les états d'une trajectoire associée à une certaine politique)  via une estimation de Monte-Carlo.
-Cela implique d'avoir accès à des trajectoires complètes de l'expert ainsi qu'à au moins un modèle génératif pour tester les politiques intermédiaires.
-Dans ce papier nous introduisons une méthode de différences temporelles, LSTD-$\mu$, pour calculer cet attribut moyen.
-Cela permet d'étendre l'apprentissage par imitation aux cas \emph{batch} et \emph{off-policy}.
+  This paper addresses the problem of apprenticeship learning, that
+  is learning control policies from demonstration by an expert. An
+  efficient framework for it is inverse reinforcement learning
+  (IRL). Based on the assumption that the expert maximizes a utility
+  function, IRL aims at learning the underlying reward from example 
+  trajectories. Many IRL algorithms
+  assume that the reward function is linearly parameterized and rely
+  on the computation of some associated \emph{feature expectations},
+  which is done through Monte Carlo simulation. However, this
+  assumes to have full trajectories for the expert policy as well as
+  at least a generative model for intermediate policies. In this
+  paper, we introduce a temporal difference method, namely LSTD-$\mu$, to compute these
+  feature expectations. This allows extending apprenticeship
+  learning to a batch and off-policy setting.
 \end{abstract}
@@
 %\bibliographystyle{named}
-%\bibliographystyle{named}
+\bibliographystyle{named}
 \bibliography{../Biblio/Biblio}

### 2011-06-21 main.tex
Version soumise à EWRL

@@
 \usepackage{natbib}
+\usepackage{graphicx}
 
@@
 
-\title{Apprentissage par imitation dans un cadre \emph{batch}, \emph{off-policy} et sans modèle}
+\title{Batch, Off-policy and Model-free Apprenticeship Learning}
 %\author{Robert}
@@
 A sequential decision problem is often framed as a Markov Decision
-Process (MDP)~\citep{puterman1994markov}, a tuple
+Process (MDP)~\citep{sutton1998reinforcement}, a tuple
 $\{S,A,P,R,\gamma\}$ with $S$ being the state space, $A$ the action
@@
 $V^\pi$, which associates to each state the expected and discounted
-cumulative reward:
-\begin{equation}
-  V^\pi(s) = E[\sum_{t=0}^\infty \gamma^t R(s_t)|s_0=s, \pi]
-\end{equation}
+cumulative reward: $V^\pi(s) = E[\sum\limits_{t=0}^\infty \gamma^t R(s_t)|s_0=s, \pi]$.
 Dynamic programming aims at finding the optimal policy $\pi^*$, that
@@
 p$. Therefore, there exists a parameter vector $\theta^*$ such that:
-\begin{equation}
-  R^*(s) = (\theta^*)^T \phi(s)
-\end{equation}
+ $R^*(s) = (\theta^*)^T \phi(s)$
 In order to ensure that rewards are bounded, we impose that
@@
 function $V^\pi(s)$ can be expressed as follows:
-\begin{align}
-  V^\pi(s) &= E[\sum_{t=0}^\infty \gamma^t \theta^T \phi(s_t)|s_0=s, \pi]
-  \\
-  &= \theta^T  E[\sum_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
-\end{align}
+\begin{equation}
+\scriptstyle
+V^\pi(s) = E[\sum\limits_{t=0}^\infty \gamma^t \theta^T \phi(s_t)|s_0=s, \pi] = \theta^T  E[\sum\limits_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
+\end{equation}
 Therefore, the value function is also linearly parameterized, with
@@
 of $\mathcal{H}_\phi$:
-\begin{align}
-  |V^{\pi_E}(s_0) - V^{\tilde{\pi}}(s_0)| &=
+\begin{equation}
+\scriptstyle
+  |V^{\pi_E}(s_0) - V^{\tilde{\pi}}(s_0)| =
   |\theta^T(\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0))|
+  \leq \|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2
 \label{eqn:vs0}
-  \\
-  &\leq \|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2
-\end{align}
+\end{equation}
 This equation uses the Cauchy-Schwarz inequality and the
@@
 estimated using a Monte Carlo approach over $m$ trajectories:
-\begin{equation}
-  \hat{\mu}_E(s_0) = \frac{1}{m} \sum_{h=1}^m \sum_{t=0}^\infty
-  \gamma^t \phi(s_t^{(h)})
-\end{equation}
+$  \hat{\mu}_E(s_0) = \frac{1}{m} \sum\limits_{h=1}^m \sum\limits_{t=0}^\infty
+  \gamma^t \phi(s_t^{(h)})$
 This approach does not hold if only transitions of the expert are
@@
 feature expectation $\mu^\pi(s)$ for some policy $\pi$:
-\begin{equation}
-  \mu_i^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi_i(s_t)|s_0=s,\pi]
-  \label{eqn:phi}
-\end{equation}
+$  \mu_i^\pi(s) = E[\sum\limits_{t=0}^\infty \gamma^t \phi_i(s_t)|s_0=s,\pi]$
 This is exactly the definition of the value function of the policy
@@
 an approximation of the following form:
-\begin{equation}
-  \hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)
-  \label{eqn:psi}
-\end{equation}
+$  \hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$
 The parameter vector $\xi_i^*$ is here the LSTD estimate:
@@
 $Q$-function extends the value function):
-\begin{equation}
-  \mu^\pi(s,a) = E[\sum_{t=0}^\infty \gamma^t
-  \phi(s_t)|s_0=s,a_0=a,\pi]
-\end{equation}
+$\mu^\pi(s,a) = E[\sum\limits_{t=0}^\infty \gamma^t
+  \phi(s_t)|s_0=s,a_0=a,\pi]$
 Compared to the classic feature expectation, this definition adds a
@@
 \begin{figure}
-\centering
+\begin{minipage}[t]{.4\linewidth}
+    \begin{center}
       \resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold}}
@@
       \label{fig:threshold}
-\end{figure}
-\begin{figure}
-\centering
+    \end{center}
+\end{minipage}
+\hfill
+\begin{minipage}[t]{.4\linewidth}
+    \begin{center}
       \resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold_EB}}
@@
       \label{fig:threshold_EB}
+    \end{center}
+\end{minipage}
 \end{figure}
+
+%% \begin{multicols}{2}
+%% \begin{figure}
+%% \centering
+%% \end{figure}
+%% \begin{figure}
+%% \centering
+%% \end{figure}
+%% \end{multicols}
 Another classic toy problem is the \emph{inverted pendulum}, used for example in \citep{lagoudakis2003least} from where we drew the exact configuration (mass of the pendulum, length of the arm, definition of the feature space, etc.).
@@
 \begin{figure}
-\centering
-\input{../Code/GridWorld/criteria_mc}
+\begin{minipage}[t]{.4\linewidth}
+    \begin{center}
+      \resizebox{\columnwidth}{!}{\input{../Code/GridWorld/criteria_mc}}
       \caption{Different criteria with respect to the number of iterations for a run of the MC variant.}
       \label{fig:A}
-\end{figure}
-\begin{figure}
-\centering
-\input{../Code/GridWorld/criteria_lstd_EB}
+    \end{center}
+\end{minipage}
+\hfill
+\begin{minipage}[t]{.4\linewidth}
+    \begin{center}
+\resizebox{\columnwidth}{!}{\input{../Code/GridWorld/criteria_lstd_EB}}
 \caption{Different criteria with respect to the number of samples from the expert, for several runs of the LSTD variant.
@@
 \label{fig:B}
+    \end{center}
+\end{minipage}
 \end{figure}

### 2011-06-22 Tangle.sh
Test pass for the TaskTransfer

@@ -1,0 +1,4 @@
+cat *.org > All.org
+emacs -batch --visit All.org --funcall org-babel-tangle --script ~/.emacs
+rm All.org
+

### 2011-09-02 main.tex
Initial commit

--- 2011-06-16 main.tex
@@
-\documentclass{llncs}
-%
-\usepackage{makeidx}  % allows for indexgeneration
+\documentclass[xcolor=x11names,12pt]{beamer}
+%\usetheme[secheader]{Boadilla} 
+\mode<presentation>{\usetheme{I6pd2}}
 
-\usepackage{amssymb}
-\usepackage{natbib}
+\usepackage{multicol} 
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+\usepackage{grffile}
+\usepackage{listings}
+\usepackage{multirow}
+\usepackage{amsmath}
+\usepackage{pifont}
+\usepackage{slashbox}
+\include{headertikz}
+\usetikzlibrary{through,shapes,arrows,decorations.pathmorphing,backgrounds,positioning,fit}
 
-% Typographie française ==================================================================
-\usepackage[english]{babel} % typographie francaise
-\usepackage[utf8]{inputenc} % saisie de caracteres accentues, etc.
+\newcommand{\argmin}{\mathop{\mathrm{min}}}
+\usepackage[orientation=portrait,size=a0,scale=1.4]{beamerposter}
 
-% Vos environnements =====================================================================
-\newtheorem{prop}{Proposition}
-\newenvironment{pf}{{\bf Proof }}{\hfill$\Box$\par}
-\newcommand{\spaceafterproof}{\vspace{1em}}
+\newcommand{\conf}[1]{\newcommand{\insertconf}{#1}}
+\newenvironment{WholeWidthBox}[1]{
+  \begin{columns}
+    \begin{column}{0.98\textwidth}
+      \begin{block}{#1}
+        \begin{hfill}
+}{
+        \end{hfill}
+      \end{block}
+    \end{column}
+  \end{columns}
+}
 
-% Corps de l'article =====================================================================
-\usepackage{color}
-% the following package is optional:
-%\usepackage{latexsym} 
-%\usepackage{amsthm}
-%% \newtheorem{theorem}{Theorem}
-%% \newtheorem{lemma}{Lemma}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{subfigure}
-%\usepackage{algorithm2e}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
+\newcommand{\TwoBoxes}[7]{%Width1 Title1 Content1 Width2 Title2 Content2 MinHeight
+  \begin{columns}
+    \begin{column}{#1\textwidth}
+      \begin{block}{#2}
+        \begin{columns}
+          \begin{column}{0.001\textwidth}
+            \vspace{#7}
+          \end{column}
+          \begin{column}{\textwidth}
+            \centering
+            #3
+          \end{column}
+        \end{columns}       
+      \end{block}
+    \end{column}
 
-\title{Apprentissage par imitation dans un cadre \emph{batch}, \emph{off-policy} et sans modèle}
-%\author{Robert}
-%   1. Supelec-Metz Campus, IMS Reasearch group, France. \texttt{prenom.nom@supelec.fr}}
-\author{Edouard Klein\inst{1}\inst{3} \and Matthieu Geist\inst{1} \and Olivier Pietquin\inst{1}\inst{2}}
- \institute{
- 1. Supélec-Metz Campus, IMS Research group, France, \texttt{prenom.nom@supelec.fr}\\
- 2. UMI 2958 CNRS - GeorgiaTech, France\\
- 3. Equipe ABC, LORIA-CNRS, France.
+    \begin{column}{#4\textwidth}
+      \begin{block}{#5}
+        \begin{columns}
+          \begin{column}{0.001\textwidth}
+            \vspace{#7}
+          \end{column}
+          \begin{column}{\textwidth}
+            \centering
+            #6
+          \end{column}
+        \end{columns}       
+      \end{block}
+    \end{column}
+  \end{columns}
 }
+
+%Start of document
+\conf{Pinky}
+\title{Blinky}
+\author{Inky}
 \date{\today}
+\institute[C.]{Clyde}
+\newlength{\columnheight}
+\setlength{\columnheight}{105cm}
 
 \begin{document}
+\begin{frame}
 
-\maketitle
-\begin{abstract}
-  This paper addresses the problem of apprenticeship learning, that
-  is learning control policies from demonstration by an expert. An
-  efficient framework for it is inverse reinforcement learning
-  (IRL). Based on the assumption that the expert maximizes a utility
-  function, IRL aims at learning the underlying reward from example 
-  trajectories. Many IRL algorithms
-  assume that the reward function is linearly parameterized and rely
-  on the computation of some associated \emph{feature expectations},
-  which is done through Monte Carlo simulation. However, this
-  assumes to have full trajectories for the expert policy as well as
-  at least a generative model for intermediate policies. In this
-  paper, we introduce a temporal difference method, namely LSTD-$\mu$, to compute these
-  feature expectations. This allows extending apprenticeship
-  learning to a batch and off-policy setting.
-\end{abstract}
-\section{Introduction}
+%%%%%%%%%%%%% DEBUT PREMIERE LIGNE %%%%%%%%%%%%%%%%%
+\begin{WholeWidthBox}{Gntroduction}
+  Ceci est mon intro
+\end{WholeWidthBox}
 
-Optimal control consists in putting a machine in control of a system
-with the goal of fulfilling a specific task, optimality being
-defined as how well the task is performed. Various solutions to this
-problem exist from automation to planification. Notably, the
-reinforcement learning (RL) paradigm~\citep{sutton1998reinforcement} is a general
-machine learning framework in which an agent learns to control
-optimally a dynamic system through interactions with it. The task is
-specified through a reward function, the agent objective being to
-take sequential decisions so as to maximize the expected cumulative
-reward.
+\vfill
+%%%%%%%%%%%% DEBUT DEUXIEME LIGNE %%%%%%%%%%%%%%%%%%
 
-However, defining optimality (through the reward function) can
-itself be a challenge. If the system can be empirically controlled
-by an expert, even though his/her behavior can be difficult to
-describe formally, apprenticeship learning is a way to have a
-machine controlling the system by mimicking the expert. Rather than
-directly mimicking the expert with some supervised learning
-approach, Inverse Reinforcement Learning (IRL)~\citep{ng2000algorithms}
-consists in learning a reward function under which the policy
-demonstrated by the expert is optimal. Mimicking the expert therefore
-ends up to learning an optimal policy according to this reward
-function. A significant advantage of such an approach is that
-expert's actions can be guessed in states which have not been
-encountered during demonstration. Firstly introduced in \citep{russell1998learning}, another advantage claimed by the author would be to
-find a compact and complete representation of the task in the form
-of the reward function.
+\TwoBoxes{.48}{Titre1}{Blah}{.48}{Titre2}{Bluh}{15cm}
+\vfill
 
-There roughly exists three families of IRL algorithms:
-feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
-margin-maximization-based~\citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
-and approaches based on the parameterization of the policy by the
-reward function~\citep{ramachandran2007bayesian,neu2007apprenticeship}. The first family
-assumes a linearly parameterized reward function. This naturally
-leads to a linearly parameterized value function, the associated
-feature vector being the so-called feature expectation (see Section~\ref{sec:back} for a formal definition).
-These approaches learn a reward function such that the feature
-expectation of the optimal policy (according to the learnt reward
-function) is close to the feature expectation of the expert policy.
-This is a sufficient condition to have close value functions (for any parameterized reward function, and therefore particularly the optimized one). The
-second family expresses IRL as a constrained optimization problem in
-which expert's examples have higher expected cumulative reward than all other
-policies by a certain margin. Moreover, suboptimality of the expert
-can be considered through the introduction of slack variables. The
-last family parameterizes policies with a reward function. Assuming
-that the expert acts according to a Gibbs policy (respectively to
-the optimal value function related to the reward function which is
-optimized), it is possible to estimate the likelihood of a set of
-state-action pairs provided by the expert. The algorithms differ in
-the way this likelihood is maximized.
+\small{References}\\
+\vspace{-1cm}
+\centering
+\begin{multicols}{2}
+  \bibliographystyle{unsrt}
+  \begin{tiny}
+    \bibliography{biblio}
+  \end{tiny}
+\end{multicols}
+\vspace{-0.2cm}
 
-This paper focuses on the first family of algorithms, and more
-precisely on the seminal work of~\citet{abbeel2004apprenticeship}. All of them
-rely on the computation of the feature expectation (which depends on
-policies but not on rewards) of (i) the expert and (ii) some
-intermediate policies. The expert's feature expectation is computed
-using a simple Monte Carlo approach (which requires full
-trajectories of the expert). Other feature expectations are either
-computed exactly (which requires knowing analytically the dynamics
-of the system) or with a Monte Carlo approach (which requires simulating the system). The contribution of this paper is LSTD-$\mu$,
-a new temporal-difference-based algorithm for estimating these feature
-expectations. It relaxes the preceding assumptions: transitions
-of the expert are sufficient (rather than full trajectories) and nor
-the model neither a simulator are necessary to compute intermediate
-feature expectations. This paper focuses on the algorithm introduced
-in~\citep{abbeel2004apprenticeship}, but the proposed
-approach can be used in other algorithms based on feature
-expectation computation.
-
-The rest of this paper is organized as follows.
-Section~\ref{sec:back} provides the necessary background,
-notably the definition of feature expectation and its use in the
-seminal IRL algorithm of~\citet{abbeel2004apprenticeship}.
-Section~\ref{sec:lstdmu} presents LSTD-$\mu$, our main contribution.
-Section~\ref{sec:exp} provides some preliminary experiments and
-Section~\ref{sec:conclusion} opens perspectives.
-
-\section{Background}
-\label{sec:back}
-A sequential decision problem is often framed as a Markov Decision
-Process (MDP)~\citep{puterman1994markov}, a tuple
-$\{S,A,P,R,\gamma\}$ with $S$ being the state space, $A$ the action
-space, $P\in\mathcal{P}(S)^{S\times A}$ the set of Markovian
-transition probabilities, $R\in\mathbb{R}^S$ the reward function
-(assumed to be absolutely bounded by 1) and $\gamma\in[0,1[$ a
-discounting factor. A policy $\pi\in A^S$ maps states to actions. The
-quality of a policy is quantified by the associated value function
-$V^\pi$, which associates to each state the expected and discounted
-cumulative reward:
-\begin{equation}
-  V^\pi(s) = E[\sum_{t=0}^\infty \gamma^t R(s_t)|s_0=s, \pi]
-\end{equation}
-Dynamic programming aims at finding the optimal policy $\pi^*$, that
-is one of the policies associated with the optimal value function,
-$V^* = \argmax_\pi V^\pi$, which maximizes the value for each state.
-If the model (that is transition probabilities and the reward
-function) is unknown, learning the optimal control through
-interactions is addressed by RL.
-
-For IRL, the problem is reversed. It is
-assumed that an expert acts according to an optimal policy $\pi_E$,
-this policy being optimal according to some unknown reward function
-$R^*$. The goal is to learn this
-reward function from sampled trajectories of the expert. This is a
-difficult and ill-posed problem~\citep{ng2000algorithms}. Apprenticeship
-learning through IRL, which is the problem at hand in this paper,
-has a somewhat weaker objective: it aims at learning a policy
-$\tilde{\pi}$ which is (approximately) as good as the expert policy
-$\pi_E$ under the unknown reward function $R^*$, for a known initial
-state $s_0$ (this condition can be easily weakened by assuming a
-distribution over initial states). Now, the approach proposed in~\citep{abbeel2004apprenticeship} is
-presented.
-
-We assume that the true reward function belongs to some hypothesis
-space $\mathcal{H}_\phi = \{\theta^T \phi(s),
-\theta\in\mathbb{R}^p\}$, of which we assume the basis functions to
-be bounded by 1: $|\phi_i(s)|\leq 1, \forall s\in S, 1\leq i \leq
-p$. Therefore, there exists a parameter vector $\theta^*$ such that:
-\begin{equation}
-  R^*(s) = (\theta^*)^T \phi(s)
-\end{equation}
-In order to ensure that rewards are bounded, we impose that
-$\|\theta\|_2\leq 1$. For any reward function belonging to
-$\mathcal{H}_\phi$ and for any policy $\pi$, the related value
-function $V^\pi(s)$ can be expressed as follows:
-\begin{align}
-  V^\pi(s) &= E[\sum_{t=0}^\infty \gamma^t \theta^T \phi(s_t)|s_0=s, \pi]
-  \\
-  &= \theta^T  E[\sum_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
-\end{align}
-Therefore, the value function is also linearly parameterized, with
-the same weights and with basis functions being grouped into the
-so-called \emph{feature expectation} $\mu^\pi$:
-\begin{equation}
-  \mu^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
-\end{equation}
-Recall that the problem is to find a policy whose performance is
-close to that of the expert's for the starting state $s_0$, on the
-unknown reward function $R^*$. In order to achieve this goal, it is
-proposed in~\citep{abbeel2004apprenticeship} to find a policy $\tilde{\pi}$ such
-that $\|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2\leq \epsilon$
-for some (small) $\epsilon>0$. Actually, this ensures that the value
-of the expert's policy and the value of the estimated policy (for
-the starting state $s_0$) are close for \emph{any} reward function
-of $\mathcal{H}_\phi$:
-\begin{align}
-  |V^{\pi_E}(s_0) - V^{\tilde{\pi}}(s_0)| &=
-  |\theta^T(\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0))|
-  \label{eqn:vs0}
-  \\
-  &\leq \|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2
-\end{align}
-This equation uses the Cauchy-Schwarz inequality and the
-assumption that $\|\theta\|_2\leq 1$. Therefore, the approach
-described here does not ensure to retrieve the true reward function
-$R^*$, but to act as well as the expert under this reward function
-(and actually under any reward function).
-
-Let us now describe the algorithm proposed in~\citep{abbeel2004apprenticeship} to
-achieve this goal:
-\begin{enumerate}
-  \item Starts with some initial policy $\pi^{(0)}$ and compute
-  $\mu^{\pi^{(0)}}(s_0)$. Set $j=1$;
-  %
-  \item Compute $t^{(j)} = \max_{\theta: \|\theta\|_2\leq 1}
-  \min_{k\in\{0,j-1\}}\theta^T(\mu^{\pi_E}(s_0) -
-  \mu^{\pi^{(k)}}(s_0))$ and let $\theta^{(j)}$ be the value
-  attaining this maximum. At this step, one searches for the reward
-  function which maximizes the distance between the value of the
-  expert at $s_0$ and the value of \emph{any} policy computed so far
-  (still at $s_0$). This optimization problem can be solved using a
-  quadratic programming approach or a projection
-  algorithm~\citep{abbeel2004apprenticeship};
-  %
-  \item if $t^{(j)}\leq \epsilon$, terminate. The algorithm outputs a
-  set of policies $\{\pi^{(0)}, \dots, \pi^{(j-1)}\}$ among which
-  the user chooses manually or automatically the closest to the
-  expert (see~\citep{abbeel2004apprenticeship} for details on how to choose this
-  policy).
-  Notice that the last policy is not necessarily the best (as illustrated in Section~\ref{sec:exp});
-  %
-  \item solve the MDP with the reward function $R^{(j)}(s) =
-  (\theta^{(j)})^T\phi(s)$ and denote $\pi^{(j)}$ the associated
-  optimal policy. Compute $\mu^{\pi^{(j)}}(s_0)$;
-  %
-  \item set $j\leftarrow j+1$ and go back to step 2.
-\end{enumerate}
-There remain three problems: computing the feature expectation of
-the expert, solving the MDP and computing feature expectations of
-intermediate policies.
-
-As suggested in~\citep{abbeel2004apprenticeship}, solving the MDP can be done
-approximately by using any appropriate reinforcement learning
-algorithm. In this paper, we use the Least-Squares Policy Iteration
-(LSPI) algorithm~\citep{lagoudakis2003least}. There remains to estimate
-feature expectations. In~\citep{abbeel2004apprenticeship}, $\mu^{\pi_E}(s_0)$ is
-estimated using a Monte Carlo approach over $m$ trajectories:
-\begin{equation}
-  \hat{\mu}_E(s_0) = \frac{1}{m} \sum_{h=1}^m \sum_{t=0}^\infty
-  \gamma^t \phi(s_t^{(h)})
-\end{equation}
-This approach does not hold if only transitions of the expert are
-available, or if trajectories are too long (in this case, it is
-still possible to truncate them). For intermediate policies, it is
-also suggested to estimate associated feature expectations using a
-Monte Carlo approach (if they cannot be computed exactly). This is
-more constraining than for the expert, as this assumes that a
-simulator of the system is available. In order to address these
-problems, we introduce a temporal-difference-based algorithm to
-estimate these feature expectations.
-
-\section{LSTD-$\mu$}
-\label{sec:lstdmu}
-Let us write the definition of the $i^\text{th}$ component of a
-feature expectation $\mu^\pi(s)$ for some policy $\pi$:
-\begin{equation}
-  \mu_i^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi_i(s_t)|s_0=s,\pi]
-  \label{eqn:phi}
-\end{equation}
-This is exactly the definition of the value function of the policy
-$\pi$ for the MDP considered with the $i^\text{th}$ basis function
-$\phi_i(s)$ as the reward function. There exist many algorithms to
-estimate a value function, any of them can be used to estimate
-$\mu_i^\pi$. Based on this remark, we propose to use specifically
-the least-squares temporal difference (LSTD)
-algorithm~\citep{bradtke1996linear} to estimate each component of the
-feature expectation (as each of these components can be understood
-as a value function related to a specific and known reward
-function).
-
-Assume that a set of transitions $\{(s_t,r_t,s_{t+1})_{1\leq t \leq
-n}\}$ sampled according to the policy $\pi$ is available. We assume
-that value functions are searched for in some  hypothesis space
-$\mathcal{H}_\psi = \{ \hat{V}_\xi(s) = \sum\limits_{i=1}^q \xi_i \psi_i(s)
-= \xi^T \psi(s), \xi\in\mathbb{R}^q\}$. As reward and value
-functions are possibly quite different, another hypothesis space is
-considered for value function estimation. But if $\mathcal{H}_\phi$
-is rich enough, one can still consider
-$\mathcal{H}_\psi=\mathcal{H}_\phi$. Therefore, we are looking for
-an approximation of the following form:
-\begin{equation}
-  \hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)
-  \label{eqn:psi}
-\end{equation}
-The parameter vector $\xi_i^*$ is here the LSTD estimate:
-\begin{equation}
-  \xi_i^* = \left(\sum_{t=1}^n
-  \psi(s_t)(\psi(s_t)-\gamma\psi(s'_{t}))^T\right)^{-1}
-  \sum_{t=1}^n \psi(s_t) \phi_i(s_t)
-\end{equation}
-For apprenticeship learning, we are interested more particularly in
-$\hat{\mu}^\pi(s_0)$. Let $\Psi = (\psi_i(s_t))_{t,i}$ be the
-$n\times q$ matrix of values predictors, $\Delta\Psi = (\psi_i(s_t)
-- \gamma\psi_i(s'_t))_{t,i}$ be the related $n\times q$ matrix and
-$\Phi = (\phi_i(s_t))_{t,i}$ the $n\times p$ matrix of reward
-predictors. It can be easily checked that $\hat{\mu}^\pi(s_0)$
-satisfies:
-\begin{equation}
-  (\hat{\mu}^\pi(s_0))^T = \psi(s_0)^T (\Psi^T \Delta\Psi)^{-1} \Psi^T \Phi
-\end{equation}
-This provides an efficient way to estimate the feature expectation
-of the expert in $s_0$.
-
-There remains to compute the feature expectations of intermediate
-policies, which should be done in an off-policy manner (that is
-without explicitly sampling trajectories according to the policy of
-interest). To do so, still interpreting each component of the
-feature expectation as a value function, we introduce a state-action
-feature expectation defined as follows (much as the classic
-$Q$-function extends the value function):
-\begin{equation}
-  \mu^\pi(s,a) = E[\sum_{t=0}^\infty \gamma^t
-  \phi(s_t)|s_0=s,a_0=a,\pi]
-\end{equation}
-Compared to the classic feature expectation, this definition adds a
-degree of freedom on the first action to be chosen before following
-the policy $\pi$. With a slightly different definition of the
-related hypothesis space, each component of this feature expectation
-can still be estimated using the LSTD algorithm (namely using the
-LSTD-Q algorithm~\citep{lagoudakis2003least}). The clear advantage of
-introducing the state-action feature expectation is that this
-additional degree of freedom allows off-policy learning.
-Extending LSTD-$\mu$ to state-action LSTD-$\mu$ is done in the same
-manner as LSTD is extended to LSTD-Q~\citep{lagoudakis2003least},
-technical details are not provided here for the clarity of
-exposition.
-
-Given the (state-action) LSTD-$\mu$ algorithm, the apprenticeship
-learning algorithm presented in Section~\ref{sec:back} can be easily extended to a batch and
-off-policy setting. The solely available data is a set of
-transitions sampled according to the expert policy  (and possibly a set of sub-optimal trajectories). The
-corresponding feature expectation for the starting state $s_0$ is
-estimated with the LSTD-$\mu$ algorithm. At step~4 of this
-algorithm, the MDP is (approximately) solved using
-LSPI~\citep{lagoudakis2003least} (an approximate policy iteration
-algorithm using LSTD-Q as the off-policy $Q$-function estimator).
-The corresponding feature expectation at state $s_0$ is estimated
-using the proposed state-action LSTD-$\mu$.
-
-Before presenting some experimental results, let us stress that
-LSTD-$\mu$ is simply the LSTD algorithm applied to a specific reward
-function. Although quite clear, the idea of using a temporal
-difference algorithm to estimate the feature expectation is new, as
-far as we know. A clear advantage of the proposed approach is that
-any theoretical result holding for LSTD also holds for LSTD-$\mu$,
-such as convergence~\citep{nedic2003least} or
-finite-sample~\citep{lazaric2010finiteLSTD} analysis for example.
-\section{Experimental benchmark}
-\label{sec:exp}
-This section provides experimental results on two complementary problems.
-Subsection~\ref{ssec:exp} details the protocol and the results while subsection~\ref{ssec:quality} inspects the meaning of the different quality criteria.
-\subsection{Experiment description and results}
-\label{ssec:exp}
-\subsubsection{GridWorld}
-\begin{figure}
-\centering
-\resizebox{\columnwidth}{!}{\input{../Code/GridWorld/both_error_EB}}
-\caption{$||\mu^{\pi}(s_0) - \mu^{\pi_E}(s_0)||_2$ with respect to the number of samples available from the expert. The error bars represent the standard deviation over 100 runs.}
-\label{fig:E}
-\end{figure}
-The first experimental benchmark chosen here is one of those proposed in \citep{ng2000algorithms}, a 5x5 grid world.
-The agent is in one of the cells of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action).
-With probability 0.9, the agent moves in the intended direction.
-With probability 0.1, the direction is randomly drawn.
-The reward optimized by the expert is 0 everywhere except in the upper-right cell, where it is 1.
-For every policy, an episode ends when the upper right cell is attained, or after 20 steps.
-At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is).
-Both the state and action spaces are finite and of small cardinality.
-Hence, the chosen feature functions $\phi$ and $\psi$ are the typical features of a tabular representation: 0 everywhere except for the component corresponding to the current state (-action pair).\\
-
-Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as the MC variant) and our adaptation (referred to as the LSTD variant) are tested side by side.
-The MDP solver of the MC variant is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly.
-Both $\mu^{\pi^{(j)}}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to make the variance negligible.
-We consider both these computations as perfect theoretical solvers for all intended purpose on this toy problem.
-We thus are in the case intended by \citet{abbeel2004apprenticeship}.
-On the other hand the LSTD variant is used without accessing a simulator.
-It uses LSPI and LSTD-$\mu$, fed only with the expert's transitions (although we could also use non expert transitions  to compute intermediate policies, if available).
-This corresponds to a real-life setting where data generation is expensive and the system cannot be controlled by an untrained machine.\\
-
-We want to compare the efficiency of both versions of the algorithm with respect to the number of samples available from the expert, as these samples usually are the bottleneck.
-Indeed as they are quite costly (in both means and human time) to generate they are often not in abundance hence the critical need for an algorithm to be expert-sample efficient.
-Having a simulator at one's disposal can also be difficult.
-For the simple problem we use this is however not an issue.
-The discussion about the choice of the performance metric has its own dedicated subsection (subsection~\ref{ssec:quality}).
-We use here the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term.
-Fig.~\ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^{(j)}}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40).
-The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons why it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations.
-The best policy is not always the last, and although it has always been observed so far with the LSTD variant, there is absolutely no way to tell whether this is a feature.
-The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is plotted here.
-We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithms quickly converge to almost the same value ; our version converges to a slightly lower error value.
-The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of tasks inaccessible to the MC variant, the requirement of a simulator being often constraining.
-\subsubsection{Inverted Pendulum}
-\begin{figure}
-\centering
-\resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold}}
-\caption{Number of balancing steps for the policies found by both variants with respect to the number of samples from a sub-optimal policy given to LSPI.
-Note the absence of middle ground in the quality.}
-\label{fig:threshold}
-\end{figure}
-\begin{figure}
-\centering
-\resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold_EB}}
-\caption{Same plot as Fig.~\ref{fig:threshold}, with mean and standard deviation over 100 runs.
-The big standard deviation stems from the absence of middle ground in the quality of the policies.
-The increasing mean w.r.t.
-the abscissa means that the proportion of good policies increases with the number of samples from a sub-optimal policy given to LSPI.}
-\label{fig:threshold_EB}
-\end{figure}
-Another classic toy problem is the \emph{inverted pendulum}, used for example in \citep{lagoudakis2003least} from where we drew the exact configuration (mass of the pendulum, length of the arm, definition of the feature space, etc.).
-In this setting, the machine must imitate an expert who maintains the unstable equilibrium of an inverted pendulum allowed to move along only one axis.
-A balancing step is considered successful (reward $0$) when the angle is less than $\pi\over 2$.
-It is considered a failure (reward $-1$) when the pendulum has fallen (i.e.
-the angle exceeds $\pi \over 2$).
-A run is stopped after a failure or after 3000 successful balancing steps.\\
-
-This problem bears two fundamuntal differences with the previous one and thus comes as complementary to it.
-First, it presents a continuous state space whereas in the GridWorld a tabular representation was possible.
-Then, the initial state is randomly drawn from a non singleton subspace of the state space.
-This last point is addressed differently by the LSTD variant and the MC variant.
-the MC variant will naturally sample the initial state distribution at each new trajectory from the expert.
-The LSTD variant, on the other hand, still does not need complete trajectories from the expert but mere samples.
-As it approximates the whole $\mu^{\pi}$ function and not just $\mu^\pi(s_0)$, it is possible to compute the mean of the approximation $\hat\mu^\pi$ over a few states drawn with the same distribution as the initial states.\\
-
-A striking finding on this toy problem is that the parameter controlling the success or the failure of the experiment is not the number of samples from the expert, but the quality of the samples available to LSPI to solve the MDP.
-\emph{A contrario} to what happened in the previous toy problem, there are fewer samples from the expert than what LSPI needs to solve the MDP as they do not cover the whole state space.
-When given only these samples, the LSTD variant fails.
-The MC variant, however, is not bound by such restrictions and can use as many samples as it needs from a simulator.
-Thus, with only one trajectory from the expert (100 samples) the MC variant is able to successfully balance the pendulum for more than 3000 steps.
-The problem here stems from the fact that we don't learn the reward, but a policy that maximizes the reward.
-If the optimal control problem was solved separately, learning the reward only from the samples of the expert would be possible.\\
-
-The sensitivity of both algorithms to the number of samples available to LSPI is extreme, as is shown in Fig.~\ref{fig:threshold}.
-It may seem nonsensical to restrict the number of samples available to LSPI for the MC variant as it can access a simulator, but this has been done to show that both variants exhibit the same behavior, hence allowing us to locate the source of this behavior in what both algorithms have in common (the use of LSPI inside the structure of \citep{abbeel2004apprenticeship}) excluding the point where they differ (LSTD-$\mu$ and Monte-Carlo).\\
-
-Fig.~\ref{fig:threshold_EB} shows that when given samples from a sub-optimal policy, the LSTD variant can sort the problem out, statistically almost always with 3000 sub-optimal samples, and sometimes with as low as 200 sub-optimal samples.
-Such a setting is still batch, off-policy and model-free.
-When given enough sub-optimal samples, both variants are successful with only one trajectory from the expert (i.e.
-100 samples, this is what is plotted here).
-Giving more trajectories from the expert does not improve the success rate.
-\subsection{Discussion about the quality criterion}
-\label{ssec:quality}
-\begin{figure}
-\centering
-\input{../Code/GridWorld/criteria_mc}
-\caption{Different criteria with respect to the number of iterations for a run of the MC variant.}
-\label{fig:A}
-\end{figure}
-\begin{figure}
-\centering
-\input{../Code/GridWorld/criteria_lstd_EB}
-\caption{Different criteria with respect to the number of samples from the expert, for several runs of the LSTD variant.
-We can see that the algorithm is blind, as all it has access to is always zero.
-The true error values, however, smoothly converge to something small.
-Knowing how many expert samples are actually needed in a real world problem is an open question.
-The error bars represent the standard deviation over 100 runs.}
-\label{fig:B}
-\end{figure}
-
-Fig.~\ref{fig:A}~and~\ref{fig:B} illustrate (using the GridWorld problem) the difficulty of choosing the quality criterion ; Fig.~\ref{fig:A} shows four different quality criteria during a run of the MC variant.
-Fig.~\ref{fig:B} shows the same criteria for several runs of the LSTD variant with a varying number of samples from the expert.
-The $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material.
-It bears an interesting relation with the difference between the expert's value function and the current value function in the initial state with respect \emph{to the current reward} (Eq.~\ref{eqn:vs0}).\\
-
-The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in Fig.~\ref{fig:A} means that, for it has access to a cheap simulator, the MC variant works as if it had access to the exact values.
-This however cannot be said of the LSTD variant, for which the two curves are indeed different (Fig.~\ref{fig:B}).
-Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for our variant, as it can introduce an error in the stopping criterion of the algorithm.\\
-
-It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in the MC variant either, as it can be low (and thus halt the algorithm) when the policy is bad.
-The best policy, however, can be easily chosen among all those created during the execution of the algorithm thanks to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which the MC variant can compute.
-When this term is low, the objective performance (that is, $V^{\pi_E}(s_0)-V^{\pi^{(j)}}(s_0)$ with respect to the unknown true reward function) is low too.
-\section{Conclusion}
-\label{sec:conclusion}
-Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning via inverse reinforcement learning to a batch, model-free, off-policy setting.
-Experimentally, there is a need for either a slightly bigger number of samples from the expert or some samples from a sub-optimal policy.
-We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach.
-Furthermore, tranferring the reward and not the policy may overcome this difficulty.
-We intend to do this in a real life setting.
-The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\
-
-%\bibliographystyle{named}
-\bibliographystyle{named}
-\bibliography{../Biblio/Biblio}
+\end{frame}
 \end{document}

### 2011-09-02 main.tex
Ajout d'un \cite dans le main pour pouvoir compiler
First version of the poster

@@
 %Start of document
-\conf{Pinky}
-\title{Blinky}
-\author{Inky}
+
+\tikzstyle{state}=[circle,
+thick,
+minimum size=1.0cm,
+draw=blue!80,
+fill=blue!20]
+\tikzstyle{action}=[rectangle,thick,
+minimum size=1.0cm,
+draw=orange!80,
+fill=orange!20]
+\tikzstyle{element}=[rectangle,
+line width=1.5mm,
+minimum size=1.0cm,
+draw=blue!80,
+fill=blue!20]
+\tikzstyle{action}=[rectangle,
+line width=1.5mm,
+minimum size=1.0cm,
+draw=orange!80,
+fill=orange!20]
+
+\conf{European workshop on reinforcement learning}
+\title{Batch, Off-policy and Model-free Apprenticeship Learning}
+\author{\underline{Edouard Klein}$^{\dag\ddag}$, Matthieu Geist$^\dag$ and Olivier Pietquin$^\dag$\\\texttt{firstname.lastname@supelec.fr}}
 \date{\today}
-\institute[C.]{Clyde}
+\institute[Supélec]{$\dag$Supélec UMI 2958 (GeorgiaTech - CNRS), France\\$\ddag$Equipe ABC UMR 7503 (LORIA-CNRS), France}
 \newlength{\columnheight}
@@
 %%%%%%%%%%%%% DEBUT PREMIERE LIGNE %%%%%%%%%%%%%%%%%
-\begin{WholeWidthBox}{Gntroduction}
-  Ceci est mon intro
+\begin{WholeWidthBox}{Introduction}
+  Apprenticeship learning can be defined as learning control policies from demonstration by an expert. An efficient framework for it is inverse reinforcement learning
+  (IRL).
+Based on the assumption that the expert maximizes a utility function, IRL aims at learning the underlying reward from example trajectories.
+Many IRL algorithms rely on the computation of \emph{feature expectations}, which is done through Monte Carlo simulation.
+In this paper, we introduce a temporal difference method, namely LSTD-$\mu$, to compute these feature expectations.
+This allows extending apprenticeship learning to a batch and off-policy setting.
 \end{WholeWidthBox}
@@
 
-\TwoBoxes{.48}{Titre1}{Blah}{.48}{Titre2}{Bluh}{15cm}
+\TwoBoxes{.48}{Setting}{
+  \begin{columns}
+    \begin{column}{.48\textwidth}
+      \includegraphics{Agent003.png}
+    \end{column}
+    \begin{column}{.48\textwidth}
+      \begin{itemize}
+      \item Expert's trace
+      \item MDP : 
+        \begin{itemize}
+        \item State space $S$
+        \item Action space $A$
+        \item Unknown reward being maximized by the expert $R$
+        \item Discount factor $\gamma$
+        \item Transition probabilities $P$
+        \end{itemize}
+      \item Agent implementing a policy $\pi : S\rightarrow A$
+      \end{itemize}
+
+    \end{column}
+  \end{columns}
+}
+{.48}{Assumptions and Goal}{
+  \begin{itemize}
+  \item Assumptions
+    \begin{itemize}
+    \item The expert is a RL agent
+    \item One can access the expert's trace
+    \item The agent has the same abilities as the expert
+    \end{itemize}
+  \item Goal : Infer the expert's reward
+    \begin{itemize}
+    \item apprenticeship of the expert's task
+    \item Generalization of the policy over never-seen-before states
+    \item Useful when rewards are hard to tune (e.g. driving)
+    \end{itemize}
+  \end{itemize}
+}
+{15cm}
 \vfill
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TROISIEME LIGNE
+\TwoBoxes{.48}{Feature Expectation}{
+  \begin{eqnarray*}
+    V^\pi(s_t) &=& E\left[\left.\sum\limits_{i}\gamma^i r_{t+i}\right|\pi\right]\\
+    V^\pi(s_t) &=& E\left[\left.\sum\limits_{i}\gamma^i \theta^T\phi(s_{t+i})\right|\pi\right]\\
+    V^\pi(s_t) &=& \theta^T\underbrace{E\left[\left.\sum\limits_{i}\gamma^i \phi(s_{t+i})\right|\pi\right]}_{\mu^\pi(s_t)}\\
+    V^\pi(s_t) &=& \theta^T\mu^\pi(s_t)
+  \end{eqnarray*}
+}{.48}{Importance of $\Delta \mu$}{
+  \begin{eqnarray*}
+    \Delta\mu &=& ||\mu_E(s_0) - \mu^\pi(s_0)||_2\\
+    \Delta V &=& |V^E(s_0) - V^\pi(s_0)|\\
+    \Delta V &=& |\theta^T\left(\mu_E(s_0) - \mu^\pi(s_0)\right)|\\
+    \Delta V &\leq& ||\theta||_2||\mu_E(s_0) - \mu^\pi(s_0)||_2\\
+    \Delta V &\leq& \Delta\mu\\
+  \end{eqnarray*}
+}
+{15cm}
+\vfill
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Quatrième ligne
+  \begin{columns}
+    \begin{column}{.48\textwidth}
+    \begin{block}{Abbeel and Ng's IRL}
+      \begin{columns}
+        \begin{column}{0.001\textwidth}
+          \vspace{8cm}
+        \end{column}
+        \begin{column}{\textwidth}
+          \centering
 
-\small{References}\\
-\vspace{-1cm}
+          \begin{tikzpicture} [scale=3]
+            \node[element] (trajE) at (0,0) {$\vcenter{\hbox{\includegraphics[height=2cm]{Expert.png}}}$Trajectories} ;
+            \node[action] (mech) at (3,-0.7) {\includegraphics[height=2cm]{Moulinette.png}} ;
+            \node[element] (policy) at (5,-0.7) {$\vcenter{\hbox{\includegraphics[height=2cm]{Pi.png}}}$Policy} ;
+            \node[action] (sim) at (8,-0.7) {$\vcenter{\hbox{\includegraphics[height=2cm]{ML.png}}}$Simulator} ;
+            \node[element] (trajA) at (0,-1.4) {$\vcenter{\hbox{\includegraphics[height=2cm]{Agent.png}}}$Trajectories} ;
+            \draw [->,line width=1.5mm] (trajE.east) .. controls (2,0) and (2,-0.7) .. (mech.west);
+            \draw [->,line width=1.5mm] (trajA.east) .. controls (2,-1.4) and (2,-0.7) .. (mech.west);  
+            \draw [->,line width=1.5mm] (mech.east) -- (policy.west);
+            \draw [->,line width=1.5mm] (policy.east) -- (sim.west);
+            \draw [->,line width=1.5mm] (sim.east) -- (10,-0.7) -- (10,-2.1) -- (0,-2.1) -- (trajA.south);
+          \end{tikzpicture}
+        \end{column}
+      \end{columns}
+
+    \end{block}
+    \begin{block}{LSTD-$\mu$}
+      \begin{columns}
+        \begin{column}{0.001\textwidth}
+          \vspace{8cm}
+        \end{column}
+        \begin{column}{\textwidth}
           \centering
-\begin{multicols}{2}
-  \bibliographystyle{unsrt}
-  \begin{tiny}
-    \bibliography{biblio}
-  \end{tiny}
-\end{multicols}
-\vspace{-0.2cm}
+          \begin{tikzpicture}[scale=3]
+            \node[element] (trajE) at (0,0) {$\vcenter{\hbox{\includegraphics[height=2cm]{Expert.png}}}$Transitions} ;
+            \node[action] (mech) at (3,-0.7) {\includegraphics[height=2cm]{Moulinette.png}} ;
+            \node[element] (policy) at (5,-0.7) {$\vcenter{\hbox{\includegraphics[height=2cm]{Pi.png}}}$Policy} ;
+            \node[element] (trajA) at (0,-1.4) {$\vcenter{\hbox{\includegraphics[height=2cm]{Agent.png}}}$Transitions} ;
+            \draw [->,line width=1.5mm] (trajE.east) .. controls (2,0) and (2,-0.7) .. (mech.west);
+            \draw [->,line width=1.5mm] (trajA.east) .. controls (2,-1.4) and (2,-0.7) .. (mech.west);  
+            \draw [->,line width=1.5mm] (mech.east) -- (policy.west);
+            \draw [->,line width=1.5mm] (policy.east) -- (10,-0.7) -- (10,-1.4) -- (3,-1.4) -- (mech.south);
+          \end{tikzpicture}
+        \end{column}
+      \end{columns}       
+    \end{block}
+    \end{column}
+
+    \begin{column}{.48\textwidth}
+\begin{block}{Update step}
+  $t = \max\limits_{\theta}\min\limits_{\pi}\theta^T(\mu^{E}(s_0)-\mu^{\pi}(s_0))$
+\end{block}
+
+      \begin{block}{Idea}
+
+        $V = V^\pi(s_t) = E\left[\left.\sum\limits_{i}\gamma^i r_{t+i}\right|\pi\right]$
+        \hfill $\mu^\pi(s_t) = E\left[\left.\sum\limits_i\gamma^i\phi(s_{t+i})\right|\pi\right]$ \\
+        \begin{center}
+          Transition: $s_t,a_t,s_{t+1},r_t$
+        \end{center}
+      \end{block}
+\begin{columns}
+\begin{column}{.5\textwidth}
+\begin{block}{LSTD}
+\label{sec-3_2_1_2}
+
+     Estimator : $\hat V^\pi(s) = \omega^T\phi(s)$
+     \begin{equation*}
+     \omega = LSTD_{\color{red}\phi}( \{s_t,a_t,s_{t+1}\}_t,\{{\color{red}r_t}\}_t)
+     \end{equation*}
+\end{block}
+\end{column}
+\begin{column}{.5\textwidth}
+\begin{block}{LSTD-$\mu$}
+\label{sec-3_2_1_3}
+
+     Estimator : $\hat \mu^\pi(s) = \xi^{T}\psi(s)$
+     \begin{equation*}
+     \xi_i = LSTD_{\color{red}\psi}( \{s_t,a_t,s_{t+1}\}_t, \{{\color{red}\phi_i(s_t)}\}_t )
+     \end{equation*}
+\end{block}
+\end{column}
+\end{columns}
+
+      %% \begin{block}{#5}
+      %%   \begin{columns}
+      %%     \begin{column}{0.001\textwidth}
+      %%       \vspace{#7}
+      %%     \end{column}
+      %%     \begin{column}{\textwidth}
+      %%       \centering
+      %%       #6
+      %%     \end{column}
+      %%   \end{columns}       
+      %% \end{block}
+    \end{column}
+  \end{columns}
+\vfill
+%%%%%%%%%%%%%%%%%%%%%%Cinquième ligne
+  \begin{columns}
+    \begin{column}{.23\textwidth}
+      \begin{block}{Quality Criterion}
+        \fontsize{11pt}{11pt}\selectfont
+        \resizebox{.9\columnwidth}{!}{\input{../CodeJFPDA/GridWorld/criteria_mc}}
+      \end{block}
+    \end{column}
+    \begin{column}{.23\textwidth}
+      \begin{block}{Gridworld}
+        \fontsize{11pt}{11pt}\selectfont
+        \resizebox{.9\columnwidth}{!}{\input{../CodeJFPDA/GridWorld/both_error_EB}}
+      \end{block}
+    \end{column}
+    \begin{column}{.23\textwidth}
+      \begin{block}{Inverted Pendulum (one run)}
+        \fontsize{11pt}{11pt}\selectfont
+        \resizebox{.9\columnwidth}{!}{\input{../CodeJFPDA/InvertedPendulum/threshold}}
+      \end{block}
+    \end{column}
+    \begin{column}{.23\textwidth}
+      \begin{block}{Inverted Pendulum}
+        \fontsize{11pt}{11pt}\selectfont
+        \resizebox{.9\columnwidth}{!}{\input{../CodeJFPDA/InvertedPendulum/threshold_EB}}
+      \end{block}
+    \end{column}
+  \end{columns}
 

### 2011-09-16 klein2011batch/klein2011batch.tex
Tout rangé dans un dossier à part

\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration

\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}

% Typographie française ==================================================================
\usepackage[english]{babel} % typographie francaise
\usepackage[utf8]{inputenc} % saisie de caracteres accentues, etc.

% Vos environnements =====================================================================
\newtheorem{prop}{Proposition}
\newenvironment{pf}{{\bf Proof }}{\hfill$\Box$\par}
\newcommand{\spaceafterproof}{\vspace{1em}}

% Corps de l'article =====================================================================
\usepackage{color}
% the following package is optional:
%\usepackage{latexsym} 
%\usepackage{amsthm}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
%\usepackage{algorithm2e}
\usepackage{algorithm}
\usepackage{algorithmic}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...

\title{Batch, Off-policy and Model-free Apprenticeship Learning}
%\author{Robert}
%   1. Supelec-Metz Campus, IMS Reasearch group, France. \texttt{prenom.nom@supelec.fr}}
\author{Edouard Klein\inst{1}\inst{3} \and Matthieu Geist\inst{1} \and Olivier Pietquin\inst{1}\inst{2}}
 \institute{
 1. Supélec-Metz Campus, IMS Research group, France, \texttt{prenom.nom@supelec.fr}\\
 2. UMI 2958 CNRS - GeorgiaTech, France\\
 3. Equipe ABC, LORIA-CNRS, France.
 }
\date{\today}

\begin{document}

\maketitle
\begin{abstract}
  This paper addresses the problem of apprenticeship learning, that
  is learning control policies from demonstration by an expert. An
  efficient framework for it is inverse reinforcement learning
  (IRL). Based on the assumption that the expert maximizes a utility
  function, IRL aims at learning the underlying reward from example 
  trajectories. Many IRL algorithms
  assume that the reward function is linearly parameterized and rely
  on the computation of some associated \emph{feature expectations},
  which is done through Monte Carlo simulation. However, this
  assumes to have full trajectories for the expert policy as well as
  at least a generative model for intermediate policies. In this
  paper, we introduce a temporal difference method, namely LSTD-$\mu$, to compute these
  feature expectations. This allows extending apprenticeship
  learning to a batch and off-policy setting.
\end{abstract}
\section{Introduction}

Optimal control consists in putting a machine in control of a system
with the goal of fulfilling a specific task, optimality being
defined as how well the task is performed. Various solutions to this
problem exist from automation to planification. Notably, the
reinforcement learning (RL) paradigm~\citep{sutton1998reinforcement} is a general
machine learning framework in which an agent learns to control
optimally a dynamic system through interactions with it. The task is
specified through a reward function, the agent objective being to
take sequential decisions so as to maximize the expected cumulative
reward.

However, defining optimality (through the reward function) can
itself be a challenge. If the system can be empirically controlled
by an expert, even though his/her behavior can be difficult to
describe formally, apprenticeship learning is a way to have a
machine controlling the system by mimicking the expert. Rather than
directly mimicking the expert with some supervised learning
approach, Inverse Reinforcement Learning (IRL)~\citep{ng2000algorithms}
consists in learning a reward function under which the policy
demonstrated by the expert is optimal. Mimicking the expert therefore
ends up to learning an optimal policy according to this reward
function. A significant advantage of such an approach is that
expert's actions can be guessed in states which have not been
encountered during demonstration. Firstly introduced in \citep{russell1998learning}, another advantage claimed by the author would be to
find a compact and complete representation of the task in the form
of the reward function.

There roughly exists three families of IRL algorithms:
feature-expectation-based~\citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum},
margin-maximization-based~\citep{ratliff2006maximum,ratliff2007imitation,ratliff2007boosting,kolter2008hierarchical}
and approaches based on the parameterization of the policy by the
reward function~\citep{ramachandran2007bayesian,neu2007apprenticeship}. The first family
assumes a linearly parameterized reward function. This naturally
leads to a linearly parameterized value function, the associated
feature vector being the so-called feature expectation (see Section~\ref{sec:back} for a formal definition).
These approaches learn a reward function such that the feature
expectation of the optimal policy (according to the learnt reward
function) is close to the feature expectation of the expert policy.
This is a sufficient condition to have close value functions (for any parameterized reward function, and therefore particularly the optimized one). The
second family expresses IRL as a constrained optimization problem in
which expert's examples have higher expected cumulative reward than all other
policies by a certain margin. Moreover, suboptimality of the expert
can be considered through the introduction of slack variables. The
last family parameterizes policies with a reward function. Assuming
that the expert acts according to a Gibbs policy (respectively to
the optimal value function related to the reward function which is
optimized), it is possible to estimate the likelihood of a set of
state-action pairs provided by the expert. The algorithms differ in
the way this likelihood is maximized.

This paper focuses on the first family of algorithms, and more
precisely on the seminal work of~\citet{abbeel2004apprenticeship}. All of them
rely on the computation of the feature expectation (which depends on
policies but not on rewards) of (i) the expert and (ii) some
intermediate policies. The expert's feature expectation is computed
using a simple Monte Carlo approach (which requires full
trajectories of the expert). Other feature expectations are either
computed exactly (which requires knowing analytically the dynamics
of the system) or with a Monte Carlo approach (which requires simulating the system). The contribution of this paper is LSTD-$\mu$,
a new temporal-difference-based algorithm for estimating these feature
expectations. It relaxes the preceding assumptions: transitions
of the expert are sufficient (rather than full trajectories) and nor
the model neither a simulator are necessary to compute intermediate
feature expectations. This paper focuses on the algorithm introduced
in~\citep{abbeel2004apprenticeship}, but the proposed
approach can be used in other algorithms based on feature
expectation computation.

The rest of this paper is organized as follows.
Section~\ref{sec:back} provides the necessary background,
notably the definition of feature expectation and its use in the
seminal IRL algorithm of~\citet{abbeel2004apprenticeship}.
Section~\ref{sec:lstdmu} presents LSTD-$\mu$, our main contribution.
Section~\ref{sec:exp} provides some preliminary experiments and
Section~\ref{sec:conclusion} opens perspectives.

\section{Background}
\label{sec:back}
A sequential decision problem is often framed as a Markov Decision
Process (MDP)~\citep{sutton1998reinforcement}, a tuple
$\{S,A,P,R,\gamma\}$ with $S$ being the state space, $A$ the action
space, $P\in\mathcal{P}(S)^{S\times A}$ the set of Markovian
transition probabilities, $R\in\mathbb{R}^S$ the reward function
(assumed to be absolutely bounded by 1) and $\gamma\in[0,1[$ a
discounting factor. A policy $\pi\in A^S$ maps states to actions. The
quality of a policy is quantified by the associated value function
$V^\pi$, which associates to each state the expected and discounted
cumulative reward: $V^\pi(s) = E[\sum\limits_{t=0}^\infty \gamma^t R(s_t)|s_0=s, \pi]$.
Dynamic programming aims at finding the optimal policy $\pi^*$, that
is one of the policies associated with the optimal value function,
$V^* = \argmax_\pi V^\pi$, which maximizes the value for each state.
If the model (that is transition probabilities and the reward
function) is unknown, learning the optimal control through
interactions is addressed by RL.

For IRL, the problem is reversed. It is
assumed that an expert acts according to an optimal policy $\pi_E$,
this policy being optimal according to some unknown reward function
$R^*$. The goal is to learn this
reward function from sampled trajectories of the expert. This is a
difficult and ill-posed problem~\citep{ng2000algorithms}. Apprenticeship
learning through IRL, which is the problem at hand in this paper,
has a somewhat weaker objective: it aims at learning a policy
$\tilde{\pi}$ which is (approximately) as good as the expert policy
$\pi_E$ under the unknown reward function $R^*$, for a known initial
state $s_0$ (this condition can be easily weakened by assuming a
distribution over initial states). Now, the approach proposed in~\citep{abbeel2004apprenticeship} is
presented.

We assume that the true reward function belongs to some hypothesis
space $\mathcal{H}_\phi = \{\theta^T \phi(s),
\theta\in\mathbb{R}^p\}$, of which we assume the basis functions to
be bounded by 1: $|\phi_i(s)|\leq 1, \forall s\in S, 1\leq i \leq
p$. Therefore, there exists a parameter vector $\theta^*$ such that:
 $R^*(s) = (\theta^*)^T \phi(s)$
In order to ensure that rewards are bounded, we impose that
$\|\theta\|_2\leq 1$. For any reward function belonging to
$\mathcal{H}_\phi$ and for any policy $\pi$, the related value
function $V^\pi(s)$ can be expressed as follows:
\begin{equation}
\scriptstyle
V^\pi(s) = E[\sum\limits_{t=0}^\infty \gamma^t \theta^T \phi(s_t)|s_0=s, \pi] = \theta^T  E[\sum\limits_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
\end{equation}
Therefore, the value function is also linearly parameterized, with
the same weights and with basis functions being grouped into the
so-called \emph{feature expectation} $\mu^\pi$:
\begin{equation}
  \mu^\pi(s) = E[\sum_{t=0}^\infty \gamma^t \phi(s_t)|s_0=s, \pi]
\end{equation}
Recall that the problem is to find a policy whose performance is
close to that of the expert's for the starting state $s_0$, on the
unknown reward function $R^*$. In order to achieve this goal, it is
proposed in~\citep{abbeel2004apprenticeship} to find a policy $\tilde{\pi}$ such
that $\|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2\leq \epsilon$
for some (small) $\epsilon>0$. Actually, this ensures that the value
of the expert's policy and the value of the estimated policy (for
the starting state $s_0$) are close for \emph{any} reward function
of $\mathcal{H}_\phi$:
\begin{equation}
\scriptstyle
  |V^{\pi_E}(s_0) - V^{\tilde{\pi}}(s_0)| =
  |\theta^T(\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0))|
  \leq \|\mu^{\pi_E}(s_0)-\mu^{\tilde{\pi}}(s_0)\|_2
\label{eqn:vs0}
\end{equation}
This equation uses the Cauchy-Schwarz inequality and the
assumption that $\|\theta\|_2\leq 1$. Therefore, the approach
described here does not ensure to retrieve the true reward function
$R^*$, but to act as well as the expert under this reward function
(and actually under any reward function).

Let us now describe the algorithm proposed in~\citep{abbeel2004apprenticeship} to
achieve this goal:
\begin{enumerate}
  \item Starts with some initial policy $\pi^{(0)}$ and compute
  $\mu^{\pi^{(0)}}(s_0)$. Set $j=1$;
  %
  \item Compute $t^{(j)} = \max_{\theta: \|\theta\|_2\leq 1}
  \min_{k\in\{0,j-1\}}\theta^T(\mu^{\pi_E}(s_0) -
  \mu^{\pi^{(k)}}(s_0))$ and let $\theta^{(j)}$ be the value
  attaining this maximum. At this step, one searches for the reward
  function which maximizes the distance between the value of the
  expert at $s_0$ and the value of \emph{any} policy computed so far
  (still at $s_0$). This optimization problem can be solved using a
  quadratic programming approach or a projection
  algorithm~\citep{abbeel2004apprenticeship};
  %
  \item if $t^{(j)}\leq \epsilon$, terminate. The algorithm outputs a
  set of policies $\{\pi^{(0)}, \dots, \pi^{(j-1)}\}$ among which
  the user chooses manually or automatically the closest to the
  expert (see~\citep{abbeel2004apprenticeship} for details on how to choose this
  policy).
  Notice that the last policy is not necessarily the best (as illustrated in Section~\ref{sec:exp});
  %
  \item solve the MDP with the reward function $R^{(j)}(s) =
  (\theta^{(j)})^T\phi(s)$ and denote $\pi^{(j)}$ the associated
  optimal policy. Compute $\mu^{\pi^{(j)}}(s_0)$;
  %
  \item set $j\leftarrow j+1$ and go back to step 2.
\end{enumerate}
There remain three problems: computing the feature expectation of
the expert, solving the MDP and computing feature expectations of
intermediate policies.

As suggested in~\citep{abbeel2004apprenticeship}, solving the MDP can be done
approximately by using any appropriate reinforcement learning
algorithm. In this paper, we use the Least-Squares Policy Iteration
(LSPI) algorithm~\citep{lagoudakis2003least}. There remains to estimate
feature expectations. In~\citep{abbeel2004apprenticeship}, $\mu^{\pi_E}(s_0)$ is
estimated using a Monte Carlo approach over $m$ trajectories:
$  \hat{\mu}_E(s_0) = \frac{1}{m} \sum\limits_{h=1}^m \sum\limits_{t=0}^\infty
  \gamma^t \phi(s_t^{(h)})$
This approach does not hold if only transitions of the expert are
available, or if trajectories are too long (in this case, it is
still possible to truncate them). For intermediate policies, it is
also suggested to estimate associated feature expectations using a
Monte Carlo approach (if they cannot be computed exactly). This is
more constraining than for the expert, as this assumes that a
simulator of the system is available. In order to address these
problems, we introduce a temporal-difference-based algorithm to
estimate these feature expectations.

\section{LSTD-$\mu$}
\label{sec:lstdmu}
Let us write the definition of the $i^\text{th}$ component of a
feature expectation $\mu^\pi(s)$ for some policy $\pi$:
$  \mu_i^\pi(s) = E[\sum\limits_{t=0}^\infty \gamma^t \phi_i(s_t)|s_0=s,\pi]$
This is exactly the definition of the value function of the policy
$\pi$ for the MDP considered with the $i^\text{th}$ basis function
$\phi_i(s)$ as the reward function. There exist many algorithms to
estimate a value function, any of them can be used to estimate
$\mu_i^\pi$. Based on this remark, we propose to use specifically
the least-squares temporal difference (LSTD)
algorithm~\citep{bradtke1996linear} to estimate each component of the
feature expectation (as each of these components can be understood
as a value function related to a specific and known reward
function).

Assume that a set of transitions $\{(s_t,r_t,s_{t+1})_{1\leq t \leq
n}\}$ sampled according to the policy $\pi$ is available. We assume
that value functions are searched for in some  hypothesis space
$\mathcal{H}_\psi = \{ \hat{V}_\xi(s) = \sum\limits_{i=1}^q \xi_i \psi_i(s)
= \xi^T \psi(s), \xi\in\mathbb{R}^q\}$. As reward and value
functions are possibly quite different, another hypothesis space is
considered for value function estimation. But if $\mathcal{H}_\phi$
is rich enough, one can still consider
$\mathcal{H}_\psi=\mathcal{H}_\phi$. Therefore, we are looking for
an approximation of the following form:
$  \hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$
The parameter vector $\xi_i^*$ is here the LSTD estimate:
\begin{equation}
  \xi_i^* = \left(\sum_{t=1}^n
  \psi(s_t)(\psi(s_t)-\gamma\psi(s'_{t}))^T\right)^{-1}
  \sum_{t=1}^n \psi(s_t) \phi_i(s_t)
\end{equation}
For apprenticeship learning, we are interested more particularly in
$\hat{\mu}^\pi(s_0)$. Let $\Psi = (\psi_i(s_t))_{t,i}$ be the
$n\times q$ matrix of values predictors, $\Delta\Psi = (\psi_i(s_t)
- \gamma\psi_i(s'_t))_{t,i}$ be the related $n\times q$ matrix and
$\Phi = (\phi_i(s_t))_{t,i}$ the $n\times p$ matrix of reward
predictors. It can be easily checked that $\hat{\mu}^\pi(s_0)$
satisfies:
\begin{equation}
  (\hat{\mu}^\pi(s_0))^T = \psi(s_0)^T (\Psi^T \Delta\Psi)^{-1} \Psi^T \Phi
\end{equation}
This provides an efficient way to estimate the feature expectation
of the expert in $s_0$.

There remains to compute the feature expectations of intermediate
policies, which should be done in an off-policy manner (that is
without explicitly sampling trajectories according to the policy of
interest). To do so, still interpreting each component of the
feature expectation as a value function, we introduce a state-action
feature expectation defined as follows (much as the classic
$Q$-function extends the value function):
$\mu^\pi(s,a) = E[\sum\limits_{t=0}^\infty \gamma^t
  \phi(s_t)|s_0=s,a_0=a,\pi]$
Compared to the classic feature expectation, this definition adds a
degree of freedom on the first action to be chosen before following
the policy $\pi$. With a slightly different definition of the
related hypothesis space, each component of this feature expectation
can still be estimated using the LSTD algorithm (namely using the
LSTD-Q algorithm~\citep{lagoudakis2003least}). The clear advantage of
introducing the state-action feature expectation is that this
additional degree of freedom allows off-policy learning.
Extending LSTD-$\mu$ to state-action LSTD-$\mu$ is done in the same
manner as LSTD is extended to LSTD-Q~\citep{lagoudakis2003least},
technical details are not provided here for the clarity of
exposition.

Given the (state-action) LSTD-$\mu$ algorithm, the apprenticeship
learning algorithm presented in Section~\ref{sec:back} can be easily extended to a batch and
off-policy setting. The solely available data is a set of
transitions sampled according to the expert policy  (and possibly a set of sub-optimal trajectories). The
corresponding feature expectation for the starting state $s_0$ is
estimated with the LSTD-$\mu$ algorithm. At step~4 of this
algorithm, the MDP is (approximately) solved using
LSPI~\citep{lagoudakis2003least} (an approximate policy iteration
algorithm using LSTD-Q as the off-policy $Q$-function estimator).
The corresponding feature expectation at state $s_0$ is estimated
using the proposed state-action LSTD-$\mu$.

Before presenting some experimental results, let us stress that
LSTD-$\mu$ is simply the LSTD algorithm applied to a specific reward
function. Although quite clear, the idea of using a temporal
difference algorithm to estimate the feature expectation is new, as
far as we know. A clear advantage of the proposed approach is that
any theoretical result holding for LSTD also holds for LSTD-$\mu$,
such as convergence~\citep{nedic2003least} or
finite-sample~\citep{lazaric2010finiteLSTD} analysis for example.
\section{Experimental benchmark}
\label{sec:exp}
This section provides experimental results on two complementary problems. Subsection~\ref{ssec:exp} details the protocol and the results while subsection~\ref{ssec:quality} inspects the meaning of the different quality criteria. 
\subsection{Experiment description and results}
\label{ssec:exp}
\subsubsection{GridWorld}
\begin{figure}
\centering
\resizebox{\columnwidth}{!}{\input{../Code/GridWorld/both_error_EB}}
\caption{$||\mu^{\pi}(s_0) - \mu^{\pi_E}(s_0)||_2$ with respect to the number of samples available from the expert. The error bars represent the standard deviation over 100 runs.}
\label{fig:E}
\end{figure}
The first experimental benchmark chosen here is one of those proposed in \citep{ng2000algorithms}, a 5x5 grid world. The agent is in one of the cells of the grid (whose coordinates is the state) and can choose at each step one of the four compass directions (the action). With probability 0.9, the agent moves in the intended direction. With probability 0.1, the direction is randomly drawn. The reward optimized by the expert is 0 everywhere except in the upper-right cell, where it is 1. For every policy, an episode ends when the upper right cell is attained, or after 20 steps. At the start of each episode, the agent is in the lower-left corner of the grid (the opposite of where the reward is). Both the state and action spaces are finite and of small cardinality. Hence, the chosen feature functions $\phi$ and $\psi$ are the typical features of a tabular representation: 0 everywhere except for the component corresponding to the current state (-action pair).\\

Both \citet{abbeel2004apprenticeship}'s algorithm (from now on referred to as the MC variant) and our adaptation (referred to as the LSTD variant) are tested side by side. The MDP solver of the MC variant is LSPI with a sample source that covers the whole grid (each state has a mean visitation count of more than 150) and draws its action randomly. Both $\mu^{\pi^{(j)}}(s_0)$ and $\mu^{\pi_E}(s_0)$ are computed thanks to a Monte-Carlo estimation with enough samples to make the variance negligible. We consider both these computations as perfect theoretical solvers for all intended purpose on this toy problem. We thus are in the case intended by \citet{abbeel2004apprenticeship}. On the other hand the LSTD variant is used without accessing a simulator. It uses LSPI and LSTD-$\mu$, fed only with the expert's transitions (although we could also use non expert transitions  to compute intermediate policies, if available). This corresponds to a real-life setting where data generation is expensive and the system cannot be controlled by an untrained machine.\\

We want to compare the efficiency of both versions of the algorithm with respect to the number of samples available from the expert, as these samples usually are the bottleneck. Indeed as they are quite costly (in both means and human time) to generate they are often not in abundance hence the critical need for an algorithm to be expert-sample efficient. Having a simulator at one's disposal can also be difficult. For the simple problem we use this is however not an issue. The discussion about the choice of the performance metric has its own dedicated subsection (subsection~\ref{ssec:quality}). We use here the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term. Fig.~\ref{fig:E} shows, for some numbers of samples from the expert, the value of $||\mu^{\pi^{(j)}}(s_0)-\mu^{\pi_E}(s_0)||_2$ after the algorithm converged or attained the maximum number of iterations (we used 40). The best policy is found by LSTD variant after one iteration only\footnote{Precise reasons why it happens are not clear now, but certainly have something to do with the fact that all the estimations are made along the same distribution of samples.} whereas in the MC variant, convergence happens after at least ten iterations. The best policy is not always the last, and although it has always been observed so far with the LSTD variant, there is absolutely no way to tell whether this is a feature. The score of the best policy (not the last) according to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ error term is plotted here. We can see that although the LSTD variant is not as good as the MC variant when very few samples are available, both algorithms quickly converge to almost the same value ; our version converges to a slightly lower error value. The fact that our variant can work in a batch, off-policy and model-free setting should make it suitable to a range of tasks inaccessible to the MC variant, the requirement of a simulator being often constraining.
\subsubsection{Inverted Pendulum}
\begin{figure}
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
      \resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold}}
      \caption{Number of balancing steps for the policies found by both variants with respect to the number of samples from a sub-optimal policy given to LSPI. Note the absence of middle ground in the quality.}
      \label{fig:threshold}
    \end{center}
\end{minipage}
\hfill
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
      \resizebox{\columnwidth}{!}{\input{../Code/InvertedPendulum/threshold_EB}}
      \caption{Same plot as Fig.~\ref{fig:threshold}, with mean and standard deviation over 100 runs. The big standard deviation stems from the absence of middle ground in the quality of the policies. The increasing mean w.r.t. the abscissa means that the proportion of good policies increases with the number of samples from a sub-optimal policy given to LSPI.}
      \label{fig:threshold_EB}
    \end{center}
\end{minipage}
\end{figure}

%% \begin{multicols}{2}
%% \begin{figure}
%% \centering
%% \end{figure}
%% \begin{figure}
%% \centering
%% \end{figure}
%% \end{multicols}
Another classic toy problem is the \emph{inverted pendulum}, used for example in \citep{lagoudakis2003least} from where we drew the exact configuration (mass of the pendulum, length of the arm, definition of the feature space, etc.). In this setting, the machine must imitate an expert who maintains the unstable equilibrium of an inverted pendulum allowed to move along only one axis. A balancing step is considered successful (reward $0$) when the angle is less than $\pi\over 2$. It is considered a failure (reward $-1$) when the pendulum has fallen (i.e. the angle exceeds $\pi \over 2$). A run is stopped after a failure or after 3000 successful balancing steps.\\

This problem bears two fundamuntal differences with the previous one and thus comes as complementary to it. First, it presents a continuous state space whereas in the GridWorld a tabular representation was possible. Then, the initial state is randomly drawn from a non singleton subspace of the state space. This last point is addressed differently by the LSTD variant and the MC variant. the MC variant will naturally sample the initial state distribution at each new trajectory from the expert. The LSTD variant, on the other hand, still does not need complete trajectories from the expert but mere samples. As it approximates the whole $\mu^{\pi}$ function and not just $\mu^\pi(s_0)$, it is possible to compute the mean of the approximation $\hat\mu^\pi$ over a few states drawn with the same distribution as the initial states.\\

A striking finding on this toy problem is that the parameter controlling the success or the failure of the experiment is not the number of samples from the expert, but the quality of the samples available to LSPI to solve the MDP. \emph{A contrario} to what happened in the previous toy problem, there are fewer samples from the expert than what LSPI needs to solve the MDP as they do not cover the whole state space. When given only these samples, the LSTD variant fails. The MC variant, however, is not bound by such restrictions and can use as many samples as it needs from a simulator. Thus, with only one trajectory from the expert (100 samples) the MC variant is able to successfully balance the pendulum for more than 3000 steps. The problem here stems from the fact that we don't learn the reward, but a policy that maximizes the reward. If the optimal control problem was solved separately, learning the reward only from the samples of the expert would be possible.\\

The sensitivity of both algorithms to the number of samples available to LSPI is extreme, as is shown in Fig.~\ref{fig:threshold}. It may seem nonsensical to restrict the number of samples available to LSPI for the MC variant as it can access a simulator, but this has been done to show that both variants exhibit the same behavior, hence allowing us to locate the source of this behavior in what both algorithms have in common (the use of LSPI inside the structure of \citep{abbeel2004apprenticeship}) excluding the point where they differ (LSTD-$\mu$ and Monte-Carlo).\\

Fig.~\ref{fig:threshold_EB} shows that when given samples from a sub-optimal policy, the LSTD variant can sort the problem out, statistically almost always with 3000 sub-optimal samples, and sometimes with as low as 200 sub-optimal samples. Such a setting is still batch, off-policy and model-free. When given enough sub-optimal samples, both variants are successful with only one trajectory from the expert (i.e. 100 samples, this is what is plotted here). Giving more trajectories from the expert does not improve the success rate.
\subsection{Discussion about the quality criterion}
\label{ssec:quality}
\begin{figure}
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
      \resizebox{\columnwidth}{!}{\input{../Code/GridWorld/criteria_mc}}
      \caption{Different criteria with respect to the number of iterations for a run of the MC variant.}
      \label{fig:A}
    \end{center}
\end{minipage}
\hfill
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
\resizebox{\columnwidth}{!}{\input{../Code/GridWorld/criteria_lstd_EB}}
\caption{Different criteria with respect to the number of samples from the expert, for several runs of the LSTD variant. We can see that the algorithm is blind, as all it has access to is always zero. The true error values, however, smoothly converge to something small. Knowing how many expert samples are actually needed in a real world problem is an open question. The error bars represent the standard deviation over 100 runs.}
\label{fig:B}
    \end{center}
\end{minipage}
\end{figure}

Fig.~\ref{fig:A}~and~\ref{fig:B} illustrate (using the GridWorld problem) the difficulty of choosing the quality criterion ; Fig.~\ref{fig:A} shows four different quality criteria during a run of the MC variant. Fig.~\ref{fig:B} shows the same criteria for several runs of the LSTD variant with a varying number of samples from the expert. The $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term is widely discussed in \citep{abbeel2004apprenticeship}'s additional material. It bears an interesting relation with the difference between the expert's value function and the current value function in the initial state with respect \emph{to the current reward} (Eq.~\ref{eqn:vs0}).\\

The fact that  the curves of the true error term $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ and  its estimate $||\hat\mu^{\pi^{(j)}}(s_0) - \hat\mu^{\pi_E}(s_0)||_2$ are indistinguishable in Fig.~\ref{fig:A} means that, for it has access to a cheap simulator, the MC variant works as if it had access to the exact values. This however cannot be said of the LSTD variant, for which the two curves are indeed different (Fig.~\ref{fig:B}). Not knowing the true value of $\mu^{\pi_E}(s_0)$ may be a problem for our variant, as it can introduce an error in the stopping criterion of the algorithm.\\

It shall be noted that although it plays its role, the halt criterion is not a good measure of the quality of the current policy in the MC variant either, as it can be low (and thus halt the algorithm) when the policy is bad. The best policy, however, can be easily chosen among all those created during the execution of the algorithm thanks to the $||\mu^{\pi^{(j)}}(s_0) - \mu^{\pi_E}(s_0)||_2$ term, which the MC variant can compute. When this term is low, the objective performance (that is, $V^{\pi_E}(s_0)-V^{\pi^{(j)}}(s_0)$ with respect to the unknown true reward function) is low too. 
\section{Conclusion}
\label{sec:conclusion}
Given some transitions generated by an expert controlling a system and maximizing in the long run an unknown reward, we ported \citet{abbeel2004apprenticeship}'s approach to apprenticeship learning via inverse reinforcement learning to a batch, model-free, off-policy setting. Experimentally, there is a need for either a slightly bigger number of samples from the expert or some samples from a sub-optimal policy. We believe this cost is not prohibitive as our approach only requires isolated samples which are often less difficult to get than whole trajectories as needed by the original approach. Furthermore, tranferring the reward and not the policy may overcome this difficulty. We intend to do this in a real life setting.
The simple idea of using LSTD to estimate the feature expectation could be applied to other algorithms as well, for example \citep{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game,ziebart2008maximum}.\\

%\bibliographystyle{named}
\bibliographystyle{named}
\bibliography{../Biblio/Biblio}
\end{document}

### 2012-02-02 klein2012/main.tex

% Exemple de fichier fonctionnnant avec la classe CAp2012.cls
% Base :  
%	Classe pf2010.cls adpatée de pf2003 par Jean CHarlet
%	Classe pf2003.cls adpatée de ic2001 par Jean CHarlet
%
% 	Classe ic2001.cls adpatée de EEGDRI3 par Jean CHarlet
%
% 	Classe EEGDRI3.cls adpatée de ic2000 par Jean CHarlet
%
% 	Classe IC'2000 (ic2000.cls) par Jean Charlet
%
% 	adaptée de la classe IC'99 (afia99.cls) développée par Fabien Torre

%\documentclass{CAp2012}
\documentclass[publibook-draft]{CAp2012}

\newcommand{\verbfichex}{CAp2012.\xspace}
\newcommand{\verbfichclass}{CAp2012.cls\xspace}
\newcommand{\verbfichbst}{CAp2012.bst\xspace}

%\usepackage[pdftex]{pstricks}

\usepackage{verbatim}
\usepackage{subfigure}
\usepackage[french]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{natbib}

\newtheorem{algorithm}{Algorithme}
\newtheorem{definition}{Définition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Titre court si le titre fait plus de 40 caractères
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\shorttitle{Batch Inverse Reinforcement Learning}

\shortouvrage{CAp 2012}

% Titre, auteur, pas de date

\title{Batch Imitation via Inverse Reinforcement Learning Without Solving the Direct Problem}

\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}\inst{2}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}\inst{3}}}

\institute{
Sup\'elec,\\
IMS Research group, France\\
\texttt{prenom.nom@supelec.fr}
\and
Equipe ABC,\\
LORIA-CNRS, France
\and
UMI 2958\\
GeorgiaTech-CNRS, France
}

\begin{document}
\maketitle

\begin{abstract}
  This paper addresses the problem of apprenticeship learning, that is learning control policies from demonstration by an expert. An efficient framework for it is inverse reinforcement learning  (IRL). Based on the assumption that the expert maximizes a utility  function, IRL aims at learning the underlying reward from example trajectories. Many IRL algorithms assume that the reward function is linearly parameterized and rely on some or all of the following elements : complete trajectories from the expert, a generative model for Monte-Carlo estimations, known transition probabilities and the ability to repeatedly solve the forward problem. In this paper we introduce a simple gradient ascent algorithm that lifts all these constraints. Relying on a temporal difference method for the estimation of the expert's {\it feature expectation} and a good set of features, our algorithm devise a reward function compatible with the expert's demonstrated behavior with a small computational and sample complexity.
  \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
\end{abstract}
\section{Introduction}
\begin{itemize}
\item Intérêt de l'Apprenticeship learning
\item Intérêt de l'IRL pour faire de l'apprenticeship learning
\item Rapide point sur les limitations des algos d'IRL actuels
\begin{itemize}
\item \cite{ng2000algorithms} : Probabilités de transitions devant être connues
\item \cite{abbeel2004apprenticeship} et toutes les approches résumées dans \cite{neu2009training}: Résolution du problème direct à chaque itération
\end{itemize}
\end{itemize}
\section{Background}
\label{back.sec}
\begin{itemize}
\item Introduction rapide, pour les notations, des MDPs, du RL et de l'IRL
\item Résumé plus détaillé que précédemment des approches IRL actuelles
\item Introduction du travail de Ratliff
\end{itemize}
\section{Loss-augmented Feature Expectation Matching}
\begin{itemize}
\item Reprise de l'équation de Ratliff
\item Reprise de l'explication sur l'approximation de $Q$ de la section \ref{back.sec}.
\item Dérivation de l'algo à partir de ces deux éléments
\item Pseudo-code
\end{itemize}
\section{Experimental results}
\subsection{GridWorld}
\begin{itemize}
\item Description du setting (GridWorld, expert, nombre de trajectoires, etc.)
\item Description des réglages de LAFEM (fonction $l$ notamment, calcul exact de $\mu_E$)
\item Présentation des résultats (pas grand chose à dire)
\end{itemize}
\begin{figure}
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
       \includegraphics[width=\textwidth]{../GridWorld/true_reward.pdf}
       \caption{Expert's reward}
       \label{nbabo}
    \end{center}
\end{minipage}
\hfill
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
       \includegraphics[width=\textwidth]{../GridWorld/retrieved_reward.pdf}
       \caption{Reward found by LAFEM}
       \label{croissnbabo}
    \end{center}
\end{minipage}\\
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
       \includegraphics[width=\textwidth]{../GridWorld/V_expert.pdf}
       \caption{Expert's reward}
       \label{nbabo}
    \end{center}
\end{minipage}
\hfill
\begin{minipage}[t]{.4\linewidth}
    \begin{center}
       \includegraphics[width=\textwidth]{../GridWorld/V_agent.pdf}
       \caption{Agent's value function}
       \label{croissnbabo}
    \end{center}
\end{minipage}\\

\end{figure}

\subsection{Inverted Pendulum}
\begin{itemize}
\item Présentation du setting (Le pendule, l'expert)
\item Réglages de LAFEM (Préciser l'utilisation de LSTD$\mu$)
\item Présentation des résultats
\item Discussion sur la variabilité de la récompense
\item Présentation des performances statistiques (expérience pas encore effectuée)/ Etude de la complexité en échantillons 
\end{itemize}
\section{Conclusion and future work}
\begin{itemize}
\item On enlève les contraintes du domaine d'un seul coup, 
\item Il ne reste que la sélection ou détection de features, mais ce n'est pas hors de portée
\item Dans la même veine, il reste la résolution du problème direct une fois la récompense obtenue, il faut peut-être creuser du côté du reward shaping
\item Faible complexité computationnelle
\item Faible complexité en échantillons
\item Il reste des tests à faire sur le Highway driving (Pour ICML ?) et sur des expériences dans la vraie vie (bras robot ?)
\end{itemize}
%
% Bibliographie
%
\bibliography{../../Biblio/Biblio}

\end{document}

### 2012-02-02 klein2012/main.tex
Papier CAP : Abstract en Français et intégration des remarques de M&O

@@
 \begin{abstract}
-  This paper addresses the problem of apprenticeship learning, that is learning control policies from demonstration by an expert.
-An efficient framework for it is inverse reinforcement learning  (IRL).
-Based on the assumption that the expert maximizes a utility  function, IRL aims at learning the underlying reward from example trajectories.
-Many IRL algorithms assume that the reward function is linearly parameterized and rely on some or all of the following elements : complete trajectories from the expert, a generative model for Monte-Carlo estimations, known transition probabilities and the ability to repeatedly solve the forward problem.
-In this paper we introduce a simple gradient ascent algorithm that lifts all these constraints.
-Relying on a temporal difference method for the estimation of the expert's {\it feature expectation} and a good set of features, our algorithm devise a reward function compatible with the expert's demonstrated behavior with a small computational and sample complexity.
+  Cette publication s'occupe du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse ({\it Inverse Reinforcement Learning, IRL}).
+Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
+L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité et cherche à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
+Les algorithmes d'IRL existant nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition et la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée.
+Notre contribution est la présentation d'un nouvel algorithme d'IRL levant toutes ces contraintes.
+En utilisant méthode supervisée dans laquelle nous introduisons la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient et possèdant une faible complexité tant en échantillons que calculatoire.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
@@
 \begin{itemize}
+\item Introduction informelle sur les processus de décision séquentiels
 \item Intérêt de l'Apprenticeship learning
@@
 \item Introduction rapide, pour les notations, des MDPs, du RL et de l'IRL
-\item Résumé plus détaillé que précédemment des approches IRL actuelles
-\item Introduction du travail de Ratliff
 \end{itemize}
@@
 \begin{itemize}
-\item Reprise de l'équation de Ratliff
+\item Reprise de l'équation de Ratliff, justification sommaire
 \item Reprise de l'explication sur l'approximation de $Q$ de la section \ref{back.sec}.
@@
 \item Présentation des performances statistiques (expérience pas encore effectuée)/ Etude de la complexité en échantillons 
+\item Pas de comparaison avec le travail de \cite{abbeel2004apprenticeship} car celui-ci ne marche tout simplement pas avec les informations dont on dispose
+\end{itemize}
+\section{Related Work}
+\begin{itemize}
+\item Résumé plus détaillé que précédemment des approches IRL actuelles
+\item Laïus sur Ratliff et son approche locale (imitation pure) et globale (plain de contraintes)
 \end{itemize}
@@
 \item Faible complexité en échantillons
-\item Il reste des tests à faire sur le Highway driving (Pour ICML ?) et sur des expériences dans la vraie vie (bras robot ?)
+\item Il reste des tests à faire sur le Highway driving (Pour ICML ?) et sur des expériences dans la vraie vie (bras robot ?) pour comparer à l'existant
 \end{itemize}

### 2012-02-02 klein2012/main.tex
Article CAp : Description de l'algorithme

@@
 \usepackage[T1]{fontenc}
+\usepackage{amsmath}
+\usepackage{amssymb}
 
 \usepackage{natbib}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+%%% francisation des algorithmes
+\renewcommand{\algorithmicrequire} {\textbf{\textsc{Entrées:}}}
+\renewcommand{\algorithmicensure}  {\textbf{\textsc{Sorties:}}}
+\renewcommand{\algorithmicwhile}   {\textbf{tantque}}
+\renewcommand{\algorithmicdo}      {\textbf{faire}}
+\renewcommand{\algorithmicendwhile}{\textbf{fin tantque}}
+\renewcommand{\algorithmicend}     {\textbf{fin}}
+\renewcommand{\algorithmicif}      {\textbf{si}}
+\renewcommand{\algorithmicendif}   {\textbf{finsi}}
+\renewcommand{\algorithmicelse}    {\textbf{sinon}}
+\renewcommand{\algorithmicthen}    {\textbf{alors}}
+\renewcommand{\algorithmicfor}     {\textbf{pour}}
+\renewcommand{\algorithmicforall}  {\textbf{pour tout}}
+\renewcommand{\algorithmicdo}      {\textbf{faire}}
+\renewcommand{\algorithmicendfor}  {\textbf{fin pour}}
+\renewcommand{\algorithmicloop}    {\textbf{boucler}}
+\renewcommand{\algorithmicendloop} {\textbf{fin boucle}}
+\renewcommand{\algorithmicrepeat}  {\textbf{répéter}}
+\renewcommand{\algorithmicuntil}   {\textbf{jusqu'à}}
 
-\newtheorem{algorithm}{Algorithme}
+\floatname{algorithm}{Algorithme}
+
+%\newtheorem{algorithm}{Algorithme}
 \newtheorem{definition}{Définition}
@@
 \section{Loss-augmented Feature Expectation Matching}
-\begin{itemize}
-\item Reprise de l'équation de Ratliff, justification sommaire
-\item Reprise de l'explication sur l'approximation de $Q$ de la section \ref{back.sec}.
-\item Dérivation de l'algo à partir de ces deux éléments
-\item Pseudo-code
-\end{itemize}
+\subsection{Principe}
+Notre algorithme cherche à minimiser la fonction de risque suivante :
+\begin{equation}
+  R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right)
+\end{equation}
+Dans le travail de \citep{ratliff2007imitation}, cette fonction de risque est introduite pour décrire un classifieur utilisant la fonction de score $q$ et qui à un élément $s$ associe la label $a = \arg\max\limits_a q(s,a)$.
+Cela est très similaire à l'équation
+  \begin{equation}
+  \label{greedy.eqn}
+  \pi^*(s) = \arg\max_{a} Q^*(s,a)
+  \end{equation}
+  qui décrit le mécanisme glouton liant fonction de valeur optimale et politique optimale.
+La politique de l'expert étant une politique optimale, elle doit respecter cette équation.
+Minimiser notre fonction de risque revient donc à chercher une fonction $q$ compatible avec la fonction de valeur optimale de l'expert.
+En effet, considérons que la fonction de perte $l$ est uniformément nulle et regardons ce qui se passe lorsque la fonction de risque $R_N(q)$ est minimisée.
+On introduit $a^* = \arg\max\limits_a(q(s_i,a))$.
+Si la fonction de risque $R_N(q)$ est positive, alors il arrive pour certains $i$ que $q(s_i,a^*) > q(s_i,a_i)$.
+Cela signifie que la fonction $q$ ne permet pas de justifier le choix de l'expert.
+A l'opposé, si l'on parvient à anuller $R_N(q)$, on a $\forall i, q(s_i,a^*) = q(s_i,a_i)$, et donc $a_i \in \arg\max\limits_a q(s_i,a)$, ce qui justifie le choix de l'expert.\\
+
+   La fonction de perte $l$ permet d'introduire de la connaissance à priori en laissant l'opérateur influer sur la forme de la fonction $q$ recherchée.\\
+
+   Nous allons profiter de la structure présente dans $Q$ pour introduire la notion de \emph{feature expectation} dans le problème.
+On rappelle qu'en IRL il est classique d'écrire $Q(s,a) = \theta^T \mu(s,a)$.
+Nous nous basons donc sur la \emph{feature expectation} de l'expert, $\mu_E$ et un vecteur de paramètres $\theta$ contrôlant la fonction de récompense que l'on recherche pour réécrire la fonction de risque que nous allons minimiser :
+  \begin{equation}
+   R_N(\theta) = {1\over N} \sum_{i=1}^N\left(\max_{a}(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
+   \end{equation}
+   Il faut noter ici l'utilisation non pas de $\mu_E : S\rightarrow \mathbb{R}^p$ mais de $\mu_E : S\times A \rightarrow \mathbb{R}^k$.
+Cette dernière partage avec la première la même relation que $Q$ partage avec $V$, c'est à dire que l'on dispose d'un degré de liberté en plus correspondant à la prochaine action, qui peut différer de celle choisie par l'expert.
+
+   De la même manière que précedemment, on note $a^* = \arg\max\limits_{a} = \theta_t ^T \mu_E(s_i,a) + l(s_i,a)$, l'expression de la fonction de risque devient alors :
+   \begin{eqnarray}
+   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*) + l(s_i,a^*) - \theta ^T \mu_E(s_i,a_i) \right)\\
+   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*) - \mu_E(s_i,a_i)\right) + l(s_i,a^*)  \right).
+   \end{eqnarray}
+
+   La présence d'un $\max$ dans cette expression (caché dans le $a^*$) nous oblige à utiliser le sous gradient dont la règle est que le sous gradient du $\max$ est le gradient de l'$\arg\max$.
+
+   Le sous gradient de cette expression est donc : 
+   \begin{equation}
+   \nabla L_N(\theta) =  {1\over N}\sum_{i=1}^N\left(\mu_E(s_i,a^*) - \mu_E(s_i,a_i)\right).
+   \end{equation}
+
+   Une descente de gradient sur le risque donne la règle d'update suivante :
+   \begin{equation}
+   \theta_{t+1} = \theta_t -\alpha_t{1\over N}\sum_{i=1}^N\left(\mu_E(s_i,a^*) - \mu_E(s_i,a_i)\right),
+   \end{equation}
+
+   avec $\alpha_t$ un pas d'apprentissage. L'algorithme complet est détaillé algorithme \ref{LAFEM.alg}.
+
+\begin{algorithm}
+\caption{LAFEM ({\it Loss-Augmented Feature Expectation Matching})}
+\label{LAFEM.alg}
+\begin{algorithmic}
+  \REQUIRE une fonction de perte $l : S\times A \rightarrow \mathbb R$
+  \REQUIRE la \emph{feature expectation} de l'expert $\mu_E : S\times A \rightarrow \mathbb R ^p$
+  \REQUIRE un pas d'apprentissage $\alpha : \mathbb N \rightarrow \mathbb R$
+  \REQUIRE des données provenant de l'expert $\{(s_i,a_i)\}_{i=1..N}$
+  \REQUIRE un vecteur de paramètres initial $\theta_0$
+  \REQUIRE un nombre maximum d'itérations $T\in \mathbb N$
+  \REQUIRE une description de l'espace d'action $A$
+
+  \FOR {T iterations}
+  \STATE $\Delta\theta \leftarrow [0...0]^T$
+    \FORALL{ $(s_i,a_i)$}
+    \STATE $a^* \leftarrow \arg\max\limits_{a} = \theta_t ^T \mu_E(s_i,a) + l(s_i,a)$
+    \STATE $\Delta\theta \leftarrow \Delta\theta + \mu_E(s_i,a^*) - \mu_E(s_i,a_i)$
+    \ENDFOR
+    \STATE $\theta_{t+1} = \theta_t -\alpha_t{1\over N}\Delta\theta$
+    \ENDFOR
+    \RETURN $\theta_T$
+\end{algorithmic}
+\end{algorithm}
 \section{Experimental results}

### 2012-02-06 klein2012/main.tex
Version de l'article CAp soumise durant le WE

@@
 \end{algorithm}
-\section{Experimental results}
-\subsection{GridWorld}
-\begin{itemize}
-\item Description du setting (GridWorld, expert, nombre de trajectoires, etc.)
-\item Description des réglages de LAFEM (fonction $l$ notamment, calcul exact de $\mu_E$)
-\item Présentation des résultats (pas grand chose à dire)
-\end{itemize}
+\subsection{Calcul de $\mu_E$}
+L'algorithme présenté repose sur le calcul de la {\it feature expectation} de l'expert, $\mu_E$. Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour la calculer.
+
+La définition du vecteur $\mu_E$ étant proche de celle de la fonction de valeur, en adaptant les algorithmes de programmation dynamique destinés au calcul de la fonction de valeur on obtient une règle de la forme :
+\begin{equation}
+\mathbf \mu^\pi = \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi
+\end{equation}
+permettant le calcul exact de la fonction de valeur pour peu que l'on dispose des probabilités de transitions.\\
+
+Dans le cas plus probable où celles-ci sont inconnues, on peut si l'on dispose d'un simulateur utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
+\begin{equation}
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
+\end{equation}
+
+Finalement si l'on ne dispose ni des probabilités de transition ni d'un simulateur, alors il est possible d'utiliser LSTD$\mu$ (\citep{klein2011batch}), un algorithme permettant l'estimation en {\it batch} de la {\it feature expectation} de l'expert.
+\section{Résultats expérimentaux}
+\subsection{Description de la tâche}
+Le problème du pendule inversé à maintenir en équilibre vertical un pendule à un degré de liberté.
+il s'agit d'un problème classique en apprentissage par renforcement ({\it Reinforcement Learning, RL}).
+Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : \S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
+
+La politique ainsi trouvée parvient systématiquement à maintenir le pendule en équlibre durant 3000 pas de temps.
+Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
+
+Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seuls transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
+Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale).
+Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.\\
+
+En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship} car ceux-ci requièrent la résolution du problème direct à chaque itération ; résolution qui ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
+Comme notre algorithme ne nécessite pas la résolution du problème direct, il ne rencontre pas ce problème.
+\subsection{Illustration des résultats}
 \begin{figure}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/true_reward.pdf}
-       \caption{Expert's reward}
-       \label{nbabo}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_true_R.pdf}
+       \caption{Récompense sur laquelle l'expert a été entraîné}
+       \label{trueR.fig}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/retrieved_reward.pdf}
-       \caption{Reward found by LAFEM}
-       \label{croissnbabo}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_lafem_R.pdf}
+       \caption{Récompense trouvée par LAFEM}
+       \label{lafemR.fig}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/V_expert.pdf}
-       \caption{Expert's reward}
-       \label{nbabo}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_Vexpert.pdf}
+       \caption{Fonction de valeur de l'expert}
+       \label{trueV.fig}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/V_agent.pdf}
-       \caption{Agent's value function}
-       \label{croissnbabo}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_Vagent.pdf}
+       \caption{Fonction de valeur de l'agent entraîné sur la récompense trouvée par LAFEM}
+       \label{lafemV.fig}
     \end{center}
 \end{minipage}\\
+\end{figure}
+Afin d'étudier le comportement de notre algorithmes, nous avons établi le protocole expérimental suivant :
+\begin{itemize}
+ \item Créer une base de données $D_0$ de transitions aléatoires.
+ \item Entraîner un expert en utilisant LSPI nourri avec ces transitions ( definissant ainsi $\pi_E : S\rightarrow A$)
+ \item Vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps
+ \item Créer une base de données $D_E$ de transitions de l'expert, contenant 10 trajectoires de 300 transitions chacune.
+ \item Définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$.
+ \item Définir $\alpha(t) = 0.1,\forall t$.
+ \item Définir $D$ à partir des transitions $D_E$ en enlevant l'état d'arrivée.
+ \item Initialiser $\omega_0 = [-1...-1]^T$.
+ \item Fixer $T=20$.
+ \item Définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E)$.
+ \item Faire tourner LAFEM.
+ \item Entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$.
+ \item Vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
+ \item Afficher la fonction de valeur de l'expert.
+ \item Afficher la fonction de valeur de l'agent.
+ \item Afficher la fonction de récompense sur laquelle l'expert a été entraîné
+ \item Afficher la fonction de récompense retrouvée par LAFEM
+\end{itemize}
 
-\end{figure}
+Si la fonction de récompense trouvée par LAFEM (figure \ref{lafemR.fig}) diffère de celle fournie à l'expert (figure \ref{trueR.fig}) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{trueV.fig} et \ref{lafemV.fig}).
+Cela est une illustration du fait que le problème de l'IRL est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
 
-\subsection{Inverted Pendulum}
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.
+
+\subsection{Etude statistique de la qualité de l'imitation}
+
+Nous souhaitons maintenant empiriquement étudier l'impact de la quantité de données fournies par l'expert sur la qualité du contrôle exhibé par l'agent.
+Pour cela nous appliquons le protocole suivant :
 \begin{itemize}
-\item Présentation du setting (Le pendule, l'expert)
-\item Réglages de LAFEM (Préciser l'utilisation de LSTD$\mu$)
-\item Présentation des résultats
-\item Discussion sur la variabilité de la récompense
-\item Présentation des performances statistiques (expérience pas encore effectuée)/ Etude de la complexité en échantillons 
-\item Pas de comparaison avec le travail de \cite{abbeel2004apprenticeship} car celui-ci ne marche tout simplement pas avec les informations dont on dispose
+ \item Créer une base de données $D_0$ de transitions aléatoires.
+ \item Entraîner un expert en utilisant LSPI nourri avec ces transitions (ce qui définit $\pi_E : S\rightarrow A$)
+ \item Vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps
+ \item Créer une base de données $D_E$ de transitions de l'expert. On la crée de 10 trajectoires de 300 transitions chacune.
+ \item Définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$.
+ \item Définir $\alpha(t) = 0.1,\forall t$ (pifomètre)
+ \item Fixer $T=20$ (pifomètre)
+ \item Pour différentes quantités de données $N$,
+\begin{itemize}
+   \item Définir $D_E^N$ en tronquant $D_E$ à ses $N$ premières transitions
+   \item Définir $D$ à partir des de $D_E^N$ en enlevant l'état d'arrivée
+   \item Définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E^N)$
+   \item Initialiser $\omega_0 = [-1...-1]^T$
+   \item Faire tourner LAFEM
+   \item Entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$
+   \item Regarder durant combien de pas de temps l'agent arrrive à faire tenir le pendule.
+   \item Tracer la courbe donnant la qualité du contrôle (nombre de pas de temps durant lequel l'agent balance le pendule) en fonction de la quantitié de données disponible ($N$)
 \end{itemize}
+\end{itemize}
+
+Cette expérience a été répétée FIXME:nombre de fois, avec les résultats illutrés FIXME:mettre une figure.
+
 \section{Related Work}

### 2012-02-07 klein2012/main.tex
Expérience avec les barres d'erreur

--- 2012-02-02 klein2012/main.tex
@@
 \end{algorithm}
-\section{Experimental results}
-\subsection{GridWorld}
-\begin{itemize}
-\item Description du setting (GridWorld, expert, nombre de trajectoires, etc.)
-\item Description des réglages de LAFEM (fonction $l$ notamment, calcul exact de $\mu_E$)
-\item Présentation des résultats (pas grand chose à dire)
-\end{itemize}
+\section{Résultats expérimentaux}
+
 \begin{figure}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/true_reward.pdf}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_true_R.pdf}
        \caption{Expert's reward}
-       \label{nbabo}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/retrieved_reward.pdf}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_lafem_R.pdf}
        \caption{Reward found by LAFEM}
-       \label{croissnbabo}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/V_expert.pdf}
-       \caption{Expert's reward}
-       \label{nbabo}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_Vexpert.pdf}
+       \caption{Expert's value function}
     \end{center}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../GridWorld/V_agent.pdf}
+       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_Vagent.pdf}
        \caption{Agent's value function}
-       \label{croissnbabo}
     \end{center}
@@
 
-\subsection{Inverted Pendulum}
 \begin{itemize}

### 2012-02-08 klein2012/main.tex
L'article klein2012 est self content
Récupération de la section Background et prise en compte des commentaires d'Olivier sur celle-ci

--- 2012-02-06 klein2012/main.tex
@@
 \begin{abstract}
-  Cette publication s'occupe du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse ({\it Inverse Reinforcement Learning, IRL}).
+  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse ({\it Inverse Reinforcement Learning, IRL}).
 Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
 L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité et cherche à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
-Les algorithmes d'IRL existant nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition et la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée.
-Notre contribution est la présentation d'un nouvel algorithme d'IRL levant toutes ces contraintes.
-En utilisant méthode supervisée dans laquelle nous introduisons la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient et possèdant une faible complexité tant en échantillons que calculatoire.
+Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée et l'accès à la strategie complète de l'expert.
+Notre contribution consiste en un nouvel algorithme d'IRL levant toutes ces contraintes.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient et possèdant une faible complexité tant en échantillons que calculatoire.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
 \end{abstract}
+%%%%%
 \section{Introduction}
+%%%%%
 \begin{itemize}
 \item Introduction informelle sur les processus de décision séquentiels
-\item Intérêt de l'Apprenticeship learning
+\item Intérêt de l'apprentissage par démonstration
 \item Intérêt de l'IRL pour faire de l'apprenticeship learning
@@
 \end{itemize}
+%%%%%
 \section{Background}
+%%%%%
 \label{back.sec}
-\begin{itemize}
-\item Introduction rapide, pour les notations, des MDPs, du RL et de l'IRL
-\end{itemize}
+%%%
+\subsection{Principe}
+%%%
+Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielle.
+La configuration du système à contrôler est alors décrite à chaque instant $t$ par un état $s_t \in S$.
+Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
+Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité $p(s_{t+1}|s_t, a_t)$.
+Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décision séquentielle.\\
+
+Dans le cadre plus restreint de l'apprentissage par renforcement, pour guider la résolution de ce problème il faut fournir à l'agent une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui lui apprendra le degré auquel l'état dans lequel il se trouve est désirable.
+Une des forces de l'apprentissage par renforcement réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte.
+On définit pour ce faire la fonction de valeur $V^\pi$ :
+\begin{eqnarray}
+\label{Vdef.eqn}
+V^\pi(s) &=& E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,\pi\right]\\
+&=& R(s) + \gamma\sum_{s'}p(s'|s,\pi(s))V^\pi(s')
+\end{eqnarray}
+où $\gamma$ est un facteur d'actualisation strictement plus petit que $1$.
+Une politique optimale est alors définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie :
+\begin{equation}
+\forall \pi, \forall s, V^*(s) \geq V^\pi(s).
+\end{equation}
+L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov ({\it Markov Decision Process, MDP}).\\
+
+Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive.
+Un exemple de ce type de tâche est la conduite d'une voiture.
+Nous accompplissons ce genre de choses au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
+Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
+
+C'est la définition de l'apprentissage par renforcement inverse.
+Le problème est de retrouver la fonction de récompense optimisée par un expert.
+Usuellement l'expert est considéré comme un agent optimal dans un MDP et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
+
+Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour lequel ce comportement est optimal.
+Particulièrement, la récompense uniformément nulle remplit ce critère dans la mesure où tout comportement est optimal vis à vis de cette récompense.
+%%%
+\subsection{Considérations techniques}
+%%%
+La recherche de la politique optimale en apprentissage par renforcement se fait souvent en utilisant une fonction voisine de la fonction de valeur, la fonction de qualité définie par :
+\begin{eqnarray}
+Q^\pi(s,a) &=& E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,a_0=a,\pi\right]\\
+&=& R(s) + \gamma\sum_{s'}p(s'|s,a)V^\pi(s').
+\end{eqnarray}
+La politique optimale se définit simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
+\begin{equation}
+\pi^*(s) \in \arg\max_a Q^*(s,a).
+\end{equation}
+Dès lors, l'approximation de la fonction de valeur (terme qui employé au sens large désigne aussi la fonction de qualité) est un thème central en apprentissage par renforcement.
+Une caractéristique répandue dans la litérature est l'utilisation d'un cadre d'approximation linéaire où, après avoir défini des attributs vectoriels $\phi: S\times A \rightarrow \mathbb{R}^k$, l'on va chercher un vecteur de paramètres $\omega$ tel que l'on ait une bonne approximation de la fonction de qualité :
+\begin{equation}
+\hat Q (s,a) = \omega^T\phi(s,a)\approx Q(s,a).
+\end{equation}
+Il est naturel d'adopter le même type d'approche pour l'apprentissage par renforcement inverse, c'est à dire de considérer que l'inconnue du problème, la récompense $R$, peut être bien approximée grâce à un vecteur de poids $\theta$ et des attributs vectoriels $\psi : S\rightarrow \mathbb{R}^p$ :
+\begin{equation}
+\hat R(s) = \theta^T\psi(s)\approx R(s).
+\end{equation}
+
+Introduire cette expression dans la définition de la fonction de valeur (Equation \ref{Vdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+\begin{eqnarray}
+V^\pi(s) &=& E\left[\left.\sum_t\gamma^t\theta^T\psi(s_t)\right|s_0=s,\pi\right]\\
+V^\pi(s) &=& \theta^TE\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0=s,\pi\right]\\
+V^\pi(s) &=& \theta^T\mu^\pi(s)
+\end{eqnarray}
+
+Le terme $\mu^\pi$ est appelé la {\it feature expectation} d'une politique.
+Il existe plusieurs variantes basées sur la même définition que nous venons de donner.
+Dans sa version la plus spécifique, notée simplement $\mu^\pi$, il s'agit en toute rigueur de l'expression :
+\begin{eqnarray}
+\mu^\pi = E\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0\sim d,\pi\right],
+\end{eqnarray}
+où $d$ est une distribution sur l'état initial.
+Il peut également être utile de le définir sous la forme $\mu^\pi(s,a)$ où son rôle par rapport à $\mu^\pi(s)$ est le même que celui de $Q$ par rapport à $V$ :
+\begin{eqnarray}
+\mu^\pi(s,a) = E\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0=s,a_0=a,\pi\right].
+\end{eqnarray}
+On peut noter que si deux politiques ont la même {\it feature expectation}, alors ces deux politiques ont la même valeur vis à vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
+En effet
+\begin{eqnarray}
+\mu^{\pi_1} = \mu^{\pi_2} &\Rightarrow& \theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2}\\
+& \Rightarrow& V^{\pi_1} = V^{\pi_2}.
+\end{eqnarray}
+La {\it feature expectation} de l'expert, $\mu_E$, pouvant être calculée sans forcément avoir accès à l'intégralité de la politique $\pi_E$, certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les {\it feature expectations} respectives de l'agent et de l'expert, afin d'obtenir un agent dont le comportement à la même valeur que celui de l'expert.
+%%%%%
 \section{Loss-augmented Feature Expectation Matching}
+%%%%%
+%%%
 \subsection{Principe}
+%%%
 Notre algorithme cherche à minimiser la fonction de risque suivante :
@@
 \end{algorithm}
+%%%%%
+\section{Résultats expérimentaux}
+%%%%%
+%%%
 \subsection{Calcul de $\mu_E$}
+%%%
 L'algorithme présenté repose sur le calcul de la {\it feature expectation} de l'expert, $\mu_E$. Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour la calculer.
@@
 Finalement si l'on ne dispose ni des probabilités de transition ni d'un simulateur, alors il est possible d'utiliser LSTD$\mu$ (\citep{klein2011batch}), un algorithme permettant l'estimation en {\it batch} de la {\it feature expectation} de l'expert.
-\section{Résultats expérimentaux}
+%%%
 \subsection{Description de la tâche}
+%%%
 Le problème du pendule inversé à maintenir en équilibre vertical un pendule à un degré de liberté.
@@
 Comme notre algorithme ne nécessite pas la résolution du problème direct, il ne rencontre pas ce problème.
+%%%
 \subsection{Illustration des résultats}
+%%%
 \begin{figure}
@@
 
+%%%
 \subsection{Etude statistique de la qualité de l'imitation}
-
+%%%
 Nous souhaitons maintenant empiriquement étudier l'impact de la quantité de données fournies par l'expert sur la qualité du contrôle exhibé par l'agent.
@@
 Cette expérience a été répétée FIXME:nombre de fois, avec les résultats illutrés FIXME:mettre une figure.
-
+%%%%%
 \section{Related Work}
+%%%%%
 \begin{itemize}
@@
 %
-\bibliography{../../Biblio/Biblio}
+\bibliography{Biblio}
 

### 2012-02-08 klein2012/main.tex
Pour que ça marche sous win

@@
 \usepackage{amssymb}
-
 \usepackage{natbib}
-\usepackage{algorithm}
-\usepackage{algorithmic}
 %%% francisation des algorithmes
-\renewcommand{\algorithmicrequire} {\textbf{\textsc{Entrées:}}}
-\renewcommand{\algorithmicensure}  {\textbf{\textsc{Sorties:}}}
-\renewcommand{\algorithmicwhile}   {\textbf{tantque}}
-\renewcommand{\algorithmicdo}      {\textbf{faire}}
-\renewcommand{\algorithmicendwhile}{\textbf{fin tantque}}
-\renewcommand{\algorithmicend}     {\textbf{fin}}
-\renewcommand{\algorithmicif}      {\textbf{si}}
-\renewcommand{\algorithmicendif}   {\textbf{finsi}}
-\renewcommand{\algorithmicelse}    {\textbf{sinon}}
-\renewcommand{\algorithmicthen}    {\textbf{alors}}
-\renewcommand{\algorithmicfor}     {\textbf{pour}}
-\renewcommand{\algorithmicforall}  {\textbf{pour tout}}
-\renewcommand{\algorithmicdo}      {\textbf{faire}}
-\renewcommand{\algorithmicendfor}  {\textbf{fin pour}}
-\renewcommand{\algorithmicloop}    {\textbf{boucler}}
-\renewcommand{\algorithmicendloop} {\textbf{fin boucle}}
-\renewcommand{\algorithmicrepeat}  {\textbf{répéter}}
-\renewcommand{\algorithmicuntil}   {\textbf{jusqu'à}}
-
-\floatname{algorithm}{Algorithme}
 
@@
 
-\begin{algorithm}
-\caption{LAFEM ({\it Loss-Augmented Feature Expectation Matching})}
-\label{LAFEM.alg}
-\begin{algorithmic}
-  \REQUIRE une fonction de perte $l : S\times A \rightarrow \mathbb R$
-  \REQUIRE la \emph{feature expectation} de l'expert $\mu_E : S\times A \rightarrow \mathbb R ^p$
-  \REQUIRE un pas d'apprentissage $\alpha : \mathbb N \rightarrow \mathbb R$
-  \REQUIRE des données provenant de l'expert $\{(s_i,a_i)\}_{i=1..N}$
-  \REQUIRE un vecteur de paramètres initial $\theta_0$
-  \REQUIRE un nombre maximum d'itérations $T\in \mathbb N$
-  \REQUIRE une description de l'espace d'action $A$
-
-  \FOR {T iterations}
-  \STATE $\Delta\theta \leftarrow [0...0]^T$
-    \FORALL{ $(s_i,a_i)$}
-    \STATE $a^* \leftarrow \arg\max\limits_{a} = \theta_t ^T \mu_E(s_i,a) + l(s_i,a)$
-    \STATE $\Delta\theta \leftarrow \Delta\theta + \mu_E(s_i,a^*) - \mu_E(s_i,a_i)$
-    \ENDFOR
-    \STATE $\theta_{t+1} = \theta_t -\alpha_t{1\over N}\Delta\theta$
-    \ENDFOR
-    \RETURN $\theta_T$
-\end{algorithmic}
-\end{algorithm}
 %%%%%
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_true_R.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_true_R.pdf}
        \caption{Récompense sur laquelle l'expert a été entraîné}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_lafem_R.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_lafem_R.pdf}
        \caption{Récompense trouvée par LAFEM}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_Vexpert.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_Vexpert.pdf}
        \caption{Fonction de valeur de l'expert}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{../InvertedPendulum/LAFEM_Exp3_Vagent.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_Vagent.pdf}
        \caption{Fonction de valeur de l'agent entraîné sur la récompense trouvée par LAFEM}

### 2012-02-08 klein2012/main.tex
Réécriture de l'explication de l'algo
Il faut cependant que je relise les remarques de M&O et que je les intègre.

@@
 \begin{equation}
+\label{greedy.eqn}
 \pi^*(s) \in \arg\max_a Q^*(s,a).
@@
 \begin{equation}
+\label{hatRdef.eqn}
 \hat R(s) = \theta^T\psi(s)\approx R(s).
@@
 V^\pi(s) &=& \theta^T\mu^\pi(s)
+\label{Vmu.eqn}
 \end{eqnarray}
@@
 \end{eqnarray}
+C'est d'ailleurs cette variante de l'expression qui sera utilisée si, au lieu d'injecter l'équation \ref{hatRdef.eqn} dans la définition de la fonction de valeur comme nous l'avons fait pour obtenir l'équation \ref{Vmu.eqn}, nous l'injectons dans la définition de la fonction de qualité.
+On obtient alors :
+\begin{eqnarray}
+Q^\pi(s,a) &=& E\left[\left.\sum_t\gamma^t\theta^T\psi(s_t)\right|s_0=s,a_0=a,\pi\right]\\
+ &=& \theta^TE\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0=s,a_0=a,\pi\right]\\
+ &=& \theta^T\mu^\pi(s,a)
+\label{Qmu.eqn}
+\end{eqnarray}
+
 On peut noter que si deux politiques ont la même {\it feature expectation}, alors ces deux politiques ont la même valeur vis à vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
@@
 %%%
-Notre algorithme cherche à minimiser la fonction de risque suivante :
+L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis à vis d'une certaine fonction de récompense que l'on cherche à retrouver.
+Cela implique, comme nous l'avons vu Equation \ref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
 \begin{equation}
-  R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right)
+\pi_E(s) \in \arg\max_aQ^E_R(s,a).
 \end{equation}
-Dans le travail de \citep{ratliff2007imitation}, cette fonction de risque est introduite pour décrire un classifieur utilisant la fonction de score $q$ et qui à un élément $s$ associe la label $a = \arg\max\limits_a q(s,a)$.
-Cela est très similaire à l'équation
+Nous allons utiliser cette propriété pour imaginer une fonction de risque dépendante d'une fonction $q$ qui devra partager avec la fonction inconnue $Q^E$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+Notons $(s_i,a_i)_i$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
 \begin{equation}
-  \label{greedy.eqn}
-  \pi^*(s) = \arg\max_{a} Q^*(s,a)
+\label{butLAFEM.eqn}
+a_i \in \arg\max_aq(s_i,a).
 \end{equation}
-  qui décrit le mécanisme glouton liant fonction de valeur optimale et politique optimale.
-La politique de l'expert étant une politique optimale, elle doit respecter cette équation.
-Minimiser notre fonction de risque revient donc à chercher une fonction $q$ compatible avec la fonction de valeur optimale de l'expert.
-En effet, considérons que la fonction de perte $l$ est uniformément nulle et regardons ce qui se passe lorsque la fonction de risque $R_N(q)$ est minimisée.
-On introduit $a^* = \arg\max\limits_a(q(s_i,a))$.
-Si la fonction de risque $R_N(q)$ est positive, alors il arrive pour certains $i$ que $q(s_i,a^*) > q(s_i,a_i)$.
-Cela signifie que la fonction $q$ ne permet pas de justifier le choix de l'expert.
-A l'opposé, si l'on parvient à anuller $R_N(q)$, on a $\forall i, q(s_i,a^*) = q(s_i,a_i)$, et donc $a_i \in \arg\max\limits_a q(s_i,a)$, ce qui justifie le choix de l'expert.\\
-
-   La fonction de perte $l$ permet d'introduire de la connaissance à priori en laissant l'opérateur influer sur la forme de la fonction $q$ recherchée.\\
-
-   Nous allons profiter de la structure présente dans $Q$ pour introduire la notion de \emph{feature expectation} dans le problème.
-On rappelle qu'en IRL il est classique d'écrire $Q(s,a) = \theta^T \mu(s,a)$.
-Nous nous basons donc sur la \emph{feature expectation} de l'expert, $\mu_E$ et un vecteur de paramètres $\theta$ contrôlant la fonction de récompense que l'on recherche pour réécrire la fonction de risque que nous allons minimiser :
+Notons 
 \begin{equation}
-   R_N(\theta) = {1\over N} \sum_{i=1}^N\left(\max_{a}(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
+a^*_i = \arg\max_aq(s_i,a),
 \end{equation}
-   Il faut noter ici l'utilisation non pas de $\mu_E : S\rightarrow \mathbb{R}^p$ mais de $\mu_E : S\times A \rightarrow \mathbb{R}^k$.
-Cette dernière partage avec la première la même relation que $Q$ partage avec $V$, c'est à dire que l'on dispose d'un degré de liberté en plus correspondant à la prochaine action, qui peut différer de celle choisie par l'expert.
-
-   De la même manière que précedemment, on note $a^* = \arg\max\limits_{a} = \theta_t ^T \mu_E(s_i,a) + l(s_i,a)$, l'expression de la fonction de risque devient alors :
+alors le critère exprimé Equation \ref{butLAFEM.eqn} s'écrit $\forall i, q(s_i,a^*_i) = q(s_i,a_i)$.
+Notre fonction de risque doit pénaliser les cas où $q(s_i,a^*_i)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \ref{greedy.eqn}) le choix $a_i$ de l'expert.\\
+Nous introduisons de surcroit une fonction de perte $l : S \times A \rightarrow \mathbb{R}_+$ dans la définition de $a^*_i$ qui devient maintenant :
+\begin{equation}
+\label{astar.def}
+a^*_i = \arg\max_aq(s_i,a) + l(s_i,a).
+\end{equation}
+La définition de notre fonction de risque peut alors être assemblée en :
   \begin{eqnarray}
-   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*) + l(s_i,a^*) - \theta ^T \mu_E(s_i,a_i) \right)\\
-   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*) - \mu_E(s_i,a_i)\right) + l(s_i,a^*)  \right).
+   R_N(q) &=& {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right)\\
+   &=& {1\over N} \sum_{i=1}^N\left(q(s_i,a_i^*) + l(s_i,a_i^*) - q(s_i,a_i) \right)
    \end{eqnarray}
 
-   La présence d'un $\max$ dans cette expression (caché dans le $a^*$) nous oblige à utiliser le sous gradient dont la règle est que le sous gradient du $\max$ est le gradient de l'$\arg\max$.
+Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance à priori dans le système.
+En effet, minimiser avec succès la fonction de risque avec un choix de $l$ judicieux permet d'imposer à $q$ une structure telle qu'il existe une marge entre l'action choisie par l'expert et la deuxième meilleure action.
+Ce type d'argument est assez commun dans la litérature.
+Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères fournis en sous-section \ref{proof.sec}.\\
 
-   Le sous gradient de cette expression est donc :
+Comme nous l'avons vu Equation \ref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
+L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q = \theta^T\mu_E(s,a)$.
+Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
+Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \ref{hatRdef.eqn}).
+   \begin{eqnarray}
+   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
+   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right) + l(s_i,a^*_i)  \right).
+   \end{eqnarray}
+   Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en sous-section \ref{proof.sec} :
    \begin{equation}
-   \nabla L_N(\theta) =  {1\over N}\sum_{i=1}^N\left(\mu_E(s_i,a^*) - \mu_E(s_i,a_i)\right).
+   \theta_{t+1} = \theta_t -\alpha_t{{1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\over ||{1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)||_2}.
    \end{equation}
-
-   Une descente de gradient sur le risque donne la règle d'update suivante :
-   \begin{equation}
-   \theta_{t+1} = \theta_t -\alpha_t{1\over N}\sum_{i=1}^N\left(\mu_E(s_i,a^*) - \mu_E(s_i,a_i)\right),
-   \end{equation}
-
-   avec $\alpha_t$ un pas d'apprentissage. L'algorithme complet est détaillé algorithme \ref{LAFEM.alg}.
-
 %%%%%
 \section{Résultats expérimentaux}
+\label{exp.sec}
 %%%%%
@@
 %%%
+\label{calculmu.sec}
 L'algorithme présenté repose sur le calcul de la {\it feature expectation} de l'expert, $\mu_E$. Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour la calculer.

### 2012-02-08 klein2012/main.tex
Pour pouvoir puller

--- 2012-02-08 klein2012/main.tex
@@
 \newtheorem{definition}{Définition}
+\newtheorem{lemme}{Lemme}
+\newtheorem{theorem}{Théorème}
+\newtheorem{hypo}{Hypothèse}
+\newtheorem{cor}{Corollaire}
+\newtheorem{prop}{Proposition}
+\newtheorem{exemple}{Exemple}
+\newtheorem{remarques}{Remarques}
+\newtheorem{remarque}{Remarque}
+\newtheorem{problem}{Problème}
+\newtheorem{algo}{Algorithme}
+\newtheorem{preuve}{Démonstration}
+\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
+\newcommand{\argmin}{\operatorname*{argmin}}
+\newcommand{\minp}{\operatorname*{min_+}}
+\newcommand{\dom}{\operatorname*{dom}}
+\newcommand{\Ker}{\operatorname*{Ker}}
+\newcommand{\trace}{\operatorname*{trace}}
+\newcommand{\cov}{\operatorname{cov}}
+\newcommand{\epi}{\operatorname{epi}}
+\newcommand{\card}{\operatorname*{Card}}
+\newcommand{\conv}{\operatorname*{Conv}}
+\newcommand{\vect}{\operatorname*{Vect}}
+\newcommand{\var}{\operatorname{Var}}
+\newcommand{\diag}{\operatorname{diag}}
+\newcommand{\erf}{\operatorname{erf}}
+\newcommand{\bound}{\operatorname*{bound}}
+\newcommand{\vpi}{\operatorname{VPI}}
+\newcommand{\gn}{\operatorname{Gain}}
+\newcommand{\p}{\operatorname{Pr}}
+\newcommand{\mlp}{\operatorname{MLP}}
+\newcommand*\tto[2]{\smash{\mathop{\longrightarrow}\limits_{#1}^{#2}}}
+\newcommand*\ntto[2]{\smash{\mathop{\nrightarrow}\limits_{#1}^{#2}}}
+\newcommand{\X}{\mathbf{X}}
+\newcommand{\Q}{\mathbf{Q}}
+\newcommand{\A}{\mathbf{A}}
+\newcommand{\Z}{\mathbf{Z}}
+\newcommand{\Y}{\mathbf{Y}}
+\newcommand{\E}{\mathbf{E}}
+\newcommand{\K}{\mathbf{K}}
+\newcommand{\F}{\mathcal{F}}
+\newcommand{\R}{\mathbf{R}}
+\newcommand{\ba}{\mathbf{a}}
+\newcommand{\bb}{\mathbf{b}}
+\newcommand{\bc}{\mathbf{c}}
+\newcommand{\bd}{\mathbf{d}}
+\newcommand{\be}{\mathbf{e}}
+\newcommand{\af}{\mathbf{f}}
+\newcommand{\bg}{\mathbf{g}}
+\newcommand{\bh}{\mathbf{h}}
+\newcommand{\bi}{\mathbf{i}}
+\newcommand{\bj}{\mathbf{j}}
+\newcommand{\bk}{\mathbf{k}}
+\newcommand{\bl}{\mathbf{l}}
+\newcommand{\bm}{\mathbf{m}}
+\newcommand{\bn}{\mathbf{n}}
+\newcommand{\bo}{\mathbf{o}}
+\newcommand{\bp}{\mathbf{p}}
+\newcommand{\bq}{\mathbf{q}}
+\newcommand{\br}{\mathbf{r}}
+\newcommand{\bs}{\mathbf{s}}
+\newcommand{\bt}{\mathbf{t}}
+\newcommand{\bu}{\mathbf{u}}
+\newcommand{\bv}{\mathbf{v}}
+\newcommand{\bw}{\mathbf{w}}
+\newcommand{\bx}{\mathbf{x}}
+\newcommand{\by}{\mathbf{y}}
+\newcommand{\bz}{\mathbf{z}}
+\newcommand{\ma}{\mathbf{A}}
+\newcommand{\mb}{\mathbf{B}}
+\newcommand{\mc}{\mathbf{C}}
+\newcommand{\md}{\mathbf{D}}
+\newcommand{\me}{\mathbf{E}}
+\newcommand{\mf}{\mathbf{F}}
+\newcommand{\mg}{\mathbf{G}}
+\newcommand{\mh}{\mathbf{H}}
+\newcommand{\mi}{\mathbf{I}}
+\newcommand{\mj}{\mathbf{J}}
+\newcommand{\mk}{\mathbf{K}}
+\newcommand{\ml}{\mathbf{L}}
+\newcommand{\mm}{\mathbf{M}}
+\newcommand{\mn}{\mathbf{N}}
+\newcommand{\mo}{\mathbf{O}}
+\newcommand{\Mp}{\mathbf{P}}
+\newcommand{\mq}{\mathbf{Q}}
+\newcommand{\mr}{\mathbf{R}}
+\newcommand{\ms}{\mathbf{S}}
+\newcommand{\mt}{\mathbf{T}}
+\newcommand{\Mu}{\mathbf{U}}
+\newcommand{\mv}{\mathbf{V}}
+\newcommand{\mw}{\mathbf{W}}
+\newcommand{\mx}{\mathbf{X}}
+\newcommand{\my}{\mathbf{Y}}
+\newcommand{\mz}{\mathbf{Z}}
+\newcommand{\tphi}{\tilde{\Phi}}
+\newcommand{\espace}{\text{ }}
+\newcommand{\x}{\mathbf{x}}
+\newcommand{\s}{\mathbf{s}}
+\newcommand{\n}{\mathbf{n}}
+\newcommand{\y}{\mathbf{y}}
+\newcommand{\I}{\mathbf{I}}
+\newcommand{\rr}{\mathbf{r}}
 
@@
 %%%%%
+\section{Preuves}\label{proof.sec}
+\subsection{Convergence de l'algorithme de sous-gradient sous réserve de l'existence d'une solution}
+\begin{definition}[Fonction convexe]
+Soient $\me$ un espace vectoriel et $\overline{\mathbb{R}}=\mathbb{R}\cup\{+\infty,-\infty\}$ la droite achevée.
+On dira qu'une fonction $\af$ de $\me$ vers $\overline{\mathbb{R}}$ est une fonction convexe si pour tout couple $(\bx,\by)\in(\dom\af)\times(\dom\af)$ et pour tout $\bt\in]0,1[$ :
+\begin{equation}
+\af((1-\bt)\bx+\bt\by)\leq(1-\bt)\af(\bx)+\bt\af(\by),
+\end{equation}
+où $\dom\af=\{\bx\in\me:\af(\bx)<+\infty\}$.
+On définit aussi l'épigraphe de $\af$ : $\epi \af=\{(\bx,\alpha)\in\me\times\mathbb{R}:\af(\bx)\leq\alpha\}$. Si ce dernier est fermé alors on dira que $\af$ est convexe fermée. 
+\end{definition}
+\begin{definition}[Sous-gradient d'une fonction convexe]
+Soit $\af$ une fonction convexe de $\me$ vers $\mathbb{R}\cup\{+\infty\}$ tel que $\dom\af$ non vide.
+On dira alors que $\af$ est une fonction convexe propre.
+On note $\conv(\me)$ l'ensemble des fonctions convexes propres.
+Lorsque $\me$ est un espace euclidien muni du produit scalaire $<,>$, on dit que la fonction $\af$ est sous-différentiable en $\bx\in\dom\af$ s'il existe un $\bx^*\in\me$ tel que:
+\begin{equation}
+\forall \by\in\me : \af(\by)\geq\af(\bx)+<\bx^*,\by-\bx>.
+\end{equation}
+On dira alors que $\bx^*$ est un sous-gradient de $\af$ en $\bx$ et on note $\partial_\bx\af$ l'ensemble des sous-gradients de $\af$ en $\bx$.   
+\end{definition}
+\begin{theorem}[Théorème issu de l'article \cite{correa1993convergence}]
+Soit $\af$ une fonction convexe, propre et fermée de $\me$ vers $\mathbb{R}\cup\{+\infty\}$.
+On suppose que $\me$ est un espace euclidien muni du produit scalaire $<,>$ et de la norme $\|.\|_2$ associée et que $\af$ prend des valeurs finies.
+Alors si la fonction admet un minimum sur $\me$, la suite définie par :
+\begin{align}
+&\bx_0\in\me,
+\\
+&\forall \bk\in\mathbb{N} : \bx_{\bk+1}=\bx_\bk-\alpha_{\bk}\frac{\gamma_{\bk}}{\|\gamma_{\bk}\|_2},
+\end{align}
+telle que :
+\begin{align}
+&\sum_{\bk=0}^{\bk=+\infty}\alpha_{\bk}=+\infty,
+\\
+&\sum_{\bk=0}^{\bk=+\infty}\alpha_{\bk}^2<+\infty,
+\end{align}
+est convergente vers un point qui minimise $\af$, où $\gamma_{\bk}\in\partial_{\bx_\bk}\af$. De plus, on a $\liminf\af({\bx_\bk})=\min_{\bx\in\me}\af(\bx)$.
+\end{theorem}
+
+%%%%%
 \section{Résultats expérimentaux}

### 2012-02-08 klein2012/main.tex
Réécriture de la sous section concernant le calcul de LSTD$\mu$

--- 2012-02-08 klein2012/main.tex
@@
 &=& R(s) + \gamma\sum_{s'}p(s'|s,\pi(s))V^\pi(s')
+\label{BellmanV.eqn}
 \end{eqnarray}
@@
 \subsection{Considérations techniques}
+\label{ConsiderationsTechniques.sec}
 %%%
@@
 \begin{eqnarray}
+\label{mupidef.eqn}
 \mu^\pi = E\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0\sim d,\pi\right],
@@
 \end{equation}
-Nous allons utiliser cette propriété pour imaginer une fonction de risque dépendante d'une fonction $q$ qui devra partager avec la fonction inconnue $Q^E$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+Nous allons utiliser cette propriété pour imaginer une fonction de risque dépendante d'une fonction $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
 Notons $(s_i,a_i)_i$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
@@
 
-Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance à priori dans le système.
+Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système.
 En effet, minimiser avec succès la fonction de risque avec un choix de $l$ judicieux permet d'imposer à $q$ une structure telle qu'il existe une marge entre l'action choisie par l'expert et la deuxième meilleure action.
@@
 Comme nous l'avons vu Equation \ref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
-L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q = \theta^T\mu_E(s,a)$.
+L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
 Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
@@
    \end{equation}
+%%%
+\subsection{Calcul de $\mu_E$}
+%%%
+\label{calculmu.sec}
+L'algorithme présenté repose sur le calcul de la {\it feature expectation} de \mbox{l'expert}, $\mu_E$.
+Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de cette {\it feature expectation}.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, la {\it feature expectation} de l'expert et non le calcul de la {\it feature expectation} d'une politique arbitraire.
+De fait cela nous épargne le besoin d'un simulateur, d'un solveur de MDP ou d'échantillons couvrant tout l'espace d'état.\\
+
+Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour calculer $\mu_E$.\\
+
+Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \ref{mupidef.eqn}) et celle de la fonction de valeur (Equation \ref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
+En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \ref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
+\begin{equation}
+\mathbf V^\pi = \mathbf R + \gamma P_\pi\mathbf V^\pi
+\end{equation}
+Il est connu qu'une règle d'update de la forme 
+\begin{equation}
+\mathbf V^\pi_{t+1} \leftarrow \mathbf R + \gamma P_\pi\mathbf V^\pi_t
+\end{equation}
+convergera vers la valeur exacte recherchée.
+Ce sont les algorithmes du type \emph{value iteration} qui exploitent cette propriété.
+Parallèlement il est possible d'utiliser une règle d'update de la forme :
+\begin{equation}
+\mathbf \mu^\pi_{t+1} \leftarrow \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi_t
+\end{equation}
+pour obtenir la valeur exacte de la {\it feature expectation}.\\
+
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
+\begin{equation}
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
+\end{equation}
+
+Finalement, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ (\citep{klein2011batch}) dans sa version \emph{on-policy}.
+Cet algorithme se base également sur la similarité entre les definitions de $\mu$ et $Q$ mais au lieu d'adapter un algorithme de programmation dynamique comme nous venons d'en donner l'exemple, il adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch} et {\it offline}.
+Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ est obtenue grâce à la solution des moindres carrés :
+\begin{equation}
+  \xi_i^* = \left(\sum_{t=1}^n
+  \psi(s_t)(\psi(s_t)-\gamma\psi(s_{t+1}))^T\right)^{-1}
+  \sum_{t=1}^n \psi(s_t) \phi_i(s_t).
+\end{equation}
+
 %%%%%
@@
 %%%%%
-%%%
-\subsection{Calcul de $\mu_E$}
-%%%
-\label{calculmu.sec}
-L'algorithme présenté repose sur le calcul de la {\it feature expectation} de l'expert, $\mu_E$. Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour la calculer.
-
-La définition du vecteur $\mu_E$ étant proche de celle de la fonction de valeur, en adaptant les algorithmes de programmation dynamique destinés au calcul de la fonction de valeur on obtient une règle de la forme :
-\begin{equation}
-\mathbf \mu^\pi = \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi
-\end{equation}
-permettant le calcul exact de la fonction de valeur pour peu que l'on dispose des probabilités de transitions.\\
-
-Dans le cas plus probable où celles-ci sont inconnues, on peut si l'on dispose d'un simulateur utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
-\begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
-\end{equation}
-
-Finalement si l'on ne dispose ni des probabilités de transition ni d'un simulateur, alors il est possible d'utiliser LSTD$\mu$ (\citep{klein2011batch}), un algorithme permettant l'estimation en {\it batch} de la {\it feature expectation} de l'expert.
 %%%
@@
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : \S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
 

### 2012-02-08 klein2012/main.tex
début travail.

--- 2012-02-08 klein2012/main.tex
@@
 \begin{definition}[Fonction convexe]
-Soient $\me$ un espace vectoriel et $\overline{\mathbb{R}}=\mathbb{R}\cup\{+\infty,-\infty\}$ la droite achevée.
-On dira qu'une fonction $\af$ de $\me$ vers $\overline{\mathbb{R}}$ est une fonction convexe si pour tout couple $(\bx,\by)\in(\dom\af)\times(\dom\af)$ et pour tout $\bt\in]0,1[$ :
+Soient ~$\me$ un espace vectoriel et ~$\overline{\mathbb{R}}=\mathbb{R}\cup\{+\infty,-\infty\}$ la droite achevée.
+On dira qu'une fonction ~$\af$ de ~$\me$ vers ~$\overline{\mathbb{R}}$ est une fonction convexe si pour tout couple ~$(\bx,\by)\in(\dom\af)\times(\dom\af)$ et pour tout ~$\bt\in]0,1[$ :
 \begin{equation}
@@
 \end{equation}
-où $\dom\af=\{\bx\in\me:\af(\bx)<+\infty\}$.
-On définit aussi l'épigraphe de $\af$ : $\epi \af=\{(\bx,\alpha)\in\me\times\mathbb{R}:\af(\bx)\leq\alpha\}$. Si ce dernier est fermé alors on dira que $\af$ est convexe fermée. 
+où ~$\dom\af=\{\bx\in\me:\af(\bx)<+\infty\}$.
+On définit aussi l'épigraphe de ~$\af$ : ~$\epi \af=\{(\bx,\alpha)\in\me\times\mathbb{R}:\af(\bx)\leq\alpha\}$. Si ce dernier est fermé alors on dira que ~$\af$ est convexe fermée.
 \end{definition}
 \begin{definition}[Sous-gradient d'une fonction convexe]
-Soit $\af$ une fonction convexe de $\me$ vers $\mathbb{R}\cup\{+\infty\}$ tel que $\dom\af$ non vide.
-On dira alors que $\af$ est une fonction convexe propre.
-On note $\conv(\me)$ l'ensemble des fonctions convexes propres.
-Lorsque $\me$ est un espace euclidien muni du produit scalaire $<,>$, on dit que la fonction $\af$ est sous-différentiable en $\bx\in\dom\af$ s'il existe un $\bx^*\in\me$ tel que:
+Soit ~$\af$ une fonction convexe de ~$\me$ vers ~$\mathbb{R}\cup\{+\infty\}$ tel que ~$\dom\af$ non vide.
+On dira alors que ~$\af$ est une fonction convexe propre.
+On note ~$\conv(\me)$ l'ensemble des fonctions convexes propres.
+Lorsque ~$\me$ est un espace euclidien muni du produit scalaire $<,>$, on dit que la fonction ~$\af$ est sous-différentiable en ~$\bx\in\dom\af$ s'il existe un ~$\bx^*\in\me$ tel que:
 \begin{equation}
@@
 \end{equation}
-On dira alors que $\bx^*$ est un sous-gradient de $\af$ en $\bx$ et on note $\partial_\bx\af$ l'ensemble des sous-gradients de $\af$ en $\bx$.   
+On dira alors que ~$\bx^*$ est un sous-gradient de ~$\af$ en ~$\bx$ et on note ~$\partial_\bx\af$ l'ensemble des sous-gradients de ~$\af$ en ~$\bx$.
 \end{definition}
-\begin{theorem}[Théorème issu de l'article \cite{correa1993convergence}]
-Soit $\af$ une fonction convexe, propre et fermée de $\me$ vers $\mathbb{R}\cup\{+\infty\}$.
-On suppose que $\me$ est un espace euclidien muni du produit scalaire $<,>$ et de la norme $\|.\|_2$ associée et que $\af$ prend des valeurs finies.
-Alors si la fonction admet un minimum sur $\me$, la suite définie par :
+\begin{theorem}[Théorème issu de l'article \cite{correa1993convergence}]\label{theoreme.sec}
+Soit ~$\af$ une fonction convexe, propre et fermée de ~$\me$ vers ~$\mathbb{R}\cup\{+\infty\}$.
+On suppose que ~$\me$ est un espace euclidien muni du produit scalaire $<,>$ et de la norme ~$\|.\|_2$ associée et que ~$\af$ prend des valeurs finies.
+Alors si la fonction admet un minimum sur ~$\me$, la suite définie par :
 \begin{align}
@@
 \end{align}
-est convergente vers un point qui minimise $\af$, où $\gamma_{\bk}\in\partial_{\bx_\bk}\af$. De plus, on a $\liminf\af({\bx_\bk})=\min_{\bx\in\me}\af(\bx)$.
+est convergente vers un point qui minimise ~$\af$, où ~$\gamma_{\bk}\in\partial_{\bx_\bk}\af$. De plus, on a ~$\liminf\af({\bx_\bk})=\min_{\bx\in\me}\af(\bx)$.
 \end{theorem}
-
+\subsection{Application à notre fonction de risque}
+On rappelle que notre fonction de risque $R_N$ est une fonction de l'espace euclidien $\mathbb{R}^p$, qu'on munira du produit scalaire canonique $<.,.>$ et de sa norme associée $\|\|_2$, vers $\mathbb{R}$ où $p$ est la dimension de l'inconnue $\theta$.
+Pour $\theta\in\mathbb{R}^p$, on a :
+\begin{eqnarray}
+   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
+   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right) + l(s_i,a^*_i)  \right).
+\end{eqnarray}
+On peut montrer que $R_N$ est définie correctement de la manière suivante.
+On définit pour chaque $i\in\{1,\dots,N\}$ et pour chaque $a\in A$, où $A$ est un ensemble fini d'actions, les fonctions $f_{(i,a)}$ qui sont des fonctions de $\mathbb{R}^p$ vers $\mathbb{R}$ par :
+\begin{equation}
+f_{(i,a)}({\theta})=\theta ^T \mu_E(s_i,a) + l(s_i,a).
+\end{equation}
+Il est évident que $f_{i,a}$ est une fonction convexe, propre et fermée à valeurs finies.
+De plus elle est continue.
+On définit aussi pour chaque $i\in\{1,\dots,N\}$, les fonctions $g_{i}$ qui sont des fonctions de $\mathbb{R}^p$ vers $\mathbb{R}$ par :
+\begin{equation}
+g_{i}({\theta})=-\theta ^T \mu_E(s_i,a_i).
+\end{equation}
+Il est évident que $g_{i}$ est une fonction convexe, propre et fermée à valeurs finies. De plus elle est continue.
+Comme $\A$ est fini, alors on peut définir pour chaque $i\in\{1,\dots,N\}$, les fonctions $h_{(i,a)}$ qui sont des fonctions de $\mathbb{R}^p$ vers $\mathbb{R}$ par :
+\begin{equation}
+h_{i}({\theta})=\max_{a \in A}\theta ^T \mu_E(s_i,a) + l(s_i,a)=\max_{a \in A}f_{(i,a)}({\theta}).
+\end{equation}
+La fonction $g_{i}$ étant défini comme un maximum sur un ensemble fini de fonctions continues, convexes, propres et fermées à valeurs finies est encore continue, convexe, propre et fermée à valeurs finies.
+Finalement il vient que l'on peut écrire pour tout $\theta$ :
+\begin{equation}
+R_N(\theta) = {1\over N} \sum_{i=1}^N g_{i}({\theta})+h_{i}({\theta}).
+\end{equation}
+Ainsi $R_N$ est la somme finie de fonctions convexes, propres, fermées et continues à valeurs finies, elle aura donc ces propriétés et est définie correctement.
+On voit donc que la fonction $R_N$ est une bonne candidate à l'application du théorème \ref{theoreme.sec} sous réserve de l'existence d'un $\theta^*$ tel que :
+\begin{equation}
+R_N(\theta^*)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta).
+\end{equation}
+Nous allons montrer l'existence de ce $\theta^*$ dans un cas particulier.
 %%%%%

### 2012-02-08 klein2012/main.tex
Pour puller le boulot de bilal

--- 2012-02-08 klein2012/main.tex
@@
 %%%
-Le problème du pendule inversé à maintenir en équilibre vertical un pendule à un degré de liberté.
-il s'agit d'un problème classique en apprentissage par renforcement ({\it Reinforcement Learning, RL}).
+Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté.
+Il s'agit d'un problème classique en apprentissage par renforcement.
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
@@
        \includegraphics[width=\textwidth]{LAFEM_Exp3_lafem_R.pdf}
-       \caption{Récompense trouvée par LAFEM}
+       \caption{Récompense trouvée par LAFEM, en vert les positions occupées par l'expert.}
        \label{lafemR.fig}

### 2012-02-08 klein2012/main.tex

--- 2012-02-08 klein2012/main.tex
@@
    \end{equation}
+%%%
+\subsection{Calcul de $\mu_E$}
+%%%
+\label{calculmu.sec}
+Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de la {\it feature expectation}.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, la {\it feature expectation} de l'expert et non le calcul de la {\it feature expectation} d'une politique arbitraire.
+De fait cela nous épargne le besoin d'un simulateur, d'un solveur de MDP ou d'échantillons couvrant tout l'espace d'état.\\
+
+Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour calculer $\mu_E$.\\
+
+Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \ref{mupidef.eqn}) et celle de la fonction de valeur (Equation \ref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
+En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \ref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
+\begin{equation}
+\mathbf V^\pi = \mathbf R + \gamma P_\pi\mathbf V^\pi
+\end{equation}
+Il est connu qu'une règle d'update de la forme 
+\begin{equation}
+\mathbf V^\pi_{t+1} \leftarrow \mathbf R + \gamma P_\pi\mathbf V^\pi_t
+\end{equation}
+convergera vers la valeur exacte recherchée.
+Ce sont les algorithmes du type \emph{value iteration} qui exploitent cette propriété.
+Parallèlement il est possible d'utiliser une règle d'update de la forme :
+\begin{equation}
+\mathbf \mu^\pi_{t+1} \leftarrow \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi_t
+\end{equation}
+pour obtenir la valeur exacte de la {\it feature expectation}.\\
+
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
+\begin{equation}
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
+\end{equation}
+
+Finalement, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ (\citep{klein2011batch}) dans sa version \emph{on-policy}.
+Cet algorithme se base également sur la similarité entre les definitions de $\mu$ et $Q$ mais au lieu d'adapter un algorithme de programmation dynamique comme nous venons d'en donner l'exemple, il adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch} et {\it offline}.
+Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ est obtenue grâce à la solution des moindres carrés :
+\begin{equation}
+  \xi_i^* = \left(\sum_{t=1}^n
+  \psi(s_t)(\psi(s_t)-\gamma\psi(s_{t+1}))^T\right)^{-1}
+  \sum_{t=1}^n \psi(s_t) \phi_i(s_t).
+\end{equation}
+
 %%%%%
@@
 Nous allons montrer l'existence de ce $\theta^*$ dans un cas particulier.
+
 %%%%%
@@
 %%%
-\subsection{Calcul de $\mu_E$}
-%%%
-\label{calculmu.sec}
-L'algorithme présenté repose sur le calcul de la {\it feature expectation} de l'expert, $\mu_E$. Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour la calculer.
-
-La définition du vecteur $\mu_E$ étant proche de celle de la fonction de valeur, en adaptant les algorithmes de programmation dynamique destinés au calcul de la fonction de valeur on obtient une règle de la forme :
-\begin{equation}
-\mathbf \mu^\pi = \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi
-\end{equation}
-permettant le calcul exact de la fonction de valeur pour peu que l'on dispose des probabilités de transitions.\\
-
-Dans le cas plus probable où celles-ci sont inconnues, on peut si l'on dispose d'un simulateur utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
-\begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
-\end{equation}
-
-Finalement si l'on ne dispose ni des probabilités de transition ni d'un simulateur, alors il est possible d'utiliser LSTD$\mu$ (\citep{klein2011batch}), un algorithme permettant l'estimation en {\it batch} de la {\it feature expectation} de l'expert.
-%%%
 \subsection{Description de la tâche}
 %%%
-Le problème du pendule inversé à maintenir en équilibre vertical un pendule à un degré de liberté.
-il s'agit d'un problème classique en apprentissage par renforcement ({\it Reinforcement Learning, RL}).
+Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté.
+Il s'agit d'un problème classique en apprentissage par renforcement.
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : \S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
 

### 2012-02-08 klein2012/main.tex
Prise en compte de toutes les remarques d'Olivier

@@
    \end{eqnarray}
-   Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en sous-section \ref{proof.sec} :
+   Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en section \ref{proof.sec} :
    \begin{equation}
@@
  \item Vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
- \item Afficher la fonction de valeur de l'expert.
- \item Afficher la fonction de valeur de l'agent.
- \item Afficher la fonction de récompense sur laquelle l'expert a été entraîné
- \item Afficher la fonction de récompense retrouvée par LAFEM
 \end{itemize}
@@
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.\\
 
-%%%
-\subsection{Etude statistique de la qualité de l'imitation}
-%%%
-Nous souhaitons maintenant empiriquement étudier l'impact de la quantité de données fournies par l'expert sur la qualité du contrôle exhibé par l'agent.
-Pour cela nous appliquons le protocole suivant :
-\begin{itemize}
- \item Créer une base de données $D_0$ de transitions aléatoires.
- \item Entraîner un expert en utilisant LSPI nourri avec ces transitions (ce qui définit $\pi_E : S\rightarrow A$)
- \item Vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps
- \item Créer une base de données $D_E$ de transitions de l'expert. On la crée de 10 trajectoires de 300 transitions chacune.
- \item Définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$.
- \item Définir $\alpha(t) = 0.1,\forall t$ (pifomètre)
- \item Fixer $T=20$ (pifomètre)
- \item Pour différentes quantités de données $N$,
-\begin{itemize}
-   \item Définir $D_E^N$ en tronquant $D_E$ à ses $N$ premières transitions
-   \item Définir $D$ à partir des de $D_E^N$ en enlevant l'état d'arrivée
-   \item Définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E^N)$
-   \item Initialiser $\omega_0 = [-1...-1]^T$
-   \item Faire tourner LAFEM
-   \item Entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$
-   \item Regarder durant combien de pas de temps l'agent arrrive à faire tenir le pendule.
-   \item Tracer la courbe donnant la qualité du contrôle (nombre de pas de temps durant lequel l'agent balance le pendule) en fonction de la quantitié de données disponible ($N$)
-\end{itemize}
-\end{itemize}
+Deux éléments méritent d'être notés.
+Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
+Ceux-ci sont portés en vert sur la Figure \ref{lafemR.fig} et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
+En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
+Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs proposés dans le problème du pendule inversé par \citet{lagoudakis2003least} consistent en un réseau de gaussiennes qui ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.\\
 
-Cette expérience a été répétée FIXME:nombre de fois, avec les résultats illutrés FIXME:mettre une figure.
+Malgré ces deux difficultés, notre algorithme parvient comme nous l'avons dit à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.
+
 %%%%%

### 2012-02-08 klein2012/main.tex
Corrections suggérées par Matthieu

@@
 
-\title{Batch Imitation via Inverse Reinforcement Learning Without Solving the Direct Problem}
+\title{Classification structurée pour l'apprentissage par renforcement inverse}
 
@@
 Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
-L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité et cherche à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
-Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée et l'accès à la strategie complète de l'expert.
+L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
+Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée ou l'accès à la strategie complète de l'expert.
 Notre contribution consiste en un nouvel algorithme d'IRL levant toutes ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient et possèdant une faible complexité tant en échantillons que calculatoire.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
@@
 \begin{itemize}
-\item Introduction informelle sur les processus de décision séquentiels
+\item Introduction informelle sur les processus de décision séquentielles
 \item Intérêt de l'apprentissage par démonstration
@@
 Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \ref{hatRdef.eqn}).
-   \begin{eqnarray}
-   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
-   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right) + l(s_i,a^*_i)  \right).
-   \end{eqnarray}
+   \begin{equation}
+   R_N(\theta) = {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)
+   \end{equation}
    Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en section \ref{proof.sec} :
@@
 
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer
+\begin{equation}
+\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]
+\end{equation}
+ l'on utilise
 \begin{equation}
@@
 Cet algorithme se base également sur la similarité entre les definitions de $\mu$ et $Q$ mais au lieu d'adapter un algorithme de programmation dynamique comme nous venons d'en donner l'exemple, il adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch} et {\it offline}.
-Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ est obtenue grâce à la solution des moindres carrés :
+Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ pour chacune des composantes $i$ de $\mu^\pi$ est obtenue grâce à la solution des moindres carrés :
 \begin{equation}
@@
 %%%%%
+\section{Related Work}
+%%%%%
+\begin{itemize}
+\item Résumé plus détaillé que précédemment des approches IRL actuelles
+\item Laïus sur Ratliff et son approche locale (imitation pure) et globale (plain de contraintes)
+\end{itemize}
+
+%%%%%
 \section{Résultats expérimentaux}
@@
 
-Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seuls transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
-Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale).
-Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.\\
+Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
+Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale).\\
 
-En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship} car ceux-ci requièrent la résolution du problème direct à chaque itération ; résolution qui ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
-Comme notre algorithme ne nécessite pas la résolution du problème direct, il ne rencontre pas ce problème.
+ Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
+En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
+Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont la {\it feature expectation} est utilisée en association à celle de l'expert.
+Cette politique est obtenue par optimisation d'une récompense sur le MDP dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
+Comme notre algorithme ne nécessite pas le calcul d'autre {\it feature expectation} que celle de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
 %%%
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{LAFEM_Exp3_true_R.pdf}
+       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_true_R.pdf}
        \caption{Récompense sur laquelle l'expert a été entraîné}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{LAFEM_Exp3_lafem_R.pdf}
+       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_lafem_R.pdf}
        \caption{Récompense trouvée par LAFEM, en vert les positions occupées par l'expert.}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{LAFEM_Exp3_Vexpert.pdf}
+       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_Vexpert.pdf}
        \caption{Fonction de valeur de l'expert}
@@
     \begin{center}
-       \includegraphics[width=\textwidth]{LAFEM_Exp3_Vagent.pdf}
+       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_Vagent.pdf}
        \caption{Fonction de valeur de l'agent entraîné sur la récompense trouvée par LAFEM}
@@
 \begin{itemize}
- \item Créer une base de données $D_0$ de transitions aléatoires.
- \item Entraîner un expert en utilisant LSPI nourri avec ces transitions ( definissant ainsi $\pi_E : S\rightarrow A$)
- \item Vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps
- \item Créer une base de données $D_E$ de transitions de l'expert, contenant 10 trajectoires de 300 transitions chacune.
- \item Définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$.
- \item Définir $\alpha(t) = 0.1,\forall t$.
- \item Définir $D$ à partir des transitions $D_E$ en enlevant l'état d'arrivée.
- \item Initialiser $\omega_0 = [-1...-1]^T$.
- \item Fixer $T=20$.
- \item Définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E)$.
- \item Faire tourner LAFEM.
- \item Entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$.
- \item Vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
+ \item créer une base de données $D_0$ de transitions aléatoires ;
+ \item entraîner un expert en utilisant LSPI nourri avec ces transitions ( definissant ainsi $\pi_E : S\rightarrow A$) ;
+ \item vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps ;
+ \item créer une base de données $D_E$ de transitions de l'expert, contenant 10 trajectoires de 300 transitions chacune ;
+ \item définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$ ;
+ \item définir $\alpha(t) = 0.1,\forall t$ ;
+ \item définir $D$ à partir des transitions $D_E$ en enlevant l'état d'arrivée ;
+ \item initialiser $\omega_0 = [-1...-1]^T$ ;
+ \item fixer $T=20$ ;
+ \item définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E)$ ;
+ \item faire tourner LAFEM ;
+ \item entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$ ;
+ \item vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
 \end{itemize}
@@
 
-Malgré ces deux difficultés, notre algorithme parvient comme nous l'avons dit à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.
+Malgré ces deux difficultés, notre algorithme parvient comme nous l'avons dit à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
 
-%%%%%
-\section{Related Work}
-%%%%%
-\begin{itemize}
-\item Résumé plus détaillé que précédemment des approches IRL actuelles
-\item Laïus sur Ratliff et son approche locale (imitation pure) et globale (plain de contraintes)
-\end{itemize}
+La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement que multiplier fonction de récompense ou fonction de valeur par une constante positive n'impacte pas les politiques.
 \section{Conclusion and future work}
@@
 \item Il ne reste que la sélection ou détection de features, mais ce n'est pas hors de portée
-\item Dans la même veine, il reste la résolution du problème direct une fois la récompense obtenue, il faut peut-être creuser du côté du reward shaping
 \item Faible complexité computationnelle

### 2012-02-09 klein2012/main.tex
preuve

--- 2012-02-08 klein2012/main.tex
@@
 &=& R(s) + \gamma\sum_{s'}p(s'|s,\pi(s))V^\pi(s')
-\label{BellmanV.eqn}
 \end{eqnarray}
@@
 \subsection{Considérations techniques}
-\label{ConsiderationsTechniques.sec}
 %%%
@@
 \begin{eqnarray}
-\label{mupidef.eqn}
 \mu^\pi = E\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0\sim d,\pi\right],
@@
 \end{equation}
-Nous allons utiliser cette propriété pour imaginer une fonction de risque dépendante d'une fonction $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+Nous allons utiliser cette propriété pour imaginer une fonction de risque dépendante d'une fonction $q$ qui devra partager avec la fonction inconnue $Q^E$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
 Notons $(s_i,a_i)_i$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
@@
 
-Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système.
+Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance à priori dans le système.
 En effet, minimiser avec succès la fonction de risque avec un choix de $l$ judicieux permet d'imposer à $q$ une structure telle qu'il existe une marge entre l'action choisie par l'expert et la deuxième meilleure action.
@@
 Comme nous l'avons vu Equation \ref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
-L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
+L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q = \theta^T\mu_E(s,a)$.
 Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
@@
    \end{equation}
-%%%
-\subsection{Calcul de $\mu_E$}
-%%%
-\label{calculmu.sec}
-Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de la {\it feature expectation}.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, la {\it feature expectation} de l'expert et non le calcul de la {\it feature expectation} d'une politique arbitraire.
-De fait cela nous épargne le besoin d'un simulateur, d'un solveur de MDP ou d'échantillons couvrant tout l'espace d'état.\\
-
-Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour calculer $\mu_E$.\\
-
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \ref{mupidef.eqn}) et celle de la fonction de valeur (Equation \ref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
-En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \ref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
-\begin{equation}
-\mathbf V^\pi = \mathbf R + \gamma P_\pi\mathbf V^\pi
-\end{equation}
-Il est connu qu'une règle d'update de la forme 
-\begin{equation}
-\mathbf V^\pi_{t+1} \leftarrow \mathbf R + \gamma P_\pi\mathbf V^\pi_t
-\end{equation}
-convergera vers la valeur exacte recherchée.
-Ce sont les algorithmes du type \emph{value iteration} qui exploitent cette propriété.
-Parallèlement il est possible d'utiliser une règle d'update de la forme :
-\begin{equation}
-\mathbf \mu^\pi_{t+1} \leftarrow \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi_t
-\end{equation}
-pour obtenir la valeur exacte de la {\it feature expectation}.\\
-
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
-\begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
-\end{equation}
-
-Finalement, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ (\citep{klein2011batch}) dans sa version \emph{on-policy}.
-Cet algorithme se base également sur la similarité entre les definitions de $\mu$ et $Q$ mais au lieu d'adapter un algorithme de programmation dynamique comme nous venons d'en donner l'exemple, il adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch} et {\it offline}.
-Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ est obtenue grâce à la solution des moindres carrés :
-\begin{equation}
-  \xi_i^* = \left(\sum_{t=1}^n
-  \psi(s_t)(\psi(s_t)-\gamma\psi(s_{t+1}))^T\right)^{-1}
-  \sum_{t=1}^n \psi(s_t) \phi_i(s_t).
-\end{equation}
-
 %%%%%
@@
 \end{align}
-est convergente vers un point qui minimise ~$\af$, où ~$\gamma_{\bk}\in\partial_{\bx_\bk}\af$. De plus, on a ~$\liminf\af({\bx_\bk})=\min_{\bx\in\me}\af(\bx)$.
+est convergente vers un point qui minimise ~$\af$, où ~$\gamma_{\bk}\in\partial_{\bx_\bk}\af$.
+De plus, on a ~$\liminf\af({\bx_\bk})=\min_{\bx\in\me}\af(\bx)$ même en cas de non existence du minimum de ~$\af$ sur ~$\me$.
 \end{theorem}
 \subsection{Application à notre fonction de risque}
-On rappelle que notre fonction de risque $R_N$ est une fonction de l'espace euclidien $\mathbb{R}^p$, qu'on munira du produit scalaire canonique $<.,.>$ et de sa norme associée $\|\|_2$, vers $\mathbb{R}$ où $p$ est la dimension de l'inconnue $\theta$.
-Pour $\theta\in\mathbb{R}^p$, on a :
+On rappelle que notre fonction de risque $R_N$ est une fonction de l'espace euclidien ~$\mathbb{R}^p$, qu'on munira du produit scalaire canonique $<.,.>$ et de sa norme associée ~$\|.\|_2$, vers ~$\mathbb{R}$ où $p$ est la dimension de l'inconnue ~$\theta$.
+Pour ~$\theta\in\mathbb{R}^p$, on a :
 \begin{eqnarray}
@@
 On peut montrer que $R_N$ est définie correctement de la manière suivante.
-On définit pour chaque $i\in\{1,\dots,N\}$ et pour chaque $a\in A$, où $A$ est un ensemble fini d'actions, les fonctions $f_{(i,a)}$ qui sont des fonctions de $\mathbb{R}^p$ vers $\mathbb{R}$ par :
+On définit pour chaque $i\in\{1,\dots,N\}$ et pour chaque $a\in A$, où $A$ est un ensemble fini d'actions, les fonctions $f_{(i,a)}$ qui sont des fonctions de ~$\mathbb{R}^p$ vers ~$\mathbb{R}$ par :
 \begin{equation}
@@
 De plus elle est continue.
-On définit aussi pour chaque $i\in\{1,\dots,N\}$, les fonctions $g_{i}$ qui sont des fonctions de $\mathbb{R}^p$ vers $\mathbb{R}$ par :
+On définit aussi pour chaque $i\in\{1,\dots,N\}$, les fonctions $g_{i}$ qui sont des fonctions de ~$\mathbb{R}^p$ vers ~$\mathbb{R}$ par :
 \begin{equation}
@@
 Il est évident que $g_{i}$ est une fonction convexe, propre et fermée à valeurs finies. De plus elle est continue.
-Comme $\A$ est fini, alors on peut définir pour chaque $i\in\{1,\dots,N\}$, les fonctions $h_{(i,a)}$ qui sont des fonctions de $\mathbb{R}^p$ vers $\mathbb{R}$ par :
+Comme ~$\A$ est fini, alors on peut définir pour chaque $i\in\{1,\dots,N\}$, les fonctions $h_{(i,a)}$ qui sont des fonctions de ~$\mathbb{R}^p$ vers ~$\mathbb{R}$ par :
 \begin{equation}
@@
 La fonction $g_{i}$ étant défini comme un maximum sur un ensemble fini de fonctions continues, convexes, propres et fermées à valeurs finies est encore continue, convexe, propre et fermée à valeurs finies.
-Finalement il vient que l'on peut écrire pour tout $\theta$ :
+Finalement il vient que l'on peut écrire pour tout ~$\theta$ :
 \begin{equation}
@@
 Ainsi $R_N$ est la somme finie de fonctions convexes, propres, fermées et continues à valeurs finies, elle aura donc ces propriétés et est définie correctement.
-On voit donc que la fonction $R_N$ est une bonne candidate à l'application du théorème \ref{theoreme.sec} sous réserve de l'existence d'un $\theta^*$ tel que :
+On voit donc que la fonction $R_N$ est une bonne candidate à l'application du théorème \ref{theoreme.sec} sous réserve de l'existence d'un ~$\theta^*$ tel que :
 \begin{equation}
@@
 \end{equation}
-Nous allons montrer l'existence de ce $\theta^*$ dans un cas particulier.
-
+Nous allons montrer l'existence de ce ~$\theta^*$ dans un cas particulier. On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
+Désormais, on imposera que notre fonction de loss vérifie :
+\begin{equation}
+\forall i\in\{1,\dots,N\}, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0,
+\end{equation}
+où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$.
+Il est facile de borné inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a :
+\begin{eqnarray}
+   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right),\\
+   &\geq& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a_i) - \mu_E(s_i,a_i)\right) + l(s_i,a_i)  \right),\\
+   &\geq& {1\over N} \sum_{i=1}^Nl(s_i,a_i).
+\end{eqnarray}
+Supposons l'existence d'un ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant :
+\begin{equation}\label{eq:existence}
+\forall i\in\{1,\dots,N\}, \hat{\theta}^T\mu_E(s_i,a_i)>\max_{A \backslash a_i}\hat{\theta}^T\mu_E(s_i,).
+\end{equation}
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,{ et ~$\hat{M}=\{\theta\in\mathbb{R}^p:\theta \text{ vérifie \ref{eq:existence} }  \}$ et posons :
+\begin{equation}
+\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,).
+\end{equation}
+Montrons qu'il existe un $x\geq0$ tel que :
+\begin{equation}
+R_N(\hat{\theta}_x)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),
+\end{equation}
+et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration  ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i), \forall \theta\in\mathbb{R}^p$.
+\begin{preuve}
+On va chercher $x\geq0$ tel que :
+\begin{align}
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
+\\
+\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, x\beta_i \geq\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i),
+\\
+\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
+\end{align}
+Or comme ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)\geq\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)+l(s_i,a))$ il est clair que :
+\begin{align}
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
+\\
+\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
+\\
+\Downarrow
+\\
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)+l(s_i,a))\geq0,
+\\
+\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)+l(s_i,a)),
+\\
+\Updownarrow
+\\
+&R_N(\hat{\theta}_x)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i).
+\end{align}
+C'est bien le résultat recherché.
+\end{preuve}
+De plus on peut montrer, toujours sous l'existence de ~$\hat{\theta}$ que $M^*\subset\hat{M}$.
+\begin{preuve}
+Soit ~$\theta^*\in\ M^*$.
+Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i), \forall \theta\in\mathbb{R}^p$, $R_N(\theta^*)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Or on a :
+\begin{align}
+&R_N(\theta^*)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),
+\\
+\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, (\theta^*)^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}((\theta^*)^T\mu_E(s_i,)+l(s_i,a))\geq0,
+\\
+\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\},\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,
+\\
+\Downarrow
+\\
+&\forall i\in\{1,\dots,N\}, (\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
+\end{align}
+D'ou ~$\theta^*\in \hat{M}$. C'est bien ce que l'on rechercher.
+\end{preuve}
+Ainsi on sait que, sous l'existence de ~$\hat{\theta}$, les minimiseurs de ~$\R_N$ existent (il y en a d'ailleurs une infinité) et vérifient l'équation \ref{eq:existence}.
+On peut donc appliquer l'algorithme de descente de sous-gradient avec l'assurance de la convergence de ce dernier vers un minimiseur de la fonction de risque qui vérifie l'équation \ref{eq:existence}.
+L'équation \ref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
 %%%%%
@@
 %%%
+\subsection{Calcul de $\mu_E$}
+%%%
+\label{calculmu.sec}
+L'algorithme présenté repose sur le calcul de la {\it feature expectation} de l'expert, $\mu_E$. Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour la calculer.
+
+La définition du vecteur $\mu_E$ étant proche de celle de la fonction de valeur, en adaptant les algorithmes de programmation dynamique destinés au calcul de la fonction de valeur on obtient une règle de la forme :
+\begin{equation}
+\mathbf \mu^\pi = \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi
+\end{equation}
+permettant le calcul exact de la fonction de valeur pour peu que l'on dispose des probabilités de transitions.\\
+
+Dans le cas plus probable où celles-ci sont inconnues, on peut si l'on dispose d'un simulateur utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ l'on utilise
+\begin{equation}
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
+\end{equation}
+
+Finalement si l'on ne dispose ni des probabilités de transition ni d'un simulateur, alors il est possible d'utiliser LSTD$\mu$ (\citep{klein2011batch}), un algorithme permettant l'estimation en {\it batch} de la {\it feature expectation} de l'expert.
+%%%
 \subsection{Description de la tâche}
 %%%
-Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté.
-Il s'agit d'un problème classique en apprentissage par renforcement.
+Le problème du pendule inversé à maintenir en équilibre vertical un pendule à un degré de liberté.
+il s'agit d'un problème classique en apprentissage par renforcement ({\it Reinforcement Learning, RL}).
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : \S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
 
@@
        \includegraphics[width=\textwidth]{LAFEM_Exp3_lafem_R.pdf}
-       \caption{Récompense trouvée par LAFEM, en vert les positions occupées par l'expert.}
+       \caption{Récompense trouvée par LAFEM}
        \label{lafemR.fig}

### 2012-02-09 klein2012/main.tex
Embryon biblio

--- 2012-02-08 klein2012/main.tex
@@
 & \Rightarrow& V^{\pi_1} = V^{\pi_2}.
+\label{memevaleur.eqn}
 \end{eqnarray}
@@
 En effet, minimiser avec succès la fonction de risque avec un choix de $l$ judicieux permet d'imposer à $q$ une structure telle qu'il existe une marge entre l'action choisie par l'expert et la deuxième meilleure action.
-Ce type d'argument est assez commun dans la litérature.
+Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
 Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères fournis en sous-section \ref{proof.sec}.\\
@@
 \section{Related Work}
+\label{biblio.sec}
 %%%%%
+Parmi les algorithmes d'apprentissage par renforcement inverse existants, certains partagent des caractéristiques communes.\\
+
+Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un optimal par une certaine marge, c'est à dire que l'action choisie par l'expert doit avoir une qualité supérieure d'une certaine marge à n'importe quelle autre action.
+Il s'agit de maximiser cette marge.
+C'est l'approche de \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical}.\\
+
+Une autre approche consiste à effectuer la recherche dans l'espace des politiques grâce à de l'inférence bayesienne sur les paires état-action fournies par l'expert, comme par exemple \cite{ramachandran2007bayesian}.\\
+
+Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
+L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \ref{memevaleur.eqn} que cela impliquait des valeurs similaires).
+Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\
+
 \begin{itemize}

### 2012-02-09 klein2012/main.tex
Tentative Biblio

@@
 
+Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
+L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \ref{memevaleur.eqn} que cela impliquait des valeurs similaires).
+Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\
+
+Une autre approche consiste à effectuer la recherche dans l'espace des politiques grâce à de l'inférence bayesienne sur les paires état-action fournies par l'expert, comme par exemple \cite{ramachandran2007bayesian}.\\
+
 Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un optimal par une certaine marge, c'est à dire que l'action choisie par l'expert doit avoir une qualité supérieure d'une certaine marge à n'importe quelle autre action.
@@
 
-Une autre approche consiste à effectuer la recherche dans l'espace des politiques grâce à de l'inférence bayesienne sur les paires état-action fournies par l'expert, comme par exemple \cite{ramachandran2007bayesian}.\\
+Particulièrement, le travail de \citet{ratliff2006maximum} partage des similarités avec le nôtre ; dans \citep{ratliff2006maximum}, les auteurs présentent une méthode d'IRL en posant le problème sous forme d'un \emph{maximum margin classification problem}.
+La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}, comme cela est expliqué dans \cite{neu2009training}.
+Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un MDP (entrée à classifier) est associée une politique (classe).
 
-Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
-L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \ref{memevaleur.eqn} que cela impliquait des valeurs similaires).
-Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\
+   Une seconde publication, \citep{ratliff2007boosting} vient compléter la précédente en incluant du \emph{boosting}.
+On y voit apparaître à nouveau une fonction de coût empirique (en omettant quelques détails comme par exemple la pondération des trajectoires) :
+   \begin{equation}
+   \label{dix.eqn}%On l'appelle comme ça parce que dans le draft et dans le papier de Ratliff elles ont le même numéro : 10.
+   {1\over n}\sum\limits_{i=1}^n \left(\theta^T\mu^E_i - \min_\pi\left(\theta^T \mu^\pi - l(\pi)\right)\right) + {\lambda \over 2} ||\theta||^2.
+   \end{equation}
 
-\begin{itemize}
-\item Résumé plus détaillé que précédemment des approches IRL actuelles
-\item Laïus sur Ratliff et son approche locale (imitation pure) et globale (plain de contraintes)
-\end{itemize}
+   On constate la présence d'une fonction de coût $l$ similaire à celle que nous utilisons.
+
+   La recherche a lieu dans l'espace des politiques via une méthode de descente de gradient sur les paramètres $\theta$ définissant la récompense, ce qui implique la résolution d'un MDP à chaque itération.
+En effet, le $\min\limits_\pi(\dots)$ présent dans la fonction de coût est analogue au $\pi^* = \arg\max\limits_\pi V^\pi$ décrivant la résolution classique d'un MDP.
+
+   Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
+Cette fonction de score $s : X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque entrée pour prendre une décision de classification i.e.
+choisir le label $y\in Y$ à appliquer à l'entrée $x\in X$ :
+   \begin{equation}
+   \label{sdef.eqn}
+   y^* = \arg\max_{y \in Y} s(x,y)
+   \end{equation}
+   Cette fonction de score est apprise dans l'espace fonctionnel en optimisant la fonction suivante, dont celle donnée en \ref{dix.eqn} est une instanciation : 
+   \begin{equation}
+   \label{rdef.eqn}
+   r[s] = {1\over N} \sum_{i=1}^N\left(\max_{y\in Y}(s(x_i,y) + l(x_i,y)) - s(x_i,y_i) \right)
+   \end{equation}
+   Cela suppose l'accès à une base de données $\{(x_i,y_i)\}_i$; i.e.
+une trace de l'expert.
+On note dans ce papier un changement par rapport aux papiers précédent : les entrées sont maintenant les états et les classes représentent les actions.
 

### 2012-02-09 klein2012/main.tex
Signed-off-by: unknown <piot_bil@PIOTLTOP.metz.supelec.local>

--- 2012-02-09 klein2012/main.tex
@@
 
-\title{Classification structurée pour l'apprentissage par renforcement inverse}
+\title{Batch Imitation via Inverse Reinforcement Learning Without Solving the Direct Problem}
 
@@
 Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
-L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
-Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée ou l'accès à la strategie complète de l'expert.
+L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité et cherche à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
+Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée et l'accès à la strategie complète de l'expert.
 Notre contribution consiste en un nouvel algorithme d'IRL levant toutes ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient et possèdant une faible complexité tant en échantillons que calculatoire.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
@@
 \begin{itemize}
-\item Introduction informelle sur les processus de décision séquentielles
+\item Introduction informelle sur les processus de décision séquentiels
 \item Intérêt de l'apprentissage par démonstration
@@
 & \Rightarrow& V^{\pi_1} = V^{\pi_2}.
-\label{memevaleur.eqn}
 \end{eqnarray}
@@
 
-Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système.
+Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance à priori dans le système.
 En effet, minimiser avec succès la fonction de risque avec un choix de $l$ judicieux permet d'imposer à $q$ une structure telle qu'il existe une marge entre l'action choisie par l'expert et la deuxième meilleure action.
-Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
+Ce type d'argument est assez commun dans la litérature.
 Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères fournis en sous-section \ref{proof.sec}.\\
@@
 Comme nous l'avons vu Equation \ref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
-L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
+L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q = \theta^T\mu_E(s,a)$.
 Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
 Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \ref{hatRdef.eqn}).
-   \begin{equation}
-   R_N(\theta) = {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)
-   \end{equation}
-   Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en section \ref{proof.sec} :
+   \begin{eqnarray}
+   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
+   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right) + l(s_i,a^*_i)  \right).
+   \end{eqnarray}
+   Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en sous-section \ref{proof.sec} :
    \begin{equation}
@@
    \end{equation}
-%%%
-\subsection{Calcul de $\mu_E$}
-%%%
-\label{calculmu.sec}
-Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de la {\it feature expectation}.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, la {\it feature expectation} de l'expert et non le calcul de la {\it feature expectation} d'une politique arbitraire.
-De fait cela nous épargne le besoin d'un simulateur, d'un solveur de MDP ou d'échantillons couvrant tout l'espace d'état.\\
-
-Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour calculer $\mu_E$.\\
-
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \ref{mupidef.eqn}) et celle de la fonction de valeur (Equation \ref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
-En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \ref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
-\begin{equation}
-\mathbf V^\pi = \mathbf R + \gamma P_\pi\mathbf V^\pi
-\end{equation}
-Il est connu qu'une règle d'update de la forme 
-\begin{equation}
-\mathbf V^\pi_{t+1} \leftarrow \mathbf R + \gamma P_\pi\mathbf V^\pi_t
-\end{equation}
-convergera vers la valeur exacte recherchée.
-Ce sont les algorithmes du type \emph{value iteration} qui exploitent cette propriété.
-Parallèlement il est possible d'utiliser une règle d'update de la forme :
-\begin{equation}
-\mathbf \mu^\pi_{t+1} \leftarrow \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi_t
-\end{equation}
-pour obtenir la valeur exacte de la {\it feature expectation}.\\
-
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer
-\begin{equation}
-\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]
-\end{equation}
- l'on utilise
-\begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
-\end{equation}
-
-Finalement, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ (\citep{klein2011batch}) dans sa version \emph{on-policy}.
-Cet algorithme se base également sur la similarité entre les definitions de $\mu$ et $Q$ mais au lieu d'adapter un algorithme de programmation dynamique comme nous venons d'en donner l'exemple, il adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch} et {\it offline}.
-Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ pour chacune des composantes $i$ de $\mu^\pi$ est obtenue grâce à la solution des moindres carrés :
-\begin{equation}
-  \xi_i^* = \left(\sum_{t=1}^n
-  \psi(s_t)(\psi(s_t)-\gamma\psi(s_{t+1}))^T\right)^{-1}
-  \sum_{t=1}^n \psi(s_t) \phi_i(s_t).
-\end{equation}
-
 %%%%%
@@
 \end{equation}
-Il est évident que $f_{i,a}$ est une fonction convexe, propre et fermée à valeurs finies.
+Il est évident que $f_({i,a})$ est une fonction convexe, propre et fermée à valeurs finies.
 De plus elle est continue.
@@
 \begin{equation}
-h_{i}({\theta})=\max_{a \in A}\theta ^T \mu_E(s_i,a) + l(s_i,a)=\max_{a \in A}f_{(i,a)}({\theta}).
+h_{i}({\theta})=\max_{a \in A}(\theta ^T \mu_E(s_i,a) + l(s_i,a))=\max_{a \in A}f_{(i,a)}({\theta}).
 \end{equation}
@@
 Ainsi $R_N$ est la somme finie de fonctions convexes, propres, fermées et continues à valeurs finies, elle aura donc ces propriétés et est définie correctement.
-On voit donc que la fonction $R_N$ est une bonne candidate à l'application du théorème \ref{theoreme.sec} sous réserve de l'existence d'un ~$\theta^*$ tel que :
+On voit donc que la fonction $R_N$ est une bonne candidate à l'application du théorème \ref{theoreme.sec} sous réserve de l'existence d'un ~$\theta^*\in\mathbb{R}^p$ tel que :
 \begin{equation}
@@
 \end{equation}
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,{ et ~$\hat{M}=\{\theta\in\mathbb{R}^p:\theta \text{ vérifie \ref{eq:existence} }  \}$ et posons :
+En toute généralité, un vecteur de récompense vérifiant l'équation \ref{eq:existence} est tel que la politique de l'expert est optimale pour cette récompense sur les états
+$s_i, \forall i\in\{1,\dots,N\}$.\\
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,{ et ~$\hat{M}=\{\theta\in\mathbb{R}^p:\theta \text{ vérifie l'équation \ref{eq:existence} }  \}$ et posons :
 \begin{equation}
@@
 \begin{equation}
-R_N(\hat{\theta}_x)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),
+R_N(\hat{\theta}_x)= {1\over N} \sum_{i=1}^N l(s_i,a_i),
 \end{equation}
-et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration  ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i), \forall \theta\in\mathbb{R}^p$.
+et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
 \begin{preuve}
@@
 \\
-\Updownarrow
+\nonumber\Updownarrow
 \\
@@
 \end{align}
-Or comme ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)\geq\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)+l(s_i,a))$ il est clair que :
+Or comme ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)~\geq~\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)~+~l(s_i,a))$, il est clair que :
 \begin{align}
@@
 \\
-\Updownarrow
+\nonumber\Updownarrow
 \\
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
 \\
-\Downarrow
+\nonumber\Downarrow
 \\
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)+l(s_i,a))\geq0,
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,
 \\
-\Updownarrow
+\nonumber\Updownarrow
 \\
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)+l(s_i,a)),
+&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a)),
 \\
-\Updownarrow
+\nonumber\Updownarrow
 \\
-&R_N(\hat{\theta}_x)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i).
+&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
 \end{align}
@@
 Soit ~$\theta^*\in\ M^*$.
-Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i), \forall \theta\in\mathbb{R}^p$, $R_N(\theta^*)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i),~\forall \theta\in\mathbb{R}^p$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Or on a :
 \begin{align}
-&R_N(\theta^*)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),
+&R_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),
 \\
-\Updownarrow
+\nonumber\Updownarrow
 \\
-&\forall i\in\{1,\dots,N\}, (\theta^*)^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}((\theta^*)^T\mu_E(s_i,)+l(s_i,a))\geq0,
+&\forall i\in\{1,\dots,N\}, (\theta^*)^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}((\theta^*)^T\mu_E(s_i,a)+l(s_i,a))\geq0,
 \\
-\Updownarrow
+\nonumber\Updownarrow
 \\
@@
 \\
-\Downarrow
+\nonumber\Downarrow
 \\
@@
 \end{align}
-D'ou ~$\theta^*\in \hat{M}$. C'est bien ce que l'on rechercher.
+D'ou ~$\theta^*\in \hat{M}$. C'est bien ce que l'on recherchait.
 \end{preuve}
@@
 L'équation \ref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
-%%%%%
-\section{Related Work}
-\label{biblio.sec}
-%%%%%
-Parmi les algorithmes d'apprentissage par renforcement inverse existants, certains partagent des caractéristiques communes.\\
-
-Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un optimal par une certaine marge, c'est à dire que l'action choisie par l'expert doit avoir une qualité supérieure d'une certaine marge à n'importe quelle autre action.
-Il s'agit de maximiser cette marge.
-C'est l'approche de \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical}.\\
-
-Une autre approche consiste à effectuer la recherche dans l'espace des politiques grâce à de l'inférence bayesienne sur les paires état-action fournies par l'expert, comme par exemple \cite{ramachandran2007bayesian}.\\
-
-Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
-L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \ref{memevaleur.eqn} que cela impliquait des valeurs similaires).
-Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\
-
-\begin{itemize}
-\item Résumé plus détaillé que précédemment des approches IRL actuelles
-\item Laïus sur Ratliff et son approche locale (imitation pure) et globale (plain de contraintes)
-\end{itemize}
-
+La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall i\in\{1,\dots,N\}$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall i\in\{1,\dots,N\}$ puisque $\theta^*$ vérifie l'équation \ref{eq:existence}.
 %%%%%
@@
 
-Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
-Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale).\\
+Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seuls transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
+Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale).
+Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.\\
 
- Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
-En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
-Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont la {\it feature expectation} est utilisée en association à celle de l'expert.
-Cette politique est obtenue par optimisation d'une récompense sur le MDP dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
-Comme notre algorithme ne nécessite pas le calcul d'autre {\it feature expectation} que celle de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
+En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship} car ceux-ci requièrent la résolution du problème direct à chaque itération ; résolution qui ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
+Comme notre algorithme ne nécessite pas la résolution du problème direct, il ne rencontre pas ce problème.
 %%%
@@
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_true_R.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_true_R.pdf}
        \caption{Récompense sur laquelle l'expert a été entraîné}
@@
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_lafem_R.pdf}
-       \caption{Récompense trouvée par LAFEM, en vert les positions occupées par l'expert.}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_lafem_R.pdf}
+       \caption{Récompense trouvée par LAFEM}
        \label{lafemR.fig}
@@
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_Vexpert.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_Vexpert.pdf}
        \caption{Fonction de valeur de l'expert}
@@
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_Vagent.pdf}
+       \includegraphics[width=\textwidth]{LAFEM_Exp3_Vagent.pdf}
        \caption{Fonction de valeur de l'agent entraîné sur la récompense trouvée par LAFEM}
@@
 \begin{itemize}
- \item créer une base de données $D_0$ de transitions aléatoires ;
- \item entraîner un expert en utilisant LSPI nourri avec ces transitions ( definissant ainsi $\pi_E : S\rightarrow A$) ;
- \item vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps ;
- \item créer une base de données $D_E$ de transitions de l'expert, contenant 10 trajectoires de 300 transitions chacune ;
- \item définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$ ;
- \item définir $\alpha(t) = 0.1,\forall t$ ;
- \item définir $D$ à partir des transitions $D_E$ en enlevant l'état d'arrivée ;
- \item initialiser $\omega_0 = [-1...-1]^T$ ;
- \item fixer $T=20$ ;
- \item définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E)$ ;
- \item faire tourner LAFEM ;
- \item entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$ ;
- \item vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
+ \item Créer une base de données $D_0$ de transitions aléatoires.
+ \item Entraîner un expert en utilisant LSPI nourri avec ces transitions ( definissant ainsi $\pi_E : S\rightarrow A$)
+ \item Vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps
+ \item Créer une base de données $D_E$ de transitions de l'expert, contenant 10 trajectoires de 300 transitions chacune.
+ \item Définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$.
+ \item Définir $\alpha(t) = 0.1,\forall t$.
+ \item Définir $D$ à partir des transitions $D_E$ en enlevant l'état d'arrivée.
+ \item Initialiser $\omega_0 = [-1...-1]^T$.
+ \item Fixer $T=20$.
+ \item Définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E)$.
+ \item Faire tourner LAFEM.
+ \item Entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$.
+ \item Vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
+ \item Afficher la fonction de valeur de l'expert.
+ \item Afficher la fonction de valeur de l'agent.
+ \item Afficher la fonction de récompense sur laquelle l'expert a été entraîné
+ \item Afficher la fonction de récompense retrouvée par LAFEM
 \end{itemize}
@@
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.\\
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.
 
-Deux éléments méritent d'être notés.
-Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{lafemR.fig} et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
-En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
-Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs proposés dans le problème du pendule inversé par \citet{lagoudakis2003least} consistent en un réseau de gaussiennes qui ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.\\
+%%%
+\subsection{Etude statistique de la qualité de l'imitation}
+%%%
+Nous souhaitons maintenant empiriquement étudier l'impact de la quantité de données fournies par l'expert sur la qualité du contrôle exhibé par l'agent.
+Pour cela nous appliquons le protocole suivant :
+\begin{itemize}
+ \item Créer une base de données $D_0$ de transitions aléatoires.
+ \item Entraîner un expert en utilisant LSPI nourri avec ces transitions (ce qui définit $\pi_E : S\rightarrow A$)
+ \item Vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps
+ \item Créer une base de données $D_E$ de transitions de l'expert. On la crée de 10 trajectoires de 300 transitions chacune.
+ \item Définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$.
+ \item Définir $\alpha(t) = 0.1,\forall t$ (pifomètre)
+ \item Fixer $T=20$ (pifomètre)
+ \item Pour différentes quantités de données $N$,
+\begin{itemize}
+   \item Définir $D_E^N$ en tronquant $D_E$ à ses $N$ premières transitions
+   \item Définir $D$ à partir des de $D_E^N$ en enlevant l'état d'arrivée
+   \item Définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E^N)$
+   \item Initialiser $\omega_0 = [-1...-1]^T$
+   \item Faire tourner LAFEM
+   \item Entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$
+   \item Regarder durant combien de pas de temps l'agent arrrive à faire tenir le pendule.
+   \item Tracer la courbe donnant la qualité du contrôle (nombre de pas de temps durant lequel l'agent balance le pendule) en fonction de la quantitié de données disponible ($N$)
+\end{itemize}
+\end{itemize}
 
-Malgré ces deux difficultés, notre algorithme parvient comme nous l'avons dit à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
-
-La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement que multiplier fonction de récompense ou fonction de valeur par une constante positive n'impacte pas les politiques.
+Cette expérience a été répétée FIXME:nombre de fois, avec les résultats illutrés FIXME:mettre une figure.
+%%%%%
+\section{Related Work}
+%%%%%
+\begin{itemize}
+\item Résumé plus détaillé que précédemment des approches IRL actuelles
+\item Laïus sur Ratliff et son approche locale (imitation pure) et globale (plain de contraintes)
+\end{itemize}
 \section{Conclusion and future work}
@@
 \item Il ne reste que la sélection ou détection de features, mais ce n'est pas hors de portée
+\item Dans la même veine, il reste la résolution du problème direct une fois la récompense obtenue, il faut peut-être creuser du côté du reward shaping
 \item Faible complexité computationnelle

### 2012-02-09 klein2012/main.tex

--- 2012-02-09 klein2012/main.tex
@@
 L'équation \ref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
+La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall i\in\{1,\dots,N\}$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall i\in\{1,\dots,N\}$ puisque $\theta^*$ vérifie l'équation \ref{eq:existence}.
 %%%%%

### 2012-02-09 klein2012/main.tex
changement
Signed-off-by: unknown <piot_bil@PIOTLTOP.metz.supelec.local>

@@
 \begin{equation}
-\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,).
+\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)>0.
 \end{equation}
@@
 C'est bien le résultat recherché.
+De plus on peut remarquer, que si l'on choisi la fonction de loss telle que :
+\begin{equation}
+\forall i\in\{1,\dots,N\}, \forall a \in A \backslash a_i, l(s_i,a)=\max_{a \in A \backslash a_i}>l(s_i,a_i)\geq0,
+\end{equation}
+alors on a ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)~=~\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)~+~l(s_i,a))$. Et donc :
+\begin{align}
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
+\\
+\nonumber\Updownarrow
+\\
+&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
+\end{align}
+Majorons les $\beta_i$ :
+\begin{equation}
+\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\leq\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2.
+\end{equation}
+Donc :
+\begin{align}
+&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i),
+\\
+\nonumber\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
+\\
+\nonumber\Downarrow
+\\
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}.
+\end{align}
+Ainsi, cela impose que la solution $\hat{\theta}_x$ soit au moins de norme $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}$.
 \end{preuve}
-De plus on peut montrer, toujours sous l'existence de ~$\hat{\theta}$ que $M^*\subset\hat{M}$.
+Revenons au premier type de fonction de loss :
+\begin{equation}
+\forall i\in\{1,\dots,N\}, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0.
+\end{equation}
+On peut montrer, toujours sous l'existence de ~$\hat{\theta}$ que $M^*\subset\hat{M}$.
 \begin{preuve}
 Soit ~$\theta^*\in\ M^*$.
-Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i),~\forall \theta\in\mathbb{R}^p$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Or on a :

### 2012-02-09 klein2012/main.tex
Signed-off-by: unknown <piot_bil@PIOTLTOP.metz.supelec.local>
changements

@@
 \begin{align}
+&&\forall \bk\in\mathbb{N}, \alpha_{\bk}>0,
+\\
 &\sum_{\bk=0}^{\bk=+\infty}\alpha_{\bk}=+\infty,
@@
 \end{equation}
-et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
 \begin{preuve}
@@
 \begin{equation}
-\forall i\in\{1,\dots,N\}, \forall a \in A \backslash a_i, l(s_i,a)=\max_{a \in A \backslash a_i}>l(s_i,a_i)\geq0,
+\forall i\in\{1,\dots,N\}, \forall a \in A \backslash a_i, l(s_i,a)=c_i=\max_{a\in A \backslash a_i}l(s_i,a)>l(s_i,a_i)\geq0, \text{ où $c_i$ est une constante réelle,} 
 \end{equation}
@@
 \begin{equation}
-\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\leq\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2.
+\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\leq\max_{A \backslash a_i}(\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2).
 \end{equation}
@@
 \end{align}
-Ainsi, cela impose que la solution $\hat{\theta}_x$ soit au moins de norme $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}$.
+Ainsi, cela impose que la solution $\hat{\theta}_x$ soit au moins de norme $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}$.\\
+Minorons les $\beta_i$ :
+\begin{equation}
+\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\geq|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|.
+\end{equation}
+Donc :
+\begin{align}
+&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i),
+\\
+\nonumber\Updownarrow
+\\
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
+\\
+\nonumber\Uparrow
+\\
+&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|}.
+\end{align}
+Ainsi cela impose que si $\hat{\theta}_x$ est de norme supérieure à $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|}$, alors il est un minimiseur de $R_N$.
 \end{preuve}

### 2012-02-09 klein2012/main.tex
Modifs simples suggérées par Matthieu

--- 2012-02-09 klein2012/main.tex
@@
 \usepackage[T1]{fontenc}
-\usepackage{amsmath}
+\usepackage{mathtools}
+\mathtoolsset{showonlyrefs=true}
 \usepackage{amssymb}
@@
 
-\shorttitle{Batch Inverse Reinforcement Learning}
+\shorttitle{Classification structurée pour l'IRL}
 
@@
 
-\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}\inst{2}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}\inst{3}}}
+\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
 
@@
 Equipe ABC,\\
-LORIA-CNRS, France
+LORIA, France
 \and
@@
 Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
-L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
+L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) à partir de  traces du comportement de l'expert.
 Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée ou l'accès à la strategie complète de l'expert.
-Notre contribution consiste en un nouvel algorithme d'IRL levant toutes ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
+Notre contribution consiste en un nouvel algorithme d'IRL levant l'ensemble de ces contraintes.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
@@
 %%%%%
-\section{Background}
+\section{Contexte}
 %%%%%
@@
 %%%
-\subsection{Principe}
+\subsection{Apprentissage par renforcement direct et inverse}
 %%%
-Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielle.
-La configuration du système à contrôler est alors décrite à chaque instant $t$ par un état $s_t \in S$.
+Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
+La configuration du système à contrôler est alors décrite à chaque instant discret $t$ par un état $s_t \in S$.
 Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
 Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité $p(s_{t+1}|s_t, a_t)$.
-Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décision séquentielle.\\
+Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité.} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
 
-Dans le cadre plus restreint de l'apprentissage par renforcement, pour guider la résolution de ce problème il faut fournir à l'agent une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui lui apprendra le degré auquel l'état dans lequel il se trouve est désirable.
-Une des forces de l'apprentissage par renforcement réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte.
+Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
+Une des forces de l'apprentissage par renforcement réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
 On définit pour ce faire la fonction de valeur $V^\pi$ :
@@
 Un exemple de ce type de tâche est la conduite d'une voiture.
-Nous accompplissons ce genre de choses au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
+Nous accomplissons ce genre de choses au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
 Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
@@
 
-Introduire cette expression dans la définition de la fonction de valeur (Equation \ref{Vdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+Introduire cette expression dans la définition de la fonction de valeur (Equation \eqref{Vdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
 \begin{eqnarray}
@@
 \end{eqnarray}
-C'est d'ailleurs cette variante de l'expression qui sera utilisée si, au lieu d'injecter l'équation \ref{hatRdef.eqn} dans la définition de la fonction de valeur comme nous l'avons fait pour obtenir l'équation \ref{Vmu.eqn}, nous l'injectons dans la définition de la fonction de qualité.
+C'est d'ailleurs cette variante de l'expression qui sera utilisée si, au lieu d'injecter l'équation \eqref{hatRdef.eqn} dans la définition de la fonction de valeur comme nous l'avons fait pour obtenir l'équation \eqref{Vmu.eqn}, nous l'injectons dans la définition de la fonction de qualité.
 On obtient alors :
@@
 L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis à vis d'une certaine fonction de récompense que l'on cherche à retrouver.
-Cela implique, comme nous l'avons vu Equation \ref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
+Cela implique, comme nous l'avons vu Equation \eqref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
 \begin{equation}
@@
 \end{equation}
-alors le critère exprimé Equation \ref{butLAFEM.eqn} s'écrit $\forall i, q(s_i,a^*_i) = q(s_i,a_i)$.
-Notre fonction de risque doit pénaliser les cas où $q(s_i,a^*_i)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \ref{greedy.eqn}) le choix $a_i$ de l'expert.\\
+alors le critère exprimé Equation \eqref{butLAFEM.eqn} s'écrit $\forall i, q(s_i,a^*_i) = q(s_i,a_i)$.
+Notre fonction de risque doit pénaliser les cas où $q(s_i,a^*_i)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.\\
 Nous introduisons de surcroit une fonction de perte $l : S \times A \rightarrow \mathbb{R}_+$ dans la définition de $a^*_i$ qui devient maintenant :
@@
 
-Comme nous l'avons vu Equation \ref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
+Comme nous l'avons vu Equation \eqref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
 L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
 Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
-Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \ref{hatRdef.eqn}).
+Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
    \begin{equation}
@@
 
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \ref{mupidef.eqn}) et celle de la fonction de valeur (Equation \ref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
-En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \ref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
+Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{mupidef.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
+En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \eqref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
 \begin{equation}
@@
 \end{equation}
-En toute généralité, un vecteur de récompense vérifiant l'équation \ref{eq:existence} est tel que la politique de l'expert est optimale pour cette récompense sur les états
+En toute généralité, un vecteur de récompense vérifiant l'équation \eqref{eq:existence} est tel que la politique de l'expert est optimale pour cette récompense sur les états
 $s_i, \forall i\in\{1,\dots,N\}$.\\
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,{ et ~$\hat{M}=\{\theta\in\mathbb{R}^p:\theta \text{ vérifie l'équation \ref{eq:existence} }  \}$ et posons :
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,{ et ~$\hat{M}=\{\theta\in\mathbb{R}^p:\theta \text{ vérifie l'équation \eqref{eq:existence} }  \}$ et posons :
 \begin{equation}
@@
 \end{preuve}
-Ainsi on sait que, sous l'existence de ~$\hat{\theta}$, les minimiseurs de ~$\R_N$ existent (il y en a d'ailleurs une infinité) et vérifient l'équation \ref{eq:existence}.
-On peut donc appliquer l'algorithme de descente de sous-gradient avec l'assurance de la convergence de ce dernier vers un minimiseur de la fonction de risque qui vérifie l'équation \ref{eq:existence}.
-L'équation \ref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
-La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall i\in\{1,\dots,N\}$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall i\in\{1,\dots,N\}$ puisque $\theta^*$ vérifie l'équation \ref{eq:existence}.
+Ainsi on sait que, sous l'existence de ~$\hat{\theta}$, les minimiseurs de ~$\R_N$ existent (il y en a d'ailleurs une infinité) et vérifient l'équation \eqref{eq:existence}.
+On peut donc appliquer l'algorithme de descente de sous-gradient avec l'assurance de la convergence de ce dernier vers un minimiseur de la fonction de risque qui vérifie l'équation \eqref{eq:existence}.
+L'équation \eqref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
+La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall i\in\{1,\dots,N\}$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall i\in\{1,\dots,N\}$ puisque $\theta^*$ vérifie l'équation \eqref{eq:existence}.
 %%%%%
@@
 Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
-L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \ref{memevaleur.eqn} que cela impliquait des valeurs similaires).
+L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \eqref{memevaleur.eqn} que cela impliquait des valeurs similaires).
 Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\

### 2012-02-09 klein2012/main.tex
Corrections suggérées par Matthieu

@@
 Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
-La configuration du système à contrôler est alors décrite à chaque instant discret $t$ par un état $s_t \in S$.
+La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$, qui respecte donc le critère de Markov.
 Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
@@
 On définit pour ce faire la fonction de valeur $V^\pi$ :
-\begin{eqnarray}
+\begin{equation}
 \label{Vdef.eqn}
-V^\pi(s) &=& E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,\pi\right]\\
-&=& R(s) + \gamma\sum_{s'}p(s'|s,\pi(s))V^\pi(s')
-\label{BellmanV.eqn}
-\end{eqnarray}
-où $\gamma$ est un facteur d'actualisation strictement plus petit que $1$.
-Une politique optimale est alors définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie :
+V^\pi(s) = E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,\pi\right]=R(s) + \gamma E[V^\pi(s')|s,\pi]
+\end{equation}
+où $\gamma$ est un facteur d'actualisation strictement plus petit que $1$. Il est également possible de définir une fonction voisine de la fonction de valeur, la fonction de qualité :
+\begin{equation}
+\label{Qdef.eqn}
+Q^\pi(s,a) = E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,a_0=a,\pi\right] = R(s) + \gamma E[Q^\pi(s',\pi(s'))|s,a].
+\end{equation}
+  Une politique optimale est définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie :
 \begin{equation}
 \forall \pi, \forall s, V^*(s) \geq V^\pi(s).
+\end{equation}
+La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
+\begin{equation}
+\label{greedy.eqn}
+\pi^*(s) \in \arg\max_a Q^*(s,a).
 \end{equation}
@@
 
-Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour lequel ce comportement est optimal.
+Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour lequel ce comportement est optimal (voir par exemple \cite{ng1999policy}).
 Particulièrement, la récompense uniformément nulle remplit ce critère dans la mesure où tout comportement est optimal vis à vis de cette récompense.
@@
 %%%
-La recherche de la politique optimale en apprentissage par renforcement se fait souvent en utilisant une fonction voisine de la fonction de valeur, la fonction de qualité définie par :
-\begin{eqnarray}
-Q^\pi(s,a) &=& E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,a_0=a,\pi\right]\\
-&=& R(s) + \gamma\sum_{s'}p(s'|s,a)V^\pi(s').
-\end{eqnarray}
-La politique optimale se définit simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
-\begin{equation}
-\label{greedy.eqn}
-\pi^*(s) \in \arg\max_a Q^*(s,a).
-\end{equation}
-Dès lors, l'approximation de la fonction de valeur (terme qui employé au sens large désigne aussi la fonction de qualité) est un thème central en apprentissage par renforcement.
-Une caractéristique répandue dans la litérature est l'utilisation d'un cadre d'approximation linéaire où, après avoir défini des attributs vectoriels $\phi: S\times A \rightarrow \mathbb{R}^k$, l'on va chercher un vecteur de paramètres $\omega$ tel que l'on ait une bonne approximation de la fonction de qualité :
-\begin{equation}
-\hat Q (s,a) = \omega^T\phi(s,a)\approx Q(s,a).
-\end{equation}
-Il est naturel d'adopter le même type d'approche pour l'apprentissage par renforcement inverse, c'est à dire de considérer que l'inconnue du problème, la récompense $R$, peut être bien approximée grâce à un vecteur de poids $\theta$ et des attributs vectoriels $\psi : S\rightarrow \mathbb{R}^p$ :
+La récompense $R$ étant l'inconnue du problème, il est naturel de l'approximer grâce à un vecteur de poids $\theta$ et des attributs vectoriels $\psi : S\rightarrow \mathbb{R}^p$ servant de fonctions de base :
 \begin{equation}
@@
 
-Introduire cette expression dans la définition de la fonction de valeur (Equation \eqref{Vdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
-\begin{eqnarray}
-V^\pi(s) &=& E\left[\left.\sum_t\gamma^t\theta^T\psi(s_t)\right|s_0=s,\pi\right]\\
-V^\pi(s) &=& \theta^TE\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0=s,\pi\right]\\
-V^\pi(s) &=& \theta^T\mu^\pi(s)
-\label{Vmu.eqn}
-\end{eqnarray}
-
-Le terme $\mu^\pi$ est appelé la {\it feature expectation} d'une politique.
-Il existe plusieurs variantes basées sur la même définition que nous venons de donner.
-Dans sa version la plus spécifique, notée simplement $\mu^\pi$, il s'agit en toute rigueur de l'expression :
-\begin{eqnarray}
-\label{mupidef.eqn}
-\mu^\pi = E\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0\sim d,\pi\right],
-\end{eqnarray}
-où $d$ est une distribution sur l'état initial.
-Il peut également être utile de le définir sous la forme $\mu^\pi(s,a)$ où son rôle par rapport à $\mu^\pi(s)$ est le même que celui de $Q$ par rapport à $V$ :
-\begin{eqnarray}
-\mu^\pi(s,a) = E\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0=s,a_0=a,\pi\right].
-\end{eqnarray}
-C'est d'ailleurs cette variante de l'expression qui sera utilisée si, au lieu d'injecter l'équation \eqref{hatRdef.eqn} dans la définition de la fonction de valeur comme nous l'avons fait pour obtenir l'équation \eqref{Vmu.eqn}, nous l'injectons dans la définition de la fonction de qualité.
-On obtient alors :
+Introduire cette expression dans la définition de la fonction de qualité (Equation \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
 \begin{eqnarray}
@@
 \end{eqnarray}
+
+Le terme $\mu^\pi$ est appelé la {\it feature expectation} ou plutôt l'attribut vectoriel moyen d'une politique.
+On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le MDP fixe la paramétrisation de $Q^\pi$.\\
 
@@
 \end{eqnarray}
-La {\it feature expectation} de l'expert, $\mu_E$, pouvant être calculée sans forcément avoir accès à l'intégralité de la politique $\pi_E$, certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les {\it feature expectations} respectives de l'agent et de l'expert, afin d'obtenir un agent dont le comportement à la même valeur que celui de l'expert.
+La {\it feature expectation} de l'expert, $\mu_E$, pouvant être calculée sans forcément avoir accès à l'intégralité de la politique $\pi_E$, certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les {\it feature expectations} respectives de l'agent et de l'expert, afin d'obtenir un agent dont le comportement à la même valeur que celui de l'expert (voir \cite{neu2009training}).
 %%%%%
@@
 \end{equation}
-Notons
+Notre fonction de risque doit donc pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.\\
+Nous introduisons de surcroit une fonction de perte $l : S \times A \rightarrow \mathbb{R}_+$ dans notre fonction de risque qui peut alors s'assembler en :
   \begin{equation}
-a^*_i = \arg\max_aq(s_i,a),
+   R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right)
    \end{equation}
-alors le critère exprimé Equation \eqref{butLAFEM.eqn} s'écrit $\forall i, q(s_i,a^*_i) = q(s_i,a_i)$.
-Notre fonction de risque doit pénaliser les cas où $q(s_i,a^*_i)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.\\
-Nous introduisons de surcroit une fonction de perte $l : S \times A \rightarrow \mathbb{R}_+$ dans la définition de $a^*_i$ qui devient maintenant :
-\begin{equation}
-\label{astar.def}
-a^*_i = \arg\max_aq(s_i,a) + l(s_i,a).
-\end{equation}
-La définition de notre fonction de risque peut alors être assemblée en :
-  \begin{eqnarray}
-   R_N(q) &=& {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right)\\
-   &=& {1\over N} \sum_{i=1}^N\left(q(s_i,a_i^*) + l(s_i,a_i^*) - q(s_i,a_i) \right)
-   \end{eqnarray}
 
@@
 Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
-Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères fournis en sous-section \ref{proof.sec}.\\
+Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans les preuves (Section \ref{proof.sec}).\\
 
@@
 Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
-   \begin{equation}
-   R_N(\theta) = {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)
-   \end{equation}
+   \begin{eqnarray}
+   R_N(\theta)  &=& {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)\\
+   &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
+\textrm{avec } a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))
+   \end{eqnarray}
    Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en section \ref{proof.sec} :
@@
 
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{mupidef.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
-En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \eqref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
+Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{Qmu.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
+En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \eqref{Vdef.eqn}) peut s'écrire sous forme matricelle :
 \begin{equation}

### 2012-02-09 klein2012/main.tex
s/feature expectation/attribut vectoriel moyen/g

@@
 
-On peut noter que si deux politiques ont la même {\it feature expectation}, alors ces deux politiques ont la même valeur vis à vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
+On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis à vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
 En effet
@@
 \end{eqnarray}
-La {\it feature expectation} de l'expert, $\mu_E$, pouvant être calculée sans forcément avoir accès à l'intégralité de la politique $\pi_E$, certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les {\it feature expectations} respectives de l'agent et de l'expert, afin d'obtenir un agent dont le comportement à la même valeur que celui de l'expert (voir \cite{neu2009training}).
+L'attribut vectoriel moyen de l'expert, $\mu_E$, pouvant être calculée sans forcément avoir accès à l'intégralité de la politique $\pi_E$, certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert, afin d'obtenir un agent dont le comportement à la même valeur que celui de l'expert (voir \cite{neu2009training}).
 %%%%%
@@
 
-Comme nous l'avons vu Equation \eqref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
-L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
+Comme nous l'avons vu Equation \eqref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
+L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
 Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
@@
 \label{calculmu.sec}
-Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de la {\it feature expectation}.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, la {\it feature expectation} de l'expert et non le calcul de la {\it feature expectation} d'une politique arbitraire.
+Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l''attribut vectoriel moyen.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul de l'attribut vectoriel moyen d'une politique arbitraire.
 De fait cela nous épargne le besoin d'un simulateur, d'un solveur de MDP ou d'échantillons couvrant tout l'espace d'état.\\
@@
 
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{Qmu.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
+Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{Qmu.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de l'attribut vectoriel moyen.\\
 En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \eqref{Vdef.eqn}) peut s'écrire sous forme matricelle :
@@
 \end{equation}
-pour obtenir la valeur exacte de la {\it feature expectation}.\\
+pour obtenir la valeur exacte de l'attribut vectoriel moyen.\\
 
@@
 Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
-L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \eqref{memevaleur.eqn} que cela impliquait des valeurs similaires).
+L'approche consiste à imiter l'expert par le biais de l'IRL ; les attributs vectoriels moyens de l'agent et de l'expert doivent être proche (l'on a vu Equation \eqref{memevaleur.eqn} que cela impliquait des valeurs similaires).
 Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\
@@
 En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
-Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont la {\it feature expectation} est utilisée en association à celle de l'expert.
+Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel moyen est utilisée en association à celle de l'expert.
 Cette politique est obtenue par optimisation d'une récompense sur le MDP dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
-Comme notre algorithme ne nécessite pas le calcul d'autre {\it feature expectation} que celle de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
+Comme notre algorithme ne nécessite pas le calcul d'autre attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
 %%%

### 2012-02-10 klein2012/main.tex
new

--- 2012-02-09 klein2012/main.tex
@@
 \usepackage[T1]{fontenc}
-\usepackage{mathtools}
-\mathtoolsset{showonlyrefs=true}
+\usepackage{amsmath}
 \usepackage{amssymb}
@@
 
-\shorttitle{Classification structurée pour l'IRL}
+\shorttitle{Batch Inverse Reinforcement Learning}
 
@@
 
-\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
+\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}\inst{2}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}\inst{3}}}
 
@@
 Equipe ABC,\\
-LORIA, France
+LORIA-CNRS, France
 \and
@@
 Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
-L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) à partir de  traces du comportement de l'expert.
+L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) avec pour entrée une trace du comportement de l'expert.
 Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée ou l'accès à la strategie complète de l'expert.
-Notre contribution consiste en un nouvel algorithme d'IRL levant l'ensemble de ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
+Notre contribution consiste en un nouvel algorithme d'IRL levant toutes ces contraintes.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
@@
 %%%%%
-\section{Contexte}
+\section{Background}
 %%%%%
@@
 %%%
-\subsection{Apprentissage par renforcement direct et inverse}
+\subsection{Principe}
 %%%
-Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
-La configuration du système à contrôler est alors décrite à chaque instant discret $t$ par un état $s_t \in S$.
+Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielle.
+La configuration du système à contrôler est alors décrite à chaque instant $t$ par un état $s_t \in S$.
 Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
 Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité $p(s_{t+1}|s_t, a_t)$.
-Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité.} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
+Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décision séquentielle.\\
 
-Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
-Une des forces de l'apprentissage par renforcement réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
+Dans le cadre plus restreint de l'apprentissage par renforcement, pour guider la résolution de ce problème il faut fournir à l'agent une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui lui apprendra le degré auquel l'état dans lequel il se trouve est désirable.
+Une des forces de l'apprentissage par renforcement réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte.
 On définit pour ce faire la fonction de valeur $V^\pi$ :
@@
 Un exemple de ce type de tâche est la conduite d'une voiture.
-Nous accomplissons ce genre de choses au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
+Nous accompplissons ce genre de choses au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
 Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
@@
 
-Introduire cette expression dans la définition de la fonction de valeur (Equation \eqref{Vdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+Introduire cette expression dans la définition de la fonction de valeur (Equation \ref{Vdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
 \begin{eqnarray}
@@
 \end{eqnarray}
-C'est d'ailleurs cette variante de l'expression qui sera utilisée si, au lieu d'injecter l'équation \eqref{hatRdef.eqn} dans la définition de la fonction de valeur comme nous l'avons fait pour obtenir l'équation \eqref{Vmu.eqn}, nous l'injectons dans la définition de la fonction de qualité.
+C'est d'ailleurs cette variante de l'expression qui sera utilisée si, au lieu d'injecter l'équation \ref{hatRdef.eqn} dans la définition de la fonction de valeur comme nous l'avons fait pour obtenir l'équation \ref{Vmu.eqn}, nous l'injectons dans la définition de la fonction de qualité.
 On obtient alors :
@@
 L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis à vis d'une certaine fonction de récompense que l'on cherche à retrouver.
-Cela implique, comme nous l'avons vu Equation \eqref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
+Cela implique, comme nous l'avons vu Equation \ref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
 \begin{equation}
@@
 \end{equation}
-alors le critère exprimé Equation \eqref{butLAFEM.eqn} s'écrit $\forall i, q(s_i,a^*_i) = q(s_i,a_i)$.
-Notre fonction de risque doit pénaliser les cas où $q(s_i,a^*_i)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.\\
+alors le critère exprimé Equation \ref{butLAFEM.eqn} s'écrit $\forall i, q(s_i,a^*_i) = q(s_i,a_i)$.
+Notre fonction de risque doit pénaliser les cas où $q(s_i,a^*_i)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \ref{greedy.eqn}) le choix $a_i$ de l'expert.\\
 Nous introduisons de surcroit une fonction de perte $l : S \times A \rightarrow \mathbb{R}_+$ dans la définition de $a^*_i$ qui devient maintenant :
@@
 
-Comme nous l'avons vu Equation \eqref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
+Comme nous l'avons vu Equation \ref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de la {\it feature expectation} de cette politique.
 L'expert ne déroge pas à cette règle et notant $\mu_E$ la {\it feature expectation} de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
 Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
-Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
+Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \ref{hatRdef.eqn}).
    \begin{equation}
@@
 
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{mupidef.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
-En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \eqref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
+Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \ref{mupidef.eqn}) et celle de la fonction de valeur (Equation \ref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de la {\it feature expectation}.\\
+En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \ref{BellmanV.eqn}) peut s'écrire sous forme matricelle :
 \begin{equation}
@@
 %%%%%
-\section{Preuves}\label{proof.sec}
-\subsection{Convergence de l'algorithme de sous-gradient sous réserve de l'existence d'une solution}
-\begin{definition}[Fonction convexe]
-Soient ~$\me$ un espace vectoriel et ~$\overline{\mathbb{R}}=\mathbb{R}\cup\{+\infty,-\infty\}$ la droite achevée.
-On dira qu'une fonction ~$\af$ de ~$\me$ vers ~$\overline{\mathbb{R}}$ est une fonction convexe si pour tout couple ~$(\bx,\by)\in(\dom\af)\times(\dom\af)$ et pour tout ~$\bt\in]0,1[$ :
+\section{Analyse}\label{proof.sec}
+\subsection{Hypothèses}
+Dans cette section, nous allons présenter le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
+On supposera que la récompense $R$ recherchée s'écrive sous la forme $R(s)=\theta_E^T\phi(s)$ où $\theta_E\in\mathbb{R}^p$.
+De plus, on supposera que notre base d'exemples est telle que $(S\times A) \subset (s_i,a_i)_{1\leq i \leq N}$.
+Finalement la récompense $R$ devra vérifier la propriété suivante :
+\begin{equation}\label{eq:existence}
+\forall 1\leq i \leq N, \theta_E^T\mu_E(s_i,a_i)>\max_{a\in A \backslash a_i}\theta_E^T\mu_E(s_i,a),
+\end{equation}
+où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$, qui est légèrement plus forte que l'équation d'optimalité de Bellman puisque l'équation \ref{eq:existence} implique :
 \begin{equation}
-\af((1-\bt)\bx+\bt\by)\leq(1-\bt)\af(\bx)+\bt\af(\by),
+\forall s\in S, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a).
 \end{equation}
-où ~$\dom\af=\{\bx\in\me:\af(\bx)<+\infty\}$.
-On définit aussi l'épigraphe de ~$\af$ : ~$\epi \af=\{(\bx,\alpha)\in\me\times\mathbb{R}:\af(\bx)\leq\alpha\}$. Si ce dernier est fermé alors on dira que ~$\af$ est convexe fermée.
-\end{definition}
-\begin{definition}[Sous-gradient d'une fonction convexe]
-Soit ~$\af$ une fonction convexe de ~$\me$ vers ~$\mathbb{R}\cup\{+\infty\}$ tel que ~$\dom\af$ non vide.
-On dira alors que ~$\af$ est une fonction convexe propre.
-On note ~$\conv(\me)$ l'ensemble des fonctions convexes propres.
-Lorsque ~$\me$ est un espace euclidien muni du produit scalaire $<,>$, on dit que la fonction ~$\af$ est sous-différentiable en ~$\bx\in\dom\af$ s'il existe un ~$\bx^*\in\me$ tel que:
+Ainsi tout vecteur $\theta$ vérifiant l'équation \ref{eq:existence} rendra la politique de l'expert optimale.
+Dans le cadre que nous venons d'établir, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\ref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation \ref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+\subsection{Résultats}
+On va montrer dans un premier temps qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \ref{eq:existence} et qui minimise $R_N$.
+On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
+Désormais, on imposera que notre fonction de perte vérifie :
 \begin{equation}
-\forall \by\in\me : \af(\by)\geq\af(\bx)+<\bx^*,\by-\bx>.
+\forall 1\leq i \leq N, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0.
 \end{equation}
-On dira alors que ~$\bx^*$ est un sous-gradient de ~$\af$ en ~$\bx$ et on note ~$\partial_\bx\af$ l'ensemble des sous-gradients de ~$\af$ en ~$\bx$.
-\end{definition}
-\begin{theorem}[Théorème issu de l'article \cite{correa1993convergence}]\label{theoreme.sec}
-Soit ~$\af$ une fonction convexe, propre et fermée de ~$\me$ vers ~$\mathbb{R}\cup\{+\infty\}$.
-On suppose que ~$\me$ est un espace euclidien muni du produit scalaire $<,>$ et de la norme ~$\|.\|_2$ associée et que ~$\af$ prend des valeurs finies.
-Alors si la fonction admet un minimum sur ~$\me$, la suite définie par :
+Il est facile de borné inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a :
+\begin{equation}
+R_N(\theta)\geq{1\over N} \geq {1\over N} \sum_{i=1}^N l(s_i,a_i).
+\end{equation}
+Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \ref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$, et ~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \ref{eq:existence} }\}$ et posons ~$\forall~1\leq i \leq N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
+Montrons qu'il existe un $x\geq0$ tel que $\forall 1\leq i \leq N, R_N(\hat{\theta}_x)= {1\over N} \sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.\\
+Avec les notations précédentes, on vérifie:
 \begin{align}
-&\bx_0\in\me,
+&\forall 1\leq i \leq N, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
 \\
-&\forall \bk\in\mathbb{N} : \bx_{\bk+1}=\bx_\bk-\alpha_{\bk}\frac{\gamma_{\bk}}{\|\gamma_{\bk}\|_2},
-\end{align}
-telle que :
-\begin{align}
-&\sum_{\bk=0}^{\bk=+\infty}\alpha_{\bk}=+\infty,
-\\
-&\sum_{\bk=0}^{\bk=+\infty}\alpha_{\bk}^2<+\infty,
-\end{align}
-est convergente vers un point qui minimise ~$\af$, où ~$\gamma_{\bk}\in\partial_{\bx_\bk}\af$.
-De plus, on a ~$\liminf\af({\bx_\bk})=\min_{\bx\in\me}\af(\bx)$ même en cas de non existence du minimum de ~$\af$ sur ~$\me$.
-\end{theorem}
-\subsection{Application à notre fonction de risque}
-On rappelle que notre fonction de risque $R_N$ est une fonction de l'espace euclidien ~$\mathbb{R}^p$, qu'on munira du produit scalaire canonique $<.,.>$ et de sa norme associée ~$\|.\|_2$, vers ~$\mathbb{R}$ où $p$ est la dimension de l'inconnue ~$\theta$.
-Pour ~$\theta\in\mathbb{R}^p$, on a :
-\begin{eqnarray}
-   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
-   &=& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right) + l(s_i,a^*_i)  \right).
-\end{eqnarray}
-On peut montrer que $R_N$ est définie correctement de la manière suivante.
-On définit pour chaque $i\in\{1,\dots,N\}$ et pour chaque $a\in A$, où $A$ est un ensemble fini d'actions, les fonctions $f_{(i,a)}$ qui sont des fonctions de ~$\mathbb{R}^p$ vers ~$\mathbb{R}$ par :
-\begin{equation}
-f_{(i,a)}({\theta})=\theta ^T \mu_E(s_i,a) + l(s_i,a).
-\end{equation}
-Il est évident que $f_({i,a})$ est une fonction convexe, propre et fermée à valeurs finies.
-De plus elle est continue.
-On définit aussi pour chaque $i\in\{1,\dots,N\}$, les fonctions $g_{i}$ qui sont des fonctions de ~$\mathbb{R}^p$ vers ~$\mathbb{R}$ par :
-\begin{equation}
-g_{i}({\theta})=-\theta ^T \mu_E(s_i,a_i).
-\end{equation}
-Il est évident que $g_{i}$ est une fonction convexe, propre et fermée à valeurs finies. De plus elle est continue.
-Comme ~$\A$ est fini, alors on peut définir pour chaque $i\in\{1,\dots,N\}$, les fonctions $h_{(i,a)}$ qui sont des fonctions de ~$\mathbb{R}^p$ vers ~$\mathbb{R}$ par :
-\begin{equation}
-h_{i}({\theta})=\max_{a \in A}(\theta ^T \mu_E(s_i,a) + l(s_i,a))=\max_{a \in A}f_{(i,a)}({\theta}).
-\end{equation}
-La fonction $g_{i}$ étant défini comme un maximum sur un ensemble fini de fonctions continues, convexes, propres et fermées à valeurs finies est encore continue, convexe, propre et fermée à valeurs finies.
-Finalement il vient que l'on peut écrire pour tout ~$\theta$ :
-\begin{equation}
-R_N(\theta) = {1\over N} \sum_{i=1}^N g_{i}({\theta})+h_{i}({\theta}).
-\end{equation}
-Ainsi $R_N$ est la somme finie de fonctions convexes, propres, fermées et continues à valeurs finies, elle aura donc ces propriétés et est définie correctement.
-On voit donc que la fonction $R_N$ est une bonne candidate à l'application du théorème \ref{theoreme.sec} sous réserve de l'existence d'un ~$\theta^*\in\mathbb{R}^p$ tel que :
-\begin{equation}
-R_N(\theta^*)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta).
-\end{equation}
-Nous allons montrer l'existence de ce ~$\theta^*$ dans un cas particulier. On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
-Désormais, on imposera que notre fonction de loss vérifie :
-\begin{equation}
-\forall i\in\{1,\dots,N\}, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0,
-\end{equation}
-où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$.
-Il est facile de borné inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a :
-\begin{eqnarray}
-   R_N(\theta) &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right),\\
-   &\geq& {1\over N} \sum_{i=1}^N\left(\theta^T\left(\mu_E(s_i,a_i) - \mu_E(s_i,a_i)\right) + l(s_i,a_i)  \right),\\
-   &\geq& {1\over N} \sum_{i=1}^Nl(s_i,a_i).
-\end{eqnarray}
-Supposons l'existence d'un ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant :
-\begin{equation}\label{eq:existence}
-\forall i\in\{1,\dots,N\}, \hat{\theta}^T\mu_E(s_i,a_i)>\max_{A \backslash a_i}\hat{\theta}^T\mu_E(s_i,).
-\end{equation}
-En toute généralité, un vecteur de récompense vérifiant l'équation \eqref{eq:existence} est tel que la politique de l'expert est optimale pour cette récompense sur les états
-$s_i, \forall i\in\{1,\dots,N\}$.\\
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,{ et ~$\hat{M}=\{\theta\in\mathbb{R}^p:\theta \text{ vérifie l'équation \eqref{eq:existence} }  \}$ et posons :
-\begin{equation}
-\forall i\in\{1,\dots,N\}, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,).
-\end{equation}
-Montrons qu'il existe un $x\geq0$ tel que :
-\begin{equation}
-R_N(\hat{\theta}_x)= {1\over N} \sum_{i=1}^N l(s_i,a_i),
-\end{equation}
-et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^Nl(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
-\begin{preuve}
-On va chercher $x\geq0$ tel que :
-\begin{align}
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
-\\
-\nonumber\Updownarrow
-\\
-&\forall i\in\{1,\dots,N\}, x\beta_i \geq\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i),
-\\
-\nonumber\Updownarrow
-\\
-&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
+\Leftrightarrow &\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
 \end{align}
@@
 \begin{align}
-&\forall i\in\{1,\dots,N\}, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
-\\
-\nonumber\Updownarrow
-\\
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
-\\
-\nonumber\Downarrow
-\\
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,
-\\
-\nonumber\Updownarrow
-\\
-&\forall i\in\{1,\dots,N\}, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a)),
+&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},\\
+\Leftrightarrow&\forall 1\leq i \leq N, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,\\
+\Rightarrow&\forall 1\leq i \leq N, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,\\
+\Leftrightarrow&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).\\
+\end{align}
+C'est bien le résultat recherché.
+De plus on peut remarquer, que si l'on choisi la fonction de perte telle que :
+\begin{equation}
+\forall 1\leq i \leq N, \forall a \in A \backslash a_i, l(s_i,a)=c_i=\max_{a\in A \backslash a_i}l(s_i,a)>l(s_i,a_i)\geq0, \text{ où $c_i$ est une constante réelle,}
+\end{equation}
+alors on a ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)~=~\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)~+~l(s_i,a))$. Et donc :
+\begin{align}
+&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
 \\
@@
 \end{align}
-C'est bien le résultat recherché.
-\end{preuve}
-De plus on peut montrer, toujours sous l'existence de ~$\hat{\theta}$ que $M^*\subset\hat{M}$.
+Majorons les $\beta_i$ :
+\begin{equation}
+\forall 1\leq i \leq N, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\leq\max_{A \backslash a_i}(\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2).
+\end{equation}
+Donc :
+\begin{align}
+&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i),
+\\
+\nonumber\Updownarrow
+\\
+&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
+\\
+\nonumber\Downarrow
+\\
+&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}.
+\end{align}
+Ainsi, cela impose que la solution $\hat{\theta}_x$ soit au moins de norme $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}$.\\
+Minorons les $\beta_i$ :
+\begin{equation}
+\forall 1\leq i \leq N, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\geq|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|.
+\end{equation}
+Donc :
+\begin{align}
+&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i),
+\\
+\nonumber\Updownarrow
+\\
+&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
+\\
+\nonumber\Uparrow
+\\
+&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|}.
+\end{align}
+Ainsi cela impose que si $\hat{\theta}_x$ est de norme supérieure à $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|}$, alors il est un minimiseur de $R_N$.
+
+Revenons au premier type de fonction de loss :
+\begin{equation}
+\forall 1\leq i \leq N, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0.
+\end{equation}
+On peut montrer, toujours sous l'existence de ~$\hat{\theta}$ que $M^*\subset\hat{M}$.
 \begin{preuve}
 Soit ~$\theta^*\in\ M^*$.
-Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i),~\forall \theta\in\mathbb{R}^p$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Or on a :
@@
 \\
-&\forall i\in\{1,\dots,N\}, (\theta^*)^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}((\theta^*)^T\mu_E(s_i,a)+l(s_i,a))\geq0,
+&\forall 1\leq i \leq N, (\theta^*)^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}((\theta^*)^T\mu_E(s_i,a)+l(s_i,a))\geq0,
 \\
@@
 \\
-&\forall i\in\{1,\dots,N\},\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,
+&\forall 1\leq i \leq N,\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,
 \\
@@
 \\
-&\forall i\in\{1,\dots,N\}, (\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
+&\forall 1\leq i \leq N, (\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
 \end{align}
@@
 \end{preuve}
-Ainsi on sait que, sous l'existence de ~$\hat{\theta}$, les minimiseurs de ~$\R_N$ existent (il y en a d'ailleurs une infinité) et vérifient l'équation \eqref{eq:existence}.
-On peut donc appliquer l'algorithme de descente de sous-gradient avec l'assurance de la convergence de ce dernier vers un minimiseur de la fonction de risque qui vérifie l'équation \eqref{eq:existence}.
-L'équation \eqref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
-La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall i\in\{1,\dots,N\}$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall i\in\{1,\dots,N\}$ puisque $\theta^*$ vérifie l'équation \eqref{eq:existence}.
+Ainsi on sait que, sous l'existence de ~$\hat{\theta}$, les minimiseurs de ~$\R_N$ existent (il y en a d'ailleurs une infinité) et vérifient l'équation \ref{eq:existence}.
+On peut donc appliquer l'algorithme de descente de sous-gradient avec l'assurance de la convergence de ce dernier vers un minimiseur de la fonction de risque qui vérifie l'équation \ref{eq:existence}.
+L'équation \ref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
+La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall 1\leq i \leq N$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall 1\leq i \leq N$ puisque $\theta^*$ vérifie l'équation \ref{eq:existence}.
 %%%%%
@@
 Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
-L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \eqref{memevaleur.eqn} que cela impliquait des valeurs similaires).
+L'approche consiste à imiter l'expert par le biais de l'IRL ; les {\it feature expectations} de l'agent et de l'expert doivent être proche (l'on a vu Equation \ref{memevaleur.eqn} que cela impliquait des valeurs similaires).
 Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\

### 2012-02-10 klein2012/main.tex
Pour puller le travail de Bilal

--- 2012-02-09 klein2012/main.tex
@@
 \usepackage{natbib}
+\usepackage{caption}
 %%% francisation des algorithmes
@@
 \begin{abstract}
-  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse ({\it Inverse Reinforcement Learning, IRL}).
+  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse.
 Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
 L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) à partir de  traces du comportement de l'expert.
-Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui du {\it Reinforcement Learning, RL}) de manière répétée ou l'accès à la strategie complète de l'expert.
+Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
 Notre contribution consiste en un nouvel algorithme d'IRL levant l'ensemble de ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it Markov Decision Process, MDP}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
   \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
@@
 %%%%%
+Le problème de l'apprentissage par renforcement inverse a pour la première fois été introduit dans \cite{russell1998learning}.
+Formellement défini dans \cite{ng2000algorithms}, les approches proposées restent intractables jusqu'à la publication de \cite{abbeel2004apprenticeship}, qui lança toute une vague d'algorithmes cherchant une politique par modifications successives d'une fonction de récompense, approches pour la plupart résumées dans le \emph{survey} \cite{neu2009training}.
 \begin{itemize}
@@
 \item Rapide point sur les limitations des algos d'IRL actuels
+
 \begin{itemize}
+
 \item \cite{ng2000algorithms} : Probabilités de transitions devant être connues
@@
 \end{equation}
-où $\gamma$ est un facteur d'actualisation strictement plus petit que $1$. Il est également possible de définir une fonction voisine de la fonction de valeur, la fonction de qualité :
+où $\gamma$ est un facteur d'actualisation positif strictement plus petit que $1$. Il est également possible de définir une fonction voisine de la fonction de valeur, la fonction de qualité :
 \begin{equation}
@@
 \end{equation}
-  Une politique optimale est définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie :
+  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie :
 \begin{equation}
@@
 
-Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour lequel ce comportement est optimal (voir par exemple \cite{ng1999policy}).
+Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour laquelle ce comportement est optimal (voir par exemple \cite{ng1999policy}).
 Particulièrement, la récompense uniformément nulle remplit ce critère dans la mesure où tout comportement est optimal vis à vis de cette récompense.
 %%%
-\subsection{Considérations techniques}
+\subsection{Attribut vectoriel moyen}
 \label{ConsiderationsTechniques.sec}
@@
 \end{eqnarray}
-L'attribut vectoriel moyen de l'expert, $\mu_E$, pouvant être calculée sans forcément avoir accès à l'intégralité de la politique $\pi_E$, certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert, afin d'obtenir un agent dont le comportement à la même valeur que celui de l'expert (voir \cite{neu2009training}).
+L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
+Certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
+Cela permet d'obtenir un agent dont le comportement a la même valeur que celui de l'expert (voir \cite{neu2009training}).
 %%%%%
-\section{Loss-augmented Feature Expectation Matching}
+\section{Classification structurée pour l'apprentissage par renforcement inverse (SCIRL)}
 %%%%%
@@
 \end{equation}
-Nous allons utiliser cette propriété pour imaginer une fonction de risque dépendante d'une fonction $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
-Notons $(s_i,a_i)_i$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
+Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+Notons $(s_i,a_i,s_{i+1})_{i=1..N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
 \begin{equation}
@@
 \end{equation}
-Notre fonction de risque doit donc pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.\\
-Nous introduisons de surcroit une fonction de perte $l : S \times A \rightarrow \mathbb{R}_+$ dans notre fonction de risque qui peut alors s'assembler en :
+Notre fonction de risque doit donc pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant la marge qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
+Nous introduisons la fonction de risque empirique :
   \begin{equation}
-   R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right)
+   R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
    \end{equation}
-
-Penchons nous sur le rôle de $l$ : cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système.
-En effet, minimiser avec succès la fonction de risque avec un choix de $l$ judicieux permet d'imposer à $q$ une structure telle qu'il existe une marge entre l'action choisie par l'expert et la deuxième meilleure action.
-Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
 Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans les preuves (Section \ref{proof.sec}).\\
 
-Comme nous l'avons vu Equation \eqref{Qmu.eqn}, en faisant l'hypothèse d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense il est possible d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
+Nous traitons le problème de l'imitation comme un problème de classification multi-classe, ce qui a déjà été envisagé (\citep{ratliff2007imitation}).
+Les approches existantes en prennent toutefois pas en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnée par la politique de l'expert.
+C'est aussi pour cette raison que les approches purement supervisées ne peuvent fonctionner pour l'imitation.\\
+
+Comme nous l'avons vu Equation \eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
+Cette paramétrisation prend en compte la dynamique imposée par $\pi_E$.
 L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
-Le calcul de ce terme $\mu_E$, bien que présentant quelques subtilités ne pose pas de problème majeurs même lorsque les seules données auquel l'algorithme a accès consistent en une trace de l'expert (cela est expliqué en sous-section \ref{calculmu.sec}).
-Nous pouvons donc maintenant exprimer notre risque non plus comme une fonction de $q$ mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
+Le calcul de ce terme $\mu_E$ est central à notre algorithmes et nous proposons quelques solutions en sous-section \ref{calculmu.sec}.
+Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
+   \begin{equation}
+     \label{structure.eqn}
+   R_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
+   \end{equation}
+   Un sous-gradient de cette expression peut se calculer à l'aide notamment de trois propriétés que nous rappelons :
+\begin{itemize}
+  \item l'opérateur de sous-gradient est linéaire,
+  \item le gradient est l'unique sous-gradient d'une fonction différentiable,
+  \item la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
+\end{itemize}
+Ainsi, l'on a :
 \begin{eqnarray}
-   R_N(\theta)  &=& {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)\\
-   &=& {1\over N} \sum_{i=1}^N\left(\theta ^T \mu_E(s_i,a^*_i) + l(s_i,a^*_i) - \theta ^T \mu_E(s_i,a_i) \right)\\
+\nabla_\theta R_N(\theta) &=& {1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\\
 \textrm{avec } a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))
 \end{eqnarray}
-   Il est possible de minimiser ce risque grâce à la règle d'update qui suit, comme cela est expliqué en section \ref{proof.sec} :
+Une descente de sous-gradient sur le risque empirique produit alors la règle de mise-à-jour suivante :
 \begin{equation}
-   \theta_{t+1} = \theta_t -\alpha_t{{1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\over ||{1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)||_2}.
+   \theta_{t+1} \leftarrow \theta_t -\alpha_t{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2}.
 \end{equation}
@@
 \label{calculmu.sec}
-Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l''attribut vectoriel moyen.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul de l'attribut vectoriel moyen d'une politique arbitraire.
-De fait cela nous épargne le besoin d'un simulateur, d'un solveur de MDP ou d'échantillons couvrant tout l'espace d'état.\\
+Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitraire.
+Une différence notable, cependant, est que nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
+Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
 
-Selon les informations dont on dispose, plusieurs méthodes peuvent être utilisées pour calculer $\mu_E$.\\
+Cette inférence est facilité par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $j=1..p$, on trouve :
+\begin{equation}
+\mu_E^j(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)\right|s_0 = s, a_0 = a, \pi_E\right],
+\end{equation}
+expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis à vis de la récompense définie par la $j$-ème composante du vecteur d'attribut $\psi$.\\
 
-Dans le cas le plus facile, on peut utiliser la proximité entre la définition du vecteur $\mu$ (Equation \eqref{Qmu.eqn}) et celle de la fonction de valeur (Equation \eqref{Vdef.eqn}) pour adapter les algorithmes d'évaluation de la fonction de valeur dans le cadre de la programmation dynamique au calcul de l'attribut vectoriel moyen.\\
-En supposant un espace d'état de cardinal fini, l'équation de Bellman (Equation \eqref{Vdef.eqn}) peut s'écrire sous forme matricelle :
-\begin{equation}
-\mathbf V^\pi = \mathbf R + \gamma P_\pi\mathbf V^\pi
-\end{equation}
-Il est connu qu'une règle d'update de la forme 
-\begin{equation}
-\mathbf V^\pi_{t+1} \leftarrow \mathbf R + \gamma P_\pi\mathbf V^\pi_t
-\end{equation}
-convergera vers la valeur exacte recherchée.
-Ce sont les algorithmes du type \emph{value iteration} qui exploitent cette propriété.
-Parallèlement il est possible d'utiliser une règle d'update de la forme :
-\begin{equation}
-\mathbf \mu^\pi_{t+1} \leftarrow \mathbf\Phi + \gamma P_\pi\mathbf\mu^\pi_t
-\end{equation}
-pour obtenir la valeur exacte de l'attribut vectoriel moyen.\\
+Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu_E$.\\
+
+Dans le cas idéal où le modèle est connu, le calcul exact de $\mu_E$ peut être effectué, les algorithmes de programmation dynamique peuvent également être utilisés.
 
@@
 \begin{equation}
-\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]
+\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \psi(s)\right|s_0 = s, a_0 = a, \pi_E\right]
 \end{equation}
@@
 \begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\phi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
 \end{equation}
 
-Finalement, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ (\citep{klein2011batch}) dans sa version \emph{on-policy}.
-Cet algorithme se base également sur la similarité entre les definitions de $\mu$ et $Q$ mais au lieu d'adapter un algorithme de programmation dynamique comme nous venons d'en donner l'exemple, il adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch} et {\it offline}.
-Une approximation de la forme $\hat{\mu}_i^\pi(s) = (\xi_i^*)^T \psi(s)$ pour chacune des composantes $i$ de $\mu^\pi$ est obtenue grâce à la solution des moindres carrés :
-\begin{equation}
-  \xi_i^* = \left(\sum_{t=1}^n
-  \psi(s_t)(\psi(s_t)-\gamma\psi(s_{t+1}))^T\right)^{-1}
-  \sum_{t=1}^n \psi(s_t) \phi_i(s_t).
-\end{equation}
+Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
+Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
+C'est la méthode que nous illustrons dans la section \ref{exp.sec}\\
 
+Finalement, il est possible de travailler avec une approximation de Monte-Carlo plus fine que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
+Cela est reporté pour des travaux futurs.
 %%%%%
@@
 %%%%%
-\section{Related Work}
+\section{Travaux Connexes}
 \label{biblio.sec}
 %%%%%
-Parmi les algorithmes d'apprentissage par renforcement inverse existants, certains partagent des caractéristiques communes.\\
+L'approche historique de \cite{ng2000algorithms}, ainsi que les algorithmes basés sur \cite{abbeel2004apprenticeship}, notamment \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée.
+Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de l'expert et l'attribut vectoriel moyen d'une politique donnée.
+Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
 
-Historiquement, la première approche générique est celle d'\cite{abbeel2004apprenticeship}, travail qui a servi de bases à d'autres algorithmes fonctionnant sur le même principe : \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}.
-L'approche consiste à imiter l'expert par le biais de l'IRL ; les attributs vectoriels moyens de l'agent et de l'expert doivent être proche (l'on a vu Equation \eqref{memevaleur.eqn} que cela impliquait des valeurs similaires).
-Cette proximité est atteinte en modifiant la récompense par itérations successives et en entraînant l'agent sur la nouvelle récompense.\\
+Cette résolution est très problématique, en effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la litérature.
+Particulièrement, l'information nécessaire à cet objectif est bien plus étendue que celle nécessaire au calcul d'uniquement l'attribut moyen de l'expert comme cela est nécessaire à notre algorithme.
+Comme cela est illustré en section \ref{exp.sec}, l'algorithme d'\citet{abbeel2004apprenticeship} est incapable de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état.\\
 
-Une autre approche consiste à effectuer la recherche dans l'espace des politiques grâce à de l'inférence bayesienne sur les paires état-action fournies par l'expert, comme par exemple \cite{ramachandran2007bayesian}.\\
+Les approches bayésiennes du type de \cite{ramachandran2007bayesian} nécessitent une connaissance du modèle pour fonctionner, modèle dont nous ne disposons pas.\\
 
-Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un optimal par une certaine marge, c'est à dire que l'action choisie par l'expert doit avoir une qualité supérieure d'une certaine marge à n'importe quelle autre action.
-Il s'agit de maximiser cette marge.
-C'est l'approche de \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical}.\\
+Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
+C'est à dire que l'action choisie par l'expert doit avoir une qualité strictement supérieure à n'importe quelle autre action.
+Les approches de planification structurée comme \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} permettent d'influer sur cette marge.\\
 
-Particulièrement, le travail de \citet{ratliff2006maximum} partage des similarités avec le nôtre ; dans \citep{ratliff2006maximum}, les auteurs présentent une méthode d'IRL en posant le problème sous forme d'un \emph{maximum margin classification problem}.
+Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
+Les auteurs présentent une méthode d'IRL en posant le problème sous forme d'un \emph{maximum margin classification problem}.
 La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}, comme cela est expliqué dans \cite{neu2009training}.
 Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un MDP (entrée à classifier) est associée une politique (classe).
-
-   Une seconde publication, \citep{ratliff2007boosting} vient compléter la précédente en incluant du \emph{boosting}.
-On y voit apparaître à nouveau une fonction de coût empirique (en omettant quelques détails comme par exemple la pondération des trajectoires) :
-   \begin{equation}
-   \label{dix.eqn}%On l'appelle comme ça parce que dans le draft et dans le papier de Ratliff elles ont le même numéro : 10.
-   {1\over n}\sum\limits_{i=1}^n \left(\theta^T\mu^E_i - \min_\pi\left(\theta^T \mu^\pi - l(\pi)\right)\right) + {\lambda \over 2} ||\theta||^2.
-   \end{equation}
-
-   On constate la présence d'une fonction de coût $l$ similaire à celle que nous utilisons.
-
-   La recherche a lieu dans l'espace des politiques via une méthode de descente de gradient sur les paramètres $\theta$ définissant la récompense, ce qui implique la résolution d'un MDP à chaque itération.
-En effet, le $\min\limits_\pi(\dots)$ présent dans la fonction de coût est analogue au $\pi^* = \arg\max\limits_\pi V^\pi$ décrivant la résolution classique d'un MDP.
+Cela impose également un cadre déterministe pour les MDP, l'algorithme de résolution utilisé étant $A^*$.\\
 
 Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
-Cette fonction de score $s : X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque entrée pour prendre une décision de classification i.e.
-choisir le label $y\in Y$ à appliquer à l'entrée $x\in X$ :
-   \begin{equation}
-   \label{sdef.eqn}
-   y^* = \arg\max_{y \in Y} s(x,y)
-   \end{equation}
-   Cette fonction de score est apprise dans l'espace fonctionnel en optimisant la fonction suivante, dont celle donnée en \ref{dix.eqn} est une instanciation : 
-   \begin{equation}
-   \label{rdef.eqn}
-   r[s] = {1\over N} \sum_{i=1}^N\left(\max_{y\in Y}(s(x_i,y) + l(x_i,y)) - s(x_i,y_i) \right)
-   \end{equation}
-   Cela suppose l'accès à une base de données $\{(x_i,y_i)\}_i$; i.e.
-une trace de l'expert.
-On note dans ce papier un changement par rapport aux papiers précédent : les entrées sont maintenant les états et les classes représentent les actions.
+Cette fonction de score $s : X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e.
+choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$.
+Notre fonction de risque est similaire à la leur, mais outre la résolution répétée du PDM liée à la manière dont ils minimisent la fonction de risque et dont nous avons déjà parlé, leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Equation \eqref{structure.eqn}).
 
 %%%%%
-\section{Résultats expérimentaux}
+\section{Illustration expérimentale}
 \label{exp.sec}
@@
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.\\
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.
+Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.
+On constate que cet espace d'état respecte le critère de Markov dont nous avons vu en section \ref{back.sec} qu'il était nécessaire.\\
 
-La politique ainsi trouvée parvient systématiquement à maintenir le pendule en équlibre durant 3000 pas de temps.
+La politique optimisant la fonction de récompense donnée figure \ref{trueR.fig} parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
 Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
@@
 Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
-Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale).\\
+Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale et son voisinage proche).\\
 
  Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
-En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir quelques transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
+En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
 Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel moyen est utilisée en association à celle de l'expert.
@@
 \begin{figure}
-  \begin{minipage}[t]{.4\linewidth}
+  \begin{minipage}[t]{.5\linewidth}
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_true_R.pdf}
-       \caption{Récompense sur laquelle l'expert a été entraîné}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_true_R.pdf}
+       \caption{}
        \label{trueR.fig}
@@
 \hfill
-\begin{minipage}[t]{.4\linewidth}
+\begin{minipage}[t]{.5\linewidth}
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_lafem_R.pdf}
-       \caption{Récompense trouvée par LAFEM, en vert les positions occupées par l'expert.}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_lafem_R.pdf}
+       \caption{}
        \label{lafemR.fig}
@@
 \end{minipage}\\
-\begin{minipage}[t]{.4\linewidth}
+\begin{minipage}[t]{.5\linewidth}
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_Vexpert.pdf}
-       \caption{Fonction de valeur de l'expert}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_Vexpert.pdf}
+       \caption{}
        \label{trueV.fig}
@@
 \hfill
-\begin{minipage}[t]{.4\linewidth}
+\begin{minipage}[t]{.5\linewidth}
     \begin{center}
-       \includegraphics[width=1.2\textwidth]{LAFEM_Exp3_Vagent.pdf}
-       \caption{Fonction de valeur de l'agent entraîné sur la récompense trouvée par LAFEM}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_Vagent.pdf}
+       \caption{}
        \label{lafemV.fig}
@@
 \end{minipage}\\
+\caption*{Illustration de l'application de notre algorithme au problème du pendule inversé.
+Un agent entraîné grâce à un algorithme d'apprentissage par renforcement sur la récompense de la Figure \ref{trueR.fig} parvient à maintenir le pendule en équilibre durant 5 minutes sans échec.
+Il joue le rôle de l'expert.
+La fonction de valeur de cet expert est illustrée Figure \ref{trueV.fig}.
+On constate la similarité avec la Figure \ref{lafemV.fig}, fonction de valeur de l'agent qui lui a été entraîné sur la récompense trouvée par notre algorithme, Figure \ref{lafemR.fig}.
+Bien que les échantillons fournis par l'expert (en vert sur la Figure \ref{lafemR.fig}) ne couvrent qu'une petite partie de l'espace d'état (position angulaire du pendule en abscisse, vitesse angulaire en ordonnées) notre algorithme parvient à l'aide de ces seules données à trouver une récompense justifiant le comportement de l'expert sur tout l'espace d'état.}
 \end{figure}
-Afin d'étudier le comportement de notre algorithmes, nous avons établi le protocole expérimental suivant :
-\begin{itemize}
- \item créer une base de données $D_0$ de transitions aléatoires ;
- \item entraîner un expert en utilisant LSPI nourri avec ces transitions ( definissant ainsi $\pi_E : S\rightarrow A$) ;
- \item vérifier que l'expert est en mesure de balancer le pendule durant 3000 pas de temps ;
- \item créer une base de données $D_E$ de transitions de l'expert, contenant 10 trajectoires de 300 transitions chacune ;
- \item définir $l$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon, $l$ est définie sur les états présents dans $D_E$ ;
- \item définir $\alpha(t) = 0.1,\forall t$ ;
- \item définir $D$ à partir des transitions $D_E$ en enlevant l'état d'arrivée ;
- \item initialiser $\omega_0 = [-1...-1]^T$ ;
- \item fixer $T=20$ ;
- \item définir $\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \phi(s)\right|s_0 = s, a_0 = a, \pi_E\right]$ comme étant en pratique $\hat\mu_E(s,a) =  \omega^T_{\pi_E}\phi(s,a), \omega_{\pi_E} = LSTD\mu(D_E)$ ;
- \item faire tourner LAFEM ;
- \item entrainer un agent sur le problème du pendule inversé, avec la récompense trouvée par LAFEM : définir $\pi : S\rightarrow A$ ;
- \item vérifier que l'agent parvient lui aussi à maintenir le pendule en équilibre durant 3000 pas de temps.
-\end{itemize}
+Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
+La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon.
+Le pas de temps est constant à $\alpha(t) = 0.1, \forall t$ et le nombre d'itérations est de $T=20$.
+Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi : S \rightarrow \mathbb{R}^p$ correspondant au mélange de gaussiennes défini dans \cite{lagoudakis2003least}.
+Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
 
@@
 En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
-Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs proposés dans le problème du pendule inversé par \citet{lagoudakis2003least} consistent en un réseau de gaussiennes qui ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.\\
+Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs proposés dans le problème du pendule inversé par \citet{lagoudakis2003least} et utilisés par nous paramétrer la récompense consistent en un maillage de gaussiennes qui ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.\\
 
@@
 \section{Conclusion and future work}
+
+L'algorithme présenté dans cette publication lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse.
+En rendant la résolution du problème inverse indépendante de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
+
+La complexité calculatoire de notre algorithme instancié avec LSTD$\mu$ reste sous contrôle. L'étape gourmande étant l'inversion d'une matrice de taille $p \times p$.
 \begin{itemize}

### 2012-02-10 klein2012/main.tex
new
new-final

@@
 \subsection{Hypothèses}
-Dans cette section, nous allons présenter le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
-On supposera que la récompense $R$ recherchée s'écrive sous la forme $R(s)=\theta_E^T\phi(s)$ où $\theta_E\in\mathbb{R}^p$.
-De plus, on supposera que notre base d'exemples est telle que $(S\times A) \subset (s_i,a_i)_{1\leq i \leq N}$.
+Ici, on présente le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
+On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$.
+De plus, on supposera que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$.
 Finalement la récompense $R$ devra vérifier la propriété suivante :
@@
 \end{equation}
-où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$, qui est légèrement plus forte que l'équation d'optimalité de Bellman puisque l'équation \ref{eq:existence} implique :
+où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$, qui est légèrement plus forte que l'équation d'optimalité de Bellman puisque l'équation \eqref{eq:existence} implique : $\forall s\in S, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
+Ainsi tout vecteur $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale.
+Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation \eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
+Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation \eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+\subsection{Résultats}
+On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $R_N$.
+On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
+Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
+Il est facile de borné inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
+Montrons qu'il existe un $x\geq0$ tel que $\forall~1~\leq~i~\leq~N,~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
+\begin{align}
+&\hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{a\in A \backslash a_i}l(s_i,a)\geq0,\\
+\Leftrightarrow&x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
+\end{align}
+Or comme la somme des $\max$ est supérieur au $\max$ d'une somme, il est clair que $\forall 1\leq i \leq N$:
+\begin{align}
+&x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},\\
+\Leftrightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{a\in A \backslash a_i}l(s_i,a)\geq0,\\
+\Rightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,\\
+\Leftrightarrow&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
+\end{align}
+C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $R_N$ vérifient bien l'équation \eqref{eq:existence}. 
+Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Or on a $\forall 1\leq i \leq N$ :
+\begin{align}
+&R_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),\\
+\Leftrightarrow&\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,\\
+\Rightarrow&(\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
+\end{align}
+D'ou ~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
+Finalement, montrons que l'algorithme de sous-gradient converge.
+Tout d'abord, il est clair que $R_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
+De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article \cite{correa1993convergence} qui nous assure que la suite :
 \begin{equation}
-\forall s\in S, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a).
+\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\frac{1}{t+1}{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2},
 \end{equation}
-Ainsi tout vecteur $\theta$ vérifiant l'équation \ref{eq:existence} rendra la politique de l'expert optimale.
-Dans le cadre que nous venons d'établir, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\ref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation \ref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
-\subsection{Résultats}
-On va montrer dans un premier temps qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \ref{eq:existence} et qui minimise $R_N$.
-On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
-Désormais, on imposera que notre fonction de perte vérifie :
-\begin{equation}
-\forall 1\leq i \leq N, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0.
-\end{equation}
-Il est facile de borné inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a :
-\begin{equation}
-R_N(\theta)\geq{1\over N} \geq {1\over N} \sum_{i=1}^N l(s_i,a_i).
-\end{equation}
-Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \ref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$, et ~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \ref{eq:existence} }\}$ et posons ~$\forall~1\leq i \leq N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
-Montrons qu'il existe un $x\geq0$ tel que $\forall 1\leq i \leq N, R_N(\hat{\theta}_x)= {1\over N} \sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x\in M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.\\
-Avec les notations précédentes, on vérifie:
-\begin{align}
-&\forall 1\leq i \leq N, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,
-\\
-\Leftrightarrow &\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
-\end{align}
-Or comme ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)~\geq~\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)~+~l(s_i,a))$, il est clair que :
-\begin{align}
-&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},\\
-\Leftrightarrow&\forall 1\leq i \leq N, \hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{A \backslash a_i}l(s_i,a)\geq0,\\
-\Rightarrow&\forall 1\leq i \leq N, \hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,\\
-\Leftrightarrow&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).\\
-\end{align}
-C'est bien le résultat recherché.
-De plus on peut remarquer, que si l'on choisi la fonction de perte telle que :
-\begin{equation}
-\forall 1\leq i \leq N, \forall a \in A \backslash a_i, l(s_i,a)=c_i=\max_{a\in A \backslash a_i}l(s_i,a)>l(s_i,a_i)\geq0, \text{ où $c_i$ est une constante réelle,}
-\end{equation}
-alors on a ~$\max_{A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,)+\max_{a\in A \backslash a_i}l(s_i,a)~=~\max_{A \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,)~+~l(s_i,a))$. Et donc :
-\begin{align}
-&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
-\\
-\nonumber\Updownarrow
-\\
-&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
-\end{align}
-Majorons les $\beta_i$ :
-\begin{equation}
-\forall 1\leq i \leq N, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\leq\max_{A \backslash a_i}(\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2).
-\end{equation}
-Donc :
-\begin{align}
-&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i),
-\\
-\nonumber\Updownarrow
-\\
-&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
-\\
-\nonumber\Downarrow
-\\
-&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}.
-\end{align}
-Ainsi, cela impose que la solution $\hat{\theta}_x$ soit au moins de norme $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\max_{A \backslash a_i}\|\mu_E(s_i,a_i)-\mu_E(s_i,a)\|_2}$.\\
-Minorons les $\beta_i$ :
-\begin{equation}
-\forall 1\leq i \leq N, \beta_i=\hat{\theta}_1^T\mu_E(s_i,a_i)-\max_{A \backslash a_i}\hat{\theta}_1^T\mu_E(s_i,a)\geq|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|.
-\end{equation}
-Donc :
-\begin{align}
-&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i),
-\\
-\nonumber\Updownarrow
-\\
-&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},
-\\
-\nonumber\Uparrow
-\\
-&\forall 1\leq i \leq N, x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|}.
-\end{align}
-Ainsi cela impose que si $\hat{\theta}_x$ est de norme supérieure à $\max_{i\in\{1,\dots,N\}}\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{|\|\mu_E(s_i,a_i)\|_2-\max_{A \backslash a_i}\|\mu_E(s_i,a)\|_2|}$, alors il est un minimiseur de $R_N$.
-
-Revenons au premier type de fonction de loss :
-\begin{equation}
-\forall 1\leq i \leq N, \forall a \in A \backslash a_i, l(s_i,a)>l(s_i,a_i)\geq0.
-\end{equation}
-On peut montrer, toujours sous l'existence de ~$\hat{\theta}$ que $M^*\subset\hat{M}$.
-\begin{preuve}
-Soit ~$\theta^*\in\ M^*$.
-Sous l'existence de ~$\hat{\theta}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
-Or on a :
-\begin{align}
-&R_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),
-\\
-\nonumber\Updownarrow
-\\
-&\forall 1\leq i \leq N, (\theta^*)^T\mu_E(s_i,a_i) + l(s_i,a_i)\geq\max_{a \in A \backslash a_i}((\theta^*)^T\mu_E(s_i,a)+l(s_i,a))\geq0,
-\\
-\nonumber\Updownarrow
-\\
-&\forall 1\leq i \leq N,\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,
-\\
-\nonumber\Downarrow
-\\
-&\forall 1\leq i \leq N, (\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
-\end{align}
-D'ou ~$\theta^*\in \hat{M}$. C'est bien ce que l'on recherchait.
-\end{preuve}
-Ainsi on sait que, sous l'existence de ~$\hat{\theta}$, les minimiseurs de ~$\R_N$ existent (il y en a d'ailleurs une infinité) et vérifient l'équation \ref{eq:existence}.
-On peut donc appliquer l'algorithme de descente de sous-gradient avec l'assurance de la convergence de ce dernier vers un minimiseur de la fonction de risque qui vérifie l'équation \ref{eq:existence}.
-L'équation \ref{eq:existence} est la formulation mathématique de ce que l'on recherche pour vecteur de récompense $\theta$.
-La politique optimale obtenue à partir de la récompense définie par un minimiseur $\theta^*$ de $R_N$, sous l'existence d'un ~$\hat{\theta}$, aura la même fonction de valeur sur les états $s_i,\forall 1\leq i \leq N$ que la politique de l'expert car cette dernière est optimale sur les états $s_i,\forall 1\leq i \leq N$ puisque $\theta^*$ vérifie l'équation \ref{eq:existence}.
-%%%%%
+est convergente vers un minimiseur $\theta^*$ de $R_N$.
+Ce minimiseur vérifiera l'équation \eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation \eqref{eq:existence} implique l'optimalité de Bellman.
 \section{Travaux Connexes}

### 2012-02-10 klein2012/main.tex
Pour pouvoir puller le travail de Bilal

--- 2012-02-10 klein2012/main.tex
@@
 
-\shorttitle{Classification structurée pour l'IRL}
+\shorttitle{Classification structurée pour l'ARI}
 
@@
 \institute{
-Sup\'elec,\\
-IMS Research group, France\\
-\texttt{prenom.nom@supelec.fr}
+Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
 \and
-Equipe ABC,\\
+Equipe ABC,
 LORIA, France
 \and
-UMI 2958\\
+UMI 2958
 GeorgiaTech-CNRS, France
@@
 \begin{abstract}
-  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse.
-Dans ce contexte, un expert accomplit une tâche qu'un agent doit essayer de reproduire.
-L'IRL part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) à partir de  traces du comportement de l'expert.
-Les algorithmes d'IRL existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
-Notre contribution consiste en un nouvel algorithme d'IRL levant l'ensemble de ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et ne nécessitant pas la résolution du problème direct.
-  \motscles{Apprentissage par renforcement, Apprentissage par renforcement inverse, Processus Decisionnel de Markov}
+  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse (\emph{ARI}).
+Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
+L'ARI part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) à partir de  traces du comportement de l'expert.
+Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
+Notre contribution consiste en un nouvel algorithme d'ARI levant l'ensemble de ces contraintes.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et surtout ne nécessitant pas la résolution du problème direct.
+  \motscles{Apprentissage par renforcement inverse, classification multi-classe, attribut vectoriel moyen}
 \end{abstract}
@@
 %%%%%
+Le problème de l'imitation se pose dans des champs variés tels la robotique, les systèmes de dialogue ou la conception d'interfaces homme-machine.
+Un expert effectue une tâche, le but est de parvenir à faire un sorte qu'un agent parvienne à accomplir la même tâche en observant l'expert.
+Dans le cadre de la prise de décisions séqueltielles, l'apprentissage par renforcement est une approche connaissant le succès.
+Cette approche consiste à guider un agent par le biais d'une récompense.
+Le mécanisme inverse peut servir de cadre au problème de l'imitation, en effet si l'on parvient à retrouver la récompense optimisée par un expert alors la résolution du problème direct donnera naissance à un comportement imitant cet expert.\\
+
 Le problème de l'apprentissage par renforcement inverse a pour la première fois été introduit dans \cite{russell1998learning}.
-Formellement défini dans \cite{ng2000algorithms}, les approches proposées restent intractables jusqu'à la publication de \cite{abbeel2004apprenticeship}, qui lança toute une vague d'algorithmes cherchant une politique par modifications successives d'une fonction de récompense, approches pour la plupart résumées dans le \emph{survey} \cite{neu2009training}.
-\begin{itemize}
-\item Introduction informelle sur les processus de décision séquentielles
-\item Intérêt de l'apprentissage par démonstration
-\item Intérêt de l'IRL pour faire de l'apprenticeship learning
-\item Rapide point sur les limitations des algos d'IRL actuels
-
-\begin{itemize}
-
-\item \cite{ng2000algorithms} : Probabilités de transitions devant être connues
-\item \cite{abbeel2004apprenticeship} et toutes les approches résumées dans \cite{neu2009training}: Résolution du problème direct à chaque itération
-\end{itemize}
-\end{itemize}
+Formellement défini dans \cite{ng2000algorithms}, les approches proposées restent intractables jusqu'à la publication de \cite{abbeel2004apprenticeship}, qui lança toute une vague d'algorithmes cherchant une politique par modifications successives d'une fonction de récompense, approches pour la plupart résumées dans la vue d'ensemble fournie par \cite{neu2009training}.
 %%%%%
@@
 Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
-La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$, qui respecte donc le critère de Markov.
+La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$.
 Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
-Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité $p(s_{t+1}|s_t, a_t)$.
+Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité markovienne $p(s_{t+1}|s_t, a_t)$.
 Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité.} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
@@
 \label{Vdef.eqn}
-V^\pi(s) = E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,\pi\right]=R(s) + \gamma E[V^\pi(s')|s,\pi]
+V^\pi(s) = E[\sum_t\gamma^tR(s_t)|s_0=s,\pi]=R(s) + \gamma E[V^\pi(s')|s,\pi]
 \end{equation}
-où $\gamma$ est un facteur d'actualisation positif strictement plus petit que $1$. Il est également possible de définir une fonction voisine de la fonction de valeur, la fonction de qualité :
+où $\gamma \in [0,1)$ est un facteur d'actualisation.
+Il est également possible de définir une fonction de qualité qui ajoute à la fonction de valeur un degré de liberté sur le choix de l'action suivante :
 \begin{equation}
 \label{Qdef.eqn}
-Q^\pi(s,a) = E\left[\left.\sum_t\gamma^tR(s_t)\right|s_0=s,a_0=a,\pi\right] = R(s) + \gamma E[Q^\pi(s',\pi(s'))|s,a].
+Q^\pi(s,a) = E[\sum_t\gamma^tR(s_t|s_0=s,a_0=a,\pi] = R(s) + \gamma E[Q^\pi(s',\pi(s'))|s,a].
 \end{equation}
-  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie :
-\begin{equation}
-\forall \pi, \forall s, V^*(s) \geq V^\pi(s).
-\end{equation}
+  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
 La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
@@
 \end{equation}
-L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov ({\it Markov Decision Process, MDP}).\\
+L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov (\emph{PDM}).\\
 
@@
 Un exemple de ce type de tâche est la conduite d'une voiture.
-Nous accomplissons ce genre de choses au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
+Nous accomplissons ce genre de chose au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite.
 Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
@@
 Le problème est de retrouver la fonction de récompense optimisée par un expert.
-Usuellement l'expert est considéré comme un agent optimal dans un MDP et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
+Usuellement l'expert est considéré comme un agent optimal dans un PDM et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
 
 Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour laquelle ce comportement est optimal (voir par exemple \cite{ng1999policy}).
-Particulièrement, la récompense uniformément nulle remplit ce critère dans la mesure où tout comportement est optimal vis à vis de cette récompense.
+Particulièrement, la récompense uniformément nulle remplit ce critère dans la mesure où tout comportement est optimal vis-à-vis de cette récompense.
+Il convient donc de contraindre l'espace des solutions.
 %%%
@@
 %%%
-La récompense $R$ étant l'inconnue du problème, il est naturel de l'approximer grâce à un vecteur de poids $\theta$ et des attributs vectoriels $\psi : S\rightarrow \mathbb{R}^p$ servant de fonctions de base :
+La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$ telles que 
 \begin{equation}
 \label{hatRdef.eqn}
-\hat R(s) = \theta^T\psi(s)\approx R(s).
+\exists \theta | R(s) = \theta^T\psi(s) = \sum_{i=1}^N\theta_i\psi_i(s).
+\end{equation}
+Comme nous allons le voir, cette contrainte peut cependant être relâchée si nécessaire.
+Introduire cette expression dans la définition de la fonction de qualité (Equation \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+\begin{equation}
+Q^\pi(s,a) = \theta^T\mu^\pi(s,a) \textrm{ avec } \mu^\pi(s,a) = E[\sum_t\gamma^t\psi(s_t)|s_0=s,a_0=a,\pi]
+\label{Qmu.eqn}
 \end{equation}
 
-Introduire cette expression dans la définition de la fonction de qualité (Equation \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
-\begin{eqnarray}
-Q^\pi(s,a) &=& E\left[\left.\sum_t\gamma^t\theta^T\psi(s_t)\right|s_0=s,a_0=a,\pi\right]\\
- &=& \theta^TE\left[\left.\sum_t\gamma^t\psi(s_t)\right|s_0=s,a_0=a,\pi\right]\\
- &=& \theta^T\mu^\pi(s,a)
-\label{Qmu.eqn}
-\end{eqnarray}
+Le terme $\mu^\pi$ est appelé la {\it feature expectation} ou plutôt l'attribut vectoriel moyen d'une politique.
+On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
 
-Le terme $\mu^\pi$ est appelé la {\it feature expectation} ou plutôt l'attribut vectoriel moyen d'une politique.
-On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le MDP fixe la paramétrisation de $Q^\pi$.\\
-
-On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis à vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
+On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
 En effet
@@
 %%%%%
-\section{Classification structurée pour l'apprentissage par renforcement inverse (SCIRL)}
+\section{Classification structurée pour l'apprentissage par renforcement inverse (SCARI)}
 %%%%%
@@
 %%%
-L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis à vis d'une certaine fonction de récompense que l'on cherche à retrouver.
+L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis-à-vis d'une certaine fonction de récompense que l'on cherche à retrouver.
 Cela implique, comme nous l'avons vu Equation \eqref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
@@
 \end{equation}
-expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis à vis de la récompense définie par la $j$-ème composante du vecteur d'attribut $\psi$.\\
+expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attribut $\psi$.\\
 
@@
 Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
-Les auteurs présentent une méthode d'IRL en posant le problème sous forme d'un \emph{maximum margin classification problem}.
+Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}.
 La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}, comme cela est expliqué dans \cite{neu2009training}.
-Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un MDP (entrée à classifier) est associée une politique (classe).
-Cela impose également un cadre déterministe pour les MDP, l'algorithme de résolution utilisé étant $A^*$.\\
+Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un PDM (entrée à classifier) est associée une politique (classe).
+Cela impose également un cadre déterministe pour les PDM, l'algorithme de résolution utilisé étant $A^*$.\\
 
@@
 Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel moyen est utilisée en association à celle de l'expert.
-Cette politique est obtenue par optimisation d'une récompense sur le MDP dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
+Cette politique est obtenue par optimisation d'une récompense sur le PDM dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
 Comme notre algorithme ne nécessite pas le calcul d'autre attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
@@
 Si la fonction de récompense trouvée par LAFEM (figure \ref{lafemR.fig}) diffère de celle fournie à l'expert (figure \ref{trueR.fig}) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{trueV.fig} et \ref{lafemV.fig}).
-Cela est une illustration du fait que le problème de l'IRL est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
+Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
 
@@
 
-La complexité calculatoire de notre algorithme instancié avec LSTD$\mu$ reste sous contrôle. L'étape gourmande étant l'inversion d'une matrice de taille $p \times p$.
-\begin{itemize}
-\item On enlève les contraintes du domaine d'un seul coup,
-\item Il ne reste que la sélection ou détection de features, mais ce n'est pas hors de portée
-\item Faible complexité computationnelle
-\item Faible complexité en échantillons
-\item Il reste des tests à faire sur le Highway driving (Pour ICML ?) et sur des expériences dans la vraie vie (bras robot ?) pour comparer à l'existant
-\end{itemize}
+La complexité calculatoire de notre algorithme instancié avec LSTD$\mu$ reste sous contrôle, létape gourmande étant l'inversion d'une matrice de taille $p \times p$.\\
+
+Comme nous l'avons illustré, les échantillons nécessaires à la réussite de notre algorithmes sont faciles à recueillir.
+Une simple trace de l'expert, même si elle ne couvre qu'une petite partie de l'espace d'état, suffit à inférer une récompense.
+Cette récompense, description compacte de la tâche effectuée par l'expert, peut alors être optimisée par un agent dont les capacités diffèrent de celles de l'expert.
+On rentre dans le cadre du \emph{transfer learning}.\\
+
+Bien que ne posant pas particulièrment problème, le point central de notre algorithme qu'est le calcul de $\mu_E$ gagnerait à être amélioré.
+L'axe d'étude que nous envisageons d'explorer serait de s'en passer purement et simplement en trouvant le moyen d'introduire la structure du PDM dans la démarche de classification sueprvisée.\\
+
+Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.
 %

### 2012-02-10 klein2012/main.tex
new-superfinal

@@
 C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $R_N$ vérifient bien l'équation \eqref{eq:existence}.
-Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$\x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Or on a $\forall 1\leq i \leq N$ :

### 2012-02-10 klein2012/main.tex
new-final+

@@
 Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
-Il est facile de borné inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Il est facile de borner inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
 Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
-Montrons qu'il existe un $x\geq0$ tel que $\forall~1~\leq~i~\leq~N,~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+Montrons qu'il existe un $x\geq0$ tel que $~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
 Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
@@
 \end{align}
-Or comme la somme des $\max$ est supérieur au $\max$ d'une somme, il est clair que $\forall 1\leq i \leq N$:
+Or comme la somme des $\max$ est supérieure au $\max$ d'une somme, il est clair que $\forall 1\leq i \leq N$:
 \begin{align}

### 2012-02-10 klein2012/main.tex
Pour puller le travail de Bilal

--- 2012-02-10 klein2012/main.tex
@@
 
-Le terme $\mu^\pi$ est appelé la {\it feature expectation} ou plutôt l'attribut vectoriel moyen d'une politique.
+Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique.
 On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
@@
 L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
-Certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
-Cela permet d'obtenir un agent dont le comportement a la même valeur que celui de l'expert (voir \cite{neu2009training}).
+La plupart des algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
+Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert (voir par exemple \cite{neu2009training}).
 %%%%%
-\section{Classification structurée pour l'apprentissage par renforcement inverse (SCARI)}
+\section{Classification structurée pour l'apprentissage par renforcement inverse}
 %%%%%
@@
 %%%
-L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis-à-vis d'une certaine fonction de récompense que l'on cherche à retrouver.
-Cela implique, comme nous l'avons vu Equation \eqref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
+Comme nous l'avons vu Equation \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue :
 \begin{equation}
@@
 Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
-Notons $(s_i,a_i,s_{i+1})_{i=1..N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
+Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
+Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons :
 \begin{equation}
@@
 \end{equation}
-Notre fonction de risque doit donc pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
-Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant la marge qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
 Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
-Nous introduisons la fonction de risque empirique :
+Nous introduisons donc la fonction de risque empirique suivante :
   \begin{equation}
@@
    \end{equation}
-Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans les preuves (Section \ref{proof.sec}).\\
+Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans l'analyse (Section \ref{proof.sec}).\\
 
@@
 \end{equation}
-expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attribut $\psi$.\\
+expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.\\
 

### 2012-02-10 klein2012/main.tex
Version soumise Vendredi soir à CAp

@@
 
-Nous traitons le problème de l'imitation comme un problème de classification multi-classe, ce qui a déjà été envisagé (\citep{ratliff2007imitation}).
-Les approches existantes en prennent toutefois pas en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnée par la politique de l'expert.
-C'est aussi pour cette raison que les approches purement supervisées ne peuvent fonctionner pour l'imitation.\\
+Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
+Notre propose n'est cependant pas d'utiliser une approche purement supervisée.
+Nous ne souhaitons pas apprendre le contrôle de l'expert mais la description de la tâche qu'il effectue (la récompense).
+Il faut pour cela prendre en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnées par la politique de l'expert.\\
 
@@
 L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
-Le calcul de ce terme $\mu_E$ est central à notre algorithmes et nous proposons quelques solutions en sous-section \ref{calculmu.sec}.
-Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
+Le calcul de ce terme $\mu_E$ est central à notre algorithmes et nous proposons quelques solutions en sous-section \ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
    \begin{equation}
@@
    \end{equation}
-   Un sous-gradient de cette expression peut se calculer à l'aide notamment de trois propriétés que nous rappelons :
-\begin{itemize}
-  \item l'opérateur de sous-gradient est linéaire,
-  \item le gradient est l'unique sous-gradient d'une fonction différentiable,
-  \item la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
-\end{itemize}
-Ainsi, l'on a :
+   Une descente de gradient devrait permettre de minimiser cette expression.
+Cependant, la présence de l'opérateur $\max$ rend les choses un peu plus complexes car l'expression n'est plus différentiable.
+Il va falloir faire appel à une généralisation plus puissante, le sous-gradient.
+Le sous-gradient d'une fonction convexe $f:W\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que :
+\begin{equation}
+\forall x' \in W, g^T(x'-x)\leq f(x')-f(x)
+\end{equation}
+Le sous-gradient n'est pas partout unique, cependant l'unique sous-gradient d'une fonction différentiable est son gradient.
+Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
+En appliquant ces règles, on obtient :
 \begin{eqnarray}
@@
 Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitraire.
-Une différence notable, cependant, est que nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
+Nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
 Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
 
-Cette inférence est facilité par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $j=1..p$, on trouve :
+Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $j=1..p$, on trouve :
 \begin{equation}
-\mu_E^j(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)\right|s_0 = s, a_0 = a, \pi_E\right],
+\mu_E^j(s,a) = E[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)|s_0 = s, a_0 = a, \pi_E],
 \end{equation}
-expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.\\
-
+expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
 Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu_E$.\\
@@
 
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a)$ (définition Equation \eqref{Qmu.eqn}) l'on utilise
 \begin{equation}
-\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \psi(s)\right|s_0 = s, a_0 = a, \pi_E\right]
-\end{equation}
- l'on utilise
-\begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\{\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, s_{t+1} \sim p(\cdot|s_t,\pi_E(s_t))\}.
 \end{equation}
@@
 Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
-C'est la méthode que nous illustrons dans la section \ref{exp.sec}\\
+C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu_E$ en utilisant LSTD.
+Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
 
-Finalement, il est possible de travailler avec une approximation de Monte-Carlo plus fine que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
+D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
 Cela est reporté pour des travaux futurs.
@@
 %%%%%
-L'approche historique de \cite{ng2000algorithms}, ainsi que les algorithmes basés sur \cite{abbeel2004apprenticeship}, notamment \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée.
-Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de l'expert et l'attribut vectoriel moyen d'une politique donnée.
-Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
+L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la prolitique optimale pour n'importe quelle récompense.
+Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de l'expert et celui d'une politique donnée.
+Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
 
-Cette résolution est très problématique, en effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la litérature.
+Cette résolution est très problématique.
+En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la litérature.
 Particulièrement, l'information nécessaire à cet objectif est bien plus étendue que celle nécessaire au calcul d'uniquement l'attribut moyen de l'expert comme cela est nécessaire à notre algorithme.
-Comme cela est illustré en section \ref{exp.sec}, l'algorithme d'\citet{abbeel2004apprenticeship} est incapable de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état.\\
-
-Les approches bayésiennes du type de \cite{ramachandran2007bayesian} nécessitent une connaissance du modèle pour fonctionner, modèle dont nous ne disposons pas.\\
+Comme cela est illustré en section \ref{exp.sec}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état.\\
 
@@
 Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
-Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}.
+Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem} dont la fonction de risque est similaire à la nôtre.
 La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}, comme cela est expliqué dans \cite{neu2009training}.
 Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un PDM (entrée à classifier) est associée une politique (classe).
-Cela impose également un cadre déterministe pour les PDM, l'algorithme de résolution utilisé étant $A^*$.\\
+Le cadre applicatif est déterministe, l'algorithme employé pour la résolution des PDM étant $A^*$.
+Il semble difficile de sortir de ce cadre.
 
@@
 choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$.
-Notre fonction de risque est similaire à la leur, mais outre la résolution répétée du PDM liée à la manière dont ils minimisent la fonction de risque et dont nous avons déjà parlé, leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Equation \eqref{structure.eqn}).
+Leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Equation \eqref{structure.eqn}).\\
 
+Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
+
+Les approches bayésiennes du type de \cite{ramachandran2007bayesian, boularias2011relative} nécessitent généralement une connaissance du modèle $\{S,A,p,\gamma\}$ ou un simulateur pour fonctionner.
 %%%%%
@@
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.
-Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.
-On constate que cet espace d'état respecte le critère de Markov dont nous avons vu en section \ref{back.sec} qu'il était nécessaire.\\
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
+Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.\\
 
-La politique optimisant la fonction de récompense donnée figure \ref{trueR.fig} parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
+La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
 Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
@@
 %%%
-\begin{figure}
-  \begin{minipage}[t]{.5\linewidth}
-    \begin{center}
-       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_true_R.pdf}
-       \caption{}
-       \label{trueR.fig}
-    \end{center}
+
+\begin{figure}[tb]
+%
+\begin{minipage}[b]{.5\linewidth}
+  \centering
+  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_true_R.pdf}}
+  \centerline{(a) Récompense de l'expert}%\medskip
 \end{minipage}
 \hfill
-\begin{minipage}[t]{.5\linewidth}
-    \begin{center}
-       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_lafem_R.pdf}
-       \caption{}
-       \label{lafemR.fig}
-    \end{center}
-\end{minipage}\\
-\begin{minipage}[t]{.5\linewidth}
-    \begin{center}
-       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_Vexpert.pdf}
-       \caption{}
-       \label{trueV.fig}
-    \end{center}
+\begin{minipage}[b]{.5\linewidth}
+  \centering
+  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_lafem_R.pdf}}
+  \centerline{(b) Récompense trouvée}%\medskip
+\end{minipage}
+\begin{minipage}[b]{.5\linewidth}
+  \centering
+  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vexpert.pdf}}
+  \centerline{(c) Fonction de valeur de l'expert}%\medskip
 \end{minipage}
 \hfill
-\begin{minipage}[t]{.5\linewidth}
-    \begin{center}
-       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_Vagent.pdf}
-       \caption{}
-       \label{lafemV.fig}
-    \end{center}
-\end{minipage}\\
-\caption*{Illustration de l'application de notre algorithme au problème du pendule inversé.
-Un agent entraîné grâce à un algorithme d'apprentissage par renforcement sur la récompense de la Figure \ref{trueR.fig} parvient à maintenir le pendule en équilibre durant 5 minutes sans échec.
-Il joue le rôle de l'expert.
-La fonction de valeur de cet expert est illustrée Figure \ref{trueV.fig}.
-On constate la similarité avec la Figure \ref{lafemV.fig}, fonction de valeur de l'agent qui lui a été entraîné sur la récompense trouvée par notre algorithme, Figure \ref{lafemR.fig}.
-Bien que les échantillons fournis par l'expert (en vert sur la Figure \ref{lafemR.fig}) ne couvrent qu'une petite partie de l'espace d'état (position angulaire du pendule en abscisse, vitesse angulaire en ordonnées) notre algorithme parvient à l'aide de ces seules données à trouver une récompense justifiant le comportement de l'expert sur tout l'espace d'état.}
+\begin{minipage}[b]{.5\linewidth}
+  \centering
+  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vagent.pdf}}
+  \centerline{(d) Fonction de valeur de l'agent}%\medskip
+\end{minipage}
+%
+\caption{Résultats fournis par notre algorithme sur le problème du pendule inversé.}
+\label{onlyFig.fig}
+%
+\vspace{-5pt}
 \end{figure}
+
 Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
@@
 
-Si la fonction de récompense trouvée par LAFEM (figure \ref{lafemR.fig}) diffère de celle fournie à l'expert (figure \ref{trueR.fig}) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{trueV.fig} et \ref{lafemV.fig}).
+Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)).
 Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.\\
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est, rappelons-le, notre critère de réussite.\\
 
@@
 Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{lafemR.fig} et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
+Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
 En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
@@
 
-Malgré ces deux difficultés, notre algorithme parvient comme nous l'avons dit à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
+Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
 
-La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement que multiplier fonction de récompense ou fonction de valeur par une constante positive n'impacte pas les politiques.
+La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement qu'appliquer une transformation affine à la fonction de récompense ou la fonction de valeur n'impacte pas les politiques.
 \section{Conclusion and future work}

### 2012-02-11 klein2012/main.tex
Intro D'olivier, Version soumise à CAp

@@
 Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
-L'ARI part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction (appelée récompense) à partir de  traces du comportement de l'expert.
+L'ARI part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
 Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
@@
 %%%%%
-Le problème de l'imitation se pose dans des champs variés tels la robotique, les systèmes de dialogue ou la conception d'interfaces homme-machine.
-Un expert effectue une tâche, le but est de parvenir à faire un sorte qu'un agent parvienne à accomplir la même tâche en observant l'expert.
-Dans le cadre de la prise de décisions séqueltielles, l'apprentissage par renforcement est une approche connaissant le succès.
-Cette approche consiste à guider un agent par le biais d'une récompense.
-Le mécanisme inverse peut servir de cadre au problème de l'imitation, en effet si l'on parvient à retrouver la récompense optimisée par un expert alors la résolution du problème direct donnera naissance à un comportement imitant cet expert.\\
+Dans cette contribution, nous nous plaçons dans le contexte de l'apprentissage par démonstration.
+Dans ce cadre, un agent artificiel apprend à reproduire une tâche grâce à l'observation d'un expert réalisant cette tâche de manière optimale.
+Pour trouver une solution à ce problème, nous nous proposons d'utiliser le paradigme de l'apprentissage par renforcement inverse.
+On suppose alors que l'expert agit de manière à être récompensé pour son comportement relativement à une fonction de récompense inconnue.
+Le but de l'agent est alors d'inférer cette fonction des démonstrations données par l'expert pour ensuite optimiser son comportement relativement à cette fonction lui aussi.\\
 
-Le problème de l'apprentissage par renforcement inverse a pour la première fois été introduit dans \cite{russell1998learning}.
-Formellement défini dans \cite{ng2000algorithms}, les approches proposées restent intractables jusqu'à la publication de \cite{abbeel2004apprenticeship}, qui lança toute une vague d'algorithmes cherchant une politique par modifications successives d'une fonction de récompense, approches pour la plupart résumées dans la vue d'ensemble fournie par \cite{neu2009training}.
-%%%%%
+La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
+Plusieurs directions ont depuis été explorées à partir de deux articles fondateurs \cite{ng2000algorithms} et \cite{abbeel2004apprenticeship}.\\
+Dans cette littérature (voir \cite{neu2009training} pour un bon survol), plusieurs verrous sont identifiés mais souvent ignorés pour proposer des algorithmes de résolution.
+Particulièrement, on suppose souvent connue la dynamique de l'environnement dans lequel évoluent l'expert et l'agent ou bien la mise à disposition de cet environnement (ou d'un simulateur) pour y tester les effets d'une politique quelconque.
+Ces contraintes sont souvent incompatibles avec la mise en oeuvre des algorithmes proposés sur des cas réels.\\
+
+Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert.
+Nous proposons une preuve de convergence de cet algorithmes et fournissons une expérimentation sur le problème du pendule inversé.
+Le choix de cette application se justifie par la difficulté  qu'elle pose pour les méthodes de résolution de la littérature.
+%%%%
 \section{Contexte}
@@
 Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité markovienne $p(s_{t+1}|s_t, a_t)$.
-Une politique de contrôle déterministe \footnote{On peut se restreindre au cas déterministe sans perte de généralité.} $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
+Une politique de contrôle déterministe $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
 
@@
 
-Ce problème est mal posé dans la mesure où étant donné un comportement, il n'y a pas unicité de la récompense pour laquelle ce comportement est optimal (voir par exemple \cite{ng1999policy}).
-Particulièrement, la récompense uniformément nulle remplit ce critère dans la mesure où tout comportement est optimal vis-à-vis de cette récompense.
+Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (voir par exemple \cite{ng1999policy}).
+Particulièrement, tout comportement est optimal vis-à-vis de la récompense uniformément nulle.
 Il convient donc de contraindre l'espace des solutions.
@@
 En effet
-\begin{eqnarray}
-\mu^{\pi_1} = \mu^{\pi_2} &\Rightarrow& \theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2}\\
-& \Rightarrow& V^{\pi_1} = V^{\pi_2}.
+\begin{equation}
+\mu^{\pi_1} = \mu^{\pi_2} \Rightarrow \theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} = V^{\pi_2}.
 \label{memevaleur.eqn}
-\end{eqnarray}
+\end{equation}
 L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
@@
 %%%
-Comme nous l'avons vu Equation \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue :
-\begin{equation}
-\pi_E(s) \in \arg\max_aQ^E_R(s,a).
-\end{equation}
+Comme nous l'avons vu Equation \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_aQ^E_R(s,a)$.\\
 Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
@@
 Il va falloir faire appel à une généralisation plus puissante, le sous-gradient.
-Le sous-gradient d'une fonction convexe $f:W\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que :
-\begin{equation}
-\forall x' \in W, g^T(x'-x)\leq f(x')-f(x)
-\end{equation}
+Le sous-gradient d'une fonction convexe $f:W\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que $\forall x' \in W, g^T(x'-x)\leq f(x')-f(x)$.
 Le sous-gradient n'est pas partout unique, cependant l'unique sous-gradient d'une fonction différentiable est son gradient.
 Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
-En appliquant ces règles, on obtient :
+En appliquant ces règles, on obtient le sous gradient et la règle de mise-à-jour associée :
 \begin{eqnarray}
 \nabla_\theta R_N(\theta) &=& {1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\\
-\textrm{avec } a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))
+a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))\\
+   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2}.
 \end{eqnarray}
-Une descente de sous-gradient sur le risque empirique produit alors la règle de mise-à-jour suivante :
-\begin{equation}
-   \theta_{t+1} \leftarrow \theta_t -\alpha_t{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2}.
-\end{equation}
 %%%
@@
 Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons $\forall~1\leq~i~\leq~N,$\\$\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
 Montrons qu'il existe un $x\geq0$ tel que $~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
@@
 
-La complexité calculatoire de notre algorithme instancié avec LSTD$\mu$ reste sous contrôle, létape gourmande étant l'inversion d'une matrice de taille $p \times p$.\\
+La complexité calculatoire de notre algorithme instancié avec LSTD$\mu$ reste sous contrôle, l'étape gourmande étant l'inversion d'une matrice de taille $p \times p$.\\
 
@@
 Bien que ne posant pas particulièrment problème, le point central de notre algorithme qu'est le calcul de $\mu_E$ gagnerait à être amélioré.
-L'axe d'étude que nous envisageons d'explorer serait de s'en passer purement et simplement en trouvant le moyen d'introduire la structure du PDM dans la démarche de classification sueprvisée.\\
+L'axe d'étude que nous envisageons d'explorer serait de s'en passer purement et simplement en trouvant le moyen d'introduire la structure du PDM dans la démarche de classification supervisée.\\
 

### 2012-02-11 klein2012/main.tex
Modifications suggérées par Matthieu. Version soumise à CAp

@@
 Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
-L'ARI part du postulat que l'expert optimise avec succès une fonction d'utilité ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
+L'ARI part du postulat que l'expert optimise avec succès une fonction de récompense ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
 Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
@@
 Particulièrement, on suppose souvent connue la dynamique de l'environnement dans lequel évoluent l'expert et l'agent ou bien la mise à disposition de cet environnement (ou d'un simulateur) pour y tester les effets d'une politique quelconque.
+De plus, ces approches supposent de résoudre le problème direct (optimiser une politique pour une récompense donnée) un grand nombre de fois pour résoudre le problème inverse (trouver une récompense expliquant la politique).
+De plus, certaines de ces approches donnent en sortie une politique généralisant celle de l'expert plutôt qu'une récompense l'expliquant.
 Ces contraintes sont souvent incompatibles avec la mise en oeuvre des algorithmes proposés sur des cas réels.\\
@@
 Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
-Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité markovienne $p(s_{t+1}|s_t, a_t)$.
+Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité de transition markovienne $p(s_{t+1}|s_t, a_t)$.
 Une politique de contrôle déterministe $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
@@
 Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
-Une des forces de l'apprentissage par renforcement réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
+Une des forces de ce cadre réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
 On définit pour ce faire la fonction de valeur $V^\pi$ :
@@
 \end{equation}
-où $\gamma \in [0,1)$ est un facteur d'actualisation.
-Il est également possible de définir une fonction de qualité qui ajoute à la fonction de valeur un degré de liberté sur le choix de l'action suivante :
+où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action :
 \begin{equation}
 \label{Qdef.eqn}
-Q^\pi(s,a) = E[\sum_t\gamma^tR(s_t|s_0=s,a_0=a,\pi] = R(s) + \gamma E[Q^\pi(s',\pi(s'))|s,a].
+Q^\pi(s,a) = E[\sum_t\gamma^tR(s_t)|s_0=s,a_0=a,\pi] = R(s) + \gamma E[V^\pi(s')|s,a].
 \end{equation}
-  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur $V^*$ (appelée fonction de valeur optimale) vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
+  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
 La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
@@
 
-Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (voir par exemple \cite{ng1999policy}).
+Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}).
 Particulièrement, tout comportement est optimal vis-à-vis de la récompense uniformément nulle.
@@
 \end{equation}
-Comme nous allons le voir, cette contrainte peut cependant être relâchée si nécessaire.
-Introduire cette expression dans la définition de la fonction de qualité (Equation \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+Cette contrainte peut cependant être relâchée si nécessaire.
+Introduire cette expression dans la définition de la fonction de qualité (Eq.
+\eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
 \begin{equation}
-Q^\pi(s,a) = \theta^T\mu^\pi(s,a) \textrm{ avec } \mu^\pi(s,a) = E[\sum_t\gamma^t\psi(s_t)|s_0=s,a_0=a,\pi]
+Q^\pi(s,a) = \theta^T\mu^\pi(s,a),\mu^\pi(s,a) = E[\sum_t\gamma^t\psi(s_t)|s_0=s,a_0=a,\pi]
 \label{Qmu.eqn}
@@
 L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
-La plupart des algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
-Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert (voir par exemple \cite{neu2009training}).
+La plupart des algorithmes de la littérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
+Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert, quelle que soit la récompense (voir par exemple \cite{neu2009training}).
 %%%%%
-\section{Classification structurée pour l'apprentissage par renforcement inverse}
+\section{Classification structurée pour l'ARI}
 %%%%%
@@
 %%%
-Comme nous l'avons vu Equation \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_aQ^E_R(s,a)$.\\
+Comme nous l'avons vu Eq.
+\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_aQ^E_R(s,a)$.\\
 Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
@@
 \end{equation}
-Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
 Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
-Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
-Nous introduisons donc la fonction de risque empirique suivante :
+Ce type d'argument est assez commun dans la littérature (voir Sec.
+\ref{biblio.sec}).
+Nous introduisons ainsi la fonction de risque empirique suivante :
   \begin{equation}
-   R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
+\label{Jdef.eqn}
+   J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
    \end{equation}
-Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans l'analyse (Section \ref{proof.sec}).\\
+Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
 
 Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
-Notre propose n'est cependant pas d'utiliser une approche purement supervisée.
+Notre propos n'est cependant pas d'utiliser une approche purement supervisée.
 Nous ne souhaitons pas apprendre le contrôle de l'expert mais la description de la tâche qu'il effectue (la récompense).
@@
 
-Comme nous l'avons vu Equation \eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
+Comme nous l'avons vu Eq.
+\eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
 Cette paramétrisation prend en compte la dynamique imposée par $\pi_E$.
-L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
-Le calcul de ce terme $\mu_E$ est central à notre algorithmes et nous proposons quelques solutions en sous-section \ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
+L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert, nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
+L'estimation de ce terme $\mu_E$ est centrale à notre algorithme et nous proposons quelques solutions en Sec.
+\ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Eq.
+\eqref{hatRdef.eqn}) :
    \begin{equation}
      \label{structure.eqn}
-   R_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
+   J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
    \end{equation}
-   Une descente de gradient devrait permettre de minimiser cette expression.
-Cependant, la présence de l'opérateur $\max$ rend les choses un peu plus complexes car l'expression n'est plus différentiable.
-Il va falloir faire appel à une généralisation plus puissante, le sous-gradient.
-Le sous-gradient d'une fonction convexe $f:W\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que $\forall x' \in W, g^T(x'-x)\leq f(x')-f(x)$.
+Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
+Nous utilisons une descente de sous-gradient (normalisé), une généralisation du gradient, pour le minimiser.
+Le sous-gradient d'une fonction convexe $f:\mathbb{R}^p\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que $\forall x' \in \mathbb{R}^p, g^T(x'-x)\leq f(x')-f(x)$.
 Le sous-gradient n'est pas partout unique, cependant l'unique sous-gradient d'une fonction différentiable est son gradient.
-Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
-En appliquant ces règles, on obtient le sous gradient et la règle de mise-à-jour associée :
+Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $\max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
+En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée :
 \begin{eqnarray}
-\nabla_\theta R_N(\theta) &=& {1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\\
+\nabla_\theta J_N(\theta) &=& {1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\\
 a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))\\
-   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2}.
+   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_\theta J_N(\theta) \over ||\nabla_\theta J_N(\theta)||_2}.
 \end{eqnarray}
@@
 \label{calculmu.sec}
-Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
+Comme signalé dans la Sec.
+\ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
 Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
@@
 
-Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $j=1..p$, on trouve :
+Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve :
 \begin{equation}
@@
 
-Dans le cas idéal où le modèle est connu, le calcul exact de $\mu_E$ peut être effectué, les algorithmes de programmation dynamique peuvent également être utilisés.
-
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a)$ (définition Equation \eqref{Qmu.eqn}) l'on utilise
+Dans le cas idéal où le modèle est connu, le calcul exact de $\mu_E$ peut être effectué grâce aux algorithmes de programmation dynamique.
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a)$ (définition Eq.
+\eqref{Qmu.eqn}) l'on utilise
 \begin{equation}
@@
 
-Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
+Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
 Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
@@
 \subsection{Hypothèses}
-Ici, on présente le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
+Ici, on présente le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $J_N$ qui rende la politique de l'expert optimale.
 On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$.
@@
 Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation \eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation \eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\eqref{eq:existence}$ qui minimise $J_N$,  que les minimiseurs de $J_N$ vérifient bien l'équation \eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
 \subsection{Résultats}
-On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $R_N$.
-On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
+On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $J_N$.
+On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:J_N(\theta)=\inf_{\theta \in \mathbb{R}^p}J_N(\theta)\}$.
 Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
-Il est facile de borner inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Il est facile de borner inférieurement le risque $J_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
 Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons $\forall~1\leq~i~\leq~N,$\\$\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
-Montrons qu'il existe un $x\geq0$ tel que $~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+Montrons qu'il existe un $x\geq0$ tel que $~J_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
 Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
@@
 \Rightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,\\
-\Leftrightarrow&R_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
+\Leftrightarrow&J_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
 \end{align}
-C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $R_N$ vérifient bien l'équation \eqref{eq:existence}.
-Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $R_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, R_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $R_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $J_N$ vérifient bien l'équation \eqref{eq:existence}.
+Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $J_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, J_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $J_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Or on a $\forall 1\leq i \leq N$ :
 \begin{align}
-&R_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),\\
+&J_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),\\
 \Leftrightarrow&\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,\\
@@
 Finalement, montrons que l'algorithme de sous-gradient converge.
-Tout d'abord, il est clair que $R_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
+Tout d'abord, il est clair que $J_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
 De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article \cite{correa1993convergence} qui nous assure que la suite :
 \begin{equation}
-\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\frac{1}{t+1}{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2},
+\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\frac{1}{t+1}{\nabla_\theta J_N(\theta) \over ||\nabla_\theta J_N(\theta)||_2},
 \end{equation}
-est convergente vers un minimiseur $\theta^*$ de $R_N$.
+est convergente vers un minimiseur $\theta^*$ de $J_N$.
 Ce minimiseur vérifiera l'équation \eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation \eqref{eq:existence} implique l'optimalité de Bellman.
@@
 %%%%%
-L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la prolitique optimale pour n'importe quelle récompense.
-Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de l'expert et celui d'une politique donnée.
+L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
+Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de \mbox{l'expert} et celui d'une politique donnée.
 Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
@@
 Cette résolution est très problématique.
-En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la litérature.
+En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la littérature.
 Particulièrement, l'information nécessaire à cet objectif est bien plus étendue que celle nécessaire au calcul d'uniquement l'attribut moyen de l'expert comme cela est nécessaire à notre algorithme.
-Comme cela est illustré en section \ref{exp.sec}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état.\\
+Comme cela est illustré dans \cite{klein2011batch}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état, ne permettant pas la résultion du problème direct en se basant uniquement sur les  données de l'expert.\\
 
@@
 Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
-Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem} dont la fonction de risque est similaire à la nôtre.
-La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}, comme cela est expliqué dans \cite{neu2009training}.
+Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}, notre fonction de risque étant similaire à la leur.
+La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}.
 Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un PDM (entrée à classifier) est associée une politique (classe).
 Le cadre applicatif est déterministe, l'algorithme employé pour la résolution des PDM étant $A^*$.
-Il semble difficile de sortir de ce cadre.
+Il semble difficile de sortir de ce cadre  en pratique (bien que l'algorithme soit introduit dans un cadre plus général).
 
+Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
+Le principe de leur approche est exactement de résoudre le risque (Eq.
+\eqref{Jdef.eqn}).
+Toutefois, ils ne prennent pas en comptent la structure du MDP, ce à quoi nous remédions avec notre contribution (voir \cite{melo2010learning} sur l'importance de prendre en compte cette structure).
 Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
@@
 choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$.
-Leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Equation \eqref{structure.eqn}).\\
+Leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Eq.
+\eqref{structure.eqn}).\\
 
 Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
-
-Les approches bayésiennes du type de \cite{ramachandran2007bayesian, boularias2011relative} nécessitent généralement une connaissance du modèle $\{S,A,p,\gamma\}$ ou un simulateur pour fonctionner.
+Il existe d'autres approches que nous ne passons pas en revue de manière exhaustive, mais à notre connaissance elles nécessitent toutes de résoudre le problème direct.
 %%%%%
@@
 %
-\caption{Résultats fournis par notre algorithme sur le problème du pendule inversé.}
+\caption{Résultats fournis par notre algorithme.}
 \label{onlyFig.fig}
@@
 La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon.
-Le pas de temps est constant à $\alpha(t) = 0.1, \forall t$ et le nombre d'itérations est de $T=20$.
+Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations est de $T=20$.
 Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi : S \rightarrow \mathbb{R}^p$ correspondant au mélange de gaussiennes défini dans \cite{lagoudakis2003least}.
@@
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est, rappelons-le, notre critère de réussite.\\
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est notre critère de réussite.\\
 
@@
 En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
-Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs proposés dans le problème du pendule inversé par \citet{lagoudakis2003least} et utilisés par nous paramétrer la récompense consistent en un maillage de gaussiennes qui ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.\\
+Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.
+Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.
+Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig.
+\ref{onlyFig.fig} Cela n'a aucune importance, une dilatation laissant la politique optimale invariante.
+\section{Conclusion}
 
-Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
-
-La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement qu'appliquer une transformation affine à la fonction de récompense ou la fonction de valeur n'impacte pas les politiques.
-\section{Conclusion and future work}
-
-L'algorithme présenté dans cette publication lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse.
+L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse.
 En rendant la résolution du problème inverse indépendante de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
-
-La complexité calculatoire de notre algorithme instancié avec LSTD$\mu$ reste sous contrôle, l'étape gourmande étant l'inversion d'une matrice de taille $p \times p$.\\
 
@@
 
-Bien que ne posant pas particulièrment problème, le point central de notre algorithme qu'est le calcul de $\mu_E$ gagnerait à être amélioré.
-L'axe d'étude que nous envisageons d'explorer serait de s'en passer purement et simplement en trouvant le moyen d'introduire la structure du PDM dans la démarche de classification supervisée.\\
-
+L'estimation de $mu_E$ est centrale dans notre approche.
+Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients que les algorithmes d'estimation d'une fonction de valeur.
+L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée.
 Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.

### 2012-02-13 klein2012/main.tex
superfinal

--- 2012-02-10 klein2012/main.tex
@@
 
-Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique.
+Le terme $\mu^\pi$ est appelé la {\it feature expectation} ou plutôt l'attribut vectoriel moyen d'une politique.
 On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
@@
 L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
-La plupart des algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
-Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert (voir par exemple \cite{neu2009training}).
+Certains algorithmes de la litérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
+Cela permet d'obtenir un agent dont le comportement a la même valeur que celui de l'expert (voir \cite{neu2009training}).
 %%%%%
-\section{Classification structurée pour l'apprentissage par renforcement inverse}
+\section{Classification structurée pour l'apprentissage par renforcement inverse (SCARI)}
 %%%%%
@@
 %%%
-Comme nous l'avons vu Equation \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue :
+L'un des postulats de base de l'apprentissage par renforcement inverse est que l'expert est un agent \emph{optimal} vis-à-vis d'une certaine fonction de récompense que l'on cherche à retrouver.
+Cela implique, comme nous l'avons vu Equation \eqref{greedy.eqn}, qu'il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à cette fonction de récompense inconnue :
 \begin{equation}
@@
 Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
-Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
-Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons :
+Notons $(s_i,a_i,s_{i+1})_{i=1..N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Il faut que pour chacun de ces échantillons l'on ait :
 \begin{equation}
@@
 \end{equation}
-Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
-Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Notre fonction de risque doit donc pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$ car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant la marge qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
 Ce type d'argument est assez commun dans la litérature (voir Section \ref{biblio.sec}).
-Nous introduisons donc la fonction de risque empirique suivante :
+Nous introduisons la fonction de risque empirique :
   \begin{equation}
@@
    \end{equation}
-Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans l'analyse (Section \ref{proof.sec}).\\
+Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Section \ref{exp.sec}) et les critères dévelopés dans les preuves (Section \ref{proof.sec}).\\
 
-Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
-Notre propose n'est cependant pas d'utiliser une approche purement supervisée.
-Nous ne souhaitons pas apprendre le contrôle de l'expert mais la description de la tâche qu'il effectue (la récompense).
-Il faut pour cela prendre en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnées par la politique de l'expert.\\
+Nous traitons le problème de l'imitation comme un problème de classification multi-classe, ce qui a déjà été envisagé (\citep{ratliff2007imitation}).
+Les approches existantes en prennent toutefois pas en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnée par la politique de l'expert.
+C'est aussi pour cette raison que les approches purement supervisées ne peuvent fonctionner pour l'imitation.\\
 
@@
 L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$.
-Le calcul de ce terme $\mu_E$ est central à notre algorithmes et nous proposons quelques solutions en sous-section \ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
+Le calcul de ce terme $\mu_E$ est central à notre algorithmes et nous proposons quelques solutions en sous-section \ref{calculmu.sec}.
+Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Equation \eqref{hatRdef.eqn}).
    \begin{equation}
@@
    \end{equation}
-   Une descente de gradient devrait permettre de minimiser cette expression.
-Cependant, la présence de l'opérateur $\max$ rend les choses un peu plus complexes car l'expression n'est plus différentiable.
-Il va falloir faire appel à une généralisation plus puissante, le sous-gradient.
-Le sous-gradient d'une fonction convexe $f:W\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que :
-\begin{equation}
-\forall x' \in W, g^T(x'-x)\leq f(x')-f(x)
-\end{equation}
-Le sous-gradient n'est pas partout unique, cependant l'unique sous-gradient d'une fonction différentiable est son gradient.
-Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
-En appliquant ces règles, on obtient :
+   Un sous-gradient de cette expression peut se calculer à l'aide notamment de trois propriétés que nous rappelons :
+\begin{itemize}
+  \item l'opérateur de sous-gradient est linéaire,
+  \item le gradient est l'unique sous-gradient d'une fonction différentiable,
+  \item la fonction convexe, différentiable par morceaux $max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
+\end{itemize}
+Ainsi, l'on a :
 \begin{eqnarray}
@@
 Comme signalé dans la sous-section \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
-Nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitraire.
+Une différence notable, cependant, est que nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
 Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
 
-Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $j=1..p$, on trouve :
+Cette inférence est facilité par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $j=1..p$, on trouve :
 \begin{equation}
-\mu_E^j(s,a) = E[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)|s_0 = s, a_0 = a, \pi_E],
+\mu_E^j(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)\right|s_0 = s, a_0 = a, \pi_E\right],
 \end{equation}
-expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
+expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attribut $\psi$.\\
+
 Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu_E$.\\
@@
 
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a)$ (définition Equation \eqref{Qmu.eqn}) l'on utilise
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer
 \begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\{\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, s_{t+1} \sim p(\cdot|s_t,\pi_E(s_t))\}.
+\mu_E(s,a) = E\left.\left[\sum\limits_{t=0}^\infty \gamma^t \psi(s)\right|s_0 = s, a_0 = a, \pi_E\right]
+\end{equation}
+ l'on utilise
+\begin{equation}
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, a_{t>0} = \pi_E(s_t).
 \end{equation}
@@
 Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
-C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu_E$ en utilisant LSTD.
-Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
+C'est la méthode que nous illustrons dans la section \ref{exp.sec}\\
 
-D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
+Finalement, il est possible de travailler avec une approximation de Monte-Carlo plus fine que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
 Cela est reporté pour des travaux futurs.
@@
 \subsection{Hypothèses}
-Ici, on présente le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
-On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$.
-De plus, on supposera que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$.
-Finalement la récompense $R$ devra vérifier la propriété suivante :
+Ici, nous présentons le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
+On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$, que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$ et que $\mu_E$ est estimée sans erreur sur $\{(s_i,a_i)\}_{1\leq i \leq N}$.
+Nous supposons également l'unicité de l'action optimale (cette hypothèse pouvant être relâchée au prix de notations plus complexes):
 \begin{equation}\label{eq:existence}
@@
 \end{equation}
-où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$, qui est légèrement plus forte que l'équation d'optimalité de Bellman puisque l'équation \eqref{eq:existence} implique : $\forall s\in S, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Ainsi tout vecteur $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale.
-Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation \eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation \eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
-\subsection{Résultats}
+où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$.
+Ainsi tout vecteur $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale.
+Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation~\eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
+Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation~$\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation~\eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+\subsection{Résultat}
 On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $R_N$.
@@
 \end{align}
-D'ou ~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
+Donc~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
 Finalement, montrons que l'algorithme de sous-gradient converge.
 Tout d'abord, il est clair que $R_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
-De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article \cite{correa1993convergence} qui nous assure que la suite :
+De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article de \cite{correa1993convergence} qui nous assure que la suite :
 \begin{equation}
-\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\frac{1}{t+1}{\nabla_\theta R_N(\theta) \over ||\nabla_\theta R_N(\theta)||_2},
+\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\alpha_t{\nabla_\theta R_N(\theta_t) \over ||\nabla_\theta R_N(\theta_t)||_2},
 \end{equation}
-est convergente vers un minimiseur $\theta^*$ de $R_N$.
-Ce minimiseur vérifiera l'équation \eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation \eqref{eq:existence} implique l'optimalité de Bellman.
+est convergente vers un minimiseur $\theta^*$ de $R_N$, où $alpha_t$ est un taux d'apprentissage satisfaisant les condition usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
+Ce minimiseur vérifiera l'équation~\eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation~\eqref{eq:existence} implique l'équation d'optimalité de Bellman.
 \section{Travaux Connexes}
@@
 %%%%%
-L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la prolitique optimale pour n'importe quelle récompense.
-Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de l'expert et celui d'une politique donnée.
-Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
+L'approche historique de \cite{ng2000algorithms}, ainsi que les algorithmes basés sur \cite{abbeel2004apprenticeship}, notamment \citep{syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée.
+Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de l'expert et l'attribut vectoriel moyen d'une politique donnée.
+Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
 
-Cette résolution est très problématique.
-En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la litérature.
+Cette résolution est très problématique, en effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la litérature.
 Particulièrement, l'information nécessaire à cet objectif est bien plus étendue que celle nécessaire au calcul d'uniquement l'attribut moyen de l'expert comme cela est nécessaire à notre algorithme.
-Comme cela est illustré en section \ref{exp.sec}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état.\\
+Comme cela est illustré en section \ref{exp.sec}, l'algorithme d'\citet{abbeel2004apprenticeship} est incapable de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état.\\
+
+Les approches bayésiennes du type de \cite{ramachandran2007bayesian} nécessitent une connaissance du modèle pour fonctionner, modèle dont nous ne disposons pas.\\
 
@@
 Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
-Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem} dont la fonction de risque est similaire à la nôtre.
+Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}.
 La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}, comme cela est expliqué dans \cite{neu2009training}.
 Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un PDM (entrée à classifier) est associée une politique (classe).
-Le cadre applicatif est déterministe, l'algorithme employé pour la résolution des PDM étant $A^*$.
-Il semble difficile de sortir de ce cadre.
+Cela impose également un cadre déterministe pour les PDM, l'algorithme de résolution utilisé étant $A^*$.\\
 
@@
 choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$.
-Leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Equation \eqref{structure.eqn}).\\
+Notre fonction de risque est similaire à la leur, mais outre la résolution répétée du PDM liée à la manière dont ils minimisent la fonction de risque et dont nous avons déjà parlé, leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Equation \eqref{structure.eqn}).
 
-Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
-
-Les approches bayésiennes du type de \cite{ramachandran2007bayesian, boularias2011relative} nécessitent généralement une connaissance du modèle $\{S,A,p,\gamma\}$ ou un simulateur pour fonctionner.
 %%%%%
@@
 Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
-Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.\\
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de bases $\phi : S\times A \rightarrow \mathbb{R}^k$ et $\psi : S \rightarrow \mathbb{R}^p$ sont conformes à ce qui est donné dans cette publication.
+Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.
+On constate que cet espace d'état respecte le critère de Markov dont nous avons vu en section \ref{back.sec} qu'il était nécessaire.\\
 
-La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
+La politique optimisant la fonction de récompense donnée figure \ref{trueR.fig} parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
 Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
@@
 %%%
-
-\begin{figure}[tb]
-%
-\begin{minipage}[b]{.5\linewidth}
-  \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_true_R.pdf}}
-  \centerline{(a) Récompense de l'expert}%\medskip
+\begin{figure}
+  \begin{minipage}[t]{.5\linewidth}
+    \begin{center}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_true_R.pdf}
+       \caption{}
+       \label{trueR.fig}
+    \end{center}
 \end{minipage}
 \hfill
-\begin{minipage}[b]{.5\linewidth}
-  \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_lafem_R.pdf}}
-  \centerline{(b) Récompense trouvée}%\medskip
-\end{minipage}
-\begin{minipage}[b]{.5\linewidth}
-  \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vexpert.pdf}}
-  \centerline{(c) Fonction de valeur de l'expert}%\medskip
+\begin{minipage}[t]{.5\linewidth}
+    \begin{center}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_lafem_R.pdf}
+       \caption{}
+       \label{lafemR.fig}
+    \end{center}
+\end{minipage}\\
+\begin{minipage}[t]{.5\linewidth}
+    \begin{center}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_Vexpert.pdf}
+       \caption{}
+       \label{trueV.fig}
+    \end{center}
 \end{minipage}
 \hfill
-\begin{minipage}[b]{.5\linewidth}
-  \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vagent.pdf}}
-  \centerline{(d) Fonction de valeur de l'agent}%\medskip
-\end{minipage}
-%
-\caption{Résultats fournis par notre algorithme sur le problème du pendule inversé.}
-\label{onlyFig.fig}
-%
-\vspace{-5pt}
+\begin{minipage}[t]{.5\linewidth}
+    \begin{center}
+       \includegraphics[width=1.3\textwidth]{LAFEM_Exp3_Vagent.pdf}
+       \caption{}
+       \label{lafemV.fig}
+    \end{center}
+\end{minipage}\\
+\caption*{Illustration de l'application de notre algorithme au problème du pendule inversé.
+Un agent entraîné grâce à un algorithme d'apprentissage par renforcement sur la récompense de la Figure \ref{trueR.fig} parvient à maintenir le pendule en équilibre durant 5 minutes sans échec.
+Il joue le rôle de l'expert.
+La fonction de valeur de cet expert est illustrée Figure \ref{trueV.fig}.
+On constate la similarité avec la Figure \ref{lafemV.fig}, fonction de valeur de l'agent qui lui a été entraîné sur la récompense trouvée par notre algorithme, Figure \ref{lafemR.fig}.
+Bien que les échantillons fournis par l'expert (en vert sur la Figure \ref{lafemR.fig}) ne couvrent qu'une petite partie de l'espace d'état (position angulaire du pendule en abscisse, vitesse angulaire en ordonnées) notre algorithme parvient à l'aide de ces seules données à trouver une récompense justifiant le comportement de l'expert sur tout l'espace d'état.}
 \end{figure}
-
 Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
@@
 
-Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)).
+Si la fonction de récompense trouvée par LAFEM (figure \ref{lafemR.fig}) diffère de celle fournie à l'expert (figure \ref{trueR.fig}) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{trueV.fig} et \ref{lafemV.fig}).
 Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est, rappelons-le, notre critère de réussite.\\
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est rappelons le notre critère de réussite.\\
 
@@
 Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
+Ceux-ci sont portés en vert sur la Figure \ref{lafemR.fig} et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
 En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
@@
 
-Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
+Malgré ces deux difficultés, notre algorithme parvient comme nous l'avons dit à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.\\
 
-La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement qu'appliquer une transformation affine à la fonction de récompense ou la fonction de valeur n'impacte pas les politiques.
+La différence dans les amplitudes des récompenses et des fonctions de valeurs n'est pas un problème, c'est en effet un résultat connu en apprentissage par renforcement que multiplier fonction de récompense ou fonction de valeur par une constante positive n'impacte pas les politiques.
 \section{Conclusion and future work}

### 2012-02-13 klein2012/main.tex

--- 2012-02-11 klein2012/main.tex
@@
 \subsection{Hypothèses}
-Ici, on présente le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $J_N$ qui rende la politique de l'expert optimale.
-On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$.
-De plus, on supposera que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$.
-Finalement la récompense $R$ devra vérifier la propriété suivante :
+Ici, nous présentons le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
+On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$, que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$ et que $\mu_E$ est estimée sans erreur sur $\{(s_i,a_i)\}_{1\leq i \leq N}$.
+Nous supposons également l'unicité de l'action optimale (cette hypothèse pouvant être relâchée au prix de notations plus complexes):
 \begin{equation}\label{eq:existence}
@@
 \end{equation}
-où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$, qui est légèrement plus forte que l'équation d'optimalité de Bellman puisque l'équation \eqref{eq:existence} implique : $\forall s\in S, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Ainsi tout vecteur $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale.
-Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation \eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation \eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation $\eqref{eq:existence}$ qui minimise $J_N$,  que les minimiseurs de $J_N$ vérifient bien l'équation \eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
-\subsection{Résultats}
-On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $J_N$.
-On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:J_N(\theta)=\inf_{\theta \in \mathbb{R}^p}J_N(\theta)\}$.
+où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$.
+Ainsi tout vecteur $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale.
+Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation~\eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
+Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation~$\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation~\eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+\subsection{Résultat}
+On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $R_N$.
+On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
 Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
-Il est facile de borner inférieurement le risque $J_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Il est facile de borner inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons $\forall~1\leq~i~\leq~N,$\\$\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
-Montrons qu'il existe un $x\geq0$ tel que $~J_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
+Montrons qu'il existe un $x\geq0$ tel que $~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
 Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
@@
 \end{align}
-D'ou ~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
+Donc~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
 Finalement, montrons que l'algorithme de sous-gradient converge.
-Tout d'abord, il est clair que $J_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
-De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article \cite{correa1993convergence} qui nous assure que la suite :
+Tout d'abord, il est clair que $R_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
+De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article de \cite{correa1993convergence} qui nous assure que la suite :
 \begin{equation}
-\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\frac{1}{t+1}{\nabla_\theta J_N(\theta) \over ||\nabla_\theta J_N(\theta)||_2},
+\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\alpha_t{\nabla_\theta R_N(\theta_t) \over ||\nabla_\theta R_N(\theta_t)||_2},
 \end{equation}
-est convergente vers un minimiseur $\theta^*$ de $J_N$.
-Ce minimiseur vérifiera l'équation \eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation \eqref{eq:existence} implique l'optimalité de Bellman.
+est convergente vers un minimiseur $\theta^*$ de $R_N$, où $alpha_t$ est un taux d'apprentissage satisfaisant les condition usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
+Ce minimiseur vérifiera l'équation~\eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation~\eqref{eq:existence} implique l'équation d'optimalité de Bellman.
 \section{Travaux Connexes}

### 2012-02-13 klein2012/main.tex
Merge branch 'NouveauxAlgos' of 193.48.224.148:/home/share/Code into NouveauxAlgos
Conflicts:
	klein2012/main.tex

@@
 a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))\\
-   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_\theta J_N(\theta) \over ||\nabla_\theta J_N(\theta)||_2}.
+   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_{\theta_t} J_N(\theta_t) \over ||\nabla_{\theta_t} J_N(\theta_t)||_2}.
 \end{eqnarray}
@@
 \subsection{Hypothèses}
-Ici, nous présentons le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
-On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$, que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$ et que $\mu_E$ est estimée sans erreur sur $\{(s_i,a_i)\}_{1\leq i \leq N}$.
+Ici, nous présentons le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $J_N$ qui rende la politique de l'expert optimale.
+On supposera que la récompense $R$ est de la forme~$R(s)~=~\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$, que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$ et que $\mu_E$ est estimée sans erreur sur $\{(s_i,a_i)\}_{1\leq i \leq N}$.
 Nous supposons également l'unicité de l'action optimale (cette hypothèse pouvant être relâchée au prix de notations plus complexes):
@@
 Ainsi tout vecteur $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale.
-Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation~\eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation~$\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation~\eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation~\eqref{eq:existence} impliquera dans ce cas : $\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
+Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation~$\eqref{eq:existence}$ qui minimise $J_N$,  que les minimiseurs de $J_N$ vérifient bien l'équation~\eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
 \subsection{Résultat}
-On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $R_N$.
-On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
+On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $J_N$.
+On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:J_N(\theta)=\inf_{\theta \in \mathbb{R}^p}J_N(\theta)\}$.
 Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
-Il est facile de borner inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Il est facile de borner inférieurement le risque $J_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
 Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
-Montrons qu'il existe un $x\geq0$ tel que $~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+Montrons qu'il existe un $x\geq0$ tel que $~J_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
 Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
@@
 C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $J_N$ vérifient bien l'équation \eqref{eq:existence}.
-Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $J_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, J_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $J_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Soit ~$\theta^*\in\ M^*$.
+Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $J_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, J_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $J_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
 Or on a $\forall 1\leq i \leq N$ :
@@
 Finalement, montrons que l'algorithme de sous-gradient converge.
-Tout d'abord, il est clair que $R_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
+Tout d'abord, il est clair que $J_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
 De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article de \cite{correa1993convergence} qui nous assure que la suite :
 \begin{equation}
-\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\alpha_t{\nabla_\theta R_N(\theta_t) \over ||\nabla_\theta R_N(\theta_t)||_2},
+\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\alpha_t{\nabla_{\theta_t} J_N(\theta_t) \over ||\nabla_{\theta_t} J_N(\theta_t)||_2},
 \end{equation}
-est convergente vers un minimiseur $\theta^*$ de $R_N$, où $alpha_t$ est un taux d'apprentissage satisfaisant les condition usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
+est convergente vers un minimiseur $\theta^*$ de $J_N$, où $\alpha_t$ est un taux d'apprentissage satisfaisant les condition usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
 Ce minimiseur vérifiera l'équation~\eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation~\eqref{eq:existence} implique l'équation d'optimalité de Bellman.

### 2012-02-14 klein2012/main.tex

@@
 \end{equation}
-est convergente vers un minimiseur $\theta^*$ de $J_N$, où $\alpha_t$ est un taux d'apprentissage satisfaisant les condition usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
+est convergente vers un minimiseur $\theta^*$ de $J_N$, où $\alpha_t$ est un taux d'apprentissage satisfaisant les conditions usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
 Ce minimiseur vérifiera l'équation~\eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation~\eqref{eq:existence} implique l'équation d'optimalité de Bellman.
@@
 
-La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
+La politique optimisant la fonction de récompense donnée Fig.
+\ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
 Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_true_R.pdf}}
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_true_R.pdf}}
   \centerline{(a) Récompense de l'expert}%\medskip
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_lafem_R.pdf}}
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_lafem_R.pdf}}
   \centerline{(b) Récompense trouvée}%\medskip
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vexpert.pdf}}
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_Vexpert.pdf}}
   \centerline{(c) Fonction de valeur de l'expert}%\medskip
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vagent.pdf}}
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_Vagent.pdf}}
   \centerline{(d) Fonction de valeur de l'agent}%\medskip
@@
 
-Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)).
+Si la fonction de récompense trouvée par LAFEM (Fig.
+\ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (Fig.
+\ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs sont très similaires (Fig.
+\ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)).
 Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
@@
 Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
+Ceux-ci sont portés en vert sur la Fig.
+\ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
 En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
@@
 
-L'estimation de $mu_E$ est centrale dans notre approche.
+L'estimation de $\mu_E$ est centrale dans notre approche.
 Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients que les algorithmes d'estimation d'une fonction de valeur.

### 2012-02-15 Highway/al_code/car.py
Ajout du code d'Umar Syed pour le Highway driving

@@ -1,0 +1,198 @@
+# Car driving simulator
+# 
+# The main parameter controlling this simulator is self.autopilot.
+#
+# If self.autopilot = "no", then the user can drive the blue car with the arrow keys:
+# 	Up = Speed up
+# 	Down = Slow down
+# 	Left = Move left
+# 	Right = Move right
+#
+# If self.autopilot = "yes", then the blue car is driven by the policy specified in self.policy_fname.
+#
+# Each line of self.policy_fname specifies the action distribution for one state.
+# Each line is a string that has the following form:
+# 	st;p1,p2,p3
+# where st is the state, and pi is the probability of taking action i in state st.
+#
+# The state of the simulator is a string describing the positions of all the cars, and the speed of the blue car.
+# Each state has the following form:
+# 	[speed, blue_car_x, [red_car_x, red_car_y]]
+# Where:
+# 	speed = Speed of the blue car.
+# 	blue_car_x = X-axis displacement of the blue car.
+# 	red_car_x = X-axis displacement of the red car.
+# 	red_car_y = Y-axis displacement of the red car.
+# The initial state is "[1, 160]". It is the only state in which the red car does not appear.
+#
+# In autopilot mode, only three actions are allowed at any given time:
+# 	In the initial state, the three actions correspond to selecting one of the three possible speeds.
+# 	In all non-initial states, the three actions correspond to staying put, moving left, and moving right.
+# So a policy can use just a single speed during a simulation. This simplification is convenient, and does not exclude any optimal policies.
+#
+
+from Tkinter import *
+import random
+
+class App:
+    def __init__(self, master):
+
+        frame = Frame(master)
+        frame.pack()
+
+        # Parameters
+        self.h = 160 + 20
+        self.w = 300
+        self.interval = 100
+	self.step = 5
+        self.speeds_list = [20, 40, 80]
+	self.speeds_names = ["Slow", "Med.", "Fast"]
+        self.autopilot = "no"
+	self.policy_fname = "policy.txt"
+        
+        # Control
+        self.speed = 1
+        self.action = "none"
+        self.policy = dict([])
+
+        # Variables
+        self.collisions = 0
+        self.offroads = 0
+
+        # Draw field of play
+        h, w = self.h, self.w
+        master.title("Car Game")
+        master.geometry(str(w) + "x" + str(h) + "+100+100")        
[…]

### 2012-02-15 Highway/al_code/opt_policy_and_feat_exp.m

@@ -1,0 +1,89 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This function computes the optimal policy with respect to a particular reward function, and simultaneously computes the "feature expectations" for that policy.
+%
+% Here's a description of the parameters:
+% 
+% Input:
+% 
+% THETA: The NA x N transition matrix, where NA is the number of state-action pairs, and A is the number of actions. THETA(i, j) is the probability of transitioning to state j under state-action pair i. NB: THETA should be sparse.
+% 
+% F: The N x K feature matrix, where N is the number of states, and K is the number of features. F(i, j) is the jth feature value for the ith state.
+% 
+% GAMMA: The discount factor, which must be a real number in [0, 1).
+% 
+% w: A K x 1 vector that is a convex combination, i.e. the components of w are nonnegative and sum to 1.
+%
+% INIT_FLAG: Specifies initial state distribution. Either 'uniform' or 'first' (i.e. concentrated at first state)
+%
+% VV: Initial per-state feature expectations. Intended to be carried over from previous invocations. 
+% 
+% Output:
+% 
+% P: The 1 x N vector that describes the optimal policy with respect to the reward function R = F*w. P(i) is the action taken in the ith state by the optimal policy.
+% 
+% M: The 1 x K vector of "feature expectations" for the optimal policy. M(i) is the expected cumulative discounted value for the ith feature when following the optimal policy (and when starting at state 1).
+%
+% VV: Final per-state feature expectations. Intended to carry over to future invocations. 
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+function [P, M, VV, ITER] = opt_policy_and_feat_exp(THETA, F, GAMMA, w, INIT_FLAG, VV)
+
+%tol = max([0.001 * abs(sum(w)), 0.001]);
+tol = 0.001;
+disp(['tol = ', num2str(tol)]);
[…]

### 2012-02-15 Highway/al_code/run_LPAL.m

@@ -1,0 +1,34 @@
+function [MA, TT] = run_LPAL
+
+% Make the feature value matrix and the transition matrix 
+THETA = make_THETA;
+F = make_F;
+
+% Setup the other parameters
+GAMMA = 0.9;
+E = [5.25, 4.15, 5];
+%E = [7.5, 5, 5];
+%E = [0, 0, 0];
+[NA, N] = size(THETA);
+A = NA / N;
+
+% Run the LPAL algorithm
+[X, MA, TT] = LPAL(THETA, F, GAMMA, E, 'first');
+
+% Construct stationary policy from occupancy measure
+P = zeros(N, A);
+for s=1:N
+	sum = 0;
+	for a=1:A
+		sa = A * (s - 1) + a;
+		sum = sum + X(sa);
+	end
+
+	for a=1:A
+		sa = A * (s - 1) + a;
+		P(s, a) = X(sa) / sum;
+	end
+end
+
+% Write out that policy
+write_out_policy(P);

### 2012-02-15 Highway/al_code/write_out_policy.m

@@ -1,0 +1,31 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This function writes a policy to a text file, in a format suitable for my Perl scripts.
+%
+% Here's a description of the parameters:
+% 
+% Input:
+% 
+% P: The 1 x N vector that describes the policy. P(i) is the action taken in the ith state by the policy.
+% 
+% --- or ---
+%
+% P: The N x A matrix that describes the policy. P(i, j) is the probability of taking the jth action in the ith state.
+% 
+% Output (file):
+%
+% policy.dat: File containing N lines, with each line containing A comma-separated values, where N is the number of states and A is the number of actions. . The jth value in the ith line is the probability of taking the jth action in the ith state.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+function write_out_policy(P)
+
+if (isvector(P))
+	A = max(P);
+
+	for a=1:A
+	   AA(a, find(P == a)) = 1;
+	end
+
+	csvwrite('policy.dat', AA');
+else
+	csvwrite('policy.dat', P)
+end

### 2012-02-20 klein2012/main.tex

--- 2012-02-13 klein2012/main.tex
@@
-% Exemple de fichier fonctionnnant avec la classe CAp2012.cls
-% Base :
-%	Classe pf2010.cls adpatée de pf2003 par Jean CHarlet
-%	Classe pf2003.cls adpatée de ic2001 par Jean CHarlet
-%
-% 	Classe ic2001.cls adpatée de EEGDRI3 par Jean CHarlet
-%
-% 	Classe EEGDRI3.cls adpatée de ic2000 par Jean CHarlet
-%
-% 	Classe IC'2000 (ic2000.cls) par Jean Charlet
-%
-% 	adaptée de la classe IC'99 (afia99.cls) développée par Fabien Torre
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\documentclass{CAp2012}
-\documentclass[publibook-draft]{CAp2012}
+% Use the following line _only_ if you're still using LaTeX 2.09.
+%\documentstyle[icml2012,epsf,natbib]{article}
+% If you rely on Latex2e packages, like most moden people use this:
+\documentclass{article}
 
-\newcommand{\verbfichex}{CAp2012.\xspace}
-\newcommand{\verbfichclass}{CAp2012.cls\xspace}
-\newcommand{\verbfichbst}{CAp2012.bst\xspace}
-
-%\usepackage[pdftex]{pstricks}
-
-\usepackage{verbatim}
+% For figures
+\usepackage{graphicx} % more modern
+%\usepackage{epsfig} % less modern
 \usepackage{subfigure}
-\usepackage[french]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
 \usepackage{mathtools}
-\mathtoolsset{showonlyrefs=true}
 \usepackage{amssymb}
 \usepackage{natbib}
-\usepackage{caption}
-%%% francisation des algorithmes
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage{hyperref}
+\newcommand{\theHalgorithm}{\arabic{algorithm}}
+\usepackage{icml2012}
+\mathtoolsset{showonlyrefs=true}
+\icmltitlerunning{Submission and Formatting Instructions for ICML 2012}
 
-%\newtheorem{algorithm}{Algorithme}
 \newtheorem{definition}{Définition}
@@
 
-\shorttitle{Classification structurée pour l'ARI}
-
-\shortouvrage{CAp 2012}
+%\shorttitle{Classification structurée pour l'ARI}
+%
+%\shortouvrage{CAp 2012}
 
@@
 
-\title{Classification structurée pour l'apprentissage par renforcement inverse}
-
-\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
-
-\institute{
-Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
-\and
-Equipe ABC,
-LORIA, France
-\and
-UMI 2958
-GeorgiaTech-CNRS, France
-}
+%\title{Classification structurée pour l'apprentissage par renforcement inverse}
+%
+%\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
+%
+%\institute{
+%Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
+%\and
+%Equipe ABC,
+%LORIA, France
+%\and
+%UMI 2958
+%GeorgiaTech-CNRS, France
+%}
 
 \begin{document}
-\maketitle
+
+\twocolumn[
+\icmltitle{Submission and Formatting Instructions for the Twenty-Eighth \\
+           International Conference on Machine Learning (ICML 2012)}
+
+% It is OKAY to include author information, even for blind
+% submissions: the style file will automatically remove it for you
+% unless you've provided the [accepted] option to the icml2012
+% package.
+\icmlauthor{Your Name}{email@yourdomain.edu}
+\icmladdress{Your Fantastic Institute,
+            314159 Pi St., Palo Alto, CA 94306 USA}
+\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
+\icmladdress{Their Fantastic Institute,
+            27182 Exp St., Toronto, ON M6H 2T1 CANADA}
+
+% You may provide any keywords that you
+% find helpful for describing your paper; these are used to populate
+% the "keywords" metadata in the PDF but will not be shown in the document
+\icmlkeywords{boring formatting information, machine learning, ICML}
+
+\vskip 0.3in
+]
 
 \begin{abstract}
-  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse (\emph{ARI}).
-Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
-L'ARI part du postulat que l'expert optimise avec succès une fonction de récompense ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
-Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
-Notre contribution consiste en un nouvel algorithme d'ARI levant l'ensemble de ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et surtout ne nécessitant pas la résolution du problème direct.
-  \motscles{Apprentissage par renforcement inverse, classification multi-classe, attribut vectoriel moyen}
+
 \end{abstract}
@@
 %%%%%
-Dans cette contribution, nous nous plaçons dans le contexte de l'apprentissage par démonstration.
+%Dans cette contribution, nous nous plaçons dans le contexte de l'apprentissage par démonstration.
 Dans ce cadre, un agent artificiel apprend à reproduire une tâche grâce à l'observation d'un expert réalisant cette tâche de manière optimale.
@@
 Le but de l'agent est alors d'inférer cette fonction des démonstrations données par l'expert pour ensuite optimiser son comportement relativement à cette fonction lui aussi.\\
-
-La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
+In this paper, we consider the apprenticeship learning problem (also called learning by watching or learning from demonstrations problem) where an artificial agent (or apprentice) is learning to perform a task thanks to the observation of an expert demonstrating the task.
+In order to find a solution to this problem, we propose to use the inverse reinforcement learning paradigm where we suppose that the expert acts in order to be rewarded by an unknown reward function.
+The aim of the agent is then to infer the unknown reward function from the demonstrations given by the expert in order to find an optimal behavior (or policy) with respect to the reward function he has just infered.
+%La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
 Plusieurs directions ont depuis été explorées à partir de deux articles fondateurs \cite{ng2000algorithms} et \cite{abbeel2004apprenticeship}.\\
-Dans cette littérature (voir \cite{neu2009training} pour un bon survol), plusieurs verrous sont identifiés mais souvent ignorés pour proposer des algorithmes de résolution.
+%Dans cette littérature (voir \cite{neu2009training} pour un bon survol), plusieurs verrous sont identifiés mais souvent ignorés pour proposer des algorithmes de résolution.
 Particulièrement, on suppose souvent connue la dynamique de l'environnement dans lequel évoluent l'expert et l'agent ou bien la mise à disposition de cet environnement (ou d'un simulateur) pour y tester les effets d'une politique quelconque.
@@
 Ces contraintes sont souvent incompatibles avec la mise en oeuvre des algorithmes proposés sur des cas réels.\\
-
-Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert.
-Nous proposons une preuve de convergence de cet algorithmes et fournissons une expérimentation sur le problème du pendule inversé.
+The literature on the subject is quite recent and finds its origin in \cite{russell1998learning}.
+Different approaches have been explored since the first articles on the matter \cite{ng2000algorithms} and \cite{abbeel2004apprenticeship}.\\
+In this literature (see \cite{neu2009training} for a survey), several difficulties are identified but often ignored in order to find algorithms.
+Specifically, we often suppose the dynamic of the environment where the expert and the agent are evolving known, or that we dispose of an environment model (via a simulator) in order to test the effects of a policy.
+Moreover, these approaches resolve the reinforcement learning problem (find an optimal policy knowing a given reward) an important number of times in order to resolve the inverse reinforcement problem (find a reward explaining the expert policy).
+Finally, some of these approaches give a policy that performs as well as the expert with respect to the unknown reward function but do not give a reward able to explain the expert policy.
+These constraints can be incompatible with the application of the proposed algorithms in real cases.
+%Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert.
+Nous proposons une preuve de convergence de cet algorithme et fournissons une expérimentation sur le problème du pendule inversé.
 Le choix de cette application se justifie par la difficulté  qu'elle pose pour les méthodes de résolution de la littérature.
+In this paper, we propose a resolution method of the inverse reinforcement problem that is free from any other knowledge than the demonstrations given by the expert.
+We propose a proof of the convergence of our algorithm and give an experimentation on the inverted pendulum problem.
+The choice of this application is justified by the fact that it is difficult to resolve it by the other methods in the literature.
 %%%%
-\section{Contexte}
+\section{Preliminaries}
 %%%%%
@@
 %%%
-\subsection{Apprentissage par renforcement direct et inverse}
+\subsection{Reinforcement Learning (RL) and Inverse Reinforcement Learning (IRL)}
 %%%
-Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
+%Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
 La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$.
@@
 Une politique de contrôle déterministe $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
-
-Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
+The general framework chosen is the sequential decision making problem posed in the Markov decision process (MDP) formalism.
+The configuration of the system is completely described at each discrete time $t$ by the state $s_t \in S$ where the agent must take a decision $a_t\in A$.
+Then, the system will evolve to the next state $s_{t+1} \in S$ with the probability $p(s_{t+1}|s_t, a_t)$.
+A deterministic policy which is a function $\pi : S\rightarrow A$ allows us to define the behavior of an agent faced to a sequential decision making problem.
+In this paper $S$ and $A$ will be finite sets and $\{P_{sa}\}_{s\in S,a\in A}$ will be a set of probability distributions, from $S$ to $[0,1]$ such that $P_{sa}(s')=p(s'|s, a)$, called transition probabilities.
+%Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
 Une des forces de ce cadre réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
 On définit pour ce faire la fonction de valeur $V^\pi$ :
+In the RL paradigm, the choice of a policy $\pi$ depends on a reward function $R : S \rightarrow \mathbb{R}$ which defines the immediate reward $R(s)$ given to the system when it attains the state $s \in S$.
+In this paradigm, the agent, following a policy $\pi$, is not going to learn to maximize the immediate reward but a criterion taking into account the futures rewards reached by the agent.
+That allows to specify the goal without specifying a way to reach it.
+The criterion commonly chosen is the value function $V^\pi$ :
 \begin{equation}
 \label{Vdef.eqn}
-V^\pi(s) = E[\sum_t\gamma^tR(s_t)|s_0=s,\pi]=R(s) + \gamma E[V^\pi(s')|s,\pi]
+V^\pi(s) = E^\pi_s[\sum_{t\geq0}\gamma^tR(s_t)]=R(s) + \gamma\sum_{s'\in S}P_{s\pi(s)} V^\pi(s'),
 \end{equation}
-où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action :
+%où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action :
+where $\gamma \in [0,1)$ is a discount factor, and $E^\pi_s$ is the expectation over the distribution of the state sequences $(s_0,s_1,\dots)$ we pass through when we execute the policy $\pi$ starting from $s_0=s$.
+It is also possible to define the action-value function $Q^\pi$ which adds a degree of freedom in the choice of the first action :
 \begin{equation}
 \label{Qdef.eqn}
-Q^\pi(s,a) = E[\sum_t\gamma^tR(s_t)|s_0=s,a_0=a,\pi] = R(s) + \gamma E[V^\pi(s')|s,a].
+Q^\pi(s,a) = E^\pi_{sa}[\sum_{t\geq0}\gamma^tR(s_t)]=R(s) + \gamma\sum_{s'\in S}P_{sa} V^\pi(s'),
 \end{equation}
-  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
+where $E^\pi_{sa}$ is the expectation over the distribution of the state sequences $(s_0,s_1,\dots)$ we pass through when we start from $s_0=s$, choose the action $a_0=a$ and then follow the policy $\pi$.
+%Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
 La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
+An optimal policy $\pi^*$ is defined as a policy which value function $V^{\pi*}=V^*=\sup_{\pi}V^{\pi}$ which means $\forall \pi, \forall s, V^{\pi*}(s) \geq V^\pi(s)$.
+An optimal policy $\pi^*$ can be computed easily thanks to the optimal action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy mechanism :
 \begin{equation}
 \label{greedy.eqn}
-\pi^*(s) \in \arg\max_a Q^*(s,a).
+\pi^*(s) \in \arg\max_{a\in A} Q^*(s,a).
 \end{equation}
-L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov (\emph{PDM}).\\
-
-Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive.
+%L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov .\\
+The tuple $M=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,R\}$ containing the finite state space $S$, the finite action space $A$, the transitions probabilities $\{P_{sa}\}_{s\in S,a\in A}$, the discount factor $\gamma$ and the reward function $R$ is called a finite Markov decision process.
+Resolving the RL problem with respect to the criterion $V^\pi$  means to find $\pi^*$.
+In a finite MDP the existence of an optimal deterministic policy is guaranteed when the reward function $R$ is bounded.\\
+%Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive.
 Un exemple de ce type de tâche est la conduite d'une voiture.
@@
 Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
-
-C'est la définition de l'apprentissage par renforcement inverse.
+Sometimes, specifying manually the reward function can be very difficult while it is natural to an operator to control the system in order to complete the desired task.
+An exemple of this kind of task is driving a car.
+We can accomplish this task very easily in our daily-life but it would be very tricky to specify the precise weights given to each important driving-parameters such as the distance we should keep between the car in front of us, or maintaining a reasonable speed and so on.
+In such cases, it is useful to infer the reward function thanks to observed behavior.\\
+%C'est la définition de l'apprentissage par renforcement inverse.
 Le problème est de retrouver la fonction de récompense optimisée par un expert.
 Usuellement l'expert est considéré comme un agent optimal dans un PDM et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
-
-Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}).
+That is the aim of the Inverse Reinforcement Learning.
+The problem is to find the reward function optimized by an expert.
+Usually, the expert is considered as an optimal agent in a MDP and we have access to the expert optimal policy $\pi_E$ or to some trajectories of this policy.
+%Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}).
 Particulièrement, tout comportement est optimal vis-à-vis de la récompense uniformément nulle.
 Il convient donc de contraindre l'espace des solutions.
+This problem is ill-posed in the sense that there is not uniqueness of the reward function for which the expert policy is optimal (\cite{ng1999policy}).
+Particularly, each policy is optimal with respect to the uniformly-zero reward function.
+One should favor some solutions.
 %%%
-\subsection{Attribut vectoriel moyen}
+\subsection{Feature expectation}
 \label{ConsiderationsTechniques.sec}
 %%%
-La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$ telles que
+%La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$  telles que
+The reward function $R$ is the unknown of the problem. We can suppose that it is possible to dispose of $p$ features $\psi_{1\leq i \leq p}$, which are functions from $S$ to $\mathbb{R}$, such  that :
 \begin{equation}
 \label{hatRdef.eqn}
-\exists \theta | R(s) = \theta^T\psi(s) = \sum_{i=1}^N\theta_i\psi_i(s).
+\exists \theta | R(s) = \theta^T\psi(s) = \sum_{i=1}^p\theta_i\psi_i(s).
 \end{equation}
-Cette contrainte peut cependant être relâchée si nécessaire.
+%Cette contrainte peut cependant être relâchée si nécessaire.
 Introduire cette expression dans la définition de la fonction de qualité (Eq.
 \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+This constraint can be relaxed if necessary.
+Putting this expression of the reward function in the definition of the action-value function (Eq.
+\eqref{Qdef.eqn}) allows us to define the term $\mu^\pi$ which is often used in the existing algorithms :
 \begin{equation}
-Q^\pi(s,a) = \theta^T\mu^\pi(s,a),\mu^\pi(s,a) = E[\sum_t\gamma^t\psi(s_t)|s_0=s,a_0=a,\pi]
+Q^\pi(s,a)=\theta^T\mu^\pi(s,a),\text{ }\mu^\pi(s,a) = E^\pi_{sa}[\sum_t\gamma^t\psi(s_t)].
 \label{Qmu.eqn}
 \end{equation}
-
-Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique.
+%Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique.
 On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
-
-On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
+The term $\mu^\pi$ is called the feature expectation of the policy $\pi$.
+We can observe that the chosen parametrization of the reward function $R$ and the dynamic resulting from the application of the policy $\pi$ in the MDP impose the parametrization of $Q^\pi$.
+To shorten notations, we may use $\mu^\pi_{sa}$ as $\mu^\pi(s,a)$.\\
+%On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
 En effet
+We see also that if two policies share the same feature expectation, then these two policies have the same value function with respect to the reward function $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$.
+Indeed :
 \begin{equation}
@@
 \end{equation}
-L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
+%L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
 La plupart des algorithmes de la littérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
 Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert, quelle que soit la récompense (voir par exemple \cite{neu2009training}).
+That is why most of the algorithms in the literature have, as we will see, the objective of minimizing a \textit{distance}  between the feature expectation of the expert and the feature expectation of the apprentice.
+That allows to obtain an apprentice which policy has the same value function as the expert policy, for a reward function of the type $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$ (see for instance \cite{neu2009training}).
 %%%%%
-\section{Classification structurée pour l'ARI}
+\section{Structured classification for IRL}
 %%%%%
 %%%
-\subsection{Principe}
+\subsection{Principle}
 %%%
-Comme nous l'avons vu Eq.
-\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_aQ^E_R(s,a)$.\\
-Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
-Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
+%Comme nous l'avons vu Eq.
+\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
+%Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+%Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
 Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons :
+As seen in Eq.
+\eqref{greedy.eqn}, there exists a greedy mechanism linking the expert policy $\pi_E$ and the optimal action-value $Q^E_R$ relative to the unknown reward function $R$: $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
+We will use this property to consider an empirical risk function $J_N$ dependent on a score function $q$ which will have to share with the unknown action-value function $Q^E_R$ the same optimal property.
+Let's note $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ the samples coming from one or several  trajectories of the expert.
+We are searching for a score function $q$ which verifies for each sample :
 \begin{equation}
 \label{butLAFEM.eqn}
-a_i \in \arg\max_aq(s_i,a).
+a_i \in \argmax_{a\in A}q(s_i,a).
 \end{equation}
-Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+%Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
 Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
@@
 Nous introduisons ainsi la fonction de risque empirique suivante :
+So, we want to introduce a risk function which should penalize the cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases the score function $q$ does not allow to explain the choice $a_i$ of the expert (see equation \eqref{greedy.eqn}).
+Moreover, we introduce a loss function $l : S \times A \rightarrow \mathbb{R}_+$; this function allows the user to put some \emph{a priori} knowledge in the system because the user will choose the gap wanted in the score function between the action chosen by the expert and the second best possible action as we can see in the following equation : $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Introducing a loss function is quite common in the literature (see Sec.
+\ref{biblio.sec}).
+We introduce then the following empirical risk function :
 \begin{equation}
 \label{Jdef.eqn}
-   J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
+J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
 \end{equation}
-Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
-
-Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
+%Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
+%
+%Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
 Notre propos n'est cependant pas d'utiliser une approche purement supervisée.
@@
 Il faut pour cela prendre en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnées par la politique de l'expert.\\
-
-Comme nous l'avons vu Eq.
+%
+%Comme nous l'avons vu Eq.
 \eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
@@
 \eqref{hatRdef.eqn}) :
+By default, choosing $l(s_i,a) = 1$ if $a\neq a_i$ and $l(s_i,a_i)=0$ works quite well as shown in the experimentations (Sec. \ref{exp.sec}) and in the analysis (Sec. \ref{proof.sec}).\\
+Here, we deal with the imitation problem as a classification problem.
+However, our purpose is not to use a purely supervised approach.
+We do not want to learn the policy of the expert but rather the description of the task he is executing (the reward function).
+In order to do that, it is necessary to take into account the MDP structure, and in particular the information on the transitions between states which are conditioned by the expert policy.\\
+As seen in Eq.
+\eqref{Qmu.eqn}, the hypothesis of a linear representation of the reward function by some features $\psi : S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) = \theta^T\mu^\pi(s,a)$.
+This parametrization takes into account the dynamic imposed by $\pi_E$.
+The expert follows the same rule and if we note $\mu^E$ the feature expectation of the expert, we can do the hypothesis that the score function $q$ that we are searching for can be written as follows : $q(s,a) = \theta^T\mu^E(s,a)$.
+Estimating the term $\mu^E$ is the key of our algorithm and we propose some solution in Sec.
+\ref{calculmu.sec}.
+We can now give the expression of our empirical risk function as a function of the parameter $\theta$ which defines the reward function (Eq.
+\eqref{hatRdef.eqn}) :
 \begin{equation}
 \label{structure.eqn}
-   J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
+J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) - \theta ^T \mu^E_{s_ia_i} \right)
 \end{equation}
-Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
+%Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
 Nous utilisons une descente de sous-gradient (normalisé), une généralisation du gradient, pour le minimiser.
@@
 En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée :
+This risk function is non linear and non differentiable everywhere because of the operator $\max$.
+But this function is clearly convex, closed, proper, continuous and piecewise-linear with finite values as $A$ is finite.
+We can then use a sub-gradient descent algorithms, which is a generalization of a gradient descent algorithm, in order to minimize the risk function.
+A sub-gradient of a convex function  $f:\mathbb{R}^p\rightarrow \mathbb{R}$ in $x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$, the set of all sub-gradients in $x$ is $\partial_xf$.
+The unique sub-gradient in $x$ of a differentiable convex function $f$ is the gradient $\nabla_xf$.
+As $A$ is finite, the convex and piecewise differentiable function $\max_{a\in A}[f(\cdot,a)]$, from $\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from $\mathbb{R}^p$ to $\mathbb{R}$,  has one of its sub-gradient in $x\in\mathbb{R}^p$ equals to $\nabla_xf(x,a^*)$ with $a^*\in\arg\max_{a\in A}f(x,a)$.
+Applying these rules, we have a sub-gradient of our risk function and an update rule for our sub-gradient algorithm :
 \begin{eqnarray}
-\nabla_\theta J_N(\theta) &=& {1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\\
-a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))\\
-   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_\theta J_N(\theta) \over ||\nabla_\theta J_N(\theta)||_2}.
+g_{\theta} &=& {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right),\\
+a^*_i &=& \arg\max_{a\in A}(\theta ^T \mu^E(s_i,a) + l(s_i,a)),\\
+g_{\theta} &\in& \partial_\theta J_N,\\
+\theta_{t+1} &\leftarrow& \theta_t -\alpha_t{g_{\theta} \over \|g_{\theta}\|_2}.
 \end{eqnarray}
 %%%
-\subsection{Calcul de $\mu_E$}
+\subsection{Estimation of $\mu^E$}
 %%%
 \label{calculmu.sec}
-Comme signalé dans la Sec.
+%Comme signalé dans la Sec.
 \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
-Nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu^E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
+Nous avons besoin de connaître $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$.
 Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
+As mentioned in \ref{ConsiderationsTechniques.sec}, some algorithms in the literature depend on the implementation of the feature expectation.
+Our algorithm, in particular, relies only on the implementation of $\mu^E$, the feature expectation of the expert, and does not require the repeated calculus of the feature expectation of another policy.
+For our algorithm, we only need to know $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$.
+That implies to infer the feature expectation of the expert on states where he has been, but on actions he has not taken.
 
-Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve :
+%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve :
+This inference is made easier by the fact that the feature expectation obeys the Bellman equation. Indeed, when we write its definition for a component $1\leq j\leq p$, we have :
 \begin{equation}
-\mu_E^j(s,a) = E[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)|s_0 = s, a_0 = a, \pi_E],
+\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t \psi_j(s)].
 \end{equation}
-expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
-Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu_E$.\\
+%expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
+Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu^E$.\\
+When we compare this expression to the equation \eqref{Qdef.eqn}, it allows us to affirm that $\mu_j^E$ can be seen as the action-value function for the policy $\pi_E$ with respect to the reward function defined by the $j$-th component of the features $\psi$.
+Depending on the prior information, several methods can be used in order to estimate $\mu^E$.\\
 
-Dans le cas idéal où le modèle est connu, le calcul exact de $\mu_E$ peut être effectué grâce aux algorithmes de programmation dynamique.
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a)$ (définition Eq.
+%Dans le cas idéal où le modèle est connu, le calcul exact de $\mu^E$ peut être effectué grâce aux algorithmes de programmation dynamique.
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu^E(s,a)$ (définition Eq.
 \eqref{Qmu.eqn}) l'on utilise
+In the ideal case where a model of the environnement is known (\textit{id est} the transition probabilities are given), the exact calculus of $\mu^E$ can be executed thanks to dynamic programmation algorithms.
+In a more realistic case, we can use a simple Monte-Carlo estimation if we dispose of a simulator and if it is possible to ask the expert what action has to be chosen in an important number of arbitrary states.The estimation of $\mu^E(s,a)$ (definition Eq.
+\eqref{Qmu.eqn}) is done over $N$ trajectories of length $2k+1$, $\{s^i_0,a^i_0,\dots,s^i_{k-1},a^i_{k-1},s^i_{k}\}_{1\leq i \leq k}$, by the following formula :
 \begin{equation}
-\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\{\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, s_{t+1} \sim p(\cdot|s_t,\pi_E(s_t))\}.
+\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k}_{t=0}\gamma^t\psi(s^i_t),
 \end{equation}
+where $s^i_0=s, a^i_0=a$ and $s^i_{t+1}$ is distributed according to $P_{s^i_t\pi_E(s^i_t)}$.
 
-Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
+%Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
 Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
-C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu_E$ en utilisant LSTD.
+C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu^E$ en utilisant LSTD.
 Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
-
-D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
+Finally, if we are in the framework where only the data at our disposition are the trajectories samples given by the expert, it is possible to use the LSTD$\mu$ algorithm \citep{klein2011batch} in its \emph{on-policy} version.
+This algorithm adapts the LSTD algorithm (\cite{bradtke1996linear}) keeping its characteristics, in particular its ability to execute the calculus in  {\it batch}-mode.
+This is the method that we show in the section \ref{exp.sec}.
+The principle is to estimate each component of $\mu^E$ using LSTD.
+Some calculus can be aggregated, giving to LSTD$\mu$ a cost of calculus which is of the same order of magnitude as LSTD.
+%D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
 Cela est reporté pour des travaux futurs.
 %%%%%
-\section{Analyse}\label{proof.sec}
-\subsection{Hypothèses}
-Ici, nous présentons le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $R_N$ qui rende la politique de l'expert optimale.
-On supposera que la récompense $R$ s'écrive sous la forme $R(s)=\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$, que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$ et que $\mu_E$ est estimée sans erreur sur $\{(s_i,a_i)\}_{1\leq i \leq N}$.
-Nous supposons également l'unicité de l'action optimale (cette hypothèse pouvant être relâchée au prix de notations plus complexes):
-\begin{equation}\label{eq:existence}
-\forall 1\leq i \leq N, \theta_E^T\mu_E(s_i,a_i)>\max_{a\in A \backslash a_i}\theta_E^T\mu_E(s_i,a),
-\end{equation}
-où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$.
-Ainsi tout vecteur $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale.
-Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation~\eqref{eq:existence} impliquera dans ce cas :$\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
-Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation~$\eqref{eq:existence}$ qui minimise $R_N$,  que les minimiseurs de $R_N$ vérifient bien l'équation~\eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
-\subsection{Résultat}
-On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $R_N$.
-On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:R_N(\theta)=\inf_{\theta \in \mathbb{R}^p}R_N(\theta)\}$.
-Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
-Il est facile de borner inférieurement le risque $R_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
-Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
-Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
-Montrons qu'il existe un $x\geq0$ tel que $~R_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$R_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
-Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
-\begin{align}
-&\hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{a\in A \backslash a_i}l(s_i,a)\geq0,\\
-\Leftrightarrow&x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
-\end{align}
-Or comme la somme des $\max$ est supérieure au $\max$ d'une somme, il est clair que $\forall 1\leq i \leq N$:
-\begin{align}
-&x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},\\
-\Leftrightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{a\in A \backslash a_i}l(s_i,a)\geq0,\\
-\Rightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,\\
-\Leftrightarrow&J_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
-\end{align}
-C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $J_N$ vérifient bien l'équation \eqref{eq:existence}.
-Soit ~$\theta^*\in\ M^*$.Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $J_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, J_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $J_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
-Or on a $\forall 1\leq i \leq N$ :
-\begin{align}
-&J_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),\\
-\Leftrightarrow&\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,\\
-\Rightarrow&(\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
-\end{align}
-Donc~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
-Finalement, montrons que l'algorithme de sous-gradient converge.
-Tout d'abord, il est clair que $R_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
-De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article de \cite{correa1993convergence} qui nous assure que la suite :
-\begin{equation}
-\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\alpha_t{\nabla_\theta R_N(\theta_t) \over ||\nabla_\theta R_N(\theta_t)||_2},
-\end{equation}
-est convergente vers un minimiseur $\theta^*$ de $R_N$, où $alpha_t$ est un taux d'apprentissage satisfaisant les condition usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
-Ce minimiseur vérifiera l'équation~\eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation~\eqref{eq:existence} implique l'équation d'optimalité de Bellman.
-\section{Travaux Connexes}
+Other estimation methods are possible, as for instance a Monte-Carlo estimation, less calculus-expensive and constraining than the first one, which would be based solely on the data given by the expert and would imply the use of an heuristic on the unknown actions.
+This estimation would be consider on future works.
+\section{Analysis}\label{proof.sec}
+\subsection{Convergence of the sub-gradient algorithm}
+
+\subsection{Quality of the solution found by the sub-gradient algorithm}
+\section{Related work}
 \label{biblio.sec}
@@
 %%%%%
-\section{Illustration expérimentale}
+\section{Experiments}
 \label{exp.sec}
@@
 %%%
-\subsection{Description de la tâche}
+\subsection{Description of the task}
 %%%
@@
 %%%
-\subsection{Illustration des résultats}
+\subsection{Illustrations}
 %%%
@@
 \bibliography{Biblio}
-
+\bibliographystyle{icml2012}
 \end{document}

### 2012-02-21 klein2012/main.tex
newone

@@
 In order to find a solution to this problem, we propose to use the inverse reinforcement learning paradigm where we suppose that the expert acts in order to be rewarded by an unknown reward function.
-The aim of the agent is then to infer the unknown reward function from the demonstrations given by the expert in order to find an optimal behavior (or policy) with respect to the reward function he has just infered.
+The aim of the agent is then to infer the unknown reward function from the demonstrations given by the expert in order to find an optimal behavior (or policy) with respect to the reward function he has just inferred.
 %La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
@@
 Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité de transition markovienne $p(s_{t+1}|s_t, a_t)$.
-Une politique de contrôle déterministe $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
+Une politique de contrôle déterministe $\pi: S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
 The general framework chosen is the sequential decision making problem posed in the Markov decision process (MDP) formalism.
@@
 Then, the system will evolve to the next state $s_{t+1} \in S$ with the probability $p(s_{t+1}|s_t, a_t)$.
-A deterministic policy which is a function $\pi : S\rightarrow A$ allows us to define the behavior of an agent faced to a sequential decision making problem.
+A deterministic policy which is a function $\pi: S\rightarrow A$ allows us to define the behavior of an agent faced to a sequential decision making problem.
 In this paper $S$ and $A$ will be finite sets and $\{P_{sa}\}_{s\in S,a\in A}$ will be a set of probability distributions, from $S$ to $[0,1]$ such that $P_{sa}(s')=p(s'|s, a)$, called transition probabilities.
-%Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
+%Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R: S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
 Une des forces de ce cadre réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
-On définit pour ce faire la fonction de valeur $V^\pi$ :
-In the RL paradigm, the choice of a policy $\pi$ depends on a reward function $R : S \rightarrow \mathbb{R}$ which defines the immediate reward $R(s)$ given to the system when it attains the state $s \in S$.
+On définit pour ce faire la fonction de valeur $V^\pi$:
+In the RL paradigm, the choice of a policy $\pi$ depends on a reward function $R: S \rightarrow \mathbb{R}$ which defines the immediate reward $R(s)$ given to the system when it attains the state $s \in S$.
 In this paradigm, the agent, following a policy $\pi$, is not going to learn to maximize the immediate reward but a criterion taking into account the futures rewards reached by the agent.
 That allows to specify the goal without specifying a way to reach it.
-The criterion commonly chosen is the value function $V^\pi$ :
+The criterion commonly chosen is the value function $V^\pi$:
 \begin{equation}
@@
 \end{equation}
-%où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action :
+%où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action:
 where $\gamma \in [0,1)$ is a discount factor, and $E^\pi_s$ is the expectation over the distribution of the state sequences $(s_0,s_1,\dots)$ we pass through when we execute the policy $\pi$ starting from $s_0=s$.
-It is also possible to define the action-value function $Q^\pi$ which adds a degree of freedom in the choice of the first action :
+It is also possible to define the action-value function $Q^\pi$ which adds a degree of freedom in the choice of the first action:
 \begin{equation}
@@
 %Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
-La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
+La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton:
 An optimal policy $\pi^*$ is defined as a policy which value function $V^{\pi*}=V^*=\sup_{\pi}V^{\pi}$ which means $\forall \pi, \forall s, V^{\pi*}(s) \geq V^\pi(s)$.
-An optimal policy $\pi^*$ can be computed easily thanks to the optimal action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy mechanism :
+An optimal policy $\pi^*$ can be computed easily thanks to the optimal action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy mechanism:
 \begin{equation}
@@
 %La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$  telles que
-The reward function $R$ is the unknown of the problem. We can suppose that it is possible to dispose of $p$ features $\psi_{1\leq i \leq p}$, which are functions from $S$ to $\mathbb{R}$, such  that :
+The reward function $R$ is the unknown of the problem. We can suppose that it is possible to dispose of $p$ features $\psi_{1\leq i \leq p}$, which are functions from $S$ to $\mathbb{R}$, such  that:
 \begin{equation}
@@
 Introduire cette expression dans la définition de la fonction de qualité (Eq.
-\eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
+\eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants:
 This constraint can be relaxed if necessary.
 Putting this expression of the reward function in the definition of the action-value function (Eq.
-\eqref{Qdef.eqn}) allows us to define the term $\mu^\pi$ which is often used in the existing algorithms :
+\eqref{Qdef.eqn}) allows us to define the term $\mu^\pi$ which is often used in the existing algorithms:
 \begin{equation}
@@
 We see also that if two policies share the same feature expectation, then these two policies have the same value function with respect to the reward function $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$.
-Indeed :
-\begin{equation}
+Indeed: \begin{equation}
 \mu^{\pi_1} = \mu^{\pi_2} \Rightarrow \theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} = V^{\pi_2}.
@@
 %Comme nous l'avons vu Eq.
-\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
+\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue: $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
 %Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
 %Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
-Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons :
+Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons:
 As seen in Eq.
@@
 Let's note $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ the samples coming from one or several  trajectories of the expert.
-We are searching for a score function $q$ which verifies for each sample :
+We are searching for a score function $q$ which verifies for each sample:
 \begin{equation}
@@
 %Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
-Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Nous introduisons de surcroit une fonction $l: S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
 Ce type d'argument est assez commun dans la littérature (voir Sec.
 \ref{biblio.sec}).
-Nous introduisons ainsi la fonction de risque empirique suivante :
+Nous introduisons ainsi la fonction de risque empirique suivante:
 So, we want to introduce a risk function which should penalize the cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases the score function $q$ does not allow to explain the choice $a_i$ of the expert (see equation \eqref{greedy.eqn}).
-Moreover, we introduce a loss function $l : S \times A \rightarrow \mathbb{R}_+$; this function allows the user to put some \emph{a priori} knowledge in the system because the user will choose the gap wanted in the score function between the action chosen by the expert and the second best possible action as we can see in the following equation : $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Moreover, we introduce a loss function $l: S \times A \rightarrow \mathbb{R}_+$; this function allows the user to put some \emph{a priori} knowledge in the system because the user will choose the gap wanted in the score function between the action chosen by the expert and the second best possible action as we can see in the following equation: $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
 Introducing a loss function is quite common in the literature (see Sec.
 \ref{biblio.sec}).
-We introduce then the following empirical risk function :
+We introduce then the following empirical risk function:
 \begin{equation}
@@
 %Comme nous l'avons vu Eq.
-\eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
+\eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi: S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
 Cette paramétrisation prend en compte la dynamique imposée par $\pi_E$.
@@
 \ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Eq.
-\eqref{hatRdef.eqn}) :
+\eqref{hatRdef.eqn}):
 By default, choosing $l(s_i,a) = 1$ if $a\neq a_i$ and $l(s_i,a_i)=0$ works quite well as shown in the experimentations (Sec. \ref{exp.sec}) and in the analysis (Sec. \ref{proof.sec}).\\
@@
 As seen in Eq.
-\eqref{Qmu.eqn}, the hypothesis of a linear representation of the reward function by some features $\psi : S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) = \theta^T\mu^\pi(s,a)$.
+\eqref{Qmu.eqn}, the hypothesis of a linear representation of the reward function by some features $\psi: S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) = \theta^T\mu^\pi(s,a)$.
 This parametrization takes into account the dynamic imposed by $\pi_E$.
-The expert follows the same rule and if we note $\mu^E$ the feature expectation of the expert, we can do the hypothesis that the score function $q$ that we are searching for can be written as follows : $q(s,a) = \theta^T\mu^E(s,a)$.
+The expert follows the same rule and if we note $\mu^E$ the feature expectation of the expert, we can do the hypothesis that the score function $q$ that we are searching for can be written as follows: $q(s,a) = \theta^T\mu^E(s,a)$.
 Estimating the term $\mu^E$ is the key of our algorithm and we propose some solution in Sec.
@@
 We can now give the expression of our empirical risk function as a function of the parameter $\theta$ which defines the reward function (Eq.
-\eqref{hatRdef.eqn}) :
+\eqref{hatRdef.eqn}):
 \begin{equation}
@@
 Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $\max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
-En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée :
+En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée:
 This risk function is non linear and non differentiable everywhere because of the operator $\max$.
@@
 As $A$ is finite, the convex and piecewise differentiable function $\max_{a\in A}[f(\cdot,a)]$, from $\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from $\mathbb{R}^p$ to $\mathbb{R}$,  has one of its sub-gradient in $x\in\mathbb{R}^p$ equals to $\nabla_xf(x,a^*)$ with $a^*\in\arg\max_{a\in A}f(x,a)$.
-Applying these rules, we have a sub-gradient of our risk function and an update rule for our sub-gradient algorithm :
+Applying these rules, we have a sub-gradient of our risk function and an update rule for our sub-gradient algorithm:
 \begin{eqnarray}
+\label{algorithm}
 g_{\theta} &=& {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right),\\
@@
 g_{\theta} &\in& \partial_\theta J_N,\\
-\theta_{t+1} &\leftarrow& \theta_t -\alpha_t{g_{\theta} \over \|g_{\theta}\|_2}.
+\theta_{t+1} &\leftarrow& \theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2}.
 \end{eqnarray}
@@
 
-%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve :
-This inference is made easier by the fact that the feature expectation obeys the Bellman equation. Indeed, when we write its definition for a component $1\leq j\leq p$, we have :
+%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve:
+This inference is made easier by the fact that the feature expectation obeys the Bellman equation. Indeed, when we write its definition for a component $1\leq j\leq p$, we have:
 \begin{equation}
@@
 In a more realistic case, we can use a simple Monte-Carlo estimation if we dispose of a simulator and if it is possible to ask the expert what action has to be chosen in an important number of arbitrary states.The estimation of $\mu^E(s,a)$ (definition Eq.
-\eqref{Qmu.eqn}) is done over $N$ trajectories of length $2k+1$, $\{s^i_0,a^i_0,\dots,s^i_{k-1},a^i_{k-1},s^i_{k}\}_{1\leq i \leq k}$, by the following formula :
-\begin{equation}
-\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k}_{t=0}\gamma^t\psi(s^i_t),
+\eqref{Qmu.eqn}) is done over $N$ trajectories of length $\{2k_i+1\}_{1\leq i \leq N}$, $\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i \leq N}$, by the following formula: \begin{equation}
+\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k_i}_{t=0}\gamma^t\psi(s^i_t),
 \end{equation}
@@
 \section{Analysis}\label{proof.sec}
+In this section we are going to prove that our algorithm (see \eqref{algorithm}) is convergent which means that $\tilde{\theta}=\lim_{t\rightarrow+\infty}\theta_t$ is well defined and that $\inf_{\theta\in\mathbb{R}^p}J_N(\theta)=J_N(\tilde{\theta})$ which means that the algorithm reaches the goal of minimizing $J_N$.
+Moreover we will determine the quality of our algorithm by giving an upper bound to the term $E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]$ where $E_{s\sim\nu}$ is the expectation over the stationary probability distribution of states $\nu$ and $\pi_A$ is the policy of the apprentice.
 \subsection{Convergence of the sub-gradient algorithm}
-
+Let $N\in\mathbb{N}^*$ and $A$ a finite set of cardinal $\card{A}$.
+Let $\{f_{i,a}\}_{1\leq i \leq N,a\in A}$ and $\{l_{i,a}\}_{1\leq i \leq N,a\in A}$ be finite sets of cardinal $N\card{A}$ which respectively each element $f_{i,a}\in\mathbb{R}^p$ and $l_{i,a}\in\mathbb{R_+}$  where $p\in\mathbb{N}^*$.
+$\mathbb{R}^p$ is an euclidian space with the scalar product $<.,.>$ and the associate euclidian norm $\|\|_2$.
+We define then the function $f_N$ from $\mathbb{R}^p$ to $\mathbb{R}$ such that $\forall \theta\in\mathbb{R}^p, f_N(\theta)=\frac{1}{N}\sum_{i=1}^{i=N}\max_{a\in A}(\theta^Tf_{i,a}+l_{i,a})$.
+It is clear, as $A$ is finite, that $f_N$ is convex, proper, closed, continuous and with finite values.
+Moreover one of its sub-gradient in $\theta\in\mathbb{R}^p$ is $g_{\theta}=\frac{1}{N}\sum_{i=1}^{i=N}f_{i,a_i^*}$ where $a_i^*\in\argmax_{a \in A}(\theta^Tf_{i,a}+l_{i,a})$.
+We also have $g_{\theta}\in\{\frac{1}{N}\sum_{i=1}^{i=N}f_{i,a}\}_{1\leq i \leq N,a\in A}=G$, $G$ is a finite set .
+Let note $g_{min}$ one element of $\argmin_{G\backslash{0}}(\frac{1}{N}\|\sum_{i=1}^{i=N}f_{i,a}\|_2)$, and
+$\overline{f}_N=\inf_{\theta\in\mathbb{R}^p}f_N(\theta)$. Finally, we suppose $f_N(0)\geq \overline{f}_N\geq0$. 
 \subsection{Quality of the solution found by the sub-gradient algorithm}
@@
 %%%%%
-L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
+L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
 Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de \mbox{l'expert} et celui d'une politique donnée.
@@
 La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}.
-Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un PDM (entrée à classifier) est associée une politique (classe).
+Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires: à un PDM (entrée à classifier) est associée une politique (classe).
 Le cadre applicatif est déterministe, l'algorithme employé pour la résolution des PDM étant $A^*$.
@@
 Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
-Cette fonction de score $s : X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e.
+Cette fonction de score $s: X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e.
 choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$.
@@
 Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations est de $T=20$.
-Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi : S \rightarrow \mathbb{R}^p$ correspondant au mélange de gaussiennes défini dans \cite{lagoudakis2003least}.
+Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange de gaussiennes défini dans \cite{lagoudakis2003least}.
 Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
@@
 Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
+Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace: celle dans laquelle le pendule est proche de la verticale.
 En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.

### 2012-02-22 klein2012/main.tex
Abstract de l'article en Anglais

@@
 \begin{abstract}
-
+  In this paper we tackle the problem of apprenticeship learning in the Inverse Reinforcement Learning (\emph{IRL}) framework.
+An expert fulfils a task to be imitated by an artificial agent.
+The IRL framework assumes that the expert is acting optimally with respect to a reward function ; the goal is to guess this function from trajectories drawn by the expert.
+Existing IRL algorithms need one or more of the following items : complete trajectories from the expert, a generative model of the environment for Monte-Carlo estimations, the knowledge of the transition probabilities, the ability to repeatedly solve the forward problem (the Reinforcement Learning (\empgh{RL}) problem), the access to the expert's strategy anywhere in the state space, or at the very least random samples covering a large part of it.
+Our contribution consists in a new IRL algorithm lifting all these constraints.
+Using a supervised technique in which the structure of the underlying Markov Decision Process (\emph{MDP}) is implicitely injected, we give rise to a subgradient descent-based algorithm whose computational and sample complexities are low.
+The sample base from the expert is the only data we need.
+A formal analysis of the algorithm as well as empirical demonstrations on two classic but challenging problems are provided.
 \end{abstract}

### 2012-02-22 klein2012/main.tex
Article: Related work en Anglais

@@
 The IRL framework assumes that the expert is acting optimally with respect to a reward function ; the goal is to guess this function from trajectories drawn by the expert.
-Existing IRL algorithms need one or more of the following items : complete trajectories from the expert, a generative model of the environment for Monte-Carlo estimations, the knowledge of the transition probabilities, the ability to repeatedly solve the forward problem (the Reinforcement Learning (\empgh{RL}) problem), the access to the expert's strategy anywhere in the state space, or at the very least random samples covering a large part of it.
+Existing IRL algorithms need one or more of the following items : complete trajectories from the expert, a generative model of the environment for Monte-Carlo estimations, the knowledge of the transition probabilities, the ability to repeatedly solve the forward problem (the Reinforcement Learning (\emph{RL}) problem), the access to the expert's strategy anywhere in the state space, or at the very least random samples covering a large part of it.
 Our contribution consists in a new IRL algorithm lifting all these constraints.
@@
 %%%%%
-L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
+%L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
 Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de \mbox{l'expert} et celui d'une politique donnée.
 Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
+The historical approach of \cite{ng2000algorithms}, as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} as summed up in \cite{neu2009training} all suffer from the same constraint: the need to repeatedly solve the forward problem, that is to say the need to repeatedly find the optimal policy with respect to an arbitrary reward.
+All these approaches try to minimize some distance between the feature expectation of the expert and the feature expectation of another policy.
+Successive updates of the vector of parameters $\theta$ will get this other policy (optimal with respect to a criterion that depends on $\theta$, this is where the \emph{RL} problem needs to be solved) closer and closer to the expert's, feature expectation-wise.\\
 
-Cette résolution est très problématique.
+%Cette résolution est très problématique.
 En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la littérature.
@@
 Comme cela est illustré dans \cite{klein2011batch}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état, ne permettant pas la résultion du problème direct en se basant uniquement sur les  données de l'expert.\\
+%FIXME:parler de boularias ici
+Solving the forward problem can be very tricky ; indeed a whole body of literature is dedicated to this problem.
+Specifically, one needs more information to solve the forward problem than to just compute the feature expectation of the expert as our algorithm does.
+As illustrated in \cite{klein2011batch}, \citep{abbeel2004apprenticeship}'s algorithm is unable to solve the inverted pendulum problem with just samlpes from the expert as those cover too small a part of the state space.
+Supplementary randomly drawn samples are needed.
+Even a recent contribution like \cite{boularias2011relative} still needs samples that cover a large portion of the state space although it does not explicitely require the ability to solve the forward problem.\\
 
-Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
+%Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
 C'est à dire que l'action choisie par l'expert doit avoir une qualité strictement supérieure à n'importe quelle autre action.
@@
 
-Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
+A recurring argument is to see the expert not only as optimal but as optimal by a certain margin.
+The action chosen by the expert should have a quality strictly greater than any other action.
+Structured planning as done in \cite{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} allow some influence on the margin.\\
+
+%Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
 Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}, notre fonction de risque étant similaire à la leur.
@@
 
-Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
+Specifically, our work shows similarities with \cite{ratliff2006maximum}.
+The authors cast the \emph{IRL} problem as a maximum margin classification problem ; our risk function is similar to theirs.
+The method they use can be seen as an iterative algorithm in the tracks of \cite{abbeel2004apprenticeship}.
+From a higher abstraction level, the classification takes place in the space of trajectories : a policy (the label) is associated with a \emph{MDP} (input to classify).
+The empirical demonstration uses a deterministic setting, as they use $A^*$ to solve the \emph{MDP}.
+Although the proposed algorithm is described in a more general setting, it seems unpractical outside the deterministic case.
+
+%Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
 Le principe de leur approche est exactement de résoudre le risque (Eq.
@@
 Toutefois, ils ne prennent pas en comptent la structure du MDP, ce à quoi nous remédions avec notre contribution (voir \cite{melo2010learning} sur l'importance de prendre en compte cette structure).
-Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
+%Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
 Cette fonction de score $s: X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e.
@@
 \eqref{structure.eqn}).\\
+Finally, \citep{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem.
+Both our approach and theirs minimize the same risk function, described Eq.
+\eqref{Jdef.eqn}, but they do not use the structure of the \emph{MDP} as we do (see \cite{melo2010learning} about the importance of taking this structure into account).\\
 
-Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
+%Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
 Il existe d'autres approches que nous ne passons pas en revue de manière exhaustive, mais à notre connaissance elles nécessitent toutes de résoudre le problème direct.
+
+Our approach is in a way between those two contributions as we propose to introduce the structure of the \emph{MDP} (this is what the first contribution do) in the same abstraction level as the second approach.
+Other techniques exist, but to the best of our knowledge none of them can work only from samples drawn by the expert.
+
 %%%%%

### 2012-02-22 klein2012/main.tex
voila

--- 2012-02-21 klein2012/main.tex
@@
 %La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$  telles que
-The reward function $R$ is the unknown of the problem. We can suppose that it is possible to dispose of $p$ features $\psi_{1\leq i \leq p}$, which are functions from $S$ to $\mathbb{R}$, such  that:
+The reward function $R$ is the unknown of the problem.
+We can suppose that it is possible to dispose of $p$ features via $\psi: S\rightarrow \mathbb{R}^p$, which is a functions from $S$ to $\mathbb{R}^p$ and each of the components of $\psi$: $\{\psi_i\}_{1\leq i \leq p}$ is one feature, such that:
 \begin{equation}
@@
 g_{\theta} &=& {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right),\\
-a^*_i &=& \arg\max_{a\in A}(\theta ^T \mu^E(s_i,a) + l(s_i,a)),\\
+a^*_i &\in& \arg\max_{a\in A}(\theta ^T \mu^E(s_i,a) + l(s_i,a)),\\
 g_{\theta} &\in& \partial_\theta J_N,\\
-\theta_{t+1} &\leftarrow& \theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2}.
+\theta_{0}&=&0,\theta_{t+1}=\theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2},\forall t\in\mathbb{N}^*,\\
+\sum_{t\geq0}\alpha_t&>&+\infty,\text{ }\sum_{t\geq0}\alpha_t^2<+\infty.
 \end{eqnarray}
@@
 In a more realistic case, we can use a simple Monte-Carlo estimation if we dispose of a simulator and if it is possible to ask the expert what action has to be chosen in an important number of arbitrary states.The estimation of $\mu^E(s,a)$ (definition Eq.
-\eqref{Qmu.eqn}) is done over $N$ trajectories of length $\{2k_i+1\}_{1\leq i \leq N}$, $\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i \leq N}$, by the following formula: \begin{equation}
+\eqref{Qmu.eqn}) is done over $N$ trajectories of length $\{2k_i+1\}_{1\leq i \leq N}$, $\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i \leq N}$, by the following formula:
+\begin{equation}
 \hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k_i}_{t=0}\gamma^t\psi(s^i_t),
@@
 where $s^i_0=s, a^i_0=a$ and $s^i_{t+1}$ is distributed according to $P_{s^i_t\pi_E(s^i_t)}$.
-
 %Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
@@
 \section{Analysis}\label{proof.sec}
-In this section we are going to prove that our algorithm (see \eqref{algorithm}) is convergent which means that $\tilde{\theta}=\lim_{t\rightarrow+\infty}\theta_t$ is well defined and that $\inf_{\theta\in\mathbb{R}^p}J_N(\theta)=J_N(\tilde{\theta})$ which means that the algorithm reaches the goal of minimizing $J_N$.
-Moreover we will determine the quality of our algorithm by giving an upper bound to the term $E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]$ where $E_{s\sim\nu}$ is the expectation over the stationary probability distribution of states $\nu$ and $\pi_A$ is the policy of the apprentice.
+In this section we are going to prove that our algorithm (see \eqref{algorithm}) is convergent, when we suppose the existence of a minimum of $J_N$,  which means that $\tilde{\theta}=\lim_{t\rightarrow+\infty}\theta_t$
+is well defined and that $\inf_{\theta\in\mathbb{R}^p}J_N(\theta)=J_N(\tilde{\theta})$ which means that the algorithm reaches the goal of minimizing $J_N$.
+Moreover we will determine the quality of our algorithm by giving an upper bound to the term $E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]$ where $E_{s\sim\nu}$ is the expectation
+over the stationary probability distribution of states $\nu$ and $\pi_A$ is the policy of the apprentice when a specific condition is fulfilled.
 \subsection{Convergence of the sub-gradient algorithm}
-Let $N\in\mathbb{N}^*$ and $A$ a finite set of cardinal $\card{A}$.
-Let $\{f_{i,a}\}_{1\leq i \leq N,a\in A}$ and $\{l_{i,a}\}_{1\leq i \leq N,a\in A}$ be finite sets of cardinal $N\card{A}$ which respectively each element $f_{i,a}\in\mathbb{R}^p$ and $l_{i,a}\in\mathbb{R_+}$  where $p\in\mathbb{N}^*$.
-$\mathbb{R}^p$ is an euclidian space with the scalar product $<.,.>$ and the associate euclidian norm $\|\|_2$.
-We define then the function $f_N$ from $\mathbb{R}^p$ to $\mathbb{R}$ such that $\forall \theta\in\mathbb{R}^p, f_N(\theta)=\frac{1}{N}\sum_{i=1}^{i=N}\max_{a\in A}(\theta^Tf_{i,a}+l_{i,a})$.
-It is clear, as $A$ is finite, that $f_N$ is convex, proper, closed, continuous and with finite values.
-Moreover one of its sub-gradient in $\theta\in\mathbb{R}^p$ is $g_{\theta}=\frac{1}{N}\sum_{i=1}^{i=N}f_{i,a_i^*}$ where $a_i^*\in\argmax_{a \in A}(\theta^Tf_{i,a}+l_{i,a})$.
-We also have $g_{\theta}\in\{\frac{1}{N}\sum_{i=1}^{i=N}f_{i,a}\}_{1\leq i \leq N,a\in A}=G$, $G$ is a finite set .
-Let note $g_{min}$ one element of $\argmin_{G\backslash{0}}(\frac{1}{N}\|\sum_{i=1}^{i=N}f_{i,a}\|_2)$, and
-$\overline{f}_N=\inf_{\theta\in\mathbb{R}^p}f_N(\theta)$. Finally, we suppose $f_N(0)\geq \overline{f}_N\geq0$. 
+$\mathbb{R}^p$ is an euclidian space with the canonical scalar product $<.,.>$ and the euclidian norm $\|\|_2$.
+Let $N\in\mathbb{N}^*$ and $A$ the finite set of actions of cardinal $\card{A}$.
+We dispose of the estimations  $\{\hat{\mu}^E_{ia}\}_{1\leq i \leq N,a\in A}$ where $\forall 1\leq i \leq N, \forall a\in A,\hat{\mu}^E_{ia}\in\mathbb{R}^p$.
+We define the loss function such that: $\forall 1\leq i \leq N,\forall a\in A\backslash{a_i}, l(s_i,a)>l(s_i,a_i)\geq0$.
+We have already defined the empirical risk function which is a function from $\mathbb{R}^p$ to $\mathbb{R}$:
+$\forall \theta\in\mathbb{R}^p, J_N(\theta)= {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \hat{\mu}^E_{s_ia} + l(s_i,a)) - \theta ^T \hat{\mu}^E_{s_ia_i} \right)$.
+It is clear, as $A$ is finite, that it is a convex, proper, closed and continuous function
+with finite values.
+Moreover one sub-gradient of $J_N$ in $\theta\in\mathbb{R}^p$ is $g_{\theta}={1\over N}\sum_{i=1}^N\left(\hat{\mu}^E_{s_ia^*_i} - \hat{\mu}^E_{s_ia_i}\right)$, where $a^*_i\in\arg\max_{a\in A}(\theta ^T \hat{\mu}^E_{s_ia} + l(s_i,a))$.
+We can then use the result from the article \cite{correa1993convergence} section 5 which allows to say, with our notations, that the sequence $\theta_0=0,\forall t\in \mathbb{N}^*,\theta_{t+1}=\theta_t-\alpha_t\frac{g_{\theta_t}}{\|g_{\theta_t}\|_2}$ is convergent
+if there exists a minimum to $J_N$, and that $\liminf_{t\rightarrow +\infty}J_N(\theta_t)=\inf_{\theta\in\mathbb{R}^p}J_N(\theta)$ under the conditions: $\sum_{t\geq0}\alpha_t>+\infty,\text{ }\sum_{t\geq0}\alpha_t^2<+\infty$.
+For a discussion on the existence
+of the minimum, one can report to the article \cite{kaufmann1988existence}.
+Finally, one can show that if there exists $\theta\in\mathbb{R}^p$ such that $\forall 1\leq i \leq N, \theta^T\mu^E_{s_ia_i}>\max_{a\in A \backslash a_i}\theta^T\mu^E_{s_i,a}$, then there exists a minimum of $J_N$ and that all the
+minimums: $\{\tilde{\theta}: \inf_{\theta\in\mathbb{R}^p}J_N(\theta)=J_N(\tilde{\theta})\}$ verify $\forall 1\leq i \leq N, \tilde{\theta}^T\mu^E_{s_ia_i}>\max_{a\in A \backslash a_i}\tilde{\theta}^T\mu^E_{s_i,a}$.
 \subsection{Quality of the solution found by the sub-gradient algorithm}
+Here we are going to bound the term $E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]$.
+First of all let us introduce the estimation errors $\{\epsilon^{\mu}_{s_ia_i}=\hat{\mu}^E_{s_ia_i}-\mu^E_{s_ia_i}\}_{1\leq i \leq N,a\in A}$ and the
+approximation error $\epsilon^R$ which results from the fact that the unknown and bounded reward function $R\in\mathbb{R}^S$ may not be a linear combination of the features $\psi_{1\leq i \leq p}$.
+Formally, let $T$ be the tuple $\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma\}$ which is
+an MDP without reward and $T(\tilde{R})=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,\tilde{R}\}$ where $\tilde{R}\in\mathbb{R}^S$.
+We note $C_R$ the reward-equivalence class of $R$: $C_R=\{\tilde{R}\in\mathbb{R}^S: \Pi(T(R))=\Pi(T(\tilde{R}))\}$ where $\Pi(T(\tilde{R}))$ is the set of optimal deterministic policies
+of $T(\tilde{R})$.
+We can now give a formal definition of $\epsilon^R$: $\epsilon^R=\|R^*-\theta^{*T}\psi\|_\infty=\max_{s\in S}|R^*(s)-\theta^{*T}\psi(s)|$ where $(R^*,\theta^*)\in\argmin_{\tilde{R}\in C_R, \theta\in\mathbb{R}^p}{\|\theta^T\psi-R\|_2}$.
+Now we suppose that our sub-gradient converges to $\tilde{\theta}$ and that $g_{\tilde{\theta}}={1\over N}\sum_{i=1}^N\left(\hat{\mu}^E_{s_ia^*_i} - \hat{\mu}^E_{s_ia_i}\right)=0$, where $a^*_i\in\arg\max_{a\in A}(\tilde{\theta}^T \hat{\mu}^E_{s_ia} + l(s_i,a))$.
+This case happen for instance when there exists $\theta\in\mathbb{R}^p$ such that $\forall 1\leq i \leq N, \theta^T\mu^E_{s_ia_i}>\max_{a\in A \backslash a_i}\theta^T\mu^E_{s_i,a}$.
+The apprentice policy $\pi_A$ is an optimal deterministic policy
+for the MDP $T(\tilde{\theta}\psi)$ and we can show that: $\forall 1\leq i\leq N, \pi_A(s_i)=a_i^*$.
+So we have: ${1\over N}\sum_{i=1}^N\left(\hat{\mu}^E_{s_i\pi_A(s_i)} - \hat{\mu}^E_{s_i\pi_E(s_i)}\right)=0$ which implies: ${1\over N}\sum_{i=1}^N(<\mu^E_{s_i\pi_A(s_i)},\theta^*>-<\mu^E_{s_i\pi_E(s_i)},\theta^*>+<\epsilon^{\mu}_{s_i\pi_A(s_i)}-\epsilon^{\mu}_{s_i\pi_E(s_i)},\theta^*>)=0$.
+Here $Q^{\pi_E}$ and $V^{\pi_E}$ are defined in function of $R^*$ and not in function of $R$, we should use the notations $Q^{\pi_E}_{R^*}$ and $V^{\pi_E}_{R^*}$.
+Let note the action-value errors $\{\epsilon^Q_{s_ia_i}=Q^{\pi_E}(s_i,a_i)-<\theta^*,\mu^E_{s_ia_i}>\}_{1\leq i \leq N,a\in A}$, we have $\forall 1\leq i\leq N, |\epsilon^Q_{s_ia_i}|=|E^{\pi_E}_{s_ia_i}[\sum_{t\geq0}\gamma^t(R^*(s_t)-<\theta^*,\psi(s_t)>)]|\leq\frac{\epsilon^R}{1-\gamma}$.
+So we have:
+${1\over N}\sum_{i=1}^N(Q^{\pi_E}(s_i,\pi_E(s_i))-\Q^{\pi_E}(s_i,\pi_A(s_i))+<\epsilon^{\mu}_{s_i\pi_E(s_i)}-\epsilon^{\mu}_{s_i\pi_A(s_i)},\theta^*>-\epsilon^{Q}_{s_i\pi_E(s_i)}+\epsilon^{Q}_{s_i\pi_A(s_i)})=0$.
+Finally, let us define $\epsilon^V$, $V$ as Vapnick, the Vapnick-error term:
+$\epsilon^V=E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]-[{1\over N}\sum_{i=1}^NQ^{\pi_E}(s_i,\pi_E(s_i))-\Q^{\pi_E}(s_i,\pi_A(s_i))]$.
+Here we suppose that the samples$\{s_i,a_i\}_{1\leq i \leq N,a\in A}$ are i.i.d and distributed over $\nu$.
+We know that the $\pi_A$ we find depends only on $\tilde{\theta}$ which depends on the samples
+but as $\tilde{\theta}$ is a linear combination of $p$ parameters it implies that the space of possible $\pi_A$ has a VC-dimension $h$ such that: $h\leq p$.
+So, we can use a Vapnick-bound for $\epsilon^V$ with probability $1-\eta$ because $\forall s\in S, 0\leq|Q^{\pi_E}(s,\pi_E(s))-\Q^{\pi_E}(s,\pi_A(s))|\leq\frac{R^*_{max}}{1-\gamma}$, where $R^*_{max}=\max_{s\in S}R^*(s)$.
+$|\epsilon_V|\leq\frac{R^*_{max}}{1-\gamma}\epsilon(N)$ where $\epsilon(N)=\sqrt{\frac{h}{N}[\ln(\frac{2N}{h})+1]-\frac{1}{N}\ln(\frac{\eta}{4})}$.
+Finally by Cauchy-Schwartz, we have the final upper-bound with probability $1-\eta$:
+$E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]\leq\frac{\epsilon^R}{1-\gamma}+2\|\theta^*\|_2\|\epsilon^{\mu}\|_{\infty}+\epsilon(N)$, where $\|\epsilon^{\mu}\|_{\infty}=\max_{1\leq i\leq N,a\in A}\|\epsilon^{\mu}_{s_ia}\|_2$.
 \section{Related work}

### 2012-02-22 klein2012/main.tex

@@
 In this paper we tackle the problem of apprenticeship learning in the Inverse Reinforcement Learning (\emph{IRL}) framework.
-An expert fulfils a task to be imitated by an artificial agent.
-The IRL framework assumes that the expert is acting optimally with respect to a reward function ; the goal is to guess this function from trajectories drawn by the expert.
-Existing IRL algorithms need one or more of the following items : complete trajectories from the expert, a generative model of the environment for Monte-Carlo estimations, the knowledge of the transition probabilities, the ability to repeatedly solve the forward problem (the Reinforcement Learning (\emph{RL}) problem), the access to the expert's strategy anywhere in the state space, or at the very least random samples covering a large part of it.
+An expert fulfils a task to be imitated by an artificial agent. The IRL framework assumes that the expert is acting optimally with respect to a reward function;
+the goal is to guess this function from trajectories drawn by the expert. Existing IRL algorithms need one or more of the following items: complete trajectories from the expert,
+a generative model of the environment for Monte-Carlo estimations, the knowledge of the transition probabilities, the ability to repeatedly solve the forward problem (the Reinforcement Learning (\emph{RL}) problem),
+the access to the expert's strategy anywhere in the state space, or at the very least random samples covering a large part of it.
 Our contribution consists in a new IRL algorithm lifting all these constraints.
-Using a supervised technique in which the structure of the underlying Markov Decision Process (\emph{MDP}) is implicitely injected, we give rise to a subgradient descent-based algorithm whose computational and sample complexities are low.
+Using a supervised technique in which the structure of the underlying Markov Decision Process (\emph{MDP}) is implicitely injected, we give rise to a subgradient descent-based algorithm whose computational
+and sample complexities are low.
 The sample base from the expert is the only data we need.
@@
 So, we want to introduce a risk function which should penalize the cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases the score function $q$ does not allow to explain the choice $a_i$ of the expert (see equation \eqref{greedy.eqn}).
-Moreover, we introduce a loss function $l: S \times A \rightarrow \mathbb{R}_+$; this function allows the user to put some \emph{a priori} knowledge in the system because the user will choose the gap wanted in the score function between the action chosen by the expert and the second best possible action as we can see in the following equation: $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
-Introducing a loss function is quite common in the literature (see Sec.
-\ref{biblio.sec}).
-We introduce then the following empirical risk function:
+Moreover, we introduce a loss function $l: S \times A \rightarrow \mathbb{R}_+$; this function allows the user to put some \emph{a priori} knowledge in the system because the user will choose the gap wanted in the score function between the action chosen by the expert and the second best possible action as we can see in the following equation:
+$\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$. Introducing a loss function is quite common in the literature (see Sec. \ref{biblio.sec}). We introduce then the following empirical risk function:
 \begin{equation}
@@
 The expert follows the same rule and if we note $\mu^E$ the feature expectation of the expert, we can do the hypothesis that the score function $q$ that we are searching for can be written as follows: $q(s,a) = \theta^T\mu^E(s,a)$.
-Estimating the term $\mu^E$ is the key of our algorithm and we propose some solution in Sec.
-\ref{calculmu.sec}.
-We can now give the expression of our empirical risk function as a function of the parameter $\theta$ which defines the reward function (Eq.
-\eqref{hatRdef.eqn}):
+Estimating the term $\mu^E$ is the key of our algorithm and we propose some solution in Sec. \ref{calculmu.sec}.
+We can now give the expression of our empirical risk function as a function of the parameter $\theta$ which defines the reward function (Eq. \eqref{hatRdef.eqn}):
 \begin{equation}
@@
 We can then use a sub-gradient descent algorithms, which is a generalization of a gradient descent algorithm, in order to minimize the risk function.
-A sub-gradient of a convex function  $f:\mathbb{R}^p\rightarrow \mathbb{R}$ in $x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$, the set of all sub-gradients in $x$ is $\partial_xf$.
+A sub-gradient of a convex function
+$f:\mathbb{R}^p\rightarrow \mathbb{R}$ in $x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$, the set of all sub-gradients in $x$ is $\partial_xf$.
 The unique sub-gradient in $x$ of a differentiable convex function $f$ is the gradient $\nabla_xf$.
-As $A$ is finite, the convex and piecewise differentiable function $\max_{a\in A}[f(\cdot,a)]$, from $\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from $\mathbb{R}^p$ to $\mathbb{R}$,  has one of its sub-gradient in $x\in\mathbb{R}^p$ equals to $\nabla_xf(x,a^*)$ with $a^*\in\arg\max_{a\in A}f(x,a)$.
+As $A$ is finite, the convex and piecewise differentiable function $\max_{a\in A}[f(\cdot,a)]$, from $\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from $\mathbb{R}^p$ to $\mathbb{R}$,
+has one of its sub-gradient in $x\in\mathbb{R}^p$ equals to $\nabla_xf(x,a^*)$ with $a^*\in\arg\max_{a\in A}f(x,a)$.
 Applying these rules, we have a sub-gradient of our risk function and an update rule for our sub-gradient algorithm:
@@
 That implies to infer the feature expectation of the expert on states where he has been, but on actions he has not taken.
-
 %Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve:
@@
 In the ideal case where a model of the environnement is known (\textit{id est} the transition probabilities are given), the exact calculus of $\mu^E$ can be executed thanks to dynamic programmation algorithms.
-In a more realistic case, we can use a simple Monte-Carlo estimation if we dispose of a simulator and if it is possible to ask the expert what action has to be chosen in an important number of arbitrary states.The estimation of $\mu^E(s,a)$ (definition Eq.
+In a more realistic case, we can use a simple Monte-Carlo estimation if we dispose of a simulator and if it is possible to ask the expert what action has to be chosen in an important number of arbitrary states.
+The estimation of $\mu^E(s,a)$ (definition Eq.
 \eqref{Qmu.eqn}) is done over $N$ trajectories of length $\{2k_i+1\}_{1\leq i \leq N}$, $\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i \leq N}$, by the following formula:
@@
 %%%
-Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté.
-Il s'agit d'un problème classique en apprentissage par renforcement.
-Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
-Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.\\
+%Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté. Il s'agit d'un problème classique en apprentissage par renforcement.
+%Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
+%Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
+%Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle)
+%la seconde étant sa vitesse angulaire.\\
+The inverted pendulum problem consists in maintaining in equilibrium a pendulum with one degree of freedom. This is a classical problem in the reinforcement learning field.
+The policy of our expert is determined by an RL-algortihm called LSPI which is explained in the article \cite{lagoudakis2003least}. The policy of the apprentice, which is the policy
+optimizing the reward function of our algorithm given Figure \ref{onlyFig.fig} (a), manages systematically  
+%La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
+%Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
 
-La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
-Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
+%Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases,
+%de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
+%Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie
+%(correspondant à la position verticale et son voisinage proche).\\
 
-Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
-Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale et son voisinage proche).\\
-
- Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
-En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
-Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel moyen est utilisée en association à celle de l'expert.
-Cette politique est obtenue par optimisation d'une récompense sur le PDM dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
-Comme notre algorithme ne nécessite pas le calcul d'autre attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
+%Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
+%En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser
+%les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}. Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel
+%moyen est utilisée en association à celle de l'expert. Cette politique est obtenue par optimisation d'une récompense sur le PDM dans lequel évolue l'expert ; cette résolution du problème
+%direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état. Comme notre algorithme ne nécessite pas le calcul d'autre
+%attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
 %%%
@@
 %%%
-
-\begin{figure}[tb]
-%
+\begin{figure}[ht]
+\vskip 0.2in
+\begin{center}
 \begin{minipage}[b]{.5\linewidth}
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_true_R.pdf}}
+  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_true_R.pdf}}
   \centerline{(a) Récompense de l'expert}%\medskip
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_lafem_R.pdf}}
+  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_lafem_R.pdf}}
   \centerline{(b) Récompense trouvée}%\medskip
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vexpert.pdf}}
+  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vexpert.pdf}}
   \centerline{(c) Fonction de valeur de l'expert}%\medskip
@@
   \centering
-  \centerline{\includegraphics[width=1.2\linewidth]{LAFEM_Exp3_Vagent.pdf}}
+  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vagent.pdf}}
   \centerline{(d) Fonction de valeur de l'agent}%\medskip
@@
 %
-\vspace{-5pt}
+\end{center}
+\vskip -0.2in
 \end{figure}
@@
 Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
-La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon.
-Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations est de $T=20$.
-Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange de gaussiennes défini dans \cite{lagoudakis2003least}.
-Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
+La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon. Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations
+est de $T=20$. Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange
+de gaussiennes défini dans \cite{lagoudakis2003least}. Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
 
-Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)).
-Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
+Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs
+sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)). Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense
+pouvant expliquer une politique.\\
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est notre critère de réussite.\\
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes,
+ce qui est notre critère de réussite.\\
 
-Deux éléments méritent d'être notés.
-Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace: celle dans laquelle le pendule est proche de la verticale.
-En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
-Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.
-Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.
-Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig.
-\ref{onlyFig.fig} Cela n'a aucune importance, une dilatation laissant la politique optimale invariante.
+Deux éléments méritent d'être notés. Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
+Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace:
+celle dans laquelle le pendule est proche de la verticale. En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
+Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer
+la fonction de récompense utilisée sans l'atteindre. Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir
+des performances similaires à celles de l'expert. Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig. \ref{onlyFig.fig} Cela n'a aucune importance, une dilatation
+laissant la politique optimale invariante.
+
 \section{Conclusion}
 
-L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse.
-En rendant la résolution du problème inverse indépendante de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
+L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse. En rendant la résolution du problème inverse indépendante
+de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
 
@@
 
-L'estimation de $mu_E$ est centrale dans notre approche.
-Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients que les algorithmes d'estimation d'une fonction de valeur.
-L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée.
-Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.
+L'estimation de $\mu_E$ est centrale dans notre approche. Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients
+que les algorithmes d'estimation d'une fonction de valeur. L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen
+d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée. Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.
 %

### 2012-02-22 klein2012/main.tex

@@
 The inverted pendulum problem consists in maintaining in equilibrium a pendulum with one degree of freedom. This is a classical problem in the reinforcement learning field.
-The policy of our expert is determined by an RL-algortihm called LSPI which is explained in the article \cite{lagoudakis2003least}. The policy of the apprentice, which is the policy
-optimizing the reward function of our algorithm given Figure \ref{onlyFig.fig} (a), manages systematically  
+The policy of our expert is determined by an RL-algortihm called LSPI which is explained in the article \cite{lagoudakis2003least}. The different settings used, from the value of the pendulum mass
+to the choice of the features, are conform to the settings of \cite{lagoudakis2003least}. In particular, the features are a gaussian network covering the two dimensional space: the first dimension
+is the angular position of the pendulum (where the verticale corresponds to the angle zero) and the second one is the angular speed.\\   
 %La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
 %Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
-
+The policy of the expert, which is the policy optimizing the reward function given Figure \ref{onlyFig.fig} (a), manages systematically the maintain the pendulum during 3000 time-steps.
+The time-steps having a value of 0.1s, this means that we are interrupting the expert after 5 minutes without the fall of the pendulum.\\
 %Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases,
@@
 %(correspondant à la position verticale et son voisinage proche).\\
-
-%Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
+The aim of this experience is to show that our algorithm is able, only from the samples of the transitions given by the expert and the definition of the features, to find a reward function
+which will allow an agent trained with this reward function to maintain the pendulum in equilibrium during 5 minutes without fall. This is not a trivial problem because the state space is
+continuous and that the data from the expert are only covering a party of it.\\
+%Ce dernier point est d'ailleurs la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
 %En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser
@@
 %%%
+This last point is the reason why the algorithm in the article \citet{abbeel2004apprenticeship} does not manage to resolve this problem only with the samples of the transitions of the expert.
 \subsection{Illustrations}

### 2012-03-27 klein2012/ICML2012_EKBPMGOP.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use the following line _only_ if you're still using LaTeX 2.09.
%\documentstyle[icml2012,epsf,natbib]{article}
% If you rely on Latex2e packages, like most moden people use this:
\documentclass{article}

% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{url}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{icml2012}
\mathtoolsset{showonlyrefs=true}

\newtheorem{definition}{Définition}
\newtheorem{lemme}{Lemma}
\newtheorem{theorem}{Théorème}
\newtheorem{hypo}{Hypothèse}
\newtheorem{cor}{Corollaire}
\newtheorem{prop}{Proposition}
\newtheorem{exemple}{Exemple}
\newtheorem{remarques}{Remarques}
\newtheorem{remarque}{Remarque}
\newtheorem{assumption}{Assumption}
\newtheorem{problem}{Problème}
\newtheorem{algo}{Algorithme}
\newtheorem{preuve}{Démonstration}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\argmin}{\operatorname*{argmin}}
\newcommand{\minp}{\operatorname*{min_+}}
\newcommand{\dom}{\operatorname*{dom}}
\newcommand{\Ker}{\operatorname*{Ker}}
\newcommand{\trace}{\operatorname*{trace}}
\newcommand{\cov}{\operatorname{cov}}
\newcommand{\epi}{\operatorname{epi}}
\newcommand{\card}{\operatorname*{Card}}
\newcommand{\conv}{\operatorname*{Conv}}
\newcommand{\vect}{\operatorname*{Vect}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\erf}{\operatorname{erf}}
\newcommand{\bound}{\operatorname*{bound}}
\newcommand{\vpi}{\operatorname{VPI}}
\newcommand{\gn}{\operatorname{Gain}}
\newcommand{\p}{\operatorname{Pr}}
\newcommand{\mlp}{\operatorname{MLP}}
\newcommand*\tto[2]{\smash{\mathop{\longrightarrow}\limits_{#1}^{#2}}}
\newcommand*\ntto[2]{\smash{\mathop{\nrightarrow}\limits_{#1}^{#2}}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bd}{\mathbf{d}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\af}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bh}{\mathbf{h}}
\newcommand{\bi}{\mathbf{i}}
\newcommand{\bj}{\mathbf{j}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bl}{\mathbf{l}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bo}{\mathbf{o}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bt}{\mathbf{t}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\ma}{\mathbf{A}}
\newcommand{\mb}{\mathbf{B}}
\newcommand{\mc}{\mathbf{C}}
\newcommand{\md}{\mathbf{D}}
\newcommand{\me}{\mathbf{E}}
\newcommand{\mf}{\mathbf{F}}
\newcommand{\mg}{\mathbf{G}}
\newcommand{\mh}{\mathbf{H}}
\newcommand{\mi}{\mathbf{I}}
\newcommand{\mj}{\mathbf{J}}
\newcommand{\mk}{\mathbf{K}}
\newcommand{\ml}{\mathbf{L}}
\newcommand{\mm}{\mathbf{M}}
\newcommand{\mn}{\mathbf{N}}
\newcommand{\mo}{\mathbf{O}}
\newcommand{\Mp}{\mathbf{P}}
\newcommand{\mq}{\mathbf{Q}}
\newcommand{\mr}{\mathbf{R}}
\newcommand{\ms}{\mathbf{S}}
\newcommand{\mt}{\mathbf{T}}
\newcommand{\Mu}{\mathbf{U}}
\newcommand{\mv}{\mathbf{V}}
\newcommand{\mw}{\mathbf{W}}
\newcommand{\mx}{\mathbf{X}}
\newcommand{\my}{\mathbf{Y}}
\newcommand{\mz}{\mathbf{Z}}
\newcommand{\tphi}{\tilde{\Phi}}
\newcommand{\espace}{\text{ }}

\newcommand{\x}{\mathbf{x}}
\newcommand{\s}{\mathbf{s}}
\newcommand{\n}{\mathbf{n}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\rr}{\mathbf{r}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Titre court si le titre fait plus de 40 caractères
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\shorttitle{Classification structurée pour l'ARI}
%
%\shortouvrage{CAp 2012}

% Titre, auteur, pas de date

%\title{Classification structurée pour l'apprentissage par renforcement inverse}
%
%\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
%
%\institute{
%Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
%\and
%Equipe ABC,
%LORIA, France
%\and
%UMI 2958
%GeorgiaTech-CNRS, France
%}
\icmltitlerunning{Structured Classification for Inverse
Reinforcement Learning}

\begin{document}

\twocolumn[ \icmltitle{Structured Classification for Inverse
Reinforcement Learning}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2012
% package.
\icmlauthor{Your Name}{email@yourdomain.edu}
\icmladdress{Your Fantastic Institute,
            314159 Pi St., Palo Alto, CA 94306 USA}
\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
\icmladdress{Their Fantastic Institute,
            27182 Exp St., Toronto, ON M6H 2T1 CANADA}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{inverse reinforcement learning, structured multi-class
classification}

\vskip 0.3in
]

\begin{abstract}
This paper addresses the Inverse Reinforcement Learning (IRL)
problem which is a particular case of learning from demonstrations.
The IRL framework assumes that an expert, demonstrating a task, is
acting optimally with respect to an unknown reward function to be
discovered. Unlike most of existing IRL algorithms, the proposed
approach doesn't require any of the following: complete trajectories
from the expert, a generative model of the environment, the
knowledge of the transition probabilities, the ability to repeatedly
solve the forward Reinforcement Learning (RL) problem, the expert's
policy anywhere in the state space. Using a classification approach
in which the structure of the underlying Markov Decision Process
(MDP) is implicitly injected, we end-up with an efficient
subgradient descent-based algorithm.
% whose computational and sample
%complexities are low.
In addition, only a small amount of expert
demonstrations (not even in the form of trajectories but simple
transitions) is required.
%A formal analysis of the algorithm as well as empirical
%demonstrations on two classic but challenging problems are provided.
\end{abstract}
%%%%%
\section{Introduction}
%%%%%
%Dans cette contribution, nous nous plaçons dans le contexte de l'apprentissage par démonstration. Dans ce cadre, un agent artificiel apprend à reproduire une tâche grâce à l'observation d'un expert réalisant cette tâche de manière optimale. Pour trouver une solution à ce problème, nous nous proposons d'utiliser le paradigme de l'apprentissage par renforcement inverse. On suppose alors que l'expert agit de manière à être récompensé pour son comportement relativement à une fonction de récompense inconnue. Le but de l'agent est alors d'inférer cette fonction des démonstrations données par l'expert pour ensuite optimiser son comportement relativement à cette fonction lui aussi.\\

This contribution addresses the problem of learning by watching or
learning from demonstrations. In this context, an artificial agent
learns to perform a task by observing an expert demonstrating the
task. In order to find a solution to this problem, the Inverse
Reinforcement Learning (IRL) paradigm has been
proposed~\cite{russell1998learning}. The fundamental idea of IRL is
that the expert acts so as to be rewarded by an unknown reward
function. The aim of the IRL agent is then to infer the unknown
reward function from the demonstrations given by the expert, the
reward being considered as the most compact representation of the
task to be transferred to the agent. This way, an optimal behavior
can be obtained in any situation by optimizing the action selection
process according to the reward function. Unlike supervised
learning, the agent may not purely imitate the expert's actions but
will try to accomplish the same task optimally.
%
%
%La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}. Plusieurs directions ont depuis été explorées à partir de deux articles fondateurs \cite{ng2000algorithms} et \cite{abbeel2004apprenticeship}.\\
%Dans cette littérature (voir \cite{neu2009training} pour un bon survol), plusieurs verrous sont identifiés mais souvent ignorés pour proposer des algorithmes de résolution. Particulièrement, on suppose souvent connue la dynamique de l'environnement dans lequel évoluent l'expert et l'agent ou bien la mise à disposition de cet environnement (ou d'un simulateur) pour y tester les effets d'une politique quelconque. De plus, ces approches supposent de résoudre le problème direct (optimiser une politique pour une récompense donnée) un grand nombre de fois pour résoudre le problème inverse (trouver une récompense expliquant la politique). De plus, certaines de ces approches donnent en sortie une politique généralisant celle de l'expert plutôt qu'une récompense l'expliquant. Ces contraintes sont souvent incompatibles avec la mise en oeuvre des algorithmes proposés sur des cas réels.\\
%
%
The literature on the subject is quite recent and finds its origin
in \cite{russell1998learning}. Different approaches have been
explored since the first articles on the matter
\cite{ng2000algorithms,abbeel2004apprenticeship}.

In this literature (see \cite{neu2009training} for a survey),
several bottlenecks are identified but often ignored when developing
algorithms. Specifically, the dynamic of the environment where the
expert and the agent are evolving is assumed to be known, or an
environment model (via a simulator) must be available so as to test
the effects of a policy. Moreover, most of existing approaches
require solving the direct reinforcement learning problem (find an
optimal policy knowing a given reward) several times. Finally, in
many cases, the output of the algorithm is a policy but not a unique
reward function which makes task transfer hard to realize. These
constraints can lead to the development of algorithms unapplicable
to real problems.
%Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert. Nous proposons une preuve de convergence de cet algorithme et fournissons une expérimentation sur le problème du pendule inversé. Le choix de cette application se justifie par la difficulté  qu'elle pose pour les méthodes de résolution de la littérature.
%
In this paper, we propose a resolution method of the inverse
reinforcement problem that is free from any other knowledge than the
demonstrations given by the expert. The proposed approach, the
Structured Classification for IRL (SCIRL), relies on a
classification paradigm in which the structure of the environment's
dynamics is implicitly introduced. Sec.~\ref{back.sec}
and~\ref{biblio.sec} provide the necessary background and review a
part of the state of the art. Sec.~\ref{sec:scirl} introduces our
contribution and Sec.~\ref{proof.sec} analyses the asymptotic
algorithm. Section~\ref{exp.sec} shows results obtained on two
classical benchmarks: inverted pendulum and the highway driving
problems.
%
%A proof of the convergence of the algorithm is provided in
%Section~\ref{proof.sec} and Section~\ref{exp.sec} shows results
%obtained on two classical benchmarks: inverted pendulum and the
%highway driving problems. The choice of the first application is
%justified by the fact that it is hard to solve by the other methods.

%%%%
\section{Background}
%%%%%
\label{back.sec}
%%%
Let's start with some background on RL and IRL and settle the
notations.

\subsection{(Inverse) Reinforcement Learning}
%%%
%Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles. La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$. Confronté à cet état, l'agent doit prendre un décision $a_t\in A$. Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité de transition markovienne $p(s_{t+1}|s_t, a_t)$. Une politique de contrôle déterministe $\pi: S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\

The problem is seen as a sequential decision making problem posed in
the Markov Decision Process (MDP) formalism. In this formalism, a
system to be controlled can reach different configurations
completely described at each discrete time $t$ by a state $s_t \in
S$. Given the knowledge of the state, the agent must take a decision
$a_t\in A$. Then, the system evolves to a next state $s_{t+1} \in S$
with the Markovian probability $p(s_{t+1}|s_t, a_t)$. %(in short
%$p(s'|s,a)$ where $s'=s_{t+1}, s=s_t$ and $a=a_t$).
A policy, which
is a function $\pi: S\rightarrow A$, defines the behavior of the
agent. In this paper all the policies considered are deterministic,
$S$ and $A$ will be finite sets and $\{P_{sa}\}_{s\in S,a\in A}$
will be a set of probability distributions, from $S$ to $[0,1]$ such
that $P_{sa}(s')=p(s'|s, a)$, called transition probabilities.
%
%Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R: S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état. Une des forces de ce cadre réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre. On définit pour ce faire la fonction de valeur $V^\pi$:
%
The RL paradigm aims at solving the problem of finding the optimal
sequence of actions in an MDP. The optimality criterion is based on
the definition of a reward function $R: S \rightarrow \mathbb{R}$
which defines the immediate reward $R(s)$ given to the agent when
the system is in the state $s \in S$. The agent should not learn to
maximize the immediate reward but a criterion taking into account
future rewards. The criterion commonly chosen is the value function
$V^\pi$:
%
\begin{align}
\label{Vdef.eqn} V^\pi(s) & = E^\pi_s[\sum_{t\geq0}\gamma^tR(s_t)]\\
& = R(s) + \gamma\sum_{s'\in S}P_{s\pi(s)}(s') V^\pi(s'),
\end{align}
%où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action:
%
where $\gamma \in [0,1)$ is a discount factor, and $E^\pi_s$ is the
expectation over the distribution of the state sequences
$(s_0,s_1,\dots)$ the agent passes through when executing the policy
$\pi$ starting from $s_0=s$. One can define the action-value
function $Q^\pi$ which adds a degree of freedom in the choice of the
first action:
%
\begin{align}
\label{Qdef.eqn} Q^\pi(s,a) & =
E^\pi_{sa}[\sum_{t\geq0}\gamma^tR(s_t)]\\
& = R(s) + \gamma\sum_{s'\in S}P_{sa}(s') V^\pi(s').
\end{align}
%
%where $E^\pi_{sa}$ is the expectation over the distribution of the
%state sequences $(s_0,s_1,\dots)$ we pass through when we start from
%$s_0=s$, choose the action $a_0=a$ and then follow the policy $\pi$.
%
%Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$. La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton:
%
An optimal policy $\pi^*$ is defined as a policy with maximal value
function, $V^{\pi*}=V^*=\sup_{\pi}V^{\pi}$, which means that
$\forall \pi, \forall s, V^{\pi*}(s) \geq V^\pi(s)$. An optimal
policy $\pi^*$ can be computed easily thanks to the optimal
action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy
mechanism:
%
\begin{equation}
\label{greedy.eqn} \pi^*(s) \in \arg\max_{a\in A} Q^*(s,a).
\end{equation}
%L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov .\\
%
The tuple $M=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,R\}$ is called
a finite MDP. Resolving the RL problem with respect to the criterion
$V^\pi$  means to find $\pi^*$, usually without the full knowledge
of the MDP but only by observing transitions and rewards. In a
finite MDP the existence of an optimal deterministic policy is
guaranteed when the reward function $R$ is bounded.
%
%Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive. Un exemple de ce type de tâche est la conduite d'une voiture. Nous accomplissons ce genre de chose au quotidien sans trop y penser mais nous serions bien en peine de préciser les poids précis que nous attribuons aux différents critères tels que la distance nous séparant de la voiture devant nous, la brutalité avec laquelle nous appuyons sur la pédale de frein lorsqu'un danger se présente et ainsi de suite. Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
%
The optimal behaviour being defined relatively to the reward, it's
observation is mandatory to learn. Sometimes, specifying manually
the reward function can be a hard job while it is natural to an
operator to achieve the desired task. For instance, driving a car is
such a task. One can accomplish this task very easily in his
daily-life but it is very tricky to specify the precise weights
given to each important driving-parameters such as the distance to
keep with the leading car, the speed below which it is unreasonable
to drive and so on. In such cases, inferring the reward function
from observed behavior would be helpful.
%
%C'est la définition de l'apprentissage par renforcement inverse. Le problème est de retrouver la fonction de récompense optimisée par un expert. Usuellement l'expert est considéré comme un agent optimal dans un PDM et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
%
It is the aim of IRL. The expert, demonstrating the task, is thus
considered as an optimal agent in a MDP and the IRL agent has access
to the expert optimal policy $\pi_E$ or to some trajectories drawn
from this policy.
%
%Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}). Particulièrement, tout comportement est optimal vis-à-vis de la récompense uniformément nulle. Il convient donc de contraindre l'espace des solutions.
The problem of finding a reward function from expert demonstrations
in the IRL framework is ill-posed. Indeed, there is not uniqueness
of the reward function for which the expert policy is
optimal~\cite{ng1999policyreward}. Particularly, each policy is
optimal with respect to the uniformly-zero reward function. Some
solutions have to be favored.
%%%
\subsection{Feature expectation}
\label{ConsiderationsTechniques.sec}
%%%
%La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$  telles que
The actual expert reward function $R_E$ is unknown. The problem is
 thus to find a reward function $R_\theta$ for which the policy of
the expert is also optimal. Let's assume that, given a set of $p$
features $\psi: S\rightarrow \mathbb{R}^p$, each components of
$\psi$: $\{\psi_i\}_{1\leq i \leq p}$ being a basis function from
$S$ to $\mathbb{R}$, such a reward can be written:
\begin{equation}
\label{hatRdef.eqn} %\exists \theta |
R_\theta(s) = \theta^T\psi(s) =
\sum_{i=1}^p\theta_i\psi_i(s).
\end{equation}
%Cette contrainte peut cependant être relâchée si nécessaire. Introduire cette expression dans la définition de la fonction de qualité (Eq. \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants:
Using this expression allows rewriting (Eq. \eqref{Qdef.eqn}) as:
\begin{equation}
Q^\pi(s,a)=\theta^T\mu^\pi(s,a),\text{ }\mu^\pi(s,a) =
E^\pi_{sa}[\sum_{t\geq 0}\gamma^t\psi(s_t)]. \label{Qmu.eqn}
\end{equation}
%Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique. On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
%
The term $\mu^\pi$ is called the feature expectation of the policy
$\pi$. We can observe that the chosen parametrization of the reward
function $R_\theta$ and the dynamics resulting from the application
of the policy $\pi$ in the MDP gives rise to an induced
parametrization for $Q^\pi$. To
shorten notations, we may use $\mu^\pi_{sa}$ as $\mu^\pi(s,a)$.\\
%
%On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci. En effet
We see also note that if two policies share the same feature
expectation, then they have the same value function with respect to
any reward function $R_\theta=\theta^T\psi$,
$\theta\in\mathbb{R}^p$. Indeed:
%
\begin{equation}  \mu^{\pi_1} = \mu^{\pi_2} \Rightarrow
\theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} =
V^{\pi_2}. \label{memevaleur.eqn}
\end{equation}
%
%L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$. La plupart des algorithmes de la littérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert. Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert, quelle que soit la récompense (voir par exemple \cite{neu2009training}).
%
That is why most of the algorithms in the literature have, as we
will see, the objective of minimizing a \textit{distance}  between
the feature expectation of the expert and the feature expectation of
the apprentice. %That allows to obtain an apprentice which policy has
%the same value function as the expert policy, for a reward function
%of the type $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$ (see for
%instance
%~\cite{neu2009training}.

\section{State of the Art}
\label{biblio.sec}

Let $\pi_\theta$ denote an optimal policy according to a reward
function $R_\theta(s)=\theta^T\psi(s)$, for some
$\theta\in\mathbb{R}^p$. With a slight abuse of notation, let us
write $\mu^\pi_{d_0}$ the mean feature expectation for a
distribution $d_0$ over starting states:
$\mu^\pi_{d_0}=E[\mu^\pi(s,\pi(s))|s\sim d_0]$. The classical
approach to IRL (or less generally to apprenticeship learning, as
the output of the algorithm may be a policy and not a reward
function), initiated in~\cite{abbeel2004apprenticeship}, consists in
finding a reward (a parameter vector) $\theta^*$ minimizing some
loss function between the feature expectation of the expert and the
one of the related optimal policy:
\begin{equation}
  \theta^* = \argmin_\theta J(\theta) \text{ with } J(\theta)=
  L(\mu^{\pi_\theta}_{d_0},\mu^E_{d_0}).\nonumber
\end{equation}
Depending on the adopted point of view,
game-theoretic-base~\cite{syed2008game}, linear
programming-based~\cite{syed2008apprenticeship}, maximum
entropy-based~\cite{ziebart2008maximum}, \textit{etc.}, the loss
function and the associated update rule differ. Most of these
approches and others are nicely reviewed in~\cite{neu2009training}.

In all cases, the solution $\theta^*$ is computed in an iterative
way. At each step, one has to compute the feature expectation
$\mu^{\pi_\theta}_{d_0}$ for some parameter vector $\theta$. This
implies computing $\pi_\theta$, the optimal policy according to the
reward $\theta^T\psi(s)$. Therefore, solving the IRL problem with
one of these methods involves solving repeatedly the direct RL
problem (some of these algorithms requiring also the model to be
known). This is a drawback we wish to get rid of.

A notable exception is~\cite{boularias2011relative}. This work,
based on a relative entropy argument, finds $\theta^*$ by performing
a subgradient ascent on some utility function. Estimating the
subgradient would require to sample trajectories according to
$\pi_\theta$, with $\theta$ being the current estimate, thus would
require to solve the direct problem. However, this can be avoided in
this specific framework by the use of importance sampling. The
gradient is estimated from trajectories sampled from some random
policy. Still, this may be a problem when the only available data
comes from the expert.

Another approach consists in casting imitation learning into a
multi-class classification problem, introducing in some way the
structure of the MDP. For example, in~\cite{melo2010learning}, the
agent policy is learnt as the solution of a classification problem.
They consider a kernel function built on some induced MDP metric,
which may be combersome to compute. Closest to our proposed
contribution is the Maximum Margin Planning (MMP)
algorithm~\cite{ratliff2006maximum}, which is a classification-based
algorithm associating optimal policies to MDPs. More precisely, the
training set is made of $N$ couples $(\mathcal{M}_i,\pi^*_i)$, with
$\mathcal{M}_i$ an MDP (without reward) and $\pi_i^*$ the associated
optimal policy. Let $\mu^\pi_{d_0,\mathcal{M}_i}$ be the feature
expectation for the policy $\pi$, the initial distribution $d_0$ and
the dynamics induced by the related MDP. Let also
$\mathcal{L}_{\mathcal{M}_i,\pi}$ be a positive and user-defined
margin function penalizing non-optimal policies. MMP minimizes the
following cost function:
\begin{equation}
  J(\theta) = \frac{1}{N}\sum_{1=1}^N \max_\pi(\theta^T
  \mu^\pi_{d_0,\mathcal{M}_i} + \mathcal{L}_{\mathcal{M}_i,\pi}) - \theta^T
  \mu^{\pi_i^*}_{d_0,\mathcal{M}_i}.\nonumber
\end{equation}
This implies solving a set of MDPs at each iteration. At the other
extreme, a similar cost function is considered
in~\cite{ratliff2007imitation}, which searches for a score function
$q$ associating expert actions to states. The training set being $N$
couples $(s_i,a_i=\pi^E(s_i))$ and $l(s,a)$ being a positive
user-defined margin function penalizing non-optimal actions, the
cost function is
\begin{equation}
  J(q) = \frac{1}{N}\sum_{i=1}^N \max_a (q(s_i,a) +
  l(s_i,a))-q(s_i,a_i).\nonumber
\end{equation}
The drawback of this approach is that it does not take into account
the structure of the underlying MDP, which may easily be shown
experimentally to be a problem (\textit{e.g.},
see~\cite{melo2010learning}).

The proposed contribution is a trade-off between the two previous
approaches. It works at the state level rather than the MDP level,
considering a similar cost function. However, it takes into account
the structure of the MDP. As shown in the next section, this allows
avoiding to solve the direct RL problem as intermediate steps and it
requires only data from the expert.

%%%%%
\section{Structured classification for IRL}
\label{sec:scirl}
%%%%%

This section introduces our contribution, the SCIRL (Structured
Classification for IRL), and discusses one of its key components,
the estimation of the expert's feature expectation.

%%%
\subsection{Principle}

First, let us review more deeply the approach
of~\cite{ratliff2007imitation}, to which we will actually add some
structure. Assume that we have a set of transitions
$\{(s_i,a_i,s_{i+1})_{1\leq i\leq N}\}$ coming from one or several
trajectories of the expert. A classic approach for generalizing the
expert policy (from a multi-class classification-based point of
view) would consist in finding some score function $q:S\rightarrow
A$ such that for all $a\neq a_i$, $q(s_i,a_i)>q(s_i,a)$. In this
framework, states are inputs and actions are labels. Moreover, one
may want to strengthen the optimality of the expert. Let $l:S\times
A\rightarrow\mathbb{R}_+$ be a user-defined margin function, which
allows to put some prior knowledge in the problem at hand.
Typically, this margin function should satisfy $l(s_i,a_i)\ll
l(s_i,a)$. For example, one may choose $l(s_i,a\neq a_i)=1$ and
$l(s_i,a_i)=0$. Then, one searches for a score function $q$
satisfying
\begin{equation}
\label{eq:inequalities_score}
  \forall a,\; q(s_i,a_i) + l(s_i,a_i) \geq q(s,a) + l(s,a).\nonumber
\end{equation}
In order to find such a score function, a natural empirical cost
function to be minimized is:
\begin{equation}
\label{Jdef.eqn} J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in
A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
\end{equation}
This has been originally introduced in~\cite{ratliff2007imitation}.
Unfortunately, it ignores a large part of data: from the transitions
$(s_i,a_i,s'_i)$ performed by the expert, only the state action
couple $(s_i,a_i)$ is used and the dynamic part (transiting to
$s'_i$ from this state action couple) is ignored.
%
Moreover, it outputs a policy, whereas we are interested in the
reward function, which is a much more compact and transferable
representation of the task at hand.

Recall that we work in a sequential decision making framework.
Therefore, there is a natural solution to optimization
problem~\eqref{Jdef.eqn}: the optimal $Q$-function according to the
unknown reward function. Now, let $\theta$ be some parameter vector.
The score function $q(s,a) = \theta^T \mu^E(s,a)$ is actually the
$Q$-function of the expert policy according to the reward
$\theta^T\psi(s)$. Moreover, assume that this parameter vector is
such that inequalities~\eqref{eq:inequalities_score} are satisfied.
The optimal $Q$-function being stable by the greedy mechanism,
see~Eq.\eqref{greedy.eqn}, this means that the expert policy $\pi_E$
is optimal respectively to the reward function $\theta^T\psi(s)$
(and $\theta^T\mu^E(s,a)$ is the associated optimal $Q$-function). A
natural cost function for computing a reward function explaining the
expert behavior is therefore the following one:
\begin{equation}
\label{structure.eqn} J_N(\theta)  = {1\over N}
\sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) -
\theta ^T \mu^E_{s_ia_i} \right).
\end{equation}
Still, we need to minimize it.

%%%%
%%Comme nous l'avons vu Eq. \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue: $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
%%Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
%%Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert. Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons:
%As seen in Eq. \eqref{greedy.eqn}, there exists a greedy mechanism
%linking the expert policy $\pi_E$ and the optimal action-value
%$Q^E_R$ relative to the unknown reward function $R$: $\pi_E(s) \in
%\arg\max_{a\in A}Q^E_R(s,a)$.
%
%We will use this property to consider an empirical risk function
%$J_N$ dependent on a score function $q$ which will have to share
%with the unknown action-value function $Q^E_R$ the same optimal
%property. Let's note $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ the samples
%coming from one or several  trajectories of the expert. We are
%searching for a score function $q$ which verifies for each sample:
%\begin{equation}
%\label{butLAFEM.eqn}
%a_i \in \argmax_{a\in A}q(s_i,a).
%\end{equation}
%%Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert. Nous introduisons de surcroit une fonction $l: S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$. Ce type d'argument est assez commun dans la littérature (voir Sec. \ref{biblio.sec}). Nous introduisons ainsi la fonction de risque empirique suivante:
%So, we want to introduce a risk function which should penalize the
%cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases
%the score function $q$ does not allow to explain the choice $a_i$ of
%the expert (see equation \eqref{greedy.eqn}). Moreover, we introduce
%a margin function $l: S \times A \rightarrow \mathbb{R}_+$; this
%function allows the user to put some \emph{a priori} knowledge in
%the system because the user will choose the gap wanted in the score
%function between the action chosen by the expert and the second best
%possible action as we can see in the following equation: $\forall a,
%q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$. Introducing a margin function
%is quite common in the literature (see Sec. \ref{biblio.sec}). We
%introduce then the following empirical risk function:
%\begin{equation}
%\label{Jdef.eqn}
%J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
%\end{equation}
%%Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
%%
%%Nous traitons le problème de l'imitation comme un problème de classification multi-classe. Notre propos n'est cependant pas d'utiliser une approche purement supervisée. Nous ne souhaitons pas apprendre le contrôle de l'expert mais la description de la tâche qu'il effectue (la récompense). Il faut pour cela prendre en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnées par la politique de l'expert.\\
%%
%%Comme nous l'avons vu Eq. \eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi: S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique. Cette paramétrisation prend en compte la dynamique imposée par $\pi_E$. L'expert ne déroge pas à cette règle et notant $\mu_E$ l'attribut vectoriel moyen de l'expert, nous pouvons donc faire l'hypothèse que la fonction $q$ que nous recherchons peut s'exprimer sous la forme $q(s,a) = \theta^T\mu_E(s,a)$. L'estimation de ce terme $\mu_E$ est centrale à notre algorithme et nous proposons quelques solutions en Sec. \ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Eq. \eqref{hatRdef.eqn}):
%By default, choosing $l(s_i,a) = 1$ if $a\neq a_i$ and
%$l(s_i,a_i)=0$ works quite well as shown in the experimentations
%(Sec. \ref{exp.sec}) and in the analysis (Sec. \ref{proof.sec}).
%
%Here, we deal with the imitation problem as a classification
%problem. However, our purpose is not to use a purely supervised
%approach. We do not want to learn the policy of the expert but
%rather the description of the task he is executing (the reward
%function). In order to do that, it is necessary to take into account
%the MDP structure, and in particular the information on the
%transitions between states which are conditioned by the expert
%policy.

%As seen in Eq. \eqref{Qmu.eqn}, the hypothesis of a linear
%representation of the reward function by some features $\psi:
%S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) =
%\theta^T\mu^\pi(s,a)$. This parametrization takes into account the
%dynamic imposed by $\pi_E$. The expert follows the same rule and if
%we note $\mu^E$ the feature expectation of the expert, we can do the
%hypothesis that the score function $q$ that we are searching for can
%be written as follows: $q(s,a) = \theta^T\mu^E(s,a)$. Estimating the
%term $\mu^E$ is the key of our algorithm and we propose some
%solution in Sec. \ref{calculmu.sec}. We can now give the expression
%of our empirical risk function as a function of the parameter
%$\theta$ which defines the reward function (Eq.
%\eqref{hatRdef.eqn}):
%\begin{equation}
%\label{structure.eqn}
%J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) - \theta ^T \mu^E_{s_ia_i} \right)
%\end{equation}
%%Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$. Nous utilisons une descente de sous-gradient (normalisé), une généralisation du gradient, pour le minimiser. Le sous-gradient d'une fonction convexe $f:\mathbb{R}^p\rightarrow \mathbb{R}$ en $x$ est défini comme un vecteur $g$ tel que $\forall x' \in \mathbb{R}^p, g^T(x'-x)\leq f(x')-f(x)$. Le sous-gradient n'est pas partout unique, cependant l'unique sous-gradient d'une fonction différentiable est son gradient. Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $\max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$. En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée:

Because of the $\max$ operator, this cost function is not linear nor
differentiable everywhere. But it is clearly convex and piecewise
linear.
%
We can then use a subgradient descent algorithm. A subgradient of a
convex function $f:\mathbb{R}^p\rightarrow \mathbb{R}$ in
$x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall
x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$. The set of all
subgradients in $x$ is noted $\partial_xf$. The unique subgradient
in $x$ of a differentiable convex function $f$ is the gradient
$\nabla_xf$. As $A$ is finite, the convex and piecewise
differentiable function $\max_{a\in A}[f(\cdot,a)]$, from
$\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex
differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from
$\mathbb{R}^p$ to $\mathbb{R}$, has one of its subgradient in
$x\in\mathbb{R}^p$ equal to $\nabla_xf(x,a^*)$ with
$a^*\in\arg\max_{a\in A}f(x,a)$. Thanks to this, we can minimize the
cost function of interest with the following update rule ($\alpha_i$
denoting a usual learning rate):% and $\theta_0$ an initial parameter vector):
\begin{align}
  \theta_{t+1}&=\theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2}
  \\
  \text{with }
  g_{\theta} &= {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right)\in \partial_\theta J_N
  \\
  \text{and } a^*_i &\in \arg\max_{a\in A}(\theta_t^T \mu^E(s_i,a) +
  l(s_i,a)).
\end{align}
The SCIRL algorithm only requires data from the expert, assuming
$\mu^E$ is known on $\{(s_i,a)_{1\leq i \leq N,a\in A}\}$.
Unfortunately, this is usually not the case.

\subsection{Estimation of $\mu^E$}
%%%
\label{calculmu.sec}

Estimating the feature expectation of the expert is a key component
of our algorithm. We provide three possible solutions, depending on
the available data.

First, notice that each component $\mu_j^E$ of the expert feature
expectation $\mu^E$ is actually the $Q$-function of the expert
policy according to the reward function $\psi_j$, the $j^\text{th}$
basis function of the considered hypothesis space:
\begin{equation}
\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t
\psi_j(s)].\label{eq:muE_bellman}
\end{equation}
Therefore, it satisfies the Bellman evaluation
equation~\eqref{Qdef.eqn}. Assuming that the expert policy and the
transition probabilities are known, the expert feature expectation
can be exactly computed.

However, this assumption is rather strong (even if it is often done
in the IRL literature). We may assume that we can simulate
trajectories according to the expert policy on demand. In this case,
one may estimate $\mu^E(s,a)$ for a given state-action couple with
Monte Carlo. Assume that we have a set of $N$ trajectories of length
$H$, $\{s^i_0,a^i_0,\dots,s^i_{H-1},a^i_{H-1},s^i_{H}\}_{1\leq i
\leq N}$, with $s^i_0 =s$ and $a^i_0=a$, sampled according to the
system dynamic and the expert policy. Then $\mu(s,a)$ is estimated
as:
\begin{equation}
\hat\mu^E(s,a) = {1\over
N}\sum_{i=1}^{N}\sum^{H}_{t=0}\gamma^t\psi(s^i_t),
\end{equation}

Yet, as claimed at the beginning of this paper, we would like to
solely use the least amount of available data, that is is the set of
transitions $\{(s_i,a_i,s'_{i})_{1\leq i\leq N}\}$ from the expert.
As each component of the feature expectation can be seen as a
$Q$-function, see Eq.~\eqref{eq:muE_bellman}, it can be estimated
solely from the available data using the on-policy
LSTD~algorithm~\cite{bradtke1996linear}. This is the basic idea of
the LSTD$\mu$ approach~\cite{klein2011batch}, which is an LSTD-like
method estimating feature expectations. As a part of the
computations can be factored, LSTD and LSTD$\mu$ computational costs
have the same order of magnitude. We use LSTD$\mu$ in the
experimental section, which is the less constraining scheme.

Other schemes may be envisioned. For example, one may use a Monte
Carlo estimate for $\mu^E(s_i,a_i)$ (using only the available data)
and some heuristic for $\mu^E(s_i,a\neq a_i)$. We left this for
future work.

%%Comme signalé dans la Sec. \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen. Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu^E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert. Nous avons besoin de connaître $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$. Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
%As mentioned in \ref{ConsiderationsTechniques.sec}, some algorithms
%in the literature depend on the implementation of the feature
%expectation. Our algorithm, in particular, relies only on the
%implementation of $\mu^E$, the feature expectation of the expert,
%and does not require the repeated calculus of the feature
%expectation of another policy. For our algorithm, we only need to
%know $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$. That implies
%to infer the feature expectation of the expert on states where he
%has been, but on actions he has not taken.
%%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve:
%This inference is made easier by the fact that the feature
%expectation obeys the Bellman equation. Indeed, when we write its
%definition for a component $1\leq j\leq p$, we have:
%\begin{equation}
%\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t \psi_j(s)].
%\end{equation}
%%expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$. Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu^E$.\\
%When we compare this expression to the equation \eqref{Qdef.eqn}, it
%allows us to affirm that $\mu_j^E$ can be seen as the action-value
%function for the policy $\pi_E$ with respect to the reward function
%defined by the $j$-th component of the features $\psi$. Depending on
%the prior information, several methods can be used in order to
%estimate $\mu^E$.\\
%
%%Dans le cas idéal où le modèle est connu, le calcul exact de $\mu^E$ peut être effectué grâce aux algorithmes de programmation dynamique. Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu^E(s,a)$ (définition Eq. \eqref{Qmu.eqn}) l'on utilise
%In the ideal case where a model of the environnement is known
%(\textit{id est} the transition probabilities are given), the exact
%calculus of $\mu^E$ can be executed thanks to dynamic programmation
%algorithms. In a more realistic case, we can use a simple
%Monte-Carlo estimation if we dispose of a simulator and if it is
%possible to ask the expert what action has to be chosen in an
%important number of arbitrary states. The estimation of $\mu^E(s,a)$
%(definition Eq. \eqref{Qmu.eqn}) is done over $N$ trajectories of
%length $\{2k_i+1\}_{1\leq i \leq N}$,
%$\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i
%\leq N}$, by the following formula:
%\begin{equation}
%\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k_i}_{t=0}\gamma^t\psi(s^i_t),
%\end{equation}
%where $s^i_0=s, a^i_0=a$ and $s^i_{t+1}$ is distributed according to $P_{s^i_t\pi_E(s^i_t)}$.
%Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}. Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}. C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu^E$ en utilisant LSTD. Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
%Finally, if we are in the framework where only the data at our
%disposition are the trajectories samples given by the expert, it is
%possible to use the LSTD$\mu$ algorithm \citep{klein2011batch} in
%its \emph{on-policy} version. This algorithm adapts the LSTD
%algorithm (\cite{bradtke1996linear}) keeping its characteristics, in
%particular its ability to execute the calculus in  {\it batch}-mode.
%This is the method that we show in the section \ref{exp.sec}. The
%principle is to estimate each component of $\mu^E$ using LSTD. Some
%calculus can be aggregated, giving to LSTD$\mu$ a cost of calculus
%which is of the same order of magnitude as LSTD.
%%D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues. Cela est reporté pour des travaux futurs.
%%%%%%
%Other estimation methods are possible, as for instance a Monte-Carlo
%estimation, less calculus-expensive and constraining than the first
%one, which would be based solely on the data given by the expert and
%would imply the use of an heuristic on the unknown actions. This
%estimation would be consider on future works.
%

\section{Asymptotic analysis}\label{proof.sec}
In this section, we prove, under two assumptions, that there is
convergence of the subgradient algorithm applied to the asymptotic
risk function $J$ (that is $J = \lim_{N\rightarrow\infty} J_N$)
defined as:
\begin{equation}
J(\theta)=\sum_{s\in S}\nu_E(s)(\max_{a \in A}(\theta^T\mu^E_{sa}+l(s,a))-\theta^T\mu^E_{s\pi_E(s)}),
\end{equation}
where $\nu_E$ is the asymptotic probability distribution on states
induced by $\pi_E$. Moreover, we prove that the result of the
subgradient algorithm, which is
$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}$, is a minimum of
$J$ and that the expert policy $\pi_E$ is the unique optimal policy
for the reward function $\tilde{\theta}^T\psi$.

\subsection{Result}
\begin{assumption}
\label{assumption1}
It is assumed that we are able to estimate perfectly the term $\{\mu^E_{sa}\}_{s\in S,a\in A}$.
\end{assumption}

\begin{assumption}
\label{assumption2}
There exists $\theta\in\mathbb{R}^p$ such that $\forall s\in S$:
\begin{equation}
\label{existence.eq}
\theta^T\mu^E_{s\pi_E(s)}>\max_{a \in A\backslash\pi_E(s)}\theta^T\mu^E_{sa}.
\end{equation}
\end{assumption}

\begin{lemme}
\label{proposition1} Under assumptions \ref{assumption1} and
\ref{assumption2}, there exists a minimum $\theta^*$ of J. The
elements of the set $M^*=\{\theta^*\in\mathbb{R}^p:
J(\theta^*)=\inf_{\theta\in\mathbb{R}^p}J(\theta)\}$ satisfy
equation \eqref{existence.eq}.
\end{lemme}
\begin{proof}
Let the margin function $l$ be such that: $\forall s\in S, \forall
a\in A\backslash\pi_E(s), l(s,a)=c>l(s,\pi_E(s))=0$. Then $\forall
\theta\in\mathbb{R}^p,J(\theta)\geq\sum_{s\in
S}\nu_E(s)l(s,\pi_E(s))=0$. Let note $\hat{\theta}$ the vector
verifying \eqref{existence.eq}, it is clear that
$\|\hat{\theta}\|_2\neq0$. Let $x>0$ and
$\hat{\theta}_x$=$x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}$, then
there exists $x$ such that $J(\hat{\theta}_x)=0$. Indeed, $\forall
s\in S$ :
\begin{align}
&J(\hat{\theta}_x)=0\\
\Leftrightarrow&\hat{\theta}_x^T\mu^E_{s\pi_E(s)}\geq\max_{a \in A\backslash\pi_E(s)}(\hat{\theta}_x^T\mu^E_{sa}+l(s,a))\\
\Leftrightarrow&x\geq\frac{c}{\hat{\theta}_1^T\mu^E_{s\pi_E(s)}-\max_{a \in A\backslash\pi_E(s)}\hat{\theta}_1^T\mu^E_{sa}}.
\end{align}
Then all the vectors $\hat{\theta}_x$ with $x\geq c_{max}=\max_{s\in
S}\frac{c}{\hat{\theta}_1^T\mu^E_{s\pi_E(s)}-\max_{a \in
A\backslash\pi_E(s)}\hat{\theta}_1^T\mu^E_{sa}}$ are such that:
$\inf_{\theta\in\mathbb{R}^p}J(\theta)\geq0=J(\hat{\theta}_x)\geq\inf_{\theta\in\mathbb{R}^p}J(\theta)$.
So $\hat{\theta}_x=\inf_{\theta\in\mathbb{R}^p}J(\theta)$ for $x\geq
c_{max}$. Finally let us show that the elements of the set
$M^*=\{\theta^*\in\mathbb{R}^p:
J(\theta^*)=\inf_{\theta\in\mathbb{R}^p}J(\theta)\}$ satisfy
equation \eqref{existence.eq}. As there exists $x$ such that
$J(\hat{\theta}_x)=0=\inf_{\theta\in\mathbb{R}^p}J(\theta)$, it
implies $J(\theta^*)=0$. So:
\begin{align}
&J(\theta^*)=0\\
\Leftrightarrow&\theta^{*T}\mu^E_{s\pi_E(s)}\geq\max_{a \in A\backslash\pi_E(s)}(\theta^{*T}\mu^E_{sa}+l(s,a))\\
\Rightarrow&\theta^{*T}\mu^E_{s\pi_E(s)}>\max_{a \in A\backslash\pi_E(s)}\theta^{*T}\mu^E_{sa}.
\end{align}
\end{proof}

\begin{prop}
Under assumptions \ref{assumption1} and \ref{assumption2}, the
subgradient algorithm defined by the sequence $\theta_0=0$,
$\theta_{t+1}=\theta_t-\alpha_t\frac{g_{\theta_t}}{\|g_{\theta_t}\|_2}$
where $g_{\theta_t}=\sum_{s\in
S}\nu_E(s)(\mu^E_{sa^*}-\mu^E_{s\pi_E(s)})$ with
$a^*\in\argmax_{a\in A}\theta_t^T\mu^E_{sa}+l(s,a)$,
$\sum_{t\geq0}\alpha_t>\infty$ and $\sum_{t\geq0}\alpha_t^2<\infty$,
is convergent:
$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$.
Moreover we have that
$J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$. Finally
the expert policy $\pi_E$ is the unique optimal policy for the
reward $\tilde{\theta}^T\psi$.
\end{prop}
\begin{proof}
The cost function  $J$ is convex, proper, closed and with finite
values (the action space being finite). Therefore, using the result
of~\cite[section~5]{correa1993convergence} allows to conclude that
$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$
and $J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)$.
%
Moreover, Lemma~\ref{proposition1} shows that
$\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$.
%
So we have
$J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$ which
means that $\tilde{\theta}\in M^*$ and satisfies equation
\eqref{existence.eq}. This implies that, $\forall s\in S$, $\forall
a\in A$:
\begin{equation}
\tilde{\theta}^T\mu^E_{s\pi_E(s)}>\tilde{\theta}^T\mu^E_{sa},
\end{equation}
which proves that the expert policy $\pi_E$ is the unique optimal
policy for the reward $\tilde{\theta}^T\psi$.
\end{proof}

\subsection{Discussion}
The assumptions \ref{assumption1} and \ref{assumption2} are quite
important and we can of course wonder what results could be
guaranteed if they were not true. Moreover, we have only sampled
transitions, so we cannot use the asymptotic risk function $J$, only
the empirical function $J_N$. Thus, in our future works, the aim
will be to find, if possible, an upper bound for the term:
$E_{s\sim\nu_E}[V^{\pi_E}(s)-V^{\pi_A}(s)]$ where $E_{s\sim\nu_E}$
is the expectation over the stationary probability distribution of
states $\nu_E$ and $\pi_A$ is the policy of the apprentice which is
optimal for the reward $\tilde{\theta}^T\psi$ found by the
subgradient algorithm. Of course, this upper-bound will depend on
the feature expectation errors
$\{\epsilon^{\mu}_{s_ia}=\hat{\mu}^E_{s_ia}-\mu^E_{s_ia}\}_{1\leq i
\leq N,a\in A}$ (which may be quantified thanks to the finite sample
analysis of LSTD~\cite{lazaric:10}) and on the approximation error
$\epsilon^R$ which results from the fact that the unknown and
bounded reward function $R_E\in\mathbb{R}^S$ may not be a linear
combination of the features $\psi_{1\leq i \leq p}$. Formally, let
$T$ be the tuple $\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma\}$ which
is an MDP without reward and $T(\tilde{R})=\{S,A,\{P_{sa}\}_{s\in
S,a\in A},\gamma,\tilde{R}\}$ where $\tilde{R}\in\mathbb{R}^S$. We
note $C_R$ the reward-equivalence class of $R$:
$C_R=\{\tilde{R}\in\mathbb{R}^S: \Pi(T(R))=\Pi(T(\tilde{R}))\}$
where $\Pi(T(\tilde{R}))$ is the set of optimal deterministic
policies of $T(\tilde{R})$ (see~\cite{ng1999policyreward} for a
characterization of $C_R$). A formal definition of $\epsilon^R$ is:
$\epsilon^R=\|R^*-\theta^{*T}\psi\|_\infty=\max_{s\in
S}|R^*(s)-\theta^{*T}\psi(s)|$ with
$(R^*,\theta^*)\in\argmin_{\tilde{R}\in C_R,
\theta\in\mathbb{R}^p}{\|\theta^T\psi-R\|_2}$. Linking the empirical
risk function $J_N$ to the true risk function $J$ could be done
using a Vapnik-like analysis. The difficult thing of course is to
have access to $V^{\pi_A}$ without resolving the MDP directly.

%\section{Related work}
%\label{biblio.sec}
%%%%%%
%%L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense. Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de \mbox{l'expert} et celui d'une politique donnée. Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
%The historical approach of \cite{ng2000algorithms}, as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} as summed up in \cite{neu2009training} all suffer from the same constraint: the need to repeatedly solve the forward problem, that is to say the need to repeatedly find the optimal policy with respect to an arbitrary reward. All these approaches try to minimize some distance between the feature expectation of the expert and the feature expectation of another policy. Successive updates of the vector of parameters $\theta$ will get this other policy (optimal with respect to a criterion that depends on $\theta$, this is where the \emph{RL} problem needs to be solved) closer and closer to the expert's, feature expectation-wise.\\
%
%%Cette résolution est très problématique. En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la littérature. Particulièrement, l'information nécessaire à cet objectif est bien plus étendue que celle nécessaire au calcul d'uniquement l'attribut moyen de l'expert comme cela est nécessaire à notre algorithme. Comme cela est illustré dans \cite{klein2011batch}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état, ne permettant pas la résultion du problème direct en se basant uniquement sur les  données de l'expert.\\
%%FIXME:parler de boularias ici
%Solving the forward problem can be very tricky ; indeed a whole body of literature is dedicated to this problem. Specifically, one needs more information to solve the forward problem than to just compute the feature expectation of the expert as our algorithm does. As illustrated in \cite{klein2011batch}, \citep{abbeel2004apprenticeship}'s algorithm is unable to solve the inverted pendulum problem with just samlpes from the expert as those cover too small a part of the state space. Supplementary randomly drawn samples are needed. Even a recent contribution like \cite{boularias2011relative} still needs samples that cover a large portion of the state space although it does not explicitely require the ability to solve the forward problem.\\
%
%%Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge. C'est à dire que l'action choisie par l'expert doit avoir une qualité strictement supérieure à n'importe quelle autre action. Les approches de planification structurée comme \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} permettent d'influer sur cette marge.\\
%
%A recurring argument is to see the expert not only as optimal but as optimal by a certain margin. The action chosen by the expert should have a quality strictly greater than any other action. Structured planning as done in \cite{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} allow some influence on the margin.\\
%
%%Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}. Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}, notre fonction de risque étant similaire à la leur. La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}. Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires: à un PDM (entrée à classifier) est associée une politique (classe). Le cadre applicatif est déterministe, l'algorithme employé pour la résolution des PDM étant $A^*$. Il semble difficile de sortir de ce cadre  en pratique (bien que l'algorithme soit introduit dans un cadre plus général).
%
%Specifically, our work shows similarities with \cite{ratliff2006maximum}. The authors cast the \emph{IRL} problem as a maximum margin classification problem ; our risk function is similar to theirs. The method they use can be seen as an iterative algorithm in the tracks of \cite{abbeel2004apprenticeship}. From a higher abstraction level, the classification takes place in the space of trajectories : a policy (the label) is associated with a \emph{MDP} (input to classify). The empirical demonstration uses a deterministic setting, as they use $A^*$ to solve the \emph{MDP}. Although the proposed algorithm is described in a more general setting, it seems unpractical outside the deterministic case.
%
%%Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe. Le principe de leur approche est exactement de résoudre le risque (Eq. \eqref{Jdef.eqn}). Toutefois, ils ne prennent pas en comptent la structure du MDP, ce à quoi nous remédions avec notre contribution (voir \cite{melo2010learning} sur l'importance de prendre en compte cette structure).
%%Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score. Cette fonction de score $s: X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e. choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$. Leur approche souffre de ne pas profiter de l'introduction de la structure dont profite la nôtre (Eq. \eqref{structure.eqn}).\\
%Finally, \citep{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem. Both our approach and theirs minimize the same risk function, described Eq. \eqref{Jdef.eqn}, but they do not use the structure of the \emph{MDP} as we do (see \cite{melo2010learning} about the importance of taking this structure into account).\\
%
%%Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde). Il existe d'autres approches que nous ne passons pas en revue de manière exhaustive, mais à notre connaissance elles nécessitent toutes de résoudre le problème direct.
%
%Our approach is in a way between those two contributions as we propose to introduce the structure of the \emph{MDP} (this is what the first contribution do) in the same abstraction level as the second approach. Other techniques exist, but to the best of our knowledge none of them can work only from samples drawn by the expert.

%%%%%
\section{Experiments}
\label{exp.sec}
%%%%%
%%%
In this section, experiments on two classical benchmarks are
proposed: the inverted pendulum and the car driving problems.

\subsection{Inverted Pendulum}
%%%
%Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté. Il s'agit d'un problème classique en apprentissage par renforcement.
%Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
%Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
%Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle)
%la seconde étant sa vitesse angulaire.\\
%
%
The inverted pendulum problem consists in maintaining in vertical
equilibrium a pendulum with one degree of freedom by moving the cart
on which it is fixed. This classical problem is extensively
described in~\cite{lagoudakis2003least} from which the parameters of
our experiment were taken. The state space is two-dimensional: the
first dimension is the angular position of the pendulum (where the
vertical position corresponds to the angle zero) and the second one
is the angular speed. The policy of the expert is determined by the
Least Square Policy Iteration (LSPI) algorithm
from~\cite{lagoudakis2003least}, using the same basis functions
(Gaussian functions).
%
%La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
%Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
The policy of the expert optimizes the reward function given Figure
\ref{onlyFig.fig} (a). It manages systematically to maintain the
pendulum during 3000 time-steps. A time-step having a value of 0.1s,
the expert is interrupted after 5 minutes of successful
interactions.
%
%Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases,
%de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
%Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie
%(correspondant à la position verticale et son voisinage proche).\\
%

The main difficulty of this experiment relies in the fact that the
state space is continuous and, above all, that expert trajectories
(shown in green on Figure~\ref{onlyFig.fig} (b)) are only covering a
very small area of the state space.
%
%Ce dernier point est d'ailleurs la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
%En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser
%les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}. Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel
%moyen est utilisée en association à celle de l'expert. Cette politique est obtenue par optimisation d'une récompense sur le PDM dans lequel évolue l'expert ; cette résolution du problème
%direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état. Comme notre algorithme ne nécessite pas le calcul d'autre
%attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
%%%
For this reason, the algorithm of \citet{abbeel2004apprenticeship}
fails in finding a solution to this problem only with the
transitions available from the expert demonstrations. Indeed, as
shown in~\citep{klein2011batch}, additional data covering the whole
state space is required.
%
%These algorithms are iterative and try to
%find a policy $\pi$ which feature expectation is compared to the
%feature expectation of the expert. Thus, these algorithms need to
%resolve the MDP and to do so need samples covering the whole state
%space. Our algorithm, only need to be able to estimate the feature
%expectation of the expert which does note require the direct
%resolution of the MDP.
\subsubsection{Results}
%%%
\begin{figure}[ht]
%\vskip 0.2in
\begin{center}
\begin{minipage}[b]{.48\linewidth}
  \centering
  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_true_R.pdf}}
  \centerline{(a) Expert reward ($R^E$)}%\medskip
\end{minipage}
%\hfill
\begin{minipage}[b]{.48\linewidth}
  \centering
  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_lafem_R.pdf}}
  \centerline{(b) SCIRL Reward ($R_\theta$)}%\medskip
\end{minipage}
\begin{minipage}[b]{.48\linewidth}
  \centering
  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vexpert.pdf}}
  \centerline{(c) Expert value function}%\medskip
\end{minipage}
\hfill
\begin{minipage}[b]{.48\linewidth}
  \centering
  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vagent.pdf}}
  \centerline{(d) SCIRL value function}%\medskip
\end{minipage}
%
\caption{SCIRL results on the inverted pendulum} \label{onlyFig.fig}
%
\end{center}
\vskip -0.2in
\end{figure}
%
%Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
%La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon. Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations
%est de $T=20$. Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange
%de gaussiennes défini dans \cite{lagoudakis2003least}. Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
%
So as to test the SCIRL algorithm, a database $D_E$ of transitions
given by the expert policy $\pi_E$ is built. It contains 10
trajectories of 300 transitions each. The margin function is $l(s,a)
= 0$ if $a=\pi_E(s)$, $1$ otherwise.
%The time step is constant,
%$\Delta t=0.1s$, and the stopping time is $T=20$.
The initial vector, $\theta_0$, is fixed at $\begin{pmatrix}   -1 &
\dots & -1
\end{pmatrix}^T$,
and the basis functions ($\psi: S \rightarrow \mathbb{R}^p$) used
for the reward function are a Gaussian network as defined
in~\cite{lagoudakis2003least}. The LSTD$\mu$
algorithm~\cite{klein2011batch} is used to estimate the feature
expectation $\mu^E(s,a)$.

%Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs
%sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)). Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense
%pouvant expliquer une politique.\\
%
Even if the reward function found by SCIRL (figure \ref{onlyFig.fig}
(b)) is different than the expert reward (figure \ref{onlyFig.fig}
(a)), the value functions are similar (figures \ref{onlyFig.fig} (c)
et \ref{onlyFig.fig} (d)). This is an illustration that the IRL
problem is ill-posed in the sense that there is not only one reward
function able to explain the optimality of a policy.

%Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes,
%ce qui est notre critère de réussite.\\
With the number of samples given by the expert (3000), the
apprentice is able to systematically maintain the pendulum in
equilibrium during 5 minutes as the expert does.
%
%Deux éléments méritent d'être notés. Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
%Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace:
%celle dans laquelle le pendule est proche de la verticale. En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
%Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer
%la fonction de récompense utilisée sans l'atteindre. Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir
%des performances similaires à celles de l'expert. Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig. \ref{onlyFig.fig} Cela n'a aucune importance, une dilatation
%laissant la politique optimale invariante.
%
Two elements deserve to be noticed. First, the expert samples are
only covering a small area of the state space (see
Figure~\ref{onlyFig.fig} (b)) and especially they are close to the
vertical position. The lack of samples in the rest of the state
space makes it impossible to infer with certainty the reward
function. Second, the \textit{actual} reward function $R^E$
(Figure~\ref{onlyFig.fig} (a)) doesn't live in the hypothesis space
span by a linear combination of the basis functions. Despite these
difficulties, SCIRL succeeds to extract a reward function leading to
perform similarly to the expert. %Notice that there is a magnitude
%difference between the SCIRL and the expert value functions on
%Figure~\ref{onlyFig.fig} but an affine transform of the value
%function does not change the optimal policies.

\subsection{Car driving}

This task is taken from~\cite{syed2008game}. It consists in
navigating a car through randomly-generated traffic on a threelane
highway. The reward features are identical to~\citet{syed2008game}
paper that is: a collision feature (0 if contact with another car,
and 0.5 otherwise), an off-road feature (0 if on the grass, and 0.5
otherwise), and a speed feature (0.5, 0.75 and 1 for each of the
three possible speeds, with higher values corresponding to higher
speeds).

\subsubsection{Results}

In this section, SCIRL is compared to the Multiplicative Weights for
Apprenticeship Learning (MWAL) algorithm proposed by
\citet{syed2008game}. This algorithm requires the perfect knowledge
of the MDP and outputs a single policy (not a reward) that imitates
the expert.

The experimental setup is as follows. The highway simulator
of~\cite{syed2008game} is used\footnote{available at \url{
http://www.cs.princeton.edu/~usyed/}}. It also contains an exact MDP
solver for the task, the transition probabilities being available.
Two reward functions were hand-tuned ($\theta$ parameters chosen by
hand) so as to generate two different expert policies: one for
high-speed driving and another for safe driving. The MDP solver is
used to obtain the expert policies given the two reward functions. A
number $N$ of trajectories of length $H$ is generated with each
strategy providing the expert data for training the SCIRL and MWAL
algorithms. The MWAL algorithm requires the knowledge of the average
feature expectation $\mu_{d0}^{\pi_E}$ for the initial state (given
a distribution $d_0$ over starting states). To reach a fair
comparison with SCIRL which only uses the expert trajectories,
$\mu_{d_0}^{\pi_E}$ is computed as the Monte Carlo estimate given
the $N$ trajectories, knowing that the initial state is always the
same:
%
$$\mu_{d_0}^{\pi_E} = \frac{1}{N}\sum_{i=1}^N \sum_{t=1}^H \gamma^t \psi(s_t^i)$$
%
For SCIRL, the LSTD$\mu$ algorithm is used again to approximate
$\mu^E(s,a)$ from the data. For each component of $\mu^E$, the basis
functions are the same as those for the reward function. SCIRL
outputs a reward function which is used to obtain a policy using the
MDP solver. The SCIRL and MWAL policies are then run on the
simulator for trajectories of length $T$. The average discounted
cumulative reward, computed with regard to the initial handcrafted
rewards, serves as the performance measure. It is an Monte Carlo
estimate of $V^\pi(s_0)$ for each policy.

This experiment is repeated $M$ times for each policy and results
are averaged. The results are shown in
Tables~\ref{table:highway_fast} and~\ref{table:highway_safe}
 for $N \in
\{1, 3, 5, 10\}$, $H = 30$, $T=50$ and $M = 10$.

%\begin{figure}
%  \includegraphics[width=\linewidth]{FastResults_EB.pdf}\\
%  \caption{Results for high-speed driving}\label{fig:highway_fast}
%\end{figure}
%
%\begin{figure}
%  \includegraphics[width=\linewidth]{SafeResults_EB.pdf}\\
%  \caption{Results for high-speed driving}\label{fig:highway_safe}
%\end{figure}

\begin{table}
\begin{center}
    \begin{tabular}{|c|c|c|c|}
    \hline
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    N     & MWAL              & SCIRL               & $V_{\text{opt}}$ \\\hline\hline
    1     & 7.25 $\pm$ 0.10   & 7.03 $\pm$ 0.09     & 7.30              \\\hline
    3     & 7.28 $\pm$ 0.06   & 7.07 $\pm$ 0.07     & 7.30              \\\hline
    5     & 7.27 $\pm$ 0.11   & 6.98 $\pm$ 0.10     & 7.30              \\\hline
    10    & 7.26 $\pm$ 0.07   & 7.05 $\pm$ 0.10     & 7.30              \\
    \hline
    \end{tabular}
    \caption{SCIRL results on the car driving
    problem (high-speed)}\label{table:highway_fast}
\end{center}
\end{table}

\begin{table}
\begin{center}
    \begin{tabular}{|c|c|c|c|}
    \hline
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    N     & MWAL              & SCIRL               & $V_{\text{opt}}$ \\\hline\hline
    1     & 4.95 $\pm$ 0.05   & 3.27 $\pm$ 0.13     & 5.00              \\\hline
    3     & 4.85 $\pm$ 0.13   & 4.90 $\pm$ 0.10     & 5.00              \\\hline
    5     & 4.90 $\pm$ 0.09   & 4.83 $\pm$ 0.08     & 5.00              \\\hline
    10    & 4.98 $\pm$ 0.03   & 4.81 $\pm$ 0.12     & 5.00              \\
    \hline
    \end{tabular}
    \caption{SCIRL results on the car driving
    problem (safe)}\label{table:highway_safe}
\end{center}
\end{table}

From these tables, one can conclude that the SCIRL method reaches
approximately the same performance as MWAL (given that enough
samples are provided in the case of safe driving). Yet, SCIRL
doesn't need any other information than the expert demonstrations
while MWAL requires solving exactly and repeatedly the MDP.

\section{Conclusion}

%L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse. En rendant la résolution du problème inverse indépendante
%de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
In this paper, we propose an algorithm for inverse reinforcement
learning (IRL) that we call SCIRL. It is based on a classification
approach in which the structure of the underlying Markov Decision
Process (MDP) is taken into account.

This approach relaxes most of the constrains of other IRL algorithms
found in the literature. It doesn't require the resolution of the
direct problem and is able to infer a reward function which is
consistent with the expert's behavior in a purely \emph{batch} mode.
%
%Comme nous l'avons illustré, les échantillons nécessaires à la réussite de notre algorithmes sont faciles à recueillir.
%Une simple trace de l'expert, même si elle ne couvre qu'une petite partie de l'espace d'état, suffit à inférer une récompense.
%Cette récompense, description compacte de la tâche effectuée par l'expert, peut alors être optimisée par un agent dont les capacités diffèrent de celles de l'expert.
%On rentre dans le cadre du \emph{transfer learning}.\\
%
A mere trajectory of the expert, even if it covers a small part of
the state space, is sufficient to infer a reward function and no
additional data nor a simulator is required. It also outputs a
reward and not a strategy which is the real goal of the IRL
paradigm. The learnt reward function, which is a compact description
of the task executed by the expert, could then be optimized by an
agent with abilities (action set) that are different from those of
the expert which places us in in the even more challenging framework
of \emph{transfer learning}.
%
%L'estimation de $\mu_E$ est centrale dans notre approche. Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients
%que les algorithmes d'estimation d'une fonction de valeur. L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen
%d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée. Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.

SCIRL is shown to compare positively to a state-of-the-art algorithm
(the MWAL algorithm) on a standard benchmark (the car driving
problem). It also performs well on a problem with a two-dimensional
continuous state space which is sparsely explored by the expert (the
inverted pendulum problem). This later problem could not be solved
with state-of-the-art algorithms without additional knowledge.

Yet, SCIRL requires the computation of $\mu^E(s,a)$ everywhere on
the expert trajectories. Here, we made use of the LSTD$\mu$
algorithm to estimate it. However LSTD$\mu$ shares the drawbacks of
value function estimation algorithms. In the future, we plan to
extend our approach so as to avoid the computation of the expert's
feature expectations. Empirical tests on more complex problems will
also be considered.

\bibliography{Biblio}
\bibliographystyle{icml2012}
\end{document}

### 2012-03-28 Nao/main.c
Code from V.parent

@@ -1,0 +1,54 @@
+#include <stdio.h>
+#include <stdlib.h>
+
+#include <gsl/gsl_matrix.h>
+#include "LSTDQ.h"
+#include "LSTDmu.h"
+#include "utils.h"
+#include "greedy.h"
+#include "LSPI.h"
+#include "RL_Globals.h"
+
+#define SAS     "../resources/sas.txt"
+#define OMEG0   "../resources/omeg0.txt"
+#define OMEG   "../resources/omeg.txt"
+#define S     "../resources/s.txt"
+#define S0     "../resources/s0.txt"
+#define A   "../resources/a.txt"
+
+#define DIM 3   //nb d'articulations consideree
+#define LONG    10  //longueur d'une trajectoire
+#define GAMMA   0.9     //gamma de l'irl
+#define K   15  //nb de fct de mapping
+#define EPSILON 0.001   //TBD : arret de l'algo
+
+gsl_matrix* g_mActions = NULL;
+
+int main()
+{
+    int i=0,j=0;
+
+    printf("Hello world!\n");
+
+
+
+
+g_mActions = file2matrix( A, g_iA );
+
+gsl_matrix* mat_sas = file2matrix(SAS, (2*g_iS+g_iA + 1 + 1)); //DIM-s DIM-a DIM-s' 1-42 1-e
+gsl_matrix* mat_lstd =proj_lstd_lspi_ANIRL(mat_sas,mat_sas);
+//gsl_matrix* mat_omega = file2matrix(OMEG, g_iK);
+
+for(i=0;i<mat_lstd->size1;i++)
+    {
+     for(j=0;j<mat_lstd->size2;j++)
+     {
+         printf("%f ",gsl_matrix_get(mat_lstd, i, j));
+
+     }
+     printf("\n");
+    }
+
+
+    return 0;
+}

### 2012-03-28 Nao/phipsi.c

@@ -1,0 +1,135 @@
+#include <gsl/gsl_matrix.h>
+#include "RL_Globals.h"
+#include <math.h>
+#include "InvertedPendulum.h"
+
+
+unsigned int g_iK = (10*15); /* dim(\phi) */
+unsigned int g_iP = (15); /* dim(\psi) */
+
+int sigma_map=2000;
+
+
+int centroids[15][3]=
+{0,-276,254,\
+0,-276,1300,\
+0,-828,777,\
+0,-1380,254,\
+0,-1380,1300,\
+-695,-276,254,\
+-695,-276,1300,\
+-695,-828,777,\
+-695,-1380,254,\
+-695,-1380,1300,\
+695,-276,254,\
+695,-276,1300,\
+695,-828,777,\
+695,-1380,254,\
+695,-1380,1300};
+
+
+
+gsl_matrix* psi( gsl_matrix* s ){
+  gsl_matrix* answer = gsl_matrix_calloc( g_iP, 1 );
+  double x;
+
+  unsigned int index = 0;
+
+double temp[g_iS];
+double  res=0;
+
+    for (int k=0;k<g_iP;k++)
+    {
+        for (int j = 0;j<g_iS;j++)
+        {
+            x= gsl_matrix_get( s, 0, j );
+
+                //calcul avec la gaussienne pour chaque dim
+                temp[j]=exp(- (x-centroids[k][j])*((x-centroids[k][j])) / (2*sigma_map*sigma_map) ) / (sigma_map * sqrt(2*PI));
+                temp[j]=pow(temp[j],2);
+        }
+
+        for (int j = 0;j<g_iS;j++)
+        {
+            res+=temp[j];
+        }
+
+        gsl_matrix_set( answer, k, 0, sqrt(res));
+    }
+
+  return answer;
+}
+
+
+
+gsl_matrix* phi( gsl_matrix* sa ){
+
+    gsl_matrix* answer = gsl_matrix_calloc( g_iK, 1 );
+
+    gsl_matrix* a = gsl_matrix_calloc( 1,g_iA );
+    gsl_matrix* s = gsl_matrix_calloc( 1,g_iS );
+
+    gsl_matrix_set(s,0,0,gsl_matrix_get(sa,0,0));
+    gsl_matrix_set(s,0,1,gsl_matrix_get(sa,0,1));
+    gsl_matrix_set(s,0,2,gsl_matrix_get(sa,0,2));
+    gsl_matrix_set(a,0,0,gsl_matrix_get(sa,0,3));
[…]

### 2012-03-28 Nao/VoidCriteria.c
Le code relatif au nao compile et tourne

@@ -1,0 +1,18 @@
+//FIXME:Corriger ANIRL versionLSTDmu pour ne plus nécessiter ce qui se trouve dans ce fichier.
+
+#include <gsl/gsl_matrix.h>
+
+double g_dBest_error = -1;
+double g_dBest_true_error = -1;
+double g_dBest_diff = -1;
+double g_dBest_t = 0;
+gsl_matrix* g_mBest_omega = NULL;
+
+
+double true_diff_norm( gsl_matrix* r){
+  return -1;
+}
+
+double true_V_diff( gsl_matrix* r){
+  return -1;
+}

### 2012-03-28 Nao/main.c

@@ -1,4 +1,5 @@
 #include <stdio.h>
+
 #include <stdlib.h>
 
 #include <gsl/gsl_matrix.h>
@@ -8,13 +9,14 @@
 #include "greedy.h"
 #include "LSPI.h"
 #include "RL_Globals.h"
+#include "phipsi.h"
 
-#define SAS     "../resources/sas.txt"
-#define OMEG0   "../resources/omeg0.txt"
-#define OMEG   "../resources/omeg.txt"
-#define S     "../resources/s.txt"
-#define S0     "../resources/s0.txt"
-#define A   "../resources/a.txt"
+#define SAS     "sas.txt"
+#define OMEG0   "omeg0.txt"
+#define OMEG   "omeg.txt"
+#define S     "s.txt"
+#define S0     "s0.txt"
+#define A   "a.txt"
 
 #define DIM 3   //nb d'articulations consideree
 #define LONG    10  //longueur d'une trajectoire
@@ -24,8 +26,39 @@
 
 gsl_matrix* g_mActions = NULL;
 
+gsl_matrix* initial_state( void );
+
+unsigned int g_iS = 3;
+unsigned int g_iA = 3;
+unsigned int g_iIt_max_lspi = 50;
+gsl_matrix* (*g_fPhi)(gsl_matrix*) = &phi;
+gsl_matrix* g_mOmega = NULL;
+double g_dLambda_lstdQ = 0.1;
+double g_dGamma_lstdq =  0.9;
+double g_dEpsilon_lspi = 0.01;
+double g_dLambda_lstdmu = 0.1;
+double g_dGamma_anirl = 0.9;
+double g_dEpsilon_anirl = 0.01;
+unsigned int g_iIt_max_anirl = 2;
+//gsl_matrix* g_mActions = NULL;
+gsl_matrix* (*g_fPsi)(gsl_matrix*) = &psi;
+gsl_matrix* (*g_fSimulator)(int) = NULL;
+gsl_matrix* (*g_fS_0)(void) = &initial_state;
+unsigned int g_iMax_episode_len = 20; //Shorter episodes are
+
+
+
+gsl_matrix* initial_state( void ){
+  gsl_matrix* answer = file2matrix("s0.txt",3);
+
+  return answer;
+}
+
+
 int main()
+
 {
+
     int i=0,j=0;
 
     printf("Hello world!\n");
@@ -51,4 +84,6 @@ for(i=0;i<mat_lstd->size1;i++)
 
 
     return 0;
+
 }
+

### 2012-03-28 Nao/phipsi.c

@@ -1,7 +1,10 @@
 #include <gsl/gsl_matrix.h>
 #include "RL_Globals.h"
 #include <math.h>
-#include "InvertedPendulum.h"
+#ifndef M_PI
+#define M_PI           3.14159265358979323846
+#endif
+
 
 
 unsigned int g_iK = (10*15); /* dim(\phi) */
@@ -10,46 +13,43 @@ unsigned int g_iP = (15); /* dim(\psi) */
 int sigma_map=2000;
 
 
-int centroids[15][3]=
-{0,-276,254,\
-0,-276,1300,\
-0,-828,777,\
-0,-1380,254,\
-0,-1380,1300,\
--695,-276,254,\
--695,-276,1300,\
--695,-828,777,\
--695,-1380,254,\
--695,-1380,1300,\
-695,-276,254,\
-695,-276,1300,\
-695,-828,777,\
-695,-1380,254,\
-695,-1380,1300};
+int centroids[15][3]= {0,-276,254,
+		       0,-276,1300,
+		       0,-828,777,
+		       0,-1380,254,
+		       0,-1380,1300,
+		       -695,-276,254,
+		       -695,-276,1300,
+		       -695,-828,777,
+		       -695,-1380,254,
+		       -695,-1380,1300,
+		       695,-276,254,
+		       695,-276,1300,
+		       695,-828,777,
+		       695,-1380,254,
+		       695,-1380,1300};
 
 
 
 gsl_matrix* psi( gsl_matrix* s ){
   gsl_matrix* answer = gsl_matrix_calloc( g_iP, 1 );
-  double x;
-
-  unsigned int index = 0;
-
-double temp[g_iS];
-double  res=0;
-
-    for (int k=0;k<g_iP;k++)
-    {
-        for (int j = 0;j<g_iS;j++)
-        {
-            x= gsl_matrix_get( s, 0, j );
-
-                //calcul avec la gaussienne pour chaque dim
-                temp[j]=exp(- (x-centroids[k][j])*((x-centroids[k][j])) / (2*sigma_map*sigma_map) ) / (sigma_map * sqrt(2*PI));
-                temp[j]=pow(temp[j],2);
-        }
-
-        for (int j = 0;j<g_iS;j++)
+  double x;  
+  double temp[g_iS];
+  double  res=0;
+  
+  for (int k=0;k<g_iP;k++){
[…]

### 2012-04-10 klein2012/main.tex
Correction de typos repérées à la relecture du papier

--- 2012-02-22 klein2012/main.tex
@@
 \usepackage{algorithmic}
+\usepackage{amsthm}
 \usepackage{hyperref}
+\usepackage{url}
 \newcommand{\theHalgorithm}{\arabic{algorithm}}
@@
 \mathtoolsset{showonlyrefs=true}
-\icmltitlerunning{Submission and Formatting Instructions for ICML 2012}
 
 \newtheorem{definition}{Définition}
-\newtheorem{lemme}{Lemme}
+\newtheorem{lemme}{Lemma}
 \newtheorem{theorem}{Théorème}
@@
 \newtheorem{remarque}{Remarque}
+\newtheorem{assumption}{Assumption}
 \newtheorem{problem}{Problème}
@@
 \newcommand{\espace}{\text{ }}
+
 \newcommand{\x}{\mathbf{x}}
@@
 %}
+\icmltitlerunning{Structured Classification for Inverse
+Reinforcement Learning}
 
@@
 
-\twocolumn[
-\icmltitle{Submission and Formatting Instructions for the Twenty-Eighth \\
-           International Conference on Machine Learning (ICML 2012)}
+\twocolumn[ \icmltitle{Structured Classification for Inverse
+Reinforcement Learning}
 
@@
 % the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{boring formatting information, machine learning, ICML}
+\icmlkeywords{inverse reinforcement learning, structured multi-class
+classification}
 
@@
 \begin{abstract}
-In this paper we tackle the problem of apprenticeship learning in the Inverse Reinforcement Learning (\emph{IRL}) framework.
-An expert fulfils a task to be imitated by an artificial agent. The IRL framework assumes that the expert is acting optimally with respect to a reward function;
-the goal is to guess this function from trajectories drawn by the expert. Existing IRL algorithms need one or more of the following items: complete trajectories from the expert,
-a generative model of the environment for Monte-Carlo estimations, the knowledge of the transition probabilities, the ability to repeatedly solve the forward problem (the Reinforcement Learning (\emph{RL}) problem),
-the access to the expert's strategy anywhere in the state space, or at the very least random samples covering a large part of it.
-Our contribution consists in a new IRL algorithm lifting all these constraints.
-Using a supervised technique in which the structure of the underlying Markov Decision Process (\emph{MDP}) is implicitely injected, we give rise to a subgradient descent-based algorithm whose computational
-and sample complexities are low.
-The sample base from the expert is the only data we need.
-A formal analysis of the algorithm as well as empirical demonstrations on two classic but challenging problems are provided.
+This paper addresses the Inverse Reinforcement Learning (IRL)
+problem which is a particular case of learning from demonstrations.
+The IRL framework assumes that an expert, demonstrating a task, is
+acting optimally with respect to an unknown reward function to be
+discovered. Unlike most of existing IRL algorithms, the proposed
+approach doesn't require any of the following: complete trajectories
+from the expert, a generative model of the environment, the
+knowledge of the transition probabilities, the ability to repeatedly
+solve the forward Reinforcement Learning (RL) problem, the expert's
+policy anywhere in the state space. Using a classification approach
+in which the structure of the underlying Markov Decision Process
+(MDP) is implicitly injected, we end-up with an efficient
+subgradient descent-based algorithm.
+% whose computational and sample
+%complexities are low.
+In addition, only a small amount of expert
+demonstrations (not even in the form of trajectories but simple
+transitions) is required.
+%A formal analysis of the algorithm as well as empirical
+%demonstrations on two classic but challenging problems are provided.
 \end{abstract}
@@
 Le but de l'agent est alors d'inférer cette fonction des démonstrations données par l'expert pour ensuite optimiser son comportement relativement à cette fonction lui aussi.\\
-In this paper, we consider the apprenticeship learning problem (also called learning by watching or learning from demonstrations problem) where an artificial agent (or apprentice) is learning to perform a task thanks to the observation of an expert demonstrating the task.
-In order to find a solution to this problem, we propose to use the inverse reinforcement learning paradigm where we suppose that the expert acts in order to be rewarded by an unknown reward function.
-The aim of the agent is then to infer the unknown reward function from the demonstrations given by the expert in order to find an optimal behavior (or policy) with respect to the reward function he has just inferred.
+
+This contribution addresses the problem of learning by watching or
+learning from demonstrations. In this context, an artificial agent
+learns to perform a task by observing an expert demonstrating the
+task. In order to find a solution to this problem, the Inverse
+Reinforcement Learning (IRL) paradigm has been
+proposed~\cite{russell1998learning}. The fundamental idea of IRL is
+that the expert acts so as to be rewarded by an unknown reward
+function. The aim of the IRL agent is then to infer the unknown
+reward function from the demonstrations given by the expert, the
+reward being considered as the most compact representation of the
+task to be transferred to the agent. This way, an optimal behavior
+can be obtained in any situation by optimizing the action selection
+process according to the reward function. Unlike supervised
+learning, the agent may not purely imitate the expert's actions but
+will try to accomplish the same task optimally.
+%
+%
 %La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
@@
 Ces contraintes sont souvent incompatibles avec la mise en oeuvre des algorithmes proposés sur des cas réels.\\
-The literature on the subject is quite recent and finds its origin in \cite{russell1998learning}.
-Different approaches have been explored since the first articles on the matter \cite{ng2000algorithms} and \cite{abbeel2004apprenticeship}.\\
-In this literature (see \cite{neu2009training} for a survey), several difficulties are identified but often ignored in order to find algorithms.
-Specifically, we often suppose the dynamic of the environment where the expert and the agent are evolving known, or that we dispose of an environment model (via a simulator) in order to test the effects of a policy.
-Moreover, these approaches resolve the reinforcement learning problem (find an optimal policy knowing a given reward) an important number of times in order to resolve the inverse reinforcement problem (find a reward explaining the expert policy).
-Finally, some of these approaches give a policy that performs as well as the expert with respect to the unknown reward function but do not give a reward able to explain the expert policy.
-These constraints can be incompatible with the application of the proposed algorithms in real cases.
+%
+%
+The literature on the subject is quite recent and finds its origin
+in \cite{russell1998learning}. Different approaches have been
+explored since the first articles on the matter
+\cite{ng2000algorithms,abbeel2004apprenticeship}.
+
+In this literature (see \cite{neu2009training} for a survey),
+several bottlenecks are identified but often ignored when developing
+algorithms. Specifically, the dynamic of the environment where the
+expert and the agent are evolving is assumed to be known, or an
+environment model (via a simulator) must be available so as to test
+the effects of a policy. Moreover, most of existing approaches
+require solving the direct reinforcement learning problem (find an
+optimal policy knowing a given reward) several times. Finally, in
+many cases, the output of the algorithm is a policy but not a unique
+reward function which makes task transfer hard to realize. These
+constraints can lead to the development of algorithms unapplicable
+to real problems.
 %Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert.
@@
 Le choix de cette application se justifie par la difficulté  qu'elle pose pour les méthodes de résolution de la littérature.
-In this paper, we propose a resolution method of the inverse reinforcement problem that is free from any other knowledge than the demonstrations given by the expert.
-We propose a proof of the convergence of our algorithm and give an experimentation on the inverted pendulum problem.
-The choice of this application is justified by the fact that it is difficult to resolve it by the other methods in the literature.
+%
+In this paper, we propose a resolution method of the inverse
+reinforcement problem that is free from any other knowledge than the
+demonstrations given by the expert. The proposed approach, the
+Structured Classification for IRL (SCIRL), relies on a
+classification paradigm in which the structure of the environment's
+dynamics is implicitly introduced. Sec.~\ref{back.sec}
+and~\ref{biblio.sec} provide the necessary background and review a
+part of the state of the art. Sec.~\ref{sec:scirl} introduces our
+contribution and Sec.~\ref{proof.sec} analyses the asymptotic
+algorithm. Section~\ref{exp.sec} shows results obtained on two
+classical benchmarks: inverted pendulum and the highway driving
+problems.
+%
+%A proof of the convergence of the algorithm is provided in
+%Section~\ref{proof.sec} and Section~\ref{exp.sec} shows results
+%obtained on two classical benchmarks: inverted pendulum and the
+%highway driving problems. The choice of the first application is
+%justified by the fact that it is hard to solve by the other methods.
+
 %%%%
-\section{Preliminaries}
+\section{Background}
 %%%%%
@@
 %%%
-\subsection{Reinforcement Learning (RL) and Inverse Reinforcement Learning (IRL)}
+Let's start with some background on RL and IRL and settle the
+notations.
+
+\subsection{(Inverse) Reinforcement Learning}
 %%%
@@
 Une politique de contrôle déterministe $\pi: S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
-The general framework chosen is the sequential decision making problem posed in the Markov decision process (MDP) formalism.
-The configuration of the system is completely described at each discrete time $t$ by the state $s_t \in S$ where the agent must take a decision $a_t\in A$.
-Then, the system will evolve to the next state $s_{t+1} \in S$ with the probability $p(s_{t+1}|s_t, a_t)$.
-A deterministic policy which is a function $\pi: S\rightarrow A$ allows us to define the behavior of an agent faced to a sequential decision making problem.
-In this paper $S$ and $A$ will be finite sets and $\{P_{sa}\}_{s\in S,a\in A}$ will be a set of probability distributions, from $S$ to $[0,1]$ such that $P_{sa}(s')=p(s'|s, a)$, called transition probabilities.
+
+The problem is seen as a sequential decision making problem posed in
+the Markov Decision Process (MDP) formalism. In this formalism, a
+system to be controlled can reach different configurations
+completely described at each discrete time $t$ by a state $s_t \in
+S$. Given the knowledge of the state, the agent must take a decision
+$a_t\in A$. Then, the system evolves to a next state $s_{t+1} \in S$
+with the Markovian probability $p(s_{t+1}|s_t, a_t)$. %(in short
+%$p(s'|s,a)$ where $s'=s_{t+1}, s=s_t$ and $a=a_t$).
+A policy, which
+is a function $\pi: S\rightarrow A$, defines the behavior of the
+agent. In this paper all the policies considered are deterministic,
+$S$ and $A$ will be finite sets and $\{P_{sa}\}_{s\in S,a\in A}$
+will be a set of probability distributions, from $S$ to $[0,1]$ such
+that $P_{sa}(s')=p(s'|s, a)$, called transition probabilities.
+%
 %Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R: S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
@@
 On définit pour ce faire la fonction de valeur $V^\pi$:
-In the RL paradigm, the choice of a policy $\pi$ depends on a reward function $R: S \rightarrow \mathbb{R}$ which defines the immediate reward $R(s)$ given to the system when it attains the state $s \in S$.
-In this paradigm, the agent, following a policy $\pi$, is not going to learn to maximize the immediate reward but a criterion taking into account the futures rewards reached by the agent.
-That allows to specify the goal without specifying a way to reach it.
-The criterion commonly chosen is the value function $V^\pi$:
-\begin{equation}
-\label{Vdef.eqn}
-V^\pi(s) = E^\pi_s[\sum_{t\geq0}\gamma^tR(s_t)]=R(s) + \gamma\sum_{s'\in S}P_{s\pi(s)} V^\pi(s'),
-\end{equation}
+%
+The RL paradigm aims at solving the problem of finding the optimal
+sequence of actions in an MDP. The optimality criterion is based on
+the definition of a reward function $R: S \rightarrow \mathbb{R}$
+which defines the immediate reward $R(s)$ given to the agent when
+the system is in the state $s \in S$. The agent should not learn to
+maximize the immediate reward but a criterion taking into account
+future rewards. The criterion commonly chosen is the value function
+$V^\pi$:
+%
+\begin{align}
+\label{Vdef.eqn} V^\pi(s) & = E^\pi_s[\sum_{t\geq0}\gamma^tR(s_t)]\\
+& = R(s) + \gamma\sum_{s'\in S}P_{s\pi(s)}(s') V^\pi(s'),
+\end{align}
 %où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action:
-where $\gamma \in [0,1)$ is a discount factor, and $E^\pi_s$ is the expectation over the distribution of the state sequences $(s_0,s_1,\dots)$ we pass through when we execute the policy $\pi$ starting from $s_0=s$.
-It is also possible to define the action-value function $Q^\pi$ which adds a degree of freedom in the choice of the first action:
-\begin{equation}
-\label{Qdef.eqn}
-Q^\pi(s,a) = E^\pi_{sa}[\sum_{t\geq0}\gamma^tR(s_t)]=R(s) + \gamma\sum_{s'\in S}P_{sa} V^\pi(s'),
-\end{equation}
-where $E^\pi_{sa}$ is the expectation over the distribution of the state sequences $(s_0,s_1,\dots)$ we pass through when we start from $s_0=s$, choose the action $a_0=a$ and then follow the policy $\pi$.
+%
+where $\gamma \in [0,1)$ is a discount factor, and $E^\pi_s$ is the
+expectation over the distribution of the state sequences
+$(s_0,s_1,\dots)$ the agent passes through when executing the policy
+$\pi$ starting from $s_0=s$. One can define the action-value
+function $Q^\pi$ which adds a degree of freedom in the choice of the
+first action:
+%
+\begin{align}
+\label{Qdef.eqn} Q^\pi(s,a) & =
+E^\pi_{sa}[\sum_{t\geq0}\gamma^tR(s_t)]\\
+& = R(s) + \gamma\sum_{s'\in S}P_{sa}(s') V^\pi(s').
+\end{align}
+%
+%where $E^\pi_{sa}$ is the expectation over the distribution of the
+%state sequences $(s_0,s_1,\dots)$ we pass through when we start from
+%$s_0=s$, choose the action $a_0=a$ and then follow the policy $\pi$.
+%
 %Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
 La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton:
-An optimal policy $\pi^*$ is defined as a policy which value function $V^{\pi*}=V^*=\sup_{\pi}V^{\pi}$ which means $\forall \pi, \forall s, V^{\pi*}(s) \geq V^\pi(s)$.
-An optimal policy $\pi^*$ can be computed easily thanks to the optimal action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy mechanism:
+%
+An optimal policy $\pi^*$ is defined as a policy with maximal value
+function, $V^{\pi*}=V^*=\sup_{\pi}V^{\pi}$, which means that
+$\forall \pi, \forall s, V^{\pi*}(s) \geq V^\pi(s)$. An optimal
+policy $\pi^*$ can be computed easily thanks to the optimal
+action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy
+mechanism:
+%
 \begin{equation}
-\label{greedy.eqn}
-\pi^*(s) \in \arg\max_{a\in A} Q^*(s,a).
+\label{greedy.eqn} \pi^*(s) \in \arg\max_{a\in A} Q^*(s,a).
 \end{equation}
 %L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov .\\
-The tuple $M=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,R\}$ containing the finite state space $S$, the finite action space $A$, the transitions probabilities $\{P_{sa}\}_{s\in S,a\in A}$, the discount factor $\gamma$ and the reward function $R$ is called a finite Markov decision process.
-Resolving the RL problem with respect to the criterion $V^\pi$  means to find $\pi^*$.
-In a finite MDP the existence of an optimal deterministic policy is guaranteed when the reward function $R$ is bounded.\\
+%
+The tuple $M=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,R\}$ is called
+a finite MDP. Resolving the RL problem with respect to the criterion
+$V^\pi$  means to find $\pi^*$, usually without the full knowledge
+of the MDP but only by observing transitions and rewards. In a
+finite MDP the existence of an optimal deterministic policy is
+guaranteed when the reward function $R$ is bounded.
+%
 %Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive.
@@
 Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
-Sometimes, specifying manually the reward function can be very difficult while it is natural to an operator to control the system in order to complete the desired task.
-An exemple of this kind of task is driving a car.
-We can accomplish this task very easily in our daily-life but it would be very tricky to specify the precise weights given to each important driving-parameters such as the distance we should keep between the car in front of us, or maintaining a reasonable speed and so on.
-In such cases, it is useful to infer the reward function thanks to observed behavior.\\
+%
+The optimal behaviour being defined relatively to the reward, its
+observation is mandatory to learn. Sometimes, specifying manually
+the reward function can be a hard job while it is natural to an
+operator to achieve the desired task. For instance, driving a car is
+such a task. One can accomplish this task very easily in his
+daily-life but it is very tricky to specify the precise weights
+given to each important driving-parameters such as the distance to
+keep with the leading car, the speed below which it is unreasonable
+to drive and so on. In such cases, inferring the reward function
+from observed behavior would be helpful.
+%
 %C'est la définition de l'apprentissage par renforcement inverse.
@@
 Usuellement l'expert est considéré comme un agent optimal dans un PDM et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
-That is the aim of the Inverse Reinforcement Learning.
-The problem is to find the reward function optimized by an expert.
-Usually, the expert is considered as an optimal agent in a MDP and we have access to the expert optimal policy $\pi_E$ or to some trajectories of this policy.
+%
+It is the aim of IRL. The expert, demonstrating the task, is thus
+considered as an optimal agent in a MDP and the IRL agent has access
+to the expert optimal policy $\pi_E$ or to some trajectories drawn
+from this policy.
+%
 %Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}).
@@
 Il convient donc de contraindre l'espace des solutions.
-This problem is ill-posed in the sense that there is not uniqueness of the reward function for which the expert policy is optimal (\cite{ng1999policy}).
-Particularly, each policy is optimal with respect to the uniformly-zero reward function.
-One should favor some solutions.
+The problem of finding a reward function from expert demonstrations
+in the IRL framework is ill-posed. Indeed, there is not uniqueness
+of the reward function for which the expert policy is
+optimal~\cite{ng1999policy}. Particularly, each policy is
+optimal with respect to the uniformly-zero reward function. Some
+solutions have to be favored.
 %%%
@@
 %La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$  telles que
-The reward function $R$ is the unknown of the problem.
-We can suppose that it is possible to dispose of $p$ features via $\psi: S\rightarrow \mathbb{R}^p$, which is a functions from $S$ to $\mathbb{R}^p$ and each of the components of $\psi$: $\{\psi_i\}_{1\leq i \leq p}$ is one feature, such that:
+The actual expert reward function $R_E$ is unknown. The problem is
+ thus to find a reward function $R_\theta$ for which the policy of
+the expert is also optimal. Let's assume that, given a set of $p$
+features $\psi: S\rightarrow \mathbb{R}^p$, each components of
+$\psi$: $\{\psi_i\}_{1\leq i \leq p}$ being a basis function from
+$S$ to $\mathbb{R}$, such a reward can be written:
 \begin{equation}
-\label{hatRdef.eqn}
-\exists \theta | R(s) = \theta^T\psi(s) = \sum_{i=1}^p\theta_i\psi_i(s).
+\label{hatRdef.eqn} %\exists \theta |
+R_\theta(s) = \theta^T\psi(s) =
+\sum_{i=1}^p\theta_i\psi_i(s).
 \end{equation}
@@
 \eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants:
-This constraint can be relaxed if necessary.
-Putting this expression of the reward function in the definition of the action-value function (Eq.
-\eqref{Qdef.eqn}) allows us to define the term $\mu^\pi$ which is often used in the existing algorithms:
+Using this expression allows rewriting (Eq. \eqref{Qdef.eqn}) as:
 \begin{equation}
-Q^\pi(s,a)=\theta^T\mu^\pi(s,a),\text{ }\mu^\pi(s,a) = E^\pi_{sa}[\sum_t\gamma^t\psi(s_t)].
-\label{Qmu.eqn}
+Q^\pi(s,a)=\theta^T\mu^\pi(s,a),\text{ }\mu^\pi(s,a) =
+E^\pi_{sa}[\sum_{t\geq 0}\gamma^t\psi(s_t)]. \label{Qmu.eqn}
 \end{equation}
@@
 On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
-The term $\mu^\pi$ is called the feature expectation of the policy $\pi$.
-We can observe that the chosen parametrization of the reward function $R$ and the dynamic resulting from the application of the policy $\pi$ in the MDP impose the parametrization of $Q^\pi$.
-To shorten notations, we may use $\mu^\pi_{sa}$ as $\mu^\pi(s,a)$.\\
+%
+The term $\mu^\pi$ is called the feature expectation of the policy
+$\pi$. We can observe that the chosen parametrization of the reward
+function $R_\theta$ and the dynamics resulting from the application
+of the policy $\pi$ in the MDP gives rise to an induced
+parametrization for $Q^\pi$. To
+shorten notations, we may use $\mu^\pi_{sa}$ as $\mu^\pi(s,a)$.\\
+%
 %On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
 En effet
-We see also that if two policies share the same feature expectation, then these two policies have the same value function with respect to the reward function $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$.
-Indeed: \begin{equation}
-\mu^{\pi_1} = \mu^{\pi_2} \Rightarrow \theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} = V^{\pi_2}.
-\label{memevaleur.eqn}
+We see also note that if two policies share the same feature
+expectation, then they have the same value function with respect to
+any reward function $R_\theta=\theta^T\psi$,
+$\theta\in\mathbb{R}^p$. Indeed:
+%
+\begin{equation}  \mu^{\pi_1} = \mu^{\pi_2} \Rightarrow
+\theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} =
+V^{\pi_2}. \label{memevaleur.eqn}
 \end{equation}
+%
 %L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
@@
 Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert, quelle que soit la récompense (voir par exemple \cite{neu2009training}).
-That is why most of the algorithms in the literature have, as we will see, the objective of minimizing a \textit{distance}  between the feature expectation of the expert and the feature expectation of the apprentice.
-That allows to obtain an apprentice which policy has the same value function as the expert policy, for a reward function of the type $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$ (see for instance \cite{neu2009training}).
+%
+That is why most of the algorithms in the literature have, as we
+will see, the objective of minimizing a \textit{distance}  between
+the feature expectation of the expert and the feature expectation of
+the apprentice. %That allows to obtain an apprentice which policy has
+%the same value function as the expert policy, for a reward function
+%of the type $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$ (see for
+%instance
+%~\cite{neu2009training}.
+
+\section{State of the Art}
+\label{biblio.sec}
+
+Let $\pi_\theta$ denote an optimal policy according to a reward
+function $R_\theta(s)=\theta^T\psi(s)$, for some
+$\theta\in\mathbb{R}^p$. With a slight abuse of notation, let us
+write $\mu^\pi_{d_0}$ the mean feature expectation for a
+distribution $d_0$ over starting states:
+$\mu^\pi_{d_0}=E[\mu^\pi(s,\pi(s))|s\sim d_0]$. The classical
+approach to IRL (or less generally to apprenticeship learning, as
+the output of the algorithm may be a policy and not a reward
+function), initiated in~\cite{abbeel2004apprenticeship}, consists in
+finding a reward (a parameter vector) $\theta^*$ minimizing some
+loss function between the feature expectation of the expert and the
+one of the related optimal policy:
+\begin{equation}
+  \theta^* = \argmin_\theta J(\theta) \text{ with } J(\theta)=
+  L(\mu^{\pi_\theta}_{d_0},\mu^E_{d_0}).\nonumber
+\end{equation}
+Depending on the adopted point of view,
+game-theoretic-base~\cite{syed2008game}, linear
+programming-based~\cite{syed2008apprenticeship}, maximum
+entropy-based~\cite{ziebart2008maximum}, \textit{etc.}, the loss
+function and the associated update rule differ. Most of these
+approches and others are nicely reviewed in~\cite{neu2009training}.
+
+In all cases, the solution $\theta^*$ is computed in an iterative
+way. At each step, one has to compute the feature expectation
+$\mu^{\pi_\theta}_{d_0}$ for some parameter vector $\theta$. This
+implies computing $\pi_\theta$, the optimal policy according to the
+reward $\theta^T\psi(s)$. Therefore, solving the IRL problem with
+one of these methods involves solving repeatedly the direct RL
+problem (some of these algorithms requiring also the model to be
+known). This is a drawback we wish to get rid of.
+
+A notable exception is~\cite{boularias2011relative}. This work,
+based on a relative entropy argument, finds $\theta^*$ by performing
+a subgradient ascent on some utility function. Estimating the
+subgradient would require to sample trajectories according to
+$\pi_\theta$, with $\theta$ being the current estimate, thus would
+require to solve the direct problem. However, this can be avoided in
+this specific framework by the use of importance sampling. The
+gradient is estimated from trajectories sampled from some random
+policy. Still, this may be a problem when the only available data
+comes from the expert.
+
+Another approach consists in casting imitation learning into a
+multi-class classification problem, introducing in some way the
+structure of the MDP. For example, in~\cite{melo2010learning}, the
+agent policy is learnt as the solution of a classification problem.
+They consider a kernel function built on some induced MDP metric,
+which may be combersome to compute. Closest to our proposed
+contribution is the Maximum Margin Planning (MMP)
+algorithm~\cite{ratliff2006maximum}, which is a classification-based
+algorithm associating optimal policies to MDPs. More precisely, the
+training set is made of $N$ couples $(\mathcal{M}_i,\pi^*_i)$, with
+$\mathcal{M}_i$ an MDP (without reward) and $\pi_i^*$ the associated
+optimal policy. Let $\mu^\pi_{d_0,\mathcal{M}_i}$ be the feature
+expectation for the policy $\pi$, the initial distribution $d_0$ and
+the dynamics induced by the related MDP. Let also
+$\mathcal{L}_{\mathcal{M}_i,\pi}$ be a positive and user-defined
+margin function penalizing non-optimal policies. MMP minimizes the
+following cost function:
+\begin{equation}
+  J(\theta) = \frac{1}{N}\sum_{1=1}^N \max_\pi(\theta^T
+  \mu^\pi_{d_0,\mathcal{M}_i} + \mathcal{L}_{\mathcal{M}_i,\pi}) - \theta^T
+  \mu^{\pi_i^*}_{d_0,\mathcal{M}_i}.\nonumber
+\end{equation}
+This implies solving a set of MDPs at each iteration. At the other
+extreme, a similar cost function is considered
+in~\cite{ratliff2007imitation}, which searches for a score function
+$q$ associating expert actions to states. The training set being $N$
+couples $(s_i,a_i=\pi^E(s_i))$ and $l(s,a)$ being a positive
+user-defined margin function penalizing non-optimal actions, the
+cost function is
+\begin{equation}
+  J(q) = \frac{1}{N}\sum_{i=1}^N \max_a (q(s_i,a) +
+  l(s_i,a))-q(s_i,a_i).\nonumber
+\end{equation}
+The drawback of this approach is that it does not take into account
+the structure of the underlying MDP, which may easily be shown
+experimentally to be a problem (\textit{e.g.},
+see~\cite{melo2010learning}).
+
+The proposed contribution is a trade-off between the two previous
+approaches. It works at the state level rather than the MDP level,
+considering a similar cost function. However, it takes into account
+the structure of the MDP. As shown in the next section, this allows
+avoiding to solve the direct RL problem as intermediate steps and it
+requires only data from the expert.
+
 %%%%%
 \section{Structured classification for IRL}
+\label{sec:scirl}
 %%%%%
+
+This section introduces our contribution, the SCIRL (Structured
+Classification for IRL), and discusses one of its key components,
+the estimation of the expert's feature expectation.
+
 %%%
 \subsection{Principle}
-%%%
-%Comme nous l'avons vu Eq.
+
+First, let us review more deeply the approach
+of~\cite{ratliff2007imitation}, to which we will actually add some
+structure. Assume that we have a set of transitions
+$\{(s_i,a_i,s_{i+1})_{1\leq i\leq N}\}$ coming from one or several
+trajectories of the expert. A classic approach for generalizing the
+expert policy (from a multi-class classification-based point of
+view) would consist in finding some score function $q:S\rightarrow
+A$ such that for all $a\neq a_i$, $q(s_i,a_i)>q(s_i,a)$. In this
+framework, states are inputs and actions are labels. Moreover, one
+may want to strengthen the optimality of the expert. Let $l:S\times
+A\rightarrow\mathbb{R}_+$ be a user-defined margin function, which
+allows to put some prior knowledge in the problem at hand.
+Typically, this margin function should satisfy $l(s_i,a_i)\ll
+l(s_i,a)$. For example, one may choose $l(s_i,a\neq a_i)=1$ and
+$l(s_i,a_i)=0$. Then, one searches for a score function $q$
+satisfying
+\begin{equation}
+\label{eq:inequalities_score}
+  \forall a,\; q(s_i,a_i) + l(s_i,a_i) \geq q(s,a) + l(s,a).\nonumber
+\end{equation}
+In order to find such a score function, a natural empirical cost
+function to be minimized is:
+\begin{equation}
+\label{Jdef.eqn} J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in
+A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
+\end{equation}
+This has been originally introduced in~\cite{ratliff2007imitation}.
+Unfortunately, it ignores a large part of data: from the transitions
+$(s_i,a_i,s'_i)$ performed by the expert, only the state action
+couple $(s_i,a_i)$ is used and the dynamic part (transiting to
+$s'_i$ from this state action couple) is ignored.
+%
+Moreover, it outputs a policy, whereas we are interested in the
+reward function, which is a much more compact and transferable
+representation of the task at hand.
+
+Recall that we work in a sequential decision making framework.
+Therefore, there is a natural solution to optimization
+problem~\eqref{Jdef.eqn}: the optimal $Q$-function according to the
+unknown reward function. Now, let $\theta$ be some parameter vector.
+The score function $q(s,a) = \theta^T \mu^E(s,a)$ is actually the
+$Q$-function of the expert policy according to the reward
+$\theta^T\psi(s)$. Moreover, assume that this parameter vector is
+such that inequalities~\eqref{eq:inequalities_score} are satisfied.
+The optimal $Q$-function being stable by the greedy mechanism,
+see~Eq.\eqref{greedy.eqn}, this means that the expert policy $\pi_E$
+is optimal respectively to the reward function $\theta^T\psi(s)$
+(and $\theta^T\mu^E(s,a)$ is the associated optimal $Q$-function). A
+natural cost function for computing a reward function explaining the
+expert behavior is therefore the following one:
+\begin{equation}
+\label{structure.eqn} J_N(\theta)  = {1\over N}
+\sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) -
+\theta ^T \mu^E_{s_ia_i} \right).
+\end{equation}
+Still, we need to minimize it.
+
+%%%%
+%%Comme nous l'avons vu Eq.
 \eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue: $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
-%Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
-%Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
+%%Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+%%Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
 Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons:
-As seen in Eq.
-\eqref{greedy.eqn}, there exists a greedy mechanism linking the expert policy $\pi_E$ and the optimal action-value $Q^E_R$ relative to the unknown reward function $R$: $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
-We will use this property to consider an empirical risk function $J_N$ dependent on a score function $q$ which will have to share with the unknown action-value function $Q^E_R$ the same optimal property.
-Let's note $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ the samples coming from one or several  trajectories of the expert.
-We are searching for a score function $q$ which verifies for each sample:
-\begin{equation}
-\label{butLAFEM.eqn}
-a_i \in \argmax_{a\in A}q(s_i,a).
-\end{equation}
-%Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+%As seen in Eq. \eqref{greedy.eqn}, there exists a greedy mechanism
+%linking the expert policy $\pi_E$ and the optimal action-value
+%$Q^E_R$ relative to the unknown reward function $R$: $\pi_E(s) \in
+%\arg\max_{a\in A}Q^E_R(s,a)$.
+%
+%We will use this property to consider an empirical risk function
+%$J_N$ dependent on a score function $q$ which will have to share
+%with the unknown action-value function $Q^E_R$ the same optimal
+%property. Let's note $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ the samples
+%coming from one or several  trajectories of the expert. We are
+%searching for a score function $q$ which verifies for each sample:
+%\begin{equation}
+%\label{butLAFEM.eqn}
+%a_i \in \argmax_{a\in A}q(s_i,a).
+%\end{equation}
+%%Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
 Nous introduisons de surcroit une fonction $l: S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
@@
 Nous introduisons ainsi la fonction de risque empirique suivante:
-So, we want to introduce a risk function which should penalize the cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases the score function $q$ does not allow to explain the choice $a_i$ of the expert (see equation \eqref{greedy.eqn}).
-Moreover, we introduce a loss function $l: S \times A \rightarrow \mathbb{R}_+$; this function allows the user to put some \emph{a priori} knowledge in the system because the user will choose the gap wanted in the score function between the action chosen by the expert and the second best possible action as we can see in the following equation:
-$\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$. Introducing a loss function is quite common in the literature (see Sec. \ref{biblio.sec}). We introduce then the following empirical risk function:
-\begin{equation}
-\label{Jdef.eqn}
-J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
-\end{equation}
-%Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
-%
-%Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
+%So, we want to introduce a risk function which should penalize the
+%cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases
+%the score function $q$ does not allow to explain the choice $a_i$ of
+%the expert (see equation \eqref{greedy.eqn}). Moreover, we introduce
+%a margin function $l: S \times A \rightarrow \mathbb{R}_+$; this
+%function allows the user to put some \emph{a priori} knowledge in
+%the system because the user will choose the gap wanted in the score
+%function between the action chosen by the expert and the second best
+%possible action as we can see in the following equation: $\forall a,
+%q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$. Introducing a margin function
+%is quite common in the literature (see Sec. \ref{biblio.sec}). We
+%introduce then the following empirical risk function:
+%\begin{equation}
+%\label{Jdef.eqn}
+%J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
+%\end{equation}
+%%Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
+%%
+%%Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
 Notre propos n'est cependant pas d'utiliser une approche purement supervisée.
@@
 Il faut pour cela prendre en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnées par la politique de l'expert.\\
-%
-%Comme nous l'avons vu Eq.
+%%
+%%Comme nous l'avons vu Eq.
 \eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi: S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
@@
 \eqref{hatRdef.eqn}):
-By default, choosing $l(s_i,a) = 1$ if $a\neq a_i$ and $l(s_i,a_i)=0$ works quite well as shown in the experimentations (Sec. \ref{exp.sec}) and in the analysis (Sec. \ref{proof.sec}).\\
-Here, we deal with the imitation problem as a classification problem.
-However, our purpose is not to use a purely supervised approach.
-We do not want to learn the policy of the expert but rather the description of the task he is executing (the reward function).
-In order to do that, it is necessary to take into account the MDP structure, and in particular the information on the transitions between states which are conditioned by the expert policy.\\
-As seen in Eq.
-\eqref{Qmu.eqn}, the hypothesis of a linear representation of the reward function by some features $\psi: S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) = \theta^T\mu^\pi(s,a)$.
-This parametrization takes into account the dynamic imposed by $\pi_E$.
-The expert follows the same rule and if we note $\mu^E$ the feature expectation of the expert, we can do the hypothesis that the score function $q$ that we are searching for can be written as follows: $q(s,a) = \theta^T\mu^E(s,a)$.
-Estimating the term $\mu^E$ is the key of our algorithm and we propose some solution in Sec. \ref{calculmu.sec}.
-We can now give the expression of our empirical risk function as a function of the parameter $\theta$ which defines the reward function (Eq. \eqref{hatRdef.eqn}):
-\begin{equation}
-\label{structure.eqn}
-J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) - \theta ^T \mu^E_{s_ia_i} \right)
-\end{equation}
-%Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
+%By default, choosing $l(s_i,a) = 1$ if $a\neq a_i$ and
+%$l(s_i,a_i)=0$ works quite well as shown in the experimentations
+%(Sec. \ref{exp.sec}) and in the analysis (Sec. \ref{proof.sec}).
+%
+%Here, we deal with the imitation problem as a classification
+%problem. However, our purpose is not to use a purely supervised
+%approach. We do not want to learn the policy of the expert but
+%rather the description of the task he is executing (the reward
+%function). In order to do that, it is necessary to take into account
+%the MDP structure, and in particular the information on the
+%transitions between states which are conditioned by the expert
+%policy.
+
+%As seen in Eq. \eqref{Qmu.eqn}, the hypothesis of a linear
+%representation of the reward function by some features $\psi:
+%S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) =
+%\theta^T\mu^\pi(s,a)$. This parametrization takes into account the
+%dynamic imposed by $\pi_E$. The expert follows the same rule and if
+%we note $\mu^E$ the feature expectation of the expert, we can do the
+%hypothesis that the score function $q$ that we are searching for can
+%be written as follows: $q(s,a) = \theta^T\mu^E(s,a)$. Estimating the
+%term $\mu^E$ is the key of our algorithm and we propose some
+%solution in Sec. \ref{calculmu.sec}. We can now give the expression
+%of our empirical risk function as a function of the parameter
+%$\theta$ which defines the reward function (Eq.
+%\eqref{hatRdef.eqn}):
+%\begin{equation}
+%\label{structure.eqn}
+%J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) - \theta ^T \mu^E_{s_ia_i} \right)
+%\end{equation}
+%%Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
 Nous utilisons une descente de sous-gradient (normalisé), une généralisation du gradient, pour le minimiser.
@@
 En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée:
-This risk function is non linear and non differentiable everywhere because of the operator $\max$.
-But this function is clearly convex, closed, proper, continuous and piecewise-linear with finite values as $A$ is finite.
-We can then use a sub-gradient descent algorithms, which is a generalization of a gradient descent algorithm, in order to minimize the risk function.
-A sub-gradient of a convex function
-$f:\mathbb{R}^p\rightarrow \mathbb{R}$ in $x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$, the set of all sub-gradients in $x$ is $\partial_xf$.
-The unique sub-gradient in $x$ of a differentiable convex function $f$ is the gradient $\nabla_xf$.
-As $A$ is finite, the convex and piecewise differentiable function $\max_{a\in A}[f(\cdot,a)]$, from $\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from $\mathbb{R}^p$ to $\mathbb{R}$,
-has one of its sub-gradient in $x\in\mathbb{R}^p$ equals to $\nabla_xf(x,a^*)$ with $a^*\in\arg\max_{a\in A}f(x,a)$.
-Applying these rules, we have a sub-gradient of our risk function and an update rule for our sub-gradient algorithm:
-\begin{eqnarray}
-\label{algorithm}
-g_{\theta} &=& {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right),\\
-a^*_i &\in& \arg\max_{a\in A}(\theta ^T \mu^E(s_i,a) + l(s_i,a)),\\
-g_{\theta} &\in& \partial_\theta J_N,\\
-\theta_{0}&=&0,\theta_{t+1}=\theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2},\forall t\in\mathbb{N}^*,\\
-\sum_{t\geq0}\alpha_t&>&+\infty,\text{ }\sum_{t\geq0}\alpha_t^2<+\infty.
-\end{eqnarray}
-%%%
+
+Because of the $\max$ operator, this cost function is not linear nor
+differentiable everywhere. But it is clearly convex and piecewise
+linear.
+%
+We can then use a subgradient descent algorithm. A subgradient of a
+convex function $f:\mathbb{R}^p\rightarrow \mathbb{R}$ in
+$x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall
+x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$. The set of all
+subgradients in $x$ is noted $\partial_xf$. The unique subgradient
+in $x$ of a differentiable convex function $f$ is the gradient
+$\nabla_xf$. As $A$ is finite, the convex and piecewise
+differentiable function $\max_{a\in A}[f(\cdot,a)]$, from
+$\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex
+differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from
+$\mathbb{R}^p$ to $\mathbb{R}$, has one of its subgradient in
+$x\in\mathbb{R}^p$ equal to $\nabla_xf(x,a^*)$ with
+$a^*\in\arg\max_{a\in A}f(x,a)$. Thanks to this, we can minimize the
+cost function of interest with the following update rule ($\alpha_i$
+denoting a usual learning rate):% and $\theta_0$ an initial parameter vector):
+\begin{align}
+  \theta_{t+1}&=\theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2}
+  \\
+  \text{with }
+  g_{\theta} &= {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right)\in \partial_\theta J_N
+  \\
+  \text{and } a^*_i &\in \arg\max_{a\in A}(\theta_t^T \mu^E(s_i,a) +
+  l(s_i,a)).
+\end{align}
+The SCIRL algorithm only requires data from the expert, assuming
+$\mu^E$ is known on $\{(s_i,a)_{1\leq i \leq N,a\in A}\}$.
+Unfortunately, this is usually not the case.
+
 \subsection{Estimation of $\mu^E$}
@@
 \label{calculmu.sec}
-%Comme signalé dans la Sec.
+
+Estimating the feature expectation of the expert is a key component
+of our algorithm. We provide three possible solutions, depending on
+the available data.
+
+First, notice that each component $\mu_j^E$ of the expert feature
+expectation $\mu^E$ is actually the $Q$-function of the expert
+policy according to the reward function $\psi_j$, the $j^\text{th}$
+basis function of the considered hypothesis space:
+\begin{equation}
+\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t
+\psi_j(s)].\label{eq:muE_bellman}
+\end{equation}
+Therefore, it satisfies the Bellman evaluation
+equation~\eqref{Qdef.eqn}. Assuming that the expert policy and the
+transition probabilities are known, the expert feature expectation
+can be exactly computed.
+
+However, this assumption is rather strong (even if it is often done
+in the IRL literature). We may assume that we can simulate
+trajectories according to the expert policy on demand. In this case,
+one may estimate $\mu^E(s,a)$ for a given state-action couple with
+Monte Carlo. Assume that we have a set of $N$ trajectories of length
+$H$, $\{s^i_0,a^i_0,\dots,s^i_{H-1},a^i_{H-1},s^i_{H}\}_{1\leq i
+\leq N}$, with $s^i_0 =s$ and $a^i_0=a$, sampled according to the
+system dynamic and the expert policy. Then $\mu(s,a)$ is estimated
+as:
+\begin{equation}
+\hat\mu^E(s,a) = {1\over
+N}\sum_{i=1}^{N}\sum^{H}_{t=0}\gamma^t\psi(s^i_t),
+\end{equation}
+
+Yet, as claimed at the beginning of this paper, we would like to
+solely use the least amount of available data, that is is the set of
+transitions $\{(s_i,a_i,s'_{i})_{1\leq i\leq N}\}$ from the expert.
+As each component of the feature expectation can be seen as a
+$Q$-function, see Eq.~\eqref{eq:muE_bellman}, it can be estimated
+solely from the available data using the on-policy
+LSTD~algorithm~\cite{bradtke1996linear}. This is the basic idea of
+the LSTD$\mu$ approach~\cite{klein2011batch}, which is an LSTD-like
+method estimating feature expectations. As a part of the
+computations can be factored, LSTD and LSTD$\mu$ computational costs
+have the same order of magnitude. We use LSTD$\mu$ in the
+experimental section, which is the less constraining scheme.
+
+Other schemes may be envisioned. For example, one may use a Monte
+Carlo estimate for $\mu^E(s_i,a_i)$ (using only the available data)
+and some heuristic for $\mu^E(s_i,a\neq a_i)$. We left this for
+future work.
+
+%%Comme signalé dans la Sec.
 \ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
@@
 Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
-As mentioned in \ref{ConsiderationsTechniques.sec}, some algorithms in the literature depend on the implementation of the feature expectation.
-Our algorithm, in particular, relies only on the implementation of $\mu^E$, the feature expectation of the expert, and does not require the repeated calculus of the feature expectation of another policy.
-For our algorithm, we only need to know $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$.
-That implies to infer the feature expectation of the expert on states where he has been, but on actions he has not taken.
-%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve:
-This inference is made easier by the fact that the feature expectation obeys the Bellman equation. Indeed, when we write its definition for a component $1\leq j\leq p$, we have:
-\begin{equation}
-\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t \psi_j(s)].
-\end{equation}
-%expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
+%As mentioned in \ref{ConsiderationsTechniques.sec}, some algorithms
+%in the literature depend on the implementation of the feature
+%expectation. Our algorithm, in particular, relies only on the
+%implementation of $\mu^E$, the feature expectation of the expert,
+%and does not require the repeated calculus of the feature
+%expectation of another policy. For our algorithm, we only need to
+%know $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$. That implies
+%to infer the feature expectation of the expert on states where he
+%has been, but on actions he has not taken.
+%%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve:
+%This inference is made easier by the fact that the feature
+%expectation obeys the Bellman equation. Indeed, when we write its
+%definition for a component $1\leq j\leq p$, we have:
+%\begin{equation}
+%\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t \psi_j(s)].
+%\end{equation}
+%%expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
 Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu^E$.\\
-When we compare this expression to the equation \eqref{Qdef.eqn}, it allows us to affirm that $\mu_j^E$ can be seen as the action-value function for the policy $\pi_E$ with respect to the reward function defined by the $j$-th component of the features $\psi$.
-Depending on the prior information, several methods can be used in order to estimate $\mu^E$.\\
-
-%Dans le cas idéal où le modèle est connu, le calcul exact de $\mu^E$ peut être effectué grâce aux algorithmes de programmation dynamique.
+%When we compare this expression to the equation \eqref{Qdef.eqn}, it
+%allows us to affirm that $\mu_j^E$ can be seen as the action-value
+%function for the policy $\pi_E$ with respect to the reward function
+%defined by the $j$-th component of the features $\psi$. Depending on
+%the prior information, several methods can be used in order to
+%estimate $\mu^E$.\\
+%
+%%Dans le cas idéal où le modèle est connu, le calcul exact de $\mu^E$ peut être effectué grâce aux algorithmes de programmation dynamique.
 Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu^E(s,a)$ (définition Eq.
 \eqref{Qmu.eqn}) l'on utilise
-In the ideal case where a model of the environnement is known (\textit{id est} the transition probabilities are given), the exact calculus of $\mu^E$ can be executed thanks to dynamic programmation algorithms.
-In a more realistic case, we can use a simple Monte-Carlo estimation if we dispose of a simulator and if it is possible to ask the expert what action has to be chosen in an important number of arbitrary states.
-The estimation of $\mu^E(s,a)$ (definition Eq.
-\eqref{Qmu.eqn}) is done over $N$ trajectories of length $\{2k_i+1\}_{1\leq i \leq N}$, $\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i \leq N}$, by the following formula:
-\begin{equation}
-\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k_i}_{t=0}\gamma^t\psi(s^i_t),
-\end{equation}
-where $s^i_0=s, a^i_0=a$ and $s^i_{t+1}$ is distributed according to $P_{s^i_t\pi_E(s^i_t)}$.
+%In the ideal case where a model of the environnement is known
+%(\textit{id est} the transition probabilities are given), the exact
+%calculus of $\mu^E$ can be executed thanks to dynamic programmation
+%algorithms. In a more realistic case, we can use a simple
+%Monte-Carlo estimation if we dispose of a simulator and if it is
+%possible to ask the expert what action has to be chosen in an
+%important number of arbitrary states. The estimation of $\mu^E(s,a)$
+%(definition Eq. \eqref{Qmu.eqn}) is done over $N$ trajectories of
+%length $\{2k_i+1\}_{1\leq i \leq N}$,
+%$\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i
+%\leq N}$, by the following formula:
+%\begin{equation}
+%\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k_i}_{t=0}\gamma^t\psi(s^i_t),
+%\end{equation}
+%where $s^i_0=s, a^i_0=a$ and $s^i_{t+1}$ is distributed according to $P_{s^i_t\pi_E(s^i_t)}$.
 %Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
@@
 Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
-Finally, if we are in the framework where only the data at our disposition are the trajectories samples given by the expert, it is possible to use the LSTD$\mu$ algorithm \citep{klein2011batch} in its \emph{on-policy} version.
-This algorithm adapts the LSTD algorithm (\cite{bradtke1996linear}) keeping its characteristics, in particular its ability to execute the calculus in  {\it batch}-mode.
-This is the method that we show in the section \ref{exp.sec}.
-The principle is to estimate each component of $\mu^E$ using LSTD.
-Some calculus can be aggregated, giving to LSTD$\mu$ a cost of calculus which is of the same order of magnitude as LSTD.
-%D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
+%Finally, if we are in the framework where only the data at our
+%disposition are the trajectories samples given by the expert, it is
+%possible to use the LSTD$\mu$ algorithm \citep{klein2011batch} in
+%its \emph{on-policy} version. This algorithm adapts the LSTD
+%algorithm (\cite{bradtke1996linear}) keeping its characteristics, in
+%particular its ability to execute the calculus in  {\it batch}-mode.
+%This is the method that we show in the section \ref{exp.sec}. The
+%principle is to estimate each component of $\mu^E$ using LSTD. Some
+%calculus can be aggregated, giving to LSTD$\mu$ a cost of calculus
+%which is of the same order of magnitude as LSTD.
+%%D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
 Cela est reporté pour des travaux futurs.
-%%%%%
-Other estimation methods are possible, as for instance a Monte-Carlo estimation, less calculus-expensive and constraining than the first one, which would be based solely on the data given by the expert and would imply the use of an heuristic on the unknown actions.
-This estimation would be consider on future works.
-\section{Analysis}\label{proof.sec}
-In this section we are going to prove that our algorithm (see \eqref{algorithm}) is convergent, when we suppose the existence of a minimum of $J_N$,  which means that $\tilde{\theta}=\lim_{t\rightarrow+\infty}\theta_t$
-is well defined and that $\inf_{\theta\in\mathbb{R}^p}J_N(\theta)=J_N(\tilde{\theta})$ which means that the algorithm reaches the goal of minimizing $J_N$.
-Moreover we will determine the quality of our algorithm by giving an upper bound to the term $E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]$ where $E_{s\sim\nu}$ is the expectation
-over the stationary probability distribution of states $\nu$ and $\pi_A$ is the policy of the apprentice when a specific condition is fulfilled.
-\subsection{Convergence of the sub-gradient algorithm}
-$\mathbb{R}^p$ is an euclidian space with the canonical scalar product $<.,.>$ and the euclidian norm $\|\|_2$.
-Let $N\in\mathbb{N}^*$ and $A$ the finite set of actions of cardinal $\card{A}$.
-We dispose of the estimations  $\{\hat{\mu}^E_{ia}\}_{1\leq i \leq N,a\in A}$ where $\forall 1\leq i \leq N, \forall a\in A,\hat{\mu}^E_{ia}\in\mathbb{R}^p$.
-We define the loss function such that: $\forall 1\leq i \leq N,\forall a\in A\backslash{a_i}, l(s_i,a)>l(s_i,a_i)\geq0$.
-We have already defined the empirical risk function which is a function from $\mathbb{R}^p$ to $\mathbb{R}$:
-$\forall \theta\in\mathbb{R}^p, J_N(\theta)= {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \hat{\mu}^E_{s_ia} + l(s_i,a)) - \theta ^T \hat{\mu}^E_{s_ia_i} \right)$.
-It is clear, as $A$ is finite, that it is a convex, proper, closed and continuous function
-with finite values.
-Moreover one sub-gradient of $J_N$ in $\theta\in\mathbb{R}^p$ is $g_{\theta}={1\over N}\sum_{i=1}^N\left(\hat{\mu}^E_{s_ia^*_i} - \hat{\mu}^E_{s_ia_i}\right)$, where $a^*_i\in\arg\max_{a\in A}(\theta ^T \hat{\mu}^E_{s_ia} + l(s_i,a))$.
-We can then use the result from the article \cite{correa1993convergence} section 5 which allows to say, with our notations, that the sequence $\theta_0=0,\forall t\in \mathbb{N}^*,\theta_{t+1}=\theta_t-\alpha_t\frac{g_{\theta_t}}{\|g_{\theta_t}\|_2}$ is convergent
-if there exists a minimum to $J_N$, and that $\liminf_{t\rightarrow +\infty}J_N(\theta_t)=\inf_{\theta\in\mathbb{R}^p}J_N(\theta)$ under the conditions: $\sum_{t\geq0}\alpha_t>+\infty,\text{ }\sum_{t\geq0}\alpha_t^2<+\infty$.
-For a discussion on the existence
-of the minimum, one can report to the article \cite{kaufmann1988existence}.
-Finally, one can show that if there exists $\theta\in\mathbb{R}^p$ such that $\forall 1\leq i \leq N, \theta^T\mu^E_{s_ia_i}>\max_{a\in A \backslash a_i}\theta^T\mu^E_{s_i,a}$, then there exists a minimum of $J_N$ and that all the
-minimums: $\{\tilde{\theta}: \inf_{\theta\in\mathbb{R}^p}J_N(\theta)=J_N(\tilde{\theta})\}$ verify $\forall 1\leq i \leq N, \tilde{\theta}^T\mu^E_{s_ia_i}>\max_{a\in A \backslash a_i}\tilde{\theta}^T\mu^E_{s_i,a}$.
-\subsection{Quality of the solution found by the sub-gradient algorithm}
-Here we are going to bound the term $E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]$.
-First of all let us introduce the estimation errors $\{\epsilon^{\mu}_{s_ia_i}=\hat{\mu}^E_{s_ia_i}-\mu^E_{s_ia_i}\}_{1\leq i \leq N,a\in A}$ and the
-approximation error $\epsilon^R$ which results from the fact that the unknown and bounded reward function $R\in\mathbb{R}^S$ may not be a linear combination of the features $\psi_{1\leq i \leq p}$.
-Formally, let $T$ be the tuple $\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma\}$ which is
-an MDP without reward and $T(\tilde{R})=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,\tilde{R}\}$ where $\tilde{R}\in\mathbb{R}^S$.
-We note $C_R$ the reward-equivalence class of $R$: $C_R=\{\tilde{R}\in\mathbb{R}^S: \Pi(T(R))=\Pi(T(\tilde{R}))\}$ where $\Pi(T(\tilde{R}))$ is the set of optimal deterministic policies
-of $T(\tilde{R})$.
-We can now give a formal definition of $\epsilon^R$: $\epsilon^R=\|R^*-\theta^{*T}\psi\|_\infty=\max_{s\in S}|R^*(s)-\theta^{*T}\psi(s)|$ where $(R^*,\theta^*)\in\argmin_{\tilde{R}\in C_R, \theta\in\mathbb{R}^p}{\|\theta^T\psi-R\|_2}$.
-Now we suppose that our sub-gradient converges to $\tilde{\theta}$ and that $g_{\tilde{\theta}}={1\over N}\sum_{i=1}^N\left(\hat{\mu}^E_{s_ia^*_i} - \hat{\mu}^E_{s_ia_i}\right)=0$, where $a^*_i\in\arg\max_{a\in A}(\tilde{\theta}^T \hat{\mu}^E_{s_ia} + l(s_i,a))$.
-This case happen for instance when there exists $\theta\in\mathbb{R}^p$ such that $\forall 1\leq i \leq N, \theta^T\mu^E_{s_ia_i}>\max_{a\in A \backslash a_i}\theta^T\mu^E_{s_i,a}$.
-The apprentice policy $\pi_A$ is an optimal deterministic policy
-for the MDP $T(\tilde{\theta}\psi)$ and we can show that: $\forall 1\leq i\leq N, \pi_A(s_i)=a_i^*$.
-So we have: ${1\over N}\sum_{i=1}^N\left(\hat{\mu}^E_{s_i\pi_A(s_i)} - \hat{\mu}^E_{s_i\pi_E(s_i)}\right)=0$ which implies: ${1\over N}\sum_{i=1}^N(<\mu^E_{s_i\pi_A(s_i)},\theta^*>-<\mu^E_{s_i\pi_E(s_i)},\theta^*>+<\epsilon^{\mu}_{s_i\pi_A(s_i)}-\epsilon^{\mu}_{s_i\pi_E(s_i)},\theta^*>)=0$.
-Here $Q^{\pi_E}$ and $V^{\pi_E}$ are defined in function of $R^*$ and not in function of $R$, we should use the notations $Q^{\pi_E}_{R^*}$ and $V^{\pi_E}_{R^*}$.
-Let note the action-value errors $\{\epsilon^Q_{s_ia_i}=Q^{\pi_E}(s_i,a_i)-<\theta^*,\mu^E_{s_ia_i}>\}_{1\leq i \leq N,a\in A}$, we have $\forall 1\leq i\leq N, |\epsilon^Q_{s_ia_i}|=|E^{\pi_E}_{s_ia_i}[\sum_{t\geq0}\gamma^t(R^*(s_t)-<\theta^*,\psi(s_t)>)]|\leq\frac{\epsilon^R}{1-\gamma}$.
-So we have:
-${1\over N}\sum_{i=1}^N(Q^{\pi_E}(s_i,\pi_E(s_i))-\Q^{\pi_E}(s_i,\pi_A(s_i))+<\epsilon^{\mu}_{s_i\pi_E(s_i)}-\epsilon^{\mu}_{s_i\pi_A(s_i)},\theta^*>-\epsilon^{Q}_{s_i\pi_E(s_i)}+\epsilon^{Q}_{s_i\pi_A(s_i)})=0$.
-Finally, let us define $\epsilon^V$, $V$ as Vapnick, the Vapnick-error term:
-$\epsilon^V=E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]-[{1\over N}\sum_{i=1}^NQ^{\pi_E}(s_i,\pi_E(s_i))-\Q^{\pi_E}(s_i,\pi_A(s_i))]$.
-Here we suppose that the samples$\{s_i,a_i\}_{1\leq i \leq N,a\in A}$ are i.i.d and distributed over $\nu$.
-We know that the $\pi_A$ we find depends only on $\tilde{\theta}$ which depends on the samples
-but as $\tilde{\theta}$ is a linear combination of $p$ parameters it implies that the space of possible $\pi_A$ has a VC-dimension $h$ such that: $h\leq p$.
-So, we can use a Vapnick-bound for $\epsilon^V$ with probability $1-\eta$ because $\forall s\in S, 0\leq|Q^{\pi_E}(s,\pi_E(s))-\Q^{\pi_E}(s,\pi_A(s))|\leq\frac{R^*_{max}}{1-\gamma}$, where $R^*_{max}=\max_{s\in S}R^*(s)$.
-$|\epsilon_V|\leq\frac{R^*_{max}}{1-\gamma}\epsilon(N)$ where $\epsilon(N)=\sqrt{\frac{h}{N}[\ln(\frac{2N}{h})+1]-\frac{1}{N}\ln(\frac{\eta}{4})}$.
-Finally by Cauchy-Schwartz, we have the final upper-bound with probability $1-\eta$:
-$E_{s\sim\nu}[V^{\pi_E}(s)-Q^{\pi_E}(s,\pi_A(s))]\leq\frac{\epsilon^R}{1-\gamma}+2\|\theta^*\|_2\|\epsilon^{\mu}\|_{\infty}+\epsilon(N)$, where $\|\epsilon^{\mu}\|_{\infty}=\max_{1\leq i\leq N,a\in A}\|\epsilon^{\mu}_{s_ia}\|_2$.
-\section{Related work}
-\label{biblio.sec}
-%%%%%
-%L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
+%%%%%%
+%Other estimation methods are possible, as for instance a Monte-Carlo
+%estimation, less calculus-expensive and constraining than the first
+%one, which would be based solely on the data given by the expert and
+%would imply the use of an heuristic on the unknown actions. This
+%estimation would be consider on future works.
+%
+
+\section{Asymptotic analysis}\label{proof.sec}
+In this section, we prove, under two assumptions, that there is
+convergence of the subgradient algorithm applied to the asymptotic
+risk function $J$ (that is $J = \lim_{N\rightarrow\infty} J_N$)
+defined as:
+\begin{equation}
+J(\theta)=\sum_{s\in S}\nu_E(s)(\max_{a \in A}(\theta^T\mu^E_{sa}+l(s,a))-\theta^T\mu^E_{s\pi_E(s)}),
+\end{equation}
+where $\nu_E$ is the asymptotic probability distribution on states
+induced by $\pi_E$. Moreover, we prove that the result of the
+subgradient algorithm, which is
+$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}$, is a minimum of
+$J$ and that the expert policy $\pi_E$ is the unique optimal policy
+for the reward function $\tilde{\theta}^T\psi$.
+
+\subsection{Result}
+\begin{assumption}
+\label{assumption1}
+It is assumed that we are able to estimate perfectly the term $\{\mu^E_{sa}\}_{s\in S,a\in A}$.
+\end{assumption}
+
+\begin{assumption}
+\label{assumption2}
+There exists $\theta\in\mathbb{R}^p$ such that $\forall s\in S$:
+\begin{equation}
+\label{existence.eq}
+\theta^T\mu^E_{s\pi_E(s)}>\max_{a \in A\backslash\pi_E(s)}\theta^T\mu^E_{sa}.
+\end{equation}
+\end{assumption}
+
+\begin{lemme}
+\label{proposition1} Under assumptions \ref{assumption1} and
+\ref{assumption2}, there exists a minimum $\theta^*$ of J. The
+elements of the set $M^*=\{\theta^*\in\mathbb{R}^p:
+J(\theta^*)=\inf_{\theta\in\mathbb{R}^p}J(\theta)\}$ satisfy
+equation \eqref{existence.eq}.
+\end{lemme}
+\begin{proof}
+Let the margin function $l$ be such that: $\forall s\in S, \forall
+a\in A\backslash\pi_E(s), l(s,a)=c>l(s,\pi_E(s))=0$. Then $\forall
+\theta\in\mathbb{R}^p,J(\theta)\geq\sum_{s\in
+S}\nu_E(s)l(s,\pi_E(s))=0$. Let note $\hat{\theta}$ the vector
+verifying \eqref{existence.eq}, it is clear that
+$\|\hat{\theta}\|_2\neq0$. Let $x>0$ and
+$\hat{\theta}_x$=$x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}$, then
+there exists $x$ such that $J(\hat{\theta}_x)=0$. Indeed, $\forall
+s\in S$ :
+\begin{align}
+&J(\hat{\theta}_x)=0\\
+\Leftrightarrow&\hat{\theta}_x^T\mu^E_{s\pi_E(s)}\geq\max_{a \in A\backslash\pi_E(s)}(\hat{\theta}_x^T\mu^E_{sa}+l(s,a))\\
+\Leftrightarrow&x\geq\frac{c}{\hat{\theta}_1^T\mu^E_{s\pi_E(s)}-\max_{a \in A\backslash\pi_E(s)}\hat{\theta}_1^T\mu^E_{sa}}.
+\end{align}
+Then all the vectors $\hat{\theta}_x$ with $x\geq c_{max}=\max_{s\in
+S}\frac{c}{\hat{\theta}_1^T\mu^E_{s\pi_E(s)}-\max_{a \in
+A\backslash\pi_E(s)}\hat{\theta}_1^T\mu^E_{sa}}$ are such that:
+$\inf_{\theta\in\mathbb{R}^p}J(\theta)\geq0=J(\hat{\theta}_x)\geq\inf_{\theta\in\mathbb{R}^p}J(\theta)$.
+So $\hat{\theta}_x=\inf_{\theta\in\mathbb{R}^p}J(\theta)$ for $x\geq
+c_{max}$. Finally let us show that the elements of the set
+$M^*=\{\theta^*\in\mathbb{R}^p:
+J(\theta^*)=\inf_{\theta\in\mathbb{R}^p}J(\theta)\}$ satisfy
+equation \eqref{existence.eq}. As there exists $x$ such that
+$J(\hat{\theta}_x)=0=\inf_{\theta\in\mathbb{R}^p}J(\theta)$, it
+implies $J(\theta^*)=0$. So:
+\begin{align}
+&J(\theta^*)=0\\
+\Leftrightarrow&\theta^{*T}\mu^E_{s\pi_E(s)}\geq\max_{a \in A\backslash\pi_E(s)}(\theta^{*T}\mu^E_{sa}+l(s,a))\\
+\Rightarrow&\theta^{*T}\mu^E_{s\pi_E(s)}>\max_{a \in A\backslash\pi_E(s)}\theta^{*T}\mu^E_{sa}.
+\end{align}
+\end{proof}
+
+\begin{prop}
+Under assumptions \ref{assumption1} and \ref{assumption2}, the
+subgradient algorithm defined by the sequence $\theta_0=0$,
+$\theta_{t+1}=\theta_t-\alpha_t\frac{g_{\theta_t}}{\|g_{\theta_t}\|_2}$
+where $g_{\theta_t}=\sum_{s\in
+S}\nu_E(s)(\mu^E_{sa^*}-\mu^E_{s\pi_E(s)})$ with
+$a^*\in\argmax_{a\in A}\theta_t^T\mu^E_{sa}+l(s,a)$,
+$\sum_{t\geq0}\alpha_t>\infty$ and $\sum_{t\geq0}\alpha_t^2<\infty$,
+is convergent:
+$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$.
+Moreover we have that
+$J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$. Finally
+the expert policy $\pi_E$ is the unique optimal policy for the
+reward $\tilde{\theta}^T\psi$.
+\end{prop}
+\begin{proof}
+The cost function  $J$ is convex, proper, closed and with finite
+values (the action space being finite). Therefore, using the result
+of~\cite[section~5]{correa1993convergence} allows to conclude that
+$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$
+and $J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)$.
+%
+Moreover, Lemma~\ref{proposition1} shows that
+$\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$.
+%
+So we have
+$J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$ which
+means that $\tilde{\theta}\in M^*$ and satisfies equation
+\eqref{existence.eq}. This implies that, $\forall s\in S$, $\forall
+a\in A$:
+\begin{equation}
+\tilde{\theta}^T\mu^E_{s\pi_E(s)}>\tilde{\theta}^T\mu^E_{sa},
+\end{equation}
+which proves that the expert policy $\pi_E$ is the unique optimal
+policy for the reward $\tilde{\theta}^T\psi$.
+\end{proof}
+
+\subsection{Discussion}
+The assumptions \ref{assumption1} and \ref{assumption2} are quite
+important and we can of course wonder what results could be
+guaranteed if they were not true. Moreover, we have only sampled
+transitions, so we cannot use the asymptotic risk function $J$, only
+the empirical function $J_N$. Thus, in our future works, the aim
+will be to find, if possible, an upper bound for the term:
+$E_{s\sim\nu_E}[V^{\pi_E}(s)-V^{\pi_A}(s)]$ where $E_{s\sim\nu_E}$
+is the expectation over the stationary probability distribution of
+states $\nu_E$ and $\pi_A$ is the policy of the apprentice which is
+optimal for the reward $\tilde{\theta}^T\psi$ found by the
+subgradient algorithm. Of course, this upper-bound will depend on
+the feature expectation errors
+$\{\epsilon^{\mu}_{s_ia}=\hat{\mu}^E_{s_ia}-\mu^E_{s_ia}\}_{1\leq i
+\leq N,a\in A}$ (which may be quantified thanks to the finite sample
+analysis of LSTD~\cite{lazaric2010finiteLSTD}) and on the approximation error
+$\epsilon^R$ which results from the fact that the unknown and
+bounded reward function $R_E\in\mathbb{R}^S$ may not be a linear
+combination of the features $\psi_{1\leq i \leq p}$. Formally, let
+$T$ be the tuple $\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma\}$ which
+is an MDP without reward and $T(\tilde{R})=\{S,A,\{P_{sa}\}_{s\in
+S,a\in A},\gamma,\tilde{R}\}$ where $\tilde{R}\in\mathbb{R}^S$. We
+note $C_R$ the reward-equivalence class of $R$:
+$C_R=\{\tilde{R}\in\mathbb{R}^S: \Pi(T(R))=\Pi(T(\tilde{R}))\}$
+where $\Pi(T(\tilde{R}))$ is the set of optimal deterministic
+policies of $T(\tilde{R})$ (see~\cite{ng1999policy} for a
+characterization of $C_R$). A formal definition of $\epsilon^R$ is:
+$\epsilon^R=\|R^*-\theta^{*T}\psi\|_\infty=\max_{s\in
+S}|R^*(s)-\theta^{*T}\psi(s)|$ with
+$(R^*,\theta^*)\in\argmin_{\tilde{R}\in C_R,
+\theta\in\mathbb{R}^p}{\|\theta^T\psi-R\|_2}$. Linking the empirical
+risk function $J_N$ to the true risk function $J$ could be done
+using a Vapnik-like analysis. The difficult thing of course is to
+have access to $V^{\pi_A}$ without resolving the MDP directly.
+
+%\section{Related work}
+%\label{biblio.sec}
+%%%%%%
+%%L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
 Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de \mbox{l'expert} et celui d'une politique donnée.
 Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
-The historical approach of \cite{ng2000algorithms}, as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} as summed up in \cite{neu2009training} all suffer from the same constraint: the need to repeatedly solve the forward problem, that is to say the need to repeatedly find the optimal policy with respect to an arbitrary reward.
+%The historical approach of \cite{ng2000algorithms}, as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} as summed up in \cite{neu2009training} all suffer from the same constraint: the need to repeatedly solve the forward problem, that is to say the need to repeatedly find the optimal policy with respect to an arbitrary reward.
 All these approaches try to minimize some distance between the feature expectation of the expert and the feature expectation of another policy.
 Successive updates of the vector of parameters $\theta$ will get this other policy (optimal with respect to a criterion that depends on $\theta$, this is where the \emph{RL} problem needs to be solved) closer and closer to the expert's, feature expectation-wise.\\
-
-%Cette résolution est très problématique.
+%
+%%Cette résolution est très problématique.
 En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la littérature.
@@
 Comme cela est illustré dans \cite{klein2011batch}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état, ne permettant pas la résultion du problème direct en se basant uniquement sur les  données de l'expert.\\
-%FIXME:parler de boularias ici
-Solving the forward problem can be very tricky ; indeed a whole body of literature is dedicated to this problem.
+%%FIXME:parler de boularias ici
+%Solving the forward problem can be very tricky ; indeed a whole body of literature is dedicated to this problem.
 Specifically, one needs more information to solve the forward problem than to just compute the feature expectation of the expert as our algorithm does.
@@
 Even a recent contribution like \cite{boularias2011relative} still needs samples that cover a large portion of the state space although it does not explicitely require the ability to solve the forward problem.\\
-
-%Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
+%
+%%Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
 C'est à dire que l'action choisie par l'expert doit avoir une qualité strictement supérieure à n'importe quelle autre action.
 Les approches de planification structurée comme \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} permettent d'influer sur cette marge.\\
-
-A recurring argument is to see the expert not only as optimal but as optimal by a certain margin.
+%
+%A recurring argument is to see the expert not only as optimal but as optimal by a certain margin.
 The action chosen by the expert should have a quality strictly greater than any other action.
 Structured planning as done in \cite{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} allow some influence on the margin.\\
-
-%Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
+%
+%%Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
 Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}, notre fonction de risque étant similaire à la leur.
@@
 Il semble difficile de sortir de ce cadre  en pratique (bien que l'algorithme soit introduit dans un cadre plus général).
-
-Specifically, our work shows similarities with \cite{ratliff2006maximum}.
+%
+%Specifically, our work shows similarities with \cite{ratliff2006maximum}.
 The authors cast the \emph{IRL} problem as a maximum margin classification problem ; our risk function is similar to theirs.
@@
 Although the proposed algorithm is described in a more general setting, it seems unpractical outside the deterministic case.
-
-%Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
+%
+%%Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
 Le principe de leur approche est exactement de résoudre le risque (Eq.
@@
 Toutefois, ils ne prennent pas en comptent la structure du MDP, ce à quoi nous remédions avec notre contribution (voir \cite{melo2010learning} sur l'importance de prendre en compte cette structure).
-%Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
+%%Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
 Cette fonction de score $s: X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e.
@@
 \eqref{structure.eqn}).\\
-Finally, \citep{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem.
+%Finally, \citep{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem.
 Both our approach and theirs minimize the same risk function, described Eq.
 \eqref{Jdef.eqn}, but they do not use the structure of the \emph{MDP} as we do (see \cite{melo2010learning} about the importance of taking this structure into account).\\
-
-%Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
+%
+%%Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
 Il existe d'autres approches que nous ne passons pas en revue de manière exhaustive, mais à notre connaissance elles nécessitent toutes de résoudre le problème direct.
-
-Our approach is in a way between those two contributions as we propose to introduce the structure of the \emph{MDP} (this is what the first contribution do) in the same abstraction level as the second approach.
+%
+%Our approach is in a way between those two contributions as we propose to introduce the structure of the \emph{MDP} (this is what the first contribution do) in the same abstraction level as the second approach.
 Other techniques exist, but to the best of our knowledge none of them can work only from samples drawn by the expert.
@@
 %%%
-\subsection{Description of the task}
+In this section, experiments on two classical benchmarks are
+proposed: the inverted pendulum and the car driving problems.
+
+\subsection{Inverted Pendulum}
 %%%
@@
 %la seconde étant sa vitesse angulaire.\\
-The inverted pendulum problem consists in maintaining in equilibrium a pendulum with one degree of freedom. This is a classical problem in the reinforcement learning field.
-The policy of our expert is determined by an RL-algortihm called LSPI which is explained in the article \cite{lagoudakis2003least}. The policy of the apprentice, which is the policy
-optimizing the reward function of our algorithm given Figure \ref{onlyFig.fig} (a), manages systematically  
+%
+%
+The inverted pendulum problem consists in maintaining in vertical
+equilibrium a pendulum with one degree of freedom by moving the cart
+on which it is fixed. This classical problem is extensively
+described in~\cite{lagoudakis2003least} from which the parameters of
+our experiment were taken. The state space is two-dimensional: the
+first dimension is the angular position of the pendulum (where the
+vertical position corresponds to the angle zero) and the second one
+is the angular speed. The policy of the expert is determined by the
+Least Square Policy Iteration (LSPI) algorithm
+from~\cite{lagoudakis2003least}, using the same basis functions
+(Gaussian functions).
+%
 %La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
 %Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
-
+The policy of the expert optimizes the reward function given Figure
+\ref{onlyFig.fig} (a). It manages systematically to maintain the
+pendulum during 3000 time-steps. A time-step having a value of 0.1s,
+the expert is interrupted after 5 minutes of successful
+interactions.
+%
 %Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases,
@@
 %(correspondant à la position verticale et son voisinage proche).\\
+%
 
-%Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
+The main difficulty of this experiment relies in the fact that the
+state space is continuous and, above all, that expert trajectories
+(shown in green on Figure~\ref{onlyFig.fig} (b)) are only covering a
+very small area of the state space.
+%
+%Ce dernier point est d'ailleurs la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
 %En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser
@@
 %%%
-\subsection{Illustrations}
+For this reason, the algorithm of \citet{abbeel2004apprenticeship}
+fails in finding a solution to this problem only with the
+transitions available from the expert demonstrations. Indeed, as
+shown in~\citep{klein2011batch}, additional data covering the whole
+state space is required.
+%
+%These algorithms are iterative and try to
+%find a policy $\pi$ which feature expectation is compared to the
+%feature expectation of the expert. Thus, these algorithms need to
+%resolve the MDP and to do so need samples covering the whole state
+%space. Our algorithm, only need to be able to estimate the feature
+%expectation of the expert which does note require the direct
+%resolution of the MDP.
+\subsubsection{Results}
 %%%
 \begin{figure}[ht]
-\vskip 0.2in
+%\vskip 0.2in
 \begin{center}
-\begin{minipage}[b]{.5\linewidth}
+\begin{minipage}[b]{.48\linewidth}
   \centering
   \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_true_R.pdf}}
-  \centerline{(a) Récompense de l'expert}%\medskip
+  \centerline{(a) Expert reward ($R^E$)}%\medskip
+\end{minipage}
+%\hfill
+\begin{minipage}[b]{.48\linewidth}
+  \centering
+  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_lafem_R.pdf}}
+  \centerline{(b) SCIRL Reward ($R_\theta$)}%\medskip
+\end{minipage}
+\begin{minipage}[b]{.48\linewidth}
+  \centering
+  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vexpert.pdf}}
+  \centerline{(c) Expert value function}%\medskip
 \end{minipage}
 \hfill
-\begin{minipage}[b]{.5\linewidth}
-  \centering
-  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_lafem_R.pdf}}
-  \centerline{(b) Récompense trouvée}%\medskip
-\end{minipage}
-\begin{minipage}[b]{.5\linewidth}
-  \centering
-  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vexpert.pdf}}
-  \centerline{(c) Fonction de valeur de l'expert}%\medskip
-\end{minipage}
-\hfill
-\begin{minipage}[b]{.5\linewidth}
+\begin{minipage}[b]{.48\linewidth}
   \centering
   \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vagent.pdf}}
-  \centerline{(d) Fonction de valeur de l'agent}%\medskip
+  \centerline{(d) SCIRL value function}%\medskip
 \end{minipage}
 %
-\caption{Résultats fournis par notre algorithme.}
-\label{onlyFig.fig}
+\caption{SCIRL results on the inverted pendulum} \label{onlyFig.fig}
 %
@@
 \end{figure}
+%
+%Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
+%La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon. Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations
+%est de $T=20$. Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange
+%de gaussiennes défini dans \cite{lagoudakis2003least}. Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
+%
+So as to test the SCIRL algorithm, a database $D_E$ of transitions
+given by the expert policy $\pi_E$ is built. It contains 10
+trajectories of 300 transitions each. The margin function is $l(s,a)
+= 0$ if $a=\pi_E(s)$, $1$ otherwise.
+%The time step is constant,
+%$\Delta t=0.1s$, and the stopping time is $T=20$.
+The initial vector, $\theta_0$, is fixed at $\begin{pmatrix}   -1 &
+\dots & -1
+\end{pmatrix}^T$,
+and the basis functions ($\psi: S \rightarrow \mathbb{R}^p$) used
+for the reward function are a Gaussian network as defined
+in~\cite{lagoudakis2003least}. The LSTD$\mu$
+algorithm~\cite{klein2011batch} is used to estimate the feature
+expectation $\mu^E(s,a)$.
 
-Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
-La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon. Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations
-est de $T=20$. Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange
-de gaussiennes défini dans \cite{lagoudakis2003least}. Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
+%Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs
+%sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)). Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense
+%pouvant expliquer une politique.\\
+%
+Even if the reward function found by SCIRL (figure \ref{onlyFig.fig}
+(b)) is different than the expert reward (figure \ref{onlyFig.fig}
+(a)), the value functions are similar (figures \ref{onlyFig.fig} (c)
+et \ref{onlyFig.fig} (d)). This is an illustration that the IRL
+problem is ill-posed in the sense that there is not only one reward
+function able to explain the optimality of a policy.
 
-Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs
-sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)). Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense
-pouvant expliquer une politique.\\
+%Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes,
+%ce qui est notre critère de réussite.\\
+With the number of samples given by the expert (3000), the
+apprentice is able to systematically maintain the pendulum in
+equilibrium during 5 minutes as the expert does.
+%
+%Deux éléments méritent d'être notés. Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
+%Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace:
+%celle dans laquelle le pendule est proche de la verticale. En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
+%Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer
+%la fonction de récompense utilisée sans l'atteindre. Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir
+%des performances similaires à celles de l'expert. Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig. \ref{onlyFig.fig} Cela n'a aucune importance, une dilatation
+%laissant la politique optimale invariante.
+%
+Two elements deserve to be noticed. First, the expert samples are
+only covering a small area of the state space (see
+Figure~\ref{onlyFig.fig} (b)) and especially they are close to the
+vertical position. The lack of samples in the rest of the state
+space makes it impossible to infer with certainty the reward
+function. Second, the \textit{actual} reward function $R^E$
+(Figure~\ref{onlyFig.fig} (a)) does not live in the hypothesis space
+span by a linear combination of the basis functions. Despite these
+difficulties, SCIRL succeeds to extract a reward function leading to
+perform similarly to the expert. %Notice that there is a magnitude
+%difference between the SCIRL and the expert value functions on
+%Figure~\ref{onlyFig.fig} but an affine transform of the value
+%function does not change the optimal policies.
 
-Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes,
-ce qui est notre critère de réussite.\\
+\subsection{Car driving}
 
-Deux éléments méritent d'être notés. Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace:
-celle dans laquelle le pendule est proche de la verticale. En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
-Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer
-la fonction de récompense utilisée sans l'atteindre. Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir
-des performances similaires à celles de l'expert. Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig. \ref{onlyFig.fig} Cela n'a aucune importance, une dilatation
-laissant la politique optimale invariante.
+This task is taken from~\cite{syed2008game}. It consists in
+navigating a car through randomly-generated traffic on a threelane
+highway. The reward features are identical to~\citet{syed2008game}
+paper that is: a collision feature (0 if contact with another car,
+and 0.5 otherwise), an off-road feature (0 if on the grass, and 0.5
+otherwise), and a speed feature (0.5, 0.75 and 1 for each of the
+three possible speeds, with higher values corresponding to higher
+speeds).
+
+\subsubsection{Results}
+
+In this section, SCIRL is compared to the Multiplicative Weights for
+Apprenticeship Learning (MWAL) algorithm proposed by
+\citet{syed2008game}. This algorithm requires the perfect knowledge
+of the MDP and outputs a single policy (not a reward) that imitates
+the expert.
+
+The experimental setup is as follows. The highway simulator
+of~\cite{syed2008game} is used\footnote{available at \url{
+http://www.cs.princeton.edu/~usyed/}}. It also contains an exact MDP
+solver for the task, the transition probabilities being available.
+Two reward functions were hand-tuned ($\theta$ parameters chosen by
+hand) so as to generate two different expert policies: one for
+high-speed driving and another for safe driving. The MDP solver is
+used to obtain the expert policies given the two reward functions. A
+number $N$ of trajectories of length $H$ is generated with each
+strategy providing the expert data for training the SCIRL and MWAL
+algorithms. The MWAL algorithm requires the knowledge of the average
+feature expectation $\mu_{d0}^{\pi_E}$ for the initial state (given
+a distribution $d_0$ over starting states). To reach a fair
+comparison with SCIRL which only uses the expert trajectories,
+$\mu_{d_0}^{\pi_E}$ is computed as the Monte Carlo estimate given
+the $N$ trajectories, knowing that the initial state is always the
+same:
+%
+$$\mu_{d_0}^{\pi_E} = \frac{1}{N}\sum_{i=1}^N \sum_{t=1}^H \gamma^t \psi(s_t^i)$$
+%
+For SCIRL, the LSTD$\mu$ algorithm is used again to approximate
+$\mu^E(s,a)$ from the data. For each component of $\mu^E$, the basis
+functions are the same as those for the reward function. SCIRL
+outputs a reward function which is used to obtain a policy using the
+MDP solver. The SCIRL and MWAL policies are then run on the
+simulator for trajectories of length $T$. The average discounted
+cumulative reward, computed with regard to the initial handcrafted
+rewards, serves as the performance measure. It is a Monte Carlo
+estimate of $V^\pi(s_0)$ for each policy.
+
+This experiment is repeated $M$ times for each policy and results
+are averaged. The results are shown in
+Tables~\ref{table:highway_fast} and~\ref{table:highway_safe}
+ for $N \in
+\{1, 3, 5, 10\}$, $H = 30$, $T=50$ and $M = 10$.
+
+%\begin{figure}
+%  \includegraphics[width=\linewidth]{FastResults_EB.pdf}\\
+%  \caption{Results for high-speed driving}\label{fig:highway_fast}
+%\end{figure}
+%
+%\begin{figure}
+%  \includegraphics[width=\linewidth]{SafeResults_EB.pdf}\\
+%  \caption{Results for high-speed driving}\label{fig:highway_safe}
+%\end{figure}
+
+\begin{table}
+\begin{center}
+    \begin{tabular}{|c|c|c|c|}
+    \hline
+    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
+    N     & MWAL              & SCIRL               & $V_{\text{opt}}$ \\\hline\hline
+    1     & 7.25 $\pm$ 0.10   & 7.03 $\pm$ 0.09     & 7.30              \\\hline
+    3     & 7.28 $\pm$ 0.06   & 7.07 $\pm$ 0.07     & 7.30              \\\hline
+    5     & 7.27 $\pm$ 0.11   & 6.98 $\pm$ 0.10     & 7.30              \\\hline
+    10    & 7.26 $\pm$ 0.07   & 7.05 $\pm$ 0.10     & 7.30              \\
+    \hline
+    \end{tabular}
+    \caption{SCIRL results on the car driving
+    problem (high-speed)}\label{table:highway_fast}
+\end{center}
+\end{table}
+
+\begin{table}
+\begin{center}
+    \begin{tabular}{|c|c|c|c|}
+    \hline
+    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
+    N     & MWAL              & SCIRL               & $V_{\text{opt}}$ \\\hline\hline
+    1     & 4.95 $\pm$ 0.05   & 3.27 $\pm$ 0.13     & 5.00              \\\hline
+    3     & 4.85 $\pm$ 0.13   & 4.90 $\pm$ 0.10     & 5.00              \\\hline
+    5     & 4.90 $\pm$ 0.09   & 4.83 $\pm$ 0.08     & 5.00              \\\hline
+    10    & 4.98 $\pm$ 0.03   & 4.81 $\pm$ 0.12     & 5.00              \\
+    \hline
+    \end{tabular}
+    \caption{SCIRL results on the car driving
+    problem (safe)}\label{table:highway_safe}
+\end{center}
+\end{table}
+
+From these tables, one can conclude that the SCIRL method reaches
+approximately the same performance as MWAL (given that enough
+samples are provided in the case of safe driving). Yet, SCIRL
+doesn't need any other information than the expert demonstrations
+while MWAL requires solving exactly and repeatedly the MDP.
 
@@
 
-L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse. En rendant la résolution du problème inverse indépendante
-de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
+%L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse. En rendant la résolution du problème inverse indépendante
+%de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
+In this paper, we propose an algorithm for inverse reinforcement
+learning (IRL) that we call SCIRL. It is based on a classification
+approach in which the structure of the underlying Markov Decision
+Process (MDP) is taken into account.
 
-Comme nous l'avons illustré, les échantillons nécessaires à la réussite de notre algorithmes sont faciles à recueillir.
-Une simple trace de l'expert, même si elle ne couvre qu'une petite partie de l'espace d'état, suffit à inférer une récompense.
-Cette récompense, description compacte de la tâche effectuée par l'expert, peut alors être optimisée par un agent dont les capacités diffèrent de celles de l'expert.
-On rentre dans le cadre du \emph{transfer learning}.\\
+This approach relaxes most of the constrains of other IRL algorithms
+found in the literature. It doesn't require the resolution of the
+direct problem and is able to infer a reward function which is
+consistent with the expert's behavior in a purely \emph{batch} mode.
+%
+%Comme nous l'avons illustré, les échantillons nécessaires à la réussite de notre algorithmes sont faciles à recueillir.
+%Une simple trace de l'expert, même si elle ne couvre qu'une petite partie de l'espace d'état, suffit à inférer une récompense.
+%Cette récompense, description compacte de la tâche effectuée par l'expert, peut alors être optimisée par un agent dont les capacités diffèrent de celles de l'expert.
+%On rentre dans le cadre du \emph{transfer learning}.\\
+%
+A mere trajectory of the expert, even if it covers a small part of
+the state space, is sufficient to infer a reward function and no
+additional data nor a simulator is required. It also outputs a
+reward and not a strategy which is the real goal of the IRL
+paradigm. The learnt reward function, which is a compact description
+of the task executed by the expert, could then be optimized by an
+agent with abilities (action set) that are different from those of
+the expert which places us in in the even more challenging framework
+of \emph{transfer learning}.
+%
+%L'estimation de $\mu_E$ est centrale dans notre approche. Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients
+%que les algorithmes d'estimation d'une fonction de valeur. L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen
+%d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée. Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.
 
-L'estimation de $\mu_E$ est centrale dans notre approche. Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients
-que les algorithmes d'estimation d'une fonction de valeur. L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen
-d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée. Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.
-%
-% Bibliographie
-%
+SCIRL is shown to compare positively to a state-of-the-art algorithm
+(the MWAL algorithm) on a standard benchmark (the car driving
+problem). It also performs well on a problem with a two-dimensional
+continuous state space which is sparsely explored by the expert (the
+inverted pendulum problem). This later problem could not be solved
+with state-of-the-art algorithms without additional knowledge.
+
+Yet, SCIRL requires the computation of $\mu^E(s,a)$ everywhere on
+the expert trajectories. Here, we made use of the LSTD$\mu$
+algorithm to estimate it. However LSTD$\mu$ shares the drawbacks of
+value function estimation algorithms. In the future, we plan to
+extend our approach so as to avoid the computation of the expert's
+feature expectations. Empirical tests on more complex problems will
+also be considered.
+
 \bibliography{Biblio}

### 2012-04-10 klein2012/main.tex
Correction d'une typo signalée dans les revues :
l. 420: The authors probably want to say "complete expert policy".
Correction d'une typo mentionnée dans les revues :
There are some broken/misdirect links in lines 365, 532.

@@
 Therefore, it satisfies the Bellman evaluation
-equation~\eqref{Qdef.eqn}. Assuming that the expert policy and the
+equation~\eqref{Qdef.eqn}. Assuming that the complete expert policy and the
 transition probabilities are known, the expert feature expectation
@@
 The cost function  $J$ is convex, proper, closed and with finite
-values (the action space being finite). Therefore, using the result
-of~\cite[section~5]{correa1993convergence} allows to conclude that
+values (the action space being finite). Therefore, using the result in section 5 
+of~\cite{correa1993convergence} allows to conclude that
 $\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$

### 2012-04-10 klein2012/main.tex
Correction d'une typo mentionnée dans les revues :
There is a strange sentence in line 491.
Correction simple suite à une remarque dans les revues :
Line 210: This warrants a reference to Abbeel & Ng 2004.

@@
 En effet
-We see also note that if two policies share the same feature
+We see also note (as made explicit in \cite{abbeel2004apprenticeship}) that if two policies share the same feature
 expectation, then they have the same value function with respect to
@@
 \theta\in\mathbb{R}^p,J(\theta)\geq\sum_{s\in
-S}\nu_E(s)l(s,\pi_E(s))=0$. Let note $\hat{\theta}$ the vector
+S}\nu_E(s)l(s,\pi_E(s))=0$. Let $\hat{\theta}$ be the vector
 verifying \eqref{existence.eq}, it is clear that

### 2012-04-10 klein2012/Rebuttal.txt
Première version du rebuttal.

We would like to thank the reviewers for their torough reading and their remarks.
We wish to raise the following points to somewhat clarify aspects of our work the reviewers showed concern about :
- LSTDmu is one of many choices for a subroutine to compute \mu_E(s,a). In its on-olicy form (the one used in our paper) it only needs some transitions from the expert to approximate mu_E(s,a) the same way LSTDQ approximate Q^\pi(s,a) for all s and a given some transitions drawn from a policy \pi.
  It can be replaced by a problem dependant more appropriate solution, such as a heuristic on the value of mu_E(s,a) when a \neq \pi_E(s).
  In (Klein et al. 2011) it was used in its off-policy form because of the requirement of the algorithm of (Abbeel et al. 2004) to compute \mu for an arbitrary policy. In problems where the state space exhibits a spatial coherence (e.g. the pendulum) it provides a very good approximation. Less so in some other spaces (e.g. the highway).

- With the pendulum experiment, we show that our method works (quantitative results given line 657) in conditions where state of the art algorithms fail due to lack of information. Quantitatively, (klein et al. 2011) have shown that the algorithm of (Abbeel and Ng 2004) needs 2000 to 3000 more arbitrary samples (not the expert's) in order to work. We lift this requirement with SCIRL.

- We show that our algorithm stands comparison when put in competition with MWAL. The latter has access to every bit of information about the problem whereas our algorithm only uses the expert's transitions.

- In addition to needing far less information than state of the art methods, by using a LSTD-like methode once and for all at the beginning, our algorithm is also computationnaly more efficient than others that require solving a MDP at each iteration. Indeed, solving a MDP implies (e.g. if one uses LSPI) using LSTD more than once. This empirically shows with a ten-fold gain on computation time on a desktop computer compared to MWAL.

- We insist on retrieving the reward because task transfer (getting an agent with different capabilities than the expert's to fulfill the same task) can not be done with a policy. With a reward function on the states, it is possible to train any agent on the problem, even if the action space is different.

- Our algorithm differs from Ratliff et al.'s work in several ways :
  - our data points are transitions, whereas MMP needs whole MDPs and associated values (transition probabilities, feature expectations, etc.),
  - we are not merely replacing mu with an approximation based on \mu_E. We introduce a new way to look at the score function q,
  - any classifier (and not only Taskar et al.'s) using such a parametrized score function could be plugged in our algorithm. The equations looks the same but our approach is different, and the results show so,
  - the way we express the score function q allows us to minimize the risk function using only the feature expectation of the expert which is easy to compute from very little data. MMP needs a lot more information to achieveits goal.

- Assumptions for the analysis are strong, but so are most of the assumptions of most of papers e.g. (Abbel and Ng 2004) also assume they can compute \mu perfectly. They provide results about error propagation, but although we formally don't, we experimentally show that this is not critical.

### 2012-04-12 klein2012/Rebuttal.txt
Version du rebuttal corrigée par Matthieu

@@
-We would like to thank the reviewers for their torough reading and their remarks.
-We wish to raise the following points to somewhat clarify aspects of our work the reviewers showed concern about :
-- LSTDmu is one of many choices for a subroutine to compute \mu_E(s,a).
-In its on-olicy form (the one used in our paper) it only needs some transitions from the expert to approximate mu_E(s,a) the same way LSTDQ approximate Q^\pi(s,a) for all s and a given some transitions drawn from a policy \pi.
-  It can be replaced by a problem dependant more appropriate solution, such as a heuristic on the value of mu_E(s,a) when a \neq \pi_E(s).
+We would like to thank the reviewers for their thorough reading and their remarks.
+We wish to raise the following points to somewhat clarify aspects of our work the reviewers showed concern about.
+
+- As mentioned in the paper, the feature expectation can be seen as a vector of Q-functions, of which each component can be estimated using LSTD.
+That's what LSTDmu does; this is just a temporal difference algorithm (and not a contribution of this paper, as rightly mentioned).
+As each component shares the same dynamic, the computational cost is the one of performing LSTD on one component (in O(p^3)), see (Klein et al.
+2011).
+Also, we have the same guarantees for estimating mu_E from expert trajectories than for estimating a Q-function with LSTD in an on-policy setting (e.g., see Lazaric et al, 2010).
 In (Klein et al.
@@
 2004) to compute \mu for an arbitrary policy.
-In problems where the state space exhibits a spatial coherence (e.g.
-the pendulum) it provides a very good approximation.
-Less so in some other spaces (e.g.
-the highway).
+The on-policy setting considered in our contribution only requires data from the expert.
 
+- Our algorithm differs from Ratliff et al.'s work in several ways.
+  - All approaches build on the structured large margin framework of Taskar et al., but they solve quite different problems.
+  - Our labels are actions, MMP's labels are policies. We only need expert transitions whereas MMP needs the whole model (for a set of MDPs, not just one). 
+  - We are not merely replacing mu with an approximation based on \mu_E (this is Klein et al).
+We introduce a new way to look at the score function q, which allows addressing two major problems of IRL.
+  - Any classifier (and not only the structured margin framework of Taskar et al. used here) using such a parameterized score function could be plugged in our algorithm.
+  - The way we express the score function q allows us to minimize the risk function using only the feature expectation of the expert, which can be estimated solely from expert transitions.
+MMP needs a lot more information to achieve its goal.
 - With the pendulum experiment, we show that our method works (quantitative results given line 657) in conditions where state of the art algorithms fail due to lack of information.
 Quantitatively, (klein et al.
-2011) have shown that the algorithm of (Abbeel and Ng 2004) needs 2000 to 3000 more arbitrary samples (not the expert's) in order to work.
+2011) have shown that the algorithm of (Abbeel and Ng 2004) needs 2000 to 3000 more arbitrary samples (not the expert's) in order to work (this is due to the off-policy learning aspect).
 We lift this requirement with SCIRL.
+
+- We recognize that the assumptions of the analysis are strong (as discussed in the paper).
+However, we show that the algorithm computes a reward for which the expert policy is the unique optimal policy, which is not a straightforward problem (even knowing the dynamic and the expert policy).
+Given that recovering a reward is the problem we address, we think it is valuable, despite the strong assumptions.
 
@@
 
-- In addition to needing far less information than state of the art methods, by using a LSTD-like methode once and for all at the beginning, our algorithm is also computationnaly more efficient than others that require solving a MDP at each iteration.
+- In addition to needing far less information than state of the art methods, by using a LSTD-like method once and for all at the beginning, our algorithm is also computationally more efficient than others that require solving one or more MDP at each iteration.
 Indeed, solving a MDP implies (e.g.
@@
 
-- We insist on retrieving the reward because task transfer (getting an agent with different capabilities than the expert's to fulfill the same task) can not be done with a policy.
+- We insist on retrieving the reward because task transfer (getting an agent with different capabilities than the expert's to fulfill the same task) cannot be done with a policy.
 With a reward function on the states, it is possible to train any agent on the problem, even if the action space is different.
-
-- Our algorithm differs from Ratliff et al.'s work in several ways :
-  - our data points are transitions, whereas MMP needs whole MDPs and associated values (transition probabilities, feature expectations, etc.),
-  - we are not merely replacing mu with an approximation based on \mu_E. We introduce a new way to look at the score function q,
-  - any classifier (and not only Taskar et al.'s) using such a parametrized score function could be plugged in our algorithm.
-The equations looks the same but our approach is different, and the results show so,
-  - the way we express the score function q allows us to minimize the risk function using only the feature expectation of the expert which is easy to compute from very little data.
-MMP needs a lot more information to achieveits goal.
-
-- Assumptions for the analysis are strong, but so are most of the assumptions of most of papers e.g.
-(Abbel and Ng 2004) also assume they can compute \mu perfectly.
-They provide results about error propagation, but although we formally don't, we experimentally show that this is not critical.

### 2012-04-12 klein2012/Rebuttal.txt
Rebuttal prenant les remarques d'Olivier en compte, trop long.

@@
 That's what LSTDmu does; this is just a temporal difference algorithm (and not a contribution of this paper, as rightly mentioned).
-As each component shares the same dynamic, the computational cost is the one of performing LSTD on one component (in O(p^3)), see (Klein et al.
-2011).
+As each component shares the same dynamic, the computational cost is the one of performing LSTD on one component (in O(p^3)), see Klein et al.
 Also, we have the same guarantees for estimating mu_E from expert trajectories than for estimating a Q-function with LSTD in an on-policy setting (e.g., see Lazaric et al, 2010).
-In (Klein et al.
-2011) it was used in its off-policy form because of the requirement of the algorithm of (Abbeel et al.
+In Klein et al.
+it was used in its off-policy form because of the requirement of the algorithm of (Abbeel et al.
 2004) to compute \mu for an arbitrary policy.
@@
 
-- Our algorithm differs from Ratliff et al.'s work in several ways.
+- As mentionned by reviewer 2, our algorithm is inspired by MMP but differs in several ways.
+We are not merely combining existing components as assessed by reviewer 3, but introducing a new way to look at the IRL problem.
  - All approaches build on the structured large margin framework of Taskar et al., but they solve quite different problems.
@@
  - Any classifier (and not only the structured margin framework of Taskar et al. used here) using such a parameterized score function could be plugged in our algorithm. 
-  - The way we express the score function q allows us to minimize the risk function using only the feature expectation of the expert, which can be estimated solely from expert transitions.
+ - The way we express the score function q allows us to minimize the risk function using only the feature expectation of the expert evaluated only in states present in the expert's dataset, which can be estimated solely from expert transitions.
 MMP needs a lot more information to achieve its goal.
-- With the pendulum experiment, we show that our method works (quantitative results given line 657) in conditions where state of the art algorithms fail due to lack of information.
-Quantitatively, (klein et al.
-2011) have shown that the algorithm of (Abbeel and Ng 2004) needs 2000 to 3000 more arbitrary samples (not the expert's) in order to work (this is due to the off-policy learning aspect).
-We lift this requirement with SCIRL.
+
+- The lack of quantitative results mentionned by reviewer 2 is addressed with the pendulum experiment, we show that our method works (quantitative results given line 657) in conditions where state of the art algorithms fail due to lack of information.
+Klein et al.'s results (not recalled in our paper due to space constraints) show that the algorithm of Abbeel and Ng needs around 3000 more arbitrary samples (not the expert's) in order to work (this is due to the off-policy learning aspect).
+We lift this requirement with SCIRL, which is to the best of our knowledge the only algorithm that works in such drastic conditions.
 
@@
 - We show that our algorithm stands comparison when put in competition with MWAL.
-The latter has access to every bit of information about the problem whereas our algorithm only uses the expert's transitions.
+The latter has access to the system's dynamics whereas our algorithm only uses the expert's transitions.
+In the light of this, marginally subpar results are indeed a contribution, indeed our algorithm can be used in cases where the lack of information would prohibit the use of MWAL.
 
@@
 if one uses LSPI) using LSTD more than once.
-This empirically shows with a ten-fold gain on computation time on a desktop computer compared to MWAL.
+We empirically notice a ten-fold gain on computation time on a desktop computer compared to MWAL.
 
-- We insist on retrieving the reward because task transfer (getting an agent with different capabilities than the expert's to fulfill the same task) cannot be done with a policy.
+- To clarify what we meant l.354, we insist on retrieving the reward because task transfer (getting an agent with different capabilities than the expert's to fulfill the same task) cannot be done with a policy.
 With a reward function on the states, it is possible to train any agent on the problem, even if the action space is different.

### 2012-04-12 klein2012/main.tex
Correction d'une typo mentionnée dans les revues :
Line 286: "1=1" should be "i=1" and N indexes the example rather than the time step here (for MMP).

Quoique je n'ai corrigé que la première partie, je comprends pas bien ce qu'il a voulu dire dans la seconde...
Mise en page d'EWRL

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Use the following line _only_ if you're still using LaTeX 2.09.
-%\documentstyle[icml2012,epsf,natbib]{article}
-% If you rely on Latex2e packages, like most moden people use this:
-\documentclass{article}
-
-% For figures
+%\documentclass[gray]{jmlr} % test grayscale version
+\documentclass[tablecaption=bottom]{jmlr}
+
+% The following packages will be automatically loaded:
+% amsmath, amssymb, natbib, graphicx, url, algorithm2e
+
+%\usepackage{rotating}% for sideways figures and tables
+%\usepackage{longtable}% for long tables
+
+% The booktabs package is used by this sample document
+% (it provides \toprule, \midrule and \bottomrule).
+% Remove the next line if you don't require it.
+\usepackage{booktabs}
+% The siunitx package is used by this sample document
+% to align numbers in a column by their decimal point.
+% Remove the next line if you don't require it.
+\usepackage[load-configurations=version-1]{siunitx} % newer version
+%\usepackage{siunitx}
 \usepackage{graphicx} % more modern
 %\usepackage{epsfig} % less modern
-\usepackage{subfigure}
+%\usepackage{subfigure}
 \usepackage{mathtools}
@@
 \usepackage{algorithmic}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{hyperref}
 \usepackage{url}
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-\usepackage{icml2012}
+%\newcommand{\theHalgorithm}{\arabic{algorithm}}
 \mathtoolsset{showonlyrefs=true}
 
-\newtheorem{definition}{Définition}
+%\newtheorem{definition}{Définition}
 \newtheorem{lemme}{Lemma}
-\newtheorem{theorem}{Théorème}
-\newtheorem{hypo}{Hypothèse}
-\newtheorem{cor}{Corollaire}
+%\newtheorem{theorem}{Théorème}
+%% \newtheorem{hypo}{Hypothèse}
+%% \newtheorem{cor}{Corollaire}
 \newtheorem{prop}{Proposition}
-\newtheorem{exemple}{Exemple}
-\newtheorem{remarques}{Remarques}
-\newtheorem{remarque}{Remarque}
+%% \newtheorem{exemple}{Exemple}
+%% \newtheorem{remarques}{Remarques}
+%% \newtheorem{remarque}{Remarque}
 \newtheorem{assumption}{Assumption}
-\newtheorem{problem}{Problème}
-\newtheorem{algo}{Algorithme}
-\newtheorem{preuve}{Démonstration}
+%% \newtheorem{problem}{Problème}
+%% \newtheorem{algo}{Algorithme}
+%% \newtheorem{preuve}{Démonstration}
 \newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Titre court si le titre fait plus de 40 caractères
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\shorttitle{Classification structurée pour l'ARI}
-%
-%\shortouvrage{CAp 2012}
-
-% Titre, auteur, pas de date
-
-%\title{Classification structurée pour l'apprentissage par renforcement inverse}
-%
-%\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
-%
-%\institute{
-%Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
-%\and
-%Equipe ABC,
-%LORIA, France
-%\and
-%UMI 2958
-%GeorgiaTech-CNRS, France
+\jmlrvolume{vol}
+\jmlryear{2012}
+\jmlrsubmitted{\today}
+\jmlrpublished{}
+
+\title[SCIRL]{Structured Classification for Inverse Reinforcement Learning}
+
+ % Use \Name{Author Name} to specify the name.
+ % If the surname contains spaces, enclose the surname
+ % in braces, e.g. \Name{John {Smith Jones}} similarly
+ % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
+ % If the first letter in the forenames is a diacritic
+ % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}
+
+ % \thanks must come after \Name{...} not inside the argument for
+ % example \Name{John Smith}\thanks{A note} NOT \Name{John
+ % Smith\thanks{A note}}
+
+ % Two authors with the same address
+  \author{\Name{Paper under review at ICML}\thanks{No author name provided so as not to breach ICML's double blind process}}
+
+ % Three or more authors with the same address:
+ % \author{\Name{Author Name1} \Email{an1@sample.com}\\
+ %  \Name{Author Name2} \Email{an2@sample.com}\\
+ %  \Name{Author Name3} \Email{an3@sample.com}\\
+ %  \Name{Author Name4} \Email{an4@sample.com}\\
+ %  \Name{Author Name5} \Email{an5@sample.com}\\
+ %  \Name{Author Name6} \Email{an6@sample.com}\\
+ %  \Name{Author Name7} \Email{an7@sample.com}\\
+ %  \Name{Author Name8} \Email{an8@sample.com}\\
+ %  \Name{Author Name9} \Email{an9@sample.com}\\
+ %  \Name{Author Name10} \Email{an10@sample.com}\\
+ %  \Name{Author Name11} \Email{an11@sample.com}\\
+ %  \Name{Author Name12} \Email{an12@sample.com}\\
+ %  \Name{Author Name13} \Email{an13@sample.com}\\
+ %  \Name{Author Name14} \Email{an14@sample.com}\\
+ %  \addr Address}
+
+ % Authors with different addresses:
+ % \author{\Name{Author Name1} \Email{abc@sample.com}\\
+ % \addr Address 1
+ % \AND
+ % \Name{Author Name2} \Email{xyz@sample.com}\\
+ % \addr Address 2
  %}
-\icmltitlerunning{Structured Classification for Inverse
-Reinforcement Learning}
+
+%\editor{Editor's name}
 
@@
 
-\twocolumn[ \icmltitle{Structured Classification for Inverse
-Reinforcement Learning}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2012
-% package.
-\icmlauthor{Your Name}{email@yourdomain.edu}
-\icmladdress{Your Fantastic Institute,
-            314159 Pi St., Palo Alto, CA 94306 USA}
-\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
-\icmladdress{Their Fantastic Institute,
-            27182 Exp St., Toronto, ON M6H 2T1 CANADA}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{inverse reinforcement learning, structured multi-class
-classification}
-
-\vskip 0.3in
-]
+\maketitle
 
@@
 \end{abstract}
+\begin{keywords}
+inverse reinforcement learning, structured multi-class
+classification
+\end{keywords}
+
 %%%%%
@@
 \begin{equation}
-  J(\theta) = \frac{1}{N}\sum_{1=1}^N \max_\pi(\theta^T
+  J(\theta) = \frac{1}{N}\sum_{i=1}^N \max_\pi(\theta^T
   \mu^\pi_{d_0,\mathcal{M}_i} + \mathcal{L}_{\mathcal{M}_i,\pi}) - \theta^T
@@
 \bibliography{Biblio}
-\bibliographystyle{icml2012}
+%\bibliographystyle{icml2012}
 \end{document}

### 2012-04-20 piot2012cascading/main_PIOT_v2.tex
Article envoyé par Bilal

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% For figures
\usepackage{graphicx} % more modern
%\usepackage[latin1]{inputenc}
%\usepackage[francais]{babel}
\usepackage{subfigure}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{dsfont}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\hypersetup{
    colorlinks,%
    citecolor=black,%
    filecolor=black,%
    linkcolor=black,%
    urlcolor=black
}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\mathtoolsset{showonlyrefs=true}
\newtheorem{hypo}{Hypothesis}
\newcommand{\argmax}{\operatorname*{argmax}}
\newcommand{\argmin}{\operatorname*{argmin}}
\newcommand{\arginf}{\operatorname*{arginf}}
\newcommand{\minp}{\operatorname*{min_+}}
\newcommand{\Ker}{\operatorname*{Ker}}
\newcommand{\trace}{\operatorname*{trace}}
\newcommand{\cov}{\operatorname{cov}}
\newcommand{\card}{\operatorname*{Card}}
\newcommand{\vect}{\operatorname*{Vect}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\erf}{\operatorname{erf}}
\newcommand{\bound}{\operatorname*{bound}}
\newcommand{\vpi}{\operatorname{VPI}}
\newcommand{\gn}{\operatorname{Gain}}
\newcommand{\p}{\operatorname{Pr}}
\newcommand{\mlp}{\operatorname{MLP}}
\newcommand*\tto[2]{\smash{\mathop{\longrightarrow}\limits_{#1}^{#2}}}
\newcommand*\ntto[2]{\smash{\mathop{\nrightarrow}\limits_{#1}^{#2}}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bd}{\mathbf{d}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\af}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bh}{\mathbf{h}}
\newcommand{\bi}{\mathbf{i}}
\newcommand{\bj}{\mathbf{j}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bl}{\mathbf{l}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bo}{\mathbf{o}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bt}{\mathbf{t}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\ma}{\mathbf{A}}
\newcommand{\mb}{\mathbf{B}}
\newcommand{\mc}{\mathbf{C}}
\newcommand{\md}{\mathbf{D}}
\newcommand{\me}{\mathbf{E}}
\newcommand{\mf}{\mathbf{F}}
\newcommand{\mg}{\mathbf{G}}
\newcommand{\mh}{\mathbf{H}}
\newcommand{\mi}{\mathbf{I}}
\newcommand{\mj}{\mathbf{J}}
\newcommand{\mk}{\mathbf{K}}
\newcommand{\ml}{\mathbf{L}}
\newcommand{\mm}{\mathbf{M}}
\newcommand{\mn}{\mathbf{N}}
\newcommand{\mo}{\mathbf{O}}
\newcommand{\Mp}{\mathbf{P}}
\newcommand{\mq}{\mathbf{Q}}
\newcommand{\mr}{\mathbf{R}}
\newcommand{\ms}{\mathbf{S}}
\newcommand{\mt}{\mathbf{T}}
\newcommand{\Mu}{\mathbf{U}}
\newcommand{\mv}{\mathbf{V}}
\newcommand{\mw}{\mathbf{W}}
\newcommand{\mx}{\mathbf{X}}
\newcommand{\my}{\mathbf{Y}}
\newcommand{\mz}{\mathbf{Z}}
\newcommand{\tphi}{\tilde{\Phi}}
\newcommand{\espace}{\text{ }}
\newcommand{\x}{\mathbf{x}}
\newcommand{\s}{\mathbf{s}}
\newcommand{\n}{\mathbf{n}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\rr}{\mathbf{r}}
\newcommand{\0}{\mathbf{0}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\am}{{\mathcal{A}_m}}
\newcommand{\amj}{{\mathcal{A}_m^{+j}}}
\newcommand{\sgn}{\operatorname{sgn}}
\begin{document}

\title{Titre}
\titlerunning{Titre}  % abbreviated title (for running head)
\author{Boul\inst{1} \and Bil\inst{2}
Boul \and Bil}
\authorrunning{} % abbreviated author list (for running head)
%
\institute{SUPELEC,\\
\email{SUPELEC},\\ WWW home page:
\texttt{SUPELEC}
\and
\\}
\maketitle              % typeset the title of the contribution
\begin{abstract}
This paper considers the problem of learning from demonstrations and more particularly the inverse reinforcement learning (IRL) problem.
The IRL framework assumes that an expert, demonstrating a task, is acting optimally in a Markov Decision Process (MDP) with respect to an unknown reward function to be discovered.
Unlike most of existing IRL algorithms, the proposed approach doesn't require any of the following: complete trajectories from an expert, a generative model of the environment, the knowledge
of the transition probabilities, the ability to compute a (near)-optimal policy for different rewards and hence resolve the forward reinforcement learning (RL) problem, the expert's policy. In addition, only a small amount of expert demonstrations (not even in the form of trajectories but simple transitions) is required. Thus, this new approach which consists in cascading two well-known supervised learning (SL) methods is a model-free and veritable IRL approach in the sense that the output of our algorithm is a non-trivial reward function for which the expert policy is near optimal. Finally, we provide an analysis which legitimates this new approach.
\end{abstract}

\section{Introduction}

Given a sequential decision making problem posed in the MDP formalism, it is often difficult to specify manually a reward function in order to realize a given task. However, observing an expert demonstrating a task is often possible and more easier~\cite{ng2000algorithms}. That is why learning from demonstrations and more particularly IRL is an interesting problem. The IRL framework, fisrt introduced in~[\cite{russell1998learning},\cite{ng2000algorithms}], assumes that an expert, demonstrating a task, is acting optimally in a MDP with respect to an unknown reward function to be discovered. Unlike direct methods of learning from demonstrations which learn a mapping from states to actions via a supervised learning~[\cite{atkeson1997robot},\cite{pomerleau1989alvinn}], IRL methods must recover a reward function which is the most compact and tranferable way to specify a task. Moreover most of the direct methods do not use the structure of the MDP (one notable excepetion is~\cite{melo2010learning}), hence mimic the policy of the expert and are not able to generalize correctly the expert policy outside state examples. In order to overcome this drawback, undirect methods known as apprenticeship learning were introduced: the first one is~\cite{ng2004feature}. These methods (see~\cite{neu2009training} for a good survey) uses the IRL framework but most of them assume the dynamic of the environment known, or an environment model (via a simulator) must be available so as to test the effects of a policy. Moreover, most of existing undirect approaches require to solve the direct reinforcement learning problem (find an optimal policy knowing a given reward) several times (one notable exception is~\cite{boularias2011relative}). Finally, in many cases, the output of the algorithm is a policy but not a reward function (see section \ref{section: related work} for an in-depth presentation). These constraints can lead to the development of algorithms unapplicable to real problems and make the problem of apprenticeship learning harder than reinforcement learning. That is why a reborn interest for direct methods which passively imitate the expert exists. More particularly the reduction of the apprenticeship learning to classification, first introduced in~\cite{zadrozny2003cost} and recently used in~\cite{melo2010learning}, is legitimated for finite horizons problems in~[\cite{syed2010reduction},\cite{ross2010efficient}]. Our approach tries to avoid the drawbacks of the direct and undirect methods by cascading two well known supervised learning(see section \ref{section: Cascading}): first a step of classification and then a step of regression which introduces the MDP structure and  outputs a reward function (see \ref{section: Cascading}). Our approach doesn't require any of the following: complete trajectories from an expert, a generative model of the environment, the knowledge of the transition probabilities, the ability to compute a (near)-optimal policy for different rewards and hence resolve the forward reinforcement learning (RL) problem, the expert's policy. In addition, only a small amount of expert demonstrations (not even in the form of trajectories but simple transitions) is required.\\
The remainder of the paper is oraganized as follows: section \ref{section: background} gives us the different notations used in the paper and presents briefly the RL and IRL frameworks, section \ref{section: Cascading} present our method which is legitimated in section \ref{section: Cascading}, \ref{section: experiments} presents an instanciation of our method and tests it on different benchmarks and finally section \ref{section: related work} compares our method to the state of the art.

\section{Background and Notations}
\label{section: background}
A (finite) MDP \cite{puterman1994markov} is a tuple $\{S,A,P,\gamma,R\}$ where $S=\{s_i\}_{1\leq i \leq N}$ is the finite state space with $N\in\mathbb{N}^*$ states, $A=\{a_k\}_{1\leq k \leq K}$ is the finite action space with $K\in\mathbb{N}^*$ actions, $P=\{P_{s,a}\}_{(s,a)\in S\times A}$ is the set of transition probabilities where $P_{s,a}$ is a distribution probability over $S$, $P$ represents the dynamics of the MDP and the notation $p(s,a,s')=P_{s,a}(s')$ is often used and quantifies the probability to reach $s'\in S$ knowing that the action $a \in A$ was taken in the state $s\in S$, $\gamma\in]0,1[$ is the discount factor and $R$ is a function from $S\times A$ to $\mathbb{R}$ called the reward function. A deterministic policy $\pi$ is a function from $S$ to $A$, $\Pi$ is the set of deterministic policies and the value function $V^\pi_R$ is a function from $S$ to $\mathbb{R}$ defined by:
\begin{equation}
V^\pi_R(s)=E^\pi_s[\sum_{t=0}^{+\infty}\gamma^tR(s_t,\pi(s_t))], \forall s \in S,
\end{equation}
where $E^\pi_s$ is the expectation over the distribution of the trajectories $(s_0,s_1,\dots)$ obtained by executing the policy $\pi$ starting from $s_0=s$.
The action-value function $Q^\pi_R$ is a function from $S\times A$ to $R$ defined by:
\begin{equation}
Q^\pi_R(s,a)=E^\pi_{s,a}[R(s_0,a)+\sum_{t=1}^{+\infty}\gamma^tR(s_t,\pi(s_t))], \forall s \in S,\forall a \in A,
\end{equation}
where $E^\pi_{s,a}$ is the expectation over the distribution of the trajectories $(s_0,s_1,\dots)$ obtained starting from $s_0=s$, executing $a$ and then following the policy $\pi$.
The classical results for finite MDPs are the well-known Bellman equations \cite{sutton1998reinforcement}:
\begin{align}
\label{equation: Bellman}
&V^{\pi}_R(s)=R(s,\pi(s))+\gamma\sum_{s'\in S}P_{s,\pi(s)}(s')V^{\pi}_R(s'), \forall s\in S,
\\
&Q^{\pi}_R(s,a)=R(s,a)+\gamma\sum_{s'\in S}P_{s,a}(s')V^{\pi}_R(s'), \forall s\in S, \forall a\in A.
\end{align}
A policy $\pi$ is said optimal for the reward $R$ when:
\begin{equation}
\label{equation:Voptimal}
V^{\pi}_R(s)\geq V^{\tilde{\pi}}_R(s) , \forall s\in S, \forall \tilde{\pi}\in\Pi.
\end{equation}
Another classical result called the Bellman optimality is: $\pi$ is an optimal policy for the reward $R$ if and only if:
\begin{equation}
\label{equation:Qoptimal}
\pi(s)\in\argmax_{a\in A} Q^\pi_R(s,a), \forall s\in S.
\end{equation}
%Inversely, a real column-vector $X=[X_i]_{1\leq i \leq N}$ can be identified to the function from $S$ to $\mathbb{R}$ that maps $s_i$ to $X_i$ for all $i\in\{1,\dots,N\}$ and the real rectangular-matrice $Y=[Y_{i,k}]_{1\leq i \leq N, 1\leq k \leq K}$ can be identified to the function from $S\times A$ to $\mathbb{R}$ that maps $(s_i,a_k)$ to $Y_{i,k}$ for all $(i,k)\in\{1,\dots,N\}\times\{1,\dots,K\}$.
For a finite-MDP $\{S,A,P,\gamma,R\}$ the existence of an optimal policy for any function $R$ is shown in \cite{bertsekas2001dynamic}.
As $S=\{s_i\}_{1\leq i \leq N}$ and $A=\{a_i\}_{1\leq i \leq K}$ are finite, a function $X$ from $S$ to $\mathbb{R}$ can be identified to the real column-vector $[X(s_i)]_{1\leq i \leq N}$ and a function $Y$ from
$S\times A$ to $\mathbb{R}$ can be identified to the real rectangular-matrice of $N$ rows and $K$ columns  $[Y(s_i,a_k)]_{1\leq i \leq N, 1\leq k \leq K}$.
Thus the notation $V^\pi_R$ depending on the context will be considered as a function or a column vector. Moreover if $Y$ is a function from $S\times A$ to $\mathbb{R}$ then the notation $Y_a$ will mean that $Y_a$ is the function from $S$ to $R$ that maps $s$ to $Y(s,a)$. So the notation $R_a$ can be a function or a column-vector depending on the context.
We will also need the following matrices and column-vectors defined for any dynamic $P$, any policy $\pi$ and reward function $R$ :$P_a=[P_{s_i,a}(s_j)]_{1\leq i,j \leq N}$, $P_\pi=[P_{s_i,\pi(s_i)}(s_j)]_{1\leq i,j \leq N}$, $R_\pi=[R(s_i,\pi(s_i))]_{1\leq i,j \leq N}$. The transposition of a real column-vector $X=[X_i]_{1\leq i \leq M}$, $M\in\mathbb{N}^*$, is the real row-vector $X^T$ and $\|X\|_{\infty}=\max_{1\leq i \leq N}|X_i|$. The transposition of the real matrice $Y=[Y_{i,j}]_{1\leq i \leq I, 1\leq j \leq J}$, $(I,J)\in\mathbb{N}^2$, will be $Y^T$ and $\|Y\|_{\text{max}}=\max_{1\leq i \leq I,1\leq j \leq J}|Y_{i,j}|$.  Thanks to these vectorial notations, we can define the Bellman operator $T^\pi_R$, for a given deterministic policy $\pi$ and a given reward $R$, which is a function from $\mathbb{R}^N$ to $\mathbb{R}^N$ such that:
\begin{equation}
T^\pi_RX=R_\pi+\gamma P_\pi X , \forall X\in \mathbb{R}^N.
\end{equation}
It is well known that the unique fixed point of the operator $T^\pi_R$ is the column-vector $V^\pi_R$ \cite{puterman1994markov}.\\
RL consists in finding an optimal policy for the reward function $R$. Notice that the state space may be too large for an exact representation of the value function (which calls for approximate
representation), that the model (P and R) may be unknown (the only information being provided through rewarded transition sampled according
to some behavioral policy), that learning can occurs in an online or offline setting, and so on. There exists good books on the subject [\cite{bertsekas2001dynamic},\cite{sutton1998reinforcement}].\\
In the classical IRL paradigm \cite{ng2000algorithms}, an MDP without reward $\{S,A,P,\gamma\}$ and a policy $\pi_E$ called expert-policy are given and the problem is to find
a reward $R^*$ for which the policy $\pi_E$ will be optimal. However this problem is clearly ill-posed in the sense that there is no uniqueness of the reward $R^*$ : many rewards functions are equivalent in the sense that they have the same optimal deterministic policies \cite{ng1999policyreward}, moreover the trivial zero-reward is a solution for any deterministic policy $\pi_E$ as it is shown in \cite{ng2000algorithms}. In the literature, some solutions are proposed in order to respond to the ill-posed nature of the problem (\cite{ng2000algorithms},\cite{ziebart2008maximum},\cite{boularias2011relative}).
In this paper, we assume that the solely available information is provided by transitions sampled according to the expert policy $\pi_E$:
\begin{equation}
\label{equation:data}
\{(s_i,a_i=\pi_E(s_i)),s_i')\}_{1\leq i \leq D}, D\in\mathbb{N}^*.
\end{equation}
The reward function is obviously unknown, but this assumption means that the dynamics ($P$) is only known through transitions $(s_i, a_i, s_i')$ and that the
policy $\pi_E$ is only known through state-action pairs $(s_i, a_i)$.\\
Finally let's recall some basic but important results of finite-Markov chain theory.
\begin{definition}
Let $Y=[Y_{i,j}]_{1\leq i,j \leq M}$ be a  stochastic square matrix of size $M\in\mathbb{N}^*$. Then $Y$ is said to be irreducible if:
\begin{equation}
U(i,j)=\sum_{k=0}^{+\infty}Y^k(i,j)>0 ,\forall (i,j)\in \{1,\dots,M\}^2.
\end{equation}
\end{definition}
\begin{theorem}
 Let $Y$  be a stochastic square matrix of size $M\in\mathbb{N}^*$. $Y$ is irreducible if and only if there exists a unique and strictly positive distribution $\mu$ over $\{1,\dots,M\}$, $\mu$ is a function from $\{1,\dots,M\}$ to $\mathbb{R}$ which can be seen as a real column vector of size $M$, such that:
\begin{equation}
\label{equation: stationarity}
\mu^T=\mu^TY.
\end{equation}
$\mu$ is called the stationary distribution of $Y$.
\end{theorem}
\begin{definition}
Let $Y$ be a stochastic square matrix of size $M\in\mathbb{N}^*$ and irreducible. It is aperiodic if for all $(i,j)\in {1,\dots,M}^2$ it exists $n_0\in \mathbb{N}^*$ such that for all $n\geq n_0$: $Y^n(i,j)>0$.
\end{definition}
\begin{theorem}
\label{theoreme : mixing exponential}
Let $Y$ be a stochastic square matrix of size $M\in\mathbb{N}^*$. If $Y$ is irreducible and aperiodic then:
\begin{equation}
\lim_{k\rightarrow +\infty}\sum_{j=1}^{j=M}[Y^k(i,j)-\mu(j)]\rightarrow 0,\forall i\in\{1,\dots,M\},
\end{equation}
and more precisely there exists $\alpha\in]0,1[$ and $C\in\mathbb{R}_+$ such that:
\begin{equation}
\sum_{j=1}^{j=M}|Y^k(i,j)-\mu(j)|\leq C\alpha^k,\forall i\in\{1,\dots,M\},\forall k\in\mathbb{N}^*.
\end{equation}
\end{theorem}
In our framework, if $P_\pi$ is irreducible, then $\mu_\pi$ will be its stationary distribution.

%Other notations will be needed in order to prove that the cascading approach is legitimate. Let's define the column-vector $\mu^{t,\pi,s}=[\mu^{t,\pi,s}_i]_{1\leq i \leq N}$ where $\mu^{t,\pi,s}_i$ is the probability of being in the state $s_i\in S$ at time $t\in\mathbb{N}$ when the start-state is $s$ and the policy followed is $\pi$ :
%\begin{equation}
%\mu^{t,\pi,s}=(\Delta_s^T(P_\pi)^t)^T,
%\end{equation}
%where $\Delta_s$ is the column-vector: $\Delta_s={\delta(s_i,s)}_{1\leq i \leq N}$ with $\delta(.,.)$ the Kronecker symbol. For a fixed policy $\pi$, the matrix $P_\pi$ is a stochastic matrix of finite dimension and if it is ergodic and recurrent positive then there exists a unique distribution probability represented by the column vector $\mu^\pi=[\mu^{\pi}_i]_{1\leq i \leq N}$ such that :
%\begin{equation}
%\mu_\pi=(\mu_\pi^T P_\pi)^T.
%\end{equation}
%In that case $\mu^\pi$ is called the stationary distribution of the policy $\pi$ and :
%\begin{equation}
%\forall s\in S \lim_{t\rightarrow+\infty}\mu_{t,\pi,s}=\mu_\pi.
%\end{equation}
%Thus in the particular case of ergodicty and recurrent positivity we can easily express $\mu_\pi^TV^\pi_R$ :
%\begin{align}
%\mu_\pi^TV^\pi_R&=\mu_\pi^T(R_\pi+\gamma P_\pi V^\pi_R)=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TP_\pi V^\pi_R,
%\\
%&=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\mu_\pi^TR_\pi.
%\end{align}
\section{Cascading Classification and Regression for IRL}
\label{section: Cascading}
In this section, we propose to cascade two supervised learning (SL) approaches (namely a classifier and a regressor) to solve the IRL problem.
\subsection{The general idea}
The available data being depicted in \eqref{equation:data}, a first problem would be to generalize the expert policy $\pi_E$ outside state examples provided in this dataset. A first approach, recently used in \cite{melo2010learning} and legitimated for finite-horizon problems in \cite{syed2010reduction}, is to frame this problem as learning a multi-classifier where each action $a \in A$ will be a class. Indeed, trained on the training set $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$, many classifiers are able to output a score function $q$ from $S\times A$ to $\mathbb{R}$. The final decision rule $\pi_C$ which is a function from $S$ to $\mathbb{R}$ and thus can be seen as a deterministic policy is constructed in order to satisfy:
\begin{equation}
\label{equation:Coptimal}
\pi_C(s)\in\argmax_{a\in A}q(s,a),\forall s\in S.
\end{equation}
Now, one can notice the similarity between \eqref{equation:Qoptimal} and \eqref{equation:Coptimal}. The score-function $q$ can then be interpreted as an action-value function and it is easy and natural to construct a reward function $R_C$ for which the policy $\pi_C$ is optimal as it will be proven in the section \ref{section: Analysis}. Let define $R^C$ as:
\begin{equation}
R^C(s,a)=q(s,a)-\gamma\sum_{s'\in S}p(s,a,s')q(s',\pi_C(s')), \forall a \in A, \forall s\in S.
\end{equation}
However, the model being unknown (contrary to $q$ and $\pi_C$) it is not possible to compute $R^C$ directly. However, it can be estimated from the available transitions (see
\eqref{equation:data}). Let us write:
\begin{equation}
\hat{r}_i=q(s_i,a_i)-\gamma q(s_i',\pi_C(s_i')), \forall i\in \{1,\dots,D\}.
\end{equation}
This is an unbiased estimate of $R_C(s_i,a_i)$, indeed:
\begin{equation}
R_C(s_i,a_i)=\sum_{s'\in S}p(s_i,a_i,s')[q(s_i,a_i)-\gamma q(s',\pi_C(s'))], \forall i\in \{1,\dots,D\}.
\end{equation}
Therefore, it is possible to build an estimate $\hat{R}_C$ of $R_C$, using a regressor trained on the dataset:
\begin{equation}
\{(s_i,a_i,s'_i,\hat{r}_i)\}_{1\leq i \leq D}.
\end{equation}
Finally, we see $\hat{R}_C$ as an approximation of $R_C$ and $\hat{\pi}_C$ will be an optimal policy for the reward $\hat{R}_C$. In order to verify that the reward function $\hat{R}_C$ is a good candidate to resolve the IRL problem, it will be proven in the section \ref{section: Analysis} that the policies $\pi_E$ and $\hat{\pi}_C$ are quite similar (confer lemma \ref{lemma : results}).
\section{Analysis} \label{section: Analysis}
This section is devoted to show, under some hypotheses, that the cascading approach is legitimate. The first result is a lemma which gives a practical way to calculate $\mu_\pi^TV^{\pi}_R$ for a given policy $\pi$ and a given reward function $R$. The second result is a theorem which gives an upper bound to the term $\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}$ where $\mu_E$ is the stationary distribution of the expert policy $\pi_E$. We also give an interpretation of the term $\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}$ and explain why being able to bound this term means that our approach is legitimate.
\subsection{Results and Discussion}
\begin{lemma}
\label{lemma: calculs V}
Let $\{S,A,P,\gamma,R\}$ be a finite MDP and $\pi$ a deterministic policy.
If $P_\pi$ is reducible, then $\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\mu_\pi^TR_\pi$.
\end{lemma}
The first lemma gives a practical tool which will be useful in order to simplify some terms in the proof of the next theorem. Moreover it is useful
to notice that the term  $\mu_\pi^TV^\pi_R$ can be reinterpreted as an expectation. Indeed we have:
\begin{equation}
\mu_\pi^TV^\pi_R=E_{s \sim \mu_\pi}[V^\pi_R(s)],
\end{equation}
where $E_{s \sim \mu_\pi}$ means that $s$ is distributed over the distribution $\mu_\pi$. All the results provided in the next theorem use the expectation $E_{s \sim \mu_E}$ which is the canonical expectation
to consider when one wants to prove a result over the state space $S$ related to the policy of the expert $\pi_E$.\\
Before giving an upper bound to $\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})=E_{s \sim \mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$, we define $\epsilon_C\in\mathbb{R}_+$ called the classification error and the function $\epsilon_R$ from $S\times A$ to $\mathbb{R}$ called the regression error such that:
\begin{align}
&\epsilon_C=\sum_{s\in S}\mu_{E}(s)\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}=E_{s \sim \mu_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}],
\\
&\epsilon_R(s,a)=\hat{R}^C(s,a)-R^C(s,a), \forall a\in A, \forall s\in S.
\end{align}
The column vectors $\epsilon^C_R=[\epsilon_R(s_i,\hat{\pi}_C(s_i))]_{1\leq i \leq N}$ and $\epsilon^E_R=[\epsilon_R(s_i,\pi_E(s_i))]_{1\leq i \leq N}$ will also be needed.
\begin{remark}
One can notice that the error classification is defined thanks to the expectation $E_{s \sim \mu_E}$. In the classical framework of classification the data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$
are generated independently according to a distribution $\mu_{\text{Data}}$ over $S$ and hence the classical classification error must be:
\begin{equation}
\epsilon_C=E_{s \sim \mu_{\text{Data}}}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
\end{equation}
Then one can suppose that our data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$ are generated independently thanks to $\mu_E$ but this assumption is quite strong and non-realistic.
However it is often the case that the data provided by the expert are trajectories. In that case the theorem \ref{theoreme : mixing exponential}, which says that the rate of convergence
to the stationary distribution is at least exponential, allows us to suppose that the data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$ are generated under $\mu_E$.
\end{remark}
\begin{theorem}
\label{theorem : results}
Let $\{S,A,P,\gamma\}$ be a finite MDP without reward and $\pi_E$ an expert policy. The only available data is depicted in \eqref{equation:data}.
The notations $q$, $\pi_C$, $\hat{\pi}_C$, $\hat{R}^C$ are introduced in the section \ref{section: Cascading}.
If $P_{\pi_E}$ is reducible, then $\mu_E$ is the stationary distribution of $\pi_E$ and:
\begin{enumerate}
\item $\pi_C$ is optimal for the reward $R^C$.
\item $\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}$.
\item $\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\|\epsilon^E_R\|_{\infty}}{1-\gamma}$.
\end{enumerate}
\end{theorem}
In order to understand why this theorem is useful, let us make some important assumptions. Let us suppose that the classification and the regression steps are perfect in the sense that $\epsilon_R(s,a)=0,\forall (s,a)\in S\times A$ and $\epsilon_C=0$. Then we obviously have, thanks to the theorem \ref{theorem : results}, that $\pi_E=\pi_C$ is optimal for $\hat{R}_C=R_C$. Thus the method is able to provide a non-trivial reward function for which the policy $\pi_E$ is optimal. Moreover if the classification step and the regression step are not perfect, the theorem \ref{theorem : results} shows, that our approach is able to provide a non trivial-reward $\hat{R}_C$ for which the policy $\pi_E$ is near-optimal in the sense that:
\begin{equation}
E_{s \sim \mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq B[\epsilon_C+\|\epsilon_R\|_{\text{max}}], B\in\mathbb{R}_+,
\end{equation}
which means that the value functions of $\pi_E$ and $\hat{\pi}_C$ are not so different in the most visited states by the policy $\pi_E$.

\subsection{Proofs}

\begin{proof}[Lemma \ref{lemma: calculs V}]
Here, we use \eqref{equation: stationarity}.
\begin{align}
\mu_\pi^TV^\pi_R&=\mu_\pi^T(R_\pi+\gamma P_\pi V^\pi_R)=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TP_\pi V^\pi_R,
\\
&=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\mu_\pi^TR_\pi.
\end{align}
\end{proof}

%\begin{remark}
%Let use the notations of the lemma \ref{lemma: calculs V}. For each $s\in S$, we define the average reward $\rho^\pi_R(s)$ such that :
%\begin{equation}
%\rho^\pi_R(s)=\lim_{n\rightarrow+\infty}\frac{1}{n}E^\pi_s[\sum_{t=0}^{n-1}\gamma^tR(s_t,\pi(s_t))]
%\end{equation}
%If $P_\pi$ is reducible then $\rho^\pi_R(s)$ does not depend on $s$ and:
%\begin{equation}
%\rho_R^\pi(s)=\rho_R^\pi=\mu_\pi^TR_\pi,\forall s\in S.
%\end{equation}
%The proof of that result is essentially an application of the ergodic theorem (see \cite{puterman1994markov}).
%So, if $P_\pi$ is reducible:
%\begin{equation}
%\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\rho_R^\pi.
%\end{equation}
%There is a clear link between the average reward and the discounted reward.
%\end{remark}

\begin{proof}[Theorem \ref{lemma : results}]
In order to prove the tree results of the lemma \ref{lemma : results}, it is interesting to introduce the function $R_E$ from $S\times A$ to $\mathbb{R}$ such that:
\begin{equation}
R^E(s,a)=q(s,a)-\gamma\sum_{s'\in S}p(s,a,s')q(s',\pi_E(s')), \forall a \in A, \forall s\in S.
\end{equation}
The first step is to show that:
\begin{align}
&q(s,\pi_C(s))=V^{\pi_C}_{R^C}(s), \forall s\in S,
\\
&q(s,\pi_E(s))=V^{\pi_E}_{R^E}(s), \forall s\in S.
\end{align}
This is quite straightforward because the column vector $q_{\pi_E}=[q(s,\pi_E(s))]_{1\leq i\leq N}$ is the fixed point of the operator $T^{\pi_E}_{R^E}$ and  $q_{\pi_C}=[q(s,\pi_C(s))]_{1\leq i\leq N}$ is the fixed point of the operator $T^{\pi_C}_{R^C}$:
\begin{align}
T^{\pi_E}_{R^E}(q_{\pi_E})&=R^E_{\pi_E}+\gamma P_{\pi_E}q_{\pi_E},
\\
&=q_{\pi_E}-\gamma P_{\pi_E}q_{\pi_E}+\gamma P_{\pi_E}q_{\pi_E}=q_{\pi_E},
\\
T^{\pi_C}_{R^C}(q_{\pi_C})&=R^C_{\pi_C}+\gamma P_{\pi_C}q_{\pi_C},
\\
&=q_{\pi_C}-\gamma P_{\pi_C}q_{\pi_C}+\gamma P_{\pi_C}q_{\pi_C}=q_{\pi_C}.
\end{align}
Moreover it is clear that $q_a=[q(s_i,a)]_{1\leq i\leq N}=Q^{\pi_C}_{R^C,a}=[Q^{\pi_C}_{R^C}(s_i,a)]_{1\leq i\leq N}$ for all $a \in A$:
\begin{align}
Q^{\pi_C}_{R^C,a}&=R^C_a+\gamma P_a V^{\pi_C}_{R^C}, \forall a\in A,
\\
&=R^C_a+\gamma P_a q_{\pi_C}=q_a-\gamma P_a q_{\pi_C} + \gamma P_a q_{\pi_C}=q_a, \forall a\in A.
\end{align}
So $q(s,a)=Q^{\pi_C}_{R^C}(s,a),\forall s\in S,\forall a\in A$ and as:
\begin{equation}
\pi_C(s)\in\argmax_{a\in A}q(s,a), \forall s\in S,
\end{equation}
$\pi_C$ is optimal for the reward $R_C$.
Now let us prove that:
\begin{equation}
\mu_E^TV^{\pi_C}_{R^C}-\mu_E^TV^{\pi_E}_{R^C}\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}.
\end{equation}
Indeed:
\begin{equation}
\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})=\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E}+V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}).
\end{equation}
And $\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})$ is such that:
\begin{align}
\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})&=\sum_{s\in S}\mu_E(s)[V^{\pi_C}_{R^C}(s)-V^{\pi_E}_{R^E}(s)],
\\
&=\sum_{s\in S}\mu_E(s)[q(s,\pi_C(s))-q(s,\pi_E(s))]
\\
&=\sum_{s\in S}\mu_E(s)[q(s,\pi_C(s))-q(s,\pi_E(s))]\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}},
\\
&\leq\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-q(s,\pi_E(s))),
\\
&\leq\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)).
\end{align}
It remains to deal with the term $\mu_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})$ using the fact that $\mu_E^TP_{\pi_E}=\mu_E^T$ and the lemma \ref{lemma: calculs V}:
\begin{align}
\mu_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\frac{1}{1-\gamma}\mu_E^T(R^E_{\pi_E}-R^C_{\pi_E}),
\\
&=\frac{1}{1-\gamma}\mu_E^T(\gamma P_{\pi_E}q_{\pi_C}-\gamma P_{\pi_E}q_{\pi_E}),
\\
&=\frac{\gamma}{1-\gamma}\mu_E^T(q_{\pi_C}-q_{\pi_E}),
\\
&=\frac{\gamma}{1-\gamma}\sum_{s\in S}\mu_E(s)[q(s,\pi_C(s))-q(s,\pi_E(s))]\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}},
\\
&\leq\frac{\gamma}{1-\gamma}\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-q(s,\pi_E(s))),
\\
&\leq\frac{\gamma}{1-\gamma}\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)).
\end{align}
Finally:
\begin{align}
\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})&=\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E}+V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}),
\\
&\leq\frac{(\gamma+1-\gamma)\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma},
\\
&=\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}.
\end{align}
In order to finish the proof it remains to show that:
\begin{equation}
\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\|\epsilon^E_R\|_{\infty}}{1-\gamma}.
\end{equation}
We notice that:
\begin{equation}
\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})=\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}+V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}+V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}).
\end{equation}
It is very easy to see that:
\begin{align}
&\|V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}\|_{+\infty}\leq\frac{\|\epsilon^C_R\|_{\infty}}{1-\gamma},
\\
&\|V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}\|_{+\infty}\leq\frac{\|\epsilon^E_R\|_{\infty}}{1-\gamma},
\end{align}
Indeed:
\begin{align}
V^{\hat{\pi}_C}(s)_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}(s)&=E^{\hat{\pi}_C}_s[\sum_{t=0}^{+\infty}\gamma^t(\hat{R}^C(s_t,\hat{\pi}_C(s_t))-R^C(s_t,\hat{\pi}_C(s_t)))], \forall s\in S,
\\
&\leq\frac{\|\epsilon^C_R\|_{\infty}}{1-\gamma}, \forall s\in S.
\\
V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}(s)&=E^{\pi_E}_s[\sum_{t=0}^{+\infty}\gamma^t(R^C(s_t,\pi_E(s_t))-\hat{R}^C(s_t,\pi_E(s_t)))], \forall s\in S,
\\
&\leq\frac{\|\epsilon^E_R\|_{\infty}}{1-\gamma}, \forall s\in S.
\end{align}
Thus:
\begin{equation}
\label{equation: cas3-1}
\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}+V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})\leq\frac{\|\epsilon^E_R\|_{\infty}+\|\epsilon^C_R\|_{\infty}}{1-\gamma}.
\end{equation}
It remains to deal with the term $\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})$:
\begin{equation}
\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})=\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C}+V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C}).
\end{equation}
As $\pi_C$ is optimal for the reward $R^C$ then:
\begin{equation}
\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C})\leq 0.
\end{equation}
So:
\begin{equation}
\label{equation: cas3-2}
\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}.
\end{equation}
Finally by regrouping the results in \eqref{equation: cas3-1} and \eqref{equation: cas3-2}:
\begin{align}
\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}&\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\|\epsilon^E_R\|_{\infty}}{1-\gamma},
\\
&\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+2\|\epsilon_R\|_{\text{max}}}{1-\gamma}.
\end{align}

\end{proof}

\section{Experiments}
\label{section: experiments}

\section{Related Work}
\label{section: related work}

\section{Conclusion}

\bibliographystyle{splncs}
\bibliography{Biblio}
\end{document}

### 2012-04-20 piot2012cascading/main_KLEIN_v3.tex
Ajout d'un embryon de partie expérimentale

--- 2012-04-20 piot2012cascading/main_PIOT_v2.tex
@@
 \begin{abstract}
-This paper considers the problem of learning from demonstrations and more particularly the inverse reinforcement learning (IRL) problem.
+This paper considers the inverse reinforcement learning (IRL) problem.
 The IRL framework assumes that an expert, demonstrating a task, is acting optimally in a Markov Decision Process (MDP) with respect to an unknown reward function to be discovered.
-Unlike most of existing IRL algorithms, the proposed approach doesn't require any of the following: complete trajectories from an expert, a generative model of the environment, the knowledge
-of the transition probabilities, the ability to compute a (near)-optimal policy for different rewards and hence resolve the forward reinforcement learning (RL) problem, the expert's policy.
-In addition, only a small amount of expert demonstrations (not even in the form of trajectories but simple transitions) is required.
-Thus, this new approach which consists in cascading two well-known supervised learning (SL) methods is a model-free and veritable IRL approach in the sense that the output of our algorithm is a non-trivial reward function for which the expert policy is near optimal.
-Finally, we provide an analysis which legitimates this new approach.
+This reward function is seen as the most succinct description of the task, allowing for task transfer from the expert to an agent with potentially different abilities.
+The proposed contribution consists in cascading a classification and a regression steps to produce a non-trivial reward function for which we show the expert policy to be near optimal.
+In addition to being generic, this approach is model-free, it does not require solving any direct reinforcement learning problem (unlike most of IRL algorithms) and it solely relies on transitions from the expert (no need to sample trajectories according to other policies).
+All of this is illustrated through nontrivial experiments.
 \end{abstract}
@@
 \begin{equation}
+\label{r.def}
 \hat{r}_i=q(s_i,a_i)-\gamma q(s_i',\pi_C(s_i')), \forall i\in \{1,\dots,D\}.
@@
 \label{section: experiments}
+\subsection{Instantiation of the algorithm}
+As hinted to FIXME(Plus insister sur cet aspect) in Section \ref{section: Cascading}, our algorithm can be instanciated in many ways, as any off-the-shelf classifier using a score function can be used in the first step, and any regression method can be used in the second step.
+This freedom of choice leaves room for problem dependant tweaks if needed.\\
 
+In our tests, we used a subgradient-descent based classifier and a simple regularized least square regression method. The classifier tries to minimize the risk function $R_N(q)$ :
+\begin{equation}
+  R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
+\end{equation}
+where $l$ is a user defined function whose role is to quantify the gap between the best and second to best actions.
+In all our tests we used the naive solution $l(s_i,a) = 1$ if $a\neq a_i$ and $l(s_i,a_i) = 0$.
+To minimize the risk, we introduce the parametrization
+\begin{equation}
+  q(s,a) = \omega^T\phi(s,a).
+\end{equation}
+The risk now becomes :
+\begin{equation}
+  R_N(\omega) = {1\over N} \sum_{i=1}^N\left(\max_{a}(\omega^T\phi(s_i,a) + l(s_i,a)) - \omega^T\phi(s_i,a_i) \right).
+\end{equation}
+
+Because of the $\max$ in the expression, we can not use a simple gradiend descent on $\omega$ to minimize the risk.
+We have to make use of the subgradient.
+Let $a^* = \max_a\omega^T\phi(s,a)$, we get the following update rule :
+\begin{equation}
+  \omega_{t+1} = \omega_t -\alpha_t{1\over N}\sum_{i=1}^N\left(\phi(s_i,a^*) - \phi(s_i,a_i)\right).
+\end{equation}
+The regressor is a classical least square fitting. Given some features $\psi : S\rightarrow \mathbb{R}^p$, we define :
+\begin{eqnarray}
+\Psi &= \begin{pmatrix}\psi(s_1)^T\\\vdots\\\psi(s_N)^T\end{pmatrix}\\
+\textrm{and}
+\hat R &= \begin{pmatrix}\hat r_1 \\\vdots\\\hat r_N\end{pmatrix}\\
+\end{eqnarray}
+with $r_i$ as in equation \ref{r.def}. We then get :
+\begin{equation}
+\theta = (\Psi^T\Psi + \lambda Id)^{-1}\Psi^T\hat R,
+\end{equation}
+so that $\hat R_C(s) = \theta^T \psi(s)$.\\
+\subsection{Experimental setting}
+For each problem we test our approach with, we compare it to the following baselines :
+\begin{itemize}
+\item an agent trained on a random reward,
+\item an agent choosing its next action randomly,
+\item the policy output by the algorithm of \cite{abbeel2004apprenticeship}.
+\end{itemize}
+
+The algorithm of \cite{abbeel2004apprenticeship} can be instanciated to work in the same dire conditions as our algorithm, that is to say using only samples from the expert.
+To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ (\cite{klein2011batch}) and LSPI (\cite{lagoudakis2003least}).\\
+
+Sometimes this particular instantiation of \cite{abbeel2004apprenticeship}'s algorithm is not able to solve the problem.
+We then fall back to an instantiation using suroutines that breaks the batch, expert-only setting in order to have something to compare to.
+Precise workarounds will be described as they are used.
+It is important to note that sometimes our algorithm is at the best of our knowledge FIXME(c'est pas vrai il y a SCIRL aussi ;) the only one able to solve the problem with such a constrained input.
+\subsection{Highway driving simulator}
+\begin{itemize}
+\item description du problème
+\item deux sets de features pour faire marcher LSPI donc ANIRL
+\item résultats
+\item Insister sur le fait qu'on est meilleur quand il y a peu de samples
+\item Insisiter sur le fait qu'en features naturelles, LSPI ne fonctionne pas
+\end{itemize}
+\subsection{Gridworld}
+\begin{itemize}
+\item description du problème
+\item Introduction du task transfert (l'expert est une sorte de tour, l'agent un roi)
+\item Résultats pour l'IRL (comparaison avec ANIRL)
+\item résultats pour le task transfer, impossible de comparer avec ANIRL car officiellement pas de récompense
+\end{itemize}
+\subsection{Pendule}
+A n'ajouter que si les résultats présentent un intérêt particulier par rapport aux deux autres problèmes.
 \section{Related Work}

### 2012-04-23 piot2012cascading/main_KLEIN_v3.tex
Figures montrant les performances du cascading
 : make Cascading_Exp1_fig1.pdf Cascading_Exp1_fig2.pdf
La légende des figures dans un fichier à part

@@
 \usepackage{subfigure}
+\usepackage{tabularx}
 \usepackage{mathtools}
@@
 \subsection{Experimental setting}
-For each problem we test our approach with, we compare it to the following baselines :
-\begin{itemize}
-\item an agent trained on a random reward,
-\item an agent choosing its next action randomly,
-\item the policy output by the algorithm of \cite{abbeel2004apprenticeship}.
-\end{itemize}
+For each problem we test our approach with, we compare it to two baselines : the mean performance of an agent trained on a random reward and the mean performance of the policy output by the algorithm of \cite{abbeel2004apprenticeship}.
+This algorithm can be instanciated to work in the same dire conditions as our algorithm, that is to say using only samples from the expert.
+To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ (\cite{klein2011batch}) and LSPI (\cite{lagoudakis2003least}).
+\subsection{Inverted pendulum}
+In \cite{klein2011batch}, it is noted that the particular instanciation of Abbeel and Ng's algorithm we use as a baseline is unable to solve the Inverted Pendulum problem (described in \cite{lagoudakis2003least}).
 
-The algorithm of \cite{abbeel2004apprenticeship} can be instanciated to work in the same dire conditions as our algorithm, that is to say using only samples from the expert.
-To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ (\cite{klein2011batch}) and LSPI (\cite{lagoudakis2003least}).\\
-
-Sometimes this particular instantiation of \cite{abbeel2004apprenticeship}'s algorithm is not able to solve the problem.
+There exist settings this particular instantiation of \cite{abbeel2004apprenticeship}'s algorithm is not able to solve.
+In \cite{klein2011batch} the authors note that this instanciation the problem.
 We then fall back to an instantiation using suroutines that breaks the batch, expert-only setting in order to have something to compare to.
@@
 \subsection{Highway driving simulator}
+Inspired from a problem seen in \cite{syed2008apprenticeship,syed2008game}, we coded a driving simulator.
+The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
+The first thing to note about this setting is that the direct problem (RL) problem is not trivial, as LSPI is unable to solve it using the natural features we first gave it.
+These feature stemmed from a discretization of the state-action space, the dimension of the features being the cardinality of this space.
+With each state-action pair $s,a$ we associate an index $i$ so that the $i$-th coordinate of the feature is $1$ and every other is $0$.
+We tried to train an agent on the reward used by the expert (which was trained using an exact dynamic programming approach), but on top of being extremely slow to run (the LSTD$Q$ subroutine implies a matrix inversion) it does not converge.\\
+As LSPI is used in \cite{abbeel2004apprenticeship}'s algorithm to solve the MDP, it is impossible to use the latter with the natural features.
+We thus have to spend quite some time engeineering new informative features to make LSPI work.
+These new features have a dimensionality of $5$, including a constant term and coordinates whose semantics is \emph{The agent should turn right}, \emph{The agent should turn left}, \emph{The agent should increase its speed} and \emph{The agent should do nothing}.
+With these features in hand, we were able to solve the IRL problem with \cite{abbeel2004apprenticeship}'s algorithm.
+The results of our experiment is given on two figures, Figure \ref{Cascading_Exp1_1.fig} and \ref{Cascading_Exp1_2.fig}.
+The data presented here is the mean, minimum and maximum value obtained over 50 runs.
+The random baseline presented here is the mean, min adn max performance of an agent over 50 different random rewards.\\
+Figure \ref{Cascading_Exp1_1.fig} shows results when employing the Cascading algorithm with the natural set of features and Figure \ref{Cascading_Exp1_2.fig} shows results when using the engineered features, comparing it the the results of \cite{abbeel2004apprenticeship}'s algorithm which could be applied in this case.
+Although the mean value with natural features is lower than with the engineered ones, we see that there is less risk of catastrophic failure (the min value is higher) and, more importantly, we were actually able to solve the problem in a satisfying manner with.
+
+ \begin{figure}
+ \begin{tabular}{ccc}
+ \subfigure[Abbeel and Ng IRL  with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig3"}\label{Cascading_Exp1_3.fig}}&\hspace{.05\linewidth}&
+ \subfigure[Cascading with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig2"}\label{Cascading_Exp1_2.fig}}\\
+ \subfigure[Cascading with natural features]{\includegraphics[width=.4\linewidth]{"Cascading_Exp1_fig1"}\label{Cascading_Exp1_1.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Cascading_Exp1_legend"}
+ \end{tabular}
+\caption{Results on the Highway Driving problem. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
+ \end{figure}
+
 \begin{itemize}

### 2012-05-03 Highway/PlayOmega2.py
Cascading_Exp1 respecte le code couleurs

@@ -1,23 +1,0 @@
-
-import sys
-sys.path+=['..']
-import Highway
-from DP import *
-from DP_mu import *
-
-omega = genfromtxt( sys.argv[1] )
-Pi = omega2pi( omega, Highway.phi2, Highway.Sgenerator(), Highway.s_index, [Highway.P( a ) for a in Highway.A ] )
-print "Mu computation..."
-Mu = DP_mu( Pi, identity( 3*9*9*3 ))
-Mu_s_0 = Mu[ Highway.s_index( Highway.S_0() )]
-
-Mu_E = genfromtxt( "Mu_E.mat" )
-Mu_E_s_0 = Mu_E[ Highway.s_index( Highway.S_0() )]
-
-perf_agent = dot( Highway.R().transpose() , Mu_s_0 )
-perf_expert = dot( Highway.R().transpose() , Mu_E_s_0 )
-print "Performance moyenne de l'expert : "
-print perf_expert
-
-print "Performance moyenne de l'agent :"
-print perf_agent

### 2012-05-04 klein2012/main.tex
Suppression d'une footnote qui ne servait à rien
Corrections des typos avant soumission de la version finale

--- 2012-04-10 klein2012/main.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Exemple de fichier fonctionnnant avec la classe CAp2012.cls
+% Base :
+%	Classe pf2010.cls adpatée de pf2003 par Jean CHarlet
+%	Classe pf2003.cls adpatée de ic2001 par Jean CHarlet
+%
+% 	Classe ic2001.cls adpatée de EEGDRI3 par Jean CHarlet
+%
+% 	Classe EEGDRI3.cls adpatée de ic2000 par Jean CHarlet
+%
+% 	Classe IC'2000 (ic2000.cls) par Jean Charlet
+%
+% 	adaptée de la classe IC'99 (afia99.cls) développée par Fabien Torre
 
-% Use the following line _only_ if you're still using LaTeX 2.09.
-%\documentstyle[icml2012,epsf,natbib]{article}
-% If you rely on Latex2e packages, like most moden people use this:
-\documentclass{article}
+%\documentclass{CAp2012}
+\documentclass[publibook-draft]{CAp2012}
 
-% For figures
-\usepackage{graphicx} % more modern
-%\usepackage{epsfig} % less modern
+\newcommand{\verbfichex}{CAp2012.\xspace}
+\newcommand{\verbfichclass}{CAp2012.cls\xspace}
+\newcommand{\verbfichbst}{CAp2012.bst\xspace}
+
+%\usepackage[pdftex]{pstricks}
+
+\usepackage{verbatim}
 \usepackage{subfigure}
+\usepackage[french]{babel}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage{mathtools}
+\mathtoolsset{showonlyrefs=true}
 \usepackage{amssymb}
 \usepackage{natbib}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage{amsthm}
-\usepackage{hyperref}
-\usepackage{url}
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-\usepackage{icml2012}
-\mathtoolsset{showonlyrefs=true}
+\usepackage{caption}
+%%% francisation des algorithmes
 
+%\newtheorem{algorithm}{Algorithme}
 \newtheorem{definition}{Définition}
-\newtheorem{lemme}{Lemma}
+\newtheorem{lemme}{Lemme}
 \newtheorem{theorem}{Théorème}
@@
 \newtheorem{remarque}{Remarque}
-\newtheorem{assumption}{Assumption}
 \newtheorem{problem}{Problème}
@@
 \newcommand{\espace}{\text{ }}
-
 \newcommand{\x}{\mathbf{x}}
@@
 
-%\shorttitle{Classification structurée pour l'ARI}
-%
-%\shortouvrage{CAp 2012}
+\shorttitle{Classification structurée pour l'ARI}
+
+\shortouvrage{CAp 2012}
 
@@
 
-%\title{Classification structurée pour l'apprentissage par renforcement inverse}
-%
-%\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
-%
-%\institute{
-%Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
-%\and
-%Equipe ABC,
-%LORIA, France
-%\and
-%UMI 2958
-%GeorgiaTech-CNRS, France
-%}
-\icmltitlerunning{Structured Classification for Inverse
-Reinforcement Learning}
+\title{Classification structurée pour l'apprentissage par renforcement inverse}
+
+\author{\fontsize{12}{12}\selectfont{Edouard Klein\inst{1}$^,$\inst{2}, Bilal Piot \inst{1}$^,$\inst{3}, Matthieu Geist\inst{1}, Olivier Pietquin\inst{1}$^,$\inst{3}}}
+
+\institute{
+Sup\'elec, IMS Research group, France, \texttt{prenom.nom@supelec.fr}
+\and
+Equipe ABC,
+LORIA, France
+\and
+UMI 2958
+GeorgiaTech-CNRS, France
+}
 
 \begin{document}
-
-\twocolumn[ \icmltitle{Structured Classification for Inverse
-Reinforcement Learning}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2012
-% package.
-\icmlauthor{Your Name}{email@yourdomain.edu}
-\icmladdress{Your Fantastic Institute,
-            314159 Pi St., Palo Alto, CA 94306 USA}
-\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
-\icmladdress{Their Fantastic Institute,
-            27182 Exp St., Toronto, ON M6H 2T1 CANADA}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{inverse reinforcement learning, structured multi-class
-classification}
-
-\vskip 0.3in
-]
+\maketitle
 
 \begin{abstract}
-This paper addresses the Inverse Reinforcement Learning (IRL)
-problem which is a particular case of learning from demonstrations.
-The IRL framework assumes that an expert, demonstrating a task, is
-acting optimally with respect to an unknown reward function to be
-discovered. Unlike most of existing IRL algorithms, the proposed
-approach doesn't require any of the following: complete trajectories
-from the expert, a generative model of the environment, the
-knowledge of the transition probabilities, the ability to repeatedly
-solve the forward Reinforcement Learning (RL) problem, the expert's
-policy anywhere in the state space. Using a classification approach
-in which the structure of the underlying Markov Decision Process
-(MDP) is implicitly injected, we end-up with an efficient
-subgradient descent-based algorithm.
-% whose computational and sample
-%complexities are low.
-In addition, only a small amount of expert
-demonstrations (not even in the form of trajectories but simple
-transitions) is required.
-%A formal analysis of the algorithm as well as empirical
-%demonstrations on two classic but challenging problems are provided.
+  Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse (\emph{ARI}).
+Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
+L'ARI part du postulat que l'expert optimise avec succès une fonction de récompense ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
+Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
+Notre contribution consiste en un nouvel algorithme d'ARI levant l'ensemble de ces contraintes.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et surtout ne nécessitant pas la résolution du problème direct.
+  \motscles{Apprentissage par renforcement inverse, classification multi-classe, attribut vectoriel moyen}
 \end{abstract}
@@
 %%%%%
-%Dans cette contribution, nous nous plaçons dans le contexte de l'apprentissage par démonstration.
+Dans cette contribution, nous nous plaçons dans le contexte de l'apprentissage par démonstration.
 Dans ce cadre, un agent artificiel apprend à reproduire une tâche grâce à l'observation d'un expert réalisant cette tâche de manière optimale.
@@
 
-This contribution addresses the problem of learning by watching or
-learning from demonstrations. In this context, an artificial agent
-learns to perform a task by observing an expert demonstrating the
-task. In order to find a solution to this problem, the Inverse
-Reinforcement Learning (IRL) paradigm has been
-proposed~\cite{russell1998learning}. The fundamental idea of IRL is
-that the expert acts so as to be rewarded by an unknown reward
-function. The aim of the IRL agent is then to infer the unknown
-reward function from the demonstrations given by the expert, the
-reward being considered as the most compact representation of the
-task to be transferred to the agent. This way, an optimal behavior
-can be obtained in any situation by optimizing the action selection
-process according to the reward function. Unlike supervised
-learning, the agent may not purely imitate the expert's actions but
-will try to accomplish the same task optimally.
-%
-%
-%La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
+La littérature sur le sujet est assez récente et trouve sa genèse dans \cite{russell1998learning}.
 Plusieurs directions ont depuis été explorées à partir de deux articles fondateurs \cite{ng2000algorithms} et \cite{abbeel2004apprenticeship}.\\
-%Dans cette littérature (voir \cite{neu2009training} pour un bon survol), plusieurs verrous sont identifiés mais souvent ignorés pour proposer des algorithmes de résolution.
+Dans cette littérature (voir \cite{neu2009training} pour un bon survol), plusieurs verrous sont identifiés mais souvent ignorés pour proposer des algorithmes de résolution.
 Particulièrement, on suppose souvent connue la dynamique de l'environnement dans lequel évoluent l'expert et l'agent ou bien la mise à disposition de cet environnement (ou d'un simulateur) pour y tester les effets d'une politique quelconque.
@@
 Ces contraintes sont souvent incompatibles avec la mise en oeuvre des algorithmes proposés sur des cas réels.\\
-%
-%
-The literature on the subject is quite recent and finds its origin
-in \cite{russell1998learning}. Different approaches have been
-explored since the first articles on the matter
-\cite{ng2000algorithms,abbeel2004apprenticeship}.
 
-In this literature (see \cite{neu2009training} for a survey),
-several bottlenecks are identified but often ignored when developing
-algorithms. Specifically, the dynamic of the environment where the
-expert and the agent are evolving is assumed to be known, or an
-environment model (via a simulator) must be available so as to test
-the effects of a policy. Moreover, most of existing approaches
-require solving the direct reinforcement learning problem (find an
-optimal policy knowing a given reward) several times. Finally, in
-many cases, the output of the algorithm is a policy but not a unique
-reward function which makes task transfer hard to realize. These
-constraints can lead to the development of algorithms unapplicable
-to real problems.
-%Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert.
+Dans cette contribution, nous proposons une méthode de résolution du problème d'apprentissage par renforcement inverse s'affranchissant de toute connaissance additionnelle aux démonstrations fournies par l'expert.
 Nous proposons une preuve de convergence de cet algorithme et fournissons une expérimentation sur le problème du pendule inversé.
 Le choix de cette application se justifie par la difficulté  qu'elle pose pour les méthodes de résolution de la littérature.
-%
-In this paper, we propose a resolution method of the inverse
-reinforcement problem that is free from any other knowledge than the
-demonstrations given by the expert. The proposed approach, the
-Structured Classification for IRL (SCIRL), relies on a
-classification paradigm in which the structure of the environment's
-dynamics is implicitly introduced. Sec.~\ref{back.sec}
-and~\ref{biblio.sec} provide the necessary background and review a
-part of the state of the art. Sec.~\ref{sec:scirl} introduces our
-contribution and Sec.~\ref{proof.sec} analyses the asymptotic
-algorithm. Section~\ref{exp.sec} shows results obtained on two
-classical benchmarks: inverted pendulum and the highway driving
-problems.
-%
-%A proof of the convergence of the algorithm is provided in
-%Section~\ref{proof.sec} and Section~\ref{exp.sec} shows results
-%obtained on two classical benchmarks: inverted pendulum and the
-%highway driving problems. The choice of the first application is
-%justified by the fact that it is hard to solve by the other methods.
-
 %%%%
-\section{Background}
+\section{Contexte}
 %%%%%
@@
 %%%
-Let's start with some background on RL and IRL and settle the
-notations.
+\subsection{Apprentissage par renforcement direct et inverse}
+%%%
+Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
+La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$.
+Confronté à cet état, l'agent doit prendre une décision $a_t\in A$.
+Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité de transition markovienne $p(s_{t+1}|s_t, a_t)$.
+Une politique de contrôle déterministe $\pi : S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
 
-\subsection{(Inverse) Reinforcement Learning}
-%%%
-%Le cadre dans lequel nous plaçons notre étude est celui de la prise de décisions séquentielles.
-La configuration du système à contrôler est alors complètement décrite à chaque instant discret $t$ par un état $s_t \in S$.
-Confronté à cet état, l'agent doit prendre un décision $a_t\in A$.
-Le système évolue alors vers l'état suivant $s_{t+1}$ selon une certaine probabilité de transition markovienne $p(s_{t+1}|s_t, a_t)$.
-Une politique de contrôle déterministe $\pi: S\rightarrow A$ définit le comportement d'un agent confronté à un tel problème de décisions séquentielles.\\
+Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R : S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
+Une des forces de ce cadre réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
+On définit pour ce faire la fonction de valeur $V^\pi$ :
+\begin{equation}
+\label{Vdef.eqn}
+V^\pi(s) = E[\sum_t\gamma^tR(s_t)|s_0=s,\pi]=R(s) + \gamma E[V^\pi(s')|s,\pi]
+\end{equation}
+où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action :
+\begin{equation}
+\label{Qdef.eqn}
+Q^\pi(s,a) = E[\sum_t\gamma^tR(s_t)|s_0=s,a_0=a,\pi] = R(s) + \gamma E[V^\pi(s')|s,a].
+\end{equation}
+  Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
+La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton :
+\begin{equation}
+\label{greedy.eqn}
+\pi^*(s) \in \arg\max_a Q^*(s,a).
+\end{equation}
+L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov (\emph{PDM}).\\
 
-The problem is seen as a sequential decision making problem posed in
-the Markov Decision Process (MDP) formalism. In this formalism, a
-system to be controlled can reach different configurations
-completely described at each discrete time $t$ by a state $s_t \in
-S$. Given the knowledge of the state, the agent must take a decision
-$a_t\in A$. Then, the system evolves to a next state $s_{t+1} \in S$
-with the Markovian probability $p(s_{t+1}|s_t, a_t)$. %(in short
-%$p(s'|s,a)$ where $s'=s_{t+1}, s=s_t$ and $a=a_t$).
-A policy, which
-is a function $\pi: S\rightarrow A$, defines the behavior of the
-agent. In this paper all the policies considered are deterministic,
-$S$ and $A$ will be finite sets and $\{P_{sa}\}_{s\in S,a\in A}$
-will be a set of probability distributions, from $S$ to $[0,1]$ such
-that $P_{sa}(s')=p(s'|s, a)$, called transition probabilities.
-%
-%Dans le cadre de l'apprentissage par renforcement,la résolution de ce problème est guidée par une fonction de récompense $R: S \rightarrow \mathbb{R}$ qui indique le degré de désirabilité de chaque état.
-Une des forces de ce cadre réside dans le fait que l'agent ne va pas apprendre à maximiser la récompense immédiate mais au contraire un critère prenant le futur en compte, ce qui permet de spécifier uniquement le but et non la façon de l'atteindre.
-On définit pour ce faire la fonction de valeur $V^\pi$:
-%
-The RL paradigm aims at solving the problem of finding the optimal
-sequence of actions in an MDP. The optimality criterion is based on
-the definition of a reward function $R: S \rightarrow \mathbb{R}$
-which defines the immediate reward $R(s)$ given to the agent when
-the system is in the state $s \in S$. The agent should not learn to
-maximize the immediate reward but a criterion taking into account
-future rewards. The criterion commonly chosen is the value function
-$V^\pi$:
-%
-\begin{align}
-\label{Vdef.eqn} V^\pi(s) & = E^\pi_s[\sum_{t\geq0}\gamma^tR(s_t)]\\
-& = R(s) + \gamma\sum_{s'\in S}P_{s\pi(s)}(s') V^\pi(s'),
-\end{align}
-%où $\gamma \in [0,1)$ est un facteur d'actualisation. Il est également possible de définir une fonction de qualité qui ajoute un degré de liberté sur le choix de la première action:
-%
-where $\gamma \in [0,1)$ is a discount factor, and $E^\pi_s$ is the
-expectation over the distribution of the state sequences
-$(s_0,s_1,\dots)$ the agent passes through when executing the policy
-$\pi$ starting from $s_0=s$. One can define the action-value
-function $Q^\pi$ which adds a degree of freedom in the choice of the
-first action:
-%
-\begin{align}
-\label{Qdef.eqn} Q^\pi(s,a) & =
-E^\pi_{sa}[\sum_{t\geq0}\gamma^tR(s_t)]\\
-& = R(s) + \gamma\sum_{s'\in S}P_{sa}(s') V^\pi(s').
-\end{align}
-%
-%where $E^\pi_{sa}$ is the expectation over the distribution of the
-%state sequences $(s_0,s_1,\dots)$ we pass through when we start from
-%$s_0=s$, choose the action $a_0=a$ and then follow the policy $\pi$.
-%
-%Une politique optimale $\pi^*$ est définie comme une politique dont la fonction de valeur (optimale) $V^*$ vérifie $\forall \pi, \forall s, V^*(s) \geq V^\pi(s)$.
-La politique optimale se calcule simplement à partir de la fonction de qualité optimale $Q^*$ via un mécanisme glouton:
-%
-An optimal policy $\pi^*$ is defined as a policy with maximal value
-function, $V^{\pi*}=V^*=\sup_{\pi}V^{\pi}$, which means that
-$\forall \pi, \forall s, V^{\pi*}(s) \geq V^\pi(s)$. An optimal
-policy $\pi^*$ can be computed easily thanks to the optimal
-action-value function $Q^{\pi*}=Q^*=\sup_{\pi}Q^\pi$ via a greedy
-mechanism:
-%
-\begin{equation}
-\label{greedy.eqn} \pi^*(s) \in \arg\max_{a\in A} Q^*(s,a).
-\end{equation}
-%L'ensemble formé par l'espace d'état $S$, l'espace d'action $A$, les probabilités de transition $p$, le facteur $\gamma$ et la fonction de récompense $R$ forme un processus décisionnel de Markov .\\
-%
-The tuple $M=\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma,R\}$ is called
-a finite MDP. Resolving the RL problem with respect to the criterion
-$V^\pi$  means to find $\pi^*$, usually without the full knowledge
-of the MDP but only by observing transitions and rewards. In a
-finite MDP the existence of an optimal deterministic policy is
-guaranteed when the reward function $R$ is bounded.
-%
-%Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive.
+Parfois, définir la fonction de récompense est une tâche ardue alors qu'il est possible à l'opérateur de convenablement contrôler le système de manière intuitive.
 Un exemple de ce type de tâche est la conduite d'une voiture.
@@
 Dans de tels cas, il est utile d'inférer la récompense à partir d'un comportement démontré.\\
-%
-The optimal behaviour being defined relatively to the reward, its
-observation is mandatory to learn. Sometimes, specifying manually
-the reward function can be a hard job while it is natural to an
-operator to achieve the desired task. For instance, driving a car is
-such a task. One can accomplish this task very easily in his
-daily-life but it is very tricky to specify the precise weights
-given to each important driving-parameters such as the distance to
-keep with the leading car, the speed below which it is unreasonable
-to drive and so on. In such cases, inferring the reward function
-from observed behavior would be helpful.
-%
-%C'est la définition de l'apprentissage par renforcement inverse.
+
+C'est la définition de l'apprentissage par renforcement inverse.
 Le problème est de retrouver la fonction de récompense optimisée par un expert.
 Usuellement l'expert est considéré comme un agent optimal dans un PDM et l'on a accès ou bien à la politique complète $\pi_E$ de l'expert ou bien à quelques trajectoires tirées selon cette politique.\\
-%
-It is the aim of IRL. The expert, demonstrating the task, is thus
-considered as an optimal agent in a MDP and the IRL agent has access
-to the expert optimal policy $\pi_E$ or to some trajectories drawn
-from this policy.
-%
-%Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}).
+
+Ce problème est mal posé dans la mesure où il n'y a pas unicité de la récompense pour laquelle un comportement est optimal (\cite{ng1999policy}).
 Particulièrement, tout comportement est optimal vis-à-vis de la récompense uniformément nulle.
 Il convient donc de contraindre l'espace des solutions.
-The problem of finding a reward function from expert demonstrations
-in the IRL framework is ill-posed. Indeed, there is not uniqueness
-of the reward function for which the expert policy is
-optimal~\cite{ng1999policy}. Particularly, each policy is
-optimal with respect to the uniformly-zero reward function. Some
-solutions have to be favored.
 %%%
-\subsection{Feature expectation}
+\subsection{Attribut vectoriel moyen}
 \label{ConsiderationsTechniques.sec}
 %%%
-%La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$  telles que
-The actual expert reward function $R_E$ is unknown. The problem is
- thus to find a reward function $R_\theta$ for which the policy of
-the expert is also optimal. Let's assume that, given a set of $p$
-features $\psi: S\rightarrow \mathbb{R}^p$, each components of
-$\psi$: $\{\psi_i\}_{1\leq i \leq p}$ being a basis function from
-$S$ to $\mathbb{R}$, such a reward can be written:
+La récompense $R$ est l'inconnue du problème. Nous supposons que l'utilisateur est en mesure de fournir $p$ fonctions de base $\psi_{1\leq i \leq N}$ telles que
 \begin{equation}
-\label{hatRdef.eqn} %\exists \theta |
-R_\theta(s) = \theta^T\psi(s) =
-\sum_{i=1}^p\theta_i\psi_i(s).
+\label{hatRdef.eqn}
+\exists \theta | R(s) = \theta^T\psi(s) = \sum_{i=1}^p\theta_i\psi_i(s).
 \end{equation}
-%Cette contrainte peut cependant être relâchée si nécessaire.
+Cette contrainte peut cependant être relâchée si nécessaire.
 Introduire cette expression dans la définition de la fonction de qualité (Eq.
-\eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants:
-Using this expression allows rewriting (Eq. \eqref{Qdef.eqn}) as:
+\eqref{Qdef.eqn}) fait apparaître un terme intéressant et dont l'usage est assez répandu dans les algorithmes existants :
 \begin{equation}
-Q^\pi(s,a)=\theta^T\mu^\pi(s,a),\text{ }\mu^\pi(s,a) =
-E^\pi_{sa}[\sum_{t\geq 0}\gamma^t\psi(s_t)]. \label{Qmu.eqn}
+Q^\pi(s,a) = \theta^T\mu^\pi(s,a),\mu^\pi(s,a) = E[\sum_t\gamma^t\psi(s_t)|s_0=s,a_0=a,\pi]
+\label{Qmu.eqn}
 \end{equation}
-%Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique.
+
+Le terme $\mu^\pi$ est appelé l'attribut vectoriel moyen ({\it feature expectation}) d'une politique.
 On peut voir que la paramétrisation choisie pour $R$ et la dynamique imposée par l'application de la politique $\pi$ dans le PDM fixe la paramétrisation de $Q^\pi$.\\
-%
-The term $\mu^\pi$ is called the feature expectation of the policy
-$\pi$. We can observe that the chosen parametrization of the reward
-function $R_\theta$ and the dynamics resulting from the application
-of the policy $\pi$ in the MDP gives rise to an induced
-parametrization for $Q^\pi$. To
-shorten notations, we may use $\mu^\pi_{sa}$ as $\mu^\pi(s,a)$.\\
-%
-%On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
+
+On peut noter que si deux politiques ont le même attribut vectoriel moyen, alors ces deux politiques ont la même valeur vis-à-vis de la récompense, quel que soit le vecteur de poids $\theta$ définissant celle-ci.
 En effet
-We see also note that if two policies share the same feature
-expectation, then they have the same value function with respect to
-any reward function $R_\theta=\theta^T\psi$,
-$\theta\in\mathbb{R}^p$. Indeed:
-%
-\begin{equation}  \mu^{\pi_1} = \mu^{\pi_2} \Rightarrow
-\theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} =
-V^{\pi_2}. \label{memevaleur.eqn}
+\begin{equation}
+\mu^{\pi_1} = \mu^{\pi_2} \Rightarrow \theta^T\mu^{\pi_1} = \theta^T\mu^{\pi_2} \Rightarrow V^{\pi_1} = V^{\pi_2}.
+\label{memevaleur.eqn}
 \end{equation}
-%
-%L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
+L'attribut vectoriel moyen de l'expert, $\mu_E$, peut être calculé sans forcément avoir accès à l'intégralité de la politique $\pi_E$.
 La plupart des algorithmes de la littérature ont, comme nous allons le voir, pour objectif de minimiser la distance entre les attributs vectoriels moyens respectifs de l'agent et de l'expert.
 Cela permet d'obtenir un agent dont le comportement a la même fonction de valeur que celui de l'expert, quelle que soit la récompense (voir par exemple \cite{neu2009training}).
-%
-That is why most of the algorithms in the literature have, as we
-will see, the objective of minimizing a \textit{distance}  between
-the feature expectation of the expert and the feature expectation of
-the apprentice. %That allows to obtain an apprentice which policy has
-%the same value function as the expert policy, for a reward function
-%of the type $R=\theta^T\psi,\forall \theta\in\mathbb{R}^p$ (see for
-%instance
-%~\cite{neu2009training}.
-
-\section{State of the Art}
-\label{biblio.sec}
-
-Let $\pi_\theta$ denote an optimal policy according to a reward
-function $R_\theta(s)=\theta^T\psi(s)$, for some
-$\theta\in\mathbb{R}^p$. With a slight abuse of notation, let us
-write $\mu^\pi_{d_0}$ the mean feature expectation for a
-distribution $d_0$ over starting states:
-$\mu^\pi_{d_0}=E[\mu^\pi(s,\pi(s))|s\sim d_0]$. The classical
-approach to IRL (or less generally to apprenticeship learning, as
-the output of the algorithm may be a policy and not a reward
-function), initiated in~\cite{abbeel2004apprenticeship}, consists in
-finding a reward (a parameter vector) $\theta^*$ minimizing some
-loss function between the feature expectation of the expert and the
-one of the related optimal policy:
+%%%%%
+\section{Classification structurée pour l'ARI}
+%%%%%
+%%%
+\subsection{Principe}
+%%%
+Comme nous l'avons vu Eq.
+\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue : $\pi_E(s) \in \arg\max_aQ^E_R(s,a)$.\\
+Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
+Notons $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
+Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons :
 \begin{equation}
-  \theta^* = \argmin_\theta J(\theta) \text{ with } J(\theta)=
-  L(\mu^{\pi_\theta}_{d_0},\mu^E_{d_0}).\nonumber
+\label{butLAFEM.eqn}
+a_i \in \arg\max_aq(s_i,a).
 \end{equation}
-Depending on the adopted point of view,
-game-theoretic-base~\cite{syed2008game}, linear
-programming-based~\cite{syed2008apprenticeship}, maximum
-entropy-based~\cite{ziebart2008maximum}, \textit{etc.}, the loss
-function and the associated update rule differ. Most of these
-approches and others are nicely reviewed in~\cite{neu2009training}.
-
-In all cases, the solution $\theta^*$ is computed in an iterative
-way. At each step, one has to compute the feature expectation
-$\mu^{\pi_\theta}_{d_0}$ for some parameter vector $\theta$. This
-implies computing $\pi_\theta$, the optimal policy according to the
-reward $\theta^T\psi(s)$. Therefore, solving the IRL problem with
-one of these methods involves solving repeatedly the direct RL
-problem (some of these algorithms requiring also the model to be
-known). This is a drawback we wish to get rid of.
-
-A notable exception is~\cite{boularias2011relative}. This work,
-based on a relative entropy argument, finds $\theta^*$ by performing
-a subgradient ascent on some utility function. Estimating the
-subgradient would require to sample trajectories according to
-$\pi_\theta$, with $\theta$ being the current estimate, thus would
-require to solve the direct problem. However, this can be avoided in
-this specific framework by the use of importance sampling. The
-gradient is estimated from trajectories sampled from some random
-policy. Still, this may be a problem when the only available data
-comes from the expert.
-
-Another approach consists in casting imitation learning into a
-multi-class classification problem, introducing in some way the
-structure of the MDP. For example, in~\cite{melo2010learning}, the
-agent policy is learnt as the solution of a classification problem.
-They consider a kernel function built on some induced MDP metric,
-which may be combersome to compute. Closest to our proposed
-contribution is the Maximum Margin Planning (MMP)
-algorithm~\cite{ratliff2006maximum}, which is a classification-based
-algorithm associating optimal policies to MDPs. More precisely, the
-training set is made of $N$ couples $(\mathcal{M}_i,\pi^*_i)$, with
-$\mathcal{M}_i$ an MDP (without reward) and $\pi_i^*$ the associated
-optimal policy. Let $\mu^\pi_{d_0,\mathcal{M}_i}$ be the feature
-expectation for the policy $\pi$, the initial distribution $d_0$ and
-the dynamics induced by the related MDP. Let also
-$\mathcal{L}_{\mathcal{M}_i,\pi}$ be a positive and user-defined
-margin function penalizing non-optimal policies. MMP minimizes the
-following cost function:
-\begin{equation}
-  J(\theta) = \frac{1}{N}\sum_{1=1}^N \max_\pi(\theta^T
-  \mu^\pi_{d_0,\mathcal{M}_i} + \mathcal{L}_{\mathcal{M}_i,\pi}) - \theta^T
-  \mu^{\pi_i^*}_{d_0,\mathcal{M}_i}.\nonumber
-\end{equation}
-This implies solving a set of MDPs at each iteration. At the other
-extreme, a similar cost function is considered
-in~\cite{ratliff2007imitation}, which searches for a score function
-$q$ associating expert actions to states. The training set being $N$
-couples $(s_i,a_i=\pi^E(s_i))$ and $l(s,a)$ being a positive
-user-defined margin function penalizing non-optimal actions, the
-cost function is
-\begin{equation}
-  J(q) = \frac{1}{N}\sum_{i=1}^N \max_a (q(s_i,a) +
-  l(s_i,a))-q(s_i,a_i).\nonumber
-\end{equation}
-The drawback of this approach is that it does not take into account
-the structure of the underlying MDP, which may easily be shown
-experimentally to be a problem (\textit{e.g.},
-see~\cite{melo2010learning}).
-
-The proposed contribution is a trade-off between the two previous
-approaches. It works at the state level rather than the MDP level,
-considering a similar cost function. However, it takes into account
-the structure of the MDP. As shown in the next section, this allows
-avoiding to solve the direct RL problem as intermediate steps and it
-requires only data from the expert.
-
-%%%%%
-\section{Structured classification for IRL}
-\label{sec:scirl}
-%%%%%
-
-This section introduces our contribution, the SCIRL (Structured
-Classification for IRL), and discusses one of its key components,
-the estimation of the expert's feature expectation.
-
-%%%
-\subsection{Principle}
-
-First, let us review more deeply the approach
-of~\cite{ratliff2007imitation}, to which we will actually add some
-structure. Assume that we have a set of transitions
-$\{(s_i,a_i,s_{i+1})_{1\leq i\leq N}\}$ coming from one or several
-trajectories of the expert. A classic approach for generalizing the
-expert policy (from a multi-class classification-based point of
-view) would consist in finding some score function $q:S\rightarrow
-A$ such that for all $a\neq a_i$, $q(s_i,a_i)>q(s_i,a)$. In this
-framework, states are inputs and actions are labels. Moreover, one
-may want to strengthen the optimality of the expert. Let $l:S\times
-A\rightarrow\mathbb{R}_+$ be a user-defined margin function, which
-allows to put some prior knowledge in the problem at hand.
-Typically, this margin function should satisfy $l(s_i,a_i)\ll
-l(s_i,a)$. For example, one may choose $l(s_i,a\neq a_i)=1$ and
-$l(s_i,a_i)=0$. Then, one searches for a score function $q$
-satisfying
-\begin{equation}
-\label{eq:inequalities_score}
-  \forall a,\; q(s_i,a_i) + l(s_i,a_i) \geq q(s,a) + l(s,a).\nonumber
-\end{equation}
-In order to find such a score function, a natural empirical cost
-function to be minimized is:
-\begin{equation}
-\label{Jdef.eqn} J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in
-A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
-\end{equation}
-This has been originally introduced in~\cite{ratliff2007imitation}.
-Unfortunately, it ignores a large part of data: from the transitions
-$(s_i,a_i,s'_i)$ performed by the expert, only the state action
-couple $(s_i,a_i)$ is used and the dynamic part (transiting to
-$s'_i$ from this state action couple) is ignored.
-%
-Moreover, it outputs a policy, whereas we are interested in the
-reward function, which is a much more compact and transferable
-representation of the task at hand.
-
-Recall that we work in a sequential decision making framework.
-Therefore, there is a natural solution to optimization
-problem~\eqref{Jdef.eqn}: the optimal $Q$-function according to the
-unknown reward function. Now, let $\theta$ be some parameter vector.
-The score function $q(s,a) = \theta^T \mu^E(s,a)$ is actually the
-$Q$-function of the expert policy according to the reward
-$\theta^T\psi(s)$. Moreover, assume that this parameter vector is
-such that inequalities~\eqref{eq:inequalities_score} are satisfied.
-The optimal $Q$-function being stable by the greedy mechanism,
-see~Eq.\eqref{greedy.eqn}, this means that the expert policy $\pi_E$
-is optimal respectively to the reward function $\theta^T\psi(s)$
-(and $\theta^T\mu^E(s,a)$ is the associated optimal $Q$-function). A
-natural cost function for computing a reward function explaining the
-expert behavior is therefore the following one:
-\begin{equation}
-\label{structure.eqn} J_N(\theta)  = {1\over N}
-\sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) -
-\theta ^T \mu^E_{s_ia_i} \right).
-\end{equation}
-Still, we need to minimize it.
-
-%%%%
-%%Comme nous l'avons vu Eq.
-\eqref{greedy.eqn}, il existe un mécanisme glouton reliant la politique de l'expert, $\pi_E$ et la fonction de qualité optimale $Q^E_R$ relative à la fonction de récompense inconnue: $\pi_E(s) \in \arg\max_{a\in A}Q^E_R(s,a)$.\\
-%%Nous allons utiliser cette propriété pour considérer un risque dépendant d'une fonction de score $q$ qui devra partager avec la fonction inconnue $Q^E_R$ la propriété d'optimalité de l'expert que nous venons d'exprimer.\\
-%%Notons $(s_i,a_i,s_{i+1})_{1\geq i\geq N}$ les échantillons représentant une ou plusieurs trajectoires de l'expert.
-Nous cherchons donc une fonction de score qui vérifie pour chacun de ces échantillons:
-%As seen in Eq. \eqref{greedy.eqn}, there exists a greedy mechanism
-%linking the expert policy $\pi_E$ and the optimal action-value
-%$Q^E_R$ relative to the unknown reward function $R$: $\pi_E(s) \in
-%\arg\max_{a\in A}Q^E_R(s,a)$.
-%
-%We will use this property to consider an empirical risk function
-%$J_N$ dependent on a score function $q$ which will have to share
-%with the unknown action-value function $Q^E_R$ the same optimal
-%property. Let's note $(s_i,a_i,s_{i+1})_{1\leq i\leq N}$ the samples
-%coming from one or several  trajectories of the expert. We are
-%searching for a score function $q$ which verifies for each sample:
-%\begin{equation}
-%\label{butLAFEM.eqn}
-%a_i \in \argmax_{a\in A}q(s_i,a).
-%\end{equation}
-%%Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
-Nous introduisons de surcroit une fonction $l: S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
+Nous souhaitons donc introduire une fonction de risque qui devrait pénaliser les cas où $\exists a, q(s_i,a)>q(s_i,a_i)$, car alors la fonction $q$ ne permet pas de justifier (au sens de l'équation \eqref{greedy.eqn}) le choix $a_i$ de l'expert.
+Nous introduisons de surcroit une fonction $l : S \times A \rightarrow \mathbb{R}_+$ ; cette fonction permet à l'opérateur d'introduire de la connaissance \emph{a priori} dans le système en précisant l'écart qu'il souhaite obtenir entre la qualité de l'action choisie par l'expert et la qualité de la seconde meilleure action selon l'inégalité $\forall a, q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$.
 Ce type d'argument est assez commun dans la littérature (voir Sec.
 \ref{biblio.sec}).
-Nous introduisons ainsi la fonction de risque empirique suivante:
-%So, we want to introduce a risk function which should penalize the
-%cases where $\exists a, q(s_i,a)>q(s_i,a_i)$, because in those cases
-%the score function $q$ does not allow to explain the choice $a_i$ of
-%the expert (see equation \eqref{greedy.eqn}). Moreover, we introduce
-%a margin function $l: S \times A \rightarrow \mathbb{R}_+$; this
-%function allows the user to put some \emph{a priori} knowledge in
-%the system because the user will choose the gap wanted in the score
-%function between the action chosen by the expert and the second best
-%possible action as we can see in the following equation: $\forall a,
-%q(s_i,a) + l(s_i,a) \leq q(s_i,a_i)$. Introducing a margin function
-%is quite common in the literature (see Sec. \ref{biblio.sec}). We
-%introduce then the following empirical risk function:
-%\begin{equation}
-%\label{Jdef.eqn}
-%J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
-%\end{equation}
-%%Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
-%%
-%%Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
+Nous introduisons ainsi la fonction de risque empirique suivante :
+  \begin{equation}
+\label{Jdef.eqn}
+   J_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
+   \end{equation}
+Par défaut, définir $l(s_i,a) = 1$ si $a\neq a_i$ et $l(s_i,a_i)=0$ fonctionne bien comme le montrent les expériences (Sec. \ref{exp.sec}) et l'analyse (Sec. \ref{proof.sec}).\\
+
+Nous traitons le problème de l'imitation comme un problème de classification multi-classe.
 Notre propos n'est cependant pas d'utiliser une approche purement supervisée.
@@
 Il faut pour cela prendre en compte la structure du PDM, notamment l'information sur les transitions entre états conditionnées par la politique de l'expert.\\
-%%
-%%Comme nous l'avons vu Eq.
-\eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi: S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
+
+Comme nous l'avons vu Eq.
+\eqref{Qmu.eqn}, l'hypothèse faite d'une représentation linéaire selon certains attributs $\psi : S\rightarrow \mathbb{R}^p$ de la fonction de récompense impose d'exprimer la fonction de qualité d'une politique en fonction de l'attribut vectoriel moyen de cette politique.
 Cette paramétrisation prend en compte la dynamique imposée par $\pi_E$.
@@
 \ref{calculmu.sec} Nous pouvons donc maintenant exprimer notre risque, non plus comme une fonction de $q$, mais comme une fonction du vecteur de paramètres $\theta$, vecteur qui définit la récompense (Eq.
-\eqref{hatRdef.eqn}):
-%By default, choosing $l(s_i,a) = 1$ if $a\neq a_i$ and
-%$l(s_i,a_i)=0$ works quite well as shown in the experimentations
-%(Sec. \ref{exp.sec}) and in the analysis (Sec. \ref{proof.sec}).
-%
-%Here, we deal with the imitation problem as a classification
-%problem. However, our purpose is not to use a purely supervised
-%approach. We do not want to learn the policy of the expert but
-%rather the description of the task he is executing (the reward
-%function). In order to do that, it is necessary to take into account
-%the MDP structure, and in particular the information on the
-%transitions between states which are conditioned by the expert
-%policy.
-
-%As seen in Eq. \eqref{Qmu.eqn}, the hypothesis of a linear
-%representation of the reward function by some features $\psi:
-%S\rightarrow \mathbb{R}^p$ imposes the relation $Q^\pi(s,a) =
-%\theta^T\mu^\pi(s,a)$. This parametrization takes into account the
-%dynamic imposed by $\pi_E$. The expert follows the same rule and if
-%we note $\mu^E$ the feature expectation of the expert, we can do the
-%hypothesis that the score function $q$ that we are searching for can
-%be written as follows: $q(s,a) = \theta^T\mu^E(s,a)$. Estimating the
-%term $\mu^E$ is the key of our algorithm and we propose some
-%solution in Sec. \ref{calculmu.sec}. We can now give the expression
-%of our empirical risk function as a function of the parameter
-%$\theta$ which defines the reward function (Eq.
-%\eqref{hatRdef.eqn}):
-%\begin{equation}
-%\label{structure.eqn}
-%J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_{a\in A}(\theta ^T \mu^E_{s_ia} + l(s_i,a)) - \theta ^T \mu^E_{s_ia_i} \right)
-%\end{equation}
-%%Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
+\eqref{hatRdef.eqn}) :
+   \begin{equation}
+     \label{structure.eqn}
+   J_N(\theta)  = {1\over N} \sum_{i=1}^N\left(\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a)) - \theta ^T \mu_E(s_i,a_i) \right)
+   \end{equation}
+Ce risque est non linéaire et n'est pas différentiable partout, en raison du $\max$.
 Nous utilisons une descente de sous-gradient (normalisé), une généralisation du gradient, pour le minimiser.
@@
 Il s'agit d'un opérateur linéaire et la fonction convexe, différentiable par morceaux $\max_y[f(x,y)]$ définie à l'aide de la fonction différentiable $f(\cdot,y)$ admet comme l'un de ses sous-gradients l'expression $\nabla_xf(x,y^*)$ avec $y^*=\arg\max_yf(x,y)$.
-En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée:
-
-Because of the $\max$ operator, this cost function is not linear nor
-differentiable everywhere. But it is clearly convex and piecewise
-linear.
-%
-We can then use a subgradient descent algorithm. A subgradient of a
-convex function $f:\mathbb{R}^p\rightarrow \mathbb{R}$ in
-$x\in\mathbb{R}^p$ is defined as a vector $g_x$ such that $\forall
-x' \in \mathbb{R}^p, g_x^T(x'-x)\leq f(x')-f(x)$. The set of all
-subgradients in $x$ is noted $\partial_xf$. The unique subgradient
-in $x$ of a differentiable convex function $f$ is the gradient
-$\nabla_xf$. As $A$ is finite, the convex and piecewise
-differentiable function $\max_{a\in A}[f(\cdot,a)]$, from
-$\mathbb{R}^p$ to $\mathbb{R}$, defined thanks to the convex
-differentiable functions $\{f(\cdot,a)\}_{a\in A}$, from
-$\mathbb{R}^p$ to $\mathbb{R}$, has one of its subgradient in
-$x\in\mathbb{R}^p$ equal to $\nabla_xf(x,a^*)$ with
-$a^*\in\arg\max_{a\in A}f(x,a)$. Thanks to this, we can minimize the
-cost function of interest with the following update rule ($\alpha_i$
-denoting a usual learning rate):% and $\theta_0$ an initial parameter vector):
-\begin{align}
-  \theta_{t+1}&=\theta_t -\alpha_t\frac{g_{\theta}}{\|g_{\theta}\|_2}
-  \\
-  \text{with }
-  g_{\theta} &= {1\over N}\sum\limits_{i=1}^N\left(\mu^E(s_i,a^*_i) - \mu^E(s_i,a_i)\right)\in \partial_\theta J_N
-  \\
-  \text{and } a^*_i &\in \arg\max_{a\in A}(\theta_t^T \mu^E(s_i,a) +
-  l(s_i,a)).
-\end{align}
-The SCIRL algorithm only requires data from the expert, assuming
-$\mu^E$ is known on $\{(s_i,a)_{1\leq i \leq N,a\in A}\}$.
-Unfortunately, this is usually not the case.
-
-\subsection{Estimation of $\mu^E$}
+En appliquant ces règles, on obtient le sous gradient et la règle de mise à jour associée :
+\begin{eqnarray}
+\nabla_\theta J_N(\theta) &=& {1\over N}\sum\limits_{i=1}^N\left(\mu_E(s_i,a^*_i) - \mu_E(s_i,a_i)\right)\\
+a^*_i &=& \arg\max_a(\theta ^T \mu_E(s_i,a) + l(s_i,a))\\
+   \theta_{t+1} &\leftarrow& \theta_t -\alpha_t{\nabla_{\theta_t} J_N(\theta_t) \over ||\nabla_{\theta_t} J_N(\theta_t)||_2}.
+\end{eqnarray}
+%%%
+\subsection{Calcul de $\mu_E$}
 %%%
 \label{calculmu.sec}
+Comme signalé dans la Sec.
+\ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
+Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu_E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
+Nous avons besoin de connaître $\mu_E(s_i,a), \forall i= 1..N,\forall a \in A$.
+Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
 
-Estimating the feature expectation of the expert is a key component
-of our algorithm. We provide three possible solutions, depending on
-the available data.
+Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve :
+\begin{equation}
+\mu_E^j(s,a) = E[\sum\limits_{t=0}^\infty \gamma^t \psi^j(s)|s_0 = s, a_0 = a, \pi_E],
+\end{equation}
+expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
+Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu_E$.\\
 
-First, notice that each component $\mu_j^E$ of the expert feature
-expectation $\mu^E$ is actually the $Q$-function of the expert
-policy according to the reward function $\psi_j$, the $j^\text{th}$
-basis function of the considered hypothesis space:
+Dans le cas idéal où le modèle est connu, le calcul exact de $\mu_E$ peut être effectué grâce aux algorithmes de programmation dynamique.
+Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu_E(s,a)$ (définition Eq.
+\eqref{Qmu.eqn}) l'on utilise
 \begin{equation}
-\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t
-\psi_j(s)].\label{eq:muE_bellman}
-\end{equation}
-Therefore, it satisfies the Bellman evaluation
-equation~\eqref{Qdef.eqn}. Assuming that the complete expert policy and the
-transition probabilities are known, the expert feature expectation
-can be exactly computed.
-
-However, this assumption is rather strong (even if it is often done
-in the IRL literature). We may assume that we can simulate
-trajectories according to the expert policy on demand. In this case,
-one may estimate $\mu^E(s,a)$ for a given state-action couple with
-Monte Carlo. Assume that we have a set of $N$ trajectories of length
-$H$, $\{s^i_0,a^i_0,\dots,s^i_{H-1},a^i_{H-1},s^i_{H}\}_{1\leq i
-\leq N}$, with $s^i_0 =s$ and $a^i_0=a$, sampled according to the
-system dynamic and the expert policy. Then $\mu(s,a)$ is estimated
-as:
-\begin{equation}
-\hat\mu^E(s,a) = {1\over
-N}\sum_{i=1}^{N}\sum^{H}_{t=0}\gamma^t\psi(s^i_t),
+\hat\mu_E(s,a) = {1\over N}\sum\limits_{i=1}^{N}\{\sum\limits_{t}\gamma^t\psi(s), s_0=s, a_0=a, s_{t+1} \sim p(\cdot|s_t,\pi_E(s_t))\}.
 \end{equation}
 
-Yet, as claimed at the beginning of this paper, we would like to
-solely use the least amount of available data, that is is the set of
-transitions $\{(s_i,a_i,s'_{i})_{1\leq i\leq N}\}$ from the expert.
-As each component of the feature expectation can be seen as a
-$Q$-function, see Eq.~\eqref{eq:muE_bellman}, it can be estimated
-solely from the available data using the on-policy
-LSTD~algorithm~\cite{bradtke1996linear}. This is the basic idea of
-the LSTD$\mu$ approach~\cite{klein2011batch}, which is an LSTD-like
-method estimating feature expectations. As a part of the
-computations can be factored, LSTD and LSTD$\mu$ computational costs
-have the same order of magnitude. We use LSTD$\mu$ in the
-experimental section, which is the less constraining scheme.
+Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
+Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
+C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu_E$ en utilisant LSTD.
+Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
 
-Other schemes may be envisioned. For example, one may use a Monte
-Carlo estimate for $\mu^E(s_i,a_i)$ (using only the available data)
-and some heuristic for $\mu^E(s_i,a\neq a_i)$. We left this for
-future work.
-
-%%Comme signalé dans la Sec.
-\ref{ConsiderationsTechniques.sec}, beaucoup d'algorithmes de la littérature reposent sur un calcul de l'attribut vectoriel moyen.
-Notre algorithme a cependant ceci de particulier qu'il ne nécessite que le calcul de $\mu^E$, l'attribut vectoriel moyen de l'expert et non le calcul répété de l'attribut vectoriel moyen d'une politique arbitrairement éloignée de celle de l'expert.
-Nous avons besoin de connaître $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$.
-Cela implique d'inférer l'attribut vectoriel moyen de l'expert sur des états où celui-ci est passé, mais pour des actions qu'il n'a pas forcément choisies.\\ %Nul homme n'est jamais assez fort pour ce calcul :)
-%As mentioned in \ref{ConsiderationsTechniques.sec}, some algorithms
-%in the literature depend on the implementation of the feature
-%expectation. Our algorithm, in particular, relies only on the
-%implementation of $\mu^E$, the feature expectation of the expert,
-%and does not require the repeated calculus of the feature
-%expectation of another policy. For our algorithm, we only need to
-%know $\mu^E(s_i,a), \forall i= 1..N,\forall a \in A$. That implies
-%to infer the feature expectation of the expert on states where he
-%has been, but on actions he has not taken.
-%%Cette inférence est facilitée par le fait que l'attribut vectoriel moyen obéit à l'équation de Bellman. En effet, lorsque l'on écrit sa définition pour une composante $1\leq j\leq p$, on trouve:
-%This inference is made easier by the fact that the feature
-%expectation obeys the Bellman equation. Indeed, when we write its
-%definition for a component $1\leq j\leq p$, we have:
-%\begin{equation}
-%\mu^E_j(s,a) = E^{\pi_E}_{sa}[\sum\limits_{t=0}^\infty \gamma^t \psi_j(s)].
-%\end{equation}
-%%expression qui lorsqu'on la rapproche de l'équation \eqref{Qdef.eqn} permet d'affirmer que $\mu^j_E$ est en fait la fonction de qualité de la politique $\pi_E$ vis-à-vis de la récompense définie par la $j$-ème composante du vecteur d'attributs $\psi$.
-Selon les informations disponibles, plusieurs méthodes peuvent donc être utilisées pour calculer $\mu^E$.\\
-%When we compare this expression to the equation \eqref{Qdef.eqn}, it
-%allows us to affirm that $\mu_j^E$ can be seen as the action-value
-%function for the policy $\pi_E$ with respect to the reward function
-%defined by the $j$-th component of the features $\psi$. Depending on
-%the prior information, several methods can be used in order to
-%estimate $\mu^E$.\\
-%
-%%Dans le cas idéal où le modèle est connu, le calcul exact de $\mu^E$ peut être effectué grâce aux algorithmes de programmation dynamique.
-Dans le cas plus réaliste où les probabilités de transitions sont inconnues, on peut, si l'on dispose d'un simulateur et qu'il est possible d'interroger l'expert en un grand nombre d'états arbitraires, utiliser une simple approximation de Monte-Carlo où pour approximer $\mu^E(s,a)$ (définition Eq.
-\eqref{Qmu.eqn}) l'on utilise
-%In the ideal case where a model of the environnement is known
-%(\textit{id est} the transition probabilities are given), the exact
-%calculus of $\mu^E$ can be executed thanks to dynamic programmation
-%algorithms. In a more realistic case, we can use a simple
-%Monte-Carlo estimation if we dispose of a simulator and if it is
-%possible to ask the expert what action has to be chosen in an
-%important number of arbitrary states. The estimation of $\mu^E(s,a)$
-%(definition Eq. \eqref{Qmu.eqn}) is done over $N$ trajectories of
-%length $\{2k_i+1\}_{1\leq i \leq N}$,
-%$\{s^i_0,a^i_0,\dots,s^i_{k_i-1},a^i_{k_i-1},s^i_{k_i}\}_{1\leq i
-%\leq N}$, by the following formula:
-%\begin{equation}
-%\hat\mu^E(s,a) = {1\over N}\sum_{i=1}^{N}\sum^{k_i}_{t=0}\gamma^t\psi(s^i_t),
-%\end{equation}
-%where $s^i_0=s, a^i_0=a$ and $s^i_{t+1}$ is distributed according to $P_{s^i_t\pi_E(s^i_t)}$.
-%Ensuite, afin d'entrer dans le cadre où les seules données qui nous sont accessibles sont des trajectoires fournies par l'expert, il est possible d'utiliser l'algorithme LSTD$\mu$ \citep{klein2011batch} dans sa version \emph{on-policy}.
-Cet algorithme adapte l'algorithme LSTD (\cite{bradtke1996linear}) dont il conserve les caractéritiques, notamment la capacité à effectuer le calcul en mode {\it batch}.
-C'est la méthode que nous illustrons dans la section \ref{exp.sec} Le principe est d'estimer chacun des composantes de $\mu^E$ en utilisant LSTD.
-Certains calculs sont factorisables, donnant ainsi à LSTD$\mu$ un coût du même ordre de grandeur que celui de LSTD.\\
-%Finally, if we are in the framework where only the data at our
-%disposition are the trajectories samples given by the expert, it is
-%possible to use the LSTD$\mu$ algorithm \citep{klein2011batch} in
-%its \emph{on-policy} version. This algorithm adapts the LSTD
-%algorithm (\cite{bradtke1996linear}) keeping its characteristics, in
-%particular its ability to execute the calculus in  {\it batch}-mode.
-%This is the method that we show in the section \ref{exp.sec}. The
-%principle is to estimate each component of $\mu^E$ using LSTD. Some
-%calculus can be aggregated, giving to LSTD$\mu$ a cost of calculus
-%which is of the same order of magnitude as LSTD.
-%%D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
+D'autres méthodes d'approximation sont envisageables, comme par exemple une approximation de Monte-Carlo moins coûteuse et moins contraignante que celle que nous venons de proposer, se basant uniquement sur les données fournies par l'expert et impliquant l'utilisation d'une heuristique pour les actions inconnues.
 Cela est reporté pour des travaux futurs.
-%%%%%%
-%Other estimation methods are possible, as for instance a Monte-Carlo
-%estimation, less calculus-expensive and constraining than the first
-%one, which would be based solely on the data given by the expert and
-%would imply the use of an heuristic on the unknown actions. This
-%estimation would be consider on future works.
-%
-
-\section{Asymptotic analysis}\label{proof.sec}
-In this section, we prove, under two assumptions, that there is
-convergence of the subgradient algorithm applied to the asymptotic
-risk function $J$ (that is $J = \lim_{N\rightarrow\infty} J_N$)
-defined as:
+%%%%%
+\section{Analyse}\label{proof.sec}
+\subsection{Hypothèses}
+Ici, nous présentons le cadre sous lequel notre algorithme de sous gradient converge vers un minimiseur de $J_N$ qui rende la politique de l'expert optimale.
+On supposera que la récompense $R$ est de la forme~$R(s)~=~\theta_E^T\psi(s)$ où $\theta_E\in\mathbb{R}^p$, que notre base d'exemples est telle que $S\subset \{s_i\}_{1\leq i \leq N}$ et que $\mu_E$ est estimée sans erreur sur $\{(s_i,a_i)\}_{1\leq i \leq N}$.
+Nous supposons également l'unicité de l'action optimale (cette hypothèse pouvant être relâchée au prix de notations plus complexes):
+\begin{equation}\label{eq:existence}
+\forall 1\leq i \leq N, \theta_E^T\mu_E(s_i,a_i)>\max_{a\in A \backslash a_i}\theta_E^T\mu_E(s_i,a),
+\end{equation}
+où la notation ~$A \backslash a_i$ désigne l'ensemble $A$ privé de l'action $a_i$.
+Ainsi tout vecteur $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale.
+Si la base d'exemples est tel que $\{s_i\}_{1\leq i \leq N}\subset S$, alors un $\theta$ vérifiant l'équation~\eqref{eq:existence} rendra la politique de l'expert optimale au moins sur $\{s_i\}_{1\leq i \leq N}$ car l'équation~\eqref{eq:existence} impliquera dans ce cas : $\forall s\in \{s_i\}_{1\leq i \leq N}, V^{\pi_E}(s)\geq\max_{a\in A}Q^{\pi_E}(s,a)$.
+Dans ce cadre, nous allons montrer qu'il existe bien un $\theta$ vérifiant l'équation~$\eqref{eq:existence}$ qui minimise $J_N$,  que les minimiseurs de $J_N$ vérifient bien l'équation~\eqref{eq:existence} et que l'algorithme de sous-gradient converge vers un minimiseur.
+\subsection{Résultat}
+On va montrer, dans un premier temps, qu'il existe bien un $\hat{\theta}$ qui vérifie l'équation \eqref{eq:existence} et qui minimise $J_N$.
+On note tout d'abord ~$M^*=\{\theta\in\mathbb{R}^p:J_N(\theta)=\inf_{\theta \in \mathbb{R}^p}J_N(\theta)\}$.
+Désormais, on imposera que notre fonction de perte vérifie : ~$\forall~1\leq~i~\leq~N,~\forall~a~\in~A~\backslash~a_i,~l(s_i,a)~>~l(s_i,a_i)~\geq~0$.
+Il est facile de borner inférieurement le risque $J_N$, pour tout ~$\theta\in\mathbb{R}^p$ on a : ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Soit ~$\hat{\theta}\in\mathbb{R}^p$ vérifiant l'équation \eqref{eq:existence}, l'existence de ce dernier est assuré par l'hypothèse sur $\theta_E$.
+Notons ~$\hat{\theta}_x=x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}\in\mathbb{R}^p$ avec $x\geq0$,~$\hat{M}=\{\hat{\theta}\in\mathbb{R}^p:\hat{\theta} \text{ vérifie l'équation \eqref{eq:existence} }\}$, et posons ~$\forall~1\leq~i~\leq~N,~\beta_i~=~\hat{\theta}_1^T~\mu_E(s_i,a_i)~-~\max_{a\in A \backslash a_i}~\hat{\theta}_1^T~\mu_E(s_i,a)~>~0~$.
+Montrons qu'il existe un $x\geq0$ tel que $~J_N(\hat{\theta}_x)~=~{1\over N}~\sum_{i=1}^N l(s_i,a_i)$, et cela nous assurera que ~$\hat{\theta}_x~\in~M^*$ d'après la minoration ~$J_N(\theta)\geq {1\over N} \sum_{i=1}^N l(s_i,a_i),~\forall~\theta~\in~\mathbb{R}^p$.
+Avec les notations précédentes, on vérifie $\forall 1\leq i \leq N$ :
+\begin{align}
+&\hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a\in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{a\in A \backslash a_i}l(s_i,a)\geq0,\\
+\Leftrightarrow&x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i}.
+\end{align}
+Or comme la somme des $\max$ est supérieure au $\max$ d'une somme, il est clair que $\forall 1\leq i \leq N$:
+\begin{align}
+&x\geq\frac{\max_{a\in A \backslash a_i}l(s_i,a)-l(s_i,a_i)}{\beta_i},\\
+\Leftrightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}\hat{\theta}_x^T\mu_E(s_i,a) + l(s_i,a_i)-\max_{a\in A \backslash a_i}l(s_i,a)\geq0,\\
+\Rightarrow&\hat{\theta}_x^T\mu_E(s_i,a_i) + l(s_i,a_i)-\max_{a \in A  \backslash a_i}(\hat{\theta}_x^T\mu_E(s_i,a)+l(s_i,a))\geq0,\\
+\Leftrightarrow&J_N(\hat{\theta}_x)={1\over N} \sum_{i=1}^N l(s_i,a_i).
+\end{align}
+C'est bien le résultat recherché. Montrons à présent que les minimiseurs de $J_N$ vérifient bien l'équation \eqref{eq:existence}.
+Soit ~$\theta^*\in\ M^*$.
+Pour ~$\hat{\theta}\in \hat{M}$, on a vu qu'il existait un ~$x>0$ tel que $J_N(\hat{\theta}_x)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$, ainsi d'après la minoration $~\forall \theta\in\mathbb{R}^p, J_N(\theta)~\geq~{1\over N}~\sum_{i=1}^Nl(s_i,a_i)$, $J_N(\theta^*)~=~{1\over N} \sum_{i=1}^N l(s_i,a_i)$.
+Or on a $\forall 1\leq i \leq N$ :
+\begin{align}
+&J_N(\theta^*)={1\over N} \sum_{i=1}^N l(s_i,a_i),\\
+\Leftrightarrow&\forall a \in A \backslash a_i,  (\theta^*)^T\mu_E(s_i,a_i)-(\theta^*)^T\mu_E(s_i,a)\geq l(s_i,a)- l(s_i,a_i)>0,\\
+\Rightarrow&(\theta^*)^T\mu_E(s_i,a_i)-\max_{a \in A \backslash a_i}(\theta^*)^T\mu_E(s_i,a)>0.
+\end{align}
+Donc~$\theta^*\in\hat{M}$. C'est bien ce que l'on recherchait. Si on récapitule, on a montré dans un premier temps que $M^*\neq\emptyset$ et dans un second temps que $M^*\subset\hat{M}$.
+Finalement, montrons que l'algorithme de sous-gradient converge.
+Tout d'abord, il est clair que $J_N$ est une fonction convexe, fermée, propre, à valeurs finies et continue en la variable $\theta\in\mathbb{R}^p$.
+De plus on sait que $M^*\neq\emptyset$, on peut donc appliquer le théorème présenté en section 5 de l'article de \cite{correa1993convergence} qui nous assure que la suite :
 \begin{equation}
-J(\theta)=\sum_{s\in S}\nu_E(s)(\max_{a \in A}(\theta^T\mu^E_{sa}+l(s,a))-\theta^T\mu^E_{s\pi_E(s)}),
+\theta_0\in\mathbb{R}^p, \forall t\in\mathbb{N}, \theta_{t+1}=\theta_t -\alpha_t{\nabla_{\theta_t} J_N(\theta_t) \over ||\nabla_{\theta_t} J_N(\theta_t)||_2},
 \end{equation}
-where $\nu_E$ is the asymptotic probability distribution on states
-induced by $\pi_E$. Moreover, we prove that the result of the
-subgradient algorithm, which is
-$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}$, is a minimum of
-$J$ and that the expert policy $\pi_E$ is the unique optimal policy
-for the reward function $\tilde{\theta}^T\psi$.
-
-\subsection{Result}
-\begin{assumption}
-\label{assumption1}
-It is assumed that we are able to estimate perfectly the term $\{\mu^E_{sa}\}_{s\in S,a\in A}$.
-\end{assumption}
-
-\begin{assumption}
-\label{assumption2}
-There exists $\theta\in\mathbb{R}^p$ such that $\forall s\in S$:
-\begin{equation}
-\label{existence.eq}
-\theta^T\mu^E_{s\pi_E(s)}>\max_{a \in A\backslash\pi_E(s)}\theta^T\mu^E_{sa}.
-\end{equation}
-\end{assumption}
-
-\begin{lemme}
-\label{proposition1} Under assumptions \ref{assumption1} and
-\ref{assumption2}, there exists a minimum $\theta^*$ of J. The
-elements of the set $M^*=\{\theta^*\in\mathbb{R}^p:
-J(\theta^*)=\inf_{\theta\in\mathbb{R}^p}J(\theta)\}$ satisfy
-equation \eqref{existence.eq}.
-\end{lemme}
-\begin{proof}
-Let the margin function $l$ be such that: $\forall s\in S, \forall
-a\in A\backslash\pi_E(s), l(s,a)=c>l(s,\pi_E(s))=0$. Then $\forall
-\theta\in\mathbb{R}^p,J(\theta)\geq\sum_{s\in
-S}\nu_E(s)l(s,\pi_E(s))=0$. Let note $\hat{\theta}$ the vector
-verifying \eqref{existence.eq}, it is clear that
-$\|\hat{\theta}\|_2\neq0$. Let $x>0$ and
-$\hat{\theta}_x$=$x\frac{\hat{\theta}}{\|\hat{\theta}\|_2}$, then
-there exists $x$ such that $J(\hat{\theta}_x)=0$. Indeed, $\forall
-s\in S$ :
-\begin{align}
-&J(\hat{\theta}_x)=0\\
-\Leftrightarrow&\hat{\theta}_x^T\mu^E_{s\pi_E(s)}\geq\max_{a \in A\backslash\pi_E(s)}(\hat{\theta}_x^T\mu^E_{sa}+l(s,a))\\
-\Leftrightarrow&x\geq\frac{c}{\hat{\theta}_1^T\mu^E_{s\pi_E(s)}-\max_{a \in A\backslash\pi_E(s)}\hat{\theta}_1^T\mu^E_{sa}}.
-\end{align}
-Then all the vectors $\hat{\theta}_x$ with $x\geq c_{max}=\max_{s\in
-S}\frac{c}{\hat{\theta}_1^T\mu^E_{s\pi_E(s)}-\max_{a \in
-A\backslash\pi_E(s)}\hat{\theta}_1^T\mu^E_{sa}}$ are such that:
-$\inf_{\theta\in\mathbb{R}^p}J(\theta)\geq0=J(\hat{\theta}_x)\geq\inf_{\theta\in\mathbb{R}^p}J(\theta)$.
-So $\hat{\theta}_x=\inf_{\theta\in\mathbb{R}^p}J(\theta)$ for $x\geq
-c_{max}$. Finally let us show that the elements of the set
-$M^*=\{\theta^*\in\mathbb{R}^p:
-J(\theta^*)=\inf_{\theta\in\mathbb{R}^p}J(\theta)\}$ satisfy
-equation \eqref{existence.eq}. As there exists $x$ such that
-$J(\hat{\theta}_x)=0=\inf_{\theta\in\mathbb{R}^p}J(\theta)$, it
-implies $J(\theta^*)=0$. So:
-\begin{align}
-&J(\theta^*)=0\\
-\Leftrightarrow&\theta^{*T}\mu^E_{s\pi_E(s)}\geq\max_{a \in A\backslash\pi_E(s)}(\theta^{*T}\mu^E_{sa}+l(s,a))\\
-\Rightarrow&\theta^{*T}\mu^E_{s\pi_E(s)}>\max_{a \in A\backslash\pi_E(s)}\theta^{*T}\mu^E_{sa}.
-\end{align}
-\end{proof}
-
-\begin{prop}
-Under assumptions \ref{assumption1} and \ref{assumption2}, the
-subgradient algorithm defined by the sequence $\theta_0=0$,
-$\theta_{t+1}=\theta_t-\alpha_t\frac{g_{\theta_t}}{\|g_{\theta_t}\|_2}$
-where $g_{\theta_t}=\sum_{s\in
-S}\nu_E(s)(\mu^E_{sa^*}-\mu^E_{s\pi_E(s)})$ with
-$a^*\in\argmax_{a\in A}\theta_t^T\mu^E_{sa}+l(s,a)$,
-$\sum_{t\geq0}\alpha_t>\infty$ and $\sum_{t\geq0}\alpha_t^2<\infty$,
-is convergent:
-$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$.
-Moreover we have that
-$J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$. Finally
-the expert policy $\pi_E$ is the unique optimal policy for the
-reward $\tilde{\theta}^T\psi$.
-\end{prop}
-\begin{proof}
-The cost function  $J$ is convex, proper, closed and with finite
-values (the action space being finite). Therefore, using the result in section 5 
-of~\cite{correa1993convergence} allows to conclude that
-$\lim_{t\rightarrow+\infty}\theta_t=\tilde{\theta}\in\mathbb{R}^p$
-and $J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)$.
-%
-Moreover, Lemma~\ref{proposition1} shows that
-$\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$.
-%
-So we have
-$J(\tilde{\theta})=\inf_{\theta\in\mathbb{R}^p}J(\theta)=0$ which
-means that $\tilde{\theta}\in M^*$ and satisfies equation
-\eqref{existence.eq}. This implies that, $\forall s\in S$, $\forall
-a\in A$:
-\begin{equation}
-\tilde{\theta}^T\mu^E_{s\pi_E(s)}>\tilde{\theta}^T\mu^E_{sa},
-\end{equation}
-which proves that the expert policy $\pi_E$ is the unique optimal
-policy for the reward $\tilde{\theta}^T\psi$.
-\end{proof}
-
-\subsection{Discussion}
-The assumptions \ref{assumption1} and \ref{assumption2} are quite
-important and we can of course wonder what results could be
-guaranteed if they were not true. Moreover, we have only sampled
-transitions, so we cannot use the asymptotic risk function $J$, only
-the empirical function $J_N$. Thus, in our future works, the aim
-will be to find, if possible, an upper bound for the term:
-$E_{s\sim\nu_E}[V^{\pi_E}(s)-V^{\pi_A}(s)]$ where $E_{s\sim\nu_E}$
-is the expectation over the stationary probability distribution of
-states $\nu_E$ and $\pi_A$ is the policy of the apprentice which is
-optimal for the reward $\tilde{\theta}^T\psi$ found by the
-subgradient algorithm. Of course, this upper-bound will depend on
-the feature expectation errors
-$\{\epsilon^{\mu}_{s_ia}=\hat{\mu}^E_{s_ia}-\mu^E_{s_ia}\}_{1\leq i
-\leq N,a\in A}$ (which may be quantified thanks to the finite sample
-analysis of LSTD~\cite{lazaric2010finiteLSTD}) and on the approximation error
-$\epsilon^R$ which results from the fact that the unknown and
-bounded reward function $R_E\in\mathbb{R}^S$ may not be a linear
-combination of the features $\psi_{1\leq i \leq p}$. Formally, let
-$T$ be the tuple $\{S,A,\{P_{sa}\}_{s\in S,a\in A},\gamma\}$ which
-is an MDP without reward and $T(\tilde{R})=\{S,A,\{P_{sa}\}_{s\in
-S,a\in A},\gamma,\tilde{R}\}$ where $\tilde{R}\in\mathbb{R}^S$. We
-note $C_R$ the reward-equivalence class of $R$:
-$C_R=\{\tilde{R}\in\mathbb{R}^S: \Pi(T(R))=\Pi(T(\tilde{R}))\}$
-where $\Pi(T(\tilde{R}))$ is the set of optimal deterministic
-policies of $T(\tilde{R})$ (see~\cite{ng1999policy} for a
-characterization of $C_R$). A formal definition of $\epsilon^R$ is:
-$\epsilon^R=\|R^*-\theta^{*T}\psi\|_\infty=\max_{s\in
-S}|R^*(s)-\theta^{*T}\psi(s)|$ with
-$(R^*,\theta^*)\in\argmin_{\tilde{R}\in C_R,
-\theta\in\mathbb{R}^p}{\|\theta^T\psi-R\|_2}$. Linking the empirical
-risk function $J_N$ to the true risk function $J$ could be done
-using a Vapnik-like analysis. The difficult thing of course is to
-have access to $V^{\pi_A}$ without resolving the MDP directly.
-
-%\section{Related work}
-%\label{biblio.sec}
-%%%%%%
-%%L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte: la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
+est convergente vers un minimiseur $\theta^*$ de $J_N$, où $\alpha_t$ est un taux d'apprentissage satisfaisant les conditions usuelles ($\sum_{t>0} \alpha_t$ diverge et $\sum_{t>0}\alpha_t^2$ converge).
+Ce minimiseur vérifiera l'équation~\eqref{eq:existence} car $M^*\subset\hat{M}$ et la politique de l'expert sera optimale pour la récompense relative à $\theta^*$ car l'équation~\eqref{eq:existence} implique l'équation d'optimalité de Bellman.
+\section{Travaux Connexes}
+\label{biblio.sec}
+%%%%%
+L'approche historique de \cite{ng2000algorithms}, ainsi que par exemple \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum}, approches résumées dans \cite{neu2009training} souffrent toutes de la même contrainte : la nécessité de résoudre le problème direct de manière répétée, c'est à dire le calcul de la politique optimale pour n'importe quelle récompense.
 Ces approches ont pour point commun de chercher à minimiser une métrique de la proximité entre l'attribut vectoriel moyen de \mbox{l'expert} et celui d'une politique donnée.
 Par mises à jour successives du vecteur de paramètres $\theta$, la politique définie selon un critère d'optimalité dépendant de $\theta$ (c'est là qu'intervient la résolution du problème direct car $\theta$ spécifie la récompense) va voir son attribut vectoriel moyen se rapprocher de celui de l'expert.\\
-%The historical approach of \cite{ng2000algorithms}, as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} as summed up in \cite{neu2009training} all suffer from the same constraint: the need to repeatedly solve the forward problem, that is to say the need to repeatedly find the optimal policy with respect to an arbitrary reward.
-All these approaches try to minimize some distance between the feature expectation of the expert and the feature expectation of another policy.
-Successive updates of the vector of parameters $\theta$ will get this other policy (optimal with respect to a criterion that depends on $\theta$, this is where the \emph{RL} problem needs to be solved) closer and closer to the expert's, feature expectation-wise.\\
-%
-%%Cette résolution est très problématique.
+
+Cette résolution est très problématique.
 En effet trouver une politique optimisant une récompense est un sujet occupant un pan entier de la littérature.
 Particulièrement, l'information nécessaire à cet objectif est bien plus étendue que celle nécessaire au calcul d'uniquement l'attribut moyen de l'expert comme cela est nécessaire à notre algorithme.
-Comme cela est illustré dans \cite{klein2011batch}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état, ne permettant pas la résultion du problème direct en se basant uniquement sur les  données de l'expert.\\
-%%FIXME:parler de boularias ici
-%Solving the forward problem can be very tricky ; indeed a whole body of literature is dedicated to this problem.
-Specifically, one needs more information to solve the forward problem than to just compute the feature expectation of the expert as our algorithm does.
-As illustrated in \cite{klein2011batch}, \citep{abbeel2004apprenticeship}'s algorithm is unable to solve the inverted pendulum problem with just samlpes from the expert as those cover too small a part of the state space.
-Supplementary randomly drawn samples are needed.
-Even a recent contribution like \cite{boularias2011relative} still needs samples that cover a large portion of the state space although it does not explicitely require the ability to solve the forward problem.\\
-%
-%%Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
+Comme cela est illustré dans \cite{klein2011batch}, l'algorithme d'\citet{abbeel2004apprenticeship} ne permet pas de résoudre le problème du pendule inversé à l'aide des seuls échantillons de l'expert car ceux-ci ne couvrent qu'une partie trop petite de l'espace d'état, ne permettant pas la résolution du problème direct en se basant uniquement sur les  données de l'expert.\\
+
+Un argument revenant parfois est celui qui consiste à voir en l'expert non pas un acteur optimal mais un acteur optimal par une certaine marge.
 C'est à dire que l'action choisie par l'expert doit avoir une qualité strictement supérieure à n'importe quelle autre action.
 Les approches de planification structurée comme \citet{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} permettent d'influer sur cette marge.\\
-%
-%A recurring argument is to see the expert not only as optimal but as optimal by a certain margin.
-The action chosen by the expert should have a quality strictly greater than any other action.
-Structured planning as done in \cite{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} allow some influence on the margin.\\
-%
-%%Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
+
+Particulièrement, notre travail présente des similarités avec celui de \citet{ratliff2006maximum}.
 Les auteurs présentent une méthode d'ARI en posant le problème sous forme d'un \emph{maximum margin classification problem}, notre fonction de risque étant similaire à la leur.
 La méthode utilisée pour résoudre ce problème peut être ramenée à un algorithme itératif dans la veine de \cite{abbeel2004apprenticeship}.
-Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires: à un PDM (entrée à classifier) est associée une politique (classe).
+Vu à un plus haut niveau d'abstraction, la classification a lieu dans l'espace des trajectoires : à un PDM (entrée à classifier) est associée une politique (classe).
 Le cadre applicatif est déterministe, l'algorithme employé pour la résolution des PDM étant $A^*$.
 Il semble difficile de sortir de ce cadre  en pratique (bien que l'algorithme soit introduit dans un cadre plus général).
-%
-%Specifically, our work shows similarities with \cite{ratliff2006maximum}.
-The authors cast the \emph{IRL} problem as a maximum margin classification problem ; our risk function is similar to theirs.
-The method they use can be seen as an iterative algorithm in the tracks of \cite{abbeel2004apprenticeship}.
-From a higher abstraction level, the classification takes place in the space of trajectories : a policy (the label) is associated with a \emph{MDP} (input to classify).
-The empirical demonstration uses a deterministic setting, as they use $A^*$ to solve the \emph{MDP}.
-Although the proposed algorithm is described in a more general setting, it seems unpractical outside the deterministic case.
-%
-%%Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
+
+Finalement, \citep{ratliff2007imitation} traitent le problème de l'imitation comme un problème de classification multi-classe.
 Le principe de leur approche est exactement de résoudre le risque (Eq.
 \eqref{Jdef.eqn}).
-Toutefois, ils ne prennent pas en comptent la structure du MDP, ce à quoi nous remédions avec notre contribution (voir \cite{melo2010learning} sur l'importance de prendre en compte cette structure).
-%%Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
-Cette fonction de score $s: X\times Y \rightarrow \mathbb{R}$ est utilisée à chaque état pour prendre une décision de classification i.e.
+Toutefois, ils ne prennent pas en compte la structure du MDP, ce à quoi nous remédions avec notre contribution (voir \cite{melo2010learning} sur l'importance de prendre en compte cette structure).
+Finalement, dans \citep{ratliff2007imitation} les auteurs utilisent un système de classification multi-classe qui met en jeu une fonction de score.
+Cette fonction de score $s : X\times Y \rightarrow \mathbb{R}$ est utilisé à chaque état pour prendre une décision de classification i.e.
 choisir le label (correspondant à une action) $y\in Y$ à appliquer à l'entrée (correspondant à un état) $x\in X$.
@@
 \eqref{structure.eqn}).\\
-%Finally, \citep{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem.
-Both our approach and theirs minimize the same risk function, described Eq.
-\eqref{Jdef.eqn}, but they do not use the structure of the \emph{MDP} as we do (see \cite{melo2010learning} about the importance of taking this structure into account).\\
-%
-%%Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
+
+Notre approche se place en quelque sorte à mi-chemin entre ces deux contributions en proposant d'introduire la structure du PDM (ce que fait la première) au niveau du choix de l'action (niveau d'abstraction de la seconde).
 Il existe d'autres approches que nous ne passons pas en revue de manière exhaustive, mais à notre connaissance elles nécessitent toutes de résoudre le problème direct.
-%
-%Our approach is in a way between those two contributions as we propose to introduce the structure of the \emph{MDP} (this is what the first contribution do) in the same abstraction level as the second approach.
-Other techniques exist, but to the best of our knowledge none of them can work only from samples drawn by the expert.
-
 %%%%%
-\section{Experiments}
+\section{Illustration expérimentale}
 \label{exp.sec}
@@
 %%%
-In this section, experiments on two classical benchmarks are
-proposed: the inverted pendulum and the car driving problems.
+\subsection{Description de la tâche}
+%%%
+Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté.
+Il s'agit d'un problème classique en apprentissage par renforcement.
+Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
+Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
+Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle) la seconde étant sa vitesse angulaire.\\
 
-\subsection{Inverted Pendulum}
+La politique optimisant la fonction de récompense donnée Fig.
+\ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
+Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
+
+Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases, de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
+Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie (correspondant à la position verticale et son voisinage proche).\\
+
+ Ce dernier point est d'ailleur la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
+En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}.
+Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel moyen est utilisée en association à celle de l'expert.
+Cette politique est obtenue par optimisation d'une récompense sur le PDM dans lequel évolue l'expert ; cette résolution du problème direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état.
+Comme notre algorithme ne nécessite pas le calcul d'autre attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
 %%%
-%Le problème du pendule inversé consiste à maintenir en équilibre vertical un pendule à un degré de liberté. Il s'agit d'un problème classique en apprentissage par renforcement.
-%Pour jouer le rôle de notre expert, nous utilisons d'ailleurs la politique trouvée par LSPI, un algorithme de RL décrit par \citet{lagoudakis2003least}.
-%Les différents réglages utilisés, de la valeur de la masse du pendule codée dans notre simulateur jusqu'au choix des fonctions de base sont conformes à ce qui est donné dans cette publication.
-%Notamment les fonctions de bases forment un réseau de gaussienne couvrant l'espace d'état à deux dimensions, la première étant la position angulaire du pendule (la verticale correspondant à un angle de valeur nulle)
-%la seconde étant sa vitesse angulaire.\\
+\subsection{Illustration des résultats}
+%%%
+
+\begin{figure}[tb]
 %
-%
-The inverted pendulum problem consists in maintaining in vertical
-equilibrium a pendulum with one degree of freedom by moving the cart
-on which it is fixed. This classical problem is extensively
-described in~\cite{lagoudakis2003least} from which the parameters of
-our experiment were taken. The state space is two-dimensional: the
-first dimension is the angular position of the pendulum (where the
-vertical position corresponds to the angle zero) and the second one
-is the angular speed. The policy of the expert is determined by the
-Least Square Policy Iteration (LSPI) algorithm
-from~\cite{lagoudakis2003least}, using the same basis functions
-(Gaussian functions).
-%
-%La politique optimisant la fonction de récompense donnée Figure \ref{onlyFig.fig} (a) parvient systématiquement à maintenir le pendule en équilibre durant 3000 pas de temps.
-%Ces derniers ayant une valeur d'un dixième de seconde, cela signifie que nous interrompons l'expert après 5 minutes de contrôle sans chute du pendule.\\
-The policy of the expert optimizes the reward function given Figure
-\ref{onlyFig.fig} (a). It manages systematically to maintain the
-pendulum during 3000 time-steps. A time-step having a value of 0.1s,
-the expert is interrupted after 5 minutes of successful
-interactions.
-%
-%Le but de cette expérience est de montrer que notre algorithme est capable, à partir des seules transitions fournies par l'expert et de la définition des fonctions de bases,
-%de retrouver une récompense telle qu'un agent entraîné sur cette récompense puisse lui aussi faire balancer le pendule durant 5 minutes sans échec.
-%Cela n'est pas trivial étant donné que l'espace d'état est continu et que les données fournies par l'expert n'en couvrent qu'une petite partie
-%(correspondant à la position verticale et son voisinage proche).\\
-%
-
-The main difficulty of this experiment relies in the fact that the
-state space is continuous and, above all, that expert trajectories
-(shown in green on Figure~\ref{onlyFig.fig} (b)) are only covering a
-very small area of the state space.
-%
-%Ce dernier point est d'ailleurs la raison pour laquelle l'algorithme d'\citet{abbeel2004apprenticeship} ne parvient pas à résoudre le problème à l'aide des seules transitions de l'expert.
-%En effet, comme cela a été montré dans \citep{klein2011batch}, à moins de fournir un certain nombre de transitions aléatoires couvrant tout l'espace d'état il est impossible d'utiliser
-%les algorithmes itératifs du type de celui d'\citet{abbeel2004apprenticeship}. Ces algorithmes procèdent par itérations à l'élaboration d'une politique $\pi$ dont l'attribut vectoriel
-%moyen est utilisée en association à celle de l'expert. Cette politique est obtenue par optimisation d'une récompense sur le PDM dans lequel évolue l'expert ; cette résolution du problème
-%direct à chaque itération ne peut se faire si l'on ne dispose que de données ne couvrant qu'une petite partie de l'espace d'état. Comme notre algorithme ne nécessite pas le calcul d'autre
-%attribut vectoriel moyen que celui de l'expert, la résolution du problème direct est inutile et nous ne rencontrons pas ce problème.
-%%%
-For this reason, the algorithm of \citet{abbeel2004apprenticeship}
-fails in finding a solution to this problem only with the
-transitions available from the expert demonstrations. Indeed, as
-shown in~\citep{klein2011batch}, additional data covering the whole
-state space is required.
-%
-%These algorithms are iterative and try to
-%find a policy $\pi$ which feature expectation is compared to the
-%feature expectation of the expert. Thus, these algorithms need to
-%resolve the MDP and to do so need samples covering the whole state
-%space. Our algorithm, only need to be able to estimate the feature
-%expectation of the expert which does note require the direct
-%resolution of the MDP.
-\subsubsection{Results}
-%%%
-\begin{figure}[ht]
-%\vskip 0.2in
-\begin{center}
-\begin{minipage}[b]{.48\linewidth}
+\begin{minipage}[b]{.5\linewidth}
   \centering
-  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_true_R.pdf}}
-  \centerline{(a) Expert reward ($R^E$)}%\medskip
-\end{minipage}
-%\hfill
-\begin{minipage}[b]{.48\linewidth}
-  \centering
-  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_lafem_R.pdf}}
-  \centerline{(b) SCIRL Reward ($R_\theta$)}%\medskip
-\end{minipage}
-\begin{minipage}[b]{.48\linewidth}
-  \centering
-  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vexpert.pdf}}
-  \centerline{(c) Expert value function}%\medskip
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_true_R.pdf}}
+  \centerline{(a) Récompense de l'expert}%\medskip
 \end{minipage}
 \hfill
-\begin{minipage}[b]{.48\linewidth}
+\begin{minipage}[b]{.5\linewidth}
   \centering
-  \centerline{\includegraphics[width=\columnwidth]{LAFEM_Exp3_Vagent.pdf}}
-  \centerline{(d) SCIRL value function}%\medskip
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_lafem_R.pdf}}
+  \centerline{(b) Récompense trouvée}%\medskip
+\end{minipage}
+\begin{minipage}[b]{.5\linewidth}
+  \centering
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_Vexpert.pdf}}
+  \centerline{(c) Fonction de valeur de l'expert}%\medskip
+\end{minipage}
+\hfill
+\begin{minipage}[b]{.5\linewidth}
+  \centering
+  \centerline{\includegraphics[width=1.15\linewidth]{LAFEM_Exp3_Vagent.pdf}}
+  \centerline{(d) Fonction de valeur de l'agent}%\medskip
 \end{minipage}
 %
-\caption{SCIRL results on the inverted pendulum} \label{onlyFig.fig}
+\caption{Résultats fournis par notre algorithme.}
+\label{onlyFig.fig}
 %
-\end{center}
-\vskip -0.2in
+\vspace{-5pt}
 \end{figure}
-%
-%Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
-%La fonction $l$ fut définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon. Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations
-%est de $T=20$. Le vecteur initial $\theta_0$ fut fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attribut $\psi: S \rightarrow \mathbb{R}^p$ correspondant au mélange
-%de gaussiennes défini dans \cite{lagoudakis2003least}. Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui fut employé.\\
-%
-So as to test the SCIRL algorithm, a database $D_E$ of transitions
-given by the expert policy $\pi_E$ is built. It contains 10
-trajectories of 300 transitions each. The margin function is $l(s,a)
-= 0$ if $a=\pi_E(s)$, $1$ otherwise.
-%The time step is constant,
-%$\Delta t=0.1s$, and the stopping time is $T=20$.
-The initial vector, $\theta_0$, is fixed at $\begin{pmatrix}   -1 &
-\dots & -1
-\end{pmatrix}^T$,
-and the basis functions ($\psi: S \rightarrow \mathbb{R}^p$) used
-for the reward function are a Gaussian network as defined
-in~\cite{lagoudakis2003least}. The LSTD$\mu$
-algorithm~\cite{klein2011batch} is used to estimate the feature
-expectation $\mu^E(s,a)$.
 
-%Si la fonction de récompense trouvée par LAFEM (figure \ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (figure \ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs
-%sont très similaires (figures \ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)). Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense
-%pouvant expliquer une politique.\\
-%
-Even if the reward function found by SCIRL (figure \ref{onlyFig.fig}
-(b)) is different than the expert reward (figure \ref{onlyFig.fig}
-(a)), the value functions are similar (figures \ref{onlyFig.fig} (c)
-et \ref{onlyFig.fig} (d)). This is an illustration that the IRL
-problem is ill-posed in the sense that there is not only one reward
-function able to explain the optimality of a policy.
+Afin d'étudier le comportement de notre algorithme, nous avons fait générer à l'expert une base de données $D_E$ de transitions contenant 10 trajectoires de 300 transitions chacune.
+La fonction $l$ est définie sur les états présents dans $D_E$ telle que $l(s,a) = 0$ si $a=\pi_E(s)$, $1$ sinon.
+Le pas de temps est constant à $\alpha_t = 0.1$ et le nombre d'itérations est de $T=20$.
+Le vecteur initial $\theta_0$ est fixé à $[-1...-1]^T$, la récompense étant paramétrée à l'aide du vecteur d'attributs $\psi : S \rightarrow \mathbb{R}^p$ correspondant au mélange de gaussiennes défini dans \cite{lagoudakis2003least}.
+Pour l'approximation de l'attribut vectoriel moyen de l'expert, c'est l'algorithme LSTD$\mu$ qui est employé.\\
 
-%Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes,
-%ce qui est notre critère de réussite.\\
-With the number of samples given by the expert (3000), the
-apprentice is able to systematically maintain the pendulum in
-equilibrium during 5 minutes as the expert does.
-%
-%Deux éléments méritent d'être notés. Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
-%Ceux-ci sont portés en vert sur la Figure \ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace:
-%celle dans laquelle le pendule est proche de la verticale. En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
-%Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer
-%la fonction de récompense utilisée sans l'atteindre. Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir
-%des performances similaires à celles de l'expert. Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig. \ref{onlyFig.fig} Cela n'a aucune importance, une dilatation
-%laissant la politique optimale invariante.
-%
-Two elements deserve to be noticed. First, the expert samples are
-only covering a small area of the state space (see
-Figure~\ref{onlyFig.fig} (b)) and especially they are close to the
-vertical position. The lack of samples in the rest of the state
-space makes it impossible to infer with certainty the reward
-function. Second, the \textit{actual} reward function $R^E$
-(Figure~\ref{onlyFig.fig} (a)) does not live in the hypothesis space
-span by a linear combination of the basis functions. Despite these
-difficulties, SCIRL succeeds to extract a reward function leading to
-perform similarly to the expert. %Notice that there is a magnitude
-%difference between the SCIRL and the expert value functions on
-%Figure~\ref{onlyFig.fig} but an affine transform of the value
-%function does not change the optimal policies.
+Si la fonction de récompense trouvée par notre algorithme (Fig.
+\ref{onlyFig.fig} (b)) diffère de celle fournie à l'expert (Fig.
+\ref{onlyFig.fig} (a)) on constate en revanche que les fonctions de valeurs sont très similaires (Fig.
+\ref{onlyFig.fig} (c) et \ref{onlyFig.fig} (d)).
+Cela est une illustration du fait que le problème de l'ARI est mal posé en ceci qu'il n'y a pas qu'une seule récompense pouvant expliquer une politique.\\
 
-\subsection{Car driving}
+Avec le nombre d'échantillons fournis par l'expert (3000 en 10 trajectoires de 300 transitions chacune), l'agent parvient systématiquement à maintenir le pendule en équilibre durant 5 minutes, ce qui est notre critère de réussite.\\
 
-This task is taken from~\cite{syed2008game}. It consists in
-navigating a car through randomly-generated traffic on a threelane
-highway. The reward features are identical to~\citet{syed2008game}
-paper that is: a collision feature (0 if contact with another car,
-and 0.5 otherwise), an off-road feature (0 if on the grass, and 0.5
-otherwise), and a speed feature (0.5, 0.75 and 1 for each of the
-three possible speeds, with higher values corresponding to higher
-speeds).
-
-\subsubsection{Results}
-
-In this section, SCIRL is compared to the Multiplicative Weights for
-Apprenticeship Learning (MWAL) algorithm proposed by
-\citet{syed2008game}. This algorithm requires the perfect knowledge
-of the MDP and outputs a single policy (not a reward) that imitates
-the expert.
-
-The experimental setup is as follows. The highway simulator
-of~\cite{syed2008game} is used\footnote{available at \url{
-http://www.cs.princeton.edu/~usyed/}}. It also contains an exact MDP
-solver for the task, the transition probabilities being available.
-Two reward functions were hand-tuned ($\theta$ parameters chosen by
-hand) so as to generate two different expert policies: one for
-high-speed driving and another for safe driving. The MDP solver is
-used to obtain the expert policies given the two reward functions. A
-number $N$ of trajectories of length $H$ is generated with each
-strategy providing the expert data for training the SCIRL and MWAL
-algorithms. The MWAL algorithm requires the knowledge of the average
-feature expectation $\mu_{d0}^{\pi_E}$ for the initial state (given
-a distribution $d_0$ over starting states). To reach a fair
-comparison with SCIRL which only uses the expert trajectories,
-$\mu_{d_0}^{\pi_E}$ is computed as the Monte Carlo estimate given
-the $N$ trajectories, knowing that the initial state is always the
-same:
-%
-$$\mu_{d_0}^{\pi_E} = \frac{1}{N}\sum_{i=1}^N \sum_{t=1}^H \gamma^t \psi(s_t^i)$$
-%
-For SCIRL, the LSTD$\mu$ algorithm is used again to approximate
-$\mu^E(s,a)$ from the data. For each component of $\mu^E$, the basis
-functions are the same as those for the reward function. SCIRL
-outputs a reward function which is used to obtain a policy using the
-MDP solver. The SCIRL and MWAL policies are then run on the
-simulator for trajectories of length $T$. The average discounted
-cumulative reward, computed with regard to the initial handcrafted
-rewards, serves as the performance measure. It is a Monte Carlo
-estimate of $V^\pi(s_0)$ for each policy.
-
-This experiment is repeated $M$ times for each policy and results
-are averaged. The results are shown in
-Tables~\ref{table:highway_fast} and~\ref{table:highway_safe}
- for $N \in
-\{1, 3, 5, 10\}$, $H = 30$, $T=50$ and $M = 10$.
-
-%\begin{figure}
-%  \includegraphics[width=\linewidth]{FastResults_EB.pdf}\\
-%  \caption{Results for high-speed driving}\label{fig:highway_fast}
-%\end{figure}
-%
-%\begin{figure}
-%  \includegraphics[width=\linewidth]{SafeResults_EB.pdf}\\
-%  \caption{Results for high-speed driving}\label{fig:highway_safe}
-%\end{figure}
-
-\begin{table}
-\begin{center}
-    \begin{tabular}{|c|c|c|c|}
-    \hline
-    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
-    N     & MWAL              & SCIRL               & $V_{\text{opt}}$ \\\hline\hline
-    1     & 7.25 $\pm$ 0.10   & 7.03 $\pm$ 0.09     & 7.30              \\\hline
-    3     & 7.28 $\pm$ 0.06   & 7.07 $\pm$ 0.07     & 7.30              \\\hline
-    5     & 7.27 $\pm$ 0.11   & 6.98 $\pm$ 0.10     & 7.30              \\\hline
-    10    & 7.26 $\pm$ 0.07   & 7.05 $\pm$ 0.10     & 7.30              \\
-    \hline
-    \end{tabular}
-    \caption{SCIRL results on the car driving
-    problem (high-speed)}\label{table:highway_fast}
-\end{center}
-\end{table}
-
-\begin{table}
-\begin{center}
-    \begin{tabular}{|c|c|c|c|}
-    \hline
-    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
-    N     & MWAL              & SCIRL               & $V_{\text{opt}}$ \\\hline\hline
-    1     & 4.95 $\pm$ 0.05   & 3.27 $\pm$ 0.13     & 5.00              \\\hline
-    3     & 4.85 $\pm$ 0.13   & 4.90 $\pm$ 0.10     & 5.00              \\\hline
-    5     & 4.90 $\pm$ 0.09   & 4.83 $\pm$ 0.08     & 5.00              \\\hline
-    10    & 4.98 $\pm$ 0.03   & 4.81 $\pm$ 0.12     & 5.00              \\
-    \hline
-    \end{tabular}
-    \caption{SCIRL results on the car driving
-    problem (safe)}\label{table:highway_safe}
-\end{center}
-\end{table}
-
-From these tables, one can conclude that the SCIRL method reaches
-approximately the same performance as MWAL (given that enough
-samples are provided in the case of safe driving). Yet, SCIRL
-doesn't need any other information than the expert demonstrations
-while MWAL requires solving exactly and repeatedly the MDP.
-
+Deux éléments méritent d'être notés.
+Tout d'abord la faible couverture de l'espace d'état offerte par les échantillons de l'expert.
+Ceux-ci sont portés en vert sur la Fig.
+\ref{onlyFig.fig} (b) et l'on constate qu'ils n'occupent qu'une toute petite partie de l'espace : celle dans laquelle le pendule est proche de la verticale.
+En l'absence de données dans le reste de l'espace d'état il n'est pas possible de pouvoir y inférer avec certitude la récompense.
+Le second point à noter est que la "vraie" récompense ne se trouve pas dans l'espace d'hypothèse que notre algorithme explore, en effet les attributs vectoriels utilisés ne peuvent qu'approximer la fonction de récompense utilisée sans l'atteindre.
+Malgré ces deux difficultés, notre algorithme parvient, comme nous l'avons dit, à extraire une récompense qui permet à un agent d'obtenir des performances similaires à celles de l'expert.
+Le lecteur attentif aura noté la différence d'amplitude des fonctions de valeur Fig.
+\ref{onlyFig.fig} Cela n'a aucune importance, une dilatation laissant la politique optimale invariante.
 \section{Conclusion}
 
-%L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse. En rendant la résolution du problème inverse indépendante
-%de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
-In this paper, we propose an algorithm for inverse reinforcement
-learning (IRL) that we call SCIRL. It is based on a classification
-approach in which the structure of the underlying Markov Decision
-Process (MDP) is taken into account.
+L'algorithme présenté dans cette contribution lève les problèmes les plus contraignants de l'apprentissage par renforcement inverse.
+En rendant la résolution du problème inverse indépendante de celle du problème direct, nous sommes en mesure d'inférer une récompense cohérente avec le comportement d'un expert en mode purement \emph{batch}.\\
 
-This approach relaxes most of the constrains of other IRL algorithms
-found in the literature. It doesn't require the resolution of the
-direct problem and is able to infer a reward function which is
-consistent with the expert's behavior in a purely \emph{batch} mode.
+Comme nous l'avons illustré, les échantillons nécessaires à la réussite de notre algorithmes sont faciles à recueillir.
+Une simple trace de l'expert, même si elle ne couvre qu'une petite partie de l'espace d'état, suffit à inférer une récompense.
+Cette récompense, description compacte de la tâche effectuée par l'expert, peut alors être optimisée par un agent dont les capacités diffèrent de celles de l'expert.
+On rentre dans le cadre du \emph{transfer learning}.\\
+
+L'estimation de $\mu_E$ est centrale dans notre approche.
+Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients que les algorithmes d'estimation d'une fonction de valeur.
+L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen d'introduire autrement la structure du PDM dans la démarche de classification que nous avons adoptée.
+Des tests empiriques sur des problèmes plus complexes sont également envisagés.
 %
-%Comme nous l'avons illustré, les échantillons nécessaires à la réussite de notre algorithmes sont faciles à recueillir.
-%Une simple trace de l'expert, même si elle ne couvre qu'une petite partie de l'espace d'état, suffit à inférer une récompense.
-%Cette récompense, description compacte de la tâche effectuée par l'expert, peut alors être optimisée par un agent dont les capacités diffèrent de celles de l'expert.
-%On rentre dans le cadre du \emph{transfer learning}.\\
+% Bibliographie
 %
-A mere trajectory of the expert, even if it covers a small part of
-the state space, is sufficient to infer a reward function and no
-additional data nor a simulator is required. It also outputs a
-reward and not a strategy which is the real goal of the IRL
-paradigm. The learnt reward function, which is a compact description
-of the task executed by the expert, could then be optimized by an
-agent with abilities (action set) that are different from those of
-the expert which places us in in the even more challenging framework
-of \emph{transfer learning}.
-%
-%L'estimation de $\mu_E$ est centrale dans notre approche. Nous disposons d'un algorithme efficace avec LSTD$\mu$, mais qui présente cependant les mêmes inconvénients
-%que les algorithmes d'estimation d'une fonction de valeur. L'axe d'étude que nous envisageons d'explorer serait de ne plus passer par l'attribut vectoriel moyen en trouvant le moyen
-%d'introduire autrment la structure du PDM dans la démarche de classification que nous avons adoptée. Des tests empiriques sur des problèmes plus comlpexes sont également envisagés.
+\bibliography{Biblio}
 
-SCIRL is shown to compare positively to a state-of-the-art algorithm
-(the MWAL algorithm) on a standard benchmark (the car driving
-problem). It also performs well on a problem with a two-dimensional
-continuous state space which is sparsely explored by the expert (the
-inverted pendulum problem). This later problem could not be solved
-with state-of-the-art algorithms without additional knowledge.
-
-Yet, SCIRL requires the computation of $\mu^E(s,a)$ everywhere on
-the expert trajectories. Here, we made use of the LSTD$\mu$
-algorithm to estimate it. However LSTD$\mu$ shares the drawbacks of
-value function estimation algorithms. In the future, we plan to
-extend our approach so as to avoid the computation of the expert's
-feature expectations. Empirical tests on more complex problems will
-also be considered.
-
-\bibliography{Biblio}
-\bibliographystyle{icml2012}
 \end{document}

### 2012-05-14 piot2012cascading/main_KLEIN_v3.tex
Début d'explication pour le Gridorld

@@
 \subsection{Instantiation of the algorithm}
-As hinted to FIXME(Plus insister sur cet aspect) in Section \ref{section: Cascading}, our algorithm can be instanciated in many ways, as any off-the-shelf classifier using a score function can be used in the first step, and any regression method can be used in the second step.
+As hinted to in Section \ref{section: Cascading}, our algorithm can be instanciated in many ways, as any off-the-shelf classifier using a score function can be used in the first step, and any regression method can be used in the second step.
 This freedom of choice leaves room for problem dependant tweaks if needed.\\
@@
 \subsection{Experimental setting}
-For each problem we test our approach with, we compare it to two baselines : the mean performance of an agent trained on a random reward and the mean performance of the policy output by the algorithm of \cite{abbeel2004apprenticeship}.
+For both problems we test our approach with, we compare it to two baselines : the mean performance of an agent trained on a random reward and the mean performance of the policy output by the algorithm of \cite{abbeel2004apprenticeship}.
 This algorithm can be instanciated to work in the same dire conditions as our algorithm, that is to say using only samples from the expert.
 To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ (\cite{klein2011batch}) and LSPI (\cite{lagoudakis2003least}).
-\subsection{Inverted pendulum}
-In \cite{klein2011batch}, it is noted that the particular instanciation of Abbeel and Ng's algorithm we use as a baseline is unable to solve the Inverted Pendulum problem (described in \cite{lagoudakis2003least}).
+When this does not work, we fall back to a setting where everything is knows, that is to say the feature expectations are computed by a dynamic programming algorithm that make use of the transition probabilities.\\
+The performance and the expert and the agents is computed with respect to the reward the expert was trained with. We use as a performance metric the mean value over the state space.
+\subsection{Gridworld}
+We first tested our approach on a simple $5\times 5$ gridworld.
+The expert and the agent can choose between 4 actions (down, left right, or up) that sends the player in the corresponding neighbooring cell.
+Trying to go off the grid will result in the player staying in its position.
+Each action has a probability $0.3$ of failing, in which case another randomly chosen different action is executed.
+The expert begins in the lower left corner and is trained to go to the higher right corner by an exact dynamic programming algorithm.
+The original reward is null everywhere but in the higher right corner where it is $1$.\\
 
-There exist settings this particular instantiation of \cite{abbeel2004apprenticeship}'s algorithm is not able to solve.
-In \cite{klein2011batch} the authors note that this instanciation the problem.
-We then fall back to an instantiation using suroutines that breaks the batch, expert-only setting in order to have something to compare to.
-Precise workarounds will be described as they are used.
-It is important to note that sometimes our algorithm is at the best of our knowledge FIXME(c'est pas vrai il y a SCIRL aussi ;) the only one able to solve the problem with such a constrained input.
 \subsection{Highway driving simulator}
@@
 Although the mean value with natural features is lower than with the engineered ones, we see that there is less risk of catastrophic failure (the min value is higher) and, more importantly, we were actually able to solve the problem in a satisfying manner with.
+\subsection{Pendule}
+A n'ajouter que si les résultats présentent un intérêt particulier par rapport aux deux autres problèmes.
 
@@
 \end{itemize}
-\subsection{Gridworld}
-\begin{itemize}
-\item description du problème
-\item Introduction du task transfert (l'expert est une sorte de tour, l'agent un roi)
-\item Résultats pour l'IRL (comparaison avec ANIRL)
-\item résultats pour le task transfer, impossible de comparer avec ANIRL car officiellement pas de récompense
-\end{itemize}
-\subsection{Pendule}
-A n'ajouter que si les résultats présentent un intérêt particulier par rapport aux deux autres problèmes.
 \section{Related Work}

### 2012-05-14 piot2012cascading/main_KLEIN_v3.tex
Passage au style NIPS

--- 2012-04-23 piot2012cascading/main_KLEIN_v3.tex
@@
-% This is LLNCS.DEM the demonstration file of
-% the LaTeX macro package from Springer-Verlag
-% for Lecture Notes in Computer Science,
-% version 2.4 for LaTeX2e as of 16. April 2010
-%
-\documentclass{llncs}
-%
+\documentclass{article} % For LaTeX2e
+\usepackage{nips12submit_e,times}
 \usepackage{makeidx}  % allows for indexgeneration
@@
 \usepackage{amssymb}
+\usepackage{amsthm}
+\newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{remark}{Remark}
 \usepackage{dsfont}
@@
 \newcommand{\sgn}{\operatorname{sgn}}
+\title{A cascading approach}
+\author{
+David S.~Hippocampus\thanks{ Use footnote for providing further information
+about author (webpage, alternative address)---\emph{not} for acknowledging
+funding agencies.} \\
+Department of Computer Science\\
+Cranberry-Lemon University\\
+Pittsburgh, PA 15213 \\
+\texttt{hippo@cs.cranberry-lemon.edu} \\
+\And
+Coauthor \\
+Affiliation \\
+Address \\
+\texttt{email} \\
+\AND
+Coauthor \\
+Affiliation \\
+Address \\
+\texttt{email} \\
+\And
+Coauthor \\
+Affiliation \\
+Address \\
+\texttt{email} \\
+\And
+Coauthor \\
+Affiliation \\
+Address \\
+\texttt{email} \\
+(if needed)\\
+}
+
+% The \author macro works with any number of authors. There are two commands
+% used to separate the names and addresses of multiple authors: \And and \AND.
+%
+% Using \And between authors leaves it to \LaTeX{} to determine where to break
+% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
+% puts 3 of 4 authors names on the first line, and the last on the second
+% line, try using \AND instead of \And before the third author name.
+
+\newcommand{\fix}{\marginpar{FIX}}
+\newcommand{\new}{\marginpar{NEW}}
+
 \begin{document}
 
-\title{Titre}
-\titlerunning{Titre}  % abbreviated title (for running head)
-\author{Boul\inst{1} \and Bil\inst{2}
-Boul \and Bil}
-\authorrunning{} % abbreviated author list (for running head)
-%
-\institute{SUPELEC,\\
-\email{SUPELEC},\\ WWW home page:
-\texttt{SUPELEC}
-\and
-\\}
 \maketitle              % typeset the title of the contribution
@@
 
- \begin{figure}
- \begin{tabular}{ccc}
- \subfigure[Abbeel and Ng IRL  with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig3"}\label{Cascading_Exp1_3.fig}}&\hspace{.05\linewidth}&
- \subfigure[Cascading with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig2"}\label{Cascading_Exp1_2.fig}}\\
- \subfigure[Cascading with natural features]{\includegraphics[width=.4\linewidth]{"Cascading_Exp1_fig1"}\label{Cascading_Exp1_1.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Cascading_Exp1_legend"}
- \end{tabular}
-\caption{Results on the Highway Driving problem. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
- \end{figure}
+%%  \begin{figure}
+%%  \begin{tabular}{ccc}
+%%  \subfigure[Abbeel and Ng IRL  with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig3"}\label{Cascading_Exp1_3.fig}}&\hspace{.05\linewidth}&
+%%  \subfigure[Cascading with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig2"}\label{Cascading_Exp1_2.fig}}\\
+%%  \subfigure[Cascading with natural features]{\includegraphics[width=.4\linewidth]{"Cascading_Exp1_fig1"}\label{Cascading_Exp1_1.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Cascading_Exp1_legend"}
+%%  \end{tabular}
+%% \caption{Results on the Highway Driving problem. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
+%%  \end{figure}
 
@@
 
-\bibliographystyle{splncs}
+\bibliographystyle{alpha}
 \bibliography{Biblio}

### 2012-05-14 piot2012cascading/GridWorld_plot.py
Données et code pour les figures relatives au GridWorld.
Les figures sotn moches, je vais faire autrement, il faut aussi corriger le problème d'abcisses entre ANIRL et le Cascading en faisant tourner à nouveau ANIRL.

@@ -1,0 +1,71 @@
+from matplotlib import rc
+rc('text', usetex=True)
+import sys
+sys.path+=['..']
+from numpy import *
+import scipy
+import pylab as pylab
+from Plot import *
+
+D_Cascading = genfromtxt( "Cascading_Exp3.mat" )
+D_C = D_Cascading[:,[0,1]]
+D_Classif = D_Cascading[:,[0,2]]
+D_ANIRL = genfromtxt( "ANIRL_Exp3.mat" )
+
+Expert = 4.11057591 #python Expert.py to get this value
+Random_mean = 0.48848324670295395#See GridWorld.org about Random.py for information on these values
+Random_min = 0.069469005947400006#python Random.py to get this value
+Random_max = 3.2720195801399998#python Random.py to get this value
+Random_var = 0.57868038965027513#python Random.py to get this value
+
+y_min = -1
+y_max = 5
+
+#Figure 1 : ANIRL, Random and Cascading (comparison to the state of the art)
+pylab.figure(1)
+pylab.clf()
+[X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_ANIRL ))
+[X,Y_mean,Y_var] = map( array, mean_var( D_ANIRL ))
+x_min = min(X)
+x_max = max(X)
+pylab.axis([x_min,x_max,y_min,y_max])
+pylab.xlabel('Number of samples from the expert')
+pylab.ylabel('${1\over card(S)}\sum\limits_{s\in S}V(s)$')
+pylab.grid(True)
+filled_mean_min_max( pylab, X, Y_mean, Y_min, Y_max, 'blue', 0.2,'--',None,None)
+filled_mean_min_max( pylab, X, Y_mean, Y_mean - Y_var, Y_mean + Y_var, 'blue', 0.4,'-.',None,None)
+
+filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), Random_min*ones(X.shape), Random_max*ones(X.shape), 'cyan',0.2,'--',"Agent trained on a random reward",None)
+filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), (Random_mean-Random_var)*ones(X.shape), (Random_mean+Random_var)*ones(X.shape), 'cyan',0.4,'-.',None,None)
+
[…]

### 2012-05-14 piot2012cascading/main_KLEIN_v3.tex
Code, données et blabla sur le GridWorld.
On a plus le saut de variance bizarre dans ANIRL, on a les mêmes abcisses et les graphes sont jolis.

@@
 To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ (\cite{klein2011batch}) and LSPI (\cite{lagoudakis2003least}).
-When this does not work, we fall back to a setting where everything is knows, that is to say the feature expectations are computed by a dynamic programming algorithm that make use of the transition probabilities.\\
-The performance and the expert and the agents is computed with respect to the reward the expert was trained with. We use as a performance metric the mean value over the state space.
+When this does not work, we fall back to a setting where everything is known, that is to say the feature expectations are computed by a dynamic programming algorithm that make use of the transition probabilities.\\
+The performance of the expert and the agents is computed with respect to the reward the expert was trained with. We use as a performance metric the mean value over the state space.
 \subsection{Gridworld}
@@
 The original reward is null everywhere but in the higher right corner where it is $1$.\\
+The results are shown in two figures.
+Figure \ref{Fig1.fig} show the baselines we compare ourselves with.
+Pure classification and the algorithm of \cite{abbeel2004apprenticeship} are on par with one another, both far better than training an agent on a random reward ; they both converge to a value close to the expert's given enough samples.
+Figure \ref{Fig2.fig} shows results of our algorithm together with the mean performance of both forementionned approaches (variance and min/max were dropped for legibility's sake).
+It shows the soundness of our cascading approach : when only a few samples are available, we provide a reward that lead to a better policy.
+When samples are widely available all approaches converge to the same value.
+  \begin{figure}
+  \begin{tabular}{ccc}
+  \subfigure[State of the art approaches on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig1"}\label{Fig1.fig}}&\hspace{.05\linewidth}&
+  \subfigure[Our new approach on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig2"}\label{Fig2.fig}}\\
+%%  \subfigure[Cascading with natural features]{\includegraphics[width=.4\linewidth]{"Cascading_Exp1_fig1"}\label{Cascading_Exp1_1.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Cascading_Exp1_legend"}
+  \end{tabular}
+ \caption{Results on the GridWorld problem. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
+  \end{figure}
 
 \subsection{Highway driving simulator}
+There are problems for which pure classification just does not cut it, as the number of samples that would be needed 
 Inspired from a problem seen in \cite{syed2008apprenticeship,syed2008game}, we coded a driving simulator.
@@
 
-%%  \begin{figure}
-%%  \begin{tabular}{ccc}
-%%  \subfigure[Abbeel and Ng IRL  with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig3"}\label{Cascading_Exp1_3.fig}}&\hspace{.05\linewidth}&
-%%  \subfigure[Cascading with engineered features]{\includegraphics[width=.45\linewidth]{"Cascading_Exp1_fig2"}\label{Cascading_Exp1_2.fig}}\\
-%%  \subfigure[Cascading with natural features]{\includegraphics[width=.4\linewidth]{"Cascading_Exp1_fig1"}\label{Cascading_Exp1_1.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Cascading_Exp1_legend"}
-%%  \end{tabular}
-%% \caption{Results on the Highway Driving problem. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
-%%  \end{figure}
-
 \begin{itemize}

### 2012-05-14 piot2012cascading/GridWorld_plot.py

@@ -21,7 +21,7 @@ Random_var = 0.57868038965027513#python Random.py to get this value
 y_min = -1
 y_max = 5
 
-#Figure 1 : ANIRL, Random and Cascading (comparison to the state of the art)
+#Figure 1 : ANIRL, Random and Classif (Concurrents...)
 pylab.figure(1)
 pylab.clf()
 [X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_ANIRL ))
@@ -40,15 +40,15 @@ filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), (Random_mean-Random_va
 
 pylab.plot(X,Expert*ones(X.shape), color='cyan',label="Expert",lw=2,ls=':')
 
-[X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_C ))
-[X,Y_mean,Y_var] = map( array, mean_var( D_C ))
-filled_mean_min_max( pylab, X, Y_mean, Y_min, Y_max, 'orange', 0.2,'--',None,None)
-filled_mean_min_max( pylab, X, Y_mean, Y_mean - Y_var, Y_mean + Y_var, 'orange', 0.4,'-.',None,None)
+[X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
+[X,Y_mean,Y_var] = map( array, mean_var( D_Classif ))
+filled_mean_min_max( pylab, X, Y_mean, Y_min, Y_max, 'yellow', 0.2,'--',None,None)
+filled_mean_min_max( pylab, X, Y_mean, Y_mean - Y_var, Y_mean + Y_var, 'yellow', 0.4,'-.',None,None)
 
 pylab.savefig("Fig1.pdf",transparent=True)
 
-#Figure 2 : Cascading et classif pure
-pylab.figure(1)
+#Figure 2 : Cascading et moyenne de ANIRL et Classif pure (comparaison aux concurrents).
+pylab.figure(2)
 pylab.clf()
 [X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
 [X,Y_mean,Y_var] = map( array, mean_var( D_Classif ))
@@ -58,10 +58,14 @@ pylab.axis([x_min,x_max,y_min,y_max])
 pylab.xlabel('Number of samples from the expert')
[…]

### 2012-05-14 piot2012cascading/main_KLEIN_v3.tex
Code et Données résultats sur le Highway.
La légende, aussi.

@@
   \subfigure[Our new approach on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig2"}\label{Fig2.fig}}\\
-%%  \subfigure[Cascading with natural features]{\includegraphics[width=.4\linewidth]{"Cascading_Exp1_fig1"}\label{Cascading_Exp1_1.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Cascading_Exp1_legend"}
+  \subfigure[All approaches on the Highway driving simulator]{\includegraphics[width=.4\linewidth]{"Fig3"}\label{Fig3.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Legend"}
   \end{tabular}
- \caption{Results on the GridWorld problem. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
+  \caption{Results on the GridWorld and Highway problems. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
   \end{figure}
@@
 \subsection{Highway driving simulator}
-There are problems for which pure classification just does not cut it, as the number of samples that would be needed 
-Inspired from a problem seen in \cite{syed2008apprenticeship,syed2008game}, we coded a driving simulator.
+There are problems for which pure classification just does not cut it, as the number of samples that would be needed to imitate the expert's control is dauntingly large.
+To illustrate this, inspired from a problem seen in \cite{syed2008apprenticeship,syed2008game}, we coded a driving simulator.
 The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
@@
 We tried to train an agent on the reward used by the expert (which was trained using an exact dynamic programming approach), but on top of being extremely slow to run (the LSTD$Q$ subroutine implies a matrix inversion) it does not converge.\\
-As LSPI is used in \cite{abbeel2004apprenticeship}'s algorithm to solve the MDP, it is impossible to use the latter with the natural features.
-We thus have to spend quite some time engeineering new informative features to make LSPI work.
-These new features have a dimensionality of $5$, including a constant term and coordinates whose semantics is \emph{The agent should turn right}, \emph{The agent should turn left}, \emph{The agent should increase its speed} and \emph{The agent should do nothing}.
-With these features in hand, we were able to solve the IRL problem with \cite{abbeel2004apprenticeship}'s algorithm.
-The results of our experiment is given on two figures, Figure \ref{Cascading_Exp1_1.fig} and \ref{Cascading_Exp1_2.fig}.
-The data presented here is the mean, minimum and maximum value obtained over 50 runs.
-The random baseline presented here is the mean, min adn max performance of an agent over 50 different random rewards.\\
-Figure \ref{Cascading_Exp1_1.fig} shows results when employing the Cascading algorithm with the natural set of features and Figure \ref{Cascading_Exp1_2.fig} shows results when using the engineered features, comparing it the the results of \cite{abbeel2004apprenticeship}'s algorithm which could be applied in this case.
-Although the mean value with natural features is lower than with the engineered ones, we see that there is less risk of catastrophic failure (the min value is higher) and, more importantly, we were actually able to solve the problem in a satisfying manner with.
-\subsection{Pendule}
-A n'ajouter que si les résultats présentent un intérêt particulier par rapport aux deux autres problèmes.
-
-\begin{itemize}
-\item description du problème
-\item deux sets de features pour faire marcher LSPI donc ANIRL
-\item résultats
-\item Insister sur le fait qu'on est meilleur quand il y a peu de samples
-\item Insisiter sur le fait qu'en features naturelles, LSPI ne fonctionne pas
-\end{itemize}
+As LSPI is used in our instanciation of \cite{abbeel2004apprenticeship}'s algorithm to solve the MDP, it is impossible to use the latter with the natural features.
+Although it proved possible to engineer features so that LSPI converges thus making the state of the art IRL algorithm work, this is outside the scope of this document.
+As a baseline we used an instanciation of \cite{abbeel2004apprenticeship}'s algorithm that imake use of all the information about the problem.
+It is displayed as an horizontal line in figure \ref{Fig3.fig} because it does not use samples from the expert, instead it uses the transition probabilities and the expert's policy to compute the exact feature expectation of the expert and every intermediate policy.\\
+As predicted, pure classification is not sample efficient enough to provide a good control whereas our approach is able to generalize from very little data and exhibits performance on par with the \emph{fully-informed} state of the art algorithm when given as few as 100 samples.\\
+Furthermore, the state of the art algorihtm make use of intermediate policies, which is computationally costly particularily when the state space gets big.
+Our approach computational cost scales with the number of samples we are given, which is as we demonstrated a small number.
+Experimentally, when producing the data for our graph, it took less time to run the cascading approach 50 times for 6 abcissa than to run \cite{abbeel2004apprenticeship}'s algorithm once.\\
 \section{Related Work}
 \label{section: related work}
+All the IRL approaches summed up in \cite{neu2009training}, namely the historical approach of \cite{ng2000algorithms} as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} share the same pivotal argument of trying to minimize some distance between the feature expectation of the expert and the feature expectation of another policy through successive updates of this intermediate policy.
+The main problem with these algorithm is that the update on the policy is done through the update of a reward function.
+This means that the forward problem of RL must be repeatedly solved ; this is often non trivial, and almost always computationally costly.\\
 
+Our algorithm avoid this pitfall, specifically, one needs more information to solve the forward problem than to just compute the feature expectation of the expert as our algorithm does.
+Even a recent contribution like \cite{boularias2011relative} still needs samples that cover a large portion of the state space although it does not explicitely require the ability to solve the forward problem.\\
+
+A recurring argument is to see the expert not only as optimal but as optimal by a certain margin.
+The action chosen by the expert should have a quality strictly greater than any other action.
+Structured planning as done in \cite{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} allow some influence on the margin.
+Through the $l$ function of our classifier, we share the same feature.\\
+
+Finally, \cite{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem.
+As seen in the experimental results, the power of generalization of our approach is greater than pure classification thanks to the regression step.
+This allows us to be more sample efficient than any other published approach known to us.
 \section{Conclusion}
+We provide a new solution to the IRL problem that avoid most pitfalls of existing approaches.
+It outputs a reward function, the most succinct and transferrable description of a task without having to repeatedly solve the forward RL problem.
+The only information our algorithm need is provided by samples from the expert, which is to the best of our knowledge a feature that can not be found in any other published algorithm.
+Even recent other approaches need randomly drawn samples.
+We provided theoretical results showing the soundness of our approach as well as experimental benchmarks comparing it to a widely known state of the art algorithm.
+In a near furure, experiments on real world robotics problems will take place.
+We hope to benchmark our algorithm against the few other approaches shown to work in such a difficult context.
 

### 2012-05-14 piot2012cascading/GridWorld_plot.py

@@ -42,8 +42,8 @@ pylab.plot(X,Expert*ones(X.shape), color='cyan',label="Expert",lw=2,ls=':')
 
 [X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
 [X,Y_mean,Y_var] = map( array, mean_var( D_Classif ))
-filled_mean_min_max( pylab, X, Y_mean, Y_min, Y_max, 'yellow', 0.2,'--',None,None)
-filled_mean_min_max( pylab, X, Y_mean, Y_mean - Y_var, Y_mean + Y_var, 'yellow', 0.4,'-.',None,None)
+filled_mean_min_max( pylab, X, Y_mean, Y_min, Y_max, 'green', 0.2,'--',None,None)
+filled_mean_min_max( pylab, X, Y_mean, Y_mean - Y_var, Y_mean + Y_var, 'green', 0.4,'-.',None,None)
 
 pylab.savefig("Fig1.pdf",transparent=True)
 
@@ -65,7 +65,7 @@ pylab.plot( X, Y_mean,color='blue',lw=2)
 
 [X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
 [X,Y_mean,Y_var] = map( array, mean_var( D_Classif ))
-pylab.plot( X, Y_mean,color='yellow',lw=2)
+pylab.plot( X, Y_mean,color='green',lw=2)
 
 [X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_C ))
 [X,Y_mean,Y_var] = map( array, mean_var( D_C ))

### 2012-05-21 piot2012cascading/main_KLEIN_v3.tex
Amélioration de détails dans le papier et nouvelles courbes pour Matthieu

@@
 \newcommand{\sgn}{\operatorname{sgn}}
-\title{A cascading approach}
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-(if needed)\\
+\title{Learning a reward function from demonstrations: a cascaded supervised
+learning approach}
+\author{Edouard Klein$^{1,2}$\\
+ 1. ABC Team\\
+ LORIA-CNRS, France.
+\And Bilal Piot$^{2}$\\
+ 2. Supélec-Metz Campus\\
+ MaLIS Research group, France\\
+\And Matthieu Geist$^1$\\
+\texttt{prenom.nom@supelec.fr}\\
+\And Olivier Pietquin$^{2,3}$\\
+3. UMI 2958 CNRS\\
+GeorgiaTech, France
 }
@@
 
-Given a sequential decision making problem posed in the MDP formalism, it is often difficult to specify manually a reward function in order to realize a given task.
-However, observing an expert demonstrating a task is often possible and more easier~\cite{ng2000algorithms}.
-That is why learning from demonstrations and more particularly IRL is an interesting problem.
-The IRL framework, fisrt introduced in~[\cite{russell1998learning},\cite{ng2000algorithms}], assumes that an expert, demonstrating a task, is acting optimally in a MDP with respect to an unknown reward function to be discovered.
-Unlike direct methods of learning from demonstrations which learn a mapping from states to actions via a supervised learning~[\cite{atkeson1997robot},\cite{pomerleau1989alvinn}], IRL methods must recover a reward function which is the most compact and tranferable way to specify a task.
-Moreover most of the direct methods do not use the structure of the MDP (one notable excepetion is~\cite{melo2010learning}), hence mimic the policy of the expert and are not able to generalize correctly the expert policy outside state examples.
-In order to overcome this drawback, undirect methods known as apprenticeship learning were introduced: the first one is~\cite{ng2004feature}.
-These methods (see~\cite{neu2009training} for a good survey) uses the IRL framework but most of them assume the dynamic of the environment known, or an environment model (via a simulator) must be available so as to test the effects of a policy.
+Given a sequential decision making problem posed in the Markov Decision Process (MDP) formalism, it is often difficult to specify manually a reward function in order to realize a given task.
+However, observing an expert demonstrating a task is often possible and easier~\cite{ng2000algorithms}.
+That is why learning from demonstrations and more particularly Inverse Reinforcement Learning (IRL) has received more and more interest this last decade.
+The IRL framework, first introduced in~\cite{russell1998learning},\cite{ng2000algorithms}, assumes that an expert, demonstrating a task, is acting optimally in a MDP with respect to an unknown reward function to be discovered.
+Unlike direct methods of learning from demonstrations which learn a mapping from states to actions via supervised learning(SL)~\cite{atkeson1997robot},\cite{pomerleau1989alvinn}, IRL methods aim at recovering a reward function which is the most compact and transferable way to specify a task.
+Moreover most of the direct methods do not use the structure of the MDP (one notable exception is~\cite{melo2010learning}), hence mimic the policy of the expert and are not able to generalize correctly the expert policy outside state examples.
+In order to overcome this drawback, undirect methods known as apprenticeship learning~\cite{abbeel2004apprenticeship} were introduced.
+These methods (see~\cite{neu2009training} for a comprehensive overview) uses the IRL framework but most of them assume the dynamic of the environment known, or an environment model (via a simulator) must be available so as to test the effects of a policy.
 Moreover, most of existing undirect approaches require to solve the direct reinforcement learning problem (find an optimal policy knowing a given reward) several times (one notable exception is~\cite{boularias2011relative}).
-Finally, in many cases, the output of the algorithm is a policy but not a reward function (see section \ref{section: related work} for an in-depth presentation).
-These constraints can lead to the development of algorithms unapplicable to real problems and make the problem of apprenticeship learning harder than reinforcement learning.
+Finally, in many cases, the output of the algorithm is a policy but not a reward function (see Section \ref{section: related work} for an in-depth presentation).
+These constraints can lead to the development of algorithms unapplicable to real-world applications and make the problem of apprenticeship learning harder than reinforcement learning.
 That is why a reborn interest for direct methods which passively imitate the expert exists.
-More particularly the reduction of the apprenticeship learning to classification, first introduced in~\cite{zadrozny2003cost} and recently used in~\cite{melo2010learning}, is legitimated for finite horizons problems in~[\cite{syed2010reduction},\cite{ross2010efficient}].
-Our approach tries to avoid the drawbacks of the direct and undirect methods by cascading two well known supervised learning(see section \ref{section: Cascading}): first a step of classification and then a step of regression which introduces the MDP structure and  outputs a reward function (see \ref{section: Cascading}).
-Our approach doesn't require any of the following: complete trajectories from an expert, a generative model of the environment, the knowledge of the transition probabilities, the ability to compute a (near)-optimal policy for different rewards and hence resolve the forward reinforcement learning (RL) problem, the expert's policy.
+More particularly the reduction of the apprenticeship learning to classification, first introduced in~\cite{zadrozny2003cost} and recently used in~\cite{melo2010learning}, is legitimated for finite horizons problems in~\cite{syed2010reduction},\cite{ross2010efficient}.
+Our approach tries to avoid the drawbacks of the direct and undirect methods by cascading two well known supervised learning(see Section \ref{section: Cascading}): first a step of classification and then a step of regression which introduces the MDP structure and  outputs a reward function (see Section \ref{section: Cascading}).
+Our approach doesn't require any of the following: complete trajectories from an expert, a generative model of the environment, the knowledge of the transition probabilities, the ability to compute a (near)-optimal policy for different rewards and hence resolve the forward reinforcement learning (RL) problem, the perfect knowledge of the expert's policy.
 In addition, only a small amount of expert demonstrations (not even in the form of trajectories but simple transitions) is required.\\
-The remainder of the paper is oraganized as follows: section \ref{section: background} gives us the different notations used in the paper and presents briefly the RL and IRL frameworks, section \ref{section: Cascading} present our method which is legitimated in section \ref{section: Cascading}, \ref{section: experiments} presents an instanciation of our method and tests it on different benchmarks and finally section \ref{section: related work} compares our method to the state of the art.
+The remainder of the paper is organized as follows: Section \ref{section: background} gives us the different notations used in the paper and presents briefly the RL and IRL frameworks, Section \ref{section: Cascading} present our method which is legitimated in Section \ref{section: Cascading}, Section \ref{section: experiments} presents an instantiation of our method and tests it on different benchmarks and finally Section \ref{section: related work} compares our method to the state of the art.
 
@@
 \label{section: background}
-A (finite) MDP \cite{puterman1994markov} is a tuple $\{S,A,P,\gamma,R\}$ where $S=\{s_i\}_{1\leq i \leq N}$ is the finite state space with $N\in\mathbb{N}^*$ states, $A=\{a_k\}_{1\leq k \leq K}$ is the finite action space with $K\in\mathbb{N}^*$ actions, $P=\{P_{s,a}\}_{(s,a)\in S\times A}$ is the set of transition probabilities where $P_{s,a}$ is a distribution probability over $S$, $P$ represents the dynamics of the MDP and the notation $p(s,a,s')=P_{s,a}(s')$ is often used and quantifies the probability to reach $s'\in S$ knowing that the action $a \in A$ was taken in the state $s\in S$, $\gamma\in]0,1[$ is the discount factor and $R$ is a function from $S\times A$ to $\mathbb{R}$ called the reward function.
+A (finite) MDP \cite{puterman1994markov} is a tuple $M=\{S,A,P,\gamma,R\}$ where $S=\{s_i\}_{1\leq i \leq N}$ is the finite state space with $N\in\mathbb{N}^*$ states, $A=\{a_k\}_{1\leq k \leq K}$ is the finite action space with $K\in\mathbb{N}^*$ actions, $P=\{P_{s,a}\}_{(s,a)\in S\times A}$ is the set of transition probabilities where $P_{s,a}$ is a distribution probability over $S$, $P$ represents the dynamics of the MDP and the notation $p(s,a,s')=P_{s,a}(s')$ is often used and quantifies the probability to reach $s'\in S$ knowing that the action $a \in A$ was taken in the state $s\in S$, $\gamma\in]0,1[$ is the discount factor and $R$ is a function from $S\times A$ to $\mathbb{R}$ called the reward function.
 A deterministic policy $\pi$ is a function from $S$ to $A$, $\Pi$ is the set of deterministic policies and the value function $V^\pi_R$ is a function from $S$ to $\mathbb{R}$ defined by:
@@
 \end{equation}
-%Inversely, a real column-vector $X=[X_i]_{1\leq i \leq N}$ can be identified to the function from $S$ to $\mathbb{R}$ that maps $s_i$ to $X_i$ for all $i\in\{1,\dots,N\}$ and the real rectangular-matrice $Y=[Y_{i,k}]_{1\leq i \leq N, 1\leq k \leq K}$ can be identified to the function from $S\times A$ to $\mathbb{R}$ that maps $(s_i,a_k)$ to $Y_{i,k}$ for all $(i,k)\in\{1,\dots,N\}\times\{1,\dots,K\}$.
-For a finite-MDP $\{S,A,P,\gamma,R\}$ the existence of an optimal policy for any function $R$ is shown in \cite{bertsekas2001dynamic}.
+For a finite-MDP $M=\{S,A,P,\gamma,R\}$, the existence of an optimal policy for any function $R$ is shown in \cite{bertsekas2001dynamic}.
 As $S=\{s_i\}_{1\leq i \leq N}$ and $A=\{a_i\}_{1\leq i \leq K}$ are finite, a function $X$ from $S$ to $\mathbb{R}$ can be identified to the real column-vector $[X(s_i)]_{1\leq i \leq N}$ and a function $Y$ from
@@
 We will also need the following matrices and column-vectors defined for any dynamic $P$, any policy $\pi$ and reward function $R$ :$P_a=[P_{s_i,a}(s_j)]_{1\leq i,j \leq N}$, $P_\pi=[P_{s_i,\pi(s_i)}(s_j)]_{1\leq i,j \leq N}$, $R_\pi=[R(s_i,\pi(s_i))]_{1\leq i,j \leq N}$.
-The transposition of a real column-vector $X=[X_i]_{1\leq i \leq M}$, $M\in\mathbb{N}^*$, is the real row-vector $X^T$ and $\|X\|_{\infty}=\max_{1\leq i \leq N}|X_i|$.
+The transposition of a real column-vector $X=[X_i]_{1\leq i \leq I}$, $I\in\mathbb{N}^*$, is the real row-vector $X^T$ and $\|X\|_{\infty}=\max_{1\leq i \leq I}|X_i|$.
 The transposition of the real matrice $Y=[Y_{i,j}]_{1\leq i \leq I, 1\leq j \leq J}$, $(I,J)\in\mathbb{N}^2$, will be $Y^T$ and $\|Y\|_{\text{max}}=\max_{1\leq i \leq I,1\leq j \leq J}|Y_{i,j}|$.
@@
 RL consists in finding an optimal policy for the reward function $R$. Notice that the state space may be too large for an exact representation of the value function (which calls for approximate
-representation), that the model (P and R) may be unknown (the only information being provided through rewarded transition sampled according
-to some behavioral policy), that learning can occurs in an online or offline setting, and so on. There exists good books on the subject [\cite{bertsekas2001dynamic},\cite{sutton1998reinforcement}].\\
-In the classical IRL paradigm \cite{ng2000algorithms}, an MDP without reward $\{S,A,P,\gamma\}$ and a policy $\pi_E$ called expert-policy are given and the problem is to find
-a reward $R^*$ for which the policy $\pi_E$ will be optimal.
-However this problem is clearly ill-posed in the sense that there is no uniqueness of the reward $R^*$ : many rewards functions are equivalent in the sense that they have the same optimal deterministic policies \cite{ng1999policyreward}, moreover the trivial zero-reward is a solution for any deterministic policy $\pi_E$ as it is shown in \cite{ng2000algorithms}.
-In the literature, some solutions are proposed in order to respond to the ill-posed nature of the problem (\cite{ng2000algorithms},\cite{ziebart2008maximum},\cite{boularias2011relative}).
-In this paper, we assume that the solely available information is provided by transitions sampled according to the expert policy $\pi_E$:
+representation), that the model ($P$ and $R$) may be unknown (the only information being provided through rewarded transitions sampled according
+to some behavioral policy), that learning can occurs in an online or off-line setting, and so on. There exists books on the subject \cite{bertsekas2001dynamic},\cite{sutton1998reinforcement}.\\
+In the classical IRL paradigm \cite{ng2000algorithms}, an MDP without reward $M\backslash R =\{S,A,P,\gamma\}$ and a policy $\pi_E$ called expert-policy are given and the problem is to find
+a reward $R^*$ for which the policy $\pi_E$ is optimal.
+However this problem is clearly ill-posed in the sense that there is not uniqueness of the reward $R^*$ : many rewards functions are equivalent in the sense that they have the same optimal deterministic policies \cite{ng1999policyreward}, moreover the trivial zero-reward is a solution for any deterministic policy $\pi_E$ as it is shown in~\cite{ng2000algorithms}.
+In the literature, some solutions are proposed in order to respond to the ill-posed nature of the problem \cite{ng2000algorithms},\cite{ziebart2008maximum},\cite{boularias2011relative}.
+In our experiments, see Section \ref{section: experiments}, we assume that the solely available information is provided by transitions sampled according to the dynamics of the environnement under $\pi_E$:
 \begin{equation}
-\label{equation:data}
-\{(s_i,a_i=\pi_E(s_i)),s_i')\}_{1\leq i \leq D}, D\in\mathbb{N}^*.
+\{(s_i,a_i=\pi_E(s_i)),s_i')\}_{1\leq i \leq D}, D\in\mathbb{N}^*,
 \end{equation}
+where $s_i'$ is sampled according to the distribution $P_{s_i,a_i}$.
 The reward function is obviously unknown, but this assumption means that the dynamics ($P$) is only known through transitions $(s_i, a_i, s_i')$ and that the
 policy $\pi_E$ is only known through state-action pairs $(s_i, a_i)$.\\
-Finally let's recall some basic but important results of finite-Markov chain theory.
+For a given deterministic policy $\pi$, a component $P_{s_i,\pi(s_i)}(s_j)$ of the matrix $P_\pi$ represents the probability to transit from $s_i$ to $s_j$ under the policy $\pi$.
+So $P_\pi$ can be seen as a transition matrix of a finite Markov-chain on the finite state space $S$.
+Thus, let us recall some basic but important results of finite-Markov chain theory.
 \begin{definition}
@@
 \end{definition}
-\begin{theorem}
+\begin{theorem}[\cite{baldi2000martingales}]
 Let $Y$  be a stochastic square matrix of size $M\in\mathbb{N}^*$.
@@
 \end{definition}
-\begin{theorem}
+\begin{theorem}[\cite{baldi2000martingales}]
 \label{theoreme : mixing exponential}
@@
 \subsection{The general idea}
-The available data being depicted in \eqref{equation:data}, a first problem would be to generalize the expert policy $\pi_E$ outside state examples provided in this dataset.
+The available data, provided by the expert, is being depicted by:
+\begin{equation}
+\label{equation:data}
+D_C=\{(s_{C,i},a_{C,i}=\pi_E(s_{C,i})),s'_{C,i})\}_{1\leq i \leq D}, D\in\mathbb{N}^*,
+\end{equation}
+where $s'_{C,i}$ is sampled according to the distribution $P_{s_{C,i},a_{C,i}}$.
+A first problem would be to generalize the expert policy $\pi_E$ outside state examples provided in this dataset.
 A first approach, recently used in \cite{melo2010learning} and legitimated for finite-horizon problems in \cite{syed2010reduction}, is to frame this problem as learning a multi-classifier where each action $a \in A$ will be a class.
-Indeed, trained on the training set $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$, many classifiers are able to output a score function $q$ from $S\times A$ to $\mathbb{R}$.
+Indeed, trained on the training set $\{s_{C,i},a_{C,i}=\pi_E(s_{C,i})\}_{1\leq i \leq D}$, many classifiers are able to output a score function $q$ from $S\times A$ to $\mathbb{R}$.
 The final decision rule $\pi_C$ which is a function from $S$ to $\mathbb{R}$ and thus can be seen as a deterministic policy is constructed in order to satisfy:
@@
 \end{equation}
-Now, one can notice the similarity between \eqref{equation:Qoptimal} and \eqref{equation:Coptimal}.
-The score-function $q$ can then be interpreted as an action-value function and it is easy and natural to construct a reward function $R_C$ for which the policy $\pi_C$ is optimal as it will be proven in the section \ref{section: Analysis}.
+Now, one can notice the similarity between equation~\eqref{equation:Qoptimal} and equation~\eqref{equation:Coptimal}.
+The score-function $q$ can then be interpreted as an action-value function and it is easy and natural to construct a reward function $R^C$ for which the policy $\pi_C$ is optimal as it will be proven in the section~\ref{section: Analysis}.
 Let define $R^C$ as:
@@
 \end{equation}
-However, the model being unknown (contrary to $q$ and $\pi_C$) it is not possible to compute $R^C$ directly. However, it can be estimated from the available transitions (see
-\eqref{equation:data}). Let us write:
+However, the model being unknown (contrary to $q$ and $\pi_C$) it is not possible to compute $R^C$ directly. But, it can be estimated from transitions of a regression data set:
 \begin{equation}
-\label{r.def}
-\hat{r}_i=q(s_i,a_i)-\gamma q(s_i',\pi_C(s_i')), \forall i\in \{1,\dots,D\}.
+D_R=\{(s_{R,i},a_{R,i},s'_{R,i})\}_{1\leq i \leq D'}, D'\in\mathbb{N}^*,
 \end{equation}
-This is an unbiased estimate of $R_C(s_i,a_i)$, indeed:
+where $s'_{R,i}$ is sampled under the probability $P_{s_{R,i},a_{R,i}}$.
+Let us write:
 \begin{equation}
-R_C(s_i,a_i)=\sum_{s'\in S}p(s_i,a_i,s')[q(s_i,a_i)-\gamma q(s',\pi_C(s'))], \forall i\in \{1,\dots,D\}.
+\label{ri.def}
+\hat{r}_i=q(s_{R,i},a_{R,i})-\gamma q(s'_{R,i},\pi_C(s'_{R,i})), \forall i\in \{1,\dots,D'\}.
 \end{equation}
-Therefore, it is possible to build an estimate $\hat{R}_C$ of $R_C$, using a regressor trained on the dataset:
+This is an unbiased estimate of $R^C(s_{R,i},a_{R,i})$, indeed:
 \begin{equation}
-\{(s_i,a_i,s'_i,\hat{r}_i)\}_{1\leq i \leq D}.
+R^C(s_{R,i},a_{R,i})=\sum_{s'\in S}p(s_{R,i},a_{R,i},s')[q(s_{R,i},a_{R,i})-\gamma q(s',\pi_C(s'))], \forall i\in \{1,\dots,D'\}.
 \end{equation}
-Finally, we see $\hat{R}_C$ as an approximation of $R_C$ and $\hat{\pi}_C$ will be an optimal policy for the reward $\hat{R}_C$.
-In order to verify that the reward function $\hat{R}_C$ is a good candidate to resolve the IRL problem, it will be proven in the section \ref{section: Analysis} that the policies $\pi_E$ and $\hat{\pi}_C$ are quite similar (confer lemma \ref{lemma : results}).
+Therefore, it is possible to build an estimate $\hat{R}^C$, which is a function from $S\times A$ to $\mathbb{R}$, of $R^C$ using a regressor trained on the dataset:
+\begin{equation}
+\{(s_{R,i},a_{R,i},\hat{r}_i)\}_{1\leq i \leq D'}.
+\end{equation}
+Finally, we see $\hat{R}^C$ as an approximation of $R^C$ and $\hat{\pi}_C$ is defined as an optimal policy for the reward $\hat{R}^C$.
+In order to verify that the reward function $\hat{R}^C$ is a good candidate to resolve the IRL problem, it will be proven in Section \ref{section: Analysis} that the policy $\pi_E$ is near-optimal for the reward $\hat{R}^C$ (confer theorem \ref{theorem : results}).
+\subsection{A particular case}
+An interesting case is  when $D_C=D_R$ which is the one chosen in Section \ref{section: experiments}.
+In this case, the solely available data are the expert data depicted by the equation \ref{equation:data}.
+The approximation function $\hat{R}^C$ obtained thanks to the regression is only accurate on the set $(s,\pi_E(s))_{\{s\in S\}}$ because of the
+regression training set.
 \section{Analysis} \label{section: Analysis}
@@
 where $E_{s \sim \mu_\pi}$ means that $s$ is distributed over the distribution $\mu_\pi$.
-All the results provided in the next theorem use the expectation $E_{s \sim \mu_E}$ which is the canonical expectation
-to consider when one wants to prove a result over the state space $S$ related to the policy of the expert $\pi_E$.\\
+All the results provided in the next theorem use the expectation $E_{s \sim \mu_E}$ which is the canonical expectation to consider when one wants to prove a result over the state space $S$ related to the policy of the expert $\pi_E$.\\
 Before giving an upper bound to $\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})=E_{s \sim \mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$, we define $\epsilon_C\in\mathbb{R}_+$ called the classification error and the function $\epsilon_R$ from $S\times A$ to $\mathbb{R}$ called the regression error such that:
@@
 \begin{remark}
-One can notice that the error classification is defined thanks to the expectation $E_{s \sim \mu_E}$. In the classical framework of classification the data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$
-are generated independently according to a distribution $\mu_{\text{Data}}$ over $S$ and hence the classical classification error must be:
+One can notice that the classification error is defined thanks to the expectation $E_{s \sim \mu_E}$.
+In the classical framework of classification, the data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$ are generated independently according to a distribution $\mu_{\text{Data}}$ over $S$ and hence the classical classification error must be:
 \begin{equation}
@@
 \label{theorem : results}
-Let $\{S,A,P,\gamma\}$ be a finite MDP without reward and $\pi_E$ an expert policy. The only available data is depicted in \eqref{equation:data}.
-The notations $q$, $\pi_C$, $\hat{\pi}_C$, $\hat{R}^C$ are introduced in the section \ref{section: Cascading}.
+Let $\{S,A,P,\gamma\}$ be a finite MDP without reward and $\pi_E$ an expert policy.
+The notations $q$, $\pi_C$, $\hat{\pi}_C$, $\hat{R}^C$ are introduced in the Section \ref{section: Cascading}.
 If $P_{\pi_E}$ is reducible, then $\mu_E$ is the stationary distribution of $\pi_E$ and:
@@
 \item $\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}$.
-\item $\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\|\epsilon^E_R\|_{\infty}}{1-\gamma}$.
+\item $\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\mu_E^T\epsilon^E_R}{1-\gamma}$.
 \end{enumerate}
@@
 Let us suppose that the classification and the regression steps are perfect in the sense that $\epsilon_R(s,a)=0,\forall (s,a)\in S\times A$ and $\epsilon_C=0$.
-Then we obviously have, thanks to the theorem \ref{theorem : results}, that $\pi_E=\pi_C$ is optimal for $\hat{R}_C=R_C$.
+Then we obviously have, thanks to the theorem \ref{theorem : results}, that $\pi_E=\pi_C$ is optimal for $\hat{R}^C=R^C$.
 Thus the method is able to provide a non-trivial reward function for which the policy $\pi_E$ is optimal.
-Moreover if the classification step and the regression step are not perfect, the theorem \ref{theorem : results} shows, that our approach is able to provide a non trivial-reward $\hat{R}_C$ for which the policy $\pi_E$ is near-optimal in the sense that:
+Moreover if the classification step and the regression step are not perfect, the theorem \ref{theorem : results} shows, that our approach is able to provide a non trivial-reward $\hat{R}^C$ for which the policy $\pi_E$ is near-optimal in the sense that:
 \begin{equation}
-E_{s \sim \mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq B[\epsilon_C+\|\epsilon_R\|_{\text{max}}], B\in\mathbb{R}_+,
+E_{s \sim \mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq B[\epsilon_C+\|\epsilon^C_R\|_{\infty}+\mu_E^T\epsilon^E_R], B\in\mathbb{R}_+.
 \end{equation}
-which means that the value functions of $\pi_E$ and $\hat{\pi}_C$ are not so different in the most visited states by the policy $\pi_E$.
-
+\begin{remark}
+It is important to be clear about this result. If the only available data are provided by the equation \eqref{equation:data}, it is possible to control
+$\epsilon_C$ and $\mu_E^T\epsilon^E_R$ because these errors depend only on the expert policy $\pi_E$. However it is not possible to control the error $\|\epsilon^C_R\|_{\infty}$
+because it depends on the policy $\hat{\pi}_C$ which can be different than the expert policy and hence do not appear in the available data \eqref{equation:data}. However
+it will be possible to obtain a control on the term $\|\epsilon^C_R\|_{\infty}$ if the data used for the regression are:
+\begin{equation}
+D_R=\{(s_{R,i},a_{R,i},s'_{R,i})\}_{1\leq i \leq D'},
+\end{equation}
+where $(s_{R,i},a_{R,i})$ are uniformly chosen on the set $S\times A$ or sampled from other policies than the expert. So, theoretically an easy way to be sure
+to control the error $\|\epsilon^C_R\|_{\infty}$ is to be able to give a data set for the regression which is sampled from the expert policy and other policies (and more particularly $\hat{\pi}_C$).
+But we give examples, see Section \ref{section: experiments}, where the regression data set given by the equation \eqref{equation:data} is sufficient to obtain good results.
+A possible argument to explain the fact that classification-regression still works when $D_C=D_R$, is that $\hat{\pi}_C$ must not be so different than $\pi_E$ but we did not manage to control
+the term $\|\epsilon^C_R\|_{\infty}$ when $D_C=D_R$.
+\end{remark}
 \subsection{Proofs}
@@
 \begin{proof}[Lemma \ref{lemma: calculs V}]
-Here, we use \eqref{equation: stationarity}.
+Here, we use \eqref{equation: stationarity}
 \begin{align}
@@
 \end{proof}
-
-%\begin{remark}
-%Let use the notations of the lemma \ref{lemma: calculs V}. For each $s\in S$, we define the average reward $\rho^\pi_R(s)$ such that :
-%\begin{equation}
-%\rho^\pi_R(s)=\lim_{n\rightarrow+\infty}\frac{1}{n}E^\pi_s[\sum_{t=0}^{n-1}\gamma^tR(s_t,\pi(s_t))]
-%\end{equation}
-%If $P_\pi$ is reducible then $\rho^\pi_R(s)$ does not depend on $s$ and:
-%\begin{equation}
-%\rho_R^\pi(s)=\rho_R^\pi=\mu_\pi^TR_\pi,\forall s\in S.
-%\end{equation}
-%The proof of that result is essentially an application of the ergodic theorem (see \cite{puterman1994markov}).
-%So, if $P_\pi$ is reducible:
-%\begin{equation}
-%\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\rho_R^\pi.
-%\end{equation}
-%There is a clear link between the average reward and the discounted reward.
-%\end{remark}
-
-\begin{proof}[Theorem \ref{lemma : results}]
-In order to prove the tree results of the lemma \ref{lemma : results}, it is interesting to introduce the function $R_E$ from $S\times A$ to $\mathbb{R}$ such that:
+\begin{proof}[Theorem \ref{theorem : results}]
+In order to prove the three results of the theorem \ref{theorem : results}, let us introduce the function $R_E$ from $S\times A$ to $\mathbb{R}$ such that:
 \begin{equation}
@@
 \end{equation}
-$\pi_C$ is optimal for the reward $R_C$.
+$\pi_C$ is optimal for the reward $R^C$.
 Now let us prove that:
@@
 \\
-&\|V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}\|_{+\infty}\leq\frac{\|\epsilon^E_R\|_{\infty}}{1-\gamma},
+&\mu_E^T(V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})\leq\frac{\mu_E^T\epsilon^E_R}{1-\gamma},
 \end{align}
@@
 \begin{align}
-V^{\hat{\pi}_C}(s)_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}(s)&=E^{\hat{\pi}_C}_s[\sum_{t=0}^{+\infty}\gamma^t(\hat{R}^C(s_t,\hat{\pi}_C(s_t))-R^C(s_t,\hat{\pi}_C(s_t)))], \forall s\in S,
+V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\hat{\pi}_C}_{R^C}(s)&=E^{\hat{\pi}_C}_s[\sum_{t=0}^{+\infty}\gamma^t(\hat{R}^C(s_t,\hat{\pi}_C(s_t))-R^C(s_t,\hat{\pi}_C(s_t)))], \forall s\in S,
 \\
 &\leq\frac{\|\epsilon^C_R\|_{\infty}}{1-\gamma}, \forall s\in S.
+\end{align}
+And, thanks to the lemma \ref{lemma: calculs V}:
+\begin{align}
+\mu_E^T(V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})&=\mu_E^T(V^{\pi_E}_{\epsilon_R}),
 \\
-V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}(s)&=E^{\pi_E}_s[\sum_{t=0}^{+\infty}\gamma^t(R^C(s_t,\pi_E(s_t))-\hat{R}^C(s_t,\pi_E(s_t)))], \forall s\in S,
-\\
-&\leq\frac{\|\epsilon^E_R\|_{\infty}}{1-\gamma}, \forall s\in S.
+&\leq\frac{\mu_E^T\epsilon^E_R}{1-\gamma}.
 \end{align}
@@
 \label{equation: cas3-1}
-\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}+V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})\leq\frac{\|\epsilon^E_R\|_{\infty}+\|\epsilon^C_R\|_{\infty}}{1-\gamma}.
+\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}+V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})\leq\frac{\mu_E^T\epsilon^E_R+\|\epsilon^C_R\|_{\infty}}{1-\gamma}.
 \end{equation}
@@
 Finally by regrouping the results in \eqref{equation: cas3-1} and \eqref{equation: cas3-2}:
-\begin{align}
-\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}&\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\|\epsilon^E_R\|_{\infty}}{1-\gamma},
-\\
-&\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+2\|\epsilon_R\|_{\text{max}}}{1-\gamma}.
-\end{align}
-
+\begin{equation}
+\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^C_R\|_{\infty}+\mu_E^T\epsilon^E_R}{1-\gamma}.
+\end{equation}
 \end{proof}
@@
 \end{eqnarray}
-with $r_i$ as in equation \ref{r.def}. We then get :
+with $\hat r_i$ as in equation \ref{ri.def}. We then get :
 \begin{equation}
@@
 \end{equation}
-so that $\hat R_C(s) = \theta^T \psi(s)$.\\
+so that $\hat R^C(s) = \theta^T \psi(s)$.\\
 \subsection{Experimental setting}
@@
 This algorithm can be instanciated to work in the same dire conditions as our algorithm, that is to say using only samples from the expert.
-To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ (\cite{klein2011batch}) and LSPI (\cite{lagoudakis2003least}).
+To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ \cite{klein2011batch} and LSPI \cite{lagoudakis2003least}.
 When this does not work, we fall back to a setting where everything is known, that is to say the feature expectations are computed by a dynamic programming algorithm that make use of the transition probabilities.\\

### 2012-05-21 piot2012cascading/main_bilal.tex

--- 2012-05-21 piot2012cascading/main_KLEIN_v3.tex
@@
-\documentclass{article} % For LaTeX2e
-\usepackage{nips12submit_e,times}
+% This is LLNCS.DEM the demonstration file of
+% the LaTeX macro package from Springer-Verlag
+% for Lecture Notes in Computer Science,
+% version 2.4 for LaTeX2e as of 16. April 2010
+%
+\documentclass{llncs}
+%
 \usepackage{makeidx}  % allows for indexgeneration
@@
 \usepackage{subfigure}
-\usepackage{tabularx}
 \usepackage{mathtools}
@@
 \usepackage{amssymb}
-\usepackage{amsthm}
-\newtheorem{definition}{Definition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{remark}{Remark}
 \usepackage{dsfont}
@@
 \newcommand{\sgn}{\operatorname{sgn}}
-\title{Learning a reward function from demonstrations: a cascaded supervised
-learning approach}
-\author{Edouard Klein$^{1,2}$\\
- 1. ABC Team\\
- LORIA-CNRS, France.
-\And Bilal Piot$^{2}$\\
- 2. Supélec-Metz Campus\\
- MaLIS Research group, France\\
-\And Matthieu Geist$^1$\\
-\texttt{prenom.nom@supelec.fr}\\
-\And Olivier Pietquin$^{2,3}$\\
-3. UMI 2958 CNRS\\
-GeorgiaTech, France
-}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to \LaTeX{} to determine where to break
-% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
-% puts 3 of 4 authors names on the first line, and the last on the second
-% line, try using \AND instead of \And before the third author name.
-
-\newcommand{\fix}{\marginpar{FIX}}
-\newcommand{\new}{\marginpar{NEW}}
-
 \begin{document}
 
+\title{Titre}
+\titlerunning{Titre}  % abbreviated title (for running head)
+\author{Boul\inst{1} \and Bil\inst{2}
+Boul \and Bil}
+\authorrunning{} % abbreviated author list (for running head)
+%
+\institute{SUPELEC,\\
+\email{SUPELEC},\\ WWW home page:
+\texttt{SUPELEC}
+\and
+\\}
 \maketitle              % typeset the title of the contribution
@@
 
-%Other notations will be needed in order to prove that the cascading approach is legitimate.
-Let's define the column-vector $\mu^{t,\pi,s}=[\mu^{t,\pi,s}_i]_{1\leq i \leq N}$ where $\mu^{t,\pi,s}_i$ is the probability of being in the state $s_i\in S$ at time $t\in\mathbb{N}$ when the start-state is $s$ and the policy followed is $\pi$ :
-%\begin{equation}
-%\mu^{t,\pi,s}=(\Delta_s^T(P_\pi)^t)^T,
-%\end{equation}
-%where $\Delta_s$ is the column-vector: $\Delta_s={\delta(s_i,s)}_{1\leq i \leq N}$ with $\delta(.,.)$ the Kronecker symbol.
-For a fixed policy $\pi$, the matrix $P_\pi$ is a stochastic matrix of finite dimension and if it is ergodic and recurrent positive then there exists a unique distribution probability represented by the column vector $\mu^\pi=[\mu^{\pi}_i]_{1\leq i \leq N}$ such that :
-%\begin{equation}
-%\mu_\pi=(\mu_\pi^T P_\pi)^T.
-%\end{equation}
-%In that case $\mu^\pi$ is called the stationary distribution of the policy $\pi$ and :
-%\begin{equation}
-%\forall s\in S \lim_{t\rightarrow+\infty}\mu_{t,\pi,s}=\mu_\pi.
-%\end{equation}
-%Thus in the particular case of ergodicty and recurrent positivity we can easily express $\mu_\pi^TV^\pi_R$ :
-%\begin{align}
-%\mu_\pi^TV^\pi_R&=\mu_\pi^T(R_\pi+\gamma P_\pi V^\pi_R)=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TP_\pi V^\pi_R,
-%\\
-%&=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\mu_\pi^TR_\pi.
-%\end{align}
 \section{Cascading Classification and Regression for IRL}
@@
 \begin{equation}
-\label{ri.def}
 \hat{r}_i=q(s_{R,i},a_{R,i})-\gamma q(s'_{R,i},\pi_C(s'_{R,i})), \forall i\in \{1,\dots,D'\}.
@@
 \subsection{Proofs}
-
 \begin{proof}[Lemma \ref{lemma: calculs V}]
@@
 \end{proof}
-
 \section{Experiments}
 \label{section: experiments}
-\subsection{Instantiation of the algorithm}
-As hinted to in Section \ref{section: Cascading}, our algorithm can be instanciated in many ways, as any off-the-shelf classifier using a score function can be used in the first step, and any regression method can be used in the second step.
-This freedom of choice leaves room for problem dependant tweaks if needed.\\
 
-In our tests, we used a subgradient-descent based classifier and a simple regularized least square regression method. The classifier tries to minimize the risk function $R_N(q)$ :
-\begin{equation}
-  R_N(q) = {1\over N} \sum_{i=1}^N\left(\max_{a}(q(s_i,a) + l(s_i,a)) - q(s_i,a_i) \right).
-\end{equation}
-where $l$ is a user defined function whose role is to quantify the gap between the best and second to best actions.
-In all our tests we used the naive solution $l(s_i,a) = 1$ if $a\neq a_i$ and $l(s_i,a_i) = 0$.
-To minimize the risk, we introduce the parametrization
-\begin{equation}
-  q(s,a) = \omega^T\phi(s,a).
-\end{equation}
-The risk now becomes :
-\begin{equation}
-  R_N(\omega) = {1\over N} \sum_{i=1}^N\left(\max_{a}(\omega^T\phi(s_i,a) + l(s_i,a)) - \omega^T\phi(s_i,a_i) \right).
-\end{equation}
-
-Because of the $\max$ in the expression, we can not use a simple gradiend descent on $\omega$ to minimize the risk.
-We have to make use of the subgradient.
-Let $a^* = \max_a\omega^T\phi(s,a)$, we get the following update rule :
-\begin{equation}
-  \omega_{t+1} = \omega_t -\alpha_t{1\over N}\sum_{i=1}^N\left(\phi(s_i,a^*) - \phi(s_i,a_i)\right).
-\end{equation}
-The regressor is a classical least square fitting. Given some features $\psi : S\rightarrow \mathbb{R}^p$, we define :
-\begin{eqnarray}
-\Psi &= \begin{pmatrix}\psi(s_1)^T\\\vdots\\\psi(s_N)^T\end{pmatrix}\\
-\textrm{and}
-\hat R &= \begin{pmatrix}\hat r_1 \\\vdots\\\hat r_N\end{pmatrix}\\
-\end{eqnarray}
-with $\hat r_i$ as in equation \ref{ri.def}. We then get :
-\begin{equation}
-\theta = (\Psi^T\Psi + \lambda Id)^{-1}\Psi^T\hat R,
-\end{equation}
-so that $\hat R^C(s) = \theta^T \psi(s)$.\\
-\subsection{Experimental setting}
-For both problems we test our approach with, we compare it to two baselines : the mean performance of an agent trained on a random reward and the mean performance of the policy output by the algorithm of \cite{abbeel2004apprenticeship}.
-This algorithm can be instanciated to work in the same dire conditions as our algorithm, that is to say using only samples from the expert.
-To do this, we estimate the feature expectations and solve the MDP in a batch, off policy manner thanks respectively to to LSTD$\mu$ \cite{klein2011batch} and LSPI \cite{lagoudakis2003least}.
-When this does not work, we fall back to a setting where everything is known, that is to say the feature expectations are computed by a dynamic programming algorithm that make use of the transition probabilities.\\
-The performance of the expert and the agents is computed with respect to the reward the expert was trained with. We use as a performance metric the mean value over the state space.
-\subsection{Gridworld}
-We first tested our approach on a simple $5\times 5$ gridworld.
-The expert and the agent can choose between 4 actions (down, left right, or up) that sends the player in the corresponding neighbooring cell.
-Trying to go off the grid will result in the player staying in its position.
-Each action has a probability $0.3$ of failing, in which case another randomly chosen different action is executed.
-The expert begins in the lower left corner and is trained to go to the higher right corner by an exact dynamic programming algorithm.
-The original reward is null everywhere but in the higher right corner where it is $1$.\\
-The results are shown in two figures.
-Figure \ref{Fig1.fig} show the baselines we compare ourselves with.
-Pure classification and the algorithm of \cite{abbeel2004apprenticeship} are on par with one another, both far better than training an agent on a random reward ; they both converge to a value close to the expert's given enough samples.
-Figure \ref{Fig2.fig} shows results of our algorithm together with the mean performance of both forementionned approaches (variance and min/max were dropped for legibility's sake).
-It shows the soundness of our cascading approach : when only a few samples are available, we provide a reward that lead to a better policy.
-When samples are widely available all approaches converge to the same value.
-  \begin{figure}
-  \begin{tabular}{ccc}
-  \subfigure[State of the art approaches on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig1"}\label{Fig1.fig}}&\hspace{.05\linewidth}&
-  \subfigure[Our new approach on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig2"}\label{Fig2.fig}}\\
-  \subfigure[All approaches on the Highway driving simulator]{\includegraphics[width=.4\linewidth]{"Fig3"}\label{Fig3.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Legend"}
-  \end{tabular}
-  \caption{Results on the GridWorld and Highway problems. Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
-  \end{figure}
-
-\subsection{Highway driving simulator}
-There are problems for which pure classification just does not cut it, as the number of samples that would be needed to imitate the expert's control is dauntingly large.
-To illustrate this, inspired from a problem seen in \cite{syed2008apprenticeship,syed2008game}, we coded a driving simulator.
-The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
-The first thing to note about this setting is that the direct problem (RL) problem is not trivial, as LSPI is unable to solve it using the natural features we first gave it.
-These feature stemmed from a discretization of the state-action space, the dimension of the features being the cardinality of this space.
-With each state-action pair $s,a$ we associate an index $i$ so that the $i$-th coordinate of the feature is $1$ and every other is $0$.
-We tried to train an agent on the reward used by the expert (which was trained using an exact dynamic programming approach), but on top of being extremely slow to run (the LSTD$Q$ subroutine implies a matrix inversion) it does not converge.\\
-As LSPI is used in our instanciation of \cite{abbeel2004apprenticeship}'s algorithm to solve the MDP, it is impossible to use the latter with the natural features.
-Although it proved possible to engineer features so that LSPI converges thus making the state of the art IRL algorithm work, this is outside the scope of this document.
-As a baseline we used an instanciation of \cite{abbeel2004apprenticeship}'s algorithm that imake use of all the information about the problem.
-It is displayed as an horizontal line in figure \ref{Fig3.fig} because it does not use samples from the expert, instead it uses the transition probabilities and the expert's policy to compute the exact feature expectation of the expert and every intermediate policy.\\
-As predicted, pure classification is not sample efficient enough to provide a good control whereas our approach is able to generalize from very little data and exhibits performance on par with the \emph{fully-informed} state of the art algorithm when given as few as 100 samples.\\
-Furthermore, the state of the art algorihtm make use of intermediate policies, which is computationally costly particularily when the state space gets big.
-Our approach computational cost scales with the number of samples we are given, which is as we demonstrated a small number.
-Experimentally, when producing the data for our graph, it took less time to run the cascading approach 50 times for 6 abcissa than to run \cite{abbeel2004apprenticeship}'s algorithm once.\\
 \section{Related Work}
 \label{section: related work}
-All the IRL approaches summed up in \cite{neu2009training}, namely the historical approach of \cite{ng2000algorithms} as well as for example \cite{abbeel2004apprenticeship,syed2008apprenticeship,syed2008game, ziebart2008maximum} share the same pivotal argument of trying to minimize some distance between the feature expectation of the expert and the feature expectation of another policy through successive updates of this intermediate policy.
-The main problem with these algorithm is that the update on the policy is done through the update of a reward function.
-This means that the forward problem of RL must be repeatedly solved ; this is often non trivial, and almost always computationally costly.\\
 
-Our algorithm avoid this pitfall, specifically, one needs more information to solve the forward problem than to just compute the feature expectation of the expert as our algorithm does.
-Even a recent contribution like \cite{boularias2011relative} still needs samples that cover a large portion of the state space although it does not explicitely require the ability to solve the forward problem.\\
+\section{Conclusion}
 
-A recurring argument is to see the expert not only as optimal but as optimal by a certain margin.
-The action chosen by the expert should have a quality strictly greater than any other action.
-Structured planning as done in \cite{ratliff2006maximum,ratliff2007imitation, ratliff2007boosting, kolter2008hierarchical} allow some influence on the margin.
-Through the $l$ function of our classifier, we share the same feature.\\
-
-Finally, \cite{ratliff2007imitation} cast the imitation learnig problem as a multi class classification problem.
-As seen in the experimental results, the power of generalization of our approach is greater than pure classification thanks to the regression step.
-This allows us to be more sample efficient than any other published approach known to us.
-\section{Conclusion}
-We provide a new solution to the IRL problem that avoid most pitfalls of existing approaches.
-It outputs a reward function, the most succinct and transferrable description of a task without having to repeatedly solve the forward RL problem.
-The only information our algorithm need is provided by samples from the expert, which is to the best of our knowledge a feature that can not be found in any other published algorithm.
-Even recent other approaches need randomly drawn samples.
-We provided theoretical results showing the soundness of our approach as well as experimental benchmarks comparing it to a widely known state of the art algorithm.
-In a near furure, experiments on real world robotics problems will take place.
-We hope to benchmark our algorithm against the few other approaches shown to work in such a difficult context.
-
-\bibliographystyle{alpha}
+\bibliographystyle{splncs}
 \bibliography{Biblio}

### 2012-05-23 klein2012/headertikz.tex
Déplacement des slides dans le bon dossier, completion des slides et correction des notations

\usepackage{pgf}
\usepackage{pgfcore}
\usepackage{pgfbaseimage}
\usepackage{pgfbaselayers}
\usepackage{pgfbasepatterns}
\usepackage{pgfbaseplot}
\usepackage{pgfbaseshapes}
\usepackage{pgfbasesnakes}
\usepackage{tikz}

\tikzset{normal/.style={ 
% The shape: 
rectangle,minimum size=6mm,rounded corners=3mm, 
% The rest 
very thick,draw=black!50, 
top color=white,bottom color=black!20, 
font=\ttfamily}}
\tikzset{arduino/.style={ 
% The shape: 
rectangle, 
% The size: 
minimum size=6mm, 
% The border: 
very thick, 
draw=blue!50!black!50, % 50% red and 50% black, 
% and that mixed with 50% white 
% The filling: 
top color=white, % a shading that is white at the top... 
bottom color=blue!50!black!20, % and something else at the bottom 
% Font 
font=\itshape 
}}
\tikzset{enceinte/.style={ 
% The shape: 
rectangle, 
% The size: 
minimum size=6mm, 
% The border: 
very thick, 
draw=yellow!90!black!50, % 50% red and 50% black, 
% and that mixed with 50% white 
% The filling: 
top color=white, % a shading that is white at the top... 
bottom color=yellow!90!black!20, % and something else at the bottom 
% Font 
font=\itshape 
}}

\tikzset{ordi/.style={ 
% The shape: 
rectangle, 
% The size: 
minimum size=6mm, 
% The border: 
very thick, 
draw=green!90!black!50, % 50% red and 50% black, 
% and that mixed with 50% white 
% The filling: 
top color=white, % a shading that is white at the top... 
bottom color=green!90!black!20, % and something else at the bottom 
% Font 
font=\itshape 
}}

### 2012-06-27 klein2012/main.tex
Last changes before EWRL12

--- 2012-04-12 klein2012/main.tex
@@
 \documentclass[tablecaption=bottom]{jmlr}
+%\usepackage{marvosym}
 
@@
  % Two authors with the same address
-  \author{\Name{Paper under review at ICML}\thanks{No author name provided so as not to breach ICML's double blind process}}
+
+\author{\Name{{\'E}douard Klein$^{\dag\ddag}$}, \Name{Bilal Piot$^{\dag\textrm{*}}$}, \Name{Matthieu Geist$^\dag$}, \Name{Olivier Pietquin$^{\dag\textrm{*}}$}\\
+\Email{firstname.lastname@supelec.fr}\\
+\addr $\dag$ {\'E}quipe IMS/MaLIS (Sup{\'e}lec), $\ddag$ {\'E}quipe ABC UMR 7503 (Loria-CNRS), * UMI 2958 (GeorgiaTech-CNRS).
+}
 

### 2012-09-03 Rapport2A/main.tex
Rapport d'avancement de these pour la reinscription

% Created 2012-06-27 Mer 16:06
\documentclass[11pt]{article}
\usepackage[plain,cm]{fullpage}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\usepackage{amsmath}
\usepackage[frenchb]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{bibtopic}
\providecommand{\alert}[1]{\textbf{#1}}

\title{Rapport Intermédiaire de Thèse en Informatique à l'Université de Lorraine (\'Ecole doctorale IAEM)\\
Découverte automatique d'attributs pour l'imitation}
\author{Edouard KLEIN}
\date{\today}
\hypersetup{
  pdfkeywords={},
  pdfsubject={},
  pdfcreator={Emacs Org-mode version 7.8.11}}

\begin{document}

\maketitle
\section*{Sujet de thèse}
L'adoption de systèmes automatisés autonomes peut réduire les coûts et les risques, cependant la complexité de la mise au point de tels systèmes gêne leur adoption généralisée. Des approches permettant à ces systèmes de prendre en charge des environnements de plus en plus difficiles font apparaître de nouveaux défis. L'un de ceux-ci est la définition de la consigne. A mesure que les tâches confiées aux systèmes automatiques se complexifient, la définition de ces dernières devient moins aisée.\\

  L'\emph{apprentissage par renforcement inverse} (ARI), cadre dans lequel s'inscrit cette thèse, a pour objet de contourner les difficultés évoquées : de la même manière que les jeunes gens n'apprennent pas à conduire en lisant le manuel de leur auto mais en observant leurs parents et leurs moniteurs de conduite puis en se mettant derrière le volant, nous comptons apprendre la tâche à effectuer en observant un \emph{expert} la réaliser. Cette démarche exploite la capacité humaine à résoudre intuitivement et rapidement des conflits qu'il serait difficile d'analyser sur papier. Pour reprendre l'exemple précédent, un automobiliste saura après un rapide coup d'œil dans son rétroviseur s'il vaut mieux qu'il pile ou qu'il déboîte en urgence et effectuera sa manœuvre dans la foulée.\\

  Notre but est de dériver, du comportement d'un expert effectuant une tâche, une description de cette tâche sous la forme d'une fonction de récompense, ce qui permet ensuite l'utilisation des outils d'apprentissage par renforcement pour apprendre cette tâche à un agent. Cela ouvrirait le champ d'application de l'apprentissage par renforcement à des tâches encore inaccessibles car trop complexes pour être ``expliquées''.\\
  L'intitulé de la thèse, \emph{Découverte automatique d'attributs pour l'imitation}, isole une partie du problème : il s'agit d'extraire de la description de la suite d'\emph{états} traversés par l'expert les informations pertinentes à l'expression de la récompense.\\

\section*{Contributions au domaine}
\label{sec-3}

  Il faut attendre 2000 pour qu'une publication pose formellement le problème. La contribution centrale intervient en 2004 et introduit la notion d'\emph{attribut moyen}, que suggérait déjà l'analyse de 2000. Après 2004, plusieurs travaux apparaissent qui utilisent cette notion d'attribut moyen .\\
  L'attribut moyen est une mesure liée à la distribution des états que traverse un agent en suivant sa politique dans l'environnement. Cette mesure tient une place centrale dans beaucoup des algorithmes existants, notamment car deux agents ayant des attributs moyens similaires rempliront une certaine tâche (\emph{i.e.}, optimiseront une certaine récompense) de manière similaire.\\

  La première contribution au domaine consiste en un mécanisme de calcul de cet attribut moyen \cite{klein2011batch}. Inspiré d'algorithmes existants pour l'approximation de fonction de valeur, thème central en apprentissage par renforcement, cette contribution apporte une méthode de calcul permettant l'évaluation \emph{off-policy} de l'attribut moyen d'une politique. Sans l'anglicisme, cela signifie que l'on peut évaluer une grandeur relative à une politique en observant \emph{une autre politique} (comme par exemple celle de l'expert). Testée en l'injectant dans l'approche centrale de 2004, cette idée a permis de résoudre les problème de l'apprentissage par renforcement inverse sur un problème jouet simple à partir des seules données issues de l'expert, mais n'a pas permis de complètement lever les obstacles imposés par la structure des algorithmes existants. Ceux-ci nécessitent en effet dans leur immense majorité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée. Cela n'est pas toujours possible uniquement avec les données de l'expert, il faut une autre source d'information.\\
  Cette contribution permet néanmoins de s'affranchir d'un simulateur pour estimer une grandeur centrale dans la majorité des approches du domaine. Cela permet d'éviter les soucis liés à la modélisation de l'environnement ou à un trop grand besoin en données (coûteuses à générer).
  Une autre contribution de plus faible envergure consiste en un apport sur la définition mathématique formelle du problème \cite{klein2011dimensionality} permettant de réduire l'espace dans lequel on doit chercher les solutions.\\
  Enfin, lors de la deuxième année de thèse, deux nouveaux algorithmes ont été trouvés qui permettent la résolution de l'ARI dans des domaines où les algorithmes existant ne fonctionnent pas. L'un de ces algorithmes a été publié dans une conférence francophone \cite{Supelec773} ainsi qu'une conférence internationale \cite{klein2012}. L'autre est en cours de relecture.

  Il m'a été permis de participer à cinq conférences en temps qu'auteur, j'ai pu rencontrer la communauté \emph{Machine Learning} au sens large à IJCAI \footnote{\href{http://ijcai-11.iiia.csic.es/}{http://ijcai-11.iiia.csic.es/}
 }, ainsi qu'à ECML\footnote{\href{http://www.ecmlpkdd2011.org/}{http://www.ecmlpkdd2011.org/}
 } qui précédait EWRL\footnote{\href{http://ewrl.wordpress.com/past-ewrl/ewrl9-2011/}{http://ewrl.wordpress.com/past-ewrl/ewrl9-2011/}
 } où j'ai pu présenter mes travaux à un public plus directement concerné du fait du sujet précis de ce \emph{workshop}. Enfin j'ai rencontré la communauté francophone à JFPDA\footnote{\href{https://zanuttini.users.greyc.fr/jfpda2011/}{https://zanuttini.users.greyc.fr/jfpda2011/}} puis l'année suivante à CAp \footnote{http://cap2012.loria.fr/}. J'ai à nouveau participé à EWRL en 2012. Les échanges formels et informels permis par ces déplacements ont enrichi ma culture et ma reflexion.

\section{Perspectives}
\label{sec-4}

Outre la rédaction de la thèse, qui sera commencée aux alentours de Février pour une remise du manuscrit en Juin et une soutenance en Septembre, cette année sera occupée par l'application de nos nouveaux algorithmes sur des problèmes complexes et tirés de cas concrets afin de démontrer empiriquement leur performance.

 \begin{btSect}[alpha]{Biblio.bib}
 \section{Mes publications}
 \nocite{klein2011dimensionality,klein2011batch,klein2011batch2,klein2011batch3,klein2012,Supelec773}
 \btPrintCited
 \end{btSect}

\end{document}

### 2012-09-13 Rapport1A/main.tex
Rajout de tous les fichier qui avaient l'air un peu important et qui trainaient par là. L'état du machin est probablement incohérent.

% Created 2012-06-15 ven. 16:27
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\usepackage{amsmath}
\providecommand{\alert}[1]{\textbf{#1}}

\title{Rapport Intermédiaire de Thèse, Janvier-Décembre 2011}
\author{Klein Edouard}
\date{\today}

\begin{document}

\maketitle

\setcounter{tocdepth}{3}
\tableofcontents
\vspace*{1cm}

\section{Introduction}
\label{sec-1}

  Ces dernières décennies, de plus en plus de tâches sont confiées à des systèmes automatiques. L'automatisation de plus en plus poussée et l'autonomie grandissante des systèmes en place réduit de manière générale les coûts et les risques associés. Diverses techniques peuvent être envisagées pour automatiser un système, qu'il s'agisse du thermostat d'un four, d'une machine-outil, d'un robot humanoïde ou même d'un système de dialogue par téléphone.\\

  L'approche la plus classique, antérieure même à l'avènement de l'informatique, est celle de l'automatique. Armé d'un modèle inverse du système à contrôler, l'automaticien parvient à créer un mécanisme soumettant le système à la consigne de l'opérateur. Le succès de cette méthode est indéniable comme le prouve sa vaste adoption encore aujourd'hui (une direction assistée est un système asservi au sens de l'automatique, par exemple). Néanmoins, les informations qu'il est nécessaire d'obtenir sur le système à contrôler restent parfois inconnues. Un modèle inverse est en effet une formulation mathématique du monde permettant de trouver l'action à effectuer pour transiter entre deux états donnés. Il n'est pas possible d'obtenir une telle généricité pour la plupart des systèmes.\\

  Une approche moins contraignante est la planification qui à partir d'informations portant sur les effets d'une action sur le système va élaborer un \emph{plan}. Cette suite d'action sera appliquée en séquence pour atteindre le but fixé. Il est beaucoup moins difficile d'établir un modèle mathématique (dit direct) d'un système permettant de prédire les effets d'une action que de créer un modèle inverse comme celui que nécessite l'automatique.

  On rentre ici dans le cadre de la prise de décision séquentielle. Un \emph{agent} est placé aux commandes du système qui devient alors l'\emph{environnement} de l'agent. A intervalles réguliers, l'agent va intervenir et appliquer une \emph{action} sur son environnement (que nous appelions \emph{système} dans les paragraphes précédents). Les outils de planifications trouvent, à l'aide du modèle direct, le chemin que devra suivre l'agent en appliquant action après action pour, sur le long terme, atteindre le but fixé.\\

  Un défaut notable des approches de planification est leur incapacité à gérer l'incertain. L'idée même de \emph{plan} suppose que tout se déroule sans accrocs. Bien des environnements de la vie réelle ne sont malheureusement pas aussi sympathiques. Il faut permettre à l'agent d'appliquer une bonne action, même si l'action précédente n'a pas atteint le but escompté ou prévu par le modèle. A cette fin, la notion de \emph{plan} est élargie en notion de \emph{politique}. A intervalles réguliers, l'agent ne se content plus d'appliquer en aveugle une action prédéfinie mais étudie l'ensemble des informations disponibles sur sa position au sein de l'environnement (appelé \emph{état})  avant de choisir l'action adaptée. Une politique ne se contente ainsi pas de tracer un chemin que l'agent doit suivre, mais propose une action cohérente pour chacun des états dans lequel l'agent peut être amené à se trouver. Il s'agit d'un mécanisme réactif, où l'agent est en interaction constante avec son environnement.

  La description de la tâche a effectuer est fournie à l'agent sous forme de récompenses, qu'il ne touche que si l'état dans lequel il se trouve est établi comme souhaitable par l'opérateur. On peut symétriquement ``punir'' l'agent en lui donnant une récompense négative lorsqu'il se trouve dans des états indésirables, par exemple ceux qui mettent l'intégrité du système en danger. Le critère maximisé par l'agent est la somme pondérée de cette récompense au long de tous les états qu'il traverse. Maximiser la somme pondérée et non la récompense directement force l'agent à apprendre un comportement cohérent sur le long terme c'est à dire à ne pas favoriser la maximisation immédiate et éphémère de la récompense seule.

  Des outils de plus en plus puissants ont été développés dans le cadre de l'\emph{apprentissage par renforcement} (AR) pour permettre à un agent d'apprendre une politique optimisant une récompense donnée. Il est maintenant possible d'apprendre une politique sans connaître même le modèle direct, l'agent le découvrant au fur et à mesure de ses interactions avec l'environnement. Mieux, l'agent peut apprendre une politique optimale (ou quasi optimale) en observant \emph{un autre agent} interagir avec l'environnement. Cela évite de placer un agent novice aux commandes d'un système coûteux. Au sein de l'équipe IMS, plusieurs chercheurs s'intéressent aux problématiques liées à l'\emph{apprentissage par renforcement}.
\section{Sujet de thèse}
\label{sec-2}

  Les outils dont nous venons de parler permettent l'apprentissage une fois la représentation du monde mise en place et une fois fournie une fonction de récompense définissant la tâche a effectuer. Cette dernière tâche n'est pas forcément la plus facile. Si, pour les tâches d'ordinaires dévolues à l'automatique, il semble facile de transposer une description textuelle d'une tâche (``Allonger le bras'') en une fonction de récompense (``Récompense donnée lorsque l'angle du coude prend la valeur 180°'') cette facilité s'estompe rapidement à mesure que la tâche devient complexe.\\

  Prenons l'exemple de la conduite. Une description textuelle de la tâche est déjà difficile à imaginer. Il est possible de partir de quelques règles du type ``Ne pas provoquer de collisions'', ``Ne pas freiner brusquement'', ``Ne pas changer de file à l'improviste''. Mais que faire lorsque ces règles entrent en conflit ? Si la voiture devant celle de l'agent pile brusquement, doit-il à tout prix éviter la collision ? Pour ce faire, faut-il qu'il change de file à l'improviste ou qu'il freine brusquement ? Il n'est pas facile de répondre \emph{a priori} à ce type de questions qui deviennent lorsque la tâche est complexe trop nombreuses pour trouver réponse satisfaisante.\\

  Nous nous trouvons maintenant dans une situation paradoxale où nous disposons d'outils extrêmement puissants et pratiques pour apprendre un comportement dans des environnements difficiles et aléatoires, mais où nous ne sommes pas capable de spécifier la tâche que nous voulons voir l'agent apprendre. Nous pouvons heureusement contourner cette difficulté par le biais de l'\emph{apprentissage par renforcement inverse} (ARI), cadre dans lequel s'inscrit cette thèse.\\

  De la même manière que les jeunes gens n'apprennent pas à conduire en lisant le manuel de leur auto mais en observant leurs parents et leurs moniteurs de conduite puis en se mettant derrière le volant, nous comptons apprendre la tâche à effectuer en observant un \emph{expert} la réaliser. Cette démarche exploite la capacité humaine à résoudre intuitivement et rapidement des conflits dont nous serions incapables d'analyser sur papier. Pour reprendre l'exemple précédent, un automobiliste saura après un rapide coup d'œil dans son rétroviseur s'il vaut mieux qu'il pile ou qu'il déboîte en urgence et effectuera sa manœuvre dans la foulée.\\

  Notre but est de dériver, du comportement d'un expert effectuant une tâche, une description de cette tâche sous la forme d'une fonction de récompense, ce qui permet ensuite l'utilisation des outils d'apprentissage par renforcement pour apprendre cette tâche à un agent. Cela ouvrirait le champ d'application de l'apprentissage par renforcement à des tâches encore inaccessibles car trop complexes pour être ``expliquées''.\\

  Les outils utilisés pour résoudre les problèmes de prise de décision séquentielle remontent jusque dans les années 50. L'apprentissage par renforcement inverse est en comparaison un problème assez jeune, posée pour la première fois en 1998, et qui a réellement pris son essor après 2004.

  Le problème de l'ARI tarde à trouver une solution à la hauteur de ce qui se fait dans le domaine de l'AR. Il s'agit notamment d'un problème dont la formulation mathématique rigoureuse laisse à désirer. Une récompense nulle dans tous les états laisse le champ libre à l'agent qui peut alors faire ce qu'il veut sans être ni pénalisé ni récompensé. Toutes les politiques sont donc optimales vis à vis de cette récompense nulle. Particulièrement, la politique de l'expert est optimale vis à vis de la récompense nulle. Une approche mathématique naïve ne permet de trouver comme récompense justifiant le comportement de l'expert que cette solution triviale et n'apportant absolument pas la réponse à notre problème.

  Un autre soucis de l'approche est la difficulté de l'appliquer à des cas réels tant elle nécessite d'information sur l'environnement. Aucun des algorithmes existants au début de la thèse ne peut se passer de disposer du modèle de l'environnement. L'interaction seule avec l'environnement est en théorie possible, mais en pratique impossible à mettre en place tant les interactions nécessaires sont nombreuses. Il faut alors avoir recours à un simulateur.
\section{Contributions au domaine}
\label{sec-3}

  La bibliographie de ce tout nouveau domaine est très fractionnée, avec des apports venue de différentes communautés. Il faut attendre 2000 pour qu'une publication pose formellement le problème. La contribution centrale intervient en 2004 et introduit la notion d'attribut moyen, sorte d'historique du passage dans l'espace d'état. Dans la foulée de cette publication, plusieurs travaux apparaissent qui utilisent cette notion d'attribut moyen. Notamment, on compare cet historique de l'expert à l'historique de l'agent, à qui l'on apprend par modification successives de la récompense à obtenir un historique proche de celui de l'expert. Bien que l'on obtienne parfois une fonction de récompense exploitable, cette approche relève plus de l'apprentissage apr imitation que de l'apprentissage par renforcement inverse au sens philosophique du terme.\\

  La première contribution au domaine consiste en un mécanisme de calcul de cet attribut moyen \cite{klein2011batch}. Inspiré d'algorithmes existants pour l'approximation de fonction de valeur, thème central en apprentissage par renforcement, cette contribution apporte une méthode de calcul permettant l'évaluation \emph{off-policy} de l'attribut moyen d'une politique. Sans l'anglicisme, cela signifie que l'on peut évaluer une grandeur relative à une politique en observant \emph{une autre politique} (comme par exemple celle de l'expert). Couplée à l'approche centrale de 2004, cette avancée a permis de partiellement lever l'un des principaux obstacles du domaine, en ceci que sur des problèmes simples il est désormais possible de retrouver une fonction de récompense non triviale justifiant le comportement de l'expert, à partir des seules données fournies par celui-ci. Plus besoin de modèle (ou de simulateur). Si l'on dispose d'agents autres que l'expert pouvant manipuler le système assez largement, alors il est possible d'exploiter ces données pour fournir résoudre le problème de l'ARI sans avoir à mettre l'agent en cours d'apprentissage au commandes du système.\\

  Une autre contribution de plus faible envergure consiste en un apport sur la définition mathématique formelle du problème \cite{klein2011dimensionality} permettant de réduire l'espace dans lequel on doit chercher les solutions.
\section{Perspectives}
\label{sec-4}

  Le problème de l'ARI n'est pas résolu, c'est à dire qu'il n'existe pas de solutions applicables facilement. Néanmoins certaines idées naissent en ce moment même dans la communauté qui rendent la résolution de moins en moins contraignante. Notre apport pour le calcul de l'attribut moyen s'inscrit dans cette direction. Nous pensons que couplé à un nouveau type d'approche ne se basant pas sur un agent imitant l'expert et améliorant petit à petit son imitation, nous pourrons arriver à résoudre l'ARI.

\bibliographystyle{alpha}
\bibliography{../../Biblio/Biblio.bib}

\end{document}

### 2012-09-13 piot2012cascading/main.tex
Rajout de tous les fichier qui avaient l'air un peu important et qui trainaient par là. L'état du machin est probablement incohérent.

% Created 2012-06-04 lun. 15:48
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\usepackage{amsmath}
\usepackage{nips12submit_e,times}
\usepackage{makeidx}  % allows for indexgeneration
% For figures
\usepackage{graphicx} % more modern
%\usepackage[latin1]{inputenc}
%\usepackage[francais]{babel}
\usepackage{subfigure}
\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\usepackage{dsfont}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\hypersetup{
colorlinks,%
citecolor=black,%
filecolor=black,%
linkcolor=black,%
urlcolor=black
}
\mathtoolsset{showonlyrefs=true}
\newtheorem{hypo}{Hypothesis}
\newcommand{\argmax}{\operatorname*{argmax}}
\newcommand{\argmin}{\operatorname*{argmin}}
\newcommand{\arginf}{\operatorname*{arginf}}
\newcommand{\minp}{\operatorname*{min_+}}
\newcommand{\Ker}{\operatorname*{Ker}}
\newcommand{\trace}{\operatorname*{trace}}
\newcommand{\cov}{\operatorname{cov}}
\newcommand{\card}{\operatorname*{Card}}
\newcommand{\vect}{\operatorname*{Vect}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\erf}{\operatorname{erf}}
\newcommand{\bound}{\operatorname*{bound}}
\newcommand{\vpi}{\operatorname{VPI}}
\newcommand{\gn}{\operatorname{Gain}}
\newcommand{\p}{\operatorname{Pr}}
\newcommand{\mlp}{\operatorname{MLP}}
\newcommand*\tto[2]{\smash{\mathop{\longrightarrow}\limits_{#1}^{#2}}}
\newcommand*\ntto[2]{\smash{\mathop{\nrightarrow}\limits_{#1}^{#2}}}
\newcommand{\X}{\mathbf{X}}
\newcommand{\Q}{\mathbf{Q}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\Z}{\mathbf{Z}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\R}{\mathbf{R}}
\newcommand{\ba}{\mathbf{a}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bd}{\mathbf{d}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\af}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bh}{\mathbf{h}}
\newcommand{\bi}{\mathbf{i}}
\newcommand{\bj}{\mathbf{j}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bl}{\mathbf{l}}
\newcommand{\bm}{\mathbf{m}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\bo}{\mathbf{o}}
\newcommand{\bp}{\mathbf{p}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bt}{\mathbf{t}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\ma}{\mathbf{A}}
\newcommand{\mb}{\mathbf{B}}
\newcommand{\mc}{\mathbf{C}}
\newcommand{\md}{\mathbf{D}}
\newcommand{\me}{\mathbf{E}}
\newcommand{\mf}{\mathbf{F}}
\newcommand{\mg}{\mathbf{G}}
\newcommand{\mh}{\mathbf{H}}
\newcommand{\mi}{\mathbf{I}}
\newcommand{\mj}{\mathbf{J}}
\newcommand{\mk}{\mathbf{K}}
\newcommand{\ml}{\mathbf{L}}
\newcommand{\mm}{\mathbf{M}}
\newcommand{\mn}{\mathbf{N}}
\newcommand{\mo}{\mathbf{O}}
\newcommand{\Mp}{\mathbf{P}}
\newcommand{\mq}{\mathbf{Q}}
\newcommand{\mr}{\mathbf{R}}
\newcommand{\ms}{\mathbf{S}}
\newcommand{\mt}{\mathbf{T}}
\newcommand{\Mu}{\mathbf{U}}
\newcommand{\mv}{\mathbf{V}}
\newcommand{\mw}{\mathbf{W}}
\newcommand{\mx}{\mathbf{X}}
\newcommand{\my}{\mathbf{Y}}
\newcommand{\mz}{\mathbf{Z}}
\newcommand{\tphi}{\tilde{\Phi}}
\newcommand{\espace}{\text{ }}
\newcommand{\x}{\mathbf{x}}
\newcommand{\s}{\mathbf{s}}
\newcommand{\n}{\mathbf{n}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\rr}{\mathbf{r}}
\newcommand{\0}{\mathbf{0}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\am}{{\mathcal{A}_m}}
\newcommand{\amj}{{\mathcal{A}_m^{+j}}}
\newcommand{\sgn}{\operatorname{sgn}}
\title{Learning a reward function from demonstrations: a cascaded supervised
learning approach}
\author{Edouard Klein$^{1,2}$\\
1. ABC Team\\
LORIA-CNRS, France.
\And Bilal Piot$^{2}$\\
2. Supélec-Metz Campus\\
MaLIS Research group, France\\
\And Matthieu Geist$^1$\\
\texttt{prenom.nom@supelec.fr}\\
\And Olivier Pietquin$^{2,3}$\\
3. UMI 2958 CNRS\\
GeorgiaTech, France
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\providecommand{\alert}[1]{\textbf{#1}}

\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Klein Edouard}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Discovering the reward function optimized by an expert in an MDP, usually referred to as the IRL problem, is addressed in this paper. The proposed generic model-free contribution consists in cascading a classification and a regression steps to produce a non-trivial reward function. We prove the expert policy to be near optimal with respect to this reward. The algorithm does not require solving any direct reinforcement learning problem (unlike most of IRL algorithms). With the help of some heuristics, it can be instantiated to solely rely on transitions from the expert (no need to sample trajectories according to other policies). All of this is illustrated through a standard benchmark.
\end{abstract}
\section{Problem statement}
\label{sec-1}

Given a sequential decision making problem framed in the Markov Decision Process (MDP) formalism, it is often difficult to manually specify a reward function in order to train an agent to fulfill a certain task. However, observing an expert demonstrating this task is often possible and easier \cite{russell1998learning}. That is why learning from demonstrations and more particularly Inverse Reinforcement Learning (IRL) has received more and more interest this last decade. The IRL framework, first introduced in \cite{russell1998learning,ng2000algorithms}, assumes that an expert, demonstrating a task, is acting optimally in an MDP with respect to an unknown reward function to be discovered. Unlike direct methods of apprenticeship learning\footnote{The definition of apprenticeship learning we use is a restriction of learning from demonstration to MDP settings, where the output of the algorithm is a control policy. } which learn a mapping from states to actions via supervised learning (SL) \cite{atkeson1997robot,pomerleau1989alvinn}, IRL methods aim at recovering a reward function. We see this reward function as the most compact and transferable way to specify a task. One can even try to derive semantic information from it, for example in biological or economical studies as hinted to in \cite{russell1998learning}.
Most of the direct methods do not use the structure of the MDP (one notable exception being \cite{melo2010learning}), hence only mimic the policy of the expert and are not able to generalize it correctly. This possibly leads to incorrect behavior outside demonstrated states.
In order to overcome this drawback, undirect methods using the IRL framework \cite{abbeel2004apprenticeship} were introduced. Most of these methods (see \cite{neu2009training} for a comprehensive overview) assume the dynamic of the environment known, or need an environment model (via a simulator) so as to test the effects of a policy. Moreover, most of existing undirect approaches require to solve the forward Reinforcement Learning (RL) problem (find an optimal policy knowing a given reward) several times (except  \cite{boularias2011relative} which however requires sampling non expert trajectories).
Finally, in many cases, the output of the algorithm is a policy but not a reward function (see Section \ref{section: related work} for an in-depth presentation). These constraints can even lead to difficulties in real-world applications and make the problem of apprenticeship learning harder than the forward RL problem. That is why a reborn interest for direct methods which passively imitate the expert exists. More particularly the reduction of the apprenticeship learning to classification, first introduced 
in \cite{zadrozny2003cost}
and recently used 
in \cite{melo2010learning}, is legitimated for finite horizons problems in \cite{syed2010reduction},\cite{ross2010efficient}.
Our approach tries to avoid the drawbacks of both direct and undirect methods by cascading two well-known supervised learning methods (see Section \ref{section: Cascading}): first a step of classification from data drawn by the expert provides us with a score function over the state-action space (i.e. for each state we are provided with a ranking of the actions), then a step of regression introduces the MDP structure and outputs a reward function.
Our approach does not require any of the following: complete trajectories from an expert, a generative model of the environment, the knowledge of the transition probabilities, the ability to compute a (near)-optimal policy for different rewards and hence solve the forward (RL) problem or even the perfect knowledge of the expert's policy. Also, we exhibit a reward function and not a control policy. Our analysis (Section \ref{section: Analysis}) shows that the expert is epsilon-optimal with respect to the resulting reward function. Moreover, our algorithm can be instantiated with any off-the-shelf score-function based classifier and any regression method. We provide such an instantiation and empirically show (Section \ref{section: experiments}) that only a small amount of expert demonstrations (not even in the form of trajectories but simple transitions) is required. Comparison to other state-of-the-art methods is provided Section \ref{section: related work}. We start with some background mathematical notations in Section \ref{section: background}.
\section{Background and notations}
\label{sec-2}

  \label{section: background}
First, we introduce some general notations. Let $X=\{x_i\}_{1\leq i \leq I}$ be a finite set of $I\in\mathbb{N}^*$ elements, a function $g\in\mathbb{R}^X$ can also be seen as the real column vector $g=[g(i)=g(x_i)]_{1\leq i \leq I}$, $g^T$ is the transposition of $g$, and $|g|=[|g|(i)=|g(i)|]_{1\leq i \leq I}$.
$\Delta_X$ is the set of probability distributions over $X$. Let $\mu\in\Delta_X$ and $f\in\mathbb{R}^X$, we define the $\mu$-weighted $l_1$-norm:
\begin{equation}
\|f\|_{\mu,1}=\sum_{x\in X}\mu(x)|f(x)|=E_{x\sim\mu}[|f(x)|]=\mu^T|f|,
\end{equation}
where the notation $E_{x\sim\mu}$ is the expectation with $x$ drawn according to $\mu$.

Now, we introduce notations specific to the MDP paradigm. A (finite) MDP \cite{puterman1994markov} is a tuple $M=\{S,A,P,\gamma,R\}$ where $S=\{s_i\}_{1\leq i \leq N}$ is the finite state space with $N\in\mathbb{N}^*$ states, $A=\{a_k\}_{1\leq k \leq K}$ is the finite action space with $K\in\mathbb{N}^*$ actions, $P=\{P_{s,a}\}_{(s,a)\in S\times A}$ is the set of transition probabilities where $P_{s,a}$ is a probability distribution over $S$. $P$ represents the dynamics of the MDP and the notation $p(s,a,s')=P_{s,a}(s')$ quantifies the probability to reach $s'\in S$ knowing that the action $a \in A$ was taken in the state $s\in S$, $\gamma\in]0,1[$ is the discount factor and $R$ is a function from $S\times A$ to $\mathbb{R}$ called the reward function. A deterministic policy $\pi$ is a function from $S$ to $A$, the value function $V^\pi_R$ is a function from $S$ to $\mathbb{R}$ defined by:
\begin{equation}
V^\pi_R(s)=E^\pi_s[\sum_{t=0}^{+\infty}\gamma^tR(s_t,\pi(s_t))], \forall s \in S,
\end{equation}
where $E^\pi_s$ is the expectation over the distribution of the trajectories $(s_0,s_1,\dots)$ obtained by executing the policy $\pi$ starting from $s_0=s$.
The action-value function $Q^\pi_R$ is a function from $S\times A$ to $R$ defined by: $\forall s\in S, \forall a\in A$, $Q^{\pi}_R(s,a)=R(s,a)+\gamma\sum_{s'\in S}P_{s,a}(s')V^{\pi}_R(s')$. The classical result for finite MDPs is the Bellman equation: $\forall s\in S$, $V^{\pi}_R(s)~=~R(s,\pi(s))~+~\gamma\sum_{s'\in S}P_{s,\pi(s)}(s')V^{\pi}_R(s')$.
A policy $\pi$ is said optimal for the reward $R$ when:
$\forall s\in S, \forall \tilde{\pi}\in A^{S}, V^{\pi}_R(s)\geq V^{\tilde{\pi}}_R(s)$.
Another classical result called the Bellman optimality is: $\pi$ is an optimal policy for the reward $R$ if and only if:
\begin{equation}
\label{equation:Qoptimal}
\forall s\in S, \pi(s)\in\argmax_{a\in A} Q^\pi_R(s,a).
\end{equation}
For a given policy $\pi$ and a given reward function $R$, we define $R_\pi=[R_\pi(i)=R(s_i,\pi(s_i))]_{1\leq i \leq N}$, the $N\times N$-matrix $P_\pi=[P_\pi(i,j)=p(s_i,\pi(s_i),s_j]_{1\leq i,j \leq N}$ and the Bellman operator $T^\pi_R$ which is a function from $\mathbb{R}^S$ to $\mathbb{R}^S$:
\begin{equation}
\forall V\in\mathbb{R}^S, T^\pi_RV=R_\pi+\gamma P_\pi V.
\end{equation}
It is well-known that the unique fixed point of the operator $T^\pi_R$ is  $V^\pi_R$ \cite{puterman1994markov}.

RL consists in finding an optimal policy for the reward function $R$. Notice that the state space may be too large for an exact representation of the value function (which calls for approximate
representation), that the model ($P$ and $R$) may be unknown (the only information being provided through rewarded transitions sampled according
to some behavioral policy), that learning can occur in an online or off-line setting, and so on. See \cite{bertsekas2001dynamic},\cite{sutton1998reinforcement} for details.

In the classical IRL paradigm \cite{ng2000algorithms}, an MDP without reward $M\backslash R =\{S,A,P,\gamma\}$ and a policy $\pi_E$ called expert-policy are given. The problem is to find
a reward $R^*$ for which the policy $\pi_E$ is optimal. However this problem is clearly ill-posed in the sense that there is not uniqueness of the reward $R^*$: many reward functions are equivalent in that they have the same optimal deterministic policies, moreover the trivial zero-reward is a solution for any deterministic policy $\pi_E$ as it is shown in \cite{ng2000algorithms}. In the literature, some solutions are proposed in order to respond to the ill-posed nature of the problem \cite{ng2000algorithms,ziebart2008maximum,boularias2011relative}.
In our experiments (see Section \ref{section: experiments}) we assume that the solely available information is provided by transitions sampled according to the dynamics of the environnement under $\pi_E$: $\{(s_i,a_i=\pi_E(s_i)),s_i')\}_{1\leq i \leq D}, D\in\mathbb{N}^*$,
where $s_i'$ is sampled according to the distribution $P_{s_i,a_i}$.
The reward function is obviously unknown, but this assumption means that the dynamics ($P$) is only known through transitions $(s_i, a_i, s_i')$ and that the
policy $\pi_E$ is only known through state-action pairs $(s_i, a_i)$.

For a given deterministic policy $\pi$, a component $P_{s_i,\pi(s_i)}(s_j)$ of the matrix $P_\pi$ represents the probability to transit from $s_i$ to $s_j$ under the policy $\pi$. So $P_\pi$ can be seen as a transition matrix of a finite Markov-chain on the finite state space $S$. If $P_\pi$ is irreducible then it exists a unique distribution $\mu_\pi$ (see \cite{baldi2000martingales}) called the stationary distribution such that:
\begin{equation}
\label{equation: stationarity}
\mu_\pi^T=\mu_\pi^T P_\pi.
\end{equation}
\label{section: background}
\section{Cascading Classification and Regression for IRL}
\label{sec-3}

\label{section: Cascading}

The first SL step in our algorithm is to train a classifer over a set
\begin{equation}
\label{equation:data}
D_C=\{(s_i,a_i=\pi_E(s_i)),s'_i)\}_{1\leq i \leq D}, D\in\mathbb{N}^*,
\end{equation} where $s'_{i}$ is sampled according to the distribution $P_{s_{i},a_{i}}$. The actions $a_i$ are seen as labels for the inputs $s_i$. We restrict ourselves to the use of score-function based classifiers for which the classification rule is of the form $\forall s \in S, \pi_C(s) \in \arg\max_{a\in A} q(s,a)$. Most classifiers, from $k$-nearest neighbours to multi-class-SVMs and structured margin approaches, fall into this category. The classification rule of this broad range of classifiers can be seen as a deterministic policy. After noticing the similarity between the definition of $\pi_C$ and equation \eqref{equation:Qoptimal}, it is easy and natural to identify the score function $q:S\times A \rightarrow \mathbb{R}$ with a state-action value function with respect to a certain reward function $R^C$ (equation \eqref{RC.eq}) for which $\pi_C$ is an optimal policy (explained and proven in Section \ref{section: Analysis}).\\
As the model is often unknown, it is not possible to directly compute $R^C$. We can however use a new dataset $D_R=\{(s_{i},a_{i},s'_{i})\}_{1\leq i \leq D'}, D'\in\mathbb{N}^*$, where $s'_{i}$ is sampled under the probability $P_{s_{i},a_{i}}$. In $D_R$, we don't necessarily have $a_i = \pi_E(s_i)$ as we had in dataset $D_C$. We can write :
\begin{equation}
\label{ri.def}
 \forall i\in \{1,\dots,D'\},\hat{r}_i=q(s_{i},a_{i})-\gamma q(s'_{i},\pi_C(s'_{i})).
\end{equation}
An unbiased estimator of $R^C(s_i,a_i)$ is $\hat r_i$. It is possible to build an estimate $\hat{R}^C$, which is a function from $S\times A$ to $\mathbb{R}$, of $R^C$ using a regressor trained on the dataset:$\{(s_{i},a_{i},\hat{r}_i)\}_{1\leq i \leq D'}$.

The whole approach is summed up algorithm \ref{algo:cascading}. We see $\hat{R}^C$ as an approximation of $R^C$ and 
$\hat\pi_{C}$ is defined as an optimal policy for the reward $\hat{R}^C$.
In order to verify that the reward function $\hat{R}^C$ is a good candidate to resolve the IRL problem, it will be proven in Section \ref{section: Analysis} that the policy $\pi_E$ is near-optimal for the reward $\hat{R}^C$ (confer theorem \ref{theorem : results}).

\begin{algorithm}%[H]
    %\small
  \caption{Cascading IRL algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi_E(s_i)),s'_i)\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{i},a_{i},s'_{i})\}_{1\leq i \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{i},a_{i},\hat{r}_i)\}_{1\leq i \leq D'}$, $\forall (s_i,a_i,s'_i) \in D_R,\hat{r}_i=q(s_{i},a_{i})-\gamma q(s'_{i},\pi_C(s'_{i}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}
\section{Analysis}
\label{sec-4}

\label{section: Analysis}
This section is devoted to show that the cascading approach is legitimate under some hypotheses. Lemma \ref{lemma: calculs V}  gives a practical way to compute $E_{s\sim\mu_\pi}[V^\pi_R]$ for a given policy $\pi$ and a given reward function $R$. Theorem \ref{theorem : results}  gives an upper bound to the term $E_{s\sim\mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]$ where $\mu_E$ is the stationary distribution of the expert policy $\pi_E$. We also give an interpretation of the term $E_{s\sim\mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]$ and explain why being able to bound this term means that our approach is legitimate.
\subsection{Results and Discussion}
\begin{lemma}
\label{lemma: calculs V}
Let $\{S,A,P,\gamma,R\}$ be a finite MDP and $\pi$ a deterministic policy.
If $P_\pi$ is irreducible, then $E_{s\sim\mu_\pi}[V^\pi_R]=\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\mu_\pi^TR_\pi$.
\end{lemma}
Lemma \ref{lemma: calculs V} gives a practical tool which will be useful in order to simplify some terms in the proof of Theorem \ref{theorem : results}.
Before giving an upper bound to $E_{s\sim\mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]$, we define $\epsilon_C\in\mathbb{R}_+$ called the classification error, $R^C\in \mathbb{R}^{S\times A}$ the reward function for which $\pi_C$ is optimal (see Theorem \ref{theorem : results}) and the function $\epsilon_R$ from $S\times A$ to $\mathbb{R}$ called the regression error such that:
\begin{align}
&\epsilon_C=\sum_{s\in S}\mu_{E}(s)\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}=E_{s \sim \mu_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}],
\\
&\forall a\in A, \forall s\in S,R^C(s,a)=\sum_{s'\in S}p(s,a,s')[q(s,a)-\gamma q(s',\pi_C(s'))],\label{RC.eq}\\
&\forall a\in A, \forall s\in S, \epsilon_R(s,a)=\hat{R}^C(s,a)-R^C(s,a).
\end{align}
We define $\epsilon^C_R=[\epsilon_R(s_i,\hat{\pi}_C(s_i))]_{1\leq i \leq N}$ and $\epsilon^E_R=[\epsilon_R(s_i,\pi_E(s_i))]_{1\leq i \leq N}$.
\begin{remark}
One can notice that the classification error is defined thanks to the expectation $E_{s \sim \mu_E}$. In the classical framework of classification, the data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$ are generated independently according to a distribution $\mu_{\text{Data}}$ over $S$ and hence the classical classification error must be $\epsilon_C=E_{s \sim \mu_{\text{Data}}}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]$.
Then one can suppose that our data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$ are generated independently thanks to $\mu_E$ but this assumption is quite strong and non-realistic.
However it is often the case that the data provided by the expert are trajectories. In that case ,if $P_{\pi_E}$ is irreducible , the rate of convergence of the data distribution
to the stationary distribution is at least exponential (this is the Doeblin theorem see \cite{stroock2005introduction}). It allows us to suppose that the data $\{s_i,a_i=\pi_E(s_i)\}_{1\leq i \leq D}$ are generated under $\mu_E$.
\end{remark}
Finally, we introduce also the first order discounted future state distribution concentration coefficient (\cite{MunosSIAM07}):
\begin{equation}
C_1=(1-\gamma)\sum_{t\geq0}\gamma^tc(t), \text{ with } c(t)=\max_{\pi_1,\dots,\pi_t,s}\frac{(\mu_E^TP_{\pi_1}\dots P_{\pi_t})(s)}{\mu_E(s)}.
\end{equation}
\begin{theorem}
\label{theorem : results}
Let $\{S,A,P,\gamma\}$ be a finite MDP without reward and $\pi_E$ an expert policy.
The notations $q$, $\pi_C$, $\hat{\pi}_C$, $\hat{R}^C$ are introduced in the Section \ref{section: Cascading}.
If $P_{\pi_E}$ is irreducible, then $\mu_E$ is the stationary distribution of $\pi_E$ and:
\begin{enumerate}
\item $\pi_C$ is optimal for the reward $R^C$.
\item $0\leq E_{s\sim\mu_E}[V^{\pi_C}_{R^C}(s)-V^{\pi_E}_{R^C}(s)]\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}$.
\item $0 \leq E_{s\sim\mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+\|\epsilon^E_R\|_{\mu_E,1}+C_1\|\epsilon^C_R\|_{\mu_E,1}}{1-\gamma}$.
\end{enumerate}
\end{theorem}
In order to understand why this theorem is useful, let us make some important assumptions. Let us suppose that the classification and the regression steps are perfect in the sense that $\forall (s,a)\in S\times A,\epsilon_R(s,a)=0$ and $\epsilon_C=0$. Then we obviously have, thanks to the theorem \ref{theorem : results}, that $\pi_E=\pi_C$ is optimal for $\hat{R}^C=R^C$. Thus the method is able to provide a non-trivial reward function for which the policy $\pi_E$ is optimal. Moreover if the classification step and the regression step are not perfect, the theorem \ref{theorem : results} shows, that our approach is able to provide a non trivial-reward $\hat{R}^C$ for which the policy $\pi_E$ is near-optimal in the sense that:
\begin{equation}
0 \leq E_{s \sim \mu_E}[V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \mathcal{O}(\frac{\epsilon_C+ \|\epsilon^C_R\|_{\mu_E,1}+\|\epsilon^E_R\|_{\mu_E,1}}{1-\gamma})
\end{equation}
\begin{remark}
It is important to be clear about this result. If the only available data are provided by equation \eqref{equation:data}, it is possible to control
$\epsilon_C$ and $\|\epsilon^E_R\|_{\mu_E,1}$ because these errors depend only on the expert policy $\pi_E$. However it is not possible to control the error $\|\epsilon^C_R\|_{\mu_E,1}$
because it depends on the policy $\hat{\pi}_C$ which can be different from the expert policy and hence do not appear in the available data \eqref{equation:data}. However
it will be possible to obtain a control on the term $\|\epsilon^C_R\|_{\mu_E,1}$ if the data used for the regression are $D_R=\{(s_i,a_i,s'_i)\}_{1\leq i \leq D'}$, where $(s_{i},a_{i})$ are uniformly chosen on the set $S\times A$ or sampled from other policies than the expert. So, theoretically an easy way to be sure
to control the error $\|\epsilon^C_R\|_{\mu_E,1}$ is to be able to give a data set for the regression which is sampled from the expert policy and other policies (and more particularly $\hat{\pi}_C$). But we give examples, see Section \ref{section: experiments}, where the regression data set given by the equation \eqref{equation:data} is sufficient to obtain good results.
A possible argument to explain the fact that classification-regression still works when $D_C=D_R$, is that $\hat{\pi}_C$ must not be so different than $\pi_E$. 
\end{remark}
\subsection{Proofs}
\begin{proof}[Lemma \ref{lemma: calculs V}]
Here, we use equation \eqref{equation: stationarity}
\begin{align}
E_{s\sim\mu_\pi}[V^\pi_R]=\mu_\pi^TV^\pi_R&=\mu_\pi^T(R_\pi+\gamma P_\pi V^\pi_R)=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TP_\pi V^\pi_R,
\\
&=\mu_\pi^TR_\pi+ \gamma\mu_\pi^TV^\pi_R=\frac{1}{1-\gamma}\mu_\pi^TR_\pi.
\end{align}
\end{proof}
\begin{proof}[Theorem \ref{theorem : results}]
In order to prove the three results of the theorem \ref{theorem : results}, let us introduce the function $R_E$ from $S\times A$ to $\mathbb{R}$ such that:
\begin{equation}
\forall a \in A, \forall s\in S, R^E(s,a)=q(s,a)-\gamma\sum_{s'\in S}p(s,a,s')q(s',\pi_E(s')).
\end{equation}
The first step is to show that: $\forall s\in S, q(s,\pi_C(s))=V^{\pi_C}_{R^C}(s)$ and $\forall s\in S, q(s,\pi_E(s))=V^{\pi_E}_{R^E}(s)$.

This is quite straightforward because $q_{\pi_E}=[q(s,\pi_E(s))]_{1\leq i\leq N}$ is the fixed point of the operator $T^{\pi_E}_{R^E}$ and  $q_{\pi_C}=[q(s,\pi_C(s))]_{1\leq i\leq N}$ is the fixed point of the operator $T^{\pi_C}_{R^C}$:
\begin{align}
T^{\pi_E}_{R^E}q_{\pi_E}&=R^E_{\pi_E}+\gamma P_{\pi_E}q_{\pi_E},
\\
&=q_{\pi_E}-\gamma P_{\pi_E}q_{\pi_E}+\gamma P_{\pi_E}q_{\pi_E}=q_{\pi_E}.
\end{align}
With the same calculus, we show that $T^{\pi_C}_{R^C}q_{\pi_C}=q_{\pi_C}$.
It is clear that $\forall s\in S,\forall a\in A, q(s,a)=Q^{\pi_C}_{R^C}(s,a)$ as $R^C(s,a)=q(s,a)-\gamma\sum_{s'\in S}p(s,a,s')q(s',\pi_C(s'))$:
\begin{align}
\forall s\in S,\forall a\in A, Q^{\pi_C}_{R^C}(s,a)&=R^C(s,a)+\gamma\sum_{s'\in S}p(s,a,s')V^{\pi_C}_{R^C}(s'),
\\
&=R^C(s,a)+\gamma\sum_{s'\in S}p(s,a,s')q(s',\pi_C(s'))
\\
&=q(s,a).
\end{align}
So $\forall s\in S,\forall a\in A, q(s,a)=Q^{\pi_C}_{R^C}(s,a)$ and as $\forall s\in S, \pi_C(s)\in\argmax_{a\in A}q(s,a)$, $\pi_C$ is optimal for the reward $R^C$ via equation \eqref{equation:Qoptimal}, which implies $E_{s\sim\mu_E}[V^{\pi_C}_{R^C}(s)-V^{\pi_E}_{R^C}(s)] \geq 0$.
Now let us prove that:
\begin{equation}
E_{s\sim\mu_E}[V^{\pi_C}_{R^C}(s)-V^{\pi_E}_{R^C}(s)]=\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}.
\end{equation}
Indeed:
\begin{equation}
\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})=\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E}+V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}).
\end{equation}
And $\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})$ is such that:
\begin{align}
\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})&=\sum_{s\in S}\mu_E(s)[V^{\pi_C}_{R^C}(s)-V^{\pi_E}_{R^E}(s)],
\\
&=\sum_{s\in S}\mu_E(s)[q(s,\pi_C(s))-q(s,\pi_E(s))]\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}},
\\
&\leq\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)).
\end{align}
It remains to deal with the term $\mu_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})$ using the fact that $\mu_E^TP_{\pi_E}=\mu_E^T$ and Lemma \ref{lemma: calculs V}:
\begin{align}
\mu_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\frac{1}{1-\gamma}\mu_E^T(R^E_{\pi_E}-R^C_{\pi_E}),
\\
&=\frac{1}{1-\gamma}\mu_E^T(\gamma P_{\pi_E}q_{\pi_C}-\gamma P_{\pi_E}q_{\pi_E}),
\\
&=\frac{\gamma}{1-\gamma}\sum_{s\in S}\mu_E(s)[q(s,\pi_C(s))-q(s,\pi_E(s))]\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}},
\\
&\leq\frac{\gamma}{1-\gamma}\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)).
\end{align}
Finally:
\begin{equation}
\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})=\mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E}+V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})\leq\frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}.
\end{equation}
In order to finish the proof it remains to show that:
\begin{equation}
0 \leq \mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+C_1\|\epsilon^C_R\|_{\infty}+\|\epsilon^E_R\|_{\infty}}{1-\gamma}.
\end{equation}
As $\hat\pi_C$ is optimal for $\hat R^C$, we get $\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}\geq 0$. We also notice that:
\begin{equation}
\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C})=\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}+V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}+V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}).
\end{equation}
It is very easy to see that:$\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C})\leq\frac{C_1\|\epsilon^E_C\|_{\mu_E,1}}{1-\gamma}$ and $\mu_E^T(V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})\leq\frac{\|\epsilon^E_R\|_{\mu_E,1}}{1-\gamma}$.

Indeed via the definition of $C_1$ ($I$ is the identity matrix):
\begin{align}
\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}(s)-V^{\hat{\pi}_C}_{R^C}(s))&=\mu_E^T(I-\gamma P_{\hat{\pi}_C})^{-1}\epsilon^E_C,
\\
&=\mu_E^T(\sum_{t\geq0}\gamma^tP_{\hat{\pi}_C}^t)\epsilon^E_C\leq\frac{C_1\|\epsilon^E_C\|_{\mu_E,1}}{1-\gamma}.
\end{align}
And, thanks to the lemma \ref{lemma: calculs V} $\mu_E^T(V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})=\mu_E^T(V^{\pi_E}_{\epsilon_R})=\frac{\mu_E^T\epsilon^E_R}{1-\gamma}\leq\frac{\|\epsilon^E_R\|_{\mu_E,1}}{1-\gamma}$. Thus:
\begin{equation}
\label{equation: cas3-1}
\mu_E^T(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}+V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})\leq\frac{C_1\|\epsilon^E_C\|_{\mu_E,1}+\|\epsilon^E_R\|_{\mu_E,1}}{1-\gamma}.
\end{equation}
It remains to deal with the term $\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})=\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C}+V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})$.
As $\pi_C$ is optimal for the reward $R^C$ then $\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C})\leq 0$ so:
\begin{equation}
\label{equation: cas3-2}
\mu_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \mu_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))}{1-\gamma}.
\end{equation}
Finally by regrouping the results in \eqref{equation: cas3-1} and \eqref{equation: cas3-2}:
\begin{equation}
\mu_E^TV^{\hat{\pi}_C}_{\hat{R}^C}-\mu_E^TV^{\pi_E}_{\hat{R}^C}\leq \frac{\epsilon_C\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))+C_1\|\epsilon^E_C\|_{\mu_E,1}+\|\epsilon^E_R\|_{\mu_E,1}}{1-\gamma}.
\end{equation}
\end{proof}

  \begin{figure}[!Ht]
  \begin{tabular}{ccc}
  \subfigure[State-of-the-art approaches on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig1"}\label{Fig1.fig}}&\hspace{.05\linewidth}&
  \subfigure[Our new approach on the GridWorld]{\includegraphics[width=.45\linewidth]{"Fig2"}\label{Fig2.fig}}\\
  \subfigure[Results on the driving simulator]{\includegraphics[width=.4\linewidth]{"Fig3"}\label{Fig3.fig}}&\hspace{.05\linewidth}&\includegraphics[width=0.45\linewidth]{"Legend"}
  \end{tabular}
  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. The lower baseline is an agent trained on a reward generated by using the same features as the cascading approach and the classifier, but with a random vector of parameters.}
  \end{figure}
\section{Experimental results}
\label{sec-5}

\label{section: experiments}

   As a classifier, we choose a structured large margin approach \cite{taskar2005learning} which solves $\min_{\omega,\zeta}\frac{1}{2}\|\omega\|^2 + \frac{\eta}{N}\sum_{i=1}^N \zeta_i$  s.t. $\forall i, q(s_i,a_i)+\zeta_i \geq \max_a q(s_i,a) + \mathcal L (s_i,a)$, with $\mathcal L(s_i,a_i)=0$ and $\mathcal L(s_i,a\neq a_i)=1$ and for some slack variables $\zeta_i$. Assuming the slack variables are tight and using the linear parametrization $q(s,a) = \omega^T\phi(s,a)$ with some feature function $\phi$ over the state-action space, we use a subgradient descent on an objective function in $J$ in which the constraints have been moved \cite{ratliff2006maximum} : $J(\theta) = {1\over D} \sum_{i=1}^D\left(\max_{a}(\theta^T\phi(s_i,a) + l(s_i,a)) - \theta^T\phi(s_i,a_i) \right)$. When comparing our algorithm to pure classification we used the output of this classifier.

As we try to devise a reward function $\hat R^C$ using \emph{only} transitions from the expert (case not covered in the analysis), we have $D_C = D_R$. A good reward function can not be learnt from this data alone as no datapoint exists for tuples $(s,a,s')$ such that $a \neq \pi_E(s)$. We can use a heuristic to get a more complete dataset for the regression step ; assuming (this is quite pessimistic) that deviating from the expert's choice is the worst thing to do, we define $\hat r_{min} = \min\limits_{1\geq i\geq D=D'} \hat r_i - \epsilon, \epsilon>0$ and create the augmented dataset $\{(s_{i},a_{i},\hat{r}_i)\}_{1\leq i \leq D=D'} \cup \{(s_{i},a,\hat{r}_{min})\}_{1\leq i \leq D=D',a\in A,a\neq a_i}$ which is then fed to a simple least-squares regressor using the same features $\phi$ as the classifier. Using straightforward matricial notations, this can be written $\theta = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R$.
Finally, we have $\hat R^C(s,a) = \theta^T \phi(s,a)$.

   We first tested our approach on a simple $5\times 5$ GridWorld. The expert and the agent can choose between 4 actions (down, left right, or up) that send the player in the corresponding neighbouring cell with probability $0.7$. Three times out of ten however, the actions fails and the player moves in another different random direction. Trying to go off the grid will result in the player staying in its position. The expert begins in the lower left corner and is trained to go to the higher right corner by an exact dynamic programming algorithm. The original reward (with respect to which the performance of the agents is computed) is null everywhere but in the higher right corner where it is $1$ for all actions.

The results are shown in two figures. Figure \ref{Fig1.fig} shows the baselines we compare ourselves with. Pure classification and the algorithm of \cite{abbeel2004apprenticeship} are on par with one another, both far better than training an agent on a random reward ; their means both converge to a value close to the expert's given enough samples. Notice that the IRL algorithm can still critically fail even when given a lot of samples. This explain the standard deviation going over the maximum. Figure \ref{Fig2.fig} shows results of our algorithm together with the mean performance of both forementioned approaches (variance and min/max were dropped for legibility's sake). It shows the soundness of our cascading approach: when only a few samples are available, we provide a reward that lead to a better (and safer, notice the decreasing standard deviation) policy. When samples are widely available all approaches converge to the same value.

To illustrate a more challenging problem with an increased state space size, we coded a driving simulator inspired from a benchmark seen in \cite{syed2008apprenticeship,syed2008game}. The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels. The expert is trained to go as fast as possible (high reward) while avoiding collisions (harshly penalised) with randomly placed slower moving vehicles and avoiding going off-road (moderately penalised). Any other situation receives a null reward. Because the forward RL problem is too hard to solve with so little information, the algorithm from \cite{abbeel2004apprenticeship} was not able to solve the problem using only samples from the expert as ours did. We still used it as a baseline, allowing it to use features expectations and intermediates optimal policies (see \cite{abbeel2004apprenticeship} for the details of the algorithm) \emph{computed perfectly with the help of the model}.

We used features stemming from a discretisation of the state-action space. The performance is assessed with respect to the uniform distribution $\mathcal{U}$, which shows the generalization ability of each method).

As expected, pure classification is not sample efficient enough to provide a good control whereas our approach is able to generalize from very little data and exhibits performance on par with the \emph{fully-informed} state-of-the-art algorithm when given as few as 100 samples. This empirically shows the usefulness of the regression step. Furthermore, the state-of-the-art algorithm makes use of intermediate policies, which is computationally costly particularly when the state space gets big. Our approach computational cost scales with the number of samples we are given, which is as we demonstrated a small number. Experimentally, when producing the data for our graph, it took less time to run the cascading approach 50 times for 6 abscissae than to run \cite{abbeel2004apprenticeship}'s algorithm once.
\section{Related work}
\label{sec-6}

\label{section: related work}

First introduced in \cite{russell1998learning} and formalized in \cite{ng2000algorithms}, IRL has sometimes been seen as a way to apprenticeship learning \cite{abbeel2004apprenticeship} with algorithms outputing a policy matching a measure of the expert's distribution in the state space. See \cite{neu2009training} for a survey. Except for \cite{dvijotham2010inverse} which is suitable to linearly solvable MDPs only and for \cite{boularias2011relative} which still requires sampling trajectories from a non expert policy (our analysis requires only samples), all IRL algorithms repeatedly solve the forward RL problem, which is a costly procedure information-wise as well as computationally.

Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs. Building on the non trivial notion of metric in an MDP, \cite{melo2010learning} built a structured kernel for a classification algorithm.
\section{Conclusion}
\label{sec-7}

  The proposed approach of cascading two supervised learning algorithms (a classifier and a regressor) is able to learn a reward function from observed behaviors for which the expert is provably near-optimal. Going further than the analysis, experiments showed the approach (with the help of a simple heuristics) to be sample-efficient enough to work with samples from the expert only, giving better results and being more computationally efficient than a state-of-the-art algorithm. Our approach will be confronted to real world robotics problems in a near future.

\bibliographystyle{plain}
\bibliography{Biblio}

\end{document}

### 2012-09-13 klein2012structured_RIA/scirl_ria.tex
Squelette au style RIA

%\newcommand{\filespath}{../../Style/}%  JANCL path

\documentclass[english,utf8]{./hermes-journal}
\usepackage[utf8]{inputenc}
% \journalName{year}{volume}{issue}
\ria{2012}{22}{1}

\firstpagenumber{9}

% \title[short header title]{title}
\title{Classification structurée pour l'apprentissage par renforcement}
\author[1]{first name 1}{last name 1}
\author[2]{first name 2}{last name 2}

% addresses are automatically numbered
\address{...\\...\\
         ..., country}
        {e-mail 1}
\address{...\\...\\
         ..., country}
        {e-mail 1}

\abstract{}

\keywords{Apprentissage par Renforcement, Apprentissage par Renforcement Inverse}

\begin{document}

\maketitle

\newpage
\cite{Ng:2000}

\section{Introduction}

\acknowledgements{...}

\bibliography{scirl_ria.bib}

\end{document}

### 2012-09-13 klein2012structured_RIA/scirl_ria.tex
Mots clefs et Abstract

@@
 % \title[short header title]{title}
-\title{Classification structurée pour l'apprentissage par renforcement}
-\author[1]{first name 1}{last name 1}
-\author[2]{first name 2}{last name 2}
+\title[SCIRL]{Classification structurée pour l'apprentissage par renforcement}
+\author[1,2]{Edouard}{Klein}
+\author[2]{Bilal}{Piot}
+\author[2,3]{Matthieu}{Geist}
+\author[2,3]{Olivier}{Pietquin}
 
 % addresses are automatically numbered
-\address{...\\...\\
-         ..., country}
-        {e-mail 1}
-\address{...\\...\\
-         ..., country}
-        {e-mail 1}
+\address{LORIA -- team ABC\\
+Nancy, France
+}{}
+\address{Supélec -- IMS-MaLIS Research Group\\
+Metz, France}
+        {prenom.nom@supelec.fr}
+\address{ UMI 2958 (GeorgiaTech-CNRS)\\
+Metz, France}{}
 
-\abstract{}
+\abstract{  This paper adresses the inverse reinforcement learning (IRL) problem, that
+  is inferring a reward for which a demonstrated expert behavior is
+  optimal. We introduce a new algorithm, SCIRL, whose principle is to use
+  the so-called feature expectation of the expert as the
+  parameterization of the score function of a multi-class
+  classifier. This approach produces a reward function for which the
+  expert policy is provably near-optimal. Contrary to most of
+  existing IRL algorithms, SCIRL does not require solving the
+  direct RL problem. Moreover, with an appropriate heuristic, it can succeed with only trajectories sampled
+  according to the expert behavior. This is illustrated on a car
+  driving simulator.}
+\resume{Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse (\emph{ARI}).
+Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
+L'ARI part du postulat que l'expert optimise avec succès une fonction de récompense ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
+Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
+Notre contribution consiste en un nouvel algorithme d'ARI levant l'ensemble de ces contraintes.
+En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et surtout ne nécessitant pas la résolution du problème direct.}
 
-\keywords{Apprentissage par Renforcement, Apprentissage par Renforcement Inverse}
+\keywords{Reinforcement Learning, Inverse Reinforcement Learning}
+\motscles{Apprentissage par Renforcement, Apprentissage par Renforcement Inverse}
 

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Réglé un problème particulièrement têtu avec les figures

@@
 \firstpagenumber{9}
+\let\chapter\section
+\usepackage[ruled,algo2e]{algorithm2e}
+\usepackage{mathtools}
+\usepackage{subfigure}
+\mathtoolsset{showonlyrefs=true}
+
+\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
+\newcommand{\argmin}{\operatorname*{argmin}}
+\newcommand{\diag}{\operatorname*{diag}}
+
+\newcommand{\p}{\mathcal{P}}
+\newcommand{\R}{\mathcal{R}}
+\newcommand{\s}{\mathcal{S}}
+\newcommand{\A}{\mathcal{A}}
+\newcommand{\X}{\mathcal{X}}
+\newcommand{\Y}{\mathcal{Y}}
+\newcommand{\D}{\mathcal{D}}
+\newcommand{\T}{\mathcal{T}}
+\newcommand{\lc}{\mathcal{L}}
+\newcommand{\E}{\mathbb{E}}
+\newcommand{\prob}{\mathbb{P}}
+\newcommand{\Mu}{\boldsymbol{\mu}}
+%\newcommand{\Xi}{\boldsymbol{\xi}}
+
+%\newtheorem{theorem}{Theorem}
 
@@
 \begin{document}
-
+\newtheorem{prop}[theorem]{Proposition}
+\cite{Ng:2000}
 \maketitle
@@
 \newpage
-\cite{Ng:2000}
 
-\section{Introduction}
+\section{Background and Notations}
+\label{sec:background}
 
-\acknowledgements{...}
+\subsection{(Inverse) Reinforcement Learning}
+\label{subsec:background:irl}
 
+A Markov Decision process (MDP)~\cite{Puterman:1994} is a tuple
+$\{\s,\A,\p,\R,\gamma\}$ where $\s$ is the finite state
+space\footnote{This work can be extended to compact state spaces, up
+to some technical aspects.}, $\A$ the finite actions space, $\p =
+\{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ the set of
+Markovian transition probabilities, $\R\in\mathbb{R}^\s$ the
+state-dependent reward function and $\gamma$ the discount factor. A
+deterministic policy $\pi\in\s^\A$ defines the behavior of an agent.
+The quality of this control is quantified by the value function
+$v_\R^\pi\in\mathbb{R}^\s$, associating to each state the cumulative
+discounted reward for starting in this state and following the
+policy $\pi$ afterwards: $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t
+\R(S_t)|S_0=s,\pi]$. An optimal policy $\pi_\R^*$ (according to the
+reward function $\R$) is a policy of associated value function
+$v^*_\R$ satisfying $v_\R^* \geq v_\R^\pi$, for any policy $\pi$ and
+componentwise.
+
+Let $P_\pi$ be the stochastic matrix $P_\pi =
+(p(s'|s,\pi(s)))_{1\leq s,s'\leq |\s|}$. With a slight abuse of
+notation, we may write $a$ the policy which associates the action
+$a$ to each state $s$. The Bellman evaluation (resp. optimality)
+operators $T^\pi_\R\text{ (resp. $T^*_\R$)}:\mathbb{R}^\s
+\rightarrow \mathbb{R}^\s$ are defined as $T^\pi_\R v = \R + \gamma
+P_\pi v$ and $T_\R^*v = \max_\pi T_\R^\pi v$.
+%\begin{equation}
+%  T^\pi_\R v = \R + \gamma P_\pi v \text{ and } T_\R^*v = \max_\pi T_\R^\pi v.
+%\end{equation}
+These operators are contractions and $v_\R^\pi$ and $v^*_\R$ are
+their respective fixed-points: $v_\R^\pi = T^\pi_\R v_\R^\pi$ and
+$v^*_\R = T^*_\R v^*_\R$. The action-value function
+$Q^\pi\in\mathbb{R}^{\s\times \A}$ adds a degree of freedom on the
+choice of the first action, it is formally defined as $Q_\R^\pi(s,a)
+= [T^a_\R v^\pi_\R](s)$. We also write $\rho_\pi$ the stationary
+distribution of the policy $\pi$ (satisfying $\rho_\pi^\top P_\pi =
+\rho_\pi^\top$).
+
+Reinforcement learning and approximate dynamic programming aim at
+estimating the optimal control policy $\pi^*_\R$ when the model
+(transition probabilities and the reward function) is unknown (but
+observed through interactions with the system to be controlled) and
+when the state space is too large to allow exact representations of
+the objects of interest (as value functions or
+policies)~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
+%
+We refer to this as the direct problem.
+%
+On the contrary, (approximate) inverse reinforcement
+learning~\cite{Ng:2000} aim at estimating a reward function for
+which an observed policy is (nearly) optimal.
+%
+Let us call this policy the expert policy, denoted $\pi_E$. We may
+assume that it optimizes some unknown reward function $\R_E$. The
+aim of IRL is to compute some reward $\hat{\R}$ such that the expert
+policy is (close to be) optimal, that is such that $v^*_{\hat{\R}}
+\approx v^{\pi_E}_{\hat{\R}}$.
+%
+We refer to this as the inverse problem.
+
+Similarly to the direct problem, the state space may be too large
+for the reward function to admit a practical exact representation.
+Therefore, we restrict our search of a good reward among linearly
+parameterized functions. Let $\phi(s) = (\phi_1(s)  \dots
+\phi_p(s))^\top$
+%\begin{pmatrix}
+%  \phi_1(s) & \dots & \phi_p(s)
+%\end{pmatrix}^\top$
+be a feature vector composed of $p$ basis function
+$\phi_i\in\mathbb{R}^\s$, we define the  parameterized reward
+functions as $\R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p
+\theta_i \phi_i(s)$.
+%\begin{equation}
+%  \R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p \theta_i \phi_i(s).
+%\end{equation}
+Searching a good reward thus reduces to searching a good parameter
+vector $\theta \in\mathbb{R}^p$. Notice that we will use
+interchangeably $\R_\theta$ and $\theta$ as subscripts
+(\textit{e.g.}, $v_\theta^\pi$ for $v_{\R_\theta}^\pi$).
+Parameterizing the reward this way implies a related
+parameterization for the action-value function:
+\begin{equation}
+  Q^\pi_\theta(s,a) = \theta^\top \mu^\pi(s,a) \text{ with }
+  \mu^\pi(s,a) = \E[\sum_{t\geq 0} \gamma^t
+  \phi(S_t)|S_0=s,A_0=a,\pi].
+  \label{eq:def:mu}
+\end{equation}
+Therefore, the action-value function shares the parameter vector of
+the reward function, with an associated feature vector $\mu^\pi$
+called the \emph{feature expectation}. This notion will be of
+primary importance for the contribution of this paper. Notice that
+each component $\mu_i^\pi$ of this feature vector is actually the
+action-value function of the policy $\pi$ assuming the reward is
+$\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$. Therefore, any
+algorithm designed for estimating an action-value function may be
+used to estimate the feature expectation, such as Monte-Carlo
+rollouts or temporal difference learning~\cite{Klein:2011}.
+
+\subsection{Classification with Linearly Parameterized Score
+Functions} \label{subsec:background:classif}
+
+Let $\X$ be a compact or a finite set (of inputs to be classified)
+and let $\Y$ be a finite set (of labels). Assume that inputs $x\in
+\X$  are drawn according to some unknown distribution $\prob(x)$ and
+that there exists some oracle which associates to each of these
+inputs a label $y\in Y$ drawn according to the unknown conditional
+distribution $\prob(y|x)$. Generally speaking, the goal of
+multi-class classification is, given a training set
+$\{(x_i,y_i)_{1\leq i \leq N}\}$ drawn according to $\prob(x,y)$, to
+produce a decision rule $g\in\Y^\X$ which aims at minimizing the
+classification error $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq
+y)$, where $\chi$ denotes the indicator function.
+
+Here, we consider a more restrictive set of classification
+algorithms. We assume that the decision rule associates to an input
+the argument which maximizes a related score function, this score
+function being linearly parameterized and the associated parameters
+being learnt by the algorithm. More formally, let $\psi(s,a) =
+(\psi_1(x,y)  \dots  \psi_d(x,y))^\top\in \mathbb{R}^d$
+%\begin{pmatrix}   \psi_1(x,y) & \dots & \psi_d(x,y)
+%\end{pmatrix}^\top \in \mathbb{R}^d$
+be a feature vector whose components are $d$ basis functions
+$\psi_i\in\mathbb{R}^{\X\times\Y}$. The linearly parameterized score
+function $s_w\in\mathbb{R}^{\X\times \Y}$ of parameter vector
+$w\in\mathbb{R}^d$ is defined as $s_w(x,y) = w^\top \psi(x,y)$. The
+associated decision rule $g_w\in{\Y^\X}$ is defined as $g_w(x) \in
+\argmax_{y\in\Y}s_w(x,y)$. Using a training set $\{(x_i,y_i)_{1\leq
+i\leq N}\}$, a linearly parameterized score function-based
+multi-class classification (MC$^2$ for short) algorithm computes a
+parameter vector $\theta_c$. The quality of the solution is
+quantified by the classification error $\epsilon_c =
+\prob(g_{\theta_c}(x)\neq y)$.
+
+We do not consider a specific MC$^2$ algorithm, as long as it
+classifies inputs by maximizing the argument of a linearly
+parameterized score function. For example, one may choose a
+multi-class support vector machine~\cite{Guermeur:2007} (taking the
+kernel induced by the feature vector) or a structured large margin
+approach~\cite{Taskar:2005}. Other choices may be possible, one can
+choose its preferred algorithm.
+
+\section{Structured Classification for Inverse Reinforcement
+Learning} \label{sec:scirl}
+
+\subsection{General Algorithm}
+\label{subsec:scirl:algo}
+
+Consider the classification framework of
+Sec.~\ref{subsec:background:classif}. The input $x$ may be seen as a
+state and the label $y$ as an action. Then, the decision rule
+$g_w(x)$ can be interpreted as a policy which is greedy according to
+the score function $w^\top \psi(x,y)$, which may itself be seen as
+an action-value function. Making the parallel with
+Eq.~\eqref{eq:def:mu}, if $\psi(x,y)$ is the feature expectation of
+some policy $\pi$ which produces labels of the training set, and if
+the classification error is small, then $w$ will be the parameter
+vector of a reward function for which we may hope the policy $\pi$
+to be near optimal. Based on these remarks, we're ready to present
+the proposed Structured Classification-based IRL (SCIRL) algorithm.
+
+Let $\pi_E$ be the expert policy from which we would like to recover
+a reward function. Assume that we have a training set $\D =
+\{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ where states are sampled
+according to the expert stationary distribution\footnote{For
+example, if the Markov chain induced by the expert policy is
+fast-mixing, sampling a trajectory will quickly lead to sample
+states according to this distribution.} $\rho_E = \rho_{\pi_E}$.
+Assume also that we have an estimate $\hat{\mu}^{\pi_E}$ of the
+expert feature expectation $\mu^{\pi_E}$ defined in
+Eq.~\eqref{eq:def:mu}. How to practically estimate this quantity is
+postponed to Sec.~\ref{subsec:practicalApproach:muE}; however,
+recall that estimating $\mu^{\pi_E}$ is simply a \emph{policy
+evaluation} problem (estimating the action-value function of a given
+policy), as noted in Sec.~\ref{subsec:background:irl}. Assume also
+that an MC$^2$ algorithm has been chosen. The proposed algorithm
+simply consists in choosing $\theta^\top\hat{\mu}^{\pi_E}(s,a)$ as
+the linearly parameterized score function, training the classifier
+on $\D$ which produces a parameter vector $\theta_c$, and outputting
+the reward function $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.
+
+\begin{algorithm2e}%[tbh]
+    %\small
+  \SetAlgoVlined
+  \caption{SCIRL algorithm}
+  \label{algo:scirl}
+  %
+  \BlankLine
+  \emph{\textbf{Given}} a training set $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
+  an estimate $\hat{\mu}^{\pi_E}$ of the expert feature expectation $\mu^{\pi_E}$ and an MC$^2$ algorithm\;
+  %
+  \BlankLine
+  \emph{\textbf{Compute}} the parameter vector $\theta_c$ using the
+  MC$^2$ algorithm
+  fed with the training set $\D$ and considering the parameterized score function
+  $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
+  %
+  \BlankLine
+  \emph{\textbf{Output}} the reward function $\R_{\theta_c}(s) = \theta_c^\top\phi(s)$ \;
+\end{algorithm2e}
+
+The proposed approach is summarized in Alg.~\ref{algo:scirl}. We
+call this Structured Classification-based IRL because using the
+(estimated) expert feature expectation as the feature vector for the
+classifier somehow implies taking into account the MDP structure
+into the classification problem and allows outputting a reward
+vector. Notice that contrary to most of existing IRL
+%and apprenticeship learning
+algorithms, SCIRL does not require solving the direct problem. If it
+possibly requires estimating the expert feature expectation, it is
+just a policy evaluation problem, less difficult than the policy
+optimization issue involved by the direct problem. This is further
+discussed in Sec.~\ref{sec:relatedWorks}.
+%
+%Now, we formally show that the proposed approach makes sens, as long
+%as $\mu^{\pi_E}$ is well estimated and as long as the classification
+%error is small.
+
+\subsection{Analysis}
+\label{subsec:scirl:analysis}
+
+In this section, we show that the expert policy $\pi_E$ is close to
+be optimal according to the reward function $\R_{\theta_c}$, more
+precisely that
+$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ is
+small. Before stating our main result, we need to introduce some
+notations and to define some objects.
+
+We will use the first order discounted future state distribution
+concentration coefficient $C_f$~\cite{Munos:2007}:
+\begin{equation}
+  C_f = (1-\gamma)\sum_{t\geq 0} \gamma^t c(t) \text{ with } c(t) =
+  \max_{\pi_1,\dots,\pi_t,s\in\s}\frac{(\rho_E^\top P_{\pi_1}\dots
+  P_{\pi_t})(s)}{\rho_E(s)}.
+\end{equation}
+We note $\pi_c$ the decision rule of the classifier: $\pi_c(s) \in
+\argmax_{a\in \A} \theta_c^\top\hat{\mu}^{\pi_E}(s,a)$. The
+classification error is therefore $\epsilon_c =
+\E_{s\sim\rho_E}[\chi_{\{\pi_c(s)\neq\pi_E(s)\}}] \in [0,1]$. We
+write $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$
+the score function computed from the training set $\D$ (which can be
+interpreted as an approximate action-value function). Let also
+$\epsilon_{\mu} = \hat{\mu}^{\pi_E} - \mu^{\pi_E}:\s\times\A
+\rightarrow  \mathbb{R}^p$ be the feature expectation error.
+Consequently, we define the action-value function error as
+$\epsilon_Q = \hat{Q}^{\pi_E}_{\theta_c} - Q^{\pi_E}_{\theta_c} =
+\theta_c^\top(\hat{\mu}^{\pi_E} - \mu^{\pi_E}) = \theta_c^\top
+\epsilon_\mu:\s\times\A\rightarrow\mathbb{R}$. We finally define the
+mean delta-max action-value function error as $\bar{\epsilon}_Q =
+\E_{s\sim\rho_E}[\max_{a\in\A}\epsilon_Q(s,a) -
+\min_{a\in\A}\epsilon_Q(s,a)]\geq 0$.
+
+\begin{theorem}
+  \label{th}
+  Let $\R_{\theta_c}$ be the reward function outputted by
+  Alg.~\ref{algo:scirl}. Let also the quantities $C_f$, $\epsilon_c$
+  and $\bar{\epsilon}_Q$ be defined as above. We have
+  %
+%  the concentration coefficient $C_f$,
+%  the classification error $\epsilon_c$ and the mean-max
+%  action-value function error $\bar{\epsilon}_Q$ be defined as
+%  above. Then, we have
+  \begin{equation}
+    0\leq
+    \E_{s\sim\rho_E}[v^*_{\R_{\theta_c}}-v^{\pi_E}_{\R_{\theta_c}}]
+    \leq \frac{C_f}{1-\gamma}\left(\bar{\epsilon}_Q +
+    \epsilon_c\frac{2\gamma\|\R_{\theta_c}\|_\infty}{1-\gamma}
+%    \epsilon_c\frac{2\gamma}{1-\gamma}\min(\|\R_{\theta_c}\|_\infty,
+%    C_p \E_{s\sim\rho_E}[\R_{\theta_c}(s)])
+    \right).
+  \end{equation}
+\end{theorem}
+\begin{proof}
+  As the proof only relies on the reward $\R_{\theta_c}$, we
+  omit the related subscripts to keep the notations simple
+  (\textit{e.g.}, $v^\pi$ for
+  $v^\pi_{\theta_c}=v^\pi_{\R_{\theta_c}}$ or $\R$ for $\R_{\theta_c}$). First, we link
+  the error $\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)]$ to the
+  Bellman residual $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s)-v^{\pi_E}(s)]$.
+  %
+  % A second step consists in bounding this residual.
+  %
+  Componentwise, we have that:
+  \begin{align}
+    v^* - v^{\pi_E} &= T^* v^*  - T^{\pi^*}v^{\pi_E} +
+    T^{\pi^*}v^{\pi_E} - T^* v^{\pi_E} + T^* v^{\pi_E} - v^{\pi_E}
+    \\
+    &\stackrel{(a)}{\leq} \gamma P_{\pi^*}(v^*-v^{\pi_E}) + T^*
+    v^{\pi_E} - v^{\pi_E}
+    %
+    \stackrel{(b)}{\leq} (I-\gamma
+    P_{\pi^*})^{-1} (T^* v^{\pi_E} - v^{\pi_E}).
+  \end{align}
+  Inequality $(a)$ holds because $T^{\pi^*} v^{\pi_E}\leq T^*
+  v^{\pi_E}$ and inequality $(b)$ holds thanks
+  to~\cite[Lemma~4.2]{Munos:2007}. Moreover, $v^*$ being optimal we
+  have that $v^*-v^{\pi_E}\geq 0$ and $T^*$ being the Bellman
+  optimality operator, we have $T^* v^{\pi_E}\geq
+  T^{\pi_E}v^{\pi_E}=v^{\pi_E}$. Additionally, remark that
+  $(I-\gamma P_{\pi^*})^{-1} = \sum_{t\geq 0}\gamma^t P_{\pi^*}^t$.
+  Therefore, using the definition of the concentration coefficient
+  $C_f$, we have that:
+  \begin{equation}
+    0\leq\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)] \leq \frac{C_f}{1-\gamma}
+    \E_{s\sim\rho_E}\left[[T^*v^{\pi_E}](s) - v^{\pi_E}(s)\right].
+    \label{eq:proof:residual}
+  \end{equation}
+  This results actually follows closely the one
+  of~\cite[Theorem~4.2]{Munos:2007}. There remains to bound the
+  Bellman residual $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
+  v^{\pi_E}(s)]$. Considering the following decomposition,
+  \begin{equation}
+    T^* v^{\pi_E} - v^{\pi_E} = T^* v^{\pi_E} - T^{\pi_c}v^{\pi_E}
+    + T^{\pi_c}v^{\pi_E}- v^{\pi_E},
+    \label{eq:proof:decomposition}
+  \end{equation}
+  we will bound $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
+  and $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]$.
+
+  The policy $\pi_c$ (the decision rule of the classifier) is greedy
+  with respect to
+  $\hat{Q}^{\pi_E}=\theta_c^\top\hat{\mu}^{\pi_E}$. Therefore, for any state-action couple
+  $(s,a)\in\s\times \A$ we have:
+  \begin{equation}
+    \hat{Q}^{\pi_E}(s,\pi_c(s))\geq
+    \hat{Q}^{\pi_E}(s,a)
+    \Leftrightarrow
+    Q^{\pi_E}(s,a) \leq Q^{\pi_E}(s,\pi_c(s)) +
+    \epsilon_Q(s,\pi_c(s)) - \epsilon_Q(s,a).
+  \end{equation}
+  By definition, $Q^{\pi_E}(s,a) = [T^a v^{\pi_E}](s)$ and
+  $Q^{\pi_E}(s,\pi_c(s)) = [T^{\pi_c} v^{\pi_E}](s)$. Therefore, for $s\in\s$:
+  \begin{align}
+    \forall a\in A,\; [T^a v^{\pi_E}](s) &\leq [T^{\pi_c}
+    v^{\pi_E}](s) + \epsilon_Q(s,\pi_c(s))-\epsilon_Q(s,a)
+    \\
+    \Rightarrow [T^* v^{\pi_E}](s) &\leq [T^{\pi_c}
+    v^{\pi_E}](s) + \max_{a\in \A}\epsilon_Q(s,a)-\min_{a\in
+    \A}\epsilon_Q(s,a).
+  \end{align}
+  Taking the expectation according to $\rho_E$ and noticing that
+  $T^* v^{\pi_E}\geq v^{\pi_E}$, we bound the first
+  term:
+  \begin{equation}
+    0 \leq \E_{s\sim\rho_E}\left[ [T^* v^{\pi_E}](s) - [T^{\pi_c}
+    v^{\pi_E}](s)\right] \leq \bar{\epsilon}_Q.
+    \label{eq:proof:b1}
+  \end{equation}
+  There finally remains to bound the term $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) -
+  v^{\pi_E}(s)]$.
+
+  Let us write $M\in\mathbb{R}^{|\s|\times |\s|}$ the diagonal matrix
+  defined as $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$. Using
+  this, the Bellman operator $T^{\pi_c}$ may be written as, for any $v\in\mathbb{R}^\s$:
+  \begin{equation}
+    T^{\pi_c}v = \R + \gamma M P_{\pi_c} v + \gamma (I-M)P_{\pi_E} v
+    = \R + \gamma P_{\pi_E} v + \gamma M (P_{\pi_c}-P_{\pi_E})v.
+  \end{equation}
+  Applying this operator to $v^{\pi_E}$ and recalling that $\R +
+  \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, we
+  get:
+  \begin{equation}
+    T^{\pi_c}v^{\pi_E} - v^{\pi_E} = \gamma M
+    (P_{\pi_c}-P_{\pi_E})v^{\pi_E}
+    \Rightarrow |\rho_E^\top (T^{\pi_c}v^{\pi_E} - v^{\pi_E})| = \gamma
+    |\rho_E^\top M (P_{\pi_c}-P_{\pi_E})v^{\pi_E}|.
+  \end{equation}
+  One can easily see that $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
+  \leq \frac{2}{1-\gamma}\|\R\|_\infty$, which allows bounding the
+  last term:
+  \begin{equation}
+    |\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]| \leq
+    \epsilon_c \frac{2\gamma}{1-\gamma} \|\R\|_\infty.
+    \label{eq:proof:b2}
+  \end{equation}
+  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
+  and~\eqref{eq:proof:b2} into Eq.~\eqref{eq:proof:residual} gives the
+  stated result.%, which concludes the proof.
+%  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
+%  and~\eqref{eq:proof:b2} into Eq.~\eqref{eq:proof:residual} using the
+%  decomposition of Eq.~\eqref{eq:proof:decomposition} gives the
+%  stated result:
+%  \begin{equation}
+%    0\leq\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)] \leq
+%    \frac{C_f}{1-\gamma} \left(\bar{\epsilon}_Q + \epsilon_c \frac{2\gamma}{1-\gamma} \|\R\|_\infty
+%    \right).
+%  \end{equation}
+%  This concludes the proof.
+\end{proof}
+
+This result shows that if the expert feature expectation is well
+estimated (in the sense that the estimation error $\epsilon_\mu$ is
+small for states sampled according to the expert stationary policy
+and for all actions) and if the classification error $\epsilon_c$ is
+small, then the proposed generic algorithm outputs a reward function
+$\R_{\theta_c}$ for which the expert policy will be near optimal. A
+direct corollary of Th.~\ref{th} is that given the true expert
+feature expectation $\mu^{\pi_E}$ and a perfect classifier
+($\epsilon_c=0$), $\pi_E$ is the unique optimal policy for
+$\R_{\theta_c}$.
+
+One may argue that this bounds trivially holds for the null reward
+function (a reward often exhibited to show that IRL is an ill-posed
+problem), obtained if $\theta_c=0$. However, recall that the
+parameter vector $\theta_c$ is computed by the classifier. With
+$\theta_c=0$, the decision rule would be a random policy and we
+would have $\epsilon_c = \frac{|\A|-1}{|\A|}$, the worst possible
+classification error. This case is really unlikely. Therefore, we
+advocate that the proposed approach somehow allows disambiguating
+the IRL problem (at least, it does not output trivial reward
+functions such as the null vector).
+%
+Also, this bound is scale-invariant: one could impose
+$\|\theta_c\|=1$ or normalize (action-) value functions by
+$\|\R_{\theta_c}\|_\infty^{-1}$.
+
+One should notice that there is a hidden dependency of the
+classification error $\epsilon_c$ to the estimated expert feature
+expectation $\hat{\mu}^{\pi_E}$. Indeed, the minimum classification
+error depends on the hypothesis space spanned by the chosen score
+function basis functions for the MC$^2$ algorithm (here
+$\hat{\mu}^{\pi_E}$). Nevertheless, provided a good representation
+for the reward function (that is a good choice of basis functions
+$\phi_i$) and a small estimation error, this should not be a
+practical problem.
+
+Finally, if our bound relies on the generalization errors
+$\epsilon_c$ and $\bar{\epsilon}_Q$, the classifier will only use
+$(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ in the training
+phase, where $s_i$ are the states from the set $\D$. It outputs
+$\theta_c$, seen as a reward function, thus the estimated feature
+expectation $\hat{\mu}^{\pi_E}$ is no longer required. Therefore,
+practically it should be sufficient to estimate well
+$\hat{\mu}^{\pi_E}$ on state-action couples $(s_i,a)_{1\leq i\leq
+N,a\in\A}$, which allows envisioning Monte-Carlo rollouts for
+example.
+
+\section{A Practical Approach}
+\label{sec:practicalApproach}
+
+\subsection{Estimating the Expert Feature Expectation}
+\label{subsec:practicalApproach:muE}
+
+SCIRL relies on an estimate $\hat{\mu}^{\pi_E}$ of the expert
+feature expectation. Basically, this is a policy evaluation problem.
+An already made key observation is that each component of
+$\mu^{\pi_E}$ is the action-value function of $\pi_E$ for a reward
+function $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
+[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. We briefly review its exact
+computation and possible estimation approaches, and consider
+possible heuristics.
+%possibly considering some heuristics.
+
+If the model is known, the feature expectation can be computed
+explicitly. Let $\Phi\in\mathbb{R}^{|\s|\times p}$ be the feature
+matrix whose rows contain the feature vectors $\phi(s)^\top$ for all
+$s\in\s$.
+%
+%(equivalently, whose columns contain the basis functions $\phi_i$
+%for $1\leq i\leq p$).
+%
+For a fixed $a\in A$, let $\Mu^{\pi_E}_a \in\mathbb{R}^{|\s|\times
+p}$ be the feature expectation matrix whose rows are the expert
+feature vectors, that is $(\mu^{\pi_E}(s,a))^\top$ for any $s\in\s$.
+With these notations, we have $\Mu_a^{\pi_E} = \Phi + \gamma
+P_a(I-\gamma P_{\pi_E})^{-1} \Phi$.
+%\begin{equation}
+%  \forall a\in\A,\; \Mu_a^{\pi_E} = \Phi + \gamma P_a(I-\gamma
+%  P_{\pi_E})^{-1} \Phi.
+%\end{equation}
+%Therefore, knowing the model (the dynamic), the expert feature
+%expectation can be easily computed.
+Moreover, the related computational cost is the same order of
+magnitude as evaluating a single policy (as the costly part,
+computing $(I-\gamma P_{\pi_E})^{-1}$, is shared by all components).
+
+If the model is unknown, any temporal difference learning algorithm
+can be used to estimate the expert feature
+expectation~\cite{Klein:2011}, as LSTD (Least-Squares Temporal
+Differences)~\cite{Bradtke:1996}. Let $\psi:\s\times \A \rightarrow
+\mathbb{R}^d$ be a feature vector composed of $d$ basis functions
+$\psi_i \in\mathbb{R}^{\s\times \A}$. Each component $\mu_i^{\pi_E}$
+of the expert feature expectation is parameterized by a vector
+$\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
+\psi(s,a)$. Assume that we have a training set
+$\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ with actions
+$a_i$ not necessarily sampled according to policy $\pi_E$
+(\textit{e.g.}, this may be obtained by sampling trajectories
+according to an expert-based $\epsilon$-greedy policy), the aim
+being to have a better variability of tuples (non-expert actions
+should be tried). Let $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
+(resp. $\tilde{\Psi}'$) be the feature matrix whose rows are the
+feature vectors $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
+Let also $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ be the feature
+matrix whose rows are the reward's feature vectors $\phi(s_i)^\top$.
+Finally, let $\Xi =
+\begin{bmatrix}   \xi_1 & \dots & \xi_p
+\end{bmatrix}\in\mathbb{R}^{d\times p}$ be the matrix of all
+parameter vectors. Applying LSTD to each component of the feature
+expectation gives the LSTD-$\mu$ algorithm~\cite{Klein:2011}: $\Xi =
+(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
+\tilde{\Psi}'))^{-1}\tilde{\Psi}^\top \tilde{\Phi}$ and
+$\hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a)$.
+%\begin{equation}
+%  \Xi = \left(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
+%  \tilde{\Psi}')\right)^{-1}\tilde{\Psi}^\top \tilde{\Phi}
+%  \text{ and } \hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a).
+%\end{equation}
+As for the exact case, the costly part (computing the inverse
+matrix) is shared by all feature expectation components, the
+computational cost is reasonable (same order as LSTD).
+
+%
+Provided a simulator and the ability to sample according to the
+expert policy, the expert feature expectation may also be estimated
+using Monte-Carlo rollouts for a given state-action pair (as noted
+in Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ need only
+be known on $(s_i,a)_{1\leq i\leq N,a\in\A}$). Assuming that $K$
+trajectories are sampled for each required state-action pair, this
+method would require $KN|\A|$ rollouts.
+
+In order to have a small error $\bar{\epsilon}_Q$, one may learn
+using transitions whose starting state is sampled according to
+$\rho_E$ and whose actions are uniformly distributed. However, it
+may happen that only transitions of the expert are available: $\T =
+\{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$. If the state-action
+couples $(s_i,a_i)$ may be used to feed the classifier, the
+transitions $(s_i,a_i,s'_i)$ are not enough to provide an accurate
+estimate of the feature expectation. In this case, we can still
+expect an accurate estimate of $\mu^{\pi_E}(s,\pi_E(s))$, but there
+is little hope for $\mu^{\pi_E}(s,a\neq\pi_E(s))$. However, one can
+still rely on some heuristic; this does not fit the analysis of
+Sec.~\ref{subsec:scirl:analysis}, but it can still provide good
+experimental results, as illustrated in Sec.~\ref{sec:experiments}.
+
+We propose such a heuristic. Assume that only data $\T$ is available
+and that we use it to provide an (accurate) estimate
+$\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (this basically means estimating a
+value function instead of an action-value function as described
+above). We may adopt an optimistic point of view by assuming that
+applying a non-expert action just delays the effect of the expert
+action. More formally, we associate to each state $s$ a virtual
+state $s_\text{v}$ for which $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ for
+any action $a$ and for which the reward feature expectation is the
+null vector, $\phi(s_v) = 0$. In this case, we have
+$\mu^{\pi_E}(s,a\neq\pi_E(s)) = \gamma \mu^{\pi_E}(s,\pi_E(s))$.
+Applying this idea to the available estimate (recalling that the
+classifiers only requires evaluating $\hat{\mu}^{\pi_E}$ on
+$(s_i,a)_{1\leq i\leq N,a\in \A}$) provides the proposed heuristic:
+for $1\leq i\leq N$, $\hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
+\hat{\mu}^{\pi_E}(s_i,a_i)$.
+%\begin{equation}
+%  \forall 1\leq i \leq N,\;
+%  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
+%  \hat{\mu}^{\pi_E}(s_i,a_i).
+%  \label{eq:heuristic}
+%\end{equation}
+
+We may even push this idea further, to get the simpler estimate of
+the expert feature expectation (but with the weakest guarantees).
+Assume that the set $\T$ consists of one long trajectory, that is
+$s'_i = s_{i+1}$ (thus $\T =
+\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). We may estimate
+$\mu^{\pi_E}(s_i,a_i)$ using the single rollout available in the
+training set and use the proposed heuristic for other actions:
+\begin{equation}
+  \forall 1\leq i \leq N,\; \hat{\mu}^{\pi_E}(s_i,a_i) =
+  \sum_{j=i}^N \gamma^{j-i}\phi(s_j) \text{ and }
+  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
+  \hat{\mu}^{\pi_E}(s_i,a_i).
+  \label{eq:mc_plus_heuristic}
+\end{equation}
+
+To sum up, the expert feature expectation may be seen as a vector of
+action-value functions (for the same policy $\pi_E$ and different
+reward functions $\phi_i$). Consequently, any action-value function
+evaluation algorithm may be used to estimate $\mu^\pi(s,a)$.
+Depending on the available data, one may have to rely on some
+heuristic to assess the feature expectation for a unexperienced
+(non-expert) action. Also, this expert feature expectation estimate
+is only required for training the classifier, so it is sufficient to
+estimate on state-action couples $(s_i,a)_{1\leq i \leq N,a\in \A}$.
+In any case, estimating $\mu^{\pi_E}$ is not harder than estimating
+the action-value function of a given policy in the on-policy case,
+which is much easier than computing an optimal policy for an
+arbitrary reward function (as required by most of existing IRL
+algorithms, see Sec.~\ref{sec:relatedWorks}).
+
+\subsection{An Instantiation}
+\label{subsec:practicalApproach:instantiation}
+
+As stated before, any MC$^2$ algorithm may be used. Here, we choose
+the structured large margin approach~\cite{Taskar:2005}. Let
+$\lc:\s\times\A\rightarrow\mathbb{R}_+$ be a user-defined margin
+function satisfying $\lc(s,\pi_E(s))\leq \lc(s,a)$ (here,
+$\lc(s_i,a_i)=0$ and $\lc(s_i,a\neq a_i)=1$). The MC$^2$ algorithm
+solves:% the following quadratic problem:
+\begin{equation}
+  \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
+  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~~~~s.t.~~~~} \forall i,
+  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
+  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a). \label{eq:qp_taskar}
+\end{equation}
+Following~\cite{Ratliff:2006}, we express the equivalent hinge-loss
+form (noting that the slack variables $\zeta_i$ are tight, which
+allows moving the constraints in the objective function):
+\begin{equation}
+  J(\theta) = \frac{1}{N}\sum_{i=1}^N \max_a \theta^\top
+  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a) -
+  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i) +
+  \frac{\lambda}{2}\|\theta\|^2.
+\end{equation}
+This objective function is minimized using a subgradient descent.
+The expert feature expectation is estimated using the scheme
+described in Eq.~\eqref{eq:mc_plus_heuristic}.
+
+%This is the instantiation we use in Sec.~\ref{sec:experiments}, but
+%our approach is more general: other choices could have been done for
+%the MC$^2$ algorithm as well as for the expert feature expectation
+%estimation. Notice that if the model is known, $\mu^{\pi_E}$ can be
+%exactly computed ($\bar{\epsilon}_Q=0$). Assuming that the reward
+%feature vector is rich enough (in the sense that there exists a
+%parameter vector such that $\pi_E$ is the unique optimal policy),
+%constraints of Eq.~\eqref{eq:qp_taskar} can be satisfied for all
+%state-action couples. This means that the proposed approach is able
+%to produce a non-trival reward for which the expert policy is the
+%unique optimal policy, and this without solving any direct problem.
+
+\section{Related Works}
+\label{sec:relatedWorks}
+
+The notion of IRL has first been introduced in~\cite{Russell:1998}
+and first been formalized in~\cite{Ng:2000}. A classic approach to
+IRL, initiated in~\cite{Abbeel:2004}, consists in finding a policy
+(through some reward function) such that its feature expectation (or
+more generally some measure of the underlying trajectories'
+distribution) matches the one of the expert policy.
+%
+See~\cite{Neu:2010} for a review.
+%
+%This can be done based on game theory~\cite{Syed:2008:game}, linear
+%programming~\cite{Syed:2008:lp}, maximum
+%entropy~\cite{Ziebart:2008}, Bayesian posterior
+%maximization~\cite{Ramachandran:2007}, \textit{etc}. Some of these
+%approaches and others are nicely reviewed in~\cite{Neu:2010}.
+%
+Notice that related algorithms are not always able to output a
+reward function, even if they may make use of IRL as an intermediate
+step. In such case, they are usually refereed to as apprenticeship
+learning algorithms.
+%
+%Notice that related algorithms do not always output a reward
+%function (\textit{e.g.}, in~\cite{Abbeel:2004}, the algorithm
+%produces a mixed policy; each intermediate policy is optimal
+%according to some reward function, but the outputted policy does not
+%correspond to a specific reward). In such case, they are refereed to
+%as ``apprenticeship learning'' algorithms (they may make use of IRL
+%as intermediate steps but do not output a reward).
+
+Closer to our contribution, some approaches also somehow introduce a
+structure in a classification
+procedure~\cite{Melo:2010}\cite{Ratliff:2006}. In~\cite{Melo:2010},
+a metric induced by the MDP is used to build a kernel which is used
+in a classification algorithm, showing improvements compared to a
+non-structured kernel. However, this approach is not an IRL
+algorithm, and more important assessing the metric of an MDP is a
+quite involved problem. In~\cite{Ratliff:2006}, a classification
+algorithm is also used to produce a reward function. However,
+instead of associating actions to states, as we do, it associates
+optimal policies (labels) to MDPs (inputs), which is how the
+structure is incorporated. This involves solving many MDPs.
+
+As far as we know, all IRL algorithms require solving the direct RL
+problem repeatedly, except~\cite{Dvij:2010,boularias:2011}.
+\cite{Dvij:2010} applies to linearly-solvable MDPs (where the
+control is done by imposing any dynamic to the system).
+In~\cite{boularias:2011}, based on a relative entropy argument, some
+utility function is maximized using a subgradient ascent. Estimating
+the subgradient requires sampling trajectories according to the
+policy being optimal for the current estimated reward. This is
+avoided thanks to the use of importance sampling. Still, this
+requires sampling trajectories according to a non-expert policy and
+the direct problem remains at the core of the approach (even if
+solving it is avoided).
+
+SCIRL does not require solving the direct problem, just estimating
+the feature expectation of the expert policy. In other words,
+instead of solving multiple policy optimization problems, we only
+solve one policy evaluation problem. This comes with theoretical
+guarantees (which is not the case of all IRL algorithms,
+\textit{e.g.}~\cite{boularias:2011}). Moreover, using heuristics
+which go beyond our analysis, SCIRL may rely solely on data provided
+by expert trajectories. We demonstrate this empirically in the next
+section. To the best of our knowledge, no other IRL algorithm can
+work in such a restrictive case.
+
+\section{Experiments}
+\label{sec:experiments}
+
+We illustrate the proposed approach on a car driving simulator,
+similar to~\cite{Abbeel:2004,Syed:2008:game}. The goal si to drive a
+car on a busy three-lane highway with randomly generated traffic
+(driving off-road is allowed on both sides). The car can move left
+and right, accelerate, decelerate and keep a constant speed. The
+expert optimizes a handcrafted reward $\R_E$ which favours speed,
+punish off-road, punish collisions even more  and is neutral
+otherwise.
+
+We compare SCIRL as instantiated in
+Sec.~\ref{subsec:practicalApproach:instantiation} to the
+unstructured classifier (using the same classification algorithm)
+and to the algorithm of~\cite{Abbeel:2004} (called here PIRL for
+Projection IRL). We also consider the optimal behavior according to
+a randomly sampled reward function as a baseline (using the same
+reward feature vector as SCIRL and PIRL, the associated parameter
+vector is randomly sampled).
+
+For SCIRL and PIRL we use a discretization of the state space as the
+reward feature vector, $\phi\in\mathbb{R}^{729}$: $9$ horizontal
+positions for the user's car, $3$ horizontal and $9$ vertical
+positions for the closest traffic's car and $3$ speeds. Notice that
+these features are much less informative than the ones used
+in~\cite{Abbeel:2004,Syed:2008:game}. Actually,
+in~\cite{Syed:2008:game} features are so informative that sampling a
+random positive parameter vector $\theta$ already gives an
+acceptable behavior. The discount factor is $\gamma = 0.9$. The
+classifier uses the same feature vector reproduced for each action.
+
+SCIRL is fed with $n$ trajectories of length $n$ (started in a
+random state) with $n$ varying from $3$ to $20$ (so fed with $9$ to
+$400$ transitions). Each experiment is repeated 50 times. The
+ classifier uses the same data. PIRL is an iterative
+algorithm, each iteration requiring to solve the MDP for some reward
+function. It is run for 70 iterations, all required objects (a
+feature expectations for a non-expert policy and an optimal policy
+according to some reward function at each iteration) are computed
+exactly using the model. We measure the performance of each approach
+with $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, where $\mathcal{U}$
+is the uniform distribution (this allows measuring the
+generalization capability of each approach for states infrequently
+encountered), $\R_E$ is the expert reward and $\pi$ is one of the
+following polices: the optimal policy for $\R_E$ (upper baseline),
+the optimal policy for a random reward (lower baseline), the optimal
+policy for $\R_{\theta_c}$ (SCIRL), the policy produced by PIRL and
+the classifier decision rule.
+
+\begin{figure}[fig:res]{Highway problem.
+The highest line is the expert value.
+For each curves, we show the mean (plain line), the standard deviation (dark color) and the min-max values (light color).
+The policy corresponding to the random reward is in blue, the policy outputted by the  classifier is in yellow and the optimal policy according the SCIRL's reward is in red.
+PIRL is the dark blue line.}
+\begin{minipage}[l]{0.45\linewidth}
+\centering
+  \centerline{\includegraphics[width=.92\linewidth]{fig_classif.pdf}}
+\end{minipage} \hfill
+\begin{minipage}[r]{0.45\linewidth}
+\centering
+  \centerline{\includegraphics[width=.92\linewidth]{fig_scirl.pdf}}
+\end{minipage}
+\end{figure}
+
+Fig.~\ref{fig:res} shows the performance of each approach as a
+number of used expert transitions (except PIRL which uses the
+model). We can see that the classifier does not work well on this
+example. Increasing the number of samples would improve its
+performance, but after 400 transitions it does not work as well as
+SCIRL with only a ten of transitions. SCIRL works pretty well here:
+after only a hundred of transitions it reaches the performance of
+PIRL, both being close to the expert value. We do not report exact
+computational times, but running SCIRL one time with $400$
+transitions is approximately hundred time faster than running PIRL
+for $70$ iteration.
+%
+% (this is due to the need to repeatedly solve MDPs for PIRL).
+
+%Recall that for SCIRL the instantiation of
+%Sec.~\ref{subsec:practicalApproach:instantiation} is considered,
+%with a single Monte-Carlo rollout for estimating
+%$\hat{\mu}^{\pi_E}(s_i,a_i)$ and using the proposed heuristic for
+%$\hat{\mu}^{\pi_E}(s_i,a\neq a_i)$, which goes beyond our analysis.
+%To our knowledge, it is the only algorithm working (at least on this
+%benchmark) only with data from the expert.
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+We have introduced a new way to perform IRL by structuring a
+linearly parameterized score function-based multi-class
+classification algorithm with an estimate of the expert feature
+expectation. This outputs a reward function for which we have shown
+the expert to be near optimal, provided a small classification error
+and a good expert feature expectation estimate. How to practically
+estimate this quantity has been  discussed and we have introduced a
+heuristic for the case where only transitions from the expert are
+available, along with a specific instantiation of the SCIRL
+algorithm. We have shown on a car driving simulator benchmark that
+the proposed approach works well (even combined with the introduced
+heuristic), much better than the unstructured classifier and as well
+as a state-of-the-art algorithm making use of the model (and with a
+much lower computational time). In the future, we plan to deepen the
+theoretical properties of SCIRL (notably regarding possible
+heuristics) and to apply it to real-world robotic problems.
+
+\acknowledgements{This research was partly funded by the EU FP7 project ILHAIRE (grant
+n$\deg$270780), by the EU INTERREG IVa project ALLEGRO and by the Région
+Lorraine (France).}
+
+\newpage
+%\small
 \bibliography{scirl_ria.bib}

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Traduction de la première sous-section

@@
 \newtheorem{prop}[theorem]{Proposition}
-\cite{Ng:2000}
 \maketitle
@@
 
-\section{Background and Notations}
+\section{Contexte et notations}
 \label{sec:background}
 
-\subsection{(Inverse) Reinforcement Learning}
+\subsection{Apprentissage par Renforcement (Inverse)}
 \label{subsec:background:irl}
 
-A Markov Decision process (MDP)~\cite{Puterman:1994} is a tuple
-$\{\s,\A,\p,\R,\gamma\}$ where $\s$ is the finite state
-space\footnote{This work can be extended to compact state spaces, up
-to some technical aspects.}, $\A$ the finite actions space, $\p =
-\{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ the set of
-Markovian transition probabilities, $\R\in\mathbb{R}^\s$ the
-state-dependent reward function and $\gamma$ the discount factor. A
-deterministic policy $\pi\in\s^\A$ defines the behavior of an agent.
-The quality of this control is quantified by the value function
-$v_\R^\pi\in\mathbb{R}^\s$, associating to each state the cumulative
-discounted reward for starting in this state and following the
-policy $\pi$ afterwards: $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t
-\R(S_t)|S_0=s,\pi]$. An optimal policy $\pi_\R^*$ (according to the
-reward function $\R$) is a policy of associated value function
-$v^*_\R$ satisfying $v_\R^* \geq v_\R^\pi$, for any policy $\pi$ and
-componentwise.
+Un Processus Décisionnel de Markov (PDM)~\cite{Puterman:1994} est un ensemble $\{\s,\A,\p,\R,\gamma\}$ avec $\s$ l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ l'espace d'action fini, $\p =
+\{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ la fonction de récompense sur les états et $\gamma$ le facteur d'oubli.
+Une politique déterministe $\pi\in\s^\A$ définit le comportement d'un agent.
+On quantifie la qualité de ce comportement par le biais de la fonction de valeur $v_\R^\pi\in\mathbb{R}^\s$ qui à chaque état associe la récompense pondérée cumulée recueillie par l'agent lorsqu'il part de cet état et suit la politique $\pi$ par la suite : $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t \R(S_t)|S_0=s,\pi]$.
+Une politique optimale $\pi_\R^*$ (vis-à-vis de la fonction de récompense $\R$) est une politique dont la fonction de valeur $v^*_\R$ satisfait $v_\R^* \geq v_\R^\pi$, composante par composante, pour toute politique $\pi$.
 
-Let $P_\pi$ be the stochastic matrix $P_\pi =
-(p(s'|s,\pi(s)))_{1\leq s,s'\leq |\s|}$. With a slight abuse of
-notation, we may write $a$ the policy which associates the action
-$a$ to each state $s$. The Bellman evaluation (resp. optimality)
-operators $T^\pi_\R\text{ (resp. $T^*_\R$)}:\mathbb{R}^\s
-\rightarrow \mathbb{R}^\s$ are defined as $T^\pi_\R v = \R + \gamma
+Soit $P_\pi$ la matrice stochastique $P_\pi =
+(p(s'|s,\pi(s)))_{1\leq s,s'\leq |\s|}$.
+Un léger abus de notation nous permet de nommer $a$ la politique associant l'action $a$ à chaque état $s$.
+Les opérateurs d'évaluation (resp.
+d'optimalité) de Bellman $T^\pi_\R\text{ (resp.
+$T^*_\R$)}:\mathbb{R}^\s
+\rightarrow \mathbb{R}^\s$ sont définis par $T^\pi_\R v = \R + \gamma
 P_\pi v$ and $T_\R^*v = \max_\pi T_\R^\pi v$.
@@
 %\end{equation}
-These operators are contractions and $v_\R^\pi$ and $v^*_\R$ are
-their respective fixed-points: $v_\R^\pi = T^\pi_\R v_\R^\pi$ and
-$v^*_\R = T^*_\R v^*_\R$. The action-value function
-$Q^\pi\in\mathbb{R}^{\s\times \A}$ adds a degree of freedom on the
-choice of the first action, it is formally defined as $Q_\R^\pi(s,a)
-= [T^a_\R v^\pi_\R](s)$. We also write $\rho_\pi$ the stationary
-distribution of the policy $\pi$ (satisfying $\rho_\pi^\top P_\pi =
-\rho_\pi^\top$).
+Ces opérateurs sont des contractions qui admettent $v_\R^\pi$ et $v^*_\R$ comme points fixes respectifs : $v_\R^\pi = T^\pi_\R v_\R^\pi$ et $v^*_\R = T^*_\R v^*_\R$.
+La fonction de qualité $Q^\pi\in\mathbb{R}^{\s\times \A}$ ajoute à $v$ un degré de liberté dans le choix de la première action, on la définit formellement par $Q_\R^\pi(s,a)
+= [T^a_\R v^\pi_\R](s)$. On note $\rho_\pi$ la distribution stationnaire de la politique $\pi$ (qui satisfait $\rho_\pi^\top P_\pi = \rho_\pi^\top$).
 
-Reinforcement learning and approximate dynamic programming aim at
-estimating the optimal control policy $\pi^*_\R$ when the model
-(transition probabilities and the reward function) is unknown (but
-observed through interactions with the system to be controlled) and
-when the state space is too large to allow exact representations of
-the objects of interest (as value functions or
-policies)~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
+L'Apprentissage par Renforcement (AR) et la programmation dynamique approchée cherchent à estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
 %
-We refer to this as the direct problem.
+On appelle cela le problème direct.
 %
-On the contrary, (approximate) inverse reinforcement
-learning~\cite{Ng:2000} aim at estimating a reward function for
-which an observed policy is (nearly) optimal.
+L'Apprentissage par Renforcement Inverse (approché)~\cite{Ng:2000} cherche à estimer la fonction de récompense pour laquelle une politique observée est optimale.
 %
-Let us call this policy the expert policy, denoted $\pi_E$. We may
-assume that it optimizes some unknown reward function $\R_E$. The
-aim of IRL is to compute some reward $\hat{\R}$ such that the expert
-policy is (close to be) optimal, that is such that $v^*_{\hat{\R}}
+On nomme cette politique la politique de l'expert, notée $\pi_E$.
+On suppose qu'elle est optimale vis-à-vis d'une certaine fonction de récompense $R_E$ inconnue.
+Le but de l'ARI est de calculer une fonction de récompense $\hat{\R}$ telle que la politique de l'expert soit (quasiment) optimale, c'est-à-dire telle que $v^*_{\hat{\R}}
 \approx v^{\pi_E}_{\hat{\R}}$.
 %
-We refer to this as the inverse problem.
+On appelle cela le problème inverse.
 
-Similarly to the direct problem, the state space may be too large
-for the reward function to admit a practical exact representation.
-Therefore, we restrict our search of a good reward among linearly
-parameterized functions. Let $\phi(s) = (\phi_1(s)  \dots
+De la même manière que pour le problème direct, l'espace d'état peut être trop grand pour que la fonction de récompense admette une représentation exacte exploitable.
+On se limite donc de fait à la recherche d'une bonne fonction de récompense paramétrée linéairement.
+Soit $\phi(s) = (\phi_1(s)  \dots
 \phi_p(s))^\top$
@@
 %\end{pmatrix}^\top$
-be a feature vector composed of $p$ basis function
-$\phi_i\in\mathbb{R}^\s$, we define the  parameterized reward
-functions as $\R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p
+un vecteur d'attributs composé de $p$ fonctions de base $\phi_i\in\mathbb{R}^\s$, on définit les fonctions de récompense paramétrées par $\R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p
 \theta_i \phi_i(s)$.
@@
 %\end{equation}
-Searching a good reward thus reduces to searching a good parameter
-vector $\theta \in\mathbb{R}^p$. Notice that we will use
-interchangeably $\R_\theta$ and $\theta$ as subscripts
-(\textit{e.g.}, $v_\theta^\pi$ for $v_{\R_\theta}^\pi$).
-Parameterizing the reward this way implies a related
-parameterization for the action-value function:
+La recherche d'une bonne récompense est donc ramenée à la recherche d'un bon vecteur de paramètres vector $\theta \in\mathbb{R}^p$.
+Nous utiliserons indifféremment $\R_\theta$ et $\theta$ comme indices (\textit{e.g.}, $v_\theta^\pi$ pour $v_{\R_\theta}^\pi$).
+Cette paramétrisation de la récompense implique une paramétrisation similaire de la fonction de qualité :
 \begin{equation}
@@
 \end{equation}
-Therefore, the action-value function shares the parameter vector of
-the reward function, with an associated feature vector $\mu^\pi$
-called the \emph{feature expectation}. This notion will be of
-primary importance for the contribution of this paper. Notice that
-each component $\mu_i^\pi$ of this feature vector is actually the
-action-value function of the policy $\pi$ assuming the reward is
-$\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$. Therefore, any
-algorithm designed for estimating an action-value function may be
-used to estimate the feature expectation, such as Monte-Carlo
-rollouts or temporal difference learning~\cite{Klein:2011}.
+Conséquemment, la fonction de qualité partage son vecteur de paramètres avec la fonction de récompense, mais en relation avec un vecteur d'attributs $\mu^\pi$ appelé l'attribut moyen.
+Cette notion est de prime importance à notre contribution.
+Chaque composante $\mu_i^\pi$ de ce vecteur d'attributs est en réalité une fonction de qualité de la politique $\pi$ où la récompense serait $\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$.
+De fait, tout algorithme d'estimation de la fonction de qualité peut être utilisé pour estimer l'attribut moyen, comme une estimation de Monte-Carlo ou un algorithme aux différences temporelles~\cite{Klein:2011}.
 
@@
 the argument which maximizes a related score function, this score
-function being linearly parameterized and the associated parameters
+function being linearly parameterized et the associated parameters
 being learnt by the algorithm. More formally, let $\psi(s,a) =
@@
 Sec.~\ref{subsec:background:classif}. The input $x$ may be seen as a
-state and the label $y$ as an action. Then, the decision rule
+state et the label $y$ as an action. Then, the decision rule
 $g_w(x)$ can be interpreted as a policy which is greedy according to
@@
 Eq.~\eqref{eq:def:mu}, if $\psi(x,y)$ is the feature expectation of
-some policy $\pi$ which produces labels of the training set, and if
+some policy $\pi$ which produces labels of the training set, et if
 the classification error is small, then $w$ will be the parameter
@@
 the linearly parameterized score function, training the classifier
-on $\D$ which produces a parameter vector $\theta_c$, and outputting
+on $\D$ which produces a parameter vector $\theta_c$, et outputting
 the reward function $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.
@@
   \emph{\textbf{Given}} a training set $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
-  an estimate $\hat{\mu}^{\pi_E}$ of the expert feature expectation $\mu^{\pi_E}$ and an MC$^2$ algorithm\;
+  an estimate $\hat{\mu}^{\pi_E}$ of the expert feature expectation $\mu^{\pi_E}$ et an MC$^2$ algorithm\;
   %
@@
   MC$^2$ algorithm
-  fed with the training set $\D$ and considering the parameterized score function
+  fed with the training set $\D$ et considering the parameterized score function
   $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
@@
 classifier somehow implies taking into account the MDP structure
-into the classification problem and allows outputting a reward
+into the classification problem et allows outputting a reward
 vector. Notice that contrary to most of existing IRL
@@
 %Now, we formally show that the proposed approach makes sens, as long
-%as $\mu^{\pi_E}$ is well estimated and as long as the classification
+%as $\mu^{\pi_E}$ is well estimated et as long as the classification
 %error is small.
@@
 small. Before stating our main result, we need to introduce some
-notations and to define some objects.
+notations et to define some objects.
 
@@
   Alg.~\ref{algo:scirl}. Let also the quantities $C_f$, $\epsilon_c$
-  and $\bar{\epsilon}_Q$ be defined as above. We have
+  et $\bar{\epsilon}_Q$ be defined as above. We have
   %
 %  the concentration coefficient $C_f$,
-%  the classification error $\epsilon_c$ and the mean-max
+%  the classification error $\epsilon_c$ et the mean-max
 %  action-value function error $\bar{\epsilon}_Q$ be defined as
@@
   Inequality $(a)$ holds because $T^{\pi^*} v^{\pi_E}\leq T^*
-  v^{\pi_E}$ and inequality $(b)$ holds thanks
+  v^{\pi_E}$ et inequality $(b)$ holds thanks
   to~\cite[Lemma~4.2]{Munos:2007}. Moreover, $v^*$ being optimal we
-  have that $v^*-v^{\pi_E}\geq 0$ and $T^*$ being the Bellman
+  have that $v^*-v^{\pi_E}\geq 0$ et $T^*$ being the Bellman
   optimality operator, we have $T^* v^{\pi_E}\geq
@@
   we will bound $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
-  and $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]$.
+  et $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]$.
 
@@
   \end{align}
-  Taking the expectation according to $\rho_E$ and noticing that
+  Taking the expectation according to $\rho_E$ et noticing that
   $T^* v^{\pi_E}\geq v^{\pi_E}$, we bound the first
@@
   \end{equation}
-  Applying this operator to $v^{\pi_E}$ and recalling that $\R +
+  Applying this operator to $v^{\pi_E}$ et recalling that $\R +
   \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, we
@@
 small for states sampled according to the expert stationary policy
-and for all actions) and if the classification error $\epsilon_c$ is
+and for all actions) et if the classification error $\epsilon_c$ is
 small, then the proposed generic algorithm outputs a reward function
@@
 direct corollary of Th.~\ref{th} is that given the true expert
-feature expectation $\mu^{\pi_E}$ and a perfect classifier
+feature expectation $\mu^{\pi_E}$ et a perfect classifier
 ($\epsilon_c=0$), $\pi_E$ is the unique optimal policy for
@@
 parameter vector $\theta_c$ is computed by the classifier. With
-$\theta_c=0$, the decision rule would be a random policy and we
+$\theta_c=0$, the decision rule would be a random policy et we
 would have $\epsilon_c = \frac{|\A|-1}{|\A|}$, the worst possible
@@
 for the reward function (that is a good choice of basis functions
-$\phi_i$) and a small estimation error, this should not be a
+$\phi_i$) et a small estimation error, this should not be a
 practical problem.
@@
 Finally, if our bound relies on the generalization errors
-$\epsilon_c$ and $\bar{\epsilon}_Q$, the classifier will only use
+$\epsilon_c$ et $\bar{\epsilon}_Q$, the classifier will only use
 $(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ in the training
@@
 [T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. We briefly review its exact
-computation and possible estimation approaches, and consider
+computation et possible estimation approaches, et consider
 possible heuristics.
@@
 %  \tilde{\Psi}')\right)^{-1}\tilde{\Psi}^\top \tilde{\Phi}
-%  \text{ and } \hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a).
+%  \text{ et } \hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a).
 %\end{equation}
@@
 %
-Provided a simulator and the ability to sample according to the
+Provided a simulator et the ability to sample according to the
 expert policy, the expert feature expectation may also be estimated
@@
 using transitions whose starting state is sampled according to
-$\rho_E$ and whose actions are uniformly distributed. However, it
+$\rho_E$ et whose actions are uniformly distributed. However, it
 may happen that only transitions of the expert are available: $\T =
@@
 state $s_\text{v}$ for which $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ for
-any action $a$ and for which the reward feature expectation is the
+any action $a$ et for which the reward feature expectation is the
 null vector, $\phi(s_v) = 0$. In this case, we have
@@
 $\mu^{\pi_E}(s_i,a_i)$ using the single rollout available in the
-training set and use the proposed heuristic for other actions:
+training set et use the proposed heuristic for other actions:
 \begin{equation}
   \forall 1\leq i \leq N,\; \hat{\mu}^{\pi_E}(s_i,a_i) =
-  \sum_{j=i}^N \gamma^{j-i}\phi(s_j) \text{ and }
+  \sum_{j=i}^N \gamma^{j-i}\phi(s_j) \text{ et }
   \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
@@
 To sum up, the expert feature expectation may be seen as a vector of
-action-value functions (for the same policy $\pi_E$ and different
+action-value functions (for the same policy $\pi_E$ et different
 reward functions $\phi_i$). Consequently, any action-value function
@@
 function satisfying $\lc(s,\pi_E(s))\leq \lc(s,a)$ (here,
-$\lc(s_i,a_i)=0$ and $\lc(s_i,a\neq a_i)=1$). The MC$^2$ algorithm
+$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). The MC$^2$ algorithm
 solves:% the following quadratic problem:
@@
 %to produce a non-trival reward for which the expert policy is the
-%unique optimal policy, and this without solving any direct problem.
+%unique optimal policy, et this without solving any direct problem.
 
@@
 %maximization~\cite{Ramachandran:2007}, \textit{etc}. Some of these
-%approaches and others are nicely reviewed in~\cite{Neu:2010}.
+%approaches et others are nicely reviewed in~\cite{Neu:2010}.
 %
@@
 non-structured kernel. However, this approach is not an IRL
-algorithm, and more important assessing the metric of an MDP is a
+algorithm, et more important assessing the metric of an MDP is a
 quite involved problem. In~\cite{Ratliff:2006}, a classification
@@
 (driving off-road is allowed on both sides). The car can move left
-and right, accelerate, decelerate and keep a constant speed. The
+and right, accelerate, decelerate et keep a constant speed. The
 expert optimizes a handcrafted reward $\R_E$ which favours speed,
-punish off-road, punish collisions even more  and is neutral
+punish off-road, punish collisions even more  et is neutral
 otherwise.
@@
 a randomly sampled reward function as a baseline (using the same
-reward feature vector as SCIRL and PIRL, the associated parameter
+reward feature vector as SCIRL et PIRL, the associated parameter
 vector is randomly sampled).
 
-For SCIRL and PIRL we use a discretization of the state space as the
+For SCIRL et PIRL we use a discretization of the state space as the
 reward feature vector, $\phi\in\mathbb{R}^{729}$: $9$ horizontal
-positions for the user's car, $3$ horizontal and $9$ vertical
-positions for the closest traffic's car and $3$ speeds. Notice that
+positions for the user's car, $3$ horizontal et $9$ vertical
+positions for the closest traffic's car et $3$ speeds. Notice that
 these features are much less informative than the ones used
@@
 function. It is run for 70 iterations, all required objects (a
-feature expectations for a non-expert policy and an optimal policy
+feature expectations for a non-expert policy et an optimal policy
 according to some reward function at each iteration) are computed
@@
 generalization capability of each approach for states infrequently
-encountered), $\R_E$ is the expert reward and $\pi$ is one of the
+encountered), $\R_E$ is the expert reward et $\pi$ is one of the
 following polices: the optimal policy for $\R_E$ (upper baseline),
@@
 The highest line is the expert value.
-For each curves, we show the mean (plain line), the standard deviation (dark color) and the min-max values (light color).
-The policy corresponding to the random reward is in blue, the policy outputted by the  classifier is in yellow and the optimal policy according the SCIRL's reward is in red.
+For each curves, we show the mean (plain line), the standard deviation (dark color) et the min-max values (light color).
+The policy corresponding to the random reward is in blue, the policy outputted by the  classifier is in yellow et the optimal policy according the SCIRL's reward is in red.
 PIRL is the dark blue line.}
@@
 %with a single Monte-Carlo rollout for estimating
-%$\hat{\mu}^{\pi_E}(s_i,a_i)$ and using the proposed heuristic for
+%$\hat{\mu}^{\pi_E}(s_i,a_i)$ et using the proposed heuristic for
 %$\hat{\mu}^{\pi_E}(s_i,a\neq a_i)$, which goes beyond our analysis.
@@
 and a good expert feature expectation estimate. How to practically
-estimate this quantity has been  discussed and we have introduced a
+estimate this quantity has been  discussed et we have introduced a
 heuristic for the case where only transitions from the expert are
@@
 the proposed approach works well (even combined with the introduced
-heuristic), much better than the unstructured classifier and as well
+heuristic), much better than the unstructured classifier et as well
 as a state-of-the-art algorithm making use of the model (and with a
@@
 theoretical properties of SCIRL (notably regarding possible
-heuristics) and to apply it to real-world robotic problems.
+heuristics) et to apply it to real-world robotic problems.
 
 \acknowledgements{This research was partly funded by the EU FP7 project ILHAIRE (grant
-n$\deg$270780), by the EU INTERREG IVa project ALLEGRO and by the Région
+n$\deg$270780), by the EU INTERREG IVa project ALLEGRO et by the Région
 Lorraine (France).}

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Traduction de la première section

@@
 
-\subsection{Classification with Linearly Parameterized Score
-Functions} \label{subsec:background:classif}
+\subsection{Classifieurs à fonction de score paramétrée linéairement} \label{subsec:background:classif}
 
-Let $\X$ be a compact or a finite set (of inputs to be classified)
-and let $\Y$ be a finite set (of labels). Assume that inputs $x\in
-\X$  are drawn according to some unknown distribution $\prob(x)$ and
-that there exists some oracle which associates to each of these
-inputs a label $y\in Y$ drawn according to the unknown conditional
-distribution $\prob(y|x)$. Generally speaking, the goal of
-multi-class classification is, given a training set
-$\{(x_i,y_i)_{1\leq i \leq N}\}$ drawn according to $\prob(x,y)$, to
-produce a decision rule $g\in\Y^\X$ which aims at minimizing the
-classification error $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq
-y)$, where $\chi$ denotes the indicator function.
+Soit $\X$ un ensemble fini ou compact (d'entrées à classifier)
+ et soit $\Y$ un ensemble fini de (de labels). Supposons que les entrées $x\in
+\X$ sont tirées selon une distribution inconnue $\prob(x)$ et qu'il existe un oracle qui à chacune de ces entrées associe un label $y\in Y$ tiré selon une distribution de probabilité conditionnelle inconnue $\prob(y|x)$.
+De manière générale, la classification multi-classe cherche étant donné un set d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tiré selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
 
-Here, we consider a more restrictive set of classification
-algorithms. We assume that the decision rule associates to an input
-the argument which maximizes a related score function, this score
-function being linearly parameterized et the associated parameters
-being learnt by the algorithm. More formally, let $\psi(s,a) =
+Nous nous préoccupons ici d'un ensemble plus réduit de classifieurs.
+Nous supposons que la règle de décision associe à l'entrée l'argument qui maximise une certaine fonction de score, celle-ci est paramétrée linéairement, les paramètres étant appris par le classifieur.
+Formellement, soit $\psi(s,a) =
 (\psi_1(x,y)  \dots  \psi_d(x,y))^\top\in \mathbb{R}^d$
@@
 %\end{pmatrix}^\top \in \mathbb{R}^d$
-be a feature vector whose components are $d$ basis functions
-$\psi_i\in\mathbb{R}^{\X\times\Y}$. The linearly parameterized score
-function $s_w\in\mathbb{R}^{\X\times \Y}$ of parameter vector
-$w\in\mathbb{R}^d$ is defined as $s_w(x,y) = w^\top \psi(x,y)$. The
-associated decision rule $g_w\in{\Y^\X}$ is defined as $g_w(x) \in
-\argmax_{y\in\Y}s_w(x,y)$. Using a training set $\{(x_i,y_i)_{1\leq
-i\leq N}\}$, a linearly parameterized score function-based
-multi-class classification (MC$^2$ for short) algorithm computes a
-parameter vector $\theta_c$. The quality of the solution is
-quantified by the classification error $\epsilon_c =
+un vecteur d'attributs dont les composantes sont $d$ fonctions de base $\psi_i\in\mathbb{R}^{\X\times\Y}$.
+La fonction de score linéairement paramétrée $s_w\in\mathbb{R}^{\X\times \Y}$ de vecteur de paramètres $w\in\mathbb{R}^d$ est définie par $s_w(x,y) = w^\top \psi(x,y)$.
+La règle de décision associée $g_w\in{\Y^\X}$ est définie par $g_w(x) \in \argmax_{y\in\Y}s_w(x,y)$.
+En se basant sur un set d'entraînement $\{(x_i,y_i)_{1\leq
+i\leq N}\}$, un classifieur multi-classe (noté MC$^2$ pour \emph{Multi-Class Classifier}) à fonction de score paramétrée linéairement choisit un vecteur de paramètres $\theta_c$.
+La qualité de ce choix est quantifiée par l'erreur de classification $\epsilon_c =
 \prob(g_{\theta_c}(x)\neq y)$.
 
-We do not consider a specific MC$^2$ algorithm, as long as it
-classifies inputs by maximizing the argument of a linearly
-parameterized score function. For example, one may choose a
-multi-class support vector machine~\cite{Guermeur:2007} (taking the
-kernel induced by the feature vector) or a structured large margin
-approach~\cite{Taskar:2005}. Other choices may be possible, one can
-choose its preferred algorithm.
+Notre étude est valable pour tout algorithme de classification tant qu'il opère en maximisant l'argument d'une fonction de score paramétrée linéairement.
+Par exemple, une machine à vecteur support multi-classe pourrait être choisie~\cite{Guermeur:2007} (en prenant le noyau induit par le vecteur de paramètres) de même qu'une approche à marge vaste et structurée~\cite{Taskar:2005}.
+D'autres choix sont possibles, chacun peut choisir son algorithme fétiche.
 
-\section{Structured Classification for Inverse Reinforcement
-Learning} \label{sec:scirl}
+\section{Classification structurée pour l'apprentissage par renforcement inverse} \label{sec:scirl}
 
-\subsection{General Algorithm}
+\subsection{Forme générale de l'algorithme}
 \label{subsec:scirl:algo}
 
-Consider the classification framework of
-Sec.~\ref{subsec:background:classif}. The input $x$ may be seen as a
-state et the label $y$ as an action. Then, the decision rule
-$g_w(x)$ can be interpreted as a policy which is greedy according to
-the score function $w^\top \psi(x,y)$, which may itself be seen as
-an action-value function. Making the parallel with
+On se place dans le cadre de la classification décrit en Sec.~\ref{subsec:background:classif}.
+L'entrée $x$ oeut être vue comme un état et le label $y$ comme une action.
+Il s'ensuit que l'on peut interpréter la règle de décision
+$g_w(x)$ comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle même être vue comme une fonction de qualité. Making the parallel with
 Eq.~\eqref{eq:def:mu}, if $\psi(x,y)$ is the feature expectation of

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Sous section de présentation de SCIRL traduite

@@
 Il s'ensuit que l'on peut interpréter la règle de décision
-$g_w(x)$ comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle même être vue comme une fonction de qualité. Making the parallel with
-Eq.~\eqref{eq:def:mu}, if $\psi(x,y)$ is the feature expectation of
-some policy $\pi$ which produces labels of the training set, et if
-the classification error is small, then $w$ will be the parameter
-vector of a reward function for which we may hope the policy $\pi$
-to be near optimal. Based on these remarks, we're ready to present
-the proposed Structured Classification-based IRL (SCIRL) algorithm.
+$g_w(x)$ comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle même être vue comme une fonction de qualité.
+On peut alors tirer le parallèle avec l'Eq.~\eqref{eq:def:mu}, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels du set d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
+Ces remarques nous permettent maintenant d'introduire notre algorithme d'ARI par classification structurée (SCIRL pour \emph{Structured Classification-based Inverse Reinforcement Learning}).
 
-Let $\pi_E$ be the expert policy from which we would like to recover
-a reward function. Assume that we have a training set $\D =
-\{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ where states are sampled
-according to the expert stationary distribution\footnote{For
-example, if the Markov chain induced by the expert policy is
-fast-mixing, sampling a trajectory will quickly lead to sample
-states according to this distribution.} $\rho_E = \rho_{\pi_E}$.
-Assume also that we have an estimate $\hat{\mu}^{\pi_E}$ of the
-expert feature expectation $\mu^{\pi_E}$ defined in
-Eq.~\eqref{eq:def:mu}. How to practically estimate this quantity is
-postponed to Sec.~\ref{subsec:practicalApproach:muE}; however,
-recall that estimating $\mu^{\pi_E}$ is simply a \emph{policy
-evaluation} problem (estimating the action-value function of a given
-policy), as noted in Sec.~\ref{subsec:background:irl}. Assume also
-that an MC$^2$ algorithm has been chosen. The proposed algorithm
-simply consists in choosing $\theta^\top\hat{\mu}^{\pi_E}(s,a)$ as
-the linearly parameterized score function, training the classifier
-on $\D$ which produces a parameter vector $\theta_c$, et outputting
-the reward function $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.
+Soit $\pi_E$ la politique de l'expert à partir de laquelle on souhaite inférer une fonction de récompense.
+Supposons l'existence d'un set d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est \emph{fast-mixing}, l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés d'après cette distribution.} $\rho_E = \rho_{\pi_E}$.
+Supposons également que nous avons à notre disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eq:def:mu}.
+Une description de la manière d'estimer cette quantité en pratique est remise à la Sec.~\ref{subsec:practicalApproach:muE}; rappelons en revanche qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé Sec.~\ref{subsec:background:irl}.
+Supposons enfin qu'un algorithme de MC$^2$ a été choisi.
+L'algorithme formant notre contribution consiste simplement à choisir $\theta^\top\hat{\mu}^{\pi_E}(s,a)$ comme la fonction de score paramétrée linéairement, puis à entraîner le classifieur sur $\D$, ce qui produit un vecteyr de paramètres $\theta_c$, et enfin à renvoyer la fonction de récompense $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.
 
@@
   \SetAlgoVlined
-  \caption{SCIRL algorithm}
+  \caption{Algorithme SCIRL}
   \label{algo:scirl}
@@
   \BlankLine
-  \emph{\textbf{Given}} a training set $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
-  an estimate $\hat{\mu}^{\pi_E}$ of the expert feature expectation $\mu^{\pi_E}$ et an MC$^2$ algorithm\;
+  \emph{\textbf{Etant donné}} un set d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
+  une estimée $\hat{\mu}^{\pi_E}$ de l'attridut moyen de l'expert $\mu^{\pi_E}$ et an un algorithme de MC$^2$\;
   %
   \BlankLine
-  \emph{\textbf{Compute}} the parameter vector $\theta_c$ using the
-  MC$^2$ algorithm
-  fed with the training set $\D$ et considering the parameterized score function
-  $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
+  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de MC$^2$ auquel sont fournis le set d'entraînement $\D$ et la paramétrisation de la fonction de score : $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
   %
   \BlankLine
-  \emph{\textbf{Output}} the reward function $\R_{\theta_c}(s) = \theta_c^\top\phi(s)$ \;
+  \emph{\textbf{Renvoyer}} la fonction de récompense $\R_{\theta_c}(s) = \theta_c^\top\phi(s)$ \;
 \end{algorithm2e}
 
-The proposed approach is summarized in Alg.~\ref{algo:scirl}. We
-call this Structured Classification-based IRL because using the
-(estimated) expert feature expectation as the feature vector for the
-classifier somehow implies taking into account the MDP structure
-into the classification problem et allows outputting a reward
-vector. Notice that contrary to most of existing IRL
-%and apprenticeship learning
-algorithms, SCIRL does not require solving the direct problem. If it
-possibly requires estimating the expert feature expectation, it is
-just a policy evaluation problem, less difficult than the policy
-optimization issue involved by the direct problem. This is further
-discussed in Sec.~\ref{sec:relatedWorks}.
+L'approche proposée est résumé en Alg.~\ref{algo:scirl}.
+Le nom de l'algorithme (Classification structurée pour l'ARI) vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
+Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct.
+Cet algorithme requiert une estimation de l'attribut moyen de l'expert mais il ne s'agit là que d'un problème d'évaluation de la politique, notoirement moins difficile que la recherche d'une politique optimale qu'implique la résolution du problème direct.
+Cela est discuté plus en détail en Sec.~\ref{sec:relatedWorks}.
 %

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Sous section analyse traduite.

@@
 
-\subsection{Analysis}
+\subsection{Analyse}
 \label{subsec:scirl:analysis}
 
-In this section, we show that the expert policy $\pi_E$ is close to
-be optimal according to the reward function $\R_{\theta_c}$, more
-precisely that
-$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ is
-small. Before stating our main result, we need to introduce some
-notations et to define some objects.
+Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi-optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément que 
+$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ est faible.
+Avant de présenter le résultat principal, il nous faut introduire quelques notations et définir quelques objets.
 
-We will use the first order discounted future state distribution
-concentration coefficient $C_f$~\cite{Munos:2007}:
+Nous allons utiliser le coefficient de concentration de la distribution pondérée du premier ordre pour l'état futur $C_f$~\cite{Munos:2007}:
 \begin{equation}
@@
 \end{equation}
-We note $\pi_c$ the decision rule of the classifier: $\pi_c(s) \in
-\argmax_{a\in \A} \theta_c^\top\hat{\mu}^{\pi_E}(s,a)$. The
-classification error is therefore $\epsilon_c =
-\E_{s\sim\rho_E}[\chi_{\{\pi_c(s)\neq\pi_E(s)\}}] \in [0,1]$. We
-write $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$
-the score function computed from the training set $\D$ (which can be
-interpreted as an approximate action-value function). Let also
+On note $\pi_c$ la règle de décision du classifieur : $\pi_c(s) \in
+\argmax_{a\in \A} \theta_c^\top\hat{\mu}^{\pi_E}(s,a)$. L'erreur de classification est donc $\epsilon_c =
+\E_{s\sim\rho_E}[\chi_{\{\pi_c(s)\neq\pi_E(s)\}}] \in [0,1]$.
+On écrit $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$ la fonction de score calculée à partir du set d'entraînement $\D$ (celle-ci peut être interprétée comme une fonction de qualité approchée).
+Soit
 $\epsilon_{\mu} = \hat{\mu}^{\pi_E} - \mu^{\pi_E}:\s\times\A
-\rightarrow  \mathbb{R}^p$ be the feature expectation error.
-Consequently, we define the action-value function error as
+\rightarrow  \mathbb{R}^p$ l'erreur sur l'estimation de l'attribut moyen.
+En conséquence, on définit l'erreur sur la fonction de qualité par 
 $\epsilon_Q = \hat{Q}^{\pi_E}_{\theta_c} - Q^{\pi_E}_{\theta_c} =
 \theta_c^\top(\hat{\mu}^{\pi_E} - \mu^{\pi_E}) = \theta_c^\top
-\epsilon_\mu:\s\times\A\rightarrow\mathbb{R}$. We finally define the
-mean delta-max action-value function error as $\bar{\epsilon}_Q =
+\epsilon_\mu:\s\times\A\rightarrow\mathbb{R}$. Finalement, on définit le delta-max moyen de l'erreur sur la fonction de qualité par $\bar{\epsilon}_Q =
 \E_{s\sim\rho_E}[\max_{a\in\A}\epsilon_Q(s,a) -
@@
   \label{th}
-  Let $\R_{\theta_c}$ be the reward function outputted by
-  Alg.~\ref{algo:scirl}. Let also the quantities $C_f$, $\epsilon_c$
-  et $\bar{\epsilon}_Q$ be defined as above. We have
+  Soit $\R_{\theta_c}$ la fonction de récompense renvoyée par l'Alg.~\ref{algo:scirl}. Soient $C_f$, $\epsilon_c$
+  et $\bar{\epsilon}_Q$ les quantités définies ci-dessus. On a
   %
@@
 \begin{proof}
-  As the proof only relies on the reward $\R_{\theta_c}$, we
-  omit the related subscripts to keep the notations simple
-  (\textit{e.g.}, $v^\pi$ for
-  $v^\pi_{\theta_c}=v^\pi_{\R_{\theta_c}}$ or $\R$ for $\R_{\theta_c}$). First, we link
-  the error $\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)]$ to the
-  Bellman residual $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s)-v^{\pi_E}(s)]$.
+  La démonstration repose uniquement sur la récompense $\R_{\theta_c}$, on omet donc par soucis de clarté des notations certains indices relatifs à la récompense 
+  (\textit{e.g.}, $v^\pi$ pour
+  $v^\pi_{\theta_c}=v^\pi_{\R_{\theta_c}}$ ou $\R$ pour $\R_{\theta_c}$).
+Tout d'abord, l'on relie  l'erreur $\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)]$ au résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s)-v^{\pi_E}(s)]$.
   %
@@
   %
-  Componentwise, we have that:
+  Composante par composante, on a :
   \begin{align}
@@
   \end{align}
-  Inequality $(a)$ holds because $T^{\pi^*} v^{\pi_E}\leq T^*
-  v^{\pi_E}$ et inequality $(b)$ holds thanks
-  to~\cite[Lemma~4.2]{Munos:2007}. Moreover, $v^*$ being optimal we
-  have that $v^*-v^{\pi_E}\geq 0$ et $T^*$ being the Bellman
-  optimality operator, we have $T^* v^{\pi_E}\geq
-  T^{\pi_E}v^{\pi_E}=v^{\pi_E}$. Additionally, remark that
+  L'inégalité $(a)$ est valable car $T^{\pi^*} v^{\pi_E}\leq T^*
+  v^{\pi_E}$ et l'inégalité $(b)$ l'est aussi grâce à~\cite[Lemma~4.2]{Munos:2007}.
+De plus, $v^*$ étant optimale, l'on voit que  $v^*-v^{\pi_E}\geq 0$ et avec $T^*$ l'opérateur d'optimalité de Bellman, on a $T^* v^{\pi_E}\geq
+  T^{\pi_E}v^{\pi_E}=v^{\pi_E}$. De plus, on remarque que 
   $(I-\gamma P_{\pi^*})^{-1} = \sum_{t\geq 0}\gamma^t P_{\pi^*}^t$.
-  Therefore, using the definition of the concentration coefficient
-  $C_f$, we have that:
+  Donc, d'après la définition du coefficient de concentration 
+  $C_f$, on a :
   \begin{equation}
@@
   \end{equation}
-  This results actually follows closely the one
-  of~\cite[Theorem~4.2]{Munos:2007}. There remains to bound the
-  Bellman residual $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
-  v^{\pi_E}(s)]$. Considering the following decomposition,
+  Ce résultat est similaire à celui de~\cite[Theorem~4.2]{Munos:2007}. Il reste à borner le résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
+  v^{\pi_E}(s)]$. Considérons la décomposition
   \begin{equation}
@@
   \end{equation}
-  we will bound $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
+  on va borner  $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
   et $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]$.
 
-  The policy $\pi_c$ (the decision rule of the classifier) is greedy
-  with respect to
-  $\hat{Q}^{\pi_E}=\theta_c^\top\hat{\mu}^{\pi_E}$. Therefore, for any state-action couple
-  $(s,a)\in\s\times \A$ we have:
+  La politique $\pi_c$ (la règle de décision du classifieur) est gloutonne vis-à-vis de 
+  $\hat{Q}^{\pi_E}=\theta_c^\top\hat{\mu}^{\pi_E}$. Donc, pour chaque couple état-action
+  $(s,a)\in\s\times \A$ on a:
   \begin{equation}
@@
   \end{equation}
-  By definition, $Q^{\pi_E}(s,a) = [T^a v^{\pi_E}](s)$ and
-  $Q^{\pi_E}(s,\pi_c(s)) = [T^{\pi_c} v^{\pi_E}](s)$. Therefore, for $s\in\s$:
+  Par définition, $Q^{\pi_E}(s,a) = [T^a v^{\pi_E}](s)$ et
+  $Q^{\pi_E}(s,\pi_c(s)) = [T^{\pi_c} v^{\pi_E}](s)$. Donc, pour $s\in\s$:
   \begin{align}
@@
   \end{align}
-  Taking the expectation according to $\rho_E$ et noticing that
-  $T^* v^{\pi_E}\geq v^{\pi_E}$, we bound the first
-  term:
+  En passant à l'espérance selon $\rho_E$ et tout en remarquant que 
+  $T^* v^{\pi_E}\geq v^{\pi_E}$, on borne le premier terme
   \begin{equation}
@@
   \end{equation}
-  There finally remains to bound the term $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) -
+  Il reste enfin à borner le terme $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) -
   v^{\pi_E}(s)]$.
 
-  Let us write $M\in\mathbb{R}^{|\s|\times |\s|}$ the diagonal matrix
-  defined as $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$. Using
-  this, the Bellman operator $T^{\pi_c}$ may be written as, for any $v\in\mathbb{R}^\s$:
+  On écrit $M\in\mathbb{R}^{|\s|\times |\s|}$ la matrice diagonale définie par $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$.
+En utilisant cela, l'opérateur de Bellman $T^{\pi_c}$ peut s'écrire, pour tout $v\in\mathbb{R}^\s$:
   \begin{equation}
@@
   \end{equation}
-  Applying this operator to $v^{\pi_E}$ et recalling that $\R +
-  \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, we
-  get:
+  En appliquant cet opérateur à  $v^{\pi_E}$ et en utilisant le fait que $\R +
+  \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, on obtient :
   \begin{equation}
@@
   \end{equation}
-  One can easily see that $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
-  \leq \frac{2}{1-\gamma}\|\R\|_\infty$, which allows bounding the
-  last term:
+  On peut facilement voir que $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
+  \leq \frac{2}{1-\gamma}\|\R\|_\infty$, ce qui permet de borner le dernier terme
   \begin{equation}
@@
   \end{equation}
-  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
-  and~\eqref{eq:proof:b2} into Eq.~\eqref{eq:proof:residual} gives the
-  stated result.%, which concludes the proof.
+  Injecter les bornes des Eqs.~\eqref{eq:proof:b1}
+  and~\eqref{eq:proof:b2} dans l'Eq.~\eqref{eq:proof:residual} montre le résultat proposé.%, which concludes the proof.
 %  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
@@
 
-This result shows that if the expert feature expectation is well
-estimated (in the sense that the estimation error $\epsilon_\mu$ is
-small for states sampled according to the expert stationary policy
-and for all actions) et if the classification error $\epsilon_c$ is
-small, then the proposed generic algorithm outputs a reward function
-$\R_{\theta_c}$ for which the expert policy will be near optimal. A
-direct corollary of Th.~\ref{th} is that given the true expert
-feature expectation $\mu^{\pi_E}$ et a perfect classifier
-($\epsilon_c=0$), $\pi_E$ is the unique optimal policy for
+Ce résultat montre que si l'attribut moyen de l'expert est bien estimé (dans le sens d'une faible erreur d'estumation $\epsilon_\mu$ pour les états échantillonnés selon la politique stationnaire de l'expert et pour toutes les actions) et si l'erreur de classification $\epsilon_c$ et également faible alors l'algorithme générique proposé fournit une fonction de récompense
+$\R_{\theta_c}$ vis-à-vis de laquelle l'expert sera quasi-optimal. Un corollaire direct du Th.~\ref{th} stipule qu'avec le vrai attribut moyen de l'expert $\mu^{\pi_E}$ et un classifieur parfait 
+($\epsilon_c=0$), $\pi_E$ est l'unique politique optimale pour 
 $\R_{\theta_c}$.
 
-One may argue that this bounds trivially holds for the null reward
-function (a reward often exhibited to show that IRL is an ill-posed
-problem), obtained if $\theta_c=0$. However, recall that the
-parameter vector $\theta_c$ is computed by the classifier. With
-$\theta_c=0$, the decision rule would be a random policy et we
-would have $\epsilon_c = \frac{|\A|-1}{|\A|}$, the worst possible
-classification error. This case is really unlikely. Therefore, we
-advocate that the proposed approach somehow allows disambiguating
-the IRL problem (at least, it does not output trivial reward
-functions such as the null vector).
+L'on pourrait arguer que ces bornes sont triviallement valables pour la fonction de récompense nulle (fonction régulièrement citée en exemple de la nature mal posée du problème de l'ARI) correspondant au cas $\theta_c=0$.
+Cependant il faut se rappeler que le vecteur de paramètres $\theta_c$ est choisi par le classifieur.
+Avec
+$\theta_c=0$, la règle de décision serait une politique aléatoire et nous aurions $\epsilon_c = \frac{|\A|-1}{|\A|}$, c'est-à-dire la pire erreur de classification possible.
+Ce cas est donc très improbable.
+De fait nous affirmons que notre approche permet, d'une certaine manière, de lever l'ambiguité du problème de l'ARI (au moins, l'algorithme ne renvoie pas de récompense triviales comme la récompense nulle).
 %
-Also, this bound is scale-invariant: one could impose
-$\|\theta_c\|=1$ or normalize (action-) value functions by
+On peut également noter que cette borne est invariante par dilatation. L'on pourrait imposer 
+$\|\theta_c\|=1$ ou normaliser la fonction de valeur (ou de qualité) par 
 $\|\R_{\theta_c}\|_\infty^{-1}$.
 
-One should notice that there is a hidden dependency of the
-classification error $\epsilon_c$ to the estimated expert feature
-expectation $\hat{\mu}^{\pi_E}$. Indeed, the minimum classification
-error depends on the hypothesis space spanned by the chosen score
-function basis functions for the MC$^2$ algorithm (here
-$\hat{\mu}^{\pi_E}$). Nevertheless, provided a good representation
-for the reward function (that is a good choice of basis functions
-$\phi_i$) et a small estimation error, this should not be a
-practical problem.
+On remarquera une dépendance caché de l'erreur de classification  $\epsilon_c$ à l'estimation de l'attribut moyen de l'expert  $\hat{\mu}^{\pi_E}$.
+En effet, l'erreur de classification minimale dépend de l'espace d'hypothèses généré par les fonctions de base de la fonction de score de l'algorithme de MC$^2$ (ici,
+$\hat{\mu}^{\pi_E}$).
+Néanmoins, avec une bonne représentation de la fonction de récompense (c'est-à-dire un choix judicieux de fonctions de base $\phi_i$) et une faible erreur d'estimation, cela ne devrait poser aucun problème en pratique.
 
-Finally, if our bound relies on the generalization errors
-$\epsilon_c$ et $\bar{\epsilon}_Q$, the classifier will only use
-$(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ in the training
-phase, where $s_i$ are the states from the set $\D$. It outputs
-$\theta_c$, seen as a reward function, thus the estimated feature
-expectation $\hat{\mu}^{\pi_E}$ is no longer required. Therefore,
-practically it should be sufficient to estimate well
-$\hat{\mu}^{\pi_E}$ on state-action couples $(s_i,a)_{1\leq i\leq
-N,a\in\A}$, which allows envisioning Monte-Carlo rollouts for
-example.
+Finalement, si notre borne se base sur les erreurs en généralisation 
+$\epsilon_c$ et $\bar{\epsilon}_Q$, le classifieur n'utilisera 
+$(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ que lors de la phase d'entraînement, où les $s_i$ sont les états présents dans $\D$. Il renvoie 
+$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire. Donc en pratique il est satisfaisant d'estimer 
+$\hat{\mu}^{\pi_E}$ correctement pour les couples état-action $(s_i,a)_{1\leq i\leq
+N,a\in\A}$, ce qui permet d'envisager une simple estimation de Monte-Carlo par exemple.
 

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Section mise en pratique traduite

@@
 
-\section{A Practical Approach}
+\section{Mise en pratique}
 \label{sec:practicalApproach}
 
-\subsection{Estimating the Expert Feature Expectation}
+\subsection{Estimation de l'attribut moyen de l'expert}
 \label{subsec:practicalApproach:muE}
 
-SCIRL relies on an estimate $\hat{\mu}^{\pi_E}$ of the expert
-feature expectation. Basically, this is a policy evaluation problem.
-An already made key observation is that each component of
-$\mu^{\pi_E}$ is the action-value function of $\pi_E$ for a reward
-function $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
-[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. We briefly review its exact
-computation et possible estimation approaches, et consider
-possible heuristics.
-%possibly considering some heuristics.
+SCIRL a besoin d'une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert. C'est un problème similaire à une évaluation de politique. Répétons l'observation-clef : chaque composante de 
+$\mu^{\pi_E}$ est une fonction de qualité pour $\pi_E$ vis-à-vis d'une fonction de récompense $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
+[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. Nous présentons une revue rapide de méthodes de calcul exacte et approchées ainsi qu'une heuristique.
 
-If the model is known, the feature expectation can be computed
-explicitly. Let $\Phi\in\mathbb{R}^{|\s|\times p}$ be the feature
-matrix whose rows contain the feature vectors $\phi(s)^\top$ for all
+Si le modèle est connu, il est possible de calculer l'attribut moyen de manière exacte.
+Soit $\Phi\in\mathbb{R}^{|\s|\times p}$ la matrice d'attributs donc les lignes contiennent les vecteurs d'attributs $\phi(s)^\top$ pour tout
 $s\in\s$.
@@
 %
-For a fixed $a\in A$, let $\Mu^{\pi_E}_a \in\mathbb{R}^{|\s|\times
-p}$ be the feature expectation matrix whose rows are the expert
-feature vectors, that is $(\mu^{\pi_E}(s,a))^\top$ for any $s\in\s$.
-With these notations, we have $\Mu_a^{\pi_E} = \Phi + \gamma
+Pour un certain $a\in A$, soit $\Mu^{\pi_E}_a \in\mathbb{R}^{|\s|\times
+p}$ la matrice des attributs moyens dont les lignes sont les attributs moyens de l'expert, c'est-à-dire  $(\mu^{\pi_E}(s,a))^\top$ pour chaque $s\in\s$.
+Ces notations nous permettent d'écrire $\Mu_a^{\pi_E} = \Phi + \gamma
 P_a(I-\gamma P_{\pi_E})^{-1} \Phi$.
@@
 %expectation can be easily computed.
-Moreover, the related computational cost is the same order of
-magnitude as evaluating a single policy (as the costly part,
-computing $(I-\gamma P_{\pi_E})^{-1}$, is shared by all components).
+Ajoutons que le coût computationel de cette méthode est du même ordre de grandeur que l'évaluation d'une seule politique, en effet la partie coûteuse (le calcul de $(I-\gamma P_{\pi_E})^{-1}$) est partagée par toutes les composantes.
 
-If the model is unknown, any temporal difference learning algorithm
-can be used to estimate the expert feature
-expectation~\cite{Klein:2011}, as LSTD (Least-Squares Temporal
-Differences)~\cite{Bradtke:1996}. Let $\psi:\s\times \A \rightarrow
-\mathbb{R}^d$ be a feature vector composed of $d$ basis functions
-$\psi_i \in\mathbb{R}^{\s\times \A}$. Each component $\mu_i^{\pi_E}$
-of the expert feature expectation is parameterized by a vector
+Si le modèle est inconnu, tous les algorithmes d'apprentissage par différences temporelles peuvent être utilisés pour obtenir une estimation de l'attribut moyen de l'expert~\cite{Klein:2011}, comme par exemple LSTD (Least-Squares Temporal
+Differences)~\cite{Bradtke:1996}. Soit $\psi:\s\times \A \rightarrow
+\mathbb{R}^d$ un vecteur d'attributs composé de $d$ fonctions de base
+$\psi_i \in\mathbb{R}^{\s\times \A}$. Chaque composante $\mu_i^{\pi_E}$
+ de l'attribut moyen de l'expert est paramétrisée par un vecteur
 $\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
-\psi(s,a)$. Assume that we have a training set
-$\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ with actions
-$a_i$ not necessarily sampled according to policy $\pi_E$
-(\textit{e.g.}, this may be obtained by sampling trajectories
-according to an expert-based $\epsilon$-greedy policy), the aim
-being to have a better variability of tuples (non-expert actions
-should be tried). Let $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
-(resp. $\tilde{\Psi}'$) be the feature matrix whose rows are the
-feature vectors $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
-Let also $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ be the feature
-matrix whose rows are the reward's feature vectors $\phi(s_i)^\top$.
-Finally, let $\Xi =
+\psi(s,a)$. Supposons que l'on dispose d'un set d'entraînement
+$\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ dont les actions 
+$a_i$ ne sont pas nécessairement échantillonnées selon la politique $\pi_E$
+(\textit{e.g.}, on peut par exemple utiliser des trajectoires obtenues par un agent suivant une politique $\epsilon$-greedy), le but est d'obtenir une meilleure représentativité des données (des actions sous optimales devraient être essayées).
+Soit $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
+(resp. $\tilde{\Psi}'$) la matrice d'attributs dont les lignes sont les vecteurs d'attributs $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
+Soit $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ la matrive d'attributs dont les lignes sont les vecteurs d'attributs de la récompense $\phi(s_i)^\top$.
+Enfin, soit $\Xi =
 \begin{bmatrix}   \xi_1 & \dots & \xi_p
-\end{bmatrix}\in\mathbb{R}^{d\times p}$ be the matrix of all
-parameter vectors. Applying LSTD to each component of the feature
-expectation gives the LSTD-$\mu$ algorithm~\cite{Klein:2011}: $\Xi =
+\end{bmatrix}\in\mathbb{R}^{d\times p}$ la matrice de tous les vecteurs de paramètres.
+Appliquer LSTD sur chacune des composantes de l'attribut moyen donne l'algorithme LSTD-$\mu$~\cite{Klein:2011}: $\Xi =
 (\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
-\tilde{\Psi}'))^{-1}\tilde{\Psi}^\top \tilde{\Phi}$ and
+\tilde{\Psi}'))^{-1}\tilde{\Psi}^\top \tilde{\Phi}$ et
 $\hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a)$.
@@
 %\end{equation}
-As for the exact case, the costly part (computing the inverse
-matrix) is shared by all feature expectation components, the
-computational cost is reasonable (same order as LSTD).
+De la même manière que pour le cas exact, la partie coûteuse de l'algorithme (inverser la matrice) est partagée par toutes les composantes. Le coût reste donc raisonnable (du même ordre que LSTD).
 
 %
-Provided a simulator et the ability to sample according to the
-expert policy, the expert feature expectation may also be estimated
-using Monte-Carlo rollouts for a given state-action pair (as noted
-in Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ need only
-be known on $(s_i,a)_{1\leq i\leq N,a\in\A}$). Assuming that $K$
-trajectories are sampled for each required state-action pair, this
-method would require $KN|\A|$ rollouts.
+Si l'on dispose d'un simulateur permettant d'échantillonner selon la politique de l'expert, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
+Si $K$
+trajectoires sont échantillonnées pour chaque couple, cette méthode requiert $KN|\A|$ simulations.
 
-In order to have a small error $\bar{\epsilon}_Q$, one may learn
-using transitions whose starting state is sampled according to
-$\rho_E$ et whose actions are uniformly distributed. However, it
-may happen that only transitions of the expert are available: $\T =
-\{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$. If the state-action
-couples $(s_i,a_i)$ may be used to feed the classifier, the
-transitions $(s_i,a_i,s'_i)$ are not enough to provide an accurate
-estimate of the feature expectation. In this case, we can still
-expect an accurate estimate of $\mu^{\pi_E}(s,\pi_E(s))$, but there
-is little hope for $\mu^{\pi_E}(s,a\neq\pi_E(s))$. However, one can
-still rely on some heuristic; this does not fit the analysis of
-Sec.~\ref{subsec:scirl:analysis}, but it can still provide good
-experimental results, as illustrated in Sec.~\ref{sec:experiments}.
+Pour minimiser l'erreur $\bar{\epsilon}_Q$, il est possible d'utiliser des transitions dont l'état de départ est tiré selon 
+$\rho_E$ et dont les actions sont uniformément distribuées. Cependant parfois seules les transitions issues de l'expert sont disponibles : $\T =
+\{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$.
+Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen.
+Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$.
+Il est heureusement possible d'utiliser une heuristique; ce cas ne rentre pas dans l'analyse présentée en
+Sec.~\ref{subsec:scirl:analysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en Sec.~\ref{sec:experiments}.
 
-We propose such a heuristic. Assume that only data $\T$ is available
-and that we use it to provide an (accurate) estimate
-$\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (this basically means estimating a
-value function instead of an action-value function as described
-above). We may adopt an optimistic point of view by assuming that
-applying a non-expert action just delays the effect of the expert
-action. More formally, we associate to each state $s$ a virtual
-state $s_\text{v}$ for which $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ for
-any action $a$ et for which the reward feature expectation is the
-null vector, $\phi(s_v) = 0$. In this case, we have
+Nous proposons une telle heuristique. Supposons que  $\T$ soient les seules données disponibles, que nous utilisons pour fournir une estimation (précise)
+$\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (cela revient à estimer non plus une fonction de qualité comme décrit ci-dessus, mais simplement une fonction de valeur).
+Un point de vue optimiste suppose que choisir une action différente de celle de l'expert ne fait que retarder l'effet de l'action de l'expert.
+Plus formellement, nous associons à chaque état $s$ un état virtuel $s_\text{v}$ pour lequel $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ pour toute action $a$ et pour lequel l'attribut (de récompense) moyen est le vecteur nul, $\phi(s_v) = 0$.
+Dans ce cas, on a
 $\mu^{\pi_E}(s,a\neq\pi_E(s)) = \gamma \mu^{\pi_E}(s,\pi_E(s))$.
-Applying this idea to the available estimate (recalling that the
-classifiers only requires evaluating $\hat{\mu}^{\pi_E}$ on
-$(s_i,a)_{1\leq i\leq N,a\in \A}$) provides the proposed heuristic:
-for $1\leq i\leq N$, $\hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
+Appliquer cette idée sur l'estimation effectivement disponible (rappelons que le classifieur n'a besoin d'évaluer  $\hat{\mu}^{\pi_E}$ qu'en
+$(s_i,a)_{1\leq i\leq N,a\in \A}$) fournit l'heuristique que nous proposons :
+pour $1\leq i\leq N$, $\hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
 \hat{\mu}^{\pi_E}(s_i,a_i)$.
@@
 
-We may even push this idea further, to get the simpler estimate of
-the expert feature expectation (but with the weakest guarantees).
-Assume that the set $\T$ consists of one long trajectory, that is
-$s'_i = s_{i+1}$ (thus $\T =
-\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). We may estimate
-$\mu^{\pi_E}(s_i,a_i)$ using the single rollout available in the
-training set et use the proposed heuristic for other actions:
+Il est également possible de pousser cette idée plus loin afin d'obtenir une estimation plus simple (mais offrant de moindres garanties) de l'attribut moyen de l'expert.
+Supposons que  $\T$ consiste en une longue trajectoire, c'est-à-dire
+$s'_i = s_{i+1}$ (donc $\T =
+\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). On peut estimer
+$\mu^{\pi_E}(s_i,a_i)$ en utilisant la seule trajectoire disponible et en utilisant l'heuristique précedente pour les autres actions :
 \begin{equation}
@@
 
-To sum up, the expert feature expectation may be seen as a vector of
-action-value functions (for the same policy $\pi_E$ et different
-reward functions $\phi_i$). Consequently, any action-value function
-evaluation algorithm may be used to estimate $\mu^\pi(s,a)$.
-Depending on the available data, one may have to rely on some
-heuristic to assess the feature expectation for a unexperienced
-(non-expert) action. Also, this expert feature expectation estimate
-is only required for training the classifier, so it is sufficient to
-estimate on state-action couples $(s_i,a)_{1\leq i \leq N,a\in \A}$.
-In any case, estimating $\mu^{\pi_E}$ is not harder than estimating
-the action-value function of a given policy in the on-policy case,
-which is much easier than computing an optimal policy for an
-arbitrary reward function (as required by most of existing IRL
-algorithms, see Sec.~\ref{sec:relatedWorks}).
+Pour résumer, l'attribut moyen de l'expert peut être vu comme un vecteur de fonctions de qualité (pour une même politique  $\pi_E$ et pour différentes fonctions de récompense $\phi_i$).
+En conséquence, tout algorithme d'évaluation de la fonction de qualité peut être utilisé pour estimer $\mu^\pi(s,a)$.
+Selon la quantité et la nature des données disponibles, l'on peut avoir recours à une heuristique pour évaluer l'attribut moyen pour une action différente de celle de l'expert.
+Cette estimation n'est nécessaire que pour entraîner les classifieur, il est donc suffisant de disposer de valeurs uniquement pour les couples état-action $(s_i,a)_{1\leq i \leq N,a\in \A}$.
+En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir Sec.~\ref{sec:relatedWorks}).
 
-\subsection{An Instantiation}
+\subsection{Instanciation}
 \label{subsec:practicalApproach:instantiation}
 
-As stated before, any MC$^2$ algorithm may be used. Here, we choose
-the structured large margin approach~\cite{Taskar:2005}. Let
-$\lc:\s\times\A\rightarrow\mathbb{R}_+$ be a user-defined margin
-function satisfying $\lc(s,\pi_E(s))\leq \lc(s,a)$ (here,
-$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). The MC$^2$ algorithm
-solves:% the following quadratic problem:
+Comme précisé précedemment, tout algorithme de MC$^2$ peut être utilisé. Ici nous choisissons l'approche à marges structurées de ~\cite{Taskar:2005}. Soit
+$\lc:\s\times\A\rightarrow\mathbb{R}_+$ une fonction de marge définie par l'utilisateur satisfaisant $\lc(s,\pi_E(s))\leq \lc(s,a)$ (ici,
+$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). L'algorithme MC$^2$ résoud :% the following quadratic problem:
 \begin{equation}
@@
 \end{equation}
-Following~\cite{Ratliff:2006}, we express the equivalent hinge-loss
-form (noting that the slack variables $\zeta_i$ are tight, which
-allows moving the constraints in the objective function):
+Comme dans~\cite{Ratliff:2006}, nous fournissons la forme \emph{hinge-loss} équivalente (avec les variables d'ajustement $\zeta_i$ serrées, ce qui permet de déplacer les contraintes dans la fonction objectif) :
 \begin{equation}
@@
 \end{equation}
-This objective function is minimized using a subgradient descent.
-The expert feature expectation is estimated using the scheme
-described in Eq.~\eqref{eq:mc_plus_heuristic}.
+La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit Eq.~\eqref{eq:mc_plus_heuristic}.
 

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Travaux connexes traduit

@@
 
-\section{Related Works}
+\section{Travaux connexes}
 \label{sec:relatedWorks}
 
-The notion of IRL has first been introduced in~\cite{Russell:1998}
-and first been formalized in~\cite{Ng:2000}. A classic approach to
-IRL, initiated in~\cite{Abbeel:2004}, consists in finding a policy
-(through some reward function) such that its feature expectation (or
-more generally some measure of the underlying trajectories'
-distribution) matches the one of the expert policy.
+La notion d'ARI a été pour la première fois introduite dans~\cite{Russell:1998}
+puis formalisée dans~\cite{Ng:2000}.
+Une approche classique, initiée par~\cite{Abbeel:2004}, consiste à trouver une politique (à travers une recherche dans l'espace des fonctions de récompense) telle que son attribut moyen (ou de manière plus générale une mesure de la distribution sous jacente aux trajectoires) s'approche de celui de la politique de l'expert.
 %
-See~\cite{Neu:2010} for a review.
+Voir~\cite{Neu:2010} pour un survey.
 %
@@
 %
-Notice that related algorithms are not always able to output a
-reward function, even if they may make use of IRL as an intermediate
-step. In such case, they are usually refereed to as apprenticeship
-learning algorithms.
+Certains algorithmes mentionnés ne sont pas capable de renvoyer une fonction de récompense, bien qu'ils utilisent l'ARI comme étape.
+On parle de généralement de ces algorithmes comme d'algorithmes d'appretissage par imitation.
 %
@@
 
-Closer to our contribution, some approaches also somehow introduce a
-structure in a classification
-procedure~\cite{Melo:2010}\cite{Ratliff:2006}. In~\cite{Melo:2010},
-a metric induced by the MDP is used to build a kernel which is used
-in a classification algorithm, showing improvements compared to a
-non-structured kernel. However, this approach is not an IRL
-algorithm, et more important assessing the metric of an MDP is a
-quite involved problem. In~\cite{Ratliff:2006}, a classification
-algorithm is also used to produce a reward function. However,
-instead of associating actions to states, as we do, it associates
-optimal policies (labels) to MDPs (inputs), which is how the
-structure is incorporated. This involves solving many MDPs.
+Plus proche de notre contribution, quelques approches introduisent également en quelque sorte de la structure dans la classification~\cite{Melo:2010}\cite{Ratliff:2006}. Dans~\cite{Melo:2010},
+une métrique induite par le PDM est utilisée pour construire un noyau qui sera utilisé par l'algorithme de classification, permettant des améliorations par rapport à un noyau non structuré.
+Cette approche n'est cependant pas un algorithme d'ARI, et plus important l'évaluation d'une métrique dans un PDM n'est pas une mince affaire.
+Dans~\cite{Ratliff:2006}, un algorithme de classification est utilisé pour fournir une fonction de récompense.
+Au lieu d'associer des actions à des états comme nous le faisons, cette algorithme associe des politiques optimales (qui jouent le rôle de labels) ) des PDM (entrées), ce qui permet d'incorporer la structure, au prix de la résolution d'un grand nombre de PDM.
 
-As far as we know, all IRL algorithms require solving the direct RL
-problem repeatedly, except~\cite{Dvij:2010,boularias:2011}.
-\cite{Dvij:2010} applies to linearly-solvable MDPs (where the
-control is done by imposing any dynamic to the system).
-In~\cite{boularias:2011}, based on a relative entropy argument, some
-utility function is maximized using a subgradient ascent. Estimating
-the subgradient requires sampling trajectories according to the
-policy being optimal for the current estimated reward. This is
-avoided thanks to the use of importance sampling. Still, this
-requires sampling trajectories according to a non-expert policy and
-the direct problem remains at the core of the approach (even if
-solving it is avoided).
+Tous les algorithmes d'ARI à notre connaissance requierent la résolution du problème direct de l'AR de manière répétée, à l'exception de~\cite{Dvij:2010,boularias:2011}.
+\cite{Dvij:2010} ne s'applique qu'aux PDM solvables linéairement (où le contrôle se fait en imposant une dynamique au système).
+Dans~\cite{boularias:2011}, en utilisant l'argument de l'entropie relative, on maximise une fonction objectif via une montée de sous gradient.
+Estimer la valeur du sous gradient demande des trajectoires échantillonnées selon la politique optimale pour la fonction de récompense courante.
+On peut contourner le problème grâce à l'\emph{importance sampling}.
+Cela requière cependant d'échantillonner des trajectoires selon une politique différente de celle de l'expert, et la du problème direct se maintient au coeur de l'approche (même si on évite sa résolution).
 
-SCIRL does not require solving the direct problem, just estimating
-the feature expectation of the expert policy. In other words,
-instead of solving multiple policy optimization problems, we only
-solve one policy evaluation problem. This comes with theoretical
-guarantees (which is not the case of all IRL algorithms,
-\textit{e.g.}~\cite{boularias:2011}). Moreover, using heuristics
-which go beyond our analysis, SCIRL may rely solely on data provided
-by expert trajectories. We demonstrate this empirically in the next
-section. To the best of our knowledge, no other IRL algorithm can
-work in such a restrictive case.
+SCIRL n'a pas besoin de résoudre le problème direct, mais uniquement d'estimer l'attribut moyen de la politique de l'expert.
+Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une fois un problème d'évaluation de la politique.
+Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI \textit{e.g.}~\cite{boularias:2011}).
+De plus par l'utilisation d'une heuristique qui dépassent le cadre de notre analyse, il est possible à SCIRL de se contenter de données fournies par l'expert.
+La prochaine section présente une démonstration empirique de cela.
+Nous ne connaissons aucun autre algorithme d'ARI en mesure de fonctionner dans des conditions si drastiques.
 

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Fin du premier jet de la traduction.

@@
 
-\section{Experiments}
+\section{Expériences}
 \label{sec:experiments}
 
-We illustrate the proposed approach on a car driving simulator,
-similar to~\cite{Abbeel:2004,Syed:2008:game}. The goal si to drive a
-car on a busy three-lane highway with randomly generated traffic
-(driving off-road is allowed on both sides). The car can move left
-and right, accelerate, decelerate et keep a constant speed. The
-expert optimizes a handcrafted reward $\R_E$ which favours speed,
-punish off-road, punish collisions even more  et is neutral
-otherwise.
+Nous illustrons l'approche proposée sur un simulateur de conduite similaire à~\cite{Abbeel:2004,Syed:2008:game}.
+Le but est de conduire une voiture sur une autoroute à trois voies dont le traffic est généré aléatoirement (les sorties de route sont possible des deux côtés).
+La voiture peut se déplacer vers la gauche ou la droite, accélérer ou ralentir et conserver sa vitesse.
+L'expert optimise une récompense définie par nos soins $\R_E$ qui récompense la vitesse, punit les sorties de route, punit sévèrement les collisions et no donne pas d'information dans les autres cas.
 
-We compare SCIRL as instantiated in
-Sec.~\ref{subsec:practicalApproach:instantiation} to the
-unstructured classifier (using the same classification algorithm)
-and to the algorithm of~\cite{Abbeel:2004} (called here PIRL for
-Projection IRL). We also consider the optimal behavior according to
-a randomly sampled reward function as a baseline (using the same
-reward feature vector as SCIRL et PIRL, the associated parameter
-vector is randomly sampled).
+Nous avons comparé SCIRL tel qu'instancié comme décrit en
+Sec.~\ref{subsec:practicalApproach:instantiation} à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au coeur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbeel:2004} (appelé ici PIRL pour \emph{
+Projection Inverse Reinforcement Learning}).
+Nous nous préoccupons également du comportement optimal vis-à-vis d'une récompense tirée aléatoirement (en utilisant la même paramétrisation que SCIRL et PIRL, le vecteur de paramètres est tiré selon une loi uniforme) afin de disposer d'un point de comparaison.
 
-For SCIRL et PIRL we use a discretization of the state space as the
-reward feature vector, $\phi\in\mathbb{R}^{729}$: $9$ horizontal
-positions for the user's car, $3$ horizontal et $9$ vertical
-positions for the closest traffic's car et $3$ speeds. Notice that
-these features are much less informative than the ones used
-in~\cite{Abbeel:2004,Syed:2008:game}. Actually,
-in~\cite{Syed:2008:game} features are so informative that sampling a
-random positive parameter vector $\theta$ already gives an
-acceptable behavior. The discount factor is $\gamma = 0.9$. The
-classifier uses the same feature vector reproduced for each action.
+Pour SCIRL et PIRL on discrétise l'espace d'état en un vecteur d'attribut pour la récompense, $\phi\in\mathbb{R}^{729}$: $9$ positions horizontales pour la voiture du joueur, $3$ positions horizontales et $9$ verticales pour la voiture la plus proche de celle du joueur et $3$ vitesses.
+Ces attributs sont bien moins informatifs que ceux utilisés dans~\cite{Abbeel:2004,Syed:2008:game}.
+Les attributs de~\cite{Syed:2008:game} sont si informatifs que tirer aléatoirement un vecteur de paramètres $\theta$ positifs pour la récompense donne lieu à un comportement acceptable.
+Le facteur d'oubli est $\gamma = 0.9$.
+Le classifieur utilise le même vecteur d'attribut reproduit pour chaque action.
 
-SCIRL is fed with $n$ trajectories of length $n$ (started in a
-random state) with $n$ varying from $3$ to $20$ (so fed with $9$ to
-$400$ transitions). Each experiment is repeated 50 times. The
- classifier uses the same data. PIRL is an iterative
-algorithm, each iteration requiring to solve the MDP for some reward
-function. It is run for 70 iterations, all required objects (a
-feature expectations for a non-expert policy et an optimal policy
-according to some reward function at each iteration) are computed
-exactly using the model. We measure the performance of each approach
-with $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, where $\mathcal{U}$
-is the uniform distribution (this allows measuring the
-generalization capability of each approach for states infrequently
-encountered), $\R_E$ is the expert reward et $\pi$ is one of the
-following polices: the optimal policy for $\R_E$ (upper baseline),
-the optimal policy for a random reward (lower baseline), the optimal
-policy for $\R_{\theta_c}$ (SCIRL), the policy produced by PIRL and
-the classifier decision rule.
+On donne à SCIRL $n$ trajectoires de longueur $n$ (débutant dans un état choisi aléatoirement), $n$ allant de  $3$ à $20$ (ce qui donne de $9$ à
+$400$ transitions).
+Chaque expérience est répétée 50 fois.
+Le classifieur utilise les mêmes données que SCIRL.
+PIRL est un algorithme itératif dont chaque itération implique la résolution du PDM pour une fonction de récompense arbitraire.
+Nous l'avons fait fonctionner pendant 70 itérations, tous les objets requis (l'attribut moyen de politiques différentes de celles de l'expert et la politique optimale pour une fonction de récompense à chaque itération) ont été calculées de manière éxacte en utilisant le modèle.
+Nous mesurons les performances de chacune des approches grâce à $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, avec $\mathcal{U}$
+la distribution uniforme (ce qui permet de tester la capacité de généralisation de chaque approche même pour des états infrequemment recontrés), $\R_E$ est la récompense de l'expert et $\pi$ est l'une des politiques suivantes: la politique optimale pour $\R_E$ (point de référence haut),
+la politique optimale pour une récompense aléatoire (point de référence bas), la politique optimale pour $\R_{\theta_c}$ (SCIRL), et la politique produite par PIRL et la règle de décision du classifieur.
 
-\begin{figure}[fig:res]{Highway problem.
-The highest line is the expert value.
-For each curves, we show the mean (plain line), the standard deviation (dark color) et the min-max values (light color).
-The policy corresponding to the random reward is in blue, the policy outputted by the  classifier is in yellow et the optimal policy according the SCIRL's reward is in red.
-PIRL is the dark blue line.}
+\begin{figure}[fig:res]{Problème de la conduite.
+La ligne la plus haute est la valeur de l'expert.
+Pour chaque courbe, nous dessinons la moyenne (trait plein), l'écart-type (foncé) et les valeurs min et max (plus clair).
+La pôlitique correspondant à une récompense aléatoire est en bleu, la politique retournée par le classifieur est en jaune et la politique optimale pour la récompense trouvée par SCIRL's est en rouge.
+La ligne bleue foncée correspond à PIRL.}
 \begin{minipage}[l]{0.45\linewidth}
@@
 
-Fig.~\ref{fig:res} shows the performance of each approach as a
-number of used expert transitions (except PIRL which uses the
-model). We can see that the classifier does not work well on this
-example. Increasing the number of samples would improve its
-performance, but after 400 transitions it does not work as well as
-SCIRL with only a ten of transitions. SCIRL works pretty well here:
-after only a hundred of transitions it reaches the performance of
-PIRL, both being close to the expert value. We do not report exact
-computational times, but running SCIRL one time with $400$
-transitions is approximately hundred time faster than running PIRL
-for $70$ iteration.
+La Fig.~\ref{fig:res} montre la performance de chacune des approche en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
+On peut voir que le classifieur ne fonctionne pas bien sur ce problème.
+Augmenter le nombre de transitions améliorerait ses performances, mais après 400 transitions, il ne fonctionne pas aussi bien que SCIRL avec uniquement une dizaine de transition.
+SCIRL fonctionne particulièrement bien ici: après seulement une centaine de transitions il atteint les performances de PIRL.
+Ces deux algorithmes sont proches de la valeur de l'expert.
+Nous ne fournissons pas de données exactes en ce qui concerne les temps de calcul, mais faire fonctionner SCIRL une fois avec $400$ transitions en entrée est environ une centaine de fois plus rapide que de faire fonctionner PIRL pour $70$ itérations.
 %
@@
 
-We have introduced a new way to perform IRL by structuring a
-linearly parameterized score function-based multi-class
-classification algorithm with an estimate of the expert feature
-expectation. This outputs a reward function for which we have shown
-the expert to be near optimal, provided a small classification error
-and a good expert feature expectation estimate. How to practically
-estimate this quantity has been  discussed et we have introduced a
-heuristic for the case where only transitions from the expert are
-available, along with a specific instantiation of the SCIRL
-algorithm. We have shown on a car driving simulator benchmark that
-the proposed approach works well (even combined with the introduced
-heuristic), much better than the unstructured classifier et as well
-as a state-of-the-art algorithm making use of the model (and with a
-much lower computational time). In the future, we plan to deepen the
-theoretical properties of SCIRL (notably regarding possible
-heuristics) et to apply it to real-world robotic problems.
+Nous avons introduit un nouveau moyen de résoudre le problème de l'ARI en structurant un classifieur à fonction de score linéairement paramétrée avec une estimée de l'attribut moyen de l'expert.
+Il renvoie une fonction de récompense pour laquelle nous avons montré que l'expert est quasi optimal, pour peu que l'erreur de classification soit faible et que l'estimation de l'attribut moyen de l'expert soit bonne.
+Des méthodes pratiques pour l'estimation de cette grandeur ont été présentées et nous avons introduit une heuristique pour le cas où les seules données disponibles proviennent de l'expert, ainsi qu'une instanciation spécifique de l'algorithme SCIRL.
+Nous avons exhibé grâce à un simulateur de conduite sur autoroute que l'approche proposée fonctionne bien (même combinée à l'heuristique discutée), bien mieux qu'un classifieur non structuré et aussi bien que la méthode de l'état de l'art qui avait accès au modèle (et avec un coût computationel moindre).
+Nous prévoyons de poursuivre notre analyse des propritétés théoriques de SCIRL (notamment en vue de l'utilisation avec des heuristiques) et l'appliquer à des problèmes robotiques réels.
 
-\acknowledgements{This research was partly funded by the EU FP7 project ILHAIRE (grant
-n$\deg$270780), by the EU INTERREG IVa project ALLEGRO et by the Région
+\acknowledgements{Ces travaux ont été en partie financés par le projet EU FP7 ILHAIRE (grant
+n$\deg$270780), par le projet EU INTERREG IVa  ALLEGRO et par la Région
 Lorraine (France).}

### 2012-09-21 klein2012structured_RIA/scirl_ria.tex
Traduction de l'intro, oublié lors du copier coller original

@@
 \newpage
+\section{Introduction}
+\label{sec:intro}
+
+L'Apprentissage par Renforcement Inverse (ARI)~\cite{Russell:1998} est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert est optimal.
+Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{sec:relatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) qui suit celle générée par l'expert.
+Souvent, cette distribution est caractérisée par ce que l'on appelle l'attribut moyen (voir Sec.~\ref{subsec:background:irl}) : étant donnée une fonction de récompense linéairement paramétrée par un vecteur d'attributs, il s'agit de l'espérance de la somme pondérée des vecteurs d'attributs sachant que l'on commence dans un certain état en choisissant une certaine action avant de suivre la politique concernée.
+
+Dans cette publication, nous choisissons une autre option.
+Il est possible d'imiter le comportement de l'expert via un algorithme d'apprentissage supervisé généralisant l'association des actions aux états.
+Nous considérons ici les classifieurs multi-classes qui à partir d'un set d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour une état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subsec:background:classif}).
+L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir Sec.~\ref{subsec:scirl:algo}).
+Le vecteur de paramètres ainsi calculé définit une fonction de récompense dont on montre qu'elle admet pour politique quasi-optimale la politique de l'expert (voir Sec.~\ref{subsec:scirl:analysis}).
+
+Un grand avantage de SCIRL est qu'il ne requiert pas, contrairement à la plupart des algorithmes d'ARI, la résolution répétée du problème direct, celui de l'apprentissage par renforcement.
+Il nécessit d'estimer l'attribut moyen de l'expert, mais ceci est équivalent un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir Sec.~\ref{sec:practicalApproach}.
+De plus, à l'aide d'une heuristique, SCIRL peut s'éxécuter en utilisant uniquement en utilisant des transitions issues de la politique de l'expert (il n'y a pas besoin d'échantilloner la dynamique complète).
+Nous illustrons cela sur un simulateur de conduite Sec.~\ref{sec:experiments}.
 

### 2012-09-21 klein2012structured_RIA/scirl_ria.tex
Modification suggérées par Matthieu après la première relecture.

@@
 \usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 % \journalName{year}{volume}{issue}
@@
 \let\chapter\section
-\usepackage[ruled,algo2e]{algorithm2e}
+\usepackage[ruled,algo2e,french]{algorithm2e}
 \usepackage{mathtools}
 \usepackage{subfigure}
+\usepackage{ gensymb }
 \mathtoolsset{showonlyrefs=true}
@@
 % \title[short header title]{title}
-\title[SCIRL]{Classification structurée pour l'apprentissage par renforcement}
+\title[SCIRL]{Classification structurée pour l'apprentissage par renforcement inverse}
 \author[1,2]{Edouard}{Klein}
+\author[2,3]{Matthieu}{Geist}
 \author[2]{Bilal}{Piot}
-\author[2,3]{Matthieu}{Geist}
 \author[2,3]{Olivier}{Pietquin}
@@
 % addresses are automatically numbered
-\address{LORIA -- team ABC\\
+\address{LORIA -- équipe ABC\\
 Nancy, France
 }{}
-\address{Supélec -- IMS-MaLIS Research Group\\
+\address{Supélec -- Groupe de recherche IMS-MaLIS\\
 Metz, France}
@@
   driving simulator.}
-\resume{Cette contribution traite du problème de l'apprentissage par imitation par le biais de l'apprentissage par renforcement inverse (\emph{ARI}).
-Dans ce contexte, un expert accomplit une tâche qu'un agent artificiel doit essayer de reproduire.
-L'ARI part du postulat que l'expert optimise avec succès une fonction de récompense ; le problème consiste à deviner cette fonction à partir de  traces du comportement de l'expert.
-Les algorithmes d'ARI existants nécessitent une ou plusieurs des conditions suivantes pour fonctionner : trajectoires complètes de la part de l'expert, un modèle génératif pour les estimations de type Monte-Carlo, la connaissance des probabilités de transition, la capacité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée ou l'accès à la strategie complète de l'expert.
-Notre contribution consiste en un nouvel algorithme d'ARI levant l'ensemble de ces contraintes.
-En utilisant une méthode supervisée dans laquelle nous introduisons implicitement la structure du processus décisionnel de Markov ({\it PDM}) sous-jacent, nous créons un algorithme basé sur une descente de sous-gradient, possèdant une faible complexité tant en échantillons que calculatoire et surtout ne nécessitant pas la résolution du problème direct.}
+\resume{Cette contribution traite le problème de l'Apprentissage par Renforcement Inverse (ARI), défini comme la recherche d'une fonction de récompense pour laquelle le comportement d'un expert (connu par le biais d'une démonstration) est optimal.
+Nous introduisons SCIRL, un nouvel algorithme qui utilise la grandeur dénommée attribut moyen de l'expert comme la paramétrisation d'une fonction de score pour un classifieur multi-classe.
+Cette approche donne une fonction de récompense pour laquelle la politique de l'expert est (nous le démontrons) quasi-optimale.
+Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct de l'Apprentissage par Renforcement.
+De plus, en utilisant une heuristique il est utilisable uniquement avec des trajectoires échantillonnées par l'expert.
+Nous illustrons cela sur un simulateur de conduite.}
 
@@
 \begin{document}
-\newtheorem{prop}[theorem]{Proposition}
 \maketitle
@@
 
-Un Processus Décisionnel de Markov (PDM)~\cite{Puterman:1994} est un ensemble $\{\s,\A,\p,\R,\gamma\}$ avec $\s$ l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ l'espace d'action fini, $\p =
+Un Processus Décisionnel de Markov (PDM)~\cite{Puterman:1994} est un tuple $\{\s,\A,\p,\R,\gamma\}$ avec $\s$ l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ l'espace d'action fini, $\p =
 \{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ la fonction de récompense sur les états et $\gamma$ le facteur d'oubli.
 Une politique déterministe $\pi\in\s^\A$ définit le comportement d'un agent.
-On quantifie la qualité de ce comportement par le biais de la fonction de valeur $v_\R^\pi\in\mathbb{R}^\s$ qui à chaque état associe la récompense pondérée cumulée recueillie par l'agent lorsqu'il part de cet état et suit la politique $\pi$ par la suite : $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t \R(S_t)|S_0=s,\pi]$.
-Une politique optimale $\pi_\R^*$ (vis-à-vis de la fonction de récompense $\R$) est une politique dont la fonction de valeur $v^*_\R$ satisfait $v_\R^* \geq v_\R^\pi$, composante par composante, pour toute politique $\pi$.
+La qualité moyenne de ce comportement est quantifiée par le biais de la fonction de valeur $v_\R^\pi\in\mathbb{R}^\s$, qui à chaque état associe la récompense pondérée cumulée recueillie par l'agent lorsqu'il part de cet état et suit la politique $\pi$ par la suite : $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t \R(S_t)|S_0=s,\pi]$.
+Une politique optimale $\pi_\R^*$ (vis-à-vis de la fonction de récompense $\R$) est une politique dont la fonction de valeur $v^*_\R$ vérifie $v_\R^* \geq v_\R^\pi$ (composante par composante) pour toute politique $\pi$.
 
-Soit $P_\pi$ la matrice stochastique $P_\pi =
+Soit $P_\pi$ la matrice stochastique définie par $P_\pi =
 (p(s'|s,\pi(s)))_{1\leq s,s'\leq |\s|}$.
@@
 \rightarrow \mathbb{R}^\s$ sont définis par $T^\pi_\R v = \R + \gamma
-P_\pi v$ and $T_\R^*v = \max_\pi T_\R^\pi v$.
+P_\pi v$ et $T_\R^*v = \max_\pi T_\R^\pi v$.
 %\begin{equation}
@@
 Ces opérateurs sont des contractions qui admettent $v_\R^\pi$ et $v^*_\R$ comme points fixes respectifs : $v_\R^\pi = T^\pi_\R v_\R^\pi$ et $v^*_\R = T^*_\R v^*_\R$.
-La fonction de qualité $Q^\pi\in\mathbb{R}^{\s\times \A}$ ajoute à $v$ un degré de liberté dans le choix de la première action, on la définit formellement par $Q_\R^\pi(s,a)
-= [T^a_\R v^\pi_\R](s)$. On note $\rho_\pi$ la distribution stationnaire de la politique $\pi$ (qui satisfait $\rho_\pi^\top P_\pi = \rho_\pi^\top$).
+La fonction de qualité $Q^\pi\in\mathbb{R}^{\s\times \A}$ ajoute un degré de liberté dans le choix de la première action, on la définit formellement par $Q_\R^\pi(s,a)
+= [T^a_\R v^\pi_\R](s)$. La distribution stationnaire de la politique $\pi$ est notée $\rho_\pi$ elle satisfait $\rho_\pi^\top P_\pi = \rho_\pi^\top$.
 
-L'Apprentissage par Renforcement (AR) et la programmation dynamique approchée cherchent à estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
+L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
 %
-On appelle cela le problème direct.
+C'est le problème direct.
 %
-L'Apprentissage par Renforcement Inverse (approché)~\cite{Ng:2000} cherche à estimer la fonction de récompense pour laquelle une politique observée est optimale.
+L'Apprentissage par Renforcement Inverse (approché)~\cite{Ng:2000} consiste à estimer la fonction de récompense pour laquelle une politique observée est optimale.
 %
-On nomme cette politique la politique de l'expert, notée $\pi_E$.
-On suppose qu'elle est optimale vis-à-vis d'une certaine fonction de récompense $R_E$ inconnue.
+Cette politique est la politique de l'expert, notée $\pi_E$.
+Elle est supposée optimale vis-à-vis d'une certaine fonction de récompense $\mathcal{R}_E$ inconnue.
 Le but de l'ARI est de calculer une fonction de récompense $\hat{\R}$ telle que la politique de l'expert soit (quasiment) optimale, c'est-à-dire telle que $v^*_{\hat{\R}}
@@
 De la même manière que pour le problème direct, l'espace d'état peut être trop grand pour que la fonction de récompense admette une représentation exacte exploitable.
-On se limite donc de fait à la recherche d'une bonne fonction de récompense paramétrée linéairement.
+La recherche est donc limitée à celle d'une bonne fonction de récompense paramétrée linéairement.
 Soit $\phi(s) = (\phi_1(s)  \dots
@@
 %\end{pmatrix}^\top$
-un vecteur d'attributs composé de $p$ fonctions de base $\phi_i\in\mathbb{R}^\s$, on définit les fonctions de récompense paramétrées par $\R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p
+un vecteur d'attributs composé de $p$ fonctions de base $\phi_i\in\mathbb{R}^\s$, les fonctions de récompense paramétrées sont définies par $\R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p
 \theta_i \phi_i(s)$.
@@
 %\end{equation}
-La recherche d'une bonne récompense est donc ramenée à la recherche d'un bon vecteur de paramètres vector $\theta \in\mathbb{R}^p$.
+La recherche d'une bonne récompense est donc ramenée à la recherche d'un bon vecteur de paramètres $\theta \in\mathbb{R}^p$.
 Nous utiliserons indifféremment $\R_\theta$ et $\theta$ comme indices (\textit{e.g.}, $v_\theta^\pi$ pour $v_{\R_\theta}^\pi$).
@@
 \begin{equation}
-  Q^\pi_\theta(s,a) = \theta^\top \mu^\pi(s,a) \text{ with }
+  Q^\pi_\theta(s,a) = \theta^\top \mu^\pi(s,a) \text{ avec }
   \mu^\pi(s,a) = \E[\sum_{t\geq 0} \gamma^t
@@
 Conséquemment, la fonction de qualité partage son vecteur de paramètres avec la fonction de récompense, mais en relation avec un vecteur d'attributs $\mu^\pi$ appelé l'attribut moyen.
-Cette notion est de prime importance à notre contribution.
-Chaque composante $\mu_i^\pi$ de ce vecteur d'attributs est en réalité une fonction de qualité de la politique $\pi$ où la récompense serait $\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$.
+Cette notion est de prime importance pour notre contribution.
+Chaque composante $\mu_i^\pi$ de ce vecteur d'attributs est en réalité la fonction de qualité de la politique $\pi$ pour la récompense $\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$.
 De fait, tout algorithme d'estimation de la fonction de qualité peut être utilisé pour estimer l'attribut moyen, comme une estimation de Monte-Carlo ou un algorithme aux différences temporelles~\cite{Klein:2011}.
@@
 Soit $\X$ un ensemble fini ou compact (d'entrées à classifier)
- et soit $\Y$ un ensemble fini de (de labels). Supposons que les entrées $x\in
+ et soit $\Y$ un ensemble fini (de labels). Supposons que les entrées $x\in
 \X$ sont tirées selon une distribution inconnue $\prob(x)$ et qu'il existe un oracle qui à chacune de ces entrées associe un label $y\in Y$ tiré selon une distribution de probabilité conditionnelle inconnue $\prob(y|x)$.
-De manière générale, la classification multi-classe cherche étant donné un set d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tiré selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
+De manière générale, la classification multi-classe cherche, étant donné un ensemble d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tiré selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
 
 Nous nous préoccupons ici d'un ensemble plus réduit de classifieurs.
-Nous supposons que la règle de décision associe à l'entrée l'argument qui maximise une certaine fonction de score, celle-ci est paramétrée linéairement, les paramètres étant appris par le classifieur.
+Nous supposons que la règle de décision associe à l'entrée l'argument qui maximise une certaine fonction de score, celle-ci étant paramétrée linéairement, les paramètres étant appris par le classifieur.
 Formellement, soit $\psi(s,a) =
@@
 La règle de décision associée $g_w\in{\Y^\X}$ est définie par $g_w(x) \in \argmax_{y\in\Y}s_w(x,y)$.
-En se basant sur un set d'entraînement $\{(x_i,y_i)_{1\leq
-i\leq N}\}$, un classifieur multi-classe (noté MC$^2$ pour \emph{Multi-Class Classifier}) à fonction de score paramétrée linéairement choisit un vecteur de paramètres $\theta_c$.
+En se basant sur un ensemble d'entraînement $\{(x_i,y_i)_{1\leq
+i\leq N}\}$, un classifieur multi-classe (noté CMC) à fonction de score paramétrée linéairement choisit un vecteur de paramètres $\theta_c$.
 La qualité de ce choix est quantifiée par l'erreur de classification $\epsilon_c =
-\prob(g_{\theta_c}(x)\neq y)$.
+\prob(g_{\theta_c}(x)\neq y)$. %Riflette
 
 Notre étude est valable pour tout algorithme de classification tant qu'il opère en maximisant l'argument d'une fonction de score paramétrée linéairement.
-Par exemple, une machine à vecteur support multi-classe pourrait être choisie~\cite{Guermeur:2007} (en prenant le noyau induit par le vecteur de paramètres) de même qu'une approche à marge vaste et structurée~\cite{Taskar:2005}.
-D'autres choix sont possibles, chacun peut choisir son algorithme fétiche.
+Par exemple, une machine à vecteur support multi-classe pourrait être choisie~\cite{Guermeur:2007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taskar:2005}.
+D'autres choix sont possibles.
 
@@
 
-On se place dans le cadre de la classification décrit en Sec.~\ref{subsec:background:classif}.
-L'entrée $x$ oeut être vue comme un état et le label $y$ comme une action.
-Il s'ensuit que l'on peut interpréter la règle de décision
-$g_w(x)$ comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle même être vue comme une fonction de qualité.
-On peut alors tirer le parallèle avec l'Eq.~\eqref{eq:def:mu}, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels du set d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
+Nous nous trouvons dans le cadre de la classification décrit en Sec.~\ref{subsec:background:classif}.
+L'entrée $x$ peut être vue comme un état et le label $y$ comme une action.
+Il s'ensuit que la règle de décision
+$g_w(x)$ est interprétable comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle-même être vue comme une fonction de qualité.
+Le parallèle avec l'Eq.~\eqref{eq:def:mu} est facilement tiré, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de l'ensemble d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
 Ces remarques nous permettent maintenant d'introduire notre algorithme d'ARI par classification structurée (SCIRL pour \emph{Structured Classification-based Inverse Reinforcement Learning}).
 
-Soit $\pi_E$ la politique de l'expert à partir de laquelle on souhaite inférer une fonction de récompense.
-Supposons l'existence d'un set d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est \emph{fast-mixing}, l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés d'après cette distribution.} $\rho_E = \rho_{\pi_E}$.
-Supposons également que nous avons à notre disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eq:def:mu}.
-Une description de la manière d'estimer cette quantité en pratique est remise à la Sec.~\ref{subsec:practicalApproach:muE}; rappelons en revanche qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé Sec.~\ref{subsec:background:irl}.
-Supposons enfin qu'un algorithme de MC$^2$ a été choisi.
-L'algorithme formant notre contribution consiste simplement à choisir $\theta^\top\hat{\mu}^{\pi_E}(s,a)$ comme la fonction de score paramétrée linéairement, puis à entraîner le classifieur sur $\D$, ce qui produit un vecteyr de paramètres $\theta_c$, et enfin à renvoyer la fonction de récompense $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.
+Soit $\pi_E$ la politique de l'expert à partir de laquelle nous souhaitons inférer une fonction de récompense.
+Supposons l'existence d'un ensemble d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est à mélange rapide (\emph{fast-mixing}), l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés d'après cette distribution.} $\rho_E = \rho_{\pi_E}$.
+Supposons également avoir à disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eq:def:mu}.
+Une description de la manière d'estimer cette quantité en pratique est reportée à la Sec.~\ref{subsec:practicalApproach:muE}; rappelons tout de même qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé Sec.~\ref{subsec:background:irl}.
+Supposons enfin qu'un algorithme de CMC a été choisi.
+L'algorithme formant notre contribution consiste simplement à choisir $\theta^\top\hat{\mu}^{\pi_E}(s,a)$ comme fonction de score paramétrée linéairement, puis à entraîner le classifieur sur $\D$, ce qui produit un vecteur de paramètres $\theta_c$, puis enfin à renvoyer la fonction de récompense $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.
 
@@
   \BlankLine
-  \emph{\textbf{Etant donné}} un set d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
-  une estimée $\hat{\mu}^{\pi_E}$ de l'attridut moyen de l'expert $\mu^{\pi_E}$ et an un algorithme de MC$^2$\;
+  \emph{\textbf{Etant donné}} un ensemble d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
+  une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ et un algorithme de CMC\;
   %
   \BlankLine
-  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de MC$^2$ auquel sont fournis le set d'entraînement $\D$ et la paramétrisation de la fonction de score : $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
+  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de CMC auquel sont fournis l'ensemble d'entraînement $\D$ et la paramétrisation de la fonction de score : $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
   %
@@
 
-L'approche proposée est résumé en Alg.~\ref{algo:scirl}.
-Le nom de l'algorithme (Classification structurée pour l'ARI) vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
+L'approche proposée est résumé par l'Alg.~\ref{algo:scirl}.
+Le nom de l'algorithme (SCRIL, anglais pour Classification structurée pour l'ARI) vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
 Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct.
@@
 
-Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi-optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément que 
-$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ est faible.
+Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi-optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément que le terme
+$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ peut être contrôlé.
 Avant de présenter le résultat principal, il nous faut introduire quelques notations et définir quelques objets.
 
-Nous allons utiliser le coefficient de concentration de la distribution pondérée du premier ordre pour l'état futur $C_f$~\cite{Munos:2007}:
+Nous allons utiliser le coefficient de concentration $C_f$~\cite{Munos:2007}:
 \begin{equation}
-  C_f = (1-\gamma)\sum_{t\geq 0} \gamma^t c(t) \text{ with } c(t) =
+  C_f = (1-\gamma)\sum_{t\geq 0} \gamma^t c(t) \text{ avec } c(t) =
   \max_{\pi_1,\dots,\pi_t,s\in\s}\frac{(\rho_E^\top P_{\pi_1}\dots
@@
 \end{equation}
-On note $\pi_c$ la règle de décision du classifieur : $\pi_c(s) \in
+La règle de décision du classifieur est notée $\pi_c(s) \in
 \argmax_{a\in \A} \theta_c^\top\hat{\mu}^{\pi_E}(s,a)$. L'erreur de classification est donc $\epsilon_c =
 \E_{s\sim\rho_E}[\chi_{\{\pi_c(s)\neq\pi_E(s)\}}] \in [0,1]$.
-On écrit $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$ la fonction de score calculée à partir du set d'entraînement $\D$ (celle-ci peut être interprétée comme une fonction de qualité approchée).
+On écrit $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$ la fonction de score calculée à partir de l'ensemble d'entraînement $\D$ (celle-ci peut être interprétée comme une fonction de qualité approchée).
 Soit
@@
 \theta_c^\top(\hat{\mu}^{\pi_E} - \mu^{\pi_E}) = \theta_c^\top
-\epsilon_\mu:\s\times\A\rightarrow\mathbb{R}$. Finalement, on définit le delta-max moyen de l'erreur sur la fonction de qualité par $\bar{\epsilon}_Q =
+\epsilon_\mu:\s\times\A\rightarrow\mathbb{R}$. Finalement, on définit l'erreur moyenne d'écart maximal sur la fonction de qualité par $\bar{\epsilon}_Q =
 \E_{s\sim\rho_E}[\max_{a\in\A}\epsilon_Q(s,a) -
@@
   Soit $\R_{\theta_c}$ la fonction de récompense renvoyée par l'Alg.~\ref{algo:scirl}. Soient $C_f$, $\epsilon_c$
-  et $\bar{\epsilon}_Q$ les quantités définies ci-dessus. On a
+  et $\bar{\epsilon}_Q$ les quantités définies ci-dessus. On a :
   %
@@
 \begin{proof}
-  La démonstration repose uniquement sur la récompense $\R_{\theta_c}$, on omet donc par soucis de clarté des notations certains indices relatifs à la récompense 
+  La démonstration repose uniquement sur la récompense $\R_{\theta_c}$, donc par soucis de clarté certains indices relatifs à la récompense sont omis des notations  
   (\textit{e.g.}, $v^\pi$ pour
   $v^\pi_{\theta_c}=v^\pi_{\R_{\theta_c}}$ ou $\R$ pour $\R_{\theta_c}$).
-Tout d'abord, l'on relie  l'erreur $\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)]$ au résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s)-v^{\pi_E}(s)]$.
+Tout d'abord, l'erreur $\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)]$ est reliée au résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s)-v^{\pi_E}(s)]$.
   %
@@
   %
-  Composante par composante, on a :
+  Composante par composante :
   \begin{align}
@@
   L'inégalité $(a)$ est valable car $T^{\pi^*} v^{\pi_E}\leq T^*
-  v^{\pi_E}$ et l'inégalité $(b)$ l'est aussi grâce à~\cite[Lemma~4.2]{Munos:2007}.
-De plus, $v^*$ étant optimale, l'on voit que  $v^*-v^{\pi_E}\geq 0$ et avec $T^*$ l'opérateur d'optimalité de Bellman, on a $T^* v^{\pi_E}\geq
+  v^{\pi_E}$ et l'inégalité $(b)$ l'est en vertu de à~\cite[Lemme~4.2]{Munos:2007}.
+De plus, $v^*$ étant optimale, il est visible que  $v^*-v^{\pi_E}\geq 0$ et avec $T^*$ l'opérateur d'optimalité de Bellman, $T^* v^{\pi_E}\geq
   T^{\pi_E}v^{\pi_E}=v^{\pi_E}$. De plus, on remarque que 
@@
   Donc, d'après la définition du coefficient de concentration 
-  $C_f$, on a :
+  $C_f$ :
   \begin{equation}
@@
   \end{equation}
-  Ce résultat est similaire à celui de~\cite[Theorem~4.2]{Munos:2007}. Il reste à borner le résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
+  Ce résultat est similaire à celui de~\cite[Theoreme~4.2]{Munos:2007}. Il reste à borner le résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
   v^{\pi_E}(s)]$. Considérons la décomposition
@@
   \end{equation}
-  on va borner  $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
+  Nous allons borner  $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
   et $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]$.
@@
   $\hat{Q}^{\pi_E}=\theta_c^\top\hat{\mu}^{\pi_E}$. Donc, pour chaque couple état-action
-  $(s,a)\in\s\times \A$ on a:
+  $(s,a)\in\s\times \A$ :
   \begin{equation}
@@
   \end{align}
-  En passant à l'espérance selon $\rho_E$ et tout en remarquant que 
-  $T^* v^{\pi_E}\geq v^{\pi_E}$, on borne le premier terme
+  En passant à l'espérance selon $\rho_E$ et tout en rappelant que 
+  $T^* v^{\pi_E}\geq v^{\pi_E}$, le premier terme est borné
   \begin{equation}
@@
 
-  On écrit $M\in\mathbb{R}^{|\s|\times |\s|}$ la matrice diagonale définie par $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$.
+   $M\in\mathbb{R}^{|\s|\times |\s|}$ est la matrice diagonale définie par $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$.
 En utilisant cela, l'opérateur de Bellman $T^{\pi_c}$ peut s'écrire, pour tout $v\in\mathbb{R}^\s$:
@@
   Injecter les bornes des Eqs.~\eqref{eq:proof:b1}
-  and~\eqref{eq:proof:b2} dans l'Eq.~\eqref{eq:proof:residual} montre le résultat proposé.%, which concludes the proof.
+  and~\eqref{eq:proof:b2} dans l'Eq.~\eqref{eq:proof:residual} achève la démonstration.%, which concludes the proof.
 %  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
@@
 
-Ce résultat montre que si l'attribut moyen de l'expert est bien estimé (dans le sens d'une faible erreur d'estumation $\epsilon_\mu$ pour les états échantillonnés selon la politique stationnaire de l'expert et pour toutes les actions) et si l'erreur de classification $\epsilon_c$ et également faible alors l'algorithme générique proposé fournit une fonction de récompense
+Ce résultat montre que si l'attribut moyen de l'expert est bien estimé (dans le sens d'une faible erreur d'estumation $\epsilon_\mu$ pour les états échantillonnés selon la politique stationnaire de l'expert et pour toutes les actions) et si l'erreur de classification $\epsilon_c$ et également faible, alors l'algorithme générique proposé fournit une fonction de récompense
 $\R_{\theta_c}$ vis-à-vis de laquelle l'expert sera quasi-optimal. Un corollaire direct du Th.~\ref{th} stipule qu'avec le vrai attribut moyen de l'expert $\mu^{\pi_E}$ et un classifieur parfait 
@@
 
-L'on pourrait arguer que ces bornes sont triviallement valables pour la fonction de récompense nulle (fonction régulièrement citée en exemple de la nature mal posée du problème de l'ARI) correspondant au cas $\theta_c=0$.
+Certains allègueraient que ces bornes sont trivialement valables pour la fonction de récompense nulle (fonction régulièrement citée comme exemple de la nature mal posée du problème de l'ARI) correspondant au cas $\theta_c=0$.
 Cependant il faut se rappeler que le vecteur de paramètres $\theta_c$ est choisi par le classifieur.
 Avec
-$\theta_c=0$, la règle de décision serait une politique aléatoire et nous aurions $\epsilon_c = \frac{|\A|-1}{|\A|}$, c'est-à-dire la pire erreur de classification possible.
+$\theta_c=0$, la règle de décision serait une politique aléatoire uniforme et nous aurions $\epsilon_c = \frac{|\A|-1}{|\A|}$, c'est-à-dire la pire erreur de classification possible.
 Ce cas est donc très improbable.
-De fait nous affirmons que notre approche permet, d'une certaine manière, de lever l'ambiguité du problème de l'ARI (au moins, l'algorithme ne renvoie pas de récompense triviales comme la récompense nulle).
+De fait, nous affirmons que notre approche permet, d'une certaine manière, de lever l'ambiguïté du problème de l'ARI (au moins, l'algorithme ne renvoie pas de récompense triviale comme la récompense nulle).
 %
-On peut également noter que cette borne est invariante par dilatation. L'on pourrait imposer 
-$\|\theta_c\|=1$ ou normaliser la fonction de valeur (ou de qualité) par 
+Cette borne est invariante par dilatation. Il est possible d'imposer 
+$\|\theta_c\|=1$ ou normaliser la fonction de valeur (et de qualité) par 
 $\|\R_{\theta_c}\|_\infty^{-1}$.
 
-On remarquera une dépendance caché de l'erreur de classification  $\epsilon_c$ à l'estimation de l'attribut moyen de l'expert  $\hat{\mu}^{\pi_E}$.
-En effet, l'erreur de classification minimale dépend de l'espace d'hypothèses généré par les fonctions de base de la fonction de score de l'algorithme de MC$^2$ (ici,
+Il existe une dépendance cachée de l'erreur de classification  $\epsilon_c$ à l'estimation de l'attribut moyen de l'expert  $\hat{\mu}^{\pi_E}$.
+En effet, l'erreur de classification minimale dépend de l'espace d'hypothèses généré par les fonctions de base de la fonction de score de l'algorithme de CMC (ici,
 $\hat{\mu}^{\pi_E}$).
-Néanmoins, avec une bonne représentation de la fonction de récompense (c'est-à-dire un choix judicieux de fonctions de base $\phi_i$) et une faible erreur d'estimation, cela ne devrait poser aucun problème en pratique.
+Néanmoins, avec une bonne représentation de la fonction de récompense (c'est-à-dire un choix judicieux de fonctions de base $\phi_i$) et une faible erreur d'estimation, cela ne devrait pas poser de problème en pratique.
 
@@
 $(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ que lors de la phase d'entraînement, où les $s_i$ sont les états présents dans $\D$. Il renvoie 
-$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire. Donc en pratique il est satisfaisant d'estimer 
-$\hat{\mu}^{\pi_E}$ correctement pour les couples état-action $(s_i,a)_{1\leq i\leq
+$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire après l'étape de CMC. Donc en pratique, il est satisfaisant d'estimer 
+$\hat{\mu}^{\pi_E}$ correctement uniquement pour les couples état-action $(s_i,a)_{1\leq i\leq
 N,a\in\A}$, ce qui permet d'envisager une simple estimation de Monte-Carlo par exemple.
@@
 
-SCIRL a besoin d'une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert. C'est un problème similaire à une évaluation de politique. Répétons l'observation-clef : chaque composante de 
-$\mu^{\pi_E}$ est une fonction de qualité pour $\pi_E$ vis-à-vis d'une fonction de récompense $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
-[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. Nous présentons une revue rapide de méthodes de calcul exacte et approchées ainsi qu'une heuristique.
+SCIRL a besoin d'une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert. C'est un problème similaire à l'évaluation d'une politique. Répétons l'observation-clef : chaque composante de 
+$\mu^{\pi_E}$ est la fonction de qualité pour $\pi_E$ vis-à-vis de la fonction de récompense $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
+[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. Nous présentons une revue rapide de méthodes de calcul exacte, et approchées ainsi qu'une heuristique.
 
 Si le modèle est connu, il est possible de calculer l'attribut moyen de manière exacte.
-Soit $\Phi\in\mathbb{R}^{|\s|\times p}$ la matrice d'attributs donc les lignes contiennent les vecteurs d'attributs $\phi(s)^\top$ pour tout
+Soit $\Phi\in\mathbb{R}^{|\s|\times p}$ la matrice d'attributs donc les lignes contiennent les vecteurs d'attributs $\phi(s)^\top$, pour tout
 $s\in\s$.
@@
 $\psi_i \in\mathbb{R}^{\s\times \A}$. Chaque composante $\mu_i^{\pi_E}$
- de l'attribut moyen de l'expert est paramétrisée par un vecteur
+ de l'attribut moyen de l'expert est paramétée par un vecteur
 $\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
-\psi(s,a)$. Supposons que l'on dispose d'un set d'entraînement
+\psi(s,a)$. Supposons que l'existence d'un ensemble d'entraînement
 $\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ dont les actions 
 $a_i$ ne sont pas nécessairement échantillonnées selon la politique $\pi_E$
-(\textit{e.g.}, on peut par exemple utiliser des trajectoires obtenues par un agent suivant une politique $\epsilon$-greedy), le but est d'obtenir une meilleure représentativité des données (des actions sous optimales devraient être essayées).
+(\textit{e.g.}, on peut utiliser des trajectoires obtenues par un agent suivant une politique $\epsilon$-gloutonne), le but étant d'obtenir une meilleure représentativité des données (des actions sous optimales devraient être essayées).
 Soit $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
 (resp. $\tilde{\Psi}'$) la matrice d'attributs dont les lignes sont les vecteurs d'attributs $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
-Soit $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ la matrive d'attributs dont les lignes sont les vecteurs d'attributs de la récompense $\phi(s_i)^\top$.
+Soit $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ la matrice d'attributs dont les lignes sont les vecteurs d'attributs de la récompense $\phi(s_i)^\top$.
 Enfin, soit $\Xi =
@@
 %
-Si l'on dispose d'un simulateur permettant d'échantillonner selon la politique de l'expert, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
+Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique de l'expert, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
 Si $K$
@@
 
-Pour minimiser l'erreur $\bar{\epsilon}_Q$, il est possible d'utiliser des transitions dont l'état de départ est tiré selon 
-$\rho_E$ et dont les actions sont uniformément distribuées. Cependant parfois seules les transitions issues de l'expert sont disponibles : $\T =
+Pour minimiser l'erreur $\bar{\epsilon}_Q$, il est nécessaire d'utiliser des transitions dont l'état de départ est tiré selon 
+$\rho_E$ et dont les actions sont uniformément distribuées. Cependant, il est possible que seules les transitions issues de l'expert soient disponibles : $\T =
 \{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$.
 Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen.
-Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$.
-Il est heureusement possible d'utiliser une heuristique; ce cas ne rentre pas dans l'analyse présentée en
+Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$, ces actions (et la dynamique résultantes) n'étant représentées dans les données.
+Il est heureusement possible de recourir à des heuristiques ; ce cas ne rentre pas dans l'analyse présentée en
 Sec.~\ref{subsec:scirl:analysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en Sec.~\ref{sec:experiments}.
 
-Nous proposons une telle heuristique. Supposons que  $\T$ soient les seules données disponibles, que nous utilisons pour fournir une estimation (précise)
+Nous proposons une telle heuristique. Supposons que  $\T$ soient les seules données disponibles, que nous utilisons pour fournir une estimation 
 $\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (cela revient à estimer non plus une fonction de qualité comme décrit ci-dessus, mais simplement une fonction de valeur).
 Un point de vue optimiste suppose que choisir une action différente de celle de l'expert ne fait que retarder l'effet de l'action de l'expert.
-Plus formellement, nous associons à chaque état $s$ un état virtuel $s_\text{v}$ pour lequel $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ pour toute action $a$ et pour lequel l'attribut (de récompense) moyen est le vecteur nul, $\phi(s_v) = 0$.
+Plus formellement, nous associons à chaque état $s$ un état virtuel $s_\text{v}$ pour lequel $p(s_v|s,a\neq \pi_E(s))=1,p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ pour toute action $a$ et pour lequel l'attribut (de récompense) moyen est le vecteur nul, $\phi(s_v) = 0$.
 Dans ce cas, on a
@@
 $s'_i = s_{i+1}$ (donc $\T =
-\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). On peut estimer
-$\mu^{\pi_E}(s_i,a_i)$ en utilisant la seule trajectoire disponible et en utilisant l'heuristique précedente pour les autres actions :
+\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$).
+$\mu^{\pi_E}(s_i,a_i)$ est estimé en utilisant la seule trajectoire disponible et en utilisant l'heuristique précedente pour les autres actions :
 \begin{equation}
@@
 En conséquence, tout algorithme d'évaluation de la fonction de qualité peut être utilisé pour estimer $\mu^\pi(s,a)$.
-Selon la quantité et la nature des données disponibles, l'on peut avoir recours à une heuristique pour évaluer l'attribut moyen pour une action différente de celle de l'expert.
-Cette estimation n'est nécessaire que pour entraîner les classifieur, il est donc suffisant de disposer de valeurs uniquement pour les couples état-action $(s_i,a)_{1\leq i \leq N,a\in \A}$.
+Selon la quantité et la nature des données disponibles, une heuristique peut-être employée pour évaluer l'attribut moyen pour une action différente de celle de l'expert.
+Cette estimation n'est nécessaire que pour entraîner le classifieur, il est donc suffisant de disposer de valeurs uniquement pour les couples état-action $(s_i,a)_{1\leq i \leq N,a\in \A}$.
 En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir Sec.~\ref{sec:relatedWorks}).
@@
 
-Comme précisé précedemment, tout algorithme de MC$^2$ peut être utilisé. Ici nous choisissons l'approche à marges structurées de ~\cite{Taskar:2005}. Soit
+Comme précisé précedemment, tout algorithme de CMC peut être utilisé. Ici nous choisissons l'approche à marges structurées de ~\cite{Taskar:2005}. Soit
 $\lc:\s\times\A\rightarrow\mathbb{R}_+$ une fonction de marge définie par l'utilisateur satisfaisant $\lc(s,\pi_E(s))\leq \lc(s,a)$ (ici,
-$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). L'algorithme MC$^2$ résoud :% the following quadratic problem:
+$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). L'algorithme CMC résout :% the following quadratic problem:
 \begin{equation}
   \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
-  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~~~~s.t.~~~~} \forall i,
+  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~~~~t.q.~~~~} \forall i,
   \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
@@
 \end{equation}
-Comme dans~\cite{Ratliff:2006}, nous fournissons la forme \emph{hinge-loss} équivalente (avec les variables d'ajustement $\zeta_i$ serrées, ce qui permet de déplacer les contraintes dans la fonction objectif) :
+De façon similaire à ~\cite{Ratliff:2006}, nous fournissons la forme \emph{hinge-loss} équivalente (avec les variables d'ajustement $\zeta_i$ serrées, ce qui permet de déplacer les contraintes dans la fonction objectif) :
 \begin{equation}
@@
 %our approach is more general: other choices could have been done for
-%the MC$^2$ algorithm as well as for the expert feature expectation
+%the CMC algorithm as well as for the expert feature expectation
 %estimation. Notice that if the model is known, $\mu^{\pi_E}$ can be
@@
 puis formalisée dans~\cite{Ng:2000}.
-Une approche classique, initiée par~\cite{Abbeel:2004}, consiste à trouver une politique (à travers une recherche dans l'espace des fonctions de récompense) telle que son attribut moyen (ou de manière plus générale une mesure de la distribution sous jacente aux trajectoires) s'approche de celui de la politique de l'expert.
+Une approche classique, initiée dans~\cite{Abbeel:2004}, consiste à trouver une politique (à travers une recherche dans l'espace des fonctions de récompense) telle que son attribut moyen (ou de manière plus générale une mesure de la distribution sous jacente aux trajectoires) s'approche de celui de la politique de l'expert.
 %
-Voir~\cite{Neu:2010} pour un survey.
+Voir~\cite{Neu:2010} pour un état de l'art partiel.
 %
@@
 Certains algorithmes mentionnés ne sont pas capable de renvoyer une fonction de récompense, bien qu'ils utilisent l'ARI comme étape.
-On parle de généralement de ces algorithmes comme d'algorithmes d'appretissage par imitation.
+Il rentrent dans le champ généralement appelé l'apprentissage par imitation.
 %
@@
 
-Plus proche de notre contribution, quelques approches introduisent également en quelque sorte de la structure dans la classification~\cite{Melo:2010}\cite{Ratliff:2006}. Dans~\cite{Melo:2010},
+Plus proche de notre contribution, quelques approches introduisent également de la structure dans la classification~\cite{Melo:2010,Ratliff:2006}. Dans~\cite{Melo:2010},
 une métrique induite par le PDM est utilisée pour construire un noyau qui sera utilisé par l'algorithme de classification, permettant des améliorations par rapport à un noyau non structuré.
-Cette approche n'est cependant pas un algorithme d'ARI, et plus important l'évaluation d'une métrique dans un PDM n'est pas une mince affaire.
+Cette approche n'est cependant pas un algorithme d'ARI, et plus important l'évaluation d'une métrique dans un PDM n'est pas triviale.
 Dans~\cite{Ratliff:2006}, un algorithme de classification est utilisé pour fournir une fonction de récompense.
-Au lieu d'associer des actions à des états comme nous le faisons, cette algorithme associe des politiques optimales (qui jouent le rôle de labels) ) des PDM (entrées), ce qui permet d'incorporer la structure, au prix de la résolution d'un grand nombre de PDM.
+Au lieu d'associer des actions à des états comme nous le faisons, cet algorithme associe des politiques optimales (qui jouent le rôle de labels) à des PDM (entrées), ce qui permet d'incorporer la structure, au prix de la résolution d'un grand nombre de PDM.
 
-Tous les algorithmes d'ARI à notre connaissance requierent la résolution du problème direct de l'AR de manière répétée, à l'exception de~\cite{Dvij:2010,boularias:2011}.
-\cite{Dvij:2010} ne s'applique qu'aux PDM solvables linéairement (où le contrôle se fait en imposant une dynamique au système).
-Dans~\cite{boularias:2011}, en utilisant l'argument de l'entropie relative, on maximise une fonction objectif via une montée de sous gradient.
-Estimer la valeur du sous gradient demande des trajectoires échantillonnées selon la politique optimale pour la fonction de récompense courante.
-On peut contourner le problème grâce à l'\emph{importance sampling}.
-Cela requière cependant d'échantillonner des trajectoires selon une politique différente de celle de l'expert, et la du problème direct se maintient au coeur de l'approche (même si on évite sa résolution).
+Tous les algorithmes d'ARI à notre connaissance requièrent la résolution du problème direct de l'AR de manière répétée, à l'exception de~\cite{Dvij:2010,boularias:2011}.
+Le travail présenté dans \cite{Dvij:2010} ne s'applique qu'aux PDM solubles linéairement (où le contrôle se fait en imposant une dynamique au système).
+Dans~\cite{boularias:2011}, en utilisant l'argument de l'entropie relative, une fonction objectif est maximisée via une montée de sous-gradient.
+Estimer la valeur du sous-gradient demande des trajectoires échantillonnées selon la politique optimale pour la fonction de récompense courante.
+On peut contourner le problème grâce à l'échantillonage préférentiel.
+Cela requiert cependant d'échantillonner des trajectoires selon une politique différente de celle de l'expert, et le problème direct se maintient au coeur de l'approche (même si on évite sa résolution).
 
 SCIRL n'a pas besoin de résoudre le problème direct, mais uniquement d'estimer l'attribut moyen de la politique de l'expert.
-Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une fois un problème d'évaluation de la politique.
-Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI \textit{e.g.}~\cite{boularias:2011}).
-De plus par l'utilisation d'une heuristique qui dépassent le cadre de notre analyse, il est possible à SCIRL de se contenter de données fournies par l'expert.
+Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une seule fois un problème d'évaluation de politique.
+Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI, par exemple~\cite{boularias:2011}).
+De plus, par l'utilisation d'une heuristique qui dépassent le cadre de notre analyse, il est possible à SCIRL de se contenter de données fournies par l'expert.
 La prochaine section présente une démonstration empirique de cela.
@@
 La voiture peut se déplacer vers la gauche ou la droite, accélérer ou ralentir et conserver sa vitesse.
-L'expert optimise une récompense définie par nos soins $\R_E$ qui récompense la vitesse, punit les sorties de route, punit sévèrement les collisions et no donne pas d'information dans les autres cas.
+L'expert optimise une récompense définie par nos soins, $\R_E$, qui récompense la vitesse, punit les sorties de route, punit sévèrement les collisions et ne donne pas d'information dans les autres cas.
 
-Nous avons comparé SCIRL tel qu'instancié comme décrit en
-Sec.~\ref{subsec:practicalApproach:instantiation} à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au coeur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbeel:2004} (appelé ici PIRL pour \emph{
+Nous avons comparé SCIRL, tel qu'instancié comme décrit en
+Sec.~\ref{subsec:practicalApproach:instantiation}, à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au coeur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbeel:2004} (appelé ici PIRL pour \emph{
 Projection Inverse Reinforcement Learning}).
-Nous nous préoccupons également du comportement optimal vis-à-vis d'une récompense tirée aléatoirement (en utilisant la même paramétrisation que SCIRL et PIRL, le vecteur de paramètres est tiré selon une loi uniforme) afin de disposer d'un point de comparaison.
+Nous nous préoccupons également du comportement optimal vis-à-vis d'une récompense tirée aléatoirement (en utilisant la même paramétrisation que SCIRL et PIRL, le vecteur de paramètres est tiré selon une loi uniforme) afin de disposer d'un autre point de comparaison.
 
@@
 Ces attributs sont bien moins informatifs que ceux utilisés dans~\cite{Abbeel:2004,Syed:2008:game}.
-Les attributs de~\cite{Syed:2008:game} sont si informatifs que tirer aléatoirement un vecteur de paramètres $\theta$ positifs pour la récompense donne lieu à un comportement acceptable.
+Les attributs de~\cite{Syed:2008:game} sont si informatifs que tirer aléatoirement un vecteur $\theta$ de paramètres positifs pour la récompense donne lieu à un comportement acceptable.
 Le facteur d'oubli est $\gamma = 0.9$.
@@
 Nous mesurons les performances de chacune des approches grâce à $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, avec $\mathcal{U}$
-la distribution uniforme (ce qui permet de tester la capacité de généralisation de chaque approche même pour des états infrequemment recontrés), $\R_E$ est la récompense de l'expert et $\pi$ est l'une des politiques suivantes: la politique optimale pour $\R_E$ (point de référence haut),
-la politique optimale pour une récompense aléatoire (point de référence bas), la politique optimale pour $\R_{\theta_c}$ (SCIRL), et la politique produite par PIRL et la règle de décision du classifieur.
+la distribution uniforme (ce qui permet de tester la capacité de généralisation de chaque approche même pour des états infréquemment recontrés), $\R_E$ étant la récompense de l'expert et $\pi$ est l'une des politiques suivantes: la politique optimale pour $\R_E$ (point de référence haut),
+la politique optimale pour une récompense aléatoire (point de référence bas), la politique optimale pour $\R_{\theta_c}$ (SCIRL), la politique produite par PIRL et la règle de décision du classifieur.
 
@@
 Pour chaque courbe, nous dessinons la moyenne (trait plein), l'écart-type (foncé) et les valeurs min et max (plus clair).
-La pôlitique correspondant à une récompense aléatoire est en bleu, la politique retournée par le classifieur est en jaune et la politique optimale pour la récompense trouvée par SCIRL's est en rouge.
+La politique correspondant à une récompense aléatoire est en bleu, la politique retournée par le classifieur est en jaune et la politique optimale pour la récompense trouvée par SCIRL est en rouge.
 La ligne bleue foncée correspond à PIRL.}
-\begin{minipage}[l]{0.45\linewidth}
+\begin{minipage}[l]{0.49\linewidth}
 \centering
-  \centerline{\includegraphics[width=.92\linewidth]{fig_classif.pdf}}
+  \centerline{\includegraphics[width=1.2\linewidth]{fig_classif.pdf}}
 \end{minipage} \hfill
-\begin{minipage}[r]{0.45\linewidth}
+\begin{minipage}[r]{0.49\linewidth}
 \centering
-  \centerline{\includegraphics[width=.92\linewidth]{fig_scirl.pdf}}
+  \centerline{\includegraphics[width=1.2\linewidth]{fig_scirl.pdf}}
 \end{minipage}
@@
 On peut voir que le classifieur ne fonctionne pas bien sur ce problème.
-Augmenter le nombre de transitions améliorerait ses performances, mais après 400 transitions, il ne fonctionne pas aussi bien que SCIRL avec uniquement une dizaine de transition.
-SCIRL fonctionne particulièrement bien ici: après seulement une centaine de transitions il atteint les performances de PIRL.
+Augmenter le nombre de transitions améliorerait ses performances, mais après 400 transitions, il ne fonctionne pas aussi bien que SCIRL avec uniquement une dizaine de transitions.
+SCIRL fonctionne particulièrement bien ici : après seulement une centaine de transitions il atteint les performances de PIRL.
 Ces deux algorithmes sont proches de la valeur de l'expert.
@@
 Nous avons introduit un nouveau moyen de résoudre le problème de l'ARI en structurant un classifieur à fonction de score linéairement paramétrée avec une estimée de l'attribut moyen de l'expert.
-Il renvoie une fonction de récompense pour laquelle nous avons montré que l'expert est quasi optimal, pour peu que l'erreur de classification soit faible et que l'estimation de l'attribut moyen de l'expert soit bonne.
+Il renvoie une fonction de récompense pour laquelle nous avons montré que l'expert est quasi-optimal, pour peu que l'erreur de classification soit faible et que l'estimation de l'attribut moyen de l'expert soit bonne.
 Des méthodes pratiques pour l'estimation de cette grandeur ont été présentées et nous avons introduit une heuristique pour le cas où les seules données disponibles proviennent de l'expert, ainsi qu'une instanciation spécifique de l'algorithme SCIRL.
-Nous avons exhibé grâce à un simulateur de conduite sur autoroute que l'approche proposée fonctionne bien (même combinée à l'heuristique discutée), bien mieux qu'un classifieur non structuré et aussi bien que la méthode de l'état de l'art qui avait accès au modèle (et avec un coût computationel moindre).
+Nous avons exhibé, grâce à un simulateur de conduite sur autoroute, que l'approche proposée fonctionne bien (même combinée à l'heuristique discutée), bien mieux qu'un classifieur non structuré et aussi bien que la méthode de l'état de l'art qui avait accès au modèle (et avec un coût computationel moindre).
 Nous prévoyons de poursuivre notre analyse des propritétés théoriques de SCIRL (notamment en vue de l'utilisation avec des heuristiques) et l'appliquer à des problèmes robotiques réels.
 
-\acknowledgements{Ces travaux ont été en partie financés par le projet EU FP7 ILHAIRE (grant
-n$\deg$270780), par le projet EU INTERREG IVa  ALLEGRO et par la Région
+\acknowledgements{Ces travaux ont été en partie financés par le projet EU FP7 ILHAIRE (Bourse
+n\degree 270780), par le projet EU INTERREG IVa  ALLEGRO et par la Région
 Lorraine (France).}

### 2012-09-28 klein2012structured_RIA/scirl_ria.tex
Corrections suggérées par MAtthieu après seconde lecture.

@@
 \author[1,2]{Edouard}{Klein}
-\author[2,3]{Matthieu}{Geist}
-\author[2]{Bilal}{Piot}
+\author[2,3]{Bilal}{Piot}
+\author[2]{Matthieu}{Geist}
 \author[2,3]{Olivier}{Pietquin}
@@
   driving simulator.}
-\resume{Cette contribution traite le problème de l'Apprentissage par Renforcement Inverse (ARI), défini comme la recherche d'une fonction de récompense pour laquelle le comportement d'un expert (connu par le biais d'une démonstration) est optimal.
+\resume{Cette contribution traite le problème de l'Apprentissage par Renforcement Inverse (ARI), défini comme la recherche d'une fonction de récompense pour laquelle le comportement d'un expert (connu par le biais de démonstrations) est optimal.
 Nous introduisons SCIRL, un nouvel algorithme qui utilise la grandeur dénommée attribut moyen de l'expert comme la paramétrisation d'une fonction de score pour un classifieur multi-classe.
@@
 Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct de l'Apprentissage par Renforcement.
-De plus, en utilisant une heuristique il est utilisable uniquement avec des trajectoires échantillonnées par l'expert.
+De plus, en utilisant une heuristique, il peut être utilisé avec uniquement des trajectoires échantillonnées par l'expert.
 Nous illustrons cela sur un simulateur de conduite.}
@@
 
-L'Apprentissage par Renforcement Inverse (ARI)~\cite{Russell:1998} est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert est optimal.
-Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{sec:relatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) qui suit celle générée par l'expert.
+L'Apprentissage par Renforcement Inverse (ARI)~\cite{Russell:1998} est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal.
+Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{sec:relatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) proche de celle générée par l'expert.
 Souvent, cette distribution est caractérisée par ce que l'on appelle l'attribut moyen (voir Sec.~\ref{subsec:background:irl}) : étant donnée une fonction de récompense linéairement paramétrée par un vecteur d'attributs, il s'agit de l'espérance de la somme pondérée des vecteurs d'attributs sachant que l'on commence dans un certain état en choisissant une certaine action avant de suivre la politique concernée.
@@
 Il est possible d'imiter le comportement de l'expert via un algorithme d'apprentissage supervisé généralisant l'association des actions aux états.
-Nous considérons ici les classifieurs multi-classes qui à partir d'un set d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour une état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subsec:background:classif}).
+Nous considérons ici les classifieurs multi-classes qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour une état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subsec:background:classif}).
 L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir Sec.~\ref{subsec:scirl:algo}).
@@
 Un grand avantage de SCIRL est qu'il ne requiert pas, contrairement à la plupart des algorithmes d'ARI, la résolution répétée du problème direct, celui de l'apprentissage par renforcement.
-Il nécessit d'estimer l'attribut moyen de l'expert, mais ceci est équivalent un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir Sec.~\ref{sec:practicalApproach}.
-De plus, à l'aide d'une heuristique, SCIRL peut s'éxécuter en utilisant uniquement en utilisant des transitions issues de la politique de l'expert (il n'y a pas besoin d'échantilloner la dynamique complète).
+Il nécessite d'estimer l'attribut moyen de l'expert, mais ceci est équivalent à un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir Sec.~\ref{sec:practicalApproach}.
+De plus, à l'aide d'une heuristique, SCIRL peut s'éxécuter en utilisant uniquement des transitions issues de la politique de l'expert (il n'y a pas besoin d'échantilloner la dynamique complète).
 Nous illustrons cela sur un simulateur de conduite Sec.~\ref{sec:experiments}.
@@
 
-Un Processus Décisionnel de Markov (PDM)~\cite{Puterman:1994} est un tuple $\{\s,\A,\p,\R,\gamma\}$ avec $\s$ l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ l'espace d'action fini, $\p =
-\{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ la fonction de récompense sur les états et $\gamma$ le facteur d'oubli.
+Un Processus Décisionnel de Markov (PDM)~\cite{Puterman:1994} est un tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ est l'espace d'action fini, $\p =
+\{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ est l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ est la fonction de récompense sur les états et $\gamma$ est le facteur d'oubli.
 Une politique déterministe $\pi\in\s^\A$ définit le comportement d'un agent.
-La qualité moyenne de ce comportement est quantifiée par le biais de la fonction de valeur $v_\R^\pi\in\mathbb{R}^\s$, qui à chaque état associe la récompense pondérée cumulée recueillie par l'agent lorsqu'il part de cet état et suit la politique $\pi$ par la suite : $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t \R(S_t)|S_0=s,\pi]$.
+La qualité de ce comportement est quantifiée par le biais de la fonction de valeur $v_\R^\pi\in\mathbb{R}^\s$, qui à chaque état associe la récompense pondérée cumulée moyenne recueillie par l'agent lorsqu'il part de cet état et suit la politique $\pi$ par la suite : $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t \R(S_t)|S_0=s,\pi]$.
 Une politique optimale $\pi_\R^*$ (vis-à-vis de la fonction de récompense $\R$) est une politique dont la fonction de valeur $v^*_\R$ vérifie $v_\R^* \geq v_\R^\pi$ (composante par composante) pour toute politique $\pi$.
@@
 %\begin{equation}
-%  T^\pi_\R v = \R + \gamma P_\pi v \text{ and } T_\R^*v = \max_\pi T_\R^\pi v.
+%  T^\pi_\R v = \R + \gamma P_\pi v \text{ et } T_\R^*v = \max_\pi T_\R^\pi v.
 %\end{equation}
 Ces opérateurs sont des contractions qui admettent $v_\R^\pi$ et $v^*_\R$ comme points fixes respectifs : $v_\R^\pi = T^\pi_\R v_\R^\pi$ et $v^*_\R = T^*_\R v^*_\R$.
-La fonction de qualité $Q^\pi\in\mathbb{R}^{\s\times \A}$ ajoute un degré de liberté dans le choix de la première action, on la définit formellement par $Q_\R^\pi(s,a)
-= [T^a_\R v^\pi_\R](s)$. La distribution stationnaire de la politique $\pi$ est notée $\rho_\pi$ elle satisfait $\rho_\pi^\top P_\pi = \rho_\pi^\top$.
+La fonction de qualité $Q^\pi\in\mathbb{R}^{\s\times \A}$ ajoute un degré de liberté dans le choix de la première action : $Q_\R^\pi(s,a)
+= [T^a_\R v^\pi_\R](s)$. La distribution stationnaire de la politique $\pi$ est notée $\rho_\pi$, elle satisfait $\rho_\pi^\top P_\pi = \rho_\pi^\top$.
 
-L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
+L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observée par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
 %
@@
 %
-On appelle cela le problème inverse.
+Nous appelons cela le problème inverse.
 
@@
 \X$ sont tirées selon une distribution inconnue $\prob(x)$ et qu'il existe un oracle qui à chacune de ces entrées associe un label $y\in Y$ tiré selon une distribution de probabilité conditionnelle inconnue $\prob(y|x)$.
-De manière générale, la classification multi-classe cherche, étant donné un ensemble d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tiré selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
+De manière générale, la classification multi-classe cherche, étant donné une base d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tirée selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
 
@@
 La règle de décision associée $g_w\in{\Y^\X}$ est définie par $g_w(x) \in \argmax_{y\in\Y}s_w(x,y)$.
-En se basant sur un ensemble d'entraînement $\{(x_i,y_i)_{1\leq
+En se basant sur une base d'entraînement $\{(x_i,y_i)_{1\leq
 i\leq N}\}$, un classifieur multi-classe (noté CMC) à fonction de score paramétrée linéairement choisit un vecteur de paramètres $\theta_c$.
@@
 $g_w(x)$ est interprétable comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle-même être vue comme une fonction de qualité.
-Le parallèle avec l'Eq.~\eqref{eq:def:mu} est facilement tiré, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de l'ensemble d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
+Le parallèle avec l'Eq.~\eqref{eq:def:mu} est aisé, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de la base d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
 Ces remarques nous permettent maintenant d'introduire notre algorithme d'ARI par classification structurée (SCIRL pour \emph{Structured Classification-based Inverse Reinforcement Learning}).
@@
 Soit $\pi_E$ la politique de l'expert à partir de laquelle nous souhaitons inférer une fonction de récompense.
-Supposons l'existence d'un ensemble d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est à mélange rapide (\emph{fast-mixing}), l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés d'après cette distribution.} $\rho_E = \rho_{\pi_E}$.
+Supposons l'existence d'une base d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est à mélange rapide (\emph{fast-mixing}), l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés selon cette distribution.} $\rho_E = \rho_{\pi_E}$.
 Supposons également avoir à disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eq:def:mu}.
@@
   \BlankLine
-  \emph{\textbf{Etant donné}} un ensemble d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
+  \emph{\textbf{Etant donnée}} une base d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
   une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ et un algorithme de CMC\;
@@
   \BlankLine
-  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de CMC auquel sont fournis l'ensemble d'entraînement $\D$ et la paramétrisation de la fonction de score : $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
+  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de CMC auquel sont fournis la base d'entraînement $\D$ et la paramétrisation de la fonction de score : $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
   %
@@
 L'approche proposée est résumé par l'Alg.~\ref{algo:scirl}.
-Le nom de l'algorithme (SCRIL, anglais pour Classification structurée pour l'ARI) vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
+Le nom de l'algorithme (SCRIL, que l'on peut traduire par Classification structurée pour l'ARI), vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
 Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct.
-Cet algorithme requiert une estimation de l'attribut moyen de l'expert mais il ne s'agit là que d'un problème d'évaluation de la politique, notoirement moins difficile que la recherche d'une politique optimale qu'implique la résolution du problème direct.
-Cela est discuté plus en détail en Sec.~\ref{sec:relatedWorks}.
+Cet algorithme requiert une estimation de l'attribut moyen de l'expert mais il ne s'agit là que d'un problème d'évaluation de la politique, notoirement moins difficile que la recherche de politiques optimales qu'implique la résolution du problème direct.
+Cela est discuté plus en détail Sec.~\ref{sec:relatedWorks}.
 %
@@
 \E_{s\sim\rho_E}[\chi_{\{\pi_c(s)\neq\pi_E(s)\}}] \in [0,1]$.
-On écrit $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$ la fonction de score calculée à partir de l'ensemble d'entraînement $\D$ (celle-ci peut être interprétée comme une fonction de qualité approchée).
+On écrit $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$ la fonction de score calculée à partir de la base d'entraînement $\D$ (celle-ci peut être interprétée comme une fonction de qualité approchée).
 Soit
@@
   L'inégalité $(a)$ est valable car $T^{\pi^*} v^{\pi_E}\leq T^*
-  v^{\pi_E}$ et l'inégalité $(b)$ l'est en vertu de à~\cite[Lemme~4.2]{Munos:2007}.
+  v^{\pi_E}$ et l'inégalité $(b)$ l'est en vertu de~\cite[Lemme~4.2]{Munos:2007}.
 De plus, $v^*$ étant optimale, il est visible que  $v^*-v^{\pi_E}\geq 0$ et avec $T^*$ l'opérateur d'optimalité de Bellman, $T^* v^{\pi_E}\geq
@@
     T^* v^{\pi_E} - v^{\pi_E} = T^* v^{\pi_E} - T^{\pi_c}v^{\pi_E}
-    + T^{\pi_c}v^{\pi_E}- v^{\pi_E},
+    + T^{\pi_c}v^{\pi_E}- v^{\pi_E}.
     \label{eq:proof:decomposition}
@@
   En passant à l'espérance selon $\rho_E$ et tout en rappelant que 
-  $T^* v^{\pi_E}\geq v^{\pi_E}$, le premier terme est borné
+  $T^* v^{\pi_E}\geq v^{\pi_E}$, le premier terme est borné :
   \begin{equation}
@@
 
-   $M\in\mathbb{R}^{|\s|\times |\s|}$ est la matrice diagonale définie par $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$.
+   Soit $M\in\mathbb{R}^{|\s|\times |\s|}$ la matrice diagonale définie par $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$.
 En utilisant cela, l'opérateur de Bellman $T^{\pi_c}$ peut s'écrire, pour tout $v\in\mathbb{R}^\s$:
@@
   En appliquant cet opérateur à  $v^{\pi_E}$ et en utilisant le fait que $\R +
-  \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, on obtient :
+  \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, nous obtenons :
   \begin{equation}
@@
   \end{equation}
-  On peut facilement voir que $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
+  Il est dacile de voir que $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
   \leq \frac{2}{1-\gamma}\|\R\|_\infty$, ce qui permet de borner le dernier terme
@@
 Cette borne est invariante par dilatation. Il est possible d'imposer 
-$\|\theta_c\|=1$ ou normaliser la fonction de valeur (et de qualité) par 
+$\|\theta_c\|=1$ ou de normaliser la fonction de valeur (et de qualité) par 
 $\|\R_{\theta_c}\|_\infty^{-1}$.
@@
 Finalement, si notre borne se base sur les erreurs en généralisation 
-$\epsilon_c$ et $\bar{\epsilon}_Q$, le classifieur n'utilisera 
+$\epsilon_c$ et $\bar{\epsilon}_Q$, le classifieur n'utilisera que 
 $(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ que lors de la phase d'entraînement, où les $s_i$ sont les états présents dans $\D$. Il renvoie 
-$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire après l'étape de CMC. Donc en pratique, il est satisfaisant d'estimer 
+$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire après l'étape de CMC. Ainsi, en pratique, il est satisfaisant d'estimer 
 $\hat{\mu}^{\pi_E}$ correctement uniquement pour les couples état-action $(s_i,a)_{1\leq i\leq
@@
 $\mu^{\pi_E}$ est la fonction de qualité pour $\pi_E$ vis-à-vis de la fonction de récompense $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
-[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. Nous présentons une revue rapide de méthodes de calcul exacte, et approchées ainsi qu'une heuristique.
+[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. Nous présentons une revue rapide de méthodes de calcul exacte et approchées, ainsi qu'une heuristique.
 
@@
 $\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
-\psi(s,a)$. Supposons que l'existence d'un ensemble d'entraînement
+\psi(s,a)$. Supposons l'existence d'une base d'entraînement
 $\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ dont les actions 
@@
 %
-Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique de l'expert, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
+Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique à imiter, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
 Si $K$
@@
 \{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$.
-Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen.
+Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ seules ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen.
 Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$, ces actions (et la dynamique résultantes) n'étant représentées dans les données.
@@
 
-Nous proposons une telle heuristique. Supposons que  $\T$ soient les seules données disponibles, que nous utilisons pour fournir une estimation 
+Nous proposons une telle heuristique. Supposons que  $\T$ contienne les seules données disponibles, que nous utilisons pour fournir une estimation 
 $\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (cela revient à estimer non plus une fonction de qualité comme décrit ci-dessus, mais simplement une fonction de valeur).
 Un point de vue optimiste suppose que choisir une action différente de celle de l'expert ne fait que retarder l'effet de l'action de l'expert.
-Plus formellement, nous associons à chaque état $s$ un état virtuel $s_\text{v}$ pour lequel $p(s_v|s,a\neq \pi_E(s))=1,p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ pour toute action $a$ et pour lequel l'attribut (de récompense) moyen est le vecteur nul, $\phi(s_v) = 0$.
+Plus formellement, nous associons à chaque état $s$ un état virtuel $s_\text{v}$ pour lequel $p(s_v|s,a\neq \pi_E(s))=1 \textrm{ et } p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ pour toute action $a$ et pour lequel l'attribut (de récompense) moyen est le vecteur nul, $\phi(s_v) = 0$.
 Dans ce cas, on a
@@
 $s'_i = s_{i+1}$ (donc $\T =
-\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$).
+\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). L'attribut moyen 
 $\mu^{\pi_E}(s_i,a_i)$ est estimé en utilisant la seule trajectoire disponible et en utilisant l'heuristique précedente pour les autres actions :
@@
 Tous les algorithmes d'ARI à notre connaissance requièrent la résolution du problème direct de l'AR de manière répétée, à l'exception de~\cite{Dvij:2010,boularias:2011}.
-Le travail présenté dans \cite{Dvij:2010} ne s'applique qu'aux PDM solubles linéairement (où le contrôle se fait en imposant une dynamique au système).
+Le travail présenté dans \cite{Dvij:2010} ne s'applique qu'aux PDM solvables linéairement (où le contrôle se fait en imposant une dynamique au système).
 Dans~\cite{boularias:2011}, en utilisant l'argument de l'entropie relative, une fonction objectif est maximisée via une montée de sous-gradient.
@@
 On peut contourner le problème grâce à l'échantillonage préférentiel.
-Cela requiert cependant d'échantillonner des trajectoires selon une politique différente de celle de l'expert, et le problème direct se maintient au coeur de l'approche (même si on évite sa résolution).
+Cela requiert cependant d'échantillonner des trajectoires selon une politique différente de celle de l'expert, et le problème direct se maintient au coeur de l'approche (même si sa résolution est évitée).
 
@@
 Le facteur d'oubli est $\gamma = 0.9$.
-Le classifieur utilise le même vecteur d'attribut reproduit pour chaque action.
+Le classifieur utilise le même vecteur d'attributs reproduit pour chaque action.
 
@@
 PIRL est un algorithme itératif dont chaque itération implique la résolution du PDM pour une fonction de récompense arbitraire.
-Nous l'avons fait fonctionner pendant 70 itérations, tous les objets requis (l'attribut moyen de politiques différentes de celles de l'expert et la politique optimale pour une fonction de récompense à chaque itération) ont été calculées de manière éxacte en utilisant le modèle.
+Nous l'avons fait fonctionner pendant 70 itérations, tous les objets requis (l'attribut moyen de politiques différentes de celles de l'expert et la politique optimale pour une fonction de récompense à chaque itération) ont été calculées de manière exacte en utilisant le modèle.
 Nous mesurons les performances de chacune des approches grâce à $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, avec $\mathcal{U}$
@@
 
-Nous avons introduit un nouveau moyen de résoudre le problème de l'ARI en structurant un classifieur à fonction de score linéairement paramétrée avec une estimée de l'attribut moyen de l'expert.
+Nous avons introduit une nouvelle approche de résolution du problème de l'ARI en structurant un classifieur à fonction de score linéairement paramétrée avec une estimée de l'attribut moyen de l'expert.
 Il renvoie une fonction de récompense pour laquelle nous avons montré que l'expert est quasi-optimal, pour peu que l'erreur de classification soit faible et que l'estimation de l'attribut moyen de l'expert soit bonne.
@@
 
-\acknowledgements{Ces travaux ont été en partie financés par le projet EU FP7 ILHAIRE (Bourse
-n\degree 270780), par le projet EU INTERREG IVa  ALLEGRO et par la Région
-Lorraine (France).}
+%\acknowledgements{Ces travaux ont été en partie financés par le projet EU FP7 ILHAIRE (Bourse
+%n\degree 270780), par le projet EU INTERREG IVa  ALLEGRO et par la Région
+%Lorraine (France).}
 

### 2012-09-28 klein2012structured_RIA/scirl_ria.tex
Version avec french comme option

@@
 
-\documentclass[english,utf8]{./hermes-journal}
+\documentclass[french,utf8]{./hermes-journal}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{subfigure}
-\usepackage{ gensymb }
+\usepackage{gensymb}
 \mathtoolsset{showonlyrefs=true}
@@
 \section{Introduction}
-\label{sec:intro}
+\label{sentro}
 
-L'Apprentissage par Renforcement Inverse (ARI)~\cite{Russell:1998} est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal.
-Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{sec:relatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) proche de celle générée par l'expert.
-Souvent, cette distribution est caractérisée par ce que l'on appelle l'attribut moyen (voir Sec.~\ref{subsec:background:irl}) : étant donnée une fonction de récompense linéairement paramétrée par un vecteur d'attributs, il s'agit de l'espérance de la somme pondérée des vecteurs d'attributs sachant que l'on commence dans un certain état en choisissant une certaine action avant de suivre la politique concernée.
+L'Apprentissage par Renforcement Inverse (ARI)~\cite{Russel998} est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal.
+Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{seelatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) proche de celle générée par l'expert.
+Souvent, cette distribution est caractérisée par ce que l'on appelle l'attribut moyen (voir Sec.~\ref{subseackgrounrl}) : étant donnée une fonction de récompense linéairement paramétrée par un vecteur d'attributs, il s'agit de l'espérance de la somme pondérée des vecteurs d'attributs sachant que l'on commence dans un certain état en choisissant une certaine action avant de suivre la politique concernée.
 
@@
 Il est possible d'imiter le comportement de l'expert via un algorithme d'apprentissage supervisé généralisant l'association des actions aux états.
-Nous considérons ici les classifieurs multi-classes qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour une état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subsec:background:classif}).
-L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir Sec.~\ref{subsec:scirl:algo}).
-Le vecteur de paramètres ainsi calculé définit une fonction de récompense dont on montre qu'elle admet pour politique quasi-optimale la politique de l'expert (voir Sec.~\ref{subsec:scirl:analysis}).
+Nous considérons ici les classifieurs multi-classes qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour une état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subseackgrounlassif}).
+L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir Sec.~\ref{subsecirlgo}).
+Le vecteur de paramètres ainsi calculé définit une fonction de récompense dont on montre qu'elle admet pour politique quasi-optimale la politique de l'expert (voir Sec.~\ref{subsecirnalysis}).
 
 Un grand avantage de SCIRL est qu'il ne requiert pas, contrairement à la plupart des algorithmes d'ARI, la résolution répétée du problème direct, celui de l'apprentissage par renforcement.
-Il nécessite d'estimer l'attribut moyen de l'expert, mais ceci est équivalent à un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir Sec.~\ref{sec:practicalApproach}.
+Il nécessite d'estimer l'attribut moyen de l'expert, mais ceci est équivalent à un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir Sec.~\ref{seracticalApproach}.
 De plus, à l'aide d'une heuristique, SCIRL peut s'éxécuter en utilisant uniquement des transitions issues de la politique de l'expert (il n'y a pas besoin d'échantilloner la dynamique complète).
-Nous illustrons cela sur un simulateur de conduite Sec.~\ref{sec:experiments}.
+Nous illustrons cela sur un simulateur de conduite Sec.~\ref{sexperiments}.
 
 \section{Contexte et notations}
-\label{sec:background}
+\label{seackground}
 
 \subsection{Apprentissage par Renforcement (Inverse)}
-\label{subsec:background:irl}
+\label{subseackgrounrl}
 
-Un Processus Décisionnel de Markov (PDM)~\cite{Puterman:1994} est un tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ est l'espace d'action fini, $\p =
+Un Processus Décisionnel de Markov (PDM)~\cite{Puterma994} est donné par le tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ est l'espace d'action fini, $\p =
 \{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ est l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ est la fonction de récompense sur les états et $\gamma$ est le facteur d'oubli.
@@
 Un léger abus de notation nous permet de nommer $a$ la politique associant l'action $a$ à chaque état $s$.
-Les opérateurs d'évaluation (resp.
-d'optimalité) de Bellman $T^\pi_\R\text{ (resp.
+Les opérateurs d'évaluation (respectivement d'optimalité) de Bellman $T^\pi_\R\text{ (resp.
 $T^*_\R$)}:\mathbb{R}^\s
@@
 
-L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observée par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
+L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observée par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertseka996,Sutto998,szepesvari2010c}.
 %
@@
 %
-L'Apprentissage par Renforcement Inverse (approché)~\cite{Ng:2000} consiste à estimer la fonction de récompense pour laquelle une politique observée est optimale.
+L'Apprentissage par Renforcement Inverse (approché)~\cite{N000} consiste à estimer la fonction de récompense pour laquelle une politique observée est optimale.
 %
@@
   \phi(S_t)|S_0=s,A_0=a,\pi].
-  \label{eq:def:mu}
+  \label{eeu}
 \end{equation}
@@
 Chaque composante $\mu_i^\pi$ de ce vecteur d'attributs est en réalité la fonction de qualité de la politique $\pi$ pour la récompense $\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$.
-De fait, tout algorithme d'estimation de la fonction de qualité peut être utilisé pour estimer l'attribut moyen, comme une estimation de Monte-Carlo ou un algorithme aux différences temporelles~\cite{Klein:2011}.
+De fait, tout algorithme d'estimation de la fonction de qualité peut être utilisé pour estimer l'attribut moyen, comme une estimation de Monte-Carlo ou un algorithme aux différences temporelles~\cite{Klei011}.
 
-\subsection{Classifieurs à fonction de score paramétrée linéairement} \label{subsec:background:classif}
+\subsection{Classifieurs à fonction de score paramétrée linéairement} \label{subseackgrounlassif}
 
@@
 Notre étude est valable pour tout algorithme de classification tant qu'il opère en maximisant l'argument d'une fonction de score paramétrée linéairement.
-Par exemple, une machine à vecteur support multi-classe pourrait être choisie~\cite{Guermeur:2007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taskar:2005}.
+Par exemple, une machine à vecteur support multi-classe pourrait être choisie~\cite{Guermeu007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taska005}.
 D'autres choix sont possibles.
 
-\section{Classification structurée pour l'apprentissage par renforcement inverse} \label{sec:scirl}
+\section{Classification structurée pour l'apprentissage par renforcement inverse} \label{secirl}
 
 \subsection{Forme générale de l'algorithme}
-\label{subsec:scirl:algo}
+\label{subsecirlgo}
 
-Nous nous trouvons dans le cadre de la classification décrit en Sec.~\ref{subsec:background:classif}.
+Nous nous trouvons dans le cadre de la classification décrit en Sec.~\ref{subseackgrounlassif}.
 L'entrée $x$ peut être vue comme un état et le label $y$ comme une action.
@@
 $g_w(x)$ est interprétable comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle-même être vue comme une fonction de qualité.
-Le parallèle avec l'Eq.~\eqref{eq:def:mu} est aisé, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de la base d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
+Le parallèle avec l'Eq.~\eqref{eeu} est aisé, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de la base d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
 Ces remarques nous permettent maintenant d'introduire notre algorithme d'ARI par classification structurée (SCIRL pour \emph{Structured Classification-based Inverse Reinforcement Learning}).
@@
 Soit $\pi_E$ la politique de l'expert à partir de laquelle nous souhaitons inférer une fonction de récompense.
-Supposons l'existence d'une base d'entraînement $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est à mélange rapide (\emph{fast-mixing}), l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés selon cette distribution.} $\rho_E = \rho_{\pi_E}$.
-Supposons également avoir à disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eq:def:mu}.
-Une description de la manière d'estimer cette quantité en pratique est reportée à la Sec.~\ref{subsec:practicalApproach:muE}; rappelons tout de même qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé Sec.~\ref{subsec:background:irl}.
+Soit $\D = \{(s_i, a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ une base d'entraînement où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est à mélange rapide (\emph{fast-mixing}), l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés selon cette distribution.} $\rho_E = \rho_{\pi_E}$.
+Supposons également avoir à disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eeu}.
+Une description de la manière d'estimer cette quantité en pratique est reportée à la Sec.~\ref{subseracticalApproacuE}; rappelons tout de même qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé Sec.~\ref{subseackgrounrl}.
 Supposons enfin qu'un algorithme de CMC a été choisi.
@@
   \caption{Algorithme SCIRL}
-  \label{algo:scirl}
+  \label{algcirl}
   %
@@
 
-L'approche proposée est résumé par l'Alg.~\ref{algo:scirl}.
+L'approche proposée est résumé par l'Alg.~\ref{algcirl}.
 Le nom de l'algorithme (SCRIL, que l'on peut traduire par Classification structurée pour l'ARI), vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
@@
 Cet algorithme requiert une estimation de l'attribut moyen de l'expert mais il ne s'agit là que d'un problème d'évaluation de la politique, notoirement moins difficile que la recherche de politiques optimales qu'implique la résolution du problème direct.
-Cela est discuté plus en détail Sec.~\ref{sec:relatedWorks}.
+Cela est discuté plus en détail Sec.~\ref{seelatedWorks}.
 %
@@
 \subsection{Analyse}
-\label{subsec:scirl:analysis}
+\label{subsecirnalysis}
 
-Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi-optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément que le terme
-$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ peut être contrôlé.
+Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi-optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément qu'il est possible de contrôler le terme
+$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$.
 Avant de présenter le résultat principal, il nous faut introduire quelques notations et définir quelques objets.
 
-Nous allons utiliser le coefficient de concentration $C_f$~\cite{Munos:2007}:
+Nous allons utiliser le coefficient de concentration $C_f$~\cite{Muno007}:
 \begin{equation}
@@
   \label{th}
-  Soit $\R_{\theta_c}$ la fonction de récompense renvoyée par l'Alg.~\ref{algo:scirl}. Soient $C_f$, $\epsilon_c$
+  Soit $\R_{\theta_c}$ la fonction de récompense renvoyée par l'Alg.~\ref{algcirl}. Soient $C_f$, $\epsilon_c$
   et $\bar{\epsilon}_Q$ les quantités définies ci-dessus. On a :
@@
   L'inégalité $(a)$ est valable car $T^{\pi^*} v^{\pi_E}\leq T^*
-  v^{\pi_E}$ et l'inégalité $(b)$ l'est en vertu de~\cite[Lemme~4.2]{Munos:2007}.
+  v^{\pi_E}$ et l'inégalité $(b)$ l'est en vertu de~\cite[Lemme~4.2]{Muno007}.
 De plus, $v^*$ étant optimale, il est visible que  $v^*-v^{\pi_E}\geq 0$ et avec $T^*$ l'opérateur d'optimalité de Bellman, $T^* v^{\pi_E}\geq
@@
     \E_{s\sim\rho_E}\left[[T^*v^{\pi_E}](s) - v^{\pi_E}(s)\right].
-    \label{eq:proof:residual}
+    \label{erooesidual}
   \end{equation}
-  Ce résultat est similaire à celui de~\cite[Theoreme~4.2]{Munos:2007}. Il reste à borner le résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
+  Ce résultat est similaire à celui de~\cite[Theoreme~4.2]{Muno007}. Il reste à borner le résidu de Bellman $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
   v^{\pi_E}(s)]$. Considérons la décomposition
@@
     + T^{\pi_c}v^{\pi_E}- v^{\pi_E}.
-    \label{eq:proof:decomposition}
+    \label{erooecomposition}
   \end{equation}
@@
     v^{\pi_E}](s)\right] \leq \bar{\epsilon}_Q.
-    \label{eq:proof:b1}
+    \label{eroo1}
   \end{equation}
@@
   \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, nous obtenons :
-  \begin{equation}
-    T^{\pi_c}v^{\pi_E} - v^{\pi_E} = \gamma M
+  \begin{align}
+    &T^{\pi_c}v^{\pi_E} - v^{\pi_E} = \gamma M
     (P_{\pi_c}-P_{\pi_E})v^{\pi_E}
-    \Rightarrow |\rho_E^\top (T^{\pi_c}v^{\pi_E} - v^{\pi_E})| = \gamma
+    \\\Rightarrow &|\rho_E^\top (T^{\pi_c}v^{\pi_E} - v^{\pi_E})| = \gamma
     |\rho_E^\top M (P_{\pi_c}-P_{\pi_E})v^{\pi_E}|.
-  \end{equation}
+  \end{align}
   Il est dacile de voir que $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
@@
     \epsilon_c \frac{2\gamma}{1-\gamma} \|\R\|_\infty.
-    \label{eq:proof:b2}
+    \label{eroo2}
   \end{equation}
-  Injecter les bornes des Eqs.~\eqref{eq:proof:b1}
-  and~\eqref{eq:proof:b2} dans l'Eq.~\eqref{eq:proof:residual} achève la démonstration.%, which concludes the proof.
-%  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
-%  and~\eqref{eq:proof:b2} into Eq.~\eqref{eq:proof:residual} using the
-%  decomposition of Eq.~\eqref{eq:proof:decomposition} gives the
+  Injecter les bornes des Eqs.~\eqref{eroo1}
+  and~\eqref{eroo2} dans l'Eq.~\eqref{erooesidual} achève la démonstration.%, which concludes the proof.
+%  Injecting bounds of Eqs.~\eqref{eroo1}
+%  and~\eqref{eroo2} into Eq.~\eqref{erooesidual} using the
+%  decomposition of Eq.~\eqref{erooecomposition} gives the
 %  stated result:
@@
 \section{Mise en pratique}
-\label{sec:practicalApproach}
+\label{seracticalApproach}
 
 \subsection{Estimation de l'attribut moyen de l'expert}
-\label{subsec:practicalApproach:muE}
+\label{subseracticalApproacuE}
 
@@
 
-Si le modèle est inconnu, tous les algorithmes d'apprentissage par différences temporelles peuvent être utilisés pour obtenir une estimation de l'attribut moyen de l'expert~\cite{Klein:2011}, comme par exemple LSTD (Least-Squares Temporal
-Differences)~\cite{Bradtke:1996}. Soit $\psi:\s\times \A \rightarrow
+Si le modèle est inconnu, tous les algorithmes d'apprentissage par différences temporelles peuvent être utilisés pour obtenir une estimation de l'attribut moyen de l'expert~\cite{Klei011}, comme par exemple LSTD (Least-Squares Temporal
+Differences)~\cite{Bradtk996}. Soit $\psi:\s\times \A \rightarrow
 \mathbb{R}^d$ un vecteur d'attributs composé de $d$ fonctions de base
@@
 \end{bmatrix}\in\mathbb{R}^{d\times p}$ la matrice de tous les vecteurs de paramètres.
-Appliquer LSTD sur chacune des composantes de l'attribut moyen donne l'algorithme LSTD-$\mu$~\cite{Klein:2011}: $\Xi =
+Appliquer LSTD sur chacune des composantes de l'attribut moyen donne l'algorithme LSTD-$\mu$~\cite{Klei011}: $\Xi =
 (\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
@@
 %
-Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique à imiter, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
+Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique à imiter, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsecirnalysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
 Si $K$
@@
 Il est heureusement possible de recourir à des heuristiques ; ce cas ne rentre pas dans l'analyse présentée en
-Sec.~\ref{subsec:scirl:analysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en Sec.~\ref{sec:experiments}.
+Sec.~\ref{subsecirnalysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en Sec.~\ref{sexperiments}.
 
@@
 %  \hat{\mu}^{\pi_E}(s_i,a_i).
-%  \label{eq:heuristic}
+%  \label{eeuristic}
 %\end{equation}
@@
   \hat{\mu}^{\pi_E}(s_i,a_i).
-  \label{eq:mc_plus_heuristic}
+  \label{ec_plus_heuristic}
 \end{equation}
@@
 Cette estimation n'est nécessaire que pour entraîner le classifieur, il est donc suffisant de disposer de valeurs uniquement pour les couples état-action $(s_i,a)_{1\leq i \leq N,a\in \A}$.
-En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir Sec.~\ref{sec:relatedWorks}).
+En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir Sec.~\ref{seelatedWorks}).
 
 \subsection{Instanciation}
-\label{subsec:practicalApproach:instantiation}
+\label{subseracticalApproacnstantiation}
 
-Comme précisé précedemment, tout algorithme de CMC peut être utilisé. Ici nous choisissons l'approche à marges structurées de ~\cite{Taskar:2005}. Soit
+Comme précisé précedemment, tout algorithme de CMC peut être utilisé. Ici nous choisissons l'approche à marges structurées de ~\cite{Taska005}. Soit
 $\lc:\s\times\A\rightarrow\mathbb{R}_+$ une fonction de marge définie par l'utilisateur satisfaisant $\lc(s,\pi_E(s))\leq \lc(s,a)$ (ici,
@@
   \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
-  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a). \label{eq:qp_taskar}
+  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a). \label{ep_taskar}
 \end{equation}
-De façon similaire à ~\cite{Ratliff:2006}, nous fournissons la forme \emph{hinge-loss} équivalente (avec les variables d'ajustement $\zeta_i$ serrées, ce qui permet de déplacer les contraintes dans la fonction objectif) :
+De façon similaire à ~\cite{Ratlif006}, nous fournissons la forme \emph{hinge-loss} équivalente (avec les variables d'ajustement $\zeta_i$ serrées, ce qui permet de déplacer les contraintes dans la fonction objectif) :
 \begin{equation}
@@
 \end{equation}
-La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit Eq.~\eqref{eq:mc_plus_heuristic}.
+La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit Eq.~\eqref{ec_plus_heuristic}.
 
-%This is the instantiation we use in Sec.~\ref{sec:experiments}, but
+%This is the instantiation we use in Sec.~\ref{sexperiments}, but
 %our approach is more general: other choices could have been done for
@@
 %parameter vector such that $\pi_E$ is the unique optimal policy),
-%constraints of Eq.~\eqref{eq:qp_taskar} can be satisfied for all
+%constraints of Eq.~\eqref{ep_taskar} can be satisfied for all
 %state-action couples. This means that the proposed approach is able
@@
 \section{Travaux connexes}
-\label{sec:relatedWorks}
+\label{seelatedWorks}
 
-La notion d'ARI a été pour la première fois introduite dans~\cite{Russell:1998}
-puis formalisée dans~\cite{Ng:2000}.
-Une approche classique, initiée dans~\cite{Abbeel:2004}, consiste à trouver une politique (à travers une recherche dans l'espace des fonctions de récompense) telle que son attribut moyen (ou de manière plus générale une mesure de la distribution sous jacente aux trajectoires) s'approche de celui de la politique de l'expert.
+La notion d'ARI a été pour la première fois introduite dans~\cite{Russel998}
+puis formalisée dans~\cite{N000}.
+Une approche classique, initiée dans~\cite{Abbee004}, consiste à trouver une politique (à travers une recherche dans l'espace des fonctions de récompense) telle que son attribut moyen (ou de manière plus générale une mesure de la distribution sous jacente aux trajectoires) s'approche de celui de la politique de l'expert.
 %
-Voir~\cite{Neu:2010} pour un état de l'art partiel.
+Voir~\cite{Ne010} pour un état de l'art partiel.
 %
-%This can be done based on game theory~\cite{Syed:2008:game}, linear
-%programming~\cite{Syed:2008:lp}, maximum
-%entropy~\cite{Ziebart:2008}, Bayesian posterior
-%maximization~\cite{Ramachandran:2007}, \textit{etc}. Some of these
-%approaches et others are nicely reviewed in~\cite{Neu:2010}.
+%This can be done based on game theory~\cite{Sye00ame}, linear
+%programming~\cite{Sye00p}, maximum
+%entropy~\cite{Ziebar008}, Bayesian posterior
+%maximization~\cite{Ramachandra007}, \textit{etc}. Some of these
+%approaches et others are nicely reviewed in~\cite{Ne010}.
 %
@@
 %Notice that related algorithms do not always output a reward
-%function (\textit{e.g.}, in~\cite{Abbeel:2004}, the algorithm
+%function (\textit{e.g.}, in~\cite{Abbee004}, the algorithm
 %produces a mixed policy; each intermediate policy is optimal
@@
 
-Plus proche de notre contribution, quelques approches introduisent également de la structure dans la classification~\cite{Melo:2010,Ratliff:2006}. Dans~\cite{Melo:2010},
+Plus proche de notre contribution, quelques approches introduisent également de la structure dans la classification~\cite{Mel010,Ratlif006}. Dans~\cite{Mel010},
 une métrique induite par le PDM est utilisée pour construire un noyau qui sera utilisé par l'algorithme de classification, permettant des améliorations par rapport à un noyau non structuré.
 Cette approche n'est cependant pas un algorithme d'ARI, et plus important l'évaluation d'une métrique dans un PDM n'est pas triviale.
-Dans~\cite{Ratliff:2006}, un algorithme de classification est utilisé pour fournir une fonction de récompense.
+Dans~\cite{Ratlif006}, un algorithme de classification est utilisé pour fournir une fonction de récompense.
 Au lieu d'associer des actions à des états comme nous le faisons, cet algorithme associe des politiques optimales (qui jouent le rôle de labels) à des PDM (entrées), ce qui permet d'incorporer la structure, au prix de la résolution d'un grand nombre de PDM.
 
-Tous les algorithmes d'ARI à notre connaissance requièrent la résolution du problème direct de l'AR de manière répétée, à l'exception de~\cite{Dvij:2010,boularias:2011}.
-Le travail présenté dans \cite{Dvij:2010} ne s'applique qu'aux PDM solvables linéairement (où le contrôle se fait en imposant une dynamique au système).
-Dans~\cite{boularias:2011}, en utilisant l'argument de l'entropie relative, une fonction objectif est maximisée via une montée de sous-gradient.
+Tous les algorithmes d'ARI à notre connaissance requièrent la résolution du problème direct de l'AR de manière répétée, à l'exception de~\cite{Dvi010,boularia011}.
+Le travail présenté dans \cite{Dvi010} ne s'applique qu'aux PDM solvables linéairement (où le contrôle se fait en imposant une dynamique au système).
+Dans~\cite{boularia011}, en utilisant l'argument de l'entropie relative, une fonction objectif est maximisée via une montée de sous-gradient.
 Estimer la valeur du sous-gradient demande des trajectoires échantillonnées selon la politique optimale pour la fonction de récompense courante.
@@
 Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une seule fois un problème d'évaluation de politique.
-Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI, par exemple~\cite{boularias:2011}).
+Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI, par exemple~\cite{boularia011}).
 De plus, par l'utilisation d'une heuristique qui dépassent le cadre de notre analyse, il est possible à SCIRL de se contenter de données fournies par l'expert.
@@
 \section{Expériences}
-\label{sec:experiments}
+\label{sexperiments}
 
-Nous illustrons l'approche proposée sur un simulateur de conduite similaire à~\cite{Abbeel:2004,Syed:2008:game}.
+Nous illustrons SCIRL sur un simulateur de conduite similaire à~\cite{Abbee004,Sye00ame}.
 Le but est de conduire une voiture sur une autoroute à trois voies dont le traffic est généré aléatoirement (les sorties de route sont possible des deux côtés).
@@
 Nous avons comparé SCIRL, tel qu'instancié comme décrit en
-Sec.~\ref{subsec:practicalApproach:instantiation}, à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au coeur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbeel:2004} (appelé ici PIRL pour \emph{
+Sec.~\ref{subseracticalApproacnstantiation}, à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au coeur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbee004} (appelé ici PIRL pour \emph{
 Projection Inverse Reinforcement Learning}).
@@
 Pour SCIRL et PIRL on discrétise l'espace d'état en un vecteur d'attribut pour la récompense, $\phi\in\mathbb{R}^{729}$: $9$ positions horizontales pour la voiture du joueur, $3$ positions horizontales et $9$ verticales pour la voiture la plus proche de celle du joueur et $3$ vitesses.
-Ces attributs sont bien moins informatifs que ceux utilisés dans~\cite{Abbeel:2004,Syed:2008:game}.
-Les attributs de~\cite{Syed:2008:game} sont si informatifs que tirer aléatoirement un vecteur $\theta$ de paramètres positifs pour la récompense donne lieu à un comportement acceptable.
+Ces attributs sont bien moins informatifs que ceux utilisés dans~\cite{Abbee004,Sye00ame}.
+Les attributs de~\cite{Sye00ame} sont si informatifs que tirer aléatoirement un vecteur $\theta$ de paramètres positifs pour la récompense donne lieu à un comportement acceptable.
 Le facteur d'oubli est $\gamma = 0.9$.
@@
 
-\begin{figure}[fig:res]{Problème de la conduite.
+\begin{figure}[fies]{Problème de la conduite.
 La ligne la plus haute est la valeur de l'expert.
@@
 
-La Fig.~\ref{fig:res} montre la performance de chacune des approche en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
+La Fig.~\ref{fies} montre la performance de chacune des approche en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
 On peut voir que le classifieur ne fonctionne pas bien sur ce problème.
@@
 %Recall that for SCIRL the instantiation of
-%Sec.~\ref{subsec:practicalApproach:instantiation} is considered,
+%Sec.~\ref{subseracticalApproacnstantiation} is considered,
 %with a single Monte-Carlo rollout for estimating
@@
 \section{Conclusion}
-\label{sec:conclusion}
+\label{seonclusion}
 

### 2012-10-01 Rapport2A/main.tex
Truc

@@
 \section*{Sujet de thèse}
-L'adoption de systèmes automatisés autonomes peut réduire les coûts et les risques, cependant la complexité de la mise au point de tels systèmes gêne leur adoption généralisée.
+L'adoption de systèmes automatisés autonomes peut réduire les coûts et les risques.
+Cependant, la complexité de leur mise au point gêne leur adoption généralisée.
 Des approches permettant à ces systèmes de prendre en charge des environnements de plus en plus difficiles font apparaître de nouveaux défis.
@@
 
-  Notre but est de dériver, du comportement d'un expert effectuant une tâche, une description de cette tâche sous la forme d'une fonction de récompense, ce qui permet ensuite l'utilisation des outils d'apprentissage par renforcement pour apprendre cette tâche à un agent.
+  Notre but est de dériver du comportement d'un expert effectuant une tâche une description de cette tâche sous la forme d'une fonction de récompense, ce qui permet ensuite l'utilisation des outils d'apprentissage par renforcement pour apprendre cette tâche à un agent.
 Cela ouvrirait le champ d'application de l'apprentissage par renforcement à des tâches encore inaccessibles car trop complexes pour être ``expliquées''.\\
-  L'intitulé de la thèse, \emph{Découverte automatique d'attributs pour l'imitation}, isole une partie du problème : il s'agit d'extraire de la description de la suite d'\emph{états} traversés par l'expert les informations pertinentes à l'expression de la récompense.\\
+  L'intitulé de la thèse, \emph{Découverte automatique d'attributs pour l'imitation}, isole une partie du problème : il s'agit d'extraire de la description de la suite d'\emph{états} traversés par l'expert les informations pertinentes pour l'expression de la récompense.\\
 
@@
 La contribution centrale intervient en 2004 et introduit la notion d'\emph{attribut moyen}, que suggérait déjà l'analyse de 2000.
-Après 2004, plusieurs travaux apparaissent qui utilisent cette notion d'attribut moyen .\\
+Après 2004, plusieurs travaux apparaissent qui utilisent cette notion d'attribut moyen.\\
   L'attribut moyen est une mesure liée à la distribution des états que traverse un agent en suivant sa politique dans l'environnement.
-Cette mesure tient une place centrale dans beaucoup des algorithmes existants, notamment car deux agents ayant des attributs moyens similaires rempliront une certaine tâche (\emph{i.e.}, optimiseront une certaine récompense) de manière similaire.\\
+Cette mesure tient une place centrale dans nombre d'algorithmes existants, car deux agents ayant des attributs moyens similaires rempliront une certaine tâche (\emph{i.e.}, optimiseront une certaine récompense) de manière similaire.\\
 
-  La première contribution au domaine consiste en un mécanisme de calcul de cet attribut moyen \cite{klein2011batch}.
+  Notre première contribution au domaine consiste en un mécanisme de calcul de cet attribut moyen \cite{klein2011batch}.
 Inspiré d'algorithmes existants pour l'approximation de fonction de valeur, thème central en apprentissage par renforcement, cette contribution apporte une méthode de calcul permettant l'évaluation \emph{off-policy} de l'attribut moyen d'une politique.
@@
 Testée en l'injectant dans l'approche centrale de 2004, cette idée a permis de résoudre les problème de l'apprentissage par renforcement inverse sur un problème jouet simple à partir des seules données issues de l'expert, mais n'a pas permis de complètement lever les obstacles imposés par la structure des algorithmes existants.
-Ceux-ci nécessitent en effet dans leur immense majorité de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée.
-Cela n'est pas toujours possible uniquement avec les données de l'expert, il faut une autre source d'information.\\
+Ceux-ci nécessitent en effet très majoritairement de résoudre le problème direct (celui de l'apprentissage par renforcement) de manière répétée.
+Cela n'est pas possible uniquement avec les données de l'expert à part sur certains problèmes jouet.
+De manière générale, les algorithmes de la littérature ont besoin d'une autre source d'information que les données de l'expert.\\
   Cette contribution permet néanmoins de s'affranchir d'un simulateur pour estimer une grandeur centrale dans la majorité des approches du domaine.
-Cela permet d'éviter les soucis liés à la modélisation de l'environnement ou à un trop grand besoin en données (coûteuses à générer).
+Cela permet d'éviter les soucis liés à la modélisation de l'environnement ou à un trop grand besoin en données (coûteuses à produire).
   Une autre contribution de plus faible envergure consiste en un apport sur la définition mathématique formelle du problème \cite{klein2011dimensionality} permettant de réduire l'espace dans lequel on doit chercher les solutions.\\
-  Enfin, lors de la deuxième année de thèse, deux nouveaux algorithmes ont été trouvés qui permettent la résolution de l'ARI dans des domaines où les algorithmes existant ne fonctionnent pas.
-L'un de ces algorithmes a été publié dans une conférence francophone \cite{Supelec773} ainsi qu'une conférence internationale \cite{klein2012}.
-L'autre est en cours de relecture.
+  Nous avons poursuivi, lors de la deuxième année de thèse, en proposant deux nouveaux algorithmes qui permettent la résolution de l'ARI dans des domaines où les algorithmes existants ne fonctionnent pas.
+Les différentes contributions autour d'un de ces algorithmes a donné lieu à trois publications, dans une conférence francophone \cite{Supelec773} ainsi que deux conférences internationales \cite{klein2012,klein20122}.
+Le travail sur l'autre algorithme se poursuit.
 
-  Il m'a été permis de participer à cinq conférences en temps qu'auteur, j'ai pu rencontrer la communauté \emph{Machine Learning} au sens large à IJCAI \footnote{\href{http://ijcai-11.iiia.csic.es/}{http://ijcai-11.iiia.csic.es/}
+  J'ai pu participé à cinq conférences en temps qu'auteur, j'ai pu rencontrer la communauté \emph{Machine Learning} au sens large à IJCAI \footnote{\href{http://ijcai-11.iiia.csic.es/}{http://ijcai-11.iiia.csic.es/}
  }, ainsi qu'à ECML\footnote{\href{http://www.ecmlpkdd2011.org/}{http://www.ecmlpkdd2011.org/}
  } qui précédait EWRL\footnote{\href{http://ewrl.wordpress.com/past-ewrl/ewrl9-2011/}{http://ewrl.wordpress.com/past-ewrl/ewrl9-2011/}
- } où j'ai pu présenter mes travaux à un public plus directement concerné du fait du sujet précis de ce \emph{workshop}.
+ } où j'ai pu présenter mes travaux à un public plus directement concerné par l'ARI du fait du sujet précis de ce \emph{workshop}.
 Enfin j'ai rencontré la communauté francophone à JFPDA\footnote{\href{https://zanuttini.users.greyc.fr/jfpda2011/}{https://zanuttini.users.greyc.fr/jfpda2011/}} puis l'année suivante à CAp \footnote{http://cap2012.loria.fr/}.
@@
 
-Outre la rédaction de la thèse, qui sera commencée aux alentours de Février pour une remise du manuscrit en Juin et une soutenance en Septembre, cette année sera occupée par l'application de nos nouveaux algorithmes sur des problèmes complexes et tirés de cas concrets afin de démontrer empiriquement leur performance.
+Outre la rédaction de la thèse, qui sera commencée aux alentours de février pour une remise du manuscrit en juin et une soutenance en septembre, cette année sera occupée par l'application de nos nouveaux algorithmes sur des problèmes complexes et tirés de cas concrets afin d'évaluer empiriquement leur performance.
 

### 2012-10-15 piot2012cascading/ECML13_cascading.tex
Starting work on the truc document

% Created 2012-10-15 lun. 14:49
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\usepackage{amsmath}
\usepackage{dsfont}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\title{Outline}
\author{Edouard Klein}
\date{\today}
\hypersetup{
  pdfkeywords={},
  pdfsubject={},
  pdfcreator={Emacs Org-mode version 7.8.11}}

\begin{document}

\maketitle

\setcounter{tocdepth}{3}
\tableofcontents
\vspace*{1cm}
\section{Abstract}
\label{sec-1}
\section{Introduction}
\label{sec-2}

  De la même manière que pour le papier de SCIRL, l'intro est une vaste annonce du papier, avec des références pointues (sous-section, voire sous,sous-section) vers les parties du papier qui expliquent tel ou tel point en détail.
\section{Background and notations}
\label{sec-3}
\begin{itemize}
\item Parler du domaine de la prise de décision séquentielle
\item Expliquer le problème direct, en profiter pour introduire l'équation d'évaluation de Bellman et l'équation d'optimalité de Bellman
\begin{itemize}
\item $RL(MDP=\{S,A,\gamma,(P),R\}) = \pi^*:S\times A \rightarrow \mathbb{R}$
\item $\forall \pi, Q^\pi(s,a) = R(s,a) + \sum_{s'}P(s'|s,a)Q^\pi(s',\pi(s'))$
\item $\pi^* = \arg\max_aQ^*(s,a)$
\end{itemize}
\item Introduire le problème inverse dans une forme mathématique relativement propre
\begin{itemize}
\item $IRL( MDP\backslash R = \{S,A,\gamma,(P)\} \cup \{\pi^E\} = R : S\times A \rightarrow \mathbb{R}$ telle que $\pi^E$ est optimale pour $R$
\item Mentionner la nature mal posée du problème sans s'appesentir dessus. Préciser toutefois qu'aucune approche ne le résoud de manière mathématiquement satisfaisante à cause de ça.
\item insister sur l'optimalité de l'expert comme hypothèse. Un expert sous optimal, c'est un autre champ de littérature, éventuellement donner deux trois refs.
\end{itemize}
\end{itemize}
We frame the sequential decision problem in a \emph{Markov Decision Process} (MDP), that is a tuple $\{S,A,P,R,\gamma\}$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ it is in to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but this dos not induce any loss of generality.}) \emph{policy} $\pi: S \rightarrow A$. The agent then transits to a new state $s_{t+1}\in S$ according to the markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The reward function $R : S\times A \rightarrow \mathbb{R}$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{|S|}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state : $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t)|s_0 = s]$. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, \forall s \in S, V^*_R(s) \geq V^\pi_R(s)$. The stationary distribution induced by a policy $\pi$ is written $\rho_\pi,\textrm{ s.t. }\rho_\pi^TP^\pi = \rho_\pi$.

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{|S|} \rightarrow  \mathbb{R}^{|S|}$ is defined by $T^\pi_RV = R + \gamma P^piV$ using the straightforward matricial notation $P\pi = (p(s'|s,\pi(s)))_{s,s' \in S}$. The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$. Both these operator are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R $V^\pi_R$ ; the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$. Optimal value function and optimal policy are also linked via the quality function $Q^\pi_R$, a state-action value function that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_Rv^\pi_R(s)$, with $a$ the policy that always returns action $a$ : the optimal policy follows a greedy mechanism with respect to its optimal quality function $\pi^*_R(s) = \argmax_aQ^*_R(s,a)$.

When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, \emph{Reinforcement Learning} (RL) or approximate dynamic programming may be used to retireve the optimal control policy. We call this the direct problem.

In this work, we wish to tackle the inverse problem. We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal. The non optimality of the expert is an interesting setting that has been discussed for example FIXME:NONOPTIMALITY but that we are not addressing here. We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is also optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see section \ref{sec:related}. 
\section{The cascading algorithm}
\label{sec-4}
\begin{itemize}
\item Mentionner (sans en parler, ce sera fait dans \hyperref[sec-7-1]{Imitation}) l'imitation. Une approche supervisée évidente est la classification, avec pour entrée un set $D_C$ de données issues de l'expert. Rappeler (expliqué pour la première fois dans \hyperref[sec-3]{Formalisme} qu'on se restreint (sans perte de généralité, le préciser) au cas des politique déterministes.
\item Tous les classifieurs utilisent une fonction de score (je n'ai pas trouvé un seul exemple d'un classifieur qui n'en utilise pas), il s'agit juste d'un manière d'écrire les choses.
\item Expliquer que l'on identifie le mécanisme de classification au mécanisme glouton qui régit la politique optimale (donc celle de l'expert).
\item Sans chercher à justifier mathématiquement la chose (ce sera fait dans la partie \hyperref[sec-5]{Analyse}), expliquer qu'intuitivement l'on va identifier fonction de score du classifieur et fonction de qualité de l'expert (qui est optimale pour une certaine récompense, rappelons-le).
\begin{itemize}
\item Repartir de l'équation d'évaluation de Bellman pour isoler le terme $R(s,a)$ et introduire $r:S\times A\in \mathbb{R}$, la récompense issue de cette équation lorsque l'on remplace $Q^E$ par la fonction de score : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi(s')) $
\item On estime ce terme à partir des transitions $D_R$ car les probabilités de transitions sont inconnues : $\hat r(s,a) = q(s,a)-\gamma \sum_{s'} {1\over N_{s'}}\sum_{(s,a,s',a')\in D_R}q(s',a')$
\item La fonction de récompense est donnée par une régression sur $D_R \cup {\hat r}$.
\end{itemize}
\item Donner l'algorithme l'environnement \texttt{algorithm2e}
\item Rapidement, l'étape de régression permet d'ajouter la structure du MDP au résultat de classif, en utilisant l'équation de Bellman.
\item Bien préciser qu'il ne s'agit ici que d'une explication qualitative, ``avec les mains'', la preuve du fait que ça résoud bien le problème est donnée dans \hyperref[sec-5]{Analyse}, une démonstration sur des problèmes si complexes que l'état de l'art ne peut les résoudre (sauf SCIRL) est donnée dans \hyperref[Experience]{Experience}. Les avantages comparés sont expliqués dans \hyperref[Autres-Approches]{Autres Approches}. Enfin, des explications plus intuitives sur les mécanismes internes de l'algorithme, ainsi que des résultats expérimentaux supplémentaires sont fournis en \hyperref[sec-9]{Annexes}.
\end{itemize}
IRL has sometimes FXME:REFIMITATION been seen as a way to cast the imitation problem, that is a way to copy the expert policy ; the reward function is then a byproduct of the analysis and not the object one is looking for. An efficient supervised way to do pure imitation is to use a classifier. The entries are the states $s_t$ and the labels are the actions $a_t$ deterministically chosen by the expert. We thus obtain a classification policy $\pi_C$.

This is the first step of our algorithm. Using a dataset $D_C = {s_i,a_i=\pi^E(s_i)}_i$ we train a score-function based classifier. We can make use of any classifier whose decision rule uses a greedy mechanism with respect to a score function $q : S \times A \rightarrow \mathbb{R}$ that quantifies how well label $a\in A$ suits entry $s\in S$. Most classifiers, from M-SVM to $k$-nearest neighboors can be cast as a score-function based classifier. This allows the use of many off-the-shelf imlpementations.

The greedy classification mechanism $\pi_C = \argmax_a q(s,a)$ is very similar to the relation between the optimal policy in a MDP and its optimal quality function : $\pi^*_R(s) = \argmax_aQ^*_R(s,a)$. As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal quality function. It is not a stretch to see $\pi_C$ as an optimal policy as it has been built from $\pi^E$ which is optimal. By inversing the Bellman evaluation equation, one can express the reward as a function of the quality function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$. This leave us with $r$, the reward function relative to our score/quality function $q$ : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s'))$.

We wish to solve the general, or approximate IRL problem, where the transition probabilities are unknown. Our reward function $r$ must be approximated to $\hat r$ with the help of a dataset $D_R = \{s_i,a_i,s'_i\}$. This gives as many datapoints $\{\hat r(s_i,a_i) = q(s_i,a_i) - \gamma q(s',\pi_C(s'))}_i $.

\section{Analyse}
\label{sec-5}

\begin{itemize}
\item On rappelle le problème posé au début.
\item Le théorème 1 montre que modulo un bonne classif et un bon régresseur, on résoud exactement le problème que l'on s'était posé
\item On en fait la démo rapide (éventuellement des bouts peuvent être déportés en \st{Pologne} \hyperref[sec-9]{Annexes}).
\item On insiste sur le fait que la récompense trouvée n'est pas triviale, on a transcendé la nature mal posée du problème.
\item On dit un mot rapide sur la complexité en échantillon (ça dépend du classifieur et du régresseur, mais \hyperref[sec-6]{Expériences} montre qu'on est bon là dessus)
\item ainsi que sur la complexité temporelle (ça dépend toujours, mais avec une M-SVM et un moinde carré ça prend O(quelque chose) (indice: pas beaucoup)) FIXME:compléter la complexité
\end{itemize}

First let introduce some notations. $\pi_C$ is the classifier policy, $\epsilon_C$ is the classification error:
\begin{equation}
\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
\end{equation}
Let $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$, we introduce $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$.
Thus we have the following vectorial equations:
\begin{align}
&R^C_a=q_a-\gamma P_aq_{\pi_C},
\\
&\epsilon^R_a=R^C_a-\hat{R}^C_a.
\end{align}
Now we are going to bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious).
We note $\hat{\pi}_C$ a deterministic optimal policy of the reward $\hat{R}^C$. We have:
\begin{equation}
V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}=(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C})+(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})+(V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}).
\end{equation}
We are going to bound each of this three terms. But, first, let $\pi$ be a given deterministic policy. We have:
\begin{equation}
V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C}=V^{\pi}_{\epsilon^R}=(I-\gamma P_\pi)^{-1}\epsilon^R_{\pi}.
\end{equation}
If $\pi=\pi_E$, $\rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\epsilon^R_{\pi_E}=\frac{1}{1-\gamma}\rho_E^T\epsilon^R_{\pi_E}\leq\frac{1}{1-\gamma}\|\epsilon^R_{\pi_E}\|_{1,\rho_E}$.
If $\pi\neq\pi_E$, we introduce the following concentration coefficient:
\begin{equation}
C_{\pi}=(1-\gamma)\sum_{t\geq0}\gamma^tc_{\pi}(t), \text{ with } c_{\pi}(t)=\max_{s\in S}\frac{(\rho_E^TP^t_\pi)(s)}{\rho_E(s)}.
\end{equation}
We have then: $\rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})\leq\frac{C_\pi}{1-\gamma}\rho_E^T\epsilon^R_{\pi}\leq\frac{C_\pi}{1-\gamma}\|\epsilon^R_{\pi}\|_{1,\rho_E}$. So, if we note, $C_*=C_{\hat{\pi}_C}$ and $\epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E}$, we are able to give an upper bound to the first and third terms:
\begin{equation}
\rho_E^T((V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})+(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}))\leq\frac{1+C_*}{1-\gamma}\epsilon_R.
\end{equation}
Now, there is a still an upper bound to find for the second term. If we introduce, $R^E_a=q_a+\gamma P_aq_{\pi_E}$, it is possible to decompose the second term:
\begin{equation}
V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}=(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C})+(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})+(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}).
\end{equation}
By construction, $\pi_C$ is optimal for $R^C$, so $V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C}\leq0$ which implies:
\begin{equation}
V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}\leq(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})+(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}).
\end{equation}
By construction, we have $V^{\pi_C}_{R^C}=q_{\pi_C}$ and $V^{\pi_E}_{R^E}=q_{\pi_E}$, thus:
\begin{equation}
\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})=\rho_E^T(q_{\pi_C}-q_{\pi_E})=\sum_{s\in S}\rho_E(s)(q(s,\pi_C(s))-q(s,\pi_E(s)))[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
\end{equation}
We note $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$ and we have:
\begin{equation}
\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})\leq\Delta q\sum_{s\in S}\rho_E(s)[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]= \Delta q \epsilon_C.
\end{equation}
Finally, we also have:
\begin{align}
\rho_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(R^E_{\pi_E}-R^E_{\pi_C}),
\\
&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\gamma P_{\pi_E}(q_{\pi_C}-q_{\pi_E}),
\\
&=\frac{\gamma}{1-\gamma}\rho_E^T(q_{\pi_C}-q_{\pi_E})\leq \frac{\gamma}{1-\gamma}\Delta q \epsilon_C.
\end{align}
So the upper bound for the second term term is: $\rho_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\Delta q}{1-\gamma}\epsilon_C$.
If we combine all of the results, we obtain the final bound:
\begin{equation}
0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\leq \frac{1}{1-\gamma}(\Delta q \epsilon_C+(1+C_*)\epsilon_R)).
\end{equation}

Few remarks: the bound is obviously true for $\hat{R}_C$, but in this case the classifier or the regressor have bad performance which it is not likely. The algorithm and the analysis are invariant by a dilatation of $q$, so of $\Delta q$. We could normalize $\Delta q$ by imposing $\Delta q=1$ (the range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$). The constant $C_*$ could be estimated a posteriori ( after $\hat{R}^C$ being calculated). We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ give a tighter final result (approximately, if $\hat{\pi_C}\approx\pi_E$ then $C_*=1$).

In order to control $\epsilon_R$, we need a precise regression over the expert stationary distribution and for all the actions. In theory, we need to be able to sample the whole dynamic (even the non expert actions). In practice, a heuristic is used (however this heuristic is not justified by the analysis).
Finally, if the classifier and the regressor are perfect ($\epsilon_C=\epsilon_R=0$), $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness suppose that $\forall s\in S, Card(\argmax_{a\in A}q(s,a))=1$ ).

\section{Expériences}
\label{sec-6}

\begin{itemize}
\item La borne fournie en \hyperref[sec-5]{Analyse} mesure le degré d'optimalité de la politique de l'expert vis à vis de la nouvelle récompense. Les expériences mesurent le degré d'optimalité de l'agent (ici entendu sous forme d'agent entraîné parfaitement sur la récompense sortie par notre algorithme) vis à vis de la vraie récompense (celle de l'expert).
\begin{itemize}
\item On se permet de faire ça car dans notre dispositif on la connait. Ce n'est pas le cas dans la vraie vie.
\item En pratique c'est un critère plus restrictif : formellement, si $\pi^E$ est optimale pour $R_C$, alors l'agent et l'expert ont les mêmes performances (wrt $R_C$), puisque l'agent est optimal pour $R_C$ par définition de l'agent. En revanche on regarde si l'agent est optimal pour $R_E$, ce qui est profondément ``injuste'', puisqu'on ne sait rien de $R_E$. On ne sait même pas si elle existe réellement.
\item Il s'agit d'une métrique qu'il n'est pas possible de mesurer sur un problème réel. C'est pourquoi la démonstration est intéressente car elle parle de quelque chose que l'on peut plus facilement évaluer ; et surtout elle parle de ce que l'on s'est proposé de résoudre à la base. Ces expériences ne sont qu'une illustration.
\end{itemize}
\item Les résultats obtenus sur ce problème sont notamment comparés à la classif, il s'agit d'une illustration du fait que l'algorithme est capable d'extraire les infos sur $P$ contenues dans les transitions, grâce à l'équation de Bellman qui est présente (modulo deux trois manips) dans l'expression de $\hat r$. Nous avons injecté la structure du MDP dans la classification (ça me rappelle quelque chose\ldots{} ;)
\item Le seul plot est celui du Highway (similaire à SCIRL). On enlève le Gridworld (reporté en annexe car trop d'infos pour si peu de place)
\item On instancie avec Taskar et un moindre carré. Ceci n'a pas la moindre importance.
\item Mentionner le fait que les autres algos ne peuvent résoudre le problème tel qu'on le pose. On s'apesentira là dessus dans \hyperref[sec-7]{Autres approches}.
\item Expliquer que quitte à tester expérimentallement, on a choisi de s'éloigner du cadre de l'analyse pour se placer dans cadre \emph{plus difficile à gérer}, ou seules les données de l'expert sont disponibles.
\begin{itemize}
\item Cela implique l'usage d'une heuristique pour la régression :
\begin{itemize}
\item sans heuristique, c'est le choix par défauit du régresseur (i.e. initialisation à 0, à R$_{\mathrm{min}}$ ou R$_{\mathrm{max}}$) qui décidera de l'optimisme ou pessimisme de l'algo
\item Pour éviter ça, on place nous même notre heuristique.
\end{itemize}
\item que l'on se contente d'expliquer à un niveau intuitif. L'analyse formelle n'a pas été effectuée, ne le sera probablement pas, on se contente de montrer qu'en pratique, ça fonctionne. Et ça fonctionne mieux que les autres algos, sauf SCIRL.
\item Cela implique aussi d'utiliser $\pi_C$ plutôt que $a'$.
\end{itemize}
\item Lien avec l'analyse : on a de bons résultats, même si le classifieur n'est pas bon. Notre borne est donc très très large.
\item La complexité en échantillons : avec 50 samples, on fait mieux que PIRL qui a accès à une matrice de taille $|S|\times|S|\times |A|$
\item La complexité en temps : on tourne 100 fois plus rapidement que PIRL
\end{itemize}
\section{Autres approches}
\label{sec-7}

\begin{itemize}
\item On reprend peu ou prou la même que SCIRL et que celle du précédent papier. On profite de la mention d'une approche pour préciser en quoi la nôtre diffère.
\end{itemize}
\subsection{Imitation}
\label{sec-7-1}

\begin{itemize}
\item Ce n'est pas le problème que l'on se pose, mais résoudre l'IRL peut permettre une forme fine d'imitation.
\item La frontière est un peu floue, exemple de PIRL
\item On peut considérer que tous les algos revus par Neu sont des algos d'imitation dans le sens où ils cherchent une proximité des $\mu$.
\item Coincidentallement, ils ont besoin de résoudre le problème direct, pas nous.
\end{itemize}
\subsection{Méthodes entropiques}
\label{sec-7-2}

\begin{itemize}
\item N'ont pas besoin de résoudre le problème direct, mais ont besoin d'autres données que celles de l'expert.
\end{itemize}
\subsection{Autres méthodes}
\label{sec-7-3}

\begin{itemize}
\item parler des MDPs solvables linéairement, des papiers mentionnés par les reviewers de NIPS, et dire du bien de SCIRL
\end{itemize}
\section{Conclusion}
\label{sec-8}

\begin{itemize}
\item Ouverture : Existe-t-il des problèmes tels que SCIRL et Cascading diffèrent en performance ?
\item Que faire quand on ne peut résoudre le RL facilement ?
\item Que faire en cas d'expert sous optimal ?
\item ouverture sur l'active learning en reprenant les notions de pessimisme ou d'optimisme en présence d'incertitude
\end{itemize}

\section{Annexes}
\label{sec-9}
\subsection{GridWorld}
\label{sec-9-1}

\begin{itemize}
\item On montre avec les deux formes de récompenses qui donnent la même fonction de valeur. On fait référence aux travaux de Ng qui résolvent exactement ce problème. Les récompenses ont la même tête, ça devrait rassurer le lecteur hésitant.
\item si on a le temps, on peut lancer une expérience où le nombre (en supposant une répartittion uniforme) des samples augmente et où la récompense tend vers ``la vraie''.
\end{itemize}
\subsection{Inverted pendulum}
\label{sec-9-2}

\begin{itemize}
\item On montre qu'on généralise sans problème aux espaces continus, on montre qu'avec des données très, très sparses vis à vis de l'espace d'état on arrive pourtant à récupérer une récompense ayant la même fonction de valeur que celle de l'expert.
\item ça montre aussi ce qui se passe quand l'espace d'hypothèse est ``mal choisi'' (i.e. la ``vraie'' récompense n'y était pas représentable).
\end{itemize}
\subsection{Highway}
\label{sec-9-3}

\begin{itemize}
\item Si on a le temps : faire conduire quelqu'un puis lancer l'imitation, voir si on reproduit les consignes données à l'opérateur humain, qui aura eu la bonté de faire quelques erreurs dans l'exécution. Ca permet de voir le comportement en cas d'expert sous optimal et de taire les questions à ce sujet.
\end{itemize}

\end{document}

### 2012-10-16 piot2012cascading/ECML13_cascading.tex
Ajout du code pour le plot du Highway et modifs sur le papier

@@
 \usepackage{hyperref}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 \tolerance=1000
@@
 \section{Background and notations}
-\label{sec-3}
-\begin{itemize}
-\item Parler du domaine de la prise de décision séquentielle
-\item Expliquer le problème direct, en profiter pour introduire l'équation d'évaluation de Bellman et l'équation d'optimalité de Bellman
-\begin{itemize}
-\item $RL(MDP=\{S,A,\gamma,(P),R\}) = \pi^*:S\times A \rightarrow \mathbb{R}$
-\item $\forall \pi, Q^\pi(s,a) = R(s,a) + \sum_{s'}P(s'|s,a)Q^\pi(s',\pi(s'))$
-\item $\pi^* = \arg\max_aQ^*(s,a)$
-\end{itemize}
-\item Introduire le problème inverse dans une forme mathématique relativement propre
-\begin{itemize}
-\item $IRL( MDP\backslash R = \{S,A,\gamma,(P)\} \cup \{\pi^E\} = R : S\times A \rightarrow \mathbb{R}$ telle que $\pi^E$ est optimale pour $R$
-\item Mentionner la nature mal posée du problème sans s'appesentir dessus. Préciser toutefois qu'aucune approche ne le résoud de manière mathématiquement satisfaisante à cause de ça.
-\item insister sur l'optimalité de l'expert comme hypothèse. Un expert sous optimal, c'est un autre champ de littérature, éventuellement donner deux trois refs.
-\end{itemize}
-\end{itemize}
+\label{sec:background}
+%% \begin{itemize}
+%% \item Parler du domaine de la prise de décision séquentielle
+%% \item Expliquer le problème direct, en profiter pour introduire l'équation d'évaluation de Bellman et l'équation d'optimalité de Bellman
+%% \begin{itemize}
+%% \item $RL(MDP=\{S,A,\gamma,(P),R\}) = \pi^*:S\times A \rightarrow \mathbb{R}$
+%% \item $\forall \pi, Q^\pi(s,a) = R(s,a) + \sum_{s'}P(s'|s,a)Q^\pi(s',\pi(s'))$
+%% \item $\pi^* = \arg\max_aQ^*(s,a)$
+%% \end{itemize}
+%% \item Introduire le problème inverse dans une forme mathématique relativement propre
+%% \begin{itemize}
+%% \item $IRL( MDP\backslash R = \{S,A,\gamma,(P)\} \cup \{\pi^E\} = R : S\times A \rightarrow \mathbb{R}$ telle que $\pi^E$ est optimale pour $R$
+%% \item Mentionner la nature mal posée du problème sans s'appesentir dessus. Préciser toutefois qu'aucune approche ne le résoud de manière mathématiquement satisfaisante à cause de ça.
+%% \item insister sur l'optimalité de l'expert comme hypothèse. Un expert sous optimal, c'est un autre champ de littérature, éventuellement donner deux trois refs.
+%% \end{itemize}
+%% \end{itemize}
 We frame the sequential decision problem in a \emph{Markov Decision Process} (MDP), that is a tuple $\{S,A,P,R,\gamma\}$.
@@
 \section{The cascading algorithm}
-\label{sec-4}
-\begin{itemize}
-\item Mentionner (sans en parler, ce sera fait dans \hyperref[sec-7-1]{Imitation}) l'imitation.
+\label{sec:algo}
+%% \begin{itemize}
+%% \item Mentionner (sans en parler, ce sera fait dans \hyperref[sec-7-1]{Imitation}) l'imitation.
 Une approche supervisée évidente est la classification, avec pour entrée un set $D_C$ de données issues de l'expert.
 Rappeler (expliqué pour la première fois dans \hyperref[sec-3]{Formalisme} qu'on se restreint (sans perte de généralité, le préciser) au cas des politique déterministes.
-\item Tous les classifieurs utilisent une fonction de score (je n'ai pas trouvé un seul exemple d'un classifieur qui n'en utilise pas), il s'agit juste d'un manière d'écrire les choses.
-\item Expliquer que l'on identifie le mécanisme de classification au mécanisme glouton qui régit la politique optimale (donc celle de l'expert).
-\item Sans chercher à justifier mathématiquement la chose (ce sera fait dans la partie \hyperref[sec-5]{Analyse}), expliquer qu'intuitivement l'on va identifier fonction de score du classifieur et fonction de qualité de l'expert (qui est optimale pour une certaine récompense, rappelons-le).
-\begin{itemize}
-\item Repartir de l'équation d'évaluation de Bellman pour isoler le terme $R(s,a)$ et introduire $r:S\times A\in \mathbb{R}$, la récompense issue de cette équation lorsque l'on remplace $Q^E$ par la fonction de score : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi(s')) $
-\item On estime ce terme à partir des transitions $D_R$ car les probabilités de transitions sont inconnues : $\hat r(s,a) = q(s,a)-\gamma \sum_{s'} {1\over N_{s'}}\sum_{(s,a,s',a')\in D_R}q(s',a')$
-\item La fonction de récompense est donnée par une régression sur $D_R \cup {\hat r}$.
-\end{itemize}
-\item Donner l'algorithme l'environnement \texttt{algorithm2e}
-\item Rapidement, l'étape de régression permet d'ajouter la structure du MDP au résultat de classif, en utilisant l'équation de Bellman.
-\item Bien préciser qu'il ne s'agit ici que d'une explication qualitative, ``avec les mains'', la preuve du fait que ça résoud bien le problème est donnée dans \hyperref[sec-5]{Analyse}, une démonstration sur des problèmes si complexes que l'état de l'art ne peut les résoudre (sauf SCIRL) est donnée dans \hyperref[Experience]{Experience}.
+%% \item Tous les classifieurs utilisent une fonction de score (je n'ai pas trouvé un seul exemple d'un classifieur qui n'en utilise pas), il s'agit juste d'un manière d'écrire les choses.
+%% \item Expliquer que l'on identifie le mécanisme de classification au mécanisme glouton qui régit la politique optimale (donc celle de l'expert).
+%% \item Sans chercher à justifier mathématiquement la chose (ce sera fait dans la partie \hyperref[sec-5]{Analyse}), expliquer qu'intuitivement l'on va identifier fonction de score du classifieur et fonction de qualité de l'expert (qui est optimale pour une certaine récompense, rappelons-le).
+%% \begin{itemize}
+%% \item Repartir de l'équation d'évaluation de Bellman pour isoler le terme $R(s,a)$ et introduire $r:S\times A\in \mathbb{R}$, la récompense issue de cette équation lorsque l'on remplace $Q^E$ par la fonction de score : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi(s')) $
+%% \item On estime ce terme à partir des transitions $D_R$ car les probabilités de transitions sont inconnues : $\hat r(s,a) = q(s,a)-\gamma \sum_{s'} {1\over N_{s'}}\sum_{(s,a,s',a')\in D_R}q(s',a')$
+%% \item La fonction de récompense est donnée par une régression sur $D_R \cup {\hat r}$.
+%% \end{itemize}
+%% \item Donner l'algorithme l'environnement \texttt{algorithm2e}
+%% \item Rapidement, l'étape de régression permet d'ajouter la structure du MDP au résultat de classif, en utilisant l'équation de Bellman.
+%% \item Bien préciser qu'il ne s'agit ici que d'une explication qualitative, ``avec les mains'', la preuve du fait que ça résoud bien le problème est donnée dans \hyperref[sec-5]{Analyse}, une démonstration sur des problèmes si complexes que l'état de l'art ne peut les résoudre (sauf SCIRL) est donnée dans \hyperref[Experience]{Experience}.
 Les avantages comparés sont expliqués dans \hyperref[Autres-Approches]{Autres Approches}.
 Enfin, des explications plus intuitives sur les mécanismes internes de l'algorithme, ainsi que des résultats expérimentaux supplémentaires sont fournis en \hyperref[sec-9]{Annexes}.
-\end{itemize}
+%% \end{itemize}
 IRL has sometimes FXME:REFIMITATION been seen as a way to cast the imitation problem, that is a way to copy the expert policy ; the reward function is then a byproduct of the analysis and not the object one is looking for.
@@
 
-This is the first step of our algorithm.
+This classification operation is the first step of our algorithm, which consists in cascading two supervised approaches.
 Using a dataset $D_C = {s_i,a_i=\pi^E(s_i)}_i$ we train a score-function based classifier.
@@
 Our reward function $r$ must be approximated to $\hat r$ with the help of a dataset $D_R = \{s_i,a_i,s'_i\}$.
-This gives as many datapoints $\{\hat r(s_i,a_i) = q(s_i,a_i) - \gamma q(s',\pi_C(s'))}_i $.
+This gives as many datapoints $\{\hat r(s_i,a_i) = q(s_i,a_i) - \gamma q(s',\pi_C(s'))\}_i$ that we then feed to a regressor to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space.
+The complete algorithm is given Alg.
+\ref{algo:cascading}.
+\begin{algorithm}%[H]
+    %\small
+  \caption{Cascading IRL algorithm}
+  \label{algo:cascading}
+  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{i},a_{i},s'_{i})\}_{1\leq i \leq D'}$\;\\
+  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
+  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{i},a_{i},\hat{r}_i)\}_{1\leq i \leq D'}$, $\forall (s_i,a_i,s'_i) \in D_R,\hat{r}_i=q(s_{i},a_{i})-\gamma q(s'_{i},\pi_C(s'_{i}))$\;\\
+  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
+\end{algorithm}
 
-\section{Analyse}
-\label{sec-5}
+Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem.
+Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with what state), but there is more to a MDP than a single policy.
+Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probability $P$.
+The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.
+As we make direct use of the available data and do not iteratively compare some metric of the stationary distribution of policies, we do not need to solve the forward problem.
+This is a huge advantage, as it greatly reduce the computational and sampling cost of the method.
 
-\begin{itemize}
-\item On rappelle le problème posé au début.
-\item Le théorème 1 montre que modulo un bonne classif et un bon régresseur, on résoud exactement le problème que l'on s'était posé
-\item On en fait la démo rapide (éventuellement des bouts peuvent être déportés en \st{Pologne} \hyperref[sec-9]{Annexes}).
-\item On insiste sur le fait que la récompense trouvée n'est pas triviale, on a transcendé la nature mal posée du problème.
-\item On dit un mot rapide sur la complexité en échantillon (ça dépend du classifieur et du régresseur, mais \hyperref[sec-6]{Expériences} montre qu'on est bon là dessus)
-\item ainsi que sur la complexité temporelle (ça dépend toujours, mais avec une M-SVM et un moinde carré ça prend O(quelque chose) (indice: pas beaucoup)) FIXME:compléter la complexité
-\end{itemize}
+By carefully choosing or tuning both supervised algorithm, it should possible to mitigate ill effects of a lack of data.
+The classification error and regression error can be acted upon with the help of the wide range of techniques available for these two well-known supervised problems, which helps control the bound we provide in Sec.
+\ref{sec:analysis}.
+The explanation in this section only is a quatitative attempt to make the reader understand the idea behind the algorithm.
+The analysis tackle things in a formal, yet less intuitive, way.
+Dire lack of data is addressed in Sec \ref{sec:experiments} where the use of an heuristic is empirically shown to be relevant.
+Data constraints are so high that to the best of our knowledge only one other published algorithm is able to solve the IRL problem in such conditions.
+Other approaches are compared with one another in Sec.
+\ref{sec:related}.
+Finally, appendices provide the reader with other experiments that exemplify the inner working of the algorithm.
+
+\section{Analysis}
+\label{sec:analysis}
+
+%% \begin{itemize}
+%% \item On rappelle le problème posé au début.
+%% \item Le théorème 1 montre que modulo un bonne classif et un bon régresseur, on résoud exactement le problème que l'on s'était posé
+%% \item On en fait la démo rapide (éventuellement des bouts peuvent être déportés en \st{Pologne} \hyperref[sec-9]{Annexes}).
+%% \item On insiste sur le fait que la récompense trouvée n'est pas triviale, on a transcendé la nature mal posée du problème.
+%% \item On dit un mot rapide sur la complexité en échantillon (ça dépend du classifieur et du régresseur, mais \hyperref[sec-6]{Expériences} montre qu'on est bon là dessus)
+%% \item ainsi que sur la complexité temporelle (ça dépend toujours, mais avec une M-SVM et un moinde carré ça prend O(quelque chose) (indice: pas beaucoup)) FIXME:compléter la complexité
+%% \end{itemize}
 
@@
 Finally, if the classifier and the regressor are perfect ($\epsilon_C=\epsilon_R=0$), $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness suppose that $\forall s\in S, Card(\argmax_{a\in A}q(s,a))=1$ ).
+This uniqueness also hint at the fact that we found a non-trivial reward (we recall that the null reward admit \emph{every} policy as optimal).
 
-\section{Expériences}
-\label{sec-6}
+The sample complexity induced by this bound depends on the sample complexity of the classification and regression subroutines.
+The litterature is wide enough for methods accomodating most of use cases to be found.
+Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach.
 
-\begin{itemize}
-\item La borne fournie en \hyperref[sec-5]{Analyse} mesure le degré d'optimalité de la politique de l'expert vis à vis de la nouvelle récompense.
+\section{Related Work}
+\label{sec:related}
+IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
+A seminal paper, \cite{abbeel2004apprenticeship} blurs the frontier between imitation and IRL, as this algorithm outputs (mixing) policies, and not explicitely a reward (though it is possible to extract one).
+Other approaches with the same philosophy of finding a policy
+(through some reward function) such that its feature expectation (or
+more generally some measure of the underlying trajectories'
+distribution) matches the one of the expert policy are summarized in \cite{neu2009training}.
+This principle implies repeatedly solving the MDP in order to get information about the feature expectation of the policy induced by the iteratively built reward.
+This has a high cost, sample-wise as well as computationally.
+The Projection IRL (PIRL) algorithm of \cite{abbeel2004apprenticeship} is used as a benchmark in our experiments Sec.\ref{sec:experiments}.
+
+Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs (which had to be solved).
+
+Using the non trivial notion of metric in an MDP, \cite{melo2010learning} build a kernel which is used
+in a classification algorithm, showing improvements compared to a
+non-structured kernel. However, this approach is not an IRL
+algorithm, and more important assessing the metric of an MDP is a
+quite involved problem.
+
+Few algorithms do not require solving the direct RL problem. \cite{Dvij:2010} only applies to linearly-solvable MDPs.
+\cite{boularias:2011} uses a subgradient ascent of a utility function based on the notion of relative entropy.
+Importance sampling is suggested as a way to avoid solving the MDP.
+This requires sampling trajectories according to a non-expert policy and
+the direct problem remains at the core of the approach (even if
+solving it is avoided).
+
+The SCIRL algorithm \cite{klein2012scirl} is able to sole the IRL problem in the same dire conditions as the cascading approach.
+However, it outputs a reward on the state space only, which can prove problematic if the expert relies on information about the cost of an action in a certain state.
+We avoid that by using a reward over the state-action space.
+
+%% \begin{itemize}
+%% \item On reprend peu ou prou la même que SCIRL et que celle du précédent papier. On profite de la mention d'une approche pour préciser en quoi la nôtre diffère.
+%% \end{itemize}
+%% \subsection{Imitation}
+%% \label{sec-7-1}
+
+%% \begin{itemize}
+%% \item Ce n'est pas le problème que l'on se pose, mais résoudre l'IRL peut permettre une forme fine d'imitation.
+%% \item La frontière est un peu floue, exemple de PIRL
+%% \item On peut considérer que tous les algos revus par Neu sont des algos d'imitation dans le sens où ils cherchent une proximité des $\mu$.
+%% \item Coincidentallement, ils ont besoin de résoudre le problème direct, pas nous.
+%% \end{itemize}
+%% \subsection{Méthodes entropiques}
+%% \label{sec-7-2}
+
+%% \begin{itemize}
+%% \item N'ont pas besoin de résoudre le problème direct, mais ont besoin d'autres données que celles de l'expert.
+%% \end{itemize}
+%% \subsection{Autres méthodes}
+%% \label{sec-7-3}
+
+%% \begin{itemize}
+%% \item parler des MDPs solvables linéairement, des papiers mentionnés par les reviewers de NIPS, et dire du bien de SCIRL
+%% \end{itemize}
+\section{Experiments}
+\label{sec:experiments}
+\subsection{Criterion}
+The theoretical bound provided in Sec.
+\ref{sec:analysis} is about the optimality of the expert with respect to the reward function output by our algorithm (this is also the way the problem was framed in \ref{sec:background}).
+For the experiments, we use a somewhat more restrictive criterion : a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the "true" reward function, the reward function of the expert.
+We compare the mean value function with respect to $R^E$for a RL agent perfectly trained to optimize $R^C$.
+We do that firstly because, for once, we can.
+The bound we provide in the analysis is something that can be measured even if $R^E$ does not actually exist.
+We make use here of the fact that $R^E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R^E$.
+The second reason why we choose this different metric in this section is because we believe the theoretical bound to be convinving, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well behaved given the use of heuristics.
+\subsection{Setting}
+The setting of the experiment is a driving simulator inspired from a benchmark seen in \cite{syed2008apprenticeship,syed2008game}.
+The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
+The expert is trained to go as fast as possible (high reward) while avoiding collisions (harshly penalised) with randomly placed slower moving vehicles and avoiding going off-road (moderately penalised).
+Any other situation receives a null reward.
+FIXME:PLUSBAVARDSURLEHIGHWAY
+
+PIRL.
+\subsection{Instanciation}
+Classif, regress, heuristique.
+\subsection{Results}
+\begin{figure}
+  \includegraphics[width=\linewidth]{"Fig3"}
+  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.
+The lower baseline is an agent trained on a reward generated by using the same features as the cascading approach and the classifier, but with a random vector of parameters.}
+  \label{fig:Highway cascading}
+\end{figure}
+
+Results are shown Fig.\ref{fig:Highway cascading}.
+It is very interesting to compare our algorithm to the behavior of a classifier alone.
+With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrate far better performance, from the start.
+This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor truly make use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
+Furthermore, the classifier whose results are diplayed here is the output of the first step of the algorithm.
+Although what we show on the plot is not exactly the classification error, it is closely related (if $\pi^C=\pi^E$ then the classification would have the same performance on our plot as the expert, which is not the case).
+This shows that our algorithm is empirically more forgiving of classification errors than our theoretical bound let us expect.
+
+PIRL is given the transition probabilities (that is, a $|S|\times |S|\times |A|$ matrix) and yet it does only slightly better than our approach with 100 samples.
+This also show that we are indeed very sample efficient, without having to repeatedly solve the MDP.
+
+%% \begin{itemize}
+%% \item La borne fournie en \hyperref[sec-5]{Analyse} mesure le degré d'optimalité de la politique de l'expert vis à vis de la nouvelle récompense.
 Les expériences mesurent le degré d'optimalité de l'agent (ici entendu sous forme d'agent entraîné parfaitement sur la récompense sortie par notre algorithme) vis à vis de la vraie récompense (celle de l'expert).
-\begin{itemize}
-\item On se permet de faire ça car dans notre dispositif on la connait. Ce n'est pas le cas dans la vraie vie.
-\item En pratique c'est un critère plus restrictif : formellement, si $\pi^E$ est optimale pour $R_C$, alors l'agent et l'expert ont les mêmes performances (wrt $R_C$), puisque l'agent est optimal pour $R_C$ par définition de l'agent.
+%% \begin{itemize}
+%% \item On se permet de faire ça car dans notre dispositif on la connait. Ce n'est pas le cas dans la vraie vie.
+%% \item En pratique c'est un critère plus restrictif : formellement, si $\pi^E$ est optimale pour $R_C$, alors l'agent et l'expert ont les mêmes performances (wrt $R_C$), puisque l'agent est optimal pour $R_C$ par définition de l'agent.
 En revanche on regarde si l'agent est optimal pour $R_E$, ce qui est profondément ``injuste'', puisqu'on ne sait rien de $R_E$.
 On ne sait même pas si elle existe réellement.
-\item Il s'agit d'une métrique qu'il n'est pas possible de mesurer sur un problème réel.
+%% \item Il s'agit d'une métrique qu'il n'est pas possible de mesurer sur un problème réel.
 C'est pourquoi la démonstration est intéressente car elle parle de quelque chose que l'on peut plus facilement évaluer ; et surtout elle parle de ce que l'on s'est proposé de résoudre à la base.
 Ces expériences ne sont qu'une illustration.
-\end{itemize}
-\item Les résultats obtenus sur ce problème sont notamment comparés à la classif, il s'agit d'une illustration du fait que l'algorithme est capable d'extraire les infos sur $P$ contenues dans les transitions, grâce à l'équation de Bellman qui est présente (modulo deux trois manips) dans l'expression de $\hat r$.
+%% \end{itemize}
+%% \item Les résultats obtenus sur ce problème sont notamment comparés à la classif, il s'agit d'une illustration du fait que l'algorithme est capable d'extraire les infos sur $P$ contenues dans les transitions, grâce à l'équation de Bellman qui est présente (modulo deux trois manips) dans l'expression de $\hat r$.
 Nous avons injecté la structure du MDP dans la classification (ça me rappelle quelque chose\ldots{} ;)
-\item Le seul plot est celui du Highway (similaire à SCIRL). On enlève le Gridworld (reporté en annexe car trop d'infos pour si peu de place)
-\item On instancie avec Taskar et un moindre carré. Ceci n'a pas la moindre importance.
-\item Mentionner le fait que les autres algos ne peuvent résoudre le problème tel qu'on le pose. On s'apesentira là dessus dans \hyperref[sec-7]{Autres approches}.
-\item Expliquer que quitte à tester expérimentallement, on a choisi de s'éloigner du cadre de l'analyse pour se placer dans cadre \emph{plus difficile à gérer}, ou seules les données de l'expert sont disponibles.
-\begin{itemize}
-\item Cela implique l'usage d'une heuristique pour la régression :
-\begin{itemize}
-\item sans heuristique, c'est le choix par défauit du régresseur (i.e. initialisation à 0, à R$_{\mathrm{min}}$ ou R$_{\mathrm{max}}$) qui décidera de l'optimisme ou pessimisme de l'algo
-\item Pour éviter ça, on place nous même notre heuristique.
-\end{itemize}
-\item que l'on se contente d'expliquer à un niveau intuitif.
+%% \item Le seul plot est celui du Highway (similaire à SCIRL). On enlève le Gridworld (reporté en annexe car trop d'infos pour si peu de place)
+%% \item On instancie avec Taskar et un moindre carré. Ceci n'a pas la moindre importance.
+%% \item Mentionner le fait que les autres algos ne peuvent résoudre le problème tel qu'on le pose. On s'apesentira là dessus dans \hyperref[sec-7]{Autres approches}.
+%% \item Expliquer que quitte à tester expérimentallement, on a choisi de s'éloigner du cadre de l'analyse pour se placer dans cadre \emph{plus difficile à gérer}, ou seules les données de l'expert sont disponibles.
+%% \begin{itemize}
+%% \item Cela implique l'usage d'une heuristique pour la régression :
+%% \begin{itemize}
+%% \item sans heuristique, c'est le choix par défauit du régresseur (i.e. initialisation à 0, à R$_{\mathrm{min}}$ ou R$_{\mathrm{max}}$) qui décidera de l'optimisme ou pessimisme de l'algo
+%% \item Pour éviter ça, on place nous même notre heuristique.
+%% \end{itemize}
+%% \item que l'on se contente d'expliquer à un niveau intuitif.
 L'analyse formelle n'a pas été effectuée, ne le sera probablement pas, on se contente de montrer qu'en pratique, ça fonctionne.
 Et ça fonctionne mieux que les autres algos, sauf SCIRL.
-\item Cela implique aussi d'utiliser $\pi_C$ plutôt que $a'$.
-\end{itemize}
-\item Lien avec l'analyse : on a de bons résultats, même si le classifieur n'est pas bon. Notre borne est donc très très large.
-\item La complexité en échantillons : avec 50 samples, on fait mieux que PIRL qui a accès à une matrice de taille $|S|\times|S|\times |A|$
-\item La complexité en temps : on tourne 100 fois plus rapidement que PIRL
-\end{itemize}
-\section{Autres approches}
-\label{sec-7}
-
-\begin{itemize}
-\item On reprend peu ou prou la même que SCIRL et que celle du précédent papier. On profite de la mention d'une approche pour préciser en quoi la nôtre diffère.
-\end{itemize}
-\subsection{Imitation}
-\label{sec-7-1}
-
-\begin{itemize}
-\item Ce n'est pas le problème que l'on se pose, mais résoudre l'IRL peut permettre une forme fine d'imitation.
-\item La frontière est un peu floue, exemple de PIRL
-\item On peut considérer que tous les algos revus par Neu sont des algos d'imitation dans le sens où ils cherchent une proximité des $\mu$.
-\item Coincidentallement, ils ont besoin de résoudre le problème direct, pas nous.
-\end{itemize}
-\subsection{Méthodes entropiques}
-\label{sec-7-2}
-
-\begin{itemize}
-\item N'ont pas besoin de résoudre le problème direct, mais ont besoin d'autres données que celles de l'expert.
-\end{itemize}
-\subsection{Autres méthodes}
-\label{sec-7-3}
-
-\begin{itemize}
-\item parler des MDPs solvables linéairement, des papiers mentionnés par les reviewers de NIPS, et dire du bien de SCIRL
-\end{itemize}
+%% \item Cela implique aussi d'utiliser $\pi_C$ plutôt que $a'$.
+%% \end{itemize}
+%% \item Lien avec l'analyse : on a de bons résultats, même si le classifieur n'est pas bon. Notre borne est donc très très large.
+%% \item La complexité en échantillons : avec 50 samples, on fait mieux que PIRL qui a accès à une matrice de taille $|S|\times|S|\times |A|$
+%% \item La complexité en temps : on tourne 100 fois plus rapidement que PIRL
+%% \end{itemize}
 \section{Conclusion}
@@
 
+\bibliographystyle{plain}
+\bibliography{Biblio}
+
 \end{document}

### 2012-10-16 piot2012cascading/Highway_plot.py

@@ -1,0 +1,55 @@
+from matplotlib import rc
+rc('text', usetex=True)
+import sys
+sys.path+=['..']
+from numpy import *
+import scipy
+import pylab as pylab
+from Plot import *
+
+D_Cascading = genfromtxt( "Cascading_Exp5.mat" )
+D_C = D_Cascading[:,[0,1]]
+D_Classif = D_Cascading[:,[0,2]]
+
+Expert = 7.74390968 #python Expert.py to get this value
+Random_mean = -1.5821833963484#See Highway.org about Random.py for information on these values
+Random_min = -4.0007295890199996#python Random.py to get this value
+Random_max = 2.7064859345599999#python Random.py to get this value
+Random_var = 1.4465398450419833#python Random.py to get this value
+ANIRL_mean = 7.2183449001
+
+y_min = -5
+y_max = 10
+
+#Figure 1 : ANIRL, Random, classif and us
+pylab.figure(1)
+pylab.clf()
+[X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
+[X,Y_mean,Y_var] = map( array, mean_var( D_Classif ))
+x_min = min(X)
+x_max = max(X)
+pylab.axis([x_min,x_max,y_min,y_max])
+pylab.xlabel('Number of samples from the expert')
+pylab.ylabel('$\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{\mathcal{R}_E}(s)]$')
+pylab.grid(True)
+
+filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), Random_min*ones(X.shape), Random_max*ones(X.shape), 'cyan',0.2,'--',"Agent trained on a random reward",None)
+filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), (Random_mean-Random_var)*ones(X.shape), (Random_mean+Random_var)*ones(X.shape), 'cyan',0.4,'-.',None,None)
+
+pylab.plot(X,Expert*ones(X.shape), color='red',label="Expert",lw=2,ls='-')
+
+[X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
+[X,Y_mean,Y_var] = map( array, mean_var( D_Classif ))
[…]

### 2012-10-16 piot2012cascading/ECML13_cascading.tex
Modifs Bilal (abstract et notations) et style ECML (juste les fichiers)

@@
 \vspace*{1cm}
-\section{Abstract}
-\label{sec-1}
+\begin{abstract}
+  This paper considers the Inverse Reinforcement Learning (IRL) problem which consists in inferring a reward function for which a demonstrated expert policy is optimal.
+We propose to break the IRL problem down into two Supervised Learning (SL) steps: this is the cascaded approach. The first step consists in a classification and the second step in a regression.
+This cascaded approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the reward function outputted by the regression step.
+Besides, we provide an algorithm which is an instantiation of the cascaded approach.
+This algorithm does not need to solve repeatedly the Reinforcement Learning (RL) problem and use only sampled transitions as an input.
+Finally, up to the use some heuristics, the algorithm may work with only transitions sampled according to the demonstrated expert policy.
+This is illustrated on a car driving simulator, on a Gridworld and on an inverted pendulum.
+  \end{abstract}
 \section{Introduction}
@@
 %% \end{itemize}
-We frame the sequential decision problem in a \emph{Markov Decision Process} (MDP), that is a tuple $\{S,A,P,R,\gamma\}$.
-At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ it is in to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but this dos not induce any loss of generality.}) \emph{policy} $\pi: S \rightarrow A$.
+First, we introduce some general notations.
+Let $E$ and $F$ be two non-empty sets, $E^F$ is the set of functions from $F$ to $E$.
+Let $X$ be a non-empty finite set. We note $\Delta_X$ the set of distributions over $X$ and $Card(X)$ its cardinal. 
+Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\leq\beta$ is a componentwise inequality.
+
+We frame the sequential decision problem in a finite \emph{Markov Decision Process} (MDP), that is a tuple $\{S,A,P,R,\gamma\}$.
+$S=\{s_i\}_{1\leq i\leq Card(S)}$ is a finite state space, $A=\{a_i\}_{1\leq i\leq Card(K)}$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the dynamic of the MDP.
+Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)=p(s,a,s')$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$.
+At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but this dos not induce any loss of generality.}) \emph{policy} $\pi\in A^S$.
 The agent then transits to a new state $s_{t+1}\in S$ according to the markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
-The reward function $R : S\times A \rightarrow \mathbb{R}$ is a local measure of the quality of the control.
-The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{|S|}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state : $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t)|s_0 = s]$.
+The reward function $R$ is a local measure of the quality of the control.
+The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t)|s_0 = s]$.
 Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, \forall s \in S, V^*_R(s) \geq V^\pi_R(s)$.
@@
 
-The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{|S|} \rightarrow  \mathbb{R}^{|S|}$ is defined by $T^\pi_RV = R + \gamma P^piV$ using the straightforward matricial notation $P\pi = (p(s'|s,\pi(s)))_{s,s' \in S}$.
+The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by $T^{\pi_R}V = R + \gamma P_\pi V$ using the straightforward matricial notation $P_\pi = (p(s'|s,\pi(s)))_{s,s' \in S}$.
 The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$.
-Both these operator are contractions.
-The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R $V^\pi_R$ ; the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$.
-Optimal value function and optimal policy are also linked via the quality function $Q^\pi_R$, a state-action value function that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_Rv^\pi_R(s)$, with $a$ the policy that always returns action $a$ : the optimal policy follows a greedy mechanism with respect to its optimal quality function $\pi^*_R(s) = \argmax_aQ^*_R(s,a)$.
+Both operators are contractions.
+The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$; the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$.
+Optimal value function and optimal policy are also linked via the quality function $Q^\pi_R\in\mathbb{R}^{S\times A}$, a state-action value function that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^aV = R + \gamma P_a V$ with $P_a = (p(s'|s,a))_{s,s' \in S}$).
+The optimal policy follows a greedy mechanism with respect to its optimal quality function $\pi^*_R(s)\subset\argmax_aQ^*_R(s,a)$.
 
-When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, \emph{Reinforcement Learning} (RL) or approximate dynamic programming may be used to retireve the optimal control policy.
+When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, \emph{Reinforcement Learning} (RL) or approximate dynamic programming may be used to retrieve the optimal control policy.
 We call this the direct problem.

### 2012-10-18 piot2012cascading/ECML13_cascading.tex
WIP sur le papier ECML

@@
 % Created 2012-10-15 lun. 14:49
-\documentclass[11pt]{article}
+\documentclass[smallextended]{svjour3}
+\smartqed 
 \usepackage[utf8]{inputenc}
@@
 \usepackage{amsmath}
+%\usepackage{amsthm}
+%\newtheorem{definition}{Definition}
+%\newtheorem{theorem}{Theorem}
 \usepackage{dsfont}
@@
 \newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
-\title{Outline}
-\author{Edouard Klein}
+\begin{document}
+\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
+\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
+\institute{
+ 1. ABC Team\\
+ LORIA-CNRS, France.\\
+\and
+2.Supélec-Metz Campus\\
+ MaLIS Research group, France\\
+ \texttt{firstname.lastname@supelec.fr}\\
+\and 
+3. UMI 2958 CNRS\\
+GeorgiaTech, France\\
+}
 \date{\today}
-\hypersetup{
-  pdfkeywords={},
-  pdfsubject={},
-  pdfcreator={Emacs Org-mode version 7.8.11}}
-
-\begin{document}
 
@@
 
-\setcounter{tocdepth}{3}
-\tableofcontents
-\vspace*{1cm}
 \begin{abstract}
-  This paper considers the Inverse Reinforcement Learning (IRL) problem which consists in inferring a reward function for which a demonstrated expert policy is optimal.
-We propose to break the IRL problem down into two Supervised Learning (SL) steps: this is the cascaded approach. The first step consists in a classification and the second step in a regression.
-This cascaded approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the reward function outputted by the regression step.
-Besides, we provide an algorithm which is an instantiation of the cascaded approach.
-This algorithm does not need to solve repeatedly the Reinforcement Learning (RL) problem and use only sampled transitions as an input.
-Finally, up to the use some heuristics, the algorithm may work with only transitions sampled according to the demonstrated expert policy.
-This is illustrated on a car driving simulator, on a Gridworld and on an inverted pendulum.
+  This paper considers the Inverse Reinforcement Learning (IRL) problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
+We propose to break the IRL problem down into two Supervised Learning (SL) steps: this is the cascaded approach.
+A classification step aiming to imitate the expert's policy gives a score function we see as an optimal action value function.
+In the second step, this score function is used together with the bellman evaluation equation to create samples then fed to a regressor that outputs a reward function.
+This cascaded approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the reward function the regression step outputs.
+We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning (RL) direct problem.
+Finally, up to the use some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy.
   \end{abstract}
@@
 
-  De la même manière que pour le papier de SCIRL, l'intro est une vaste annonce du papier, avec des références pointues (sous-section, voire sous,sous-section) vers les parties du papier qui expliquent tel ou tel point en détail.
+Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} aims at inferring a reward function for which a demonstrated expert policy is optimal.
+An important part of the literature (see Section~\ref{sec:related} for a brief review) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution) which matches the one induced by the expert: this is called apprenticeship learning.
+Most algorithms of apprenticeship learning use Reinforcement
+Learning (RL) algorithms as subroutines, thus their performance depends strongly on the quality of these subroutines. Consequently, apprenticeship learning algorithms suffer from the
+same challenges of large state spaces, exploration vs. exploitation trade-offs as RL. In order to avoid to solve repeatedly RL problems, we decide to take a different point of view.
+
+Having in mind that there is a one to one relation between a reward function and its associated optimal quality function (via the inverse Bellman equation), it is worth thinking of a method which is able to output a quality function for which the greedy policy is the demonstrated expert policy.
+Thus, the demonstrated expert policy will be optimal for the reward function associated to the quality function.
+As a method which is able to output a quality function for which the greedy policy is the demonstrated expert policy, we propose a score-based classification (see Section~\ref{sec:algo}).
+Besides, in order to retrieve, via the inverse Bellman equation, the reward associated to the score function the classification outputs, we propose a regression (see Section~\ref{sec:algo}).
+That is why the method is called the Cascaded Supervised Approach (CSA).
+This method is justified by a theoretical analysis (see Section~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
+
+Finally, an instantiation of CSA is proposed~\ref{sec:experiments}.
+This algorithm does not need to solve iteratively an RL problem and needs only sampled transitions from expert and non-experts policies as inputs.
+Moreover, up to the use of some heuristics, the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
+The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm, to a pure classification algorithm and to a recent successfull IRL method.
+
 \section{Background and notations}
@@
 Let $E$ and $F$ be two non-empty sets, $E^F$ is the set of functions from $F$ to $E$.
-Let $X$ be a non-empty finite set. We note $\Delta_X$ the set of distributions over $X$ and $Card(X)$ its cardinal. 
-Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\leq\beta$ is a componentwise inequality.
+We note $\Delta_X$ the set of distributions over $X$.
+Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\leq\beta \Leftrightarrow \forall x\in X, \alpha(x) \leq \beta(x)$.
+We will often slightly abuse the notations and consider (where applicable) most objects as if they were matrices and vectors indexed by the set they operate upon.
 
 We frame the sequential decision problem in a finite \emph{Markov Decision Process} (MDP), that is a tuple $\{S,A,P,R,\gamma\}$.
-$S=\{s_i\}_{1\leq i\leq Card(S)}$ is a finite state space, $A=\{a_i\}_{1\leq i\leq Card(K)}$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the dynamic of the MDP.
-Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)=p(s,a,s')$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$.
-At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but this dos not induce any loss of generality.}) \emph{policy} $\pi\in A^S$.
+$S$ is a finite state space, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the dynamic of the MDP.
+Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$.
+At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one deterministic optimal policy.}) \emph{policy} $\pi\in A^S$.
 The agent then transits to a new state $s_{t+1}\in S$ according to the markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
@@
 
-The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by $T^{\pi_R}V = R + \gamma P_\pi V$ using the straightforward matricial notation $P_\pi = (p(s'|s,\pi(s)))_{s,s' \in S}$.
+The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by $T^{\pi_R}V = R + \gamma P_\pi V$ where $P_\pi$ is $(p(s'|s,\pi(s)))_{s,s' \in S}$.
 The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$.
@@
 Optimal value function and optimal policy are also linked via the quality function $Q^\pi_R\in\mathbb{R}^{S\times A}$, a state-action value function that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^aV = R + \gamma P_a V$ with $P_a = (p(s'|s,a))_{s,s' \in S}$).
-The optimal policy follows a greedy mechanism with respect to its optimal quality function $\pi^*_R(s)\subset\argmax_aQ^*_R(s,a)$.
+The optimal policy follows a greedy mechanism with respect to its optimal quality function
+\begin{equation}
+  \label{eq:greedy}
+  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
+\end{equation}
 
-When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, \emph{Reinforcement Learning} (RL) or approximate dynamic programming may be used to retrieve the optimal control policy.
+When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to retrieve the optimal control policy \cite{sutton1998reinforcement}.
 We call this the direct problem.
@@
 We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal.
-The non optimality of the expert is an interesting setting that has been discussed for example FIXME:NONOPTIMALITY but that we are not addressing here.
-We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is also optimal.
+The non optimality of the expert is an interesting setting that has been discussed for example in \cite{meloXXXanalysis} but that we are not addressing here.
+We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is also near-optimal.
 The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal).
@@
 %% \end{itemize}
-IRL has sometimes FXME:REFIMITATION been seen as a way to cast the imitation problem, that is a way to copy the expert policy ; the reward function is then a byproduct of the analysis and not the object one is looking for.
-An efficient supervised way to do pure imitation is to use a classifier.
-The entries are the states $s_t$ and the labels are the actions $a_t$ deterministically chosen by the expert.
-We thus obtain a classification policy $\pi_C$.
+Our first step in the quest for a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (MC$^2$).
+This classifier learn a score function $q:S\times A \rightarrow \mathbb{R}$ that rates the association of a certain label $a\in A$ with a certain input $s\in S$.
+The classification rule $\pi_C: S \rightarrow A$ is simply to select (one of) the label(s) that achieve the highest score for the given input :
+\begin{equation}
+  \label{eq:greedy2}
+\pi_C(s) \in \argmax_a q(s,a).
+\end{equation}
+Most notably, \emph{Multi-class Support Vector Machines} can be seen as MC$^2$ algorithms.
+We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in subsection \ref{sec:instantiation}.
 
-This classification operation is the first step of our algorithm, which consists in cascading two supervised approaches.
-Using a dataset $D_C = {s_i,a_i=\pi^E(s_i)}_i$ we train a score-function based classifier.
-We can make use of any classifier whose decision rule uses a greedy mechanism with respect to a score function $q : S \times A \rightarrow \mathbb{R}$ that quantifies how well label $a\in A$ suits entry $s\in S$.
-Most classifiers, from M-SVM to $k$-nearest neighboors can be cast as a score-function based classifier.
-This allows the use of many off-the-shelf imlpementations.
-
-The greedy classification mechanism $\pi_C = \argmax_a q(s,a)$ is very similar to the relation between the optimal policy in a MDP and its optimal quality function : $\pi^*_R(s) = \argmax_aQ^*_R(s,a)$.
-As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal quality function.
-It is not a stretch to see $\pi_C$ as an optimal policy as it has been built from $\pi^E$ which is optimal.
+Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier.
+Apart from discussion and comparison purpose, we do not make use of the classification policy $\pi_C$.
+What is of interest to us is the score function $q$ itself.
+One can easily notice the similarity between equations \ref{eq:greedy2} and \ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal quality function.
+As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal quality function for the classifier policy $\pi_C$.
 By inversing the Bellman evaluation equation, one can express the reward as a function of the quality function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$.
@@
 
-We wish to solve the general, or approximate IRL problem, where the transition probabilities are unknown.
-Our reward function $r$ must be approximated to $\hat r$ with the help of a dataset $D_R = \{s_i,a_i,s'_i\}$.
-This gives as many datapoints $\{\hat r(s_i,a_i) = q(s_i,a_i) - \gamma q(s',\pi_C(s'))\}_i$ that we then feed to a regressor to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space.
+We wish to solve the general, or approximate, IRL problem where the transition probabilities are unknown.
+For this we use another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to.
+Action $a_j$ need not be chosen by any particular policy.
+The dataset $D_R$ brings us information about the dynamics of the system.
+Our reward function $r$ must be approximated to $\hat r$ with the help of the information in $D_R$.
+This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor FIXME:citation 36/15 papier de munos to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space.
 The complete algorithm is given Alg.
@@
   \label{algo:cascading}
-  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{i},a_{i},s'_{i})\}_{1\leq i \leq D'}$\;\\
+  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
   \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
-  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{i},a_{i},\hat{r}_i)\}_{1\leq i \leq D'}$, $\forall (s_i,a_i,s'_i) \in D_R,\hat{r}_i=q(s_{i},a_{i})-\gamma q(s'_{i},\pi_C(s'_{i}))$\;\\
+  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{j},a_{j},\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
   \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
@@
 The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.
-As we make direct use of the available data and do not iteratively compare some metric of the stationary distribution of policies, we do not need to solve the forward problem.
-This is a huge advantage, as it greatly reduce the computational and sampling cost of the method.
 
-By carefully choosing or tuning both supervised algorithm, it should possible to mitigate ill effects of a lack of data.
-The classification error and regression error can be acted upon with the help of the wide range of techniques available for these two well-known supervised problems, which helps control the bound we provide in Sec.
-\ref{sec:analysis}.
+As the analysis (Sec \ref{sec:analysis}) reveals, both classification error and regression error should be controlled.
+By carefully choosing or tuning both supervised algorithm, it should possible to mitigate any ill effects of the specific settings the cascading algorithm is used in.
+If one operates in a well-known domain where a lo of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings.
+If data is lacking, one could for example have to trade off between bias and variance.
+The possibility to use and leverage the wide range of techniques developped for these two well known supervised problems is a strong advantage of the proposed cascading approach.
+%Reformuler ce passage
 The explanation in this section only is a quatitative attempt to make the reader understand the idea behind the algorithm.
-The analysis tackle things in a formal, yet less intuitive, way.
-Dire lack of data is addressed in Sec \ref{sec:experiments} where the use of an heuristic is empirically shown to be relevant.
+The analysis of the next section tackle things in a formal, yet less intuitive, way.
+In Sec \ref{sec:experiments}, the use of heuristics (not tackled in the analysis) is empirically shown to be relevant.
 Data constraints are so high that to the best of our knowledge only one other published algorithm is able to solve the IRL problem in such conditions.
@@
 
-First let introduce some notations. $\pi_C$ is the classifier policy, $\epsilon_C$ is the classification error:
+In this Section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ outputted by the regression step.
+We formulate this result in a more formal way in Theorem \ref{theorem: theorem analysis}
+
+\begin{theorem}
+\label{theorem: theorem analysis}
+Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward outputted by the regression step. We have:
+\begin{equation}
+0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\leq \frac{1}{1-\gamma}(\Delta q \epsilon_C+(1+C_*)\epsilon_R)).
+\end{equation}
+$\Delta q$, and $C_*$ are constants and are defined in the proof.
+$\epsilon_C$ represents the classification error and $\epsilon_R$ represents the regression error: both terms are well-defined in the proof.
+\end{theorem}
+\begin{proof}
+First let introduce some notations.
+$q\in\mathbb{R}^{S\times A}$ is the score function outputted by the classification step, $\forall s \in S,\pi_C\in\argmax{a\in A}q(s,a)$ is a deterministic classifier policy and $\epsilon_C$ is the classification error:
 \begin{equation}
 \epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
+\end{equation}
+Moreover we recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
+\begin{equation}
+\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) +\gamma\sum_{s'\in S}P(s'|s,a)q(\pi_C(s'),a),
+\end{equation}
+and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function outputted by the regression step.
+The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$, $\epsilon^R\in\mathbb{R}^{S\times A}$ must not be mistaken for the regression error $\epsilon_R\in\mathbb{R}$ defined below.
+We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof:
+\begin{equation}
+\forall (s,a)\in S\times A, R^E(s,a)=q(s,a) +\gamma\sum_{s'\in S}P(s'|s,a)q(\pi_E(s'),a).
 \end{equation}
@@
 &R^C_a=q_a-\gamma P_aq_{\pi_C},
+\\
+&R^E_a=q_a+\gamma P_aq_{\pi_E},
 \\
@@
 \end{equation}
-Now, there is a still an upper bound to find for the second term. If we introduce, $R^E_a=q_a+\gamma P_aq_{\pi_E}$, it is possible to decompose the second term:
+Now, there is a still an upper bound to find for the second term. It is possible to decompose the second term:
 \begin{equation}
@@
 \begin{align}
-\rho_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(R^E_{\pi_E}-R^E_{\pi_C}),
+\rho_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(R^E_{\pi_E}-R^C_{\pi_E}),
 \\
@@
 \end{equation}
+\end{proof}
 
-Few remarks: the bound is obviously true for $\hat{R}_C$, but in this case the classifier or the regressor have bad performance which it is not likely.
-The algorithm and the analysis are invariant by a dilatation of $q$, so of $\Delta q$.
+Now, let us comment this results with some remarks.
+the bound is obviously true for $\hat{R}_C$, but in this case the classifier or the regressor have bad performance which is not likely.
+The algorithm and the analysis are invariant by  dilatation of $q$, hence of $\Delta q$.
 We could normalize $\Delta q$ by imposing $\Delta q=1$ (the range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$).
 The constant $C_*$ could be estimated a posteriori ( after $\hat{R}^C$ being calculated).
-We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ give a tighter final result (approximately, if $\hat{\pi_C}\approx\pi_E$ then $C_*=1$).
+We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (approximately, if $\hat{\pi_C}\approx\pi_E$ then $C_*=1$).
 
@@
 In theory, we need to be able to sample the whole dynamic (even the non expert actions).
-In practice, a heuristic is used (however this heuristic is not justified by the analysis).
+In practice, a heuristic is used (see Section~\ref{sec:experiments}) (however this heuristic is not justified by the analysis).
 Finally, if the classifier and the regressor are perfect ($\epsilon_C=\epsilon_R=0$), $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness suppose that $\forall s\in S, Card(\argmax_{a\in A}q(s,a))=1$ ).
@@
 \subsection{Setting}
-The setting of the experiment is a driving simulator inspired from a benchmark seen in \cite{syed2008apprenticeship,syed2008game}.
-The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
-The expert is trained to go as fast as possible (high reward) while avoiding collisions (harshly penalised) with randomly placed slower moving vehicles and avoiding going off-road (moderately penalised).
+The setting of the experiment is a driving simulator inspired from a benchmark seen in \cite{syed2008apprenticeship,syed2008game}.The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
+At all times, there will be one car in one of the three lanes.
+Even at the lowest speed, the player's car moves faster than the others.
+When the other car disappear at the bottom of the screen, another one appear at the top in a randomly chosen lane.
+It takes two transitions to completely change lanes, as the player can move left or right for half a lane's length at a time.
+At the highest speed setting, if the other car appear in the lane the player is in, it is not possible to avoid the collision.
+The main difference between the original benchmark and ours is that we made the MDP more ergodic by allowing the player to change speed whenever he wishes so, not just during the first transition.
+If anything, by adding two actions, we enlarged the state-action space and thus made the problem harder.
+FIXME:METTRE UNE CAPTURE DECRAN.
+The expert is trained to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
 Any other situation receives a null reward.
-FIXME:PLUSBAVARDSURLEHIGHWAY
 
-PIRL.
+FIXME:Comparer à SCIRL et à PIRL
+We need some baseline to compare our new approach to.
+The natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
+Another natural benchmark is to compute the mean value function of the classifier policy.
+Also there is no concept of reward here, it shows that even when talking only about imitation, IRL is a relevant method.
 \subsection{Instanciation}
-Classif, regress, heuristique.
+The cascading algorithm can be instanciated with any classification or regression algorithm.
+The choice of such subroutines may be dictated by the kind and amount of available data, by ease of use or by computational complexity, for example.
+
+We referred in Sec.\ref{sec:algo} to \emph{score-function based multi-class classifiers} and explained how the classification rule is similar to the greedy mechanism that exists between an optimal action-value function and an optimal policy in a MDP.
+Most classification algorithm can be seen as such a classifier.
+In a simple $k$-nearest neighboor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighboors of $s$.
+The generic M-SVM model makes the score function explicit \cite{guermeur2011generic} (equation 1).
+In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning}.
+FIXME:Recopier le QP
+
+Ease of implementation lead us to choose a simple least-square regressor for our experiments.
+Using the same features $\phi$ as our classifier, it can be written (using straightforward matricial notations) as :  $\theta = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R$.
+The reward function is : $\hat R^C(s,a) = \theta^T \phi(s,a)$.
+
+It is possible to get imaginative in the last step.
+For example, using a gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable the use of fuzzy reward reinforcement learning.
+FIXME:Trouver une ref pour les fuzzy rewards
+
 \subsection{Results}
@@
 The lower baseline is an agent trained on a reward generated by using the same features as the cascading approach and the classifier, but with a random vector of parameters.}
-  \label{fig:Highway cascading}
+  \label{fig:Highway}
 \end{figure}
 
-Results are shown Fig.\ref{fig:Highway cascading}.
+Results are shown Fig.\ref{fig:Highway}.
 It is very interesting to compare our algorithm to the behavior of a classifier alone.
@@
 \section{Conclusion}
-\label{sec-8}
-
-\begin{itemize}
-\item Ouverture : Existe-t-il des problèmes tels que SCIRL et Cascading diffèrent en performance ?
-\item Que faire quand on ne peut résoudre le RL facilement ?
-\item Que faire en cas d'expert sous optimal ?
-\item ouverture sur l'active learning en reprenant les notions de pessimisme ou d'optimisme en présence d'incertitude
-\end{itemize}
+\label{sec:conclusion}
+We have introduced a new way to perform IRL by cascading two well-known supervised approaches, namely classification and regression.
+The expert is shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.
+A few practical examples of classifiers and regressors have been given, and one combination has been empirically shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
+On the chosen benchmark (a car driving simulator), our algorithm is shown to outperform other state of the art approaches, and to be on-par with the very recent SCIRL algorithm.
+We plan on deepening the analysis of the theoretical properties of our approach and on applying it to real world robotics problems.
+%% \begin{itemize}
+%% \item Ouverture : Existe-t-il des problèmes tels que SCIRL et Cascading diffèrent en performance ?
+%% \item Que faire quand on ne peut résoudre le RL facilement ?
+%% \item Que faire en cas d'expert sous optimal ?
+%% \item ouverture sur l'active learning en reprenant les notions de pessimisme ou d'optimisme en présence d'incertitude
+%% \end{itemize}
 

### 2012-10-19 piot2012cascading/ECML13_cascading.tex
Presque fini

@@
 
-We frame the sequential decision problem in a finite \emph{Markov Decision Process} (MDP), that is a tuple $\{S,A,P,R,\gamma\}$.
+We frame the sequential decision problem in a finite \emph{Markov Decision Process} (MDP) \cite{puterman1994markov}, that is a tuple $\{S,A,P,R,\gamma\}$.
 $S$ is a finite state space, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the dynamic of the MDP.
@@
 We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal.
-The non optimality of the expert is an interesting setting that has been discussed for example in \cite{meloXXXanalysis} but that we are not addressing here.
+The non optimality of the expert is an interesting setting that has been discussed for example in \cite{melo2012analysis} but that we are not addressing here.
 We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is also near-optimal.
@@
 Most notably, \emph{Multi-class Support Vector Machines} can be seen as MC$^2$ algorithms.
-We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in subsection \ref{sec:instantiation}.
+We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in subsection \ref{subsec:instanciation}.
 
@@
 Our reward function $r$ must be approximated to $\hat r$ with the help of the information in $D_R$.
-This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor FIXME:citation 36/15 papier de munos to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space.
+This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor \cite{vapnik1996nature} to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space.
 The complete algorithm is given Alg.
@@
 The analysis of the next section tackle things in a formal, yet less intuitive, way.
-In Sec \ref{sec:experiments}, the use of heuristics (not tackled in the analysis) is empirically shown to be relevant.
-Data constraints are so high that to the best of our knowledge only one other published algorithm is able to solve the IRL problem in such conditions.
-Other approaches are compared with one another in Sec.
-\ref{sec:related}.
-Finally, appendices provide the reader with other experiments that exemplify the inner working of the algorithm.
 
@@
 
-In this Section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ outputted by the regression step.
+In this Section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
 We formulate this result in a more formal way in Theorem \ref{theorem: theorem analysis}
@@
 \label{theorem: theorem analysis}
-Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward outputted by the regression step. We have:
+Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward the cascading algorithm outputs. We have:
 \begin{equation}
@@
 
-Now, let us comment this results with some remarks.
-the bound is obviously true for $\hat{R}_C$, but in this case the classifier or the regressor have bad performance which is not likely.
+The bound is obviously true for $\hat{R}_C = 0$, but for this to happen the classifier or the regressor must have very bad performance, which is not likely.
+
 The algorithm and the analysis are invariant by  dilatation of $q$, hence of $\Delta q$.
 We could normalize $\Delta q$ by imposing $\Delta q=1$ (the range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$).
-The constant $C_*$ could be estimated a posteriori ( after $\hat{R}^C$ being calculated).
-We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (approximately, if $\hat{\pi_C}\approx\pi_E$ then $C_*=1$).
+
+The constant $C_*$ could be estimated a posteriori ( after $\hat{R}^C$ is computed).
+We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (approximately, if $\hat{\pi_C}\approx\pi_E$ then $C_* \approx 1$).
 
@@
 In theory, we need to be able to sample the whole dynamic (even the non expert actions).
-In practice, a heuristic is used (see Section~\ref{sec:experiments}) (however this heuristic is not justified by the analysis).
-Finally, if the classifier and the regressor are perfect ($\epsilon_C=\epsilon_R=0$), $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness suppose that $\forall s\in S, Card(\argmax_{a\in A}q(s,a))=1$ ).
+In practice heuristics can be used (see Section~\ref{sec:experiments}) however this heuristics is not justified by the analysis.
+
+If the classifier and the regressor are perfect ($\epsilon_C=\epsilon_R=0$), $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness suppose that $\forall s\in S, Card(\argmax_{a\in A}q(s,a))=1$, with $Card(X)$ denoting the cardinality of the set $X$).
 This uniqueness also hint at the fact that we found a non-trivial reward (we recall that the null reward admit \emph{every} policy as optimal).
 
-The sample complexity induced by this bound depends on the sample complexity of the classification and regression subroutines.
-The litterature is wide enough for methods accomodating most of use cases to be found.
+The bound can be controlled by teh quality of the classification and regression subroutines.
+The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off etc.) to be found.
 Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach.
+
+Other differences between existing IRL or apprenticeship learning approaches and the proposed cascading algorithm are further examined in the following section.
 
@@
 IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
-A seminal paper, \cite{abbeel2004apprenticeship} blurs the frontier between imitation and IRL, as this algorithm outputs (mixing) policies, and not explicitely a reward (though it is possible to extract one).
-Other approaches with the same philosophy of finding a policy
-(through some reward function) such that its feature expectation (or
+A seminal paper, \cite{abbeel2004apprenticeship} blurs the frontier between imitation and IRL, as this algorithm outputs (mixing) policies, and not explicitely a reward although IRL is the core of this method.
+This work introduced the general method of finding a reward function such that the corresponding optimal policy's feature expectation (or
 more generally some measure of the underlying trajectories'
-distribution) matches the one of the expert policy are summarized in \cite{neu2009training}.
-This principle implies repeatedly solving the MDP in order to get information about the feature expectation of the policy induced by the iteratively built reward.
-This has a high cost, sample-wise as well as computationally.
-The Projection IRL (PIRL) algorithm of \cite{abbeel2004apprenticeship} is used as a benchmark in our experiments Sec.\ref{sec:experiments}.
+distribution) matches the one of the expert policy.
+By iteratively changing the considered reward function, one can get the feature expectation of the expert and of the optimal policy to match within a certain bound.
+This implies using RL as a subroutine to solve the direct problem at each iteration.
+Some of those approaches are summed up in \cite{neu2009training}.
+The need to solve a MDP at each step is very demanding, both sample-wise and computationally.
+The cascading approach being able to output a reward function without having to solve the MDP is thus a significant improvement.
+The Projection IRL (PIRL) algorithm of \cite{abbeel2004apprenticeship} is used as a benchmark in our experiments, see subection \ref{subsec:setting}.
 
-Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs (which had to be solved).
+Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
 
@@
 
-Few algorithms do not require solving the direct RL problem. \cite{Dvij:2010} only applies to linearly-solvable MDPs.
+Few algorithms do not require solving the direct RL problem. \cite{Dvij:2010} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
+
 \cite{boularias:2011} uses a subgradient ascent of a utility function based on the notion of relative entropy.
@@
 the direct problem remains at the core of the approach (even if
-solving it is avoided).
+solving it is avoided). As our experiments show, we are able to tackle the IRL problem with only samples from the expert.
 
-The SCIRL algorithm \cite{klein2012scirl} is able to sole the IRL problem in the same dire conditions as the cascading approach.
-However, it outputs a reward on the state space only, which can prove problematic if the expert relies on information about the cost of an action in a certain state.
-We avoid that by using a reward over the state-action space.
+The LPAL algorithm of \cite{syed2008apprenticeship} also does not require solving the direct problem, but it requires knowing the transition probabilities $P$ of the MDP (which we don't need).
+
+Finally, the SCIRL algorithm \cite{klein2012scirl} is able to solve the IRL problem in the same dire conditions the cascading approach is tested in.
+However, it needs a good estimate of the expert's feature expectation in the states visited by the expert, but for all actions not just the expert's.
+The use of some heuristics can save the day, but it is not known wether such heuristics translate well on a more difficult problem.
+We also use it as a benchmark in our experiments, where it shows very slightly better results than the cascading algorithm.
 
@@
 \label{sec:experiments}
+In this section, we put our approach under stress to empirically demonstrate its behavior.
+After explaining the criterion we consider and its rationale (subsection \ref{subsec:criterion}), we describe the benchmark's setting along with the other algorithms we use for comparison purposes (subsection \ref{subsec:setting}), before giving and anlyzing the results (subsection \ref{subsec:results}), we explain the details of our choices for the classification and regression subroutines and give some hints for other possibilities (subsection \ref{subsec:instanciation}).
+
 \subsection{Criterion}
+\label{subsec:criterion}
 The theoretical bound provided in Sec.
@@
 For the experiments, we use a somewhat more restrictive criterion : a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the "true" reward function, the reward function of the expert.
-We compare the mean value function with respect to $R^E$for a RL agent perfectly trained to optimize $R^C$.
-We do that firstly because, for once, we can.
-The bound we provide in the analysis is something that can be measured even if $R^E$ does not actually exist.
+We compare the mean (w.r.t.
+the uniform distribution over the states) value function with respect to $R^E$ for a RL agent perfectly trained to optimize $R^C$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
+We do that firstly because, for once, we can : the bound we provide in the analysis is something that can be measured even if $R^E$ does not actually exist.
 We make use here of the fact that $R^E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R^E$.
-The second reason why we choose this different metric in this section is because we believe the theoretical bound to be convinving, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well behaved given the use of heuristics.
+The second reason why we choose this different metric in this section is because we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well behaved given the use of heuristics.
 \subsection{Setting}
+\label{subsec:setting}
 The setting of the experiment is a driving simulator inspired from a benchmark seen in \cite{syed2008apprenticeship,syed2008game}.The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
@@
 If anything, by adding two actions, we enlarged the state-action space and thus made the problem harder.
-FIXME:METTRE UNE CAPTURE DECRAN.
+%FIXME:METTRE UNE CAPTURE DECRAN.
 The expert is trained to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
@@
 
-FIXME:Comparer à SCIRL et à PIRL
 We need some baseline to compare our new approach to.
 The natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
-Another natural benchmark is to compute the mean value function of the classifier policy.
-Also there is no concept of reward here, it shows that even when talking only about imitation, IRL is a relevant method.
+Another natural benchmark we use is to compute the mean value function of the classifier policy.
+We also compare the proposed cascading method to the PIRL algorithm of \cite{abbeel2004apprenticeship} and the SCIRL algorithm of \cite{klein2012scirl}.
+It is worth noting that the random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
+PIRL is instanciated in a way that make use of the transition probability matrix $P$ to exactly compute all relevant objects (feature expectations, optimal policies) via exact dynamic programming.
+It is left to run for 70 iterations.
 \subsection{Instanciation}
-The cascading algorithm can be instanciated with any classification or regression algorithm.
+\label{subsec:instanciation}
+The cascading algorithm can be instanciated with almost any classification and any regression algorithm.
 The choice of such subroutines may be dictated by the kind and amount of available data, by ease of use or by computational complexity, for example.
@@
 In a simple $k$-nearest neighboor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighboors of $s$.
-The generic M-SVM model makes the score function explicit \cite{guermeur2011generic} (equation 1).
-In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning}.
-FIXME:Recopier le QP
+The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}, equation 1).
+In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi : S\times A \rightarrow \mathbb{R}^p$ solves:
+\begin{equation}
+  \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
+  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~s.t.~} \forall i,
+  \theta^\top{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
+  {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
+\end{equation}
+With $l(s_i,a) = 1\textrm{ if } a\neq a_i,l(s_i,a_i)=0$ and the natural tabular features for as the $\phi$ vector ($0$ everywhere but on the unique index relative to the argument, where the component in $1$).
+The score function of this classifier is simply $q(s,a) = \theta^\top \phi(s,a)$.
 
 Ease of implementation lead us to choose a simple least-square regressor for our experiments.
-Using the same features $\phi$ as our classifier, it can be written (using straightforward matricial notations) as :  $\theta = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R$.
-The reward function is : $\hat R^C(s,a) = \theta^T \phi(s,a)$.
+Using the same features $\phi$ as our classifier, it can be written (using straightforward matricial notations) as :  $\omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R$.
+The reward function is : $\hat R^C(s,a) = \omega^T \phi(s,a)$.
 
 It is possible to get imaginative in the last step.
-For example, using a gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable the use of fuzzy reward reinforcement learning.
-FIXME:Trouver une ref pour les fuzzy rewards
+For example, using a gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable (nonwithstanding a nontrivial amount of work) the use of reward-uncertain reinforcement learning \cite{regan2011robust}.
 
 \subsection{Results}
+\label{subsec:results}
 \begin{figure}
   \includegraphics[width=\linewidth]{"Fig3"}
-  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.
-The lower baseline is an agent trained on a reward generated by using the same features as the cascading approach and the classifier, but with a random vector of parameters.}
+  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
   \label{fig:Highway}
@@
 Results are shown Fig.\ref{fig:Highway}.
+Performance is on par with the PRIL and SCIRL algorithm, slightly below the performance of the expert himself.
+Very few samples (100) are needed to reliably achieve expert-level performance.
+
 It is very interesting to compare our algorithm to the behavior of a classifier alone.
-With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrate far better performance, from the start.
-This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor truly make use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
+With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrate far better performance from the start.
+This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor and outputting not a policy, but a reward function that can be optimized on the MDP truly make use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
+
 Furthermore, the classifier whose results are diplayed here is the output of the first step of the algorithm.
-Although what we show on the plot is not exactly the classification error, it is closely related (if $\pi^C=\pi^E$ then the classification would have the same performance on our plot as the expert, which is not the case).
-This shows that our algorithm is empirically more forgiving of classification errors than our theoretical bound let us expect.
+The classification performance is obviously not that good, this shows that our algorithm is empirically more forgiving of classification errors than our theoretical bound let us expect.
 
@@
 This also show that we are indeed very sample efficient, without having to repeatedly solve the MDP.
+
+Finally, only the SCIRL algorithm operates with the same input as our proposed approach, with very slightly better results.
 
@@
 
-\section{Annexes}
-\label{sec-9}
-\subsection{GridWorld}
-\label{sec-9-1}
+%% \section{Annexes}
+%% \label{sec-9}
+%% \subsection{GridWorld}
+%% \label{sec-9-1}
 
-\begin{itemize}
-\item On montre avec les deux formes de récompenses qui donnent la même fonction de valeur.
+%% \begin{itemize}
+%% \item On montre avec les deux formes de récompenses qui donnent la même fonction de valeur.
 On fait référence aux travaux de Ng qui résolvent exactement ce problème.
 Les récompenses ont la même tête, ça devrait rassurer le lecteur hésitant.
-\item si on a le temps, on peut lancer une expérience où le nombre (en supposant une répartittion uniforme) des samples augmente et où la récompense tend vers ``la vraie''.
-\end{itemize}
-\subsection{Inverted pendulum}
-\label{sec-9-2}
+%% \item si on a le temps, on peut lancer une expérience où le nombre (en supposant une répartittion uniforme) des samples augmente et où la récompense tend vers ``la vraie''.
+%% \end{itemize}
+%% \subsection{Inverted pendulum}
+%% \label{sec-9-2}
 
-\begin{itemize}
-\item On montre qu'on généralise sans problème aux espaces continus, on montre qu'avec des données très, très sparses vis à vis de l'espace d'état on arrive pourtant à récupérer une récompense ayant la même fonction de valeur que celle de l'expert.
-\item ça montre aussi ce qui se passe quand l'espace d'hypothèse est ``mal choisi'' (i.e. la ``vraie'' récompense n'y était pas représentable).
-\end{itemize}
-\subsection{Highway}
-\label{sec-9-3}
+%% \begin{itemize}
+%% \item On montre qu'on généralise sans problème aux espaces continus, on montre qu'avec des données très, très sparses vis à vis de l'espace d'état on arrive pourtant à récupérer une récompense ayant la même fonction de valeur que celle de l'expert.
+%% \item ça montre aussi ce qui se passe quand l'espace d'hypothèse est ``mal choisi'' (i.e. la ``vraie'' récompense n'y était pas représentable).
+%% \end{itemize}
+%% \subsection{Highway}
+%% \label{sec-9-3}
 
-\begin{itemize}
-\item Si on a le temps : faire conduire quelqu'un puis lancer l'imitation, voir si on reproduit les consignes données à l'opérateur humain, qui aura eu la bonté de faire quelques erreurs dans l'exécution.
+%% \begin{itemize}
+%% \item Si on a le temps : faire conduire quelqu'un puis lancer l'imitation, voir si on reproduit les consignes données à l'opérateur humain, qui aura eu la bonté de faire quelques erreurs dans l'exécution.
 Ca permet de voir le comportement en cas d'expert sous optimal et de taire les questions à ce sujet.
-\end{itemize}
+%% \end{itemize}
 

### 2012-10-26 piot2012cascading/ECML13_cascading.tex
version 2 du draft

@@
 \usepackage{algorithmic}
+\usepackage{subfigure}
 \tolerance=1000
@@
 \newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
+\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
 \begin{document}
@@
 \author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
+\titlerunning{Cascading approach to IRL}
 \institute{
@@
 \and
-2.Supélec-Metz Campus\\
- MaLIS Research group, France\\
+2.Supélec\\
+ IMS-MaLIS Research group, France\\
  \texttt{firstname.lastname@supelec.fr}\\
 \and 
-3. UMI 2958 CNRS\\
-GeorgiaTech, France\\
+3. UMI 2958 GeorgiaTech-CNRS\\
+France\\
 }
@@
 \begin{abstract}
-  This paper considers the Inverse Reinforcement Learning (IRL) problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
-We propose to break the IRL problem down into two Supervised Learning (SL) steps: this is the cascaded approach.
-A classification step aiming to imitate the expert's policy gives a score function we see as an optimal action value function.
-In the second step, this score function is used together with the bellman evaluation equation to create samples then fed to a regressor that outputs a reward function.
-This cascaded approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the reward function the regression step outputs.
-We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning (RL) direct problem.
-Finally, up to the use some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy.
+  This paper considers the Inverse Reinforcement Learning problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
+We propose to break the IRL problem down into two generic Supervised Learning steps: this is the cascaded approach. A classification step followed by a regression step output a reward function.
+This approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the output reward function.
+We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning direct problem.
+This makes the proposed approach very competitive to the existing algorithms.
+Finally, up to the use of some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy.
+We quantitavely demonstrate the soundness of this approach on a highway driving simulator, and illustrate some aspects of the proposed algorithm on the classical inverted pendulum benchmark.
   \end{abstract}
@@
 Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} aims at inferring a reward function for which a demonstrated expert policy is optimal.
-An important part of the literature (see Section~\ref{sec:related} for a brief review) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution) which matches the one induced by the expert: this is called apprenticeship learning.
-Most algorithms of apprenticeship learning use Reinforcement
-Learning (RL) algorithms as subroutines, thus their performance depends strongly on the quality of these subroutines. Consequently, apprenticeship learning algorithms suffer from the
-same challenges of large state spaces, exploration vs. exploitation trade-offs as RL. In order to avoid to solve repeatedly RL problems, we decide to take a different point of view.
+An important part of the literature (see Sec.~\ref{sec:related} for a brief overview) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution, typically, the feature expectation) starting in a certain state space\footnote{Formally, there is a distribution over the starting state.} which matches the one induced by the expert.
+Most of these algorithms use Reinforcement
+Learning (RL) algorithms as subroutines as finding the optimal policy for a reward function is a necessary step before computing a measure of the previously mentionned distribution.
+Thus, their performance depends strongly on the quality of these subroutines.
+Consequently, they suffer from the same challenges of large state spaces, lack of data etc.
+as RL.
+In order to avoid repeatedly solving RL problems, we decide to take a different point of view.
 
-Having in mind that there is a one to one relation between a reward function and its associated optimal quality function (via the inverse Bellman equation), it is worth thinking of a method which is able to output a quality function for which the greedy policy is the demonstrated expert policy.
-Thus, the demonstrated expert policy will be optimal for the reward function associated to the quality function.
-As a method which is able to output a quality function for which the greedy policy is the demonstrated expert policy, we propose a score-based classification (see Section~\ref{sec:algo}).
-Besides, in order to retrieve, via the inverse Bellman equation, the reward associated to the score function the classification outputs, we propose a regression (see Section~\ref{sec:algo}).
+Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the inverse of the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy.
+Thus, the demonstrated expert policy will be optimal for the reward function associated to the action-value function.
+As a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy, we propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}).
+Besides, in order to retrieve, via the inverse of the Bellman optimality equation, the reward associated to the score function the classification outputs, we introduce a regression step (see Sec.~\ref{sec:algo}).
 That is why the method is called the Cascaded Supervised Approach (CSA).
-This method is justified by a theoretical analysis (see Section~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
+This method is justified by a theoretical analysis (see Sec.~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
 
-Finally, an instantiation of CSA is proposed~\ref{sec:experiments}.
-This algorithm does not need to solve iteratively an RL problem and needs only sampled transitions from expert and non-experts policies as inputs.
+Finally, an instantiation of CSA is proposed Sec.~\ref{sec:experiments}.
+This algorithm does not need to iteratively solve a RL problem and needs only sampled transitions from expert and non-experts policies as inputs.
 Moreover, up to the use of some heuristics, the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm, to a pure classification algorithm and to a recent successfull IRL method.
+The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm \cite{abbeel2004apprenticeship}, to a pure classification algorithm \cite{taskar2005learning} and to a recent successfull IRL method \cite{klein2012scirl}.
 
@@
 
-We frame the sequential decision problem in a finite \emph{Markov Decision Process} (MDP) \cite{puterman1994markov}, that is a tuple $\{S,A,P,R,\gamma\}$.
-$S$ is a finite state space, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the dynamic of the MDP.
+We work with finite \emph{Markov Decision Processes} (MDPs) \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$.
+The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP.
 Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$.
-At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one deterministic optimal policy.}) \emph{policy} $\pi\in A^S$.
-The agent then transits to a new state $s_{t+1}\in S$ according to the markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
+At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$.
+The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
 The reward function $R$ is a local measure of the quality of the control.
-The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t)|s_0 = s]$.
-Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, \forall s \in S, V^*_R(s) \geq V^\pi_R(s)$.
+The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s]$.
+Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$.
 The stationary distribution induced by a policy $\pi$ is written $\rho_\pi,\textrm{ s.t.
-}\rho_\pi^TP^\pi = \rho_\pi$.
+}\rho_\pi^TP^\pi = \rho_\pi^T$.
 
-The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by $T^{\pi_R}V = R + \gamma P_\pi V$ where $P_\pi$ is $(p(s'|s,\pi(s)))_{s,s' \in S}$.
+The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
+\begin{equation}
+  T^{\pi}_RV = R_\pi + \gamma P_\pi V
+  \end{equation}
+where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$ and $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$.
 The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$.
 Both operators are contractions.
-The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$; the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$.
-Optimal value function and optimal policy are also linked via the quality function $Q^\pi_R\in\mathbb{R}^{S\times A}$, a state-action value function that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^aV = R + \gamma P_a V$ with $P_a = (p(s'|s,a))_{s,s' \in S}$).
-The optimal policy follows a greedy mechanism with respect to its optimal quality function
+The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$, the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$.
+Another object object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$) and $R_a$ = $(R(s,a))_{s\in S}$.
+Once expanded, this gives what we refer to as the Bellman optimality equation :
+\begin{equation}
+  Q^*_R = R_a + \gamma P_{\pi^*_R}V^*_R
+  \label{eq:bellman1}
+\end{equation}
+
+An optimal policy follows a greedy mechanism with respect to its optimal action-value function 
 \begin{equation}
@@
 
-When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to retrieve the optimal control policy \cite{sutton1998reinforcement}.
+When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
 We call this the direct problem.
@@
 We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal.
-The non optimality of the expert is an interesting setting that has been discussed for example in \cite{melo2012analysis} but that we are not addressing here.
-We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is also near-optimal.
+The suboptimality of the expert is an interesting setting, that has been discussed for example in \cite{melo2010analysis,syed2012reduction}, but that we are not addressing here due to the numerous complications it introduces.
+We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal.
 The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal).
-Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see section \ref{sec:related}.
+Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.
 \section{The cascading algorithm}
@@
 Our first step in the quest for a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (MC$^2$).
-This classifier learn a score function $q:S\times A \rightarrow \mathbb{R}$ that rates the association of a certain label $a\in A$ with a certain input $s\in S$.
-The classification rule $\pi_C: S \rightarrow A$ is simply to select (one of) the label(s) that achieve the highest score for the given input :
+This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$.
+The classification rule $\pi_C\in A^S$ is simply to select (one of) the label(s) that achieves the highest score for the given inputs :
 \begin{equation}
@@
 \end{equation}
-Most notably, \emph{Multi-class Support Vector Machines} can be seen as MC$^2$ algorithms.
-We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in subsection \ref{subsec:instanciation}.
+For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as MC$^2$ algorithms, we however used a structured margin approach.
+We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in Subsec.~\ref{subsec:instanciation}.
 
 Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier.
-Apart from discussion and comparison purpose, we do not make use of the classification policy $\pi_C$.
-What is of interest to us is the score function $q$ itself.
-One can easily notice the similarity between equations \ref{eq:greedy2} and \ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal quality function.
-As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal quality function for the classifier policy $\pi_C$.
-By inversing the Bellman evaluation equation, one can express the reward as a function of the quality function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$.
-This leave us with $r$, the reward function relative to our score/quality function $q$ : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s'))$.
-
-We wish to solve the general, or approximate, IRL problem where the transition probabilities are unknown.
+The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL).
+What is of particular interest to us is the score function $q$ itself.
+One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal action-value function.
+As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$ (which can itself be seen as a good approximation of $\pi_E$).
+By inversing the Bellman equation, one can express the reward as a function of the action-value function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$.
+This leave us with $R^C$, the reward function relative to our score/action-value function $q$ :
+\begin{equation}
+  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
+\end{equation}
+We wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown.
 For this we use another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to.
@@
 The dataset $D_R$ brings us information about the dynamics of the system.
-Our reward function $r$ must be approximated to $\hat r$ with the help of the information in $D_R$.
-This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor \cite{vapnik1996nature} to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space.
-The complete algorithm is given Alg.
-\ref{algo:cascading}.
+Our reward function $R^C$ must be approximated with the help of the information in $D_R$.
+This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor (a simple least-square approximator can do) to obtain $\hat R^C$, a generalisation of $\{(\hat r_j)_j\}$ over the whole state-action space.
+The complete algorithm is given Alg.~\ref{algo:cascading}.
 \begin{algorithm}%[H]
     %\small
-  \caption{Cascading IRL algorithm}
+  \caption{CSA-IRL algorithm}
   \label{algo:cascading}
@@
 Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem.
-Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with what state), but there is more to a MDP than a single policy.
-Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probability $P$.
+Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with which state), but there is more to a MDP than a single policy.
+Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probabilities $P$.
 The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.
 
-As the analysis (Sec \ref{sec:analysis}) reveals, both classification error and regression error should be controlled.
-By carefully choosing or tuning both supervised algorithm, it should possible to mitigate any ill effects of the specific settings the cascading algorithm is used in.
-If one operates in a well-known domain where a lo of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings.
-If data is lacking, one could for example have to trade off between bias and variance.
-The possibility to use and leverage the wide range of techniques developped for these two well known supervised problems is a strong advantage of the proposed cascading approach.
-%Reformuler ce passage
-The explanation in this section only is a quatitative attempt to make the reader understand the idea behind the algorithm.
-The analysis of the next section tackle things in a formal, yet less intuitive, way.
+As the analysis (Sec.~\ref{sec:analysis}) reveals, both classification error and regression error should be controlled.
+By carefully choosing or tuning both supervised learning algorithms, it should be possible to mitigate any ill effects of the specific settings the cascading algorithm is used in.
+If one operates in a well-known domain where a lot of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings.
+If data is lacking, one could for example have to overcome bias, variance or generalization problems in the regression step (more about that in Subsec.\ref{subsec:instanciation}).
+The possibility to use and leverage the wide range of techniques developped for these two well-known supervised problems is a strong advantage of the proposed cascading approach.
+
+The explanation in this section is only a quatitative attempt to make the reader understand the idea behind the algorithm.
+The analysis of the next section tackles things in a formal, yet less intuitive, way.
 
@@
 
-In this Section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
-We formulate this result in a more formal way in Theorem \ref{theorem: theorem analysis}
+In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs, more formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]$ is small by bounding it by a term that depends on :
+\begin{itemize}
+  \item the regression and classification errors, respectively $\epsilon_R$ and $\epsilon_C$, these terms are small if the regressor and the classifier are good;
+  \item $C_*$, a concentration coefficient assessing the closeness of the stationary distribution of $\pi_E$ and $\hat \pi_C$, the latter being the optimal policy for the reward $\hat R^C$ output by the algorithm ;
+  \item $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$, which could be normalized to $1$ without loss of generality.
+  \end{itemize}
 
@@
 \begin{equation}
-0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\leq \frac{1}{1-\gamma}(\Delta q \epsilon_C+(1+C_*)\epsilon_R)).
+0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \frac{1}{1-\gamma}\left(\epsilon_C\Delta q +\epsilon_R(1+C_*)\right).
 \end{equation}
-$\Delta q$, and $C_*$ are constants and are defined in the proof.
-$\epsilon_C$ represents the classification error and $\epsilon_R$ represents the regression error: both terms are well-defined in the proof.
 \end{theorem}
 \begin{proof}
-First let introduce some notations.
-$q\in\mathbb{R}^{S\times A}$ is the score function outputted by the classification step, $\forall s \in S,\pi_C\in\argmax{a\in A}q(s,a)$ is a deterministic classifier policy and $\epsilon_C$ is the classification error:
+First let's introduce some notations.
+$q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\forall s \in S,\pi_C\in\argmax_{a\in A}q(s,a)$ is a deterministic classifier policy.
+We recall that $\epsilon_C$ is the classification error:
 \begin{equation}
@@
 \begin{equation}
-\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) +\gamma\sum_{s'\in S}P(s'|s,a)q(\pi_C(s'),a),
+\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s')),
 \end{equation}
-and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function outputted by the regression step.
-The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$, $\epsilon^R\in\mathbb{R}^{S\times A}$ must not be mistaken for the regression error $\epsilon_R\in\mathbb{R}$ defined below.
+and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function output by the regression step.
+
+The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$, $\epsilon^R\in\mathbb{R}^{S\times A}$ must not be mistaken for the regression error $\epsilon_R\in\mathbb{R}$ that appears in the final result.
 We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof:
 \begin{equation}
-\forall (s,a)\in S\times A, R^E(s,a)=q(s,a) +\gamma\sum_{s'\in S}P(s'|s,a)q(\pi_E(s'),a).
+\forall (s,a)\in S\times A, R^E(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_E(s')).
 \end{equation}
@@
 \end{align}
-Now we are going to bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious).
-We note $\hat{\pi}_C$ a deterministic optimal policy of the reward $\hat{R}^C$. We have:
+Now we are going to upper bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious as $V^*$ is optimal).
+We note $\hat{\pi}_C$ a deterministic optimal policy of the reward $\hat{R}^C$. First, the term $V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}$ is decomposed:
 \begin{equation}
@@
 \end{equation}
-We are going to bound each of this three terms. But, first, let $\pi$ be a given deterministic policy. We have:
+We are going to bound each of these three terms. But, first, let $\pi$ be a given deterministic policy. We have:
 \begin{equation}
@@
 \end{equation}
-If $\pi=\pi_E$, $\rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\epsilon^R_{\pi_E}=\frac{1}{1-\gamma}\rho_E^T\epsilon^R_{\pi_E}\leq\frac{1}{1-\gamma}\|\epsilon^R_{\pi_E}\|_{1,\rho_E}$.
+We use $\|.\|_{1,\mu}$ to denote the $\mu$-weighted $L_1$ norm : $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$. 
+If $\pi=\pi_E$, we have, thanks to the power series expression of $(I-\gamma P_{\pi_E})^{-1}$ and the definition of $\rho_E$:
+\begin{equation}
+  \rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\epsilon^R_{\pi_E}=\frac{1}{1-\gamma}\rho_E^T\epsilon^R_{\pi_E}\leq\frac{1}{1-\gamma}\|\epsilon^R_{\pi_E}\|_{1,\rho_E}.\end{equation}
+
 If $\pi\neq\pi_E$, we introduce the following concentration coefficient:
@@
 \end{equation}
-Now, there is a still an upper bound to find for the second term. It is possible to decompose the second term:
+Now, there is still an upper bound to find for the second term. It is possible to decompose the second term as follows:
 \begin{equation}
@@
 By construction, we have $V^{\pi_C}_{R^C}=q_{\pi_C}$ and $V^{\pi_E}_{R^E}=q_{\pi_E}$, thus:
-\begin{equation}
-\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})=\rho_E^T(q_{\pi_C}-q_{\pi_E})=\sum_{s\in S}\rho_E(s)(q(s,\pi_C(s))-q(s,\pi_E(s)))[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
-\end{equation}
-We note $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$ and we have:
+\begin{eqnarray}
+\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})&=&\rho_E^T(q_{\pi_C}-q_{\pi_E})\\
+&=&\sum_{s\in S}\rho_E(s)(q(s,\pi_C(s))-q(s,\pi_E(s)))[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
+\end{eqnarray}
+We recall that $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$. We have:
 \begin{equation}
@@
 \begin{equation}
-0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\leq \frac{1}{1-\gamma}(\Delta q \epsilon_C+(1+C_*)\epsilon_R)).
+0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \frac{1}{1-\gamma}(\epsilon_C\Delta q +\epsilon_R(1+C_*))).
 \end{equation}
@@
 
-The bound is obviously true for $\hat{R}_C = 0$, but for this to happen the classifier or the regressor must have very bad performance, which is not likely.
+The first remark one can make about this result is that the bound is tight if both the classifier and the regressor are perfect (i.e.
+if the associated error terms are null).
+In this case, the expert's policy  $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness implies that $\forall s\in S, \Card(\argmax_{a\in A}q(s,a))=1$, with $\Card(X)$ denoting the cardinality of the set $X$).
+This uniqueness also hints at the fact that we found a non-trivial reward (we recall that the null reward admit \emph{every} policy as optimal).
+Therefore, obtaining  $\hat{R}_C = 0$ (for which the bound is obviously true: the bounded term is $0$, the bounding term is positive) is unlikely as long as the classifier and the regressor exhibit decent performance.
 
-The algorithm and the analysis are invariant by  dilatation of $q$, hence of $\Delta q$.
+In order to control $\epsilon_R$, we need a precise regression over the expert stationary distribution and for all the actions.
+In theory, we need to be able to sample the whole dynamics (even the non expert actions).
+In practice heuristics can be used (see Sec.~\ref{sec:experiments}) however this heuristics is not justified by the analysis.
+
+As mentionned earlier, the algorithm and the analysis are invariant by  dilatation of $q$, hence of $\Delta q$.
 We could normalize $\Delta q$ by imposing $\Delta q=1$ (the range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$).
@@
 The constant $C_*$ could be estimated a posteriori ( after $\hat{R}^C$ is computed).
-We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (approximately, if $\hat{\pi_C}\approx\pi_E$ then $C_* \approx 1$).
+We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (approximately, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
 
-In order to control $\epsilon_R$, we need a precise regression over the expert stationary distribution and for all the actions.
-In theory, we need to be able to sample the whole dynamic (even the non expert actions).
-In practice heuristics can be used (see Section~\ref{sec:experiments}) however this heuristics is not justified by the analysis.
-
-If the classifier and the regressor are perfect ($\epsilon_C=\epsilon_R=0$), $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness suppose that $\forall s\in S, Card(\argmax_{a\in A}q(s,a))=1$, with $Card(X)$ denoting the cardinality of the set $X$).
-This uniqueness also hint at the fact that we found a non-trivial reward (we recall that the null reward admit \emph{every} policy as optimal).
-
-The bound can be controlled by teh quality of the classification and regression subroutines.
+The bound can be fully controlled by the quality of the classification and regression subroutines.
 The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off etc.) to be found.
-Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach.
+Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach over existing algorithms.
 
@@
 IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
-A seminal paper, \cite{abbeel2004apprenticeship} blurs the frontier between imitation and IRL, as this algorithm outputs (mixing) policies, and not explicitely a reward although IRL is the core of this method.
+The seminal paper \cite{abbeel2004apprenticeship} blurs the frontier between imitation and IRL, as this algorithm outputs (mixing) policies, and not explicitely a reward although IRL is the core of this method.
 This work introduced the general method of finding a reward function such that the corresponding optimal policy's feature expectation (or
@@
 The cascading approach being able to output a reward function without having to solve the MDP is thus a significant improvement.
-The Projection IRL (PIRL) algorithm of \cite{abbeel2004apprenticeship} is used as a benchmark in our experiments, see subection \ref{subsec:setting}.
+The Projection IRL (PIRL) algorithm of \cite{abbeel2004apprenticeship} is used as a benchmark in our experiments, see Subsec.~\ref{subsec:setting}.
 
 Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
+Classification on the same level as the first step of the CSA-IRL method (with actions as labels and states as inputs) can be found in \cite{ratliff2007imitation} with the help of a structured margin method, but it did not include the structure of the MDP as the CSA-IRL algorithm does.
 
-Using the non trivial notion of metric in an MDP, \cite{melo2010learning} build a kernel which is used
+Using the non trivial notion of metric in an MDP, \cite{melo2010learning} builds a kernel which is used
 in a classification algorithm, showing improvements compared to a
@@
 
-Few algorithms do not require solving the direct RL problem. \cite{Dvij:2010} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
-
-\cite{boularias:2011} uses a subgradient ascent of a utility function based on the notion of relative entropy.
+Few algorithms do not require solving the direct RL problem.
+The approach of \cite{syed2008apprenticeship} requires knowing the transition probabilities $P$ of the MDP (which we don't need) and outputs a policy (and not a reward).
+The algorithm in \cite{dvijotham2010inverse} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
+Closer to our use-case is the idea presented in \cite{boularias2011relative} to use a subgradient ascent of a utility function based on the notion of relative entropy.
 Importance sampling is suggested as a way to avoid solving the MDP.
@@
 
-The LPAL algorithm of \cite{syed2008apprenticeship} also does not require solving the direct problem, but it requires knowing the transition probabilities $P$ of the MDP (which we don't need).
-
-Finally, the SCIRL algorithm \cite{klein2012scirl} is able to solve the IRL problem in the same dire conditions the cascading approach is tested in.
-However, it needs a good estimate of the expert's feature expectation in the states visited by the expert, but for all actions not just the expert's.
-The use of some heuristics can save the day, but it is not known wether such heuristics translate well on a more difficult problem.
-We also use it as a benchmark in our experiments, where it shows very slightly better results than the cascading algorithm.
+Finally, the SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} is able to solve the IRL problem in the same dire conditions the cascading approach is tested in.
+It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation as feature function.
+This necessity of a linarly parametrized approach makes the SCIRL algorithm less convenient than CSA-IRL, which can make use on non-parametrized supervised subroutines.
+SCIRL is used for comparison purposes on the Highway driving problem.
 
@@
 In this section, we put our approach under stress to empirically demonstrate its behavior.
-After explaining the criterion we consider and its rationale (subsection \ref{subsec:criterion}), we describe the benchmark's setting along with the other algorithms we use for comparison purposes (subsection \ref{subsec:setting}), before giving and anlyzing the results (subsection \ref{subsec:results}), we explain the details of our choices for the classification and regression subroutines and give some hints for other possibilities (subsection \ref{subsec:instanciation}).
+After explaining the criterion we consider and its rationale (Subsec.~\ref{subsec:criterion}), we describe the benchmark's setting along with the other algorithms we use for comparison purposes (Subsec.~\ref{subsec:setting}).
+Before giving and anlyzing the results (Subsec.~\ref{subsec:results}), we explain the details of our choices for the classification and regression subroutines and give some hints for other possibilities (Subsec.~\ref{subsec:instanciation}).
+We also give the details of our heuristics.
+
+Finally, Subsec.~\ref{subsec:IP} provides some results on a continuous state space setting (the inverted pendulum) in order to provide the reader with a practical illustration of the workings of our algorithm.
+There is no comparison with other methods, these results are provided to help the reader form an intuition about the way the CSA-IRL algorithm works.
 
@@
 \label{subsec:criterion}
-The theoretical bound provided in Sec.
-\ref{sec:analysis} is about the optimality of the expert with respect to the reward function output by our algorithm (this is also the way the problem was framed in \ref{sec:background}).
+The theoretical bound provided in Sec.~\ref{sec:analysis} is about the optimality of the expert with respect to the reward function computed by our algorithm (this is also the way the problem was framed in Sec.~\ref{sec:background}).
 For the experiments, we use a somewhat more restrictive criterion : a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the "true" reward function, the reward function of the expert.
-We compare the mean (w.r.t.
-the uniform distribution over the states) value function with respect to $R^E$ for a RL agent perfectly trained to optimize $R^C$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
-We do that firstly because, for once, we can : the bound we provide in the analysis is something that can be measured even if $R^E$ does not actually exist.
-We make use here of the fact that $R^E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R^E$.
-The second reason why we choose this different metric in this section is because we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well behaved given the use of heuristics.
+We compare the mean (over the uniform distribution) value function with respect to $R_E$ of a policy $\pi$ learned by optimizing $\hat R^C$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
+We do that firstly because, for once, we can : the bound we provide in the analysis is something that can be measured even if $R_E$ does not actually exist.
+We make use here of the fact that $R_E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R_E$.
+The second reason for this different metric is that we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well-behaved given the use of heuristics.
 \subsection{Setting}
 \label{subsec:setting}
-The setting of the experiment is a driving simulator inspired from a benchmark seen in \cite{syed2008apprenticeship,syed2008game}.The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
+The setting of the experiment is a driving simulator inspired from a benchmark already used in \cite{syed2008apprenticeship,syed2008game}.
+The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
 At all times, there will be one car in one of the three lanes.
 Even at the lowest speed, the player's car moves faster than the others.
-When the other car disappear at the bottom of the screen, another one appear at the top in a randomly chosen lane.
+When the other car disappears at the bottom of the screen, another one appear at the top in a randomly chosen lane.
 It takes two transitions to completely change lanes, as the player can move left or right for half a lane's length at a time.
@@
 %FIXME:METTRE UNE CAPTURE DECRAN.
-The expert is trained to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
+The reward function $R_E$ trains the expert to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
 Any other situation receives a null reward.
@@
 We also compare the proposed cascading method to the PIRL algorithm of \cite{abbeel2004apprenticeship} and the SCIRL algorithm of \cite{klein2012scirl}.
-It is worth noting that the random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
+The random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
 PIRL is instanciated in a way that make use of the transition probability matrix $P$ to exactly compute all relevant objects (feature expectations, optimal policies) via exact dynamic programming.
 It is left to run for 70 iterations.
+We could have displayed PIRL's data sensitivity by making $\pi_E$ known only through observations, but we decided against it because we want this baseline to show the performance of a fully informed well-established appreticeship algorithm.
+
+Both datasets $D_C$ and $D_R$ will be constructed from samples drawn with the expert's policy.
+The expert drives the car in the simulator for $n$ runs of $n$ transitions each, with $n$ varying in $\{3,7,10,15,20\}$.
+Apart from SCIRL, to the best of our knowledge, no other algorithm can infer a reward function only from expert data.
+Recall from Sec.~\ref{sec:related} that most existing approaches either need to solve the direct RL problem repeatedly (which can not be done on this setting from expert data alone) or rely on a sampling of the dynamics of the whole MDP.
+The reward functions found by SCIRL and CSA-IRL are then optimized using a dynamic programming algorithm to show the performance of the associated policy.
+
 \subsection{Instanciation}
 \label{subsec:instanciation}
-The cascading algorithm can be instanciated with almost any classification and any regression algorithm.
+The cascading algorithm can be instanciated with some standard classification algorithms and any regression algorithm.
 The choice of such subroutines may be dictated by the kind and amount of available data, by ease of use or by computational complexity, for example.
@@
 In a simple $k$-nearest neighboor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighboors of $s$.
-The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}, equation 1).
+The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}).
 In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi : S\times A \rightarrow \mathbb{R}^p$ solves:
-\begin{equation}
-  \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
-  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~s.t.~} \forall i,
+\begin{eqnarray*}
+  &\min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
+  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \\
+  &\text{~s.t.~} \forall i,
   \theta^\top{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
   {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
-\end{equation}
-With $l(s_i,a) = 1\textrm{ if } a\neq a_i,l(s_i,a_i)=0$ and the natural tabular features for as the $\phi$ vector ($0$ everywhere but on the unique index relative to the argument, where the component in $1$).
+\end{eqnarray*}
+with $l(s_i,a) = 1\textrm{ if } a\neq a_i,l(s_i,a_i)=0$ and the natural tabular features as the $\phi$ vector ($0$ everywhere but on the unique index relative to the argument, where the component in $1$).
 The score function of this classifier is simply $q(s,a) = \theta^\top \phi(s,a)$.
+This is the classifier our implementation of SCIRL \cite{klein2012scirl} is based on.
+We chose to use it in CSA-IRL to provide a fair comparison between both algorithms.
 
-Ease of implementation lead us to choose a simple least-square regressor for our experiments.
+Ease of implementation leads us to choose a simple least-square regressor for our experiments.
 Using the same features $\phi$ as our classifier, it can be written (using straightforward matricial notations) as :  $\omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R$.
@@
 It is possible to get imaginative in the last step.
-For example, using a gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable (nonwithstanding a nontrivial amount of work) the use of reward-uncertain reinforcement learning \cite{regan2011robust}.
+For example, using a Gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable (nonwithstanding a nontrivial amount of work) the use of reward-uncertain reinforcement learning \cite{regan2011robust}.
 
+As we only had transitions $\{s_j,\pi_E(s_j),s'_j\}$from the expert to construct the samples $\{s_j,a_j,r_j\}$ fed to the regressor, we had to use a heuristics.
+We assert that disagreeing with the expert's choice is undesirable.
+Mathematically, we do that by artificially introducting samples $\{s_j,a,r_{min}\}_{a\neq \pi_E(s_j)}$ where $r_{min} = \min_j\hat r_j - 1$.
+Although this heuristics was not analyzed in Sec.~ \ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown here demonstrate its soundness.
 \subsection{Results}
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{"Fig3"}
-  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs.}
+   \subfigure[PIRL and Random Baselines.
+There is no abcissa as these methods do not need data input.
+PIRL uses the transition probabilities matrix $P$.]{\includegraphics[width=.5\linewidth]{"Fig1-NoX"}}
+   \subfigure[Cascading and Classification.]{\includegraphics[width=.5\linewidth]{"Fig2-ClassifAndUs"}}
+  \subfigure[SCIRL and Cascading mean value.]{\includegraphics[width=.5\linewidth]{"Fig3-SCIRLAndUs"}}
+  \subfigure[Legend]{\includegraphics[width=.5\linewidth]{"Fig4-Legend"}}
+
+  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. See Subsec.~\ref{subsec:setting} for detailled information about the experimental setting.}
   \label{fig:Highway}
@@
 
-Results are shown Fig.\ref{fig:Highway}.
-Performance is on par with the PRIL and SCIRL algorithm, slightly below the performance of the expert himself.
+Results are shown Fig.~\ref{fig:Highway}.
+We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL and CSA-IRL, the policy $\pi_C$ of the classifier (the very one the classification step of CSA-IRL outputs), the output of the PIRL algorithm, the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
+Performance for CSA-IRL is on par with the PIRL and SCIRL algorithms, slightly below the performance of the expert himself.
 Very few samples (100) are needed to reliably achieve expert-level performance.
@@
 It is very interesting to compare our algorithm to the behavior of a classifier alone.
-With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrate far better performance from the start.
-This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor and outputting not a policy, but a reward function that can be optimized on the MDP truly make use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
+With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrates far better performance from the start.
+This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor and outputting not a policy, but a reward function that can be optimized on the MDP truly makes use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
 
@@
 PIRL is given the transition probabilities (that is, a $|S|\times |S|\times |A|$ matrix) and yet it does only slightly better than our approach with 100 samples.
-This also show that we are indeed very sample efficient, without having to repeatedly solve the MDP.
+This also shows that SCA-IRL is indeed very sample efficient, without having to repeatedly solve the MDP.
 
-Finally, only the SCIRL algorithm operates with the same input as our proposed approach, with very slightly better results.
+Finally, only the SCIRL algorithm operates on the same data as our proposed approach, with very slightly better results.
 
+\subsection{Inverted Pendulum}
+\label{subsec:IP}
+Finally, we provide results on the inverted pendulum problem.
+We do not compare our approaches with others as we did on the highway driving problem.
+The information given here serves an illustrative purpose.
+
+The inverted pendulum (also known as the cart pole) is a classical benchmark in the RL community, it is fully described in \cite{lagoudakis2003least}.
+The goal is to maintain a pole in a unstable equilibrium position with the help of a cart moving along one axis.
+The state space is continuous, it consists in the angular position an speed of the pole.
+It is possible to use a RL algorithm to learn how to balance the pendule in an upright position by using the following reward : as long as the angular position of the pole is (in absolute value) less than $\pi \over 2$ the reward is $0$, the reward is $-1$ if the pole falls.
+Three actions are available to the agent: it can choose to apply a leftward or rightward force on the cart or no force at all.
+We used the feature parametrization of \cite{lagoudakis2003least}, that is a grid of 9 Gaussian functions and a constant.
+These are used to approximate the value function as well as for the regression step of the CSA-IRL algorithm.
+The instanciation of the algorithm is the same as for the highway driving problem except for the feature functions (which we just described).
+
+\begin{figure}
+   \subfigure[Reward on which the expert was trained]{\label{toto}\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_true_R"}}
+   \subfigure[Reward found by CSA-IRL. The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_lafem_R"}}
+  \subfigure[Value function of the expert (i.e. optimal value function for the reward displayed Fig.~\ref{toto})]{\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_Vexpert"}}
+  \subfigure[Optimal value function for the reward function found by CSA-IRL (Fig.~\ref{tata})]{\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_Vagent"}}
+
+  \caption{Results on the inverted pendulum benchmark, illsutrating the dissimilarity of the rewards, but the similarity of the value functions.
+The state space is represented on a position-speed plan.}
+  \label{fig:IP}
+\end{figure}
+
+As with the previous benchmark, CSA-IRL only used data from the expert to devise a reward function.
+The results are shown Fig.~\ref{fig:IP}.
+In the left column, we show the reward and value functions of the expert.
+The expert (a RL agent trained on the specified reward) is able to balance the pendule indefinitely.
+With as little as 300 transitions (30 seconds of balancing), CSA-IRL is able to find a reward that, once optimized by a RL agent, will reliably lead to a control as good as the expert's.
+As Fig.~\ref{tata} illustrates, the data we get from the expert only covers a tiny fraction of the state space.
+Furthermore, the feature functions of our linear approximation scheme for the reward did not permit to accurately approximate the true reward\footnote{The true reward function is a function in $\mathbb{R^S}$, CSA-IRL outputs a reward in $\mathbb{R}^{S\times A}$, but we plot here $R(s,\pi^*(s))$ with $\pi^*$ the optimal policy for $R$.}.
+Despite these two setbacks, CSA-IRL manage to find a reward for which not only the expert's control is near-optimal (that is what the analysis of Sec.~\ref{sec:analysis} garantees under the assumption that a reasonable amount of data is available in order to keep both classification and regression errors down) but that also lead to a control policy as good as the expert's \emph{with respect to the (theoretically) unknown expert's reward function}.
 %% \begin{itemize}
@@
 \label{sec:conclusion}
-We have introduced a new way to perform IRL by cascading two well-known supervised approaches, namely classification and regression.
+We have introduced a new way to perform IRL by cascading two supervised approaches.
 The expert is shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.

### 2012-10-30 piot2012cascading/4plots.py
Plot code

@@ -1,0 +1,104 @@
+from matplotlib import rc
+rc('text', usetex=True)
+import sys
+sys.path+=['..']
+from numpy import *
+import scipy
+import pylab as pylab
+from Plot import *
+
+D_Cascading = genfromtxt( "Cascading_Exp5.mat" )
+D_C = D_Cascading[:,[0,1]]
+D_Classif = D_Cascading[:,[0,2]]
+D_SCIRL = genfromtxt("SCIRL_Exp3.mat")
+
+
+Expert = 7.74390968 #python Expert.py to get this value
+Random_mean = -1.5821833963484#See Highway.org about Random.py for information on these values
+Random_min = -4.0007295890199996#python Random.py to get this value
+Random_max = 2.7064859345599999#python Random.py to get this value
+Random_var = 1.4465398450419833#python Random.py to get this value
+ANIRL_mean = 7.2183449001
+
+y_min = -5
+y_max = 10
+#Figure 1 : No X-axis, random, Expert and PIRL
+pylab.figure(1)
+pylab.clf()
+X = array([0,1])
+filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), Random_min*ones(X.shape), Random_max*ones(X.shape), 'cyan',0.2,'--',"Agent trained on a random reward",None)
+filled_mean_min_max( pylab, X, Random_mean*ones(X.shape), (Random_mean-Random_var)*ones(X.shape), (Random_mean+Random_var)*ones(X.shape), 'cyan',0.4,'-.',None,None)
+pylab.axes().get_xaxis().set_visible(False)
+pylab.ylabel('$\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{\mathcal{R}_E}(s)]$')
+pylab.plot(X,[ANIRL_mean,ANIRL_mean],color='blue',label="Abbeen \& Ng IRL",lw=2,linestyle='-')
+pylab.plot(X,[Expert,Expert],color='purple',label="Expert",lw=2,linestyle='-')
+pylab.ylim((-5,10))
+pylab.savefig("Fig1-NoX.pdf",transparent=True)
+
+
+#Figure 2 : classif and us
+pylab.figure(2)
+pylab.clf()
+[X,Y_mean,Y_min,Y_max] = map( array, mean_min_max( D_Classif ))
[…]

### 2012-10-30 piot2012cascading/ECML13_cascading.tex
Grands changements paris

@@
 %\newtheorem{theorem}{Theorem}
+\newtheorem{cor}{Corollary}
 \usepackage{dsfont}
+\usepackage{mathtools}
+\mathtoolsset{showonlyrefs=true}
 \providecommand{\alert}[1]{\textbf{#1}}
@@
   This paper considers the Inverse Reinforcement Learning problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
-We propose to break the IRL problem down into two generic Supervised Learning steps: this is the cascaded approach. A classification step followed by a regression step output a reward function.
-This approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the output reward function.
-We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning direct problem.
-This makes the proposed approach very competitive to the existing algorithms.
-Finally, up to the use of some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy.
-We quantitavely demonstrate the soundness of this approach on a highway driving simulator, and illustrate some aspects of the proposed algorithm on the classical inverted pendulum benchmark.
+We propose to break the IRL problem down into two generic Supervised Learning steps: this is the Cascaded Supervised IRL (CSI) approach.
+A classification step that defines a score function is followed by a regression step that outputs a reward function.
+An analysis of this approach shows that the demonstrated expert policy is near-optimal for the computed reward function.
+Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach.
+It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics.
+For this purpose two classical benchmarks (a highway driving simulator and a inverted pendulum simulator) are used.
   \end{abstract}
@@
 \label{sec-2}
+The sequential decision making problem consists in taking the appropriate action pondering the available data in order to maximize a certain criterion.
+When framed in an MDP (see Sec.~\ref{sec:background}), the criterion to maximize is the expected sum of discounted rewards.
+The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which we tackle here, aims at inferring a reward function for which a demonstrated expert policy is optimal.
 
-Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} aims at inferring a reward function for which a demonstrated expert policy is optimal.
-An important part of the literature (see Sec.~\ref{sec:related} for a brief overview) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution, typically, the feature expectation) starting in a certain state space\footnote{Formally, there is a distribution over the starting state.} which matches the one induced by the expert.
-Most of these algorithms use Reinforcement
-Learning (RL) algorithms as subroutines as finding the optimal policy for a reward function is a necessary step before computing a measure of the previously mentionned distribution.
-Thus, their performance depends strongly on the quality of these subroutines.
-Consequently, they suffer from the same challenges of large state spaces, lack of data etc.
-as RL.
-In order to avoid repeatedly solving RL problems, we decide to take a different point of view.
+IRL is often used as a way to do Apprenticeshipe Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function.
+The reward function nevertheless is of interest in and of itself.
+As mentionned in \cite{ng2000algorithms}, its semantics can be analyzed in biology or econometrics.
+Practically, the reward can be seen as a succinct description of a task.
+Discovering it removes the coupling between understanding the task and learning how to fulfill it that exists in AL.
+IRL allows the use of (Approximate) Dynamic programming ((A)DP) or Reinforcement Learning (RL) techniques to learn the task from the computed reward function.
+Although some approaches use it as a byproduct, most AL algorithm make not use of the reward function.
+A very straightforward way to do AL is to use a multi-class classifier to directly learn the expert policy.
+We compare this idea with our IRL algorithm in the experiments (Sec.~\ref{sec:experiments}).
 
-Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the inverse of the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy.
-Thus, the demonstrated expert policy will be optimal for the reward function associated to the action-value function.
-As a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy, we propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}).
-Besides, in order to retrieve, via the inverse of the Bellman optimality equation, the reward associated to the score function the classification outputs, we introduce a regression step (see Sec.~\ref{sec:algo}).
-That is why the method is called the Cascaded Supervised Approach (CSA).
-This method is justified by a theoretical analysis (see Sec.~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
+A lot of existing approaches in either IRL or AL need to repeatedly solve the underlying MDP.
+Thus, their performance depends strongly on the quality of the associated subroutine.
+Consequently, they suffer from the same challenges of scalability, data scarcity, etc., as RL and (A)DP.
+In order to avoid repeatedly solving such problems, we adopt a different point of view.
 
-Finally, an instantiation of CSA is proposed Sec.~\ref{sec:experiments}.
-This algorithm does not need to iteratively solve a RL problem and needs only sampled transitions from expert and non-experts policies as inputs.
-Moreover, up to the use of some heuristics, the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm \cite{abbeel2004apprenticeship}, to a pure classification algorithm \cite{taskar2005learning} and to a recent successfull IRL method \cite{klein2012scirl}.
+Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy.
+Thus, the demonstrated expert policy will be optimal for the related reward function.
+We propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function.
+Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}).
+That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI).
+This method is analyzed, see Sec.~\ref{sec:analysis} where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
 
+This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-experts policies as inputs.
+Moreover, up to the use of some heuristics (see Subsubsec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
+A specific instantiation of CSI (proposed Subsec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Subsec~\ref{subsec:IP}) and a car driving simulator (Subsec~\ref{subsec:Highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.
+
+Differences and similarities with existing AL or IRL approaches are succinctly exposed in Sec.~\ref{sec:related}.
 \section{Background and notations}
 \label{sec:background}
-%% \begin{itemize}
-%% \item Parler du domaine de la prise de décision séquentielle
-%% \item Expliquer le problème direct, en profiter pour introduire l'équation d'évaluation de Bellman et l'équation d'optimalité de Bellman
-%% \begin{itemize}
-%% \item $RL(MDP=\{S,A,\gamma,(P),R\}) = \pi^*:S\times A \rightarrow \mathbb{R}$
-%% \item $\forall \pi, Q^\pi(s,a) = R(s,a) + \sum_{s'}P(s'|s,a)Q^\pi(s',\pi(s'))$
-%% \item $\pi^* = \arg\max_aQ^*(s,a)$
-%% \end{itemize}
-%% \item Introduire le problème inverse dans une forme mathématique relativement propre
-%% \begin{itemize}
-%% \item $IRL( MDP\backslash R = \{S,A,\gamma,(P)\} \cup \{\pi^E\} = R : S\times A \rightarrow \mathbb{R}$ telle que $\pi^E$ est optimale pour $R$
-%% \item Mentionner la nature mal posée du problème sans s'appesentir dessus. Préciser toutefois qu'aucune approche ne le résoud de manière mathématiquement satisfaisante à cause de ça.
-%% \item insister sur l'optimalité de l'expert comme hypothèse. Un expert sous optimal, c'est un autre champ de littérature, éventuellement donner deux trois refs.
-%% \end{itemize}
-%% \end{itemize}
 First, we introduce some general notations.
@@
 We note $\Delta_X$ the set of distributions over $X$.
-Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\leq\beta \Leftrightarrow \forall x\in X, \alpha(x) \leq \beta(x)$.
+Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\geq\beta \Leftrightarrow \forall x\in X, \alpha(x) \geq \beta(x)$.
 We will often slightly abuse the notations and consider (where applicable) most objects as if they were matrices and vectors indexed by the set they operate upon.
 
-We work with finite \emph{Markov Decision Processes} (MDPs) \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$.
+We work with finite MDPs \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$.
 The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP.
@@
 The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
+The stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
+\begin{equation}
+  \rho_\pi^TP_\pi = \rho_\pi^T,
+  \end{equation}
+where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$. The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.
+
 The reward function $R$ is a local measure of the quality of the control.
-The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s]$.
-Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$.
-The stationary distribution induced by a policy $\pi$ is written $\rho_\pi,\textrm{ s.t.
-}\rho_\pi^TP^\pi = \rho_\pi^T$.
+The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the discounted cumulative reward for following policy $\pi$ from this state:
+\begin{equation}
+V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s].
+\end{equation}
+This long-term criterion is what is being optimized when solving an MDP.
+Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$..
 
@@
   \end{equation}
-where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$ and $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$.
-The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$.
-Both operators are contractions.
-The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$, the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$.
-Another object object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$) and $R_a$ = $(R(s,a))_{s\in S}$.
-Once expanded, this gives what we refer to as the Bellman optimality equation :
+where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally
 \begin{equation}
-  Q^*_R = R_a + \gamma P_{\pi^*_R}V^*_R
+  T^*_RV = \max_\pi T^\pi_RV.
+\end{equation}
+  Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R :
+  \begin{eqnarray}
+    V^\pi_R &=& T^\pi_R V^\pi_R\\
+    V^\pi_R &=& R_\pi + \gamma P_\pi V^\pi_R.
+  \end{eqnarray}
+The Bellman optimality operator $T^*_R$ also admits a fixed point, the optimal value function $V_R^*$ with respect to reward $R$.
+
+Another object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$ and $R_a$ = $(R(s,a))_{s\in S}$).
+The value function $V^\pi_R$ and the action-value function $Q^\pi_R$ are quite directly related : $\forall s \in S, V^\pi_R(s) = Q^\pi_R(s,\pi(s))$.
+The Bellman evaluation equation for $Q^\pi_R$ is therefore :
+\begin{equation}
+  Q^\pi_R(s,a) = R(s,a) + \gamma \sum_{s'\in S}P(s'|s,a) Q(s',\pi(s')).
   \label{eq:bellman1}
@@
 
-An optimal policy follows a greedy mechanism with respect to its optimal action-value function 
+An optimal policy follows a greedy mechanism with respect to its optimal action-value function $Q^*_R$
 \begin{equation}
@@
 
-When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
-We call this the direct problem.
+When the state space is too large to allow matricial representation or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
 
-In this work, we wish to tackle the inverse problem.
-We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal.
-The suboptimality of the expert is an interesting setting, that has been discussed for example in \cite{melo2010analysis,syed2012reduction}, but that we are not addressing here due to the numerous complications it introduces.
-We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal.
+We recall that solving the MDP is the direct problem.
+This contribution aims at solving the inverse one.
+We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi_E$, assuming that there exists some unknown reward $R_E$ for which the expert is optimal.
+The suboptimality of the expert is an interesting setting that has been discussed for example in \cite{melo2010analysis,syed2010reduction}, but that we are not addressing here.
+We do not try to find this unknown reward $R_E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal.
 The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal).
@@
 \label{sec:algo}
-%% \begin{itemize}
-%% \item Mentionner (sans en parler, ce sera fait dans \hyperref[sec-7-1]{Imitation}) l'imitation.
-Une approche supervisée évidente est la classification, avec pour entrée un set $D_C$ de données issues de l'expert.
-Rappeler (expliqué pour la première fois dans \hyperref[sec-3]{Formalisme} qu'on se restreint (sans perte de généralité, le préciser) au cas des politique déterministes.
-%% \item Tous les classifieurs utilisent une fonction de score (je n'ai pas trouvé un seul exemple d'un classifieur qui n'en utilise pas), il s'agit juste d'un manière d'écrire les choses.
-%% \item Expliquer que l'on identifie le mécanisme de classification au mécanisme glouton qui régit la politique optimale (donc celle de l'expert).
-%% \item Sans chercher à justifier mathématiquement la chose (ce sera fait dans la partie \hyperref[sec-5]{Analyse}), expliquer qu'intuitivement l'on va identifier fonction de score du classifieur et fonction de qualité de l'expert (qui est optimale pour une certaine récompense, rappelons-le).
-%% \begin{itemize}
-%% \item Repartir de l'équation d'évaluation de Bellman pour isoler le terme $R(s,a)$ et introduire $r:S\times A\in \mathbb{R}$, la récompense issue de cette équation lorsque l'on remplace $Q^E$ par la fonction de score : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi(s')) $
-%% \item On estime ce terme à partir des transitions $D_R$ car les probabilités de transitions sont inconnues : $\hat r(s,a) = q(s,a)-\gamma \sum_{s'} {1\over N_{s'}}\sum_{(s,a,s',a')\in D_R}q(s',a')$
-%% \item La fonction de récompense est donnée par une régression sur $D_R \cup {\hat r}$.
-%% \end{itemize}
-%% \item Donner l'algorithme l'environnement \texttt{algorithm2e}
-%% \item Rapidement, l'étape de régression permet d'ajouter la structure du MDP au résultat de classif, en utilisant l'équation de Bellman.
-%% \item Bien préciser qu'il ne s'agit ici que d'une explication qualitative, ``avec les mains'', la preuve du fait que ça résoud bien le problème est donnée dans \hyperref[sec-5]{Analyse}, une démonstration sur des problèmes si complexes que l'état de l'art ne peut les résoudre (sauf SCIRL) est donnée dans \hyperref[Experience]{Experience}.
-Les avantages comparés sont expliqués dans \hyperref[Autres-Approches]{Autres Approches}.
-Enfin, des explications plus intuitives sur les mécanismes internes de l'algorithme, ainsi que des résultats expérimentaux supplémentaires sont fournis en \hyperref[sec-9]{Annexes}.
-%% \end{itemize}
-Our first step in the quest for a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (MC$^2$).
+Our first step towards a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (SFMC$^2$ for short).
 This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$.
-The classification rule $\pi_C\in A^S$ is simply to select (one of) the label(s) that achieves the highest score for the given inputs :
+The classification rule $\pi_C\in A^S$ simply selects (one of) the label(s) that achieves the highest score for the given inputs:
 \begin{equation}
@@
 \end{equation}
-For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as MC$^2$ algorithms, we however used a structured margin approach.
-We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in Subsec.~\ref{subsec:instanciation}.
+For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} which we consider in the experimental setting.
+Other algorithms may be envisionned (see Subsec.~\ref{subsec:instanciation}).
 
@@
 One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal action-value function.
-As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$ (which can itself be seen as a good approximation of $\pi_E$).
-By inversing the Bellman equation, one can express the reward as a function of the action-value function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$.
-This leave us with $R^C$, the reward function relative to our score/action-value function $q$ :
+It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$.
+By inversing the Bellman equation Eq.~\ref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$ :
 \begin{equation}
+  \label{eq:rc}
   R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
 \end{equation}
-We wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown.
-For this we use another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to.
-Action $a_j$ need not be chosen by any particular policy.
+As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system.
+We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to.
+Action $a_j$ needs not be chosen by any particular policy.
 The dataset $D_R$ brings us information about the dynamics of the system.
-Our reward function $R^C$ must be approximated with the help of the information in $D_R$.
-This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor (a simple least-square approximator can do) to obtain $\hat R^C$, a generalisation of $\{(\hat r_j)_j\}$ over the whole state-action space.
-The complete algorithm is given Alg.~\ref{algo:cascading}.
+From it, we construct datapoints
+\begin{equation}
+  \label{eq:rj}
+  \{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j.
+\end{equation}
+As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$.
+A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space.
+The complete algorithm is given in Alg.~\ref{algo:cascading}.
 \begin{algorithm}%[H]
     %\small
-  \caption{CSA-IRL algorithm}
+  \caption{CSI algorithm}
   \label{algo:cascading}
@@
   \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
-  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{j},a_{j},\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
+  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
   \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
@@
 Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem.
-Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with which state), but there is more to a MDP than a single policy.
-Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probabilities $P$.
-The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.
+Indeed, mere classification only takes into account information from the expert (i.e., which action goes with which state) whereas using the Bellman equation in the expression of $\hat r_j$ makes use of the information lying in the transitions $(s_j,a_j,s'_j)$, namely information about the transition probabilities $P$.
+The final regression step is just a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.
 
-As the analysis (Sec.~\ref{sec:analysis}) reveals, both classification error and regression error should be controlled.
-By carefully choosing or tuning both supervised learning algorithms, it should be possible to mitigate any ill effects of the specific settings the cascading algorithm is used in.
-If one operates in a well-known domain where a lot of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings.
-If data is lacking, one could for example have to overcome bias, variance or generalization problems in the regression step (more about that in Subsec.\ref{subsec:instanciation}).
-The possibility to use and leverage the wide range of techniques developped for these two well-known supervised problems is a strong advantage of the proposed cascading approach.
-
-The explanation in this section is only a quatitative attempt to make the reader understand the idea behind the algorithm.
-The analysis of the next section tackles things in a formal, yet less intuitive, way.
+Being able to alleviate the ill effects of scalability or data scarcity by leveraging the wide range of techniques developped for the classification and regression problems is a strong advantage of the CSI approach.
 
@@
 
-%% \begin{itemize}
-%% \item On rappelle le problème posé au début.
-%% \item Le théorème 1 montre que modulo un bonne classif et un bon régresseur, on résoud exactement le problème que l'on s'était posé
-%% \item On en fait la démo rapide (éventuellement des bouts peuvent être déportés en \st{Pologne} \hyperref[sec-9]{Annexes}).
-%% \item On insiste sur le fait que la récompense trouvée n'est pas triviale, on a transcendé la nature mal posée du problème.
-%% \item On dit un mot rapide sur la complexité en échantillon (ça dépend du classifieur et du régresseur, mais \hyperref[sec-6]{Expériences} montre qu'on est bon là dessus)
-%% \item ainsi que sur la complexité temporelle (ça dépend toujours, mais avec une M-SVM et un moinde carré ça prend O(quelque chose) (indice: pas beaucoup)) FIXME:compléter la complexité
-%% \end{itemize}
-
-In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs, more formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]$ is small by bounding it by a term that depends on :
+In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
+More formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on :
 \begin{itemize}
-  \item the regression and classification errors, respectively $\epsilon_R$ and $\epsilon_C$, these terms are small if the regressor and the classifier are good;
-  \item $C_*$, a concentration coefficient assessing the closeness of the stationary distribution of $\pi_E$ and $\hat \pi_C$, the latter being the optimal policy for the reward $\hat R^C$ output by the algorithm ;
-  \item $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$, which could be normalized to $1$ without loss of generality.
+  \item the classification error
+\begin{equation}  
+\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}];
+\end{equation}
+\item the regression error
+  \begin{equation}
+    \epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E},
+  \end{equation}
+  with :
+  \begin{itemize}
+  \item the subscript notation already used in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are repectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
+  \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
+  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm : $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$ ;
+  \end{itemize}
+\item the concentration coefficient $C_* = C_{\hat \pi_C}$ with :
+  \begin{itemize}
+  \item $C_{\pi}=(1-\gamma)\sum_{t\geq0}\gamma^tc_{\pi}(t), \text{ with } c_{\pi}(t)=\max_{s\in S}\frac{(\rho_E^TP^t_\pi)(s)}{\rho_E(s)}$ ;
+  \item $\hat \pi_C$, the optimal policy for the reward $\hat R^C$ output by the algorithm ;
+  \end{itemize}
+  The constant $C_*$ could be estimated \emph{a posteriori} ( after $\hat{R}^C$ is computed).
+We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
+\item $\Delta q$, defined by 
+  \begin{eqnarray}
+    \Delta q &= &\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))\\
+    &=&\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)),
+    \end{eqnarray} which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
   \end{itemize}
@@
 \begin{theorem}
-\label{theorem: theorem analysis}
+\label{thm}
 Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward the cascading algorithm outputs. We have:
@@
 \begin{proof}
-First let's introduce some notations.
-$q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\forall s \in S,\pi_C\in\argmax_{a\in A}q(s,a)$ is a deterministic classifier policy.
-We recall that $\epsilon_C$ is the classification error:
-\begin{equation}
-\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
-\end{equation}
-Moreover we recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
+First let's introduce some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$ is a deterministic classifier policy.
+We recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
 \begin{equation}
@@
 
-The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$, $\epsilon^R\in\mathbb{R}^{S\times A}$ must not be mistaken for the regression error $\epsilon_R\in\mathbb{R}$ that appears in the final result.
+The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$.
 We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof:
@@
 \end{equation}
-Let $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$, we introduce $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$.
-Thus we have the following vectorial equations:
+
+We can now introduce the following vectorial equations :
 \begin{align}
@@
 \\
-&R^E_a=q_a+\gamma P_aq_{\pi_E},
+&R^E_a=q_a-\gamma P_aq_{\pi_E},
 \\
@@
 \end{align}
-Now we are going to upper bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious as $V^*$ is optimal).
-We note $\hat{\pi}_C$ a deterministic optimal policy of the reward $\hat{R}^C$. First, the term $V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}$ is decomposed:
+Now, we are going to upper bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious as $V^*$ is optimal).
+Recall that $\hat{\pi}_C$ is a deterministic optimal policy of the reward $\hat{R}^C$. First, the term $V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}$ is decomposed:
 \begin{equation}
@@
 \end{equation}
-We are going to bound each of these three terms. But, first, let $\pi$ be a given deterministic policy. We have:
+We are going to bound each of these three terms. First, let $\pi$ be a given deterministic policy. We have:
 \begin{equation}
@@
 \end{equation}
-We use $\|.\|_{1,\mu}$ to denote the $\mu$-weighted $L_1$ norm : $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$. 
 If $\pi=\pi_E$, we have, thanks to the power series expression of $(I-\gamma P_{\pi_E})^{-1}$ and the definition of $\rho_E$:
@@
 
-If $\pi\neq\pi_E$, we introduce the following concentration coefficient:
+If $\pi\neq\pi_E$, we use the coefficient $C_\pi$. We have then:
 \begin{equation}
-C_{\pi}=(1-\gamma)\sum_{t\geq0}\gamma^tc_{\pi}(t), \text{ with } c_{\pi}(t)=\max_{s\in S}\frac{(\rho_E^TP^t_\pi)(s)}{\rho_E(s)}.
+  \rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})\leq\frac{C_\pi}{1-\gamma}\rho_E^T\epsilon^R_{\pi}\leq\frac{C_\pi}{1-\gamma}\|\epsilon^R_{\pi}\|_{1,\rho_E}.
   \end{equation}
-We have then: $\rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})\leq\frac{C_\pi}{1-\gamma}\rho_E^T\epsilon^R_{\pi}\leq\frac{C_\pi}{1-\gamma}\|\epsilon^R_{\pi}\|_{1,\rho_E}$.
-So, if we note, $C_*=C_{\hat{\pi}_C}$ and $\epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E}$, we are able to give an upper bound to the first and third terms:
+So, using the notations introduced before we stated the theorem, we are able to give an upper bound to the first and third terms:
 \begin{equation}
@@
 \end{equation}
-Now, there is still an upper bound to find for the second term. It is possible to decompose the second term as follows:
+Now, there is still an upper bound to find for the second term. It is possible to decompose it as follows:
 \begin{equation}
@@
 \end{eqnarray}
-We recall that $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$. We have:
+Using $\Delta q$, we have:
 \begin{equation}
@@
 
-The first remark one can make about this result is that the bound is tight if both the classifier and the regressor are perfect (i.e.
-if the associated error terms are null).
-In this case, the expert's policy  $\pi_E$ is the only deterministic optimal policy of $\hat{R}^C=R^C$ (the uniqueness implies that $\forall s\in S, \Card(\argmax_{a\in A}q(s,a))=1$, with $\Card(X)$ denoting the cardinality of the set $X$).
-This uniqueness also hints at the fact that we found a non-trivial reward (we recall that the null reward admit \emph{every} policy as optimal).
+\begin{corollary}
+  \label{cor}
+  If the bound of Theorem~\ref{thm} is tight (if and only if $\epsilon_C = 0$ and $\epsilon_R = 0$), $\pi_E$ is the only optimal policy for $\hat R^C$.
+\end{corollary}
+\begin{proof}
+  To get rid of technicalities we assume that $\forall s \in S, \rho_E(s) \neq 0$. If not we create a new MDP $\{S',A,\gamma,P,\hat R^C\}$ with $S' = \{s \in S | \rho_E(s) \neq 0\}$ and work in it.
+
+  The function $q$ is the action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\ref{eq:rc}).
+
+  As $\epsilon_C = 0$, we have $\pi_C = \pi_E$.
+This means that $\forall s, \pi_E(s)$ is the only element of the set $\argmax_{a\in A}q(sa)$.
+Therefore, $\pi_C = \pi_E$ is the unique optimal policy for $R^C$.
+
+  As $\epsilon_R = 0$, we have $\hat R^C = R^C$, hence the result.
+\end{proof}
+The uniqueness of $\pi_E$ as an optimal policy for $\hat R^C$ in the conditions of corollary~\ref{cor} hints at the fact that we found a non-trivial reward (we recall that the null reward admits \emph{every} policy as optimal).
 Therefore, obtaining  $\hat{R}_C = 0$ (for which the bound is obviously true: the bounded term is $0$, the bounding term is positive) is unlikely as long as the classifier and the regressor exhibit decent performance.
 
-In order to control $\epsilon_R$, we need a precise regression over the expert stationary distribution and for all the actions.
-In theory, we need to be able to sample the whole dynamics (even the non expert actions).
-In practice heuristics can be used (see Sec.~\ref{sec:experiments}) however this heuristics is not justified by the analysis.
+The only constraints the bound of Thm.~\ref{thm} implies on datasets $D_R$ and $D_C$ is that they provide enough information to the supervised algorithms to keep both error terms $\epsilon_C$ and $\epsilon_R$ low.
+In Sec.~\ref{sec:experiments} we deal with a dire lack of data in dataset $D_R$.
+We address the problem with the use of heuristics (Subsubsec.~\ref{subsubsec:heuristics}) in order to show the behavior of the CSI algorithm in somewhat realistic conditions.
 
-As mentionned earlier, the algorithm and the analysis are invariant by  dilatation of $q$, hence of $\Delta q$.
-We could normalize $\Delta q$ by imposing $\Delta q=1$ (the range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$).
-
-The constant $C_*$ could be estimated a posteriori ( after $\hat{R}^C$ is computed).
-We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (approximately, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
-
-The bound can be fully controlled by the quality of the classification and regression subroutines.
+More generally, the error terms  $\epsilon_C$ and $\epsilon_R$ can be reduced by a wise choice for the classification and regression subroutines.
 The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off etc.) to be found.
@@
 
-\section{Related Work}
-\label{sec:related}
-IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
-The seminal paper \cite{abbeel2004apprenticeship} blurs the frontier between imitation and IRL, as this algorithm outputs (mixing) policies, and not explicitely a reward although IRL is the core of this method.
-This work introduced the general method of finding a reward function such that the corresponding optimal policy's feature expectation (or
-more generally some measure of the underlying trajectories'
-distribution) matches the one of the expert policy.
-By iteratively changing the considered reward function, one can get the feature expectation of the expert and of the optimal policy to match within a certain bound.
-This implies using RL as a subroutine to solve the direct problem at each iteration.
-Some of those approaches are summed up in \cite{neu2009training}.
-The need to solve a MDP at each step is very demanding, both sample-wise and computationally.
-The cascading approach being able to output a reward function without having to solve the MDP is thus a significant improvement.
-The Projection IRL (PIRL) algorithm of \cite{abbeel2004apprenticeship} is used as a benchmark in our experiments, see Subsec.~\ref{subsec:setting}.
-
-Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
-Classification on the same level as the first step of the CSA-IRL method (with actions as labels and states as inputs) can be found in \cite{ratliff2007imitation} with the help of a structured margin method, but it did not include the structure of the MDP as the CSA-IRL algorithm does.
-
-Using the non trivial notion of metric in an MDP, \cite{melo2010learning} builds a kernel which is used
-in a classification algorithm, showing improvements compared to a
-non-structured kernel. However, this approach is not an IRL
-algorithm, and more important assessing the metric of an MDP is a
-quite involved problem.
-
-Few algorithms do not require solving the direct RL problem.
-The approach of \cite{syed2008apprenticeship} requires knowing the transition probabilities $P$ of the MDP (which we don't need) and outputs a policy (and not a reward).
-The algorithm in \cite{dvijotham2010inverse} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
-Closer to our use-case is the idea presented in \cite{boularias2011relative} to use a subgradient ascent of a utility function based on the notion of relative entropy.
-Importance sampling is suggested as a way to avoid solving the MDP.
-This requires sampling trajectories according to a non-expert policy and
-the direct problem remains at the core of the approach (even if
-solving it is avoided). As our experiments show, we are able to tackle the IRL problem with only samples from the expert.
-
-Finally, the SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} is able to solve the IRL problem in the same dire conditions the cascading approach is tested in.
-It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation as feature function.
-This necessity of a linarly parametrized approach makes the SCIRL algorithm less convenient than CSA-IRL, which can make use on non-parametrized supervised subroutines.
-SCIRL is used for comparison purposes on the Highway driving problem.
-
-%% \begin{itemize}
-%% \item On reprend peu ou prou la même que SCIRL et que celle du précédent papier. On profite de la mention d'une approche pour préciser en quoi la nôtre diffère.
-%% \end{itemize}
-%% \subsection{Imitation}
-%% \label{sec-7-1}
-
-%% \begin{itemize}
-%% \item Ce n'est pas le problème que l'on se pose, mais résoudre l'IRL peut permettre une forme fine d'imitation.
-%% \item La frontière est un peu floue, exemple de PIRL
-%% \item On peut considérer que tous les algos revus par Neu sont des algos d'imitation dans le sens où ils cherchent une proximité des $\mu$.
-%% \item Coincidentallement, ils ont besoin de résoudre le problème direct, pas nous.
-%% \end{itemize}
-%% \subsection{Méthodes entropiques}
-%% \label{sec-7-2}
-
-%% \begin{itemize}
-%% \item N'ont pas besoin de résoudre le problème direct, mais ont besoin d'autres données que celles de l'expert.
-%% \end{itemize}
-%% \subsection{Autres méthodes}
-%% \label{sec-7-3}
-
-%% \begin{itemize}
-%% \item parler des MDPs solvables linéairement, des papiers mentionnés par les reviewers de NIPS, et dire du bien de SCIRL
-%% \end{itemize}
+%% Feel good comment
 \section{Experiments}
@@
 In this section, we put our approach under stress to empirically demonstrate its behavior.
-After explaining the criterion we consider and its rationale (Subsec.~\ref{subsec:criterion}), we describe the benchmark's setting along with the other algorithms we use for comparison purposes (Subsec.~\ref{subsec:setting}).
-Before giving and anlyzing the results (Subsec.~\ref{subsec:results}), we explain the details of our choices for the classification and regression subroutines and give some hints for other possibilities (Subsec.~\ref{subsec:instanciation}).
-We also give the details of our heuristics.
+We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data is given on Subsubsec.~\ref{subsubsec:data} and a precise description of the way we instanciated the CSI algorithm is given on subsubsections \ref{subsubsec:subroutines} and \ref{subsubsec:heuristics}.
 
-Finally, Subsec.~\ref{subsec:IP} provides some results on a continuous state space setting (the inverted pendulum) in order to provide the reader with a practical illustration of the workings of our algorithm.
-There is no comparison with other methods, these results are provided to help the reader form an intuition about the way the CSA-IRL algorithm works.
+We then use a classical RL benchmark, the Inverted Pendulum (Subsec.~\ref{subsec:IP}), to provide a qualitative illustration of the proposed CSI approach.
 
-\subsection{Criterion}
-\label{subsec:criterion}
-The theoretical bound provided in Sec.~\ref{sec:analysis} is about the optimality of the expert with respect to the reward function computed by our algorithm (this is also the way the problem was framed in Sec.~\ref{sec:background}).
-For the experiments, we use a somewhat more restrictive criterion : a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the "true" reward function, the reward function of the expert.
-We compare the mean (over the uniform distribution) value function with respect to $R_E$ of a policy $\pi$ learned by optimizing $\hat R^C$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
-We do that firstly because, for once, we can : the bound we provide in the analysis is something that can be measured even if $R_E$ does not actually exist.
-We make use here of the fact that $R_E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R_E$.
-The second reason for this different metric is that we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well-behaved given the use of heuristics.
-\subsection{Setting}
-\label{subsec:setting}
-The setting of the experiment is a driving simulator inspired from a benchmark already used in \cite{syed2008apprenticeship,syed2008game}.
-The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
-At all times, there will be one car in one of the three lanes.
-Even at the lowest speed, the player's car moves faster than the others.
-When the other car disappears at the bottom of the screen, another one appear at the top in a randomly chosen lane.
-It takes two transitions to completely change lanes, as the player can move left or right for half a lane's length at a time.
-At the highest speed setting, if the other car appear in the lane the player is in, it is not possible to avoid the collision.
-The main difference between the original benchmark and ours is that we made the MDP more ergodic by allowing the player to change speed whenever he wishes so, not just during the first transition.
-If anything, by adding two actions, we enlarged the state-action space and thus made the problem harder.
-%FIXME:METTRE UNE CAPTURE DECRAN.
-The reward function $R_E$ trains the expert to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
-Any other situation receives a null reward.
-
-We need some baseline to compare our new approach to.
-The natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
-Another natural benchmark we use is to compute the mean value function of the classifier policy.
-We also compare the proposed cascading method to the PIRL algorithm of \cite{abbeel2004apprenticeship} and the SCIRL algorithm of \cite{klein2012scirl}.
-The random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
-PIRL is instanciated in a way that make use of the transition probability matrix $P$ to exactly compute all relevant objects (feature expectations, optimal policies) via exact dynamic programming.
-It is left to run for 70 iterations.
-We could have displayed PIRL's data sensitivity by making $\pi_E$ known only through observations, but we decided against it because we want this baseline to show the performance of a fully informed well-established appreticeship algorithm.
-
-Both datasets $D_C$ and $D_R$ will be constructed from samples drawn with the expert's policy.
-The expert drives the car in the simulator for $n$ runs of $n$ transitions each, with $n$ varying in $\{3,7,10,15,20\}$.
-Apart from SCIRL, to the best of our knowledge, no other algorithm can infer a reward function only from expert data.
-Recall from Sec.~\ref{sec:related} that most existing approaches either need to solve the direct RL problem repeatedly (which can not be done on this setting from expert data alone) or rely on a sampling of the dynamics of the whole MDP.
-The reward functions found by SCIRL and CSA-IRL are then optimized using a dynamic programming algorithm to show the performance of the associated policy.
-
+We finally provide quantitative results and comparison with state-of-the-art approaches on a Highway driving simulator (Subsec.~\ref{subsec:highway}).
 \subsection{Instanciation}
 \label{subsec:instanciation}
+\subsubsection{Data scarcity}
+\label{subsubsec:data}
+In order to create a somewhat realistic environment, we decided that the CSI algorithm could neither use the transition probabilities nor a simulator.
+The only data it could access is sampled according to the expert policy.
+This corresponds for example to a situation where a costly system can only be controlled by a trained operator as a bad control sequence could lead to a system breakage.
+
+More precisely, the expert controls the system for $n$ runs of length $n$, with values of $n$ in $\{3,7,10,15,20\}$, giving $n^2$ samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$.
+The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms :
+\begin{equation}
+  D_C = \{(s_i = s_k,a_i=a_k)_i\}.
+\end{equation}
+There is no particuliar problem for the classification step.
+It is however not reasonable to construct the dataset $D_R$ only from expert transitions and expect a small regression error term $\epsilon_R$.
+Indeed, the dataset $D_E$ only sample the dynamics induced by the expert's policy and not the whole dynamics of the MDP.
+The regression step also needs samples $((s_j,a_j),\hat r_j)_j$ drawn from non expert actions.
+We address this problem by using a heuristics.
+\subsubsection{Heuristics}
+\label{subsubsec:heuristics}
+From the raw expert dataset $D_E$ we can construct samples $((s_j=s_k,a_j=a_k),\hat r_j)$ with $\hat r_j$ as specified in Eq.~\ref{eq:rj}.
+To compensate for the lack of information about the dynamics of the whole MDP, we assert that disagreeing with the expert's choice is undesirable.
+Mathematically, we do that by artificially introducting samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
+\begin{equation}
+  r_{min} = \min_j\hat r_j - 1.
+\end{equation}
+Although this heuristics was not analyzed in Sec.~ \ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
+\subsubsection{Subroutines choices}
+\label{subsubsec:subroutines}
 The cascading algorithm can be instanciated with some standard classification algorithms and any regression algorithm.
@@
 \end{eqnarray*}
-with $l(s_i,a) = 1\textrm{ if } a\neq a_i,l(s_i,a_i)=0$ and the natural tabular features as the $\phi$ vector ($0$ everywhere but on the unique index relative to the argument, where the component in $1$).
+with $l(s_i,a) = 1\textrm{ if } a\neq a_i,l(s_i,a_i)=0$.
 The score function of this classifier is simply $q(s,a) = \theta^\top \phi(s,a)$.
-This is the classifier our implementation of SCIRL \cite{klein2012scirl} is based on.
-We chose to use it in CSA-IRL to provide a fair comparison between both algorithms.
+The implementation of SCIRL \cite{klein2012scirl} is based on this.
+We chose to use it in our CSI implementation to provide a fair comparison between both algorithms.
 
 Ease of implementation leads us to choose a simple least-square regressor for our experiments.
-Using the same features $\phi$ as our classifier, it can be written (using straightforward matricial notations) as :  $\omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R$.
-The reward function is : $\hat R^C(s,a) = \omega^T \phi(s,a)$.
+Using the same features $\phi$ as our classifier, we introduce $\lambda$ a regularization coefficient (set to $0.1$ in our experiments), $Id$ the identity matrix and
+\begin{eqnarray}
+  \Phi = \begin{bmatrix}
+    \vdots\\
+    \phi(s_j,a_j)\\
+    \vdots
+  \end{bmatrix},\\
+  \hat R = \begin{bmatrix}
+    \vdots\\
+    \hat r_j\\
+    \vdots
+  \end{bmatrix}.
+\end{eqnarray}
+
+Our regressor can be written:
+\begin{equation}
+  \omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R.
+\end{equation}
+The reward function is : $\hat R^C(s,a) = \omega^T \phi(s,a)$
 
@@
 
-As we only had transitions $\{s_j,\pi_E(s_j),s'_j\}$from the expert to construct the samples $\{s_j,a_j,r_j\}$ fed to the regressor, we had to use a heuristics.
-We assert that disagreeing with the expert's choice is undesirable.
-Mathematically, we do that by artificially introducting samples $\{s_j,a,r_{min}\}_{a\neq \pi_E(s_j)}$ where $r_{min} = \min_j\hat r_j - 1$.
-Although this heuristics was not analyzed in Sec.~ \ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown here demonstrate its soundness.
-\subsection{Results}
-\label{subsec:results}
+Our complete instanciation of CSI is summed up in Alg.~\ref{alg:CSIimp}.
+\begin{algorithm}%[H]
+    %\small
+  \caption{A CSI instanciation}
+  \label{alg:CSIimp}
+  \emph{\textbf{Given}} a dataset  $D_E = (s_k,a_k=\pi_E(a_k),s'_k)_k$\;\\
+  \emph{\textbf{Given}} a feature function $\phi: S\times A \rightarrow \mathbb{R}^p$\;\\
+  \emph{\textbf{Construct}} the dataset $D_C=\{(s_i=s_k,a_i=\pi^E(s_i))=a_k\}$ \;\\
+  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$ by solving \begin{eqnarray*}
+  &\min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
+  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \\
+  &\text{~s.t.~} \forall i,
+  \theta^\top{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
+  {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
+\end{eqnarray*}\;\\
+\emph{\textbf{Construct}} the dataset $\{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}$ with $\hat r_j = q(s_j,a_j) - \gamma q(s'_j=s'_k,\pi_C(s'_j=s'_k))$\;\\
+\emph{\textbf{Set}} $r_{min} = \min_j\hat r_j - 1.$\;\\
+\emph{\textbf{Construct}} the training set  $D_R = \{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}\cup\{((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}\}$ \;\\
+\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$ by computing :
+\begin{equation*}
+  \omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R.
+\end{equation*}\;\\
+\emph{\textbf{Output}} the reward function $\hat R^{C}: (s,a) \mapsto \omega^T \phi(s,a)$ \;
+\end{algorithm}
+
+\subsection{Inverted Pendulum}
+\label{subsec:IP}
+We provide here qualitative results on the inverted pendulum problem.
+Quantitative results obtained by following a procedure very similar to Subsec.~\ref{subsec:highway} were of the same nature and justified the same analysis as those provided in Subsubsec.~\ref{subsubsec:hresults}.
+We thus omit them here for the sake of avoiding redundancy.
+Instead, we use the inverted pendulum problem for illustrative purposes.
+\subsubsection{Setting}
+The inverted pendulum (also known as the cart pole) is a classical benchmark in the RL community, it is fully described in \cite{lagoudakis2003least}.
+The goal is to maintain a pole in a unstable equilibrium position with the help of a cart moving along one axis.
+The state space consists in the angular position an speed of the pole.
+It is possible to use an RL algorithm to learn how to balance the pendule in an upright position by using the following reward: as long as the angular position of the pole is (in absolute value) less than $\pi \over 2$ the reward is $0$, the reward is $-1$ if the pole falls ; this reward function is plotted on Fig.~\ref{toto}.
+Three actions are available to the agent: it can choose to apply a leftward or rightward force on the cart or no force at all.
+
+The fact that the state space is continuous is not a problem to our particular instanciation of the CSI approach (see Alg.~\ref{alg:CSIimp}) as long as we provide a feature function $\phi : S\times A \rightarrow \mathbb{R}^p$.
+Here $p=30$ as we use the exact same feature functions as proposed in \cite{lagoudakis2003least}.
+The state space is featured with a $3\times 3$ grid of Gaussian functions concatenated with a constant component to construct a feature vector $\psi:S \rightarrow \mathbb{R}^{10}$.
+The eature function $\phi$ is defined as :
+\begin{equation}
+  \phi(s,a) = \begin{pmatrix}
+     \delta_{a = LEFT}\psi(s)\\
+     \delta_{a = NONE}\psi(s)\\
+     \delta_{a = RIGHT}\psi(s)
+  \end{pmatrix}
+\end{equation}
+
+\subsubsection{Results}
+The results are shown Fig.~\ref{fig:IP}.
+In the left column, we show the reward and value functions of the expert.
+The expert (a RL agent trained on the  reward shown on Fig.~\ref{toto}) is able to balance the pendule indefinitely.
+With as little as 300 transitions (30 seconds of balancing), CSI is able to find a reward that, once optimized by an RL agent, will reliably lead to a control as good as the expert's.
+As Fig.~\ref{tata} illustrates, the data we get from the expert only covers a tiny fraction of the state space.
+Furthermore, the feature function $\phi$ did not permit to accurately approximate the true reward\footnote{The true reward function is a function in $\mathbb{R}^S$, CSI outputs a reward in $\mathbb{R}^{S\times A}$, but we plot here $R(s,\pi^*(s))$ with $\pi^*$ the optimal policy for $R$.}.
+Despite these two setbacks, CSI manage to find a reward for which not only the expert's control is near-optimal (that is what the analysis of Sec.~\ref{sec:analysis} garantees under the assumption that a reasonable amount of data is available in order to keep both classification and regression errors down) but that also lead to a control policy as good as the expert's \emph{with respect to the (theoretically) unknown expert's reward function}.
+
+\begin{figure}
+   \subfigure[Reward on which the expert was trained]{\label{toto}\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_true_R"}\end{minipage}}
+   \subfigure[Reward found by CSA-IRL.
+The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\begin{minipage}{.45\linewidth}\hspace{-3em}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_lafem_R"}\end{minipage}}
+  \subfigure[Value function of the expert (i.e.
+optimal value function for the reward displayed Fig.~\ref{toto})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vexpert"}\end{minipage}}
+  \subfigure[Optimal value function for the reward function found by CSA-IRL (Fig.~\ref{tata})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vagent"}\end{minipage}}
+
+  \caption{Results on the inverted pendulum benchmark, illsutrating the dissimilarity of the rewards, but the similarity of the value functions.
+The state space is represented on a position-speed plan.}
+  \label{fig:IP}
+\end{figure}
+
+\subsection{Highway driving simulator}
+\label{subsec:highway}
+Let's now move from a qualitative illustration to a quantitative assessment of the performance of the CSI algorithm.
+The Highway driving problem is described Subsubsec.~\ref{subsubsec:hsetting}, the criterion we use is explained Subsubsec.~\ref{subsubsec:hcriterion} and finally results and comparison to other approaches are presented Subsubsec.~\ref{subsubsec:hresults}.
+\subsubsection{Setting}
+\label{subsubsec:hsetting}
+The setting of the experiment is a driving simulator inspired from a benchmark already used in \cite{syed2008apprenticeship,syed2008game}.
+The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
+At all times, there will be one car in one of the three lanes.
+Even at the lowest speed, the player's car moves faster than the others.
+When the other car disappears at the bottom of the screen, another one appear at the top in a randomly chosen lane.
+It takes two transitions to completely change lanes, as the player can move left or right for half a lane's length at a time.
+At the highest speed setting, if the other car appear in the lane the player is in, it is not possible to avoid the collision.
+The main difference between the original benchmark \cite{syed2008apprenticeship,syed2008game} and ours is that we made the MDP more ergodic by allowing the player to change speed whenever he wishes so, not just during the first transition.
+If anything, by adding two actions, we enlarged the state-action space and thus made the problem harder.
+%FIXME:METTRE UNE CAPTURE DECRAN.
+The reward function $R_E$ trains the expert to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
+Any other situation receives a null reward.
+
+The feature function $\phi$ for this setting is the natural tabular feature function where each state action couple $(s,a)$ is associated with a unique index $i_{(s,a)}$.
+The only non-zero component of $\phi(s,a)$ is then the component at index $i_{(s,a)}$.
+
+We need some baseline to compare our new approach to.
+The natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
+Another natural benchmark we use is to compute the mean value function of the classifier policy.
+We also compare the proposed cascading method to the PIRL algorithm of \cite{abbeel2004apprenticeship} and the SCIRL algorithm of \cite{klein2012scirl}.
+The random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
+PIRL is instanciated in a way that makes use of the transition probability matrix $P$ to exactly compute all relevant objects (feature expectations, optimal policies) via exact dynamic programming.
+It is left to run for 70 iterations.
+We could have displayed PIRL's data sensitivity by making $\pi_E$ known only through observations, but we decided against it because we want this baseline to show the performance of a fully informed well-established apprenticeship algorithm.
+
+Apart from SCIRL, to the best of our knowledge, no other algorithm can infer a reward function only from expert data.
+Recall from Sec.~\ref{sec:related} that most existing approaches either need to solve the direct RL problem repeatedly (which can not be done on this setting from expert data alone) or rely on a sampling of the dynamics of the whole MDP.
+The reward functions found by SCIRL and CSI are then optimized using a dynamic programming algorithm to show the performance of the associated policy.
+
+\subsubsection{Criterion}
+\label{subsubsec:hcriterion}
+The theoretical bound provided in Sec.~\ref{sec:analysis} is about the optimality of the expert with respect to the reward function computed by our algorithm (which is also the way the problem was framed in Sec.~\ref{sec:background}).
+For this experiment, we use a somewhat more restrictive criterion : a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the ``true'' reward function, the reward function of the expert.
+We use the mean (over the uniform distribution) value function with respect to $R_E$ of the policy $\hat \pi_C$ learned by optimizing $\hat R^C$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
+We do that firstly because, for once, we can : the bound we provide in the analysis is something that can be measured even if $R_E$ does not actually exist.
+We make use here of the fact that $R_E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R_E$.
+The second reason for this different metric is that we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well-behaved given the use of heuristics.
+\
+\subsubsection{Results}
+\label{subsubsec:hresults}
+Results are shown Fig.~\ref{fig:Highway}.
+We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL and CSI, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), the output of the PIRL algorithm, the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
+Performance for CSA-IRL is on par with the PIRL and SCIRL algorithms, slightly below the performance of the expert himself.
+Very few samples (100) are needed to reliably achieve expert-level performance.
+
+It is very interesting to compare our algorithm to the behavior of a classifier alone (respectively red and green plots on Fig.~\ref{tutu}).
+With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrates far better performance from the start.
+This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor and outputting not a policy, but a reward function that can be optimized on the MDP truly makes use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
+
+Furthermore, the classifier whose results are diplayed here is the output of the first step of the algorithm.
+The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound let us expect.
+
+PIRL is given the transition probabilities (that is, a $|S|\times |S|\times |A|$ matrix) and yet it does only slightly better than our approach with 100 samples.
+This also shows that CSI is indeed very sample efficient, without having to repeatedly solve the MDP.
+
+Finally, only the SCIRL algorithm operates on the same data as our proposed approach, with very slightly better results.
+
 \begin{figure}
@@
 PIRL uses the transition probabilities matrix $P$.]{\includegraphics[width=.5\linewidth]{"Fig1-NoX"}}
-   \subfigure[Cascading and Classification.]{\includegraphics[width=.5\linewidth]{"Fig2-ClassifAndUs"}}
+   \subfigure[Cascading and Classification.]{\label{tutu}\includegraphics[width=.5\linewidth]{"Fig2-ClassifAndUs"}}
   \subfigure[SCIRL and Cascading mean value.]{\includegraphics[width=.5\linewidth]{"Fig3-SCIRLAndUs"}}
@@
 
-Results are shown Fig.~\ref{fig:Highway}.
-We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL and CSA-IRL, the policy $\pi_C$ of the classifier (the very one the classification step of CSA-IRL outputs), the output of the PIRL algorithm, the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
-Performance for CSA-IRL is on par with the PIRL and SCIRL algorithms, slightly below the performance of the expert himself.
-Very few samples (100) are needed to reliably achieve expert-level performance.
+\section{Related Work}
+\label{sec:related}
+IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
+Approaches summarized in \cite{neu2009training} can be seen as iteratively contructing a reward function, solving an MDP at each iteration.
+Some of these algorithms are IRL algorithms while others fall in the AL category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship}\footnote{We call it PIRL for Projection IRL.} which we used in our benchmarks (Sec.~\ref{sec:experiments}).
+In both cases the need to solve an MDP at each step may be very demanding, both sample-wise and computationally.
+CSI being able to output a reward function without having to solve the MDP is thus a significant improvement.
 
-It is very interesting to compare our algorithm to the behavior of a classifier alone.
-With \emph{the exact same data}, albeit the use of a very simple heuristics, the cascading approach demonstrates far better performance from the start.
-This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor and outputting not a policy, but a reward function that can be optimized on the MDP truly makes use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
+AL via classification has been proposed in \cite{ratliff2007imitation} with the help of a structured margin method.
+Using the non trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
+in a classification algorithm, showing improvements compared to a
+non-structured kernel.
 
-Furthermore, the classifier whose results are diplayed here is the output of the first step of the algorithm.
-The classification performance is obviously not that good, this shows that our algorithm is empirically more forgiving of classification errors than our theoretical bound let us expect.
+Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
+The SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} which we used Subsec.~\ref{subsec:highway} is able to solve the IRL problem (without having to solve any MDP) in the same dire conditions the cascading approach has been tested in.
+It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation as feature function.
+This necessity of a linarly parametrized approach and the need to compute the expert's feature exepctations makes the SCIRL algorithm less convenient than the CSI approach.
 
-PIRL is given the transition probabilities (that is, a $|S|\times |S|\times |A|$ matrix) and yet it does only slightly better than our approach with 100 samples.
-This also shows that SCA-IRL is indeed very sample efficient, without having to repeatedly solve the MDP.
+Few IRL or AL algorithms do not require solving an MDP.
+The approach of \cite{syed2008apprenticeship} requires knowing the transition probabilities of the MDP (which CSI does not need) and outputs a policy (and not a reward).
+The algorithm in \cite{dvijotham2010inverse} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
+Closer to our use-case is the idea presented in \cite{boularias2011relative} to use a subgradient ascent of a utility function based on the notion of relative entropy.
+Importance sampling is suggested as a way to avoid solving the MDP.
+This requires sampling trajectories according to a non-expert policy and
+the direct problem remains at the core of the approach (even if
+solving it is avoided).
 
-Finally, only the SCIRL algorithm operates on the same data as our proposed approach, with very slightly better results.
-
-\subsection{Inverted Pendulum}
-\label{subsec:IP}
-Finally, we provide results on the inverted pendulum problem.
-We do not compare our approaches with others as we did on the highway driving problem.
-The information given here serves an illustrative purpose.
-
-The inverted pendulum (also known as the cart pole) is a classical benchmark in the RL community, it is fully described in \cite{lagoudakis2003least}.
-The goal is to maintain a pole in a unstable equilibrium position with the help of a cart moving along one axis.
-The state space is continuous, it consists in the angular position an speed of the pole.
-It is possible to use a RL algorithm to learn how to balance the pendule in an upright position by using the following reward : as long as the angular position of the pole is (in absolute value) less than $\pi \over 2$ the reward is $0$, the reward is $-1$ if the pole falls.
-Three actions are available to the agent: it can choose to apply a leftward or rightward force on the cart or no force at all.
-We used the feature parametrization of \cite{lagoudakis2003least}, that is a grid of 9 Gaussian functions and a constant.
-These are used to approximate the value function as well as for the regression step of the CSA-IRL algorithm.
-The instanciation of the algorithm is the same as for the highway driving problem except for the feature functions (which we just described).
-
-\begin{figure}
-   \subfigure[Reward on which the expert was trained]{\label{toto}\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_true_R"}}
-   \subfigure[Reward found by CSA-IRL. The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_lafem_R"}}
-  \subfigure[Value function of the expert (i.e. optimal value function for the reward displayed Fig.~\ref{toto})]{\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_Vexpert"}}
-  \subfigure[Optimal value function for the reward function found by CSA-IRL (Fig.~\ref{tata})]{\includegraphics[width=.5\linewidth]{"LAFEM_Exp3_Vagent"}}
-
-  \caption{Results on the inverted pendulum benchmark, illsutrating the dissimilarity of the rewards, but the similarity of the value functions.
-The state space is represented on a position-speed plan.}
-  \label{fig:IP}
-\end{figure}
-
-As with the previous benchmark, CSA-IRL only used data from the expert to devise a reward function.
-The results are shown Fig.~\ref{fig:IP}.
-In the left column, we show the reward and value functions of the expert.
-The expert (a RL agent trained on the specified reward) is able to balance the pendule indefinitely.
-With as little as 300 transitions (30 seconds of balancing), CSA-IRL is able to find a reward that, once optimized by a RL agent, will reliably lead to a control as good as the expert's.
-As Fig.~\ref{tata} illustrates, the data we get from the expert only covers a tiny fraction of the state space.
-Furthermore, the feature functions of our linear approximation scheme for the reward did not permit to accurately approximate the true reward\footnote{The true reward function is a function in $\mathbb{R^S}$, CSA-IRL outputs a reward in $\mathbb{R}^{S\times A}$, but we plot here $R(s,\pi^*(s))$ with $\pi^*$ the optimal policy for $R$.}.
-Despite these two setbacks, CSA-IRL manage to find a reward for which not only the expert's control is near-optimal (that is what the analysis of Sec.~\ref{sec:analysis} garantees under the assumption that a reasonable amount of data is available in order to keep both classification and regression errors down) but that also lead to a control policy as good as the expert's \emph{with respect to the (theoretically) unknown expert's reward function}.
-%% \begin{itemize}
-%% \item La borne fournie en \hyperref[sec-5]{Analyse} mesure le degré d'optimalité de la politique de l'expert vis à vis de la nouvelle récompense.
-Les expériences mesurent le degré d'optimalité de l'agent (ici entendu sous forme d'agent entraîné parfaitement sur la récompense sortie par notre algorithme) vis à vis de la vraie récompense (celle de l'expert).
-%% \begin{itemize}
-%% \item On se permet de faire ça car dans notre dispositif on la connait. Ce n'est pas le cas dans la vraie vie.
-%% \item En pratique c'est un critère plus restrictif : formellement, si $\pi^E$ est optimale pour $R_C$, alors l'agent et l'expert ont les mêmes performances (wrt $R_C$), puisque l'agent est optimal pour $R_C$ par définition de l'agent.
-En revanche on regarde si l'agent est optimal pour $R_E$, ce qui est profondément ``injuste'', puisqu'on ne sait rien de $R_E$.
-On ne sait même pas si elle existe réellement.
-%% \item Il s'agit d'une métrique qu'il n'est pas possible de mesurer sur un problème réel.
-C'est pourquoi la démonstration est intéressente car elle parle de quelque chose que l'on peut plus facilement évaluer ; et surtout elle parle de ce que l'on s'est proposé de résoudre à la base.
-Ces expériences ne sont qu'une illustration.
-%% \end{itemize}
-%% \item Les résultats obtenus sur ce problème sont notamment comparés à la classif, il s'agit d'une illustration du fait que l'algorithme est capable d'extraire les infos sur $P$ contenues dans les transitions, grâce à l'équation de Bellman qui est présente (modulo deux trois manips) dans l'expression de $\hat r$.
-Nous avons injecté la structure du MDP dans la classification (ça me rappelle quelque chose\ldots{} ;)
-%% \item Le seul plot est celui du Highway (similaire à SCIRL). On enlève le Gridworld (reporté en annexe car trop d'infos pour si peu de place)
-%% \item On instancie avec Taskar et un moindre carré. Ceci n'a pas la moindre importance.
-%% \item Mentionner le fait que les autres algos ne peuvent résoudre le problème tel qu'on le pose. On s'apesentira là dessus dans \hyperref[sec-7]{Autres approches}.
-%% \item Expliquer que quitte à tester expérimentallement, on a choisi de s'éloigner du cadre de l'analyse pour se placer dans cadre \emph{plus difficile à gérer}, ou seules les données de l'expert sont disponibles.
-%% \begin{itemize}
-%% \item Cela implique l'usage d'une heuristique pour la régression :
-%% \begin{itemize}
-%% \item sans heuristique, c'est le choix par défauit du régresseur (i.e. initialisation à 0, à R$_{\mathrm{min}}$ ou R$_{\mathrm{max}}$) qui décidera de l'optimisme ou pessimisme de l'algo
-%% \item Pour éviter ça, on place nous même notre heuristique.
-%% \end{itemize}
-%% \item que l'on se contente d'expliquer à un niveau intuitif.
-L'analyse formelle n'a pas été effectuée, ne le sera probablement pas, on se contente de montrer qu'en pratique, ça fonctionne.
-Et ça fonctionne mieux que les autres algos, sauf SCIRL.
-%% \item Cela implique aussi d'utiliser $\pi_C$ plutôt que $a'$.
-%% \end{itemize}
-%% \item Lien avec l'analyse : on a de bons résultats, même si le classifieur n'est pas bon. Notre borne est donc très très large.
-%% \item La complexité en échantillons : avec 50 samples, on fait mieux que PIRL qui a accès à une matrice de taille $|S|\times|S|\times |A|$
-%% \item La complexité en temps : on tourne 100 fois plus rapidement que PIRL
-%% \end{itemize}
 \section{Conclusion}
@@
 The expert is shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.
-A few practical examples of classifiers and regressors have been given, and one combination has been empirically shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
-On the chosen benchmark (a car driving simulator), our algorithm is shown to outperform other state of the art approaches, and to be on-par with the very recent SCIRL algorithm.
+Practical examples of classifiers and regressors have been given, and one combination has been empirically (on two classical benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
+On one of the benchmarks (a car driving simulator), our algorithm is shown to outperform other state of the art approaches, and to be on-par with the very recent SCIRL algorithm.
 We plan on deepening the analysis of the theoretical properties of our approach and on applying it to real world robotics problems.
-%% \begin{itemize}
-%% \item Ouverture : Existe-t-il des problèmes tels que SCIRL et Cascading diffèrent en performance ?
-%% \item Que faire quand on ne peut résoudre le RL facilement ?
-%% \item Que faire en cas d'expert sous optimal ?
-%% \item ouverture sur l'active learning en reprenant les notions de pessimisme ou d'optimisme en présence d'incertitude
-%% \end{itemize}
-
-%% \section{Annexes}
-%% \label{sec-9}
-%% \subsection{GridWorld}
-%% \label{sec-9-1}
-
-%% \begin{itemize}
-%% \item On montre avec les deux formes de récompenses qui donnent la même fonction de valeur.
-On fait référence aux travaux de Ng qui résolvent exactement ce problème.
-Les récompenses ont la même tête, ça devrait rassurer le lecteur hésitant.
-%% \item si on a le temps, on peut lancer une expérience où le nombre (en supposant une répartittion uniforme) des samples augmente et où la récompense tend vers ``la vraie''.
-%% \end{itemize}
-%% \subsection{Inverted pendulum}
-%% \label{sec-9-2}
-
-%% \begin{itemize}
-%% \item On montre qu'on généralise sans problème aux espaces continus, on montre qu'avec des données très, très sparses vis à vis de l'espace d'état on arrive pourtant à récupérer une récompense ayant la même fonction de valeur que celle de l'expert.
-%% \item ça montre aussi ce qui se passe quand l'espace d'hypothèse est ``mal choisi'' (i.e. la ``vraie'' récompense n'y était pas représentable).
-%% \end{itemize}
-%% \subsection{Highway}
-%% \label{sec-9-3}
-
-%% \begin{itemize}
-%% \item Si on a le temps : faire conduire quelqu'un puis lancer l'imitation, voir si on reproduit les consignes données à l'opérateur humain, qui aura eu la bonté de faire quelques erreurs dans l'exécution.
-Ca permet de voir le comportement en cas d'expert sous optimal et de taire les questions à ce sujet.
-%% \end{itemize}
 

### 2012-10-30 piot2012cascading/ECML13_cascading.tex
corrections de fomre mineures

@@
 This long-term criterion is what is being optimized when solving an MDP.
-Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$..
+Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states: $\forall \pi, V^*_R\geq V^\pi_R$..
 
@@
 \end{equation}
-  Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R :
+  Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R:
   \begin{eqnarray}
@@
 Another object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$ and $R_a$ = $(R(s,a))_{s\in S}$).
-The value function $V^\pi_R$ and the action-value function $Q^\pi_R$ are quite directly related : $\forall s \in S, V^\pi_R(s) = Q^\pi_R(s,\pi(s))$.
-The Bellman evaluation equation for $Q^\pi_R$ is therefore :
+The value function $V^\pi_R$ and the action-value function $Q^\pi_R$ are quite directly related: $\forall s \in S, V^\pi_R(s) = Q^\pi_R(s,\pi(s))$.
+The Bellman evaluation equation for $Q^\pi_R$ is therefore:
 \begin{equation}
@@
 What is of particular interest to us is the score function $q$ itself.
-One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal action-value function.
+One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function.
 It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$.
-By inversing the Bellman equation Eq.~\ref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$ :
+By inversing the Bellman equation Eq.~\ref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
 \begin{equation}
@@
 In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
-More formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on :
+More formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on:
 \begin{itemize}
@@
   \end{equation}
-  with :
+  with:
   \begin{itemize}
@@
   \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
-  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm : $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$ ;
+  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$ ;
   \end{itemize}
-\item the concentration coefficient $C_* = C_{\hat \pi_C}$ with :
+\item the concentration coefficient $C_* = C_{\hat \pi_C}$ with:
   \begin{itemize}
@@
 
-We can now introduce the following vectorial equations :
+We can now introduce the following vectorial equations:
 \begin{align}
@@
 More precisely, the expert controls the system for $n$ runs of length $n$, with values of $n$ in $\{3,7,10,15,20\}$, giving $n^2$ samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$.
-The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms :
+The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
 \begin{equation}
@@
 
-We referred in Sec.\ref{sec:algo} to \emph{score-function based multi-class classifiers} and explained how the classification rule is similar to the greedy mechanism that exists between an optimal action-value function and an optimal policy in a MDP.
+We referred in Sec.\ref{sec:algo} to \emph{score-function based multi-class classifiers} and explained how the classification rule is similar to the greedy mechanism that exists between an optimal action-value function and an optimal policy in an MDP.
 Most classification algorithm can be seen as such a classifier.
@@
 The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}).
-In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi : S\times A \rightarrow \mathbb{R}^p$ solves:
+In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi: S\times A \rightarrow \mathbb{R}^p$ solves:
 \begin{eqnarray*}
@@
 \end{equation}
-The reward function is : $\hat R^C(s,a) = \omega^T \phi(s,a)$
+The reward function is: $\hat R^C(s,a) = \omega^T \phi(s,a)$
 
@@
 \emph{\textbf{Construct}} the training set  $D_R = \{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}\cup\{((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}\}$ \;\\
-\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$ by computing :
+\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$ by computing:
 \begin{equation*}
@@
 
-The fact that the state space is continuous is not a problem to our particular instanciation of the CSI approach (see Alg.~\ref{alg:CSIimp}) as long as we provide a feature function $\phi : S\times A \rightarrow \mathbb{R}^p$.
+The fact that the state space is continuous is not a problem to our particular instanciation of the CSI approach (see Alg.~\ref{alg:CSIimp}) as long as we provide a feature function $\phi: S\times A \rightarrow \mathbb{R}^p$.
 Here $p=30$ as we use the exact same feature functions as proposed in \cite{lagoudakis2003least}.
 The state space is featured with a $3\times 3$ grid of Gaussian functions concatenated with a constant component to construct a feature vector $\psi:S \rightarrow \mathbb{R}^{10}$.
-The eature function $\phi$ is defined as :
+The eature function $\phi$ is defined as:
 \begin{equation}
@@
    \subfigure[Reward on which the expert was trained]{\label{toto}\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_true_R"}\end{minipage}}
-   \subfigure[Reward found by CSA-IRL.
+   \subfigure[Reward found by CSI.
 The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\begin{minipage}{.45\linewidth}\hspace{-3em}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_lafem_R"}\end{minipage}}
@@
 optimal value function for the reward displayed Fig.~\ref{toto})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vexpert"}\end{minipage}}
-  \subfigure[Optimal value function for the reward function found by CSA-IRL (Fig.~\ref{tata})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vagent"}\end{minipage}}
+  \subfigure[Optimal value function for the reward function found by CSI (Fig.~\ref{tata})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vagent"}\end{minipage}}
 
@@
 The theoretical bound provided in Sec.~\ref{sec:analysis} is about the optimality of the expert with respect to the reward function computed by our algorithm (which is also the way the problem was framed in Sec.~\ref{sec:background}).
-For this experiment, we use a somewhat more restrictive criterion : a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the ``true'' reward function, the reward function of the expert.
-We use the mean (over the uniform distribution) value function with respect to $R_E$ of the policy $\hat \pi_C$ learned by optimizing $\hat R^C$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
-We do that firstly because, for once, we can : the bound we provide in the analysis is something that can be measured even if $R_E$ does not actually exist.
+For this experiment, we use a somewhat more restrictive criterion: a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the ``true'' reward function, the reward function of the expert.
+We use the mean (over the uniform distribution) value function with respect to $R_E$ of the policy $\hat \pi_C$ learned by optimizing $\hat R^C$: $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
+We do that firstly because, for once, we can: the bound we provide in the analysis is something that can be measured even if $R_E$ does not actually exist.
 We make use here of the fact that $R_E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R_E$.
@@
 We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL and CSI, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), the output of the PIRL algorithm, the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
-Performance for CSA-IRL is on par with the PIRL and SCIRL algorithms, slightly below the performance of the expert himself.
+Performance for CSI is on par with the PIRL and SCIRL algorithms, slightly below the performance of the expert himself.
 Very few samples (100) are needed to reliably achieve expert-level performance.

### 2012-10-31 piot2012cascading/ECML13_cascading.tex
corrections de fomre mineures

@@
 \tolerance=1000
-\usepackage{amsmath}
+%\usepackage{amsmath}
 %\usepackage{amsthm}
@@
 
-Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy.
+Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\eqref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy.
 Thus, the demonstrated expert policy will be optimal for the related reward function.
@@
 What is of particular interest to us is the score function $q$ itself.
-One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function.
+One can easily notice the similarity between Eq.~\eqref{eq:greedy2} and Eq.~\eqref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function.
 It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$.
-By inversing the Bellman equation Eq.~\ref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
+By inversing the Bellman equation Eq.~\eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
 \begin{equation}
@@
 
-  The function $q$ is the action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\ref{eq:rc}).
+  The function $q$ is the action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\eqref{eq:rc}).
 
@@
 \label{subsubsec:heuristics}
-From the raw expert dataset $D_E$ we can construct samples $((s_j=s_k,a_j=a_k),\hat r_j)$ with $\hat r_j$ as specified in Eq.~\ref{eq:rj}.
+From the raw expert dataset $D_E$ we can construct samples $((s_j=s_k,a_j=a_k),\hat r_j)$ with $\hat r_j$ as specified in Eq.~\eqref{eq:rj}.
 To compensate for the lack of information about the dynamics of the whole MDP, we assert that disagreeing with the expert's choice is undesirable.

### 2012-11-04 piot2012cascading/ECML13_cascading.tex

@@
 \begin{document}
-\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
+\title{A cascaded supervised learning approach to inverse reinforcement learning}
 \author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
@@
 \begin{abstract}
-  This paper considers the Inverse Reinforcement Learning problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
+  This paper considers the Inverse Reinforcement Learning (IRL) problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
 We propose to break the IRL problem down into two generic Supervised Learning steps: this is the Cascaded Supervised IRL (CSI) approach.
-A classification step that defines a score function is followed by a regression step that outputs a reward function.
-An analysis of this approach shows that the demonstrated expert policy is near-optimal for the computed reward function.
+A classification step that defines a score function is followed by a regression step providing a reward function.
+A theoretical analysis shows that the demonstrated expert policy is near-optimal for the computed reward function.
 Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach.
 It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics.
-For this purpose two classical benchmarks (a highway driving simulator and a inverted pendulum simulator) are used.
+This is exemplified on two classical benchmarks (a highway driving simulator and an inverted pendulum simulator).
   \end{abstract}
@@
 \label{sec-2}
-The sequential decision making problem consists in taking the appropriate action pondering the available data in order to maximize a certain criterion.
-When framed in an MDP (see Sec.~\ref{sec:background}), the criterion to maximize is the expected sum of discounted rewards.
+Sequential decision making consists in choosing the appropriate action given the available data in order to maximize a certain criterion.
+When framed in a Markov Decision Process (MDP) (see Sec.~\ref{sec:background}), the criterion to maximize is the expected sum of discounted rewards.
 The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which we tackle here, aims at inferring a reward function for which a demonstrated expert policy is optimal.
@@
 This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-experts policies as inputs.
-Moreover, up to the use of some heuristics (see Subsubsec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-A specific instantiation of CSI (proposed Subsec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Subsec~\ref{subsec:IP}) and a car driving simulator (Subsec~\ref{subsec:Highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.
+Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
+A specific instantiation of CSI (proposed Sec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Sec.~\ref{subsec:IP}) and a car driving simulator (Sec.~\ref{subsec:Highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.
 
@@
 For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} which we consider in the experimental setting.
-Other algorithms may be envisionned (see Subsec.~\ref{subsec:instanciation}).
+Other algorithms may be envisionned (see Sec.~\ref{subsec:instanciation}).
 
@@
 In Sec.~\ref{sec:experiments} we deal with a dire lack of data in dataset $D_R$.
-We address the problem with the use of heuristics (Subsubsec.~\ref{subsubsec:heuristics}) in order to show the behavior of the CSI algorithm in somewhat realistic conditions.
+We address the problem with the use of heuristics (Sec.~\ref{subsubsec:heuristics}) in order to show the behavior of the CSI algorithm in somewhat realistic conditions.
 
@@
 In this section, we put our approach under stress to empirically demonstrate its behavior.
-We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data is given on Subsubsec.~\ref{subsubsec:data} and a precise description of the way we instanciated the CSI algorithm is given on subsubsections \ref{subsubsec:subroutines} and \ref{subsubsec:heuristics}.
+We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data is given on Sec.~\ref{subsubsec:data} and a precise description of the way we instanciated the CSI algorithm is given on Sec.~\ref{subsubsec:subroutines} and \ref{subsubsec:heuristics}.
 
-We then use a classical RL benchmark, the Inverted Pendulum (Subsec.~\ref{subsec:IP}), to provide a qualitative illustration of the proposed CSI approach.
+We then use a classical RL benchmark, the Inverted Pendulum (Sec.~\ref{subsec:IP}), to provide a qualitative illustration of the proposed CSI approach.
 
-We finally provide quantitative results and comparison with state-of-the-art approaches on a Highway driving simulator (Subsec.~\ref{subsec:highway}).
+We finally provide quantitative results and comparison with state-of-the-art approaches on a Highway driving simulator (Sec.~\ref{subsec:highway}).
 \subsection{Instanciation}
@@
 We provide here qualitative results on the inverted pendulum problem.
-Quantitative results obtained by following a procedure very similar to Subsec.~\ref{subsec:highway} were of the same nature and justified the same analysis as those provided in Subsubsec.~\ref{subsubsec:hresults}.
+Quantitative results obtained by following a procedure very similar to Sec.~\ref{subsec:highway} were of the same nature and justified the same analysis as those provided in Sec.~\ref{subsubsec:hresults}.
 We thus omit them here for the sake of avoiding redundancy.
@@
 Let's now move from a qualitative illustration to a quantitative assessment of the performance of the CSI algorithm.
-The Highway driving problem is described Subsubsec.~\ref{subsubsec:hsetting}, the criterion we use is explained Subsubsec.~\ref{subsubsec:hcriterion} and finally results and comparison to other approaches are presented Subsubsec.~\ref{subsubsec:hresults}.
+The Highway driving problem is described Sec.~\ref{subsubsec:hsetting}, the criterion we use is explained Sec.~\ref{subsubsec:hcriterion} and finally results and comparison to other approaches are presented Sec.~\ref{subsubsec:hresults}.
 \subsubsection{Setting}
@@
 
-  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. See Subsec.~\ref{subsec:setting} for detailled information about the experimental setting.}
+  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. See Sec.~\ref{subsec:setting} for detailled information about the experimental setting.}
   \label{fig:Highway}
@@
 Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
-The SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} which we used Subsec.~\ref{subsec:highway} is able to solve the IRL problem (without having to solve any MDP) in the same dire conditions the cascading approach has been tested in.
+The SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} which we used Sec.~\ref{subsec:highway} is able to solve the IRL problem (without having to solve any MDP) in the same dire conditions the cascading approach has been tested in.
 It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation as feature function.

### 2012-11-07 piot2012cascading/ECML13_cascading.tex
Dernieres modifs avant soumission

@@
 \and
-2.Supélec\\
+2. Supélec\\
  IMS-MaLIS Research group, France\\
@@
 \and 
-3. UMI 2958 GeorgiaTech-CNRS\\
+3. UMI 2958 (GeorgiaTech-CNRS)\\
 France\\
@@
 Sequential decision making consists in choosing the appropriate action given the available data in order to maximize a certain criterion.
-When framed in a Markov Decision Process (MDP) (see Sec.~\ref{sec:background}), the criterion to maximize is the expected sum of discounted rewards.
-The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which we tackle here, aims at inferring a reward function for which a demonstrated expert policy is optimal.
+When framed in a Markov Decision Process (MDP) (see Sec.~\ref{sec:background}), (Approximate) Dynamic programming ((A)DP) or Reinforcement Learning (RL) are often used to solve the problem by maximizing the expected sum of discounted rewards.
+The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which is addressed here, aims at inferring a reward function for which a demonstrated expert policy is optimal.
 
-IRL is often used as a way to do Apprenticeshipe Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function.
-The reward function nevertheless is of interest in and of itself.
-As mentionned in \cite{ng2000algorithms}, its semantics can be analyzed in biology or econometrics.
+IRL is one of many ways to perform Apprenticeship Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function.
+The reward function nevertheless is of interest in its own right.
+As mentionned in \cite{russell1998learning}, its semantics can be analyzed in biology or econometrics for instance.
 Practically, the reward can be seen as a succinct description of a task.
 Discovering it removes the coupling between understanding the task and learning how to fulfill it that exists in AL.
-IRL allows the use of (Approximate) Dynamic programming ((A)DP) or Reinforcement Learning (RL) techniques to learn the task from the computed reward function.
-Although some approaches use it as a byproduct, most AL algorithm make not use of the reward function.
-A very straightforward way to do AL is to use a multi-class classifier to directly learn the expert policy.
-We compare this idea with our IRL algorithm in the experiments (Sec.~\ref{sec:experiments}).
+IRL allows the use of (A)DP or RL techniques to learn the task from the computed reward function.
+Although some approaches use it as a byproduct, most AL algorithm makes no use of the reward function, or have nothing to do with IRL altogether.
+A very straightforward non-IRL way to do AL is for example to use a multi-class classifier to directly learn the expert policy.
+We provide in the experiments (Sec.~\ref{sec:experiments}) a comparison between AL and IRL algorithms by using IRL as a way to do AL.
 
-A lot of existing approaches in either IRL or AL need to repeatedly solve the underlying MDP.
+A lot of existing approaches in either IRL or IRL-based AL need to repeatedly solve the underlying MDP to find the optimal policies of intermediate reward functions.
 Thus, their performance depends strongly on the quality of the associated subroutine.
 Consequently, they suffer from the same challenges of scalability, data scarcity, etc., as RL and (A)DP.
-In order to avoid repeatedly solving such problems, we adopt a different point of view.
+In order to avoid \emph{repeatedly} solving such problems, we adopt a different point of view.
 
@@
 Thus, the demonstrated expert policy will be optimal for the related reward function.
-We propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function.
+We propose to use a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function.
 Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}).
 That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI).
-This method is analyzed, see Sec.~\ref{sec:analysis} where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
+This method is analyzed in Sec.~\ref{sec:analysis}, where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.
 
@@
 Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-A specific instantiation of CSI (proposed Sec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Sec.~\ref{subsec:IP}) and a car driving simulator (Sec.~\ref{subsec:Highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.
+A specific instantiation of CSI (proposed Sec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Sec.~\ref{subsec:IP}) and a car driving simulator (Sec.~\ref{subsec:highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.
 
@@
 We work with finite MDPs \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$.
-The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP.
-Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$.
+The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamics of the MDP.
+Thus, for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing action $a$ in state $s$.
 At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$.
@@
 The reward function $R$ is a local measure of the quality of the control.
-The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the discounted cumulative reward for following policy $\pi$ from this state:
+The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the expected discounted cumulative reward for following policy $\pi$ from this state:
 \begin{equation}
@@
 This long-term criterion is what is being optimized when solving an MDP.
-Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states: $\forall \pi, V^*_R\geq V^\pi_R$..
+Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states: $\forall \pi, V^*_R\geq V^\pi_R$.
 
@@
   \end{equation}
-where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally
+where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally:
 \begin{equation}
@@
   \begin{eqnarray}
-    V^\pi_R &=& T^\pi_R V^\pi_R\\
-    V^\pi_R &=& R_\pi + \gamma P_\pi V^\pi_R.
+    V^\pi_R &=& T^\pi_R V^\pi_R \Leftrightarrow V^\pi_R = R_\pi + \gamma P_\pi V^\pi_R.
   \end{eqnarray}
@@
 
-An optimal policy follows a greedy mechanism with respect to its optimal action-value function $Q^*_R$
+An optimal policy follows a greedy mechanism with respect to its optimal action-value function $Q^*_R$:
 \begin{equation}
@@
 
-When the state space is too large to allow matricial representation or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
+When the state space is too large to allow matricial representations or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
 
@@
 This contribution aims at solving the inverse one.
-We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi_E$, assuming that there exists some unknown reward $R_E$ for which the expert is optimal.
+We observe trajectories drawn from an expert's deterministic\footnotemark[\value{footnote}] policy $\pi_E$, assuming that there exists some unknown reward $R_E$ for which the expert is optimal.
 The suboptimality of the expert is an interesting setting that has been discussed for example in \cite{melo2010analysis,syed2010reduction}, but that we are not addressing here.
@@
 
-Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier.
+Given a dataset $D_C = \{(s_i,a_i=\pi_E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier.
 The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL).
@@
 One can easily notice the similarity between Eq.~\eqref{eq:greedy2} and Eq.~\eqref{eq:greedy} that describes the relation between the optimal policy in an MDP and its optimal action-value function.
-It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$.
+The score function $q$ of the classifier can thus be viewed as some kind of optimal action-value function for the classifier policy $\pi_C$.
 By inversing the Bellman equation Eq.~\eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
@@
   \label{algo:cascading}
-  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
+  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi_E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
   \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
@@
 Indeed, mere classification only takes into account information from the expert (i.e., which action goes with which state) whereas using the Bellman equation in the expression of $\hat r_j$ makes use of the information lying in the transitions $(s_j,a_j,s'_j)$, namely information about the transition probabilities $P$.
-The final regression step is just a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.
+The final regression step is a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.
 
@@
 In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs.
-More formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on:
+More formally, recalling from Sec.~\ref{sec:background} that $\rho_E$ is the stationary distribution of the expert policy, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]$ is bounded by a term that depends on:
 \begin{itemize}
-  \item the classification error
+  \item the classification error defined as
 \begin{equation}  
@@
 \end{equation}
-\item the regression error
+\item the regression error defined as 
   \begin{equation}
@@
   \begin{itemize}
-  \item the subscript notation already used in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are repectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
+  \item the subscript notation already used for $R_\pi$ and $P_\pi$ in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are repectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
   \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
@@
   \end{itemize}
-  The constant $C_*$ could be estimated \emph{a posteriori} ( after $\hat{R}^C$ is computed).
+  The constant $C_*$ could be estimated \emph{a posteriori} (after $\hat{R}^C$ is computed).
 We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
@@
   \begin{eqnarray}
-    \Delta q &= &\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))\\
-    &=&\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)),
+    \Delta q &= &\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)),
     \end{eqnarray} which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
@@
 \begin{proof}
-First let's introduce some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$ is a deterministic classifier policy.
+First let's introduce some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$.
 We recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
@@
 \end{equation}
-We are going to bound each of these three terms. First, let $\pi$ be a given deterministic policy. We have:
+We are going to bound each of these three terms. First, let $\pi$ be a given deterministic policy. We have, using $\epsilon^R = R^C - \hat R^C$:
 \begin{equation}
@@
 
-If $\pi\neq\pi_E$, we use the coefficient $C_\pi$. We have then:
+If $\pi\neq\pi_E$, we use the concentration coefficient $C_\pi$. We have then:
 \begin{equation}
@@
   \end{equation}
-So, using the notations introduced before we stated the theorem, we are able to give an upper bound to the first and third terms:
+So, using the notations introduced before we stated the theorem, we are able to give an upper bound to the first and third terms (recall also the notation $C_{\hat \pi_C} = C_*$):
 \begin{equation}
@@
 \end{equation}
+\qed
 \end{proof}
-
+A direct corollary of this theorem is that, given a perfect classifier and regressor, CSI produces a reward function for which $\pi_E$ is the \emph{unique} optimal policy.
 \begin{corollary}
   \label{cor}
-  If the bound of Theorem~\ref{thm} is tight (if and only if $\epsilon_C = 0$ and $\epsilon_R = 0$), $\pi_E$ is the only optimal policy for $\hat R^C$.
+  Assume that $\rho_E>0$ and that the classifier and the regressor are perfect (that is $\epsilon_C = 0$ and $\epsilon_R = 0$). Then $\pi_E$ is the \emph{unique} optimal policy for $\hat R^C$.
 \end{corollary}
 \begin{proof}
-  To get rid of technicalities we assume that $\forall s \in S, \rho_E(s) \neq 0$. If not we create a new MDP $\{S',A,\gamma,P,\hat R^C\}$ with $S' = \{s \in S | \rho_E(s) \neq 0\}$ and work in it.
 
-  The function $q$ is the action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\eqref{eq:rc}).
-
+  The function $q$ is the optimal action-value function for $\pi_C$ with respect to the reward $R^C$, by definition (see Eq.~\eqref{eq:rc}).
   As $\epsilon_C = 0$, we have $\pi_C = \pi_E$.
-This means that $\forall s, \pi_E(s)$ is the only element of the set $\argmax_{a\in A}q(sa)$.
+This means that $\forall s, \pi_E(s)$ is the only element of the set $\argmax_{a\in A}q(s,a)$.
 Therefore, $\pi_C = \pi_E$ is the unique optimal policy for $R^C$.
-
   As $\epsilon_R = 0$, we have $\hat R^C = R^C$, hence the result.
+  \qed
 \end{proof}
-The uniqueness of $\pi_E$ as an optimal policy for $\hat R^C$ in the conditions of corollary~\ref{cor} hints at the fact that we found a non-trivial reward (we recall that the null reward admits \emph{every} policy as optimal).
+This corollary hints at the fact that we found a non-trivial reward (we recall that the null reward admits \emph{every} policy as optimal).
 Therefore, obtaining  $\hat{R}_C = 0$ (for which the bound is obviously true: the bounded term is $0$, the bounding term is positive) is unlikely as long as the classifier and the regressor exhibit decent performance.
 
-The only constraints the bound of Thm.~\ref{thm} implies on datasets $D_R$ and $D_C$ is that they provide enough information to the supervised algorithms to keep both error terms $\epsilon_C$ and $\epsilon_R$ low.
-In Sec.~\ref{sec:experiments} we deal with a dire lack of data in dataset $D_R$.
-We address the problem with the use of heuristics (Sec.~\ref{subsubsec:heuristics}) in order to show the behavior of the CSI algorithm in somewhat realistic conditions.
+The only constraints the bound of theorem~\ref{thm} implies on datasets $D_R$ and $D_C$ is that they provide enough information to the supervised algorithms to keep both error terms $\epsilon_C$ and $\epsilon_R$ low.
+In Sec.~\ref{sec:experiments} we deal with a lack of data in dataset $D_R$.
+We address the problem with the use of heuristics (Sec.~\ref{subsubsec:heuristics}) in order to show the behavior of the CSI algorithm in somewhat more realistic (but difficult) conditions.
 
 More generally, the error terms  $\epsilon_C$ and $\epsilon_R$ can be reduced by a wise choice for the classification and regression subroutines.
-The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off etc.) to be found.
-Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach over existing algorithms.
+The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off, etc.) to be found.
+Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach over existing IRL algorithms.
 
@@
 In this section, we put our approach under stress to empirically demonstrate its behavior.
-We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data is given on Sec.~\ref{subsubsec:data} and a precise description of the way we instanciated the CSI algorithm is given on Sec.~\ref{subsubsec:subroutines} and \ref{subsubsec:heuristics}.
-
+We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data and a precise description of the way we instanciated the CSI algorithm are given on Sec.~\ref{subsec:instanciation}.
 We then use a classical RL benchmark, the Inverted Pendulum (Sec.~\ref{subsec:IP}), to provide a qualitative illustration of the proposed CSI approach.
-
-We finally provide quantitative results and comparison with state-of-the-art approaches on a Highway driving simulator (Sec.~\ref{subsec:highway}).
+We finally provide quantitative results and comparison with state-of-the-art approaches on a highway driving simulator (Sec.~\ref{subsec:highway}).
 \subsection{Instanciation}
@@
 The only data it could access is sampled according to the expert policy.
-This corresponds for example to a situation where a costly system can only be controlled by a trained operator as a bad control sequence could lead to a system breakage.
+This corresponds for example to a situation where a costly system can only be controlled by a trained operator as a bad control sequence could lead to a system breakdown.
 
@@
 It is however not reasonable to construct the dataset $D_R$ only from expert transitions and expect a small regression error term $\epsilon_R$.
-Indeed, the dataset $D_E$ only sample the dynamics induced by the expert's policy and not the whole dynamics of the MDP.
+Indeed, the dataset $D_E$ only samples the dynamics induced by the expert's policy and not the whole dynamics of the MDP.
 The regression step also needs samples $((s_j,a_j),\hat r_j)_j$ drawn from non expert actions.
@@
 \end{equation}
-Although this heuristics was not analyzed in Sec.~ \ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
+Although this heuristics was not analyzed in Sec.~\ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
 \subsubsection{Subroutines choices}
@@
   &\text{~s.t.~} \forall i,
-  \theta^\top{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
+  \theta^T{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^T
   {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
 \end{eqnarray*}
-with $l(s_i,a) = 1\textrm{ if } a\neq a_i,l(s_i,a_i)=0$.
-The score function of this classifier is simply $q(s,a) = \theta^\top \phi(s,a)$.
-The implementation of SCIRL \cite{klein2012scirl} is based on this.
+with $l(s_i,a) = 1\textrm{ if } a\neq a_i \textrm{ and } l(s_i,a_i)=0$.
+The score function of this classifier is simply $q(s,a) = \theta^T \phi(s,a)$.
+The implementation of the Structured Classification for IRL (SCIRL) algorithm \cite{klein2012scirl} (see Sec.~\ref{sec:related}) is based on this.
 We chose to use it in our CSI implementation to provide a fair comparison between both algorithms.
@@
 Ease of implementation leads us to choose a simple least-square regressor for our experiments.
-Using the same features $\phi$ as our classifier, we introduce $\lambda$ a regularization coefficient (set to $0.1$ in our experiments), $Id$ the identity matrix and
+Using the same features $\phi$ as our classifier, we introduce $\lambda$ a regularization coefficient (set to $0.1$ in our experiments), $I$ the identity matrix and
 \begin{eqnarray}
@@
     \vdots
-  \end{bmatrix},\\
+  \end{bmatrix} \in \mathbb{R}^{n|A|\times p}\textrm{ and }
   \hat R = \begin{bmatrix}
@@
     \vdots
-  \end{bmatrix}.
+  \end{bmatrix}\in \mathbb{R}^{n|A|},
 \end{eqnarray}
-
-Our regressor can be written:
+with $r_j = r_{min}$ if $a_j \neq \pi_E(s_j)$. Our regressor can be written:
 \begin{equation}
-  \omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R.
+  \omega = (\Phi^T\Phi + \lambda I)^{-1}\Phi^T\hat R.
 \end{equation}
@@
     %\small
-  \caption{A CSI instanciation}
+  \caption{A CSI instanciation with heuristics}
   \label{alg:CSIimp}
@@
   &\text{~s.t.~} \forall i,
-  \theta^\top{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
+  \theta^T{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^T
   {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
@@
 We provide here qualitative results on the inverted pendulum problem.
-Quantitative results obtained by following a procedure very similar to Sec.~\ref{subsec:highway} were of the same nature and justified the same analysis as those provided in Sec.~\ref{subsubsec:hresults}.
+Quantitative results obtained by following a procedure very similar to Sec.~\ref{subsec:highway} were of the same nature and allowed drawing the same conclusions as those provided in Sec.~\ref{subsubsec:hresults}.
 We thus omit them here for the sake of avoiding redundancy.
@@
 The goal is to maintain a pole in a unstable equilibrium position with the help of a cart moving along one axis.
-The state space consists in the angular position an speed of the pole.
-It is possible to use an RL algorithm to learn how to balance the pendule in an upright position by using the following reward: as long as the angular position of the pole is (in absolute value) less than $\pi \over 2$ the reward is $0$, the reward is $-1$ if the pole falls ; this reward function is plotted on Fig.~\ref{toto}.
+The state space consists in the angular position and speed of the pole.
+It is possible to use an RL algorithm to learn how to balance the pendule in an upright position by using the following reward: as long as the angular position of the pole is (in absolute value) less than $\pi \over 2$ the reward is $0$, the reward is $-1$ if the pole falls; this reward function is plotted on Fig.~\ref{toto}.
 Three actions are available to the agent: it can choose to apply a leftward or rightward force on the cart or no force at all.
@@
   \phi(s,a) = \begin{pmatrix}
-     \delta_{a = LEFT}\psi(s)\\
-     \delta_{a = NONE}\psi(s)\\
-     \delta_{a = RIGHT}\psi(s)
+     \delta_{a = \textsc{left}}\psi(s)\\
+     \delta_{a = \textsc{none}}\psi(s)\\
+     \delta_{a = \textsc{right}}\psi(s)
   \end{pmatrix}
@@
 In the left column, we show the reward and value functions of the expert.
-The expert (a RL agent trained on the  reward shown on Fig.~\ref{toto}) is able to balance the pendule indefinitely.
+The expert (an RL agent trained on the  reward shown on Fig.~\ref{toto}) is able to balance the pendule indefinitely.
 With as little as 300 transitions (30 seconds of balancing), CSI is able to find a reward that, once optimized by an RL agent, will reliably lead to a control as good as the expert's.
 As Fig.~\ref{tata} illustrates, the data we get from the expert only covers a tiny fraction of the state space.
-Furthermore, the feature function $\phi$ did not permit to accurately approximate the true reward\footnote{The true reward function is a function in $\mathbb{R}^S$, CSI outputs a reward in $\mathbb{R}^{S\times A}$, but we plot here $R(s,\pi^*(s))$ with $\pi^*$ the optimal policy for $R$.}.
-Despite these two setbacks, CSI manage to find a reward for which not only the expert's control is near-optimal (that is what the analysis of Sec.~\ref{sec:analysis} garantees under the assumption that a reasonable amount of data is available in order to keep both classification and regression errors down) but that also lead to a control policy as good as the expert's \emph{with respect to the (theoretically) unknown expert's reward function}.
+Furthermore, the feature function $\phi$ can not permit to accurately approximate the true reward\footnote{The true reward function is a function in $\mathbb{R}^S$, CSI outputs a reward in $\mathbb{R}^{S\times A}$, but we plot here $R(s,\pi^*(s))$ with $\pi^*$ the optimal policy for $R$.}.
+Despite these two setbacks, CSI manages to find a reward for which not only the expert's control is near-optimal (that is what the analysis of Sec.~\ref{sec:analysis} garantees under the assumption that a reasonable amount of data is available in order to keep both classification and regression errors down) but that also leads to a control policy as good as the expert's \emph{with respect to the (theoretically) unknown expert's reward function}.
+This phenomenon appears clearly when comparing figures \ref{fifi} and \ref{riri}: although the reward functions were different, the associated optimal value functions are (up to the scale, which does not actually matters) very similar.
 
 \begin{figure}
-   \subfigure[Reward on which the expert was trained]{\label{toto}\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_true_R"}\end{minipage}}
+  \begin{minipage}{.4\linewidth}
+   \subfigure[Reward on which the expert was trained]{\label{toto}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_true_R"}}
+   \end{minipage}
+  \hfill
+  \begin{minipage}{.4\linewidth}
     \subfigure[Reward found by CSI.
-The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\begin{minipage}{.45\linewidth}\hspace{-3em}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_lafem_R"}\end{minipage}}
-  \subfigure[Value function of the expert (i.e.
-optimal value function for the reward displayed Fig.~\ref{toto})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vexpert"}\end{minipage}}
-  \subfigure[Optimal value function for the reward function found by CSI (Fig.~\ref{tata})]{\begin{minipage}{.45\linewidth}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vagent"}\end{minipage}}
-
+The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\hspace{-3em}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_lafem_R"}}
+  \end{minipage}\\
+  \begin{minipage}{.4\linewidth}
+    \subfigure[Value function of the expert (i.e. optimal value function for the reward displayed Fig.~\ref{toto})]{\label{riri}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vexpert"}}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{.4\linewidth}
+  \subfigure[Optimal value function for the reward function found by CSI (Fig.~\ref{tata})]{\label{fifi}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vagent"}}
+  \end{minipage}
   \caption{Results on the inverted pendulum benchmark, illsutrating the dissimilarity of the rewards, but the similarity of the value functions.
@@
 The agent controls a car that can switch between the three lanes of the road, go off-road on either side and modulate between three speed levels.
-At all times, there will be one car in one of the three lanes.
+At all timesteps, there will be one car in one of the three lanes.
 Even at the lowest speed, the player's car moves faster than the others.
-When the other car disappears at the bottom of the screen, another one appear at the top in a randomly chosen lane.
+When the other car disappears at the bottom of the screen, another one appears at the top in a randomly chosen lane.
 It takes two transitions to completely change lanes, as the player can move left or right for half a lane's length at a time.
-At the highest speed setting, if the other car appear in the lane the player is in, it is not possible to avoid the collision.
+At the highest speed setting, if the other car appears in the lane the player is in, it is not possible to avoid the collision.
 The main difference between the original benchmark \cite{syed2008apprenticeship,syed2008game} and ours is that we made the MDP more ergodic by allowing the player to change speed whenever he wishes so, not just during the first transition.
-If anything, by adding two actions, we enlarged the state-action space and thus made the problem harder.
+If anything, by adding two actions, we enlarged the state-action space and thus made the problem tougher.
 %FIXME:METTRE UNE CAPTURE DECRAN.
@@
 Another natural benchmark we use is to compute the mean value function of the classifier policy.
-We also compare the proposed cascading method to the PIRL algorithm of \cite{abbeel2004apprenticeship} and the SCIRL algorithm of \cite{klein2012scirl}.
+We also compare the proposed cascading method to the projection version of the seminal algorithms in \cite{abbeel2004apprenticeship}\footnote{We call it PIRL for Projection IRL.} and the SCIRL algorithm of \cite{klein2012scirl}.
 The random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
@@
 Apart from SCIRL, to the best of our knowledge, no other algorithm can infer a reward function only from expert data.
-Recall from Sec.~\ref{sec:related} that most existing approaches either need to solve the direct RL problem repeatedly (which can not be done on this setting from expert data alone) or rely on a sampling of the dynamics of the whole MDP.
+We will see in Sec.~\ref{sec:related} that most existing approaches either need to solve the direct RL problem repeatedly (which can not be done on this setting from expert data alone) or rely on a sampling of the dynamics of the whole MDP and do not propose (as CSI or SCIRL) the use of heuristics to avoid this problem.
 The reward functions found by SCIRL and CSI are then optimized using a dynamic programming algorithm to show the performance of the associated policy.
@@
 The second reason for this different metric is that we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well-behaved given the use of heuristics.
-\
+Finally, it is quite fair to use a single objective criterion than to just compare all approaches with one another.
+
 \subsubsection{Results}
@@
 This also shows that CSI is indeed very sample efficient, without having to repeatedly solve the MDP.
+Furthermore, it is much faster to run our implementation of CSI on so little data than to run PIRL with the dynamics for 70 iterations.
 
@@
 
-  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. See Sec.~\ref{subsec:setting} for detailled information about the experimental setting.}
+  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. See Sec.~\ref{subsubsec:hsetting} for detailled information about the experimental setting.}
   \label{fig:Highway}
@@
 Approaches summarized in \cite{neu2009training} can be seen as iteratively contructing a reward function, solving an MDP at each iteration.
-Some of these algorithms are IRL algorithms while others fall in the AL category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship}\footnote{We call it PIRL for Projection IRL.} which we used in our benchmarks (Sec.~\ref{sec:experiments}).
+Some of these algorithms are IRL algorithms while others fall in the AL category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship} which we used in our benchmarks (Sec.~\ref{sec:experiments}).
 In both cases the need to solve an MDP at each step may be very demanding, both sample-wise and computationally.
@@
 
-AL via classification has been proposed in \cite{ratliff2007imitation} with the help of a structured margin method.
+AL via classification has been proposed in for example \cite{ratliff2007imitation} with the help of a structured margin method.
 Using the non trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
@@
 The SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} which we used Sec.~\ref{subsec:highway} is able to solve the IRL problem (without having to solve any MDP) in the same dire conditions the cascading approach has been tested in.
-It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation as feature function.
-This necessity of a linarly parametrized approach and the need to compute the expert's feature exepctations makes the SCIRL algorithm less convenient than the CSI approach.
+It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation (the expected discounted sum of the a feature function evaluated over the states of the trajoctories the expert follow) as feature function.
+This necessity of a linarly parametrized approach and the need to compute the expert's feature expectations makes the SCIRL algorithm less convenient than the CSI approach.
 
@@
 We have introduced a new way to perform IRL by cascading two supervised approaches.
-The expert is shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.
-Practical examples of classifiers and regressors have been given, and one combination has been empirically (on two classical benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
+The expert is theoretically shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.
+Practical examples of classifiers and regressors have been given, and one combination has been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
 On one of the benchmarks (a car driving simulator), our algorithm is shown to outperform other state of the art approaches, and to be on-par with the very recent SCIRL algorithm.

### 2012-11-07 piot2012cascading/4plots.py

@@ -91,7 +91,7 @@ pylab.savefig("Fig3-SCIRLAndUs.pdf",transparent=True)
 fig = pylab.figure(4)
 figlegend = pylab.figure(figsize=(4,2.5))
 ax = fig.add_subplot(111)
-lines = ax.plot([-1,-1],[-1,-2], color='red',label="Cascading IRL",lw=2,linestyle='-')
+lines = ax.plot([-1,-1],[-1,-2], color='red',label="CSI",lw=2,linestyle='-')
 lines += ax.plot([-1,-1],[-1,-2], color='green',label="Pure classification",lw=2,linestyle='-')
 lines += ax.plot([-1,-1],[-1,-2], color='blue',label="Abbeen \& Ng IRL",lw=2,linestyle='-')
 lines += ax.plot([-1,-1],[-1,-2], color='cyan',label="Agent trained on a random reward",lw=2,linestyle='-')
@@ -99,6 +99,6 @@ lines += ax.plot([-1,-1],[-1,-2], color='orange',label="SCIRL",lw=2,linestyle='-
 lines += ax.plot([-1,-1],[-1,-2], color='purple',label="Expert",lw=2,linestyle='-')
 lines += ax.plot([-1,-1],[-1,-2], color='black',label="min, max",lw=1,linestyle='--')
 lines += ax.plot([-1,-1],[-1,-2], color='black',label="Standard deviation",lw=1,linestyle='-.')
-figlegend.legend(lines,("Cascading IRL","Pure Classification","Abbeel \& Ng IRL","Agent trained on a random reward","SCIRL","Expert","min, max","Standard deviation"),"center")
+figlegend.legend(lines,("CSI","Pure Classification","Abbeel \& Ng IRL","Agent trained on a random reward","SCIRL","Expert","min, max","Standard deviation"),"center")
 figlegend.savefig("Fig4-Legend.pdf")
 

### 2012-11-23 klein2012structured_NIPS/scirl_nips_cr.tex

\documentclass{article} % For LaTeX2e
\usepackage{nips12submit_e,times}
%\documentstyle[nips12submit_09,times,art10]{article} % For LaTeX 2.09

\usepackage[latin1]{inputenc}
% \usepackage[francais]{babel}
\usepackage[T1]{fontenc}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}
%\usepackage{mathtools}

\usepackage[ruled,algo2e]{algorithm2e}
\usepackage{hyperref}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amscd}

\usepackage{mathtools}
\mathtoolsset{showonlyrefs=true}

\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\argmin}{\operatorname*{argmin}}
\newcommand{\diag}{\operatorname*{diag}}

\newcommand{\p}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\s}{\mathcal{S}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\lc}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\prob}{\mathbb{P}}
\newcommand{\Mu}{\boldsymbol{\mu}}
%\newcommand{\Xi}{\boldsymbol{\xi}}

\newtheorem{theorem}{Theorem}
\newtheorem{prop}[theorem]{Proposition}

\title{Inverse Reinforcement Learning \\ through Structured Classification}

\author{
Edouard Klein$^{1,2}$\\
$^{1}$LORIA -- team ABC\\
Nancy, France\\
\texttt{edouard.klein@supelec.fr} \And
Matthieu Geist$^{2}$\\
$^{2}$Supélec -- IMS-MaLIS Research Group\\
Metz, France\\
\texttt{matthieu.geist@supelec.fr}
\And Bilal Piot$^{2,3}$, Olivier Pietquin$^{2,3}$\\
$^{3}$ UMI 2958 (GeorgiaTech-CNRS)\\
Metz, France\\
\texttt{$\{$bilal.piot,olivier.pietquin$\}$@supelec.fr} }

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\begin{abstract}
  This paper adresses the inverse reinforcement learning (IRL) problem, that
  is inferring a reward for which a demonstrated expert behavior is
  optimal. We introduce a new algorithm, SCIRL, whose principle is to use
  the so-called feature expectation of the expert as the
  parameterization of the score function of a multi-class
  classifier. This approach produces a reward function for which the
  expert policy is provably near-optimal. Contrary to most of
  existing IRL algorithms, SCIRL does not require solving the
  direct RL problem. Moreover, with an appropriate heuristic, it can succeed with only trajectories sampled
  according to the expert behavior. This is illustrated on a car
  driving simulator.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Inverse reinforcement learning (IRL)~\cite{Russell:1998} consists in
finding a reward function such that a demonstrated expert behavior
is optimal. Many IRL algorithms (to be briefly reviewed in
Sec.~\ref{sec:relatedWorks}) search for a reward function such that
the associated optimal policy induces a distribution over
trajectories (or some measure of this distribution) which matches
the one induced by the expert. Often, this distribution is
characterized by the so-called feature expectation (see
Sec.~\ref{subsec:background:irl}): given a reward function linearly
parameterized by some feature vector, it is the expected discounted
cumulative feature vector for starting in a given state, applying a
given action and following the related policy.

In this paper, we take a  different route. The expert behavior could
be mimicked by a supervised learning algorithm generalizing the
mapping from states to actions. Here, we consider generally
multi-class classifiers which compute from a training set the
parameters of a linearly parameterized score function; the decision
rule for a given state is the argument (the action) which maximizes
the score function for this state (see
Sec.~\ref{subsec:background:classif}). The basic idea of our SCIRL
(Structured Classification-based IRL) algorithm is simply to take an
estimate of the expert feature expectation as the parameterization
of the score function (see Sec.~\ref{subsec:scirl:algo}). The
computed parameter vector actually defines a reward function for
which we show the expert policy to be near-optimal
(Sec.~\ref{subsec:scirl:analysis}).

Contrary to most existing IRL algorithms, a clear advantage of SCIRL
is that it does not require solving repeatedly the direct
reinforcement learning (RL) problem. It requires estimating the
expert feature expectation, but this is roughly a policy evaluation
problem (for an observed policy, so is less involved than repeated
policy optimization problems), see Sec.~\ref{sec:practicalApproach}.
Moreover, up to the use of some heuristic, SCIRL may be trained
solely from transitions sampled from the expert policy (no need to
sample the whole dynamic). We illustrate this on a car driving
simulator in Sec.~\ref{sec:experiments}.

\section{Background and Notations}
\label{sec:background}

\subsection{(Inverse) Reinforcement Learning}
\label{subsec:background:irl}

A Markov Decision process (MDP)~\cite{Puterman:1994} is a tuple
$\{\s,\A,\p,\R,\gamma\}$ where $\s$ is the finite state
space\footnote{This work can be extended to compact state spaces, up
to some technical aspects.}, $\A$ the finite actions space, $\p =
\{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ the set of
Markovian transition probabilities, $\R\in\mathbb{R}^\s$ the
state-dependent reward function and $\gamma$ the discount factor. A
deterministic policy $\pi\in\s^\A$ defines the behavior of an agent.
The quality of this control is quantified by the value function
$v_\R^\pi\in\mathbb{R}^\s$, associating to each state the cumulative
discounted reward for starting in this state and following the
policy $\pi$ afterwards: $v_\R^\pi(s) = \E[\sum_{t\geq 0} \gamma^t
\R(S_t)|S_0=s,\pi]$. An optimal policy $\pi_\R^*$ (according to the
reward function $\R$) is a policy of associated value function
$v^*_\R$ satisfying $v_\R^* \geq v_\R^\pi$, for any policy $\pi$ and
componentwise.

Let $P_\pi$ be the stochastic matrix $P_\pi =
(p(s'|s,\pi(s)))_{1\leq s,s'\leq |\s|}$. With a slight abuse of
notation, we may write $a$ the policy which associates the action
$a$ to each state $s$. The Bellman evaluation (resp. optimality)
operators $T^\pi_\R\text{ (resp. $T^*_\R$)}:\mathbb{R}^\s
\rightarrow \mathbb{R}^\s$ are defined as $T^\pi_\R v = \R + \gamma
P_\pi v$ and $T_\R^*v = \max_\pi T_\R^\pi v$.
%\begin{equation}
%  T^\pi_\R v = \R + \gamma P_\pi v \text{ and } T_\R^*v = \max_\pi T_\R^\pi v.
%\end{equation}
These operators are contractions and $v_\R^\pi$ and $v^*_\R$ are
their respective fixed-points: $v_\R^\pi = T^\pi_\R v_\R^\pi$ and
$v^*_\R = T^*_\R v^*_\R$. The action-value function
$Q^\pi\in\mathbb{R}^{\s\times \A}$ adds a degree of freedom on the
choice of the first action, it is formally defined as $Q_\R^\pi(s,a)
= [T^a_\R v^\pi_\R](s)$. We also write $\rho_\pi$ the stationary
distribution of the policy $\pi$ (satisfying $\rho_\pi^\top P_\pi =
\rho_\pi^\top$).

Reinforcement learning and approximate dynamic programming aim at
estimating the optimal control policy $\pi^*_\R$ when the model
(transition probabilities and the reward function) is unknown (but
observed through interactions with the system to be controlled) and
when the state space is too large to allow exact representations of
the objects of interest (as value functions or
policies)~\cite{Bertsekas:1996,Sutton:1998,szepesvari2010c}.
%
We refer to this as the direct problem.
%
On the contrary, (approximate) inverse reinforcement
learning~\cite{Ng:2000} aim at estimating a reward function for
which an observed policy is (nearly) optimal.
%
Let us call this policy the expert policy, denoted $\pi_E$. We may
assume that it optimizes some unknown reward function $\R_E$. The
aim of IRL is to compute some reward $\hat{\R}$ such that the expert
policy is (close to be) optimal, that is such that $v^*_{\hat{\R}}
\approx v^{\pi_E}_{\hat{\R}}$.
%
We refer to this as the inverse problem.

Similarly to the direct problem, the state space may be too large
for the reward function to admit a practical exact representation.
Therefore, we restrict our search of a good reward among linearly
parameterized functions. Let $\phi(s) = (\phi_1(s)  \dots
\phi_p(s))^\top$
%\begin{pmatrix}
%  \phi_1(s) & \dots & \phi_p(s)
%\end{pmatrix}^\top$
be a feature vector composed of $p$ basis function
$\phi_i\in\mathbb{R}^\s$, we define the  parameterized reward
functions as $\R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p
\theta_i \phi_i(s)$.
%\begin{equation}
%  \R_\theta(s) = \theta^\top \phi(s) = \sum_{i=1}^p \theta_i \phi_i(s).
%\end{equation}
Searching a good reward thus reduces to searching a good parameter
vector $\theta \in\mathbb{R}^p$. Notice that we will use
interchangeably $\R_\theta$ and $\theta$ as subscripts
(\textit{e.g.}, $v_\theta^\pi$ for $v_{\R_\theta}^\pi$).
Parameterizing the reward this way implies a related
parameterization for the action-value function:
\begin{equation}
  Q^\pi_\theta(s,a) = \theta^\top \mu^\pi(s,a) \text{ with }
  \mu^\pi(s,a) = \E[\sum_{t\geq 0} \gamma^t
  \phi(S_t)|S_0=s,A_0=a,\pi].
  \label{eq:def:mu}
\end{equation}
Therefore, the action-value function shares the parameter vector of
the reward function, with an associated feature vector $\mu^\pi$
called the \emph{feature expectation}. This notion will be of
primary importance for the contribution of this paper. Notice that
each component $\mu_i^\pi$ of this feature vector is actually the
action-value function of the policy $\pi$ assuming the reward is
$\phi_i$: $\mu_i^\pi(s,a) = Q^\pi_{\phi_i}(s,a)$. Therefore, any
algorithm designed for estimating an action-value function may be
used to estimate the feature expectation, such as Monte-Carlo
rollouts or temporal difference learning~\cite{Klein:2011}.

\subsection{Classification with Linearly Parameterized Score
Functions} \label{subsec:background:classif}

Let $\X$ be a compact or a finite set (of inputs to be classified)
and let $\Y$ be a finite set (of labels). Assume that inputs $x\in
\X$  are drawn according to some unknown distribution $\prob(x)$ and
that there exists some oracle which associates to each of these
inputs a label $y\in Y$ drawn according to the unknown conditional
distribution $\prob(y|x)$. Generally speaking, the goal of
multi-class classification is, given a training set
$\{(x_i,y_i)_{1\leq i \leq N}\}$ drawn according to $\prob(x,y)$, to
produce a decision rule $g\in\Y^\X$ which aims at minimizing the
classification error $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq
y)$, where $\chi$ denotes the indicator function.

Here, we consider a more restrictive set of classification
algorithms. We assume that the decision rule associates to an input
the argument which maximizes a related score function, this score
function being linearly parameterized and the associated parameters
being learnt by the algorithm. More formally, let $\psi(s,a) =
(\psi_1(x,y)  \dots  \psi_d(x,y))^\top\in \mathbb{R}^d$
%\begin{pmatrix}   \psi_1(x,y) & \dots & \psi_d(x,y)
%\end{pmatrix}^\top \in \mathbb{R}^d$
be a feature vector whose components are $d$ basis functions
$\psi_i\in\mathbb{R}^{\X\times\Y}$. The linearly parameterized score
function $s_w\in\mathbb{R}^{\X\times \Y}$ of parameter vector
$w\in\mathbb{R}^d$ is defined as $s_w(x,y) = w^\top \psi(x,y)$. The
associated decision rule $g_w\in{\Y^\X}$ is defined as $g_w(x) \in
\argmax_{y\in\Y}s_w(x,y)$. Using a training set $\{(x_i,y_i)_{1\leq
i\leq N}\}$, a linearly parameterized score function-based
multi-class classification (MC$^2$ for short) algorithm computes a
parameter vector $\theta_c$. The quality of the solution is
quantified by the classification error $\epsilon_c =
\prob(g_{\theta_c}(x)\neq y)$.

We do not consider a specific MC$^2$ algorithm, as long as it
classifies inputs by maximizing the argument of a linearly
parameterized score function. For example, one may choose a
multi-class support vector machine~\cite{Guermeur:2007} (taking the
kernel induced by the feature vector) or a structured large margin
approach~\cite{Taskar:2005}. Other choices may be possible, one can
choose its preferred algorithm.

\section{Structured Classification for Inverse Reinforcement
Learning} \label{sec:scirl}

\subsection{General Algorithm}
\label{subsec:scirl:algo}

Consider the classification framework of
Sec.~\ref{subsec:background:classif}. The input $x$ may be seen as a
state and the label $y$ as an action. Then, the decision rule
$g_w(x)$ can be interpreted as a policy which is greedy according to
the score function $w^\top \psi(x,y)$, which may itself be seen as
an action-value function. Making the parallel with
Eq.~\eqref{eq:def:mu}, if $\psi(x,y)$ is the feature expectation of
some policy $\pi$ which produces labels of the training set, and if
the classification error is small, then $w$ will be the parameter
vector of a reward function for which we may hope the policy $\pi$
to be near optimal. Based on these remarks, we're ready to present
the proposed Structured Classification-based IRL (SCIRL) algorithm.

Let $\pi_E$ be the expert policy from which we would like to recover
a reward function. Assume that we have a training set $\D =
\{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ where states are sampled
according to the expert stationary distribution\footnote{For
example, if the Markov chain induced by the expert policy is
fast-mixing, sampling a trajectory will quickly lead to sample
states according to this distribution.} $\rho_E = \rho_{\pi_E}$.
Assume also that we have an estimate $\hat{\mu}^{\pi_E}$ of the
expert feature expectation $\mu^{\pi_E}$ defined in
Eq.~\eqref{eq:def:mu}. How to practically estimate this quantity is
postponed to Sec.~\ref{subsec:practicalApproach:muE}; however,
recall that estimating $\mu^{\pi_E}$ is simply a \emph{policy
evaluation} problem (estimating the action-value function of a given
policy), as noted in Sec.~\ref{subsec:background:irl}. Assume also
that an MC$^2$ algorithm has been chosen. The proposed algorithm
simply consists in choosing $\theta^\top\hat{\mu}^{\pi_E}(s,a)$ as
the linearly parameterized score function, training the classifier
on $\D$ which produces a parameter vector $\theta_c$, and outputting
the reward function $\R_{\theta_c}(s) = \theta_c^\top \phi(s)$.

\begin{algorithm2e}%[tbh]
    %\small
  \SetAlgoVlined
  \caption{SCIRL algorithm}
  \label{algo:scirl}
  %
  \BlankLine
  \emph{\textbf{Given}} a training set $\D = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
  an estimate $\hat{\mu}^{\pi_E}$ of the expert feature expectation $\mu^{\pi_E}$ and an MC$^2$ algorithm\;
  %
  \BlankLine
  \emph{\textbf{Compute}} the parameter vector $\theta_c$ using the
  MC$^2$ algorithm
  fed with the training set $\D$ and considering the parameterized score function
  $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
  %
  \BlankLine
  \emph{\textbf{Output}} the reward function $\R_{\theta_c}(s) = \theta_c^\top\phi(s)$ \;
\end{algorithm2e}

The proposed approach is summarized in Alg.~\ref{algo:scirl}. We
call this Structured Classification-based IRL because using the
(estimated) expert feature expectation as the feature vector for the
classifier somehow implies taking into account the MDP structure
into the classification problem and allows outputting a reward
vector. Notice that contrary to most of existing IRL
%and apprenticeship learning
algorithms, SCIRL does not require solving the direct problem. If it
possibly requires estimating the expert feature expectation, it is
just a policy evaluation problem, less difficult than the policy
optimization issue involved by the direct problem. This is further
discussed in Sec.~\ref{sec:relatedWorks}.
%
%Now, we formally show that the proposed approach makes sens, as long
%as $\mu^{\pi_E}$ is well estimated and as long as the classification
%error is small.

\subsection{Analysis}
\label{subsec:scirl:analysis}

In this section, we show that the expert policy $\pi_E$ is close to
be optimal according to the reward function $\R_{\theta_c}$, more
precisely that
$\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$ is
small. Before stating our main result, we need to introduce some
notations and to define some objects.

We will use the first order discounted future state distribution
concentration coefficient $C_f$~\cite{Munos:2007}:
\begin{equation}
  C_f = (1-\gamma)\sum_{t\geq 0} \gamma^t c(t) \text{ with } c(t) =
  \max_{\pi_1,\dots,\pi_t,s\in\s}\frac{(\rho_E^\top P_{\pi_1}\dots
  P_{\pi_t})(s)}{\rho_E(s)}.
\end{equation}
We note $\pi_c$ the decision rule of the classifier: $\pi_c(s) \in
\argmax_{a\in \A} \theta_c^\top\hat{\mu}^{\pi_E}(s,a)$. The
classification error is therefore $\epsilon_c =
\E_{s\sim\rho_E}[\chi_{\{\pi_c(s)\neq\pi_E(s)\}}] \in [0,1]$. We
write $\hat{Q}^{\pi_E}_{\theta_c} = \theta_c^\top \hat{\mu}^{\pi_E}$
the score function computed from the training set $\D$ (which can be
interpreted as an approximate action-value function). Let also
$\epsilon_{\mu} = \hat{\mu}^{\pi_E} - \mu^{\pi_E}:\s\times\A
\rightarrow  \mathbb{R}^p$ be the feature expectation error.
Consequently, we define the action-value function error as
$\epsilon_Q = \hat{Q}^{\pi_E}_{\theta_c} - Q^{\pi_E}_{\theta_c} =
\theta_c^\top(\hat{\mu}^{\pi_E} - \mu^{\pi_E}) = \theta_c^\top
\epsilon_\mu:\s\times\A\rightarrow\mathbb{R}$. We finally define the
mean delta-max action-value function error as $\bar{\epsilon}_Q =
\E_{s\sim\rho_E}[\max_{a\in\A}\epsilon_Q(s,a) -
\min_{a\in\A}\epsilon_Q(s,a)]\geq 0$.

\begin{theorem}
  \label{th}
  Let $\R_{\theta_c}$ be the reward function outputted by
  Alg.~\ref{algo:scirl}. Let also the quantities $C_f$, $\epsilon_c$
  and $\bar{\epsilon}_Q$ be defined as above. We have
  %
%  the concentration coefficient $C_f$,
%  the classification error $\epsilon_c$ and the mean-max
%  action-value function error $\bar{\epsilon}_Q$ be defined as
%  above. Then, we have
  \begin{equation}
    0\leq
    \E_{s\sim\rho_E}[v^*_{\R_{\theta_c}}-v^{\pi_E}_{\R_{\theta_c}}]
    \leq \frac{C_f}{1-\gamma}\left(\bar{\epsilon}_Q +
    \epsilon_c\frac{2\gamma\|\R_{\theta_c}\|_\infty}{1-\gamma}
%    \epsilon_c\frac{2\gamma}{1-\gamma}\min(\|\R_{\theta_c}\|_\infty,
%    C_p \E_{s\sim\rho_E}[\R_{\theta_c}(s)])
    \right).
  \end{equation}
\end{theorem}
\begin{proof}
  As the proof only relies on the reward $\R_{\theta_c}$, we
  omit the related subscripts to keep the notations simple
  (\textit{e.g.}, $v^\pi$ for
  $v^\pi_{\theta_c}=v^\pi_{\R_{\theta_c}}$ or $\R$ for $\R_{\theta_c}$). First, we link
  the error $\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)]$ to the
  Bellman residual $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s)-v^{\pi_E}(s)]$.
  %
  % A second step consists in bounding this residual.
  %
  Componentwise, we have that:
  \begin{align}
    v^* - v^{\pi_E} &= T^* v^*  - T^{\pi^*}v^{\pi_E} +
    T^{\pi^*}v^{\pi_E} - T^* v^{\pi_E} + T^* v^{\pi_E} - v^{\pi_E}
    \\
    &\stackrel{(a)}{\leq} \gamma P_{\pi^*}(v^*-v^{\pi_E}) + T^*
    v^{\pi_E} - v^{\pi_E}
    %
    \stackrel{(b)}{\leq} (I-\gamma
    P_{\pi^*})^{-1} (T^* v^{\pi_E} - v^{\pi_E}).
  \end{align}
  Inequality $(a)$ holds because $T^{\pi^*} v^{\pi_E}\leq T^*
  v^{\pi_E}$ and inequality $(b)$ holds thanks
  to~\cite[Lemma~4.2]{Munos:2007}. Moreover, $v^*$ being optimal we
  have that $v^*-v^{\pi_E}\geq 0$ and $T^*$ being the Bellman
  optimality operator, we have $T^* v^{\pi_E}\geq
  T^{\pi_E}v^{\pi_E}=v^{\pi_E}$. Additionally, remark that
  $(I-\gamma P_{\pi^*})^{-1} = \sum_{t\geq 0}\gamma^t P_{\pi^*}^t$.
  Therefore, using the definition of the concentration coefficient
  $C_f$, we have that:
  \begin{equation}
    0\leq\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)] \leq \frac{C_f}{1-\gamma}
    \E_{s\sim\rho_E}\left[[T^*v^{\pi_E}](s) - v^{\pi_E}(s)\right].
    \label{eq:proof:residual}
  \end{equation}
  This results actually follows closely the one
  of~\cite[Theorem~4.2]{Munos:2007}. There remains to bound the
  Bellman residual $\E_{s\sim\rho_E}[[T^*v^{\pi_E}](s) -
  v^{\pi_E}(s)]$. Considering the following decomposition,
  \begin{equation}
    T^* v^{\pi_E} - v^{\pi_E} = T^* v^{\pi_E} - T^{\pi_c}v^{\pi_E}
    + T^{\pi_c}v^{\pi_E}- v^{\pi_E},
    \label{eq:proof:decomposition}
  \end{equation}
  we will bound $\E_{s\sim\rho_E}[[T^* v^{\pi_E}](s) - [T^{\pi_c}v^{\pi_E}](s)]$
  and $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]$.

  The policy $\pi_c$ (the decision rule of the classifier) is greedy
  with respect to
  $\hat{Q}^{\pi_E}=\theta_c^\top\hat{\mu}^{\pi_E}$. Therefore, for any state-action couple
  $(s,a)\in\s\times \A$ we have:
  \begin{equation}
    \hat{Q}^{\pi_E}(s,\pi_c(s))\geq
    \hat{Q}^{\pi_E}(s,a)
    \Leftrightarrow
    Q^{\pi_E}(s,a) \leq Q^{\pi_E}(s,\pi_c(s)) +
    \epsilon_Q(s,\pi_c(s)) - \epsilon_Q(s,a).
  \end{equation}
  By definition, $Q^{\pi_E}(s,a) = [T^a v^{\pi_E}](s)$ and
  $Q^{\pi_E}(s,\pi_c(s)) = [T^{\pi_c} v^{\pi_E}](s)$. Therefore, for $s\in\s$:
  \begin{align}
    \forall a\in A,\; [T^a v^{\pi_E}](s) &\leq [T^{\pi_c}
    v^{\pi_E}](s) + \epsilon_Q(s,\pi_c(s))-\epsilon_Q(s,a)
    \\
    \Rightarrow [T^* v^{\pi_E}](s) &\leq [T^{\pi_c}
    v^{\pi_E}](s) + \max_{a\in \A}\epsilon_Q(s,a)-\min_{a\in
    \A}\epsilon_Q(s,a).
  \end{align}
  Taking the expectation according to $\rho_E$ and noticing that
  $T^* v^{\pi_E}\geq v^{\pi_E}$, we bound the first
  term:
  \begin{equation}
    0 \leq \E_{s\sim\rho_E}\left[ [T^* v^{\pi_E}](s) - [T^{\pi_c}
    v^{\pi_E}](s)\right] \leq \bar{\epsilon}_Q.
    \label{eq:proof:b1}
  \end{equation}
  There finally remains to bound the term $\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) -
  v^{\pi_E}(s)]$.

  Let us write $M\in\mathbb{R}^{|\s|\times |\s|}$ the diagonal matrix
  defined as $M = \diag (\chi_{\{\pi_c(s)\neq\pi_E(s)\}})$. Using
  this, the Bellman operator $T^{\pi_c}$ may be written as, for any $v\in\mathbb{R}^\s$:
  \begin{equation}
    T^{\pi_c}v = \R + \gamma M P_{\pi_c} v + \gamma (I-M)P_{\pi_E} v
    = \R + \gamma P_{\pi_E} v + \gamma M (P_{\pi_c}-P_{\pi_E})v.
  \end{equation}
  Applying this operator to $v^{\pi_E}$ and recalling that $\R +
  \gamma P_{\pi_E} v^{\pi_E} = T^{\pi_E} v^{\pi_E} = v^{\pi_E}$, we
  get:
  \begin{equation}
    T^{\pi_c}v^{\pi_E} - v^{\pi_E} = \gamma M
    (P_{\pi_c}-P_{\pi_E})v^{\pi_E}
    \Rightarrow |\rho_E^\top (T^{\pi_c}v^{\pi_E} - v^{\pi_E})| = \gamma
    |\rho_E^\top M (P_{\pi_c}-P_{\pi_E})v^{\pi_E}|.
  \end{equation}
  One can easily see that $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
  \leq \frac{2}{1-\gamma}\|\R\|_\infty$, which allows bounding the
  last term:
  \begin{equation}
    |\E_{s\sim\rho_E}[[T^{\pi_c}v^{\pi_E}](s) - v^{\pi_E}(s)]| \leq
    \epsilon_c \frac{2\gamma}{1-\gamma} \|\R\|_\infty.
    \label{eq:proof:b2}
  \end{equation}
  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
  and~\eqref{eq:proof:b2} into Eq.~\eqref{eq:proof:residual} gives the
  stated result.%, which concludes the proof.
%  Injecting bounds of Eqs.~\eqref{eq:proof:b1}
%  and~\eqref{eq:proof:b2} into Eq.~\eqref{eq:proof:residual} using the
%  decomposition of Eq.~\eqref{eq:proof:decomposition} gives the
%  stated result:
%  \begin{equation}
%    0\leq\E_{s\sim\rho_E}[v^*(s)-v^{\pi_E}(s)] \leq
%    \frac{C_f}{1-\gamma} \left(\bar{\epsilon}_Q + \epsilon_c \frac{2\gamma}{1-\gamma} \|\R\|_\infty
%    \right).
%  \end{equation}
%  This concludes the proof.
\end{proof}

This result shows that if the expert feature expectation is well
estimated (in the sense that the estimation error $\epsilon_\mu$ is
small for states sampled according to the expert stationary policy
and for all actions) and if the classification error $\epsilon_c$ is
small, then the proposed generic algorithm outputs a reward function
$\R_{\theta_c}$ for which the expert policy will be near optimal. A
direct corollary of Th.~\ref{th} is that given the true expert
feature expectation $\mu^{\pi_E}$ and a perfect classifier
($\epsilon_c=0$), $\pi_E$ is the unique optimal policy for
$\R_{\theta_c}$.

One may argue that this bounds trivially holds for the null reward
function (a reward often exhibited to show that IRL is an ill-posed
problem), obtained if $\theta_c=0$. However, recall that the
parameter vector $\theta_c$ is computed by the classifier. With
$\theta_c=0$, the decision rule would be a random policy and we
would have $\epsilon_c = \frac{|\A|-1}{|\A|}$, the worst possible
classification error. This case is really unlikely. Therefore, we
advocate that the proposed approach somehow allows disambiguating
the IRL problem (at least, it does not output trivial reward
functions such as the null vector).
%
Also, this bound is scale-invariant: one could impose
$\|\theta_c\|=1$ or normalize (action-) value functions by
$\|\R_{\theta_c}\|_\infty^{-1}$.

One should notice that there is a hidden dependency of the
classification error $\epsilon_c$ to the estimated expert feature
expectation $\hat{\mu}^{\pi_E}$. Indeed, the minimum classification
error depends on the hypothesis space spanned by the chosen score
function basis functions for the MC$^2$ algorithm (here
$\hat{\mu}^{\pi_E}$). Nevertheless, provided a good representation
for the reward function (that is a good choice of basis functions
$\phi_i$) and a small estimation error, this should not be a
practical problem.

Finally, if our bound relies on the generalization errors
$\epsilon_c$ and $\bar{\epsilon}_Q$, the classifier will only use
$(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ in the training
phase, where $s_i$ are the states from the set $\D$. It outputs
$\theta_c$, seen as a reward function, thus the estimated feature
expectation $\hat{\mu}^{\pi_E}$ is no longer required. Therefore,
practically it should be sufficient to estimate well
$\hat{\mu}^{\pi_E}$ on state-action couples $(s_i,a)_{1\leq i\leq
N,a\in\A}$, which allows envisioning Monte-Carlo rollouts for
example.

\section{A Practical Approach}
\label{sec:practicalApproach}

\subsection{Estimating the Expert Feature Expectation}
\label{subsec:practicalApproach:muE}

SCIRL relies on an estimate $\hat{\mu}^{\pi_E}$ of the expert
feature expectation. Basically, this is a policy evaluation problem.
An already made key observation is that each component of
$\mu^{\pi_E}$ is the action-value function of $\pi_E$ for a reward
function $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. We briefly review its exact
computation and possible estimation approaches, and consider
possible heuristics.
%possibly considering some heuristics.

If the model is known, the feature expectation can be computed
explicitly. Let $\Phi\in\mathbb{R}^{|\s|\times p}$ be the feature
matrix whose rows contain the feature vectors $\phi(s)^\top$ for all
$s\in\s$.
%
%(equivalently, whose columns contain the basis functions $\phi_i$
%for $1\leq i\leq p$).
%
For a fixed $a\in A$, let $\Mu^{\pi_E}_a \in\mathbb{R}^{|\s|\times
p}$ be the feature expectation matrix whose rows are the expert
feature vectors, that is $(\mu^{\pi_E}(s,a))^\top$ for any $s\in\s$.
With these notations, we have $\Mu_a^{\pi_E} = \Phi + \gamma
P_a(I-\gamma P_{\pi_E})^{-1} \Phi$.
%\begin{equation}
%  \forall a\in\A,\; \Mu_a^{\pi_E} = \Phi + \gamma P_a(I-\gamma
%  P_{\pi_E})^{-1} \Phi.
%\end{equation}
%Therefore, knowing the model (the dynamic), the expert feature
%expectation can be easily computed.
Moreover, the related computational cost is the same order of
magnitude as evaluating a single policy (as the costly part,
computing $(I-\gamma P_{\pi_E})^{-1}$, is shared by all components).

If the model is unknown, any temporal difference learning algorithm
can be used to estimate the expert feature
expectation~\cite{Klein:2011}, as LSTD (Least-Squares Temporal
Differences)~\cite{Bradtke:1996}. Let $\psi:\s\times \A \rightarrow
\mathbb{R}^d$ be a feature vector composed of $d$ basis functions
$\psi_i \in\mathbb{R}^{\s\times \A}$. Each component $\mu_i^{\pi_E}$
of the expert feature expectation is parameterized by a vector
$\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
\psi(s,a)$. Assume that we have a training set
$\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ with actions
$a_i$ not necessarily sampled according to policy $\pi_E$
(\textit{e.g.}, this may be obtained by sampling trajectories
according to an expert-based $\epsilon$-greedy policy), the aim
being to have a better variability of tuples (non-expert actions
should be tried). Let $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
(resp. $\tilde{\Psi}'$) be the feature matrix whose rows are the
feature vectors $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
Let also $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ be the feature
matrix whose rows are the reward's feature vectors $\phi(s_i)^\top$.
Finally, let $\Xi =
\begin{bmatrix}   \xi_1 & \dots & \xi_p
\end{bmatrix}\in\mathbb{R}^{d\times p}$ be the matrix of all
parameter vectors. Applying LSTD to each component of the feature
expectation gives the LSTD-$\mu$ algorithm~\cite{Klein:2011}: $\Xi =
(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
\tilde{\Psi}'))^{-1}\tilde{\Psi}^\top \tilde{\Phi}$ and
$\hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a)$.
%\begin{equation}
%  \Xi = \left(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
%  \tilde{\Psi}')\right)^{-1}\tilde{\Psi}^\top \tilde{\Phi}
%  \text{ and } \hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a).
%\end{equation}
As for the exact case, the costly part (computing the inverse
matrix) is shared by all feature expectation components, the
computational cost is reasonable (same order as LSTD).

%
Provided a simulator and the ability to sample according to the
expert policy, the expert feature expectation may also be estimated
using Monte-Carlo rollouts for a given state-action pair (as noted
in Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ need only
be known on $(s_i,a)_{1\leq i\leq N,a\in\A}$). Assuming that $K$
trajectories are sampled for each required state-action pair, this
method would require $KN|\A|$ rollouts.

In order to have a small error $\bar{\epsilon}_Q$, one may learn
using transitions whose starting state is sampled according to
$\rho_E$ and whose actions are uniformly distributed. However, it
may happen that only transitions of the expert are available: $\T =
\{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$. If the state-action
couples $(s_i,a_i)$ may be used to feed the classifier, the
transitions $(s_i,a_i,s'_i)$ are not enough to provide an accurate
estimate of the feature expectation. In this case, we can still
expect an accurate estimate of $\mu^{\pi_E}(s,\pi_E(s))$, but there
is little hope for $\mu^{\pi_E}(s,a\neq\pi_E(s))$. However, one can
still rely on some heuristic; this does not fit the analysis of
Sec.~\ref{subsec:scirl:analysis}, but it can still provide good
experimental results, as illustrated in Sec.~\ref{sec:experiments}.

We propose such a heuristic. Assume that only data $\T$ is available
and that we use it to provide an (accurate) estimate
$\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (this basically means estimating a
value function instead of an action-value function as described
above). We may adopt an optimistic point of view by assuming that
applying a non-expert action just delays the effect of the expert
action. More formally, we associate to each state $s$ a virtual
state $s_\text{v}$ for which $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ for
any action $a$ and for which the reward feature expectation is the
null vector, $\phi(s_v) = 0$. In this case, we have
$\mu^{\pi_E}(s,a\neq\pi_E(s)) = \gamma \mu^{\pi_E}(s,\pi_E(s))$.
Applying this idea to the available estimate (recalling that the
classifiers only requires evaluating $\hat{\mu}^{\pi_E}$ on
$(s_i,a)_{1\leq i\leq N,a\in \A}$) provides the proposed heuristic:
for $1\leq i\leq N$, $\hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
\hat{\mu}^{\pi_E}(s_i,a_i)$.
%\begin{equation}
%  \forall 1\leq i \leq N,\;
%  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
%  \hat{\mu}^{\pi_E}(s_i,a_i).
%  \label{eq:heuristic}
%\end{equation}

We may even push this idea further, to get the simpler estimate of
the expert feature expectation (but with the weakest guarantees).
Assume that the set $\T$ consists of one long trajectory, that is
$s'_i = s_{i+1}$ (thus $\T =
\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). We may estimate
$\mu^{\pi_E}(s_i,a_i)$ using the single rollout available in the
training set and use the proposed heuristic for other actions:
\begin{equation}
  \forall 1\leq i \leq N,\; \hat{\mu}^{\pi_E}(s_i,a_i) =
  \sum_{j=i}^N \gamma^{j-i}\phi(s_j) \text{ and }
  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
  \hat{\mu}^{\pi_E}(s_i,a_i).
  \label{eq:mc_plus_heuristic}
\end{equation}

To sum up, the expert feature expectation may be seen as a vector of
action-value functions (for the same policy $\pi_E$ and different
reward functions $\phi_i$). Consequently, any action-value function
evaluation algorithm may be used to estimate $\mu^\pi(s,a)$.
Depending on the available data, one may have to rely on some
heuristic to assess the feature expectation for a unexperienced
(non-expert) action. Also, this expert feature expectation estimate
is only required for training the classifier, so it is sufficient to
estimate on state-action couples $(s_i,a)_{1\leq i \leq N,a\in \A}$.
In any case, estimating $\mu^{\pi_E}$ is not harder than estimating
the action-value function of a given policy in the on-policy case,
which is much easier than computing an optimal policy for an
arbitrary reward function (as required by most of existing IRL
algorithms, see Sec.~\ref{sec:relatedWorks}).

\subsection{An Instantiation}
\label{subsec:practicalApproach:instantiation}

As stated before, any MC$^2$ algorithm may be used. Here, we choose
the structured large margin approach~\cite{Taskar:2005}. Let
$\lc:\s\times\A\rightarrow\mathbb{R}_+$ be a user-defined margin
function satisfying $\lc(s,\pi_E(s))\leq \lc(s,a)$ (here,
$\lc(s_i,a_i)=0$ and $\lc(s_i,a\neq a_i)=1$). The MC$^2$ algorithm
solves:% the following quadratic problem:
\begin{equation}
  \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~~~~s.t.~~~~} \forall i,
  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a). \label{eq:qp_taskar}
\end{equation}
Following~\cite{Ratliff:2006}, we express the equivalent hinge-loss
form (noting that the slack variables $\zeta_i$ are tight, which
allows moving the constraints in the objective function):
\begin{equation}
  J(\theta) = \frac{1}{N}\sum_{i=1}^N \max_a \theta^\top
  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a) -
  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i) +
  \frac{\lambda}{2}\|\theta\|^2.
\end{equation}
This objective function is minimized using a subgradient descent.
The expert feature expectation is estimated using the scheme
described in Eq.~\eqref{eq:mc_plus_heuristic}.

%This is the instantiation we use in Sec.~\ref{sec:experiments}, but
%our approach is more general: other choices could have been done for
%the MC$^2$ algorithm as well as for the expert feature expectation
%estimation. Notice that if the model is known, $\mu^{\pi_E}$ can be
%exactly computed ($\bar{\epsilon}_Q=0$). Assuming that the reward
%feature vector is rich enough (in the sense that there exists a
%parameter vector such that $\pi_E$ is the unique optimal policy),
%constraints of Eq.~\eqref{eq:qp_taskar} can be satisfied for all
%state-action couples. This means that the proposed approach is able
%to produce a non-trival reward for which the expert policy is the
%unique optimal policy, and this without solving any direct problem.

\section{Related Works}
\label{sec:relatedWorks}

The notion of IRL has first been introduced in~\cite{Russell:1998}
and first been formalized in~\cite{Ng:2000}. A classic approach to
IRL, initiated in~\cite{Abbeel:2004}, consists in finding a policy
(through some reward function) such that its feature expectation (or
more generally some measure of the underlying trajectories'
distribution) matches the one of the expert policy.
%
See~\cite{Neu:2010} for a review.
%
%This can be done based on game theory~\cite{Syed:2008:game}, linear
%programming~\cite{Syed:2008:lp}, maximum
%entropy~\cite{Ziebart:2008}, Bayesian posterior
%maximization~\cite{Ramachandran:2007}, \textit{etc}. Some of these
%approaches and others are nicely reviewed in~\cite{Neu:2010}.
%
Notice that related algorithms are not always able to output a
reward function, even if they may make use of IRL as an intermediate
step. In such case, they are usually refereed to as apprenticeship
learning algorithms.
%
%Notice that related algorithms do not always output a reward
%function (\textit{e.g.}, in~\cite{Abbeel:2004}, the algorithm
%produces a mixed policy; each intermediate policy is optimal
%according to some reward function, but the outputted policy does not
%correspond to a specific reward). In such case, they are refereed to
%as ``apprenticeship learning'' algorithms (they may make use of IRL
%as intermediate steps but do not output a reward).

Closer to our contribution, some approaches also somehow introduce a
structure in a classification
procedure~\cite{Melo:2010}\cite{Ratliff:2006}. In~\cite{Melo:2010},
a metric induced by the MDP is used to build a kernel which is used
in a classification algorithm, showing improvements compared to a
non-structured kernel. However, this approach is not an IRL
algorithm, and more important assessing the metric of an MDP is a
quite involved problem. In~\cite{Ratliff:2006}, a classification
algorithm is also used to produce a reward function. However,
instead of associating actions to states, as we do, it associates
optimal policies (labels) to MDPs (inputs), which is how the
structure is incorporated. This involves solving many MDPs.

As far as we know, all IRL algorithms require solving the direct RL
problem repeatedly, except~\cite{Dvij:2010,boularias:2011}.
\cite{Dvij:2010} applies to linearly-solvable MDPs (where the
control is done by imposing any dynamic to the system).
In~\cite{boularias:2011}, based on a relative entropy argument, some
utility function is maximized using a subgradient ascent. Estimating
the subgradient requires sampling trajectories according to the
policy being optimal for the current estimated reward. This is
avoided thanks to the use of importance sampling. Still, this
requires sampling trajectories according to a non-expert policy and
the direct problem remains at the core of the approach (even if
solving it is avoided).

SCIRL does not require solving the direct problem, just estimating
the feature expectation of the expert policy. In other words,
instead of solving multiple policy optimization problems, we only
solve one policy evaluation problem. This comes with theoretical
guarantees (which is not the case of all IRL algorithms,
\textit{e.g.}~\cite{boularias:2011}). Moreover, using heuristics
which go beyond our analysis, SCIRL may rely solely on data provided
by expert trajectories. We demonstrate this empirically in the next
section. To the best of our knowledge, no other IRL algorithm can
work in such a restrictive case.

\section{Experiments}
\label{sec:experiments}

We illustrate the proposed approach on a car driving simulator,
similar to~\cite{Abbeel:2004,Syed:2008:game}. The goal si to drive a
car on a busy three-lane highway with randomly generated traffic
(driving off-road is allowed on both sides). The car can move left
and right, accelerate, decelerate and keep a constant speed. The
expert optimizes a handcrafted reward $\R_E$ which favours speed,
punish off-road, punish collisions even more  and is neutral
otherwise.

We compare SCIRL as instantiated in
Sec.~\ref{subsec:practicalApproach:instantiation} to the
unstructured classifier (using the same classification algorithm)
and to the algorithm of~\cite{Abbeel:2004} (called here PIRL for
Projection IRL). We also consider the optimal behavior according to
a randomly sampled reward function as a baseline (using the same
reward feature vector as SCIRL and PIRL, the associated parameter
vector is randomly sampled).

For SCIRL and PIRL we use a discretization of the state space as the
reward feature vector, $\phi\in\mathbb{R}^{729}$: $9$ horizontal
positions for the user's car, $3$ horizontal and $9$ vertical
positions for the closest traffic's car and $3$ speeds. Notice that
these features are much less informative than the ones used
in~\cite{Abbeel:2004,Syed:2008:game}. Actually,
in~\cite{Syed:2008:game} features are so informative that sampling a
random positive parameter vector $\theta$ already gives an
acceptable behavior. The discount factor is $\gamma = 0.9$. The
classifier uses the same feature vector reproduced for each action.

SCIRL is fed with $n$ trajectories of length $n$ (started in a
random state) with $n$ varying from $3$ to $20$ (so fed with $9$ to
$400$ transitions). Each experiment is repeated 50 times. The
 classifier uses the same data. PIRL is an iterative
algorithm, each iteration requiring to solve the MDP for some reward
function. It is run for 70 iterations, all required objects (a
feature expectations for a non-expert policy and an optimal policy
according to some reward function at each iteration) are computed
exactly using the model. We measure the performance of each approach
with $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, where $\mathcal{U}$
is the uniform distribution (this allows measuring the
generalization capability of each approach for states infrequently
encountered), $\R_E$ is the expert reward and $\pi$ is one of the
following polices: the optimal policy for $\R_E$ (upper baseline),
the optimal policy for a random reward (lower baseline), the optimal
policy for $\R_{\theta_c}$ (SCIRL), the policy produced by PIRL and
the classifier decision rule.

\begin{figure}
\begin{minipage}[l]{0.5\linewidth}
\centering
  \centerline{\includegraphics[width=.92\linewidth]{fig_classif.pdf}}
\end{minipage} \hfill
\begin{minipage}[r]{0.5\linewidth}
\centering
  \centerline{\includegraphics[width=.92\linewidth]{fig_scirl.pdf}}
\end{minipage}
  \caption{Highway problem. The highest line is the expert value. For each curves,
  we show the mean (plain line), the standard deviation (dark color) and the min-max values (light color).
  The policy corresponding to the random reward is in blue, the policy outputted by the  classifier is
  in yellow and the optimal policy according the SCIRL's reward is in red. PIRL is the dark blue line.}
  \label{fig:res}
\end{figure}

Fig.~\ref{fig:res} shows the performance of each approach as a
number of used expert transitions (except PIRL which uses the
model). We can see that the classifier does not work well on this
example. Increasing the number of samples would improve its
performance, but after 400 transitions it does not work as well as
SCIRL with only a ten of transitions. SCIRL works pretty well here:
after only a hundred of transitions it reaches the performance of
PIRL, both being close to the expert value. We do not report exact
computational times, but running SCIRL one time with $400$
transitions is approximately hundred time faster than running PIRL
for $70$ iteration.
%
% (this is due to the need to repeatedly solve MDPs for PIRL).

%Recall that for SCIRL the instantiation of
%Sec.~\ref{subsec:practicalApproach:instantiation} is considered,
%with a single Monte-Carlo rollout for estimating
%$\hat{\mu}^{\pi_E}(s_i,a_i)$ and using the proposed heuristic for
%$\hat{\mu}^{\pi_E}(s_i,a\neq a_i)$, which goes beyond our analysis.
%To our knowledge, it is the only algorithm working (at least on this
%benchmark) only with data from the expert.

\section{Conclusion}
\label{sec:conclusion}

We have introduced a new way to perform IRL by structuring a
linearly parameterized score function-based multi-class
classification algorithm with an estimate of the expert feature
expectation. This outputs a reward function for which we have shown
the expert to be near optimal, provided a small classification error
and a good expert feature expectation estimate. How to practically
estimate this quantity has been  discussed and we have introduced a
heuristic for the case where only transitions from the expert are
available, along with a specific instantiation of the SCIRL
algorithm. We have shown on a car driving simulator benchmark that
the proposed approach works well (even combined with the introduced
heuristic), much better than the unstructured classifier and as well
as a state-of-the-art algorithm making use of the model (and with a
much lower computational time). In the future, we plan to deepen the
theoretical properties of SCIRL (notably regarding possible
heuristics) and to apply it to real-world robotic problems.

\paragraph*{Acknowledgments.}

This research was partly funded by the EU FP7 project ILHAIRE (grant
n°270780), by the EU INTERREG IVa project ALLEGRO and by the Région
Lorraine (France).

\newpage
%\small
\bibliographystyle{plain}%{unsrt}
\bibliography{scirl_nips}
\end{document}

### 2012-11-23 klein2012structured_EWRL/Rebuttal.txt
Déplacement du dossier klein2012 en klein2012structured_EWRL

--- 2012-04-12 klein2012/Rebuttal.txt
@@
-We would like to thank the reviewers for their thorough reading and their remarks.
-We wish to raise the following points to somewhat clarify aspects of our work the reviewers showed concern about.
+We would like to thank the reviewers for their thorough reading and remarks. We wish to somewhat clarify some points the reviewers showed concern about.
 
@@
 As each component shares the same dynamic, the computational cost is the one of performing LSTD on one component (in O(p^3)), see Klein et al.
-Also, we have the same guarantees for estimating mu_E from expert trajectories than for estimating a Q-function with LSTD in an on-policy setting (e.g., see Lazaric et al, 2010).
-In Klein et al.
-it was used in its off-policy form because of the requirement of the algorithm of (Abbeel et al.
-2004) to compute \mu for an arbitrary policy.
-The on-policy setting considered in our contribution only requires data from the expert.
+We have the same guarantees for estimating mu_E from expert trajectories than for estimating a Q-function with LSTD in an on-policy setting (e.g., see Lazaric et al.).
+Klein et al.
+used it in its off-policy form because of the requirement of the algorithm of (Abbeel et al.
+2004) to compute mu for an arbitrary policy.
+The on-policy setting considered in our work only requires data from the expert.
 
-- As mentionned by reviewer 2, our algorithm is inspired by MMP but differs in several ways.
-We are not merely combining existing components as assessed by reviewer 3, but introducing a new way to look at the IRL problem.
+- As mentioned by reviewer 2, our algorithm is inspired by MMP but differs in several ways. We are not merely combining existing components as assessed by reviewer 3.
  - All approaches build on the structured large margin framework of Taskar et al., but they solve quite different problems.
@@
  - We are not merely replacing mu with an approximation based on \mu_E (this is Klein et al).
-We introduce a new way to look at the score function q, which allows addressing two major problems of IRL.
+We introduce a new way to look at the score function q, which allows us to minimize the risk function using only \mu_E evaluated only in states present in the expert's dataset, which can be estimated solely from expert transitions.
+MMP needs a lot more information.
  - Any classifier (and not only the structured margin framework of Taskar et al. used here) using such a parameterized score function could be plugged in our algorithm.
- - The way we express the score function q allows us to minimize the risk function using only the feature expectation of the expert evaluated only in states present in the expert's dataset, which can be estimated solely from expert transitions.
-MMP needs a lot more information to achieve its goal.
 
-- The lack of quantitative results mentionned by reviewer 2 is addressed with the pendulum experiment, we show that our method works (quantitative results given line 657) in conditions where state of the art algorithms fail due to lack of information.
+- The lack of quantitative results mentionned by reviewer 2 is addressed lines 612 and 657, we show that SCIRL works in conditions where state of the art algorithms fail due to lack of information.
 Klein et al.'s results (not recalled in our paper due to space constraints) show that the algorithm of Abbeel and Ng needs around 3000 more arbitrary samples (not the expert's) in order to work (this is due to the off-policy learning aspect).
-We lift this requirement with SCIRL, which is to the best of our knowledge the only algorithm that works in such drastic conditions.
+We lift this requirement with SCIRL which is to the best of our knowledge the only algorithm that works in such drastic conditions.
 
-- We recognize that the assumptions of the analysis are strong (as discussed in the paper).
+- As discussed in the paper, we recognize that the assumptions of the analysis are strong.
 However, we show that the algorithm computes a reward for which the expert policy is the unique optimal policy, which is not a straightforward problem (even knowing the dynamic and the expert policy).
-Given that recovering a reward is the problem we address, we think it is valuable, despite the strong assumptions.
+Given that recovering a reward is the problem we address, we think it is valuable despite the strong assumptions.
 
 - We show that our algorithm stands comparison when put in competition with MWAL.
-The latter has access to the system's dynamics whereas our algorithm only uses the expert's transitions.
-In the light of this, marginally subpar results are indeed a contribution, indeed our algorithm can be used in cases where the lack of information would prohibit the use of MWAL.
+The latter needs access to the system's dynamics whereas our algorithm only uses the expert's transitions.
+Marginally subpar results thus are a contribution, indeed SCIRL can be used in cases where the lack of information would prohibit the use of MWAL.
 
 - In addition to needing far less information than state of the art methods, by using a LSTD-like method once and for all at the beginning, our algorithm is also computationally more efficient than others that require solving one or more MDP at each iteration.
-Indeed, solving a MDP implies (e.g.
-if one uses LSPI) using LSTD more than once.
 We empirically notice a ten-fold gain on computation time on a desktop computer compared to MWAL.

### 2012-11-23 klein2012structured_EWRL/jmlr-sample.tex
Déplacement du dossier klein2012 en klein2012structured_EWRL

%\documentclass[gray]{jmlr} % test grayscale version
\documentclass[tablecaption=bottom]{jmlr}

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

%\usepackage{rotating}% for sideways figures and tables
%\usepackage{longtable}% for long tables

% The booktabs package is used by this sample document
% (it provides \toprule, \midrule and \bottomrule).
% Remove the next line if you don't require it.
\usepackage{booktabs}
% The siunitx package is used by this sample document
% to align numbers in a column by their decimal point.
% Remove the next line if you don't require it.
\usepackage[load-configurations=version-1]{siunitx} % newer version
%\usepackage{siunitx}

% The following command is just for this sample document:
\newcommand{\cs}[1]{\texttt{\char`\\#1}}

\jmlrvolume{vol}
\jmlryear{2010}
\jmlrsubmitted{submission date}
\jmlrpublished{publication date}

\title[Short Title]{Full Title of Article}

 % Use \Name{Author Name} to specify the name.
 % If the surname contains spaces, enclose the surname
 % in braces, e.g. \Name{John {Smith Jones}} similarly
 % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
 % If the first letter in the forenames is a diacritic
 % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

 % \thanks must come after \Name{...} not inside the argument for
 % example \Name{John Smith}\thanks{A note} NOT \Name{John
 % Smith\thanks{A note}}

 % Two authors with the same address
  \author{\Name{Author Name1}\thanks{A note} \Email{abc@sample.com}\and
   \Name{Author Name2} \Email{xyz@sample.com}\\
   \addr Address}

 % Three or more authors with the same address:
 % \author{\Name{Author Name1} \Email{an1@sample.com}\\
 %  \Name{Author Name2} \Email{an2@sample.com}\\
 %  \Name{Author Name3} \Email{an3@sample.com}\\
 %  \Name{Author Name4} \Email{an4@sample.com}\\
 %  \Name{Author Name5} \Email{an5@sample.com}\\
 %  \Name{Author Name6} \Email{an6@sample.com}\\
 %  \Name{Author Name7} \Email{an7@sample.com}\\
 %  \Name{Author Name8} \Email{an8@sample.com}\\
 %  \Name{Author Name9} \Email{an9@sample.com}\\
 %  \Name{Author Name10} \Email{an10@sample.com}\\
 %  \Name{Author Name11} \Email{an11@sample.com}\\
 %  \Name{Author Name12} \Email{an12@sample.com}\\
 %  \Name{Author Name13} \Email{an13@sample.com}\\
 %  \Name{Author Name14} \Email{an14@sample.com}\\
 %  \addr Address}

 % Authors with different addresses:
 % \author{\Name{Author Name1} \Email{abc@sample.com}\\
 % \addr Address 1
 % \AND
 % \Name{Author Name2} \Email{xyz@sample.com}\\
 % \addr Address 2
 %}

\editor{Editor's name}

\begin{document}

\maketitle

\begin{abstract}
This is the abstract for this article.
\end{abstract}
\begin{keywords}
List of keywords
\end{keywords}

\section{Introduction}
\label{sec:intro}

This is a sample article that uses the \textsf{jmlr} class with
the \texttt{wcp} class option.  Please follow the guidelines in
this sample document as it can help to reduce complications when
combining the articles into a book. Please avoid using obsolete
commands, such as \verb|\rm|, and obsolete packages, such as
\textsf{epsfig}.\footnote{See
\url{http://www.ctan.org/pkg/l2tabu}} Please also ensure
that your document will compile with PDF\LaTeX.

\subsection{Sub-sections}

Sub-sections are produced using \verb|\subsection|.

\subsubsection{Sub-sub-sections}

Sub-sub-sections are produced using \verb|\subsubsection|.

\paragraph{Sub-sub-sub-sections}

Sub-sub-sub-sections are produced using \verb|\paragraph|.
These are unnumbered with a running head.

\subparagraph{Sub-sub-sub-sub-sections}

Sub-sub-sub-sub-sections are produced using \verb|\subparagraph|.
These are unnumbered with a running head.

\section{Cross-Referencing}

Always use \verb|\label| and \verb|\ref| (or one of the commands
described below) when cross-referencing.  For example, the next
section is Section~\ref{sec:math}. The \textsf{jmlr} class
provides some convenient cross-referencing commands:
\verb|\sectionref|, \verb|\equationref|, \verb|\tableref|,
\verb|\figureref|, \verb|\algorithmref|, \verb|\theoremref|,
\verb|\lemmaref|, \verb|\remarkref|, \verb|\corollaryref|,
\verb|\definitionref|, \verb|\conjectureref|, \verb|\axiomref|,
\verb|\exampleref| and \verb|\appendixref|. The argument of these
commands may either be a single label or a comma-separated list
of labels. Examples:

Referencing sections: \sectionref{sec:math} or
\sectionref{sec:intro,sec:math} or
\sectionref{sec:intro,sec:math,sec:tables,sec:figures}.

Referencing equations: \equationref{eq:trigrule} or
\equationref{eq:trigrule,eq:df} or
\equationref{eq:trigrule,eq:f,eq:df,eq:y}.

Referencing tables: \tableref{tab:operatornames} or
\tableref{tab:operatornames,tab:example} or
\tableref{tab:operatornames,tab:example,tab:example-booktabs}.

Referencing figures: \figureref{fig:nodes} or
\figureref{fig:nodes,fig:teximage} or
\figureref{fig:nodes,fig:teximage,fig:subfigex} or
\figureref{fig:circle,fig:square}.

Referencing algorithms: \algorithmref{alg:gauss} or
\algorithmref{alg:gauss,alg:moore} or
\algorithmref{alg:gauss,alg:moore,alg:net}.

Referencing theorem-like environments: \theoremref{thm:eigenpow},
\lemmaref{lem:sample}, \remarkref{rem:sample}, 
\corollaryref{cor:sample}, \definitionref{def:sample},
\conjectureref{con:sample}, \axiomref{ax:sample} and
\exampleref{ex:sample}.

Referencing appendices: \appendixref{apd:first} or
\appendixref{apd:first,apd:second}.

\section{Equations}
\label{sec:math}

The \textsf{jmlr} class loads the \textsf{amsmath} package, so
you can use any of the commands and environments defined there.
(See the \textsf{amsmath} documentation for further
details.\footnote{Either \texttt{texdoc amsmath} or
\url{http://www.ctan.org/pkg/amsmath}})

Unnumbered single-lined equations should be displayed using
\verb|\[| and \verb|\]|. For example:
\[E = m c^2\]
Numbered single-line equations should be displayed using the
\texttt{equation} environment. For example:
\begin{equation}\label{eq:trigrule}
\cos^2\theta + \sin^2\theta \equiv 1
\end{equation}
This can be referenced using \verb|\label| and \verb|\equationref|.
For example, \equationref{eq:trigrule}.

Multi-lined numbered equations should be displayed using the
\texttt{align} environment. For example:
\begin{align}
f(x) &= x^2 + x\label{eq:f}\\
f'(x) &= 2x + 1\label{eq:df}
\end{align}
Unnumbered multi-lined equations should be displayed using the
\texttt{align*} environment. For example:
\begin{align*}
f(x) &= (x+1)(x-1)\\
&= x^2 - 1
\end{align*}
If you want to mix numbered with unnumbered lines use the
align environment and suppress unwanted line numbers with
\verb|\nonumber|. For example:
\begin{align}
y &= x^2 + 3x - 2x + 1\nonumber\\
&= x^2 + x + 1\label{eq:y}
\end{align}
An equation that is too long to fit on a single line can be
displayed using the \texttt{split} environment. 
Text can be embedded in an equation using \verb|\text| or
\verb|\intertext| (as used in \theoremref{thm:eigenpow}).
See the \textsf{amsmath} documentation for further details.

\subsection{Operator Names}
\label{sec:op}

Predefined operator names are listed in \tableref{tab:operatornames}.
For additional operators, either use \verb|\operatorname|,
for example $\operatorname{var}(X)$ or declare it with
\verb|\DeclareMathOperator|, for example
\begin{verbatim}
\DeclareMathOperator{\var}{var}
\end{verbatim}
and then use this new command. If you want limits that go above and
below the operator (like \verb|\sum|) use the starred versions
(\verb|\operatorname*| or \verb|\DeclareMathOperator*|).

\begin{table*}[htbp]
\floatconts
  {tab:operatornames}%
  {\caption{Predefined Operator Names (taken from 
   \textsf{amsmath} documentation)}}%
  {%
\begin{tabular}{rlrlrlrl}
\cs{arccos} & $\arccos$ &  \cs{deg} & $\deg$ &  \cs{lg} & $\lg$ &  \cs{projlim} & $\projlim$ \\
\cs{arcsin} & $\arcsin$ &  \cs{det} & $\det$ &  \cs{lim} & $\lim$ &  \cs{sec} & $\sec$ \\
\cs{arctan} & $\arctan$ &  \cs{dim} & $\dim$ &  \cs{liminf} & $\liminf$ &  \cs{sin} & $\sin$ \\
\cs{arg} & $\arg$ &  \cs{exp} & $\exp$ &  \cs{limsup} & $\limsup$ &  \cs{sinh} & $\sinh$ \\
\cs{cos} & $\cos$ &  \cs{gcd} & $\gcd$ &  \cs{ln} & $\ln$ &  \cs{sup} & $\sup$ \\
\cs{cosh} & $\cosh$ &  \cs{hom} & $\hom$ &  \cs{log} & $\log$ &  \cs{tan} & $\tan$ \\
\cs{cot} & $\cot$ &  \cs{inf} & $\inf$ &  \cs{max} & $\max$ &  \cs{tanh} & $\tanh$ \\
\cs{coth} & $\coth$ &  \cs{injlim} & $\injlim$ &  \cs{min} & $\min$ \\
\cs{csc} & $\csc$ &  \cs{ker} & $\ker$ &  \cs{Pr} & $\Pr$
\end{tabular}\par
\begin{tabular}{rlrl}
\cs{varlimsup} & $\varlimsup$ 
& \cs{varinjlim} & $\varinjlim$\\
\cs{varliminf} & $\varliminf$ 
& \cs{varprojlim} & $\varprojlim$
\end{tabular}
}
\end{table*}

\section{Vectors and Sets}
\label{sec:vec}

Vectors should be typeset using \cs{vec}. For example $\vec{x}$.
The \textsf{jmlr} class also provides \cs{set} to typeset a
set. For example $\set{S}$.

\section{Floats}
\label{sec:floats}

Floats, such as figures, tables and algorithms, are moving
objects and are supposed to float to the nearest convenient
location. Please don't force them to go in a particular place. In
general it's best to use the \texttt{htbp} specifier and don't
put the figure or table in the middle of a paragraph (that is
make sure there's a paragraph break above and below the float).
Floats are supposed to have a little extra space above and below
them to make them stand out from the rest of the text. This extra
spacing is put in automatically and shouldn't need modifying.

To ensure consistency, please \emph{don't} try changing the format of the caption by doing
something like:
\begin{verbatim}
\caption{\textit{A Sample Caption.}}
\end{verbatim}
or
\begin{verbatim}
\caption{\em A Sample Caption.}
\end{verbatim}
You can, of course, change the font for individual words or 
phrases, for example:
\begin{verbatim}
\caption{A Sample Caption With Some \emph{Emphasized Words}.}
\end{verbatim}

\subsection{Tables}
\label{sec:tables}

Tables should go in the \texttt{table} environment. Within this
environment use \verb|\floatconts| (defined by \textsf{jmlr})
to set the caption correctly and center the table contents.

\begin{table}[htbp]
% The first argument is the label.
% The caption goes in the second argument, and the table contents
% go in the third argument.
\floatconts
  {tab:example}%
  {\caption{An Example Table}}%
  {\begin{tabular}{ll}
  \bfseries Dataset & \bfseries Result\\
  Data1 & 0.12345\\
  Data2 & 0.67890\\
  Data3 & 0.54321\\
  Data4 & 0.09876
  \end{tabular}}
\end{table}

If you want horizontal rules you can use the \textsf{booktabs}
package which provides the commands \verb|\toprule|, 
\verb|\midrule| and \verb|\bottomrule|. For example, see
\tableref{tab:example-booktabs}.

\begin{table}[hbtp]
\floatconts
  {tab:example-booktabs}
  {\caption{A Table With Horizontal Lines}}
  {\begin{tabular}{ll}
  \toprule
  \bfseries Dataset & \bfseries Result\\
  \midrule
  Data1 & 0.12345\\
  Data2 & 0.67890\\
  Data3 & 0.54321\\
  Data4 & 0.09876\\
  \bottomrule
  \end{tabular}}
\end{table}

If you want vertical lines as well, you can't use the
\textsf{booktabs} commands as there'll be some unwanted gaps.
Instead you can use \LaTeX's \verb|\hline|, but the rows may
appear a bit cramped.  You can add extra space above or below a
row using \verb|\abovestrut| and \verb|\belowstrut|. For example,
see \tableref{tab:example-hline}.

\begin{table}[htbp]
\floatconts
  {tab:example-hline}
  {\caption{A Table With Horizontal and Vertical Lines}}%
  {%
    \begin{tabular}{|l|l|}
    \hline
    \abovestrut{2.2ex}\bfseries Dataset & \bfseries Result\\\hline
    \abovestrut{2.2ex}Data1 & 0.12345\\
    Data2 & 0.67890\\
    Data3 & 0.54321\\
    \belowstrut{0.2ex}Data4 & 0.09876\\\hline
    \end{tabular}
  }
\end{table}

If you want to align numbers on their decimal point, you can
use the \textsf{siunitx} package. For example, see
\tableref{tab:example-siunitx}. For further details see the
\textsf{siunitx} documentation\footnote{Either \texttt{texdoc
siunitx} or \url{http://www.ctan.org/pkg/siunitx}}.

\begin{table}[htbp]
\floatconts
  {tab:example-siunitx}
  {\caption{A Table With Numbers Aligned on the Decimal Point}}
  {\begin{tabular}{lS[tabformat=3.5]}
  \bfseries Dataset & {\bfseries Result}\\
  Data1 & 0.12345\\
  Data2 & 10.6789\\
  Data3 & 50.543\\
  Data4 & 200.09876
  \end{tabular}}
\end{table}

If the table is too wide, you can adjust the inter-column
spacing by changing the value of \verb|\tabcolsep|. For
example:
\begin{verbatim}
\setlength{\tabcolsep}{3pt}
\end{verbatim}
If the table is very wide but not very long, you can use the
\texttt{sidewaystable} environment defined in the
\textsf{rotating} package (so use \verb|\usepackage{rotating}|).
If the table is too long to fit on a page, you should use the
\texttt{longtable} environment defined in the \textsf{longtable}
package (so use \verb|\usepackage{longtable}|).

\subsection{Figures}
\label{sec:figures}

Figures should go in the \texttt{figure} environment. Within this
environment, use \verb|\floatconts| to correctly position the
caption and center the image. Use \verb|\includegraphics|
for external graphics files but omit the file extension. Do not
use \verb|\epsfig| or \verb|\psfig|. If you want to scale the
image, it's better to use a fraction of the line width rather
than an explicit length. For example, see \figureref{fig:nodes}.

\begin{figure}[htbp]
% Caption and label go in the first argument and the figure contents
% go in the second argument
\floatconts
  {fig:nodes}
  {\caption{Example Image}}
  {\includegraphics[width=0.5\linewidth]{images/nodes}}
\end{figure}

If your image is made up of \LaTeX\ code (for example, commands
provided by the \textsf{pgf} package) you can include it using
\cs{includeteximage} (defined by the \textsf{jmlr} class). This
can be scaled and rotated in the same way as \cs{includegraphics}.
For example, see \figureref{fig:teximage}.

\begin{figure}[htbp]
\floatconts
  {fig:teximage}
  {\caption{Image Created Using \LaTeX\ Code}}
  {\includeteximage[angle=45]{images/teximage}}
\end{figure}

If the figure is too wide to fit on the page, you can use the
\texttt{sidewaysfigure} environment defined in the
\textsf{rotating} package.

Don't use \verb|\graphicspath|. If the images are contained in
a subdirectory, specify this when you include the image, for
example \verb|\includegraphics{figures/mypic}|.

\subsubsection{Sub-Figures}
\label{sec:subfigures}

Sub-figures can be created using \verb|\subfigure|, which is
defined by the \textsf{jmlr} class. The optional argument allows
you to provide a subcaption. The label should be placed in the
mandatory argument of \verb|\subfigure|. You can reference the
entire figure, for example \figureref{fig:subfigex}, or you can
reference part of the figure using \verb|\figureref|, for example
\figureref{fig:circle}. Alternatively you can reference the
subfigure using \verb|\subfigref|, for example
\subfigref{fig:circle,fig:square} in \figureref{fig:subfigex}.

\begin{figure}[htbp]
\floatconts
  {fig:subfigex}
  {\caption{An Example With Sub-Figures.}}
  {%
    \subfigure[A Circle]{\label{fig:circle}%
      \includegraphics[width=0.2\linewidth]{images/circle}}%
    \qquad
    \subfigure[A Square]{\label{fig:square}%
      \includegraphics[width=0.2\linewidth]{images/square}}
  }
\end{figure}

By default, the sub-figures are aligned on the baseline.
This can be changed using the second optional argument
of \verb|\subfigure|. This may be \texttt{t} (top), \texttt{c}
(centered) or \texttt{b} (bottom). For example, the subfigures
\subfigref{fig:circle2,fig:square2} in \figureref{fig:subfigex2}
both have \verb|[c]| as the second optional argument.

\begin{figure}[htbp]
\floatconts
  {fig:subfigex2}
  {\caption{Another Example With Sub-Figures.}}
  {%
    \subfigure[A Small Circle][c]{\label{fig:circle2}%
      \includegraphics[width=0.1\linewidth]{images/circle}}%
    \qquad
    \subfigure[A Square][c]{\label{fig:square2}%
      \includegraphics[width=0.2\linewidth]{images/square}}
  }
\end{figure}

\subsection{Sub-Tables}
\label{sec:subtables}
There is an analogous command \verb|\subtable| for sub-tables.
It has the same syntax as \verb|\subfigure| described above.
You can reference the table using \verb|\tableref|, for example
\tableref{tab:subtabex} or you can reference part of the table,
for example \tableref{tab:ab}. Alternatively you can reference the
subtable using \verb|\subtabref|, for example
\subtabref{tab:ab,tab:cd} in \tableref{tab:subtabex}.

\begin{table}[htbp]
\floatconts
 {tab:subtabex}
 {\caption{An Example With Sub-Tables}}
 {%
   \subtable{%
     \label{tab:ab}%
     \begin{tabular}{cc}
     \bfseries A & \bfseries B\\
     1 & 2
     \end{tabular}
   }\qquad
   \subtable{%
     \label{tab:cd}%
     \begin{tabular}{cc}
     \bfseries C & \bfseries D\\
     3 & 4\\
     5 & 6
     \end{tabular}
   }
 }
\end{table}

By default, the sub-tables are aligned on the top.
This can be changed using the second optional argument
of \verb|\subtable|. This may be \texttt{t} (top), \texttt{c}
(centered) or \texttt{b} (bottom). For example, the sub-tables
\subtabref{tab:ab2,tab:cd2} in \tableref{tab:subtabex2}
both have \verb|[c]| as the second optional argument.

\begin{table}[htbp]
\floatconts
 {tab:subtabex2}
 {\caption{Another Example With Sub-Tables}}
 {%
   \subtable[][c]{%
     \label{tab:ab2}%
     \begin{tabular}{cc}
     \bfseries A & \bfseries B\\
     1 & 2
     \end{tabular}
   }\qquad
   \subtable[][c]{%
     \label{tab:cd2}%
     \begin{tabular}{cc}
     \bfseries C & \bfseries D\\
     3 & 4\\
     5 & 6
     \end{tabular}
   }
 }
\end{table}

\subsection{Algorithms}
\label{sec:algorithms}

Enumerated textual algorithms can be displayed using the
\texttt{algorithm} environment. Within this environment, use
\verb|\caption| to set the caption and you can use an
\texttt{enumerate} or nested \texttt{enumerate} environments.
For example, see \algorithmref{alg:gauss}. Note that algorithms
float like figures and tables.

\begin{algorithm}[htbp]
\floatconts
  {alg:gauss}%
  {\caption{The Gauss-Seidel Algorithm}}%
{%
\begin{enumerate}
  \item For $k=1$ to maximum number of iterations
    \begin{enumerate}
      \item For $i=1$ to $n$
        \begin{enumerate}
        \item $x_i^{(k)} = 
          \frac{b_i - \sum_{j=1}^{i-1}a_{ij}x_j^{(k)}
          - \sum_{j=i+1}^{n}a_{ij}x_j^{(k-1)}}{a_{ii}}$
        \item If $\|\vec{x}^{(k)}-\vec{x}^{(k-1)} < \epsilon\|$,
          where $\epsilon$ is a specified stopping criteria, stop.
      \end{enumerate}
    \end{enumerate}
\end{enumerate}
}%
\end{algorithm}

If you'd rather have the same numbering throughout the algorithm
but still want the convenient indentation of nested 
\texttt{enumerate} environments, you can use the
\texttt{enumerate*} environment provided by the \textsf{jmlr}
class. For example, see \algorithmref{alg:moore}.

\begin{algorithm}
\floatconts
  {alg:moore}%
  {\caption{Moore's Shortest Path}}%
{%
Given a connected graph $G$, where the length of each edge is 1:
\begin{enumerate*}
  \item Set the label of vertex $s$ to 0
  \item Set $i=0$
  \begin{enumerate*}
    \item \label{step:locate}Locate all unlabelled vertices 
          adjacent to a vertex labelled $i$ and label them $i+1$
    \item If vertex $t$ has been labelled,
    \begin{enumerate*}
      \item[] the shortest path can be found by backtracking, and 
      the length is given by the label of $t$.
    \end{enumerate*}
    otherwise
    \begin{enumerate*}
      \item[] increment $i$ and return to step~\ref{step:locate}
    \end{enumerate*}
  \end{enumerate*}
\end{enumerate*}
}%
\end{algorithm}

Pseudo code can be displayed using the \texttt{algorithm2e}
environment. This is defined by the \textsf{algorithm2e} package
(which is automatically loaded) so check the \textsf{algorithm2e}
documentation for further details.\footnote{Either \texttt{texdoc
algorithm2e} or \url{http://www.ctan.org/pkg/algorithm2e}}
For an example, see \algorithmref{alg:net}.

\begin{algorithm2e}
\caption{Computing Net Activation}
\label{alg:net}
% older versions of algorithm2e have \dontprintsemicolon instead
% of the following:
%\DontPrintSemicolon
% older versions of algorithm2e have \linesnumbered instead of the
% following:
%\LinesNumbered
\KwIn{$x_1, \ldots, x_n, w_1, \ldots, w_n$}
\KwOut{$y$, the net activation}
$y\leftarrow 0$\;
\For{$i\leftarrow 1$ \KwTo $n$}{
  $y \leftarrow y + w_i*x_i$\;
}
\end{algorithm2e}

\section{Description Lists}

The \textsf{jmlr} class also provides a description-like 
environment called \texttt{altdescription}. This has an
argument that should be the widest label in the list. Compare:
\begin{description}
\item[add] A method that adds two variables.
\item[differentiate] A method that differentiates a function.
\end{description}
with
\begin{altdescription}{differentiate}
\item[add] A method that adds two variables.
\item[differentiate] A method that differentiates a function.
\end{altdescription}

\section{Theorems, Lemmas etc}
\label{sec:theorems}

The following theorem-like environments are predefined by
the \textsf{jmlr} class: \texttt{theorem}, \texttt{example},
\texttt{lemma}, \texttt{proposition}, \texttt{remark}, 
\texttt{corollary}, \texttt{definition}, \texttt{conjecture}
and \texttt{axiom}. You can use the \texttt{proof} environment
to display the proof if need be, as in \theoremref{thm:eigenpow}.

\begin{theorem}[Eigenvalue Powers]\label{thm:eigenpow}
If $\lambda$ is an eigenvalue of $\vec{B}$ with eigenvector
$\vec{\xi}$, then $\lambda^n$ is an eigenvalue of $\vec{B}^n$
with eigenvector $\vec{\xi}$.
\begin{proof}
Let $\lambda$ be an eigenvalue of $\vec{B}$ with eigenvector
$\xi$, then
\begin{align*}
\vec{B}\vec{\xi} &= \lambda\vec{\xi}
\intertext{premultiply by $\vec{B}$:}
\vec{B}\vec{B}\vec{\xi} &= \vec{B}\lambda\vec{\xi}\\
\Rightarrow \vec{B}^2\vec{\xi} &= \lambda\vec{B}\vec{\xi}\\
&= \lambda\lambda\vec{\xi}\qquad
\text{since }\vec{B}\vec{\xi}=\lambda\vec{\xi}\\
&= \lambda^2\vec{\xi}
\end{align*}
Therefore true for $n=2$. Now assume true for $n=k$:
\begin{align*}
\vec{B}^k\vec{\xi} &= \lambda^k\vec{\xi}
\intertext{premultiply by $\vec{B}$:}
\vec{B}\vec{B}^k\vec{\xi} &= \vec{B}\lambda^k\vec{\xi}\\
\Rightarrow \vec{B}^{k+1}\vec{\xi} &= \lambda^k\vec{B}\vec{\xi}\\
&= \lambda^k\lambda\vec{\xi}\qquad
\text{since }\vec{B}\vec{\xi}=\lambda\vec{\xi}\\
&= \lambda^{k+1}\vec{\xi}
\end{align*}
Therefore true for $n=k+1$. Therefore, by induction, true for all
$n$.
\end{proof}
\end{theorem}

\begin{lemma}[A Sample Lemma]\label{lem:sample}
This is a lemma.
\end{lemma}

\begin{remark}[A Sample Remark]\label{rem:sample}
This is a remark.
\end{remark}

\begin{corollary}[A Sample Corollary]\label{cor:sample}
This is a corollary.
\end{corollary}

\begin{definition}[A Sample Definition]\label{def:sample}
This is a definition.
\end{definition}

\begin{conjecture}[A Sample Conjecture]\label{con:sample}
This is a conjecture.
\end{conjecture}

\begin{axiom}[A Sample Axiom]\label{ax:sample}
This is an axiom.
\end{axiom}

\begin{example}[An Example]\label{ex:sample}
This is an example.
\end{example}

\section{Color vs Grayscale}
\label{sec:color}

It's helpful if authors supply grayscale versions of their
images in the event that the article is to be incorporated into
a black and white printed book. With external PDF, PNG or JPG
graphic files, you just need to supply a grayscale version of the
file. For example, if the file is called \texttt{myimage.png},
then the gray version should be \texttt{myimage-gray.png} or
\texttt{myimage-gray.pdf} or \texttt{myimage-gray.jpg}. You don't
need to modify your code. The \textsf{jmlr} class checks for
the existence of the grayscale version if it is print mode 
(provided you have used \verb|\includegraphics| and haven't
specified the file extension).

You can use \verb|\ifprint| to determine which mode you are in.
For example, in \figureref{fig:nodes}, the 
\ifprint{dark gray}{purple} ellipse represents an input and the
\ifprint{light gray}{yellow} ellipse represents an output.
Another example: {\ifprint{\bfseries}{\color{red}}important text!}

You can use the class option \texttt{gray} to see how the
document will appear in gray scale mode. \textcolor{blue}{Colored
text} will automatically be converted to gray scale.

The \textsf{jmlr} class loads the \textsf{xcolor}
package, so you can also define your own colors. For example:
\ifprint
  {\definecolor{myred}{gray}{0.5}}%
  {\definecolor{myred}{rgb}{0.5,0,0}}%
\textcolor{myred}{XYZ}.

The \textsf{xcolor} class is loaded with the \texttt{x11names}
option, so you can use any of the x11 predefined colors (listed
in the \textsf{xcolor} documentation\footnote{either 
\texttt{texdoc xcolor} or \url{http://www.ctan.org/pkg/xcolor}}).

\section{Citations and Bibliography}
\label{sec:cite}

The \textsf{jmlr} class automatically loads \textsf{natbib}.
This sample file has the citations defined in the accompanying
BibTeX file \texttt{jmlr-sample.bib}. For a parenthetical
citation use \verb|\citep|. For example
\citep{guyon-elisseeff-03}. For a textual citation use
\verb|\citet|. For example \citet{guyon2007causalreport}. 
Both commands may take a comma-separated list, for example
\citet{guyon-elisseeff-03,guyon2007causalreport}.

These commands have optional arguments and have a starred
version. See the \textsf{natbib} documentation for further
details.\footnote{Either \texttt{texdoc natbib} or
\url{http://www.ctan.org/pkg/natbib}}

The bibliography is displayed using \verb|\bibliography|.

\acks{Acknowledgements go here.}

\bibliography{jmlr-sample}

\appendix

\section{First Appendix}\label{apd:first}

This is the first appendix.

\section{Second Appendix}\label{apd:second}

This is the second appendix.

\end{document}

### 2012-11-23 klein2012structured_EWRL/klein2012structured_poster.tex
Déplacement du dossier klein2012 en klein2012structured_EWRL

\documentclass[xcolor=x11names,12pt]{beamer}
%\usetheme[secheader]{Boadilla} 
\mode<presentation>{\usetheme{I6pd2}}

\usepackage{multicol} 
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{grffile}
\usepackage{listings}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{pifont}
\usepackage{slashbox}
\usepackage{marvosym}
\usepackage[ruled]{algorithm2e}
\include{headertikz}
\usetikzlibrary{through,shapes,arrows,decorations.pathmorphing,backgrounds,positioning,fit}

\newcommand{\argmin}{\mathop{\mathrm{min}}}
\usepackage[orientation=portrait,size=a0,scale=1.4]{beamerposter}

\newcommand{\conf}[1]{\newcommand{\insertconf}{#1}}
\newenvironment{WholeWidthBox}[1]{
  \begin{columns}
    \begin{column}{0.98\textwidth}
      \begin{block}{#1}
        \begin{hfill}
}{
        \end{hfill}
      \end{block}
    \end{column}
  \end{columns}
}

\newcommand{\TwoBoxes}[7]{%Width1 Title1 Content1 Width2 Title2 Content2 MinHeight
  \begin{columns}
    \begin{column}{#1\textwidth}
      \begin{block}{#2}
        \begin{columns}
          \begin{column}{0.001\textwidth}
            \vspace{#7}
          \end{column}
          \begin{column}{\textwidth}
            \centering
            #3
          \end{column}
        \end{columns}       
      \end{block}
    \end{column}

    \begin{column}{#4\textwidth}
      \begin{block}{#5}
        \begin{columns}
          \begin{column}{0.001\textwidth}
            \vspace{#7}
          \end{column}
          \begin{column}{\textwidth}
            \centering
            #6
          \end{column}
        \end{columns}       
      \end{block}
    \end{column}
  \end{columns}
}

%Start of document

\tikzstyle{state}=[circle,
thick,
minimum size=1.0cm,
draw=blue!80,
fill=blue!20]
\tikzstyle{action}=[rectangle,thick,
minimum size=1.0cm,
draw=orange!80,
fill=orange!20]
\tikzstyle{element}=[rectangle,
line width=1.5mm,
minimum size=1.0cm,
draw=blue!80,
fill=blue!20]
\tikzstyle{action}=[rectangle,
line width=1.5mm,
minimum size=1.0cm,
draw=orange!80,
fill=orange!20]

\conf{Journée d'automne d'IAEM. APIL2012}
\title{Structured Classification for Inverse Reinforcement Learning}
\author{\underline{Edouard Klein}}
\date{\today}
\institute[Supélec]{Equipe IMS/MaLIS (Supélec), France\\Equipe ABC UMR 7503 (Loria-CNRS), France}
\newlength{\columnheight}
\setlength{\columnheight}{105cm}

\begin{document}
\begin{frame}

%%%%%%%%%%%%% DEBUT PREMIERE LIGNE %%%%%%%%%%%%%%%%%
\begin{WholeWidthBox}{1. Introduction}
  The inverse reinforcement learning (IRL) problem can be defined as inferring a reward for which a demonstrated expert behavior is optimal. We introduce a new algorithm, SCIRL, whose principle is to use the so-called feature expectation of the expert as the parameterization of the score function of a multi-class classifier. This approach produces a reward function for which the expert policy is provably near-optimal. Contrary to most of existing IRL algorithms, SCIRL does not require solving a single time the direct RL problem. Moreover, up to the use of some heuristic, it may work with only trajectories sampled according to the expert behavior. This is illustrated on toy problems.
\end{WholeWidthBox}

\vfill
%%%%%%%%%%%% DEBUT DEUXIEME LIGNE %%%%%%%%%%%%%%%%%%

\TwoBoxes{.48}{2. Setting}{
  \begin{columns}
    \begin{column}{.48\textwidth}
      \includegraphics{Agent003.png}
    \end{column}
    \begin{column}{.48\textwidth}
      \begin{itemize}
      \item Expert's trace
      \item $MDP\backslash R $
      \item $\pi^*(s) \in \arg\max_{a\in A}Q^*(s,a)$
      \item $R(s) = \theta^T\psi(s)$
      \end{itemize}

    \end{column}
  \end{columns}
}
{.48}{3. Assumptions and Goal}{
  \begin{itemize}
  \item Assumptions
    \begin{itemize}
    \item The expert is a RL agent
    \item One can access the expert's trace
    \item The agent has the same abilities as the expert
    \end{itemize}
  \item Goal : Infer the expert's reward
    \begin{itemize}
    \item apprenticeship of the expert's task
    \item Generalization of the policy over never-seen-before states
    \item Useful when rewards are hard to tune ({\it e.g.}, driving)
    \end{itemize}
  \end{itemize}
}
{15cm}
\vfill
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TROISIEME, quatrième, cinquième et sixième LIGNES
\begin{columns}
  \begin{column}{.48\textwidth}
    \begin{block}{4. Expert}
      \begin{columns}
        \begin{column}{0.001\textwidth}
          \vspace{8cm}
        \end{column}
        \begin{column}{\textwidth}
          \begin{itemize}
          \item The expert is optimal : $\pi_E(s) = \arg\max\limits_{a} Q^{\pi_E}(s,a)$
          \item w.r.t. a linearly parametrized reward : $Q^\pi(s_t,a) = \theta^T\mu^\pi(s_t,a)$
          \item thus : \tikz{\node[fill=red!30,draw=red,ultra thick,rounded corners]{$\pi_E(s) \in \arg\max\limits_{a} \theta^T\mu^{\pi_E}(s,a)$};}
          \end{itemize}
        \end{column}
      \end{columns}
    \end{block}
    \vspace{1cm}
    \begin{block}{6. Linearly parametrized score function based classifiers}
      \begin{columns}
        \begin{column}{0.001\textwidth}
          \vspace{8cm}
        \end{column}
        \begin{column}{\textwidth}
          \begin{itemize}
          \item A score function : $g_s(x) \in \arg\max\limits_{y\in\mathcal{Y}}s(x,y)$
          \item which is linearly parametrized : $s(x,y) = \theta^T \phi(x,y)$
          \item gives : \tikz{\node[fill=red!30,draw=red,ultra thick,rounded corners]{$g_s(x) \in \arg\max\limits_{y\in\mathcal{Y}}\theta^T\phi(x,y)$};}
          \end{itemize}
        \end{column}
      \end{columns}             
    \end{block}
    \vspace{1cm}
    %% \begin{block}{Putting it all together}
    %%   \begin{equation*}
    %%     \mathcal{X} \equiv S, \mathcal{Y} \equiv A, s\equiv Q^{\pi_E} \Rightarrow \phi \equiv \mu^{\pi_E}
    %%   \end{equation*}
    %% \end{block}
    \vfill
    \begin{block}{7. Pseudo-code}
      \resizebox{.95\columnwidth}{!}{
        \begin{algorithm}[H]%[tbh]
          %\small
          %\SetVline
          \caption{SCIRL algorithm}
          \label{algo:scirl}
          %
          \BlankLine
          \emph{\textbf{Given}} a training set $\mathcal{D} = \{(s_i,a_i=\pi_E(s_i))_{1\leq i\leq N}\}$,
          an estimate $\hat{\mu}^{\pi_E}$ of the expert feature expectation $\mu^{\pi_E}$ and a classification algorithm\;
          %
          \BlankLine
          \emph{\textbf{Compute}} the parameter vector $\theta_c$ using the
          classification algorithm
          fed with the training set $\mathcal{D}$ and considering the parameterized score function
          $\theta^T\hat{{ \mu}}^{\pi_E}(s,a)$\;
          %
          \BlankLine
          \emph{\textbf{Output}} the reward function $R_{\theta_c}(s) = \theta_c^T\psi(s)$ \;
        \end{algorithm}
      }
    \end{block}
  \end{column}
  \begin{column}{.48\textwidth}
    \begin{block}{5. Feature Expectation}
        \begin{equation*}
          Q^\pi(s_t,a) = E\left[\left.\sum\limits_{i}\gamma^i r_{t+i}\right|\pi,s_t,a\right] = \theta^T\underbrace{E\left[\left.\sum\limits_{i}\gamma^i \psi(s_{t+i})\right|\pi,s_t,a\right]}_{\color{red}\mu^\pi(s_t,a)}
        \end{equation*}
    \end{block}
    \vfill
    \begin{block}{8. Error bound}
      With concentration coefficient $C_f$, classification error $\epsilon_C$ and $\bar{\epsilon}_Q$ a measure of the quality of our estimate of $\mu^{\pi_E}$ :
      \begin{equation*}
        0\leq
        \mathop{E}_{s\sim\rho_E}[V^*_{R_{\theta_c}}-V^{\pi_E}_{R_{\theta_c}}]
        \leq \frac{C_f}{1-\gamma}\left(\bar{\epsilon}_Q +
        \epsilon_c\frac{2\gamma\|R_{\theta_c}\|_\infty}{1-\gamma}
        \right)
      \end{equation*}
    \end{block}
    \vfill
    \begin{block}{9. Highway}
      \begin{columns}
        \begin{column}{0.001\textwidth}
          \vspace{7cm}
        \end{column}
        \begin{column}{\textwidth}
          \centering
          \fontsize{11pt}{11pt}\selectfont
          \resizebox{.95\columnwidth}{!}{\includegraphics[width=0.47\textwidth]{Cascading_Exp5_fig2}
            \includegraphics[width=0.47\textwidth]{SCIRL_Exp3_fig1}}
        \end{column}
      \end{columns}             
    \end{block}
  \end{column}
\end{columns}
\vfill
%%%%%%%%%%%%% Septième ligne
  \begin{columns}
    \begin{column}{.48\textwidth}
      \begin{block}{10. GridWorld}
        \centering
%        \fontsize{11pt}{11pt}\selectfont
%        \resizebox{.95\columnwidth}{!}{
%         Bli\\
\vspace{-70pt}
          \includegraphics[width=.52\textwidth]{true_reward.pdf}\hspace{-100pt}\includegraphics[width=.52\textwidth]{SCIRL_Exp1_R.pdf}\\
%          Blah
\vspace{-110pt}
          \includegraphics[width=.52\textwidth]{V_expert.pdf}\hspace{-100pt}\includegraphics[width=.52\textwidth]{SCIRL_Exp1_V.pdf}
\vspace{-90pt}
%}
      \end{block}
    \end{column}
    \begin{column}{.48\textwidth}
      \begin{block}{11. Inverted Pendulum}
        \centering
        \fontsize{11pt}{11pt}\selectfont
        \resizebox{.95\columnwidth}{!}{

}
\vspace{-70pt}
          \includegraphics[width=.52\textwidth]{LAFEM_Exp3_true_R}\hspace{-100pt}\includegraphics[width=.52\textwidth]{LAFEM_Exp3_lafem_R}\\
%          Blah
\vspace{-110pt}
          \includegraphics[width=.52\textwidth]{LAFEM_Exp3_Vexpert}\hspace{-100pt}\includegraphics[width=.52\textwidth]{LAFEM_Exp3_Vagent}
\vspace{-90pt}
%}
      \end{block}
    \end{column}
  \end{columns}

\end{frame}
\end{document}

### 2012-11-23 klein2012structured_NIPS/klein2012structured_poster.tex
Récupérations d'anciennes valeurs (EWRL) pour l poster NIPS

--- 2012-11-23 klein2012structured_EWRL/klein2012structured_poster.tex
@@
 
-\conf{Journée d'automne d'IAEM. APIL2012}
+\conf{European workshop on reinforcement learning}
 \title{Structured Classification for Inverse Reinforcement Learning}
-\author{\underline{Edouard Klein}}
+\author{\underline{Edouard Klein}$^{\dag\ddag}$, Bilal Piot$^{\dag\textrm{\Ankh}}$, Matthieu Geist$^\dag$ and Olivier Pietquin$^{\dag\textrm{\Ankh}}$\\\texttt{firstname.lastname@supelec.fr}}
 \date{\today}
-\institute[Supélec]{Equipe IMS/MaLIS (Supélec), France\\Equipe ABC UMR 7503 (Loria-CNRS), France}
+\institute[Supélec]{$\dag$Equipe IMS/MaLIS (Supélec), France\\$\ddag$Equipe ABC UMR 7503 (Loria-CNRS), France\\\Ankh UMI 2958 (GeorgiaTech-CNRS)}
 \newlength{\columnheight}

### 2012-11-27 klein2012structured_NIPS/klein2012structured_poster.tex
Poster tel qu'envoyé à l'impression
Dernières modifs sr le poster (titre et ordre auteurs et typo dans la borne)

@@
 
-\conf{European workshop on reinforcement learning}
-\title{Structured Classification for Inverse Reinforcement Learning}
-\author{\underline{Edouard Klein}$^{\dag\ddag}$, Bilal Piot$^{\dag\textrm{\Ankh}}$, Matthieu Geist$^\dag$ and Olivier Pietquin$^{\dag\textrm{\Ankh}}$\\\texttt{firstname.lastname@supelec.fr}}
+\conf{NIPS 2012}
+\title{Inverse Reinforcement Learning through Structured Classification}
+\author{\underline{Edouard Klein}$^{\dag\ddag}$, Matthieu Geist$^\dag$, Bilal Piot$^{\dag\textrm{\Ankh}}$ and Olivier Pietquin$^{\dag\textrm{\Ankh}}$\\\texttt{firstname.lastname@supelec.fr}}
 \date{\today}
@@
         0\leq
-        \mathop{E}_{s\sim\rho_E}[V^*_{R_{\theta_c}}-V^{\pi_E}_{R_{\theta_c}}]
+        \mathop{E}_{s\sim\rho_E}[V^*_{R_{\theta_c}}(s)-V^{\pi_E}_{R_{\theta_c}}(s)]
         \leq \frac{C_f}{1-\gamma}\left(\bar{\epsilon}_Q +

### 2013-01-11 klein2012structured_RIA/revisions.tex
Prise en compte des remarques du reviewer 1 pour RIA

\documentclass[11pt, a4paper]{letter}
\usepackage[plain,cm]{fullpage}
\usepackage[utf8] {inputenc}
\usepackage[frenchb]  {babel}
\usepackage  {color}

% \usepackage{lmodern}
% \renewcommand*\familydefault{\ttdefault} %% Only if the base font of the document is to be typewriter style
\usepackage[T1]{fontenc}

% \usepackage[default]{frcursive}

% \usepackage{inconsolata}
% % \renewcommand{\familydefault}{\ttdefault}
% % \renewcommand{\ttdefault}{pcr}
% \usepackage[T1]{fontenc}

\usepackage{anysize}
\marginsize{2.5cm}{2.5cm}{2cm}{2cm}

\definecolor{blue}{rgb}{0.2,0.3,1}

\begin{document}

% \pagestyle{plain}

%\tt
\color{blue}

\address{Edouard Klein,  Matthieu Geist, \\Bilal Piot, Olivier Pietquin \\ IMS – MaLIS Research Group, Sup\'elec \\ 2 rue Edouard Belin, 57070 Metz (France)}
%\signature{Lucie Daubigney, \\ Matthieu Geist, \\ Senthilkumar Chandramohan, \\ Olivier Pietquin.}
\date{Metz, \today. \vspace{1.5cm}}

\begin{letter}{\large \textbf{Révisions sur le manuscrit ``Classification structurée pour l'apprentissage par renforcement inverse''}}
\bigskip
\opening{Chers éditeurs, Chers reviewers}

\bigskip

Nous vous remercions pour le temps passé à relire notre manuscrit. Les remarques des reviewers étant largement orthogonales, nous les prenons en compte une par une. Les changements apportés au manuscrit sont reportés ici en bleu.

\newpage
\color{black}
\begin{large} \textbf{Reviewer 1:} \\ \end{large}
\textcolor{blue}{
Toutes les corrections suggérées ont été appliquées.
}

Some comments:
\begin{enumerate}
\item The abstract should make it more clear that your aim here is to show KTD works on both on- and off-policy settings, and with nonlinear parameterisations.

\textcolor{blue}{ The abstract has been corrected to list the qualities a good DM algorithm should exhibit.}

\item The title doesn't read very well. I would consider replacing "sole" with "single". Possibly also make it clear that you are addressing both on- and off-policy training with a single algorithm.

\textcolor{blue}{ The title has been changed to : ``An All-in-One
Reinforcement Learning Algorithm for Dialogue Management
Optimisation''.}

\item Abstract: ``Now, for each different issue was proposed a different algorithm  …". Rephrase. Maybe start "Different algorithm have been proposed to handle each different issue, making… ''.

\textcolor{blue}{We rephrased this sentence in the abstract: ``.
However, each algorithm often solves a single issue at a time while
dialogue systems exhibit all the problems at once''}

\item The paper would be much stronger with human user experiments. Simulations are known to be a poor reflection of reality. \textcolor{blue}{ Humans experiments have been lead with both DIPPER and HIS systems. The references to the related works have been added.}
\end{enumerate}

%\vspace{.5cm}
\newpage
\begin{large} \textbf{Reviewer 2:} \\ \end{large}
 Recommendation: Major Revision\\
 Comments:\\
 Summary: This paper presents an application of the recently proposed Kalman
 Temporal Differences (KTD) framework to optimize spoken dialog systems
 (SDS).  The KTD framework is claimed to enjoy benefits of (i) enabling
 off-policy learning, (ii) improving sample efficiency, (iii) facilitating
 guided exploration, (iv) adapting to nonstationary environments, and (v)
 handling nonlinear approximation architectures.  Motivated by Kalman
 filters, KTD is essentially a graphical, generative model for TD errors in
 which one is interested in inferring the (unobservable) value function.

   Two SDS are considered in experiments --- DIPPER and HIS, on which KTD is
 compared to other algoirhtms including fitted Q-learning, LSPI, and
 Q-learning.

 The KTD framework appears to be a reasonable fit to optimal control in SDS,
 given the design objectives of the framework.  Since KTD is an existing
 technique, the main contribution and novelty is therefore in its
 application to SDS.  The paper does a good job at explaining the
 motivation, and the level of technical details is appropriate for non-RL
 expert to understand most of the main idea.  Overall, the paper is quite
 easy to follow, except for some typos (see below).

 Since KTD is not new, it becomes important to have an extensive comparison
 to alternative approaches to demonstrate its pros and cons.  While a few
 natural candidates have been included, some others are ignored, and some
 may not be implemented in the best way.  Details:

\begin{enumerate}
\item  Many Bayesian RL algorithms share all (or most) claimed benefits
 of KTD.  (In fact, I think KTD may be viewed as an instance of Bayesian RL
 algorithm.)  So one wonders naturally how KTD compares to other Bayesian
 methods, say, those by Yaakov Engel (a series of papers at ICML/NIPS) and
 by Jan Peters (at ACC-08).  I am particularly wondering whether the real
 benefits here come uniquely from KTD, or from Bayesian RL in general.

\textcolor{blue}{  Precisions about how Bayesian approches differ
from the KTD approaches are given at the beginning of the Sec. III :
``It is important to notice that Kalman filtering differs from the
Bayesian filtering. Indeed, the Bayesian filtering aims at computing
the complete posterior distribution of the hidden variable while
Kalman filtering only focuses on the first and second moments of
this distribution with constrained linear update. They are equal
only when Gaussian distributions and linear parameterisations are
concerned.''. A comparison with a Bayesian algorithm (namely GPTD
proposed by Engel \textit{et al.}) has been added in the paper
Sec.~V-C (on Fig.~3 and Fig.~6).}

\item  LSPI: first, it is possible to extend it to online learning, such
 as in the AAMAS-09 paper by Lihong Li et al.

\textcolor{blue}{Thank you for the reference, we added it to our
bibliography.}

Second, ill-conditioning is
 reported in the experiments, but it is common practice to to L2-regularized
 LSTD inside LSPI.  It is not surprising that, with poor conditioning,
 LSTD/LSPI become very unstable and give bad results.  One thus wonders
 whether these algorithms are implemented in the 'right' (or robust) way to
 allow for fair comparisons.

\textcolor{blue}{Precisions about the experimental conditions when
using LSPI have been given. In particular, regularisation has
actually been tested and led to poor strategies. It is now precised}

\item Fitted Q-learning: essentially the same concerns apply as in the
 case of LSPI.

\item  The empirical results do *not* seem to be impressive.  Unless I
 misinterpret the plots, KTD-style algorithms only beat incremental
 algorithms like Q-learning and Sarsa.  Compared to other batch algorithms,
 KTD shows no significant advantage.

\textcolor{blue}{The performance of KTD is not effectively better
than those of the other batch algorithms but our goal in this
article is to propose a unique algorithm to solve several learning
issues with comparable performance. Moreover KTD processes samples
only once while other methods process them many times.}

\item In general, there is no free lunch when we come to algorithm
 selections.  KTD achieves some of the benefits at the cost of other
 disadvantages.  The paper will be more valuable to readers if the 'cons'
 are also discussed, such as convergence and optimality guarantees, to name
 a few.

\textcolor{blue}{A proof of convergence of the KTD algorithm under
certain hypothesis have been proposed (in the JAIR paper). A
difficulty appears also as far as the choice of the KTD parameters
are concerned. A paragraph about how to choose them has been added
Sec. III: `` The $\hat{\theta}_{0|0}$ vector can be initialised to a
default value or, if known, a value near the optimal one. The
$P_{0|0}$ matrix translates the certainty the user has in the prior
$\hat{\theta}_{0|0}$. It appears in Eq. (2) as a regularisation
term. A common initialisation for this matrix is $\lambda
\mathrm{Id}$. Only one parameter ($\lambda$) has thus to be chosen
instead of $p^2$ parameters ($p$ being the number of parameters to
learn).  The lower $\lambda$ the less certain is. The observation
noise $P_{n_t}$ represents the confidence about the ability of the
chosen parameterisation to represent the true Q-function.
Additionally, the evolution noise is chosen so that $P_{v_t} = \eta
P_{\theta_{t-1|t-1}}$ with $0 < \eta \ll 1$. The effect of this
heuristic is to put more emphasis on more recent data (it acts as a
forgetting factor).''. It has to be noticed that only an order of
magnitude has to be chosen. The KTD algorithm is not sensitive to
small changes in the parameters.}

\item  Furthermore, while KTD represents a reasonable algorithm, some of the
 algorithmic choices seem arbitrary and should be justified.  For example,
 the noise terms in eqn 6 are unspecified.  It may be tempting (as probably
 is done in experiments) to assume the noises are normal, but this modeling
 assumption can be quite wrong in many RL problems.  Also, more algorithmic
 details about KTD would make the paper more self-contained.  The Kalman
 gain, for instance, is quite mysterious --- how about giving a formula for
 computing that matrix after equation 7?

 \textcolor{blue}{ The KTD algorithm is now explained Sec. III: `` The KTD algorithm is done in four steps: an \textit{initialisation} step, a \textit{prediction} step, a computational phase of \textit{statistics of interest} and finally a \textit{correction} step. [$\ldots$]''. The detail of each step is then given. As precised previously, a paragraph has also been added to explain the role of the different parameters. }

\item  As said, the performance of KTD compared to fitted Q-learning or
 LSPI is not impressive.  In fact, KTD seems to converge to slightly worse
 performance.  The incremental vs. batch explanation does not make sense,
 since KTD is pretty much a batch algorithm as well --- see the authors' own
 comments on the sample efficiency of KTD below eqn 6.

\textcolor{blue}{The KTD algorithm is not a batch algorithm. The
update of the gain is done each time a sample is processed. The aim
of the article was only to propose a single algorithm to solve
different problems with performance that compares to specialised
algorithms. }

\item  What is LSPI not in any of the plots?  What is more, given the
 competitiveness of fitted Q-learning (and LSPI), I find it surprising to
 see the comparisons to them in only one experiments, but not in all (like
 in HIS).

 \textcolor{blue}{LSPI has been added to the set of curves for both systems in Fig.~1 and Fig.~2.}

\item  One of the claimed benefits of KTD is the ability to handle
 nonlinear function approximators.  In the experiments, in particular figs 5
 and 6, it is a bit sad to see nonlinear alternatives provide no benefits
 (but actually cause slower learning speed).  One then asks, at least in the
 context of SDS in this paper, why should one care about nonlinear KTD.

\textcolor{blue}{ Curves about the non-linear parameterisation have
been added to better show what happens during the learning with this
parameterisation (Fig.~4 and Fig.~8). The following explanation has
been given: ``The different runs used to compute the mean are added
on the graph. They show the detailed behaviour of the ANN. Contrary
to the linear case, the learning is either very good or very poor.
The curve representing the average results presents thus poor
results with small sets of data.''. A method should then be found to
automatically remove the worst runs like said in the conclusion.}

\item  One final comment is about the authors' prior work published at IJCAI 2011.
 It seems some results reported here have appeared in that conference
 paper.  It would avoid future confusions if the connection is made clearer.

\textcolor{blue}{The results presented in the IJCAI paper present
only an off-policy approach. This has been clarified in the
article.}

\end{enumerate}

\vspace{.5cm}
\newpage
\begin{large} \textbf{Reviewer 3:} \\ \end{large}
 Recommendation: Major Revision\\
 Comments: This paper presents how the Kalman Temporal Differences framework
 (published at JAIR by the same authors) can be applied to dialogue
 management. To show that the authors apply it to two dialogue systems one
 developed using DIPPER and one developed using the HIS dialogue management
 framework. The idea of applying the KTD framework to dialogue management is
 certainly novel and promising, however, the paper is poorly written and the
 evaluation needs to be seriously extended. The argument that because
 the algorithm performs well in simulation it should also perform well with
 real users is very weak especially for a journal submission. If things were
 so simple nobody would ever bother perform experiments with real users.

\textcolor{blue}{Our aim was not to assess that leading experiments
with simulated users is enough to prove the efficiency of the
algorithms tested. The detailed explanation of our motivation are
given Sec.~IV-A1. We precise that experiments with two differents
systems, with different state spaces, different reward functions,
different user simulation methods \textit{etc} give a good hint
about the efficiency of them and the transferability to real world
applications. In addition, DIPPER and HIS trained with the methods
we used as baselines have been tested with real users (the
references have been precised previously and added in the paper).}

 As I mentioned above the writing is not good. A lot of important details
 are missing, even simple things, like the definition of the reward
 function.

\textcolor{blue}{Indeed, we forgot to give some important details
about how the two dialogue management frameworks are cast into an
MDP. We added an entire paragraph to describe each of them (Sec.~4-A
2 \& 3). }

It is not clear to me why you compare with other existing
 algorithms only using the DIPPER-based system and not the HIS-based system.
 Is it because the HIS-based system is more complex?

\textcolor{blue}{We added others algorithms to the HIS-based system
comparison. Now, the KTD algorithm performance is compared to those
of \textit{Gaussian Process Temporal Differences} (GPTD) and
\textit{Least Square Policy Iteration} (LSPI). The GPTD algorithm is
one of the more recent and efficient algorithm for dialogue
management (references given in the article). The results are given
Fig.~2, Fig.~3 and Fig.~6.}

But this is exactly the case where one would want to use a superior algorithm. Is it because
 for example LSPI would have problems in the HIS-based system because of the
 large table that would have to be inverted? Please explain.

\textcolor{blue}{We added a description of the state space for the
HIS system in Sec.~IV-A3 and a description of the Q-function
parameterisation in Sec.~IV-C2. The number of parameters is not so
large thus we do not have matrix computation problems.}

Furthermore, I don't think anyone can manage to reproduce this experiment given that
 important details are missing. For example, the section about the
 Q-function representation is very confusing. It is unclear to me what the
 structures of the Radial Basis Function networks (one per action) are.

\textcolor{blue}{Indeed, an important part of the experimental
settings was missing. We tried to make it clearer by giving an
accurate description of them. The experimental settings are given
Sec.~V-A. The value of the parameters used for the different
algorithms have been precised for both systems so as the number of
runs to plot the different runs. We also give more details about the
Q-function representation in Sec.~IV-C. We precisely provide the
mathematical descriptions of the parameterisations.}

 Anyway, what I'm saying here, is that the authors should not expect that
 readers should have to go back and read the JAIR article because the
 missing point here is not the algorithm itself but how it is applied to
 dialogue management. Thus even if someone had understood the algorithm it's
 not clear that he would manage to apply it to dialogue management and
 reproduce the experiments presented in this paper.

\textcolor{blue}{A more accurate description of the KTD algorithm
has been given Sec.~III. The different steps of the algorithm have
been presented. A description of how the dialogue management problem
is cast into the MDP framework is in Sec.~IV-A for both systems.}

 Here are my detailed comments:

\begin{enumerate}
\item  Title: The part of the title "with a sole algorithm" sounds very weird.
 What the authors want to say is that they have developed an algorithm that
 can solve a lot of problems that previously could be solved using a variety
 of algorithms. I'm not sure what the title should be but it should
 definitely be more informative.

\textcolor{blue}{We tried to make the title clearer about what the
article is about.}

\item Introduction: Here the authors talk about different approaches to learning,
 e.g. dynamic programming, off-policy learning, on-policy learning, etc. I
 suggest they also mention the work of Henderson et al. "Hybrid
 Reinforcement/Supervised Learning of Dialogue Policies from Fixed
 Datasets", Computational Libguistics 2008. Here the authors use a hybrid
 model of SARSA-lambda and supervised learning to deal with the fact that
 there wasn't much data available for their complex state space.

\textcolor{blue}{ Thank you for the reference. We added it to the
introduction and the bibliography.}

\item Page 13, line 48: DIPPER is not a dialogue system. It is a tool that allows
 someone to write dialogue management rules. In the same way HIS is a
 dialogue management framework. So I suggest the authors say something like
 DIPPER-based and HIS-based dialogue systems instead of DIPPER and HIS
 dialogue systems. In any case, please explain the difference in the paper.

\textcolor{blue}{We modified the text accordingly.}

\item  Background: In this section, I've noticed problems with notation that can
 be confusing especially for the non-expert reader. You refer to the
 Sutton-Barto book so one would think that you would use the notation of
 Sutton-Barto. So I suggest you use r\_(t+1) instead of r\_t when you refer to
 the reward that occurs after the agent performs action a and transitions
 from state s\_t to state s\_(t+1). I'm not saying that using r\_t is wrong,
 only that it may confuse most readers that are familiar with the
 Sutton-Barto notation. Also, you use notation R(s, a, s') but you don't
 explain that it is the same as r\_(t+1). Please try to be consistent. For
 example, in equations (2) and (4) (Bellman equations) you use the notation
 R(s, a, s') but then later in section 3 you use again the Bellman equations
 with r\_t (which in my opinion should become r\_(t+1)) instead of R(s, a,
 s'). Also, sometimes you use s' and sometimes s\_(t+1). Anyway, please try
 to be consistent.

\textcolor{blue}{ The notations have been corrected to be
consistent. All the different notations introduced are now defined.}

\item Equation 4 is wrong. Near the end it should be "... gamma*max\_a'Q*(s', a')"
 (see section 3.8 in the Sutton-Barto book).

\textcolor{blue}{Thank you for mentionning this typo. We corrected
it.}

\item  In section III you need to do a better job of explaining how the KTD
 framework is applied to the dialogue management problem. For example the
 text says "the term K\_t is the Kalman gain, which depends..." Please
 explain how you calculate this. I know there is some explanation in the
 JAIR paper but you need to provide some hints in the current paper too.

\textcolor{blue}{ Some important details have been added to better
understand how the KTD algorithm works in Sec.~III. Now, the
different steps of the algorithm are precised. The definition of the
Kalman gain has been added. This gain is the one which minimises the
mean square error $J_t(\theta)$. The different parameters for the
algorithm are also given.}

\item  Then in page 17, line 38 the paper provides a very brief explanation (2
 lines) why the algorithm is sample efficient. Again, the same issue as
 before. Please provide some more information. This is one of your stronger
 points that this algorithm is sample efficient, so please explain in simple
 words why this is the case. The bottom line is that the description of the
 algorithm should be as self-sufficient as possible, minimizing the need to
 read the whole JAIR paper.

\textcolor{blue}{The classical Q-learning and SARSA algorithms are
gradient descents (first order) while KTD minimises a second error
term (second order algorithm). The convergence in the last case is
thus faster. We added this explanation at the end of the Sec.~III.}

\item In section IV, the paper says that there are no experiments with
 non-stationarity handling and yet the authors claim that this is one of the
 most important features of the algorithm. So I suggest you perform
 experiments both in simulation (with a simulated user that can change its
 goal even if it is a hand-crafted one) and with real users that change
 their goals and see how the algorithm performs. Just saying that this was
 shown in the JAIR paper is not enough because that wasn't for dialogue
 management.

\textcolor{blue}{Precisions about handling non-stationarity are
added in Sec.~III and are experimentally presented Fig.~3. When a
controlled and on-policy approach is used, handling the
non-stationarity is a good feature. Indeed, the policy used for
control varies which introduce non-stationaties.}

\item In section IV, please provide a better description of the two dialogue
 systems. For example, how many possible states do we have in each one of
 them? What are the slots and the actions used?

\textcolor{blue}{ A paragraph has been added to describe precisely
each system and how the dialogue problem is cast into the MDP
framework (state space, action space, reward function, discount
factor) in Sec.~IV-A 2 \& 3.}

\item  As I mentioned above, section IV-B is incomprehensible. You really need to
 show examples with the structure of the networks, perhaps some figures
 would help. In the same action I don't understand how you came up with the
 number 144. Also what do you mean by dimensions here, features of the
 state? Which are the 2 continuous dimensions and which are the 2 discrete
 ones?

\textcolor{blue}{We also provide now a precise description of the
two systems in Sec.~IV-A. We also added precise descriptions of the
features used to represent the Q-function in Sec.~IV-C.}

\item In section V-A the text says ``random actions are chosen among reasonable
 actions''. Could you please explain what you mean by ``reasonable actions''?

\textcolor{blue}{We precised in Sec.~V-B how the handcrafted policy
works. For the DIPPER system, a set of hand crafted rules prevents
the system from choosing actions inconsistent with the current
dialogue state. For example, it avoids to confirm a slot while it
has never has been asked before.}

\item In section V-A the text talks about comparing with LSPI but I can't see
 that in Figure 1. The LSPI graph is missing.

\textcolor{blue}{ We previously made a mistake while compiling the
graphs. We added LSPI for both DIPPER and HIS systems.}

What is the reward function
 used in the experiments of Figure 1?

\textcolor{blue}{A complete description of the rewards functions are
given Sec.~IV-A. For the DIPPER system, each dialogue turn is
penalised with a reward $-1$. At the end of the dialogue, the reward
is provided the following: $-75$ for incorrect slot filling,  $+25$
per slot filled correclty, and $-300$ if the slot is empty. For the
HIS system, the reward function penalises each dialogue turn with
$-1$. At the end of the dialogue, a reward of $+20$ is given if the
DM managed to fulfil the user request. In both cases, the dialogue
is stopped if it has more than $100$ turns. The definition of the
reward function is not an easy task. Other solutions can be
considered.}

 How come there is no variance in the
 hand-crafted graph? One would expect that the variety in the performance of
 the n-gram based simulated user would cause differences in the resulting
 dialogues.

\textcolor{blue}{The variance has not been plotted on Fig.~1 because
the error bars were too large. It has been precised in the comments
of the graphs.}

Of course that depends on what the reward function is which is
 important information not provided in the paper.

\item  For the HIS-based system, why don't you compare with LSPI, etc.? Is it
 because LSPI would be in trouble because of the larger state space? If you
 want to show that this algorithm is better than existing algorithms you'd
 better show that in a more complex system than the DIPPER-based system. You
 could perhaps try the LSPI with fast feature selection to deal with the
 large state space, see the paper by Li et al. "Reinforcement Learning for
 Dialog Management using Least-Squares Policy Iteration and Fast Feature
 Selection" Interspeech 2009.

\textcolor{blue}{Our aim was to show that one algorithm (KTD) can
perform as well as standard other ones (and sometime better) while
being generic. We added the LSPI comparison for both systems (Fig.~1
and Fig.~2). Thank you very much for he reference. We added it to
the bibliography. }

\item In Figure 2 you only show one of the behavioural policies and the KTD one.
 Is this correct? You say that there are two behavioural policies used (I
 believe the graph shows the one with greedy-exploration). Why don't you
 show both behavioural policies? And what is the reward function here?

\textcolor{blue}{We modified the figures hoping the illustration of
our points will be clearer. We also change the caption of the curves
to be more accurate.}

\item Sometimes one algorithm is trained 50 times, another 250 times, etc. Is
 there a reason you are not being consistent?

\textcolor{blue}{We precised in the article at the end of the
Sec.~V-A that there is a difference in the number of trainings
because the learning wich KTD, LSPI and GPTD algorithms takes much
more time than with Q-learning and SARSA because of matrix
inversions.}

 Also, in some figures the
 x-axis has turns and in others dialogues. Again, please try to be
 consistent or at least explain why you are not being consistent.

\textcolor{blue}{The original choice was made different for the two
systems. We precised how to make the link between the two notations
at the beginning of Sec.~V-A. For the HIS system, the length of the
dialogue can be deduced from the reward. The number of dialogue can
thus be onverted in number of samples. For example, if the mean
discounted cumulative reward is $14$ for a set of $50$ training
dialogues, since the maximum reward that can be obtained is $20$ and
that each turn a penalty of $-1$ is given, the average length of the
dialogue is $20 - 14 = 6$ turns. The number of samples is thus
approximately $6 \cdot 50 = 300$.}

\item  In Figure 3, why don't you show the hand-crafted policy?

\textcolor{blue}{Most of the figures have been modified to better
present the result. Each figure corresponds now to a particular
aspect of the KTD algorithm. We particularly paid attention to the
caption of the figures.}

 What is the reward function? Figure 4 has the same problems as Figure 2. For on-policy learning for the graphs you use a sliding windows. Do you use that for the off-policy graphs too?

\textcolor{blue}{A sliding window is only used for the controlled
approaches (Fig.~3,4,6,7,8). It has been precised in the article
when a sliding window is used. }

\item  In section V-C you talk about 22 user actions and you say "the number of
 action for the action space is 12", but earlier when you describe the
 HIS-based system you mention
 that there are 22 system actions. Please explain.
Also, in the same section you perform an experiment on handling non-linearities with the HIS-based
 system. I'd like to see that with the DIPPER-based system too.
Also, in Figure 5 what is the reward function?

\textcolor{blue}{Detailed description of the DIPPER ans HIS systems
has been given Sec.~IV.}

\end{enumerate}

\color{blue} We hope that the modifications improve the quality of
the paper. Whatever the final decision is, we would like to thank
again the reviewers for their careful reading of this paper and for
the precise and detailed comments they suggested.

Best regards,

\begin{flushright}
Lucie Daubigney, Matthieu Geist, \\ Senthilkumar Chandramohan, Olivier Pietquin.
\end{flushright}

%\closing{Best regards,}

\end{letter}
\end{document}

### 2013-01-11 klein2012structured_RIA/scirl_ria.tex
Prise en compte des remarques du reviewer 1 pour RIA

@@
 Il est possible d'imiter le comportement de l'expert via un algorithme d'apprentissage supervisé généralisant l'association des actions aux états.
-Nous considérons ici les classifieurs multi-classes qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour une état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subseackgrounlassif}).
+Nous considérons ici les classifieurs multi-classes qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour un état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subseackgrounlassif}).
 L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir Sec.~\ref{subsecirlgo}).
@@
 
-L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observée par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertseka996,Sutto998,szepesvari2010c}.
+L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertseka996,Sutto998,szepesvari2010c}.
 %
@@
 Notre étude est valable pour tout algorithme de classification tant qu'il opère en maximisant l'argument d'une fonction de score paramétrée linéairement.
-Par exemple, une machine à vecteur support multi-classe pourrait être choisie~\cite{Guermeu007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taska005}.
+Par exemple, un séparateur à vaste marge multi-classe pourrait être choisie~\cite{Guermeu007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taska005}.
 D'autres choix sont possibles.
@@
 \end{theorem}
-\begin{proof}
+\begin{proof*}[Preuve du Théorème \ref{th}]
   La démonstration repose uniquement sur la récompense $\R_{\theta_c}$, donc par soucis de clarté certains indices relatifs à la récompense sont omis des notations  
@@
   \end{align}
-  Il est dacile de voir que $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
+  Il est facile de voir que $\|(P_{\pi_c}-P_{\pi_E})v^{\pi_E}\|_\infty
   \leq \frac{2}{1-\gamma}\|\R\|_\infty$, ce qui permet de borner le dernier terme
@@
   \end{equation}
-  Injecter les bornes des Eqs.~\eqref{eroo1}
+  Injecter les bornes des Eq.~\eqref{eroo1}
   and~\eqref{eroo2} dans l'Eq.~\eqref{erooesidual} achève la démonstration.%, which concludes the proof.
@@
 %  This concludes the proof.
-\end{proof}
+\end{proof*}
 
@@
 Finalement, si notre borne se base sur les erreurs en généralisation 
-$\epsilon_c$ et $\bar{\epsilon}_Q$, le classifieur n'utilisera que 
+$\epsilon_c$ et $\bar{\epsilon}_Q$, le classifieur n'utilisera 
 $(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ que lors de la phase d'entraînement, où les $s_i$ sont les états présents dans $\D$. Il renvoie 
@@
 $\hat{\mu}^{\pi_E}$ correctement uniquement pour les couples état-action $(s_i,a)_{1\leq i\leq
-N,a\in\A}$, ce qui permet d'envisager une simple estimation de Monte-Carlo par exemple.
+N,a\in\A}$, ce qui permet d'envisager, par exemple, une simple estimation de Monte-Carlo.
 
@@
 Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ seules ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen.
-Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$, ces actions (et la dynamique résultantes) n'étant représentées dans les données.
+Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$, ces actions (et la dynamique résultante) n'étant représentées dans les données.
 Il est heureusement possible de recourir à des heuristiques ; ce cas ne rentre pas dans l'analyse présentée en
@@
 \end{equation}
-La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit Eq.~\eqref{ec_plus_heuristic}.
+La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit par Eq.~\eqref{ec_plus_heuristic}.
 
@@
 Certains algorithmes mentionnés ne sont pas capable de renvoyer une fonction de récompense, bien qu'ils utilisent l'ARI comme étape.
-Il rentrent dans le champ généralement appelé l'apprentissage par imitation.
+Ils rentrent dans le champ généralement appelé l'apprentissage par imitation.
 %
@@
 Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI, par exemple~\cite{boularia011}).
-De plus, par l'utilisation d'une heuristique qui dépassent le cadre de notre analyse, il est possible à SCIRL de se contenter de données fournies par l'expert.
+De plus, par l'utilisation d'une heuristique qui dépasse le cadre de notre analyse, il est possible à SCIRL de se contenter de données fournies par l'expert.
 La prochaine section présente une démonstration empirique de cela.
@@
 Nous illustrons SCIRL sur un simulateur de conduite similaire à~\cite{Abbee004,Sye00ame}.
-Le but est de conduire une voiture sur une autoroute à trois voies dont le traffic est généré aléatoirement (les sorties de route sont possible des deux côtés).
+Le but est de conduire une voiture sur une autoroute à trois voies dont le traffic est généré aléatoirement (les sorties de route sont possibles des deux côtés).
 La voiture peut se déplacer vers la gauche ou la droite, accélérer ou ralentir et conserver sa vitesse.
@@
 PIRL est un algorithme itératif dont chaque itération implique la résolution du PDM pour une fonction de récompense arbitraire.
-Nous l'avons fait fonctionner pendant 70 itérations, tous les objets requis (l'attribut moyen de politiques différentes de celles de l'expert et la politique optimale pour une fonction de récompense à chaque itération) ont été calculées de manière exacte en utilisant le modèle.
+Nous l'avons fait fonctionner pendant 70 itérations, tous les objets requis (l'attribut moyen de politiques différentes de celles de l'expert et la politique optimale pour une fonction de récompense à chaque itération) ont été calculés de manière exacte en utilisant le modèle.
 Nous mesurons les performances de chacune des approches grâce à $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, avec $\mathcal{U}$
@@
 
-La Fig.~\ref{fies} montre la performance de chacune des approche en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
+La Fig.~\ref{fies} montre la performance de chacune des approches en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
 On peut voir que le classifieur ne fonctionne pas bien sur ce problème.

### 2013-01-11 klein2012structured_RIA/revisions.tex
Prise en compte des suggestions des reviewers 2 et 3 de RIA.

@@
 \definecolor{blue}{rgb}{0.2,0.3,1}
+
+\newcommand{\R}{\mathcal{R}}
 
@@
 
-Some comments:
-\begin{enumerate}
-\item The abstract should make it more clear that your aim here is to show KTD works on both on- and off-policy settings, and with nonlinear parameterisations.
-
-\textcolor{blue}{ The abstract has been corrected to list the qualities a good DM algorithm should exhibit.}
-
-\item The title doesn't read very well. I would consider replacing "sole" with "single". Possibly also make it clear that you are addressing both on- and off-policy training with a single algorithm.
-
-\textcolor{blue}{ The title has been changed to : ``An All-in-One
-Reinforcement Learning Algorithm for Dialogue Management
-Optimisation''.}
-
-\item Abstract: ``Now, for each different issue was proposed a different algorithm  …". Rephrase. Maybe start "Different algorithm have been proposed to handle each different issue, making… ''.
-
-\textcolor{blue}{We rephrased this sentence in the abstract: ``.
-However, each algorithm often solves a single issue at a time while
-dialogue systems exhibit all the problems at once''}
-
-\item The paper would be much stronger with human user experiments.
-Simulations are known to be a poor reflection of reality.
-\textcolor{blue}{ Humans experiments have been lead with both DIPPER and HIS systems.
-The references to the related works have been added.}
-\end{enumerate}
-
-%\vspace{.5cm}
 \newpage
 \begin{large} \textbf{Reviewer 2:} \\ \end{large}
- Recommendation: Major Revision\\
- Comments:\\
- Summary: This paper presents an application of the recently proposed Kalman
- Temporal Differences (KTD) framework to optimize spoken dialog systems
- (SDS).  The KTD framework is claimed to enjoy benefits of (i) enabling
- off-policy learning, (ii) improving sample efficiency, (iii) facilitating
- guided exploration, (iv) adapting to nonstationary environments, and (v)
- handling nonlinear approximation architectures.  Motivated by Kalman
- filters, KTD is essentially a graphical, generative model for TD errors in
- which one is interested in inferring the (unobservable) value function.
-
-   Two SDS are considered in experiments --- DIPPER and HIS, on which KTD is
- compared to other algoirhtms including fitted Q-learning, LSPI, and
- Q-learning.
-
- The KTD framework appears to be a reasonable fit to optimal control in SDS,
- given the design objectives of the framework.  Since KTD is an existing
- technique, the main contribution and novelty is therefore in its
- application to SDS.  The paper does a good job at explaining the
- motivation, and the level of technical details is appropriate for non-RL
- expert to understand most of the main idea.  Overall, the paper is quite
- easy to follow, except for some typos (see below).
-
- Since KTD is not new, it becomes important to have an extensive comparison
- to alternative approaches to demonstrate its pros and cons.  While a few
- natural candidates have been included, some others are ignored, and some
- may not be implemented in the best way.  Details:
-
-\begin{enumerate}
-\item  Many Bayesian RL algorithms share all (or most) claimed benefits
- of KTD.  (In fact, I think KTD may be viewed as an instance of Bayesian RL
- algorithm.)  So one wonders naturally how KTD compares to other Bayesian
- methods, say, those by Yaakov Engel (a series of papers at ICML/NIPS) and
- by Jan Peters (at ACC-08).  I am particularly wondering whether the real
- benefits here come uniquely from KTD, or from Bayesian RL in general.
-
-\textcolor{blue}{  Precisions about how Bayesian approches differ
-from the KTD approaches are given at the beginning of the Sec. III :
-``It is important to notice that Kalman filtering differs from the
-Bayesian filtering. Indeed, the Bayesian filtering aims at computing
-the complete posterior distribution of the hidden variable while
-Kalman filtering only focuses on the first and second moments of
-this distribution with constrained linear update. They are equal
-only when Gaussian distributions and linear parameterisations are
-concerned.''. A comparison with a Bayesian algorithm (namely GPTD
-proposed by Engel \textit{et al.}) has been added in the paper
-Sec.~V-C (on Fig.~3 and Fig.~6).}
-
-\item  LSPI: first, it is possible to extend it to online learning, such
- as in the AAMAS-09 paper by Lihong Li et al.
-
-\textcolor{blue}{Thank you for the reference, we added it to our
-bibliography.}
-
-Second, ill-conditioning is
- reported in the experiments, but it is common practice to to L2-regularized
- LSTD inside LSPI.  It is not surprising that, with poor conditioning,
- LSTD/LSPI become very unstable and give bad results.  One thus wonders
- whether these algorithms are implemented in the 'right' (or robust) way to
- allow for fair comparisons.
-
-\textcolor{blue}{Precisions about the experimental conditions when
-using LSPI have been given. In particular, regularisation has
-actually been tested and led to poor strategies. It is now precised}
-
-\item Fitted Q-learning: essentially the same concerns apply as in the
- case of LSPI.
-
-\item  The empirical results do *not* seem to be impressive.  Unless I
- misinterpret the plots, KTD-style algorithms only beat incremental
- algorithms like Q-learning and Sarsa.  Compared to other batch algorithms,
- KTD shows no significant advantage.
-
-\textcolor{blue}{The performance of KTD is not effectively better
-than those of the other batch algorithms but our goal in this
-article is to propose a unique algorithm to solve several learning
-issues with comparable performance. Moreover KTD processes samples
-only once while other methods process them many times.}
-
-\item In general, there is no free lunch when we come to algorithm
- selections.  KTD achieves some of the benefits at the cost of other
- disadvantages.  The paper will be more valuable to readers if the 'cons'
- are also discussed, such as convergence and optimality guarantees, to name
- a few.
-
-\textcolor{blue}{A proof of convergence of the KTD algorithm under
-certain hypothesis have been proposed (in the JAIR paper). A
-difficulty appears also as far as the choice of the KTD parameters
-are concerned. A paragraph about how to choose them has been added
-Sec. III: `` The $\hat{\theta}_{0|0}$ vector can be initialised to a
-default value or, if known, a value near the optimal one. The
-$P_{0|0}$ matrix translates the certainty the user has in the prior
-$\hat{\theta}_{0|0}$. It appears in Eq. (2) as a regularisation
-term. A common initialisation for this matrix is $\lambda
-\mathrm{Id}$. Only one parameter ($\lambda$) has thus to be chosen
-instead of $p^2$ parameters ($p$ being the number of parameters to
-learn).  The lower $\lambda$ the less certain is. The observation
-noise $P_{n_t}$ represents the confidence about the ability of the
-chosen parameterisation to represent the true Q-function.
-Additionally, the evolution noise is chosen so that $P_{v_t} = \eta
-P_{\theta_{t-1|t-1}}$ with $0 < \eta \ll 1$. The effect of this
-heuristic is to put more emphasis on more recent data (it acts as a
-forgetting factor).''. It has to be noticed that only an order of
-magnitude has to be chosen. The KTD algorithm is not sensitive to
-small changes in the parameters.}
-
-\item  Furthermore, while KTD represents a reasonable algorithm, some of the
- algorithmic choices seem arbitrary and should be justified.  For example,
- the noise terms in eqn 6 are unspecified.  It may be tempting (as probably
- is done in experiments) to assume the noises are normal, but this modeling
- assumption can be quite wrong in many RL problems.  Also, more algorithmic
- details about KTD would make the paper more self-contained.  The Kalman
- gain, for instance, is quite mysterious --- how about giving a formula for
- computing that matrix after equation 7?
-
- \textcolor{blue}{ The KTD algorithm is now explained Sec.
-III: `` The KTD algorithm is done in four steps: an \textit{initialisation} step, a \textit{prediction} step, a computational phase of \textit{statistics of interest} and finally a \textit{correction} step.
-[$\ldots$]''.
-The detail of each step is then given.
-As precised previously, a paragraph has also been added to explain the role of the different parameters.
-}
-
-\item  As said, the performance of KTD compared to fitted Q-learning or
- LSPI is not impressive.  In fact, KTD seems to converge to slightly worse
- performance.  The incremental vs. batch explanation does not make sense,
- since KTD is pretty much a batch algorithm as well --- see the authors' own
- comments on the sample efficiency of KTD below eqn 6.
-
-\textcolor{blue}{The KTD algorithm is not a batch algorithm. The
-update of the gain is done each time a sample is processed. The aim
-of the article was only to propose a single algorithm to solve
-different problems with performance that compares to specialised
-algorithms. }
-
-\item  What is LSPI not in any of the plots?  What is more, given the
- competitiveness of fitted Q-learning (and LSPI), I find it surprising to
- see the comparisons to them in only one experiments, but not in all (like
- in HIS).
-
- \textcolor{blue}{LSPI has been added to the set of curves for both systems in Fig.~1 and Fig.~2.}
-
-\item  One of the claimed benefits of KTD is the ability to handle
- nonlinear function approximators.  In the experiments, in particular figs 5
- and 6, it is a bit sad to see nonlinear alternatives provide no benefits
- (but actually cause slower learning speed).  One then asks, at least in the
- context of SDS in this paper, why should one care about nonlinear KTD.
-
-\textcolor{blue}{ Curves about the non-linear parameterisation have
-been added to better show what happens during the learning with this
-parameterisation (Fig.~4 and Fig.~8). The following explanation has
-been given: ``The different runs used to compute the mean are added
-on the graph. They show the detailed behaviour of the ANN. Contrary
-to the linear case, the learning is either very good or very poor.
-The curve representing the average results presents thus poor
-results with small sets of data.''. A method should then be found to
-automatically remove the worst runs like said in the conclusion.}
-
-\item  One final comment is about the authors' prior work published at IJCAI 2011.
- It seems some results reported here have appeared in that conference
- paper.  It would avoid future confusions if the connection is made clearer.
-
-\textcolor{blue}{The results presented in the IJCAI paper present
-only an off-policy approach. This has been clarified in the
-article.}
-
-\end{enumerate}
-
+\begin{itemize}
+\item  On sent bien qu'il faudrait une normalisation (c'est plus ou moins esquissée par les auteurs). Sans cela la
+propriété démontrée n'est pas hyper convaincante. J'encourage les auteurs à développer cet aspect esquissé page
+17 (avant la section 4).\\
+\textcolor{blue}{La phrase ``Ce cas est donc très improbable.'' (en parlant d'un classifieur qui renverrait un vecteur de récompense $\theta_C=0$) a été changée pour : ``Des mécanismes de régularisation dépendant du classifieur choisi peuvent être employés pour éviter cela.''.
+La normalisation mentionnée immédiatement après ne l'est que pour désamorcer les interrogations du lecteur quant à la présence dans la borne du terme $||\R_{\theta_c}||_\infty$.
+Le sujet de la régularisation de la récompense dans le cadre de l'IRL est assez subtil (du fait de la relation complexe entre récompense et valeur) et mérite un papier à part entière.
+Nous n'avons pas introduit ici les outils qui nous permettraient de mener cette discussion.}
+\item Une troisième limitation est que je souhaiterais être convaincu de la pertinence de l'ARI. L'application à la conduite automobile est sympathique, mais que fait-on des résultats obtenus ?\\
+  \textcolor{blue}{La première phrase de l'introduction a été augmentée afin de pointer plus finement vers la référence historique de l'ARI ainsi qu'une référence récente, qui expliquent l'intérêt de l'approche mieux que nous n'avons la place de le faire : ``historiquement proposée dans (Russell, 1998),
+ce mécanisme trouve des applications dans divers champs, de la biologie à la neuropsychologie en passant par l’économie et plus récemment la robotique (Abbeel et al., 2010)''.}
+\item Regrettons l'absence d'un état de l'art des résultats de complexité (même succinct) pour justifier le fait que l'estimation de valeur de police est plus facile que l'apprentissage par renforcement.
+\\
+  \textcolor{blue}{Dans la section 5 (comparaison à l'existant), la note de bas de page suivante a été ajoutée : ``La résolution d'un MDP (en utilisant par exemple un algorithme d'\emph{itération de la valeur}) implique en effet de calculer de manière répétée la valeur d'une politique arbitraire.
+Résoudre plusieurs fois le MDP pour des récompenses arbitraires est donc beaucoup plus dur que d'estimer une fois pour toutes l'attribut moyen de l'expert.''}
+\end{itemize}\vspace{1.5 em}
+  \textcolor{blue}{Les typos relevées ont été corrigées.}
 \vspace{.5cm}
@@
 \begin{large} \textbf{Reviewer 3:} \\ \end{large}
- Recommendation: Major Revision\\
- Comments: This paper presents how the Kalman Temporal Differences framework
- (published at JAIR by the same authors) can be applied to dialogue
- management. To show that the authors apply it to two dialogue systems one
- developed using DIPPER and one developed using the HIS dialogue management
- framework. The idea of applying the KTD framework to dialogue management is
- certainly novel and promising, however, the paper is poorly written and the
- evaluation needs to be seriously extended. The argument that because
- the algorithm performs well in simulation it should also perform well with
- real users is very weak especially for a journal submission. If things were
- so simple nobody would ever bother perform experiments with real users.
+\begin{itemize}
+\item Alg. 1 pas clair.
+\textcolor{blue}{Nous avons rajouté le nom de la fonction de score dans l'algorithme :  $s_\theta(s,a) = \theta^\top\hat{\mu}^{\pi_E}(s,a)$.
+Mais nous ne sommes pas sûrs que cela améliore vraiment les choses.
+Il semble difficile de rendre l'algorithme plus clair que ce qu'il est déjà.}
+\item Les auteurs ont mégoté sur les définitions, ce qui n'est jamais bon signe; définir $\pi_1..\pi_t$.Je suppose qu'il s'agit des actions choisies aux t premiers pas de temps.
+Donner l'intuition de c(t); clarifier ce qui se passe si la distribution stationnaire ne passe pas en s\\
+\textcolor{blue}{Les définitions sont présentes dans le papier de Munos.
+Les contraintes de place et de clarté ne nous permettent pas de reprendre toute l'explication développée par celui-ci.
+Comme la présence des $\pi_1 etc.$ sous le signe $\max$ le laisse entendre, il s'agit de variables muettes, il peut s'agir de n'importe quelle série de politiques.
+La discussion sur la signification des coefficients de concentrations est présente dans le papier cité, il apparaît délicat de la paraphraser ici sans alourdir la lecture.}
+\item Lemme 4.2 de Munos n'est pas ça du tout (version Hal).\\
+\textcolor{blue}{Pour respecter les contraintes de places, il est attendu du lecteur qu'il fasse le développement suivant afin de justifier l'inégalité (b) :
+\begin{eqnarray}
+  v^{*} - v^{\pi_E} \leq \gamma P_{\pi^*}(v^*-v^{\pi_E}) + T^*
+    v^{\pi_E} - v^{\pi_E}\\
+    v^{*} - v^{\pi_E} - \gamma P_{\pi^*}(v^*-v^{\pi_E}) \leq T^*
+    v^{\pi_E} - v^{\pi_E}\\
+    (I- \gamma P_{\pi^*})(v^*-v^{\pi_E}) \leq T^*
+    v^{\pi_E} - v^{\pi_E}\\
+    v^*-v^{\pi_E} \leq (I- \gamma P_{\pi^*}) ^{-1}(T^*
+    v^{\pi_E} - v^{\pi_E})
+ \end{eqnarray}
+(1) est donné par l'inégalité (a) de notre contribution, le passage de (3) à (4) se fait grâce au Lemme 4.2 du papier de Munos (dans sa version HAL).
+}
+\item En pratique il est satisfaisant ... p. 17. non; à la rigueur, il est suffisant...\\
+  \textcolor{blue}{Changement effectué.}
+\item L'erreur sur la projection des états ds l'espace des attributs moyens ; ce paragraphe pourrait être affiné en considérant la marge du cmc sous-jacent.\\
+  \textcolor{blue}{Nous perdrions alors en généralité, nous considérons aussi les cmc sans marges.}
+\item 4.1.
+le modèle; préciser.\\
+ \textcolor{blue}{ ``Le modèle ...'' devient ``Le modèle (les probabilités de transition $P$) ...''.}
+ \item ligne suivante, charabia.\\
+ \textcolor{blue}{ ``Soit $\Phi\in R^{|S|\times p}$ la matrice d'attributs donc les lignes contiennent les vecteurs d'attributs $\phi(s)^T$, pour tout $s\in S$.'' devient ``Soit $\Phi\in R^{|S|\times p}$ la matrice d'attributs dont les lignes sont indexées par $s\in S$ et contiennent les vecteurs d'attributs $\phi(s)^T$''.}
+ \item Discuter le rapport entre gamma et la longueur N des trajectoires disponibles.\\
+ \textcolor{blue}{Il n'y a pas de rapport.
+Nous avons pris $\gamma = 0.9$ comme le veut la coutume.
+Une étude superficielle (comprendre, on a joué avec sur quelques runs) ne montre pas de sensibilité évidente de l'algorithme à ce paramètre.
+Une étude systématique n'a pas été entreprise et ne nous semble pas extrêmement pertinente.}
+\item Si on a une bonne description psi des états, pourquoi a-t-on besoin des $\mu$ ?
+\textcolor{blue}{Car $\mu$ contient de l'information quant à la dynamique engendrée par la politique de l'expert.}
+\item
+  omettre note 1 ou ajouter des détails.
+notations 2.3 $\psi(s,a)= (psi1(s,a),...)$
+d'autres choix sont possibles:non. dire lesquels ou omettre la phrase.
+dacile, estumation, p. 16
+recontres
+\textcolor{blue}{Corrections effectuées telles que suggérées.
+}
+\end{itemize}
 
-\textcolor{blue}{Our aim was not to assess that leading experiments
-with simulated users is enough to prove the efficiency of the
-algorithms tested. The detailed explanation of our motivation are
-given Sec.~IV-A1. We precise that experiments with two differents
-systems, with different state spaces, different reward functions,
-different user simulation methods \textit{etc} give a good hint
-about the efficiency of them and the transferability to real world
-applications. In addition, DIPPER and HIS trained with the methods
-we used as baselines have been tested with real users (the
-references have been precised previously and added in the paper).}
+\newpage
 
- As I mentioned above the writing is not good. A lot of important details
- are missing, even simple things, like the definition of the reward
- function.
+\color{blue} Encore une fois, nous remercions les reviewers et les éditeurs pour le temps passé à nous aider à améliorer notre contribution.
 
-\textcolor{blue}{Indeed, we forgot to give some important details
-about how the two dialogue management frameworks are cast into an
-MDP. We added an entire paragraph to describe each of them (Sec.~4-A
-2 \& 3). }
-
-It is not clear to me why you compare with other existing
- algorithms only using the DIPPER-based system and not the HIS-based system.
- Is it because the HIS-based system is more complex?
-
-\textcolor{blue}{We added others algorithms to the HIS-based system
-comparison. Now, the KTD algorithm performance is compared to those
-of \textit{Gaussian Process Temporal Differences} (GPTD) and
-\textit{Least Square Policy Iteration} (LSPI). The GPTD algorithm is
-one of the more recent and efficient algorithm for dialogue
-management (references given in the article). The results are given
-Fig.~2, Fig.~3 and Fig.~6.}
-
-But this is exactly the case where one would want to use a superior algorithm. Is it because
- for example LSPI would have problems in the HIS-based system because of the
- large table that would have to be inverted? Please explain.
-
-\textcolor{blue}{We added a description of the state space for the
-HIS system in Sec.~IV-A3 and a description of the Q-function
-parameterisation in Sec.~IV-C2. The number of parameters is not so
-large thus we do not have matrix computation problems.}
-
-Furthermore, I don't think anyone can manage to reproduce this experiment given that
- important details are missing. For example, the section about the
- Q-function representation is very confusing. It is unclear to me what the
- structures of the Radial Basis Function networks (one per action) are.
-
-\textcolor{blue}{Indeed, an important part of the experimental
-settings was missing. We tried to make it clearer by giving an
-accurate description of them. The experimental settings are given
-Sec.~V-A. The value of the parameters used for the different
-algorithms have been precised for both systems so as the number of
-runs to plot the different runs. We also give more details about the
-Q-function representation in Sec.~IV-C. We precisely provide the
-mathematical descriptions of the parameterisations.}
-
- Anyway, what I'm saying here, is that the authors should not expect that
- readers should have to go back and read the JAIR article because the
- missing point here is not the algorithm itself but how it is applied to
- dialogue management. Thus even if someone had understood the algorithm it's
- not clear that he would manage to apply it to dialogue management and
- reproduce the experiments presented in this paper.
-
-\textcolor{blue}{A more accurate description of the KTD algorithm
-has been given Sec.~III. The different steps of the algorithm have
-been presented. A description of how the dialogue management problem
-is cast into the MDP framework is in Sec.~IV-A for both systems.}
-
- Here are my detailed comments:
-
-\begin{enumerate}
-\item  Title: The part of the title "with a sole algorithm" sounds very weird.
- What the authors want to say is that they have developed an algorithm that
- can solve a lot of problems that previously could be solved using a variety
- of algorithms. I'm not sure what the title should be but it should
- definitely be more informative.
-
-\textcolor{blue}{We tried to make the title clearer about what the
-article is about.}
-
-\item Introduction: Here the authors talk about different approaches to learning,
- e.g. dynamic programming, off-policy learning, on-policy learning, etc. I
- suggest they also mention the work of Henderson et al. "Hybrid
- Reinforcement/Supervised Learning of Dialogue Policies from Fixed
- Datasets", Computational Libguistics 2008. Here the authors use a hybrid
- model of SARSA-lambda and supervised learning to deal with the fact that
- there wasn't much data available for their complex state space.
-
-\textcolor{blue}{ Thank you for the reference. We added it to the
-introduction and the bibliography.}
-
-\item Page 13, line 48: DIPPER is not a dialogue system. It is a tool that allows
- someone to write dialogue management rules. In the same way HIS is a
- dialogue management framework. So I suggest the authors say something like
- DIPPER-based and HIS-based dialogue systems instead of DIPPER and HIS
- dialogue systems. In any case, please explain the difference in the paper.
-
-\textcolor{blue}{We modified the text accordingly.}
-
-\item  Background: In this section, I've noticed problems with notation that can
- be confusing especially for the non-expert reader. You refer to the
- Sutton-Barto book so one would think that you would use the notation of
- Sutton-Barto. So I suggest you use r\_(t+1) instead of r\_t when you refer to
- the reward that occurs after the agent performs action a and transitions
- from state s\_t to state s\_(t+1). I'm not saying that using r\_t is wrong,
- only that it may confuse most readers that are familiar with the
- Sutton-Barto notation. Also, you use notation R(s, a, s') but you don't
- explain that it is the same as r\_(t+1). Please try to be consistent. For
- example, in equations (2) and (4) (Bellman equations) you use the notation
- R(s, a, s') but then later in section 3 you use again the Bellman equations
- with r\_t (which in my opinion should become r\_(t+1)) instead of R(s, a,
- s'). Also, sometimes you use s' and sometimes s\_(t+1). Anyway, please try
- to be consistent.
-
-\textcolor{blue}{ The notations have been corrected to be
-consistent. All the different notations introduced are now defined.}
-
-\item Equation 4 is wrong. Near the end it should be "... gamma*max\_a'Q*(s', a')"
- (see section 3.8 in the Sutton-Barto book).
-
-\textcolor{blue}{Thank you for mentionning this typo. We corrected
-it.}
-
-\item  In section III you need to do a better job of explaining how the KTD
- framework is applied to the dialogue management problem. For example the
- text says "the term K\_t is the Kalman gain, which depends..." Please
- explain how you calculate this. I know there is some explanation in the
- JAIR paper but you need to provide some hints in the current paper too.
-
-\textcolor{blue}{ Some important details have been added to better
-understand how the KTD algorithm works in Sec.~III. Now, the
-different steps of the algorithm are precised. The definition of the
-Kalman gain has been added. This gain is the one which minimises the
-mean square error $J_t(\theta)$. The different parameters for the
-algorithm are also given.}
-
-\item  Then in page 17, line 38 the paper provides a very brief explanation (2
- lines) why the algorithm is sample efficient. Again, the same issue as
- before. Please provide some more information. This is one of your stronger
- points that this algorithm is sample efficient, so please explain in simple
- words why this is the case. The bottom line is that the description of the
- algorithm should be as self-sufficient as possible, minimizing the need to
- read the whole JAIR paper.
-
-\textcolor{blue}{The classical Q-learning and SARSA algorithms are
-gradient descents (first order) while KTD minimises a second error
-term (second order algorithm). The convergence in the last case is
-thus faster. We added this explanation at the end of the Sec.~III.}
-
-\item In section IV, the paper says that there are no experiments with
- non-stationarity handling and yet the authors claim that this is one of the
- most important features of the algorithm. So I suggest you perform
- experiments both in simulation (with a simulated user that can change its
- goal even if it is a hand-crafted one) and with real users that change
- their goals and see how the algorithm performs. Just saying that this was
- shown in the JAIR paper is not enough because that wasn't for dialogue
- management.
-
-\textcolor{blue}{Precisions about handling non-stationarity are
-added in Sec.~III and are experimentally presented Fig.~3. When a
-controlled and on-policy approach is used, handling the
-non-stationarity is a good feature. Indeed, the policy used for
-control varies which introduce non-stationaties.}
-
-\item In section IV, please provide a better description of the two dialogue
- systems. For example, how many possible states do we have in each one of
- them? What are the slots and the actions used?
-
-\textcolor{blue}{ A paragraph has been added to describe precisely
-each system and how the dialogue problem is cast into the MDP
-framework (state space, action space, reward function, discount
-factor) in Sec.~IV-A 2 \& 3.}
-
-\item  As I mentioned above, section IV-B is incomprehensible. You really need to
- show examples with the structure of the networks, perhaps some figures
- would help. In the same action I don't understand how you came up with the
- number 144. Also what do you mean by dimensions here, features of the
- state? Which are the 2 continuous dimensions and which are the 2 discrete
- ones?
-
-\textcolor{blue}{We also provide now a precise description of the
-two systems in Sec.~IV-A. We also added precise descriptions of the
-features used to represent the Q-function in Sec.~IV-C.}
-
-\item In section V-A the text says ``random actions are chosen among reasonable
- actions''. Could you please explain what you mean by ``reasonable actions''?
-
-\textcolor{blue}{We precised in Sec.~V-B how the handcrafted policy
-works. For the DIPPER system, a set of hand crafted rules prevents
-the system from choosing actions inconsistent with the current
-dialogue state. For example, it avoids to confirm a slot while it
-has never has been asked before.}
-
-\item In section V-A the text talks about comparing with LSPI but I can't see
- that in Figure 1. The LSPI graph is missing.
-
-\textcolor{blue}{ We previously made a mistake while compiling the
-graphs. We added LSPI for both DIPPER and HIS systems.}
-
-What is the reward function
- used in the experiments of Figure 1?
-
-\textcolor{blue}{A complete description of the rewards functions are
-given Sec.~IV-A. For the DIPPER system, each dialogue turn is
-penalised with a reward $-1$. At the end of the dialogue, the reward
-is provided the following: $-75$ for incorrect slot filling,  $+25$
-per slot filled correclty, and $-300$ if the slot is empty. For the
-HIS system, the reward function penalises each dialogue turn with
-$-1$. At the end of the dialogue, a reward of $+20$ is given if the
-DM managed to fulfil the user request. In both cases, the dialogue
-is stopped if it has more than $100$ turns. The definition of the
-reward function is not an easy task. Other solutions can be
-considered.}
-
- How come there is no variance in the
- hand-crafted graph? One would expect that the variety in the performance of
- the n-gram based simulated user would cause differences in the resulting
- dialogues.
-
-\textcolor{blue}{The variance has not been plotted on Fig.~1 because
-the error bars were too large. It has been precised in the comments
-of the graphs.}
-
-Of course that depends on what the reward function is which is
- important information not provided in the paper.
-
-\item  For the HIS-based system, why don't you compare with LSPI, etc.? Is it
- because LSPI would be in trouble because of the larger state space? If you
- want to show that this algorithm is better than existing algorithms you'd
- better show that in a more complex system than the DIPPER-based system. You
- could perhaps try the LSPI with fast feature selection to deal with the
- large state space, see the paper by Li et al. "Reinforcement Learning for
- Dialog Management using Least-Squares Policy Iteration and Fast Feature
- Selection" Interspeech 2009.
-
-\textcolor{blue}{Our aim was to show that one algorithm (KTD) can
-perform as well as standard other ones (and sometime better) while
-being generic. We added the LSPI comparison for both systems (Fig.~1
-and Fig.~2). Thank you very much for he reference. We added it to
-the bibliography. }
-
-\item In Figure 2 you only show one of the behavioural policies and the KTD one.
- Is this correct? You say that there are two behavioural policies used (I
- believe the graph shows the one with greedy-exploration). Why don't you
- show both behavioural policies? And what is the reward function here?
-
-\textcolor{blue}{We modified the figures hoping the illustration of
-our points will be clearer. We also change the caption of the curves
-to be more accurate.}
-
-\item Sometimes one algorithm is trained 50 times, another 250 times, etc. Is
- there a reason you are not being consistent?
-
-\textcolor{blue}{We precised in the article at the end of the
-Sec.~V-A that there is a difference in the number of trainings
-because the learning wich KTD, LSPI and GPTD algorithms takes much
-more time than with Q-learning and SARSA because of matrix
-inversions.}
-
- Also, in some figures the
- x-axis has turns and in others dialogues. Again, please try to be
- consistent or at least explain why you are not being consistent.
-
-\textcolor{blue}{The original choice was made different for the two
-systems. We precised how to make the link between the two notations
-at the beginning of Sec.~V-A. For the HIS system, the length of the
-dialogue can be deduced from the reward. The number of dialogue can
-thus be onverted in number of samples. For example, if the mean
-discounted cumulative reward is $14$ for a set of $50$ training
-dialogues, since the maximum reward that can be obtained is $20$ and
-that each turn a penalty of $-1$ is given, the average length of the
-dialogue is $20 - 14 = 6$ turns. The number of samples is thus
-approximately $6 \cdot 50 = 300$.}
-
-\item  In Figure 3, why don't you show the hand-crafted policy?
-
-\textcolor{blue}{Most of the figures have been modified to better
-present the result. Each figure corresponds now to a particular
-aspect of the KTD algorithm. We particularly paid attention to the
-caption of the figures.}
-
- What is the reward function? Figure 4 has the same problems as Figure 2. For on-policy learning for the graphs you use a sliding windows. Do you use that for the off-policy graphs too?
-
-\textcolor{blue}{A sliding window is only used for the controlled
-approaches (Fig.~3,4,6,7,8). It has been precised in the article
-when a sliding window is used. }
-
-\item  In section V-C you talk about 22 user actions and you say "the number of
- action for the action space is 12", but earlier when you describe the
- HIS-based system you mention
- that there are 22 system actions. Please explain.
-Also, in the same section you perform an experiment on handling non-linearities with the HIS-based
- system. I'd like to see that with the DIPPER-based system too.
-Also, in Figure 5 what is the reward function?
-
-\textcolor{blue}{Detailed description of the DIPPER ans HIS systems
-has been given Sec.~IV.}
-
-\end{enumerate}
-
-\color{blue} We hope that the modifications improve the quality of
-the paper. Whatever the final decision is, we would like to thank
-again the reviewers for their careful reading of this paper and for
-the precise and detailed comments they suggested.
-
-Best regards,
+Cordialement,
 
 \begin{flushright}
-Lucie Daubigney, Matthieu Geist, \\ Senthilkumar Chandramohan, Olivier Pietquin.
+Edouard Klein, Matthieu Geist, \\ Bilal Piot, Olivier Pietquin.
 \end{flushright}

### 2013-01-11 klein2012structured_RIA/scirl_ria.tex
Prise en compte des suggestions des reviewers 2 et 3 de RIA.

@@
 
-L'Apprentissage par Renforcement Inverse (ARI)~\cite{Russel998} est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal.
+L'Apprentissage par Renforcement Inverse (ARI) est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal ; historiquement proposé dans~\cite{Russel998}, ce mécanisme trouve des applications dans divers champs, de la biologie à la neuropsychologie en passant par l'économie et plus récemment la robotique~\cite{abbeel2010autonomous}.
 Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{seelatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) proche de celle générée par l'expert.
@@
 
-Un Processus Décisionnel de Markov (PDM)~\cite{Puterma994} est donné par le tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini\footnote{Cette contribution peut être étendue aux espaces d'état compacts moyennant le traitement de certains détails techniques.}, $\A$ est l'espace d'action fini, $\p =
+Un Processus Décisionnel de Markov (PDM)~\cite{Puterma994} est donné par le tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini, $\A$ est l'espace d'action fini, $\p =
 \{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ est l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ est la fonction de récompense sur les états et $\gamma$ est le facteur d'oubli.
@@
 Nous supposons que la règle de décision associe à l'entrée l'argument qui maximise une certaine fonction de score, celle-ci étant paramétrée linéairement, les paramètres étant appris par le classifieur.
-Formellement, soit $\psi(s,a) =
+Formellement, soit $\psi(x,y) =
 (\psi_1(x,y)  \dots  \psi_d(x,y))^\top\in \mathbb{R}^d$
@@
 Par exemple, un séparateur à vaste marge multi-classe pourrait être choisie~\cite{Guermeu007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taska005}.
-D'autres choix sont possibles.
 
@@
   \BlankLine
-  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de CMC auquel sont fournis la base d'entraînement $\D$ et la paramétrisation de la fonction de score : $\theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
+  \emph{\textbf{Calculer}} le vecteur de paramètres $\theta_c$ en utilisant l'algorithme de CMC auquel sont fournis la base d'entraînement $\D$ et la paramétrisation de la fonction de score : $s_\theta(s,a) = \theta^\top\hat{\mu}^{\pi_E}(s,a)$\;
   %
@@
 $\theta_c=0$, la règle de décision serait une politique aléatoire uniforme et nous aurions $\epsilon_c = \frac{|\A|-1}{|\A|}$, c'est-à-dire la pire erreur de classification possible.
-Ce cas est donc très improbable.
+Des mécanismes de régularisation dépendant du classifieur choisi peuvent être employés pour éviter cela.
 De fait, nous affirmons que notre approche permet, d'une certaine manière, de lever l'ambiguïté du problème de l'ARI (au moins, l'algorithme ne renvoie pas de récompense triviale comme la récompense nulle).
@@
 $(\hat{\mu}^{\pi_E}(s_i,a))_{1\leq i\leq N,a\in\A}$ que lors de la phase d'entraînement, où les $s_i$ sont les états présents dans $\D$. Il renvoie 
-$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire après l'étape de CMC. Ainsi, en pratique, il est satisfaisant d'estimer 
+$\theta_c$, vu comme une fonction de récompense, donc l'estimée de l'attribut moyen $\hat{\mu}^{\pi_E}$ n'est plus nécessaire après l'étape de CMC. Ainsi, en pratique, il est suffisant d'estimer 
 $\hat{\mu}^{\pi_E}$ correctement uniquement pour les couples état-action $(s_i,a)_{1\leq i\leq
@@
 
-Si le modèle est connu, il est possible de calculer l'attribut moyen de manière exacte.
-Soit $\Phi\in\mathbb{R}^{|\s|\times p}$ la matrice d'attributs donc les lignes contiennent les vecteurs d'attributs $\phi(s)^\top$, pour tout
-$s\in\s$.
+Si le modèle (les probabilités de transition $P$) est connu, il est possible de calculer l'attribut moyen de manière exacte.
+Soit $\Phi\in\mathbb{R}^{|\s|\times p}$ la matrice d'attributs dont les lignes sont indexées par $s\in\s$ et contiennent les vecteurs d'attributs $\phi(s)^\top$.
+
 %
@@
 SCIRL n'a pas besoin de résoudre le problème direct, mais uniquement d'estimer l'attribut moyen de la politique de l'expert.
-Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une seule fois un problème d'évaluation de politique.
+Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une seule fois un problème d'évaluation de politique\footnote{La résolution d'un MDP (en utilisant par exemple un algorithme d'\emph{itération de la valeur}) implique en effet de calculer de manière répétée la valeur d'une politique arbitraire.
+Résoudre plusieurs fois le MDP pour des récompenses arbitraires est donc beaucoup plus dur que d'estimer une fois pour toutes l'attribut moyen de l'expert.}.
 Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI, par exemple~\cite{boularia011}).
@@
 Nous mesurons les performances de chacune des approches grâce à $\E_{s\sim \mathcal{U}}[v^\pi_{\R_E}(s)]$, avec $\mathcal{U}$
-la distribution uniforme (ce qui permet de tester la capacité de généralisation de chaque approche même pour des états infréquemment recontrés), $\R_E$ étant la récompense de l'expert et $\pi$ est l'une des politiques suivantes: la politique optimale pour $\R_E$ (point de référence haut),
+la distribution uniforme (ce qui permet de tester la capacité de généralisation de chaque approche même pour des états infréquemment rencontrés), $\R_E$ étant la récompense de l'expert et $\pi$ est l'une des politiques suivantes: la politique optimale pour $\R_E$ (point de référence haut),
 la politique optimale pour une récompense aléatoire (point de référence bas), la politique optimale pour $\R_{\theta_c}$ (SCIRL), la politique produite par PIRL et la règle de décision du classifieur.

### 2013-01-11 klein2012structured_RIA/revisions.tex
Modifs de Matthieu et Olivier sur la letter de révisions.

@@
 
-\address{Edouard Klein,  Matthieu Geist, \\Bilal Piot, Olivier Pietquin \\ IMS – MaLIS Research Group, Sup\'elec \\ 2 rue Edouard Belin, 57070 Metz (France)}
+\address{Edouard Klein,  Bilal Piot,\\Matthieu Geist, Olivier Pietquin \\ IMS – Groupe de recherche MaLIS, Sup\'elec \\ 2 rue Edouard Belin, 57070 Metz (France)}
 %\signature{Lucie Daubigney, \\ Matthieu Geist, \\ Senthilkumar Chandramohan, \\ Olivier Pietquin.}
@@
 
-\begin{letter}{\large \textbf{Révisions sur le manuscrit ``Classification structurée pour l'apprentissage par renforcement inverse''}}
+\begin{letter}{\large \textbf{Révisions du manuscrit ``Classification structurée pour l'apprentissage par renforcement inverse'' (RIA, N$^o$ spécial JFPDA 11+12)}}
 \bigskip
-\opening{Chers éditeurs, Chers reviewers}
+\opening{Chers éditeurs, chers relecteurs}
 
@@
 
-Nous vous remercions pour le temps passé à relire notre manuscrit.
-Les remarques des reviewers étant largement orthogonales, nous les prenons en compte une par une.
-Les changements apportés au manuscrit sont reportés ici en bleu.
+Tout d'abord, nous souhaitons vivement vous remercier pour le temps passé à relire et commenter notre manuscrit.
+L'ensemble de vos judicieuses suggestions ont été prises en compte, nous espérons que cela améliorera le manuscrit.
+
+Le reste de cette lettre précise les modifications effectées et répond aux questions et commentaires des relecteurs. Dans un soucis de clarté, nous reproduisons les relectures originales en noir.
+
+Dans l'espoir d'un avis de publication favorable, veuillez agréer nos salutations.
+
+\begin{flushright}
+Edouard Klein, Bilal Piot, \\ Matthieu Geist, Olivier Pietquin.
+\end{flushright}
 
@@
 \color{black}
-\begin{large} \textbf{Reviewer 1:} \\ \end{large}
+\begin{large} \textbf{Relecteur 1:} \\ \end{large}
+RESUME : LE PAPIER SE PLACE DANS LE CADRE DE L’APPRENTISSAGE PAR RENFORCEMENT INVERSE (ARI)
+DONT LE BUT EST D’APPRENDRE UNE FONCTION DE RECOMPENSE POUR LAQUELLE LE COMPORTEMENT D’UN
+EXPERT EST (QUASI) OPTIMAL. UN NOUVEL ALGORITHME, SCIRL, EST PROPOSE ; LA PARTICULARITE
+PRINCIPALE DE SCIRL EST DE NE PAS AVOIR BESOIN DE RESOUDRE DE PROBLEME D’AR DIRECT (INFERER
+UNE POLITIQUE OPTIMALE A PARTIR DES DONNEES). UNE ANALYSE THEORIQUE EST PROPOSEE, ET DES
+RESULTATS EXPERIMENTAUX OBTENUS SUR UN PROBLEME INTERESSANT SONT EGALEMENT FOURNIS.
+LA CONTRIBUTION DE CE PAPIER EST CLAIREMENT IDENTIFIEE ET INTERESSANTE. LE PAPIER EST AGREABLE
+A LIRE (EN DEHORS DE QUELQUES COQUILLES DETAILLEES CI-DESSOUS), ET LES RESULTATS THEORIQUES ET
+EMPIRIQUES TRES PROMETTEURS. ON NE SAURAIT QU’ENCOURAGER LES AUTEURS A POURSUIVRE LEURS
+EXPERIMENTATIONS DE SCIRL SUR DES PROBLEMES ENCORE PLUS COMPLEXES (COMME SUGGERE DANS LA
+CONCLUSION).
+COQUIILLES :\\
+P.11 : « LA REGLE DE DECISION POUR UNE ETAT » : […] UN ETAT ;\\
+P.12 : « (MAIS OBSERVEE PAR INTERACTION […] » : (MAIS OBSERVE PAR INTERACTION […] ;\\
+P.13 « UNE MACHINE A VECTEUR SUPPORT » : IL SEMBLERAIT QU’ON UTILISE PLUS SOUVENT LA TRADUCTION
+« SEPARATEUR A VASTE MARGE » ;\\
+P.15 : ON A D’ABORD « THEOREME 1 » PUIS « PREUVE 2 ». IL POURRAIT ETRE JUDICIEUX SOIT D’ENLEVER
+LE 2 SOIT DE LE REMPLACER PAR UN 1 ;\\
+P.16 : « IL EST DACILE » : IL EST FACILE ;\\
+P.16 : « INJECTER LES BORNES DES EQS. » : EN FRANÇAIS, LES ABREVIATIONS NE SE METTENT PAS AU
+PLURIEL ;\\
+P.17 : « LE CLASSIFIEUR N’UTILISE QUE [FORMULE MATH] QUE LORS […] » : IL Y A UN « QUE » EN TROP ;\\
+P.17 : « CE QUI PERMET D’ENVISAGER UNE SIMPLE ESTIMATION DE MONTE-CARLO PAR EXEMPLE » : CE QUI
+PERMET D’ENVISAGER, PAR EXEMPLE, UNE SIMPLE ESTIMATION MONTE CARLO ;\\
+P.18 : « (ET LA DYNAMIQUE RESULTANTES) » : ET LA DYNAMIQUE RESULTANTE
+P.20 : « EN UTILISANT LE PRINCIPE DECRIT EQ. (5) » : EN UTILISANT LE PRINCIPE DECRIT PAR EQ. (5)\\
+P.20 : « IL RENTRENT DANS LE CHAMP GENERALEMENT APPELE […] » : ILS RENTRENT […]\\
+P.21 : « PAR L’UTILISATION D’UNE HEURISTIQUE QUI DEPASSENT » : […] QUI DEPASSE\\
+P.21 : « LES SORTIES DE ROUTE SONT POSSIBLE » : […] SONT POSSIBLES\\
+P.21 : « TOUS LES OBJETS REQUIS (…) ONT ETE CALCULEES » : […] ONT ETE CALCULES\\
+P.22 : « LA PERFORMANCE DE CHACUNE DES APPROCHE » : […] DES APPROCHES\\
+DANS LES REFERENCES :\\
+- HARMONISER LA MISE DE MAJUSCULES DANS LES TITRES (CERTAINES REFERENCES ONT DES MAJUSCULES
+A CHAQUE MOT, D’AUTRE NON – EN PRINCIPE LA REGLE EST : MAJUSCULES A CHAQUE MOT POUR LES LIVRES
+ET LES THESES, PAS POUR LES AUTRES)\\
+- METTRE DES MAJUSCULES POUR LES NOMS DES CONFERENCES ET DES JOURNAUX\\
+- METTRE LES ACRONYMES DES CONFERENCES EN MAJUSCULES (ICML, ECML, NIPS, AISTATS, COLT)\\
+- GUERNEUR Y. (2007) : « VC THOERY » : VC THEORY\\
 \textcolor{blue}{
@@
 \newpage
-\begin{large} \textbf{Reviewer 2:} \\ \end{large}
-\begin{itemize}
-\item  On sent bien qu'il faudrait une normalisation (c'est plus ou moins esquissée par les auteurs). Sans cela la
+\begin{large} \textbf{Relecteur 2:} \\ \end{large}
+La première limitation est le critère retenu pour valider une fonction de reward comme étant correcte pour le
+problème d'ARI.
+On établit ici que le reward a la propriété suivante:
+"la stratégie observée est quasi optimale pour ce reward"
+Comme le disent les auteurs, c'est vrai aussi pour la fonction nulle.
+On sent bien qu'il faudrait une normalisation (c'est plus ou moins esquissée par les auteurs). Sans cela la  
 propriété démontrée n'est pas hyper convaincante. J'encourage les auteurs à développer cet aspect esquissé page
 17 (avant la section 4).\\
-\textcolor{blue}{La phrase ``Ce cas est donc très improbable.'' (en parlant d'un classifieur qui renverrait un vecteur de récompense $\theta_C=0$) a été changée pour : ``Des mécanismes de régularisation dépendant du classifieur choisi peuvent être employés pour éviter cela.''.
+\textcolor{blue}{La phrase ``Ce cas est donc très improbable.'' (en parlant d'un classifieur qui renverrait un vecteur de récompense $\theta_C=0$) a été changée pour : ``Ce cas est très improbable, l'objectif du classifieur étant de minimiser $\epsilon_C$''.
 La normalisation mentionnée immédiatement après ne l'est que pour désamorcer les interrogations du lecteur quant à la présence dans la borne du terme $||\R_{\theta_c}||_\infty$.
-Le sujet de la régularisation de la récompense dans le cadre de l'IRL est assez subtil (du fait de la relation complexe entre récompense et valeur) et mérite un papier à part entière.
-Nous n'avons pas introduit ici les outils qui nous permettraient de mener cette discussion.}
-\item Une troisième limitation est que je souhaiterais être convaincu de la pertinence de l'ARI. L'application à la conduite automobile est sympathique, mais que fait-on des résultats obtenus ?\\
-  \textcolor{blue}{La première phrase de l'introduction a été augmentée afin de pointer plus finement vers la référence historique de l'ARI ainsi qu'une référence récente, qui expliquent l'intérêt de l'approche mieux que nous n'avons la place de le faire : ``historiquement proposée dans (Russell, 1998),
-ce mécanisme trouve des applications dans divers champs, de la biologie à la neuropsychologie en passant par l’économie et plus récemment la robotique (Abbeel et al., 2010)''.}
-\item Regrettons l'absence d'un état de l'art des résultats de complexité (même succinct) pour justifier le fait que l'estimation de valeur de police est plus facile que l'apprentissage par renforcement.
-\\
-  \textcolor{blue}{Dans la section 5 (comparaison à l'existant), la note de bas de page suivante a été ajoutée : ``La résolution d'un MDP (en utilisant par exemple un algorithme d'\emph{itération de la valeur}) implique en effet de calculer de manière répétée la valeur d'une politique arbitraire.
-Résoudre plusieurs fois le MDP pour des récompenses arbitraires est donc beaucoup plus dur que d'estimer une fois pour toutes l'attribut moyen de l'expert.''}
-\end{itemize}\vspace{1.5 em}
-  \textcolor{blue}{Les typos relevées ont été corrigées.}
+Dans l'absolu, tous nos résultats et algorithmes sont invariants par dilatation, cette norme peut être imposée par l'utilisateur.}\\
+La seconde limitation à ma recommandation d'acceptation ferme est que je ne suis pas expert d'apprentissage par
+renforcement inverse; si je suis convaincu par la justesse de l'analyse et la pertinence de l'algorithme, je suis mal
+placé pour garantir la nouveauté.
+Une troisième limitation est que je souhaiterais être convaincu de la pertinence de l'ARI. L'application à la conduite automobile est sympathique, mais que fait-on des résultats obtenus ?\\
+  \textcolor{blue}{Considérant l'AR comme pertinent, nous sommes convaincus que trouver la bonne récompense à optimiser est un problème intéressant.
+Afin de l'expliquer, la première phrase de l'introduction a été augmentée.
+Elle pointe plus finement vers la référence historique de l'ARI ainsi qu'une référence récente, qui expliquent l'intérêt de l'approche mieux que nous n'avons la place de le faire : ``historiquement proposée dans (Russell, 1998),
+ce mécanisme trouve des applications dans divers champs, de la biologie à la neuropsychologie en passant par l’économie et plus récemment la robotique (Abbeel et al., 2010)''.}\\
+On utilise ici des fonctions de reward pas trop informatifs; il est précisé dans l'article que des travaux existants.
+Regrettons l'absence d'un état de l'art des résultats de complexité (même succinct) pour justifier le fait que l'estimation de valeur de police est plus facile que l'apprentissage par renforcement. \\
+  \textcolor{blue}{Un état de l'art détaillé est difficile à effectuer, tout dépendant de la méthode utilisée pour résoudre le MDP.
+Pour expliquer les grandes lignes, dans la section 5 (comparaison à l'existant), la note de bas de page suivante a été ajoutée : ``La résolution d'un MDP (en utilisant par exemple un algorithme d'\emph{itération de la valeur}) implique en effet de calculer de manière répétée la valeur de politiques arbitraires.
+Résoudre plusieurs fois le MDP pour des récompenses arbitraires est donc beaucoup plus dur que d'estimer une fois pour toutes l'attribut moyen de l'expert (sans mentionner le problème de l'apprentissage \emph{off-policy}).''}\\
+
+  Quelques typos:
+"L'approche proposée est résuméE"
+"estUmation"\\
+  \textcolor{blue}{Les problèmes d'ordre orthographique ou gramatical relevés ont été corrigés.}
 \vspace{.5cm}
 \newpage
-\begin{large} \textbf{Reviewer 3:} \\ \end{large}
+\begin{large} \textbf{Relecteur 3:} \\ \end{large}
 \begin{itemize}
 \item Alg. 1 pas clair.
-\textcolor{blue}{Nous avons rajouté le nom de la fonction de score dans l'algorithme :  $s_\theta(s,a) = \theta^\top\hat{\mu}^{\pi_E}(s,a)$.
-Mais nous ne sommes pas sûrs que cela améliore vraiment les choses.
-Il semble difficile de rendre l'algorithme plus clair que ce qu'il est déjà.}
+\textcolor{blue}{Nous avons rajouté le nom de la fonction de score dans l'algorithme :  $s_\theta(s,a) = \theta^\top\hat{\mu}^{\pi_E}(s,a)$. Nous espérons que cela rend l'algorithme plus clair.}
 \item Les auteurs ont mégoté sur les définitions, ce qui n'est jamais bon signe; définir $\pi_1..\pi_t$.Je suppose qu'il s'agit des actions choisies aux t premiers pas de temps.
 Donner l'intuition de c(t); clarifier ce qui se passe si la distribution stationnaire ne passe pas en s\\
-\textcolor{blue}{Les définitions sont présentes dans le papier de Munos.
+\textcolor{blue}{Les définitions sont présentes dans (Munos, 2007), l'article cité.
 Les contraintes de place et de clarté ne nous permettent pas de reprendre toute l'explication développée par celui-ci.
-Comme la présence des $\pi_1 etc.$ sous le signe $\max$ le laisse entendre, il s'agit de variables muettes, il peut s'agir de n'importe quelle série de politiques.
-La discussion sur la signification des coefficients de concentrations est présente dans le papier cité, il apparaît délicat de la paraphraser ici sans alourdir la lecture.}
+Comme la présence des $\pi_i$ sous le signe $\max$ le laisse entendre, il s'agit de variables muettes, il peut s'agir de n'importe quelle série de politiques.
+La discussion sur la signification des coefficients de concentration (notion relativement standard en apprentissage par renforcement) est présente dans le papier cité, il apparaît délicat de la paraphraser ici sans alourdir la lecture.
+Le lecteur intéressé par les preuves peut se référer à ce papier fondamental en la matière.}
 \item Lemme 4.2 de Munos n'est pas ça du tout (version Hal).\\
@@
  \textcolor{blue}{Il n'y a pas de rapport.
-Nous avons pris $\gamma = 0.9$ comme le veut la coutume.
-Une étude superficielle (comprendre, on a joué avec sur quelques runs) ne montre pas de sensibilité évidente de l'algorithme à ce paramètre.
-Une étude systématique n'a pas été entreprise et ne nous semble pas extrêmement pertinente.}
+Nous avons pris $\gamma = 0.9$.
+Une étude superficielle ne montre pas de sensibilité évidente de l'algorithme à ce paramètre.
+Une étude systématique n'a pas été entreprise, mais nous l'envisageaons, merci pour la suggestion.}
 \item Si on a une bonne description psi des états, pourquoi a-t-on besoin des $\mu$ ?
-\textcolor{blue}{Car $\mu$ contient de l'information quant à la dynamique engendrée par la politique de l'expert.}
+\textcolor{blue}{Car $\mu$ contient de l'information quant à la dynamique engendrée par la politique de l'expert.
+Cela est clairement illustré expérimentalement.
+Nous avons d'ailleurs augmenté le dernier paragraphe de la section 6 de la phrase suivante ``Cela illustre la pertinence d'utiliser $\mu$ et non $\phi$ dans la paramétrisation de la fonction de score, l'attribut moyen contient en effet de l’information quant à la dynamique engendrée par la politique de l’expert.''}
 \item
@@
 
-\newpage
-
-\color{blue} Encore une fois, nous remercions les reviewers et les éditeurs pour le temps passé à nous aider à améliorer notre contribution.
-
-Cordialement,
-
-\begin{flushright}
-Edouard Klein, Matthieu Geist, \\ Bilal Piot, Olivier Pietquin.
-\end{flushright}
-
-%\closing{Best regards,}
-
 \end{letter}

### 2013-01-11 klein2012structured_RIA/scirl_ria.tex
Modifs de Matthieu et Olivier sur la letter de révisions.

@@
 $\theta_c=0$, la règle de décision serait une politique aléatoire uniforme et nous aurions $\epsilon_c = \frac{|\A|-1}{|\A|}$, c'est-à-dire la pire erreur de classification possible.
-Des mécanismes de régularisation dépendant du classifieur choisi peuvent être employés pour éviter cela.
+Ce cas est très improbable, l'objectif du classifieur étant de minimiser $\epsilon_C$.
 De fait, nous affirmons que notre approche permet, d'une certaine manière, de lever l'ambiguïté du problème de l'ARI (au moins, l'algorithme ne renvoie pas de récompense triviale comme la récompense nulle).
@@
 SCIRL n'a pas besoin de résoudre le problème direct, mais uniquement d'estimer l'attribut moyen de la politique de l'expert.
-Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une seule fois un problème d'évaluation de politique\footnote{La résolution d'un MDP (en utilisant par exemple un algorithme d'\emph{itération de la valeur}) implique en effet de calculer de manière répétée la valeur d'une politique arbitraire.
-Résoudre plusieurs fois le MDP pour des récompenses arbitraires est donc beaucoup plus dur que d'estimer une fois pour toutes l'attribut moyen de l'expert.}.
+Autrement dit, plutôt que de résoudre plusieurs fois le problème de l'optimisation d'une politique, nous ne résolvons qu'une seule fois un problème d'évaluation de politique\footnote{La résolution d'un MDP (en utilisant par exemple un algorithme d'\emph{itération de la valeur}) implique en effet de calculer de manière répétée la valeur de politiques arbitraires.
+Résoudre plusieurs fois le MDP pour des récompenses arbitraires est donc beaucoup plus dur que d'estimer une fois pour toutes l'attribut moyen de l'expert (sans mentionner le problème de l'apprentissage \emph{off-policy}).}.
 Cela amène des garanties théoriques (ce qui n'est pas le cas de tous les algorithmes d'ARI, par exemple~\cite{boularia011}).
@@
 Augmenter le nombre de transitions améliorerait ses performances, mais après 400 transitions, il ne fonctionne pas aussi bien que SCIRL avec uniquement une dizaine de transitions.
+Cela illustre la pertinence d'utiliser $\mu$ et non $\phi$ dans la paramétrisation de la fonction de score, l'attribut moyen contient en effet de l’information quant à la dynamique engendrée par la politique de l’expert.
 SCIRL fonctionne particulièrement bien ici : après seulement une centaine de transitions il atteint les performances de PIRL.

### 2013-03-26 piot2012cascading/ECML13_cascading.tex
Update piot2012cascading for ECML submission.

@@
 Practically, the reward can be seen as a succinct description of a task.
-Discovering it removes the coupling between understanding the task and learning how to fulfill it that exists in AL.
+Discovering it removes the coupling that exists in AL between understanding the task and learning how to fulfill it.
 IRL allows the use of (A)DP or RL techniques to learn the task from the computed reward function.
@@
 Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\eqref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy.
-Thus, the demonstrated expert policy will be optimal for the related reward function.
+Thus, the demonstrated expert policy will be optimal for the corresponding reward function.
 We propose to use a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function.
-Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}).
+Besides, in order to retrieve via the Bellman equation the reward associated with the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}).
 That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI).
@@
 
-This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-experts policies as inputs.
+This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-expert policies as inputs.
 Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
@@
 
-Differences and similarities with existing AL or IRL approaches are succinctly exposed in Sec.~\ref{sec:related}.
+Differences and similarities with existing AL or IRL approaches are succinctly discussed in Sec.~\ref{sec:related}.
 \section{Background and notations}
@@
 We work with finite MDPs \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$.
-The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamics of the MDP.
+The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in (0,1)$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamics of the MDP.
 Thus, for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing action $a$ in state $s$.
@@
 The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
-The stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
+Given that $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$, the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
 \begin{equation}
-  \rho_\pi^TP_\pi = \rho_\pi^T,
+  \rho_\pi^TP_\pi = \rho_\pi^T\textrm{, with }X^T\textrm{ being the transpose of }X.
   \end{equation}
-where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$. The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.
+The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.
 
@@
 
-When the state space is too large to allow matricial representations or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
+When the state space is too large to allow matrix representations or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
 
@@
 The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal).
-Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.
+Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although a lot of progress have been made, see Sec.~\ref{sec:related}.
 \section{The cascading algorithm}
@@
 Our first step towards a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (SFMC$^2$ for short).
-This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$.
-The classification rule $\pi_C\in A^S$ simply selects (one of) the label(s) that achieves the highest score for the given inputs:
+This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given action\footnote{Here, actions play the role of what is known as {\it labels} or {\it categories} when talking about classifiers.} $a\in A$ with a certain input $s\in S$.
+The classification rule $\pi_C\in A^S$ simply selects (one of) the action(s) that achieves the highest score for the given inputs:
 \begin{equation}
@@
 As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system.
-We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to.
-Action $a_j$ needs not be chosen by any particular policy.
+We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to.
+Action $a_j$ need not be chosen by any particular policy.
 The dataset $D_R$ brings us information about the dynamics of the system.
@@
 The complete algorithm is given in Alg.~\ref{algo:cascading}.
+
+There is no particular constraint on $D_C$ and $D_R$.
+Evidently, there is a direct link between various qualities of those two sets (amount of data, statistical representativity, etc.) and the classification and regression error.
+The exact nature of the relationship between these quantities depends on which classifier and regressor are chosen.
+The theoretical analysis of Sec.\ref{sec:analysis} abstracts itself from the choice of a regressor and a classifier and from the composition of $D_C$ and $D_R$ by reasoning with the classification and regression error.
+In Sec.\ref{sec:experiments}, %FIXME subsec heuristic
+the use of a single dataset to create both $D_C$ and $D_R$ is thoroughly explored.
 \begin{algorithm}%[H]
@@
   \begin{itemize}
-  \item the subscript notation already used for $R_\pi$ and $P_\pi$ in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are repectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
+  \item the subscript notation already used for $R_\pi$ and $P_\pi$ in Sec.~\ref{sec:background} meaning that, given an $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$,  $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ are respectively such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$ ;
   \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
-  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$ ;
+  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$, one property of which is that $\forall X, \mu^TX \leq \|X\|_{1,\mu}$;
   \end{itemize}
@@
   \end{itemize}
-  The constant $C_*$ could be estimated \emph{a posteriori} (after $\hat{R}^C$ is computed).
-We could upper bound $C_*$ by a more usual and general concentration coefficient but $C_*$ gives a tighter final result (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
+  The constant $C_*$ can be estimated \emph{a posteriori} (after $\hat{R}^C$ is computed).
+{\it A priori}, $C_*$ can be upper bound by a more usual and general concentration coefficient but $C_*$ gives a tighter final result: one can informally see $C_*$ as a measure of the similarity between the distributions indeuced by $\hat\pi_C$ and $\pi_E$ (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
 \item $\Delta q$, defined by 
@@
   \end{itemize}
-
+Readers familiar with the work presented in \cite{klein2012scirl} will see some similarities between the theoretical analyses of SCIRL and CSI.
+As both algorithms solve the same problem, both theorems naturally bound the same criterion.
+Another shared feature is the use of the score function $q$ of the classifier as a proxy for the action-value function of the expert.
+The same structure can therefore be found in both analyses as there is only so many ways to study error propagation in IRL algorithms.\\
+The attentive reader, however, will percieve that similarities stop there.
+The error terms occurring in the two bounds are not related to one another.
+As CSI makes no use of the feature expectation of the expert, what is known as $\bar\epsilon_Q$ in \cite{klein2012scirl} does not appear in this analysis.
+Likewise, the regression error $\epsilon_R$ of this paper does not appear in the analysis of SCIRL, which does not use a regressor.
+Perhaps more subtly, the classification error and classification policy known in both papers as $\epsilon_C$ and $\pi_C$ are not the same.
+The classification policy of SCIRL is no tantamount to what is called $\pi_C$ here.
+For SCIRL, $\pi_C$ is the greedy policy for an approximation of the value function of the expert with respect to the reward output by the algorithm.
+For CSI, $\pi_C$ is the decision rule of the classifier, an object that is not aware of the structure of the MDP.
 \begin{theorem}
@@
 \\
-&=\frac{\gamma}{1-\gamma}\rho_E^T(q_{\pi_C}-q_{\pi_E})\leq \frac{\gamma}{1-\gamma}\Delta q \epsilon_C.
+&=\frac{\gamma}{1-\gamma}\rho_E^T(q_{\pi_C}-q_{\pi_E})\leq \frac{\gamma}{1-\gamma}\Delta q \epsilon_C \leq  \frac{1}{1-\gamma}\Delta q \epsilon_C\textrm{ as }\gamma<1.
 \end{align}
-So the upper bound for the second term term is: $\rho_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\Delta q}{1-\gamma}\epsilon_C$.
+So the upper bound for the second term is: $\rho_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\Delta q}{1-\gamma}\epsilon_C$.
 If we combine all of the results, we obtain the final bound:
@@
 In this section, we put our approach under stress to empirically demonstrate its behavior.
-We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data and a precise description of the way we instanciated the CSI algorithm are given on Sec.~\ref{subsec:instanciation}.
-We then use a classical RL benchmark, the Inverted Pendulum (Sec.~\ref{subsec:IP}), to provide a qualitative illustration of the proposed CSI approach.
-We finally provide quantitative results and comparison with state-of-the-art approaches on a highway driving simulator (Sec.~\ref{subsec:highway}).
-\subsection{Instanciation}
-\label{subsec:instanciation}
+We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data, the rationale behing the heuristics we use to compensate for the dire data scarcity and a quick word about the contender CSI is compared to are given Sec.~\ref{subsec:generalities}.
+We supply quantitative results and comparison of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) %FIXME: subsec moutnain car
+and then on a more difficult highway driving simulator (Sec.~\ref{subsec:highway}).
+\subsection{Generalities}
+\label{subsec:generalities}
 \subsubsection{Data scarcity}
 \label{subsubsec:data}
-In order to create a somewhat realistic environment, we decided that the CSI algorithm could neither use the transition probabilities nor a simulator.
-The only data it could access is sampled according to the expert policy.
+The CSI algorithm was designed not to repeatedly solve the RL problem.
+This feature makes it particularly well-suited to environments where sampling the MDP is difficult or costly.
+In the experiments, CSIis fed only with data sampled according to the expert policy.
 This corresponds for example to a situation where a costly system can only be controlled by a trained operator as a bad control sequence could lead to a system breakdown.
 
-More precisely, the expert controls the system for $n$ runs of length $n$, with values of $n$ in $\{3,7,10,15,20\}$, giving $n^2$ samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$.
-The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
+More precisely, the expert controls the system for $M$ runs of lengths ${L_i}_{1\leq i \leq M}$, %with values of $n$ in $\{3,7,10,15,20\}$,
+giving samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$. The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
 \begin{equation}
@@
 \end{equation}
-There is no particuliar problem for the classification step.
-It is however not reasonable to construct the dataset $D_R$ only from expert transitions and expect a small regression error term $\epsilon_R$.
-Indeed, the dataset $D_E$ only samples the dynamics induced by the expert's policy and not the whole dynamics of the MDP.
-The regression step also needs samples $((s_j,a_j),\hat r_j)_j$ drawn from non expert actions.
-We address this problem by using a heuristics.
 \subsubsection{Heuristics}
 \label{subsubsec:heuristics}
-From the raw expert dataset $D_E$ we can construct samples $((s_j=s_k,a_j=a_k),\hat r_j)$ with $\hat r_j$ as specified in Eq.~\eqref{eq:rj}.
-To compensate for the lack of information about the dynamics of the whole MDP, we assert that disagreeing with the expert's choice is undesirable.
-Mathematically, we do that by artificially introducting samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
+
+There is no particuliar problem for the classification step.
+It is however not reasonable to construct the dataset $D_R = \{((s_k,a_k=\pi_E(s_k)),\hat r_k)_k\}$ only from expert transitions and expect a small regression error term $\epsilon_R$.
+Indeed, the dataset $D_E$ only samples the dynamics induced by the expert's policy and not the whole dynamics of the MDP.
+This means that for a certain state $s_k$ we only know the corresponding expert action $a_k=\pi_E(s_k)$ and the following state $s'_k$ sampled according to the MDP dynamics : $s'_k \sim P(\cdot|s_k,a_k)$.
+For the regression to be meaningful, we need samples associating the same state $s_k$ and a different action $a\neq a_k$ with a datapoint $\hat r \neq \hat r_k$.\\
+Recall that $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ (Eq.~\ref{eq:rj}), without knowing $s'_k \sim P(\cdot|s_k,a\neq a_k)$, we can not provide the regressor with a datapoint to associate with $(s_k,a\neq a_k)$.
+We will artificially augment the dataset $D_R$ with samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
 \begin{equation}
@@
 \end{equation}
+This heuristics intructs the regressor to associate a state-action tuple disagreeing with the expert ({\it i.e.
+}$(s_k,a\neq a_k)$) with a reward strictly inferior to any of those associated with expert state action tuples ({\it i.e.
+}$(s_k,a_k=\pi_E(s_k))$).
+Semantically, we are asserting that disagreeing with the expert in states the expert visited is a bad idea, thus the shape of the reward function should reflect it.
+This heuristics says nothing about states absent from the expert dataset.
+For such states the generalization capabilities of the regressor and, later on, the exploration of the MDP by an agent optimizing the reward will solve the problem.
 Although this heuristics was not analyzed in Sec.~\ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
+\subsubsection{Comparison with state of the art approaches}
+The similar looking yet fundamentally different algorithm SCIRL is an obvious choice as a contender to CSI as it advertises the same ability to work with very little data, without repeatedly solving the RL problem.
+In both experiments we give the exact same data to CSI and SCIRL.
+
+The algorithm of \cite{dvijotham2010inverse} also advertises not having to solve the RL problem, but needs to deal with linearly solvable MDPs, therefore we could not include it in our tests.
+The Relative Entropy (RE) method of \cite{boularias2011relative} has no such need, so we included it in our benchmarks.
+It could not, however, work with the small amount of data we provided SCIRL and CSI with, so to allow for importance sampling, we created another dataset $D_{random}$ that was used by RE but not by SCIRL nor CSI.
+
+Finally, the classification policy $\pi_C$ output by the classification step of CSI was evaluated as well.
+Comparing classification and IRL algorithms makes no sense if the object of interest is the reward itself as can be envisionned in a biological or economical context.
+It is however perfectly sound to do so in an imitation context where what matters is the performance of the agent with respect to some objective criterion.
+Both experiments use such a criterion.
+Calssification algorithms don't have to optimize any reward, the classification policy can directly be used in the environment.
+IRL algorithms, on the other hand, output a reward that must then be plugged in a RL or DP algorithm to get a policy.
+In each benchmark we used the same (benchmark-dependant) algorithm to get a policy from each of the three rewards output by SCIRL, CSI and RE.
+It is these policies whose performance we show.
+Finding a policy from a reward if of course a non-trivial problem that should not be swept under the rug, nevertheless we choose not to concern ourselves with it here as we wish to focus on IRL algorithms, not RL or DP algorithms.
+In this regard, using a classifier that directly outputs a policy may seem a much simpler solution, but we hope that the reader will be convinced that the gap in performance between classification and IRL is worth the trouble.
+
+We do not compare CSI to other IRL algorithms requiring repeatedly solving the MDP : as we would need to provide them with enough data to do so, the comparison makes little sense.
 \subsubsection{Subroutines choices}
@@
 In a simple $k$-nearest neighboor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighboors of $s$.
-The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}).
-In our experiments, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi: S\times A \rightarrow \mathbb{R}^p$ solves:
+The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}) (we use a SVM in the mountain car experiment Sec.~\ref{sec:mountaincar}).
+In the highway experiment, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi: S\times A \rightarrow \mathbb{R}^p$ solves:
 \begin{eqnarray*}
@@
 The score function of this classifier is simply $q(s,a) = \theta^T \phi(s,a)$.
-The implementation of the Structured Classification for IRL (SCIRL) algorithm \cite{klein2012scirl} (see Sec.~\ref{sec:related}) is based on this.
-We chose to use it in our CSI implementation to provide a fair comparison between both algorithms.
-
-Ease of implementation leads us to choose a simple least-square regressor for our experiments.
+We chose a SVR as a regressor in the mountain car experiment and a simple least-square regressor on the highway.
 Using the same features $\phi$ as our classifier, we introduce $\lambda$ a regularization coefficient (set to $0.1$ in our experiments), $I$ the identity matrix and
@@
   \emph{\textbf{Given}} a dataset  $D_E = (s_k,a_k=\pi_E(a_k),s'_k)_k$\;\\
-  \emph{\textbf{Given}} a feature function $\phi: S\times A \rightarrow \mathbb{R}^p$\;\\
   \emph{\textbf{Construct}} the dataset $D_C=\{(s_i=s_k,a_i=\pi^E(s_i))=a_k\}$ \;\\
-  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$ by solving \begin{eqnarray*}
-  &\min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
-  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \\
-  &\text{~s.t.~} \forall i,
-  \theta^T{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^T
-  {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
-\end{eqnarray*}\;\\
+  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$ 
 \emph{\textbf{Construct}} the dataset $\{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}$ with $\hat r_j = q(s_j,a_j) - \gamma q(s'_j=s'_k,\pi_C(s'_j=s'_k))$\;\\
@@
 \emph{\textbf{Construct}} the training set  $D_R = \{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}\cup\{((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}\}$ \;\\
-\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$ by computing:
-\begin{equation*}
-  \omega = (\Phi^T\Phi + \lambda Id)^{-1}\Phi^T\hat R.
-\end{equation*}\;\\
+\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$ 
 \emph{\textbf{Output}} the reward function $\hat R^{C}: (s,a) \mapsto \omega^T \phi(s,a)$ \;
 \end{algorithm}
+\subsection{Mountain car}
+The mountain car is a classical toy problem in RL : an underpowered car is tasked with climding a steep hill.
+In order to do so, it has to first move away from the target and climb the slope on its left, and then is moves right, gaining enough speed to climb the hill on the right on top of which is the target.
+We used standard parameters for this problem, as can for example be found in \cite{sutton1998reinforcement}.
+When training an RL agent, the reward is, for example, $1$ if the car's position is greater than $0.5$ and $0$ anywhere else.
+The expert was a very simple hand crafted policy in which the agent uses the power of the car to go in the direction it already moves ({\it i.e.} go left when the speed is negative, right when it is positive)\\
 
-\subsection{Inverted Pendulum}
-\label{subsec:IP}
-We provide here qualitative results on the inverted pendulum problem.
-Quantitative results obtained by following a procedure very similar to Sec.~\ref{subsec:highway} were of the same nature and allowed drawing the same conclusions as those provided in Sec.~\ref{subsubsec:hresults}.
-We thus omit them here for the sake of avoiding redundancy.
-Instead, we use the inverted pendulum problem for illustrative purposes.
-\subsubsection{Setting}
-The inverted pendulum (also known as the cart pole) is a classical benchmark in the RL community, it is fully described in \cite{lagoudakis2003least}.
-The goal is to maintain a pole in a unstable equilibrium position with the help of a cart moving along one axis.
-The state space consists in the angular position and speed of the pole.
-It is possible to use an RL algorithm to learn how to balance the pendule in an upright position by using the following reward: as long as the angular position of the pole is (in absolute value) less than $\pi \over 2$ the reward is $0$, the reward is $-1$ if the pole falls; this reward function is plotted on Fig.~\ref{toto}.
-Three actions are available to the agent: it can choose to apply a leftward or rightward force on the cart or no force at all.
+The initial position of the car was uniformly randomly picked in $[-1.2;-0.9]$ and its speed uniformly randomly picked in $[-0.07;0]$.
+From this position, the hand-crafted policy was left to play until the car reached the objective (with a position greater than $0.5$) at which point the episode ended.
+Enough episodes were played (and the last one was truncated) so that the dataset $D_E$ contained exactly $n$ samples, with $n$ successively equat to $10$, $30$, $100$ and $300$.
+With these parameters, the expert is always able to reach the top on the first time it tries to climb the hill on the right.
+Therefore a whole part of the state space (when the position is on the hill on the right and the speed is negative) is not visited by the expert.
+This hole in the state space will be dealt with differently by the classifer and the IRL algorithms.
+The classifier will use its generalization power to find a default action in this part of the state space, while the IRL algorithms will devise a default reward ; a (potentially untrained) RL agent finding itself in this part of the state space will use the reward signal to decide what to do, making use of new data available at that time.
 
-The fact that the state space is continuous is not a problem to our particular instanciation of the CSI approach (see Alg.~\ref{alg:CSIimp}) as long as we provide a feature function $\phi: S\times A \rightarrow \mathbb{R}^p$.
-Here $p=30$ as we use the exact same feature functions as proposed in \cite{lagoudakis2003least}.
-The state space is featured with a $3\times 3$ grid of Gaussian functions concatenated with a constant component to construct a feature vector $\psi:S \rightarrow \mathbb{R}^{10}$.
-The eature function $\phi$ is defined as:
-\begin{equation}
-  \phi(s,a) = \begin{pmatrix}
-     \delta_{a = \textsc{left}}\psi(s)\\
-     \delta_{a = \textsc{none}}\psi(s)\\
-     \delta_{a = \textsc{right}}\psi(s)
-  \end{pmatrix}
-\end{equation}
+In order to get a policy from the rewards given by SCIRL, CSI and RE the RL problem was solved by LSPI fed with a dataset $D_{random}$ of $1000$ episodes of length $5$ with a starting point uniformly randomly chosen in the whole state space and actions picked at random.
+This dataset was also used by RE (and not by SCIRL nor CSI).
 
-\subsubsection{Results}
-The results are shown Fig.~\ref{fig:IP}.
-In the left column, we show the reward and value functions of the expert.
-The expert (an RL agent trained on the  reward shown on Fig.~\ref{toto}) is able to balance the pendule indefinitely.
-With as little as 300 transitions (30 seconds of balancing), CSI is able to find a reward that, once optimized by an RL agent, will reliably lead to a control as good as the expert's.
-As Fig.~\ref{tata} illustrates, the data we get from the expert only covers a tiny fraction of the state space.
-Furthermore, the feature function $\phi$ can not permit to accurately approximate the true reward\footnote{The true reward function is a function in $\mathbb{R}^S$, CSI outputs a reward in $\mathbb{R}^{S\times A}$, but we plot here $R(s,\pi^*(s))$ with $\pi^*$ the optimal policy for $R$.}.
-Despite these two setbacks, CSI manages to find a reward for which not only the expert's control is near-optimal (that is what the analysis of Sec.~\ref{sec:analysis} garantees under the assumption that a reasonable amount of data is available in order to keep both classification and regression errors down) but that also leads to a control policy as good as the expert's \emph{with respect to the (theoretically) unknown expert's reward function}.
-This phenomenon appears clearly when comparing figures \ref{fifi} and \ref{riri}: although the reward functions were different, the associated optimal value functions are (up to the scale, which does not actually matters) very similar.
+The objective criterion for success is the number of steps needed to attain the goal when starting from a state picked at random ; the lesser the better.
+We can see Fig.~\ref{fig:MountainCar} that the optimal policies for the rewards found by SCIRL and CSI very rapidly attain expert-level performance and outperform the optimal policy for the reward of RE and the classification policy.
+\begin{figure}
+  \includegraphics[width=\textwidth]{Exp11.pdf}
+  \caption{Performance of various policies on the mountain car problem. This is the mean over 50 runs.}
+  \label{fig:MountainCar}
+\end{figure}
 
-\begin{figure}
-  \begin{minipage}{.4\linewidth}
-   \subfigure[Reward on which the expert was trained]{\label{toto}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_true_R"}}
-   \end{minipage}
-  \hfill
-  \begin{minipage}{.4\linewidth}
-    \subfigure[Reward found by CSI.
-The green zone shows where the data available to the algorithm lies in the state space.]{\label{tata}\hspace{-3em}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_lafem_R"}}
-  \end{minipage}\\
-  \begin{minipage}{.4\linewidth}
-    \subfigure[Value function of the expert (i.e. optimal value function for the reward displayed Fig.~\ref{toto})]{\label{riri}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vexpert"}}
-  \end{minipage}
-  \hfill
-  \begin{minipage}{.4\linewidth}
-  \subfigure[Optimal value function for the reward function found by CSI (Fig.~\ref{tata})]{\label{fifi}\includegraphics[width=1.2\linewidth]{"LAFEM_Exp3_Vagent"}}
-  \end{minipage}
-  \caption{Results on the inverted pendulum benchmark, illsutrating the dissimilarity of the rewards, but the similarity of the value functions.
-The state space is represented on a position-speed plan.}
-  \label{fig:IP}
-\end{figure}
+The classifier for CSI was an off-the-shelf SVM\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}}, the regressor was an off-the-shelf SVR\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVR.html}}.
 
@@
 \label{subsec:highway}
-Let's now move from a qualitative illustration to a quantitative assessment of the performance of the CSI algorithm.
-The Highway driving problem is described Sec.~\ref{subsubsec:hsetting}, the criterion we use is explained Sec.~\ref{subsubsec:hcriterion} and finally results and comparison to other approaches are presented Sec.~\ref{subsubsec:hresults}.
-\subsubsection{Setting}
-\label{subsubsec:hsetting}
 The setting of the experiment is a driving simulator inspired from a benchmark already used in \cite{syed2008apprenticeship,syed2008game}.
@@
 %FIXME:METTRE UNE CAPTURE DECRAN.
-The reward function $R_E$ trains the expert to go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
+The reward function $R_E$ the expert is trained by a DP algorithm on makes it go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
 Any other situation receives a null reward.
@@
 
-We need some baseline to compare our new approach to.
-The natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
-Another natural benchmark we use is to compute the mean value function of the classifier policy.
-We also compare the proposed cascading method to the projection version of the seminal algorithms in \cite{abbeel2004apprenticeship}\footnote{We call it PIRL for Projection IRL.} and the SCIRL algorithm of \cite{klein2012scirl}.
-The random baseline and PIRL do not use samples (with obvious reasons for the random baseline).
-PIRL is instanciated in a way that makes use of the transition probability matrix $P$ to exactly compute all relevant objects (feature expectations, optimal policies) via exact dynamic programming.
-It is left to run for 70 iterations.
-We could have displayed PIRL's data sensitivity by making $\pi_E$ known only through observations, but we decided against it because we want this baseline to show the performance of a fully informed well-established apprenticeship algorithm.
+The performance criterion for a policy $\pi$ is the mean (over the uniform distribution) value function with respect to $R_E$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
 
-Apart from SCIRL, to the best of our knowledge, no other algorithm can infer a reward function only from expert data.
-We will see in Sec.~\ref{sec:related} that most existing approaches either need to solve the direct RL problem repeatedly (which can not be done on this setting from expert data alone) or rely on a sampling of the dynamics of the whole MDP and do not propose (as CSI or SCIRL) the use of heuristics to avoid this problem.
-The reward functions found by SCIRL and CSI are then optimized using a dynamic programming algorithm to show the performance of the associated policy.
+Expert performance averages to $7.74$ ; we also show the natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
+The reward functions found by SCIRL, CSI and RE are then optimized using a DP algorithm.
+The dataset $D_{random}$ needed by RE (and neither by CSI nor SCIRL) is made of $100$ episodes of length $10$ starting randomly in the state space and following a random policy.
+the dataset $D_E$ is made of $n$ episodes of length $n$, with $n \in \{3,7,10,15,20\}$.
 
-\subsubsection{Criterion}
-\label{subsubsec:hcriterion}
-The theoretical bound provided in Sec.~\ref{sec:analysis} is about the optimality of the expert with respect to the reward function computed by our algorithm (which is also the way the problem was framed in Sec.~\ref{sec:background}).
-For this experiment, we use a somewhat more restrictive criterion: a measure of the optimality of an agent trained on the reward output by our algorithm, with respect to the ``true'' reward function, the reward function of the expert.
-We use the mean (over the uniform distribution) value function with respect to $R_E$ of the policy $\hat \pi_C$ learned by optimizing $\hat R^C$: $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
-We do that firstly because, for once, we can: the bound we provide in the analysis is something that can be measured even if $R_E$ does not actually exist.
-We make use here of the fact that $R_E$ not only exists, but is known to us as we use, as an expert, a RL agent perfectly trained on a handcrafted $R_E$.
-The second reason for this different metric is that we believe the theoretical bound to be convincing, and what we wanted to illustrate here is not that the theoretical bound holds, but that under really dire sample scarcity (a case not covered by the analysis) our algorithm is well-behaved given the use of heuristics.
-Finally, it is quite fair to use a single objective criterion than to just compare all approaches with one another.
-
-\subsubsection{Results}
-\label{subsubsec:hresults}
-Results are shown Fig.~\ref{fig:Highway}.
-We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL and CSI, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), the output of the PIRL algorithm, the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
-Performance for CSI is on par with the PIRL and SCIRL algorithms, slightly below the performance of the expert himself.
+Results are shown Fig.~\ref{fig:Highway} and Tab.~\ref{tab:highway}.
+We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL, CSI and RE, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
+Performance for CSI is slightly but definitively higher than for SCIRL (see the variance in Tab.~\ref{tab:highway}), slightly below the performance of the expert itself.
 Very few samples (100) are needed to reliably achieve expert-level performance.
@@
 The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound let us expect.
+\begin{figure}
+      \begin{tabular}{|l|l|l|l|l|}
+        \hline
+        &\bf CSI& \bf SCIRL& \bf RE& \bf Classif\\
+        \hline \multicolumn{5}{|c|}{$n = 9$}\\ \hline
+        Var& 5.139& 5.684& 4.686& 0.045 \\
+        Min& -2.708& -3.175& -2.717& -1.564 \\
+        Mean& 2.419& 2.762& 2.286& -1.149 \\
+        Max& 7.050& 7.060& 7.027& -0.596 \\
+        \hline \multicolumn{5}{|c|}{$n = 49$}\\ \hline
+        Var& 0.214& 0.517& 0.701& 0.121 \\
+        Min& 5.406& 4.652& 4.214& -1.148 \\
+        Mean& 7.012& 6.784& 6.314& -0.158 \\
+        Max& 7.454& 7.529& 7.465& 0.619 \\
+        \hline \multicolumn{5}{|c|}{$n = 100$}\\ \hline
+        Var& 0.011& 0.269& 0.486& 0.154 \\
+        Min& 7.078& 5.724& 5.487& -0.574 \\
+        Mean& 7.364& 7.194& 6.778& 0.456 \\
+        Max& 7.526& 7.562& 7.509& 1.483 \\ \hline
+        \end{tabular}
+      \begin{tabular}{|l|l|l|l|l|}
+        \hline
+        &\bf CSI& \bf SCIRL& \bf RE& \bf Classif\\
+        \hline \multicolumn{5}{|c|}{$n = 225$}\\ \hline
+        Var& 0.006& 0.029& 0.146& 0.167 \\
+        Min& 7.172& 5.882& 5.612& 0.128 \\
+        Mean& 7.459& 7.377& 7.215& 1.151 \\
+        Max& 7.563& 7.572& 7.578& 2.220 \\
+        \hline \multicolumn{5}{|c|}{$n = 400$}\\ \hline
+        Var& 0.002& 0.002& 0.005& 0.267 \\
+        Min& 7.321& 7.247& 7.144& 0.041 \\
+        Mean& 7.495& 7.369& 7.279& 1.744 \\
+        Max& 7.552& 7.557& 7.519& 2.761 \\
+        \hline
+      \end{tabular}
+      \caption{Raw results on the Highway driving problem, see how the variance of SCIRL and CSI decreases.}
+     \label{tab:highway}
+      \end{figure}
+  \begin{figure}
+   \subfigure[Mean performance over 100 runs on the Highway driving problem.]{\label{tutu}\includegraphics[width=.5\linewidth]{"Exp14"}}
+   \subfigure[Zoom of Fig~\ref{tutu} showing the ranking of the three IRL algorithms.]{\includegraphics[width=.5\linewidth]{"Exp14_zoom"}}
+  \caption{Results on the highway driving problem.}
+  \label{fig:Highway}
 
-PIRL is given the transition probabilities (that is, a $|S|\times |S|\times |A|$ matrix) and yet it does only slightly better than our approach with 100 samples.
-This also shows that CSI is indeed very sample efficient, without having to repeatedly solve the MDP.
-Furthermore, it is much faster to run our implementation of CSI on so little data than to run PIRL with the dynamics for 70 iterations.
-
-Finally, only the SCIRL algorithm operates on the same data as our proposed approach, with very slightly better results.
-
-\begin{figure}
-   \subfigure[PIRL and Random Baselines.
-There is no abcissa as these methods do not need data input.
-PIRL uses the transition probabilities matrix $P$.]{\includegraphics[width=.5\linewidth]{"Fig1-NoX"}}
-   \subfigure[Cascading and Classification.]{\label{tutu}\includegraphics[width=.5\linewidth]{"Fig2-ClassifAndUs"}}
-  \subfigure[SCIRL and Cascading mean value.]{\includegraphics[width=.5\linewidth]{"Fig3-SCIRLAndUs"}}
-  \subfigure[Legend]{\includegraphics[width=.5\linewidth]{"Fig4-Legend"}}
-
-  \caption{Data is shown with mean, standard deviation, minimum and maximum value over 50 runs. See Sec.~\ref{subsubsec:hsetting} for detailled information about the experimental setting.}
-  \label{fig:Highway}
 \end{figure}
@@
 Approaches summarized in \cite{neu2009training} can be seen as iteratively contructing a reward function, solving an MDP at each iteration.
-Some of these algorithms are IRL algorithms while others fall in the AL category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship} which we used in our benchmarks (Sec.~\ref{sec:experiments}).
+Some of these algorithms are IRL algorithms while others fall in the Apprenticeship Learning (AL) category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship}.
 In both cases the need to solve an MDP at each step may be very demanding, both sample-wise and computationally.
@@
 Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
-The SCIRL (for Structured Classification for IRL) algorithm \cite{klein2012scirl} which we used Sec.~\ref{subsec:highway} is able to solve the IRL problem (without having to solve any MDP) in the same dire conditions the cascading approach has been tested in.
-It injects the structure of the MDP in a linearly parametrized score-function based classifier by using the expert's feature expectation (the expected discounted sum of the a feature function evaluated over the states of the trajoctories the expert follow) as feature function.
-This necessity of a linarly parametrized approach and the need to compute the expert's feature expectations makes the SCIRL algorithm less convenient than the CSI approach.
+It may be unclear how SCIRL \cite{klein2012scirl} relates to the proposed approach of his paper.
+Both algorithms use the score function of a classifier as a proxy to the action-value function of the expert with respect to the (unknown) true reward : $Q^{\pi_E}_R$.
+The way this proxy is constructed and used, however, fundamentally differs in the two algorithms.
+This difference will cause the theoretical analysis of both approaches (see Sec.~\ref{sec:analysis}) to be distinct.
+In SCIRL, the score function of the classifier is approximated {\it via} a linear parametrization that relies on the feature expectation of the expert $\mu^E(s) = E[\sum_{t\geq 0}\gamma^t\phi(s_t)|s_0=s,\pi_E]$.
+This entails the use of a specific kind of classifier (namely linearly-parametrized-score-function-based classifiers) and of a method of approximation of $\mu^E$.
+By contrast, almost any off-the-shelf classifier can be used in the first step of the cascading approach of this paper.
+The classification step of CSI is unaware of the structure of the MDP whereas SCIRL knows about it thanks to the use of $\mu^E$.
+In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \ref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.\\
+The regressor of CSI can be chosen off-the-shelf.
+One can argue that this and not having to approximate $\mu^E$ increases the ease-of-use of CSI over SCIRL and makes for a more versatile algorithm.
+In practice, as seen in Sec.~\ref{sec:experiments}, perfomance of SCIRL and CSI are very close to one another thus CSI may be a better choice as it is easier to deploy.
+We would like to emphasize the fact that although SCIRL and CSI share similarities, there exist key differences that we just explained.
+Neither approach is a generalization of the other.
 
@@
 The expert is theoretically shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.
-Practical examples of classifiers and regressors have been given, and one combination has been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
-On one of the benchmarks (a car driving simulator), our algorithm is shown to outperform other state of the art approaches, and to be on-par with the very recent SCIRL algorithm.
+Practical examples of classifiers and regressors have been given, and two combinations have been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
+On one of the benchmarks (a highway driving simulator), our algorithm is shown to outperform other state of the art approaches.
 We plan on deepening the analysis of the theoretical properties of our approach and on applying it to real world robotics problems.

### 2013-03-27 piot2012cascading/ECML13_cascading.tex
Version ECML envoyée à M&O

@@
 It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics.
-This is exemplified on two classical benchmarks (a highway driving simulator and an inverted pendulum simulator).
+This is exemplified on two classical benchmarks (the mountain car problem and a highway driving simulator).
   \end{abstract}
@@
 Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-A specific instantiation of CSI (proposed Sec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Sec.~\ref{subsec:IP}) and a car driving simulator (Sec.~\ref{subsec:highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.
+A specific instantiation of CSI (proposed Sec.~\ref{subsubsec:subroutines}) is tested on the mountain car problem (Sec.~\ref{subsec:mountaincar}) and a highway driving simulator (Sec.~\ref{subsec:highway}) where we compare it with a pure classification algorithm \cite{taskar2005learning} and with two recent successfull IRL methods \cite{klein2012scirl} as well as with a random baseline.
 
@@
 The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
-Given that $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$, the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
+Given that $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$ the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
 \begin{equation}
@@
 \end{equation}
-For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} which we consider in the experimental setting.
-Other algorithms may be envisionned (see Sec.~\ref{subsec:instanciation}).
+For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} both of which we consider in the experimental setting.
+Other algorithms may be envisionned (see Sec.~\ref{subsubsec:subroutines}).
 
@@
 As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$.
-A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space.
+A regressor (a simple least-square approximator can do but other solutions could also be envisionned, see Sec.~\ref{subsubsec:subroutines}) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space.
 The complete algorithm is given in Alg.~\ref{algo:cascading}.
@@
 As both algorithms solve the same problem, both theorems naturally bound the same criterion.
-Another shared feature is the use of the score function $q$ of the classifier as a proxy for the action-value function of the expert.
+Another shared feature is the use of the score function $q$ of the classifier as a proxy for the action-value function of the expert $Q^{\pi_E}_{R_E}$.
 The same structure can therefore be found in both analyses as there is only so many ways to study error propagation in IRL algorithms.\\
@@
 \begin{proof}
-First let's introduce some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$.
-We recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
+First let's recall some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$, $R^C\in\mathbb{R}^{S\times A}$ is:
 \begin{equation}
@@
 \end{equation}
-and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function output by the regression step.
+and $\hat{R}^C\in\mathbb{R}^{S\times A}$ is the reward function output by the regression step.
 
 The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$.
-We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof:
+We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof, not to be confused with $R_E$ the unknown reward function the expert optimizes:
 \begin{equation}
@@
 
-Other differences between existing IRL or apprenticeship learning approaches and the proposed cascading algorithm are further examined in the following section.
+Other differences between existing IRL or apprenticeship learning approaches and the proposed cascading algorithm are further examined Sec.~\ref{sec:related}.
 
@@
 We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data, the rationale behing the heuristics we use to compensate for the dire data scarcity and a quick word about the contender CSI is compared to are given Sec.~\ref{subsec:generalities}.
-We supply quantitative results and comparison of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) %FIXME: subsec moutnain car
-and then on a more difficult highway driving simulator (Sec.~\ref{subsec:highway}).
+We supply quantitative results and comparison of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) Sec.~\ref{subsec:mountaincar} and then on a more difficult highway driving simulator (Sec.~\ref{subsec:highway}).
 \subsection{Generalities}
@@
 This feature makes it particularly well-suited to environments where sampling the MDP is difficult or costly.
-In the experiments, CSIis fed only with data sampled according to the expert policy.
+In the experiments, CSI is fed only with data sampled according to the expert policy.
 This corresponds for example to a situation where a costly system can only be controlled by a trained operator as a bad control sequence could lead to a system breakdown.
 
-More precisely, the expert controls the system for $M$ runs of lengths ${L_i}_{1\leq i \leq M}$, %with values of $n$ in $\{3,7,10,15,20\}$,
-giving samples $(s_k,a_k=\pi_E(a_k),s'_k)_k = D_E$. The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
+More precisely, the expert controls the system for $M$ runs of lengths $\{L_i\}_{1\leq i \leq M}$, %with values of $n$ in $\{3,7,10,15,20\}$,
+giving samples $\{(s_k,a_k=\pi_E(a_k),s'_k)_k\} = D_E$. The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
 \begin{equation}
@@
 For the regression to be meaningful, we need samples associating the same state $s_k$ and a different action $a\neq a_k$ with a datapoint $\hat r \neq \hat r_k$.\\
-Recall that $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ (Eq.~\ref{eq:rj}), without knowing $s'_k \sim P(\cdot|s_k,a\neq a_k)$, we can not provide the regressor with a datapoint to associate with $(s_k,a\neq a_k)$.
+Recall that $\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))$ (Eq.~\eqref{eq:rj}), without knowing $s'_k \sim P(\cdot|s_k,a\neq a_k)$, we can not provide the regressor with a datapoint to associate with $(s_k,a\neq a_k)$.
 We will artificially augment the dataset $D_R$ with samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
 \begin{equation}
-  r_{min} = \min_j\hat r_j - 1.
+  r_{min} = \min_k\hat r_k - 1.
 \end{equation}
@@
 }$(s_k,a_k=\pi_E(s_k))$).
-Semantically, we are asserting that disagreeing with the expert in states the expert visited is a bad idea, thus the shape of the reward function should reflect it.
+Semantically, we are asserting that disagreeing with the expert in states the expert visited is a bad idea.
 This heuristics says nothing about states absent from the expert dataset.
@@
 Both experiments use such a criterion.
-Calssification algorithms don't have to optimize any reward, the classification policy can directly be used in the environment.
-IRL algorithms, on the other hand, output a reward that must then be plugged in a RL or DP algorithm to get a policy.
+Classification algorithms don't have to optimize any reward, the classification policy can directly be used in the environment.
+IRL algorithms on the other hand output a reward that must then be plugged in a RL or DP algorithm to get a policy.
 In each benchmark we used the same (benchmark-dependant) algorithm to get a policy from each of the three rewards output by SCIRL, CSI and RE.
 It is these policies whose performance we show.
-Finding a policy from a reward if of course a non-trivial problem that should not be swept under the rug, nevertheless we choose not to concern ourselves with it here as we wish to focus on IRL algorithms, not RL or DP algorithms.
-In this regard, using a classifier that directly outputs a policy may seem a much simpler solution, but we hope that the reader will be convinced that the gap in performance between classification and IRL is worth the trouble.
+Finding a policy from a reward is of course a non-trivial problem that should not be swept under the rug, nevertheless we choose not to concern ourselves with it here as we wish to focus on IRL algorithms, not RL or DP algorithms.
+In this regard, using a classifier that directly outputs a policy may seem a much simpler solution, but we hope that the reader will be convinced that the gap in performance between classification and IRL is worth the trouble of solving the RL problem (once and for all, and not repeatedly as a subroutine like some other IRL algorithms).
 
@@
 In a simple $k$-nearest neighboor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighboors of $s$.
-The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}) (we use a SVM in the mountain car experiment Sec.~\ref{sec:mountaincar}).
-In the highway experiment, we choose to use a structured margin classification approach \cite{taskar2005learning} that, given a feature vector $\phi: S\times A \rightarrow \mathbb{R}^p$ solves:
-\begin{eqnarray*}
-  &\min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
-  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \\
-  &\text{~s.t.~} \forall i,
-  \theta^T{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^T
-  {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
-\end{eqnarray*}
-with $l(s_i,a) = 1\textrm{ if } a\neq a_i \textrm{ and } l(s_i,a_i)=0$.
-The score function of this classifier is simply $q(s,a) = \theta^T \phi(s,a)$.
+The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}) (we use a SVM in the mountain car experiment Sec.~\ref{subsec:mountaincar}).
+In the highway experiment, we choose to use a structured margin classification approach \cite{taskar2005learning}.
+%% that, given a feature vector $\phi: S\times A \rightarrow \mathbb{R}^p$ solves:
+%% \begin{eqnarray*}
+%%   &\min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
+%%   \frac{\eta}{N}\sum_{i=1}^N \zeta_i \\
+%%   &\text{~s.t.~} \forall i,
+%%   \theta^T{\phi}(s_i,a_i)+\zeta_i \geq \max_a \theta^T
+%%   {\phi}(s_i,a) + l(s_i,a). \label{eq:qp_taskar}
+%% \end{eqnarray*}
+%% with $l(s_i,a) = 1\textrm{ if } a\neq a_i \textrm{ and } l(s_i,a_i)=0$. The score function of this classifier is simply $q(s,a) = \theta^T \phi(s,a)$.
 We chose a SVR as a regressor in the mountain car experiment and a simple least-square regressor on the highway.
-Using the same features $\phi$ as our classifier, we introduce $\lambda$ a regularization coefficient (set to $0.1$ in our experiments), $I$ the identity matrix and
-\begin{eqnarray}
-  \Phi = \begin{bmatrix}
-    \vdots\\
-    \phi(s_j,a_j)\\
-    \vdots
-  \end{bmatrix} \in \mathbb{R}^{n|A|\times p}\textrm{ and }
-  \hat R = \begin{bmatrix}
-    \vdots\\
-    \hat r_j\\
-    \vdots
-  \end{bmatrix}\in \mathbb{R}^{n|A|},
-\end{eqnarray}
-with $r_j = r_{min}$ if $a_j \neq \pi_E(s_j)$. Our regressor can be written:
-\begin{equation}
-  \omega = (\Phi^T\Phi + \lambda I)^{-1}\Phi^T\hat R.
-\end{equation}
-The reward function is: $\hat R^C(s,a) = \omega^T \phi(s,a)$
+%% Using the same features $\phi$ as our classifier, we introduce $\lambda$ a regularization coefficient (set to $0.1$ in our experiments), $I$ the identity matrix and 
+%% \begin{eqnarray}
+%%   \Phi = \begin{bmatrix}
+%%     \vdots\\
+%%     \phi(s_j,a_j)\\
+%%     \vdots
+%%   \end{bmatrix} \in \mathbb{R}^{n|A|\times p}\textrm{ and }
+%%   \hat R = \begin{bmatrix}
+%%     \vdots\\
+%%     \hat r_j\\
+%%     \vdots
+%%   \end{bmatrix}\in \mathbb{R}^{n|A|},
+%% \end{eqnarray}
+%% with $r_j = r_{min}$ if $a_j \neq \pi_E(s_j)$. Our regressor can be written:
+%% \begin{equation}
+%%   \omega = (\Phi^T\Phi + \lambda I)^{-1}\Phi^T\hat R.
+%% \end{equation}
+%% The reward function is: $\hat R^C(s,a) = \omega^T \phi(s,a)$
 
@@
   \emph{\textbf{Construct}} the dataset $D_C=\{(s_i=s_k,a_i=\pi^E(s_i))=a_k\}$ \;\\
-  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$ 
+  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$ \;\\
 \emph{\textbf{Construct}} the dataset $\{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}$ with $\hat r_j = q(s_j,a_j) - \gamma q(s'_j=s'_k,\pi_C(s'_j=s'_k))$\;\\
@@
 \emph{\textbf{Construct}} the training set  $D_R = \{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}\cup\{((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}\}$ \;\\
-\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$ 
+\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$\;\\
 \emph{\textbf{Output}} the reward function $\hat R^{C}: (s,a) \mapsto \omega^T \phi(s,a)$ \;
@@
 \subsection{Mountain car}
-The mountain car is a classical toy problem in RL : an underpowered car is tasked with climding a steep hill.
-In order to do so, it has to first move away from the target and climb the slope on its left, and then is moves right, gaining enough speed to climb the hill on the right on top of which is the target.
+\label{subsec:mountaincar}
+The mountain car is a classical toy problem in RL : an underpowered car is tasked with climbing a steep hill.
+In order to do so, it has to first move away from the target and climb the slope on its left, and then it moves right, gaining enough momentum to climb the hill on the right on top of which lies the target.
 We used standard parameters for this problem, as can for example be found in \cite{sutton1998reinforcement}.
@@
 The initial position of the car was uniformly randomly picked in $[-1.2;-0.9]$ and its speed uniformly randomly picked in $[-0.07;0]$.
-From this position, the hand-crafted policy was left to play until the car reached the objective (with a position greater than $0.5$) at which point the episode ended.
-Enough episodes were played (and the last one was truncated) so that the dataset $D_E$ contained exactly $n$ samples, with $n$ successively equat to $10$, $30$, $100$ and $300$.
+From this position, the hand-crafted policy was left to play until the car reached the objective ({\it i.e.} a position greater than $0.5$) at which point the episode ended.
+Enough episodes were played (and the last one was truncated) so that the dataset $D_E$ contained exactly $n$ samples, with $n$ successively equal to $10$, $30$, $100$ and $300$.
 With these parameters, the expert is always able to reach the top on the first time it tries to climb the hill on the right.
 Therefore a whole part of the state space (when the position is on the hill on the right and the speed is negative) is not visited by the expert.
-This hole in the state space will be dealt with differently by the classifer and the IRL algorithms.
+This hole about the state space in the data will be dealt with differently by the classifier and the IRL algorithms.
 The classifier will use its generalization power to find a default action in this part of the state space, while the IRL algorithms will devise a default reward ; a (potentially untrained) RL agent finding itself in this part of the state space will use the reward signal to decide what to do, making use of new data available at that time.
 
-In order to get a policy from the rewards given by SCIRL, CSI and RE the RL problem was solved by LSPI fed with a dataset $D_{random}$ of $1000$ episodes of length $5$ with a starting point uniformly randomly chosen in the whole state space and actions picked at random.
+In order to get a policy from the rewards given by SCIRL, CSI and RE, the RL problem was solved by LSPI fed with a dataset $D_{random}$ of $1000$ episodes of length $5$ with a starting point uniformly randomly chosen in the whole state space and actions picked at random.
 This dataset was also used by RE (and not by SCIRL nor CSI).
@@
 
-The feature function $\phi$ for this setting is the natural tabular feature function where each state action couple $(s,a)$ is associated with a unique index $i_{(s,a)}$.
+%% The feature function $\phi$ for this setting is the natural tabular feature function where each state action couple $(s,a)$ is associated with a unique index $i_{(s,a)}$.
 The only non-zero component of $\phi(s,a)$ is then the component at index $i_{(s,a)}$.
@@
 
-Expert performance averages to $7.74$ ; we also show the natural random baseline consists in drawing a random reward vector (with a uniform law) and training an agent on it.
+Expert performance averages to $7.74$ ; we also show the natural random baseline that consists in drawing a random reward vector (with a uniform law) and training an agent on it.
 The reward functions found by SCIRL, CSI and RE are then optimized using a DP algorithm.
@@
 Results are shown Fig.~\ref{fig:Highway} and Tab.~\ref{tab:highway}.
-We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL, CSI and RE, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), the expert's policy $\pi_E$ and the optimal policy for a randomly drawn reward.
+We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL, CSI and RE, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), and the optimal policy for a randomly drawn reward.
 Performance for CSI is slightly but definitively higher than for SCIRL (see the variance in Tab.~\ref{tab:highway}), slightly below the performance of the expert itself.
@@
 Furthermore, the classifier whose results are diplayed here is the output of the first step of the algorithm.
-The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound let us expect.
+The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound lets us expect.
 \begin{figure}
@@
 The classification step of CSI is unaware of the structure of the MDP whereas SCIRL knows about it thanks to the use of $\mu^E$.
-In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \ref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.\\
+In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \eqref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.\\
 The regressor of CSI can be chosen off-the-shelf.

### 2013-03-28 piot2012cascading/ECML13_cascading.tex
First modifs before JFPDA13

@@
 
-IRL is one of many ways to perform Apprenticeship Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function.
+IRL is one of many ways to perform Apprenticeship Learning (AL): imitating a demonstrated expert policy, without necessarily explicitely looking for the reward function.
 The reward function nevertheless is of interest in its own right.
@@
 
-This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-expert policies as inputs.
+This algorithm does not need to iteratively solve an MDP and requires only sampled transitions from expert and non-expert policies as inputs.
 Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-A specific instantiation of CSI (proposed Sec.~\ref{subsubsec:subroutines}) is tested on the mountain car problem (Sec.~\ref{subsec:mountaincar}) and a highway driving simulator (Sec.~\ref{subsec:highway}) where we compare it with a pure classification algorithm \cite{taskar2005learning} and with two recent successfull IRL methods \cite{klein2012scirl} as well as with a random baseline.
+A specific instantiation of CSI (proposed Sec.~\ref{subsubsec:subroutines}) is tested on the mountain car problem (Sec.~\ref{subsec:mountaincar}) and on a highway driving simulator (Sec.~\ref{subsec:highway}) where we compare it with a pure classification algorithm \cite{taskar2005learning} and with two recent successfull IRL methods \cite{klein2012scirl} as well as with a random baseline.
 
@@
 The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
-Given that $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$ the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
+Given that $P_\pi = (P(s'|s,\pi(s)))_{s,s' \in S}$ is the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
 \begin{equation}
@@
 \begin{equation}
-V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s].
+V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s,\pi].
 \end{equation}
@@
   \end{equation}
-where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally:
+where $R_\pi = (R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally:
 \begin{equation}
@@
 The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal).
-Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although a lot of progress have been made, see Sec.~\ref{sec:related}.
+Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although a lot of progress has been made, see Sec.~\ref{sec:related}.
 \section{The cascading algorithm}

### 2013-03-29 piot2012cascading/ECML13_cascading.tex
Suites modifs pour JFPDA/ECML

@@
 Given a dataset $D_C = \{(s_i,a_i=\pi_E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier.
-The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL).
+The classification policy $\pi_C$ is not the end product we are looking for (that would be mere supervised imitation of the expert, not IRL).
 What is of particular interest to us is the score function $q$ itself.
@@
 The score function $q$ of the classifier can thus be viewed as some kind of optimal action-value function for the classifier policy $\pi_C$.
-By inversing the Bellman equation Eq.~\eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
+By inversing the Bellman equation \eqref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$:
 \begin{equation}
@@
 \end{equation}
-As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system.
+As we wish to approximately solve the general IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system.
 We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to.
@@
 \end{equation}
-As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$.
+As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s_j,a_j)$.
 A regressor (a simple least-square approximator can do but other solutions could also be envisionned, see Sec.~\ref{subsubsec:subroutines}) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space.
@@
 There is no particular constraint on $D_C$ and $D_R$.
-Evidently, there is a direct link between various qualities of those two sets (amount of data, statistical representativity, etc.) and the classification and regression error.
+Clearly, there is a direct link between various qualities of those two sets (amount of data, statistical representativity, etc.) and the classification and regression errors.
 The exact nature of the relationship between these quantities depends on which classifier and regressor are chosen.
-The theoretical analysis of Sec.\ref{sec:analysis} abstracts itself from the choice of a regressor and a classifier and from the composition of $D_C$ and $D_R$ by reasoning with the classification and regression error.
-In Sec.\ref{sec:experiments}, %FIXME subsec heuristic
+The theoretical analysis of Sec.~\ref{sec:analysis} abstracts itself from the choice of a regressor and a classifier and from the composition of $D_C$ and $D_R$ by reasoning with the classification and regression errors.
+In Sec.~\ref{sec:experiments}, %FIXME subsec heuristic
 the use of a single dataset to create both $D_C$ and $D_R$ is thoroughly explored.
@@
 
-Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem.
+Cascading two supervised approaches like we do is a way to inject the MDP structure into the resolution of the problem.
 Indeed, mere classification only takes into account information from the expert (i.e., which action goes with which state) whereas using the Bellman equation in the expression of $\hat r_j$ makes use of the information lying in the transitions $(s_j,a_j,s'_j)$, namely information about the transition probabilities $P$.
@@
   \item $\epsilon^R_\pi=R^C_\pi-\hat{R}^C_\pi$ ;
-  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$, one property of which is that $\forall X, \mu^TX \leq \|X\|_{1,\mu}$;
+  \item $\|.\|_{1,\mu}$ the $\mu$-weighted $L_1$ norm: $\|f\|_{1,\mu} = \E_{x\sim \mu}[|f(x)|]$;
   \end{itemize}
@@
   The constant $C_*$ can be estimated \emph{a posteriori} (after $\hat{R}^C$ is computed).
-{\it A priori}, $C_*$ can be upper bound by a more usual and general concentration coefficient but $C_*$ gives a tighter final result: one can informally see $C_*$ as a measure of the similarity between the distributions indeuced by $\hat\pi_C$ and $\pi_E$ (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
+{\it A priori}, $C_*$ can be upper-bounded by a more usual and general concentration coefficient but $C_*$ gives a tighter final result: one can informally see $C_*$ as a measure of the similarity between the distributions induced by $\hat\pi_C$ and $\pi_E$ (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
 \item $\Delta q$, defined by 
@@
   \end{itemize}
-Readers familiar with the work presented in \cite{klein2012scirl} will see some similarities between the theoretical analyses of SCIRL and CSI.
-As both algorithms solve the same problem, both theorems naturally bound the same criterion.
-Another shared feature is the use of the score function $q$ of the classifier as a proxy for the action-value function of the expert $Q^{\pi_E}_{R_E}$.
-The same structure can therefore be found in both analyses as there is only so many ways to study error propagation in IRL algorithms.\\
-The attentive reader, however, will percieve that similarities stop there.
-The error terms occurring in the two bounds are not related to one another.
-As CSI makes no use of the feature expectation of the expert, what is known as $\bar\epsilon_Q$ in \cite{klein2012scirl} does not appear in this analysis.
-Likewise, the regression error $\epsilon_R$ of this paper does not appear in the analysis of SCIRL, which does not use a regressor.
-Perhaps more subtly, the classification error and classification policy known in both papers as $\epsilon_C$ and $\pi_C$ are not the same.
-The classification policy of SCIRL is no tantamount to what is called $\pi_C$ here.
-For SCIRL, $\pi_C$ is the greedy policy for an approximation of the value function of the expert with respect to the reward output by the algorithm.
-For CSI, $\pi_C$ is the decision rule of the classifier, an object that is not aware of the structure of the MDP.
 \begin{theorem}
@@
 \end{equation}
-If $\pi=\pi_E$, we have, thanks to the power series expression of $(I-\gamma P_{\pi_E})^{-1}$ and the definition of $\rho_E$:
+If $\pi=\pi_E$, we have, thanks to the power series expression of $(I-\gamma P_{\pi_E})^{-1}$, the definition of $\rho_E$ and the definition of the $\mu$-weighted $L_1$ norm, one property of which is that $\forall X, \mu^TX \leq \|X\|_{1,\mu}$:
 \begin{equation}
@@
 \end{proof}
-A direct corollary of this theorem is that, given a perfect classifier and regressor, CSI produces a reward function for which $\pi_E$ is the \emph{unique} optimal policy.
+Readers familiar with the work presented in \cite{klein2012scirl} will see some similarities between the theoretical analyses of SCIRL and CSI as both study error propagation in IRL algorithms.
+Another shared feature is the use of the score function $q$ of the classifier as a proxy for the action-value function of the expert $Q^{\pi_E}_{R_E}$.
+\\
+The attentive reader, however, will percieve that similarities stop there.
+The error terms occurring in the two bounds are not related to one another.
+As CSI makes no use of the feature expectation of the expert, what is known as $\bar\epsilon_Q$ in \cite{klein2012scirl} does not appear in this analysis.
+Likewise, the regression error $\epsilon_R$ of this paper does not appear in the analysis of SCIRL, which does not use a regressor.
+Perhaps more subtly, the classification error and classification policy known in both papers as $\epsilon_C$ and $\pi_C$ are not the same.
+The classification policy of SCIRL is no tantamount to what is called $\pi_C$ here.
+For SCIRL, $\pi_C$ is the greedy policy for an approximation of the value function of the expert with respect to the reward output by the algorithm.
+For CSI, $\pi_C$ is the decision rule of the classifier, an object that is not aware of the structure of the MDP.
+We shall also mention that the error terms appearing in the CSI bound are more standard than the ones of SCIRL ({\it e.g.}, regression error vs feature expectation estimation error) thus they may be easier to control.
+A direct corollary of this theorem is that, given perfect classifier and regressor, CSI produces a reward function for which $\pi_E$ is the \emph{unique} optimal policy.
 \begin{corollary}
   \label{cor}
-  Assume that $\rho_E>0$ and that the classifier and the regressor are perfect (that is $\epsilon_C = 0$ and $\epsilon_R = 0$). Then $\pi_E$ is the \emph{unique} optimal policy for $\hat R^C$.
+  Assume that $\rho_E>0$ and that the classifier and the regressor are perfect ($\epsilon_C = 0$ and $\epsilon_R = 0$). Then $\pi_E$ is the \emph{unique} optimal policy for $\hat R^C$.
 \end{corollary}
@@
 
-The only constraints the bound of theorem~\ref{thm} implies on datasets $D_R$ and $D_C$ is that they provide enough information to the supervised algorithms to keep both error terms $\epsilon_C$ and $\epsilon_R$ low.
+The only constraints the bound of Th.~\ref{thm} implies on datasets $D_R$ and $D_C$ is that they provide enough information to the supervised algorithms to keep both error terms $\epsilon_C$ and $\epsilon_R$ low.
 In Sec.~\ref{sec:experiments} we deal with a lack of data in dataset $D_R$.
@@
 
-More generally, the error terms  $\epsilon_C$ and $\epsilon_R$ can be reduced by a wise choice for the classification and regression subroutines.
+More generally, the error terms  $\epsilon_C$ and $\epsilon_R$ can be reduced by a wise choice for the classification and regression algorithms.
 The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off, etc.) to be found.
@@
 \label{sec:experiments}
-In this section, we put our approach under stress to empirically demonstrate its behavior.
-We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data, the rationale behing the heuristics we use to compensate for the dire data scarcity and a quick word about the contender CSI is compared to are given Sec.~\ref{subsec:generalities}.
-We supply quantitative results and comparison of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) Sec.~\ref{subsec:mountaincar} and then on a more difficult highway driving simulator (Sec.~\ref{subsec:highway}).
+In this section, we empirically demonstrate the behavior of our approach.
+We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data, the rationale behind the heuristics we use to compensate for the dire data scarcity and a quick word about the contenders CSI is compared to are given Sec.~\ref{subsec:generalities}.
+We supply quantitative results and comparisons of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) Sec.~\ref{subsec:mountaincar} and then on a more difficult highway driving simulator (Sec.~\ref{subsec:highway}).
 \subsection{Generalities}
@@
 
-There is no particuliar problem for the classification step.
+There is no particular problem for the classification step.
 It is however not reasonable to construct the dataset $D_R = \{((s_k,a_k=\pi_E(s_k)),\hat r_k)_k\}$ only from expert transitions and expect a small regression error term $\epsilon_R$.
@@
 For the regression to be meaningful, we need samples associating the same state $s_k$ and a different action $a\neq a_k$ with a datapoint $\hat r \neq \hat r_k$.\\
-Recall that $\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))$ (Eq.~\eqref{eq:rj}), without knowing $s'_k \sim P(\cdot|s_k,a\neq a_k)$, we can not provide the regressor with a datapoint to associate with $(s_k,a\neq a_k)$.
-We will artificially augment the dataset $D_R$ with samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
+Recall that $\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))$ (Eq.~\eqref{eq:rj}); without knowing $s'_k \sim P(\cdot|s_k,a\neq a_k)$, we cannot provide the regressor with a datapoint to associate with $(s_k,a\neq a_k)$.
+We artificially augment the dataset $D_R$ with samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
 \begin{equation}
@@
 \end{equation}
-This heuristics intructs the regressor to associate a state-action tuple disagreeing with the expert ({\it i.e.
-}$(s_k,a\neq a_k)$) with a reward strictly inferior to any of those associated with expert state action tuples ({\it i.e.
-}$(s_k,a_k=\pi_E(s_k))$).
+This heuristics intructs the regressor to associate a state-action tuple disagreeing with the expert ({\it i.e.}, $(s_k,a\neq a_k)$) with a reward strictly inferior to any of those associated with expert state action tuples ({\it i.e.}, $(s_k,a_k=\pi_E(s_k))$).
 Semantically, we are asserting that disagreeing with the expert in states the expert visited is a bad idea.
@@
 Although this heuristics was not analyzed in Sec.~\ref{sec:analysis} (where the availability of a more complete dataset $D_R$ was assumed), the results shown in the next two subsections demonstrate its soundness.
-\subsubsection{Comparison with state of the art approaches}
-The similar looking yet fundamentally different algorithm SCIRL is an obvious choice as a contender to CSI as it advertises the same ability to work with very little data, without repeatedly solving the RL problem.
+\subsubsection{Comparison with state-of-the-art approaches}
+The similar looking yet fundamentally different algorithm SCIRL \cite{klein2012scirl} is an obvious choice as a contender to CSI as it advertises the same ability to work with very little data, without repeatedly solving the RL problem.
 In both experiments we give the exact same data to CSI and SCIRL.
 
-The algorithm of \cite{dvijotham2010inverse} also advertises not having to solve the RL problem, but needs to deal with linearly solvable MDPs, therefore we could not include it in our tests.
+The algorithm of \cite{dvijotham2010inverse} also advertises not having to solve the RL problem, but needs to deal with linearly solvable MDPs, therefore we do not include it in our tests.
 The Relative Entropy (RE) method of \cite{boularias2011relative} has no such need, so we included it in our benchmarks.
@@
 Comparing classification and IRL algorithms makes no sense if the object of interest is the reward itself as can be envisionned in a biological or economical context.
-It is however perfectly sound to do so in an imitation context where what matters is the performance of the agent with respect to some objective criterion.
+It is however sound to do so in an imitation context where what matters is the performance of the agent with respect to some objective criterion.
 Both experiments use such a criterion.
 Classification algorithms don't have to optimize any reward, the classification policy can directly be used in the environment.
-IRL algorithms on the other hand output a reward that must then be plugged in a RL or DP algorithm to get a policy.
+IRL algorithms on the other hand output a reward that must then be plugged in an RL or DP algorithm to get a policy.
 In each benchmark we used the same (benchmark-dependant) algorithm to get a policy from each of the three rewards output by SCIRL, CSI and RE.
@@
 
-We do not compare CSI to other IRL algorithms requiring repeatedly solving the MDP : as we would need to provide them with enough data to do so, the comparison makes little sense.
-\subsubsection{Subroutines choices}
+We do not compare CSI to other IRL algorithms requiring repeatedly solving the MDP: as we would need to provide them with enough data to do so, the comparison makes little sense.
+\subsubsection{Supervised Steps}
 \label{subsubsec:subroutines}
@@
 \label{subsec:mountaincar}
-The mountain car is a classical toy problem in RL : an underpowered car is tasked with climbing a steep hill.
+The mountain car is a classical toy problem in RL: an underpowered car is tasked with climbing a steep hill.
 In order to do so, it has to first move away from the target and climb the slope on its left, and then it moves right, gaining enough momentum to climb the hill on the right on top of which lies the target.
@@
 This hole about the state space in the data will be dealt with differently by the classifier and the IRL algorithms.
-The classifier will use its generalization power to find a default action in this part of the state space, while the IRL algorithms will devise a default reward ; a (potentially untrained) RL agent finding itself in this part of the state space will use the reward signal to decide what to do, making use of new data available at that time.
+The classifier will use its generalization power to find a default action in this part of the state space, while the IRL algorithms will devise a default reward; a (potentially untrained) RL agent finding itself in this part of the state space will use the reward signal to decide what to do, making use of new data available at that time.
 
-In order to get a policy from the rewards given by SCIRL, CSI and RE, the RL problem was solved by LSPI fed with a dataset $D_{random}$ of $1000$ episodes of length $5$ with a starting point uniformly randomly chosen in the whole state space and actions picked at random.
+In order to get a policy from the rewards given by SCIRL, CSI and RE, the RL problem was solved by LSPI fed with a dataset $D_{random}$ of $1000$ episodes of length $5$ with a starting point uniformly and randomly chosen in the whole state space and actions picked at random.
 This dataset was also used by RE (and not by SCIRL nor CSI).
@@
 At the highest speed setting, if the other car appears in the lane the player is in, it is not possible to avoid the collision.
-The main difference between the original benchmark \cite{syed2008apprenticeship,syed2008game} and ours is that we made the MDP more ergodic by allowing the player to change speed whenever he wishes so, not just during the first transition.
+The main difference between the original benchmark \cite{syed2008apprenticeship,syed2008game} and ours is that we made the problem more ergodic by allowing the player to change speed whenever he wishes so, not just during the first transition.
 If anything, by adding two actions, we enlarged the state-action space and thus made the problem tougher.
@@
 
-AL via classification has been proposed in for example \cite{ratliff2007imitation} with the help of a structured margin method.
+AL via classification has been proposed for example in \cite{ratliff2007imitation}, with the help of a structured margin method.
 Using the non trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
@@
 It may be unclear how SCIRL \cite{klein2012scirl} relates to the proposed approach of his paper.
-Both algorithms use the score function of a classifier as a proxy to the action-value function of the expert with respect to the (unknown) true reward : $Q^{\pi_E}_R$.
+Both algorithms use the score function of a classifier as a proxy to the action-value function of the expert with respect to the (unknown) true reward: $Q^{\pi_E}_R$.
 The way this proxy is constructed and used, however, fundamentally differs in the two algorithms.
@@
 The classification step of CSI is unaware of the structure of the MDP whereas SCIRL knows about it thanks to the use of $\mu^E$.
-In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \eqref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.\\
+In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \eqref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.
 The regressor of CSI can be chosen off-the-shelf.
@@
 Practical examples of classifiers and regressors have been given, and two combinations have been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
-On one of the benchmarks (a highway driving simulator), our algorithm is shown to outperform other state of the art approaches.
+On one of the benchmarks (a highway driving simulator), our algorithm is shown to outperform other state-of-the-art approaches.
 We plan on deepening the analysis of the theoretical properties of our approach and on applying it to real world robotics problems.

### 2013-03-29 Exp11.py
p-values of mean equality test between CSI and SCIRL on the Highway and the mountain car

@@ -59,6 +59,19 @@ savefig("Exp11.pdf")
 
 # <codecell>
 
+mean(CSI_data[0])
+
+# <codecell>
+
+from scipy.stats import ks_2samp
+for i in range(0,len(X)):
+    y_csi = CSI_data[i]
+    y_scirl = SCIRL_data[i]
+    pvalue = ks_2samp(y_csi,y_scirl)[1]
+    print "Pour X = "+str(X[i])+(", \tKS mean equality test p-value : %.2e "%pvalue)+"\t\t"+("REJECTED" if pvalue < 0.01 else "")
+
+# <codecell>
+
 #plotting the mountain car data
 def filled_mean_min_max(X, Y_mean, Y_min, Y_max, color, _alpha, style, lblmain,lblminmax ):
     "Plot data, with bold mean line, and a light color fill betwee the min and max"

### 2013-03-29 Plot15.py

@@ -38,16 +38,13 @@ grid()
 axis([0,410,-2.2,8])
 savefig("Exp14.pdf")
 
-name = ["Var", "Min","Mean","Max"]
-for i in range(0,len(Y_CSI)):
-    print "\hline \multicolumn{4}{|c|}{$n = "+str(Abcissas[i])+"$}\\\\ \hline"
-    j=0
-    for func in [var, min, mean, max]:
-        print name[j]+"&",
-        j+=1
-        for dataset in [Y_CSI,Y_SCIRL,Y_RE, Y_Classif]:
-            print "%.3f" %func(dataset[i])+"&",
-        print "\\\\"
+# <codecell>
+
+from scipy.stats import ks_2samp
+for i in range(0,len(Abcissas)):
+    y_csi = Y_CSI[i]
+    y_scirl = Y_SCIRL[i]
+    print "Pour X = "+str(Abcissas[i])+", \tKS mean equality test p-value : %.2e "%ks_2samp(y_csi,y_scirl)[1]
 
 # <codecell>
 

### 2013-03-29 piot2012cascading/ECML13_cascading.tex
Fin corrections Matthieu.

@@
 \usepackage{subfigure}
+% scientific notation, 1\e{9} will print as 1x10^9
+\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
+\usepackage{booktabs} % Fancy tables
 \tolerance=1000
@@
 IRL allows the use of (A)DP or RL techniques to learn the task from the computed reward function.
-Although some approaches use it as a byproduct, most AL algorithm makes no use of the reward function, or have nothing to do with IRL altogether.
 A very straightforward non-IRL way to do AL is for example to use a multi-class classifier to directly learn the expert policy.
@@
 
+\section{Related Work}
+\label{sec:related}
+IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
+Approaches summarized in \cite{neu2009training} can be seen as iteratively contructing a reward function, solving an MDP at each iteration.
+Some of these algorithms are IRL algorithms while others fall in the Apprenticeship Learning (AL) category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship}.
+In both cases the need to solve an MDP at each step may be very demanding, both sample-wise and computationally.
+CSI being able to output a reward function without having to solve the MDP is thus a significant improvement.
+
+AL via classification has been proposed for example in \cite{ratliff2007imitation}, with the help of a structured margin method.
+Using the non trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
+in a classification algorithm, showing improvements compared to a
+non-structured kernel.
+
+Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
+It may be unclear how SCIRL \cite{klein2012scirl} relates to the proposed approach of his paper.
+Both algorithms use the score function of a classifier as a proxy to the action-value function of the expert with respect to the (unknown) true reward: $Q^{\pi_E}_R$.
+The way this proxy is constructed and used, however, fundamentally differs in the two algorithms.
+This difference will cause the theoretical analysis of both approaches (see Sec.~\ref{sec:analysis}) to be distinct.
+In SCIRL, the score function of the classifier is approximated {\it via} a linear parametrization that relies on the feature expectation of the expert $\mu^E(s) = E[\sum_{t\geq 0}\gamma^t\phi(s_t)|s_0=s,\pi_E]$.
+This entails the use of a specific kind of classifier (namely linearly-parametrized-score-function-based classifiers) and of a method of approximation of $\mu^E$.
+By contrast, almost any off-the-shelf classifier can be used in the first step of the cascading approach of this paper.
+The classification step of CSI is unaware of the structure of the MDP whereas SCIRL knows about it thanks to the use of $\mu^E$.
+In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \eqref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.
+The regressor of CSI can be chosen off-the-shelf.
+One can argue that this and not having to approximate $\mu^E$ increases the ease-of-use of CSI over SCIRL and makes for a more versatile algorithm.
+In practice, as seen in Sec.~\ref{sec:experiments}, perfomance of SCIRL and CSI are very close to one another thus CSI may be a better choice as it is easier to deploy.
+We would like to emphasize the fact that although SCIRL and CSI share similarities, there exist key differences that we just explained.
+Neither approach is a generalization of the other.
+
+Few IRL or AL algorithms do not require solving an MDP.
+The approach of \cite{syed2008apprenticeship} requires knowing the transition probabilities of the MDP (which CSI does not need) and outputs a policy (and not a reward).
+The algorithm in \cite{dvijotham2010inverse} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
+Closer to our use-case is the idea presented in \cite{boularias2011relative} to use a subgradient ascent of a utility function based on the notion of relative entropy.
+Importance sampling is suggested as a way to avoid solving the MDP.
+This requires sampling trajectories according to a non-expert policy and
+the direct problem remains at the core of the approach (even if
+solving it is avoided).
+
 %% Feel good comment
@@
 We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data, the rationale behind the heuristics we use to compensate for the dire data scarcity and a quick word about the contenders CSI is compared to are given Sec.~\ref{subsec:generalities}.
-We supply quantitative results and comparisons of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) Sec.~\ref{subsec:mountaincar} and then on a more difficult highway driving simulator (Sec.~\ref{subsec:highway}).
+We supply quantitative results and comparisons of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) Sec.~\ref{subsec:mountaincar} and then on a highway driving simulator (Sec.~\ref{subsec:highway}).
 \subsection{Generalities}
@@
 When training an RL agent, the reward is, for example, $1$ if the car's position is greater than $0.5$ and $0$ anywhere else.
-The expert was a very simple hand crafted policy in which the agent uses the power of the car to go in the direction it already moves ({\it i.e.} go left when the speed is negative, right when it is positive)\\
+The expert was a very simple hand crafted policy in which the agent uses the power of the car to go in the direction it already moves ({\it i.e.}, go left when the speed is negative, right when it is positive).\\
 
 The initial position of the car was uniformly randomly picked in $[-1.2;-0.9]$ and its speed uniformly randomly picked in $[-0.07;0]$.
-From this position, the hand-crafted policy was left to play until the car reached the objective ({\it i.e.} a position greater than $0.5$) at which point the episode ended.
+From this position, the hand-crafted policy was left to play until the car reached the objective ({\it i.e.}, a position greater than $0.5$) at which point the episode ended.
 Enough episodes were played (and the last one was truncated) so that the dataset $D_E$ contained exactly $n$ samples, with $n$ successively equal to $10$, $30$, $100$ and $300$.
 With these parameters, the expert is always able to reach the top on the first time it tries to climb the hill on the right.
-Therefore a whole part of the state space (when the position is on the hill on the right and the speed is negative) is not visited by the expert.
+Therefore, a whole part of the state space (when the position is on the hill on the right and the speed is negative) is not visited by the expert.
 This hole about the state space in the data will be dealt with differently by the classifier and the IRL algorithms.
@@
 
+The classifier for CSI was an off-the-shelf SVM\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}} which also was the classifier we evaluate, the regressor of CSI was an off-the-shelf SVR\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVR.html}}.
+RE and SCIRL need features over the state space, we used the same evenly-spaced hand-tuned $7\times 7$ RBF network for both algorithms.
+
 The objective criterion for success is the number of steps needed to attain the goal when starting from a state picked at random ; the lesser the better.
 We can see Fig.~\ref{fig:MountainCar} that the optimal policies for the rewards found by SCIRL and CSI very rapidly attain expert-level performance and outperform the optimal policy for the reward of RE and the classification policy.
+When very few samples are available, CSI does better than SCIRL (with such a low $p$-value for $n=10$, see Tab.~\ref{tab:mc}, the hypothesis that the mean performance is equal can be rejected); SCIRL catches up when more samples are available.
+Furthermore, CSI required very little engineering as we cascaded two off-the-shelf implementations whereas SCIRL used hand-tuned features and a custom classifier.
+\begin{table}
+	\begin{center}
+		\begin{tabular}{cc}
+		\toprule
+			Number of expert samples & $p$-value\\
+			\midrule
+			$10$ & $4.5195\e{-14}$\\
+			$30$ & $0.0131123$\\
+			$100$ & $0.139975$\\
+			$300$ & $0.893752$\\
+		\bottomrule
+		\end{tabular}
+	\end{center}
+        \caption{Kolmogorov–Smirnov mean equality test $p$-values for CSI and SCIRL on the mountain car. High values (>0.01) means that the hypothesis that the means are equal cannot be rejected.}
+        \label{tab:mc}
+\end{table}
 \begin{figure}
   \includegraphics[width=\textwidth]{Exp11.pdf}
-  \caption{Performance of various policies on the mountain car problem. This is the mean over 50 runs.}
+  \caption{Performance of various policies on the mountain car problem. This is the mean over 100 runs.}
   \label{fig:MountainCar}
 \end{figure}
-
-The classifier for CSI was an off-the-shelf SVM\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}}, the regressor was an off-the-shelf SVR\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVR.html}}.
 
@@
 
-Results are shown Fig.~\ref{fig:Highway} and Tab.~\ref{tab:highway}.
+Results are shown Fig.~\ref{fig:Highway}.
 We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL, CSI and RE, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), and the optimal policy for a randomly drawn reward.
-Performance for CSI is slightly but definitively higher than for SCIRL (see the variance in Tab.~\ref{tab:highway}), slightly below the performance of the expert itself.
+Performance for CSI is slightly but definitively higher than for SCIRL (see the $p$-values for the mean equality test in Tab.~\ref{tab:highway}), slightly below the performance of the expert itself.
 Very few samples (100) are needed to reliably achieve expert-level performance.
@@
 The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound lets us expect.
-\begin{figure}
-      \begin{tabular}{|l|l|l|l|l|}
-        \hline
-        &\bf CSI& \bf SCIRL& \bf RE& \bf Classif\\
-        \hline \multicolumn{5}{|c|}{$n = 9$}\\ \hline
-        Var& 5.139& 5.684& 4.686& 0.045 \\
-        Min& -2.708& -3.175& -2.717& -1.564 \\
-        Mean& 2.419& 2.762& 2.286& -1.149 \\
-        Max& 7.050& 7.060& 7.027& -0.596 \\
-        \hline \multicolumn{5}{|c|}{$n = 49$}\\ \hline
-        Var& 0.214& 0.517& 0.701& 0.121 \\
-        Min& 5.406& 4.652& 4.214& -1.148 \\
-        Mean& 7.012& 6.784& 6.314& -0.158 \\
-        Max& 7.454& 7.529& 7.465& 0.619 \\
-        \hline \multicolumn{5}{|c|}{$n = 100$}\\ \hline
-        Var& 0.011& 0.269& 0.486& 0.154 \\
-        Min& 7.078& 5.724& 5.487& -0.574 \\
-        Mean& 7.364& 7.194& 6.778& 0.456 \\
-        Max& 7.526& 7.562& 7.509& 1.483 \\ \hline
+\begin{table}[ht]
+	\begin{center}
+		\begin{tabular}{cc}
+		\toprule
+			Number of expert samples & $p$-value\\
+			\midrule
+			$9$ & $0.193042$\\
+			$49$ & $0.00178467$\\
+			$100$ & $0.00178467$\\
+			$225$ & $6.66654\e{-22}$\\
+			$400$ & $3.70654\e{-36}$\\
+		\bottomrule
 		\end{tabular}
-      \begin{tabular}{|l|l|l|l|l|}
-        \hline
-        &\bf CSI& \bf SCIRL& \bf RE& \bf Classif\\
-        \hline \multicolumn{5}{|c|}{$n = 225$}\\ \hline
-        Var& 0.006& 0.029& 0.146& 0.167 \\
-        Min& 7.172& 5.882& 5.612& 0.128 \\
-        Mean& 7.459& 7.377& 7.215& 1.151 \\
-        Max& 7.563& 7.572& 7.578& 2.220 \\
-        \hline \multicolumn{5}{|c|}{$n = 400$}\\ \hline
-        Var& 0.002& 0.002& 0.005& 0.267 \\
-        Min& 7.321& 7.247& 7.144& 0.041 \\
-        Mean& 7.495& 7.369& 7.279& 1.744 \\
-        Max& 7.552& 7.557& 7.519& 2.761 \\
-        \hline
-      \end{tabular}
-      \caption{Raw results on the Highway driving problem, see how the variance of SCIRL and CSI decreases.}
+	\end{center}
+        \caption{Kolmogorov–Smirnov mean equality test $p$-values for CSI and SCIRL on the highway simulator.
+High values (>0.01) means that the hypothesis that the means are equal cannot be rejected.}
         \label{tab:highway}
-      \end{figure}
+\end{table}
   \begin{figure}
@@
 
-\section{Related Work}
-\label{sec:related}
-IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
-Approaches summarized in \cite{neu2009training} can be seen as iteratively contructing a reward function, solving an MDP at each iteration.
-Some of these algorithms are IRL algorithms while others fall in the Apprenticeship Learning (AL) category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship}.
-In both cases the need to solve an MDP at each step may be very demanding, both sample-wise and computationally.
-CSI being able to output a reward function without having to solve the MDP is thus a significant improvement.
-
-AL via classification has been proposed for example in \cite{ratliff2007imitation}, with the help of a structured margin method.
-Using the non trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
-in a classification algorithm, showing improvements compared to a
-non-structured kernel.
-
-Classification and IRL have met in the past in \cite{ratliff2006maximum}, but the labels were complete optimal policies rather than actions and the inputs were MDPs, which had to be solved.
-It may be unclear how SCIRL \cite{klein2012scirl} relates to the proposed approach of his paper.
-Both algorithms use the score function of a classifier as a proxy to the action-value function of the expert with respect to the (unknown) true reward: $Q^{\pi_E}_R$.
-The way this proxy is constructed and used, however, fundamentally differs in the two algorithms.
-This difference will cause the theoretical analysis of both approaches (see Sec.~\ref{sec:analysis}) to be distinct.
-In SCIRL, the score function of the classifier is approximated {\it via} a linear parametrization that relies on the feature expectation of the expert $\mu^E(s) = E[\sum_{t\geq 0}\gamma^t\phi(s_t)|s_0=s,\pi_E]$.
-This entails the use of a specific kind of classifier (namely linearly-parametrized-score-function-based classifiers) and of a method of approximation of $\mu^E$.
-By contrast, almost any off-the-shelf classifier can be used in the first step of the cascading approach of this paper.
-The classification step of CSI is unaware of the structure of the MDP whereas SCIRL knows about it thanks to the use of $\mu^E$.
-In CSI, the structure of the MDP is injected by reversing the Bellman equation prior to the regression step (Eq.~\ref{eq:rc} and \eqref{eq:rj}), a step that does not exist in SCIRL as SCIRL directly outputs the parameter vector found by its linearly-parametrized-score-function-based classifier.
-The regressor of CSI can be chosen off-the-shelf.
-One can argue that this and not having to approximate $\mu^E$ increases the ease-of-use of CSI over SCIRL and makes for a more versatile algorithm.
-In practice, as seen in Sec.~\ref{sec:experiments}, perfomance of SCIRL and CSI are very close to one another thus CSI may be a better choice as it is easier to deploy.
-We would like to emphasize the fact that although SCIRL and CSI share similarities, there exist key differences that we just explained.
-Neither approach is a generalization of the other.
-
-Few IRL or AL algorithms do not require solving an MDP.
-The approach of \cite{syed2008apprenticeship} requires knowing the transition probabilities of the MDP (which CSI does not need) and outputs a policy (and not a reward).
-The algorithm in \cite{dvijotham2010inverse} only applies to linearly-solvable MDPs whereas our approach does not place such restrictions.
-Closer to our use-case is the idea presented in \cite{boularias2011relative} to use a subgradient ascent of a utility function based on the notion of relative entropy.
-Importance sampling is suggested as a way to avoid solving the MDP.
-This requires sampling trajectories according to a non-expert policy and
-the direct problem remains at the core of the approach (even if
-solving it is avoided).
-
 \section{Conclusion}
@@
 Practical examples of classifiers and regressors have been given, and two combinations have been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
-On one of the benchmarks (a highway driving simulator), our algorithm is shown to outperform other state-of-the-art approaches.
+On both benchmarks, our algorithm is shown to outperform other state-of-the-art approaches (although SCIRL catches up on the mountain car).
 We plan on deepening the analysis of the theoretical properties of our approach and on applying it to real world robotics problems.

### 2013-03-29 piot2012cascading_JFPDA/Cascading.tex
Version JFPDA du papier sur CSI.

--- 2013-03-29 piot2012cascading/ECML13_cascading.tex
@@
-% Created 2012-10-15 lun. 14:49
-\documentclass[smallextended]{svjour3}
-\smartqed 
+\documentclass{jfpda2013}
+
+\newcommand{\verbfichex}{jfpda2013.\xspace}
+\newcommand{\verbfichclass}{jfpda2013.cls\xspace}
+\newcommand{\verbfichbst}{jfpda2013.bst\xspace}
+
+\usepackage{amsthm}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{subfigure}
+\usepackage{natbib}
 % scientific notation, 1\e{9} will print as 1x10^9
@@
 \tolerance=1000
-%\usepackage{amsmath}
-%\usepackage{amsthm}
+\usepackage{amsmath}
 %\newtheorem{definition}{Definition}
-%\newtheorem{theorem}{Theorem}
-\newtheorem{cor}{Corollary}
+\newtheorem{theorem}{Theorem}
+\newtheorem{corollary}{Corollary}
 \usepackage{dsfont}
@@
 \newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
-\begin{document}
-\title{A cascaded supervised learning approach to inverse reinforcement learning}
-\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
-\titlerunning{Cascading approach to IRL}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% fournir un titre court si le titre fait plus de 40 caractères
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\shorttitle{Approche cascadée de l'ARI}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% donner le titre de l'ouvrage selon la conférence IC, CAp, JFPDA ou RJCIA
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\shortouvrage{CAp 2007}
+%\shortouvrage{IC 2007}
+%\shortouvrage{JFPDA 2007}
+%\shortouvrage{RJCIA 2007}
+\shortouvrage{JFPDA 2013}
+
+% Titre, auteur, pas de date
+
+\title{Apprentissage par renforcement inverse en cascadant classification et régression
+}
+
+\author{Edouard Klein\inst{1,}\inst{2}, Bilal Piot\inst{2,}\inst{3}, Matthieu Geist\inst{2}, Olivier Pietquin\inst{2,}\inst{3}}
+
 \institute{
@@
 3. UMI 2958 (GeorgiaTech-CNRS)\\
-France\\
-}
+France\\}
+
+\begin{document}
+
 \date{\today}
@@
 \maketitle
+
+\begin{abstract}
+  Cette contribution traite du problème de l'apprentissage par renforcement inverse, c'est à dire la découverte d'une fonction de récompense pour laquelle la politique de l'expert que l'on observe est optimale.
+Nous poposons de diviser le problème en deux étapes d'apprentissage supervisé : c'est l'approche en cascade (CSI pour {\it Cascaded Supervised Inverse reinforcement learning}).
+Une étape de classification qui définit une fonction de score est suivie par une étape de régression qui donne une fonction de récompense.
+Une analyse théorique démondre que la politique de l'expert observée est quasi-optimale pour la fonction de récompense trouvée.
+Deux avantages de cette approche sont de ne pas avoir recours à la résolution répétée du Processus Decisionnel de Markov (MDP) et la possibilité de profiter des techniques existantes en classification et régression.
+Nous montrons de plus empiriquement que CSI peut fonctionner avec pour seule information (à une heuristique près) des échantillons recueillis par la politique de l'expert.
+Nous testons cela sur deux banc d'essais classiques, le {\it mountain car} et un simulateur de conduite sur autoroute.
+  \end{abstract}
 
@@
 \end{equation}
-\qed
+%\qed
 \end{proof}
@@
   As $\epsilon_R = 0$, we have $\hat R^C = R^C$, hence the result.
-  \qed
+  %\qed
 \end{proof}
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 \bibliography{Biblio}

### 2013-04-03 Exp11.py
KS test

@@ -59,16 +59,24 @@ savefig("Exp11.pdf")
 
 # <codecell>
 
-mean(CSI_data[0])
+from matrix2latex import matrix2latex
+m = [[1, 1], [2, 4], [3, 9]] # python nested list
+t = matrix2latex(m)
+print t
+X
 
 # <codecell>
 
 from scipy.stats import ks_2samp
-for i in range(0,len(X)):
+table = zeros((4,2))
+for i in range(0,len(X[:4])):
     y_csi = CSI_data[i]
     y_scirl = SCIRL_data[i]
     pvalue = ks_2samp(y_csi,y_scirl)[1]
-    print "Pour X = "+str(X[i])+(", \tKS mean equality test p-value : %.2e "%pvalue)+"\t\t"+("REJECTED" if pvalue < 0.01 else "")
+    table[i,0] = X[i]
+    table[i,1] = pvalue
+t = matrix2latex(table, headerRow=["Number of expert samples","$p$-value"])
+print t
 
 # <codecell>
 

### 2013-04-03 Plot15.py

@@ -41,10 +41,15 @@ savefig("Exp14.pdf")
 # <codecell>
 
 from scipy.stats import ks_2samp
+from matrix2latex import matrix2latex
+table = zeros((len(Abcissas),2))
 for i in range(0,len(Abcissas)):
     y_csi = Y_CSI[i]
     y_scirl = Y_SCIRL[i]
-    print "Pour X = "+str(Abcissas[i])+", \tKS mean equality test p-value : %.2e "%ks_2samp(y_csi,y_scirl)[1]
+    table[i][0] = Abcissas[i]
+    table[i][1] = ks_2samp(y_csi,y_scirl)[1]
+t = matrix2latex(table, headerRow=["Number of expert samples","$p$-value"])
+print t
 
 # <codecell>
 

### 2013-04-03 matrix2latex/test.py
matrix2latex

@@ -1,0 +1,193 @@
+"""This file is part of matrix2latex.
+
+matrix2latex is free software: you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation, either version 3 of the License, or
+(at your option) any later version.
+
+matrix2latex is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with matrix2latex. If not, see <http://www.gnu.org/licenses/>.
+"""
+
+# tests for matrix2latex.py
+from matrix2latex import matrix2latex
+m = [[1, 2, 3], [4, 5, 6]]
+
+f = open('test.tex')
+answers = dict()
+for line in f:
+    if line.startswith('%%%'):
+        name = line[3:-1]               # ignore %%% and \n
+        answers[name] = ''
+    else:
+        answers[name] += line
+
+def loopTwoLists(x, y):
+    for ix in range(max([len(x), len(y)])):
+        try: a = x[ix].strip()
+        except: a = ''
+        try: b = y[ix].strip()
+        except: b = ''
+        yield a, b
+
+def assertEqual(x, name):
+    # assert each line is equal, ignoring leading and trailing spaces
+    print(x)
+    y = answers[name]
+    x = x.split('\n')
+    y = y.split('\n')
+    correct = True
+    for a, b in loopTwoLists(x, y):
+        if a != b:
+            correct = False # found 1 or more error
+            
+    if not(correct):
+        for a, b in loopTwoLists(x, y):
+            print(a,b)
+        raise AssertionError
+
+def test_simple():
+    t = matrix2latex(m)
+    assertEqual(t, "simple")
+
+def test_transpose1():
+    t = matrix2latex(m, transpose=True)
+    assertEqual(t, "transpose1")
+
+def test_transpose2():
+    cl = ["a", "b"]
+    t = matrix2latex(m, transpose=True, headerRow=cl)
+    assertEqual(t, "transpose2")
+
+def test_file():
+    matrix2latex(m, 'tmp.tex')
+    f = open('tmp.tex')
[…]

### 2013-04-03 matrix2latex/test_compatibility.py

@@ -1,0 +1,53 @@
+# Tries to run testsuit with python version
+# python<major>.<minor>
+# where major and minor varies between 2-3 and 0-10.
+# (expand this if you have say python1.11 or python4.0)
+# Now also looks for pypy implementations.
+"""This file is part of matrix2latex.
+
+matrix2latex is free software: you can redistribute it and/or modify
+it under the terms of the GNU General Public License as published by
+the Free Software Foundation, either version 3 of the License, or
+(at your option) any later version.
+
+matrix2latex is distributed in the hope that it will be useful,
+but WITHOUT ANY WARRANTY; without even the implied warranty of
+MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+GNU General Public License for more details.
+
+You should have received a copy of the GNU General Public License
+along with matrix2latex. If not, see <http://www.gnu.org/licenses/>.
+"""
+
+import os
+from subprocess import call
+from matrix2latex import matrix2latex
+
+table = list()
+cl = list()
+
+def test(python, table, cl):
+    ret = call(python + " test.py", shell=True, stdout=file(os.devnull, "w"))
+
+    print(python)            
+    print(ret)
+    if ret == 0: table.append(["True"])
+    else: table.append(["False"])
+    cl.append(python)
+
+
+for major in range(2, 3+1):
+    for minor in range(0, 10):
+        python = 'python%d.%d' % (major, minor)
+        
+        if call(python + " -c ''", shell=True, stderr=file(os.devnull, "w")) == 0:
+            test(python, table, cl)
+
+if call("pypy-c" + " -c 'pass'", shell=True, stderr=file(os.devnull, "w")) == 0:
+    test("pypy-c", table, cl)
+
+c = "Does 'python test.py' return 0?"
+compatibleTable = matrix2latex(table, 'doc/compatibleTable',
+                               headerColumn=cl, headerRow=['Compatible'],
+                               caption=c)
+print compatibleTable

### 2013-04-05 piot2012cascading_JFPDA/Cascading.tex
Version soumise à JFPDA2013

@@
 			\midrule
-			$10$ & $4.5195\e{-14}$\\
-			$30$ & $0.0131123$\\
-			$100$ & $0.139975$\\
-			$300$ & $0.893752$\\
+                        $ 10$ & $\mathbf{1.5e-12}$\\
+                        $ 30$ & $3.8e-01$\\
+                        $100$ & $1.3e-02$\\
+                        $300$ & $7.4e-01$\\
                         \bottomrule
@@
 	\end{center}
-        \caption{Kolmogorov–Smirnov mean equality test $p$-values for CSI and SCIRL on the mountain car. High values (>0.01) means that the hypothesis that the means are equal cannot be rejected.}
+        \caption{Student or Welch test of mean equality (depending on wether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the mountain car.
+High values ($>1.0\e{-02}$) means that the hypothesis that the means are equal cannot be rejected.
+As highlighted here, the means are different when the algorithms are fed only $10$ expert samples in which case CSI outperforms SCIRL.}
         \label{tab:mc}
@@
 			\midrule
-			$9$ & $0.193042$\\
-			$49$ & $0.00178467$\\
-			$100$ & $0.00178467$\\
-			$225$ & $6.66654\e{-22}$\\
-			$400$ & $3.70654\e{-36}$\\
+                        $  9$ & $3.0e-01$\\
+                        $ 49$ & $\mathbf{8.9e-03}$\\
+                        $100$ & $\mathbf{1.8e-03}$\\
+                        $225$ & $\mathbf{2.4e-05}$\\
+                        $400$ & $\mathbf{2.0e-50}$\\
                         \bottomrule
@@
 	\end{center}
-        \caption{Kolmogorov–Smirnov mean equality test $p$-values for CSI and SCIRL on the highway simulator.
-High values (>0.01) means that the hypothesis that the means are equal cannot be rejected.}
+        \caption{Student or Welch test of mean equality (depending on wether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the driving simulator.
+High values ($>1.0\e{-02}$) means that the hypothesis that the means are equal cannot be rejected.
+As highlighted here, the means are different from 49 samples on : CSI slightly but definitively outperforms SCIRL.}
         \label{tab:highway}

### 2013-04-05 piot2012cascading_ECML/Cascading.tex
Version ECML du cascading

--- 2013-03-29 piot2012cascading/ECML13_cascading.tex
@@
-% Created 2012-10-15 lun. 14:49
-\documentclass[smallextended]{svjour3}
-\smartqed 
+\documentclass{llncs}
+\usepackage{makeidx}  % allows for indexgeneration
+
+%\usepackage{amsthm}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{subfigure}
+%\usepackage{natbib}
 % scientific notation, 1\e{9} will print as 1x10^9
@@
 \tolerance=1000
-%\usepackage{amsmath}
-%\usepackage{amsthm}
+\usepackage{amsmath}
 %\newtheorem{definition}{Definition}
 %\newtheorem{theorem}{Theorem}
-\newtheorem{cor}{Corollary}
+%\newtheorem{corollary}{Corollary}
 \usepackage{dsfont}
@@
 \newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
+
 \begin{document}
 \title{A cascaded supervised learning approach to inverse reinforcement learning}
-\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
 \titlerunning{Cascading approach to IRL}
+\authorrunning{Klein, Piot, Geist and Pietquin}
+\author{Edouard Klein\inst{1,}\inst{2}, Bilal Piot\inst{2,}\inst{3}, Matthieu Geist\inst{2}, Olivier Pietquin\inst{2,}\inst{3}}
+
 \institute{
@@
 3. UMI 2958 (GeorgiaTech-CNRS)\\
-France\\
-}
+France\\}
+
 \date{\today}
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{splncs03}
 \bibliography{Biblio}

### 2013-04-05 piot2012cascading_ECML/Cascading.tex
Version ECML sur 16 pages.

@@
 %\usepackage{amsthm}
+\usepackage{subfig}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{algorithmic}
-\usepackage{subfigure}
+%\usepackage{subfigure}
 %\usepackage{natbib}
@@
 The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
-Given that $P_\pi = (P(s'|s,\pi(s)))_{s,s' \in S}$ is the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
-\begin{equation}
-  \rho_\pi^TP_\pi = \rho_\pi^T\textrm{, with }X^T\textrm{ being the transpose of }X.
-  \end{equation}
+Given that $P_\pi = (P(s'|s,\pi(s)))_{s,s' \in S}$ is the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies $\rho_\pi^TP_\pi = \rho_\pi^T\textrm{, with }X^T\textrm{ being the transpose of }X$.
 The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.
@@
 The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the expected discounted cumulative reward for following policy $\pi$ from this state:
-\begin{equation}
-V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s,\pi].
-\end{equation}
+$V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s,\pi]$.
 This long-term criterion is what is being optimized when solving an MDP.
@@
 The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
-\begin{equation}
-  T^{\pi}_RV = R_\pi + \gamma P_\pi V
-  \end{equation}
+$  T^{\pi}_RV = R_\pi + \gamma P_\pi V$
 where $R_\pi = (R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally:
-\begin{equation}
-  T^*_RV = \max_\pi T^\pi_RV.
-\end{equation}
+ $ T^*_RV = \max_\pi T^\pi_RV$.
   Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R:
-  \begin{eqnarray}
-    V^\pi_R &=& T^\pi_R V^\pi_R \Leftrightarrow V^\pi_R = R_\pi + \gamma P_\pi V^\pi_R.
-  \end{eqnarray}
+    $V^\pi_R = T^\pi_R V^\pi_R \Leftrightarrow V^\pi_R = R_\pi + \gamma P_\pi V^\pi_R$.
 The Bellman optimality operator $T^*_R$ also admits a fixed point, the optimal value function $V_R^*$ with respect to reward $R$.
@@
   \item the classification error defined as
-\begin{equation}  
-\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}];
-\end{equation}
+$\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]$;
 \item the regression error defined as 
-  \begin{equation}
-    \epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E},
-  \end{equation}
+ $   \epsilon_R=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\rho_E}$,
   with:
@@
 {\it A priori}, $C_*$ can be upper-bounded by a more usual and general concentration coefficient but $C_*$ gives a tighter final result: one can informally see $C_*$ as a measure of the similarity between the distributions induced by $\hat\pi_C$ and $\pi_E$ (roughly, if $\hat{\pi}_C\approx\pi_E$ then $C_* \approx 1$).
-\item $\Delta q$, defined by 
-  \begin{eqnarray}
-    \Delta q &= &\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a)),
-    \end{eqnarray} which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
+\item $\Delta q = \max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$,
+    which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
   \end{itemize}
@@
 \begin{proof}
-First let's recall some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$, $R^C\in\mathbb{R}^{S\times A}$ is:
-\begin{equation}
-\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s')),
-\end{equation}
+First let's recall some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$, $R^C\in\mathbb{R}^{S\times A}$ is so that $\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s'))$,
 and $\hat{R}^C\in\mathbb{R}^{S\times A}$ is the reward function output by the regression step.
@@
 
-We can now introduce the following vectorial equations:
-\begin{align}
-&R^C_a=q_a-\gamma P_aq_{\pi_C},
-\\
-&R^E_a=q_a-\gamma P_aq_{\pi_E},
-\\
-&\epsilon^R_a=R^C_a-\hat{R}^C_a.
-\end{align}
+We no have the following vectorial equalities $
+R^C_a=q_a-\gamma P_aq_{\pi_C} ;
+R^E_a=q_a-\gamma P_aq_{\pi_E};
+\epsilon^R_a=R^C_a-\hat{R}^C_a$.
 Now, we are going to upper bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious as $V^*$ is optimal).
@@
 We are going to bound each of these three terms. First, let $\pi$ be a given deterministic policy. We have, using $\epsilon^R = R^C - \hat R^C$:
-\begin{equation}
-V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C}=V^{\pi}_{\epsilon^R}=(I-\gamma P_\pi)^{-1}\epsilon^R_{\pi}.
-\end{equation}
+$V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C}=V^{\pi}_{\epsilon^R}=(I-\gamma P_\pi)^{-1}\epsilon^R_{\pi}$.
 If $\pi=\pi_E$, we have, thanks to the power series expression of $(I-\gamma P_{\pi_E})^{-1}$, the definition of $\rho_E$ and the definition of the $\mu$-weighted $L_1$ norm, one property of which is that $\forall X, \mu^TX \leq \|X\|_{1,\mu}$:
@@
 So, using the notations introduced before we stated the theorem, we are able to give an upper bound to the first and third terms (recall also the notation $C_{\hat \pi_C} = C_*$):
-\begin{equation}
-\rho_E^T((V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})+(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}))\leq\frac{1+C_*}{1-\gamma}\epsilon_R.
-\end{equation}
+$\rho_E^T((V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})+(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}))\leq\frac{1+C_*}{1-\gamma}\epsilon_R$.
 Now, there is still an upper bound to find for the second term. It is possible to decompose it as follows:
@@
 Using $\Delta q$, we have:
-\begin{equation}
-\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})\leq\Delta q\sum_{s\in S}\rho_E(s)[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]= \Delta q \epsilon_C.
-\end{equation}
+$\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})\leq\Delta q\sum_{s\in S}\rho_E(s)[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]= \Delta q \epsilon_C$.
 Finally, we also have:
 \begin{align}
-\rho_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(R^E_{\pi_E}-R^C_{\pi_E}),
-\\
-&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\gamma P_{\pi_E}(q_{\pi_C}-q_{\pi_E}),
+\rho_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(R^E_{\pi_E}-R^C_{\pi_E})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\gamma P_{\pi_E}(q_{\pi_C}-q_{\pi_E}),
 \\
@@
 So the upper bound for the second term is: $\rho_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\Delta q}{1-\gamma}\epsilon_C$.
-If we combine all of the results, we obtain the final bound:
-\begin{equation}
-0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \frac{1}{1-\gamma}(\epsilon_C\Delta q +\epsilon_R(1+C_*))).
-\end{equation}
+If we combine all of the results, we obtain the final bound as stated in the theorem.
 %\qed
@@
 giving samples $\{(s_k,a_k=\pi_E(a_k),s'_k)_k\} = D_E$. The dataset $D_C$ fed to the classifier is straightforwardly constructed from $D_E$ by dropping the $s'_k$ terms:
-\begin{equation}
-  D_C = \{(s_i = s_k,a_i=a_k)_i\}.
-\end{equation}
+$  D_C = \{(s_i = s_k,a_i=a_k)_i\}$.
 \subsubsection{Heuristics}
@@
 We artificially augment the dataset $D_R$ with samples $((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}$ where
-\begin{equation}
-  r_{min} = \min_k\hat r_k - 1.
-\end{equation}
+$  r_{min} = \min_k\hat r_k - 1$.
 This heuristics intructs the regressor to associate a state-action tuple disagreeing with the expert ({\it i.e.}, $(s_k,a\neq a_k)$) with a reward strictly inferior to any of those associated with expert state action tuples ({\it i.e.}, $(s_k,a_k=\pi_E(s_k))$).
@@
 \begin{table}
-	\begin{center}
+  \small
+  \subfloat[Mountain Car\label{tab:mc}]{
     \begin{tabular}{cc}
@@
     \end{tabular}
-	\end{center}
-        \caption{Student or Welch test of mean equality (depending on wether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the mountain car.
-High values ($>1.0\e{-02}$) means that the hypothesis that the means are equal cannot be rejected.
-As highlighted here, the means are different when the algorithms are fed only $10$ expert samples in which case CSI outperforms SCIRL.}
-        \label{tab:mc}
+  }
+  \hfill
+  \subfloat[Highway Driving\label{tab:highway}]{
+    \begin{tabular}{cc}
+      \toprule
+      Number of expert samples & $p$-value\\
+      \midrule
+      $  9$ & $3.0e-01$\\
+      $ 49$ & $\mathbf{8.9e-03}$\\
+      $100$ & $\mathbf{1.8e-03}$\\
+      $225$ & $\mathbf{2.4e-05}$\\
+      $400$ & $\mathbf{2.0e-50}$\\
+      \bottomrule
+    \end{tabular}
+  }
+  \caption{Student or Welch test of mean equality (depending on wether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the mountain car (\ref{tab:mc}) and the highway driving simulator (\ref{tab:highway}).
+High values ($>1.0\e{-02}$) means that the hypothesis that the means are equal cannot be rejected.}
 \end{table}
 \begin{figure}
-  \includegraphics[width=\textwidth]{Exp11.pdf}
+  \center
+  \includegraphics[width=.71\textwidth]{Exp11.pdf}
   \caption{Performance of various policies on the mountain car problem. This is the mean over 100 runs.}
   \label{fig:MountainCar}
+\end{figure}
+
+  \begin{figure}
+   \subfloat[Mean performance over 100 runs on the Highway driving problem.]{\label{tutu}\includegraphics[width=.5\linewidth]{"Exp14"}}
+   \subfloat[Zoom of Fig~\ref{tutu} showing the ranking of the three IRL algorithms.]{\includegraphics[width=.5\linewidth]{"Exp14_zoom"}}
+  \caption{Results on the highway driving problem.}
+  \label{fig:Highway}
+
 \end{figure}
@@
 We give the values of $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$ with $\pi$ being in turn the optimal policy for the rewards given by SCIRL, CSI and RE, the policy $\pi_C$ of the classifier (the very one the classification step of CSI outputs), and the optimal policy for a randomly drawn reward.
-Performance for CSI is slightly but definitively higher than for SCIRL (see the $p$-values for the mean equality test in Tab.~\ref{tab:highway}), slightly below the performance of the expert itself.
+Performance for CSI is slightly but definitively higher than for SCIRL (see the $p$-values for the mean equality test in Tab.~\ref{tab:highway}, from 49 samples on), slightly below the performance of the expert itself.
 Very few samples (100) are needed to reliably achieve expert-level performance.
@@
 The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound lets us expect.
-\begin{table}[ht]
-	\begin{center}
-		\begin{tabular}{cc}
-		\toprule
-			Number of expert samples & $p$-value\\
-			\midrule
-                        $  9$ & $3.0e-01$\\
-                        $ 49$ & $\mathbf{8.9e-03}$\\
-                        $100$ & $\mathbf{1.8e-03}$\\
-                        $225$ & $\mathbf{2.4e-05}$\\
-                        $400$ & $\mathbf{2.0e-50}$\\
-                        \bottomrule
-		\end{tabular}
-	\end{center}
-        \caption{Student or Welch test of mean equality (depending on wether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the driving simulator.
-High values ($>1.0\e{-02}$) means that the hypothesis that the means are equal cannot be rejected.
-As highlighted here, the means are different from 49 samples on : CSI slightly but definitively outperforms SCIRL.}
-        \label{tab:highway}
-\end{table}
-  \begin{figure}
-   \subfigure[Mean performance over 100 runs on the Highway driving problem.]{\label{tutu}\includegraphics[width=.5\linewidth]{"Exp14"}}
-   \subfigure[Zoom of Fig~\ref{tutu} showing the ranking of the three IRL algorithms.]{\includegraphics[width=.5\linewidth]{"Exp14_zoom"}}
-  \caption{Results on the highway driving problem.}
-  \label{fig:Highway}
-
-\end{figure}
 

### 2013-04-18 piot2012cascading_ECML/Cascading.tex
Version soumise à ECML track conf.

@@
 Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach.
-It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics.
+It is furthermore empirically demonstrated to compare positively to state-of-the-art approaches when using only transitions sampled according to the expert policy, up to the use of some heuristics.
 This is exemplified on two classical benchmarks (the mountain car problem and a highway driving simulator).
@@
 \\
-&=\frac{\gamma}{1-\gamma}\rho_E^T(q_{\pi_C}-q_{\pi_E})\leq \frac{\gamma}{1-\gamma}\Delta q \epsilon_C \leq  \frac{1}{1-\gamma}\Delta q \epsilon_C\textrm{ as }\gamma<1.
+&=\frac{\gamma}{1-\gamma}\rho_E^T(q_{\pi_C}-q_{\pi_E})\leq \frac{\gamma}{1-\gamma}\Delta q \epsilon_C.
 \end{align}
-So the upper bound for the second term is: $\rho_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq \frac{\Delta q}{1-\gamma}\epsilon_C$.
+So the upper bound for the second term is: $\rho_E^T(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})\leq (\Delta q +\frac{\gamma}{1-\gamma}\Delta q) \epsilon_C =   \frac{\Delta q}{1-\gamma}\epsilon_C$.
 If we combine all of the results, we obtain the final bound as stated in the theorem.

### 2013-04-19 klein2012structured_RIA/scirl_ria.tex
Dernières modifs avant publication de SCIRL à RIA

@@
 \documentclass[french,utf8]{./hermes-journal}
+\usepackage[frenchb]{babel}
 \usepackage[utf8]{inputenc}
@@
 % \journalName{year}{volume}{issue}
-\ria{2012}{22}{1}
-
-\firstpagenumber{9}
+\ria{2013}{27}{2}
+\firstpagenumber{155}
+%\lastpagenumber{179}
 \let\chapter\section
@@
 Nancy, France
+\vspace{-2em}
 }{}
@@
 \address{ UMI 2958 (GeorgiaTech-CNRS)\\
-Metz, France}{}
+Metz, France
+\vspace{-2em}
+}{}
 
@@
   the so-called feature expectation of the expert as the
-  parameterization of the score function of a multi-class
+  parameterization of the score function of a multiclasse
   classifier. This approach produces a reward function for which the
@@
   driving simulator.}
-\resume{Cette contribution traite le problème de l'Apprentissage par Renforcement Inverse (ARI), défini comme la recherche d'une fonction de récompense pour laquelle le comportement d'un expert (connu par le biais de démonstrations) est optimal.
-Nous introduisons SCIRL, un nouvel algorithme qui utilise la grandeur dénommée attribut moyen de l'expert comme la paramétrisation d'une fonction de score pour un classifieur multi-classe.
-Cette approche donne une fonction de récompense pour laquelle la politique de l'expert est (nous le démontrons) quasi-optimale.
-Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct de l'Apprentissage par Renforcement.
-De plus, en utilisant une heuristique, il peut être utilisé avec uniquement des trajectoires échantillonnées par l'expert.
+\resume{Cette contribution traite le problème de l'apprentissage par renforcement inverse (ARI), défini comme la recherche d'une fonction de récompense pour laquelle le comportement d'un expert (connu par le biais de démonstrations) est optimal.
+Nous introduisons SCIRL, un nouvel algorithme qui utilise la grandeur dénommée attribut moyen de l'expert comme la paramétrisation d'une fonction de score pour un classifieur multiclasse.
+Cette approche donne une fonction de récompense pour laquelle la politique de l'expert est (nous le démontrons) quasi optimale.
+Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct de l'apprentissage par renforcement.
+De plus, en utilisant une heuristique, il fonctionne avec uniquement des trajectoires échantillonnées par l'expert.
 Nous illustrons cela sur un simulateur de conduite.}
 
-\keywords{Reinforcement Learning, Inverse Reinforcement Learning}
-\motscles{Apprentissage par Renforcement, Apprentissage par Renforcement Inverse}
+\keywords{reinforcement learning, inverse reinforcement learning.}
+\motscles{apprentissage par renforcement, apprentissage par renforcement inverse.}
 
@@
 
-L'Apprentissage par Renforcement Inverse (ARI) est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal ; historiquement proposé dans~\cite{Russel998}, ce mécanisme trouve des applications dans divers champs, de la biologie à la neuropsychologie en passant par l'économie et plus récemment la robotique~\cite{abbeel2010autonomous}.
-Beaucoup d'algorithmes d'ARI (que nous décrirons succintement Sec.~\ref{seelatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) proche de celle générée par l'expert.
-Souvent, cette distribution est caractérisée par ce que l'on appelle l'attribut moyen (voir Sec.~\ref{subseackgrounrl}) : étant donnée une fonction de récompense linéairement paramétrée par un vecteur d'attributs, il s'agit de l'espérance de la somme pondérée des vecteurs d'attributs sachant que l'on commence dans un certain état en choisissant une certaine action avant de suivre la politique concernée.
+L'apprentissage par renforcement inverse (ARI) est le problème lié à la recherche, à partir de démonstrations d'un expert, d'une fonction de récompense telle que le comportement de l'expert soit optimal ; historiquement proposé dans~\cite{Russel998}, ce mécanisme trouve des applications dans divers champs, de la biologie à la neuropsychologie en passant par l'économie et plus récemment la robotique~\cite{abbeel2010autonomous}.
+Beaucoup d'algorithmes d'ARI (que nous décrirons succinctement section~\ref{seelatedWorks}) cherchent une fonction de récompense dont la politique optimale associée génère une distribution sur les trajectoires (ou une grandeur liée à cette distribution) proche de celle générée par l'expert.
+Souvent, cette distribution est caractérisée par ce que l'on appelle l'attribut moyen (voir section~\ref{subseackgrounrl}) : étant donnée une fonction de récompense linéairement paramétrée par un vecteur d'attributs, il s'agit de l'espérance de la somme pondérée des vecteurs d'attributs sachant que l'on commence dans un certain état en choisissant une certaine action avant de suivre la politique concernée.
 
@@
 Il est possible d'imiter le comportement de l'expert via un algorithme d'apprentissage supervisé généralisant l'association des actions aux états.
-Nous considérons ici les classifieurs multi-classes qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour un état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir Sec.~\ref{subseackgrounlassif}).
-L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir Sec.~\ref{subsecirlgo}).
-Le vecteur de paramètres ainsi calculé définit une fonction de récompense dont on montre qu'elle admet pour politique quasi-optimale la politique de l'expert (voir Sec.~\ref{subsecirnalysis}).
+Nous considérons ici les classifieurs multiclasses qui à partir d'une base d'entraînement calculent les paramètres d'une fonction de score paramétrée linéairement ; la règle de décision pour un état est l'argument (l'action) qui maximise la fonction de score pour cet état (voir section~\ref{subseackgrounlassif}).
+L'idée de base de l'algorithme SCIRL que nous proposons est simplement d'estimer l'attribut moyen de l'expert comme la paramétrisation de la fonction de score (voir section~\ref{subsecirlgo}).
+Le vecteur de paramètres ainsi calculé définit une fonction de récompense dont on montre qu'elle admet pour politique quasi optimale la politique de l'expert (voir section~\ref{subsecirnalysis}).
 
 Un grand avantage de SCIRL est qu'il ne requiert pas, contrairement à la plupart des algorithmes d'ARI, la résolution répétée du problème direct, celui de l'apprentissage par renforcement.
-Il nécessite d'estimer l'attribut moyen de l'expert, mais ceci est équivalent à un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir Sec.~\ref{seracticalApproach}.
-De plus, à l'aide d'une heuristique, SCIRL peut s'éxécuter en utilisant uniquement des transitions issues de la politique de l'expert (il n'y a pas besoin d'échantilloner la dynamique complète).
-Nous illustrons cela sur un simulateur de conduite Sec.~\ref{sexperiments}.
+Il nécessite d'estimer l'attribut moyen de l'expert, mais ceci est équivalent à un problème d'évaluation de la politique (pour une politique observée, ce qui est moins problématique que l'optimisation répétée d'une politique), voir section~\ref{seracticalApproach}.
+De plus, à l'aide d'une heuristique, SCIRL peut s'éxécuter en utilisant uniquement des transitions issues de la politique de l'expert (il n'y a pas besoin d'échantillonner la dynamique complète).
+Nous illustrons cela sur un simulateur de conduite section~\ref{sexperiments}.
 
@@
 
-\subsection{Apprentissage par Renforcement (Inverse)}
+\subsection{Apprentissage par renforcement (inverse)}
 \label{subseackgrounrl}
 
-Un Processus Décisionnel de Markov (PDM)~\cite{Puterma994} est donné par le tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini, $\A$ est l'espace d'action fini, $\p =
+Un processus décisionnel de Markov (PDM)~\cite{Puterma994} est donné par le tuple $\{\s,\A,\p,\R,\gamma\}$ où $\s$ est l'espace d'état fini, $\A$ est l'espace d'action fini, $\p =
 \{P_a = (p(s'|s,a))_{1\leq s,s'\leq |\s|}, a\in\A\}$ est l'ensemble des probabilités de transition markoviennes, $\R\in\mathbb{R}^\s$ est la fonction de récompense sur les états et $\gamma$ est le facteur d'oubli.
@@
 
-L'objectif de l'Apprentissage par Renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertseka996,Sutto998,szepesvari2010c}.
+L'objectif de l'apprentissage par renforcement (AR) est d'estimer la politique de contrôle optimale lorsque le modèle (les probabilités de transition et la fonction de récompense) est inconnu (mais observé par interaction avec le système à contrôler) et quand l'espace d'état est trop grand pour qu'une représentation exacte des objets concernés (fonctions de valeur, politiques) soit possible~\cite{Bertseka996,Sutto998,szepesvari2010c}.
 %
@@
 %
-L'Apprentissage par Renforcement Inverse (approché)~\cite{N000} consiste à estimer la fonction de récompense pour laquelle une politique observée est optimale.
+L'apprentissage par renforcement inverse (approché)~\cite{N000} consiste à estimer la fonction de récompense pour laquelle une politique observée est optimale.
 %
@@
 \X$ sont tirées selon une distribution inconnue $\prob(x)$ et qu'il existe un oracle qui à chacune de ces entrées associe un label $y\in Y$ tiré selon une distribution de probabilité conditionnelle inconnue $\prob(y|x)$.
-De manière générale, la classification multi-classe cherche, étant donné une base d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tirée selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
+De manière générale, la classification multiclasse cherche, étant donné une base d'entraînement $\{(x_i,y_i)_{1\leq i \leq N}\}$ tirée selon $\prob(x,y)$, une règle de décision $g\in\Y^\X$ minimisant l'erreur de classification $\E[\chi_{\{g(x)\neq y\}}] = \prob(g(x)\neq y)$, où $\chi$ est la fonction indicatrice.
 
@@
 En se basant sur une base d'entraînement $\{(x_i,y_i)_{1\leq
-i\leq N}\}$, un classifieur multi-classe (noté CMC) à fonction de score paramétrée linéairement choisit un vecteur de paramètres $\theta_c$.
+i\leq N}\}$, un classifieur multiclasse (noté CMC) à fonction de score paramétrée linéairement choisit un vecteur de paramètres $\theta_c$.
 La qualité de ce choix est quantifiée par l'erreur de classification $\epsilon_c =
@@
 Notre étude est valable pour tout algorithme de classification tant qu'il opère en maximisant l'argument d'une fonction de score paramétrée linéairement.
-Par exemple, un séparateur à vaste marge multi-classe pourrait être choisie~\cite{Guermeu007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taska005}.
+Par exemple, un séparateur à vaste marge multiclasse pourrait être choisie~\cite{Guermeu007} (en considérant le noyau induit par le vecteur d'attributs) ou encore une approche à vaste marge structurée~\cite{Taska005}.
 
@@
 
-Nous nous trouvons dans le cadre de la classification décrit en Sec.~\ref{subseackgrounlassif}.
+Nous nous trouvons dans le cadre de la classification décrit en section~\ref{subseackgrounlassif}.
 L'entrée $x$ peut être vue comme un état et le label $y$ comme une action.
@@
 $g_w(x)$ est interprétable comme une politique gloutonne vis-à-vis de la fonction de score $w^\top \psi(x,y)$, qui peut elle-même être vue comme une fonction de qualité.
-Le parallèle avec l'Eq.~\eqref{eeu} est aisé, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de la base d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi-optimale.
+Le parallèle avec l'équation~\eqref{eeu} est aisé, si $\psi(x,y)$ est l'attribut moyen d'une politique $\pi$ qui a fourni les labels de la base d'entraînement, et si l'erreur de classification est faible, alors $w$ est le vecteur de paramètres de la fonction de récompense vis-à-vis de laquelle on espère que la politique $\pi$ est quasi optimale.
 Ces remarques nous permettent maintenant d'introduire notre algorithme d'ARI par classification structurée (SCIRL pour \emph{Structured Classification-based Inverse Reinforcement Learning}).
@@
 Soit $\D = \{(s_i, a_i=\pi_E(s_i))_{1\leq i\leq N}\}$ une base d'entraînement où les états sont échantillonnés selon la distribution stationnaire de l'expert\footnote{Par exemple, si la chaîne de Markov induite par la politique de l'expert est à mélange rapide (\emph{fast-mixing}), l'échantillonnage d'une trajectoire donnera rapidement des échantillons tirés selon cette distribution.} $\rho_E = \rho_{\pi_E}$.
-Supposons également avoir à disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation Eq.~\eqref{eeu}.
-Une description de la manière d'estimer cette quantité en pratique est reportée à la Sec.~\ref{subseracticalApproacuE}; rappelons tout de même qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé Sec.~\ref{subseackgrounrl}.
+Supposons également avoir à disposition une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert $\mu^{\pi_E}$ défini à l'équation~\eqref{eeu}.
+Une description de la manière d'estimer cette quantité en pratique est reportée à la section~\ref{subseracticalApproacuE}; rappelons tout de même qu'estimer $\mu^{\pi_E}$ est simplement un problème d'\emph{évaluation de la politique} (estimation de la fonction de qualité d'une politique), comme signalé section~\ref{subseackgrounrl}.
 Supposons enfin qu'un algorithme de CMC a été choisi.
@@
   \SetAlgoVlined
-  \caption{Algorithme SCIRL}
+  \caption{{\it SCIRL}}
   \label{algcirl}
@@
 
-L'approche proposée est résumé par l'Alg.~\ref{algcirl}.
-Le nom de l'algorithme (SCRIL, que l'on peut traduire par Classification structurée pour l'ARI), vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
+L'approche proposée est résumée par l'algorithme~\ref{algcirl}.
+Le nom de l'algorithme (SCI\-RL, que l'on peut traduire par Classification structurée pour l'ARI), vient de l'utilisation de l'attribut moyen de l'expert dans le classifieur, ce qui revient d'une certaine manière à prendre en compte la structure du MDP dans le problème de classification et permet le calcul du vecteur de paramètres de la récompense.
 Contrairement à la plupart des algorithmes d'ARI existants, SCIRL n'a pas besoin de résoudre le problème direct.
 Cet algorithme requiert une estimation de l'attribut moyen de l'expert mais il ne s'agit là que d'un problème d'évaluation de la politique, notoirement moins difficile que la recherche de politiques optimales qu'implique la résolution du problème direct.
-Cela est discuté plus en détail Sec.~\ref{seelatedWorks}.
+Cela est discuté plus en détail section~\ref{seelatedWorks}.
 %
@@
 
-Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi-optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément qu'il est possible de contrôler le terme
+Dans cette section, nous montrons que la politique de l'expert $\pi_E$ est quasi optimale vis-à-vis de la fonction de récompense $\R_{\theta_c}$, plus précisément qu'il est possible de contrôler le terme
 $\E_{s\sim\rho_E}[v^*_{\theta_c}(s)-v^{\pi_E}_{\theta_c}(s)]$.
@@
 \begin{proof*}[Preuve du Théorème \ref{th}]
-  La démonstration repose uniquement sur la récompense $\R_{\theta_c}$, donc par soucis de clarté certains indices relatifs à la récompense sont omis des notations  
+  La démonstration repose uniquement sur la récompense $\R_{\theta_c}$, donc par souci de clarté certains indices relatifs à la récompense sont omis des notations  
   (\textit{e.g.}, $v^\pi$ pour
@@
   \end{equation}
-  Injecter les bornes des Eq.~\eqref{eroo1}
-  and~\eqref{eroo2} dans l'Eq.~\eqref{erooesidual} achève la démonstration.%, which concludes the proof.
-%  Injecting bounds of Eqs.~\eqref{eroo1}
-%  and~\eqref{eroo2} into Eq.~\eqref{erooesidual} using the
-%  decomposition of Eq.~\eqref{erooecomposition} gives the
+  Injecter les bornes des équations~\eqref{eroo1}
+  et~\eqref{eroo2} dans l'équation~\eqref{erooesidual} achève la démonstration.%, which concludes the proof.
+%  Injecting bounds of l'équation.~\eqref{eroo1}
+%  and~\eqref{eroo2} into l'équation~\eqref{erooesidual} using the
+%  decomposition of l'équation~\eqref{erooecomposition} gives the
 %  stated result:
@@
 Ce résultat montre que si l'attribut moyen de l'expert est bien estimé (dans le sens d'une faible erreur d'estumation $\epsilon_\mu$ pour les états échantillonnés selon la politique stationnaire de l'expert et pour toutes les actions) et si l'erreur de classification $\epsilon_c$ et également faible, alors l'algorithme générique proposé fournit une fonction de récompense
-$\R_{\theta_c}$ vis-à-vis de laquelle l'expert sera quasi-optimal. Un corollaire direct du Th.~\ref{th} stipule qu'avec le vrai attribut moyen de l'expert $\mu^{\pi_E}$ et un classifieur parfait 
+$\R_{\theta_c}$ vis-à-vis de laquelle l'expert sera quasi optimal. Un corollaire direct du théorème~\ref{th} stipule qu'avec le vrai attribut moyen de l'expert $\mu^{\pi_E}$ et un classifieur parfait 
 ($\epsilon_c=0$), $\pi_E$ est l'unique politique optimale pour 
@@
 %expectation can be easily computed.
-Ajoutons que le coût computationel de cette méthode est du même ordre de grandeur que l'évaluation d'une seule politique, en effet la partie coûteuse (le calcul de $(I-\gamma P_{\pi_E})^{-1}$) est partagée par toutes les composantes.
+Ajoutons que le coût computationnel de cette méthode est du même ordre de grandeur que l'évaluation d'une seule politique, en effet la partie coûteuse (le calcul de $(I-\gamma P_{\pi_E})^{-1}$) est partagée par toutes les composantes.
 
-Si le modèle est inconnu, tous les algorithmes d'apprentissage par différences temporelles peuvent être utilisés pour obtenir une estimation de l'attribut moyen de l'expert~\cite{Klei011}, comme par exemple LSTD (Least-Squares Temporal
-Differences)~\cite{Bradtk996}. Soit $\psi:\s\times \A \rightarrow
+Si le modèle est inconnu, tous les algorithmes d'apprentissage par différences temporelles peuvent être utilisés pour obtenir une estimation de l'attribut moyen de l'expert~\cite{Klei011}, comme par exemple LSTD ({\it Least-Squares Temporal
+Differences})~\cite{Bradtk996}. Soit $\psi:\s\times \A \rightarrow
 \mathbb{R}^d$ un vecteur d'attributs composé de $d$ fonctions de base
@@
 $a_i$ ne sont pas nécessairement échantillonnées selon la politique $\pi_E$
-(\textit{e.g.}, on peut utiliser des trajectoires obtenues par un agent suivant une politique $\epsilon$-gloutonne), le but étant d'obtenir une meilleure représentativité des données (des actions sous optimales devraient être essayées).
+(\textit{e.g.}, on peut utiliser des trajectoires obtenues par un agent suivant une politique $\epsilon$-gloutonne), le but étant d'obtenir une meilleure représentativité des données (des actions sous-optimales devraient être essayées).
 Soit $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
@@
 %
-Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique à imiter, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsecirnalysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
+Si l'on dispose d'un simulateur permettant notammment d'échantillonner selon la politique à imiter, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en section~\ref{subsecirnalysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$).
 Si $K$
@@
 Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ seules ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen.
-Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$, ces actions (et la dynamique résultante) n'étant représentées dans les données.
+Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$, ces actions (et la dynamique résultante) n'étant pas représentées dans les données.
 Il est heureusement possible de recourir à des heuristiques ; ce cas ne rentre pas dans l'analyse présentée en
-Sec.~\ref{subsecirnalysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en Sec.~\ref{sexperiments}.
+section~\ref{subsecirnalysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en section~\ref{sexperiments}.
 
@@
 Cette estimation n'est nécessaire que pour entraîner le classifieur, il est donc suffisant de disposer de valeurs uniquement pour les couples état-action $(s_i,a)_{1\leq i \leq N,a\in \A}$.
-En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir Sec.~\ref{seelatedWorks}).
+En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir section~\ref{seelatedWorks}).
 
@@
 \end{equation}
-La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit par Eq.~\eqref{ec_plus_heuristic}.
+La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit par l'équation~\eqref{ec_plus_heuristic}.
 
-%This is the instantiation we use in Sec.~\ref{sexperiments}, but
+%This is the instantiation we use in section~\ref{sexperiments}, but
 %our approach is more general: other choices could have been done for
@@
 %parameter vector such that $\pi_E$ is the unique optimal policy),
-%constraints of Eq.~\eqref{ep_taskar} can be satisfied for all
+%constraints of l'équation~\eqref{ep_taskar} can be satisfied for all
 %state-action couples. This means that the proposed approach is able
@@
 %
-Certains algorithmes mentionnés ne sont pas capable de renvoyer une fonction de récompense, bien qu'ils utilisent l'ARI comme étape.
+Certains algorithmes mentionnés ne sont pas capables de renvoyer une fonction de récompense, bien qu'ils utilisent l'ARI comme étape.
 Ils rentrent dans le champ généralement appelé l'apprentissage par imitation.
@@
 
-Plus proche de notre contribution, quelques approches introduisent également de la structure dans la classification~\cite{Mel010,Ratlif006}. Dans~\cite{Mel010},
+Plus proches de notre contribution, quelques approches introduisent également de la structure dans la classification~\cite{Mel010,Ratlif006}. Dans~\cite{Mel010},
 une métrique induite par le PDM est utilisée pour construire un noyau qui sera utilisé par l'algorithme de classification, permettant des améliorations par rapport à un noyau non structuré.
@@
 Estimer la valeur du sous-gradient demande des trajectoires échantillonnées selon la politique optimale pour la fonction de récompense courante.
-On peut contourner le problème grâce à l'échantillonage préférentiel.
+On peut contourner le problème grâce à l'échantillonnage préférentiel.
 Cela requiert cependant d'échantillonner des trajectoires selon une politique différente de celle de l'expert, et le problème direct se maintient au coeur de l'approche (même si sa résolution est évitée).
@@
 Nous illustrons SCIRL sur un simulateur de conduite similaire à~\cite{Abbee004,Sye00ame}.
-Le but est de conduire une voiture sur une autoroute à trois voies dont le traffic est généré aléatoirement (les sorties de route sont possibles des deux côtés).
+Le but est de conduire une voiture sur une autoroute à trois voies dont le trafic est généré aléatoirement (les sorties de route sont possibles des deux côtés).
 La voiture peut se déplacer vers la gauche ou la droite, accélérer ou ralentir et conserver sa vitesse.
@@
 Nous avons comparé SCIRL, tel qu'instancié comme décrit en
-Sec.~\ref{subseracticalApproacnstantiation}, à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au coeur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbee004} (appelé ici PIRL pour \emph{
+section~\ref{subseracticalApproacnstantiation}, à un classifieur non structuré (en utilisant le même algorithme de classification que celui placé au c{\oe}ur de SCIRL) ainsi qu'à l'algorithme de~\cite{Abbee004} (appelé ici PIRL pour \emph{
 Projection Inverse Reinforcement Learning}).
@@
 Pour chaque courbe, nous dessinons la moyenne (trait plein), l'écart-type (foncé) et les valeurs min et max (plus clair).
-La politique correspondant à une récompense aléatoire est en bleu, la politique retournée par le classifieur est en jaune et la politique optimale pour la récompense trouvée par SCIRL est en rouge.
-La ligne bleue foncée correspond à PIRL.}
+La politique correspondant à une récompense aléatoire occupe le bas des deux graphes, la politique retournée par le classifieur donne la courbe croissante du graphe de gauche et la politique optimale pour la récompense trouvée par SCIRL donne la courbe croissante du graphe de droite.
+La ligne solitaire constante à gauche correspond à PIRL.}
 \begin{minipage}[l]{0.49\linewidth}
@@
 
-La Fig.~\ref{fies} montre la performance de chacune des approches en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
+La figure~\ref{fies} montre la performance de chacune des approches en fonction du nombre de transitions de l'expert utilisées (sauf pour PIRL qui a utilisé le modèle).
 On peut voir que le classifieur ne fonctionne pas bien sur ce problème.
@@
 %Recall that for SCIRL the instantiation of
-%Sec.~\ref{subseracticalApproacnstantiation} is considered,
+%section~\ref{subseracticalApproacnstantiation} is considered,
 %with a single Monte-Carlo rollout for estimating
@@
 Nous avons introduit une nouvelle approche de résolution du problème de l'ARI en structurant un classifieur à fonction de score linéairement paramétrée avec une estimée de l'attribut moyen de l'expert.
-Il renvoie une fonction de récompense pour laquelle nous avons montré que l'expert est quasi-optimal, pour peu que l'erreur de classification soit faible et que l'estimation de l'attribut moyen de l'expert soit bonne.
+Il renvoie une fonction de récompense pour laquelle nous avons montré que l'expert est quasi optimal, pour peu que l'erreur de classification soit faible et que l'estimation de l'attribut moyen de l'expert soit bonne.
 Des méthodes pratiques pour l'estimation de cette grandeur ont été présentées et nous avons introduit une heuristique pour le cas où les seules données disponibles proviennent de l'expert, ainsi qu'une instanciation spécifique de l'algorithme SCIRL.
-Nous avons exhibé, grâce à un simulateur de conduite sur autoroute, que l'approche proposée fonctionne bien (même combinée à l'heuristique discutée), bien mieux qu'un classifieur non structuré et aussi bien que la méthode de l'état de l'art qui avait accès au modèle (et avec un coût computationel moindre).
-Nous prévoyons de poursuivre notre analyse des propritétés théoriques de SCIRL (notamment en vue de l'utilisation avec des heuristiques) et l'appliquer à des problèmes robotiques réels.
+Nous avons exhibé, grâce à un simulateur de conduite sur autoroute, que l'approche proposée fonctionne bien (même combinée à l'heuristique discutée), bien mieux qu'un classifieur non structuré et aussi bien que la méthode de l'état de l'art qui avait accès au modèle (et avec un coût computationnel moindre).
+Nous prévoyons de poursuivre notre analyse des propritétés théoriques de SCIRL (notamment en vue de l'utilisation avec des heuristiques) et de l'appliquer à des problèmes robotiques réels.
 
@@
 
-\newpage
+%\newpage
 %\small
+
+\renewcommand\bibname{Bibliographie}
 \bibliography{scirl_ria.bib}

### 2013-05-30 Manuscrit/Figures/Mountain_car_state_space.tex
Nouvelles figures (mountain car espace d'état, features et mu)

   \node (plotCenter) at (-2,-2.4){};
   \node [scale=1,anchor=south] at (plotCenter.center){\includegraphics{Figures/Mountain_car_Expert_traj_length.pdf}};
   \node[red] (plotMark) at ($(plotCenter.center) + (0.59,8)$) {\huge \bf x};
   \draw [thick,red,dotted](plotMark.center) -- ++(0,-6.6);
   \draw [thick,red,dotted](plotMark.center) -- ++(3,0);

   \node (mcSW) at (4,-1){};
  %Mountain car
  \draw  [ultra thick] (mcSW)++(0,6) sin ++(5,-5) cos ++(5,5) sin ++(5,5);
  \fill[green!50] (mcSW.center)-- ++(0,6) sin ++(5,-5) cos ++(5,5) sin ++(5,5) -- ++(0,-11) --cycle;
  %\draw (0,-1) grid (20,10);
 \node (tuture)[scale=1,rotate=55,anchor=south]at ($(mcSW)+(10.38,6.5)$) {\usebox\tuture};
 \node [scale=1,rotate=55,anchor=south,scale=1.4](speedArrow)at ($(mcSW)+(9.8,6.9)$) {\speedarrow{48}};
 \draw [very thin] (mcSW.center) -- ++(15,0);
 \foreach \x in {0,1,...,15}{
   \draw (mcSW.center) ++(\x,0) -- ++(0,.05);
 }
 \foreach \x in {-1,-0.5,0.0,0.5}{
   \draw (mcSW.center) ++($10*(\x+1,0)$) -- ++(0,.15);
   \node at ($(mcSW.center) +10*(\x+1,-.03)$) {$\x$};
 }

 \draw [thick,red,dotted]($(tuture.center)+(0,-0.15)$) -- ++(0,-6.5);
 \draw [red,<->]($(mcSW.center)+(10,-0.6)$) -- ++(0,-1) -- ($(plotMark.center)+(0,-8.2)$) -- ++(0,1);
 \draw [thick,red,decorate,decoration=brace,rotate=55] ($(speedArrow)+(-.7,0.2)$) -- ++(1.4,0);

 \draw [red,<->]($(speedArrow.center)+(-.7,0.2)$)  -- ($(plotMark.center)+(3.5,0)$);

### 2013-05-30 Plot19.py

@@ -1,0 +1,275 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+#Des graphes sur le mountain car pour le manuscrit
+from matplotlib import rc
+rc('font',**{'family':'sans-serif','sans-serif':['Helvetica']})
+## for Palatino and other serif fonts use:
+#rc('font',**{'family':'serif','serif':['Palatino']})
+rc('text', usetex=True)
+#Code from Exp5
+#Mountain Car
+from stuff import *
+from pylab import *
+from random import *
+import pickle
+import numpy
+from rl import *
+import sys
+
+NB_SAMPLES=100
+#NB_SAMPLES=int(sys.argv[1])
+RAND_STRING=str(int(rand()*10000000000))
+
+
+ACTION_SPACE=[-1,0,1]
+
+
+
+def mountain_car_next_state(state,action):
+    position,speed=state
+    next_speed = squeeze(speed+action*0.001+cos(3*position)*(-0.0025))
+    next_position = squeeze(position+next_speed)
+    if not -0.07 <= next_speed <= 0.07:
+        next_speed = sign(next_speed)*0.07
+    if not -1.2 <= next_position <= 0.6:
+        next_speed=0.
+        next_position = -1.2 if next_position < -1.2 else 0.6
+    return array([next_position,next_speed])
+
+def mountain_car_uniform_state():
+    return array([numpy.random.uniform(low=-1.2,high=0.6),numpy.random.uniform(low=-0.07,high=0.07)])
+
+mountain_car_mu_position, mountain_car_mu_speed = meshgrid(linspace(-1.2,0.6,7),linspace(-0.07,0.07,7))
+
+mountain_car_sigma_position = 2*pow((0.6+1.2)/10.,2)
+mountain_car_sigma_speed = 2*pow((0.07+0.07)/10.,2)
+
+def mountain_car_single_psi(state):
+    position,speed=state
+    psi=[]
+    for mu in zip_stack(mountain_car_mu_position, mountain_car_mu_speed).reshape(7*7,2):
+        psi.append(exp( -pow(position-mu[0],2)/mountain_car_sigma_position 
+                        -pow(speed-mu[1],2)/mountain_car_sigma_speed))
+    psi.append(1.)
+    return array(psi).reshape((7*7+1,1))
+
[…]

### 2013-06-05 Plot19.py
Nouvelles figures et Chapitre 4.

@@ -184,7 +184,6 @@ for i in range(0,7):
 
 # <codecell>
 
-#FIXME:foutre la légende à part sur les deux graphes (la même pour tous les plots, d'ailleurs)
 k = 0
 for states in zip_stack(mountain_car_mu_position,mountain_car_mu_speed):
     for s in states:
@@ -212,6 +211,52 @@ for states in zip_stack(mountain_car_mu_position,mountain_car_mu_speed):
         #plot(s[0],s[1],marker='o',color='pink')
         savefig("Mountain_car_mu_{k}.pdf".format(k=k),bbox_inches='tight')
         k+=1
+
+# <codecell>
+
+pos = linspace(-1.2,0.6,30)
+speed = linspace(-0.07,0.07,30)
+pos,speed = meshgrid(pos,speed)
+states=zip_stack(pos,speed)
+dic_mu_E = {}
+for s in states.reshape(-1,2):
+    data_MC = []
+    reward = 0
+    state = s
+    while reward == 0:
+        action = mountain_car_manual_policy(state)
+        next_state = mountain_car_next_state(state, action)
+        reward = mountain_car_reward(hstack([state, action, next_state]))
+        data_MC.append(hstack([state, action, next_state, reward]))
+        state=next_state
+    data_MC = array(data_MC)
+    GAMMAS = range(0,len(data_MC))
+    GAMMAS = array([el for el in map( lambda x: pow(GAMMA,x), GAMMAS)])
+    state_action = data_MC[0,:3]
+    state = data_MC[0,:2]
+    action = data_MC[0,2]
+    mu = None
+    if len(data_MC) > 1:
+        mu = dot( GAMMAS,squeeze(mountain_car_psi(data_MC[:,:2])))
+    else:
+        mu = squeeze(mountain_car_psi(squeeze(data_MC[:,:2])))
+    dic_mu_E[str(s)] = mu
+
+# <codecell>
+
+dic_mu_E
+f_mu = lambda s:dic_mu_E[str(s)]
+vf_mu = non_scalar_vectorize(f_mu,(2,),(50,1)) 
+Z = squeeze(vf_mu(states.reshape(-1,2)))
+for i in range(0,7):
+    for j in range(0,7):
+        k = j+i*7
+        figure(figsize=(1,1))
+        axis('off')
[…]

### 2013-06-07 DP.py
WIP sur 5 : scirl

@@ -18,19 +18,20 @@ def weighted_choice(choices):
 #end of code from StackOverflow
 class MDP:
     def __init__(self, P, R):
-        print "MDP OK3"
+        print("MDP OK3")
         self.cardSA,self.cardS = P.shape
-        self.cardA = self.cardSA/self.cardS
+        assert self.cardSA%self.cardS == 0, "P's shape is not good"
+        self.cardA = int(self.cardSA/self.cardS)
         self.gamma=0.9
         self.P = P
         assert all(f_eq(self.P.sum(axis=1), 1)), "Probability matrix unproperly conditionned"
         self.R = R
         #self.A = range(0,self.cardA)
-        print "Shape of P "+str(P.shape)
-        print "Shape of R "+str(R.shape)
-        print "Card of S "+str(self.cardS)
-        print "Card of A "+str(self.cardA)
-        print "Card of SA "+str(self.cardSA)
+        print("Shape of P "+str(P.shape))
+        print("Shape of R "+str(R.shape))
+        print("Card of S "+str(self.cardS))
+        print("Card of A "+str(self.cardA))
+        print("Card of SA "+str(self.cardSA))
         
 
     def optimal_policy(self):
@@ -50,7 +51,7 @@ class MDP:
             assert allclose(V, dot(Pi,self.R) + self.gamma*dot(dot(Pi,self.P),V)), "Bellman equation"
             Q = self.R + self.gamma*dot( self.P,V)
             Pi = self.Q2Pi( Q )
-            print "Iteration "+str(T)+", "+str(sum(Pi!=oldPi))+"\tactions changed."
+            print("Iteration "+str(T)+", "+str(sum(Pi!=oldPi))+"\tactions changed.")
             T+=1
             if( all( Pi == oldPi ) ):
                 break

### 2013-06-25 piot2012cascading_ECML/Cascading.tex
Corrections ECML pour le Camera Ready

@@
 
-IRL is one of many ways to perform Apprenticeship Learning (AL): imitating a demonstrated expert policy, without necessarily explicitely looking for the reward function.
+IRL is one of many ways to perform Apprenticeship Learning (AL): imitating a demonstrated expert policy, without necessarily explicitly looking for the reward function.
 The reward function nevertheless is of interest in its own right.
-As mentionned in \cite{russell1998learning}, its semantics can be analyzed in biology or econometrics for instance.
+As mentioned in \cite{russell1998learning}, its semantics can be analyzed in biology or econometrics for instance.
 Practically, the reward can be seen as a succinct description of a task.
 Discovering it removes the coupling that exists in AL between understanding the task and learning how to fulfill it.
-IRL allows the use of (A)DP or RL techniques to learn the task from the computed reward function.
+IRL allows the use of (A)DP or RL techniques to learn how to do the task from the computed reward function.
 A very straightforward non-IRL way to do AL is for example to use a multi-class classifier to directly learn the expert policy.
@@
 Moreover, up to the use of some heuristics (see Sec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy.
-A specific instantiation of CSI (proposed Sec.~\ref{subsubsec:subroutines}) is tested on the mountain car problem (Sec.~\ref{subsec:mountaincar}) and on a highway driving simulator (Sec.~\ref{subsec:highway}) where we compare it with a pure classification algorithm \cite{taskar2005learning} and with two recent successfull IRL methods \cite{klein2012scirl} as well as with a random baseline.
-
+A specific instantiation of CSI (proposed in Sec.~\ref{subsubsec:subroutines}) is tested on the mountain car problem (Sec.~\ref{subsec:mountaincar}) and on a highway driving simulator (Sec.~\ref{subsec:highway}) where we compare it with a pure classification algorithm \cite{taskar2005learning} and with two recent successful IRL methods \cite{klein2012scirl} as well as with a random baseline.
 Differences and similarities with existing AL or IRL approaches are succinctly discussed in Sec.~\ref{sec:related}.
-\section{Background and notations}
+\section{Background and notation}
 \label{sec:background}
-First, we introduce some general notations.
+First, we introduce some general notation.
 Let $E$ and $F$ be two non-empty sets, $E^F$ is the set of functions from $F$ to $E$.
@@
 Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\geq\beta \Leftrightarrow \forall x\in X, \alpha(x) \geq \beta(x)$.
-We will often slightly abuse the notations and consider (where applicable) most objects as if they were matrices and vectors indexed by the set they operate upon.
+We will often slightly abuse the notation and consider (where applicable) most objects as if they were matrices and vectors indexed by the set they operate upon.
 
@@
 The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$.
-Given that $P_\pi = (P(s'|s,\pi(s)))_{s,s' \in S}$ is the irreductible transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies $\rho_\pi^TP_\pi = \rho_\pi^T\textrm{, with }X^T\textrm{ being the transpose of }X$.
-The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.
+Given that $P_\pi = (P(s'|s,\pi(s)))_{s,s' \in S}$ is the transition probability matrix, the stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies $\rho_\pi^TP_\pi = \rho_\pi^T\textrm{, with }X^T\textrm{ being the transpose of }X$.
+The stationary distribution relative to the expert policy $\pi_E$ is $\rho_E$.
 
@@
 
-When the state space is too large to allow matrix representations or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
+When the state space is too large to allow matrix representations or when the transition probabilities or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.
 
@@
 For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} both of which we consider in the experimental setting.
-Other algorithms may be envisionned (see Sec.~\ref{subsubsec:subroutines}).
+Other algorithms may be envisioned (see Sec.~\ref{subsubsec:subroutines}).
 
@@
   \label{eq:rc}
-  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
+  R^C(s,a) =q(s,a) -\gamma \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
 \end{equation}
@@
 \end{equation}
-As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s_j,a_j)$.
-A regressor (a simple least-square approximator can do but other solutions could also be envisionned, see Sec.~\ref{subsubsec:subroutines}) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space.
+As $s'_j$ is sampled according to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s_j,a_j)$.
+A regressor (a simple least-square approximator can do but other solutions could also be envisioned, see Sec.~\ref{subsubsec:subroutines}) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalization of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space.
 The complete algorithm is given in Alg.~\ref{algo:cascading}.
@@
 The final regression step is a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.
-
-Being able to alleviate the ill effects of scalability or data scarcity by leveraging the wide range of techniques developped for the classification and regression problems is a strong advantage of the CSI approach.
+Being able to alleviate the ill effects of scalability or data scarcity by leveraging the wide range of techniques developed for the classification and regression problems is a strong advantage of the CSI approach.
 
@@
 \item $\Delta q = \max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$,
-    which could be normalized to $1$ without loss of generality. The range of variation of $q$ induces the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
+    which could be normalized to $1$ without loss of generality.
+The range of variation of $q$ is tied to the one of $R^C$, $\hat{R}^C$ and $V^{\pi_C}_{\hat{R^C}}$.
+What matters with these objects is the relative values for different state action couples, not the objective range.
+They can be shifted and positively scaled without consequence.
   \end{itemize}
@@
 \begin{proof}
-First let's recall some notations, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$, $R^C\in\mathbb{R}^{S\times A}$ is so that $\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s'))$,
+First let's recall some notation, $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\pi_C$ is a deterministic classifier policy so that $\forall s \in S,\pi_C(s)\in\argmax_{a\in A}q(s,a)$, $R^C\in\mathbb{R}^{S\times A}$ is so that $\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s'))$,
 and $\hat{R}^C\in\mathbb{R}^{S\times A}$ is the reward function output by the regression step.
 
-The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$.
-We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof, not to be confused with $R_E$ the unknown reward function the expert optimizes:
+The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C\in\mathbb{R}^{S\times A}$.
+We also introduce the reward function $\mathfrak{R}^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof, not to be confused with $R_E$ the unknown reward function the expert optimizes:
 \begin{equation}
-\forall (s,a)\in S\times A, R^E(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_E(s')).
+\forall (s,a)\in S\times A, \mathfrak{R}^E(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_E(s')).
 \end{equation}
 
-We no have the following vectorial equalities $
+We now have the following vectorial equalities $
 R^C_a=q_a-\gamma P_aq_{\pi_C} ;
-R^E_a=q_a-\gamma P_aq_{\pi_E};
+\mathfrak{R}^E_a=q_a-\gamma P_aq_{\pi_E};
 \epsilon^R_a=R^C_a-\hat{R}^C_a$.
@@
   \end{equation}
-So, using the notations introduced before we stated the theorem, we are able to give an upper bound to the first and third terms (recall also the notation $C_{\hat \pi_C} = C_*$):
+So, using the notation introduced before we stated the theorem, we are able to give an upper bound to the first and third terms (recall also the notation $C_{\hat \pi_C} = C_*$):
 $\rho_E^T((V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C})+(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C}))\leq\frac{1+C_*}{1-\gamma}\epsilon_R$.
@@
 \begin{equation}
-V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}=(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C})+(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})+(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}).
+V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}=(V^{\hat{\pi}_C}_{R^C}-V^{\pi_C}_{R^C})+(V^{\pi_C}_{R^C}-V^{\pi_E}_{\mathfrak{R}^E})+(V^{\pi_E}_{\mathfrak{R}^E}-V^{\pi_E}_{R^C}).
 \end{equation}
@@
 \begin{equation}
-V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}\leq(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})+(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C}).
+V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C}\leq(V^{\pi_C}_{R^C}-V^{\pi_E}_{\mathfrak{R}^E})+(V^{\pi_E}_{\mathfrak{R}^E}-V^{\pi_E}_{R^C}).
 \end{equation}
-By construction, we have $V^{\pi_C}_{R^C}=q_{\pi_C}$ and $V^{\pi_E}_{R^E}=q_{\pi_E}$, thus:
+By construction, we have $V^{\pi_C}_{R^C}=q_{\pi_C}$ and $V^{\pi_E}_{\mathfrak{R}^E}=q_{\pi_E}$, thus:
 \begin{eqnarray}
-\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})&=&\rho_E^T(q_{\pi_C}-q_{\pi_E})\\
+\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{\mathfrak{R}^E})&=&\rho_E^T(q_{\pi_C}-q_{\pi_E})\\
 &=&\sum_{s\in S}\rho_E(s)(q(s,\pi_C(s))-q(s,\pi_E(s)))[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
@@
 Using $\Delta q$, we have:
-$\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{R^E})\leq\Delta q\sum_{s\in S}\rho_E(s)[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]= \Delta q \epsilon_C$.
+$\rho_E^T(V^{\pi_C}_{R^C}-V^{\pi_E}_{\mathfrak{R}^E})\leq\Delta q\sum_{s\in S}\rho_E(s)[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}]= \Delta q \epsilon_C$.
 Finally, we also have:
 \begin{align}
-\rho_E^T(V^{\pi_E}_{R^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(R^E_{\pi_E}-R^C_{\pi_E})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\gamma P_{\pi_E}(q_{\pi_C}-q_{\pi_E}),
+\rho_E^T(V^{\pi_E}_{\mathfrak{R}^E}-V^{\pi_E}_{R^C})&=\rho_E^T(I-\gamma P_{\pi_E})^{-1}(\mathfrak{R}^E_{\pi_E}-R^C_{\pi_E})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\gamma P_{\pi_E}(q_{\pi_C}-q_{\pi_E}),
 \\
@@
 If we combine all of the results, we obtain the final bound as stated in the theorem.
-%\qed
+\qed
 \end{proof}
@@
 \\
-The attentive reader, however, will percieve that similarities stop there.
+The attentive reader, however, will perceive that similarities stop there.
 The error terms occurring in the two bounds are not related to one another.
@@
 Perhaps more subtly, the classification error and classification policy known in both papers as $\epsilon_C$ and $\pi_C$ are not the same.
-The classification policy of SCIRL is no tantamount to what is called $\pi_C$ here.
+The classification policy of SCIRL is not tantamount to what is called $\pi_C$ here.
 For SCIRL, $\pi_C$ is the greedy policy for an approximation of the value function of the expert with respect to the reward output by the algorithm.
@@
 More generally, the error terms  $\epsilon_C$ and $\epsilon_R$ can be reduced by a wise choice for the classification and regression algorithms.
-The litterature is wide enough for methods accomodating most of use cases (lack of data, fast computation, bias/variance trade-off, etc.) to be found.
+The literature is wide enough for methods accommodating most of use cases (lack of data, fast computation, bias/variance trade-off, etc.) to be found.
 Being able to leverage such common algorithms as multi-class classifiers and regressors is a big advantage of our cascading approach over existing IRL algorithms.
 
-Other differences between existing IRL or apprenticeship learning approaches and the proposed cascading algorithm are further examined Sec.~\ref{sec:related}.
+Other differences between existing IRL or apprenticeship learning approaches and the proposed cascading algorithm are further examined in Sec.~\ref{sec:related}.
 
@@
 IRL was first introduced in \cite{russell1998learning} and then formalized in \cite{ng2000algorithms}.
-Approaches summarized in \cite{neu2009training} can be seen as iteratively contructing a reward function, solving an MDP at each iteration.
+Approaches summarized in \cite{neu2009training} can be seen as iteratively constructing a reward function, solving an MDP at each iteration.
 Some of these algorithms are IRL algorithms while others fall in the Apprenticeship Learning (AL) category, as for example the projection version of the algorithm in \cite{abbeel2004apprenticeship}.
@@
 AL via classification has been proposed for example in \cite{ratliff2007imitation}, with the help of a structured margin method.
-Using the non trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
+Using the non-trivial notion of metric in an MDP, the authors of \cite{melo2010learning} build a kernel which is used
 in a classification algorithm, showing improvements compared to a
@@
 One can argue that this and not having to approximate $\mu^E$ increases the ease-of-use of CSI over SCIRL and makes for a more versatile algorithm.
-In practice, as seen in Sec.~\ref{sec:experiments}, perfomance of SCIRL and CSI are very close to one another thus CSI may be a better choice as it is easier to deploy.
-We would like to emphasize the fact that although SCIRL and CSI share similarities, there exist key differences that we just explained.
+In practice, as seen in Sec.~\ref{sec:experiments}, performance of SCIRL and CSI are very close to one another thus CSI may be a better choice as it is easier to deploy.
 Neither approach is a generalization of the other.
@@
 In this section, we empirically demonstrate the behavior of our approach.
-We begin by providing information pertaining to both benchmarks: an explanation about the amount and source of the available data, the rationale behind the heuristics we use to compensate for the dire data scarcity and a quick word about the contenders CSI is compared to are given Sec.~\ref{subsec:generalities}.
-We supply quantitative results and comparisons of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) Sec.~\ref{subsec:mountaincar} and then on a highway driving simulator (Sec.~\ref{subsec:highway}).
+We begin by providing information pertaining to both benchmarks.
+An explanation about the amount and source of the available data, the rationale behind the heuristics we use to compensate for the dire data scarcity and a quick word about the contenders CSI is compared to are given Sec.~\ref{subsec:generalities}.
+We supply quantitative results and comparisons of CSI with state-of-the art approaches on first a classical RL benchmark (the mountain car) in Sec.~\ref{subsec:mountaincar} and then on a highway driving simulator (Sec.~\ref{subsec:highway}).
 \subsection{Generalities}
@@
 \label{subsubsec:data}
-The CSI algorithm was designed not to repeatedly solve the RL problem.
+The CSI algorithm was designed to avoid repeatedly solving the RL problem.
 This feature makes it particularly well-suited to environments where sampling the MDP is difficult or costly.
@@
 
-There is no particular problem for the classification step.
-It is however not reasonable to construct the dataset $D_R = \{((s_k,a_k=\pi_E(s_k)),\hat r_k)_k\}$ only from expert transitions and expect a small regression error term $\epsilon_R$.
+It is not reasonable to construct the dataset $D_R = \{((s_k,a_k=\pi_E(s_k)),\hat r_k)_k\}$ only from expert transitions and expect a small regression error term $\epsilon_R$.
 Indeed, the dataset $D_E$ only samples the dynamics induced by the expert's policy and not the whole dynamics of the MDP.
@@
 $  r_{min} = \min_k\hat r_k - 1$.
-This heuristics intructs the regressor to associate a state-action tuple disagreeing with the expert ({\it i.e.}, $(s_k,a\neq a_k)$) with a reward strictly inferior to any of those associated with expert state action tuples ({\it i.e.}, $(s_k,a_k=\pi_E(s_k))$).
+This heuristics instructs the regressor to associate a state-action tuple disagreeing with the expert ({\it i.e.}, $(s_k,a\neq a_k)$) with a reward strictly inferior to any of those associated with expert state action tuples ({\it i.e.}, $(s_k,a_k=\pi_E(s_k))$).
 Semantically, we are asserting that disagreeing with the expert in states the expert visited is a bad idea.
@@
 The Relative Entropy (RE) method of \cite{boularias2011relative} has no such need, so we included it in our benchmarks.
-It could not, however, work with the small amount of data we provided SCIRL and CSI with, so to allow for importance sampling, we created another dataset $D_{random}$ that was used by RE but not by SCIRL nor CSI.
+It could not, however, work with the small amount of data we provided SCIRL and CSI with, and so to allow for importance sampling, we created another dataset $D_{random}$ that was used by RE but not by SCIRL nor CSI.
 
 Finally, the classification policy $\pi_C$ output by the classification step of CSI was evaluated as well.
-Comparing classification and IRL algorithms makes no sense if the object of interest is the reward itself as can be envisionned in a biological or economical context.
+Comparing classification and IRL algorithms makes no sense if the object of interest is the reward itself as can be envisioned in a biological or economical context.
 It is however sound to do so in an imitation context where what matters is the performance of the agent with respect to some objective criterion.
 Both experiments use such a criterion.
-Classification algorithms don't have to optimize any reward, the classification policy can directly be used in the environment.
+Classification algorithms don't have to optimize any reward since the classification policy can directly be used in the environment.
 IRL algorithms on the other hand output a reward that must then be plugged in an RL or DP algorithm to get a policy.
-In each benchmark we used the same (benchmark-dependant) algorithm to get a policy from each of the three rewards output by SCIRL, CSI and RE.
+In each benchmark we used the same (benchmark-dependent) algorithm to get a policy from each of the three rewards output by SCIRL, CSI and RE.
 It is these policies whose performance we show.
-Finding a policy from a reward is of course a non-trivial problem that should not be swept under the rug, nevertheless we choose not to concern ourselves with it here as we wish to focus on IRL algorithms, not RL or DP algorithms.
+Finding a policy from a reward is of course a non-trivial problem that should not be swept under the rug; nevertheless we choose not to concern ourselves with it here as we wish to focus on IRL algorithms, not RL or DP algorithms.
 In this regard, using a classifier that directly outputs a policy may seem a much simpler solution, but we hope that the reader will be convinced that the gap in performance between classification and IRL is worth the trouble of solving the RL problem (once and for all, and not repeatedly as a subroutine like some other IRL algorithms).
 
-We do not compare CSI to other IRL algorithms requiring repeatedly solving the MDP: as we would need to provide them with enough data to do so, the comparison makes little sense.
+We do not compare CSI to other IRL algorithms requiring repeatedly solving the MDP. As we would need to provide them with enough data to do so, the comparison makes little sense.
 \subsubsection{Supervised Steps}
 \label{subsubsec:subroutines}
-The cascading algorithm can be instanciated with some standard classification algorithms and any regression algorithm.
+The cascading algorithm can be instantiated with some standard classification algorithms and any regression algorithm.
 The choice of such subroutines may be dictated by the kind and amount of available data, by ease of use or by computational complexity, for example.
@@
 We referred in Sec.\ref{sec:algo} to \emph{score-function based multi-class classifiers} and explained how the classification rule is similar to the greedy mechanism that exists between an optimal action-value function and an optimal policy in an MDP.
-Most classification algorithm can be seen as such a classifier.
-In a simple $k$-nearest neighboor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighboors of $s$.
+Most classifications algorithms can be seen as such a classifier.
+In a simple $k$-nearest neighbor approach, for example, the score function $q(s,a)$ is the number of elements of class $a$ among the $k$-nearest neighbors of $s$.
 The generic M-SVM model makes the score function explicit (see \cite{guermeur2011generic}) (we use a SVM in the mountain car experiment Sec.~\ref{subsec:mountaincar}).
@@
 It is possible to get imaginative in the last step.
-For example, using a Gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable (nonwithstanding a nontrivial amount of work) the use of reward-uncertain reinforcement learning \cite{regan2011robust}.
+For example, using a Gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable (notwithstanding a nontrivial amount of work) the use of reward-uncertain reinforcement learning \cite{regan2011robust}.
 
-Our complete instanciation of CSI is summed up in Alg.~\ref{alg:CSIimp}.
+Our complete instantiation of CSI is summed up in Alg.~\ref{alg:CSIimp}.
 \begin{algorithm}%[H]
     %\small
-  \caption{A CSI instanciation with heuristics}
+  \caption{A CSI instantiation with heuristics}
   \label{alg:CSIimp}
@@
 \emph{\textbf{Construct}} the training set  $D_R = \{((s_{j}=s_k,a_{j}=a_k),\hat r_j)_j\}\cup\{((s_j=s_k,a),r_{min})_{j;\forall a\neq \pi_E(s_j) = a_k}\}$ \;\\
-\emph{\textbf{Learn}} a reward function $\hat R^C$ from the trainsing set $D_R$\;\\
+\emph{\textbf{Learn}} a reward function $\hat R^C$ from the training set $D_R$\;\\
 \emph{\textbf{Output}} the reward function $\hat R^{C}: (s,a) \mapsto \omega^T \phi(s,a)$ \;
@@
 When training an RL agent, the reward is, for example, $1$ if the car's position is greater than $0.5$ and $0$ anywhere else.
-The expert was a very simple hand crafted policy in which the agent uses the power of the car to go in the direction it already moves ({\it i.e.}, go left when the speed is negative, right when it is positive).\\
+The expert policy was a very simple hand crafted policy that uses the power of the car to go in the direction it already moves ({\it i.e.}, go left when the speed is negative, right when it is positive).\\
 
@@
   }
-  \caption{Student or Welch test of mean equality (depending on wether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the mountain car (\ref{tab:mc}) and the highway driving simulator (\ref{tab:highway}).
+  \caption{Student or Welch test of mean equality (depending on whether a Bartlett test of variance equality succeeds) $p$-values for CSI and SCIRL on the mountain car (\ref{tab:mc}) and the highway driving simulator (\ref{tab:highway}).
 High values ($>1.0\e{-02}$) means that the hypothesis that the means are equal cannot be rejected.}
@@
 %FIXME:METTRE UNE CAPTURE DECRAN.
-The reward function $R_E$ the expert is trained by a DP algorithm on makes it go as fast as possible (high reward) while avoiding collisions (harshly penalised) and avoiding going off-road (moderately penalised).
+The reward function $R_E$ the expert is trained by a DP algorithm on makes it go as fast as possible (high reward) while avoiding collisions (harshly penalized) and avoiding going off-road (moderately penalised).
 Any other situation receives a null reward.
@@
 
-Furthermore, the classifier whose results are diplayed here is the output of the first step of the algorithm.
-The classification performance is obviously not that good, this hints to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound lets us expect.
+Furthermore, the classifier whose results are displayed here is the output of the first step of the algorithm.
+The classification performance is obviously not that good, which points to the fact that our algorithm may be empirically more forgiving of classification errors than our theoretical bound lets us expect.
 
@@
 The expert is theoretically shown to be near-optimal for the reward function the proposed algorithm outputs, given small classification and regression errors.
-Practical examples of classifiers and regressors have been given, and two combinations have been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), provided the use of simple heuristics.
-On both benchmarks, our algorithm is shown to outperform other state-of-the-art approaches (although SCIRL catches up on the mountain car).
+Practical examples of classifiers and regressors have been given, and two combinations have been empirically (on two classic benchmarks) shown to be very resilient to dire lack of data on the input (only data from the expert was used to retrieve the reward function), with the help of simple heuristics.
+On both benchmarks, our algorithm is shown to outperform other state-of-the-art approaches although SCIRL catches up on the mountain car.
 We plan on deepening the analysis of the theoretical properties of our approach and on applying it to real world robotics problems.

### 2013-06-26 piot2012cascading_ECML/Cascading.tex
Modifications après revues ECML

@@
 
+\setlength\floatsep{1.25\baselineskip plus 3pt minus 2pt}
+\setlength\textfloatsep{1.25\baselineskip plus 3pt minus 2pt}
+\setlength\intextsep{1\baselineskip plus 3pt minus 2 pt}
+
 \begin{document}
@@
 \authorrunning{Klein, Piot, Geist and Pietquin}
-\author{Edouard Klein\inst{1,}\inst{2}, Bilal Piot\inst{2,}\inst{3}, Matthieu Geist\inst{2}, Olivier Pietquin\inst{2,}\inst{3}}
+\author{Edouard Klein\inst{1,}\inst{2}, Bilal Piot\inst{2,}\inst{3}, Matthieu Geist\inst{2}, Olivier Pietquin\inst{2,}\inst{3}\thanks{The research leading to these results has received funding from the European Union Seventh Framework Programme (FP7/2007-2013) under grant agreement n\textdegree270780.}}
 
 \institute{
- 1. ABC Team\\
- LORIA-CNRS, France.\\
+ ABC Team\, LORIA-CNRS, France.\\
 \and
-2. Supélec\\
- IMS-MaLIS Research group, France\\
+Supélec, IMS-MaLIS Research group, France\\
  \texttt{firstname.lastname@supelec.fr}\\
 \and
-3. UMI 2958 (GeorgiaTech-CNRS)\\
-France\\}
+UMI 2958 (GeorgiaTech-CNRS), France\\
+}
 
-\date{\today}
+%\date{\today}
 
 \maketitle
+%\renewcommand*\abstractname{Acknowledgement.}
+%\begin{abstract}
+%The research leading to these results has received funding from the European Union Seventh Framework Programme (FP7/2007-2013) under grant agreement n\textdegree270780.
+%\end{abstract}
 
+\renewcommand*\abstractname{Abstract.}
 \begin{abstract}
@@
 For example, using a Gaussian process regressor \cite{rasmussen2006gaussian} that outputs both expectation and variance can enable (notwithstanding a nontrivial amount of work) the use of reward-uncertain reinforcement learning \cite{regan2011robust}.
-
 Our complete instantiation of CSI is summed up in Alg.~\ref{alg:CSIimp}.
@@
 The performance criterion for a policy $\pi$ is the mean (over the uniform distribution) value function with respect to $R_E$ : $\mathbf{E}_{s\sim\mathcal{U}}[V^{\pi}_{R_E}(s)]$.
-
 Expert performance averages to $7.74$ ; we also show the natural random baseline that consists in drawing a random reward vector (with a uniform law) and training an agent on it.
@@
 This is a clear illustration of the fact that using the Bellman equation to construct the data fed to the regressor and outputting not a policy, but a reward function that can be optimized on the MDP truly makes use of the information that the transitions $(s,a,s')$ bear (we recall that the classifier only uses $(s,a)$ couples).
-
 Furthermore, the classifier whose results are displayed here is the output of the first step of the algorithm.

### 2013-07-01 Manuscrit/titlepage.tex
Title page

\renewcommand\maketitle{
\newgeometry{left=3cm}
\thispagestyle{empty} \noindent
\vspace*{-2cm}

\begin{tabular}{l r}
\begin{minipage}{0.60\linewidth}
  \includegraphics[scale=.35]{figures/universite.png}
\end{minipage} &
\begin{minipage}{0.40\linewidth}
   \begin{flushright}
     \normalsize{Ecole Doctorale IAEM}\\
     \normalsize{Département Informatique}
\end{flushright}
\end{minipage}
\end{tabular}

\vspace{2.5cm}

\begin{center}
  \Large{Thèse de Doctorat}\\
  \large{\textbf{Edouard Klein}}\\
  \vspace{3cm}
  \LARGE{\textbf{Contributions à l'apprentissage par renforcement inverse}} \\
  \vspace{5cm}
\end{center}

\vspace{0.5cm}
% \hfill
\vfill

\noindent \textbf{Membres du jury}\noindent\\ \noindent
\begin{tabularx}{18cm}{p{2.6cm}p{3.6cm}p{10.5cm}}
Examinateurs & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy (directeur de thèse)}\\
 &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz (codirecteur de thèse)}\\
 &Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
 &Manuel Lopes&\small{Equipe Flowers, INRIA -- Bordeaux}\\
 Raporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
 &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec} 
\end{tabularx}

\vspace{5em}

\begin{center}
\begin{tabular}{c c}
\begin{minipage}{0.50\linewidth}
  \begin{center}
    \includegraphics[scale=.25]{figures/supelec.png}\\
    \textbf{Equipe IMS-MaLIS}\\
    \textbf{Supélec -- Metz}\\
  \end{center}
\end{minipage} &

\begin{minipage}{0.50\linewidth}
  \begin{center}
   \includegraphics[scale=.25]{figures/loria.png}\\
   \textbf{Equipe ABC} \\
   \textbf{Loria-CNRS -- Nancy}\\
  \end{center}
\end{minipage}
\end{tabular}
\end{center}
\vspace{-2cm}
\restoregeometry
% \begin{center}
% \normalsize{Metz, septembre 2013}\\
% \end{center}

%é
}

### 2013-07-01 Exp11.py

@@ -55,7 +55,7 @@ legend()
 xlabel(r"Number of samples from the expert")
 ylabel("Average length of episode")
 grid()
-savefig("Exp11.pdf")
+savefig("Exp11.pdf",transparent=True)
 
 # <codecell>
 

### 2013-07-23 Manuscrit/titlepage.tex
Title page

@@
 
-\vspace{0.5cm}
-% \hfill
-\vfill
+%\vspace{0.5cm}
 
 \noindent \textbf{Membres du jury}\noindent\\ \noindent
-\begin{tabularx}{18cm}{p{2.6cm}p{3.6cm}p{10.5cm}}
-Examinateurs & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy (directeur de thèse)}\\
- &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz (codirecteur de thèse)}\\
- &Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
+\begin{tabularx}{18cm}{p{3cm}p{4cm}p{10.5cm}}
+Directeurs de thèse & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy}\\
+ &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
+ Raporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
+ &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec}\\ 
+Examinateurs &Sylvain Contassot-Vivier&\small{Loria-Université de Lorraine -- Nancy}\\
  &Manuel Lopes&\small{Equipe Flowers, INRIA -- Bordeaux}\\
- Raporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
- &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec} 
+&Guillaume Laurent&\small{Institut FEMTO-ST / ENSMM, Besançon}\\
+&Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}
 \end{tabularx}
 
-\vspace{5em}
+\vspace{4em}
 

### 2013-07-23 Plot15.py

@@ -25,18 +25,18 @@ rc('text', usetex=True)
 rcParams['text.usetex'] = True
 rcParams['font.family'] = 'serif'
 rcParams['legend.fontsize'] = 'medium'
-plot(Abcissas,map(mean,Y_CSI),"ro-",mec="r", mfc="w", lw=5, mew=3, ms=10, label=r"CSI")
-plot(Abcissas,map(mean,Y_SCIRL),"v--",color='orange',mec="orange", mfc="w", lw=3, mew=3, ms=9, label="SCIRL")
-plot(Abcissas,map(mean,Y_RE),"^-.",color='blue',mec="blue", mfc="w", lw=3, mew=3, ms=9, label=r"Relative Entropy")
-plot(Abcissas,map(mean,Y_Classif),"s:",color='green',mec="green", mfc="w", lw=3, mew=3, ms=9, label="Classification")
-plot(Abcissas,map(mean,Y_random),"D-",color='gray',mec="gray", mfc="w", lw=3, mew=3, ms=9, label="Random")
+plot(Abcissas,[x for x in map(mean,Y_CSI)],"ro-",mec="r", mfc="w", lw=5, mew=3, ms=10, label=r"CSI")
+plot(Abcissas,[x for x in map(mean,Y_SCIRL)],"v--",color='orange',mec="orange", mfc="w", lw=3, mew=3, ms=9, label="SCIRL")
+plot(Abcissas,[x for x in map(mean,Y_RE)],"^-.",color='blue',mec="blue", mfc="w", lw=3, mew=3, ms=9, label=r"Relative Entropy")
+plot(Abcissas,[x for x in map(mean,Y_Classif)],"s:",color='green',mec="green", mfc="w", lw=3, mew=3, ms=9, label="Classification")
+plot(Abcissas,[x for x in map(mean,Y_random)],"D-",color='gray',mec="gray", mfc="w", lw=3, mew=3, ms=9, label="Random")
 #axis([0,310,40,250])
 legend(bbox_to_anchor=(1, 0.85))
 xlabel(r"Number of samples from the expert")
 ylabel("Average performance")
 grid()
 axis([0,410,-2.2,8])
-savefig("Exp14.pdf")
+savefig("Exp14.pdf",transparent=True)
 
 # <codecell>
 
@@ -53,22 +53,29 @@ print t
 
 # <codecell>
 
+import scipy
[…]

### 2013-07-25 Manuscrit/titlepage.tex
Typo title page

--- 2013-07-01 Manuscrit/titlepage.tex
@@
 
-\vspace{0.5cm}
-% \hfill
-\vfill
+%\vspace{0.5cm}
 
 \noindent \textbf{Membres du jury}\noindent\\ \noindent
-\begin{tabularx}{18cm}{p{2.6cm}p{3.6cm}p{10.5cm}}
-Examinateurs & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy (directeur de thèse)}\\
- &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz (codirecteur de thèse)}\\
- &Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
+\begin{tabularx}{18cm}{p{3cm}p{4cm}p{10.5cm}}
+Directeurs de thèse & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy}\\
+ &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
+ Rapporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
+ &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec}\\ 
+Examinateurs &Sylvain Contassot-Vivier&\small{Loria-Université de Lorraine -- Nancy}\\
  &Manuel Lopes&\small{Equipe Flowers, INRIA -- Bordeaux}\\
- Raporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
- &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec} 
+&Guillaume Laurent&\small{Institut FEMTO-ST / ENSMM, Besançon}\\
+&Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}
 \end{tabularx}
 
-\vspace{5em}
+\vspace{4em}
 

### 2013-07-26 Manuscrit/titlepage.tex
Version soumise "pour a dernière fois" à M&O

--- 2013-07-01 Manuscrit/titlepage.tex
@@
 
-\vspace{0.5cm}
-% \hfill
-\vfill
+%\vspace{0.5cm}
 
 \noindent \textbf{Membres du jury}\noindent\\ \noindent
-\begin{tabularx}{18cm}{p{2.6cm}p{3.6cm}p{10.5cm}}
-Examinateurs & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy (directeur de thèse)}\\
- &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz (codirecteur de thèse)}\\
- &Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
+\begin{tabularx}{18cm}{p{3.5cm}p{4cm}p{10.5cm}}
+Directeur de thèse & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy}\\
+Co-directeur de thèse &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
+ Rapporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
+ &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec}\\ 
+Examinateurs &Sylvain Contassot-Vivier&\small{Loria-Université de Lorraine -- Nancy}\\
  &Manuel Lopes&\small{Equipe Flowers, INRIA -- Bordeaux}\\
- Raporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
- &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec} 
+&Guillaume Laurent&\small{Institut FEMTO-ST / ENSMM, Besançon}\\
+&Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}
 \end{tabularx}
 
-\vspace{5em}
+\vspace{4em}
 

### 2013-08-01 Manuscrit/titlepage.tex
Titres et fonctions sur la page de garde

--- 2013-07-23 Manuscrit/titlepage.tex
@@
 \renewcommand\maketitle{
-\newgeometry{left=3cm}
 \thispagestyle{empty} \noindent
+\newgeometry{left=1cm}
 \vspace*{-2cm}
@@
   \LARGE{\textbf{Contributions à l'apprentissage par renforcement inverse}} \\
-  \vspace{5cm}
+  \vspace{4cm}
 \end{center}
@@
 \noindent \textbf{Membres du jury}\noindent\\ \noindent
-\begin{tabularx}{18cm}{p{3cm}p{4cm}p{10.5cm}}
-Directeurs de thèse & Yann Guermeur &\small{Equipe ABC, Loria-CNRS -- Nancy}\\
- &Matthieu Geist &\small{Equipe IMS-MaLIS, Supélec -- Metz}\\
- Raporteurs&Rachid Alami &\small{LAAS /  CNRS -- Toulouse}\\
- &Brahim  Chaib-draa&\small{DAMAS, Université Laval -- Québec}\\ 
-Examinateurs &Sylvain Contassot-Vivier&\small{Loria-Université de Lorraine -- Nancy}\\
- &Manuel Lopes&\small{Equipe Flowers, INRIA -- Bordeaux}\\
-&Guillaume Laurent&\small{Institut FEMTO-ST / ENSMM, Besançon}\\
-&Olivier Pietquin &\small{Equipe IMS-MaLIS, Supélec -- Metz}
+\begin{tabularx}{18cm}{p{3.5cm}p{4cm}p{10.5cm}}
+Directeur de thèse & Dr. Yann Guermeur &\small{Directeur de recherche, Equipe ABC, Loria-CNRS -- Nancy}\\
+Co-directeur de thèse &Dr. Matthieu Geist &\small{Directeur de recherche, Equipe IMS-MaLIS, Supélec -- Metz}\\
+ Rapporteurs&Dr. Rachid Alami &\small{Directeur de recherche, LAAS /  CNRS -- Toulouse}\\
+ &Prof. Brahim  Chaib-draa&\small{Professeur, DAMAS, Université Laval -- Québec}\\ 
+Examinateurs &Prof. Sylvain Contassot-Vivier&\small{Professeur, Loria-Université de Lorraine -- Nancy}\\
+&Dr. Guillaume Laurent&\small{Maître de conférences, Institut FEMTO-ST / ENSMM -- Besançon}\\
+ &Dr. Manuel Lopes&\small{Chargé de recherche, Equipe Flowers, INRIA -- Bordeaux}\\
+&Dr. Olivier Pietquin &\small{Directeur de recherche, Equipe IMS-MaLIS / UMI 2958, Supélec -- Metz}
 \end{tabularx}

### 2013-08-03 DP_mu.py
Blop

@@ -1,0 +1,33 @@
+
+from numpy import *
+import scipy
+
+#FIXME: Cette fonction n'a rien a faire la, mais je la rangerai plus tard
+def monte_carlo_mu( traj, gamma, psi ):
+    #try:
+    assert( len( traj.shape ) == 2 )
+    #except AssertionError: #Trajectory with only one transition
+    #    return psi( traj )
+    mGamma = map( lambda i: pow( gamma,i),range(0,traj.shape[0]) )
+    mPsi = map( psi, traj )
+    mDiscountedPsi = map( lambda gamma, psi : gamma*psi, mGamma, mPsi )
+    return sum( mDiscountedPsi, 0 )
+    
+
+
+g_fGamma = 0.9
+
+def DP_mu( pi, Phi ):
+    "Returns the matrix corresponding to the feature expectation of the given policy."
+    global g_fGamma
+    answer = scipy.rand( Phi.shape[0], Phi.shape[1] )
+    changed = True
+    while changed:
+        new_answer = Phi + g_fGamma*dot(pi,answer)
+        diff = sum( abs( new_answer - answer ) )
+        answer = new_answer
+        if diff > 0.001:
+            changed = True
+        else:
+            changed = False
+    return answer

### 2013-08-03 Exp12_nxn.py

@@ -1,0 +1,212 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+#SCIRL on the Highway
+from DP import *
+from stuff import *
+from pylab import *
+from random import *
+import numpy
+from rl import *
+P = genfromtxt("Highway_P.mat")
+R = genfromtxt("Highway_R.mat")
+Gamma = 0.9
+ACTION_SPACE = range(0,5)
+
+# <codecell>
+
+%load_ext autoreload
+%autoreload 2
+
+# <codecell>
+
+Highway = MDP(P,R)
+mPi_E, V_E, Pi_E = Highway.optimal_policy()
+
+# <codecell>
+
+rho = lambda : int(rand()*729) #uniform distribtion over S
+l_D_E = [array(Highway.D_func(Pi_E, 1, 10,  rho)) for i in range(0,10)]
+D_E = vstack(l_D_E)
+
+# <codecell>
+
+#SCIRL
+#Precomputing mu with MC and heuristics
+def MC_mu(episode):
+    answer = zeros((729,1))
+    for i in range(0,len(episode)):
+        answer[episode[i]] += pow(Gamma,i)
+    return answer
+
+feature_expectations_MC = {}
+d_mu_MC = {}
+for episode in l_D_E:
+    for start_index in range(0,len(episode)):
+        data=episode[start_index:,:2]
+        state_action = data[0,:2]
+        state = data[0,0]
+        action = data[0,1]
+        mu = MC_mu(data)
+        try:
+            d_mu_MC[str(state_action)].append(mu)
+        except KeyError:
+            d_mu_MC[str(state_action)] = [mu]
+        for other_action in [a for a in ACTION_SPACE if a != action]:
+            state_action=hstack([state,other_action])
+            try:
+                d_mu_MC[str(state_action)].append(Gamma*mu)
+            except KeyError:
+                d_mu_MC[str(state_action)]=[Gamma*mu]
+for sa in d_mu_MC.keys():
+    feature_expectations_MC[sa] = mean(d_mu_MC[sa],axis=0)
+ 
+
+# <codecell>
+
+#Structured Classifier
+class GradientDescent(object):
+    
+   def alpha( self, t ):
+      raise NotImplementedError, "Cannot call abstract method"
+
+   theta_0=None
+   Threshold=None
+   T = -1
+   sign = None
+        
+   def run( self, f_grad, f_proj=None, b_norm=False ): #grad is a function of theta
[…]

### 2013-08-03 Manuscrit/Resume.tex
Blip

\documentclass[11pt]{article}
\pagestyle{empty}
\usepackage[francais]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{url}

\usepackage{graphicx}
\usepackage{fancybox}

\setlength{\textwidth}{160mm}

\setlength{\oddsidemargin}{0mm}
% \setlength{\evensidemargin}{30mm}

\begin{document}
\section*{Résumé}
Cette thèse, intitulée "Contributions à l'apprentissage par renforcement inverse", fournit trois contributions majeures au domaine.
La première est une méthode d'estimation de l'attribut moyen,
une quantité exploitée par la grande majorité des approches constituant
l'état de l'art. Elle a permis d'étendre ces approches au cadre {\em batch}
et {\em off-policy}.
La seconde contribution majeure est un algorithme d'apprentissage par
renforcement inverse, 
{\em structured classification for inverse reinforcement learning}
(SCIRL), qui relâche une contrainte standard du domaine, la résolution répétée
d'un processus décisionnel de Markov en introduisant la structure temporelle (par le biais de l'attribut moyen) de ce processus dans
un algorithme de classification structurée. 
Les garanties théoriques qui lui sont attachées et ses bonnes performances
en pratique ont permis sa présentation dans une conférence internationale
prestigieuse~: NIPS.
Enfin, la troisième contribution est constituée par la méthode
{\em cascaded supervised learning for inverse reinforcement learning}
(CSI) consistant à apprendre le comportement de l'expert par une méthode supervisée puis à introduire la structure temporelle du MDP par une régression mettant en jeu la fonction de score du classifieur utilisé. Cette méthode offre des garanties théoriques de même nature que celles
de SCIRL tout en présentant l'aventage d'utiliser des composants standards pour la classification et la régression,
ce qui simplifie la mise en \oe uvre. Ce travail sera présenté dans une autre
conférence internationale prestigieuse~: ECML.\\
{\bf Mots-clefs :} Apprentissage par renforcement inverse, Imitation, Apprentissage à partir de démonstrations, Processus décisionnels de Markov, Prédiction structurée.
\section*{Summary}
This thesis, "Contributions à l'apprentissage par renforcement inverse", brings three major contributions to the community.
The first one is a method for estimating the feature expectation, a quantity involved in most of state-of-the-art approaches which were thus extended to a batch off-policy setting.
The second major contribution is an Inverse Reinforcement Learning algorithm, 
{\em structured classification for inverse reinforcement learning}
(SCIRL), which relaxes a standard constraint in the field, the repeated solving of a Markov Decision Process, by introducing the temporal structure (using the feature expectation) of this process into a structured margin classification algorithm.
The afferent theoritical guarantee and the good empirical performance it exhibited allowed it to be presentend in a good international conference: NIPS.
Finally, the third contribution is {\em cascaded supervised learning for inverse reinforcement learning}
(CSI) a method consisting in learning the expert's behavior via a supervised learning approach, and then introducing the temporal structure of the MDP via a regression involving the score function of the classifier. This method presents the same type of theoretical guarantee as SCIRL, but uses standard components for classification and regression, which makes its use simpler. This work will be presented in another good international conference: ECML.

\end{document}

### 2013-08-03 Plot20.py

@@ -1,0 +1,198 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+
+from numpy import *
+import scipy
+import sys
+from DP import *
+from DP_mu import *
+from RWC import *
+import random
+
+class GridWorld(MDP):
+    card_S = 25
+    card_A = 4
+    
+    def Sgenerator(self):
+        for x in range(0,5):
+            for y in range(0,5):
+                yield [x,y]
+    S = None
+    A = range(0,4)
+    
+    def s_index(self, state ):
+        x = state[0]
+        y = state[1]
+        index = y*5 + x
+        return int(index)
+
+    def sa_index(self, state, action ):
+        x = state[0]
+        y = state[1]
+        a = action
+        index = y*5*4 + x*4 + a
+        return int(index)
+    
+    def P(self, a):
+        "Returns the matrix of transition probability for action a."
+        P_a = zeros((5*5,5*5))
+        for state in self.Sgenerator():
+            current_index = self.s_index( state )
+            states = next_states( state, a )
+            for sdash,w in states:
+                index_dash = self.s_index( sdash )
+                P_a[current_index, index_dash] += w
+        return P_a
+
+    def R(self):
+        reward = zeros((5*5*4,1))
+        indices = [self.sa_index([4,0],a) for a in self.A]
+        reward[indices] = 1.
+        return reward
+
+    
+    def __init__(self):
+        self.S = [s for s in self.Sgenerator()]
+        m_P = vstack([self.P(a) for a in self.A])
+        R = self.R()
+        super().__init__(m_P,R)
+
+
+
+def psi( s ):
+    answer = zeros(( 5*5, 1 ))
+    answer[ s_index( s )] = 1.
+    return answer
+
+def phi( s,a ):
+    answer = zeros(( 5*5*4, 1 ))
+    answer[ sa_index( s, a )] = 1.
+    return answer 
+
+
+def next_states( state, action ):
+    "Returns the list [[s,w],...] of next possible states and associated probability"
+    x = state[0]
+    y = state[1]
+    x_south = x
+    y_south = y + 1 if y!=4 else 4
[…]

### 2013-08-03 Plot21.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+%load_ext autoreload
+%autoreload 2
+from DP import *
+from stuff import *
+from pylab import *
+from random import *
+import numpy
+from rl import *
+
+def next_states( state, action ):
+    "Returns the list [[s,w],...] of next possible states and associated probability"
+    x = state[0]
+    y = state[1]
+    x_south = x
+    y_south = y + 1 if y!=4 else 4
+    x_west = x - 1 if x!=0 else 0
+    y_west = y
+    x_east = x + 1 if x!=4 else 4
+    y_east = y 
+    x_north = x
+    y_north = y - 1 if y!=0 else 0
+    weights = zeros((1,4)) + .1
+    weights[0,action] = 0.7
+    assert abs(sum(weights) - 1.) < 0.00001
+    states = map( array, [[x_south,y_south],[x_west,y_west],[x_east,y_east],[x_north,y_north]]) #Same order as specified in the textual description of the action space
+    return zip( states, weights[0] )
+def P( a ):
+    "Returns the matrix of transition probability for action a."
+    P_a = zeros((5*5,5*5))
+    for state in Sgenerator():
+        current_index = s_index( state )
+        states = next_states( state, a )
+        for sdash,w in states:
+            index_dash = s_index( sdash )
+            P_a[current_index, index_dash] += w
+    return P_a
+def Sgenerator( ):
+    for x in range(0,5):
+        for y in range(0,5):
+            yield [x,y]
+
+def s_index( state ):
+    x = state[0]
+    y = state[1]
+    index = y*5 + x
+    return int(index)
+def sa_index( state, action ):
+    x = state[0]
+    y = state[1]
+    a = action
+    index = y*5*4 + x*4 + a
+    return int(index)
+P = vstack([P(a) for a in range(0,4)])
+def R():
+    reward = zeros((5*5*4,1))
+    indices = [sa_index([4,0],a) for a in range(0,4)]
+    reward[indices] = 1.
+    return reward
+GW = MDP(P,R())
+GW.optimal_policy()
+

### 2013-08-03 RWC.py

@@ -1,0 +1,13 @@
+
+#code from http://stackoverflow.com/questions/3679694/a-weighted-version-of-random-choice
+import random
+
+def weighted_choice(choices):
+   total = sum(w for c,w in choices)
+   r = random.uniform(0, total)
+   upto = 0
+   for c, w in choices:
+      if upto+w > r:
+         return c
+      upto += w
+   assert False, "Shouldn't get here"

### 2013-08-09 Manuscrit/titlepage.tex
Envoi aux trois

@@
 Directeur de thèse & Dr. Yann Guermeur &\small{Directeur de recherche, Equipe ABC, Loria-CNRS -- Nancy}\\
-Co-directeur de thèse &Dr. Matthieu Geist &\small{Directeur de recherche, Equipe IMS-MaLIS, Supélec -- Metz}\\
+Co-directeur de thèse &Dr. Matthieu Geist &\small{Enseignant-chercheur, Equipe IMS-MaLIS, Supélec -- Metz}\\
  Rapporteurs&Dr. Rachid Alami &\small{Directeur de recherche, LAAS /  CNRS -- Toulouse}\\
@@
  &Dr. Manuel Lopes&\small{Chargé de recherche, Equipe Flowers, INRIA -- Bordeaux}\\
-&Dr. Olivier Pietquin &\small{Directeur de recherche, Equipe IMS-MaLIS / UMI 2958, Supélec -- Metz}
+&Dr. Olivier Pietquin &\small{Enseignant-chercheur, Equipe IMS-MaLIS / UMI 2958, Supélec -- Metz}
 \end{tabularx}

### 2013-08-26 piot2012cascading_ECML/poster.tex
Poster ECML, proposition

\documentclass[xcolor=x11names,12pt]{beamer}
%\usetheme[secheader]{Boadilla} 
\mode<presentation>{\usetheme{I6pd2}}

\usepackage{multicol} 
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{grffile}
\usepackage{listings}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{pifont}
\usepackage{slashbox}
\usepackage{marvosym}
\usepackage{stmaryrd}
\usepackage[ruled]{algorithm2e}
\newcommand{\classifpolicy}{\pi^C}
\newcommand{\classifscorefunc}{q}
\newcommand{\expertdistrib}{\rho_E}
\newcommand{\apprRc}{\hat{R}^C}
\newcommand{\Rc}{R^C}
\newcommand{\actionspace}{\mathcal{A}}
\newcommand{\statespace}{\mathcal{S}}
\newcommand{\discount}{\gamma}
\newcommand{\actionBis}{a}
\newcommand{\datasetindex}{i}
\newcommand{\nbsamples}{N}
\newcommand{\rsample}{\hat{r}}
\newcommand{\state}{s}
\newcommand{\optimalpolicy}[1]{\pi^*_{#1}}
\newcommand{\expertpolicy}{\pi_E}
\newcommand{\expertreward}{R^E}
\newcommand{\satrace}[1]{D^{#1}_{sa}}
\newcommand{\ccoeffpi}[1]{C_{#1}}
\newcommand{\rlvalue}[2]{V^{#1}_{#2}}
\newcommand{\sastrace}[1]{D^{#1}_{sas}}
\newcommand{\expectationknowing}[2]{\mathbf{E}\left[\left.#1\right|#2\right]}
\newcommand{\transprobfunceval}[3]{p\left(\left.#3\right|#1,#2\right)}
\newcommand{\quality}[2]{Q^{#1}_{#2}}
\include{headertikz}
\usetikzlibrary{through,shapes,arrows,decorations.pathmorphing,backgrounds,positioning,fit}

\newcommand{\argmin}{\mathop{\mathrm{min}}}
\usepackage[orientation=portrait,size=a0,scale=1.4]{beamerposter}

\newcommand{\conf}[1]{\newcommand{\insertconf}{#1}}
\newenvironment{WholeWidthBox}[1]{
  \begin{columns}
    \begin{column}{0.98\textwidth}
      \begin{block}{#1}
        \begin{hfill}
}{
        \end{hfill}
      \end{block}
    \end{column}
  \end{columns}
}

\newcommand{\TwoBoxes}[7]{%Width1 Title1 Content1 Width2 Title2 Content2 MinHeight
  \begin{columns}
    \begin{column}{#1\textwidth}
      \begin{block}{#2}
        \begin{columns}
          \begin{column}{0.001\textwidth}
            \vspace{#7}
          \end{column}
          \begin{column}{\textwidth}
            \centering
            #3
          \end{column}
        \end{columns}       
      \end{block}
    \end{column}

    \begin{column}{#4\textwidth}
      \begin{block}{#5}
        \begin{columns}
          \begin{column}{0.001\textwidth}
            \vspace{#7}
          \end{column}
          \begin{column}{\textwidth}
            \centering
            #6
          \end{column}
        \end{columns}       
      \end{block}
    \end{column}
  \end{columns}
}

%Start of document

\tikzstyle{state}=[circle,
thick,
minimum size=1.0cm,
draw=blue!80,
fill=blue!20]
\tikzstyle{action}=[rectangle,thick,
minimum size=1.0cm,
draw=orange!80,
fill=orange!20]
\tikzstyle{element}=[rectangle,
line width=1.5mm,
minimum size=1.0cm,
draw=blue!80,
fill=blue!20]
\tikzstyle{action}=[rectangle,
line width=1.5mm,
minimum size=1.0cm,
draw=orange!80,
fill=orange!20]

\conf{ECMLPKDD 2013}
\title{\Large{A cascaded supervised learning approach to inverse reinforcement learning}}
\author{{Edouard Klein}$^{\dag\ddag}$, \underline{Bilal Piot}$^{\dag\textrm{\Ankh}}$, Matthieu Geist$^\dag$ and Olivier Pietquin$^{\dag\textrm{\Ankh}}$\\\texttt{firstname.lastname@supelec.fr}}
\date{\today}
\institute[Supélec]{$\dag$Equipe IMS/MaLIS (Supélec), France\\$\ddag$Equipe ABC UMR 7503 (Loria-CNRS), France\\\Ankh UMI 2958 (GeorgiaTech-CNRS)}
\newlength{\columnheight}
\setlength{\columnheight}{105cm}

\begin{document}
\begin{frame}

%%%%%%%%%%%%% DEBUT PREMIERE LIGNE %%%%%%%%%%%%%%%%%
\begin{WholeWidthBox}{1. Introduction}
The Inverse Reinforcement Learning (IRL) problem can be defined as inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two generic Supervised Learning steps: this is the Cascaded Supervised IRL (CSI) approach. A classification step that defines a score function is followed by a regression step providing a reward function.
A theoretical analysis shows that the demonstrated expert policy is near-optimal for the computed reward function.
Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach. It is furthermore empirically demonstrated to compare positively to state-of-the-art approaches when using only transitions sampled according to the expert policy, up to the use of some heuristics. This is exemplified on two classical benchmarks (the mountain car problem and a highway driving simulator).
\end{WholeWidthBox}

\vfill
%%%%%%%%%%%% DEBUT DEUXIEME LIGNE %%%%%%%%%%%%%%%%%%

\TwoBoxes{.48}{2. Setting}{
  \begin{columns}
    \begin{column}{.48\textwidth}
      \includegraphics{Agent003.png}
    \end{column}
    \begin{column}{.48\textwidth}
      \begin{itemize}
      \item Expert's trace
      \item $MDP\backslash R $
      \item $\pi^*(s) \in \arg\max_{a\in A}Q^*(s,a)$
      \end{itemize}

    \end{column}
  \end{columns}
}
{.48}{3. Assumptions and Goal}{
  \begin{itemize}
  \item Assumptions
    \begin{itemize}
    \item The expert is a RL agent
    \item One can access the expert's trace
    \item The agent has the same abilities as the expert
    \end{itemize}
  \item Goal : Infer the expert's reward
    \begin{itemize}
    \item apprenticeship of the expert's task
    \item Generalization of the policy over never-seen-before states
    \item Useful when rewards are hard to tune ({\it e.g.}, driving)
    \end{itemize}
  \end{itemize}
}
{15cm}
\vfill
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TROISIEME, quatrième, cinquième et sixième LIGNES
\begin{columns}
  \begin{column}{.48\textwidth}
    \begin{block}{4. Expert}
      \begin{columns}
        \begin{column}{0.001\textwidth}
          \vspace{8cm}
        \end{column}
        \begin{column}{\textwidth}
      $\pi_E(s) = \arg\max\limits_{a\in\mathcal{A}} Q^{\pi_E}(s,a)$ \hspace{12em} $\pi^C(s) \in \arg\max\limits_{a\in\mathcal{A}}q(s,a)$

\begin{align*}
\quality{\expertpolicy}{\expertreward}(\state,\actionBis) &= \expertreward(\state,\actionBis) + \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\actionBis}{\state'} \quality{\expertpolicy}{\expertreward}(\state',\expertpolicy(\state'))\\
\expertreward(\state,\actionBis) &= \quality{\expertpolicy}{\expertreward}(\state,\actionBis) - \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\actionBis}{\state'} \quality{\expertpolicy}{\expertreward}(\state',\expertpolicy(\state'))\\
\Rc(\state,\actionBis) &= \classifscorefunc(\state,\actionBis) - \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\actionBis}{\state'} \classifscorefunc(\state',\classifpolicy(\state'))
\end{align*}

\vspace{1em}
$\classifpolicy$ is optimal for $\Rc$ and $\classifpolicy \approx \expertpolicy$, ergo we would be happy to find $\Rc$.
\vspace{1em}

$\sastrace{\sim} = \{\state_{\datasetindex},\actionBis_{\datasetindex},\state'_{\datasetindex}\}_{0\leq\datasetindex\leq\nbsamples}$\hspace{12em}
$\rsample_{\datasetindex} = q(s_{\datasetindex},a_{\datasetindex}) - \gamma q(s'_{\datasetindex},\classifpolicy(s'_{\datasetindex}))$ 
        \end{column}
      \end{columns}
    \end{block}
    \vspace{1cm}
    \begin{block}{6. Heuristics}
      \begin{columns}
        \begin{column}{0.001\textwidth}
          \vspace{8cm}
        \end{column}
        \begin{column}{\textwidth}
          $\sastrace{\expertpolicy} = \{(s_i,a_i,s'_i)_{1\leq i \leq N}\}$ \hspace{5em} $\left(s_{\datasetindex},\forall \actionBis \neq\expertpolicy(\state_{\datasetindex})\right) ,\rsample_{min} = \min_{\datasetindex\in \llbracket 1;\nbsamples\rrbracket}\rsample_{\datasetindex} - 1$
        \end{column}
      \end{columns}             
    \end{block}
    \vspace{1cm}
    %% \begin{block}{Putting it all together}
    %%   \begin{equation*}
    %%     \mathcal{X} \equiv S, \mathcal{Y} \equiv A, s\equiv Q^{\pi_E} \Rightarrow \phi \equiv \mu^{\pi_E}
    %%   \end{equation*}
    %% \end{block}
    \vfill
    \begin{block}{7. Pseudo-code}
      \resizebox{.95\columnwidth}{!}{

        \begin{algorithm}[H]
    %\small
  \caption{CSI algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $\satrace{\expertpolicy}=\{(s_i,a_i=\pi_E(s_i))\}_{1\leq i \leq \nbsamples}$\\and another training set $\sastrace{\sim}=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq \nbsamples'}$\;
  \emph{\textbf{Train}} a score function-based classifier on $\satrace{\expertpolicy}$, obtaining decision rule $\pi^C$ and score function $q:\statespace\times \actionspace \rightarrow \mathbb R$\;
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq \nbsamples'}$, $\forall (s_j,a_j,s'_j) \in \sastrace{\sim},\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}
      }
    \end{block}
  \end{column}
  \begin{column}{.48\textwidth}
    \vfill
    \begin{block}{8. Error bound}
      With concentration coefficient $\ccoeffpi{\optimalpolicy{\apprRc}}$, classification error $\epsilon_C$ and regression error $\epsilon_R$ :
\begin{equation}
0\leq\expectationknowing{\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}(s)-\rlvalue{\expertpolicy}{\apprRc}(s)}{s\sim\expertdistrib}\leq \frac{1}{1-\gamma}\left(\epsilon_C\Delta q +\epsilon_R(1+\ccoeffpi{\optimalpolicy{\apprRc}})\right).
\end{equation}
    \end{block}
    \vfill
    \begin{block}{9. Highway}
      \begin{columns}
        \begin{column}{0.001\textwidth}
          \vspace{7cm}
        \end{column}
        \begin{column}{\textwidth}
          \centering
          \fontsize{11pt}{11pt}\selectfont
          \resizebox{.95\columnwidth}{!}{\includegraphics[width=0.47\textwidth]{Exp14}
            \includegraphics[width=0.47\textwidth]{Exp14_zoom}}
        \end{column}
      \end{columns}             
    \end{block}
    \begin{block}{11. Mountain Car}
      \centering

      \includegraphics[width=\textwidth]{Exp11}
      %}
    \end{block}
  \end{column}
\end{columns}
\vfill

\end{frame}
\end{document}

### 2013-09-12 Manuscrit/main_for_diff.tex
Mise en place d'une regle diff

% Created 2013-09-12 Jeu 20:53
\documentclass[frenchb,a4paper,justified,notoc]{tufte-book}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\usepackage{amsmath}
\usepackage{tikz}
\def\E{\operatornamewithlimits{%
\mathchoice{\vcenter{\hbox{\huge \ensuremath{\mathbb{E}}}}}
{\vcenter{\hbox{\Large  \ensuremath{\mathbb{E}}}}}
{ \ensuremath{\mathbb{E}}}
{ \ensuremath{\mathbb{E}}}}}
\usepackage{dsfont}
\usepackage[xindy,toc,acronym,nomain,sort=def]{glossaries}
\newcommand{\dash}{\ensuremath{'}}
\newcommand{\apos}{'}
\newglossary[nlg]{notation}{not}{ntn}{Notations}

\newglossaryentry{rewardfunc}{type=notation,
name={\ensuremath{R}},
description={Fonction de récompense},
}
\newcommand{\rewardfunc}{\gls{rewardfunc}}

\newglossaryentry{paramclassif}{type=notation,
name={\ensuremath{\omega}},
description={Vecteur de paramètres pour la classification},
}
\newcommand{\paramclassif}{\gls{paramclassif}}

\newglossaryentry{Rc}{type=notation,
name={\ensuremath{R^{C}}},
description={Fonction de récompense induite par la fonction de score d{\apos}un classifieur},
}
\newcommand{\Rc}{\gls{Rc}}

\newglossaryentry{empiricalclassiferror}{type=notation,
name={\ensuremath{\epsilon_C^{emp}}},
description={Erreur empirique de classification},
}
\newcommand{\empiricalclassiferror}{\gls{empiricalclassiferror}}

\newglossaryentry{structuredcost}{type=notation,
name={\ensuremath{J}},
description={Fonction de coût de la classification structurée},
}
\newcommand{\structuredcost}{\gls{structuredcost}}

\newglossaryentry{featurestatefunc}{type=notation,
name={\ensuremath{\psi}},
description={Fonction d{\apos}attribut sur l{\apos}espace d{\apos}état},
}
\newcommand{\featurestatefunc}{\gls{featurestatefunc}}

\newglossaryentry{discount}{type=notation,
name={\ensuremath{\gamma}},
description={Facteur d{\apos}amortissement},
}
\newcommand{\discount}{\gls{discount}}

\newglossaryentry{kernel}{type=notation,
name={\ensuremath{\kappa}},
description={Noyau pour une SVM},
}
\newcommand{\kernel}{\gls{kernel}}

\newglossaryentry{paramirl}{type=notation,
name={\ensuremath{\theta}},
description={Vecteur de paramètres pour l{\apos}ARI},
}
\newcommand{\paramirl}{\gls{paramirl}}

\newglossaryentry{rctrace}{type=notation,
name={\ensuremath{D_{sar}^{q}}},
description={Base d{\apos}entraînement du régresseur dans CSI},
}
\newcommand{\rctrace}{\gls{rctrace}}

\newglossaryentry{subgrad}{type=notation,
name={\ensuremath{\nabla}},
description={Sous-gradient d{\apos}une fonction},
}
\newcommand{\subgrad}{\gls{subgrad}}

\newglossaryentry{dimpsi}{type=notation,
name={\ensuremath{d_{\psi}}},
description={Dimension de l{\apos}espace d{\apos}attributs sur l{\apos}espace d{\apos}état},
}
\newcommand{\dimpsi}{\gls{dimpsi}}

\newglossaryentry{expertmu}{type=notation,
name={\ensuremath{\mu^E}},
description={Attribut moyen de l{\apos}expert},
}
\newcommand{\expertmu}{\gls{expertmu}}

\newglossaryentry{datasetindex}{type=notation,
name={\ensuremath{i}},
description={Indice des exemples dans une base},
}
\newcommand{\datasetindex}{\gls{datasetindex}}

\newglossaryentry{apprRc}{type=notation,
name={\ensuremath{\hat{R}^{C}}},
description={Approximation de $R^C$},
}
\newcommand{\apprRc}{\gls{apprRc}}

\newglossaryentry{classifscorefunc}{type=notation,
name={\ensuremath{q}},
description={Fonction de score pour la classification},
}
\newcommand{\classifscorefunc}{\gls{classifscorefunc}}

\newglossaryentry{expertreward}{type=notation,
name={\ensuremath{R^E}},
description={Récompense optimisée par l{\apos}expert},
}
\newcommand{\expertreward}{\gls{expertreward}}

\newglossaryentry{state}{type=notation,
name={\ensuremath{s}},
description={État},
}
\newcommand{\state}{\gls{state}}

\newglossaryentry{transprobfunc}{type=notation,
name={\ensuremath{p}},
description={Probabilités de transition},
}
\newcommand{\transprobfunc}{\gls{transprobfunc}}

\newglossaryentry{action}{type=notation,
name={\ensuremath{a}},
description={Une action},
}
\newcommand{\action}{\gls{action}}

\newglossaryentry{slack}{type=notation,
name={\ensuremath{\zeta}},
description={Variable d{\apos}écart},
}
\newcommand{\slack}{\gls{slack}}

\newglossaryentry{statespace}{type=notation,
name={\ensuremath{\mathcal{S}}},
description={Espace d{\apos}état},
}
\newcommand{\statespace}{\gls{statespace}}

\newglossaryentry{nbsamples}{type=notation,
name={\ensuremath{N}},
description={Nombre d{\apos}exemples dans une base},
}
\newcommand{\nbsamples}{\gls{nbsamples}}

\newglossaryentry{reg}{type=notation,
name={\ensuremath{\lambda}},
description={Coefficient de régularisation},
}
\newcommand{\reg}{\gls{reg}}

\newglossaryentry{expertpolicy}{type=notation,
name={\ensuremath{\pi^E}},
description={Politique de l{\apos}expert},
}
\newcommand{\expertpolicy}{\gls{expertpolicy}}

\newglossaryentry{ccoeff}{type=notation,
name={\ensuremath{C_f}},
description={Coefficient de concentration dans le pire des cas},
}
\newcommand{\ccoeff}{\gls{ccoeff}}

\newglossaryentry{margin}{type=notation,
name={\ensuremath{\mathfrak{l}}},
description={Fonction de marge dans le classifieur à marge},
}
\newcommand{\margin}{\gls{margin}}

\newglossaryentry{classifpolicy}{type=notation,
name={\ensuremath{\pi^C}},
description={Politique issue d{\apos}un classifieur},
}
\newcommand{\classifpolicy}{\gls{classifpolicy}}

\newglossaryentry{timehorizon}{type=notation,
name={\ensuremath{\infty}},
description={Horizon temporel},
}
\newcommand{\timehorizon}{\gls{timehorizon}}

\newglossaryentry{dimindexbis}{type=notation,
name={\ensuremath{k}},
description={Entier indexant les dimensions d{\apos}un espace},
}
\newcommand{\dimindexbis}{\gls{dimindexbis}}

\newglossaryentry{reals}{type=notation,
name={\ensuremath{\mathbb{R}}},
description={Le corps des réels},
}
\newcommand{\reals}{\gls{reals}}

\newglossaryentry{actionspace}{type=notation,
name={\ensuremath{\mathcal{A}}},
description={Espace d{\apos}action},
}
\newcommand{\actionspace}{\gls{actionspace}}

\newglossaryentry{policy}{type=notation,
name={\ensuremath{\pi}},
description={Une politique},
}
\newcommand{\policy}{\gls{policy}}

\newglossaryentry{dimstate}{type=notation,
name={\ensuremath{d_{\mathcal{S}}}},
description={Dimension de l{\apos}espace d{\apos}état},
}
\newcommand{\dimstate}{\gls{dimstate}}

\newglossaryentry{timeindex}{type=notation,
name={\ensuremath{t}},
description={Indice temporel},
}
\newcommand{\timeindex}{\gls{timeindex}}

\newglossaryentry{mudistance}{type=notation,
name={\ensuremath{d}},
description={Distance entre deux attributs moyens},
}
\newcommand{\mudistance}{\gls{mudistance}}

\newglossaryentry{spacedistrib}{type=notation,
name={\ensuremath{\rho}},
description={Loi de probabilité ou fonction de poids},
}
\newcommand{\spacedistrib}{\gls{spacedistrib}}

\newglossaryentry{dimindex}{type=notation,
name={\ensuremath{j}},
description={Entier indexant les dimensions d{\apos}un espace},
}
\newcommand{\dimindex}{\gls{dimindex}}

\newglossaryentry{paramlstdmu}{type=notation,
name={\ensuremath{\zeta}},
description={Matrice de paramètres pour LSTD-$\mu$},
}
\newcommand{\paramlstdmu}{\gls{paramlstdmu}}

\newglossaryentry{expertdistrib}{type=notation,
name={\ensuremath{\rho_E}},
description={Distribution stationnaire de l{\apos}expert},
}
\newcommand{\expertdistrib}{\gls{expertdistrib}}

\newglossaryentry{mdpbis}{type=notation,
name={\ensuremath{\mathcal{M}}},
description={Un \glstext{mdp}},
}
\newcommand{\mdpbis}{\gls{mdpbis}}

\newglossaryentry{gaussiancenter}{type=notation,
name={\ensuremath{\mathfrak{m}}},
description={Centre d{\apos}une gaussienne},
}
\newcommand{\gaussiancenter}{\gls{gaussiancenter}}

\newglossaryentry{muupdate}{type=notation,
name={\ensuremath{u}},
description={Règle de mise à jour pour les algorithme d{\apos}ARI itératifs},
}
\newcommand{\muupdate}{\gls{muupdate}}

\newglossaryentry{munbtraj}{type=notation,
name={\ensuremath{M}},
description={Nombre de trajectoires},
}
\newcommand{\munbtraj}{\gls{munbtraj}}

\newglossaryentry{naturals}{type=notation,
name={\ensuremath{\mathbb{N}}},
description={Entiers naturels},
}
\newcommand{\naturals}{\gls{naturals}}

\newglossaryentry{dimphi}{type=notation,
name={\ensuremath{d_{\phi}}},
description={Dimension de l{\apos}espace d{\apos}attributs état-action},
}
\newcommand{\dimphi}{\gls{dimphi}}

\newglossaryentry{paramrl}{type=notation,
name={\ensuremath{\xi}},
description={Vecteur de paramètres pour l{\apos}AR},
}
\newcommand{\paramrl}{\gls{paramrl}}

\newglossaryentry{gaussianvar}{type=notation,
name={\ensuremath{\sigma}},
description={Ecart-type d{\apos}une gaussienne},
}
\newcommand{\gaussianvar}{\gls{gaussianvar}}

\newglossaryentry{muhorizon}{type=notation,
name={\ensuremath{L}},
description={Longueur d{\apos}une trajectoire},
}
\newcommand{\muhorizon}{\gls{muhorizon}}

\newglossaryentry{spaceindex}{type=notation,
name={\ensuremath{m}},
description={Entier indexant un ensemble},
}
\newcommand{\spaceindex}{\gls{spaceindex}}

\newglossaryentry{rsample}{type=notation,
name={\ensuremath{\hat{r}}},
description={Echantillon pour l{\apos}estimation de $R^C$},
}
\newcommand{\rsample}{\gls{rsample}}

\newglossaryentry{featurestateactionfunc}{type=notation,
name={\ensuremath{\phi}},
description={Fonction d{\apos}attributs état-action},
}
\newcommand{\featurestateactionfunc}{\gls{featurestateactionfunc}}

\newglossaryentry{featurestateactionspace}{type=notation,
name={\ensuremath{\Phi}},
description={Espace d{\apos}attributs état-action},
}
\newcommand{\featurestateactionspace}{\gls{featurestateactionspace}}

\newcommand{\rlvalue}[2]{\ensuremath{\glsadd{FORCDONOTUSErlvalue}V^{#1}_{#2}}}
\newglossaryentry{FORCDONOTUSErlvalue}{type=notation,
name={\ensuremath{V^{\pi}_R}},
description={Fonction de valeur pour la récompense $R$ lorsqu{\apos}on suit la politique $\pi$},
}

\newcommand{\best}[1]{\ensuremath{\glsadd{FORCDONOTUSEbest}{#1}^{*}}}
\newglossaryentry{FORCDONOTUSEbest}{type=notation,
name={\ensuremath{x^*}},
description={Element issu d{\apos}un $\arg\max_x$},
}

\newcommand{\bellmanevalopeval}[3]{\ensuremath{\glsadd{FORCDONOTUSEbellmanevalopeval}B^{#1}_{#2}{#3}}}
\newglossaryentry{FORCDONOTUSEbellmanevalopeval}{type=notation,
name={\ensuremath{B^{\pi}_{R}}},
description={Opérateur d{\apos}évaluation de Bellman},
}

\newcommand{\bellmanoptopeval}[2]{\ensuremath{\glsadd{FORCDONOTUSEbellmanoptopeval}B^{*}_{#1}{#2}}}
\newglossaryentry{FORCDONOTUSEbellmanoptopeval}{type=notation,
name={\ensuremath{B^{*}_{R}}},
description={Opérateur d{\apos}optimalité de Bellman},
}

\newcommand{\quality}[2]{\ensuremath{\glsadd{FORCDONOTUSEquality}Q^{#1}_{#2}}}
\newglossaryentry{FORCDONOTUSEquality}{type=notation,
name={\ensuremath{Q^{\pi}_{R}}},
description={Fonction de qualité de la politique $\pi$ pour la récompense $R$},
}

\newcommand{\appr}[1]{\ensuremath{\glsadd{FORCDONOTUSEappr}\widehat{#1}}}
\newglossaryentry{FORCDONOTUSEappr}{type=notation,
name={\ensuremath{\hat X}},
description={Approximation de $X$ à partir de données},
}

\newcommand{\stationarydistrib}[1]{\ensuremath{\glsadd{FORCDONOTUSEstationarydistrib}\rho_{#1}}}
\newglossaryentry{FORCDONOTUSEstationarydistrib}{type=notation,
name={\ensuremath{\rho_\pi}},
description={Distribution stationnaire induite par la politique $\pi$},
}

\newcommand{\expectationknowing}[2]{\ensuremath{\glsadd{FORCDONOTUSEexpectationknowing}\mathbb{E} \left[\left. #1\right|#2\right]}}
\newglossaryentry{FORCDONOTUSEexpectationknowing}{type=notation,
name={\ensuremath{\E \left[\left. f(x)\right| x \sim \rho \right]}},
description={Espérance de $f(x)$ pour $x$ tiré selon $\rho$},
}

\newcommand{\dimx}[1]{\ensuremath{\glsadd{FORCDONOTUSEdimx}d_{#1}}}
\newglossaryentry{FORCDONOTUSEdimx}{type=notation,
name={\ensuremath{d_{f}}},
description={Dimension de l{\apos}espace engendré par $f$},
}

\newcommand{\classiferror}[1]{\ensuremath{\glsadd{FORCDONOTUSEclassiferror}\epsilon_C^{#1}}}
\newglossaryentry{FORCDONOTUSEclassiferror}{type=notation,
name={\ensuremath{\epsilon_C^{\spacedistrib}}},
description={Risque de classification sur la distribution $\spacedistrib$},
}

\newcommand{\transprobfunceval}[3]{\ensuremath{\glsadd{FORCDONOTUSEtransprobfunceval}p\left(#3|#1,#2\right)}}
\newglossaryentry{FORCDONOTUSEtransprobfunceval}{type=notation,
name={\ensuremath{p(s\apos|s,a)}},
description={Probabilité qu{\apos}un agent transite en $s\apos$ après avoir choisi l{\apos}action $a$ dans l{\apos}état $s$},
}

\newcommand{\transpose}[1]{\ensuremath{\glsadd{FORCDONOTUSEtranspose}{#1}^{T}}}
\newglossaryentry{FORCDONOTUSEtranspose}{type=notation,
name={\ensuremath{X^T}},
description={Transposée de la matrice $X$},
}

\newcommand{\transprobmat}[1]{\ensuremath{\glsadd{FORCDONOTUSEtransprobmat}P^{#1}}}
\newglossaryentry{FORCDONOTUSEtransprobmat}{type=notation,
name={\ensuremath{P^{\pi}}},
description={Matrice des probabilités de transition induites par la politique $\policy$},
}

\newcommand{\gaussian}[3]{\ensuremath{\glsadd{FORCDONOTUSEgaussian}\mathfrak{G}^{#1}_{#2}(#3)}}
\newglossaryentry{FORCDONOTUSEgaussian}{type=notation,
name={\ensuremath{\mathfrak{G}^{m}_{\sigma}}},
description={Fonction gaussienne de centre $m$ et d{\apos}écart-type $\sigma$},
}

\newcommand{\optimalpolicy}[1]{\ensuremath{\glsadd{FORCDONOTUSEoptimalpolicy}\pi^*_{#1}}}
\newglossaryentry{FORCDONOTUSEoptimalpolicy}{type=notation,
name={\ensuremath{\pi^*_R}},
description={Une politique optimale pour la fonction de récompense $R$},
}

\newcommand{\sasrtrace}[1]{\ensuremath{\glsadd{FORCDONOTUSEsasrtrace}D_{sasr}^{#1}}}
\newglossaryentry{FORCDONOTUSEsasrtrace}{type=notation,
name={\ensuremath{D_{sasr}^{\rewardfunc}}},
description={Base de type $s,a,s\prime,r$},
}

\newcommand{\satrace}[1]{\ensuremath{\glsadd{FORCDONOTUSEsatrace}D_{sa}^{#1}}}
\newglossaryentry{FORCDONOTUSEsatrace}{type=notation,
name={\ensuremath{D_{sa}^{\policy}}},
description={Base de type $s,a$ obtenue en suivant la politique $\policy$},
}

\newcommand{\twosetsfunctions}[2]{\ensuremath{\glsadd{FORCDONOTUSEtwosetsfunctions}{#2}^{#1}}}
\newglossaryentry{FORCDONOTUSEtwosetsfunctions}{type=notation,
name={\ensuremath{B^A}},
description={Ensemble des applications de $A$ dans $B$},
}

\newcommand{\greedy}[1]{\ensuremath{\glsadd{FORCDONOTUSEgreedy}g\left(#1\right)}}
\newglossaryentry{FORCDONOTUSEgreedy}{type=notation,
name={\ensuremath{g(\policy)}},
description={Politique gloutonne vis-à-vis de la fonction de qualité de $\pi$},
}

\newcommand{\sastrace}[1]{\ensuremath{\glsadd{FORCDONOTUSEsastrace}D_{sas}^{#1}}}
\newglossaryentry{FORCDONOTUSEsastrace}{type=notation,
name={\ensuremath{D_{sas}^{\policy}}},
description={Base de type $s,a,s\dash$ obtenue en suivant la politique $\pi$},
}

\newcommand{\component}[2]{\ensuremath{\glsadd{FORCDONOTUSEcomponent}{#1}_{#2}}}
\newglossaryentry{FORCDONOTUSEcomponent}{type=notation,
name={\ensuremath{X_j}},
description={Composante $j$ du vecteur $X$},
}

\newcommand{\indicatorfunc}[1]{\ensuremath{\glsadd{FORCDONOTUSEindicatorfunc}\mathds{1}(#1)}}
\newglossaryentry{FORCDONOTUSEindicatorfunc}{type=notation,
name={\ensuremath{\mathds{1}}},
description={Fonction indicatrice},
}

\newcommand{\scalarprod}[2]{\ensuremath{\glsadd{FORCDONOTUSEscalarprod}\left\langle{#1} , {#2}\right\rangle}}
\newglossaryentry{FORCDONOTUSEscalarprod}{type=notation,
name={\ensuremath{\langle x,y\rangle}},
description={Produit scalaire de $x$ et $y$},
}

\newcommand{\norm}[2]{\ensuremath{\glsadd{FORCDONOTUSEnorm}\left|\left|{#2}\right|\right|_{#1}}}
\newglossaryentry{FORCDONOTUSEnorm}{type=notation,
name={\ensuremath{||\cdot||_x}},
description={Norme $x$},
}

\newcommand{\rlmu}[1]{\ensuremath{\glsadd{FORCDONOTUSErlmu}\mu^{#1}}}
\newglossaryentry{FORCDONOTUSErlmu}{type=notation,
name={\ensuremath{\mu^{\pi}}},
description={Attribut moyen de la politique $\pi$},
}

\newcommand{\gaussperdim}[1]{\ensuremath{\glsadd{FORCDONOTUSEgaussperdim}g_{#1}}}
\newglossaryentry{FORCDONOTUSEgaussperdim}{type=notation,
name={\ensuremath{g_i}},
description={Nombre de gaussiennes pour la dimension $i$ dans un vecteur d{\apos}attributs basé sur un réseau de gaussiennes},
}

\newcommand{\matrixbyterm}[3]{\ensuremath{\glsadd{FORCDONOTUSEmatrixbyterm}\left({#1}\right)_{{#2},{#3}}}}
\newglossaryentry{FORCDONOTUSEmatrixbyterm}{type=notation,
name={\ensuremath{\left(f(i,j)\right)_{i,j}}},
description={Matrice dont l{\apos}élément de ligne $i$ et de colonne $j$ est $f(i,j)$},
}

\newcommand{\card}[1]{\ensuremath{\glsadd{FORCDONOTUSEcard}\left|#1\right|}}
\newglossaryentry{FORCDONOTUSEcard}{type=notation,
name={\ensuremath{|A|}},
description={Cardinal de l{\apos}ensemble A},
}

\newcommand{\ccoeffpi}[1]{\ensuremath{\glsadd{FORCDONOTUSEccoeffpi}C_{#1}}}
\newglossaryentry{FORCDONOTUSEccoeffpi}{type=notation,
name={\ensuremath{C_{\pi}}},
description={Coefficient de concentration entre une politique et la politique experte},
}

\usepackage{tikz} 
\usetikzlibrary{shapes.arrows}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.pathreplacing}
\tikzset{for this and nested ones/.style={#1,every picture/.style={#1}}}
\input{titlepage}   
\usepackage{csquotes}
\bibliography{Biblio,BiblioKlein}
\renewcommand{\theoremautorefname}{théorème}
\renewcommand{\appendixautorefname}{annexe}
\hypersetup{ % config hyperref pour virer les box/color affreux...
colorlinks,%
citecolor=black,%
filecolor=black,%
linkcolor=black,%
urlcolor=black
} 
\usepackage{subfigure}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[frenchb]{babel}
\usepackage{tabularx}
\usepackage{amsthm}
\usepackage{thmtools}
\renewcommand{\listtheoremname}{Liste des théorèmes et propositions} 
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Théorème}
\newtheorem{corollary}{Corollaire}
\newtheorem{lemma}{Lemme}
\newglossary[angl]{anglicisme}{aot}{atn}{Anglicismes}
\newcommand{\newangl}[3]{\newglossaryentry{#1}{type=anglicisme,name={\emph{#2}},description={#3}}}
\renewcommand{\acronymname}{Acronymes}
\makeglossaries
\usepackage{tocbibind}
\usepackage[french,ruled,vlined,noend,linesnumbered]{algorithm2e}
\newcommand{\listofalgorithmes}{\tocfile{Liste des algorithmes}{loa}}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\sgn}{sign}
% scientific notation, 1\e{9} will print as 1x10^9
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\newacronym{mdp}{PDM}{Processus D{é}cisionnel de Markov}
\newacronym{irl}{ARI}{Apprentissage par Renforcement Inverse}
\newacronym{dp}{PD}{Programmation Dynamique}
\newacronym{rl}{AR}{Apprentissage par Renforcement}
\newacronym{lspi}{LSPI}{\emph{Least Squares Policy Iteration}}
\newacronym{pirl}{PIRL}{\emph{Projection Inverse Reinforcement Learning}}
\newacronym{mmp}{MMP}{\emph{Maximum Margin Planning}}
\newacronym{pm}{PM}{\emph{Policy Matching}}
\newacronym{mwal}{MWAL}{\emph{Multiplicative Weights for Apprenticeship Learning}}
\newacronym{maxent}{MaxEnt}{\emph{Maximum Entropy}}
\newacronym{relent}{Rel\-Ent}{\emph{Relative Entropy}}
\newacronym{lpal}{LPAL}{\emph{Linear Programming for Apprenticeship Learning}}
\newacronym{birl}{BIRL}{\emph{Bayesian Inverse Reinforcement Learning}}
\newacronym{gpirl}{GPIRL}{\emph{Gaussian Processes Inverse Reinforcement Learning}}
\newacronym{firl}{FIRL}{\emph{Feature Inverse Reinforcement Learning}}
\newacronym{irlgp}{FIRL}{\emph{Inverse Reinforcement Learning with Gaussian Processes}}
\newacronym{lstdmu}{LSTD-$\mu$}{\emph{Least Squares Temporal Differences feature expectations}}
\newacronym{lstd}{LSTD}{\emph{Least Squares Temporal Differences}}
\newacronym{scirl}{SCIRL}{\emph{Structured Classification for Inverse Reinforcement Learning}}
\newacronym{cascading}{CSI}{\emph{Cascaded Supervised learning for Inverse reinforcement learning}}
\newacronym{cnn}{CNN}{\emph{Convolutional Neural Network}}
\newacronym{svm}{SVM}{\emph{Support Vector Machine} (Machine à Vecteurs Support)}
\newacronym{gmm}{GMM}{\emph{Gaussian Mixture Model} (Modèle à base de Mélange de Gaussiennes)}
\newacronym{alvinn}{ALVINN}{\emph{Autonomous Land Vehicle In a Neural Network}}
\newacronym{churps}{CHURPs}{\emph{Compressed Heuristic Universal Reaction Planners}}
\newacronym{knn}{$k$-NN}{$k$-\emph{plus proches voisins} (en anglais \emph{k-Nearest Neighboors})}
\newacronym{gp}{PG}{Processus Gaussiens}
\newacronym{smile}{SMILe}{Stochastic Mixing Iterative Learning}
\newacronym{dagger}{DAGGER}{Dataset Aggregation}
\newangl{batch}{batch}{Par paquet}
\newangl{offpolicy}{off-policy}{Signifie que la politique qui contrôle le système n'est pas celle qui est évaluée}
\newangl{onpolicy}{on-policy}{\`A la différence du {\it off-policy}, la politique évaluée est celle qui contrôle le système}
\newangl{gridworld}{gridworld}{Echiquier, damier}
\newangl{rewardshaping}{reward shaping}{Transformation de la récompense ne changeant pas les politiques optimales}
\newangl{boosting}{boosting}{Augmentation du pouvoir descriptif de la représentation}
\newangl{mountaincar}{mountain-car}{Problème jouet où une voiture doit sortir d'une vallée}
\newangl{highway}{highway}{Problème jouet où une voiture doit doubler d'autres voitures sur une autoroute}
\providecommand{\alert}[1]{\textbf{#1}}

\title{Contributions à l'apprentissage par renforcement inverse}
\author{Edouard KLEIN}
\date{\today}
\hypersetup{
  pdfkeywords={},
  pdfsubject={},
  pdfcreator={Emacs Org-mode version 7.9.3f}}

\begin{document}

\maketitle

\setcounter{tocdepth}{3}
\tableofcontents
\vspace*{1cm}

\listoffigures
\listoftables
\listofalgorithmes
\listoftheorems
\printglossaries
\newsavebox\tuture
\begin{lrbox}{\tuture}
   \begin{tikzpicture}
    \fill[yellow] (0,0) -- (1.2,0) -- (1.2,.5) -- (0,.5) --cycle;
    \draw[black] (0,0) -- (1.2,0) -- (1.2,.5) -- (0,.5) --cycle;
    \node[shape=circle,draw=black,fill=green,scale=1] at (0.3,0) {};
    \node[shape=circle,draw=black,fill=green,scale=1] at (.9,0) {};
    \node[red] (carMark) at (.6,.2) {\bf x};
  \end{tikzpicture}
\end{lrbox}
\newcommand{\speedarrow}[1]{
  \begin{tikzpicture}
    \node [single arrow,fill=blue!50,minimum height=#1,single arrow head indent=3] at(0,0) {};
  \end{tikzpicture}
}
\chapter{Introduction}
\label{sec-1}

Ce chapitre expose les motivations de nos travaux et situe nos contributions. Nous regroupons la liste de ces contributions ainsi que des publications auxquelles elles ont donné lieu en annexe. Nous terminons ce chapitre par une vue d'ensemble du manuscrit.
\section{Motivations}
\label{sec-1-1}

Dans les pays développés, peu de travaux confiés à des humains peuvent aujourd'hui être confiés à des machines\footnote{Dans les pays en développement ou non développés, il est plus rentable de payer un travailleur non qualifié que d'utiliser un robot ou un ordinateur.
 }. Les limitations des systèmes automatiques ne sont pas mécaniques, les machines aujourd'hui peuvent être plus puissantes et robustes que n'importe quel animal et plus précises et plus fiables qu'un bon artisan. La généralisation de l'automatisation est freinée par l'impossibilité pour une machine d'assumer la charge cognitive liée à la gestion des environnements dynamiques et incertains dans lesquels les humains parviennent à évoluer.

Le problème du \emph{contrôle optimal} consiste à influencer le comportement d'un système dynamique pour qu'il remplisse une tâche. Le concepteur d'un système informatique résolvant le problème du contrôle optimal doit faire preuve d'une double expertise :
\begin{itemize}
\item il doit comprendre comment réaliser la tâche, c'est-à-dire prendre en compte les critères de réussite et d'échec, les risques associés à un échec catastrophique, les compromis entre les différents coûts (en temps, en énergie, en matières premières etc.) ;
\item il doit également savoir exprimer ces contraintes dans le cadre d'une méthode de recherche du contrôle optimal.
\end{itemize}

Pour des tâches complexes il est coûteux de réunir ces deux types d'expertise dans une même équipe de développement. Pour réduire l'expertise nécessaire dans le domaine du contrôle optimal, nous allons, à partir d'une \emph{démonstration} du contrôle, extraire automatiquement une \emph{description des contraintes de la tâche} exploitable par des méthodes de recherche du contrôle optimal. Ce paradigme, l'\emph{apprentissage par imitation}, vise à permettre à un utilisateur non qualifié dans le domaine informatique, mais performant en ce qui concerne la tâche à effectuer, d'enseigner de nouveaux comportements à une machine simplement en en faisant l'exemple.

L'application de ce paradigme à des domaines dynamiques dont le modèle est inconnu et possiblement non linéaire nous pousse à choisir l'\gls{rl} comme méthode de recherche du contrôle optimal. L'apprentissage par renforcement permet de manière générique\footnote{C'est-à-dire qu'un même algorithme peut être utilisé pour apprendre à gérer deux domaines complètement différents.
 } d'apprendre le contrôle optimal à partir d'un signal de récompense, information locale sur la qualité du contrôle, en exploitant l'information issue d'interactions avec le système. Nos contributions se situent donc dans le domaine de l'\gls{irl}, qui cherche à obtenir un signal de récompense cohérent avec une démonstration de la tâche que fournit un expert.
\section{Vue d'ensemble du manuscrit}
\label{sec-1-2}

 Le \textbf{\autoref{hier-formalisme}} explore la problématique de l'apprentissage par imitation. Nous commençons par formaliser la notion de politique déterministe markovienne : à un état contenant toutes les données nécessaires à la prise d'une décision, l'agent associe de manière déterministe une action. Nous remarquons alors que le problème d'apprentissage supervisé d'une politique se réduit au problème bien connu de la classification multi-classe, les états étant les entrées à classifier et les actions étant les catégories à associer aux entrées. Nous définissons le risque empirique et le risque réel, qui mesurent l'erreur de classification respectivement sur les données dont nous disposons et sur la distribution sur laquelle il nous importe d'apprendre la politique experte. Nous nous attardons ensuite sur les modèles linéaires, de grande importance pour nous car nous les utiliserons tout au long du manuscrit comme modèle d'approximation des fonctions de valeur, de score, de récompense et d'attribut moyen de l'expert. Nous proposons deux choix de fonction d'attributs génériques, que nous utiliserons dans nos expériences. Ces prolégomènes nous amènent à étudier plus particulièrement deux méthodes de classification, une méthode à marge structurée que nous retrouverons au \autoref{hier-scirl} lors de l'exposé de l'algorithme \gls{scirl} et les \gls{svm}, qui nous serons utiles au \autoref{hier-cascading} comme routine de l'algorithme \gls{cascading}.

Après un état de l'art des méthodes d'imitation supervisée, nous définirons le cadre des \gls{mdp} dans lequel se situera le reste du manuscrit. La notion de dynamique temporelle, formalisée comme les probabilités de transition vers un état d'arrivée d'un agent choisissant une certaine action dans un état de départ, y est fondamentale. Elle permet de dépasser le cadre de la politique comme simple association état-action que manipulent les méthodes supervisées en introduisant des critères long terme sur la qualité d'une politique. La fonction de valeur, espérance de la somme pondérée d'une fonction de récompense (signal localisé sur la qualité du contrôle) permet d'ordonner les politiques. La \gls{dp} est utilisée pour rechercher une politique optimale via la résolution des équations de Bellman. Nous nous intéresserons plutôt à l'\gls{rl}, qui résout le même problème sans connaître explicitement les probabilités de transition, mais en interagissant avec le système à contrôler. D'un intérêt particulier pour nous sont les algorithmes pouvant fonctionner à partir d'une base de données recueillies une fois pour toutes, nous donnons l'exemple de \emph{Fitted-Q} (\autoref{alg:fittedq}) et \gls{lspi} (\autoref{alg:lspi}), plus efficaces en échantillons que les algorithmes en ligne.

L'\gls{rl} permet de trouver une politique optimale à partir d'information localisée sur la qualité du contrôle présentée sous la forme d'une fonction de récompense. Le problème inverse, l'\gls{irl}, consiste à trouver une fonction de récompense pour laquelle un comportement dont nous observons l'exemple est optimal. L'optimisation de cette fonction par un algorithme d'\gls{rl} fournira une politique résolvant le problème de l'apprentissage par imitation. Nous fournissons deux critères de résolution du problème de l'\gls{irl} (formules \ref{eq:aricriterebis} et \ref{eq:aricritere}). L'un mesure l'optimalité de la politique d'imitation pour la récompense (normalement inconnue) optimisée par l'expert, l'autre mesure l'optimalité de l'expert pour la récompense trouvée par l'algorithme d'\gls{irl}. Enfin, nous terminerons le chapitre par la définition de l'attribut moyen. L'existence de cette grandeur découle du choix d'une modélisation linéaire de la fonction de récompense. Elle joue un rôle capital dans la plupart des approches de l'état de l'art, car deux politiques d'attributs moyens proches auront (pour toute récompense) des valeurs proches (\autoref{thm:closemuclosev}).

Le \textbf{\autoref{hier-etatdelart}} propose un état de l'art de l'\gls{irl}. Nous justifions tout d'abord notre choix de considérer la fonction de récompense comme une fonction de l'espace joint état-action dans $\reals$, puis étudions les transformations qu'il est possible d'appliquer à une fonction de récompense sans modifier les politiques optimales associées (\gls{rewardshaping}). Nous faisons ensuite l'état de l'art de l'\gls{irl} proprement dit. La grande majorité des approches obéit à une structure commune que nous exposons \autoref{alg:repetee}. Dans cette approche itérative, la fonction de récompense est approximée par un modèle linéaire. L'attribut moyen de la politique optimale pour la récompense courante est comparé à l'attribut moyen de l'expert. Le vecteur de paramètres de la fonction de récompense est ensuite mis à jour afin de faire diminuer une certaine distance entre ces deux attributs moyens. Les algorithmes suivant ce schéma diffèrent par la définition de la distance entre deux attributs moyens et par la règle de mise à jour. Toutes ces approchent souffrent des problèmes intrinsèquement liés à leur structure itérative. Il faut de manière répétée résoudre un \gls{mdp}, c'est-à-dire trouver la politique optimale pour une récompense arbitraire. Il faut ensuite estimer l'attribut moyen de cette politique optimale. Ces deux étapes sont gourmandes en données et en temps de calcul. Plus récemment, de nouvelles approches ce sont détachées de cette structure itérative. Certaines abandonnent également l'approximation linéaire de la fonction de récompense, la remplaçant par exemple par des \gls{gp}. La plupart de ces nouvelles approches nécessitent cependant pour fonctionner au moins autant d'information que les approches itératives précédentes. L'une d'entre elle cependant, \gls{relent} (\autoref{alg:relent}) est particulièrement peu gourmande en données. Elle ne réclame dans sa version échantillonnée qu'une base d'exemples de l'expert et des échantillons tirés en application d'une politique aléatoire, couvrant tout l'espace d'état. De ce fait, elle nous servira de point de comparaison aux chapitres \ref{hier-scirl} et \ref{hier-cascading} lorsque nous testerons nos nouveaux algorithmes d'\gls{irl} sur des problèmes jouet.

Le \textbf{\autoref{hier-lstdmu}} décrit notre première contribution, l'algorithme \gls{lstdmu}. Nous commençons par rappeler la définition de l'attribut moyen et constatons que, composante par composante, l'attribut moyen est une fonction de valeur (avec pour récompense la composante correspondante du vecteur d'attribut). Pour remplacer la méthode usuelle de calcul de l'attribut moyen (une estimation de Monte-Carlo, \autoref{alg:mcmu}) nous nous proposons d'exploiter le constat que nous venons de faire, non pas en utilisant les outils de la \gls{dp} (\autoref{alg:dpmu}), qui nécessitent de connaître les probabilités de transition, mais en adaptant un algorithme d'estimation de la fonction de valeur bien connu en \gls{rl}, \gls{lstd}. L'algorithme obtenu, \gls{lstdmu}, est une généralisation vectorielle de \gls{lstd}-$Q$. La complexité algorithmique asymptotique des deux algorithmes est la même car l'opération coûteuse, une inversion de matrice, n'a besoin d'être effectuée qu'une seule fois pour toutes les composantes de l'attribut moyen. Nous montrons comment les résultats théoriques disponibles sur \gls{lstd} peuvent être adaptés à \gls{lstdmu}. Nous plaçons ensuite cette contribution dans le cadre des approches d'\gls{irl} itératives dont la structure a été isolée au chapitre précédent. Les algorithmes d'\gls{rl} tels \gls{lspi} permettant de résoudre un \gls{mdp} grâce à une base de données recueillie à l'avance, nous pouvons maintenant grâce à \gls{lstdmu}, qui peut estimer l'attribut moyen d'une politique arbitraire grâce à cette même base de données, faire fonctionner ce type d'algorithmes d'\gls{irl} de manière \gls{batch} et \gls{offpolicy}. Ce mode de fonctionnement nous permet d'espérer les appliquer à des problèmes où le modèle dynamique du système n'est pas disponible. Nous validons empiriquement notre démarche sur le problème jouet du pendule inversé, et constatons que l'estimation fournie par \gls{lstdmu} permet à un algorithme d'\gls{irl} itératif d'atteindre les mêmes performances que lorsqu'il utilise l'estimation de l'attribut moyen fournie par un oracle ayant accès à un simulateur du système. La source d'erreurs semble donc se situer au niveau de la résolution du \gls{mdp} à chaque itération. Il apparaît important de définir des algorithmes d'\gls{irl} ne nécessitant pas une telle résolution répétée.

Nous proposons un tel algorithme au \textbf{\autoref{hier-scirl}}. Rappelant les notions vues aux chapitres \ref{hier-formalisme} et \ref{hier-etatdelart}, nous établissons un parallèle entre la fonction de score d'un classifieur et la fonction de qualité de l'expert. Pour trouver l'action (l'étiquette) à associer à un état (une entrée), les classifieurs à fonction de score choisissent l'action qui associée à l'état obtient le score le plus haut. L'expert mettant en œuvre un politique optimale, il va utiliser le même mécanisme glouton de sélection d'action, non pas sur une fonction de score mais sur la fonction de qualité de sa politique. Grâce à la modélisation linéaire de la fonction de récompense (objet de la recherche menée par les algorithmes d'\gls{irl}), cette fonction de qualité peut s'exprimer comme le produit scalaire du vecteur de paramètres de la récompense et de l'attribut moyen de l'expert. Il est dont naturel d'utiliser également pour la fonction de score du classifieur une paramétrisation linéaire mettant en jeu l'attribut moyen de l'expert. Le principe de \gls{scirl}, le nouvel algorithme d'\gls{irl} que nous proposons, est simplement d'apprendre directement la fonction de récompense grâce à un classifieur à fonction de score linéaire. Il suffit d'utiliser l'attribut moyen de l'expert dans la paramétrisation linéaire. L'algorithme de classification structurée étudié au début du manuscrit se prête bien à cette tâche. Nous proposons de plus une heuristique qui, en conjonction avec une méthode \gls{onpolicy} d'estimation de l'attribut moyen de l'expert (comme par exemple l'algorithme \gls{lstdmu} du chapitre précédent\footnote{\gls{lstdmu} peut en effet fonctionner, comme \gls{lstd}, au choix en mode \gls{onpolicy} ou \gls{offpolicy}.
 }), permet d'utiliser \gls{scirl} uniquement à partir de données échantillonnées par l'expert, ce qu'à notre connaissance aucun autre algorithme d'\gls{irl} ne peut faire sur un problème non trivial.

Nous fournissons \autoref{thm:scirl} une borne théorique sur les performances de \gls{scirl}. Cette borne garantit que lorsque l'erreur d'estimation de l'attribut moyen de l'expert et l'erreur de classification sont faibles, alors la politique de l'expert est quasi-optimale pour la récompense trouvée par \gls{scirl}, ce qui est un des critères de réussite exprimés au \autoref{hier-formalisme}. Le \autoref{thm:coroscirl} laisse espérer que \gls{scirl} évite les solutions dégénérées au problème de l'\gls{irl}. Les performances de \gls{scirl} vis-à-vis du second critère de réussite du \autoref{hier-formalisme} font l'objet d'une étude empirique. Notre instanciation de \gls{scirl}, avec l'algorithme de classification structurée discuté plus tôt en conjonction avec l'heuristique permettant le fonctionnement à partir de données expertes seules, est testée sur les problèmes jouet du \gls{mountaincar} et du \gls{highway}. Sur le premier, \gls{scirl} obtient des résultats significativement meilleurs que ceux de \gls{relent} et ceux d'un algorithme d'imitation supervisée. Sur le second, les deux approches d'\gls{irl} montrent de meilleures performances que la classification, et \gls{scirl} reste légèrement meilleur que \gls{relent}.

Basé sur le même constat de proximité entre fonction de score d'un classifieur et la fonction de qualité de l'expert, l'algorithme \gls{cascading} fait l'objet du \textbf{\autoref{hier-cascading}}. Plutôt que d'introduire la structure temporelle du \gls{mdp} par le biais de l'attribut moyen de l'expert, nous utilisons une version échantillonnée de l'équation de Bellman, où la fonction de qualité de l'expert est remplacée par la fonction de score d'un classifieur entraîné sur les données expertes. La classification intervient ici comme sous-routine. Elle est la première étape d'une cascade. La seconde est une étape de régression où l'équation de Bellman inversée permet d'obtenir des données grâce auxquelles un régresseur apprend la fonction de récompense. Une nouvelle heuristique est proposée. En utilisant cette heuristique il est possible de définir les données fournies à l'étape de régression uniquement à partir de données expertes. Nous reproduisons \autoref{thm:csi} les résultats théoriques obtenus par un autre doctorant de l'équipe MaLIS. Si l'erreur de classification et l'erreur de régression sont faibles, alors la politique de l'expert est quasi-optimale pour la récompense trouvée par \gls{cascading}. L'autre critère, à savoir l'optimalité vis-à-vis de la récompense de l'expert\footnote{Normalement inconnue, mais nous travaillons sur des problèmes jouet.
 } de l'agent entraîné sur la récompense trouvée par \gls{cascading} est là aussi l'objet d'une étude expérimentale sur les mêmes problèmes jouet que pour \gls{scirl}. La classification s'avérant être moins performante que les méthodes d'\gls{irl}, nous ne gardons dans cette étude que \gls{relent} comme point de comparaison. Les performances de \gls{cascading} sont sensiblement équivalentes à celles de \gls{scirl}, un test d'égalité des moyennes montre qu'elles sont par endroit légèrement supérieures.

Le \textbf{\autoref{hier-conclusion}} résume les contributions apportées et propose des perspectives de recherche, l'étude de certaines ayant été entamée sans avoir pu être menée à terme dans le cadre de cette thèse. Il serait notamment utile de renforcer l'étude empirique de nos deux nouveaux algorithmes d'\gls{irl} en les appliquant à des problèmes réels. Les difficultés rencontrées lors des travaux préliminaires en ce sens (voir la \autoref{hier-autrestravaux}) sont liées à l'optimisation de la récompense trouvée par les algorithmes d'\gls{irl} et non à la recherche de la récompense elle-même. Une solution à ce problème serait le développement d'algorithmes d'\gls{irl} fournissant une politique de contrôle et non simplement une récompense.

Les trois annexes regroupent pour l'\textbf{\autoref{hier-demo}} la preuve des principaux résultats théoriques, pour l'\textbf{\autoref{hier-jouets}} une description des problèmes jouet sur lesquels nous évaluons empiriquement les nouvelles méthodes que nous proposons et pour l'\textbf{\autoref{hier-contributions}} une liste des publications et communications auxquelles cette thèse a donné lieu.
\chapter{Apprentissage par imitation}
\label{sec-2}
\label{hier-formalisme}

  L'\gls{irl} permet de transformer l'apprentissage de tâches complexes à partir d'un signal de récompense en apprentissage par imitation, en déduisant de démonstrations d'un expert un signal de récompense cohérent avec ces démonstrations. Ce chapitre formalise les notions qui nous seront utiles tout au long du manuscrit. Nous abordons l'apprentissage par imitation par le biais de l'apprentissage supervisé (discrimination). Ces méthodes ne passent pas par une fonction de récompense, mais apprennent directement l'association que fait l'expert entre entrées sensorielles et actions sur l'environnement. Nous en présentons un état de l'art. Nous précisons ensuite le cadre de l'\gls{rl}, puis de l'\gls{irl}, dont nous ferons l'état de l'art au \autoref{hier-etatdelart}.
\section{Imitation supervisée}
\label{sec-2-1}
\label{hier-nonari}

Nous réduisons dans cette section le problème de l'imitation supervisée à celui de la classification. Après une étude rapide des notions centrales d'attribut et de généralisation, nous étudions deux cadres de classification que nous retrouverons aux chapitres \ref{hier-scirl} et \ref{hier-cascading} lorsque nous présenterons nos deux nouveaux algorithmes d'\gls{irl}. Nous terminons par un état de l'art de l'imitation supervisée.
\subsection{Formalisme}
\label{sec-2-1-1}

   Un agent, qu'il soit artificiel ou humain, qu'il s'agisse de l'expert, qu'il soit en cours d'apprentissage ou même qu'il agisse de manière aléatoire (utile pour l'exploration), met en œuvre une politique. D'un point de vue mathématique, une politique $\policy$ est formellement définie comme une application d'un espace d'état vers un espace d'action\footnote{Une autre manière de formaliser les choses repose sur les politiques stochastiques définies dans $\twosetsfunctions{\statespace\times\actionspace}{[0,1]}$ (ignorons le formalisme des espaces probabilisés pour cette discussion). Cela complexifie un peu l'analyse, et la perte de généralité lorsque l'on se cantonne aux politiques déterministes est minimale (dans un cadre markovien).
 } :
\begin{equation}
\policy \in \twosetsfunctions{\statespace}{\actionspace}.
\end{equation}

   Cette politique encode le comportement de l'agent : dans un état $\state \in \statespace$, l'agent choisira l'action $\policy(\state) \in \actionspace$. On constate que ce formalisme implique que pour choisir son action, l'agent n'utilise que les informations contenues dans l'état. Il faudra donc en pratique veiller à ce que celui-ci contienne toutes les informations utiles à la prise de décision, c'est-à-dire par exemple pour un système physique, non seulement les valeurs courantes des capteurs, mais aussi peut-être certaines valeurs passées afin de pouvoir calculer des taux de variation.

   Les espaces d'état et d'action ne sont généralement pas des espaces vectoriels : la plage de valeurs que peut prendre une composante est rarement illimitée. Il s'agit même parfois d'ensembles finis. Dans ce cas, on s'autorisera à manipuler les états comme les $|\statespace|$ premiers entiers, puisqu'il existe un isomorphisme entre $\statespace$ et $\llbracket 1;|\statespace|\rrbracket$. C'est dans les problèmes qui nous intéressent quasiment toujours le cas pour l'espace d'action. Nous considérons par défaut qu'il s'agit d'un ensemble fini de petit cardinal. La plupart des notations que nous utiliserons supposent un espace d'état également fini, mais l'usage d'attributs judicieusement choisis (\autoref{hier-attributs}) permet dans la plupart des cas l'extension pratique au cas continu.
\subsection{Classification}
\label{sec-2-1-2}
\label{hier-classification}

   Il est possible de voir le problème de l'imitation comme celui de la recherche d'une politique correspondant à celle de l'expert, si possible pour chaque état ou au mieux en fonction des contraintes en mémoire et en temps de l'agent. Il est en effet évident que si les politiques des deux agents sont identiques alors ces deux agents accomplissent la même tâche avec le même degré d'efficacité. Même lorsque la politique de l'expert (notée $\expertpolicy$) est intégralement connue, cette formulation n'est pas forcément dénuée d'intérêt, il peut en effet être souhaitable de remplacer l'expert par un agent moins coûteux mais probablement plus limité. Apprendre parfaitement (ou apprendre au mieux avec quelques erreurs) la politique de l'expert est alors sensé.

   Bien souvent, cependant, il est impossible de connaître intégralement la politique de l'expert, ne fut-ce que parce que l'espace d'état est trop grand. Il faut alors se contenter d'exemples sur un certain nombre $\nbsamples$ d'états \footnote{L'indice $sa$ de $\satrace{\expertpolicy}$ indique quels éléments (ici des couples $(\state_{\datasetindex},\action_{\datasetindex})$) composent la base. A l'\autoref{eq:sastracedef}, par exemple, ce sont des quadruplets $(\state_i,a_i,s'_i,r_i)$ qui sont utilisés.
 }:
\begin{equation}
\satrace{\expertpolicy} = \{(\state_{\datasetindex},\action_{\datasetindex}=\expertpolicy(\state_{\datasetindex})) | \datasetindex \in \llbracket 1;\nbsamples\rrbracket\}.
\end{equation}

   Le problème de l'imitation se trouve ainsi réduit à celui de la classification. Étant donné que l'espace d'action est fini et de petit cardinal, chaque action est vue comme une étiquette à appliquer à un état. La démonstration de l'expert fournit la base d'entraînement.

La classification est un problème plus subtil que celui de naïvement apprendre par cœur la base d'entraînement. Ce que nous cherchons à optimiser n'est pas la performance sur la base d'entraînement fournie, mais la performance sur l'espace d'état en général. Plus précisément, certains états nous intéressent plus que d'autres. Pour une justification intuitive, il suffit de penser aux jeux de plateau, où bien agir dans les quelques états qui apparaissent souvent en début de partie est beaucoup plus intéressant que bien agir dans un état improbable que l'on ne rencontrera peut-être jamais (d'où par exemple le travail sur les ouvertures aux échecs). Pour mesurer l'importance accordée à un état, on définit une fonction de poids homogène à une densité de probabilité\footnote{Par abus de notation, nous allons identifier loi de probabilité et densité de probabilité, ce qui nous amènera à écrire des choses comme $s\sim\rho$, même si $\rho$ est défini comme une densité et non comme une loi.
 } et qui donc somme à un : $\spacedistrib \in \twosetsfunctions{\statespace}{[0,1]}$ telle que $\sum\limits_{\state \in \statespace} \spacedistrib(\state) = 1$. La mesure de l'erreur d'une politique de classification $\classifpolicy$ se basant uniquement sur la base d'entraînement (avec $\indicatorfunc$ la fonction indicatrice) :
\begin{equation}
\empiricalclassiferror = {1\over \nbsamples}\sum_{(\state_{\datasetindex},\action_{\datasetindex}) \in \satrace{\expertpolicy}} \indicatorfunc{\classifpolicy(\state_{\datasetindex}) \neq \action_{\datasetindex}}
\end{equation}
est différente de celle que l'on cherche réellement à optimiser :
\begin{align}
\classiferror{\spacedistrib} &= \sum_{\state \in \statespace} \spacedistrib(\state)\indicatorfunc{\classifpolicy(\state) \neq \expertpolicy(\state)}\\
&= \expectationknowing{ \indicatorfunc{\classifpolicy(\state) \neq \expertpolicy(\state)}}{\state\sim\spacedistrib}.
\label{eq:classiferror}
\end{align}
Il existe des conditions de consistance du principe de minimisation empirique du risque. Cependant lorsque la taille de la base d'apprentissage et la complexité du modèle ne sont pas compatibles, les problèmes de sur-apprentissage peuvent apparaître.

L'hypothèse de base en discrimination est que la distribution $\spacedistrib$ pour laquelle on cherche à minimiser l'erreur de classification est la même que celle selon laquelle on échantillonne les données. Il est alors possible de se baser sur le risque empirique pour résoudre le problème de la classification.

Il peut être nécessaire de choisir une autre distribution $\spacedistrib$ sur laquelle optimiser la classification. On peut, comme on l'a vu intuitivement, favoriser les états de départ. Il est possible de tenter d'estimer à partir d'une base d'exemples la vraie distribution des états qui seront soumis au contrôle de l'agent (qui peut être différente de la distribution à laquelle l'expert est confronté). On peut également, pour certains environnements, accorder plus d'importance à certains états critiques où une erreur aurait des conséquences fâcheuses (si l'on dispose de cette connaissance \emph{a priori}).
\subsection{Attributs}
\label{sec-2-1-3}
\label{hier-attributs}

La capacité d'une méthode d'imitation à extrapoler un contrôle satisfaisant à partir de données dont la distribution n'est pas forcément celle sur laquelle il nous importe que l'agent soit performant s'appelle la capacité de généralisation. La capacité de généralisation d'un modèle est affectée par la manière dont l'espace d'état lui apparaît. Les descripteurs des données sont souvent choisis de manière à permettre une modélisation linéaire\footnote{Comme nous le verrons tout au long du manuscrit, la méthode d'approximation que nous nous préparons à expliquer peut être utilisée pour la fonction de score d'un classifieur, la fonction de valeur ou de qualité en \gls{rl} et pour la fonction de récompense et l'attribut moyen en \gls{irl}.
 }  , ceci afin de faciliter leur exploitation. Rien ne garantit qu'une fonction linéaire dans l'espace d'état-action $\statespace\times\actionspace$ dispose d'un pouvoir de représentation en adéquation avec la complexité des objets à approximer. Pour élargir l'espace d'hypothèses, on va choisir de travailler avec des fonctions linéaires dans un espace d'attributs $\featurestateactionspace$ de dimension $\dimphi$ qui est l'image de l'espace d'état-action $\statespace\times\actionspace$ par une fonction vectorielle d'attribut $\featurestateactionfunc \in \twosetsfunctions{\statespace \times \actionspace}{(\reals^{\dimphi})}$. Par exemple, le classifieur que nous étudierons \autoref{hier-taskar} recherche une fonction de score telle que :
\begin{equation}
q_{\paramclassif} = \transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}.
\end{equation}
Les approximations linéaires comme celle-ci ont plusieurs propriétés qui les rendent populaires :
\begin{itemize}
\item elles permettent de remplacer la recherche dans un espace fonctionnel par une recherche dans l'espace des paramètres $\reals^{\dimphi}$ (elles ne sont bien sûr pas seules à permettre cela) ;
\item deux vecteurs de paramètres $\paramclassif$ et $\paramclassif'$ proches dans cet espace (tels que $||\paramclassif - \paramclassif'||_2$ soit petit) correspondent à deux fonctions $q_{\paramclassif}$ et $q_{\paramclassif'}$ également proches. Par contraste, dans les réseaux de neurones par exemple, la non linéarité des fonctions d'activation peut faire que deux jeux de paramètres semblables correspondent à deux fonctions radicalement différentes ;
\item les fonctions linéaires sont aisément dérivables en fonction du vecteur de paramètre. Le gradient est le vecteur d'attribut :
\end{itemize}
\begin{equation}
\subgrad_{\paramclassif} q_{\paramclassif} = \subgrad_{\paramclassif} (\transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}) = \featurestateactionfunc_{\classifscorefunc};
\end{equation}
\begin{itemize}
\item la mise en œuvre informatique est triviale, il s'agit du produit scalaire du vecteur de paramètres et du vecteur d'attributs.
\end{itemize}
Si l'on dispose d'une fonction d'attribut $\featurestatefunc \in \twosetsfunctions{\statespace}{(\reals^{\dimpsi})}$ sur l'espace d'état, une technique classique pour obtenir une fonction d'attribut sur l'espace d'état-action consiste à distribuer la représentation sur les différentes actions. D'un vecteur de dimension $\dimpsi$, on passe à un vecteur de dimension $\dimphi = \card{A}\dimpsi$ (où $\card{\cdot}$ dénote le cardinal d'un ensemble) en définissant :
\begin{equation}
\featurestateactionfunc(\state,\action) = \begin{pmatrix}
\indicatorfunc{\action=\action_1}\featurestatefunc(\state)\\
\vdots\\
\indicatorfunc{\action=\action_{\card{\actionspace}}}\featurestatefunc(\state)\\
\end{pmatrix}.
\end{equation}

Le choix d'une bonne fonction d'attribut sur l'espace d'état est extrêmement dépendant du problème, néanmoins dans de nombreux cas deux techniques simples donnent de bons résultats. Dans le cas d'un espace d'état fini de taille raisonnable, il est possible de définir une fonction d'attribut binaire en associant un unique indice à chaque état. Le vecteur d'attribut d'un état est nul partout sauf en l'indice associé à l'état :
\begin{equation}
\featurestatefunc(\state) = \begin{pmatrix}
\indicatorfunc{\state=\state_1}\\
\vdots\\
\indicatorfunc{\state=\state_{\card{\statespace}}}\\
\end{pmatrix}.
\label{eq:attributsdiscrets}
\end{equation}
Un avantage de ce schéma est qu'il permet une représentation exacte de la fonction de score. En effet le produit $q_{\paramclassif}(\state,\action) = \transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}(\state,\action)$ revient à isoler la composante de $\paramclassif$ correspondant à l'unique indice associé au couple $(\state, \action)$. Les deux gros désavantages sont l'incapacité de ce schéma à passer à l'échelle et l'absence totale de structure : on aura beau disposer d'énormément d'information sur les ``voisins'' d'un élément de l'espace, tant que l'on aura pas vu précisément cet élément dans la base d'exemple, c'est la valeur d'initialisation de la coordonnée correspondante dans $\paramclassif$ qui sera utilisée.

Pour les espaces continus, une paramétrisation usuelle consiste à paver l'espace de fonctions à base radiale, telles les gaussiennes. On assigne un nombre $\gaussperdim{\dimindex}$ à chacune des dimensions $0 < \dimindex \leq \dimstate$ de l'espace d'état et l'on construit un maillage de $\dimpsi = \prod_{\dimindex=1}^{\dimstate}\gaussperdim{\dimindex}$ points $\gaussiancenter_{\dimindexbis}$, $0<\dimindexbis\leq\dimpsi$ équirépartis dans l'espace qui seront les centres des $\dimpsi$ composantes gaussiennes de la fonction d'attribut. L'écart-type pour une dimension $\dimindex$ peut être choisi par exemple comme :
\begin{equation}
\gaussianvar^{\dimindex} = { \max(\component{\state}{\dimindex})-\min(\component{\state}{\dimindex})\over 2 \gaussperdim{\dimindex}},\textrm{ avec $\component{\state}{\dimindex}$ la $\dimindex^\textrm{ième}$ composante de $\state$.}
\end{equation}
En notant :
\begin{equation}
\gaussian{\gaussiancenter}{\gaussianvar}{s} = \exp\left(-\sum_{\dimindex=1}^{\dimstate}{(\component{s}{\dimindex}-\component{\gaussiancenter}{\dimindex})^2\over
2(\component{\gaussianvar}{\dimindex})^2}\right),
\end{equation}
on obtient finalement la fonction d'attribut suivante après l'ajout d'une composante constante\footnote{Formellement, on utilise du fait de l'ajout de ce terme constant une modélisation affine plutôt que linéaire.
 } :
\begin{equation}
\featurestatefunc(s) = \begin{pmatrix}
\gaussian{\gaussiancenter_{1}}{\sigma}{s}\\
\vdots\\
\gaussian{\gaussiancenter_{\dimpsi}}{\sigma}{s}\\
1
\end{pmatrix}.
\label{eq:gaussianrbf}
\end{equation}
Contrairement à la fonction d'attribut binaire précédente, celle-ci possède une structure spatiale. Les scores de deux états topologiquement proches subiront l'influence de la même composante du vecteur de paramètres. Le nombre de gaussiennes du réseau croît de manière exponentielle avec le nombre de dimensions. Une illustration de ce type de vecteur d'attributs sur l'espace d'état bidimensionnel du \gls{mountaincar} \footnote{Il s'agit d'un problème jouet où une voiture doit sortir d'une vallée, mais n'est pas assez puissante pour gravir la côte du premier coup. Ce problème est décrit en \autoref{hier-mountaincar}.
 } est présentée Figure 2.1.

Lorsque cela s'avère nécessaire, l'opérateur peut effectuer un travail d'ingénierie en construisant un vecteur d'attribut spécifique au problème courant, qui tient compte des informations \emph{a priori} disponibles sur celui-ci. Par exemple pour la fonction de valeur d'un agent sur le problème du \gls{mountaincar}, on pourrait envisager de calculer l'énergie cinétique et l'énergie potentielle. Ce n'est cependant pas la route que nous allons suivre. Notre but étant d'établir des modèles génériques, nous utiliserons dans nos expériences les attributs gaussiens et binaires que nous venons de décrire. 
\begin{figure*}
\begin{tikzpicture}[scale=2.3]
%\draw [help lines] (0,0) grid (6,-6);
\node at (0,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x0}};
\node at (0,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x0}};
\node at (0,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x0}};
\node at (0,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x0}};
\node at (0,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x0}};
\node at (0,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x0}};
\node at (0,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x0}};

\node at (1,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x1}};
\node at (1,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x1}};
\node at (1,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x1}};
\node at (1,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x1}};
\node at (1,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x1}};
\node at (1,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x1}};
\node at (1,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x1}};

\node at (2,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x2}};
\node at (2,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x2}};
\node at (2,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x2}};
\node at (2,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x2}};
\node at (2,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x2}};
\node at (2,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x2}};
\node at (2,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x2}};

\node at (3,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x3}};
\node at (3,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x3}};
\node at (3,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x3}};
\node at (3,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x3}};
\node at (3,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x3}};
\node at (3,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x3}};
\node at (3,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x3}};

\node at (4,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x4}};
\node at (4,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x4}};
\node at (4,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x4}};
\node at (4,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x4}};
\node at (4,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x4}};
\node at (4,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x4}};
\node at (4,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x4}};

\node at (5,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x5}};
\node at (5,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x5}};
\node at (5,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x5}};
\node at (5,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x5}};
\node at (5,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x5}};
\node at (5,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x5}};
\node at (5,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x5}};

\node at (6,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_6x6}};
\node at (6,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_5x6}};
\node at (6,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_4x6}};
\node at (6,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_3x6}};
\node at (6,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_2x6}};
\node at (6,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_1x6}};
\node at (6,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_psi_0x6}};
\end{tikzpicture}
\caption[Attributs gaussiens sur le problème du \glstext{mountaincar}]{Attributs gaussiens sur le problème du \gls{mountaincar}. L'espace d'état est pavé de $7\times 7 = 49$ gaussiennes elliptiques dont les centres sont répartis à équidistance. Les variances sont les mêmes pour toutes les gaussiennes et dépendent de la plage de valeurs sur une dimension. Toute fonction aux variations raisonnables peut être approximée de manière correcte par une somme pondérée de ces gaussiennes. Cette représentation est à comparer à la Figure 2.2 
% \autoref{fig:mountain_car_mu}
où est présenté l'attribut moyen qui, lui, prend en compte la dynamique temporelle du problème.}
\label{fig:mountain_car_psi}
\end{figure*}
\clearpage
\subsection{Classification à marge structurée}
\label{sec-2-1-4}
\label{hier-taskar}

Illustrons ce propos par l'étude d'une approche de classification qui utilise une fonction de score linéairement paramétrée sur l'espace d'attribut état-action $\featurestateactionspace$ \citep{taskar2005learning}.

Le principe quasi-ubiquitaire en classification\footnote{Les arbres de décision formant un contre-exemple notable \citep{safavian1991survey}.
 } de la fonction de score est le suivant : à chaque couple état-action une fonction $\classifscorefunc$ associe un score. Pour associer une action à un état, le classifieur passe simplement en revue toutes les actions (on voit donc l'intérêt d'un petit espace d'action) et choisit celle qui associée à cet état obtient le score le plus haut :
\begin{align}
\classifscorefunc &\in \twosetsfunctions{\statespace \times \actionspace}{\reals},\\
\forall \state, \classifpolicy(\state) &\in \arg\max_{\action \in \actionspace} \classifscorefunc(\state,\action).
\label{eq:classifscore}
\end{align}
Apprendre une bonne fonction de score permet donc de résoudre le problème de classification. L'approche proposée dans \citep{taskar2005learning} prend le parti d'une fonction de score linéaire :
\begin{equation}
q_{\paramclassif}(\state,\action) = \transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}(\state,\action).
\end{equation}
Ce que nous cherchons maintenant est donc un bon vecteur de paramètres $\paramclassif$.

Considérant que nous disposons d'un vecteur d'attributs permettant de continuer, examinons à des fins d'illustration le problème que \citet{taskar2005learning} se proposent de résoudre, problème que nous retrouverons \autoref{hier-scirl}. Il s'agit d'un problème d'optimisation convexe sous contraintes\footnote{Pour voir que les variables $\slack_{\datasetindex}$ sont positives, on peut poser $a=a_i$ (et considérer que $\margin(s_i,a_i)=0$).
 } :
\begin{align}
&\min\limits_{\slack \in \reals_{+}} \frac{\lambda}{2}\|\omega\|^2+{1\over \nbsamples}\sum\limits_{\datasetindex=1}^{\nbsamples}\slack_{\datasetindex}\\
\textrm{sous les contraintes }\forall \datasetindex, \classifscorefunc_{\paramclassif}(\state_{\datasetindex},\action_{\datasetindex}) \geq &\max\limits_{\action \in \actionspace}(\classifscorefunc_{\paramclassif}(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action)) - \slack_{\datasetindex}.
\end{align}
La fonction objectif cherche naturellement à réduire les variables d'écart $\slack_{\datasetindex}$ tandis que les contraintes sont telles que la valeur prise par la fonction de score $\classifscorefunc_{\paramclassif}$ en un couple $(\state_{\datasetindex}, \action_{\datasetindex})$ correspondant à une décision experte est supérieure ou égale au meilleur score. Elle doit être non seulement supérieure au score seul, mais supérieure d'une certaine marge $\margin(\state_{\datasetindex},\action)$ qui contribue à donner à ce classifieur sa capacité de généralisation. En effet, on constate que si $\margin$ est uniformément nulle, alors parvenir à minimiser parfaitement la fonction de coût se réduit à apprendre par cœur la base d'exemples, c'est-à-dire à probablement subir les effets du sur-apprentissage. Fixer 
\begin{equation}
\margin(\state_{\datasetindex},\action) = \begin{cases}
0 &\textrm{si }\action = \action_{\datasetindex}\\
1 &\textrm{si }\action \neq \action_{\datasetindex}
\end{cases}
\label{eq:margin}
\end{equation}
permet de donner aux choix de l'expert un score strictement supérieur aux scores des autres choix. \citet{taskar2005learning} précisent qu'il est possible d'adapter la marge $\margin$ en fonction de la qualité des choix alternatifs, un bon choix correspondant à une petite marge. Nous verrons qu'en pratique la marge binaire que nous venons de suggérer fonctionne assez bien pour les problèmes qui nous intéressent.

La résolution informatique d'un problème d'optimisation convexe sous contraintes pouvant être malaisée, nous en cherchons une formulation plus simple (qui revient à celle employée dans \citep{ratliff2007imitation}) en constatant que si une contrainte de bon classement est violée, alors compte-tenu de la nature de la contribution empirique à la fonction objectif, la valeur de la variable d'écart correspondante est obtenue en traitant ladite contrainte comme une contrainte-égalité :
\begin{equation}
\slack_{\datasetindex} =  \max\limits_{\action \in \actionspace}(\classifscorefunc_{\paramclassif}(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action)) - \classifscorefunc_{\paramclassif}(\state_{\datasetindex},\action_{\datasetindex}).
\end{equation}
En posant :
\begin{align}
\best{\action}_{\datasetindex} &= \arg\max_{\action \in \actionspace}(\classifscorefunc_{\paramclassif}(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action))\\
 &= \arg\max_{\action \in \actionspace} (\transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action)),
 \end{align}
et en faisant monter les contraintes dans la fonction objectif, on obtient alors une simple fonction de coût à minimiser :
\begin{align}
\structuredcost(\paramclassif) &= {1\over \nbsamples} \sum_{\datasetindex=1}^{\nbsamples} (\classifscorefunc_{\paramclassif}(\state_{\datasetindex},\best{\action}_{\datasetindex}) + \margin(\state_{\datasetindex},\best{\action}_{\datasetindex}) - \classifscorefunc_{\paramclassif}(\state_{\datasetindex},\action_{\datasetindex})) +\frac{\lambda}{2}\|\omega\|^2\\  &= {1\over \nbsamples} \sum_{\datasetindex=1}^{\nbsamples} (\transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}(\state_{\datasetindex},\best{\action}_{\datasetindex}) + \margin(\state_{\datasetindex},\best{\action}_{\datasetindex}) - \transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}(\state_{\datasetindex},\action_{\datasetindex}))+\frac{\lambda}{2}\|\omega\|^2.
\end{align}
Cette fonction n'est pas différentiable à cause de l'opérateur non-linéaire $\max$ caché dans le terme $\best{\action}_{\datasetindex}$, mais la généralisation du gradient qu'est le sous-gradient permet de contourner cette difficulté.
Le sous-gradient de la fonction de coût est :
\begin{equation}
\subgrad\structuredcost(\paramclassif) = \sum_{\datasetindex=1}^{N}(\featurestateactionfunc_{\classifscorefunc}(\state_{\datasetindex},\best{\action}_{\datasetindex}) - \featurestateactionfunc_{\classifscorefunc}(\state_{\datasetindex},\action_{\datasetindex}))+\lambda \omega,
\end{equation}
il est donc possible de résoudre le problème d'optimisation original en effectuant une simple descente de sous-gradient pour minimiser la fonction de coût $J(\paramclassif)$, comme cela est décrit dans l'\autoref{alg:taskar}.
\begin{algorithm}
\Entree{
Une base d'entraînement établie par l'expert $\satrace{\expertpolicy}$\;}
\Donnees{
Une fonction d'attribut $\featurestateactionfunc_{\classifscorefunc}$\;
}
\Sortie{Une règle de décision $\classifpolicy$}
Initialiser $\paramclassif$ arbitrairement\;
Procéder à la descente de sous-gradient dont le sous-gradient est :
\begin{equation*}
\subgrad\structuredcost(\paramclassif) = \sum_{\datasetindex=1}^{N}\left[\featurestateactionfunc_{\classifscorefunc}\left(\state_{\datasetindex},\max_{\action\in\actionspace}\transpose{\paramclassif}\featurestateactionfunc(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action)\right) - \featurestateactionfunc_{\classifscorefunc}(\state_{\datasetindex},\action_{\datasetindex})\right] + \lambda \omega\;
\end{equation*}
\Retour{\begin{equation*}
\classifpolicy : \state \rightarrow \classifpolicy(\state) \in \arg\max_{\action\in\actionspace} \transpose{\paramclassif}\featurestateactionfunc_{\classifscorefunc}(\state,\action)
\end{equation*}}
\caption{Algorithme de classification structurée}
\label{alg:taskar}
\end{algorithm}

Nous avons présenté cette technique de classification plus en détail à des fins d'illustration de l'importance du choix des attributs, et également car nous la retrouverons \autoref{hier-scirlinstan} lorsque nous nous intéresserons à une nouvelle technique d'\gls{irl}.

Il existe bien d'autres moyens de faire de la classification, comme par exemple les \gls{svm}.
\subsection{\glsentrytext{svm}}
\label{sec-2-1-5}
\label{hier-svm}

    Une \gls{svm} est un séparateur linéaire. Plus précisément, une \gls{svm} va chercher, dans un espace de représentation des entrées, l'hyperplan de marge maximale séparant les données : d'un côté de l'hyperplan se trouvent toutes les entrées associées à une certaine étiquette, et de l'autre côté se trouvent les entrées associées à l'autre étiquette. Il faut positionner l'hyperplan dans l'espace de représentation de manière à maximiser la distance entre celui-ci et les données.

    Il est illusoire de croire que l'espace de description est construit d'une manière qui permette à un hyperplan de séparer les exemples sans erreur. Le classifieur à fonction de score paramétrée linéairement que nous avons décrit précédemment compense cela par une projection dans un espace d'attributs (ou espace de représentation) $\featurestateactionspace$. Les \gls{svm} utilisent une technique similaire, l'idée est de remplacer les produits scalaires $\scalarprod{\state}{\state'}_{\statespace}$ qui apparaitraient dans l'algorithme par un produit scalaire $\scalarprod{\featurestatefunc(\state)}{\featurestatefunc(\state')}_{\Psi}$ dans l'espace d'attributs $\Psi$, dans l'espoir que les données soient linéairement séparables dans cet espace. L'astuce, dite astuce du noyau \citep{aizerman1964theoretical,boser1992training} est d'observer que les données n'apparaissent dans l'algorithme que par le biais de leur produit scalaire ; plutôt que de définir explicitement cet attribut $\featurestatefunc$ et de calculer le produit scalaire associé, un noyau $\kernel$ est défini. Pour peu que $\kernel$ satisfasse certaines conditions\footnote{\`A savoir être continu, symétrique et semi-défini positif.
 }, alors il existe \citep{mercer1909functions,berlinet2004reproducing} une fonction $\featurestatefunc$ telle que \footnote{Le noyau $\kappa$ engendre un \emph{Reproducing Kernel Hilbert Space} (RKHS) \citep{berlinet2004reproducing}.
 }:
\begin{equation}
\kernel(\state,\state') = \scalarprod{\featurestatefunc(\state)}{\featurestatefunc(\state')}_{\Psi}.
\end{equation}

Un noyau populaire, le noyau gaussien :
\begin{equation}
\kernel(\state,\state')=\exp\left(- \frac{\|\state - \state'\|^2}{2 \sigma^2}\right),
\end{equation}
permet de manipuler le produit scalaire $\scalarprod{\featurestatefunc(\state)}{\featurestatefunc(\state')}_\Psi$ où  l'espace d'attributs $\Psi$ engendré par $\featurestatefunc$ est de dimension infinie.

Le modèle que nous venons de décrire est celui de la \gls{svm} à marge dure \citep{boser1992training}. Comme dans le cas du classifieur à marge de la \autoref{hier-taskar}, il est possible d'introduire des variables d'écart. On passe alors au modèle de la \gls{svm} à marge douce \citep{cortes1995support}.

Pour l'imitation, la multiplicité des actions possibles appelle l'utilisation d'une \gls{svm} multi-classe (M-SVM). Un modèle unifiant les machines de ce type est proposé dans \citep{guermeur2012generic}. Pour simplifier l'exposé nous nous sommes cantonné au cas bi-classe. Le classifieur à marge de la sous-section précédente peut être instancié (si l'on choisit convenablement les valeurs des hyper-paramètres) comme un cas particulier d'une \gls{svm}, mais les deux modèles ne sont pas équivalents dans leur forme générique.

Les \gls{svm} sont très populaires et de nombreux logiciels les mettent en œuvre. Il est donc facile de les mettre en œuvre sur les problèmes jouet que nous allons traiter ; Ce sont des classifieurs pour lesquels la sélection de modèle est assez bien maîtrisée (\citet{bonidal2013analyse} a récemment présenté une thèse abordant ce sujet).
\subsection{Imitation par apprentissage supervisé de la politique}
\label{sec-2-1-6}
\label{hier-myopie}

Apprendre la politique de l'expert de manière supervisée à l'aide d'une base d'exemples peut s'avérer efficace, comme le démontrent plusieurs approches. Dans \citep{ratliff2007imitation}, les auteurs utilisent le classifieur à marge décrit plus haut
pour apprendre une politique experte pour un problème de locomotion quadrupède et sur un problème de manipulation d'objets. Le choix des attributs est automatisé grâce à une technique de \gls{boosting} similaire à \citep{friedman2001greedy,mason1999functional}.

Le boosting permet de déplacer de manière intelligente le problème du choix des attributs, sans le régler totalement. Il reste en effet à construire l'espace où choisir les nouveaux attributs. Un espace trop simple ne permettrait pas de minimiser efficacement la fonction de coût, tandis qu'un espace trop riche permettrait de l'annuler, mais sans doute au prix d'un sur-apprentissage aux conséquences fâcheuses. C'est donc cet espace qui doit être construit afin de donner au classifieur ses capacités de généralisation. \citet{ratliff2007imitation} proposent d'utiliser un réseau de neurones.

Plus brutale\footnote{Puisqu'elle repose sur l'utilisation d'une base d'exemple exhaustive.
 }, l'approche de \citet{lecun2006off} utilise un \gls{cnn} (réseau de convolution) à 6 couches pour apprendre une association directe entre une image (stéréo) et un angle de braquage (la tâche à apprendre est la conduite d'un véhicule en terrain libre). Le problème de la généralisation est résolu en exigeant une base d'entraînement couvrant au maximum l'espace d'état. Les auteurs ne cachent pas la difficulté de constituer une telle base qui doit réunir des conditions de terrain et d'illumination variées tout en exigeant un comportement extrêmement cohérent et prédictible de la part de l'opérateur humain et ce sur un grand nombre de trajectoires (il faut réunir près d'une centaine de milliers d'échantillons). En contrepartie de ces efforts, la technique proposée est robuste et ne nécessite aucun travail d'ingénierie au niveau des attributs, puisque la politique apprise associe directement la sortie du capteur à la consigne de l'actuateur du robot. Bien que cela soit moins problématique aujourd'hui avec l'augmentation de puissance des équipements embarqués, elle semble également plus rapide (dans l'exploitation, non dans l'apprentissage) que l'état de l'art de l'époque. Elle améliore les résultats notamment par rapport à \gls{alvinn} \citep{pomerleau1993knowledge} en ceci que la résolution des caméras peut être augmentée sans une trop grande explosion du nombre de connexions du réseau grâce à l'usage de la convolution et non d'un réseau complètement connecté, et que la tâche apprise est plus difficile, il s'agit de conduire en terrain libre et non de suivre une route.

Nous venons de voir deux outils différents (\gls{boosting} et réseau de convolution) permettant d'apprendre une politique sans contrôler le système, de manière supervisée, avec une intervention humaine minimale : soit l'on dispose de suffisamment de données pour que le risque empirique soit proche du risque réel, soit l'on construit des attributs tels que l'apprentissage au mieux (en minimisant une fonction de coût exprimée sur les données) ne soit pas un apprentissage par cœur, mais un apprentissage généralisant sur tout l'espace d'état. Apprendre une politique de manière locale, c'est-à-dire en se concentrant trop sur une base de données ne couvrant pas tout l'espace d'état, n'est pas satisfaisant. Cela donne un résultat fragile, l'agent sera en effet pris au dépourvu s'il a à contrôler le système dans une configuration différente de celles sur lesquelles il a été entraîné : il ne dispose ni d'information relative au comportement de l'expert dans une telle situation, ni d'information sur la tâche à accomplir qui lui permettraient de déduire ce que pourrait être ce comportement.

L'idée de demander ces échantillons de manière interactive a été proposée afin de minimiser la quantité de données nécessaire à l'apprentissage de la politique experte. Un exemple d'une telle approche est décrit par \citet{chernova2007confidence}. Des \gls{gmm} sont appris à partir d'une base de données experte de départ, puis l'agent applique la politique apprise tout en demandant à l'expert, lorsque l'incertitude est trop grande, de lui fournir un échantillon supplémentaire. Cette approche permet de limiter la redondance de la base d'entraînement et de guider l'échantillonnage vers les zones intéressantes de l'espace d'état, ce qui est également une solution au problème de la généralisation : quand l'agent ne sait pas généraliser, il demande à l'expert. Cela est connu sous le nom d'apprentissage actif \citep{cohn1996active}.

Une autre approche d'apprentissage actif est l'algorithme \gls{smile} proposé par \citet{ross2010efficient}. A chaque itération $\timeindex$ une politique mixante $\policy_{\timeindex}$ \footnote{Une politique mixante est une sorte de méta-politique qui choisit aléatoirement quelle politique exécuter parmi un ensemble de politiques, généralement de manière non uniforme, certaines politiques étant préférées à d'autres.
 } est définie à l'aide d'une politique de classification $\classifpolicy$ apprise sur une base de données experte. Cette base de données est construite sur la distribution d'états induite par l'application de la politique mixante de l'itération précédente, $\policy_{\timeindex-1}$. La politique mixante à l'itération $\timeindex$ est une pondération de trois politiques : la politique mixante de l'itération précédente $\policy_{\timeindex-1}$, la politique apprise par le classifieur à l'itération courante $\classifpolicy$ et la politique initiale $\policy_{0}$ qui se réfère systématiquement à l'expert. L'algorithme \gls{dagger} (\citet{ross2010reduction}) maintient une base d'entraînement de plus en plus grande grâce à une politique mixante entre une politique de classification et la politique de l'expert. Les états-actions traversés par la politique mixante courante ayant donné lieu à une décision de l'expert sont ajoutés à la base d'entraînement et à l'itération suivante le classifieur est entraîné sur la nouvelle base. Le poids accordé à la politique de l'expert diminue pour laisser petit à petit le contrôle à l'agent. La logique derrière ces deux approches est d'éviter de faire de l'apprentissage supervisé sur la distribution fixe de l'expert $\expertdistrib$, pour la remplacer par la distribution que l'agent rencontrera vraiment, et qui tient compte des erreurs qu'il peut parfois commettre. Grâce au schéma interactif où l'expert donne peu à peu le contrôle à l'agent, l'expert est parfois entraîné loin de sa distribution et peut alors montrer à l'agent comment corriger l'erreur que celui-ci a commise.

L'apprentissage direct de la politique experte est parfois intégré à un cadre plus large, où les notions de hiérarchie et de but apparaissent.

La classification par arbre de décision a été appliquée à l'apprentissage d'un plan de vol par \citet{sammut1992learning}. L'application est ambitieuse ; \emph{apprendre} \footnote{La conception par des humains d'un système de pilotage automatique précède l'ordinateur puisque la démonstration d'un prototype a été faite à Paris en 1914.
 } à piloter un avion, même en simulation, n'est pas une mince affaire puisqu'il faut en temps réel prendre en compte un grand nombre de facteurs pour décider d'une action parmi un éventail assez large. L'apprentissage automatique nécessite un grand nombre d'échantillons. Un comportement cohérent est exigé de l'expert humain (à un point tel que les démonstrations de deux experts ne peuvent être mélangées en une seule base d'entraînement). L'aspect automatique de l'approche est limité à l'apprentissage d'une politique par phase de vol. La détection de la phase de vol courante et donc le choix de la politique de contrôle à appliquer est effectué par des règles d'origine humaine.

De fait, cette approche a été le point de départ de nombreuses améliorations. Le travail présenté par \citet{stirling1995churps} (appelé \gls{churps}) consiste à déduire un contrôleur à partir d'une description du modèle d'évolution du système et du but à atteindre. Pour automatiser la création de ces descriptions, tâche réclamant un travail difficile car nécessitant de décrire des mécanismes précis à l'aide d'un langage contraignant, \citet{bain2000framework} proposent d'utiliser les données de l'expert. Les règles complexes ainsi apprises étant ajoutées à l'espace d'action, il est possible d'apprendre de manière automatique un classifieur plus concis que celui de \citet{sammut1992learning}, et nécessitant moins de données expertes. L'architecture proposée utilise la logique du premier ordre et donc le raisonnement symbolique. Cela permet d'introduire explicitement des connaissances expertes dans le système. Ces connaissances peuvent être acquises semi-automatiquement : les prédicats sont construits à la main et les paramètres sont appris grâce aux données de vol comme le proposent \citet{srinivasan1998inductive}. La sémantique des symboles (ici, virage, altitude, trajectoire de vol, etc.) est très liée au problème concerné. Retrouver la puissance des techniques d'apprentissage symbolique sur un autre problème nécessite d'effectuer de nouveau le difficile travail de définition des symboles et prédicats. D'autres éléments potentiellement rédhibitoires sont la mauvaise gestion des fluctuations aléatoires dues à la dynamique et la difficulté d'exprimer la tâche à accomplir en utilisant un langage symbolique. Dans une approche hybride symbolique/numérique, \citet{shiraz1997combining} proposent à l'expert soit de décrire la tâche symboliquement, soit d'en démontrer l'exécution. Les phases les plus délicates (par exemple l'atterrissage) n'ont pu être décrites et ont été démontrées. La facilité d'exploitation des règles symboliques rentre en conflit avec la difficulté qu'il y a à les définir, à l'inverse la relative facilité de génération d'une base d'exemples se heurte à la difficulté qu'il y a à généraliser à partir de celle-ci.

Une autre approche utilise les notions de hiérarchie et de but, mais de manière quelque peu différente. Plutôt que d'utiliser la logique des prédicats, ce sont les principes de programmation impérative qui se voient assistés par l'apprentissage supervisé. Dans \citep{saunders2006teaching}, ce sont les \gls{knn} qui sont utilisés pour l'apprentissage supervisé d'une politique. Les attributs sont construits à la main à partir des valeurs de sortie des capteurs du robot, et portent une sémantique forte et explicite (distance, angle), donc pratique pour l'exploitation par un opérateur humain. Les problèmes de généralisation de l'apprentissage supervisé sont contournés par l'intégration dans un cadre beaucoup plus riche : l'opérateur humain peut élargir l'espace d'action à volonté, soit en définissant une séquence d'actions qui seront exécutées en série de manière déterministe, soit en proposant des exemples du comportement souhaité en précisant ou non un état-but correspondant à la situation dans laquelle on souhaite voir le robot une fois la politique exécutée. Ces exemples servent alors à l'apprentissage d'une politique de manière supervisée, cette politique est ajoutée en tant qu'action et son exécution pourra être déclenchée dans le cadre d'une autre politique, de niveau d'abstraction plus grand. Cette hiérarchisation des comportements permet de limiter l'effort humain, d'optimiser l'utilisation des exemples et de rapidement mettre en place des comportements complexes par la création de nouveaux niveaux d'abstraction.

L'apprentissage supervisé est dans les approches que nous venons de citer utilisé comme sous routine d'un système beaucoup plus large, dans lequel l'expertise humaine explicite reste le moyen central permettant la généralisation des comportements.

Le principal problème de l'apprentissage direct de la politique de manière supervisé est, pour reprendre le terme de \citep{ratliff2009learning}, sa \emph{myopie}. Pour compenser le fait que l'on travaille au niveau d'abstraction le plus bas, celui du choix immédiat d'une action en fonction des informations contenues dans un état transitoire, les approches que nous venons de détailler font apparaître en filigrane la notion de but : l'expert n'agit en effet pas à tâtons mais dirige le système en fonction de critères qu'il paraît difficile d'exprimer au niveau d'une simple politique. On se repose donc sur une formulation plus ou moins explicite (dans le choix des attributs, dans la définition de la base d'exemples, dans l'introduction de règles logiques ou dans la définition d'une hiérarchie) de ce but, mais toujours d'origine humaine. Nous allons voir qu'il est possible de formaliser cette notion de but tout en continuant à travailler avec une politique et des échantillons semblables à ceux auxquels nous sommes habitués. Nous verrons par la suite que le but de l'expert, formalisé de cette manière, peut alors être automatiquement déduit d'une base d'échantillons acquise une fois pour toutes. 
\section{Cadre des \glsentrytext{mdp} pour la prise de décisions séquentielles}
\label{sec-2-2}
\label{hier-cadre}

Pour comprendre ce but de l'expert qu'il nous importe de connaître, ce n'est pas au niveau du choix état-action que décrit la politique qu'il faut regarder, mais à un niveau d'abstraction plus élevé : la dynamique que la politique de l'expert impose au système. La notion qui nous manque pour entamer le raisonnement est celle de l'effet d'une action. Nous ne nous sommes préoccupés que du choix de l'action en fonction de l'état courant sans nous soucier de ce que ce choix allait imposer comme contraintes sur le prochain état que l'agent va rencontrer. Afin de pouvoir considérer la politique de l'expert non plus comme un ensemble décousu d'associations état-action, mais comme un outil capable de produire des séquences d'actions pertinentes au sens d'un critère long terme, nous formalisons la notion de dynamique temporelle.
\subsection{Dynamique temporelle}
\label{sec-2-2-1}
\label{hier-tempdyn}

L'agent (qu'il s'agisse de l'expert ou d'un agent artificiel que l'on entraîne) aux manettes du système contrôle celui-ci non pas ponctuellement de temps à autres mais de manière cohérente sur un laps de temps durant lequel il devra opérer des actions de contrôle les unes après les autres. Il est donc naturel d'indexer ces actions et les états traversés par un indice temporel $\timeindex \in \llbracket 0;\infty\rrbracket$. Par exemple, cette formulation n'impose pas de pas d'échantillonnage constant, il s'agit ici d'ordonner les états et actions par ordre de causalité, ce qui incidemment correspond\footnote{\`A moins que \emph{Doctor Who} et \emph{Retour vers le futur} ne soient des documentaires.
 } à un indice temporel croissant, non pas de transcrire avec quelque fidélité les problèmes de l'échantillonnage temporel. Qui plus est cette formulation correspond à la réalité du contrôle numérique, intrinsèquement discret.

Pour prendre en compte les imperfections de la modélisation ou plus simplement parfois la nature réellement stochastique du problème, les effets d'une action qui sont décrits par une loi de probabilité, qui, informée d'un état $\state_{\timeindex}$ et d'une action $\action_{\timeindex}$, prédit vers quel état $\state_{\timeindex+1}$ le système va transiter. On note cela (encore une fois en identifiant loi de probabilité et densité de probabilité)\footnote{Avec $f\in\twosetsfunctions{A}{B\times C\times D}$, la notation $f(b,\cdot,d)$ désigne la fonction de $C$ dans $A$ : $c \rightarrow f(b,c,d)$.
 } :
\begin{equation}
\state_{\timeindex+1}\sim \transprobfunceval{\state_{\timeindex}}{\action_{\timeindex}}{\cdot}, \transprobfunc \in \twosetsfunctions{\statespace \times \actionspace \times \statespace}{[0;1]}.
\end{equation}
On constate que l'information sur l'état vers lequel on transite ne dépend que de l'état courant et de l'action courante, et non de la trajectoire. C'est la propriété de Markov qui donne son nom au \gls{mdp}.
La répétition du cycle consistant à choisir une action puis à transiter vers un nouvel état où l'agent choisit une action qui le fera transiter vers un nouvel état (etc.) forme une trajectoire. Les probabilités de transitions contraintes par une politique $\policy$ peuvent être dans le cas fini $\card{\statespace}<\infty$ représentées\footnote{En utilisant l'isomorphisme naturel entre $\statespace$ et $\llbracket 1,|\statespace|\rrbracket$.
 } par une matrice stochastique de taille $\card{\statespace}\times\card{\statespace}$ :
\begin{equation}
\transprobmat{\policy} = \matrixbyterm{\transprobfunceval{\state}{\policy(\state)}{\state'}}{\state}{\state'},
\end{equation}
où l'on note: $\left(f(i,j)\right)_{i,j}$ la matrice dont le terme à la ligne $i$ et à la colonne $j$ est $f(i,j)$.
Un agent de politique $\policy$ va visiter certains états plus que d'autres. Pour quantifier cela, il est possible d'utiliser la matrice des probabilités de transition que nous venons juste de définir. Le terme à la ligne $\state$ et à la colonne $\state'$ est la probabilité $\transprobfunceval{\state}{\policy(\state)}{\state'}$ que l'agent se trouve dans l'état $\state'$ au temps $\timeindex+1$ s'il était en $\state$ au temps $\timeindex$. Si l'on multiplie la matrice $\transprobmat{\policy}$ par elle même, le terme général (ligne $\state$, colonne $\state''$) du résultat est :
\begin{equation}
\sum_{\state'\in\statespace} \transprobfunceval{\state}{\policy(\state)}{\state'} \transprobfunceval{\state'}{\policy(\state')}{\state''},
\end{equation}
il s'agit de la probabilité pour l'agent de passer de $\state$ à $\state''$ en deux pas de temps. Si (et seulement si) l'agent peut espérer se trouver en chacun des états en temps fini, alors ce que l'on appelle la distribution stationnaire $\stationarydistrib{\policy}$ existe et est unique \citep{norris1998markov}. Elle est définie telle que :
\begin{equation}
\transpose{\stationarydistrib{\policy}} \transprobmat{\policy} = \transpose{\stationarydistrib{\policy}}.
\end{equation}
On peut donc la voir comme la probabilité pour l'agent de se trouver en un certain état, après un temps infini, quel que soit l'état de départ. Cette condition d'ergodicité que nous plaçons généralement sur le \gls{mdp} contraint par la politique de l'expert implique l'existence de la distribution stationnaire de l'expert $\expertdistrib$. Un \gls{mdp} contraint par une politique correspond à une chaine de Markov.
\subsection{Récompense et valeur}
\label{sec-2-2-2}

Dans les approches vues précédemment, le but était défini comme des valeurs spécifiques que doivent prendre certaines composantes de l'état (par exemple pour le pilotage, une certaine altitude) ou certaines valeurs de commandes acceptables dans certaines zones de l'espace d'état. Il est au premier abord assez naturel de définir une consigne comme cela. Pour peu que l'espace d'état-action soit construit d'une manière qui permet l'analyse sémantique, l'opérateur humain n'a pas trop de mal à exprimer ce qu'il souhaite que la machine fasse en définissant les états et actions désirables et ceux qu'il faut éviter. Charge à la machine de trouver comment se placer dans les états désirables en évitant les états problématiques. Nous formalisons cela sous la forme d'une fonction de récompense. Il s'agit d'un jugement local de l'intérêt qu'il y a à se trouver en un certain état ou appliquer certaines actions en certains états :
\begin{equation}
\rewardfunc \in \twosetsfunctions{\statespace\times \actionspace}{\reals}.
\end{equation}
Nous reviendrons \autoref{hier-recompense} sur la définition de la fonction de récompense.

Il faut maintenant que ce critère local donne lieu à un comportement globalement intéressant. Comment, à l'échelle d'une politique choisissant une action pour un état, parvenir à un contrôle tenant compte de la dynamique complète du système ? Il faut qu'une politique $\policy$ soit jugée dans son ensemble sur la trajectoire qu'elle impose à l'agent. Mathématiquement nous souhaitons optimiser la valeur de la politique\footnote{Pour simplifier, à l'avenir nous noterons le conditionnement par la dynamique induite par une politique comme cela : $\expectationknowing{\cdot}{s_0=s,\policy}$
 } :
\begin{align}
\label{eq:Vdefsum}
\rlvalue{\policy}{\rewardfunc}(\state,\action)&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\rewardfunc(\state_{\timeindex},\action_{\timeindex})}{\state_0 = s}\\
\textrm{avec }\forall \timeindex \in \llbracket 1;\infty\rrbracket, \state_{\timeindex} &\sim \transprobfunceval{\state_{\timeindex-1}}{\policy(\state_{\timeindex-1})}{\cdot}.
\end{align}
Comme l'horizon temporel est infini, pour s'assurer de la convergence de la somme, le facteur d'amortissement $\discount \in [0;1[$ est introduit. Une conséquence est la diminution d'attrait des récompenses loin dans le futur au profit de récompenses accessibles plus rapidement. Cela permet de récompenser l'agent qui effectue rapidement sa tâche.

La fonction de valeur, et un peu plus loin la fonction de qualité (\autoref{eq:qdef}) sont paramétrées par la fonction de récompense $\rewardfunc$ car par la suite nous serons amenés à étudier la valeur selon une certaine récompense d'une politique qui est optimale pour une autre récompense.

L'ensemble de l'espace d'état $\statespace$, de l'espace d'action $\actionspace$, des probabilités de transitions $\transprobfunc$, de la fonction de récompense $\rewardfunc$ et du facteur d'amortissement $\discount$ forment un \gls{mdp} $\mdpbis$ \citep{puterman1994markov}
\begin{equation}
\mdpbis = \left\{\statespace, \actionspace, \transprobfunc, \rewardfunc, \discount\right\}
\end{equation}
dans lequel le problème de la prise de décision séquentielle pour le contrôle optimal peut être formulé.

Nous recherchons une politique optimale $\optimalpolicy{\rewardfunc}$ telle qu'en tout état sa valeur soit supérieure ou égale à celle de tout autre politique $\pi$ :
\begin{equation}
\forall \state, \rlvalue{\optimalpolicy{\rewardfunc}}{\rewardfunc}(\state) \geq \rlvalue{\policy}{\rewardfunc}(\state).
\label{eq:optimalite}
\end{equation}
Pour résoudre ce problème, intéressons nous de plus près à l'expression de la valeur d'une politique, dont la définition, donnée \autoref{eq:Vdefsum}, peut être transformée en une expression récursive (grâce à la propriété de Markov) :
\begin{equation}
\label{eq:BellmanEval}
\rlvalue{\policy}{\rewardfunc}(\state) = \rewardfunc(\state,\policy(\state)) + \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\policy(\state)}{\state'} \rlvalue{\policy}{\rewardfunc}(\state').
\end{equation}
C'est l'équation d'évaluation de \citet{bellman2003dynamic} qui est à l'origine de l'opérateur d'évaluation de Bellman :
\begin{align}
\textrm{En définissant : }R_\pi(s) &= R(s,\pi(s))\\
\bellmanevalopeval{\policy}{\rewardfunc}{} &\in \twosetsfunctions{\twosetsfunctions{\statespace}{\reals}}{(\twosetsfunctions{\statespace}{\reals})}\\
\forall V \in \twosetsfunctions{\statespace}{\reals}, \bellmanevalopeval{\policy}{\rewardfunc}{V} &= R_\pi + \discount \transprobmat{\policy} V. 
\label{eq:bellmanevop}
\end{align}

Cette opérateur est contractant, par conséquence il possède un unique point fixe. L'équation de définition de ce point fixe :
\begin{equation}
V = \bellmanevalopeval{\policy}{\rewardfunc}{V}
\end{equation}
est exactement la même que l'équation d'évaluation de Bellman (\autoref{eq:BellmanEval}). L'unique point fixe de l'opérateur $\bellmanevalopeval{\policy}{\rewardfunc}{}$ est donc la fonction de valeur de la politique : $\rlvalue{\policy}{\rewardfunc}$. Lorsque l'on dispose des probabilités de transition il est possible de calculer la valeur d'une politique en inversant l'équation de Bellman :
\begin{equation}
\rlvalue{\policy}{\rewardfunc} = (I-\discount\transprobmat{\policy})^{-1}\rewardfunc_\pi
\end{equation}

Dans l'équation d'évaluation de Bellman, l'action qui fait passer de $\state$ à $\state'$ est explicitement donnée comme étant $\policy(\state)$. Les actions suivantes sont également choisies par la politique $\policy$ comme l'indique le terme $\rlvalue{{\color{red}\policy}}{\rewardfunc}$. Imaginons maintenant que connaissant la valeur d'une politique $\policy$, nous soyons chargé pour l'état $\state$ de choisir la meilleur action $\action$, qui peut être différente de $\policy(\state)$, mais qu'ensuite la politique $\policy$ reprenne le contrôle. C'est le degré de liberté décrit par la fonction de qualité $\quality{\policy}{\rewardfunc}$ :
\begin{align}
\label{eq:qualitydef}
\quality{\policy}{\rewardfunc} &\in \twosetsfunctions{\statespace \times \actionspace}{\reals}\\
\quality{\policy}{\rewardfunc}(\state,\action) &= \rewardfunc(\state,\action) + \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\action}{\state'} \rlvalue{\policy}{\rewardfunc}(\state').
\label{eq:qdef}
\end{align}
Notre meilleur choix consisterait à maximiser la fonction de qualité, c'est-à-dire à rendre le contrôle à $\policy$ dans l'état $\state'$ dans lequel sa valeur est maximale. En effectuant ce choix sur chacun des états de $\statespace$, on définit une politique gloutonne :
\begin{equation}
\greedy{\policy}: \state\rightarrow \arg\max_{\action\in\actionspace}\quality{\policy}{\rewardfunc}(\state,\action). 
\label{eq:greedydef}
\end{equation}
Cette agglomération de choix localement optimisés permet un optimisation beaucoup plus générale. La politique gloutonne que nous venons de définir est le meilleur choix pour un problème d'optimisation plus large :
\begin{equation}
\greedy{\policy} = \arg\max_{\policy'}\bellmanevalopeval{\policy'}{\rewardfunc}{\rlvalue{\policy}{\rewardfunc}}.
\end{equation}
L'opérateur associé :
\begin{equation}
\bellmanoptopeval{\rewardfunc}{V} = \max_{\policy}\bellmanevalopeval{\policy}{\rewardfunc}{V}
\label{eq:bellmanoptop}
\end{equation}
est l'opérateur d'optimalité de Bellman. Contractant lui aussi, il admet donc un unique point fixe qui se trouve être la fonction de valeur optimale $\rlvalue{\optimalpolicy{\rewardfunc}}{\rewardfunc}$. Toute politique gloutonne vis-à-vis de la politique optimale associée est également une politique optimale. Ainsi : 
\begin{align}
\optimalpolicy{\rewardfunc} &\in \greedy{\optimalpolicy{\rewardfunc}}\label{eq:greedy01}\\
\forall \state \in \statespace, \optimalpolicy{\rewardfunc}(\state) &\in \arg\max_{\action\in \actionspace}\quality{\optimalpolicy{\rewardfunc}}{\rewardfunc}(\state,\action).
\label{eq:rlargmax}
\end{align}
Il est intéressant de noter que, grâce à la fonction de valeur, l'optimisation ``myope'' répétée à l'échelle du choix d'une action pour un état mène à une optimisation à l'échelle de l'espace d'état complet, au niveau de la politique. Grâce à la prise en compte des probabilités de transitions, la fonction de valeur fait le lien entre le court et le long terme.

Lorsque les probabilités de transitions sont connues sur un espace d'état fini, on peut de manière exacte résoudre le problème du contrôle optimal grâce à la \gls{dp} \citep{puterman1994markov}. Les choses se corsent lorsque ces probabilités de transition sont inconnues ou que les algorithmes de \gls{dp} deviennent non tractables (espace d'état trop grand, principalement). Il faut alors avoir recours à l'\gls{rl} \citep{sutton1998reinforcement}. L'\gls{rl} permet, par interactions répétées avec le système, d'apprendre à contrôler celui-ci. Les probabilités de transition étant souvent difficiles à exprimer, nous allons porter notre attention sur l'\gls{rl} de préférence à la \gls{dp}.
\subsection{Algorithmes d'\glstext{rl}}
\label{sec-2-2-3}
\label{hier-algosrl}

La \gls{dp} supposant les probabilités de transition connues, si l'on connaît la valeur d'une politique, il est possible de calculer la fonction de qualité associée grâce à l'\autoref{eq:qdef}, et donc d'en déduite un contrôle glouton grâce à l'\autoref{eq:greedydef}.

Puisqu'en \gls{rl} la dynamique ne peut être connue que par interaction avec le système, on apprend directement la fonction de qualité $Q$ afin de pouvoir en déduire un contrôle glouton sans avoir à passer par l'\autoref{eq:qdef}.

Il existe plusieurs schémas d'\gls{rl}. Certains comme par exemple $Q$-learning \citep{watkins1989learning} adaptent le principe d'itération de la valeur \citep{bellman2003dynamic}. Une amélioration permettant le fonctionnement hors ligne, à partir d'une base de données acquise une fois pour toutes est présentée par \citet{gordon1995stable}, c'est l'algorithme \emph{fitted}-$Q$, que nous exposons \autoref{alg:fittedq}. Le choix du régresseur de l'étape de la \autoref{alg:fittedq:regression} étant laissé libre, \emph{fitted}-$Q$ a donné lieu à de nombreux travaux exploitant cette liberté. Citons par exemple l'utilisation de réseaux de neurones par \citet{riedmiller2005neuralQ} ou d'arbres par \citet{ernst2005tree}.

\begin{algorithm}
\Entree{Un ensemble d'entraînement $\sasrtrace{\rewardfunc}$\;}
\Sortie{Une politique presque optimale}
Initialiser $Q$ arbitrairement \;
\Pour{$t\in \llbracket1;T\rrbracket$}{
Initialiser $D_{saq}^Q \leftarrow \varnothing$\;
\Pour{$(s_i,a_i,s'_i,r_i)\in \sasrtrace{\rewardfunc}$}{
$D_{saq}^Q = D_{saq}^Q \cup \{\left(\state_{\datasetindex},\action_{\datasetindex}\right) , r_i + \max\limits_{a'}Q(s'_{\datasetindex},a')\}$\;
}
$Q\leftarrow$ régresseur entraîné sur $D_{saq}^Q$\nllabel{alg:fittedq:regression}\;
}
\Retour{$s\rightarrow \arg\max_{a\in \actionspace}Q(s,a)$}
\caption{Fitted-$Q$}
\label{alg:fittedq}
\end{algorithm}

 Nous allons nous intéresser à une autre famille, les algorithmes d'itération de la politique \footnote{L'idée vient de la \gls{dp} : \citep{howard1960dynamic}
 }. Il s'agit de répéter un cycle consistant à évaluer la fonction de qualité d'une politique puis de remplacer la politique par la politique gloutonne vis-à-vis de la fonction de qualité que l'on vient de calculer.

L'évaluation de la politique peut être effectuée grâce à \gls{lstd} de \citet{bradtke1996linear}. \gls{lstd} utilise un schéma classique, celui de l'approximation linéaire de la fonction de valeur\footnote{On utilise parfois le terme de ``fonction de valeur'' indistinctement pour la fonction de valeur et la fonction de qualité.
 }. On réduit alors le problème du contrôle optimal au choix du vecteur de paramètres $\paramrl$ donnant la meilleure approximation de la fonction de qualité optimale :
\begin{equation}
\hat{Q}^{\optimalpolicy{\rewardfunc}}_{\rewardfunc}(\state,\action) = \transpose{\paramrl}\featurestateactionfunc_{Q}(\state, \action).
\label{eq:paramrl}
\end{equation}
Comme pour la classification, le choix des fonctions d'attributs $\featurestateactionfunc_{Q}$ n'est pas anodin car il va fortement influencer la qualité du contrôle.

Pour trouver les paramètres $\paramrl$ de la qualité d'une politique $\policy$, \gls{lstd} minimise (dans le sens des moindres carrés) la distance entre $\bellmanevalopeval{\policy}{\rewardfunc}{}\hat{Q}^\pi_R$ et son projeté dans l'espace d'hypothèse induit par $\featurestateactionfunc_Q$.

L'utilisation de \gls{lstd} dans un schéma d'itération de la politique donne l'algorithme d'\gls{rl} \gls{lspi} proposé par \citet{lagoudakis2003least} et résumé \autoref{alg:lspi}.

\begin{algorithm}
\Entree{Une base de données $\sasrtrace{\rewardfunc}$\;
Un critère d'arrêt $\epsilon$\;}
\Donnees{Un coefficient de régularisation $\reg$}
\Sortie{
Une politique quasi-optimale $\policy$\;
}
$\Delta \leftarrow \infty$\;
Initialiser $\paramrl$ arbitrairement\;
Initialiser $\policy$ arbitrairement\;
\Tq{$\Delta>\epsilon$}{
$\paramrl'\leftarrow \paramrl$\;
$\paramrl \leftarrow (A+\reg I)^{-1}b$ avec : 
\begin{equation*}
A=  \sum_{(s_i,a_i,s'_i,r_i)\in\sasrtrace{\rewardfunc}}\featurestateactionfunc_{Q}(\state_i,\action_i)\transpose{\left(
\featurestateactionfunc_{Q}(\state_i,\action_i) - \discount \featurestateactionfunc_{Q}(\state'_i,\policy(\state'_i))
\right)}
\end{equation*}
et 
\begin{equation*}
b = \sum_{(s_i,a_i,s'_i,r_i)\in\sasrtrace{\rewardfunc}}\featurestateactionfunc_{Q}(\state_i,\action_i)r_i
\end{equation*}
$\policy \leftarrow \greedy{\policy}: \state\rightarrow \arg\max_{\action\in\actionspace}\hat Q^{\policy}_{\rewardfunc}(\state,\action) $\;
$\Delta \leftarrow \|\paramrl'-\paramrl\|_2$\;
}
\Retour{\policy}
\caption{\glstext{lspi}}
\label{alg:lspi}
\end{algorithm}

Les algorithmes \ref{alg:fittedq} et \ref{alg:lspi} n'ont besoin que d'une base de données recueillies une fois pour toutes\footnote{La base $\sasrtrace{\rewardfunc}$ devant être représentative de la dynamique du \gls{mdp}, elle est généralement basée sur une base $\sastrace{\sim}$ recueillie par une politique aléatoire à laquelle on ajoute l'information de récompense. Du point de vue des notations, l'exposant n'est donc plus la politique d'échantillonnage comme par exemple dans la base $\satrace{\expertpolicy} = \{(s_i,a_i)|i\in \llbracket 1;\nbsamples\rrbracket\}$ mais la fonction de récompense $R$ telle que $r_i=R(s_i,a_i)$.
 }. Il ne s'agit pas exactement d'une base $\satrace{\expertpolicy}$ comme la classification en utilise, mais d'une base plus contraignante à réunir, qui contient l'état suivant l'action (afin d'obtenir de l'information sur la dynamique du système) et le signal de récompense (afin d'obtenir de l'information sur la qualité du contrôle) :
\begin{equation}
\sasrtrace{\rewardfunc} = \{(\state_{\datasetindex},\action_{\datasetindex}, \state'_{\datasetindex}, r_{\datasetindex} = \rewardfunc(\state_{\datasetindex},a_i)) | \datasetindex \in \llbracket 1;\nbsamples\rrbracket\}.
\label{eq:sastracedef}
\end{equation}

Les échantillons de $\sasrtrace{\rewardfunc}$ n'ont pas besoin de se suivre, on peut avoir $\state_{\datasetindex+1} \neq \state'_{\datasetindex}$. Il importe que la base d'échantillons soit représentative de la dynamique du \gls{mdp} pour chaque action. Une technique pratique et de réunir plusieurs trajectoires d'une politique aléatoire, en la faisant démarrer de différents états.

Les deux algorithmes \emph{fitted}-$Q$ et \gls{lspi} sont \gls{batch} et \gls{offpolicy}. Cela signifie respectivement qu'ils peuvent utiliser des données échantillonnées une fois pour toutes et n'ont pas besoin de contrôler le système. Ils sont donc plus efficaces en échantillons que les algorithmes en ligne (mais plus coûteux en temps de calcul) et potentiellement plus sûrs aussi, puisque l'on a pas à placer un agent au contrôle du système avant que son entraînement n'ait eu lieu.
\section{Définition de l'\glsentrytext{irl}}
\label{sec-2-3}

Nous voilà donc armés pour résoudre (de manière approchée) le problème du contrôle optimal, pour peu que l'on dispose d'une description de la tâche à accomplir sous la forme d'une fonction de récompense $\rewardfunc$ et d'échantillons représentatifs du problème.
\subsection{Définition du problème}
\label{sec-2-3-1}
\label{hier-problemdef}

La question qui se pose maintenant est de savoir dans quelle mesure il est possible d'automatiquement déduire d'une trace du comportement de l'expert une description de son but sous la forme d'une fonction de récompense. Cette méthode d'imitation a été suggérée pour la première fois par \citet{russell1998learning}. L'état de l'art sera traité au chapitre suivant, nous nous attelons ici à trouver une formulation mathématique du problème et à exposer les notions nécessaires à l'analyse des méthodes existantes et à l'introduction des nouvelles approches que nous proposons.

Outre l'intérêt intrinsèque de la découverte des motivations de l'expert, apprendre une fonction de récompense correspondant au comportement de l'expert permet de soigner les méthodes d'imitation de leur ``myopie'' (\autoref{hier-myopie}), en guidant l'agent vers une imitation non pas de la manière dont l'expert accomplit la tâche, comme le fait l'imitation directe de la politique, mais du but de l'expert.

Dans le formalisme des \gls{mdp} introduit plus tôt, l'expression du problème de l'\gls{irl} est de prime abord simple : on suppose que la politique de l'expert $\expertpolicy$ est une politique optimale pour une certaine fonction de récompense $\expertreward$ qu'il s'agit de trouver.
Les choses se corsent malheureusement extrêmement rapidement. Tout d'abord, cette formulation n'est pas un problème mathématique bien posé au sens d'\citet{hadamard1902problemes} : il existe en effet de multiples solutions, dont au moins une est dégénérée, la récompense uniformément nulle. Tous les comportements (donc celui de l'expert) ont la même valeur ($0$) pour cette récompense, donc tous sont optimaux. En inversant, la politique de l'expert est optimale pour la récompense nulle, qui est donc solution du problème.

Plus subtilement, utiliser le formalisme des \gls{mdp} présuppose que certaines conditions sont réunies, notamment en ce qui concerne les espaces d'état et d'action. Il faut être en mesure de définir un espace d'état markovien, afin qu'un agent puisse prendre sa décision quant à l'action en se basant uniquement sur les informations de l'état. Il faut également être en mesure d'obtenir une trace de l'expert (le problème se pose déjà pour l'imitation directe de la politique). L'espace d'action doit rester discret et de faible cardinal si l'on veut que l'immense majorité des approches d'\gls{irl} y restent tractables. Enfin, il faut que l'expert soit effectivement optimal, l'introduction d'erreurs\footnote{Certaines approches d'\gls{irl} relâchent la contrainte d'optimalité de l'expert par exemple par l'introduction de variables d'écart.
 } dans la démonstration pouvant poser problème.

Ces conditions, qu'il est difficile de satisfaire en totalité, ne vont cependant pas nous préoccuper. Nous nous attelons à trouver de nouvelles solutions au problème formulé par \citet{russell1998learning}. Pour mesurer notre succès, nous envisageons deux mesures de la qualité de la récompense $\appr{\rewardfunc}$ trouvée par un algorithme d'\gls{irl}. L'une est objective :
\begin{equation}
\expectationknowing{{\color{blue}V^{\pi^E}_{R^E}(s)} - {\color{brown}V^{\pi^*_{\hat{R}}}_{R^E}(s)}}{\state\sim\expertdistrib},
\label{eq:aricriterebis}
\end{equation}
l'autre est estimable :
\begin{equation}
\expectationknowing{{\color{blue}V^{\pi^*_{\hat{R}}}_{\hat{R}}} - {\color{brown}V^{\pi^E}_{\hat{R}}(s)}}{\state \sim \expertdistrib}.
\label{eq:aricritere}
\end{equation}

Ces deux critères sont positifs ou au mieux nuls, il s'agit de les minimiser. Les deux termes $\textrm{\color{blue}bleus}$ sont des fonctions de valeur optimales (puisque $\expertpolicy$ est optimale pour $\expertreward$ et que $\optimalpolicy{\appr{\rewardfunc}}$ est optimale pour $\appr{\rewardfunc}$). Ces deux critères sont donc liés à l'optimalité des valeurs $\textrm{\color{brown}marrons}$ et tendront vers $0$ quand, pour le premier la politique apprise en optimisant la récompense $\appr{\rewardfunc}$ est optimale pour la récompense inconnue $\expertreward$ et quand pour le second la récompense $\appr{\rewardfunc}$ est telle qu'elle admet la politique expert comme politique optimale.

La distribution en jeu $\spacedistrib$ est également importante. La distribution selon laquelle les données de l'expert sont tirées, $\expertdistrib$, est celle considérée par défaut, comme c'est le cas pour les méthodes supervisées (voir explication en \autoref{hier-classification}). L'\gls{irl} présente cependant sur les méthodes supervisées l'avantage de pouvoir bien mieux généraliser aux parties de l'espace d'état sur lesquelles l'expert n'a pas fourni de démonstrations\footnote{Cela est principalement dû au fait qu'une fois la récompense trouvée par un algorithme d'\gls{irl}, on l'optimise sur tout l'espace d'état grâce à un algorithme d'\gls{rl} ou de \gls{dp}.
 }, il peut donc être intéressant d'évaluer la qualité du contrôle non pas sur la distribution correspondant aux démonstrations de l'expert, mais justement sur une distribution à support plus large.

Le premier critère est celui qui nous intéresse vraiment. Il juge le comportement issu de l'optimisation par un algorithme d'\gls{rl} ou de \gls{dp} de la récompense $\appr{\rewardfunc}$ à l'aune de la description de la tâche confiée à l'expert, à savoir $\expertreward$. Cela permet dans les cas où plusieurs bonnes solutions existent de ne pas pénaliser un agent qui aurait fait un choix différent de celui de l'expert si cela importe peu. Il s'agit donc d'une mesure plus fine que l'erreur de classification $\classiferror{\expertdistrib}$, qui sanctionne toute divergence d'opinion avec l'expert. Lors du test d'algorithmes d'\gls{irl} sur des problèmes jouet où la récompense $\expertreward$ nous est connue, nous étudierons ce critère ou un critère équivalent. Sur de vrais problèmes, en revanche, il est impossible de l'estimer : il faudrait paradoxalement résoudre le problème de l'\gls{irl} pour savoir si on a bien résolu le problème de l'\gls{irl}.

Le second critère a le défaut d'être optimisé par la récompense nulle. Nous verrons cependant, en analysant les différentes approches d'\gls{irl}, les mécanismes mis en place afin d'éviter les solutions dégénérées. Il a l'avantage de pouvoir être estimé puisqu'à l'inverse de $\expertreward$, la fonction de récompense $\appr{\rewardfunc}$ est connue. Un algorithme comme \gls{lstd} peut fournir une approximation \gls{offpolicy} d'une fonction de valeur, c'est-à-dire évaluer une politique arbitraire, différente de celle qui contrôle le système.
\subsection{Attribut moyen}
\label{sec-2-3-2}
\label{hier-attribut}

    Le fonction de récompense étant l'objet recherché celle-ci sera, pour ne pas bouder une méthode éprouvée, approximée par un modèle linéaire :
    \begin{equation}
    \rewardfunc_{\paramirl}(\state,\action) = \transpose{\paramirl}\featurestateactionfunc_{\rewardfunc}(\state,\action).
    \label{eq:paramirl}
    \end{equation}
Là encore, le choix des attributs est important. Il faut qu'ils permettent à la fonction de récompense de décrire la tâche effectuée par l'expert.

Dans le contexte de l'approximation linéaire de la fonction de récompense, l'expression de la fonction de valeur d'une politique $\policy$ fait apparaître un terme que nous appellerons $\rlmu{\policy}$ dont le prochain chapitre révèlera l'importance :
\begin{align}
\label{eq:mudef}
\rlvalue{\policy}{\appr{\rewardfunc}}(\state)&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\appr{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0=s,\policy}\\
&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\transpose{\paramirl}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0=s,\policy}\\
&=\transpose{\paramirl}\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0=s,\policy}\\
&=\transpose{\paramirl}\rlmu{\policy}(\state)
\label{eq:vthetamu}
\end{align}

Ce terme, l'attribut moyen d'une politique, est une fonction vectorielle qui porte la structure temporelle du \gls{mdp} contraint par la politique. C'est, que l'on nous pardonne l'oxymore, une trace du futur qui est liée à la distribution qu'occupera l'agent à partir d'un certain état. La présence du facteur d'amortissement $\discount$ dans l'expression empêche l'identification complète à une distribution, ce qui est heureux car sans cela l'expression tendrait vers la distribution stationnaire quel que soit l'état de départ (à la projection dans $\featurestateactionspace$ près), or on souhaite que l'attribut moyen ne soit pas une fonction constante de l'espace d'état-action.

De la même manière que la fonction de qualité donne à la fonction de valeur un degré de liberté supplémentaire, il est possible de donner une action en argument à l'attribut moyen avec le même effet :
\begin{align}
\rlmu{\policy}(\state)&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0 = \state,\policy}\\
\rlmu{\policy}(\state,\action)&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0 = \state,\action_0=\action,\policy}.
\label{eq:mudefsum}
\end{align}

L'attribut moyen donne une idée des états que traversera l'agent. Il contient donc des informations sur la dynamique induite par sa politique. Cela va avoir un impact important sur la qualité des diverses approximations linéaires dont nous avons parlé si l'attribut moyen est utilisé en tant qu'attribut tout court, comme il le sera au \autoref{hier-scirl}. La présence de la dynamique du \gls{mdp} dans l'attribut moyen est illustrée  Figure 2.2.

Une propriété immédiate de l'attribut moyen\footnote{Exposée par exemple dans \citep{abbeel2004apprenticeship}.
 } est exploitée par une classe d'algorithmes d'\gls{irl} (que nous décrirons plus en détail en \autoref{hier-repetee}). Cette propriété découle directement de la définition de l'attribut moyen :
\begin{proposition}[Proximité de l'attribut moyen et valeur]
\label{thm:closemuclosev}
Soit $\policy_1$ et $\policy_2$ deux politiques sur un même \gls{mdp}, dont les attributs moyens respectifs $\rlmu{\policy_1}$ et $\rlmu{\policy_2}$ sont proches (pour la norme $\norm{2}{\cdot}$) :
\begin{equation}
\forall \state \in \statespace, \norm{2}{\rlmu{\policy_1}(s)-\rlmu{\policy_2}(s)} \leq \epsilon \in \reals.
\end{equation}
Alors quelle que soit la récompense $\rewardfunc_{\paramirl} = \transpose{\paramirl}\featurestatefunc$, les valeurs des deux politiques $\policy_1$ et $\policy_2$ sont également "proches" :
\begin{equation}
\forall s\in \statespace, \left|\rlvalue{\policy_1}{\rewardfunc_{\paramirl}}(s)-\rlvalue{\policy_2}{\rewardfunc_{\paramirl}}(s)\right| \leq \norm{2}{\paramirl}\epsilon.
\end{equation}
\end{proposition}
\begin{proof}
D'après l'\autoref{eq:vthetamu} : 
\begin{align}
\forall \state \in \statespace, \left|\rlvalue{\policy_1}{\rewardfunc_{\paramirl}}(s)-\rlvalue{\policy_2}{\rewardfunc_{\paramirl}}(s)\right|
 &= \left|\transpose{\paramirl}\rlmu{\policy_1}(s)-\transpose{\paramirl}\rlmu{\policy_2}(s)\right|\\
&= \left|\transpose{\paramirl}(\rlmu{\policy_1}(s)-\rlmu{\policy_2}(s))\right|.\\
\end{align}
En utilisant l'inégalité de Cauchy-Schwartz, on conclut : 
\begin{align}
\forall \state \in \statespace, \left|\rlvalue{\policy_1}{\rewardfunc_{\paramirl}}(s)-\rlvalue{\policy_2}{\rewardfunc_{\paramirl}}(s)\right| &\leq \norm{2}{\paramirl}\left\|\rlmu{\policy_1}(s)-\rlmu{\policy_2}(s)\right\|_2\\
&\leq \norm{2}{\paramirl}\epsilon.
\end{align}
\end{proof}
Ce résultat est assez intuitif, si deux politiques induisent les mêmes trajectoires, elles auront même valeur. La manière dont l'apprentissage supervisé résout le problème de l'imitation donne ce type de résultat. Il est intéressant de noter que dans certains cas la réciproque est fausse. Il est donc théoriquement possible de résoudre le problème de l'\gls{irl} en s'éloignant des solutions que l'apprentissage supervisé peut proposer. Il est en effet possible d'imaginer qu'un algorithme d'\gls{irl} renvoie une récompense qui, une fois optimisée, donnera une politique dont l'attribut moyen est différent de celui de la politique de l'expert, mais de valeur semblable.

\begin{figure*}
\begin{tikzpicture}[scale=2.3]
%\draw [help lines] (0,0) grid (6,-6);
\node at (0,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x0}};
\node at (0,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x0}};
\node at (0,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x0}};
\node at (0,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x0}};
\node at (0,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x0}};
\node at (0,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x0}};
\node at (0,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x0}};

\node at (1,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x1}};
\node at (1,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x1}};
\node at (1,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x1}};
\node at (1,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x1}};
\node at (1,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x1}};
\node at (1,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x1}};
\node at (1,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x1}};

\node at (2,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x2}};
\node at (2,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x2}};
\node at (2,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x2}};
\node at (2,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x2}};
\node at (2,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x2}};
\node at (2,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x2}};
\node at (2,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x2}};

\node at (3,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x3}};
\node at (3,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x3}};
\node at (3,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x3}};
\node at (3,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x3}};
\node at (3,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x3}};
\node at (3,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x3}};
\node at (3,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x3}};

\node at (4,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x4}};
\node at (4,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x4}};
\node at (4,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x4}};
\node at (4,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x4}};
\node at (4,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x4}};
\node at (4,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x4}};
\node at (4,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x4}};

\node at (5,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x5}};
\node at (5,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x5}};
\node at (5,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x5}};
\node at (5,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x5}};
\node at (5,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x5}};
\node at (5,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x5}};
\node at (5,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x5}};

\node at (6,-0) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_6x6}};
\node at (6,-1) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_5x6}};
\node at (6,-2) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_4x6}};
\node at (6,-3) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_3x6}};
\node at (6,-4) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_2x6}};
\node at (6,-5) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_1x6}};
\node at (6,-6) {\includegraphics[width=.13\textwidth]{Figures/Mountain_car_mu_0x6}};

\end{tikzpicture}
\caption[Attribut moyen de l'expert sur le \glstext{mountaincar}]{\`A comparer à la Figure 2.1
%\autoref{fig:mountain_car_psi}
, l'attribut moyen (ici celui de l'expert sur le problème du \gls{mountaincar}) est un reflet de la dynamique du système contraint par la politique. La forme en escargot, caractéristique de trajectoires en va-et-vient à l'apogée de plus en plus haute, a clairement laissé sa trace sur l'attribut moyen.}
\label{fig:mountain_car_mu}
\end{figure*}
\chapter{Problématique de l'\glstext{irl} et état de l'art}
\label{sec-3}
\label{hier-etatdelart}

  Avant d'analyser les approches existantes d'\gls{irl}, intéressons nous plus en détail à la définition précise de la fonction de récompense. Nous allons voir quelles transformations il est possible de lui appliquer sans changer les politiques optimales, une technique connue sous le nom de \gls{rewardshaping}. 
\section{Fonction de récompense}
\label{sec-3-1}
\label{hier-recompense}
\subsection{Espace de départ}
\label{sec-3-1-1}

En \gls{rl}, on utilise souvent la transition $\state,\action,\state',r$ comme unité d'information\footnote{L'algorithme SARSA utilise comme son nom l'indique une séquence augmentée de l'action $a'$ suivant l'état d'arrivée $s'$.
 }. On y trouve de l'information sur la dynamique du système : l'agent ayant choisi $a$ en $s$, il est arrivé en $s'$, ainsi que le signal de récompense. Il peut être envisagé d'utiliser une récompense dépendant de la transition complète $s,a,s'$, mais on peut facilement définir une récompense parfaitement équivalente sur $\statespace\times \actionspace$ :
\begin{equation}
R(s,a) = \sum_{\state'\in \statespace}\transprobfunceval{s}{a}{s'}R'(s,a,s').
\end{equation}

Une récompense sur les états seuls est envisageable, et permet de transférer une tâche d'un agent à l'autre sans tenir compte de l'espace d'action de chacun des agents, pour peu que tous deux aient le même espace d'état\footnote{Par exemple sur un échiquier où la récompense est définie sur l'état (la case où la pièce se trouve) on peut entraîner indistinctement une tour, une reine ou un cavalier à se déplacer à l'endroit voulu bien que ces pièces ne se déplacent pas de la même manière.
 }. Cela est cependant problématique car il y a une perte de généralité entre une récompense sur $\statespace$ et une récompense sur $\statespace\times \actionspace$ qu'il est impossible de compenser sans changer le \gls{mdp} \footnote{Par exemple sur un \gls{mdp} à deux états et deux actions : \emph{rester} et \emph{changer}, il est impossible de définir une récompense sur les états seuls qui admette pour politique optimale celle consistant à toujours choisir l'action \emph{changer}.
 }. De plus, les récompenses sont généralement définies sur $\statespace\times\actionspace$ dans la littérature.

Pour toutes ces raisons, nous considèrerons les récompenses comme définies dans $\statespace\times\actionspace$ par défaut.
\subsection{\Glstext{rewardshaping}}
\label{sec-3-1-2}

L'étude de la relation entre la condition d'optimalité d'une politique (comme celle de l'expert) et la fonction de récompense associée (que l'on cherche à connaître) nous amène à découvrir d'autres propriétés importantes pour l'analyse du problème de l'\gls{irl}. Nous nous intéressons particulièrement aux transformations qu'il est possible d'appliquer à la récompense sans perturber la relation d'ordre existant entre les valeurs des différentes politiques (relation d'ordre sur laquelle est basée la notion d'optimalité \autoref{eq:optimalite}). Nous ne souhaitons pas voir ces transformations en mesure de minimiser la fonction objectif d'un algorithme d'\gls{irl}. En effet, cela signifierait que l'on modifie la récompense sans modifier le sens qu'elle porte et donc qu'en réalité on ne résout pas le problème de l'\gls{irl} mais une formulation dégénérée de celui-ci.

Un résultat immédiat, qui découle du fait que la condition d'optimalité d'une politique est basée sur une relation d'ordre sur les valeurs, qui sont par définition linéaires vis-à-vis de la récompense, est qu'il est possible d'appliquer n'importe quelle transformation affine de coefficient strictement positif à une récompense sans en changer la signification. Ainsi notamment nous pourrons normaliser une récompense sans perte de généralité.

L'étude des transformations n'ayant pas d'impact sur les politiques optimales a été menée dans le cadre de l'\gls{rl}. L'objectif est de faciliter la recherche de politiques optimales sans modifier celles-ci. \citet{ng1999policy} trouvent une élégante condition nécessaire et suffisante à la préservation de politiques optimales par changement de récompense : à une récompense $\rewardfunc$ définie sur le triple espace état-action-état $\statespace \times \actionspace \times \statespace$, on peut ajouter toute fonction $\rewardfunc'$ définie sous la forme : 
\begin{equation}
\label{eq:rewardshaping}
\rewardfunc'(\state,\action,\state_{+1}) = \discount\rewardfunc_{\statespace}(s_{+1}) -  \rewardfunc_{\statespace}(s),
\end{equation}
où $\rewardfunc_{\statespace}$ est définie sur l'espace d'état. Cette condition est suffisante dans le sens que toute fonction de récompense définie de la sorte peut être ajoutée à une fonction de récompense de $\twosetsfunctions{\statespace \times \actionspace \times \statespace}{\reals}$ sans en changer les politiques optimales. Elle est nécessaire en ceci qu'une fonction de récompense $\rewardfunc'$ pourra être ajoutée à une autre si il existe une fonction $\rewardfunc_{\statespace}$ qui permette de l'écrire sous la forme donnée en \autoref{eq:rewardshaping}. Si une telle fonction n'existe pas, il est possible que les probabilités de transition soient telles que la modification de la fonction de récompense entraînera une modification des politiques optimales.

Maintenant que nous disposons d'une meilleure compréhension de l'objet que nous cherchons, voyons comment les approches existantes mènent cette recherche.
\section{Premières formulations du problème}
\label{sec-3-2}

   Dans la dernière partie de sa contribution, \citet{russell1998learning} pose le problème de l'\gls{irl} en termes généraux. Il ne propose pas de solution, mais explique l'intérêt du problème et isole des questions-clefs dont certaines restent encore ouvertes aujourd'hui.

   C'est dans \citep{ng2000algorithms} que les premières solutions apparaissent. Afin de trouver la récompense, les auteurs inversent l'équation de Bellman et formulent ainsi une condition nécessaire et suffisante sur la récompense pour qu'elle soit solution du problème de l'\gls{irl}. Cette condition n'écarte pas les solutions dégénérées comme la récompense nulle. En conséquence, pour les \gls{mdp} finis où les probabilités de transition sont connues, les auteurs ajoutent à cette condition nécessaire et suffisante une contrainte de pénalisation du désaccord avec l'expert ainsi qu'une contrainte de régularisation en norme $\ell_1$ et obtiennent un programme linéaire. Pour les \gls{mdp} dont l'espace d'état est grand ou infini, une paramétrisation linéaire de la fonction de récompense est naturellement choisie (rares sont les approches qui choisissent un autre schéma). De fait, et bien qu'il ne porte pas encore ce nom, l'attribut moyen de l'expert apparaît dans le développement. Une version échantillonnée du programme linéaire découlant du schéma d'approximation linéaire de la récompense est proposé. La contrainte d'optimalité repose non plus sur la connaissance des probabilités de transition, mais sur la recherche d'une récompense donnant à la politique de l'expert une plus grande valeur qu'aux politiques optimales trouvées en optimisant la récompense telle que connue pendant les itérations précédentes de l'algorithme (il faut donc résoudre un \gls{mdp} à chaque itération).

   On retrouve dans ce travail plusieurs éléments communs aux approches ultérieures. Il y a le constat primordial que contrairement à ce qui se passe pour l'\gls{rl} où l'on cherche le point fixe d'un opérateur contractant (quand on estime une fonction de valeur) et qui donc existe et est unique, on a ici affaire à un problème mal posé, et il convient d'introduire des contraintes supplémentaires afin d'éviter les solutions dégénérées. Le choix et la justification de ces contraintes est un bon moyen de différencier les approches existantes. Il y a le schéma presque ubiquitaire de l'approximation linéaire de la fonction de récompense, qui entraîne l'apparition de l'attribut moyen. Enfin, on retrouve la nécessité pour l'algorithme d'\gls{irl} de résoudre le problème de l'\gls{rl} de manière répétée.
\section{Méthodes nécessitant la résolution répétée d'un \glsentrytext{mdp}}
\label{sec-3-3}
\label{hier-repetee}

   Cette nécessité est lourde de conséquences quant à l'usage pratique d'un algorithme d'\gls{irl}. La résolution du problème de l'\gls{rl} nécessite en effet soit beaucoup de données pour sa résolution \gls{offpolicy}, soit l'accès à un simulateur ou au système réel pour sa résolution \gls{onpolicy}.

   \citet{abbeel2004apprenticeship} proposent l'algorithme \gls{pirl} dans un travail séminal dont la structure va être reprise (plus ou moins consciemment) par plusieurs autres approches. La récompense est paramétrée linéairement, de fait la notion d'attribut moyen apparaît assez logiquement (\autoref{eq:mudef}). Elle obtient même une place centrale dans cette approche itérative de l'\gls{irl}. L'idée (résumée \autoref{alg:repetee}) est, à chaque itération, d'apprendre la politique optimale pour la fonction de récompense courante (via un algorithme d'\gls{rl}). L'attribut moyen de cette politique est alors évalué. La fonction de coût que minimise l'approche joue sur une définition de la proximité $d$ entre l'attribut moyen de cette politique et l'attribut moyen de la politique de l'expert. A chaque itération cette proximité est évaluée, et le vecteur de paramètre $\paramirl$ de la récompense change dans une direction (donnée par une règle de  mise à jour $u$) qui va chercher à réduire cette distance. La justification est que deux politiques d'attributs moyens semblables auront des fonctions de valeur semblables pour toute récompense, y compris donc la récompense inconnue optimisée par l'expert (\autoref{thm:closemuclosev}). Le seul bémol de cette logique est que la réciproque peut être fausse, deux attributs extrêmement différents peuvent mener à la même valeur. Plusieurs autres approches sont basées sur le même schéma itératif. Une revue en est proposée par \citet{neu2009training}.

\begin{algorithm}
\Entree{
Une démonstration de la politique experte $\expertpolicy$\;
Un critère d'arrêt $\epsilon$\;}
\Donnees{
Une notion de distance $\mudistance$ permettant d'évaluer la similarité de deux attributs moyens\;
Une règle de mise à jour $\muupdate$ permettant de réduire la distance\;
Une méthode d'évaluation de l'attribut moyen d'une politique arbitraire\;
Une méthode de résolution d'un \gls{mdp}\;
}
\Sortie{Un vecteur de paramètres $\paramirl$ pour la récompense $\rewardfunc_{\paramirl}$}
Calculer l'attribut moyen de l'expert $\expertmu$ à partir de la démonstration\;
Initialiser $\theta$ arbitrairement\;
\Tq{$\mudistance(\expertmu,\hat{\mu^\pi}) > \epsilon$}{
$\policy \leftarrow $ Une politique (quasi) optimale pour la récompense $\appr{\rewardfunc}_{\paramirl}$\nllabel{alg:l:mdpsolver}\;
$\hat{\mu^\pi} \leftarrow$ Une approximation de l'attribut moyen de la politique $\policy$\nllabel{alg:l:musolver}\;
$\theta \leftarrow \muupdate(\theta,\expertmu,\hat{\mu^\pi})$\;
}
\Retour{$\theta$}
\caption{Structure commune à la plupart des algorithmes d'\glstext{irl}}
\label{alg:repetee}
\end{algorithm}

\gls{pirl} est intrinsèquement un algorithme d'imitation plus que d'\gls{irl}, en ce sens que la sortie est une politique stochastique (plus précisément un mélange de politiques déterministes) et non une fonction de récompense. Il est cependant facile de le ``convertir'' en extrayant la meilleure fonction de récompense, celle liée à la plus petite distance entre attribut moyen de la politique optimale et attribut moyen de l'expert.

   La technique proposée dans \citep{neu2007apprenticeship} (\gls{pm}) repose sur une réduction des désaccords entre la politique de l'expert et la politique à l'itération courante, via une recherche dans l'espace des paramètres $\paramirl$ de la récompense. Le lien unissant récompense et politique, qui est lié à la résolution d'un \gls{mdp}, est ici caractérisé par des outils d'analyse fonctionnelle afin d'extraire une solution analytique du problème qui soit également calculable. Cette contribution présente une plus grand robustesse aux changements d'échelle des attributs. On a vu en \autoref{hier-recompense} que l'ensemble des politiques optimales est stable par dilatation de la récompense, et donc également par dilatation des attributs. Cette plus grande robustesse est donc signe d'une formulation plus saine de l'\gls{irl}. Le bruit dans les attributs est également mieux géré.

   Basé sur la théorie des jeux dans un contexte où l'agent définit une politique qui doit être meilleure que celle de l'expert pour une récompense choisie par l'adversaire, l'algorithme \gls{mwal} de \citet{syed2008game} tombe sur un os\footnote{Os à \gls{mwal}.
 }. L'inclusion de savoir \emph{a priori} quant à la contribution positive ou néfaste de certains attributs à la réalisation de la tâche donne un algorithme d'\gls{irl} dont le résultat peut s'avérer meilleur que l'expert ou, encore plus paradoxal, qui peut fonctionner même en l'absence d'expert (sic). En contrepartie de la nécessité de disposer de ces connaissances \emph{a priori}, il est plus rapide à l'exécution que \gls{pirl} et capable de gérer la non optimalité de l'expert. Il exige malheureusement de connaître la dynamique du \gls{mdp}. La définition de l'\gls{irl} comme un problème distinct de l'imitation y gagne en clarté car cette contribution mentionne explicitement les problèmes liés aux raisonnements sur des politiques mixées et non des récompenses, sans pour autant les résoudre. Les auteurs rendent aussi explicites les difficultés soulevées en \autoref{hier-problemdef} quant à la définition des espaces d'état et d'action, notamment le respect de la propriété de Markov.

   Ce travail est étendu par \citet{syed2008apprenticeship} qui, en formulant la résolution du MDP comme un programme linéaire, assainissent la formulation de \gls{pirl} et \gls{mwal} en leur permettant de raisonner sur une politique stationnaire et sans mélange. En formulant également l'\gls{irl} comme un programme linéaire, les auteurs proposent \gls{lpal}, qui retourne un politique (non une récompense). La formulation de ce programme linéaire met en jeu des grandeurs qui correspondent à l'attribut moyen de politiques optimales pour des récompenses arbitraires, la structure itérative est donc toujours présente.

   L'approche de classification à marge structurée décrite en \autoref{hier-attributs} et utilisée en \citep{ratliff2007imitation} pour l'imitation supervisée apparaît également dans \citep{ratliff2006maximum}. Plutôt que d'associer une action à un état comme dans l'approche supervisée, l'algorithme \gls{mmp} proposé associe une politique optimale à un \gls{mdp}. Bien que les auteurs s'en défendent, cette formulation est compatible avec la structure familière des autres algorithmes d'\gls{irl} de cette période (\autoref{alg:repetee}), comme le démontrent \citet{neu2009training}. La philosophie à la base de l'approche reste cependant différente, et bien que les applications envisagées sont variées et non triviales (voir \citep{ratliff2009learning}, où elles sont toutes réunies et la méthode de boosting proposée en \citep{ratliff2007boosting} expliquée plus en profondeur), d'autres problèmes apparaissent comme la nécessité de résoudre de multiples \gls{mdp} de manière tractable, et de formuler le problème du contrôle non pas comme un \gls{mdp}, mais comme de multiples \gls{mdp} ``compatibles'' entre eux\footnote{La résolution se fait avec par exemple $A^*$, il faut que les actions des différents \gls{mdp} aient le même type d'effet sur les états.
 }.

   La formulation probabiliste de \gls{maxent} de \citet{ziebart2008maximum} est intéressante car elle formule un critère (l'entropie) pour choisir entre deux politiques qui jusqu'à présent étaient équivalentes (même valeur ou même attribut moyen). Structurellement, cependant, les différences restent minces avec toujours présentes la résolution répétée d'un MDP et l'estimation de l'attribut moyen de politiques arbitraires.

   Curieusement semblables à une approche non officiellement \gls{irl} et plus ancienne, \citep{chajewska2001learning}, les travaux de  \citet{ramachandran2007bayesian} posent l'\gls{irl} comme un problème d'inférence bayésienne où l'on cherche la récompense la plus vraisemblable au vu de la démonstration de l'expert. Il en résulte une formulation plutôt saine de l'\gls{irl} permettant par exemple d'incorporer de la connaissance \emph{a priori} sans que cela soit pour autant nécessaire, ou de qualifier le niveau de confiance dans la démonstration de l'expert, qui peut alors être sous-optimal. Des arguments similaires à ceux développés pour \gls{maxent} permettent de lever les ambiguïtés induites par le fait que l'\gls{irl} est un problème mal posé. Ces travaux sont étendus par \citet{lopes2009active} qui en extrayant une information d'incertitude proposent un schéma d'échantillonnage interactif selon la même logique, et avec les mêmes avantages que ceux développés par \citeauthor{chernova2007confidence} (voir la \autoref{hier-nonari}) pour l'approche supervisée (voir \autoref{hier-myopie}). Bien que cette approche diffère des autres dans l'exposition du raisonnement, elle reste très similaire dans la mise en œuvre, puisque'il faut toujours calculer des politiques optimales et obtenir des échantillons de cette politique (pour calculer des \emph{a posteriori} bayésiens et non plus des attributs moyens).

   Signalons finalement les travaux de \citet{jin2010gaussian}, qui par l'usage des \gls{gp} plutôt que d'un schéma linéaire pour l'approximation de la récompense, ainsi que par l'usage d'algorithmes d'\gls{rl} mettant les \gls{gp} en jeu \citep{rasmussen2004gaussian,deisenroth2009gaussian} permet de ne plus avoir à concevoir des attributs pour appliquer le schéma itératif classique de l'\gls{irl}, se basant sur les échantillons comme points supports.

   La plupart de ces approches sont résumées dans \citep{neu2009training}. Ces différentes contributions ont le mérite d'observer le problème sous plusieurs angles, de se placer aux limites du problème (expert non optimal, attributs bruités, etc.) et d'aborder des questions fondamentales (notion de distance entre politiques ou récompenses). Le manque d'harmonisation du domaine de l'\gls{irl} (encore jeune) se fait sentir. Chacun redéfinit le problème à sa manière. Toutes ces approches sont malgré cela structurellement très similaires  : schéma itératif avec résolution répétée de MDP et approximation de l'attribut moyen de la politique courante puis comparaison à celui de l'expert. Les approches les plus tardives font apparaître les difficultés soulevées par la recherche d'un politique mixée, d'une politique au lieu d'une récompense, de l'absence d'un critère commun, de la résolution répétée du MDP et de l'approximation de l'attribut moyen. Si de bonnes solutions aux deux premiers problèmes sont proposées, les autres soucis ne trouvent en revanche pas encore de réponse.

   Outre les applications variées développées au fur et à mesure par \citet{ratliff2009learning}, il est intéressant de constater l'efficacité des algorithmes de cette veine que \citet{abbeel2010autonomous} ont démontrée par le contrôle d'un hélicoptère radiocommandé en vol acrobatique. L'\gls{irl} ne joue cependant qu'un petit rôle dans cette expérience, où la contribution principale est l'apprentissage du modèle dynamique de l'hélicoptère. Ce modèle est ensuite utilisé pour la résolution du problème de l'\gls{rl}, et donc également de manière répétée lors de la résolution de l'\gls{irl}. Développer un simulateur est une tâche lourde.
\section{Méthodes ne nécessitant pas la résolution répétée d'un \glsentrytext{mdp}}
\label{sec-3-4}
\label{hier-nouveauxalgos}

   Les problèmes intrinsèquement liés à la structure itérative des premiers algorithmes d'\gls{irl} étant identifiés, un regain de créativité a permis l'émergence d'approches hétérogènes ces dernières années.

   Un schéma itératif (mais sans lien évident avec le schéma itératif classique de la section précédente) permet à \citet{levine2010feature} d'effectuer de la sélection d'attributs tout en apprenant la récompense, à l'aide d'une méthode semblable dans sa motivation au \gls{boosting} opéré dans \citep{ratliff2007boosting}. La méthode de \citet{levine2010feature} semble cependant plus générique, et l'idée qu'une récompense ``simple'' est préférable y est enracinée. La tractabilité n'est malheureusement pas rendez-vous.

   Présenté comme une approche supervisée de l'imitation, la méthode de \citet{melo2010learning} repose sur une notion particulière de distance dans un \gls{mdp}. Cette distance n'est pas uniquement basée sur la définition de l'espace d'état (comme l'est par exemple la distance euclidienne standard) mais prend en compte la dynamique du \gls{mdp} afin que deux états semblables pour le problème qui nous préoccupe (permettant d'accéder aux mêmes états en effectuant la même action) aient une distance nulle. Le calcul de cette distance cache la résolution du \gls{mdp} par des méthodes de \gls{dp}, on trouve le point fixe d'un opérateur par son application répétée. Bien que la récompense n'apparaisse pas explicitement dans l'exposition de la méthode, le raisonnement qui définit l'échantillonnage interactif et la politique renvoyée par l'algorithme est similaire a ce qui a été proposé par \citet{lopes2009active}.

   Raisonnant selon la même logique bayésienne que \citeauthor{ramachandran2007bayesian}, mais présentant la particularité de ne pas faire appel à une approximation linéaire de la fonction de récompense, le travail de \citet{levine2011nonlinear} utilise des \gls{gp} dans une approche itérative (appelée \gls{gpirl}) qui apprend en même temps la fonction de récompense et le noyau permettant d'en représenter la structure. Elle souffre d'un problème de tractabilité que les auteurs résolvent en partie en proposant une version échantillonnée sur une partie de l'espace d'état, ce qui induit d'autres problèmes, notamment au niveau de la qualité de la généralisation. Cette approche pourrait facilement être augmentée d'un schéma d'échantillonnage actif en raison de la présence de l'information d'incertitude dans les \gls{gp}.

   Les \gls{gp} sont aussi présents chez \citet{qiao2011inverse}, dans une approche également bayésienne mais beaucoup plus analytique. Il ne s'agit pas simplement de calculer un maximum \emph{a posteriori} sur une classe de fonctions de récompenses : le modèle graphique bayésien est sémantiquement plus riche (et donc malheureusement plus lourd), chaque choix de l'expert rajoute une relation dans un modèle bayésien sur les $Q$ valeurs. Cette approche semble ne pas souffrir des problèmes de généralisation qui pénalisent \gls{gpirl}, mais paraît encore moins tractable et est réservée aux \gls{mdp} discrets.

   L'algorithme proposé par \citet{dvijotham2010inverse} remplace la résolution du problème de l'\gls{rl} par l'inversion d'un \gls{mdp} soluble linéairement. Les résultats obtenus sont aussi bon ou meilleurs que ceux obtenus avec les algorithmes itératifs classiques, et le sont plus rapidement. L'astuce réside dans le fait que pour convertir un \gls{mdp} arbitraire en \gls{mdp} soluble linéairement, il faut connaître les probabilités de transition.

   Finalement, \gls{relent} de \citet{boularias2011relative} reste dans la lignée des algorithmes itératifs classiques en ceci que son argumentaire ressemble à celui qui amène \gls{maxent} et qu'il tient compte des progrès fait dans la définition du problème de l'\gls{irl} : il s'agit de trouver une récompense (non une politique), sans accès complet à la dynamique du système (mais en ayant accès à des trajectoires) et en observant la politique de l'expert (sans la connaître totalement). La méthode cependant est plus efficace, l'échantillonnage préférentiel permet à l'algorithme de fonctionner en utilisant uniquement des échantillons de l'expert et des échantillons obtenus grâce à une politique aléatoire. La résolution répétée du problème de l'\gls{rl} n'est plus nécessaire. L'approche échantillonnée est décrite \autoref{alg:relent}. C'est cette approche, la plus efficace de l'état de l'art en terme d'information nécessaire à son fonctionnement, qui nous servira de point de comparaison lorsque nous testerons empiriquement nos approches des chapitres \ref{hier-scirl} et \ref{hier-cascading}.

\begin{algorithm}
\Entree{Une base de données $\sastrace{\expertpolicy}$ organisée en $\munbtraj$ trajectoires de longueurs $\muhorizon_{\spaceindex}, 1\leq m\leq M$\;
Une base de données échantillonnées aléatoirement $\sastrace{\sim}$ organisée en $\munbtraj'$ trajectoires de longueurs $\muhorizon'_{\spaceindex}$\;}
\Sortie{Un vecteur de paramètres $\paramirl$ pour la récompense $\rewardfunc_{\paramirl} = \transpose{\paramirl}\featurestateactionfunc$}
Initialiser $\paramirl$ arbitrairement\;
Calculer 
\begin{equation*}
\bar\expertmu = 
{1\over \munbtraj}\sum_{\spaceindex = 1}^{\munbtraj}\sum\limits_{\timeindex = 0}^{\muhorizon_{\spaceindex}-1} \discount^{\timeindex}\featurestateactionfunc(\state_{\timeindex,m},a_{t,m})\textrm{\;}
\end{equation*}
Calculer
\begin{equation*}
^{\spaceindex}\mu = \sum\limits_{\timeindex = 0}^{\muhorizon'_{\spaceindex}-1} \discount^{\timeindex}\featurestateactionfunc(\state_{\timeindex,m},a_{t,m})
\end{equation*}
pour chaque $\spaceindex \in \llbracket 1;\munbtraj'\rrbracket$ \;
Procéder à la descente de sous-gradient dont le sous-gradient est :
\begin{equation*}
\subgrad_{\paramirl} = \bar\expertmu - \frac{\sum\limits_{\spaceindex=1}^{\munbtraj'}\exp(\transpose{\paramirl}(^{\spaceindex}\mu))(^{\spaceindex}\mu)}{\sum\limits_{\spaceindex=1}^{\munbtraj'}\exp(\transpose{\paramirl}(^{\spaceindex}\mu))} - E
\end{equation*}
avec $E$ le vecteur de terme général :
\begin{equation*}
\component{E}{\dimindex} = \sgn(\component{\paramirl}{\dimindex})\epsilon
\end{equation*} 
où $\epsilon$ est un méta-paramètre à régler\;
\Retour{$\paramirl$}
\caption{\glsentrytext{relent} en version échantillonnée}
\label{alg:relent}
\end{algorithm}

   Ces nouvelles approches abandonnent la structure itérative classique, évitant ainsi les problèmes qu'elle implique, à savoir rendre l'algorithme lent et gourmand en données, et d'un point de vue théorique faire apparaître des termes d'erreur mal maîtrisés.

\chapter{Calcul de l'attribut moyen : \glsentrytext{lstdmu}}
\label{sec-4}
\label{hier-lstdmu}

  De par la sémantique dont il est porteur, illustrée Figure 2.2,
et de par le rôle qu'il tient dans les algorithmes itératifs énumérés en \autoref{hier-repetee}, l'attribut moyen est une notion centrale en \gls{irl}.

Nous cherchons à résoudre le problème de l'\gls{irl} dans un cadre où la collecte de données est difficile. Cela signifie qu'un simulateur n'est pas disponible. Or un simulateur est nécessaire à l'application de la méthode de Monte-Carlo (\autoref{alg:mcmu}) que les méthodes actuelles utilisent. Il nous faut donc trouver une nouvelle méthode d'approximation de l'attribut moyen, c'est la première (chronologiquement) contribution de cette thèse, parue dans \citep{klein2011batch,klein2011batch1,klein2011batch2}. Cette méthode est exposée (ainsi que deux alternatives) \autoref{hier-lstdmuprincipe}. La pertinence de cette méthode dans le cadre ouvert par les approches récentes de l'\gls{irl} est finalement étudiée \autoref{hier-lstdmuavantages}, qui présage des défis relevés par les contributions des chapitres \ref{hier-scirl} et \ref{hier-cascading}.
\section{Principe}
\label{sec-4-1}
\label{hier-lstdmuprincipe}

   Les approches discutées en \autoref{hier-repetee} font toutes appel, du fait de leur structure commune (\autoref{alg:repetee}), à une procédure d'estimation de l'attribut moyen de l'expert ainsi que, de manière répétée, à une procédure d'estimation de l'attribut moyen\footnote{Souvent, c'est un terme plus simple à estimer, $\expectationknowing{\mu^\pi(s)}{s\sim\rho_0}$ où $\rho_0$ est par exemple une distribution des états de départ, qu'il est nécessaire d'approximer.
 } d'une politique arbitraire $\policy$ :
\begin{align}
\rlmu{\policy} &\in \twosetsfunctions{\statespace \times \actionspace}{(\reals^{\dimx{\featurestatefunc_{R}}})}\\
\rlmu{\policy}(\state,\action)&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0 = \state,\action_0=\action,\policy}
\end{align}

La méthode préconisée pour le calcul de l'attribut moyen est l'utilisation d'un simulateur pour obtenir les échantillons nécessaires à une estimation de Monte-Carlo comme cela est formalisé dans l'\autoref{alg:mcmu}. Cet estimateur non biaisé possède de surcroît l'avantage d'être simple à mettre en œuvre… pour peu que l'on dispose d'un simulateur.

   L'horizon $\muhorizon$ après lequel on tronque la trajectoire peut être choisi en fonction du facteur d'amortissement $\discount$ : au bout d'un moment le terme $\discount^{\timeindex}$ en facteur de l'attribut devient négligeable, poursuivre le calcul de la somme n'a donc plus d'intérêt\footnote{A noter que la somme $\sum_{t=0}^\infty \gamma^t$ converge vers un réel (en l'occurrence $1\over 1-\gamma$).
 }. Le nombre de trajectoires $\munbtraj$ est le fruit d'un compromis entre variance et temps de calcul.

   Il est à noter que la méthode est sans biais (au fait de tronquer les trajectoires près), pour peu que le simulateur soit fidèle à la réalité. Dans le cas contraire, on obtient un estimateur non biaisé d'une grandeur ne correspondant pas à ce que l'on cherche, c'est le biais de modélisation. Il reste toujours la possibilité d'utiliser des données en provenance d'un système réel, mais cela implique de pouvoir placer le système dans un état arbitraire et de laisser une politique potentiellement très mauvaise aux manettes sans provoquer de casse, le tout pour un coût en temps ou ressources raisonnable.

\begin{algorithm}
\Entree{Le couple $(\state,\action)$ en lequel on souhaite évaluer $\rlmu{\policy}$\;}
\Donnees{La politique $\policy$ et un simulateur du système à contrôler\;}
\Res{\rlmu{\policy}(\state,\action)\;}
$\satrace{\policy} \leftarrow \varnothing$\;
\SetKwFor{nfois}{répéter}{fois}{fin répéter}
\nfois{$\munbtraj$}{
Simuler une trajectoire $D$ de longueur $\muhorizon$, commençant par $\state$ et $\action$, puis suivant $\policy$ par la suite\;
$\satrace{\policy} \leftarrow \satrace{\policy}\cup D$\;
}
\Retour{\begin{equation*}
{1\over \munbtraj}\sum_{\datasetindex = 1}^{\munbtraj}\sum\limits_{\timeindex = 0}^{\muhorizon-1} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex,i},\action_{\timeindex,i})
\end{equation*}}
\caption{Estimation de $\rlmu{\policy}$ par une méthode de Monte-Carlo}
\label{alg:mcmu}
\end{algorithm}

Si l'on compare la définition de l'attribut moyen (\autoref{eq:mudefsum}) que nous venons de rappeler à la définition de la fonction de valeur (\autoref{eq:Vdefsum}), on constate que les deux expressions sont quasiment identiques. La fonction de valeur est définie par l'espérance d'un cumul $\discount$-pondéré de récompenses, tandis que l'attribut moyen est l'espérance d'un cumul $\discount$-pondéré d'attributs. La sémantique des ces deux objets n'est pas la même puisque l'attribut ne porte généralement pas d'information immédiatement exploitable quant à la qualité du contrôle mais, mathématiquement, les différences sont minimes. La seule nuance entre l'attribut moyen et la fonction de valeur est la dimension. La fonction de valeur travaille sur une récompense réelle tandis ce qui est sommé dans l'attribut moyen est une grandeur vectorielle (de dimension $\dimx{\featurestateactionfunc_{\rewardfunc}}$).

Si l'on considère l'attribut moyen composante par composante, alors cette dissimilitude s'estompe. La composante $\dimindex$ de l'attribut moyen est exactement le même objet que la fonction de valeur pour la récompense donnée par la $\dimindex^\textrm{ième}$ composante de l'attribut $\featurestateactionfunc_{\rewardfunc}$ :
\begin{equation}
\component{\rlmu{\policy}}{\dimindex} = \rlvalue{\policy}{\component{\featurestateactionfunc_{\rewardfunc}}{\dimindex}}.
\label{eq:muvaleurvec}
\end{equation}
Sachant cela, il paraît naturel, pour calculer l'attribut moyen d'une politique, d'adapter les techniques existantes permettant de calculer la valeur de cette politique. Nous n'aurons, comme nous allons le voir, même pas à utiliser la décomposition en composante comme nous venons de le faire, le fait que l'expression soit vectorielle plutôt que scalaire ne posant pas de problèmes insurmontables.

Dans le cas idyllique où l'espace d'état est discret de faible cardinal et la matrice des probabilités de transition $\transprobmat{\policy}$ induites par la politique $\policy$ est connue, un algorithme d'évaluation de la politique tiré de la \gls{dp} convient tout à fait à l'estimation de l'attribut moyen, comme l'explicite l'\autoref{alg:dpmu}.

\begin{algorithm}
\Entree{La matrice $\transprobmat{\policy}$ des probabilités de transition $\transprobfunceval{\state}{\policy(\state)}{\state'}$\;
La fonction d'attributs $\featurestateactionfunc_{\rewardfunc}$\;
} 
\Res{La valeur exacte de $\rlmu{\policy}$}
Pour chaque composante $\dimindex$ de $\rlmu{\policy}$ :
\begin{equation*}
\component{\rlmu{\policy}}{\dimindex}\leftarrow (I-\discount \transprobmat{\policy})^{-1}\component{(\featurestateactionfunc_{\rewardfunc})}{\dimindex}
\end{equation*}
\Retour{$\rlmu{\policy}$}
\caption{Estimation de $\rlmu{\policy}$ par une méthode de \glstext{dp}}
\label{alg:dpmu}
\end{algorithm}

Maintenant que le principe d'estimer l'attribut moyen par une méthode de calcul de la fonction de valeur est établi, voyons comment il est possible d'adapter non plus un algorithme de \gls{dp} mais un algorithme d'\gls{rl}. La \gls{dp} suppose connues les probabilités de transition. Cela est en général difficile à réaliser sur les systèmes réels, notamment les systèmes robotiques. Le passage à un algorithme d'\gls{rl} permettrait d'apprendre l'attribut moyen au travers d'interactions avec le système réel, en minimisant le travail d'ingénierie.

Un algorithme d'estimation de la fonction de valeur particulièrement approprié est \gls{lstd} de \citet{bradtke1996linear}. Dans sa variante proposée par \citet{lagoudakis2003least} (voir \autoref{alg:lspi}), \gls{lstd}-$Q$, il permet d'estimer la fonction de qualité d'une politique de manière \gls{batch} et \gls{offpolicy}. Cela signifie que le fonctionnement de l'algorithme est complètement découplé de la commande du système. \gls{lstd}-$Q$ peut en effet fonctionner à partir de données recueillies une fois pour toutes (c'est l'aspect \gls{batch}, ou \emph{par lots}) par une politique qui n'est pas celle que l'on évalue (c'est l'aspect \gls{offpolicy}).

Cet algorithme, comme beaucoup d'autres en \gls{rl} repose sur une approximation linéaire de la fonction de qualité (\autoref{eq:paramrl}). Il va donc nous falloir utiliser un schéma d'approximation linéaire pour estimer l'attribut moyen qui lui même est issu de la définition d'un schéma d'approximation linéaire pour la récompense\footnote{Yo dawg, I heard you like linear approximation schemes\ldots{}
 } (\autoref{eq:mudef}). On suppose ici que la récompense est approximée selon le schéma de l'\autoref{eq:paramirl} :
\begin{equation}
\rewardfunc_{\paramirl}(\state,\action) = \transpose{\paramirl}\featurestateactionfunc_{\rewardfunc}(\state,\action).
\end{equation}
L'attribut moyen est donc l'espérance du cumul $\discount$-pondéré de l'attribut $\phi_R$. Nous allons l'approximer dans un espace d'hypothèse engendré par un autre attribut : $\featurestateactionfunc_{\mu}$ :
\begin{equation}
%\hat{\mu^\pi}_{\paramlstdmu}(\state,\action) = \transpose{\paramlstdmu}\featurestateactionfunc(\state,\action).
\hat{\mu}^\pi_{\zeta}(\state,\action) = \transpose{\paramlstdmu}\featurestateactionfunc_{\mu}(\state,\action).
\label{eq:paramlstdmu}
\end{equation}

Ceci posé, l'adaptation de l'algorithme est relativement simple, le calcul de la matrice de paramètres $\paramlstdmu$ est décrit \autoref{alg:lstdmu}. La principale différence avec \gls{lstd}-$Q$ réside dans la dimension du vecteur de paramètres. Pour \gls{lstd}-$Q$, il s'agit d'un vecteur de dimension $\dimx{\phi_V}$, tandis que pour \gls{lstdmu}, il s'agit d'une matrice de taille $\dimx{\featurestateactionfunc_{\mu}} \times \dimx{\featurestateactionfunc_{\rewardfunc}}$. Pour l'évaluation de $\hat\mu^{\pi}_\zeta$ en un couple $(\state,\action)$, il suffit d'utiliser l'\autoref{eq:paramlstdmu}. L'inversion de matrice au cœur de l'algorithme est la même pour toutes les composantes du vecteur d'attribut moyen, il n'y a donc à la faire qu'une seule fois. La complexité algorithmique de \gls{lstdmu} est donc (asymptotiquement) la même que celle de \gls{lstd}-$Q$.

\begin{algorithm}
\Entree{Un base de données $\sasrtrace{\featurestateactionfunc_{\rewardfunc}}$\;
Une fonction d'attribut $\featurestateactionfunc_{\mu}$\;
La politique $\policy$\;
Un paramètre de régularisation $\reg$\;}
\Res{La matrice $\paramlstdmu$ de paramètres pour l'approximation linéaire de l'attribut moyen\;}
$A\leftarrow 0$\tcc*{Matrice de taille $\dimx{\featurestateactionfunc_{\mu}}\times\dimx{\featurestateactionfunc_{\mu}}$}
$b \leftarrow 0$\tcc*{Matrice de taille $\dimx{\featurestateactionfunc_{\mu}}\times\dimx{\featurestateactionfunc_{\rewardfunc}}$}
\PourCh{$\state,\action,\state',\featurestateactionfunc_{\rewardfunc}(\state,\action)\in\sasrtrace{\featurestateactionfunc_{\rewardfunc}}$}{
$A\leftarrow A+\featurestateactionfunc_{\mu}(\state,\action)\transpose{\left(
\featurestateactionfunc_{\mu}(\state,\action) - \discount \featurestateactionfunc_{\mu}(\state',\policy(\state'))
\right)}$\nllabel{alg:lstdmu:A}\;
$b\leftarrow b + \featurestateactionfunc_{\mu}(\state,\action)\transpose{\featurestateactionfunc_{\rewardfunc}(\state)}$\nllabel{alg:lstdmu:b}\;
}
$\paramlstdmu \leftarrow (A+\reg I)^{-1}b$\;
\Retour{$\paramlstdmu$}
\caption{Estimation de $\rlmu{\policy}$ par \glstext{lstdmu}}
\label{alg:lstdmu}
\end{algorithm}
\section{Erreurs d'approximation}
\label{sec-4-2}
\label{hier-lstdmuerror}

\citet{lazaric2010finiteLSTD} proposent une borne sur la performance de \gls{lstd} lorsque celui-ci est employé pour l'estimation d'un fonction de valeur à l'aide d'une base de $\nbsamples$ échantillons. Avec probabilité $1-\delta$, on a :
\begin{align}
\|\hat{V}^\pi_R - \rlvalue{\policy}{\rewardfunc}\|_N &\leq \epsilon_{\phi_V} + O\left(\sqrt{\frac{ \dimx{\phi_V}\ln(\frac{\dimx{\phi_V}}{\delta})}{\nbsamples}}\right)\\
\textrm{avec :}\|f\|_N &= \frac{1}{\nbsamples}\sum_{\timeindex=1}^{\nbsamples}f(s_{\timeindex})^2
\end{align}
Chaque composante de l'attribut moyen étant exactement une fonction de valeur (\autoref{eq:muvaleurvec}) cette borne s'applique donc directement. La différence entre la valeur recherchée $\rlvalue{\policy}{\component{\featurestateactionfunc_{\rewardfunc}}{\dimindex}}$ et la valeur trouvée par \gls{lstdmu} $(\hat{\mu}^\pi_{\zeta})_j$ est donc bornée par le terme d'erreur relatif au biais inductif $\epsilon_{\phi_\mu}$ qui dépend de la différence entre la projection empirique de $\rlvalue{\policy}{\component{\featurestateactionfunc_{\rewardfunc}}{\dimindex}}$ dans l'espace d'hypothèse induit par $\featurestateactionfunc_\mu$ et $\rlvalue{\policy}{\component{\featurestateactionfunc_{\rewardfunc}}{\dimindex}}$ elle-même. \`A cela s'ajoute un terme dépendant de la taille $\nbsamples$ de la base d'entraînement.
\begin{equation}
\|\rlvalue{\policy}{\component{\featurestateactionfunc_{\rewardfunc}}{\dimindex}} - (\hat{\mu}^\pi_{\zeta})_j\|_N = \|\mu^\pi_j - (\hat{\mu}^\pi_{\zeta})_j\|_N \leq \epsilon_{\phi_\mu}^j + O\left(\sqrt{\frac{ \dimx{\phi_\mu}\ln(\frac{\dimx{\phi_\mu}}{\delta})}{\nbsamples}}\right)
\end{equation}
En appliquant ce résultat à chacune des $\dimx{\phi_R}$ composantes de la fonction vectorielle $\mu^\pi$ avec probabilité $1-\delta'$, nous pourrions borner l'écart maximum avec un probabilité $1-\dimx{\phi_R}\delta'$ (application d'une borne d'union). En posant $\delta' = \frac{\delta}{\dimx{\phi_R}}$, on obtient donc avec probabilité $1-\delta$ : 
\begin{align}
\|\hat{\mu}^\pi_{\zeta} - \rlmu{\policy}\| &\leq \max_{\dimindex \in \llbracket1,\dimx{\phi_R}\rrbracket}\left[\epsilon_{\phi_\mu}^j\right] + O\left(\sqrt{\frac{\dimx{\phi_\mu}\ln(\frac{ \dimx{\phi_\mu}\dimx{\phi_R}}{\delta})}{\nbsamples}}\right),\label{eq:lstdmuborne}\\
\textrm{avec : } \|\mu\| &= \max_{\dimindex \in \llbracket1,\dimx{\phi_R}\rrbracket} \|\mu_j\|_N.
\end{align}

Lors des évaluations empiriques des algorithmes que nous proposons, nous prendrons :
\begin{equation}
\phi_\mu = \phi_R.
\end{equation}
Afin de ne pas fournir d'information \emph{ad-hoc} aux problèmes sur lesquels nous testons nos approches, nous utiliserons les attributs standards définis équations \ref{eq:attributsdiscrets} et \ref{eq:gaussianrbf}.

Pour minimiser les termes d'erreurs $\epsilon_{\phi_\mu}^j$ il est cependant possible de réfléchir à un choix d'attributs plus performants. Il est même possible d'estimer chaque composante de $\rlmu{\policy}$ séparément, et donc de choisir des fonctions d'attribut différentes pour chaque composante. Les autres termes d'erreur diminuent lorsqu'on augmente la taille $\nbsamples$ de la base d'entraînement.

Cette borne portant sur les échantillons disponibles pour calculer l'approximation, elle est particulièrement adaptée à l'usage que nous ferons de \gls{lstdmu} au \autoref{hier-scirl}, lorsqu'il sera utilisé pour fournir à \gls{scirl} une estimation de l'attribut moyen de l'expert sur les données que celui-ci a échantillonné.

Une borne en généralisation a également été fournie par \citeauthor{lazaric2010finiteLSTD}. Elle s'applique lorsque la chaine de Markov issue du \gls{mdp} contraint par la politique que l'on souhaite évaluer admet une distribution stationnaire $\rho$ et que les échantillons sont acquis après que le régime stationnaire ait été atteint.
\begin{equation}
\|\hat{V}^\pi_R - \rlvalue{\policy}{\rewardfunc}\|_\rho \leq \epsilon_{\phi_V}^\rho + O\left(\sqrt{\frac{ \dimx{\phi_V}\ln(\frac{N\dimx{\phi_V}}{\delta})}{\nbsamples}}\right)
\end{equation}
Là encore, il est possible d'étendre cette borne à l'attribut moyen :
\begin{align}
\|\hat{\mu}^\pi_{\zeta} - \rlmu{\policy}\| &\leq \max_{\dimindex \in \llbracket1,\dimx{\phi_R}\rrbracket}\left[\epsilon_{\phi_\mu}^{\rho,j}\right] + O\left(\sqrt{\frac{\dimx{\phi_\mu}\ln(\frac{N\dimx{\phi_\mu}\dimx{\phi_R}}{\delta})}{\nbsamples}}\right),\\
\textrm{avec : } \|\mu\| &= \max_{\dimindex \in \llbracket1,\dimx{\phi_R}\rrbracket} \|\mu_j\|_\rho.
\end{align}
Les différences principales avec la borne de l'\autoref{eq:lstdmuborne} sont l'utilisation de la norme sur la distribution $\rho$ et la présence du terme $\epsilon^\rho_{\phi}$ qui dépend non plus de la projection empirique de l'objet recherché sur l'espace d'hypothèses, mais simplement de sa projection.
\section{Avantages dans le cadre des approches existantes}
\label{sec-4-3}
\label{hier-lstdmuavantages}

   Dans le schéma itératif standard des premiers algorithmes d'\gls{irl} (\autoref{alg:repetee}), les problèmes techniques se cachent dans les étapes des lignes \ref{alg:l:mdpsolver} et \ref{alg:l:musolver}, à savoir la résolution du \gls{mdp} et l'estimation de l'attribut moyen. Au delà de l'aspect computationnel, qui pose de moins en moins de problèmes à mesure que le temps passe et que croît la puissance des machines, le goulot d'étranglement réside dans la collecte des données nécessaires à la résolution de ces deux étapes.

   Les auteurs des approches mentionnées en \autoref{hier-repetee} présupposent la disponibilité d'un simulateur. Dans bien des cas, la conception de ce simulateur représente un défi. Par exemple dans les travaux déjà mentionnés de \citet{abbeel2010autonomous}, l'apprentissage du modèle est si peu évident qu'il est l'objet principal de la contribution.

   Dans les cas où il est impossible ou incommode de construire un tel simulateur, recueillir des données reste cependant souvent faisable. Nous nous plaçons dans le cas où l'expert peut contrôler le système, amassant ainsi une démonstration experte $\sastrace{\expertpolicy}$ tout en pouvant aller explorer d'autres zones de l'espace d'état que celles dans lesquelles il souhaite se trouver, soit que l'on perturbe sa démonstration de manière aléatoire, soit qu'il agisse volontairement de manière sous optimale dans un but d'exploration. Ces données non expertes sont recueillies dans une trace $\sastrace{\sim}$. Il s'agit là d'une structure extrêmement contrainte, mais qui correspond par exemple aux problèmes de locomotion robotique ou d'Interaction Homme-Machine où les modèles sont imparfaits et la collecte de données coûteuse.

   Dans le cadre de l'\gls{rl}, l'approche de \citet{lagoudakis2003least} peut fonctionner sous de telles contraintes. Elle découple, comme nous l'avons dit sections \ref{hier-algosrl} et \ref{hier-lstdmuprincipe}, la collecte de données et la résolution approchée du \gls{mdp}. L'aspect de traitement par lot, couplé à l'aspect \gls{offpolicy}, permet de réutiliser tous les échantillons pour évaluer une politique différente de celle qui était au commandes du système lors de la collecte. L'algorithme \gls{lspi}, qui utilise $\gls{lstd}$-$Q$ dans un schéma d'itération de la politique permet donc de trouver la solution approchée d'un \gls{mdp} à partir d'une trace $\sasrtrace{\rewardfunc}$ \footnote{On rappelle que $D^{\color{blue}R}_{sasr} = \{(s_i,a_i,s'_i,{\color{blue}r_i=R(s_i,a_i)})|\datasetindex\in \llbracket 1;\nbsamples\rrbracket\}$
 }. Comme, pour créer \gls{lstdmu}, nous nous sommes inspirés de \gls{lstd}-$Q$, les mêmes avantages sont transférés à l'estimation de l'attribut moyen. Nous pouvons donc facilement imaginer porter le schéma itératif classique de l'\gls{irl} décrit dans l'\autoref{alg:repetee} à un cadre où la collecte de données est coûteuse. C'est ce que présente l'\autoref{alg:repeteelstd}.

\begin{algorithm}
\Entree{
Une démonstration $\sastrace{\expertpolicy}$ de la politique expert $\expertpolicy$\;
Un critère d'arrêt $\epsilon$\;}
\Donnees{
Une notion de distance $\mudistance$ permettant d'évaluer la similarité de deux attributs moyens\;
Une règle de mise à jour $\muupdate$ permettant de réduire cette distance\;
{\color{blue}Des données non-expertes $\sastrace{\sim}$}\;
}
\Sortie{Un vecteur de paramètres $\paramirl$ pour la récompense $\appr{\rewardfunc}_{\paramirl}$}
Calculer l'attribut moyen de l'expert $\expertmu$ à partir de la démonstration\;
Initialiser $\theta$ arbitrairement\;
\Tq{$\mudistance(\expertmu,\hat{\mu^\pi}) > \epsilon$}{
{\color{blue}
Construire la trace $\sasrtrace{\hat\rewardfunc_{\paramirl}}$ à partir de $\sastrace{\sim}$ et $\hat\rewardfunc_{\paramirl}$\;
Construire la trace $\sasrtrace{\featurestateactionfunc_{\rewardfunc}}$ à partir de $\sastrace{\sim}$ et $\featurestateactionfunc_{\rewardfunc}$\;
$\policy \leftarrow LSPI(\sasrtrace{\hat\rewardfunc_{\paramirl}})$\;
$\hat{\mu^\pi} \leftarrow LSTD\mu(\sasrtrace{\featurestateactionfunc_{\rewardfunc}})$\;}
$\theta \leftarrow \muupdate(\theta,\expertmu,\hat{\mu^\pi})$\;
}
\Retour{$\theta$}

\caption[\glstext{irl} itératif avec méthodes \glstext{lstd}]{Instanciation du schéma itératif de l'\gls{irl} avec des méthodes \gls{lstd} (en bleu les changements par rapport à l'\autoref{alg:repetee})}
\label{alg:repeteelstd}
\end{algorithm}

L'estimation de l'attribut moyen par \gls{lstdmu} est imparfaite, comme nous l'avons étudié \autoref{hier-lstdmuerror}. Une erreur trop grande empêcherait l'algorithme d'\gls{irl} de converger. Dans le schéma de l'\autoref{alg:repeteelstd}, les sources d'erreurs principales sont l'estimation de l'attribut moyen par \gls{lstdmu} et la résolution du \gls{mdp} par \gls{lspi}. Les erreurs introduites par ces deux routines dépendent largement de la qualité et de la quantité des données contenues dans la trace $\sastrace{\sim}$ de données non expertes. Elles dépendent aussi bien sûr du choix des fonctions d'attribut $\featurestateactionfunc_R$ et $\featurestateactionfunc_\mu$.

Intuitivement, on peut argumenter que le schéma d'itération de la politique qu'est \gls{lspi}, qui implique donc l'évaluation répétée de politiques arbitraires, sera plus problématique qu'une simple évaluation de la politique (même sur plusieurs composantes) comme \gls{lstdmu}. Une expérience permet de confirmer empiriquement cette intuition.
\section{Validation empirique}
\label{sec-4-4}

Cette expérience fut pour la première fois présentée dans notre article \citep{klein2011batch}. Le schéma en jeu est celui proposé dans l'\autoref{alg:repeteelstd} avec la notion de distance $\mudistance$ et la règle de mise à jour $\muupdate$ choisis conformément aux travaux de \citet{abbeel2004apprenticeship}.
L'idée est de comparer ce schéma à un autre qui, toutes choses étant égales par ailleurs, fait appel à un oracle pour l'estimation de $\hat{\mu^\pi}$. La comparaison des deux approches va porter sur leur comportement en fonction de la taille de la trace non experte $\sastrace{\sim}$. Lorsque cette trace ne contient pas assez de données, \gls{lspi} ne peut résoudre le \gls{mdp} et \gls{lstdmu} ne peut estimer correctement l'attribut moyen. Au fur et à mesure que la taille de $\sastrace{\sim}$ augmente, \gls{lspi} et \gls{lstdmu} seront de mieux en mieux à même de remplir leur mission. Si le besoin en données est le même pour \gls{lstdmu} et \gls{lspi}, la version de l'algorithme d'\gls{irl} qui profite d'un oracle pour l'attribut moyen devrait voir ses performances augmenter plus vite en fonction de la taille de $\sastrace{\sim}$ que la version utilisant \gls{lstdmu}, celui-ci étant censé introduire des erreurs significatives. Si, à l'inverse, les performances des deux versions s'avèrent similaires, cela signifie que les besoins en données de \gls{lspi} sont supérieurs à ceux de \gls{lstdmu}. Quand \gls{lspi} parvient à résoudre le \gls{mdp}, \gls{lstdmu} a déjà suffisamment d'information à sa disposition pour fournir une estimation $\hat\mu^\pi_{\paramlstdmu}$ de l'attribut moyen aussi bonne (pour l'usage que l'on veut en faire) que celle de l'oracle.

Le problème jouet (le pendule inversé, \autoref{hier-pendule}) sur lequel ces deux approches sont testées permettant de définir un simulateur, l'oracle est un estimateur de Monte-Carlo tel que décrit par l'\autoref{alg:mcmu} dont les paramètres $\muhorizon$ et $\munbtraj$ sont réglés de manière à obtenir une variance négligeable.

L'expert est la politique apprise par \gls{lspi} lorsque celui-ci est entraîné sur une récompense nulle partout, mais négative lorsque le pendule tombe. La paramétrisation $\featurestateactionfunc_Q$ de \gls{lspi} est celle suggérée par \citet{lagoudakis2003least} \footnote{Le pendule inversé est en effet l'un des bancs de test dans la contribution qui introduit \gls{lspi}.
 }, à savoir un réseau de $3\times 3$ gaussiennes. Afin de minimiser le travail d'ingénierie et de ne pas introduire de savoir \emph{a priori} en utilisant des attributs porteurs de sens, on utilise les mêmes attributs pour la paramétrisation de la récompense et de l'attribut moyen :
\begin{equation}
\featurestateactionfunc_Q = \featurestateactionfunc_R = \featurestateactionfunc_\mu.
\end{equation}

La trace non experte $\sastrace{\sim}$ provient d'une politique aléatoire mise au contrôle du pendule alors qu'il se trouve dans une position proche de l'équilibre, jusqu'à ce qu'il tombe. On commence alors un nouvel épisode, jusqu'à ce que la quantité d'échantillons voulue soit atteinte. L'expert fournit $100$ échantillons (soit $10$ secondes de démonstration) dans la base $\sastrace{\expertpolicy}$. Cela est suffisant pour obtenir une bonne approximation de $\expertmu$ par Monte-Carlo. L'expérience est répétée $100$ fois afin d'obtenir des informations de variance.

 Le résultat de l'expérience est donné Figure 4.1
. On constate que la taille de $\sastrace{\sim}$ \footnote{La bonne répartition des données dans l'espace d'état-action est également importante. Ici, grâce à l'usage d'une politique aléatoire, $\sastrace{\sim}$ est bien représentative de la dynamique.
 }est bien le facteur déterminant de la qualité de l'imitation. On constate surtout que les deux approches manifestent le même comportement. Cela pourrait signifier que quelle qu'ait pu être l'erreur introduite par l'utilisation de \gls{lstdmu}, elle est gommée par les besoins de \gls{lspi} en données. Quand ce dernier dispose de suffisamment de données pour résoudre le \gls{mdp} suffisamment bien pour que l'algorithme d'\gls{irl} converge vers une bonne solution, \gls{lstdmu} possède assez d'information pour fournir une estimation de l'attribut moyen qui, pour l'usage qui en est fait, est aussi bonne que celle de l'oracle.

\begin{figure}
\centering
\subfigure[Oracle : une seule expérience.]{
\centering
\includegraphics[width=.45\linewidth]{Figures/LSTDMU_a.pdf}
}
\subfigure[\glstext{lstdmu} : une seule expérience.]{
\centering
\includegraphics[width=.45\linewidth]{Figures/LSTDMU_b.pdf}
}\\
\subfigure[Oracle : moyenne et écart-type sur 100 expériences.]{
\centering
\includegraphics[width=.45\linewidth]{Figures/LSTDMU_c.pdf}
}
\subfigure[\glstext{lstdmu} : moyenne et écart-type sur 100 expériences.]{
\centering
\includegraphics[width=.45\linewidth]{Figures/LSTDMU_d.pdf}
}
\caption[\glstext{irl} itératif et \glstext{lstdmu} sur le pendule]{Comparaison de deux approches itératives de l'\gls{irl} sur le problème du pendule inversé. L'ordonnée est le nombre de pas de temps durant lesquels la politique évaluée est en mesure de maintenir le pendule en équilibre, avec un maximum à $3000$. L'une est décrite \autoref{alg:repeteelstd} (LSTD sur les courbes), l'autre est identique sauf en ce qui concerne l'estimation de l'attribut moyen, qui est faite par un oracle (MC sur les courbes). On constate que les deux approches ont le même comportement. Dès lors que la quantité de données obtenues aléatoirement est suffisante pour que \gls{lspi} résolve le \gls{mdp} correctement, l'algorithme d'\gls{irl} converge vers une bonne solution. Cela démontre empiriquement que les erreurs introduites par la résolution répétée du \gls{mdp} sont plus importantes que celles liées à l'estimation de l'attribut moyen par notre algorithme \gls{lstdmu}.}
\label{fig:lstdmuexp}
\end{figure}

Le goulot d'étranglement des approches itératives est donc bien la résolution répétée du \gls{mdp}, il importe de s'affranchir de cette contrainte si l'on souhaite pouvoir utiliser l'\gls{irl} sur des problèmes complexes sans avoir à déployer des trésors d'ingénierie.

Un algorithme de ce type est l'approche de \citet{boularias2011relative}, \gls{relent}, qui peut fonctionner dans le même cadre contraint que celui que nous venons de décrire, à savoir dans le cadre où les seules données dont on dispose sont une trace de l'expert $\sastrace{\expertpolicy}$ et une trace non experte $\sastrace{\sim}$. Il serait cependant souhaitable d'aller encore plus loin et de pouvoir résoudre le problème de l'\gls{irl} armé uniquement de données expertes. Ce sont en effet les données les plus naturelles à rassembler.
\section{Conclusion}
\label{sec-4-5}

L'algorithme \gls{lstdmu} que nous proposons nécessite pour fonctionner moins de données que l'algorithme d'\gls{rl} \gls{lspi}. Les garanties théoriques applicables à \gls{lstd} se transfèrent à \gls{lstdmu}. De manière générale, on peut imaginer adapter d'autres résultats de \gls{lstd} à \gls{lstdmu}, comme par exemple l'utilisation de traces d'éligibilité \citep{geist2012recursive}, l'utilisation de projections aléatoires \citep{ghavamzadeh2010lstd} ou l'introduction de méthodes parcimonieuses \citep{kolter2009regularization,geist2012dantzig}, entre autres.
\chapter{Classification structurée pour l'apprentissage par renforcement inverse}
\label{sec-5}
\label{hier-scirl}

Comme nous venons de le voir sur l'exemple précédent, c'est bien la résolution répétée du \gls{mdp} qui rend les algorithmes d'\gls{irl} itératifs (\autoref{alg:repetee}) si gourmands en information. Il faut donc trouver un moyen de résoudre le problème de l'\gls{irl} sans avoir à résoudre de \gls{mdp}. Parmi les algorithmes les plus récents (\autoref{hier-nouveauxalgos}), la plupart ont besoin de données qui permettraient de résoudre le \gls{mdp}, et ne présentent donc pas, du point de vue de la facilité d'utilisation dans un cadre où la collecte de données est difficile, un avantage décisif sur les approches itératives.

Un de ces algorithmes, cependant, sort du lot puisqu'il ne nécessite pas la résolution du \gls{mdp} et peut fonctionner à l'aide de simplement deux bases d'entraînement $\sastrace{\expertpolicy}$ et $\sastrace{\sim}$ comme notre \autoref{alg:repeteelstd}. Il s'agit de \gls{relent}. Comme il ne nécessite ni de connaître les probabilités de transition, ni de résoudre un \gls{mdp} de manière répétée, il sera celui auquel nous comparerons les approches définies dans ce chapitre et le suivant, approches qui constituent la contribution principale de cette thèse et qui sont (à notre connaissance) les seules capables de résoudre le problème de l'\gls{irl} uniquement grâce à des données expertes \footnote{Nonobstant l'utilisation d'heuristiques.
 }.
\section{Liens entre classification et \glsentrytext{irl}}
\label{sec-5-1}
\label{hier-classifirl}

Les approches supervisées d'imitation (\autoref{hier-myopie}) apprennent directement la politique de l'expert et non, comme les approches d'\gls{irl}, une description succincte de la tâche sous la forme d'une fonction de récompense. L'imitation supervisée peut être vue comme moins problématique que l'\gls{irl}, disons certainement plus directe, puisque ce que ces méthodes cherchent à apprendre (la politique de l'expert) est de même nature que ce qui est présent dans les données. A l'inverse, pour les approches d'\gls{irl} il s'agit de comprendre la motivation à long terme de l'expert, et ce à partir d'exemples de décisions à portée immédiate. Il faut résoudre un problème de contexte temporel entre ce qui est démontré et ce que l'on cherche.

Les approches supervisées n'ont pas besoin pour fonctionner de données non expertes ou des probabilités de transition. Elles peuvent être appliquées lorsqu'on dispose uniquement de données expertes. Il est plus dur de fonctionner avec si peu de données lorsque l'on résout le problème de l'\gls{irl}, puisque l'objet recherché est plus complexe à trouver. Pourtant il existe, entre les outils développés dans le cadre de l'\gls{rl} (\autoref{hier-cadre}) et certaines approches de classification, des similitudes qui vont nous permettre de faire un parallèle entre les deux philosophies.

Dans le cas de la classification à fonction de score (\autoref{eq:classifscore}), le classifieur apprend, à partir des données expertes, une fonction $\classifscorefunc$ qui permet de définir simplement la politique de classification $\classifpolicy$ par une optimisation de ce score sur les actions (\autoref{eq:classifscore}) :
\begin{equation}
\forall \state, \classifpolicy(\state) \in \arg\max_{\action \in \actionspace} \classifscorefunc(\state,\action).
\end{equation}
Or la politique de l'expert, qui par hypothèse est une politique optimale, obéit à une relation du même type vis-à-vis de la fonction de qualité optimale pour la récompense inconnue $\expertreward$ dont l'expert cherche à maximiser le cumul (\autoref{eq:rlargmax}) :
\begin{equation}
\forall \state \in \statespace, \expertpolicy(\state) \in \arg\max_{\action\in \actionspace}\quality{\expertpolicy}{\expertreward}(\state,\action).
\end{equation}
La fonction de score $\classifscorefunc$ apprise par la plupart des approches d'imitation supervisée peut donc être vue comme un objet de même type qu'une fonction de qualité, dont nous avons dit en \autoref{hier-cadre} qu'elle fait le lien entre une optimisation myope, à l'échelle de la décision immédiate, et le critère à optimiser sur le long terme, le cumul des récompenses.

Il est problématique que l'apprentissage supervisé ne tient pas compte de cette structure temporelle dans la résolution du problème de l'imitation. Tel qu'il est posé, le problème de classification suppose que les attributs choisis encodent l'information nécessaire à la prise de décision pour chaque état or, comme tend à le montrer ce lien entre classification et \gls{rl}, l'optimisation doit avoir lieu sur un critère à long terme qui prend la dynamique du système en compte.

Il est possible d'effectuer un travail d'ingénierie sur les attributs (\autoref{hier-attributs}) qui permette aux méthodes supervisées de bien généraliser, mais nous souhaitons ne pas trop avoir à faire appel à l'expertise humaine pour bâtir des systèmes intelligents. Il faut trouver un moyen d'informer les méthodes supervisées de la structure temporelle du problème, afin d'éviter le décalage actuel entre un problème d'optimisation mettant en jeu une fonction de score qui remplit le rôle d'une fonction de qualité, et qui pourtant provient de l'exploitation directe de données concernant les décisions immédiates de l'expert.

L'algorithme \gls{cascading} du chapitre suivant et \gls{scirl}, que nous présentons ici, font précisément cela. Nous exposons la logique derrière \gls{scirl} dans la \autoref{hier-scirldesc}, puis en présentons une justification théorique (\autoref{hier-scirltheo}) et expérimentale (\autoref{hier-scirlexp}).
\section{Description}
\label{sec-5-2}
\label{hier-scirldesc}

Nous décrivons les principe général de \gls{scirl} \autoref{hier-scirlprinciple}, puis présentons une instanciation particulière\footnote{Qui grâce à une heuristique peut fonctionner en utilisant uniquement des données expertes.
 } (\autoref{hier-scirlinstan}) qui sera celle utilisée pour les expériences (\autoref{hier-scirlexp}).
\subsection{Principe général}
\label{sec-5-2-1}
\label{hier-scirlprinciple}

   L'un des résultats principaux de la \autoref{hier-cadre} est la notion de politique gloutonne, qui en optimisant à l'échelle de la décision immédiate la fonction de qualité parvient à optimiser le critère long terme du cumul des récompenses. Les approches de classification à fonction de score souffrent de ce que la fonction $\classifscorefunc$ ne présente pas la même structure temporelle qu'une fonction de qualité bien qu'elle remplit un rôle similaire, comme nous venons de le découvrir.

   Afin de permettre aux techniques de classification d'exploiter cette structure temporelle qui leur manque, nous allons étudier comment introduire l'attribut moyen dans la fonction de score. Nous avons vu \autoref{eq:muvaleurvec} que l'attribut moyen est la généralisation vectorielle de la fonction de valeur, qui porte cette structure temporelle indispensable à la réconciliation de l'optimisation locale et long terme. De plus, de par sa définition, il est naturellement présent dans la fonction de qualité $\quality{\expertpolicy}{\rewardfunc_{\paramirl}}$ pour la récompense paramétrée linéairement $\rewardfunc_{\paramirl}$ :
\begin{align}
\quality{\expertpolicy}{\rewardfunc_{\paramirl}}(s,a)&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\rewardfunc_{\paramirl}(\state_{\timeindex},\action_{\timeindex})}{\state_0 = \state,a_0=a,\policy}\\
&=\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\transpose{\paramirl}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0 = \state,\action_0=\action,\policy}\\
&=\transpose{\paramirl}\expectationknowing{\sum\limits_{\timeindex = 0}^{\infty} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex},\action_{\timeindex})}{\state_0 = \state,\action_0=\action,\policy}\\
&=\transpose{\paramirl} \expertmu(\state,\action).
\label{eq:qemuparam}
\end{align}

Sachant tout cela, nous choisissons un classifieur à fonction de score $\classifscorefunc_{\paramirl}$ qui puisse être paramétrée linéairement. Comme nous avons identifié le rôle de la fonction de score et celui de la fonction de qualité optimale, nous utilisons comme attribut pour la paramétrisation linéaire de la fonction de score une approximation $\hat\expertmu$ de l'attribut moyen de l'expert $\expertmu$ afin d'obtenir une expression identique à l'\autoref{eq:qemuparam} :
\begin{equation}
\classifscorefunc_{\paramirl} = \transpose{\paramirl}\hat{\mu}^E.
\end{equation}
L'algorithme \gls{lstdmu} du \autoref{hier-lstdmu} est tout indiqué pour fournir cette estimation.

Ce que le classifieur apprend en réalité quand il apprend la fonction de score est le vecteur de paramètres $\paramirl$, qui est également le vecteur de paramètres de la fonction de récompense (\autoref{eq:paramirl}). Ainsi, en résolvant de manière supervisée le problème de l'imitation, on trouve une récompense (qui est l'objet d'intérêt de l'\gls{irl}), simplement en paramétrant la classification avec un objet porteur de la structure temporelle du \gls{mdp}. Tout cela est résumé \autoref{alg:scirlgenerique}.

\begin{algorithm}
\Entree{
Une base d'entraînement établie par l'expert $\sastrace{\expertpolicy}$\;
\Donnees{
Une approximation $\hat{\mu}^E$ de l'attribut moyen de l'expert\;
Un classifieur à fonction de score paramétrée linéairement\;
}
\Sortie{Un vecteur de paramètres $\paramirl$ pour la récompense $\rewardfunc_{\paramirl} = \transpose{\paramirl}\featurestateactionfunc_{\rewardfunc}$}
Paramétrer la fonction de score $\classifscorefunc_{\paramirl}$ du classifieur comme cela :
\begin{equation}
\classifscorefunc_{\paramirl} = \transpose{\paramirl}\hat{\mu}^E\;
\end{equation}
Entraîner le classifieur sur la base de données $\sastrace{\expertpolicy}$\;
}
\Retour{$\paramirl$}
\caption{Principe générique de \glstext{scirl}}
\label{alg:scirlgenerique}
\end{algorithm}
\subsection{Instanciation avec heuristique}
\label{sec-5-2-2}
\label{hier-scirlinstan}

Si la majorité des approches de classification peuvent être vue comme utilisant une fonction de score, celle-ci n'est pas forcément paramétrée linéairement d'une manière qui nous permette d'appliquer \gls{scirl}. Fort heureusement, nous avons étudié un classifieur se prêtant bien à cet exercice en \autoref{hier-attributs}. L'apprentissage se réduit à une descente de sous-gradient :
\begin{align}
\subgrad\structuredcost(\paramirl) &= \sum_{\datasetindex=1}^{N}\left(\hat{\mu}^E(\state_{\datasetindex},\best{\action}_{\datasetindex}) - \hat{\mu}^E(\state_{\datasetindex},\action_{\datasetindex})\right) + \lambda \theta,\\
\textrm{avec : }
\best{\action}_{\datasetindex} &=  \arg\max_{\action \in \actionspace} \transpose{\paramirl}\hat{\mu}^E(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action),
\end{align}
et $\margin$ telle que définie \autoref{eq:margin}.

Cette méthode de classification est aussi celle employée par \citet{ratliff2006maximum} dans leur algorithme \gls{mmp}. Une différence importante entre \gls{scirl} et \gls{mmp} est que ce dernier prend la structure temporelle du problème en compte en associant des politiques à des \gls{mdp}. Cela fonctionne mais implique de devoir résoudre ces \gls{mdp} et fait donc de \gls{mmp} une approche itérative conforme au schéma de l'\autoref{alg:repetee}. Ici ce sont les actions sont associées aux états, comme dans les approches supervisées (\autoref{hier-nonari}). La structure temporelle est introduite par l'attribut moyen.

Le calcul du sous-gradient $\subgrad\structuredcost(\paramirl)$ implique de pouvoir évaluer l'attribut moyen de l'expert $\expertmu$ en tout état présent dans la base d'entraînement $\sastrace{\expertpolicy}$, pour toute action.

Si les données de l'expert sont organisées en trajectoires, sur une trajectoire de longueur $L$, il est possible d'estimer $\expertmu$ sur tous les couples $(\state_i,\action_i)$ de la trajectoire grâce à une simple estimation de Monte-Carlo :
\begin{equation}
\hat{\mu}^E(\state_{\datasetindex},\action_{\datasetindex}) = \sum\limits_{\timeindex = 0}^{\muhorizon-1-\datasetindex} \discount^{\timeindex}\featurestateactionfunc_{\rewardfunc}(\state_{\timeindex+\datasetindex},\action_{\timeindex+\datasetindex}).
\label{eq:montecarlodupauvre}
\end{equation}
Il faut noter que la qualité de l'estimation risque de diminuer à mesure que l'on s'approche de la fin de la trajectoire, puisque le nombre d'échantillons participant au calcul diminue. Cette méthode fonctionne bien sur les problèmes quasi-déterministes mais risque de poser problème sur les dynamiques plus stochastiques (la variance sera grande car une seule trajectoire est utilisée). Utiliser \gls{lstdmu} est la solution la plus souple. Si seules les données de l'expert sont disponibles, une version \gls{onpolicy} de l'algorithme permettra d'obtenir une estimation $\hat{\mu}^E$ sur les couples $(s_i,a_i)$ présents dans la base d'entraînement. Si des données non expertes sont également disponibles, alors \gls{lstdmu} dans sa version \gls{offpolicy} saura les exploiter (si l'on peut interroger l'expert sur un état arbitraire).

Le calcul de $\best{\action_{\datasetindex}}$ dans le sous-gradient implique, pour un état de la base d'entraînement, de connaître l'attribut moyen de l'expert pour toute action, et non juste celle choisie par l'expert. En l'absence de donnée, il est impossible de pallier ce problème d'une manière qui soit formellement juste. Nous pouvons en revanche employer une heuristique dont l'étude expérimentale (\autoref{hier-scirlexp}) montrera la validité.

Pour introduire intuitivement l'heuristique, imaginons dans une voiture à double commande telles celles employées par les auto-écoles un expert au volant, la tâche étant de conduire. Imaginons maintenant qu'un plaisantin dans le siège passager applique un instant une force sur le volant, ou appuie brièvement et légèrement sur une des pédales. On voit qu'à l'échelle de la tâche cette perturbation est minime et le conducteur expert a tôt fait de poursuivre tel qu'il l'aurait fait si la perturbation n'avait pas eu lieu. De la même manière, on suppose qu'une perturbation causée par une action parasite lorsqu'on doit évaluer $\expertmu(\state,\action\neq\expertpolicy(\state))$ est immédiatement absorbée par une action correctrice de l'expert, et donc que la seule conséquence de la perturbation est de faire perdre un pas de temps à l'expert. Ce qui se traduit mathématiquement par :
\begin{equation}
\hat{\mu}^E(\state_i,\action \neq \action) = \discount \hat{\mu}^E(\state_i,a_i)
\label{eq:heuristiquescirl}
\end{equation}

Armés de notre méthode de classification à fonction de score linéairement paramétrée ainsi que de notre heuristique pour le calcul de l'attribut moyen de l'expert, il est maintenant possible d'instancier \gls{scirl} afin qu'il apprenne une récompense uniquement nourri de données expertes (\autoref{alg:scirlonlyexpert}).

\begin{algorithm}
\Entree{
Une base d'entraînement établie par l'expert $\sastrace{\expertpolicy}$\;}
\Donnees{
Une approximation $\hat{\mu}^E$ de l'attribut moyen sur tous les états présents dans la base $\sastrace{\expertpolicy}$\;
L'extension de cette approximation à toutes les actions grâce à l'heuristique de l'\autoref{eq:heuristiquescirl}\;
}
\Sortie{Un vecteur de paramètres $\paramirl$ pour la récompense $\rewardfunc_{\paramirl} = \transpose{\paramirl}\featurestateactionfunc_{\rewardfunc}$}
Initialiser $\paramirl$ arbitrairement\;
Procéder à la descente de sous-gradient dont le sous-gradient est :
\begin{equation*}
\subgrad\structuredcost(\paramirl) = \sum_{\datasetindex=1}^{N}\hat{\mu}^E\left(\state_{\datasetindex},\arg\max_{\action\in\actionspace}\transpose{\paramirl}\hat{\mu}^E(\state_{\datasetindex},\action) + \margin(\state_{\datasetindex},\action)\right) - \hat{\mu}^E(\state_{\datasetindex},\action_{\datasetindex}) + \lambda \theta\;
\end{equation*}
\Retour{$\paramirl$}
\caption{\glstext{scirl} uniquement sur des données expertes}
\label{alg:scirlonlyexpert}
\end{algorithm}
\section{Validation théorique}
\label{sec-5-3}
\label{hier-scirltheo}

Maintenant que le principe de \gls{scirl} ainsi qu'une instanciation particulière utilisant l'\autoref{alg:taskar} et \gls{lstdmu} sont en place, Nous fournissons des garanties théoriques sur les performances d'imitation de \gls{scirl} \footnote{Chronologiquement postérieures à celles concernant l'algorithme \gls{cascading}, qui sont dues au travail d'un autre doctorant de l'équipe MaLIS, Bilal Piot.
 }.

   Lorsque nous avons posé le problème de l'\gls{irl}, nous avons discuté des critères permettant d'évaluer la qualité de la solution. Dans l'analyse théorique de \gls{scirl}, c'est le critère de l'\autoref{eq:aricritere} qui nous intéresse. Il mesure l'optimalité de la politique de l'expert pour la récompense renvoyée par \gls{scirl}. Nous allons le borner par un terme qui dépend d'objets dont certains n'ont pas encore été introduits.

   Le premier de ces objet est le coefficient de concentration $\ccoeff$ \citep{munos2007performance}, qui mesure l'écart entre les distributions induites par toutes les politiques. Ici il s'agit d'un maximum qui sélectionne la séquence de politiques dont la distribution est la plus ``différente'' possible de celle de l'expert :
\begin{align}
  \ccoeff &= (1-\gamma)\sum_{t\geq 0} \gamma^t c(t)\label{eq:ccoeffdef}\\
\text{ avec } c(t) &=
  \max_{\pi_1,\dots,\pi_t,s\in\statespace}\frac{(\expertdistrib^\top \transprobmat{\pi_1}\dots
  \transprobmat{\pi_t})(s)}{\expertdistrib(s)}.
\label{eq:ctdef}
\end{align}

Nous rappelons également la définition de l'erreur de classification $\classiferror{\expertdistrib}$ (\autoref{eq:classiferror}) :
\begin{equation}
\classiferror{\expertdistrib} = \expectationknowing{ \indicatorfunc{\classifpolicy(\state) \neq \expertpolicy(\state)}}{\state\sim\expertdistrib}.
\end{equation}

L'erreur d'estimation de l'attribut moyen de l'expert est quantifiée par :
\begin{equation}
\epsilon_{\mu} = \hat{\expertmu} - \expertmu\in\twosetsfunctions{\statespace\times\actionspace}{
(\mathbb{R}^{\dimx{\featurestateactionfunc_{\rewardfunc}}})}.
\end{equation}

En conséquence, L'erreur entre la fonction de score apprise par le classifieur $\classifscorefunc_{\paramirl}$ et la fonction de qualité de l'expert $\quality{\expertpolicy}{\rewardfunc_{\paramirl}}$ pour la récompense de \gls{scirl} vaut :
\begin{align}
\epsilon_Q &=  \classifscorefunc_{\paramirl}-\quality{\expertpolicy}{\rewardfunc_{\paramirl}}\\
& =\transpose{\theta}(\hat{\expertmu} - \expertmu)\\
& = \transpose{\theta}\epsilon_\mu\in\twosetsfunctions{\statespace\times\actionspace}{\mathbb{R}}.
\end{align}
C'est l'espérance sur la distribution stationnaire de l'expert de l'écart maximum entre les actions de ce dernier terme qui intervient dans la borne :
\begin{equation}
\bar{\epsilon}_Q  = \expectationknowing{ \max_{a\in\actionspace}\epsilon_Q(s,a) -
\min_{a\in\actionspace}\epsilon_Q(s,a)}{\state\sim\expertdistrib}.
\end{equation}
Nous avons maintenant introduit les notions nécessaires à l'expression de la borne théorique sur la qualité de la récompense fournie par \gls{scirl} :
\begin{theorem}[Borne des performances de \glstext{scirl}]
  \label{thm:scirl}
  Soit $R_{\theta}$ la fonction de récompense renvoyée par l'\autoref{alg:scirlgenerique}. Soient $C_f$, $\classiferror{\expertdistrib}$
  et $\bar{\epsilon}_Q$ les quantités définies ci-dessus. On a :
  \begin{equation}
    0\leq
    \expectationknowing{V^{\pi^*_{R_{\theta}}}_{R_{\theta}}-V^{\expertpolicy}_{R_{\theta}}}{s\sim\expertdistrib}
    \leq \frac{C_f}{1-\gamma}\left(\bar{\epsilon}_Q +
    \classiferror{\expertdistrib}\frac{2\gamma\|R_{\theta}\|_\infty}{1-\gamma}
    \right).
  \end{equation}
\end{theorem}
La preuve est reportée en \autoref{hier-demo}.

Le terme $\|R_{\paramirl}\|_\infty$ (Le choix de la norme est d'une importance limitée) n'est pas problématique car nous savons depuis la \autoref{hier-recompense} qu'il est possible de normaliser $\paramirl$ sans perte de généralité.

Le terme le moins contrôlable de cette borne est l'erreur $\bar\epsilon_{Q}$, qui dépend de la qualité de l'estimation de l'attribut moyen $\hat\expertmu$. L'usage de \gls{lstdmu} permet cependant de profiter des garanties offertes \autoref{hier-lstdmuerror}.

Lorsque les deux termes d'erreur sont nuls, nous découvrons un corollaire intéressant :
\begin{corollary}[\glstext{scirl} dans le cas idéal]
\label{thm:coroscirl}
Si $\classiferror{\expertdistrib} = 0$ et $\bar\epsilon_Q = 0$, alors $\expertpolicy$ est l'unique politique optimale pour la récompense $\rewardfunc_{\paramirl}$ renvoyée par \gls{scirl}.
\end{corollary}
\begin{proof}
Soit $\policy^H$ une politique optimale pour $\rewardfunc_{\paramirl}$. Comme $\policy^H$ et $\expertpolicy$ sont toutes deux optimales, alors d'après l'\autoref{eq:greedy01} :
\begin{align}
\policy^H &\in \greedy{\expertpolicy}\\
\Rightarrow \forall \state \in \statespace, \policy^H(s) &= \arg\max_{\action \in \actionspace}\quality{\expertpolicy}{\rewardfunc_\theta}(s,a)\\
&=\arg\max_{\action \in \actionspace}\transpose{\paramirl}\expertmu(s,a).
\end{align}
Comme $\bar\epsilon_Q = 0$,
\begin{align}
\forall \state \in \statespace, \policy^H(s) &= \arg\max_{\action \in \actionspace}\classifscorefunc_{\paramirl}(s,a)\\
&=\classifpolicy(s).
\end{align}
Et comme $\classiferror{\expertdistrib} = 0$,
\begin{equation}
\forall \state \in \statespace, \policy^H(s) = \expertpolicy(s).
\end{equation}
\end{proof}

Une autre propriété de \gls{scirl} est d'éviter les solutions dégénérées, en effet la solution dégénérée par excellence, la récompense nulle, minimise le critère utilisé dans la borne\footnote{Pour $\paramirl=0$, $\expertpolicy$ est optimale pour $\rewardfunc_{\paramirl}$, de même que n'importe quelle autre politique.
 } mais implique que la politique de classification $\classifpolicy$ est la politique aléatoire uniforme, atteignant l'erreur de classification maximale $\classiferror{\expertdistrib} = \frac{\card{\actionspace}-1}{\card{\actionspace}}$. Le classifieur cherchant à minimiser $\classiferror{\expertdistrib}$, il est peu probable que ce cas advienne.
\section{Validation empirique}
\label{sec-5-4}
\label{hier-scirlexp}

   Si l'agent est aux commandes d'un système où la moindre déviation par rapport au contrôle optimal entraîne un échec irrécupérable\footnote{Par échec irrécupérable, il est entendu une déviation vers une partie de l'espace d'état qui ne communique pas ou peu avec le reste, et de fait garde l'agent prisonnier sans possibilité de sortie. L'existence d'une telle zone dont la sortie serait rigoureusement impossible violerait la condition d'ergodicité dont nous avons parlé \autoref{hier-tempdyn}.
 }, alors l'heuristique proposée \autoref{eq:heuristiquescirl} ne permet probablement pas une bonne approximation de l'attribut moyen de l'expert, puisqu'elle est intuitivement basée sur le principe que l'expert est capable de revenir sur la bonne trajectoire après qu'un mauvais choix d'action a été effectué. Il faut donc que l'opérateur, en bâtissant la représentation du problème, soit vigilant à ce phénomène. Nous illustrons empiriquement dans cette section que les propriétés nécessaires au bon fonctionnement de \gls{scirl} ne sont pas difficiles à réunir.
\subsection{Introduction}
\label{sec-5-4-1}

   Contrairement à la section précédente, qui étudie le critère de l'\autoref{eq:aricritere} quantifiant le degré d'optimalité de la politique experte vis-à-vis de la récompense trouvée par \gls{scirl}, nous utiliserons ici un oracle pouvant observer des critères semblables à celui de l'\autoref{eq:aricriterebis}, qui mesure l'optimalité de la politique optimale pour la récompense trouvée par \gls{scirl}, par rapport à la récompense inconnue optimisée par l'expert. Nous profitons en cela de ce que la récompense optimisée par l'expert nous est connue sur ces problèmes jouet.

   La politique dont l'optimalité est mesurée est issue de l'optimisation d'un \gls{mdp}, puisqu'elle est optimale pour la récompense $\rewardfunc_{\paramirl}$ trouvée par \gls{scirl}. Il faut bien comprendre que cette étape de résolution d'un \gls{mdp} ne fait pas partie de \gls{scirl} et est indépendante de la résolution du problème de l'\gls{irl}. L'objet d'intérêt de l'\gls{irl} est la récompense $\rewardfunc_{\paramirl}$, nous ne l'optimisons que dans un but d'évaluation, par un procédé dont nous préciserons la nature dans la description des expériences, mais qui n'a pas de réel intérêt. Dans une approche pratique d'imitation par le biais de l'\gls{irl}, il faudrait bien entendu tenir compte de la contrainte qu'est la nécessité de résoudre le \gls{mdp} une fois la récompense trouvée. Il existe de multiples manières de faire\footnote{En l'absence de simulateur, il faudra avoir recours à des méthodes \gls{batch} et \gls{offpolicy}.
 }, le choix de la technique utilisée dépend énormément du contexte et rejoint le problème vaste de l'apprentissage du contrôle optimal. Nous choisissons ici d'exposer des approches efficaces de l'\gls{irl}, facilement applicables à de multiples domaines, sans nous soucier de savoir comment la récompense sera optimisée, tant cette dernière étape est dépendante du problème. Nous la voyons comme un problème parallèle à celui de l'\gls{irl}.

   Nous testons \gls{scirl} dans l'instanciation proposée \autoref{alg:scirlonlyexpert}, en utilisant l'\autoref{eq:montecarlodupauvre} pour l'estimation de l'attribut moyen de l'expert sur deux problèmes jouet, le \gls{mountaincar} \autoref{hier-expmountaincar} et le \gls{highway} \autoref{hier-exphighway}. Dans les deux cas, nous le comparons d'une part à une approche d'imitation supervisée et d'autre part à l'algorithme d'\gls{irl} existant à notre connaissance le mieux adapté à un scénario où la collecte de données est coûteuse, \gls{relent}, dont la version échantillonnée que nous utilisons est décrite \autoref{alg:relent}. La représentation choisie (attributs, \autoref{hier-attributs}, ou noyau, \autoref{hier-svm}) est toujours une représentation naturelle et simple à mettre en œuvre, ne présentant aucune caractéristique \emph{ad-hoc}.

   Rappelons que la capacité de généralisation des méthodes supervisées dépendant en grande partie du soin apporté à la définition de cette représentation, ces expériences devraient montrer la supériorité des méthodes d'\gls{irl} face aux méthodes supervisées en ce qui concerne la généralisation à des parties de l'espace d'état n'ayant pas été explorées par l'expert. Grâce à l'extraction de la fonction de récompense, description succincte de la tâche à effectuer, les algorithmes d'\gls{irl} sont capables de guider l'agent lorsqu'il apprend le contrôle optimal dans la zone inexplorée, tandis que les méthodes supervisées restent myopes et ne peuvent utiliser toute l'information contenue dans les échantillons de l'expert.

   Ces deux expériences devraient également montrer la validité de l'heuristique qui permet à \gls{scirl} de fonctionner avec pour seule information les échantillons provenant de l'expert. Il faut que les performances de \gls{scirl} soient au moins aussi bonnes que celle de \gls{relent}, qui lui dispose d'échantillons aléatoires lui donnant des informations sur la dynamique du système (et ne peut utiliser notre heuristique). 
\subsection{\glstext{scirl} appliqué au \glstext{mountaincar}}
\label{sec-5-4-2}
\label{hier-expmountaincar}

    Comme nous l'avons précisé \autoref{hier-mountaincar}, la démonstration de l'expert ignore toute une partie de l'espace d'état, le quadrant Sud-Est où la voiture est proche du but, mais s'en éloigne. Les données de l'expert sont collectées, en choisissant un état de départ telle que la position est dans l'intervalle $[-1.2;-0.9]$ et la vitesse dans $[-0.07;0]$. La trajectoire est poursuivie jusqu'à ce que l'expert atteigne le but. Autant de trajectoires que nécessaire sont effectuées pour obtenir des ensembles de $10$, $30$, $100$ et $300$ échantillons. La dernière trajectoire est tronquée pour atteindre exactement le nombre d'échantillons voulu (cela signifie typiquement que pour les ensemble de $10$ et $30$ échantillons, l'expert ne fournit même pas une trajectoire complète).

    L'algorithme \gls{relent} ne fonctionnant pas avec uniquement des données de l'expert, un base d'entraînement supplémentaire $\sastrace{\sim}$ lui sera fournie. Cette base est constituée de $1000$ trajectoires de longueur $5$, débutant dans un état choisi au hasard uniformément dans tout l'espace d'état, les actions étant choisies également au hasard. Il est intéressant de noter que cette base contient d'un à deux ordres de grandeur plus d'échantillons que la base experte.

    Cette même base $\sastrace{\sim}$ sera utilisée par \gls{lspi} pour optimiser les récompenses trouvées par \gls{scirl} et \gls{relent}.

    Les attributs utilisés par \gls{lspi}, \gls{scirl} et \gls{relent} sont les mêmes, il s'agit d'un maillage de gaussiennes tel que décrit \autoref{eq:gaussianrbf} et illustré Figure 2.1.

    Le classifieur dont les performances sont illustrées est une \gls{svm} à noyau gaussien dont les hyper-paramètres sont choisis automatiquement par l'implémentation utilisée\footnote{\href{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}{Module SVM de Scikit-learn}
 }. Les performances de cette modèle étant très légèrement supérieures à celles du classifieur structuré de \citet{taskar2005learning} (\autoref{alg:taskar}), ce sont celles que nous affichons ici.

    En ordonnée sur la Figure 5.1,
 nous représentons le nombre de pas nécessaire à l'agent pour qu'il atteigne le but (plafonné à $300$), lorsque l'état de départ est tiré aléatoirement dans l'espace d'état entier. Les valeurs affichées sont la moyenne et l'écart-type sur $100$ expériences. Ce critère est directement lié à la valeur de la politique pour la récompense inconnue $\expertreward$ : si l'expert touche une récompense de $1$ lorsqu'il atteint l'objectif, et que la récompense est nulle partout ailleurs, du fait de l'amortissement temporel la valeur de sa politique en un état vaut $\discount^T$ où $T$ est le nombre de pas qui le sépare de l'objectif. En conséquence, ce que nous traçons est le logarithme en base $\discount$ de la valeur moyenne de la politique.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/mountain_car_scirl.pdf}
\caption[Algorithmes d'imitation sur le \glstext{mountaincar}]{Algorithmes d'imitation sur le \glstext{mountaincar}. La politique de l'expert est également évaluée afin d'offrir un point de référence.}
\label{fig:scirlmountaincar}
\end{figure}

Les résultats présentés Figure 5.1
 sont conformes à ce qui était espéré, en ce sens que les deux algorithmes d'\gls{irl} sont plus performants que la méthode supervisée. On constate cependant que les performances de \gls{relent} sont beaucoup plus proches de celles de la classification qu'elles ne le sont de celles de \gls{scirl}. Ce dernier parvient dès $30$ échantillons à des performances meilleures que celles des deux autres méthodes quelle que soit la quantité de données expertes. Quand au moins une trajectoire complète est présente dans les données, la politique issue de la récompense trouvée par \gls{scirl} est presque aussi bonne que celle de l'expert.

Le calcul de l'attribut moyen de l'expert pour \gls{scirl} est ici effectué avec l'\autoref{eq:montecarlodupauvre} car le problème est déterministe.

En utilisant un peu plus\footnote{La classification utilise une base $\satrace{\expertpolicy}$ tandis que \gls{scirl} utilise une base $\sastrace{\expertpolicy}$ où il convient d'enregistrer l'état d'arrivée.
 } d'information que la méthode de classification, et moins que l'autre méthode d'\gls{irl}, \gls{scirl} est en mesure de trouver une récompense menant à une imitation de qualité.
\subsection{\glstext{scirl} appliqué au \glstext{highway}}
\label{sec-5-4-3}
\label{hier-exphighway}

    Le problème du simulateur de conduite sur autoroute est défini \autoref{hier-highway}. L'espace est discret, mais grand. Les attributs choisis sont les attributs standards décrits \autoref{eq:attributsdiscrets}. La méthode supervisée de classification structurée considérée cette fois ci (\autoref{alg:taskar}) utilise ces attributs, et ne va donc pas disposer d'information particulière sur la dynamique ou la sémantique du problème. Les démonstrations de l'expert étant fort courtes, il va falloir attendre que celui-ci explore une vaste partie de l'espace d'état avant que la classification ne parvienne à apprendre un contrôle correct. L'expert fournit une base de données $\sastrace{\expertpolicy}$ organisée sous la forme de $\nbsamples$ trajectoires de longueur $\nbsamples$, avec $\nbsamples$ prenant successivement pour valeur $3$, $7$, $10$, $15$ et $20$.

    L'expert est un agent optimal entraîné sur une récompense pénalisant les collisions ($-1$ par collision) et les sorties de route ($-0.5$ par sortie de route) mais exigeant une grande vitesse ($1$ si la vitesse est maximale).

    \gls{relent} nécessite pour fonctionner une autre base de données $\sastrace{\sim}$. Celle-ci est constituée de $100$ marches aléatoires de longueur $10$, démarrant dans un état choisi aléatoirement. Il y a cette fois-ci encore besoin d'un nombre d'échantillons d'un à deux ordres de grandeur plus grand que les démonstrations expertes.

    \gls{scirl}, pour calculer l'attribut moyen de l'expert, utilise la méthode de l'\autoref{eq:montecarlodupauvre} car l'inversion de matrice nécessaire au fonctionnement de \gls{lstdmu} est, lorsque répétée sur les $100$ expériences que nous avons menées pour obtenir l'information de variance, un petit peu longue à effectuer. Il serait possible de régler ce problème en choisissant les attributs avec plus de soin, mais nous souhaitons montrer que \gls{scirl} est une méthode générique qui ne nécessite que peu ou pas de travail de la part de l'opérateur.

    L'espace d'état étant discret et la dynamique connue, nous pouvons optimiser les récompenses trouvées par les algorithmes d'\gls{irl} avec une méthode de \gls{dp}. Le critère que nous utilisons pour comparer les différentes politiques est la moyenne sur tous les états de la valeur de la politique. Même les démonstrations les plus longues de l'expert ($400$ échantillons) n'explorent qu'une petite partie de l'espace d'état-action (de cardinal $3645$), il est donc primordial de bien généraliser pour obtenir un contrôle correct.

    Les résultats obtenus sont donnés Figure 5.2.
Les résultats présentent bien la hiérarchie à laquelle nous nous attendions : l'algorithme supervisé ne parvient à apprendre un bon contrôle que petit à petit, à mesure que la taille de la base de données experte croît. L'ordre est le même que pour le problème du \gls{mountaincar}, mais les performances de \gls{relent} diffèrent en ceci qu'elles sont maintenant beaucoup plus proches de celles de \gls{scirl} (qui reste malgré tout au dessus) que de la classification.

    Dans un domaine où les attributs (discrets) n'ont pas été en mesure de fournir à la méthode supervisée la moindre information \emph{a priori} sur le système (précédemment pour le \gls{mountaincar}, le noyau gaussien de la \gls{svm} possède la propriété de cohérence spatiale\footnote{Deux états proches auront des attributs proches.
 }), les méthodes d'\gls{irl} présentent de bien meilleures capacités de généralisation.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/highway_scirl.pdf}
\caption[Algorithmes d'imitation sur le \glstext{highway}]{Algorithmes d'imitation sur le \glstext{highway}. En plus de la valeur de l'expert comme dans la Figure 5.1,
% \autoref{fig:scirlmountaincar},
la valeur d'une politique apprise sur une récompense aléatoire donne un point de référence de valeur minimale. On montre ici l'écart-type et la moyenne sur $100$ expériences.}
\label{fig:scirlhighway}
\end{figure}
\section{Conclusion}
\label{sec-5-5}

    Dans les deux expériences que nous venons de décrire, les méthodes supervisées ne peuvent qu'imiter l'expert là où sa démonstration donne des informations. La capacité de généralisation est sévèrement limitée par l'absence de savoir \emph{a priori} dans la définition du problème. Les algorithmes d'\gls{irl}, à l'inverse, sont capables, à partir d'une démonstration locale, d'extraire une description satisfaisante de la tâche accomplie par l'expert.

    \gls{scirl} en particulier a pu fonctionner à partir des mêmes données que la méthode supervisée (à l'ajout de l'état d'arrivée $s'$ près), sans données supplémentaires, à l'inverse de \gls{relent}. Il a su exploiter les informations contenues dans les transitions pour introduire la dynamique du système dans la résolution du problème de l'imitation. L'heuristique que nous avons présentée \autoref{eq:heuristiquescirl} et que la théorie ne permet pas facilement d'analyser a fait ses preuves sur des problèmes déterministes. \gls{scirl} est moins gourmand en temps de calcul et en données qu'un algorithme d'\gls{rl} comme par exemple \gls{lspi}. Il faut tenir compte de cela, puisque contrairement aux méthodes supervisées qui renvoient directement une politique de contrôle, \gls{scirl} renvoie une récompense. Cette récompense devra par la suite être optimisée, en utilisant par exemple un algorithme d'\gls{rl} auquel il faut fournir de l'information dont \gls{scirl} n'a pas besoin. Le gain en performance justifie probablement l'effort lié à la résolution du \gls{mdp}, des méthodes \emph{ad-hoc} permettant de traiter les problèmes qui ne manqueront pas de se poser.

    Les résultats théoriques et pratiques montrent que \gls{scirl} résout bien le problème de l'\gls{irl} tel que nous l'avion posé en \autoref{hier-problemdef}. Les soucis liés au choix des attributs, bien que toujours présents, sont moins préoccupants que pour les approches supervisées grâce à l'utilisation de l'attribut moyen, qui prend la dynamique en compte. L'estimation de celui-ci lorsqu'uniquement des données expertes sont disponibles est rendue possible grâce à l'usage d'une heuristique qui semble bien fonctionner en pratique. Les solutions dégénérées sont évitées grâce à la structure de \gls{scirl}, qui adapte une méthode de classification. L'algorithme de l'état de l'art le plus à même de suivre \gls{scirl} sur les problèmes où peu d'information est disponible, \gls{relent}, est moins performant et nécessite plus d'information. Si cependant cette information est facilement disponible, \gls{scirl} peut en profiter grâce à l'emploi de \gls{lstdmu}.
\chapter{Apprentissage par renforcement inverse en cascadant classification et régression}
\label{sec-6}
\label{hier-cascading}

Les avantages présentés par \gls{scirl} découlent de l'identification de la fonction de score d'un classifieur à la fonction de qualité de l'expert. Cela permet d'introduire la dynamique du système dans une approche supervisée, combinant les avantages en terme de généralisation caractéristiques des approches d'\gls{irl} et la simplicité d'utilisation des approches supervisées.

Cependant \gls{scirl} nécessite l'emploi d'un classifieur à fonction de score linéairement paramétrée, et l'estimation de l'attribut moyen de l'expert. Il existe comme nous allons le voir dans cette section un autre moyen de profiter de la similitude entre fonction de score et fonction de qualité. Nous présentons \autoref{hier-csidesc} l'algorithme \gls{cascading}, pour lequel il existe également une garantie théorique (\autoref{hier-csitheo}) ainsi qu'une étude empirique permettant de comparer ses performances à celles de \gls{scirl} (\autoref{hier-csiempirique}) et de valider l'heuristique qui, là encore, permet de dépasser le cadre de l'analyse théorique en n'utilisant que des données expertes.
\section{Description}
\label{sec-6-1}
\label{hier-csidesc}

   Notre nouvelle approche de l'\gls{irl}, \gls{cascading}, est basée à l'instar de \gls{scirl} sur le constat (\autoref{hier-classifirl}) de similitude entre la fonction de score $\classifscorefunc$ d'un classifieur et la fonction de qualité $\quality{\expertpolicy}{\expertreward}$ de l'expert (qui par hypothèse est un agent optimal) pour la récompense inconnue $\expertreward$ que justement, nous aimerions connaître.
\subsection{Principe général}
\label{sec-6-1-1}

   Là où cette approche diverge de \gls{scirl}, c'est dans la manière d'introduire la dynamique du \gls{mdp} en utilisant la fonction de score $\classifscorefunc$ apprise par un classifieur. Nous n'allons pas ici utiliser une paramétrisation astucieuse de cette fonction de score. Nous allons utiliser l'équation de Bellman (\autoref{eq:qdef}) qui définit de manière récursive la fonction de qualité, et isoler la récompense. On rappelle que $\expertpolicy(s) \in \arg\max_{\action \in \actionspace}\quality{\expertpolicy}{\expertreward}(s,a)$ car $\expertpolicy$ est optimale. La fonction de qualité de l'expert\footnote{Qui est donc une fonction de qualité optimale pour $\expertreward$
 } obéit donc alors à l'équation d'optimalité de Bellman :
\begin{equation}
\quality{\expertpolicy}{\expertreward}(\state,\action) = \expertreward(\state,\action) + \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\action}{\state'} \quality{\expertpolicy}{\expertreward}(\state',\expertpolicy(\state')),
\end{equation}
donc la récompense $\expertreward$ que l'on cherche suit :
\begin{equation}
\expertreward(\state,\action) = \quality{\expertpolicy}{\expertreward}(\state,\action) - \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\action}{\state'} \quality{\expertpolicy}{\expertreward}(\state',\expertpolicy(\state')).
\end{equation}

La fonction de score $q$ d'un classifieur joue un rôle similaire à celui d'une fonction de qualité optimale :
\begin{equation}
\classifpolicy(s) \in \arg\max_{\action \in \actionspace}q(s,a)
\label{eq:toto1}
\end{equation}

En remplaçant la fonction de qualité de l'expert par son \emph{alter-ego} la fonction de score d'un classifieur imitant l'expert, nous définissons la ``récompense du classifieur'' $\Rc$ \footnote{La politique du classifieur $\classifpolicy$ est une politique optimale pour la récompense $\Rc$ puisque du fait de l'\autoref{eq:toto1} et de la définition de $\Rc$, $q$ vérifie l'équation d'optimalité de Bellman pour $\Rc$.
 }:
\begin{align}
\Rc(\state,\action) &= \classifscorefunc(\state,\action) - \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\action}{\state'} \classifscorefunc(\state',\max_{a'}q(s',a')),\\
 &= \classifscorefunc(\state,\action) - \discount \sum_{\state'\in \statespace}\transprobfunceval{\state}{\action}{\state'} \classifscorefunc(\state',\classifpolicy(\state')).
\label{eq:rc}
\end{align}

La présence dans cette équation des probabilités de transition ne nous permet pas d'espérer évaluer directement $\Rc$, nous souhaitons en effet que \gls{cascading} fonctionne dans un cadre où les données sont difficiles à réunir.

Supposons la disponibilité d'une base d'échantillons non experte $\sastrace{\sim}$ :
\begin{equation}
\sastrace{\sim} = \{\state_{\datasetindex},\action_{\datasetindex},\state'_{\datasetindex}\}_{0\leq\datasetindex\leq\nbsamples}.
\end{equation}
L'état d'arrivée $\state'_{\datasetindex}$ est tiré selon la distribution $\transprobfunceval{\state_{\datasetindex}}{\action_{\datasetindex}}{\cdot}$, dès lors il est possible d'établir une version échantillonnée de l'\autoref{eq:rc} :
\begin{equation}
\rsample_{\datasetindex} = q(s_{\datasetindex},a_{\datasetindex}) - \gamma q(s'_{\datasetindex},\classifpolicy(s'_{\datasetindex})).
\end{equation}
L'idée générale de \gls{cascading} et qui lui donne le nom de \emph{cascadée}, est, après avoir appris la fonction de score $\classifscorefunc$ et la politique de classification $\classifpolicy$ en entraînant un classifieur sur une base de données experte $\satrace{\expertpolicy}$, de fournir à un régresseur la base de données
\begin{equation}
\rctrace = \{\left(\state_{\datasetindex},\action_{\datasetindex}\right),\rsample_{\datasetindex}|\state_{\datasetindex},\action_{\datasetindex},\state'_{\datasetindex}\in\sastrace{\sim}\},
\label{eq:csinormalrctrace}
\end{equation}
afin qu'il apprenne $\apprRc$, une approximation de la version échantillonnée de la récompense induite par la fonction de score du classifieur $\Rc$. L'\autoref{alg:csigenerique} résume cela.

\begin{algorithm}
\Entree{
Une base d'entraînement établie par l'expert $\satrace{\expertpolicy}$\;
Une base d'entraînement non experte $\sastrace{\sim}$\;
} 
\Donnees{
Une méthode de classification à fonction de score\;
Une méthode de régression\;
}
\Sortie{Une fonction de récompense $\apprRc$}
Entraîner le classifieur sur la base de données experte \satrace{\expertpolicy}, obtenant ainsi $\classifscorefunc$ et $\classifpolicy$\;
Établir la base d'entraînement
\begin{equation*}
\rctrace = \{\left(\state_{\datasetindex},\action_{\datasetindex}\right),\rsample_{\datasetindex} = q(s_{\datasetindex},a_{\datasetindex}) - \gamma q(s'_{\datasetindex},\classifpolicy(s'_{\datasetindex}))|\state_{\datasetindex},\action_{\datasetindex},\state'_{\datasetindex}\in\sastrace{\sim}\}\textrm{\;}
\end{equation*}
Entraîner le régresseur sur la base $\rctrace$, obtenant ainsi $\apprRc$\;
\Retour{$\apprRc$}
\caption{Principe générique de \glstext{cascading}}
\label{alg:csigenerique}
\end{algorithm}

Grâce à l'utilisation de l'équation de Bellman, cascader deux approches supervisées permet (comme nous allons le montrer dans les deux sections qui suivent) de trouver une récompense. La plupart des méthodes de classification utilisent une fonction de score. N'importe quelle méthode de régression peut être utilisée pour la seconde étape. Cette grande latitude dans le choix d'un algorithme ou d'un autre pour chacune des étapes est une force de \gls{cascading} qui peut ainsi profiter du large éventail de techniques de l'apprentissage supervisé afin de limiter l'effort de l'opérateur lorsque cela est possible, ou de permettre à celui-ci d'adapter chaque étape au problème courant lorsque cela est nécessaire.

\subsection{Heuristique pour étendre \glstext{cascading} au cas où seules les données expertes sont disponibles}
\label{sec-6-1-2}

    La version générique de \gls{cascading} repose sur la disponibilité de deux bases de données, une experte et l'autre non, comme c'est aussi le cas pour par exemple \gls{relent}. Nous avons vu que \gls{scirl} peut, grâce à une heuristique, fonctionner à l'aide d'uniquement des données expertes. Il est souhaitable d'étendre cette capacité à \gls{cascading}. Voyons comment l'absence de base d'entraînement $\sastrace{\sim}$ impacte l'algorithme, et voyons surtout comment compenser ce manque.

    Si l'on ne dispose que d'une base d'entraînement experte $\sastrace{\expertpolicy}$ et que l'on essaye de bâtir la base $\rctrace$ avec, nous ne disposerons que d'échantillons :
\begin{equation}
\left(s_{\datasetindex},\expertpolicy(\state_{\datasetindex})\right) , \rsample_{\datasetindex} = q(s_{\datasetindex},\expertpolicy(\state_{\datasetindex})) - \gamma q(s'_{\datasetindex},\classifpolicy(s'_{\datasetindex})).
\end{equation}
Ces données, les seules dont on dispose, sont tirées selon la dynamique induite par la politique experte. Or un \gls{mdp} est plus riche que la seule chaîne de Markov induite par une politique. Particulièrement, il n'y a parmi ces échantillons aucune information concernant les couples $(\state_{\datasetindex},\action \neq \expertpolicy(\state_{\datasetindex}))$. Entraîner un régresseur sur ces seules données revient à espérer que le régresseur parviendra à généraliser à des actions qu'il n'a jamais vues. \`A moins de disposer d'information \emph{a priori} et de parvenir à les intégrer au régresseur, cela est probablement une mauvaise idée. Typiquement, rien n'empêche le régresseur de décider que les couples $(\state_{\datasetindex},\action \neq \expertpolicy(\state_{\datasetindex}))$ sont associés à une valeur plus grande que $\rsample_{\datasetindex}$.

Pour contraindre les solutions au problème de l'\gls{irl}, dont on a vu qu'il en existait un nombre infini dont certaines dégénérées, beaucoup d'algorithmes existants (\citep{abbeel2004apprenticeship,syed2008apprenticeship,ratliff2009learning,ramachandran2007bayesian}) prennent le parti d'affirmer que le choix d'une action effectué par l'expert doit être meilleur (selon un certain critère) que les autre choix possibles, par une certaine marge. Notre heuristique adopte le même parti pris en injectant dans la base $\rctrace$ les échantillons :
\begin{align}
\left(s_{\datasetindex},\forall \action \neq\expertpolicy(\state_{\datasetindex})\right) ,\rsample_{min} = \min_{\datasetindex\in \llbracket 1;\nbsamples\rrbracket}\rsample_{\datasetindex} - 1.
\label{eq:csiheuristique}
\end{align}
Cela signifie en effet que la fonction apprise par le régresseur ne doit pas associer à un choix en désaccord avec une démonstration de l'expert une récompense supérieure ou égale à n'importe quel autre choix expert.

La récapitulation de l'approche \gls{cascading} en utilisant cette heuristique est fournie \autoref{alg:csiheuristique}.

\begin{algorithm}
\Entree{
Une base d'entraînement établie par l'expert $\sastrace{\expertpolicy}$\;
} 
\Donnees{
Une méthode de classification à fonction de score\;
Une méthode de régression\;
}
\Sortie{Une fonction de récompense $\apprRc$}
Entraîner le classifieur sur la base de données experte \sastrace{\expertpolicy}, obtenant ainsi $\classifscorefunc$ et $\classifpolicy$\;
Établir la base d'entraînement
\begin{align*}
\rctrace &= \left\{\left(\state_{\datasetindex},\action_{\datasetindex} = \expertpolicy(\state_{\datasetindex})\right) , \rsample_{\datasetindex} = q(s_{\datasetindex},\expertpolicy(\state_{\datasetindex})) - \gamma q(s'_{\datasetindex},\classifpolicy(s'_{\datasetindex}))|\state_{\datasetindex},\action_{\datasetindex},\state'_{\datasetindex}\in\sastrace{\expertpolicy}\right\}\\
&\cup\left\{\left(s_{\datasetindex},\forall \action \neq\expertpolicy(\state_{\datasetindex})\right) , \rsample_{min} = \min_{\datasetindex\in \llbracket 1;\nbsamples\rrbracket}\rsample_{\datasetindex} - 1\right\} \textrm{\;}
\end{align*}
Entraîner le régresseur sur la base $\rctrace$, obtenant ainsi $\apprRc$\;
\Retour{$\apprRc$}
\caption{\glstext{cascading} dans sa version utilisant l'heuristique}
\label{alg:csiheuristique}
\end{algorithm}
\section{Validation théorique}
\label{sec-6-2}
\label{hier-csitheo}

   La borne que nous présentons quant aux performances de \gls{cascading} est comme celle de \gls{scirl} basée sur le critère de l'\autoref{eq:aricritere}. Elle montre qu'en contrôlant les erreurs de chacune des deux étapes supervisées, la récompense fournie par \gls{cascading} devient telle que l'expert est optimal vis-à-vis d'elle.

   Pour exprimer cette borne, nous avons besoin de l'erreur de classification, dont la définition est toujours la même :
\begin{equation}
\classiferror{\expertdistrib} = \expectationknowing{ \indicatorfunc{\classifpolicy(\state) \neq \expertpolicy(\state)}}{\state\sim\expertdistrib}.
\end{equation}

Nous utiliserons également l'erreur $\epsilon_R$ due au régresseur : 
\begin{align}
\epsilon_R&=\max_{\pi\in A^S}\|\epsilon_\pi^R\|_{1,\expertdistrib}\\
\textrm{avec : }\epsilon^R_\pi(\cdot)&=R^C(\cdot,\pi(\cdot))-\hat{R}^C(\cdot,\pi(\cdot)),\\
\textrm{et la norme $\ell_1$ $\rho$-pondérée : }\|f\|_{1,\rho} &= \expectationknowing{|f(x)|}{x\sim\rho},
\end{align}
ainsi que l'écart de score :
\begin{equation}
\Delta q = \max_{s\in \statespace}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\classifpolicy(s))-\min_{a\in A}q(s,a)),
\end{equation}
que nous pouvons normaliser, puisque le choix de l'action se fait sur les valeurs comparées et non absolues des scores.

Enfin, nous introduisons un coefficient de concentration plus fin que celui que nous avons utilisé pour borner la performance de \gls{scirl}, qui montre le désaccord de distributions induite par une politique $\pi$ et la politique de l'expert :
\begin{align}
\ccoeffpi{\pi}&=(1-\gamma)\sum_{t\geq0}\gamma^t c_\pi(t)\\ 
\textrm{avec : }c_{\pi}(t)&=\max_{s\in S}\frac{(\transpose{\expertdistrib}(\transprobmat{\pi})^t)(s)}{\expertdistrib(s)}
\end{align}

\begin{theorem}[Borne des performances de \glstext{cascading}]
\label{thm:csi}
Soit $\Rc$ la récompense retournée par l'\autoref{alg:csigenerique}. Soient $\ccoeffpi{\optimalpolicy{\apprRc}}$, $\epsilon_C$ et $\epsilon_R$ les quantités définies ci-dessus. On a :
\begin{equation}
0\leq\expectationknowing{\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}(s)-\rlvalue{\expertpolicy}{\apprRc}(s)}{s\sim\expertdistrib}\leq \frac{1}{1-\gamma}\left(\epsilon_C\Delta q +\epsilon_R(1+\ccoeffpi{\optimalpolicy{\apprRc}})\right).
\end{equation}
\end{theorem}

Prenons le temps d'analyser les différences entre cette borne et celle du \autoref{thm:scirl}. Le terme $\bar\epsilon_{Q}$ n'apparaît pas plus dans cette borne que le terme $\epsilon_R$ présent ici n'apparaît pour \gls{scirl}. Il faut également se méfier de l'erreur de classification $\classiferror{\expertdistrib}$. Pour \gls{scirl}, la classification est informée de la structure temporelle du \gls{mdp} par l'utilisation de l'attribut moyen. Dans \gls{cascading}, l'introduction de la structure temporelle du \gls{mdp} n'intervient qu'après l'étape de classification, le terme d'erreur de classification $\classiferror{\expertdistrib}$ présent ici traite d'une étape ``myope'', tandis que celui de \gls{scirl} traite d'une approximation de la fonction de qualité de l'expert, paramétrée linéairement grâce à l'attribut moyen.

Le terme $\Delta_q$ peut cacher un facteur en $\frac{1}{1-\discount}$, on retrouve le même type de constantes que dans le \autoref{thm:scirl}.

Enfin, les coefficients de concentration diffèrent légèrement. Celui présent ici est plus petit, il ne correspond pas à un pire des cas comme celui de \gls{scirl}. On peut le voir comme une mesure de la similitude entre $\expertpolicy$ et $\optimalpolicy{\apprRc}$. Si $\optimalpolicy{\apprRc} \approx \expertpolicy$ alors $\ccoeffpi{\optimalpolicy{\apprRc}}\approx 1$ \footnote{Il est nécessaire de connaître $\apprRc$, le résultat de \gls{cascading}, pour calculer $\ccoeffpi{\optimalpolicy{\apprRc}}$. Si l'on souhaite disposer d'une borne à priori alors il est possible de relâcher la borne en utilisant le coefficient $\max_\pi C_\pi$, qui reste plus petit que le coefficient de la borne de \gls{scirl} (\autoref{eq:ccoeffdef}).
 }.

La preuve, due au travail de Bilal Piot\footnote{Co-auteur des contributions sur \gls{cascading} \citep{klein2012cascading,klein2012cascading1}.
 }, est reportée en \autoref{hier-demo}.

Les deux principaux termes d'erreurs présents dans cette borne, $\epsilon_C$ et $\epsilon_R$ sont conformes aux termes d'erreur usuels en apprentissage supervisé. Utiliser les nombreuses astuces et adaptations rendues possibles par l'usage d'un classifieur et d'un régresseur de l'état de l'art permet d'espérer minimiser ces deux termes. Les résultats théoriques de l'apprentissage supervisé peuvent être facilement intégrés à \gls{cascading}. Il est plus raisonnable d'espérer contrôler le terme $\epsilon_R$ présent dans cette borne que le terme $\bar\epsilon_Q$ de la borne du \autoref{thm:scirl}.

Cette borne est également rassurante en ce qui concerne l'apparition de solutions dégénérées ; si les deux termes d'erreur supervisés sont annulés on obtient un résultat similaire à celui du \autoref{thm:coroscirl}.
\begin{corollary}[\glstext{cascading} dans le cas idéal]
  Si $\expertdistrib>0$ et le classifieur et le régresseur sont parfaits ($\epsilon_C = 0$ et $\epsilon_R = 0$), alors $\expertpolicy$ est l'\emph{unique} politique optimale pour $\hat R^C$.
\label{thm:cor}
\end{corollary}

\begin{proof}
  La fonction $\classifscorefunc$ est la fonction de qualité optimale pour la politique $\classifpolicy$ vis-à-vis de la récompense $R^C$, par définition (voir \autoref{eq:rc}). Si $\epsilon_C = 0$, alors $\classifpolicy = \expertpolicy$. Cela signifie que $\forall s, \expertpolicy(s)$ est le seul élément de l'ensemble $\arg\max_{a\in \actionspace}q(s,a)$. Du coup, $\classifpolicy = \expertpolicy$ est l'unique politique optimale pour $R^C$.
  Et puisque $\epsilon_R = 0$, on a $\apprRc = \Rc$, d'où le résultat.
\end{proof}
\section{Validation empirique}
\label{sec-6-3}
\label{hier-csiempirique}

La borne théorique donnée à la section précédente ne prend pas l'heuristique proposée \autoref{eq:csiheuristique} en compte, puisque celle-ci introduit des échantillons non conformes à l'\autoref{eq:csinormalrctrace}. Nous allons donc, comme nous l'avons fait pour \gls{scirl} \autoref{hier-scirlexp} tester empiriquement la version heuristique de \gls{cascading} afin de montrer que l'heuristique est valable.
\subsection{Introduction}
\label{sec-6-3-1}

Nous retrouvons ici les bancs d'essai utilisés \autoref{hier-scirlexp}, le problème du \gls{mountaincar} et celui du \gls{highway}. La classification s'étant avérée moins performante que les méthodes d'\gls{irl}, nous nous concentrons ici sur celles-ci, et comparons \gls{scirl}, \gls{cascading} et \gls{relent}. Le critère utilisé est le même que celui des expériences précédentes, à savoir un critère donnant une idée de l'optimalité de la politique optimale pour $\apprRc$ par rapport à la récompense inconnue $\expertreward$.

L'instanciation de \gls{cascading} utilisée est celle de l'\autoref{alg:csiheuristique}. Nous préciserons dans les sous-sections le choix du classifieur et du régresseur.
\subsection{\glstext{cascading} appliqué au \glstext{mountaincar}}
\label{sec-6-3-2}

Le protocole expérimental est le même que dans la \autoref{hier-expmountaincar}. Le classifieur choisi pour \gls{cascading} est le même que celui dont les performances sont illustrées Figure 6.1.
Il s'agit de d'une SVM\footnote{\href{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}{Module SVM de Scikit-learn}
 }. Le noyau est gaussien. Nous sommes dans un cas où l'ingénierie déployée par l'opérateur est minimale. Le régresseur est choisi dans le même esprit, il s'agit d'une SVR (\emph{Support Vector Regression}, Régresseur à Vecteur Support)\footnote{\href{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVR.html}{Module SVR de Scikit-learn}
 } utilisée avec ses paramètres par défaut.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/mountain_car_csi.pdf}
\caption{Algorithmes d'\glstext{irl} sur le \glstext{mountaincar}.}
\label{fig:csimountaincar}
\end{figure}

Les résultats obtenus après avoir répété $100$ fois l'expérience sont donnés Figure 6.1,
il est clair que les deux nouveaux algorithmes d'\gls{irl} \gls{scirl} et \gls{cascading} sont plus performants que \gls{relent} (qui était déjà lui-même plus performant que l'approche supervisée). Il n'est pas clair en revanche qui de \gls{scirl} ou \gls{cascading} est le plus performant. Pour en avoir le cœur net, nous allons procéder à un test statistique d'égalité des moyennes. Si un test d'égalité des variances de Bartlett réussit, un test de Student est réalisé, si le Bartlett échoue, c'est un test de Welch. Les $p$-valeurs sont données table 6.1. On voit que la seule abscisse pour laquelle on peut affirmer que les moyennes sont inégales est la première, où \gls{cascading} est plus performant que \gls{scirl} (et aussi que \gls{relent}). Pour les abscisses suivantes, \gls{scirl} et \gls{cascading} ont peu ou prou les mêmes performances, très proches des performances expertes.

\begin{table}
\centering
\begin{tabular}{cc}
\toprule
Taille de la base experte & $p$-valeur\\
\midrule
$ 10$ & $\mathbf{1.5e-12}$\\
$ 30$ & $3.8e-01$\\
$100$ & $1.3e-02$\\
$300$ & $7.4e-01$\\
\bottomrule
\end{tabular}
\caption[Départager \glstext{scirl} et \glstext{cascading} sur le \glstext{mountaincar}]{Test d'égalité des moyennes de Student ou de Welch concernant \gls{cascading} et \gls{scirl} sur le problème du \gls{mountaincar}. Des $p$-valeurs élevées ($>1.0\e{-02}$) signifient que l'hypothèse d'égalité des moyennes ne peut être rejetée.}
\label{tab:mountaincar}
\end{table}
\subsection{\glstext{cascading} appliqué au \glstext{highway}}
\label{sec-6-3-3}

Là encore, le protocole expérimental est le même que celui que nous avons suivi \autoref{hier-exphighway}. L'approche supervisée, l'\autoref{alg:taskar} armé des attributs standard définis \autoref{eq:attributsdiscrets} étant moins performante que les algorithmes d'\gls{irl}, nous ne donnons pas ses performances ici. Elle constitue la première étape de \gls{cascading}. La seconde étape, la régression, est effectuée par un régresseur aux moindres carrés (\autoref{alg:lsr}), utilisant également les attributs de l'\autoref{eq:attributsdiscrets}.
\begin{algorithm}
\Entree{Une base d'entraînement $\rctrace$\;
}
\Donnees{
Une fonction d'attribut $\featurestateactionfunc$ sur l'espace d'état action\; 
}
\Sortie{Une approximation $\apprRc$}
Initialiser la matrice $X$ de taille $\nbsamples\times \dimphi$ :
\begin{equation}
X = \matrixbyterm{\component{\featurestateactionfunc}{\dimindex}(\state_{\datasetindex},\action_{\datasetindex})}{\datasetindex}{\dimindex}\textrm{\;}
\end{equation}
Initialiser le vecteur $y$ de taille $\nbsamples\times 1$ :
 \begin{equation}
\component{y}{\datasetindex} = \rsample_{\datasetindex}\textrm{\;}
\end{equation}
Calculer le vecteur de paramètres $\beta$ :
\begin{equation}
\beta = (\transpose{X}X)^{-1}\transpose{X} y
\end{equation}
\Retour{La fonction $\transpose{\beta}\featurestateactionfunc$}
\caption{Régresseur aux moindres carrés pour \glstext{cascading}}
\label{alg:lsr}
\end{algorithm}

Les résultats obtenus sont donnés Figure 6.2.
Les performances des algorithmes d'\gls{irl} étant très proches les unes des autres, nous offrons un grossissement Figure 6.3
 qui montre que \gls{cascading} semble disposer d'un avantage sur les deux autres approches. Cela est confirmé par un test statistique d'égalité des moyennes (comme précédemment, il s'agit d'un test de Student ou de Welch selon qu'un test de Bartlett d'égalité des variances réussit ou non) dont les résultats sont fournis table 6.2 : pour les abscisses de la Figure 6.3,
les performances de \gls{cascading} sont bel et bien supérieures à celles de \gls{scirl}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/highway_csi1.pdf}
\caption{Algorithmes d'\glstext{irl} sur le \glstext{highway}.}
\label{fig:csihighway1}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figures/highway_csi2.pdf}
\caption{Grossissement de la Figure 6.2.}%\autoref{fig:csihighway1}.}
\label{fig:csihighway2}
\end{figure}

\begin{table}
\centering
\begin{tabular}{cc}
\toprule
Taille de la base experte & $p$-valeur\\
\midrule
      $  9$ & $3.0e-01$\\
      $ 49$ & $\mathbf{8.9e-03}$\\
      $100$ & $\mathbf{1.8e-03}$\\
      $225$ & $\mathbf{2.4e-05}$\\
      $400$ & $\mathbf{2.0e-50}$\\
\bottomrule
\end{tabular}
\caption[Départager \glstext{scirl} et \glstext{cascading} sur le \glstext{highway}]{Test d'égalité des moyennes de Student ou de Welch concernant \gls{cascading} et \gls{scirl} sur le problème du \gls{highway}. Des $p$-valeurs élevées ($>1.0\e{-02}$) signifient que l'hypothèse d'égalité des moyennes ne peut être rejetées.}
\label{tab:highway}
\end{table}
\section{Conclusion}
\label{sec-6-4}

   L'algorithme \gls{cascading} que nous venons de présenter est basé sur un principe commun avec \gls{scirl}, qui consiste à voir la fonction de score d'un classifieur comme une fonction de qualité optimale. L'introduction de la structure temporelle du \gls{mdp} dans l'approche diffère cependant. L'utilisation cascadée de deux approches supervisée offre en effet des garanties théoriques plus serrées, avec des termes d'erreur plus faciles à maîtriser.

   La souplesse offerte par la liberté de choix des deux approches supervisées laisse à l'opérateur la possibilité d'adapter l'approche au problème qu'il cherche à résoudre. Une heuristique est offerte qui permet à \gls{cascading} de fonctionner avec une expertise humaine minimale en n'utilisant que des données expertes. Cela est illustré sur les mêmes bancs d'essai que ceux qui ont permis la validation empirique de \gls{scirl}. Les performances empiriques offertes par \gls{cascading} sont légèrement supérieures à celles de \gls{scirl}. Les deux approches sont plus performantes que l'approche supervisée au prix cependant de devoir résoudre le \gls{mdp} une fois la récompense obtenue.
\chapter{Conclusions et perspectives}
\label{sec-7}
\label{hier-conclusion}
\section{Rappel des contributions}
\label{sec-7-1}

L'un des freins à l'adoption généralisée de la robotique dans la société est le coût élevé des dispositifs automatiques. Du fait de l'augmentation exponentielle des capacités techniques du matériel, le goulot d'étranglement se situe de plus en plus au niveau du coût de conception des systèmes intelligents. Le développement d'algorithmes génériques pouvant s'adapter à une vaste gamme de problèmes est l'un des moyens par lesquels la réduction de ces coûts de développement pourrait être réalisée.

L'\gls{rl} est un cadre de travail offrant de tels algorithmes. Malheureusement, ils nécessitent pour fonctionner une description formelle de la tâche qu'ils doivent accomplir sous la forme d'une fonction de récompense. La conception d'une bonne fonction de récompense est un problème non trivial nécessitant des réglages fins de la part de l'opérateur.

Savoir effectuer la tâche (ou savoir comment faire, sans pour autant savoir le faire\footnote{Pour jongler, lancez les balles en l'air sans les laisser tomber au sol.
 }) n'est malheureusement pas suffisant pour pouvoir définir rapidement une bonne fonction de récompense, les capacités d'introspection de l'humain étant notoirement mauvaises.

La nécessité d'un travail non trivial de la part d'un opérateur qualifié rend l'application des algorithmes d'\gls{rl} difficile voire impossible en dehors d'un laboratoire\footnote{De recherche ou de conception.
 }. Pouvoir établir la fonction de récompense non plus formellement mais à l'aide de démonstrations permettrait, nous l'espérons, la commercialisation de systèmes intuitifs où le client n'aurait qu'à réaliser quelques fois la tâche pour que le robot l'imite.

Le \autoref{hier-formalisme} a introduit le formalisme de l'apprentissage par imitation et traité dans un premier temps des approches supervisées, qui n'apprennent pas de fonction de récompense mais imitent directement le comportement de l'expert. L'importance du choix des attributs a également été abordée et un état de l'art de l'apprentissage par imitation par le biais d'approches supervisées a été fourni. Dans un second temps, le formalisme des \gls{mdp} et de l'\gls{rl} a été introduit, nous permettant de préciser notamment les notions de fonction de récompense et de fonction de valeur. Une fois cela posé, nous avons pu définir le problème de l'\gls{irl}, ainsi que les deux critères d'évaluation que nous avons utilisés respectivement dans notre analyse théorique et dans nos évaluations empiriques.

Le \autoref{hier-etatdelart} propose un état de l'art des méthodes d'\gls{irl}. Nous y avons fait apparaître la nécessité pour la plupart des algorithmes d'\gls{irl} de disposer au moins d'un simulateur du système à contrôler. Cette nécessité vient de la structure commune à ces algorithmes, qui implique la résolution répétée du \gls{mdp} concerné. La quantité d'information nécessaire est bien plus importante que pour les méthodes supervisées étudiées au chapitre précédent, qui ne nécessitent que des données produites par l'expert. Certaines méthodes récentes d'\gls{irl} s'affranchissent de la structure itérative des premières approches, mais la plupart ne peuvent passer à l'échelle, ou nécessitent de connaître les probabilités de transition du \gls{mdp}. Un algorithme (\gls{relent}) sort cependant du lot, puisqu'il peut fonctionner en utilisant uniquement des données produites par l'expert et des données produites par une politique aléatoire, afin d'avoir un échantillonnage représentatif du \gls{mdp}.

Grâce à la première de nos contributions, \gls{lstdmu} (\autoref{hier-lstdmu}), il est possible de faire fonctionner les algorithmes itératifs de la littérature à partir des mêmes entrées que \gls{relent}. En effet, \gls{lstdmu} permet l'estimation de l'attribut moyen d'une politique arbitraire en mode \gls{batch} et \gls{offpolicy}, en adaptant l'algorithme d'estimation de la valeur \gls{lstd} à l'estimation de l'attribut moyen, dont nous remarquons qu'il s'agit de la généralisation vectorielle d'une fonction de valeur. Pour peu que l'on puisse résoudre le \gls{mdp} avec un algorithme d'\gls{rl} présentant les mêmes caractéristiques (par exemple \gls{lspi}), alors il est possible de trouver la politique optimale pour une récompense arbitraire et d'estimer l'attribut moyen de cette politique grâce à une seule base de données, échantillonnées une fois pour toutes.

Nous proposons \autoref{hier-scirl} un nouvel algorithme d'\gls{irl}, \gls{scirl}, qui diffère du schéma itératif. En utilisant l'attribut moyen de l'expert (qu'il est possible d'estimer en utilisant \gls{lstdmu}, notre contribution du chapitre précédent) pour paramétrer la fonction de score d'un classifieur à marge structurée (décrit au \autoref{hier-formalisme}), nous obtenons directement une fonction de récompense. L'analyse théorique que nous fournissons démontre que la politique de l'expert tend à être optimale pour cette fonction de récompense quand l'erreur de classification et l'erreur d'estimation de l'attribut moyen sont petites. Nous proposons une heuristique permettant de s'affranchir de la nécessité de disposer d'une base de données représentative du \gls{mdp}, et illustrons empiriquement les bonnes performances de \gls{scirl} comparativement à une méthode supervisée et à \gls{relent}. Les capacités de généralisation des algorithmes d'\gls{irl} sont empiriquement supérieures à celles de la méthode supervisée. Il faut cependant tenir compte du fait que la sortie d'un algorithme d'\gls{irl} est une récompense, qu'il faut optimiser pour obtenir une politique de contrôle tandis que les algorithmes supervisés renvoient directement une politique.

Plus souple, l'algorithme \gls{cascading} présenté au \autoref{hier-cascading} ne nécessite pas d'estimer l'attribut moyen de l'expert, et consiste en la cascade de deux méthode supervisées : une étape de classification permettant d'apprendre une fonction de score et une étape de régression permettant, grâce à une inversion de l'équation de Bellman, d'apprendre la récompense associée à cette fonction de score. Cet algorithme dispose de garanties théoriques : la politique de l'expert est optimale pour la récompense apprise si les erreurs de classification et de régression sont petites. La panoplie d'approches de regression et de classification fournie par la littérature nous permet de contrôler ces deux termes plus facilement que ceux présents dans la borne de \gls{scirl}. Nous proposons une nouvelle heuristique, qui là encore nous permet de nous affranchir de la nécessité de données autres que celles échantillonnées par l'expert, comme l'illustrent les expériences que nous menons. Sur ces expériences, les performances de \gls{cascading} sont légèrement supérieures à celles de \gls{scirl}.

Nous démontrons empiriquement que les deux algorithmes d'\gls{irl} que nous proposons peuvent fonctionner avec peu de données. Bien que les problèmes sur lesquels nous les testons soient des problèmes jouet, nos algorithmes sont à notre connaissance les seuls algorithmes d'\gls{irl} pouvant fonctionner uniquement avec des données expertes\footnote{Moyennant l'utilisation d'heuristiques
 }. Les résultats obtenus montrent des performances supérieures à celles des algorithmes supervisés et à une approche récente d'\gls{irl} ; il faut cependant trouver le moyen d'optimiser la récompense pour obtenir une politique tandis que les algorithmes supervisés la fournissent directement.\\

Les systèmes réels sont souvent fragiles et interagir avec peut être coûteux. Il est également difficile de les simuler. On comprend alors l'intérêt de développer des algorithmes moins gourmands en données.
\section{Perspectives de recherche}
\label{sec-7-2}

Les deux nouveaux algorithmes d'\gls{irl} que nous proposons disposent de garanties théoriques et une étude empirique illustre le fait qu'ils sont plus performants que les méthodes de l'état de l'art. Il serait souhaitable de renforcer cette étude empirique en les appliquant à d'autres problèmes, que ce soit des problèmes déjà résolus par d'autres méthodes, à des fins comparatives, ou des problèmes ouverts. Notamment, deux domaines ont commencé à être explorés, mais les travaux n'ont pu être menés à terme et de fait n'apparaissent pas dans ce manuscrit. Le premier est l'\emph{Arcade Learning Environment} (ALE) : il s'agit d'un émulateur d'Atari 2600 permettant de jouer à une vaste collection de jeux vidéos. Le second est l'application d'algorithmes d'\gls{irl} à la commande d'un bras robot à partir de signaux issus d'une interface cerveau/machine (\emph{Brain Computer Interface}, BCI)\footnote{Voir la \autoref{hier-autrestravaux} pour une description des travaux effectués sur la BCI.
 }.  Dans ces deux cas, la difficulté n'est pas dans l'application des algorithmes d'\gls{irl} eux-même, mais dans l'optimisation de la récompense qu'ils trouvent afin d'obtenir une politique de contrôle. Nous avons dans nos travaux considéré que cette étape ne posait pas problème, afin de nous concentrer sur le problème de l'\gls{irl} tel qu'envisagé par la communauté. L'application à des problèmes pratiques doit faire face à l'optimisation de la récompense, il faut alors faire appel à des méthodes de l'état de l'art en \gls{rl} et des adaptations \emph{ad-hoc} au problème (par exemple dans le choix des attributs).

Pour remédier de manière plus générale à ce problème\footnote{Qui n'affecte pas les approches supervisées.
 }, nous pourrions également tenter de développer un cadre pour les algorithmes d'\gls{irl} dans lequel l'optimisation finale de la récompense est simplifiée ou contournée. La politique de classification du classifieur paramétré par l'attribut moyen de l'expert qui constitue \gls{scirl} est par exemple une bonne politique de départ pour un algorithme d'\gls{rl} en ligne. Les données nécessaires à l'apprentissage de la politique de contrôle (dont \gls{scirl} n'a pas besoin, grâce à son heuristique, pour trouver la récompense) pourraient alors être recueillies par la politique de classification, qui convergerait alors petit à petit grâce à l'algorithme d'\gls{rl} en ligne vers la politique optimale pour la récompense trouvée par \gls{scirl}. La politique issue de la première étape supervisée de \gls{cascading} pourrait être utilisée de la même manière. Il est également possible de poursuivre l'approche envisagée dans \citep{geist2013cascading}, consistant à instancier \gls{cascading} avec une \gls{svm} et une SVR et à réunir les deux problèmes d'optimisation en un seul, dont la résolution permettrait d'obtenir directement une politique prenant en compte la structure de la récompense.

Finalement, il faudrait intégrer l'\gls{rl} et l'\gls{irl} dans des systèmes informatiques réels aux mains d'utilisateurs non qualifiés. Cela nécessitera sans doute d'importants travaux concernant l'intégration de multiples algorithmes dans un schéma hiérarchique : avant d'apprendre à un robot à faire un mojito, il faut qu'il apprenne à saisir un verre et une bouteille. Mais lorsqu'on lui apprend la recette, on ne veut pas avoir à se soucier des détails de la saisie manuelle d'objet, on fonctionne à un niveau d'abstraction plus élevé. Il faut donc qu'une politique par exemple ``Saisir un verre'' faisant appel à des actions telles que ``Fermer doigt 1 de la main gauche'' et ``Tourner poignet gauche dans le sens direct'' puisse être utilisée directement comme une action par une politique de plus haut niveau, à côté d'actions telles que ``Piler la glace'' et ``Placer la rondelle de citron''. Il faudra également effectuer des progrès dans le domaine de l'interface homme/machine pour parvenir à gérer intuitivement des systèmes aussi complexes. Au-delà de l'aspect robotique grand public, les chercheurs et ingénieurs ont eux aussi beaucoup à gagner à manipuler intuitivement les machines, mais cela n'est pas facilement compatible avec la nécessité de rapidement changer de multiples paramètres.

\appendix
\chapter{Démonstrations}
\label{sec-8}
\label{hier-demo}
\section{Borne sur la performance de \glstext{scirl}}
\label{sec-8-1}

Avant de démontrer le \autoref{thm:scirl}, rappelons un résultat dont nous allons avoir besoin :
\begin{lemma}[{\citep[Lemme 4.2]{munos2007performance}}]
Soit $A$ une matrice inversible telle que tous les éléments de $A^{-1}$ sont positifs. Les solutions de l'inégalité $Au\leq b$ sont aussi solution de $u\leq A^{-1}b$.
\end{lemma}

\begin{proof}[Preuve du \autoref{thm:scirl}]
Nous allons décomposer le terme $V^{\pi^*_{R_{\theta}}}_{R_{\theta}}-V^{\expertpolicy}_{R_{\theta}}$ (qui est positif ou, au mieux, nul, puisque $\pi^*_{R_{\theta}}$ est optimale) afin de faire apparaître le résidu de Bellman $\expectationknowing{\bellmanoptopeval{\rewardfunc_{\theta}}{V^{\expertpolicy}_{R_{\theta}}}(s)-V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}$. On rappelle que les définitions des opérateurs d'optimalité et d'évaluation de Bellman $\bellmanoptopeval{\rewardfunc}{}$ et $\bellmanevalopeval{\pi}{\rewardfunc}{}$ sont données équations \ref{eq:bellmanoptop} et \ref{eq:bellmanevop}.
  \begin{equation}
    V^{\pi^*_{R_{\theta}}}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}} =  V^{\pi^*_{R_{\theta}}}_{R_{\theta}}  - \bellmanevalopeval{\pi^*_{R_{\theta}}}{R_{\theta}}V^{\expertpolicy}_{R_{\theta}} +
    {\bellmanevalopeval{\pi^*_{R_{\theta}}}{R_{\theta}}V^{\expertpolicy}_{R_{\theta}} - \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}} + \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}}
\end{equation}
or $\bellmanevalopeval{\pi^*_{R_{\theta}}}{R_{\theta}} V^{\expertpolicy}_{R_{\theta}}\leq \bellmanoptopeval{\rewardfunc_{\theta}}{}
  V^{\expertpolicy}_{R_{\theta}}$, donc ${\bellmanevalopeval{\pi^*_{R_{\theta}}}{R_{\theta}} V^{\expertpolicy}_{R_{\theta}}- \bellmanoptopeval{\rewardfunc_{\theta}}{}
  V^{\expertpolicy}_{R_{\theta}}\leq 0}$, donc :
\begin{align}
    V^{\pi^*_{R_{\theta}}}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}} &\leq  V^{\pi^*_{R_{\theta}}}_{R_{\theta}}  - \bellmanevalopeval{\pi^*_{R_{\theta}}}{R_{\theta}}V^{\expertpolicy}_{R_{\theta}} + \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}}\\
     &= \bellmanevalopeval{\optimalpolicy{\rewardfunc_{\paramirl}}}{\rewardfunc_{\theta}}{} V^{\pi^*_{R_{\theta}}}_{R_{\theta}}  - \bellmanevalopeval{\pi^*_{R_{\theta}}}{R_{\theta}}V^{\expertpolicy}_{R_{\theta}} + \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}}\\
     &= \rewardfunc_{\theta}+\discount \transprobmat{\pi^*_{R_{\theta}}} V^{\pi^*_{R_{\theta}}}_{R_{\theta}}  - \rewardfunc_{\theta}-\discount \transprobmat{\pi^*_{R_{\theta}}} V^{\expertpolicy}_{R_{\theta}} + \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}}\\
     &= \discount \transprobmat{\pi^*_{R_{\theta}}} V^{\pi^*_{R_{\theta}}}_{R_{\theta}}  -\discount \transprobmat{\pi^*_{R_{\theta}}} V^{\expertpolicy}_{R_{\theta}} + \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}}\\
     &= \discount \transprobmat{\pi^*_{R_{\theta}}}\left( V^{\pi^*_{R_{\theta}}}_{R_{\theta}}  - V^{\expertpolicy}_{R_{\theta}}\right) + \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}},\\
\Rightarrow (I-\gamma \transprobmat{\pi^*_{R_{\theta}}})(V^{\pi^*_{R_{\theta}}}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}})&\leq\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}}
\end{align}
En vertu de~\citep[Lemme~4.2]{munos2007performance}, car $(I-\gamma \transprobmat{\pi^*_{R_{\theta}}}) = \sum_{t\geq 0}(\transprobmat{\pi^*_{R_{\theta}}})^t$ a tous ses éléments positifs, 
\begin{equation}
  V^{\pi^*_{R_{\theta}}}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}} \leq (I-\gamma \transprobmat{\pi^*_{R_{\theta}}})^{-1} (\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}- V^{\expertpolicy}_{R_{\theta}}).
 \end{equation}
Comme $(I-\gamma\transprobmat{\pi^*_{R_{\theta}}} )^{-1} = \sum\limits_{t\geq 0}\gamma^t (\transprobmat{\pi^*_{R_{\theta}}})^t$,
\begin{equation}
V^{\pi^*_{R_{\theta}}}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}} \leq \sum\limits_{t\geq 0}\gamma^t (\transprobmat{\pi^*_{R_{\theta}}})^t (\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}- V^{\expertpolicy}_{R_{\theta}}).
\end{equation}

D'après la définition de $c(t)$ (\autoref{eq:ctdef}) et du coefficient de concentration $\ccoeff$ (\autoref{eq:ccoeffdef}) :
  \begin{align}
    0\leq\expectationknowing{V^{\pi^*_{R_{\theta}}}_{R_{\theta}}(s)-V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib} &\leq \sum_{t\geq 0}c(t)
    \expectationknowing{\bellmanoptopeval{\rewardfunc_{\theta}}{}V^{\expertpolicy}_{R_{\theta}}(s) - V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}\\
    0\leq\expectationknowing{V^{\pi^*_{R_{\theta}}}_{R_{\theta}}(s)-V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib} &\leq \frac{C_f}{1-\gamma}
    \expectationknowing{\bellmanoptopeval{\rewardfunc_{\theta}}{}V^{\expertpolicy}_{R_{\theta}}(s) - V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}\\
    \label{erooesidual}
  \end{align}

Il nous faut maintenant borner le résidu de Bellman 
\begin{equation}
\expectationknowing{\bellmanoptopeval{\rewardfunc_{\theta}}{}V^{\expertpolicy}_{R_{\theta}}(s) -
  V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib},
\end{equation}
que nous décomposons en :
\begin{equation}
  \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}} = \bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}} - \bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}} + \bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}}- V^{\expertpolicy}_{R_{\theta}}.
    \label{erooecomposition}
  \end{equation}
  Nous allons borner séparément 
\begin{align}
\expectationknowing{\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}(s) - \bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}\textrm{ et }\expectationknowing{\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}}(s) - V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}.
\end{align}

  La politique de classification $\classifpolicy$ est gloutonne vis-à-vis de $\classifscorefunc_{\paramirl}=\transpose{\paramirl}\hat{\mu}^{\expertpolicy}$. Donc, pour chaque couple état-action
  $(s,a)\in\statespace\times \actionspace$ :
  \begin{equation}
    \classifscorefunc_{\paramirl}(s,\classifpolicy(s))\geq
    \classifscorefunc_{\paramirl}(s,a)
    \Leftrightarrow
    \quality{\expertpolicy}{\rewardfunc_{\paramirl}}(s,a) \leq \quality{\expertpolicy}{\rewardfunc_{\paramirl}}(s,\classifpolicy(s)) +
    \epsilon_Q(s,\classifpolicy(s)) - \epsilon_Q(s,a).
  \end{equation}
  Par définition, $\quality{\expertpolicy}{\rewardfunc_{\paramirl}}(s,a) = [\bellmanevalopeval{a}{\rewardfunc_{\paramirl}}{} V^{\expertpolicy}_{R_{\theta}}](s)$ et
  $\quality{\expertpolicy}{\rewardfunc_{\paramirl}}(s,\classifpolicy(s)) = [\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{} V^{\expertpolicy}_{R_{\theta}}](s)$. Donc, pour $s\in\statespace$:
  \begin{align}
    \forall a\in A,\; [\bellmanevalopeval{a}{\rewardfunc_{\paramirl}}{} V^{\expertpolicy}_{R_{\theta}}](s) &\leq [\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}
    V^{\expertpolicy}_{R_{\theta}}](s) + \epsilon_Q(s,\classifpolicy(s))-\epsilon_Q(s,a)
    \\
    \Rightarrow [\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}](s) &\leq [\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}
    V^{\expertpolicy}_{R_{\theta}}](s) + \max_{a\in \actionspace}\epsilon_Q(s,a)-\min_{a\in
    \actionspace}\epsilon_Q(s,a).
  \end{align}
  En passant à l'espérance selon $\expertdistrib$ et tout en rappelant que 
  $\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}\geq V^{\expertpolicy}_{R_{\theta}}$, le premier terme est borné :
  \begin{equation}
    0 \leq \expectationknowing{\bellmanoptopeval{\rewardfunc_{\theta}}{} V^{\expertpolicy}_{R_{\theta}}(s) - \bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}
    V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib} \leq \bar{\epsilon}_Q.
    \label{eroo1}
  \end{equation}
  Il reste finalement à borner le terme $\expectationknowing{\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}}(s) -
  V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}$. \`A cet effet nous introduisons $M\in\mathbb{R}^{|\statespace|\times |\statespace|}$ la matrice diagonale définie par $M = \diag (\indicatorfunc{\classifpolicy(s)\neq\expertpolicy(s)})$. Grâce à cela, l'opérateur de Bellman $\bellmanevalopeval{\classifpolicy}{\rewardfunc}{}$ peut s'écrire, pour tout $V\in\mathbb{R}^{\statespace}$:
  \begin{equation}
    \bellmanevalopeval{\classifpolicy}{\rewardfunc}{}V = R + \gamma M \transprobmat{\classifpolicy} V + \gamma (I-M)\transprobmat{\expertpolicy} V
    = R + \gamma \transprobmat{\expertpolicy} V + \gamma M (\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V.
  \end{equation}
  En appliquant cet opérateur à  $V^{\expertpolicy}_{R_{\theta}}$ et en utilisant le fait que $R +
  \gamma \transprobmat{\expertpolicy} V^{\expertpolicy}_{R_{\theta}} = \bellmanevalopeval{\expertpolicy}{\rewardfunc_{\paramirl}}{} V^{\expertpolicy}_{R_{\theta}} = V^{\expertpolicy}_{R_{\theta}}$, nous obtenons :
  \begin{align}
    \bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}} &= \gamma M
    (\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}
    \\\Rightarrow |\expertdistrib^\top (\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}} - V^{\expertpolicy}_{R_{\theta}})| &= \gamma
    |\transpose{\expertdistrib} M (\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}    |.
    % &=|\scalarprod{\transpose{\expertdistrib} M}{(\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}}|
  \end{align}
  D'après l'inégalité de Hoelder\footnote{Qui stipule que si $1\leq p,q\leq\infty,\frac{1}{p}+\frac{1}{q} = 1$, alors \begin{equation*}|<f,g>|\leq \|f\|_p\|g\|_q.\end{equation*}
 } appliquée ici avec $p=1$ et $q=\infty$ :
\begin{align}
|\transpose{\expertdistrib} M (\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}|&\leq \|\transpose{\expertdistrib} M \|_1\|(\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}\|_{\infty}\\
&=\classiferror{\expertdistrib}\|(\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}\|_{\infty}
\end{align}
Bornons le terme $(\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}$ pour un état quelconque :
\begin{align}
(\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}(s)&=|\sum_{s'}\left(\transprobfunceval{s}{\classifpolicy(s)}{s'}-\transprobfunceval{s}{\expertpolicy(s)}{s'}\right)V^{\expertpolicy}_{R_{\theta}}(s)|\\
&\leq \sum_{s'}|\left(\transprobfunceval{s}{\classifpolicy(s)}{s'}-\transprobfunceval{s}{\expertpolicy(s)}{s'}\right)||V^{\expertpolicy}_{R_{\theta}}(s)|\\
&= \sum_{s'}|\left(\transprobfunceval{s}{\classifpolicy(s)}{s'}-\transprobfunceval{s}{\expertpolicy(s)}{s'}\right)|\frac{\|R_{\theta}\|_{\infty}}{1-\discount}\\
&\leq\left(\sum_{s'}\transprobfunceval{s}{\classifpolicy(s)}{s'}+\sum_{s'}\transprobfunceval{s}{\expertpolicy(s)}{s'}\right)\frac{\|R_{\theta}\|_{\infty}}{1-\discount}\\
&\leq\frac{2}{1-\gamma}\|R_{\theta}\|_{\infty}.
\end{align} 
Ce terme ne dépend pas de $s$ donc la norme infinie est identique, et finalement on a $\|(\transprobmat{\classifpolicy}-\transprobmat{\expertpolicy})V^{\expertpolicy}_{R_{\theta}}\|_\infty
  \leq \frac{2}{1-\gamma}\|R_{\paramirl}\|_\infty$, ce qui permet de borner le dernier terme
  \begin{equation}
    \left|\expectationknowing{\bellmanevalopeval{\classifpolicy}{\rewardfunc_{\paramirl}}{}V^{\expertpolicy}_{R_{\theta}}(s) - V^{\expertpolicy}_{R_{\theta}}(s)}{s\sim\expertdistrib}\right| \leq
    \classiferror{\expertdistrib} \frac{2\gamma}{1-\gamma} \|R_{\paramirl}\|_\infty.
    \label{eroo2}
  \end{equation}
  Injecter les bornes des équations~\eqref{eroo1}
  et~\eqref{eroo2} dans l'équation~\eqref{erooesidual} achève la démonstration.%, which concludes the proof.
\end{proof}
\section{Borne sur la performance de \glstext{cascading}}
\label{sec-8-2}

\begin{proof}[Preuve du \autoref{thm:csi}]
%The difference between $\Rc$ and $\apprRc$ is noted $\epsilon^R=\Rc-\apprRc$.
%We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof, not to be confused with $R_E$ the unknown reward %function the expert optimizes:
%\begin{equation}
%\forall (s,a)\in S\times A, R^E(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\expertpolicy(s')).
%\end{equation}
%We no have the following vectorial equalities $
%\Rc_a=q_a-\gamma P_aq_{\classifpolicy} ;
%R^E_a=q_a-\gamma P_aq_{\expertpolicy};
%\epsilon^R_a=\Rc_a-\apprRc_a$.
Nous allons borner l'expression : 
\begin{equation}
\expectationknowing{\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}(s)-\rlvalue{\expertpolicy}{\apprRc}(s)}{s\sim\expertdistrib}
\end{equation}
qui est positive ou, au mieux, nulle puisque $\optimalpolicy{\apprRc}$ est optimale. Pour cela, décomposons le terme dans l'espérance : 
\begin{equation}
\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}-\rlvalue{\expertpolicy}{\apprRc}=
(\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}-\rlvalue{\optimalpolicy{\apprRc}}{\Rc})+
(\rlvalue{\optimalpolicy{\apprRc}}{\Rc}-\rlvalue{\expertpolicy}{\Rc})+
(\rlvalue{\expertpolicy}{\Rc}-\rlvalue{\expertpolicy}{\apprRc}).
\end{equation}
Nous allons borner ces trois termes en deux fois.
%First, let $\pi$ be a given deterministic policy. We have, using $\epsilon^R = \Rc - \hat \Rc$:
%$V^{\pi}_{\Rc}-V^{\pi}_{\apprRc}=V^{\pi}_{\epsilon^R}=(I-\gamma P_\pi)^{-1}\epsilon^R_{\pi}$.
%If $\pi=\expertpolicy$, we have, thanks to the power series expression of $(I-\gamma \transprobmat{\expertpolicy})^{-1}$, the definition of $\expertdistrib$ and the definition of the %$\mu$-weighted $L_1$ norm, one property of which is that $\forall X, \mu^TX \leq \|X\|_{1,\mu}$:

Pour toute politique $\policy$,
\begin{align}
  \transpose{\expertdistrib}(\rlvalue{\pi}{\Rc}-\rlvalue{\pi}{\apprRc})&=\transpose{\expertdistrib}\rlvalue{\policy}{(\Rc-\apprRc)}\\
&= \transpose{\expertdistrib}(I-\gamma \transprobmat{\policy})^{-1} \epsilon^R_{\policy}\\
&\leq\frac{\ccoeffpi{\policy}}{1-\gamma}\transpose{\expertdistrib}\epsilon^R_{\policy}\\
&\leq\frac{\ccoeffpi{\policy}}{1-\gamma}\|\epsilon^R_{\policy}\|_{1,\expertdistrib}.
\end{align}
Le passage de la deuxième à la troisième ligne vient de la décomposition en série entière de $(I-\gamma \transprobmat{\policy})^{-1}$, et le passage à la dernière ligne est possible car l'une des propriétés de la norme $L_1$ $\rho$-pondérée est que $\forall X, \rho^TX \leq \|X\|_{1,\rho}$.

Nous sommes maintenant en mesure de borner le premier et dernier terme :
\begin{align}
\transpose{\expertdistrib}(\rlvalue{\expertpolicy}{\Rc}-\rlvalue{\expertpolicy}{\apprRc}) &\leq \frac{\ccoeffpi{\expertpolicy}}{1-\discount}\epsilon_R\\
 &= \frac{1}{1-\discount}\epsilon_R\\
\textrm{et }\transpose{\expertdistrib}(\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}-\rlvalue{\optimalpolicy{\apprRc}}{\Rc}) &\leq\frac{\ccoeffpi{\optimalpolicy{\apprRc}}}{1-\gamma}\epsilon_R\\
\Rightarrow \transpose{\expertdistrib}\left((\rlvalue{\expertpolicy}{\Rc}-\rlvalue{\expertpolicy}{\apprRc}) + (\rlvalue{\optimalpolicy{\apprRc}}{\apprRc}-\rlvalue{\optimalpolicy{\apprRc}}{\Rc})\right) &\leq \frac{1+\ccoeffpi{\optimalpolicy{\apprRc}}}{1-\gamma}\epsilon_R
\label{eq:csiproof1}
\end{align}
Il nous reste maintenant à trouver une borne pour le deuxième terme. Dans cet objectif, nous introduisons la récompense :
\begin{equation}
 \bar\Rc(s,a)=q(s,a) -\gamma\sum_{s'\in \statespace}\transprobfunceval{s}{a}{s'}q(s',\expertpolicy(s')),
\end{equation}
ce qui nous permet de décomposer le deuxième terme en :
\begin{equation}
\rlvalue{\optimalpolicy{\apprRc}}{\Rc}-\rlvalue{\expertpolicy}{\Rc}=(\rlvalue{\optimalpolicy{\apprRc}}{\Rc}-V^{\classifpolicy}_{\Rc})+(V^{\classifpolicy}_{\Rc}-V^{\expertpolicy}_{\bar\Rc})+(V^{\expertpolicy}_{\bar\Rc}-\rlvalue{\expertpolicy}{\Rc}).
\end{equation}
En conséquence directe de leurs définitions, $\classifpolicy$ est optimale pour $\Rc$, donc :
\begin{equation}
\rlvalue{\optimalpolicy{\apprRc}}{\Rc}-V^{\classifpolicy}_{\Rc}\leq0,
\end{equation}
et conséquemment :
\begin{equation}
\rlvalue{\optimalpolicy{\apprRc}}{\Rc}-\rlvalue{\expertpolicy}{\Rc}\leq(\rlvalue{\classifpolicy}{\Rc}-\rlvalue{\expertpolicy}{\bar\Rc})+(\rlvalue{\expertpolicy}{\bar\Rc}-\rlvalue{\expertpolicy}{\Rc}).
\end{equation}
Toujours par définition, nous avons $\rlvalue{\classifpolicy}{\Rc}=q(\cdot,\classifpolicy(\cdot))$ et $\rlvalue{\expertpolicy}{\bar\Rc}=q(\cdot,\expertpolicy(\cdot))$, donc:
\begin{align}
\transpose{\expertdistrib}(\rlvalue{\classifpolicy}{\Rc}-\rlvalue{\expertpolicy}{\bar\Rc})&=\transpose{\expertdistrib}(q(\cdot,\classifpolicy(\cdot))-q(\cdot,\expertpolicy(\cdot)))\\
&=\sum_{s\in S}\expertdistrib(s)(q(s,\classifpolicy(s))-q(s,\expertpolicy(s)))\indicatorfunc{\classifpolicy(s)\neq\expertpolicy(s)}\\
&\leq\Delta q\sum_{s\in S}\expertdistrib(s)[\mathds{1}_{\{\classifpolicy(s)\neq\expertpolicy(s)\}}]\\
&\leq \Delta q \epsilon_C.
\end{align}
Pour conclure, constatons que :
\begin{align}
\bar\Rc(\state,\expertpolicy(\state))-\Rc(\state,\expertpolicy(\state)) &=  (q(\state,\expertpolicy(\state)) -\gamma\transprobmat{\expertpolicy}q(\state,\expertpolicy(\state)))\\
&- (\classifscorefunc(\state,\expertpolicy(\state)) - \discount\transprobmat{\expertpolicy} \classifscorefunc(\state,\classifpolicy(\state)))\\
&= \discount\transprobmat{\expertpolicy} \classifscorefunc(\state,\classifpolicy(\state)) -  \discount\transprobmat{\expertpolicy} \classifscorefunc(\state,\expertpolicy(\state)))\\
&=\gamma\transprobmat{\expertpolicy}( \classifscorefunc(\state,\classifpolicy(\state)) - q(\state,\expertpolicy(\state))),
\end{align}
ce qui nous permet de borner le dernier terme (de l'extension du deuxième terme) :
\begin{align}
\transpose{\expertdistrib}(\rlvalue{\expertpolicy}{\bar\Rc}-\rlvalue{\expertpolicy}{\Rc})&=\transpose{\expertdistrib}(I-\gamma \transprobmat{\expertpolicy})^{-1}(\bar\Rc(\cdot,\expertpolicy(\cdot))-\Rc(\cdot,\expertpolicy(\cdot)))\\
&=\transpose{\expertdistrib}(I-\gamma \transprobmat{\expertpolicy})^{-1}\gamma \transprobmat{\expertpolicy}(\classifscorefunc(\cdot,\classifpolicy(\cdot)) - q(\cdot,\expertpolicy(\cdot)))\\
&=\frac{\gamma}{1-\gamma}\transpose{\expertdistrib}(\classifscorefunc(\cdot,\classifpolicy(\cdot)) - q(\cdot,\expertpolicy(\cdot)))\\
&\leq \frac{\gamma}{1-\gamma}\Delta q \epsilon_C.
\end{align}

Finalement la borne pour le deuxième terme est : 
\begin{align}
\transpose{\expertdistrib}(\rlvalue{\optimalpolicy{\apprRc}}{\Rc}-\rlvalue{\expertpolicy}{\Rc})&\leq (\Delta q +\frac{\gamma}{1-\gamma}\Delta q) \epsilon_C\\
& \leq \frac{\Delta q}{1-\gamma}\epsilon_C.
\label{eq:csiproof2}
\end{align}
En combinant les équations \ref{eq:csiproof1} et \ref{eq:csiproof2}, nous obtenons la borne du théorème.
\end{proof}
\chapter{Problèmes jouet}
\label{sec-9}
\label{hier-jouets}

   Pour accompagner les explications nous étudierons trois problèmes jouet. Ces problèmes sont suffisamment simples pour qu'une interprétation des grandeurs en jeu soit possible, mais suffisamment complexes pour présenter un défi aux approches existantes. Ils serviront également de bancs d'essai lors de nos études empiriques des nouveaux algorithmes que nous proposons.
\section{Problème du pendule inversé}
\label{sec-9-1}
\label{hier-pendule}

Le problème du pendule inversé consiste à maintenir en équilibre une masse située au bout d'un bras en commandant les mouvements du chariot à la base du bras. Il s'agit d'un problème continu, la position du pendule étant décrite par l'angle du bras, à valeurs dans $[-\frac{\pi}{2},\frac{\pi}{2}]$ et sa vitesse à valeurs dans $\reals$.

\begin{marginfigure}
\centering
\includegraphics[width=\textwidth]{Figures/Pendule.png}
\caption[Pendule inversé]{Schéma de principe du pendule inversé : le chariot est libre de bouger selon l'axe $x$, le but est de maintenir le pendule en équilibre en commandant le chariot.}
\end{marginfigure}

Ce problème a été utilisé comme banc de test de l'algorithme d'\gls{rl} \gls{lspi}. \citet{lagoudakis2003least} en décrivent la dynamique, basée sur la résolution des équations différentielles du mouvement par la méthode d'intégration d'Euler.
\begin{itemize}
\item On trouve l'accélération du pendule en utilisant les équations du mouvement de Newton.
\end{itemize}
\begin{equation}
a = {g\cdot \sin(\theta) -{m\cdot l\cdot\dot{\theta}^2\cdot\sin(2\cdot\theta)\over 2(m+M)} - 
                    {\cos(\theta)\cdot F\over (m+M)} \over {4\cdot {l\over 3} - {m\cdot l\cdot\cos(\theta)^2\over m+M}}}
\end{equation}
\begin{itemize}
\item A chaque pas de temps, on trouve la nouvelle position avec :
\end{itemize}
\begin{equation}
\theta \leftarrow \theta + \dot{\theta}\cdot \Delta t, \Delta t = 0.1 s
\end{equation}
\begin{itemize}
\item La vitesse est ensuite également mise à jour :
\end{itemize}
\begin{equation}
\dot{\theta} \leftarrow \dot{\theta} + a \cdot \Delta t, \Delta t = 0.1s
\end{equation}

L'expert pour ce problème est un agent entraîné avec \gls{lspi} (voir aussi \autoref{hier-algosrl}); il parvient à maintenir le pendule en équilibre indéfiniment, mais la simulation s'arrête après $3000$ pas de temps, soit 5 minutes de temps simulé.
\section{Problème du \glstext{mountaincar}}
\label{sec-9-2}
\label{hier-mountaincar}

Dans le problème du \gls{mountaincar}, une voiture placée dans un creux doit gravir une côte trop pentue pour qu'elle puisse la surmonter du premier coup. La solution du problème consiste à reculer, donc à s'éloigner de l'objectif, en remontant la pente opposée puis à profiter de l'élan pour gravir la pente. Le processus peut-être répété si besoin. Cette nécessité contre-intuitive de devoir s'éloigner de l'objectif pour pouvoir l'atteindre fait de ce problème jouet un bon test pour les algorithmes d'\gls{rl}. Il a été proposé pour la première fois par \citet{moore1990efficient} et a été popularisé lors son introduction dans le livre de \citet{sutton1998reinforcement}.

La Figure 1.2 
décrit les variables qui représentent l'information dont dispose l'agent pour ce problème. Il y a bien sûr la position de la voiture : le point le plus à gauche a une coordonnée de $-1.2$, l'objectif, situé le plus à droite, se trouve à la coordonnée $0.6$. Le fond de la vallée à la coordonnée $-0.5$. La deuxième information est la vitesse de la voiture, elle peut aller de $-0.07$ à $0.07$ où un chiffre positif désigne naturellement un déplacement vers la droite et négatif vers la gauche. La vitesse est plafonnée, et si la voiture atteint l'extrémité gauche, sa vitesse est annulée et elle reste en place.

La trajectoire de l'expert représentée Figure 1.2 
débute dans le quadrant Sud-Ouest de l'espace d'état, correspondant à une position éloignée de l'objectif et une vitesse dans la direction opposée à l'objectif. La dynamique de l'expert renverse ensuite la vitesse (quadrant Nord-Ouest) rapprochant la voiture de l'objectif (quadrant Nord-Est), jusqu'à ce qu'elle l'atteigne. Dans nos études empiriques, les démonstrations de l'expert suivront ce modèle : le point de départ sera choisi aléatoirement dans le quadrant Sud-Ouest \footnote{Plus précisément dans une partie du quadrant Sud-Ouest telle que l'expert n'a pas à passer dans le quadrant Sud-Est pour atteindre l'objectif.
 }, et donc la dynamique entraînera l'exploration de ce quadrant et des deux quadrants Nord. Le quadrant Sud-Est, lui, sera ignoré par l'expert. Les algorithmes d'imitation ne disposeront donc pas d'information concernant le comportement de l'expert dans cet partie de l'espace d'état. La performance de la politique de l'agent imitant l'expert sera bien évidemment testée sur tout l'espace d'état, afin de voir comment l'agent a été capable d'extrapoler.
\begin{figure*}
\begin{tikzpicture}[for this and nested ones={scale=.6},transform shape]
\input{Figures/Mountain_car_state_space.tex}
\end{tikzpicture} 
\caption[Espace d'état du \glstext{mountaincar}]{L'espace d'état du \glstext{mountaincar} : la position de la voiture est représentée en abscisse et sa vitesse en ordonnée. Représenté sur la figure de gauche par des points : un exemple d'une trajectoire d'une bonne politique de contrôle où l'expert poursuit la montée de la pente opposée à l'objectif avant de la redescendre et de parvenir à gravir la pente à droite. La couleur des points indique l'action effectuée par l'expert : en bleu, il accélère vers la gauche, en rouge vers la droite. La surface en couleur représente le nombre de pas de temps nécessaire à cet expert pour parvenir à l'objectif. La forme en escargot est caractéristique de ce problème, nous la retrouvons Figure 2.2.}
% \autoref{fig:mountain_car_mu}.}
\label{fig:mcstatespace}
\end{figure*}
\section{Problème du \glstext{highway}}
\label{sec-9-3}
\label{hier-highway}

Utilisé sous diverses variantes comme banc d'essai de plusieurs algorithmes d'\gls{irl} \citep{abbeel2004apprenticeship,syed2008game,syed2008apprenticeship}, le simulateur de conduite sur autoroute place l'agent au volant d'une voiture (bleue) plus rapide que le trafic, et qui de fait doit éviter les voitures plus lentes (rouges) présentes sur la route. La voiture peut choisir parmi trois voies de circulation, et peut également rouler dans les deux bas-côtés de chaque côté de la route. Il y a trois vitesses possibles (toutes plus élevées que celles des voitures à éviter), une seule autre voiture est visible, quand l'agent la dépasse une autre apparaît dans une voie choisie aléatoirement par le simulateur. \`A la vitesse maximale, l'agent ne peut éviter la voiture qui arrive si elle est située dans la même voie que lui. \`A chaque instant, il peut choisir de se décaler à droite ou à gauche, ou d'accélérer ou ralentir, il peut également ne rien changer.

Ce simulateur est intéressant car beaucoup de comportements différents sont envisageables, pouvant présenter un subtil compromis entre vitesse et sécurité par exemple. Il est possible de conduire raisonnablement sans aucune collision et sans sortir de la route, à condition de ne pas aller vite. Il est également possible d'aller vite et d'éviter les collisions, à condition que l'on accepte de sortir de la route. Arriver à identifier les motivations précises d'un expert en observant sa conduite est donc un défi.

Contrairement au problème du \gls{mountaincar}, l'espace d'état est de dimension trop large pour être représenté. On y trouve en effet la vitesse de l'agent, sa position horizontale, et la position verticale et horizontale de l'autre voiture.
\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{Figures/Highway.png}
\caption[Capture d'écran du problème du \gls{highway}]{Capture d'écran du problème du \gls{highway}.}
\label{fig:highwayscreenshot}
\end{figure}
\chapter{Contributions}
\label{sec-10}
\label{hier-contributions}

L'évaluation d'une méthode d'apprentissage par imitation peut se faire selon deux critères :
\begin{itemize}
\item la nature et la quantité d'information qu'il faut lui donner pour qu'elle fonctionne;
\item la qualité du contrôle qu'elle offre, comparée à la qualité du contrôle de l'expert ayant effectué la démonstration.
\end{itemize}

Les méthodes d'apprentissage numérique supervisé excellent quant au premier critère, elles ne nécessitent pour fonctionner que de disposer d'une base d'exemples du comportement expert. Les méthodes d'apprentissage par renforcement inverse de l'état de l'art ne sont pas aussi souples. Elles nécessitent\footnote{Sauf une à notre connaissance, \gls{relent}, que nous utiliserons de fait comme point de comparaison.
 } pour fonctionner au moins un simulateur du système dynamique à contrôler. Les contributions de cette thèse visent à s'affranchir de ce besoin. Les méthodes d'\gls{irl} sont intrinsèquement plus difficiles à utiliser en pratique que les méthodes d'imitation supervisée car après avoir trouvé une récompense, il faut l'optimiser pour obtenir une politique de contrôle. Nous verrons cependant que si l'on peut surmonter cette contrainte, alors les méthodes d'\gls{irl} que nous proposons ici présentent expérimentalement une meilleure capacité de généralisation que les méthodes supervisées.
\section{\glstext{lstdmu}}
\label{sec-10-1}

La première contribution, l'algorithme \gls{lstdmu}, auquel le \autoref{hier-lstdmu} est consacré, permet d'estimer une grandeur centrale en \gls{irl}, l'attribut moyen, sans avoir à utiliser un simulateur du système :

\begin{itemize}
\item Elle a donné lieu à une première communication dans un atelier d'IJCAI portant sur l'apprentissage machine grâce à des instructeurs humains :
\end{itemize}
\fullcite{klein2011batch2}
\begin{itemize}
\item Nous avons ensuite rencontré la communauté d'apprentissage francophone aux JFPDA :
\end{itemize}
\fullcite{klein2011batch1}
\begin{itemize}
\item Ces travaux ont donné lieu à une publication avec actes à la conférence EWRL, qui regroupe la communauté internationale d'apprentissage par renforcement.
\end{itemize}
 \fullcite{klein2011batch}
\section{\glstext{scirl}}
\label{sec-10-2}

Vient ensuite le nouvel algorithme d'\gls{irl} \gls{scirl} que nous décrivons et étudions au \autoref{hier-scirl} et qui peut tirer parti des avancées offertes par \gls{lstdmu}. Grâce à une heuristique, il est capable de trouver une récompense à partir uniquement de données expertes.

\begin{itemize}
\item Une version préliminaire de ces travaux a donné lieu à une communication sans actes à EWRL.
\end{itemize}
\fullcite{klein2012structured3}
\begin{itemize}
\item Nous avons par la suite présenté nos travaux à la communauté francophone à CAp.
\end{itemize}
\fullcite{klein2012structured2}
\begin{itemize}
\item Une version aboutie de ces travaux a été publiée à NIPS.
\end{itemize}
\fullcite{klein2012structured}
\begin{itemize}
\item Une version étendue de l'article présenté à CAp a été publiée dans la revue francophone RIA.
\end{itemize}
\fullcite{klein2012structured1}
\section{\glstext{cascading}}
\label{sec-10-3}

Notre dernière contribution majeure est l'algorithme \gls{cascading} (\autoref{hier-cascading}), il s'agit d'un autre algorithme d'\gls{irl} un peu plus souple que \gls{scirl}. Les deux algorithmes ont été développés parallèlement, mais les publications concernant \gls{cascading} sont plus récentes.
\begin{itemize}
\item L'algorithme a été présenté aux JFPDA cette année.
\end{itemize}
\fullcite{klein2012cascading1}
\begin{itemize}
\item Il sera également présenté à Prague à ECML/PKDD
\end{itemize}
\fullcite{klein2012cascading}
\begin{itemize}
\item Enfin, une communication étudiant les liens entre classification et \gls{irl}, regroupant ces travaux et ceux menés sur \gls{scirl} aura lieu à RLDM2013 :
\end{itemize}
\fullcite{geist2013around}

L'analyse théorique de \gls{cascading} est due à un autre doctorant de l'équipe MaLIS, Bilal Piot. Elle précède chronologiquement l'analyse de \gls{scirl}, qui s'en inspire.
\section{Autres travaux}
\label{sec-10-4}
\label{hier-autrestravaux}

D'autres travaux, parallèles à ceux présentés ici, n'ont pas donné de résultats satisfaisants. Les résultats préliminaires avaient cependant donné lieu à une publication. L'idée est de réduire la dimension de l'espace d'hypothèses dans lequel on cherche une récompense en exploitant les transformations (translations, multiplications par un facteur positif) qu'il est possible, sur tout \gls{mdp}, d'appliquer à une récompense sans changer ses politiques optimales. Une modification de l'algorithme du simplexe permet, dans cet espace de dimension réduite, de trouver les récompenses creuses (différentes de zéro en seulement quelques points) expliquant le comportement de l'expert. Malheureusement l'existence de telles solutions n'est pas garantie et la complexité de l'algorithme croît exponentiellement avec le cardinal de l'espace d'état. \\
\fullcite{klein2011dimentionality}

Lors de la 8$^{\textrm{ème}}$ école d'été sur les interfaces multimodales (eNTERFACE'12), nous avons tenté d'appliquer l'\gls{irl} à un problème réel et de grande dimension : la commande d'un bras robot par le biais d'une interface cerveau/machine (\emph{Brain Computer Interface}, BCI). En observant un oracle associer la sortie de la BCI (une fois filtrée) à un ordre (avancer, droite, gauche, etc.), le but est de trouver une récompense qui, une fois optimisée, permettrait à un agent de remplacer l'oracle. L'utilisation de l'\gls{rl} permettrait d'obtenir un agent s'améliorant au fur et à mesure que le nombre d'interactions augmente, et qui s'adapte à l'utilisateur. Les résultats ont été prometteurs, en ce sens que l'algorithme \gls{scirl} (que nous présentons au \autoref{hier-scirl}) a été en mesure de trouver une fonction de récompense. Nous avons cependant manqué de temps pour collecter les données nécessaires à l'optimisation de cette récompense, et n'avons de fait pas pu évaluer la qualité de la récompense fournie par \gls{scirl}.\\
\fullcite{bougrain2012inverse}

\begin{fullwidth}
\printbibliography
\end{fullwidth}

\end{document}

### 2013-10-29 Manuscrit/Resume_fac.tex
Résumé pour la fac

\documentclass[11pt]{article}
\pagestyle{empty}
\usepackage[francais]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{url}

\usepackage{graphicx}
\usepackage{fancybox}

\setlength{\textwidth}{160mm}

\setlength{\oddsidemargin}{0mm}
% \setlength{\evensidemargin}{30mm}

\begin{document}
\section*{Résumé}
Humains et animaux évoluent dans des environnement incertains et dont la dynamique peut être non linéaire. Donner à une machine ou un robot la capacité d'accomplir une tâche dans un tel environnement constitue le problème du contrôle optimal. Résoudre ce problème d'une manière qui ne soit pas \emph{ad-hoc} à un environnement particulier est un problème ouvert.

Un paradigme efficace pour la résolution générique du problème du contrôle optimal est l'Apprentissage par Renforcement (AR), qui à partir d'un signal de récompense défini dans un Processus Décisionnel de Markov (PDM), trouve (éventuellement de manière approchée) par interactions avec l'environnement le comportement maximisant le cumul des récompenses sur le long terme. La conception par l'opérateur humain de ce signal de récompense reste une tâche ardue nécessitant d'une part des connaissances en AR et d'autre part une expertise quant au problème que la machine doit résoudre.

Afin de contourner la difficulté que représente la conception de la fonction de récompense, nous étudions les techniques dites d'Apprentissage par Renforcement Inverse (ARI) qui extraient la récompense de démonstrations de la tâche à effectuer accomplies par un expert. Une fois la récompense extraite des démonstrations expertes, elle peut être optimisée par un algorithme d'AR fournissant ainsi une politique de contrôle imitant le comportement de l'expert.

Ces méthodes rentrent dans le cadre plus large de l'apprentissage par imitation, où l'on trouve notamment les algorithmes d'imitation supervisée qui n'utilisent pas de description de la tâche à effectuer (rôle joué en AR par la fonction de récompense) mais apprennent l'association démontrée par l'expert entre la situation de la machine dans l'environnement et les actions qu'il effectue. Cet apprentissage, qui correspond au problème de la classification multi-classe, ne tient pas compte de la structure temporelle présente dans les données. Il présente l'avantage de ne nécessiter que des données expertes au prix de difficultés à généraliser le comportement de l'expert dans les zones de l'environnement où celui-ci n'a pas fourni de démonstrations.

L'ARI demande à l'inverse des données plus difficiles à réunir mais, grâce à l'optimisation de la fonction de récompense par un algorithme d'AR une fois celle-ci extraite des données, promet une meilleur capacité de généralisation. Les méthode d'ARI de l'état de l'art obéissent pour la plupart à un schéma itératif qui compare de manière répétée la politique optimale pour la récompense courante à la politique de l'expert, par le biais de l'attribut moyen. Le calcul répété d'une politique optimale et de son attribut moyen rendent ces approches inopérantes en l'absence d'un simulateur de l'environnement considéré. Une méthode d'ARI plus moderne ne nécessite pour fonctionner que les données expertes et des données représentatives (i.e. tirées par une politique aléatoire) du MDP afin d'effectuer de l'échantillonnage préférentiel.

Notre thèse porte sur la définition d'algorithmes d'ARI efficaces en échantillons. Le but est de bénéficier de la souplesse d'utilisation des algorithmes supervisés, qui ne nécessitent pour fonctionner que des données faciles à réunir dans la pratique, tout en conservant la bonne capacité de généralisation des algorithmes d'ARI. A cet effet nous fournissons un algorithme d'estimation de l'attribut moyen qui permet de faire fonctionner les algorithmes itératifs de la littérature sans avoir besoin d'un simulateur, celui-ci étant remplacé par des données représentatives du MDP.

Nous proposons ensuite deux nouveaux algorithmes d'ARI. Le premier introduit la dynamique temporelle du MDP dans une méthode de classification structurée par le biais de l'attribut moyen (qui peut être évalué par l'algorithme d'estimation que nous proposons). L'attribut moyen intervient dans la définition de la fonction de score du classifieur. Le second algorithme, plus souple, cascade une méthode de classification avec une étape de régression, qui en inversant l'équation de Bellman permet d'obtenir une fonction de récompense à partir de la fonction de score du classifieur.

Ces deux nouvelles méthodes disposent de garanties théoriques portant sur l'optimalité de la politique experte vis-à-vis de la fonction de récompense renvoyée par l'algorithme d'ARI. Armant chacune d'une heuristique qui lui est propre, nous illustrons empiriquement leur capacité à déduire une fonction de récompense à partir de démonstrations expertes uniquement. Les performances de nos algorithmes sont supérieures à celles d'algorithmes d'imitation supervisée et à un algorithme d'ARI moderne disposant de données non expertes supplémentaires.

\end{document}

### 2013-10-30 Manuscrit/Resume_fac.tex
Résumé pour la fac, remarques M&O

@@
-\documentclass[11pt]{article}
+\documentclass[9pt]{article}
 \pagestyle{empty}
@@
 \usepackage{url}
-
+\usepackage{fullpage}
 \usepackage{graphicx}
@@
 \section*{Résumé}
-Humains et animaux évoluent dans des environnement incertains et dont la dynamique peut être non linéaire.
-Donner à une machine ou un robot la capacité d'accomplir une tâche dans un tel environnement constitue le problème du contrôle optimal.
-Résoudre ce problème d'une manière qui ne soit pas \emph{ad-hoc} à un environnement particulier est un problème ouvert.
+Ce manuscrit s'intéresse au problème du contrôle optimal, qui consiste à trouver un comportement maximisant un critère défini par l'opérateur.
+Il traite plus particulièrement des méthodes permettant de trouver le contrôle optimal de manière générique et non \emph{ad-hoc} à un environnement particulier, lorsque le modèle dynamique de cet environnement est inconnu.
 
-Un paradigme efficace pour la résolution générique du problème du contrôle optimal est l'Apprentissage par Renforcement (AR), qui à partir d'un signal de récompense défini dans un Processus Décisionnel de Markov (PDM), trouve (éventuellement de manière approchée) par interactions avec l'environnement le comportement maximisant le cumul des récompenses sur le long terme.
+Un paradigme efficace pour la résolution générique et sans modèle du problème du contrôle optimal est l'Apprentissage par Renforcement (AR).
+Les critères de réussite de la tâche sont fournis par l'opérateur sous la forme d'un signal de récompense défini dans un Processus Décisionnel de Markov (PDM).
+Les algorithmes d'AR trouvent (éventuellement de manière approchée) par interactions avec l'environnement le comportement maximisant le cumul des récompenses sur le long terme.
 La conception par l'opérateur humain de ce signal de récompense reste une tâche ardue nécessitant d'une part des connaissances en AR et d'autre part une expertise quant au problème que la machine doit résoudre.
 
-Afin de contourner la difficulté que représente la conception de la fonction de récompense, nous étudions les techniques dites d'Apprentissage par Renforcement Inverse (ARI) qui extraient la récompense de démonstrations de la tâche à effectuer accomplies par un expert.
-Une fois la récompense extraite des démonstrations expertes, elle peut être optimisée par un algorithme d'AR fournissant ainsi une politique de contrôle imitant le comportement de l'expert.
+Afin de contourner cette étape difficile, les techniques dites d'Apprentissage par Renforcement Inverse (ARI) sont apparues.
+Elles extraient la récompense de démonstrations de la tâche à effectuer, démonstrations accomplies par un expert.
+Cette récompense peut alors être optimisée par un algorithme d'AR fournissant ainsi une politique de contrôle imitant le comportement de l'expert.
 
@@
 
-L'ARI demande à l'inverse des données plus difficiles à réunir mais, grâce à l'optimisation de la fonction de récompense par un algorithme d'AR une fois celle-ci extraite des données, promet une meilleur capacité de généralisation.
-Les méthode d'ARI de l'état de l'art obéissent pour la plupart à un schéma itératif qui compare de manière répétée la politique optimale pour la récompense courante à la politique de l'expert, par le biais de l'attribut moyen.
-Le calcul répété d'une politique optimale et de son attribut moyen rendent ces approches inopérantes en l'absence d'un simulateur de l'environnement considéré.
-Une méthode d'ARI plus moderne ne nécessite pour fonctionner que les données expertes et des données représentatives (i.e.
-tirées par une politique aléatoire) du MDP afin d'effectuer de l'échantillonnage préférentiel.
+L'ARI demande à l'inverse des données plus difficiles à réunir mais, grâce à l'optimisation de la fonction de récompense par un algorithme d'AR une fois celle-ci extraite des données, promet une meilleure capacité de généralisation.
+Les méthodes d'ARI de l'état de l'art obéissent pour la plupart à un schéma itératif qui compare de manière répétée la politique optimale pour la récompense courante à la politique de l'expert, par le biais de l'attribut moyen.
+Le calcul répété d'une politique optimale et de son attribut moyen rendent ces approches inopérantes lorsque le modèle est inconnu.
+Une méthode d'ARI plus moderne ne nécessite pour fonctionner que les données expertes et des données tirées par une politique aléatoire afin d'effectuer de l'échantillonnage préférentiel.
 
-Notre thèse porte sur la définition d'algorithmes d'ARI efficaces en échantillons.
+Cette thèse porte sur la définition d'algorithmes d'ARI efficaces en échantillons.
 Le but est de bénéficier de la souplesse d'utilisation des algorithmes supervisés, qui ne nécessitent pour fonctionner que des données faciles à réunir dans la pratique, tout en conservant la bonne capacité de généralisation des algorithmes d'ARI.
-A cet effet nous fournissons un algorithme d'estimation de l'attribut moyen qui permet de faire fonctionner les algorithmes itératifs de la littérature sans avoir besoin d'un simulateur, celui-ci étant remplacé par des données représentatives du MDP.
+A cet effet ce document fournit un algorithme d'estimation de l'attribut moyen qui permet de faire fonctionner les algorithmes itératifs de la littérature sans avoir besoin d'un simulateur, celui-ci étant remplacé par des données représentatives du PDM.
 
-Nous proposons ensuite deux nouveaux algorithmes d'ARI.
-Le premier introduit la dynamique temporelle du MDP dans une méthode de classification structurée par le biais de l'attribut moyen (qui peut être évalué par l'algorithme d'estimation que nous proposons).
+Deux nouveaux algorithmes d'ARI sont ensuite proposés.
+Tous deux reposent sur un constat de similarité entre le rôle de la fonction de qualité de l'expert et celui de la fonction de score d'un classifieur.
+Le premier algorithme introduit la dynamique temporelle du PDM dans une méthode de classification structurée par le biais de l'attribut moyen (qui peut être évalué par l'algorithme d'estimation précédent).
 L'attribut moyen intervient dans la définition de la fonction de score du classifieur.
@@
 Ces deux nouvelles méthodes disposent de garanties théoriques portant sur l'optimalité de la politique experte vis-à-vis de la fonction de récompense renvoyée par l'algorithme d'ARI.
-Armant chacune d'une heuristique qui lui est propre, nous illustrons empiriquement leur capacité à déduire une fonction de récompense à partir de démonstrations expertes uniquement.
-Les performances de nos algorithmes sont supérieures à celles d'algorithmes d'imitation supervisée et à un algorithme d'ARI moderne disposant de données non expertes supplémentaires.
+Armée chacune d'une heuristique qui lui est propre, leur capacité à déduire une fonction de récompense à partir de démonstrations expertes uniquement est illustrée empiriquement.
+Les performances de ces algorithmes sont supérieures à celles d'algorithmes d'imitation supervisée et à un algorithme d'ARI moderne disposant de données non expertes supplémentaires.
 
